---
title: 'OHI `r format(Sys.Date(), "%Y")` - Tourism and Recreation '
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html' 
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---


[REFERENCE RMD FILE](http://ohi-science.org/ohiprep_v2023/globalprep/tr/v2023/tr_data_prep.html)


# Summary
This document describes the steps for obtaining the data used to calculate the tourism and recreation goal for the 2023 global assessment.

The general calculation is:
tr = Ap * Sr 
and
Xtr = tr/90th quantile across regions

* Ap = Proportion of overnight tourist arrivals to total arrivals
* Sr = (S-1)/5; Sustainability of tourism


## The following data are used:

* Numbers of tourist arrivals, used in the calculation of the proportion of arrivals to area of coastline to population: obtained through the [UNWTO](https://www.unwto.org/tourism-statistics/key-tourism-statistics) (in the form of thousands of people). More info on tourism terms [here](https://www.unwto.org/glossary-tourism-terms). Range: 1995-2021 (only 2008-2021 is used)
* Tourism sustainability: World Economic Forum. The Travel & Tourism Development Index 2021 dataset (version 24 May 2022). 2022. [TTDI](https://www.weforum.org/reports/travel-and-tourism-development-index-2021/downloads-510eb47e12#report-nav)
* Per capita GDP: (World Bank with gaps filled using CIA data), used to gapfill missing values in Tourism sustainability


# Updates from previous assessment

## Tourism sustainability
None in v2023. Copied data from v2022.

## Tourist arrivals (formerly tourism employment)
The data for this layer has been paywalled. Because of this, we have replaced the WTTC data with UNWTO data. This change causes us to be a year behind on information relative to the old data source, but this is unavoidable. We have also decided to use UNWTO tourist arrivals data instead of UNWTO tourism employment data due to the employment data containing a significant amount of missing data and there being a lack of sufficient gapfilling methods.


**We were able to update the following data:**

* Proportion of tourist arrivals to area of coastline to population - UNWTO data on thousands of tourist arrivals, reported until 2021 (downloaded [here](https://www.unwto.org/tourism-statistics/key-tourism-statistics) (dataset: "Total arrivals" under Inbound Tourism; we use the "Overnights visitors (tourists)" categorization as arrivals where possible) on 8/10/2023.


## Initial set-up code

```{r setup, message=FALSE, warning=FALSE, results="hide"}
# library(devtools)
# devtools::install_github("ohi-science/ohicore@dev") # dont worry about devtools
library(ohicore)
library(tidyverse)
library(stringr)
library(WDI)
library(here)
library(janitor)
library(plotly)
library(readxl)
library(naniar)
library(countrycode)

version_year <- "2023"
prev_ver_yr <- as.character(as.numeric(version_year) - 1)

source(paste0("http://ohi-science.org/ohiprep_v", version_year, "/workflow/R/common.R"))
region_data()
regions_shape()
#source(here(paste0("globalprep/tr/v", version_year, "/R/tr_fxns.R"))) # not used presently
```


# Ap: Proportion of tourist arrivals to area of coastline to population

We use international arrivals data from the [United Nations World Tourism Organization (UNWTO)](https://www.unwto.org/). Up until the current assessment, we accessed data from the [World Travel & Tourism Council (WTTC)](http://www.wttc.org/), but this is no longer a viable option. 

To address missing values in arrivals, specifically referring to "Overnight visitors (tourists)," we employ a two-step process. First, we attempt to fill the gaps by subtracting "Same-day visitors (excursionists)" from "Total arrivals" if the latter is available. If this is not feasible, we resort to interpolating or extrapolating based on historical data spanning from 1995 to 2019, employing a linear model to estimate increases or decreases on a regional level.

However, in light of the Covid-19 pandemic, we have adopted a distinct approach for the years 2020 and 2021. We calculate the global average proportionate change from the preceding year, apply this percentage change to the previous year's arrivals or total values, and then add the result to the corresponding previous year's arrivals or total value.
 - So 2020 was gapfilled with the global average decrease proportion of ~-0.70 for both arrivals and totals. Meaning that we took the 2019 value and multiplied it by -0.7 and then addd that to the 2019 value. 2019*-0.7 + 2019 = 2020
 - 2021 was gapfilled with the global average increase proportion of ~0.2. Meaning that we took the 2020 value and multiplied it by 0.2 and then addd that to the 2020 value. 2020*0.2 + 2020 = 2021


### Source cleaned and gapfilled arrivals and population data sources

```{r, eval=FALSE}
# source in cleaned UNWTO data for current version year (make sure to download from website and put on Mazu in the UNWTO folder first)
source(here(paste0("globalprep/tr/v", version_year, "/R/process_UNWTO_arrivals.R"))) # outputs unwto_dupe_fix_downup_gf

coastal_pop_data <- read.csv(here("globalprep/mar_prs_population/v2021/output/mar_pop_25mi.csv")) ## read in coastal population data from other data layer

coastal_pop_data_fill <- coastal_pop_data %>%
  filter(year == 2020) %>%
  mutate(year = 2021) %>%
  rbind(coastal_pop_data) # add 2021 data year, just using year 2020
```

### Divide tourist arrival count by total arrival count

 - In 2023, we considered rescaling by coastal area and/or coastal population, however, we decided to scrap this, as it was unduly harming scores for larger regions. We will just use the tourist arrival/total arrivals metric.
 - Because of some gapfilling and uncertainty in the data, some regions reported higher tourist arrivals than total arrivals. Due to this, we sometimes have ratios >1. To deal with this, we just give those regions a ratio of 1. 
 - Some regions reported only tourist arrivals, and no total arrivals. Because of this, we gapfilled any that didn't have total arrivals with the tourist arrivals. This isn't ideal, but was the best we could do. 

```{r, eval=FALSE}
# divide the number of tourist arrivals by area of coastline and population to get the proportion
tourism_props <- unwto_all_gf %>%
    mutate(year = as.numeric(year)) %>%
 # left_join(offshore_data, by = c("rgn_id", "year")) %>%
  left_join(coastal_pop_data_fill, by = c("rgn_id", "year")) %>%
  # mutate(Ap = ((tourism_arrivals_ct/total_inland_offshore_area)/popsum)) %>%
  mutate(total = ifelse(is.na(total), tourism_arrivals_ct, total)) %>%
  mutate(Ap = tourism_arrivals_ct/total) %>%
  filter(year >= 2008) %>% # filter to the years we are interested in for any data that isn't filtered yet
  filter(popsum > 0) %>% # filter out no pop regions
  left_join(rgns_eez) %>%
  mutate(Ap = ifelse(Ap > 1, 1, Ap)) # if greater than 1 make 1

# check out things so far
summary(tourism_props) # should be 0 NAs for Ap if gapfilling worked
hist(tourism_props$Ap)

```

```{r, eval=FALSE}

# make rescaled column the actual value column
tourism_props_rescaled <- tourism_props %>%
  dplyr::select(rgn_id, year, arrivals_method, arrivals_gapfilled, Ap)
  
```

### Removing low population / uninhabited regions

```{r, eval=FALSE}
### after gap-filling, make sure low/uninhabited regions are NA
# create df for unpopulated/low populated regions
low_pop()
low_pop <- low_pop %>%
  filter(est_population < 3000 | is.na(est_population)) %>%  # filter out regions that have populations > or equal to 3000 and keep NA values 
  rename(rgn_label = rgn_nam)

summary(tourism_props_rescaled)
# v2020 371 NAs
# v2022 114 NAs
# v2023 0 NAs (because of gapfilling)

# make sure all the NAs are uninhabited regions
tourism_props_nas <- tourism_props_rescaled %>% 
  filter(is.na(Ap)) %>% 
  select(rgn_id, year) %>% # v2023: was rgn_id, year, r1_label, r2_label, rgn_label but we did not gapfill by georegion so did not have the last 3
  left_join(low_pop, by = "rgn_id") # v2023: was by = c("rgn_id", "rgn_label") but we did not gapfill by georegion so did not have rgn_label

tourism_props_nas %>% 
  filter(Inhabited == 0 & !is.na(est_population)) %>% 
  nrow() # 0 ✓

max(tourism_props_nas$est_population, na.rm=TRUE) < 3000 # should be true

# make sure all the uninhabited regions are NA and then drop them (all NA regions get added back as a last step just in case its relevant)
tourism_props_rescaled <- tourism_props_rescaled %>% 
  mutate(Ap = ifelse(rgn_id %in% low_pop$rgn_id, NA, Ap)) 


# check NAs once more 
summary(tourism_props_rescaled)
# v2019: Adding the low pop df identifies 13 additional regions that should be NA instead of gapfilled, taking the total number of NAs in the data set from 245 to 700
# v2020: Adding the low pop df takes the total number of NAs in the data set from 371 to 832
# v2022: Adding the low pop df takes the total number of NAs in the data set from 14 to 40
# v2023: Adding the low pop df takes the total number of NAs in the data set from 0 to 14

# after checking NAs, get rid of them
tourism_props_rescaled <- tourism_props_rescaled %>% 
  drop_na(Ap)
```

### Write output files

```{r, eval=FALSE}
# we want to make sure all OHI regions are present in the data, even if could not calculate Ap (this will add back in what we removed above as well as any region not present in the data)
year_range <- unique(tourism_props_rescaled$year) # get the year range of Ap
year_range_df <- data.frame(year = year_range) # make it a dataframe

# save gapfill info
tourism_props_gf_to_write <- tourism_props_rescaled %>%
  select(-Ap) # don't need actual values for the gapfill information

write_csv(tourism_props_gf_to_write, here(paste0("globalprep/tr/v", version_year, "/output/tr_arrivals_props_tourism_gf.csv")))

# save gap-filled data
tourism_props_to_write <- tourism_props_rescaled %>%
  select(rgn_id, year, Ap) # don't need gf info here, just the values

write_csv(tourism_props_to_write, here(paste0("globalprep/tr/v", version_year, "/output/tr_arrivals_props_tourism.csv")))
```

### Look at changes in recent years

We would expect for tourism jobs to decrease across the board from 2019 and 2020 given the pandemic, and likely see a rebound to some extent between 2020 and 2021 — let's make sure that's reflected in our results.

```{r, eval=FALSE}
tourism_props_compare <- tourism_props_to_write %>%
  mutate(year = as.numeric(as.character(year))) %>%
  filter(year >= 2019) %>%
  pivot_wider(names_from = year, values_from = Ap)

# compare 2019 and 2020
plot(tourism_props_compare$"2019", tourism_props_compare$"2020",
     xlab = "v2023 2019 Arrivals Proportion", ylab = "v2023 2020 Arrivals Proportion")
abline(0, 1) # more data below the line

# compare 2020 and 2021
plot(tourism_props_compare$"2020", tourism_props_compare$"2021",
     xlab = "v2023 2020 Arrivals Proportion", ylab = "v2023 2021 Arrivals Proportion")
abline(0, 1) # more data above the line
```

Everything looks reasonable.

### Look at changes vs. previous data source (v2023)

```{r, eval=FALSE}
new_data <- read_csv(paste0("globalprep/tr/v", version_year, "/output/tr_arrivals_props_tourism.csv"))
old_data <- read_csv(paste0("globalprep/tr/v", prev_ver_yr, "/output/tr_jobs_pct_tourism.csv"))

compare_common_data <- new_data %>%
  left_join(old_data, by = c("rgn_id", "year")) %>%
  drop_na()

plot(compare_common_data$Ep, compare_common_data$Ap,
     xlab = "v2022 Employment Proportion", ylab = "v2023 Arrivals Proportion")
abline(0, 1)




compare_common_data_2021 <- new_data %>%
  left_join(old_data, by = c("rgn_id", "year")) %>%
  drop_na() %>%
  filter(year == 2021)

plot(compare_common_data_2021$Ap, compare_common_data_2021$Ep,
     xlab = "v2023 Arrivals Proportion", ylab = "v2022 Employment Proportion")
abline(0, 1)

compare_common_data_2020 <- new_data %>%
  left_join(old_data, by = c("rgn_id", "year")) %>%
  drop_na() %>%
  filter(year == 2020)

plot(compare_common_data_2020$Ap, compare_common_data_2020$Ep,
     xlab = "v2023 Arrivals Proportion", ylab = "v2022 Employment Proportion")
abline(0, 1)

compare_common_data_2019 <- new_data %>%
  left_join(old_data, by = c("rgn_id", "year")) %>%
  drop_na() %>%
  filter(year == 2019)

plot(compare_common_data_2019$Ap, compare_common_data_2019$Ep,
     xlab = "v2023 Arrivals Proportion", ylab = "v2022 Employment Proportion")
abline(0, 1)

compare_common_data_2015 <- new_data %>%
  left_join(old_data, by = c("rgn_id", "year")) %>%
  drop_na() %>%
  filter(year == 2015)

plot(compare_common_data_2015$Ap, compare_common_data_2015$Ep,
     xlab = "v2023 Arrivals Proportion", ylab = "v2022 Employment Proportion")
abline(0, 1)
```

### Check out some specific countries (v2023) -- this was for exploring changes in methodology, can skip or use/edit parts in future years

```{r, eval=FALSE}
# check some countries that changed a lot in v2023's first push to global
check_countries_graph <- tourism_props %>% 
  filter(rgn_id == 24 | rgn_id == 51 | rgn_id == 189 | rgn_id == 118 | rgn_id == 31) %>%
  mutate(rgn_id_plot = as.factor(rgn_id),
         year_plot = year)

Ap_graph <- ggplot(check_countries_graph, aes(x = year_plot, y = Ap, color = rgn_id_plot)) +
  geom_line() +
  theme_minimal() +
  labs(x = "",
       color = "Region ID")


Ap_graph

countries_in_2021 <- check_countries_graph %>%
  filter(year == "2021") %>%
  left_join(rgns_eez, by = "rgn_id") %>%
  select(-year_plot)

library(kableExtra)
kable(countries_in_2021) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```



# Ts: Tourism sustainability

These data are from the World Economic Forum's "Travel and Tourism Development Index" (https://www.weforum.org/reports/travel-and-tourism-development-index-2021/downloads-510eb47e12) See mazu: _raw_data/WEF-Economics/ for more details and the raw data.

The TTDI was formerly the TTCI which was a similar index, but unfortunately not comparable. The TTDI only extends back to 2019. 

These data are gapfilled using gdppcppp and UN georegion information (see next section for obtaining and preparing these data).

```{r WEF processing, eval=FALSE}
# update to latest file name
ttdi_file <- "WEF_TTDI_2021_data_for_download.xlsx"

ttdi_raw <- read_excel(paste0(dir_M, "/git-annex/globalprep/_raw_data/WEF-Economics/d", version_year, "/", ttdi_file),
                 skip = 2) 

# move up column names from first row while keeping the full country names as columns too
names(ttdi_raw)[1:9] <- as.character(ttdi_raw[1, 1:9])

# filtering for sustainability scores, selecting needed columns, and pivoting to tidy format
ttdi <- ttdi_raw %>%
  filter(Title == "T&T Sustainability subindex, 1-7 (best)",
         Attribute == "Score") %>% 
  select(year = Edition, Albania:Zambia) %>% 
  # currently Zambia is the last country column
  pivot_longer(cols = Albania:Zambia, names_to = "country",
               values_to = "score") %>% 
  mutate(score = as.numeric(score))


# Changing names that are not recognized by ohicore
ttdi <- ttdi %>%
    mutate(country = ifelse(str_detect(country, "Ivoire"), "Ivory Coast", country))
  
  
ttdi_rgn <- name_2_rgn(df_in = ttdi, 
                       fld_name='country')

## Duplicated regions weighted mean
weight_data <- data.frame(country = c("China", "Hong Kong SAR"),
                          # pop values from World Bank 2021 estimates - updated v2022
                          population = c(1412360000, 7413100))


ttdi_rgn <- ttdi_rgn %>%
  arrange(country) %>%
  left_join(weight_data, by = "country") %>%
  mutate(population = ifelse(is.na(population), 1, population)) %>%
  group_by(rgn_id, rgn_name, year) %>%
  summarize(score = weighted.mean(score, population)) %>%
  select(year, rgn_id, rgn_name, score)

# compare with old dataframe to make sure only the duplicated region scores changed 

head(ttdi_rgn, 10)

### Save TTDI data file
write.csv(ttdi_rgn, here(paste0("globalprep/tr/v", version_year, "/intermediate/wef_ttdi.csv")), row.names = FALSE)

```

## Preparing the gdppcppp data:
These data are used to gapfill missing values in tourism sustainability.  Most of the data are from the World Bank, but CIA data fill some gaps (CIA data is available for only the most recent year).

The Artisanal Opportunities goal uses gdppcppp data, so we will get the data that was processed for that goal.


```{r worldbank, eval=FALSE}
wb <- read.csv(here(paste0("globalprep/ao/v", version_year, "/intermediate/gdppcppp_ohi.csv"))) %>%
  dplyr::select(rgn_id, year, value)

```

CIA data are used to fill in missing gaps in the gdppcppp data (https://www.cia.gov/the-world-factbook/field/real-gdp-per-capita/country-comparison)

Downloaded: 07/05/2022

See README on the raw folder for instructions on how to download this data. 

The following code is used to prepare these data for OHI:

```{r cia gdp, eval=FALSE}

cia_gdp <- read.csv(here(paste0("globalprep/tr/v", version_year, "/raw/cia_gdp_pc_ppp.csv"))) %>% 
  # remove dollar signs and commas and convert to numeric
  mutate(value = as.numeric(gsub("[$,]", "", value))) %>% 
  select(name, value) %>% 
  rename(country = name, pcgdp_cia = value)

 ## Data reported in a lower resolution than OHI regions
splits <- data.frame(country = "Saint Helena, Ascension, and Tristan da Cunha", 
                     country2 = c("Saint Helena", "Ascension","Tristan da Cunha"))

cia_gdp <- cia_gdp %>%
  left_join(splits, by='country') %>%
  mutate(country2 = ifelse(is.na(country2), country, country2)) %>%
  select(country = country2, pcgdp_cia)

cia_gdp_rgn <- name_2_rgn(df_in = cia_gdp, 
                       fld_name='country')

### Duplicated regions: Collapse regions after weighting by population (regions we include as a single region) - 

population_weights <- data.frame(country = c("Virgin Islands", "Puerto Rico",
                                             "China", "Hong Kong", "Macau",
                                             "Guam", "Northern Mariana Islands"),
                                 # from world bank - updated v2022
                                 population = c(105870, 3263584, 1412360000,
                                                7413100, 658391, 170184, 57910))

cia_gdp_rgn <- cia_gdp_rgn %>%
  left_join(population_weights, by="country") %>%
  mutate(population = ifelse(is.na(population), 1, population)) %>%
  group_by(rgn_id) %>%
  summarize(pcgdp_cia = weighted.mean(pcgdp_cia, population)) %>%
  ungroup() %>%
  filter(rgn_id <= 250) %>%
  select(rgn_id, pcgdp_cia)

write.csv(cia_gdp_rgn, here(paste0("globalprep/tr/v", version_year, "/intermediate/wb_rgn_cia_GDPPCPPP.csv")), row.names=FALSE)

```

The following code combines the two gdp datasets and gapfills missing regions using UN georegions.

If there is no World Bank gdppcppp data (pcgdp), the CIA data is used (pcgdp_cia).  The pcgdp2 variable includes both the World Bank and CIA data (with CIA data only used if there is not World Bank data).  The remaining data are estimated using UN geopolitical regions.  Ideally, the mean gdppcppp value is calculated at the r2 scale (gdp_pred_r2) using regions within each class with gdppcppp data.  If there were not enough regions with data at the r2 scale, the average at the r1 scale was used (gdp_pred_r1). The gdp_all variable combines all estimates using the following heirarchy:  World Bank -> CIA -> estimated using mean from r2 UN geopolitical regions -> estimated using mean from r1 UN geopolitical regions.    

```{r gapfill gdp, eval=FALSE}

### world bank gdp data
gdppcppp <- wb %>%
  select(rgn_id, year, pcgdp = value)

### cia gdp data
gdppcppp2 <- read.csv(here(paste0("globalprep/tr/v", version_year, "/intermediate/wb_rgn_cia_GDPPCPPP.csv")))


### Use WB data, but if missing, use pcgdp_cia.
### combine with UN georegion data
years <- data.frame(year = min(gdppcppp$year):max(gdppcppp$year))

georegions <- ohicore::georegions

regions <- georegions %>%
  left_join(georegion_labels, by = 'rgn_id')

gdp_raw <- merge(years, regions, by=NULL) %>%
   left_join(gdppcppp, by = c('rgn_id', 'year')) %>%
  left_join(gdppcppp2, by = c("rgn_id")) 

## quick compare to make sure the CIA and World Bank data are compatible
plot(gdp_raw$pcgdp[gdp_raw$year==2021], gdp_raw$pcgdp_cia[gdp_raw$year==2021])
abline(0,1, col="red")
# a few minor outliers but overall looks good

gdp_raw <- gdp_raw %>%
  mutate(pcgdp2 = ifelse(is.na(pcgdp), pcgdp_cia, pcgdp))

## Calculating the means across different geopolitical levels (e.g. r2, r1)
gdp_raw <- gdp_raw %>%
  group_by(r2, year) %>%
  mutate(gdp_pred_r2 = mean(pcgdp2, na.rm=TRUE)) %>%
  ungroup() %>%
  group_by(r1, year) %>%
  mutate(gdp_pred_r1 = mean(pcgdp2, na.rm=TRUE)) %>%
  ungroup() 

gdp_raw_gf <- gdp_raw %>%
  mutate(gdp_all = ifelse(is.na(pcgdp2), gdp_pred_r2, pcgdp2)) %>%
  mutate(gdp_all = ifelse(is.na(gdp_all), gdp_pred_r1, gdp_all)) %>%
  mutate(gapfilled = ifelse(is.na(pcgdp2) & !is.na(gdp_all), "gapfilled", NA)) %>%
  mutate(method = ifelse(is.na(pcgdp2) & !is.na(gdp_pred_r2), "UN georegion (r2)", NA)) %>%
  mutate(method = ifelse(is.na(pcgdp2) & is.na(gdp_pred_r2) & !is.na(gdp_pred_r1), "UN georegion (r1)", method)) 

write_csv(gdp_raw_gf, here(paste0("globalprep/tr/v", version_year, "/intermediate/gdp_raw_gf.csv")))

gdp_data_gf <- gdp_raw_gf %>%
  select(rgn_id, year, gapfilled, method) 

write_csv(gdp_data_gf, here(paste0("globalprep/tr/v", version_year, "/intermediate/gdp_gf.csv")))

gdp_data <- gdp_raw_gf %>%
  select(rgn_id, year, pcgdp = gdp_all)

write_csv(gdp_data, here(paste0("globalprep/tr/v", version_year, "/intermediate/gdp.csv")))

```


The final step is gapfilling the Sustainability data using a linear model with gdppcppp and UN geopolitical regions as predictor variables.  

```{r, eval=FALSE}

sust <- read.csv(here(paste0("globalprep/tr/v", version_year, "/intermediate/wef_ttdi.csv")), stringsAsFactors = FALSE)

### don't need to gapfill data without tourism data:
## Most recent tourism data is 2019.  

ap_gf <- read.csv(here(paste0("globalprep/tr/v", version_year, "/output/tr_arrivals_props_tourism.csv"))) %>%
  # filter(year == 2021) %>%
  select(rgn_id, Ap, year) %>%
  filter(!is.na(Ap))

# gdp dataframe prepared above (World Bank, CIA, and gapfilled gdp data)
gdp_raw_gf <- read.csv(here(paste0("globalprep/tr/v", version_year, "/intermediate/gdp_raw_gf.csv")), stringsAsFactors = FALSE) %>% 
  # filter(year == 2021) %>%
  select(rgn_id, r0_label, r1_label, r2_label, rgn_label,
         pcgdp, pcgdp_cia, pcgdp2, gdp_all, year) 

tr_sust <- gdp_raw_gf %>%
           left_join(sust, by = c("rgn_id", "year")) %>%
          left_join(ap_gf, by = c("rgn_id", "year")) %>%
          rename(S_score = score) %>%
          filter(rgn_id != 213)

### Add gapfill flag variable 
## Reminder:
## pcgdp2: includes both the World Bank and CIA data (with CIA data only used if there is not World Bank data)
## Ep: Proportion of workforce directly employed in tourism
## S_score: tourism sustainability score

tr_sust_gf <- tr_sust %>%
  mutate(gapfilled = ifelse(is.na(S_score) & !is.na(Ap), "gapfilled", NA)) %>%
  mutate(method = ifelse(is.na(S_score) & !is.na(Ap) & is.na(pcgdp2), "lm georegion + gdppcppp, with est. gdppcppp", NA)) %>%
  mutate(method = ifelse(is.na(S_score) & !is.na(Ap) & !is.na(pcgdp2), "lm georegion + gdppcppp", method)) %>%
  select(rgn_id, gapfilled, method, year)

write.csv(tr_sust_gf, here(paste0("globalprep/tr/v", version_year, "/output/tr_sustainability_gf.csv")), row.names=FALSE)

```
  

### Gapfilling
Linear models using gdppcppp and UN geopolitical regions as predictor variables. However if there is no gdppc data we estimate the gdppc using the UN georegions and then used in the linear model to gapfill the sustainability score.

```{r, eval=FALSE}

### Gapfill S using r1 and/or r2 regional data and PPP-adjusted per-capita GDP
### Looked at models with a year variable, but wasn't significant and decided to exclude

mod3 <- lm(S_score ~ as.factor(r2_label) + gdp_all, data=tr_sust, na.action = na.exclude)
summary(mod3)
anova(mod3)

mod4 <- lm(S_score ~ as.factor(r1_label) + gdp_all, data=tr_sust, na.action = na.exclude)
summary(mod4)
anova(mod4)

plot(predict(mod3), tr_sust$S_score)
abline(0,1)
plot(predict(mod4), tr_sust$S_score)
abline(0,1)


## Estimate missing data and gapfill
# Some of the r1 levels do not have data and consequently causes a fail. This chunk of code drops these levels so an NA is returned

# Select only r2 column
new_data <- tr_sust %>% 
  dplyr::select(r2_label, gdp_all)

unique(tr_sust$r2_label)

r2_w_data <- unique(tr_sust$r2_label[!is.na(tr_sust$S_score)])
  
new_data_r2 <- new_data %>%
  mutate(r2_label = ifelse(r2_label %in% r2_w_data, r2_label, NA))

# Predict sustainability scores using linear model 3 (using r2 data)
tr_sust <- tr_sust %>% 
  dplyr::mutate(S_score_pred_r2 = predict(mod3, newdata = new_data_r2))


# Select only r1 column
new_data <- tr_sust %>% 
  dplyr::select(r1_label, gdp_all)

unique(tr_sust$r1_label)

r1_w_data <- unique(tr_sust$r1_label[!is.na(tr_sust$S_score)])

new_data_r1 <- new_data %>%
  mutate(r1_label = ifelse(r1_label %in% r1_w_data, r1_label, NA))

# Predict sustainability scores using linear model 4 (using r1 data)
tr_sust <- tr_sust %>% 
  dplyr::mutate(S_score_pred_r1 = predict(mod4, newdata = new_data_r1))



## some are missing the r1 predictions, but none of these have Ep scores, so not relevant
View(filter(tr_sust, is.na(S_score_pred_r1)))

tr_sust <- tr_sust %>%
  mutate(S_score_2 = ifelse(is.na(S_score), S_score_pred_r2, S_score)) %>%
  mutate(S_score_2 = ifelse(is.na(S_score_2), S_score_pred_r1, S_score_2)) %>%
  filter(year %in% c(2019, 2021)) %>%
  select(rgn_id, year, S_score=S_score_2)

summary(tr_sust)

write_csv(tr_sust, here(paste0("globalprep/tr/v", version_year, "/output/tr_sustainability.csv")))
```


## Compare with previous year of data 

```{r, eval=FALSE}
tr_sust <- read_csv(here(paste0("globalprep/tr/v", version_year, "/output/tr_sustainability.csv")))

prev_year <- (as.numeric(version_year) - 1) %>% 
  as.character()

compare <- tr_sust %>% 
  pivot_wider(names_from = year, values_from = S_score)

# current vs previous year of data
plot(compare$"2021", compare$"2019")
abline(0, 1, col="red")
# looks good

```


# Tw: Travel warnings

 - Travel warnings were deleted from the v2020 assessment. 

