---
title: 'OHI `r format(Sys.time(), "%Y")`: Coral Harvest Pressure'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---

[REFERENCE RMD FILE: http://ohi-science.org/ohiprep_v2023/globalprep/prs_coral_harvest/v2023/prs_coral_harvest.html]

# Summary
This analysis converts FAO commodities data into data layers used to calculate OHI 2023 global coral harvest pressure.


## 2024 Updates from previous assessment
- New year of FAO data (1976-2022; new year 2022)
- Added code to remove flag column (was done inherently or upstream before)
- Built on programmatic year usage
- More updates to name_2_rgn
- Restructured file paths for cross-platform capability, primarily using `{here}` package
- Updated `raw/commodities2products.csv` to include new coral category
- Updated script to work with `janitor::clean_names()` instead of raw column names
- Updated `read.csv()` and `write.csv()` to `read_csv()` and `write_csv()` when appropriate

# 2023 Updates from previous assessment
- New 2 years of FAO data (1976-2021; new years 2020 and 2021)
- Added/edited code for the Datacheck section
- Added a tic/toc for timing purposes
- Added more programmatic year usage
- Swapped np_split_antilles for split_regions (more generalized usage)
- Updates to name_2_rgn reduce the code length of the Rmd
- Cleaned up code and comments

***

# Data Source

**Reference**:
    https://www.fao.org/fishery/statistics-query/en/trade
    App release date: March 2024
FAO raw commodities quantity 1976_2022
FAO raw commodities value 1976_2022
FAO raw commodities metadata found [here](https://www.fao.org/fishery/en/collection/global_commodity_prod) and overall FAO data directory found [here](https://www.fao.org/fishery/en/fishstat)

**Downloaded**: August 5, 2024

**Description**:  Quantity (tonnes) and value (USD) of raw commodities (Exports only) for each country, taxa, year. The FAO data is subset to include commodities in these categories: shells, corals, ornamental fish, fish oil, seaweed and plants, sponges (see: `raw/commodities2products.csv` for details).

**Time range**: 1976-2022


Flag codes:

E: Estimate
N: Not significant (below 0.5)
X: Value from international organization


**Summary of data layer from methodology**:
`hd_coral`

*Pressure*

**Category**: ecological

**Subcategory**: habitat destruction

>The total tonnes of coral harvest were determined for each region using export 
data from the FAO Global Commodities database. The tonnes of ornamental fishing 
was divided by the area of coral, taken from the hab_coral layer, to get the 
intensity of coral harvest per region. Following this, we set the reference 
value as the 95th quantile of coral harvest intensity. We then divided the 
intensity by the reference intensity. Anything that scored above 1 received a[n] 
intensity pressure score of 1. To incorporate the health of the coral, we then 
multiplied the intensity pressure score by the health of the coral, to get the 
final pressure score.


I think instead of ornamental fishing, it's tons of corals commodities 
***
  
# Methods
```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(eval = FALSE)

# ==== Load packages ====
if (!require(librarian)) {
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  ohicore, #devtools::install_github('ohi-science/ohicore@dev')
  dplyr,
  stringr,
  tidyr,
  here,
  tidyverse,
  zoo,
  ggplot2,
  plotly,
  tictoc,
  RColorBrewer
)

# ==== Set file paths ====
scen_year <- "2024"
v_scen_year <- paste0("v", scen_year)
data_dir_year <- paste0("d", scen_year)
prev_scen_year <- as.character(as.numeric(scen_year) - 1)
common_year <- as.character(as.numeric(scen_year) - 4) # this depends on what the latest data year is; there's often a 3 year lag so the common year between the current and previous version year would be 4 years prior to the current version year; this time there is a 2 year lag but the common year is still 4 years prior at the moment
# define common directories to shorten file paths in the script
current_prs_folder <- here::here("globalprep", "prs_coral_harvest", v_scen_year)
current_int_folder <- here::here(current_prs_folder, "int")

# ==== Load FAO-specific user-defined functions ====
source(here::here("workflow", "R", "fao_fxn.R")) # function for cleaning FAO files
source(here::here("workflow", "R", "common.R")) # directory locations
source(here::here(current_prs_folder, "R", "np_fxn.R"))
```

# Import Raw Data: FAO Commodities

Simultaneously read and process FAO commodities value and quantity data.
```{r}
## NOTE: This can be run as a loop, but the "value" and "quant" datasets need to be run individually to make sure
##  there are no problems (after this check, they can be looped for efficiency)

# describe where the raw data are located:
dir_fao_data <- here::here(dir_M, "git-annex", "globalprep", "_raw_data", "FAO_commodities", data_dir_year)

# list files included in d2024 folder (value and quant datasets)
files <- list.files(dir_fao_data, pattern = glob2rx('*.csv'), full.names = TRUE)

## To compare to old data:
# dir_fao_data <- file.path(dir_M, 'git-annex/globalprep/_raw_data/FAO_commodities/d2020')
# files <- list.files(dir_fao_data, pattern=glob2rx('*.csv'), full.names=T)
## To compare to 2023
#old_dir_fao_data <- file.path(dir_M, 'git-annex/globalprep/_raw_data/FAO_commodities/d2023')
#old_files <- list.files(old_dir_fao_data, pattern=glob2rx('*.csv'), full.names=T)
#test_old_quant <- read.csv(old_files[1], check.names = FALSE, strip.white = TRUE, stringsAsFactors = FALSE)

# ========== Data processing loop ===================
tic()
for (f in files) {  
  #f = files[2] # uncomment to run line-by-line to test each file
  # Print name of file in current iteration
  cat(sprintf('\n\n\n====\nfile: %s\n', basename(f)))
  
  # ---- Read in data ----
  d <- readr::read_csv(f) # testing out tibble option for reading in data
  #d <- read.csv(f, check.names = FALSE, strip.white = TRUE, stringsAsFactors = FALSE) # stringsAsFactors=T
  # checks names syntactically, strips leading and trailing whitespace, prevents conversion of characters to factors 
  
  # ---- Specify units ----
  ## Specifies that units are tonnes if we are reading in the Commodities Quantity data csv, and usd if we are reading in the Commodities Value data csv
  units <- c('tonnes','usd')[str_detect(f, c('quant','value'))] # detect unit name using lowercase American English

  # ---- Clean and tidy data ----
  ## gather into long format and clean up FAO-specific data foibles
  ## warning: attributes are not identical across measure variables; they will be dropped: this is fine (didn't get this warning in 2024)
  m <- d %>% 
    janitor::clean_names() %>% 
    dplyr::select(-c(unit_name)) %>% # "Tonnes – net product weight" == TPW
    rename(country = reporting_country_name,
           commodity = commodity_name,
           trade = trade_flow_name) %>%
    rename_with(~ gsub("x", "", .)) %>% # tidy up year column names (clean_names() added "x"s)
    pivot_longer(cols = -c(country, commodity, trade, unit),
                   names_to = "year", values_to = "value") 
  
  # ---- Filter to only include "Exports" data
  m <- m %>%
    filter(trade == "Exports")

  # ---- Run fao data cleaning function ----
  # cleans up flags, swaps out FAO-specific codes for analysis
  m <- m %>%
    fao_clean_data_new() %>%  # swaps out FAO-specific codes. NOTE: optional parameter 'sub_N' can be passed to control how an 'N' code is interpreted.
    dplyr::select(-c(trade, unit)) %>% # eliminate 'trade' and 'unit' columns
    arrange(country, commodity, is.na(value), year)

  # 2024 note -- warning message: NAs introduced by coercion (in value = as.numeric(as.character(value)) from line 65 in fao_fxn.R)
  
  # ---- Products join ----
  ## attach product categories from com2prod, and filter out all entries that 
  ## do not match a product category.
  ## Note: commodity_lookup is user-defined function to compare commodities in
  ## data vs commodities in lookup table
  
  # Load lookup for converting commodities to products
  #com2prod <- read.csv(here::here(current_prs_folder, "raw", "commodities2products_weighting.csv"), na.strings = '')
  com2prod <- readr::read_csv(here::here(current_prs_folder, "raw", "commodities2products_weighting.csv"), na = '')
  ## Check the current commodity-to-product lookup table.  If necessary, make changes to "raw/commodities2products_weighting.csv"
  ## For prs_coral_harvest, The main product we're concerned about is "corals". The other natural product commodities are more important for the np goal and should be inspected more carefully for updating the lookup table for that goal
  np_commodity_lookup(m, com2prod) 
  ## v2024: added shells,Powder and waste of shells; corals,Miscellaneous corals and shells
    
  # Perform inner join
  ## inner_join will attach product names to matching commodities according to
  ## lookup table 'com2prod', and eliminate all commodities that do not appear in the lookup table.
  m <- m %>%
      inner_join(com2prod, by = 'commodity')
    
    
  # ---- Run name_2_rgn ----
  # Prep:
  ## Special case: user-defined function deals with 
  ##   breaking up Antilles into separate reported rgns
  m <- ohicore::split_regions(m) # used to use np_split_antilles(); this incorporates more than just antilles
  
  # Run name_2_rgn to filter to OHI regions, assign rgn_ids, standardize rgn_names
  m_rgn <- ohicore::name_2_rgn(df_in = m,
                               fld_name = "country", 
                               flds_unique = c("commodity", "product", "year"))
  
# v2023 removed countries:
#   These data were removed for not having any match in the lookup tables:
# 
#           yugoslavia sfr 
#                        1 (not used and only has data in 1984-1988)
    
# v2023 duplicates:  [1] "China"                 "China, Hong Kong SAR"  "China, Macao SAR"      "Guadeloupe"           
# [5] "Martinique"            "Montenegro"            "Russian Federation"    "Serbia and Montenegro"
# [9] "Sudan"                 "Sudan (former)"        "Un. Sov. Soc. Rep."     - these are all fixed in the group by and summarise below  
  
  # v2024: removed yugoslavia sfr in both quantity and value
  # ascension saint helena and tristan da cunha in quantity and value -- updated in ohicore::split_regions()
  # 85,World,Europe,Northern Europe,Ascension
  # 86,World,Europe,Northern Europe,Saint Helena
  # 88,World,Europe,Northern Europe,Tristan da Cunha
  # these do not appear in the previous year's data, and there is no coral data within this aggregation of groups. Updated ohicore::split_regions() to include the new case of this group so that it is properly addressed in the np goal's dataprep (the region does have some data for other commodity groups including fish_oil and ornamental fish)
  
  
  # v2024: had the same duplicates as v2023; all are fixed in the group by and summarize (aggregating groups by ohi rgn_id among other variables, which sums the "duplicates" above into appropriate rgn_id groups)
  
  # ---- Aggregate: combine composite regions ----
  ## When summarizing the dataset, this function provides a modified way to sum the value column while maintaining NA values when both variables are NA (rather than turning to zero values). The function will sum non-NA values normally.
  sum_function <- function(x) {
    if (sum(is.na(x)) == length(x)) 
      return(NA)
    return(sum(x, na.rm = T))}
  
  # Apply sum function
  m_rgn <- m_rgn %>%
    dplyr::group_by(rgn_id, rgn_name, commodity, product, year) %>%
    dplyr::summarize(value = sum_function(value)) %>% # alternate: .groups = "drop (currently experimental)
    dplyr::ungroup()

  # ---- Write out to int folder ----
  # prep units: rename value field to units based on filename
  names(m_rgn)[names(m_rgn) == 'value'] <- units  
  
  # output to .csv - should create two csv files (tonnes.csv and usd.csv)
  harvest_out <- sprintf(here(paste0('globalprep/prs_coral_harvest/v', scen_year, '/int/%s.csv')), units)
  readr::write_csv(m_rgn, harvest_out, na = '')
  #write.csv(m_rgn, harvest_out, row.names = FALSE, na = '')
}
toc()
# v2023: 152.934 sec elapsed
# v2024: 96.25 sec elapsed
```


### Testing issue with gapfilling:
```{r ,eval=FALSE}
# ==== Data investigation ====
# there appear to be no NAs in the "value" dataset -- this is causing issues
# during gapfilling, as the np_regr_coef() function uses 
# np_regr_fill(), which gapfills NAs (and expects there to be NAs)
#test_na[,1]


quant_file <- files[1]
# value_file <- files[2]

# these methods yield identical datasets
quant_raw <- read.csv(quant_file, check.names = FALSE, strip.white = TRUE, stringsAsFactors = FALSE)
quant_raw_new <- readr::read_csv(quant_file)

units <- c('tonnes','usd')[str_detect(quant_file, c('quant','value'))] # detect unit name using lowercase American English

quant_raw_clean <- quant_raw_new %>% janitor::clean_names()

quant_raw_tidy <- quant_raw_clean %>% 
  dplyr::select(-c(unit_name)) %>% 
  dplyr::rename(country = reporting_country_name,
         commodity = commodity_name,
         trade = trade_flow_name) %>%
  dplyr::rename_with(~ gsub("x", "", .)) %>% # tidy year columns
  dplyr::mutate_if(is.numeric, as.character) %>% # setting datatypes for pivoting (we want to include Flag columns for use in processing)
  tidyr::pivot_longer(cols = -c(country, commodity, trade, unit),
                   names_to = "year", values_to = "value") 

```


# Data Wrangle

Read in the tonnes and usd data that was completed in the Natural Products ornamentals dataprep (and above). Combining the quantity and value data and a bit of cleaning to remove data prior to first reporting year for coral commodities and regions. 
```{r, eval=FALSE}
## Read in quant dataset from intermediate folder
h_tonnes <- readr::read_csv(here::here(current_int_folder, "tonnes.csv"))

## Read in value dataset from intermediate folder
h_usd <- readr::read_csv(here::here(current_int_folder, "usd.csv"))

## concatenates h_tonnes and h_usd data
## h includes rgn_name, rgn_id, commodity, product, year, tonnes, and usd.
h <- h_usd %>%
    full_join(h_tonnes, by = c("rgn_name", "rgn_id", "commodity", "product", "year")) %>%
    mutate(commodity = as.character(commodity)) %>%
    arrange(rgn_id, product, commodity, year) # %>%
  # dplyr::filter(product == "corals") ##filter for our commodities of interest

## clips out years prior to first reporting year, for each commodity per region
h_clip <- h %>% np_harvest_preclip()
```


```{r , eval=FALSE}
# test plot
library(RColorBrewer)

h_agg_plot_df <- h_clip %>% 
  dplyr::ungroup() %>% 
  dplyr::group_by(rgn_id, rgn_name, product, year) %>% 
  summarize(total_usd = sum(usd, na.rm = TRUE),
            total_tonnes = sum(tonnes, na.rm = TRUE),
            .groups = "drop")

h_product_agg_df <- h_clip %>% 
  dplyr::ungroup() %>% 
  dplyr::group_by(product, year) %>% 
  summarize(total_usd = sum(usd, na.rm = TRUE),
            total_tonnes = sum(tonnes, na.rm = TRUE),
            .groups = "drop")

corals_quant_agg <- h_agg_plot_df %>% 
  filter(product == "corals")

quantity_rgn_plot <- ggplot(data = h_agg_plot_df, aes(x = year, color = rgn_name)) +
  geom_line(aes(y = total_tonnes, color = rgn_name)) +
  facet_wrap(~product) + 
  theme_bw() +
  labs(title = "Aquatic Trade Quantity per Product Category Per Region over Time",
       y = "Tonnes")

coral_agg_plot <- ggplot(data = corals_quant_agg, aes(x = year)) +
  geom_line(aes(y = total_tonnes, color = rgn_name)) +
  theme_bw() +
  labs(title = "Aquatic Trade Quantity of Corals over Time",
       y = "Tonnes")

plotly::ggplotly(coral_agg_plot)

```



# Gapfilling 

Summary of gapfilling that is performed:

  * Zero-fill: for observations with NAs for both values (tonnes & usd), fill both as zero. Also cross-fills zeros where one value is zero, other is NA.
  * Regression fill, first pass: Where enough non-zero paired observations exist at the country level, use country-level data to create regression models (tonnes ~ usd and vice versa) for gapfilling.  About 25% success. 
  * Regression fill, second pass: Where pass 1 failed, and enough non-zero paired observations exist at georegional level, use georegional-level data to create regression models (tonnes ~ usd + year, and vice versa) for gapfilling.  About 90% success. 
 * Regression fill third pass: Where passes 1 and 2 failed, use global-scale data to create  regression models (tonnes ~ usd + year, and vice versa) for gapfilling.  100% success.
 * End-fill:  For years where NAs still exist in final year, carry forward data from prior year (after other gapfilling techniques).

```{r, eval = FALSE}  
# Flag for gap-filling
h_gapflag <- h_clip %>% np_harvest_gapflag()
## Adds flag for required gap-filling, based upon NAs in data. 
## NOTE: Does not perform any gap-filling.
## At this point, h_gapflag includes: 
##    rgn_name   rgn_id   commodity   product   year   tonnes   usd   gapfill
## 'gapfill' will be in (zerofill, endfill, tbd, none)

data_check <- h_gapflag %>% np_datacheck()
## for each commodity within each region, creates (but doesn't save...) summary info:
##   num_years:        the length of the data series for this commodity in this region
##   usd_unique_nz:    (or 'tns') number of unique non-zero values for usd or tonnes 
##   usd_na & tns_na:  number of NA occurrences
##   paired_obs:       number of non-zero paired observations
##   usd_unique_pairs: (or 'tns') within set of paired observations, count of unique usd and tonnes
##   unique_pairs:     lesser of usd_unique_pairs and tns_unique_pairs
##   count_no_data:    number of paired NAs - years with no value reported

# Fill with zeroes when appropriate
h_zerofill <- h_gapflag %>% np_zerofill()
## for post-reporting years with NA for both tonnes and USD, fill zero - 
##    assumes that non-reporting indicates zero harvest to report.
## Also cross-fills zeros where one side is 0, other is NA (not flagged as gapfill)

h_lowdata_filter <- h_zerofill %>% np_lowdata_filter()
## Exclude commodities (within a region) that have few non-zero data points.
## Optional parameter with default: nonzero_h_yr_min = 4
## NOTE: This filter has consequences for the regression, but also has meaning in terms of 
##    not inflicting a penalty on regions trying, and then stopping, an experimental harvest.

## Melanie's script to add a georegional ID tag based on country keys and IDs.
h <- h_lowdata_filter %>% add_georegion_id()

# Gapfill quantity (tonnes) if value (usd) present based on model (tonnes ~ usd) or (tonnes ~ usd + year)
h <- h %>% np_regr_fill(years_back = 10, vars = 'td', scope = 'rgn_id')
h <- h %>% np_regr_fill(vars = 'tdy', scope = 'georgn_id')
h <- h %>% np_regr_fill(vars = 'tdy', scope = 'global')
## np_regr_fill() is a generalized regression gapfill function. Parameters (with defaults):
## * years_back=50 (int):    This determines how far back in the time series to include within the regression.
## * min_paired_obs=4 (int): This determines how many paired observations are required to attempt a regression.
## * scope = 'rgn_id' (str): ('rgn_id', 'georgn_id', 'global') Determines grouping scale for regression.
## * vars = 'tdy' (str):     ('td', 'tdy') Determines model: (tonnes ~ usd) or (tonnes ~ usd + year) [and vice versa]

# For final year of data, if both usd and tonnes originally reported as NA, pull forward values for usd and tonnes from the previous year. This should happen after regression fill.
h <- h %>% np_end_fill()


# Store commodity-level data, before moving on to the product-level smoothing.
h_comm <- h

## Output gapfilling report to .csv files.
## Very few usd gapfilling, and none in recent years (data used to weight contributions), so will ignore this: gapfill=="r2_u_gr"
h_gap <- h %>%
  dplyr::mutate(gapfill = ifelse(gapfill == "r2_u_gr", "none", gapfill)) %>% # focusing only on tonnes gapfilling
  dplyr::select(rgn_id, commodity, product, year, gapfill) %>%
  dplyr::filter(product == "corals")

#write.csv(h_gap, here(current_prs_folder, "output", "prs_coral_gf.csv"), row.names = FALSE, na = '')
readr::write_csv(h_gap, here(current_prs_folder, "output", "prs_coral_gf.csv"), na = '')

```

# Final Data Wranglng

## Summarize values

Summarize each product per country per year, e.g., all corals in Albania in 2011. And, do some error checking.
```{r, eval = FALSE}
# Summarize each product per country per year
h_prod <- h_comm %>%
  dplyr::filter(product == "corals") %>%
  dplyr::group_by(rgn_name, rgn_id, product, year) %>%
  summarize(tonnes = sum(tonnes, na.rm = TRUE), 
            usd = sum(usd, na.rm = TRUE))
          
# Error-checking and table exports to see if there are duplicates
stopifnot(sum(duplicated(h_prod[ , c('rgn_id', 'product', 'year')])) == 0)
```

## Quick Data Check

Look at wide format with all commodities and product subtotal (where commodity column value is "Z_TOTAL"), compared with the input data prior to summing.

```{r, eval = FALSE}
# wide format for tonnes
h_x_tonnes <- h_comm %>% 
  bind_rows(mutate(h_prod, commodity = 'Z_TOTAL')) %>%
  dplyr::select(rgn_name, rgn_id, commodity, product, year, tonnes) %>%
  dplyr::arrange(rgn_name, product, commodity, year) %>%
  pivot_wider(names_from = year, values_from = tonnes)

# wide format for usd
h_x_usd <- h_comm %>% 
  bind_rows(mutate(h_prod, commodity = 'Z_TOTAL')) %>%
  dplyr::select(rgn_name, rgn_id, commodity, product, year, usd) %>%
  dplyr::arrange(rgn_name, product, commodity, year) %>%
  pivot_wider(names_from = year, values_from = usd)

# Check a random country and commodity
australia <- h_x_usd %>% dplyr::filter(product == "corals", rgn_name == "Australia") 
australia ## perfect



# Can open up in Excel to compare subtotals per country-product-year
readr::write_csv(h_x_tonnes, here::here(current_int_folder, "coral_harvest_tonnes_wide.csv"), na = "NA")
readr::write_csv(h_x_usd, here::here(current_int_folder, "coral_harvest_usd_wide.csv"), na = "NA")
```


## Calculate Rolling Averages
Determine rolling averages for tonnes and USD in order to determine peak values.  This is based upon total harvests by product group, not individual commodity.  

```{r, eval = FALSE}
# Find max year in the summarized data table
year_max <- max(h_prod$year)
# 2022 in v2024

# Determine rolling averages for tonnes and USD
roll_prod <- h_prod %>%
  dplyr::arrange(rgn_id, product, year) %>%
  dplyr::group_by(rgn_id, product) %>% # group by region and product
  # determine rolling averages for tonnes and USD
  dplyr::mutate(
    tonnes_rollmean = rollapply(tonnes, width = 4, FUN = mean, align = "right", partial = TRUE, na.rm = FALSE),
    usd_rollmean = rollapply(usd, width = 4, FUN = mean, align = "right", partial = TRUE, na.rm = FALSE)) %>%
  dplyr::rename(
    tonnes_orig = tonnes, # prevent overwriting of reported and gapfilled values
    usd_orig = usd) %>% # prevent overwriting of reported and gapfilled values
  dplyr::mutate(
    tonnes = ifelse(!is.na(tonnes_rollmean), tonnes_rollmean, tonnes_orig),
    usd = ifelse(!is.na(usd_rollmean), usd_rollmean, usd_orig)) %>%
  dplyr::select(rgn_id, rgn_name, product, year, tonnes, usd, tonnes_orig, usd_orig)

# check grouping status
#is.grouped_df(roll_prod)

readr::write_csv(roll_prod, here::here(current_int_folder, "tonnes_coral_harvest.csv"))
```

## Calculate pressure score
 - Divide the harvest by the area of coral and take the 95th quantile of harvest 
 - Anything above the 95th quantile recieves a pressure score of 1, otherwise what it was before. 
 - Multiply the pressure score by the health score to get the final pressure score.
```{r, eval = FALSE}
# Read in production harvest data 
roll_prod <- readr::read_csv(here::here(current_int_folder, "tonnes_coral_harvest.csv"))

# check grouping status
#is.grouped_df(roll_prod)

# Read in coral extent data
coral_ext <- readr::read_csv(here::here(
  "globalprep", "hab_coral",
# paste0("v", (as.numeric(scen_year) - 2)), # error in v2024: most recent update to this folder is 2021
  "v2021", # temporary hard-coded fix (UPDATE TO MOST RECENT VERSION!!)
  "data", "habitat_extent_coral_updated.csv")) %>% 
  dplyr::select(-c(habitat, year)) %>%
  filter(km2 != 0)

# Read in coral health data 
# coral_health <- readr::read_csv(paste0("globalprep/hab_coral/v", (as.numeric(scen_year) - 2), "/data/habitat_health_coral_updated.csv")) %>% # UPDATE TO MOST RECENT VERSION
#   dplyr::select(-habitat, -year)
coral_health <- readr::read_csv(here::here(
  "globalprep", "hab_coral",
  # paste0("v", (as.numeric(scen_year) - 2)), # error in v2024: most recent update to this folder is 2021... 
  "v2021", # temporary hard-coded fix (UPDATE TO MOST RECENT VERSION!!)
  "data", "habitat_health_coral_updated.csv")) %>% 
  dplyr::select(-c(habitat, year))

# Join coral harvest data with the coral health and extent
coral_harvest_join <- roll_prod %>%
  left_join(coral_ext, by = "rgn_id") %>%
  left_join(coral_health, by = "rgn_id")

# Calculate harvest intensity 
coral_harvest <- coral_harvest_join %>%
  mutate(intensity = tonnes/km2)

# Find the 95th quantile for a reference point
ref <- quantile(coral_harvest$intensity, probs = 0.95, na.rm = TRUE) 

# check grouping status
#is.grouped_df(coral_harvest)
# if it isn't ungrouped, the select statement in the next step adds "product" 
# because it's still one of the grouping variables

# Calculate pressures for final coral harvest df
coral_harvest_prs <- coral_harvest %>%
  mutate(pressure_no_health = ifelse(intensity > ref, 1, (intensity / ref))) %>%
  mutate(pressure_health = pressure_no_health * health) %>% # calculate the pressure score
  filter(!is.na(pressure_health)) %>%
  dplyr::ungroup() %>% 
  dplyr::select(rgn_id, year, pressure_score = pressure_health) 

# test <- coral_harvest %>%
#   filter(is.na(km2))
# 
# unique(test$rgn_id)

# # highest pressure
# high_prs_c_h <- coral_harvest_prs %>% 
#   dplyr::arrange(desc(pressure_score))
```

## Save data layer
```{r, eval = FALSE}
#write.csv(coral_harvest, paste0("globalprep/prs_coral_harvest/v", scen_year, "/output/prs_coral_harvest.csv"), row.names = FALSE)
readr::write_csv(coral_harvest_prs, here::here(current_prs_folder, "output", "prs_coral_harvest.csv"))
```


Datacheck

```{r}
# get current version and previous version data and then combine for comparison
current_version_coral <- readr::read_csv(here::here(current_prs_folder, "output", "prs_coral_harvest.csv")) %>%
  rename("new_prs" = "pressure_score")

previous_version_coral <- readr::read_csv(here::here("globalprep", "prs_coral_harvest",
                                              paste0("v", prev_scen_year),
                                              "output", "prs_coral_harvest.csv")) %>%
  rename("old_prs" = "pressure_score") 

combined_coral <- previous_version_coral %>%
  left_join(current_version_coral, by = c("rgn_id", "year")) 

# if wanting to look at highest differences
changes <- combined_coral %>%
  mutate(diff = abs(old_prs - new_prs))

# select highest differences
highest_changes <- changes %>% dplyr::arrange(desc(diff))
# 208,World,Asia,South-Eastern Asia,Singapore
# 40,World,Asia,Southern Asia,Sri Lanka
# 14,World,Asia,Eastern Asia,Taiwan

# see what rows have NAs and take note if anything is odd
na_regions <- combined_coral[!complete.cases(combined_coral), ]
na_regions %>% select(rgn_id) %>% unique()
# v2023: no NAs
# v2024:
# NAs for region 127 from 1999-2021 (127: World,Latin America and the Caribbean,Caribbean,Saint Vincent and the Grenadines)
# rgn 212 from 2017-2021 (212: World,Oceania,Micronesia,Kiribati)

# check all data from previous version vs. all data from current version
plot(combined_coral$old_prs, combined_coral$new_prs,
     xlab = paste0("Old Pressure All Years v", prev_scen_year),
     ylab = paste0("New Pressure All Years v", scen_year))
abline(0,1, col = "red")

# explore some deviations v2023 & v2024
previous_version_coral_latest_yr <- previous_version_coral %>%
  filter(year == common_year) # year 2020 in v2024

current_version_coral_latest_yr <- current_version_coral %>%
  filter(year == as.character(as.numeric(scen_year) - 2)) # set to latest year of data

combined_latest_yr <- previous_version_coral_latest_yr %>%
  left_join(current_version_coral_latest_yr, by = "rgn_id") %>%
  dplyr::select(-year.x, -year.y) # 2020 and 2022 in v2024

plot(combined_latest_yr$old_prs, combined_latest_yr$new_prs,
     xlab = paste0("Old Pressure Data Year ", common_year, " v", prev_scen_year),
     ylab = paste0("New Pressure Data Year ", (as.numeric(scen_year) - 2), " v", scen_year))
abline(0,1, col = "red")

# compare within the new data itself 
# 2020 vs. 2021 since both years are new v2023
# 2021 vs. 2022 in v2024
current_version_coral_prior_yr <- current_version_coral %>%
  filter(year == as.character(as.numeric(scen_year) - 3)) # set to 1 prior to latest year of data

combined_two_recent_yrs <- current_version_coral_prior_yr %>%
  rename(prior_prs = new_prs) %>%
  left_join(current_version_coral_latest_yr, by = "rgn_id") %>%
  select(-year.x, -year.y)

plot(combined_two_recent_yrs$prior_prs, combined_two_recent_yrs$new_prs,
     xlab = paste0("New Pressure Data Year ", (as.numeric(scen_year) - 3), " v", scen_year),
     ylab = paste0("New Pressure Data Year ", (as.numeric(scen_year) - 2), " v", scen_year))
abline(0,1, col = "red")

# check the most recent year in common in both versions
combined_coral_common_yr <- combined_coral %>%
  filter(year == common_year)

plot(combined_coral_common_yr$old_prs, combined_coral_common_yr$new_prs,
     xlab = paste0("Old Pressure Data Year ", (as.numeric(prev_scen_year) - 3), " v", prev_scen_year),
     ylab = paste0("New Pressure Data Year ", (as.numeric(scen_year) - 4), " v", scen_year))
abline(0,1, col = "red")

# comparing changes in the data for the same year
data_yr_changes <- combined_coral_common_yr %>%
  mutate(diff = abs(old_prs - new_prs))

data_yr_top_diff <- data_yr_changes %>% dplyr::arrange(desc(diff))
#data_yr_top_diff

# glance at rolling production for one region
roll_prod_check <- readr::read_csv(here::here(current_int_folder, "tonnes_coral_harvest.csv")) %>%
  filter(rgn_id == 207)
```

