---
title: 'OHI 2020: Natural Products: weighting data prep'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

[REFERENCE RMD FILE: http://ohi-science.org/ohiprep_v2020/globalprep/np/v2020/np_dataprep.html]

# Summary
This analysis converts FAO commodities tonnes data for ornamentals, Watson fisheries tonnes data for FOFM, and FAO mariculture tonnes data for seasweeds into a weighting scheme per each ohi region, based off of 5 year averages of tonnes and values (in usd) for each product. This weighting scheme will be applied to the scores when updating in ohi global. 


***

# Data Source

**Reference**:      
    http://www.fao.org/fishery/statistics/software/fishstatj/en#downlApp
    App release date: July 2019 
FAO raw commodities quantity 1976_2017
FAO raw commodities value 1976_2017
FAO metadata found [here](http://www.fao.org/fishery/statistics/global-commodities-production/en)

**Downloaded**: May 3 2020

**Description**:  Quantity (tonnes) and value (USD) of raw commodities (Exports only) for each country, taxa, year.  The FAO data is subset to include commodities in these categories: shells, corals, ornamental fish, fish oil, seaweed and plants, sponges (see: raw/commodities2products.csv for details).

**Time range**: 1976-2017


**Reference**: Watson, R. A. and Tidd, A. 2019. Mapping nearly a century and a half of global marine fishing: 1869–2017. Marine Policy, 93, pp. 171-177. [(Paper URL)](https://doi.org/10.1016/j.marpol.2018.04.023)

**Downloaded**: December 11, 2019 from [IMAS portal](http://data.imas.utas.edu.au/portal/search?uuid=ff1274e1-c0ab-411b-a8a2-5a12eb27f2c0) - click on download tab, step 3

**Description**:  Global fisheries landings data. 

**Native data resolution**:   

**Time range**: 1950 - 2017

**Format**:  CSV format

**Additional Information**: [Metadata](http://metadata.imas.utas.edu.au/geonetwork/srv/eng/metadata.show), [Supplementary Material](https://ars.els-cdn.com/content/image/1-s2.0-S0308597X18300605-mmc1.docx)


**Reference**: Froehlich, H.E., Jacobsen, N.S., Essington, T.E., Clavelle, T., and Halpern, B.S. (2018). Avoiding the ecological limits of forage fish for fed aquaculture. Nature Sustainability 1, 298.

**Downloaded**: July 7, 2020. Obtained from Melanie Frazier (NCEAS). 

**Description**:  List of FOFM species from Watson v3 data. 

**Native data resolution**:   

**Format**:  CSV format


**Reference**: [RAM Legacy Stock Assessment Database](http://ramlegacy.org) v4.491

**Downloaded**: 06/10/2020

**Description**: B/Bmsy value by stock and year (other data, which we do not use, are also available in the database)

**Native data resolution**: stock (fish stock, species and region specific)

**Time range**: 1950 - 2016

**Format**: CSV format 

**Additional Information**: We use the finalized b/bmsy layer from OHI-global for this data prep. We do not actually read in the raw RAM data here. 

* DOI: 10.5281/zenodo.2542919


**Reference**: [FAO Mariculture data](http://www.fao.org/fishery/statistics/software/fishstatj/en#downlApp)
Release date: March 2019
FAO Global Aquaculture Production Quantity 1950_2018
FAO metadata found [here](http://www.fao.org/fishery/statistics/global-aquaculture-production/en)

**Downloaded**: 5/11/2020

**Description**:  Quantity (tonnes) of mariculture for each country, species, year.

**Time range**: 1950-2018 
***

## Setup

```{r setup, warning=FALSE, message=FALSE}

knitr::opts_chunk$set(eval=FALSE)

library(here)
library(tidyverse)
source(here('workflow/R/common.R'))
source(here('globalprep/np/v2020/R/np_fxn.R'))

```
  
## Methods

Import prepped FAO commodities data set and find 5 year aaverage $USD / Tonne average value per product per region. Gapfill these based on the UN geopolitical region/global values. 
```{r}
## read in the prepped FAO commodities data set 
harvest_tonnes_usd <- read_csv("int/np_harvest_tonnes_usd.csv") %>%
  filter(rgn_id != 213)

max_year = max(harvest_tonnes_usd$year)
min_year = max_year - 4 
years = c(min_year:max_year)
years_df = data.frame(year = min_year:max_year) ## so that we can get 5 year average

region_data()

rgns_eez <- rgns_eez %>%
  dplyr::select(rgn_id) %>%
  filter(rgn_id != 213)

## make a data frame with every region and every product
products <- data.frame(product = c("seaweeds", "ornamentals", "fish_oil"))

region_product <- full_join(rgns_eez, products, by = character()) %>%
  full_join(years_df, by = character())

## calculate $/tonne for each ohi region and product and gapfill based on UN geopolitical region 

harvest_tonnes_usd_geo_rgn <- harvest_tonnes_usd %>%
    full_join(region_product, by = c("rgn_id", "product", "year")) %>%
  add_georegion_id() %>%
  dplyr::filter(year %in% years) %>%
  arrange(rgn_id)
  
  
geo_rgn_values_df <- harvest_tonnes_usd_geo_rgn %>%
  dplyr::group_by(georgn_id, year, product) %>%
  dplyr::mutate(values = usd/tonnes) %>%
    mutate(values = ifelse(values == "Inf", NA, values)) %>% ## assign Inf as NA
  summarise(georgn_values = mean(values, na.rm = TRUE)) %>% 
  ungroup() ## now we have a 5 year average value for geo region

global_values_df <- harvest_tonnes_usd_geo_rgn %>%
  dplyr::group_by(product, year) %>%
  mutate(values = usd/tonnes) %>%
  mutate(values = ifelse(values == "Inf", NA, values)) %>% ## assign Inf as NA
  summarise(global_values = mean(values, na.rm = TRUE)) ## now we have a 5 year average global value

harvest_tonnes_usd_values_gf <- harvest_tonnes_usd_geo_rgn %>%
  left_join(geo_rgn_values_df, by = c("georgn_id", "product", "year")) %>%
  left_join(global_values_df, by = c("product", "year")) %>%
  mutate(values = usd/tonnes) %>%
  mutate(values = ifelse(values == "Inf", NA, 
                         ifelse(values == "NaN", NA, values))) %>% ## assign Inf and NaNs as NA
  mutate(values_final = ifelse(is.na(values) & georgn_values != "NaN", georgn_values,
                               ifelse(is.na(values) & georgn_values == "NaN", global_values, values))) %>%## gapfill so that when there is no regional value, take the geo regional value, and if there is no geo regional value, take the global value. 
    mutate(values_gf_description = ifelse(is.na(values) & georgn_values != "NaN", "geo_region",
                               ifelse(is.na(values) & georgn_values == "NaN", "global", "none"))) %>%
  mutate(values_gf = ifelse(is.na(values) & georgn_values != "NaN", 1,
                               ifelse(is.na(values) & georgn_values == "NaN", 1, 0))) ## now we have a dataset with gapfilled average values and calculated yearly values... now we will save this, and then calculate the 5 year averages

## save this data frame to int
write.csv(harvest_tonnes_usd_values_gf, "int/harvest_tonnes_usd_weighting_gf.csv", row.names = FALSE)

## now calculate 5 year averages and assign year = 2020
weighting_usd_values_final <- harvest_tonnes_usd_values_gf %>%
  dplyr::select(rgn_id, product, year, values_final) %>%
  group_by(rgn_id, product) %>%
  summarise(value_per_tonne = mean(values_final)) %>%
  ungroup() 

#save this df
write.csv(weighting_usd_values_final, "int/harvest_weighting_values.csv", row.names = FALSE)

```

Now prep the tonnes data from each product to be per each region for the 5 year average (2013-2017). 
```{r}
## Read in FOFM tonnes data
fofm_tonnes <- read_csv("int/mean_catch_FOFM.csv") %>%
  dplyr::filter(year %in% years) %>%
  mutate(product = "fish_oil") %>%
  group_by(rgn_id, year, product) %>%
  summarise(tonnes = sum(catch)*0.3) %>% ## Note: multiply by 0.3 to account for water loss when converting fish to fish oil.. about 30% of fish are water and 70% are fish oil
  ungroup() %>%
  group_by(rgn_id, product) %>%
  summarise(tonnes = mean(tonnes)) %>% ##calculate 5 year mean of catch
  ungroup()


## Read in ornamentals tonnes data
np_tonnes <- read_csv("int/np_harvest_tonnes_usd.csv") %>% 
  filter(rgn_id != 213)

orn_fill_df <- region_product %>%
  filter(product == "ornamentals")
  
orn_tonnes <- np_tonnes %>%
  dplyr::filter(year %in% years, product == "ornamentals") %>%
  full_join(orn_fill_df, by = c("rgn_id", "year", "product")) %>% ## full join to fill in the rest of the regions with 0
  mutate(tonnes = ifelse(is.na(tonnes), 0, tonnes),
         usd = ifelse(is.na(usd), 0, usd)) %>% ## gapfill the NAs to be 0
  select(rgn_id, year, product, tonnes) %>%
  group_by(rgn_id, product) %>%
  summarise(tonnes = mean(tonnes)) %>% ## calculate 5 year mean of catch
  ungroup()
  
## Read in seaweed tonnes data
sw_fill_df <- region_product %>%
  filter(product == "seaweeds")

sw_tonnes <- read_csv("int/np_seaweeds_tonnes_weighting.csv") %>%
  mutate(product = "seaweeds") %>%
  group_by(rgn_id, year, product) %>%
  summarise(tonnes = sum(tonnes, na.rm = TRUE)) %>%
  dplyr::filter(year %in% years) %>%
  full_join(sw_fill_df, by = c("rgn_id", "year", "product")) %>%
    mutate(tonnes = ifelse(is.na(tonnes), 0, tonnes)) %>% ## gapfill the NAs to be 0
  dplyr::select(rgn_id, year, product, tonnes) %>%
  ungroup() %>%
  group_by(rgn_id, product) %>%
  summarise(tonnes = mean(tonnes)) %>% ## calculate 5 year average
  ungroup()


```

Now we will calculate the weights per each product. To do this we need to multiply our average $ value for each product * tonnes of each product, and then divide by the total per each region. We will also assign year = 2020 so that this can be read into OHI-global (2020 corresponds to the year that these weights were calculated).

```{r}
harvest_weighting_values <- read_csv("int/harvest_weighting_values.csv") %>%
  dplyr::select(rgn_id, product, value_per_tonne)

prod_weights <- orn_tonnes %>%
  bind_rows(sw_tonnes) %>%
  bind_rows(fofm_tonnes) %>% 
  left_join(harvest_weighting_values, by = c("rgn_id", "product")) %>%
  arrange(rgn_id) %>%
  mutate(usd_product = tonnes*value_per_tonne) %>%
  group_by(rgn_id) %>%
  mutate(total_usd = sum(usd_product)) %>%
  ungroup() %>%
  mutate(weight = usd_product/total_usd) %>%
  dplyr::select(-3, -4, -5, -6) %>%
  mutate(year = 2020) %>%
  dplyr::select(rgn_id, year, product, weight)

write.csv(prod_weights, "output/np_product_weights.csv", row.names = FALSE)
```

```{r}
##datacheck
old_prod_weights <- read_csv("../v2020/output/np_harvest_weights_from_usd.csv")

check <- prod_weights %>%
  rename("new_weight" = "weight") %>%
  left_join(old_prod_weights, by = c("rgn_id", "product"))

plot(check$new_weight, check$weight)
abline(0,1, col="red")

check_sw <- check %>%
  filter(product == "seaweeds")

plot(check_sw$new_weight, check_sw$weight)
abline(0,1, col="red")

check_orn <- check %>%
  filter(product == "ornamentals")

plot(check_orn$new_weight, check_orn$weight)
abline(0,1, col="red")

check_fofm <- check %>%
  filter(product == "fish_oil")

plot(check_fofm$new_weight, check_fofm$weight)
abline(0,1, col="red")

```

