---
title: "OHI 2021: STEP5: Calculate prop of manure leaching and volaizing for manure applied to crops"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---

## Summary
Creates rasters for 1) the proportion manure that leaches after being applied to soils and 2) the proportion of manure that volatizes after being applied to soils

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(countrycode)
library(stringr)
library(vroom)
library(doParallel)

source(here('workflow/R/common.R'))

region_data()




nutrient_d2021_anx <- file.path("/home/shares/ohi/git-annex/globalprep/_raw_data/FAOSTAT/manure_nutrient/d2021/")
git_manure <- file.path(here(), "globalprep/prs_land-based_nutrient/v2021/STEP2_manure")
location_tifs_path <- file.path("/home/shares/ohi/git-annex/globalprep/_raw_data/halpern_food_systems/d2021/animal_farm/location_tifs")
prep <- file.path("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2021")
git_crop <- file.path(here(), "globalprep/prs_land-based_nutrient/v2021/STEP1_crop")

source(here(git_crop, "R/common_spatial.R")) # read in necessary spatial files 

un <- read_csv(here("globalprep/spatial/v2021/UNSD_Methodology.csv")) %>% 
  dplyr::select(iso3c, Global_Name, Region_Name, Sub_region_Name, Intermediate_Region_Name)
```


From Bouwmen et all 2009: "in most industrialized countries 50% of stored and avalibale manure is applied to arable land and the remainder to grassland. In most developing countries, 95% of the avaliable manure is assumed to be applied to cropland and 5% to grassland."

Need to calculate:
  (1) withdraw; table V NUE Bouwman et al 2005
  (2) volatization NH3; table 5 Bouwman
  (3) denitrification N2O = 0.01, N2 = *10.2 for N2O + N2
  (4) leaching will be whatever N is left
  
  
So we will do the leaching first:
  
1. Withdraw  
```{r}

bouw_2005_tb_5 <- read_csv(here::here(git_crop, "raw/bouwman_2005_tbl5.csv"))

withdraw <-  bouw_2005_tb_5 %>% 
  mutate(adj_sys_recov = system_n_recovery/100) %>% 
  select(iso3c, adj_sys_recov) %>% 
  right_join(food_rgns_xy) %>%
    arrange(x,y) %>%
  mutate(cell_id = 1:9331200) # create unique cell_id to join by 


```

Volatization
```{r}

bouw_2002_tb_5 <- vroom::vroom(here::here(git_crop, "raw/bouwman_2002_tbl5.csv"))

volatize <- left_join(food_rgns_xy, bouw_2002_tb_5) %>% 
  select(x,y,iso3c, prop_nh3_n_cropsandrice) %>% 
    arrange(x,y) %>%
  mutate(cell_id = 1:9331200) # create unique cell_id to join by 


```
Denitrification
```{r}

denitr <- food_rgns_tif
denitr[denitr > 0] <- 10.2*0.008

denitr <- raster_df(denitr) %>% 
  right_join(food_rgns_xy) %>% 
  rename(denitri = food_rgns) %>%
    arrange(x,y) %>%
  mutate(cell_id = 1:9331200) # create unique cell_id to join by 


# check
unique(denitr$denitri)


```

## Leaching
Calculate how much N is left to leach
```{r}

files <- list.files(file.path(prep, "animal_farm_int/STEP4_N_prop_manure_applied/applied_to_crops/"), full.names = TRUE)

for(file in files){

  ############  
  ## Get the mapped head counts for each system.
  ## (only really used in step 1, to check data)
  
     # file <- files[18]
  
    ## grab the names of all the parts from the file
    animal_name <- str_split(str_split(file, pattern = "//")[[1]][2], pattern = "_")[[1]][1]
    system_name <- str_split(str_split(file, pattern = "//")[[1]][2], pattern = "_")[[1]][2]
    product_name <- str_split(str_split(file, pattern = "//")[[1]][2], pattern = "_")[[1]][3]
    
    N_applied <- read_csv(file, col_types = c("ddcn")) %>%
        arrange(x,y) %>%
  mutate(cell_id = 1:9331200) # create unique cell_id to join by 

    
    prop_leached <- left_join(N_applied, withdraw, by = "cell_id") %>% 
      left_join(volatize, by = "cell_id") %>% 
      left_join(denitr, by = "cell_id") %>% 
      dplyr::select(x = x.x, y=y.x, iso3c = iso3c.x, prop_N_applied_crops, withdraw = adj_sys_recov, volatize =prop_nh3_n_cropsandrice, denitrification= denitri) %>%
      mutate(leach = (1-(withdraw+volatize+denitrification)),
             leach = ifelse(leach <0, 0, leach),
             prop_leach = prop_N_applied_crops*leach)
    
    # test <- prop_leached %>%
    #   dplyr::select(x,y,prop_leach) %>%
    #   rasterFromXYZ(crs=food_crs)
    # plot(log(test+1))

    prop_leached_gf <- left_join(prop_leached, un, by = "iso3c") %>% 
      group_by(Intermediate_Region_Name) %>% 
      mutate(prop_leach = ifelse(is.na(prop_leach), mean(prop_leach, na.rm = TRUE), prop_leach)) %>% 
      ungroup()%>% 
      group_by(Sub_region_Name) %>% 
      mutate(prop_leach = ifelse(is.na(prop_leach), mean(prop_leach, na.rm = TRUE), prop_leach)) %>% 
      ungroup() %>% 
      group_by(Region_Name) %>% 
      mutate(prop_leach = ifelse(is.na(prop_leach), mean(prop_leach, na.rm = TRUE), prop_leach)) %>% 
      ungroup() %>% 
      group_by(Global_Name) %>% 
      mutate(prop_leach = ifelse(is.na(prop_leach), mean(prop_leach, na.rm = TRUE), prop_leach)) %>% 
      ungroup() %>% 
      dplyr::select(x,y, iso3c, N_prop_leach_applied = prop_leach)
    
    write_csv(prop_leached_gf,
                file.path(paste0(prep, "/animal_farm_int/STEP5_N_leached_prop_applied_crops/",
                                 animal_name, "_",
                                 system_name,"_",
                                 product_name,
                                 "_n_leached_prop_applied_crops.csv")))
    
}

```

Check to see if anything weird happened
```{r}
df_list <- list.files(file.path(prep,"animal_farm_int/STEP5_N_leached_prop_applied_crops/"), full.names = TRUE)


df <- df_list[5]

test <- read_csv(df) %>%
  dplyr::select(x,y,N_prop_leach_applied) %>%
  rasterFromXYZ(crs = food_crs)
plot(test)

df_check <- data.frame(iso3c = NA, animal = NA, system = NA, product = NA, N_prop_leach_applied = NA)
  
for(df in df_list){
  
#  df <- df_list[11]
    ## grab the names of all the parts from the file
    animal_name <- str_split(str_split(df, pattern = "//")[[1]][2], pattern = "_")[[1]][1]
    system_name <- str_split(str_split(df, pattern = "//")[[1]][2], pattern = "_")[[1]][2]
     
  ## values are the same throughout the country so a median or mean function won't make a difference
    country_sum <- read_csv(df) %>% 
      group_by(iso3c) %>%
      mutate(animal = animal_name,
             system = system_name,
             product = product_name) %>% 
      dplyr::mutate(N_prop_leach_applied = mean(N_prop_leach_applied, na.rm = TRUE)) %>% 
       dplyr::select(iso3c, N_prop_leach_applied, animal, system, product) %>%
      distinct()
  
    df_check <- rbind(df_check, country_sum) %>% 
      filter(!is.na(iso3c)) 
    
}

## look at it

ggplot(df_check, aes(x = animal, y = N_prop_leach_applied, color = system) ) +
  geom_point(position = "jitter") +
  labs(title = "Proportion of all manure leached from applied to soils")


```


Let's look at just the USA cows-mixed-milk, cows-grassland-milk, chicken-industrial

```{r}
df_list <- list.files(file.path(prep, "animal_farm_int/STEP5_N_leached_prop_applied_crops/"), full.names = TRUE)
df_list <- str_subset(pattern = "cows", df_list)
df_file <- df_list[4]

    ## grab the names of all the parts from the file
    category_name <- str_split(df_file, pattern = "//")[[1]][2]
  
  ## values are the same throughout the country so a median or mean function won't make a difference
    country_sum_mixed <- read_csv(df_file, col_types = "ddcn") %>% 
      group_by(iso3c) %>% 
      dplyr::summarise(prop_applied_leaching = mean(N_prop_leach_applied, na.rm = TRUE)) %>% 
      mutate(category = category_name,
             animal_system_product = word(category, 1,3, sep="_")) %>% 
       dplyr::select(iso3c,prop_applied_leaching, animal_system_product) %>% 
      filter(iso3c == "USA")
    
    df_file <- df_list[2]

    ## grab the names of all the parts from the file
    category_name <- str_split(df_file, pattern = "//")[[1]][2]
  
  ## values are the same throughout the country so a median or mean function won't make a difference
    country_sum_pasture <- read_csv(df_file, col_types = "ddcn") %>% 
      group_by(iso3c) %>% 
      dplyr::summarise(prop_applied_leaching = mean(N_prop_leach_applied, na.rm = TRUE)) %>% 
      mutate(category = category_name,
             animal_system_product = word(category, 1,3, sep="_")) %>% 
       dplyr::select(iso3c,prop_applied_leaching, animal_system_product) %>% 
      filter(iso3c == "USA")
    
  country_sum <- rbind(country_sum_pasture, country_sum_mixed)  

write_csv(country_sum, here::here(git_manure, "data_check/usa_cows_leached_applied_crops.csv"))

```


## Volatizing
Calculate how much N is volatizing
```{r}

files <- list.files(file.path(prep, "animal_farm_int/STEP4_N_prop_manure_applied/applied_to_crops/"), full.names = TRUE)

registerDoParallel(5)   
foreach(file = files) %dopar% {

  ############  
  ## Get the mapped head counts for each system.
  ## (only really used in step 1, to check data)
  
     # file <- files[6]
  
    ## grab the names of all the parts from the file
    animal_name <- str_split(str_split(file, pattern = "//")[[1]][2], pattern = "_")[[1]][1]
    system_name <- str_split(str_split(file, pattern = "//")[[1]][2], pattern = "_")[[1]][2]
    product_name <- str_split(str_split(file, pattern = "//")[[1]][2], pattern = "_")[[1]][3]
    
    N_applied <- read_csv(file, col_types = c("ddcn")) %>%
              arrange(x,y) %>%
  mutate(cell_id = 1:9331200) # create unique cell_id to join by 
    
    prop_volt <- left_join(N_applied, volatize, by = "cell_id") %>% 
      dplyr::select(x=x.x,y=y.x, iso3c=iso3c.x, prop_N_applied_crops, volatize =prop_nh3_n_cropsandrice) %>%
      mutate(prop_volt = prop_N_applied_crops*volatize) %>% 
      dplyr::select(x,y, iso3c, prop_volt)

    prop_volt_gf <- left_join(prop_volt, un, by = "iso3c") %>% 
      group_by(Intermediate_Region_Name) %>% 
      mutate(prop_volt = ifelse(is.na(prop_volt), mean(prop_volt, na.rm = TRUE), prop_volt)) %>% 
      ungroup()%>% 
      group_by(Sub_region_Name) %>% 
      mutate(prop_volt = ifelse(is.na(prop_volt), mean(prop_volt, na.rm = TRUE), prop_volt)) %>% 
      ungroup() %>% 
      group_by(Region_Name) %>% 
      mutate(prop_volt = ifelse(is.na(prop_volt), mean(prop_volt, na.rm = TRUE), prop_volt)) %>% 
      ungroup() %>% 
      group_by(Global_Name) %>% 
      mutate(prop_volt = ifelse(is.na(prop_volt), mean(prop_volt, na.rm = TRUE), prop_volt)) %>% 
      ungroup() %>% 
      dplyr::select(x,y, iso3c, N_prop_volt_applied = prop_volt)
    
    write_csv(prop_volt_gf,
                file.path(paste0(prep, "/animal_farm_int/STEP5_N_volatizing_prop_applied_crops/",
                                 animal_name, "_",
                                 system_name,"_",
                                 product_name,
                                 "_n_volatized_prop_applied_crops.csv")))
    
}

```

Check to see if anything weird happened
```{r}
df_list <- list.files(file.path(prep,"animal_farm_int/STEP5_N_volatizing_prop_applied_crops/"), full.names = TRUE)

df_check <- data.frame(iso3c = NA, animal = NA, system = NA, product = NA, N_prop_volt_applied = NA)
  
for(df in df_list){
  
#  df <- df_list[4]
    ## grab the names of all the parts from the file
    animal_name <- str_split(str_split(df, pattern = "//")[[1]][2], pattern = "_")[[1]][1]
    system_name <- str_split(str_split(df, pattern = "//")[[1]][2], pattern = "_")[[1]][2]
    product_name <- str_split(str_split(df, pattern = "//")[[1]][2], pattern = "_")[[1]][3]
     
  ## values are the same throughout the country so a median or mean function won't make a difference
    country_sum <- read_csv(df) %>% 
      group_by(iso3c) %>%
      dplyr::mutate(N_prop_volt_applied = mean(N_prop_volt_applied, na.rm = TRUE)) %>% 
      mutate(animal = animal_name,
             system = system_name,
             product = product_name) %>% 
       dplyr::select(iso3c, N_prop_volt_applied, animal, system, product) %>%
      distinct()
  
    df_check <- rbind(df_check, country_sum) %>% 
      filter(!is.na(iso3c)) 
    
}

## look at it

ggplot(df_check, aes(x = animal, y = N_prop_volt_applied, color = system) ) +
  geom_point(position = "jitter") +
  labs(title = "Proportion of all manure voltized from applied to soils")


```