---
title: "OHI 2021: STEP5 Calculating leaching and volatization from manure applied to pastures"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---

## Summary
Creates rasters for 1) the proportion manure that leaches after manure that has been through mms is then applied to pastures  2) the proportion manure that volatizes after manure that has been through mms is then applied to pastures

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(countrycode)
library(stringr)
library(vroom)
library(parallel)
library(doParallel)

nutrient_d2021_anx <- file.path("/home/shares/ohi/git-annex/globalprep/_raw_data/FAOSTAT/manure_nutrient/d2021/")
git_manure <- file.path(here(), "globalprep/prs_land-based_nutrient/v2021/STEP2_manure")
location_tifs_path <- file.path("/home/shares/ohi/git-annex/globalprep/_raw_data/halpern_food_systems/d2021/animal_farm/location_tifs")
prep <- file.path("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2021")
git_crop <- file.path(here(), "globalprep/prs_land-based_nutrient/v2021/STEP1_crop")

source(here(git_crop, "R/common_spatial.R")) # read in necessary spatial files 

un <- read_csv(here("globalprep/spatial/v2021/UNSD_Methodology.csv")) %>% 
  dplyr::select(iso3c, Global_Name, Region_Name, Sub_region_Name, Intermediate_Region_Name)

```


Volatization
```{r}

bouw_2002_tb_5 <- vroom::vroom(here::here(git_crop, "raw/bouwman_2002_tbl5.csv"))

volatize <- left_join(food_rgns_xy, bouw_2002_tb_5) %>% 
  dplyr::select(x,y, iso3c, volt = prop_nh3_n_grassland) %>%
  arrange(x,y) %>%
  mutate(cell_id = 1:9331200) # create unique cell_id to join by 


test <- volatize %>%
  dplyr::select(x,y,volt) %>%
  rasterFromXYZ(crs=food_crs)
plot(test) # so far so good


```
Denitrification
```{r}

denitr <- food_rgns_tif
denitr[denitr > 0] <- 9.125*0.008 # so far so good

denitr <- raster_df(denitr) %>% 
  right_join(food_rgns_xy) %>% 
  rename(denitri = food_rgns) %>%
    arrange(x,y) %>%
  mutate(cell_id = 1:9331200) # create unique cell_id to join by 


unique(denitr$denitri) #    NA 0.073
colnames(denitr)
test <- denitr %>%
  dplyr::select(x,y,denitri) %>%
  rasterFromXYZ(crs=food_crs)
plot(test) # so far so good
test


```

## Leaching
Calculate how much N is left to leach
```{r}

files <- list.files(file.path(prep, "animal_farm_int/STEP4_N_prop_manure_applied/applied_to_pastures/"), full.names = TRUE)

for(file in files){

  ############  
  ## Get the mapped head counts for each system.
  ## (only really used in step 1, to check data)
  
     # file <- files[11]
  
    ## grab the names of all the parts from the file
    animal_name <- str_split(str_split(file, pattern = "//")[[1]][2], pattern = "_")[[1]][1]
    system_name <- str_split(str_split(file, pattern = "//")[[1]][2], pattern = "_")[[1]][2]
    product_name <- str_split(str_split(file, pattern = "//")[[1]][2], pattern = "_")[[1]][3]


    
    
    N_applied_pasture <- read_csv(file, col_types= c("ddcn")) %>% 
      arrange(x,y) %>%
      mutate(cell_id = 1:9331200) %>% # create unique cell_id to join by 
      dplyr::left_join(volatize, by = "cell_id") %>% 
      dplyr::left_join(denitr, by = "cell_id") %>%
      dplyr::select(x = x.x, y = y.x, iso3c = iso3c.x, cell_id, volt, denitri, prop_N_applied_pastures) %>%
      mutate(withdraw = (1-volt)*0.6,
             leaching = (1 - (withdraw+volt+denitri)),
             leaching = ifelse(leaching <0, 0, leaching),
             prop_leach = leaching*prop_N_applied_pastures)
    
    # take a look to make sure nothing screwed up
  
    # test_volt <- N_applied_pasture %>%
    #   dplyr::select(x,y,volt) %>%
    #   rasterFromXYZ(crs=food_crs)
    # plot(test_volt)
    # 
    # test_deni <- N_applied_pasture %>%
    #   dplyr::select(x,y,denitri) %>%
    #   rasterFromXYZ(crs=food_crs)
    # plot(test_deni)
    # 
    # test_N <- N_applied_pasture %>%
    #   dplyr::select(x,y,withdraw) %>%
    #   rasterFromXYZ(crs=food_crs)
    # plot(test_N)
    # 


    prop_leached_gf <- left_join(N_applied_pasture, un, by = "iso3c") %>% 
      group_by(Intermediate_Region_Name) %>% 
      mutate(prop_leach = ifelse(is.na(prop_leach), mean(prop_leach, na.rm = TRUE), prop_leach)) %>% 
      ungroup()%>% 
      group_by(Sub_region_Name) %>% 
      mutate(prop_leach = ifelse(is.na(prop_leach), mean(prop_leach, na.rm = TRUE), prop_leach)) %>% 
      ungroup() %>% 
      group_by(Region_Name) %>% 
      mutate(prop_leach = ifelse(is.na(prop_leach), mean(prop_leach, na.rm = TRUE), prop_leach)) %>% 
      ungroup() %>% 
      group_by(Global_Name) %>% 
      mutate(prop_leach = ifelse(is.na(prop_leach), mean(prop_leach, na.rm = TRUE), prop_leach)) %>% 
      ungroup() %>% 
      dplyr::select(x,y, iso3c, N_prop_leach_applied_pastures = prop_leach)

# test <- prop_leached_gf %>%
#   dplyr::select(x,y,N_prop_leach_applied_pastures) %>%
#   rasterFromXYZ(crs=food_crs)
# plot(test)
    
    write_csv(prop_leached_gf,
                file.path(paste0(prep, "/animal_farm_int/STEP5.5_N_leached_prop_applied_pastures/",
                                 animal_name, "_",
                                 system_name,"_",
                                 product_name,
                                 "_n_leached_prop_applied_pastures.csv")))

    
}

```

Check to see if anything weird happened
```{r}
df_list <- list.files(file.path(prep,"animal_farm_int/STEP5.5_N_leached_prop_applied_pastures/"), full.names = TRUE)

df_check <- data.frame(iso3c = NA, animal = NA, system = NA, product = NA, N_prop_leach_applied = NA)
  
for(df in df_list){
  
#  df <- df_list[23]
    ## grab the names of all the parts from the file
    animal_name <- str_split(str_split(df, pattern = "//")[[1]][2], pattern = "_")[[1]][1]
    system_name <- str_split(str_split(df, pattern = "//")[[1]][2], pattern = "_")[[1]][2]
    product_name <- str_split(str_split(df, pattern = "//")[[1]][2], pattern = "_")[[1]][3]
     
  ## values are the same throughout the country so a median or mean function won't make a difference
    country_sum <- read_csv(df) %>% 
      group_by(iso3c) %>%
      mutate(animal = animal_name,
             system = system_name,
             product = product_name) %>% 
       dplyr::select(iso3c, N_prop_leach_applied = N_prop_leach_applied_pastures , animal, system, product) %>%
      distinct()
  
    df_check <- rbind(df_check, country_sum) %>% 
      filter(!is.na(iso3c)) 
    
}

## look at it

ggplot(df_check, aes(x = animal, y = N_prop_leach_applied, color = system) ) +
  geom_point(position = "jitter") +
  labs(title = "Proportion of all manure leached from on applied to pastures")



```


Let's look at just the USA cows-mixed-milk, cows-grassland-milk, chicken-industrial

```{r}
df_list <- list.files(file.path(prep, "animal_farm_int/STEP5.5_N_leached_prop_applied_pastures/"), full.names = TRUE)
df_list <- str_subset(pattern = "cows", df_list)
df_file <- df_list[4]

    ## grab the names of all the parts from the file
    category_name <- str_split(df_file, pattern = "//")[[1]][2]
  
  ## values are the same throughout the country so a median or mean function won't make a difference
    country_sum_mixed <- read_csv(df_file, col_types = "ddcn") %>% 
      group_by(iso3c) %>% 
      dplyr::summarise(prop_applied_leaching = mean(N_prop_leach_applied_pastures, na.rm = TRUE)) %>% 
      mutate(category = category_name,
             animal_system_product = word(category, 1,3, sep="_")) %>% 
       dplyr::select(iso3c,prop_applied_leaching, animal_system_product) %>% 
      filter(iso3c == "USA")
    
    df_file <- df_list[2]

    ## grab the names of all the parts from the file
    category_name <- str_split(df_file, pattern = "//")[[1]][2]
  
  ## values are the same throughout the country so a median or mean function won't make a difference
    country_sum_pasture <- read_csv(df_file, col_types = "ddcn") %>% 
      group_by(iso3c) %>% 
      dplyr::summarise(prop_applied_leaching = mean(N_prop_leach_applied_pastures, na.rm = TRUE)) %>% 
      mutate(category = category_name,
             animal_system_product = word(category, 1,3, sep="_")) %>% 
       dplyr::select(iso3c,prop_applied_leaching, animal_system_product) %>% 
      filter(iso3c == "USA")
    
  country_sum <- rbind(country_sum_pasture, country_sum_mixed)  

write_csv(country_sum, here::here(git_manure, "data_check/usa_cows_leached_applied_pastures.csv"))

```


## prop volatized
Calculate how much N is volatized
```{r}
files <- list.files(file.path(prep, "animal_farm_int/STEP4_N_prop_manure_applied/applied_to_pastures/"), full.names = TRUE)

registerDoParallel(5)   
foreach(file = files) %dopar% {

  ############  
  ## Get the mapped head counts for each system.
  ## (only really used in step 1, to check data)
  
     # file <- files[6]
  
    ## grab the names of all the parts from the file
    animal_name <- str_split(str_split(file, pattern = "//")[[1]][2], pattern = "_")[[1]][1]
    system_name <- str_split(str_split(file, pattern = "//")[[1]][2], pattern = "_")[[1]][2]
    product_name <- str_split(str_split(file, pattern = "//")[[1]][2], pattern = "_")[[1]][3]
    
    prop_volt_df <- read_csv(file, col_types= c("ddcn")) %>%
      arrange(x,y) %>%
      mutate(cell_id = 1:9331200) %>% # create unique cell_id to join by 
      left_join(volatize, by = "cell_id") %>% 
      dplyr::select(x = x.x, y = y.x, iso3c = iso3c.x, cell_id, volt, prop_N_applied_pastures) %>%
      mutate(prop_volatized = volt*prop_N_applied_pastures) 
    
    # test <- prop_volt_df %>%
    #   dplyr::select(x,y,prop_volatized) %>%
    #   rasterFromXYZ(crs=food_crs)
    # plot(test)

    prop_volt_gf <- left_join(prop_volt_df, un, by = "iso3c") %>% 
      group_by(Intermediate_Region_Name) %>% 
      mutate(prop_volatized = ifelse(is.na(prop_volatized), mean(prop_volatized, na.rm = TRUE), prop_volatized)) %>% 
      ungroup()%>% 
      group_by(Sub_region_Name) %>% 
      mutate(prop_volatized = ifelse(is.na(prop_volatized), mean(prop_volatized, na.rm = TRUE), prop_volatized)) %>% 
      ungroup() %>% 
      group_by(Region_Name) %>% 
      mutate(prop_volatized = ifelse(is.na(prop_volatized), mean(prop_volatized, na.rm = TRUE), prop_volatized)) %>% 
      ungroup() %>% 
      group_by(Global_Name) %>% 
      mutate(prop_volatized = ifelse(is.na(prop_volatized), mean(prop_volatized, na.rm = TRUE), prop_volatized)) %>% 
      ungroup() %>% 
      mutate(prop_volatized = ifelse(is.na(prop_volatized), 0, prop_volatized)) %>% 
      dplyr::select(x,y, iso3c, N_prop_volt_applied_pastures = prop_volatized)
    
    write_csv(prop_volt_gf,
                file.path(paste0(prep, "/animal_farm_int/STEP5.5_N_volatizing_prop_applied_pasture/",
                                 animal_name, "_",
                                 system_name,"_",
                                 product_name,
                                 "_n_volatized_prop_applied_pastures.csv")))

    
}

```
