---
title: 'OHI 2022 - Exctracting nutrient pressure data '
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../../workflow/templates/ohi_hdr.html' 
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---



# Summary
This prep rescales the new land-based nutrient plume layers to reference points, and extracts the scores per each OHI region for 3nm and entire EEZs. 


# Initial Setup Code
```{r setup, eval = FALSE}
library(raster)
library(rgdal)
library(dplyr)
library(stringr)
library(here)
library(parallel)
library(foreach)
library(doParallel)
library(fasterize)
library(sf)
library(mapview)
library(terra)

source(here('workflow/R/common.R'))
region_data()
regions_shape()
ohi_rasters()
prs_int_dir <- "/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2022/int"

```

# Methods

```{r, eval = FALSE}

save_loc_data <- "globalprep/prs_land-based_nutrient/v2022"


rast_loc <- file.path(dir_M, 
                       "git-annex/globalprep/prs_land-based_nutrient/v2022/output/N_plume")

```



## Mask ocean raster to make consistent with standard OHI global region file

```{r, eval = FALSE}
files <- list.files(rast_loc, pattern = "pourpoints", full=TRUE)

### mask out the land 

files <- list.files(file.path(dir_M, "git-annex/globalprep/prs_land-based_nutrient/v2022/output/N_plume"), 
         pattern = "pourpoints", full=TRUE)



registerDoParallel(5)
foreach(file = files) %dopar% { #file = files[16]
  name <- basename(file)
  name <- sub('\\.tif$', '', name)

  tmp <- rast(file) 
  
  ## need to extend tmp file to ocean extent
  
  tmp_ext <- extend(tmp, ocean)
  
      mask(tmp_ext, ocean, 
         filename = file.path(dir_M, 
                              sprintf("git-annex/globalprep/prs_land-based_nutrient/v2022/output/N_plume_masked/%s", basename(file))),
         overwrite = TRUE)
  
}

# check to see if we lost data
test <- rast(list.files("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2022/output/N_plume_masked", full.names = TRUE))
tmp <- rast(files[1])

global(tmp, "sum") # 16083871
global(test, "sum", na.rm = TRUE) # 16082433
```

## Reclassify 0's 
However, to calculate the quantiles, we don't want any of the 0's that are way out in the high seas, or on land to effect those quantiles, so we need to remove them. 

```{r, eval = FALSE}

## make all 0's NA
files <- list.files(file.path(dir_M, "git-annex/globalprep/prs_land-based_nutrient/v2022/output/N_plume_masked"), 
         pattern = "pourpoints", full=TRUE)


registerDoParallel(4)
foreach(file = files) %dopar% { #file = files[1]
  name <- basename(file)
  name <- sub('\\.tif$', '', name)

  tmp <- rast(file)
  
classify(tmp, cbind(-Inf, 0, NA), right=NA, filename = file.path(dir_M, 
                              sprintf("git-annex/globalprep/prs_land-based_nutrient/v2022/output/N_plume_no_zero/%s", basename(file))), 
         overwrite = TRUE)

  
}

test <- rast(file)
plot(test)
global(test, "sum", na.rm = TRUE) # 16082433

global(rast("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2022/output/N_plume_no_zero/pourpoints_crop_manure_leached_volt_N_2005_joined.tif"), "sum", na.rm = TRUE) # 16082433

plot(rast("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2022/output/N_plume_no_zero/pourpoints_crop_manure_leached_volt_N_2005_joined.tif")) 

```

## Calculate the 99th quantile to use as the reference level. 

```{r, eval = FALSE}
### Collect quantile data for regulars for the 99th quantile 

files_unlog <- list.files(file.path(dir_M, 'git-annex/globalprep/prs_land-based_nutrient/v2022/output/N_plume_no_zero'), full.names = TRUE)

quantiles <- data.frame(plumeData = basename(files_unlog), quantile_99=NA)

for(file in files_unlog) { #file = files_unlog[1]

  tmp <- raster(file)
  quantiles$quantile_99[quantiles$plumeData == basename(file)] <- quantile(tmp, .99)
  
}

write.csv(quantiles, 
          "globalprep/prs_land-based_nutrient/v2022/STEP6_prep_layers/int/quantiles_99.csv",
          row.names = FALSE)

## reference point
## 15.66116
## This was derived using all 99th quantile from all years of data for 2022 OHI assessment (fao data from 2005 to 2020)

quantiles <- read.csv(file.path(here(), "globalprep/prs_land-based_nutrient/v2022/STEP6_prep_layers/int/quantiles_99.csv"))

quant_ref <- mean(quantiles$quantile_99)


ref_point <- read.csv(here("globalprep/supplementary_information/v2022/reference_points_pressures.csv")) %>% 
  mutate(ref_point = ifelse(pressure == "Fertilizer plume data", quant_ref, ref_point)) %>%
  mutate(data_years = ifelse(pressure == "Fertilizer plume data", "2005-2020", data_years),
         method = ifelse(pressure == "Fertilizer plume data", "99 quantile across full time range", method), 
         ref_year = ifelse(pressure == "Fertilizer plume data", "N/A", ref_year))


write.csv(ref_point, here("globalprep/supplementary_information/v2022/reference_points_pressures.csv"))


ref_point <- read.csv(here("globalprep/supplementary_information/v2022/reference_points_pressures.csv")) %>%
  dplyr::filter(pressure == "Fertilizer plume data") %>%
  dplyr::select(ref_point) %>%
  data.frame()
ref_point_fert <- as.numeric(as.character(ref_point$ref_point))

```

## Rescale the rasters to the 99th quantile reference point
```{r, eval = FALSE}
files <- list.files(file.path(dir_M, 'git-annex/globalprep/prs_land-based_nutrient/v2022/output/N_plume_no_zero'), 
                    full.names = TRUE)

registerDoParallel(cores = 2)
foreach(file = files) %dopar% { #file = files[1]

  year <- str_sub(file,-15,-12)
  
  tmp <- rast(file)
  
  app(tmp, fun=function(x){ifelse(x>ref_point_fert, 1, x/ref_point_fert)},
       filename = file.path(dir_M, 
  sprintf("git-annex/globalprep/prs_land-based_nutrient/v2022/output/N_rescaled_99/nutrient_%s_N_rescaled_mol.tif", year)), 
       overwrite=TRUE)
}


#### Data check 
## take a look at one year of data and zoom in on some areas

test <- rast(file.path("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2022/output/N_rescaled_99/nutrient_2005_N_rescaled_mol.tif"))
plot(test)


uk_extent <- c(-800000, 200000, 5800000, 6900000)
uk <- terra::crop(test, uk_extent)



fl_extent <- c(-7900000, -7300000, 3000000, 3800000)
fl <- terra::crop(test, fl_extent)


med_extent <- c(1900000, 3300000, 3700000, 4900000)
med <- terra::crop(test, med_extent)



cn_extent <- c(10000000, 11400000, 2500000, 4400000)
cn <- terra::crop(test, cn_extent)



par(mfrow=c(2,2))
plot(uk)
plot(fl)
plot(med)
plot(cn)

## compare to old data
testing <- rast("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2017/int/global_plumes_fert_2013_log.tif")
plot(testing)



testing <- rast(file.path("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2017/int/nutrient_2005_rescaled_mol.tif"))



uk_extent <- c(-800000, 200000, 5800000, 6900000)
uk <- terra::crop(testing, uk_extent)



fl_extent <- c(-7900000, -7300000, 3000000, 3800000)
fl <- terra::crop(testing, fl_extent)


med_extent <- c(1900000, 3300000, 3700000, 4900000)
med <- terra::crop(testing, med_extent)



cn_extent <- c(10000000, 11400000, 2500000, 4400000)
cn <- terra::crop(testing, cn_extent)


par(mfrow=c(2,2))
plot(uk)
plot(fl)
plot(med)
plot(cn)
```



## extract by OHI zone

```{r, eval = FALSE}

par(mfrow=c(1,1))

files <- list.files(file.path(dir_M, "git-annex/globalprep/prs_land-based_nutrient/v2022/output/N_rescaled_99"), pattern = "rescaled", full=TRUE)

tmp <- raster(files[1])

eez_land_raster <- fasterize::fasterize(regions, tmp, field="rgn_ant_id")
plot(eez_land_raster) ## this is what I want to use to extract


registerDoParallel(cores = 4)
foreach(file = files) %dopar% {
 # file = files[1]
  
  yr <- str_sub(file,-23,-20)
  
  tmp <- raster(file)
  

regions_stats <- zonal(tmp,  eez_land_raster, fun="mean", na.rm=TRUE)

regions_stats2 <- data.frame(regions_stats)

# print(setdiff(regions_stats2$zone, rgns_all$rgn_id)) #should be none.. or those above 255
# print(setdiff(rgns_all$rgn_id, regions_stats2$zone)) #should be 213.. antarctica

data <- merge(rgns_all, regions_stats2, all.y=TRUE, by.x="rgn_id", by.y="zone") %>% 
  mutate(year = yr)

write.csv(data, file.path(here(), save_loc_data, sprintf("STEP6_prep_layers/eez_int/nutrients_plume_data_%s.csv", yr)), row.names=FALSE)

}

myMergedData <- 
  do.call(rbind,
          lapply(list.files(file.path(here(), save_loc_data, "STEP6_prep_layers/eez_int"), pattern = "plume", full = TRUE), read.csv)) %>%
  filter(rgn_type == "eez") %>%
  dplyr::select(rgn_id, year, pressure_score=mean) %>%
  dplyr::arrange(rgn_id) %>%
  mutate(pressure_score = ifelse(is.na(pressure_score), 0, pressure_score))


write.csv(myMergedData, file.path(here(), save_loc_data, "output/cw_fertilizers_score_updated.csv"), row.names=FALSE)



## extract at 3 nm (in addition to a pressure, this will be used for CW and the CW trends

### first we need to make a 3nm raster + the inland raster. Because some of our plumes are a little bit inland, we don't want those to get lost, so we will extract by everything on land and everything 3nm offshore. 
# rgns_3nm_tif <- raster(file.path("/home/shares/ohi/git-annex/globalprep/spatial/v2018/rgns_3nm_offshore_mol.tif"))
# plot(rgns_3nm_tif)
# 
# inland <- regions %>%
#   filter(rgn_type == "land")
# 
# inland_tif <- fasterize(inland, tmp, field = "rgn_id")
# plot(inland_tif)
# 
# combine_inland_3nm <- merge(rgns_3nm_tif, inland_tif, progress = "text")
# plot(combine_inland_3nm)
# 
# writeRaster(combine_inland_3nm, file.path("/home/shares/ohi/git-annex/globalprep/spatial/v2021/inland_3nm_rast.tif"))

combine_inland_3nm <- raster(file.path("/home/shares/ohi/git-annex/globalprep/spatial/v2021/inland_3nm_rast.tif"))

## now do zonal stats to extract the scores 
files <- list.files(file.path(dir_M, "git-annex/globalprep/prs_land-based_nutrient/v2022/output/N_rescaled_99"), pattern = "rescaled", full=TRUE)


registerDoParallel(cores = 4)
foreach(file = files) %dopar% {
  # file = files[1]
  
   yr <- str_sub(file,-23,-20)
   
  tmp <- raster(file)
  

regions_stats_3nm <- zonal(tmp,  combine_inland_3nm, fun="mean", na.rm=TRUE, progress="text")

regions_stats_3nm_2 <- data.frame(regions_stats_3nm) %>% 
  rename("rgn_id" = "zone", "pressure_score" = "mean") %>%
  mutate(year = yr)

# mean(regions_stats_3nm_2$pressure_score)
# cellStats(tmp, "mean")

write.csv(regions_stats_3nm_2, file.path(here(), save_loc_data, sprintf("STEP6_prep_layers/3nm_int/nutrients_plume_data_offshore_3nm_%s.csv", yr)), row.names=FALSE)
}


myMergedData <- 
  do.call(rbind,
          lapply(list.files(file.path(here(), save_loc_data, "STEP6_prep_layers/3nm_int"), pattern = "plume", full = TRUE), read.csv)) %>%
  dplyr::select(rgn_id, year, pressure_score) %>%
  dplyr::arrange(rgn_id) %>%
  mutate(pressure_score = ifelse(is.na(pressure_score), 0, pressure_score))

write.csv(myMergedData, file.path(here(), save_loc_data, "output/cw_fertilizers_score_3nm_updated.csv"), row.names=FALSE)

test <- myMergedData %>%
  left_join(rgns_eez)
```


## TREND calculations

```{r, eval = FALSE}
## trend should be calculated on 3nm (not eez)

trend_data <- read.csv(file.path(here(), save_loc_data, "output/cw_fertilizers_score_3nm_updated.csv"))

for(scenario_year in 2009:2020){ #scenario_year=2009


trend_years <- (scenario_year-4):(scenario_year)
adj_trend_year <- as.numeric(min(trend_years))


  trends <- trend_data %>%
    filter(!is.na(pressure_score)) %>%
    group_by(rgn_id) %>%
    do(mdl = lm(pressure_score ~ year, data=., subset=year %in% trend_years),
       adjust_trend = .$pressure_score[.$year == adj_trend_year]) %>%
    summarize(rgn_id, trend = ifelse(coef(mdl)['year']==0, 0, coef(mdl)['year']/adjust_trend * 5)) %>%
    ungroup() %>%
    mutate(trend = ifelse(trend>1, 1, trend)) %>%
    mutate(trend = ifelse(trend<(-1), (-1), trend)) %>%
    mutate(trend = round(trend, 4)) %>%
    dplyr::select(rgn_id, trend) 
  
  write.csv(trends, file.path(here(), save_loc_data, sprintf('int/cw_fertilizers_trend_%s_new.csv', scenario_year)), row.names=FALSE)
}


data <- data.frame()

for (year in 2009:2020){ # year = 2012
  trend <- read.csv(file.path(here(), sprintf("globalprep/prs_land-based_nutrient/v2022/int/cw_fertilizers_trend_%s_new.csv", year)))
  
  trend <- trend %>%
    mutate(year = year) %>%
    select(rgn_id, year, trend)
  
  data <- rbind(data, trend) %>%
    mutate(trend = ifelse(is.na(trend), 0, trend))
  
}


write.csv(data, file.path(here(), "globalprep/prs_land-based_nutrient/v2022/output/cw_fertilizers_trend_updated.csv"),
          row.names=FALSE)
```

## compare new and old trend values:

```{r, eval = FALSE}
  old <- read.csv(file.path(here(), 'globalprep/prs_land-based_nutrient/v2021/output/cw_fertilizers_trend_updated.csv')) %>%
    dplyr::select(rgn_id, old_trend = trend)
new <- read.csv(file.path(save_loc_data, 'output/cw_fertilizers_trend_updated.csv')) %>%
  left_join(old, by="rgn_id") 
new
plot(new$old_trend, new$trend)
abline(0, 1, col="red")



  old <- read.csv(file.path(here(), 'globalprep/prs_land-based_nutrient/v2021/output/cw_fertilizers_score_3nm_updated.csv')) %>%
    filter(year == 2016) %>%
    dplyr::select(rgn_id, old_pressure = pressure_score)
new <- read.csv(file.path(save_loc_data, 'output/cw_fertilizers_score_3nm_updated.csv')) %>%
  filter(year == 2016) %>%
  left_join(old, by="rgn_id") 
 # mutate(pressure_score = ifelse(is.na(pressure_score), 0, pressure_score))
new
plot(new$old_pressure, new$pressure_score)
abline(0, 1, col="red")
    
```


    