---
title: 'OHI 2021: Seagrass trend'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---


```{r setup, include=FALSE, eval = FALSE}

library(readr)      # for read_csv()
library(raster)
library(here)
library(sf)
library(fasterize)
library(tidyverse)
library(broom)
library(mapview)
library(ggplot2)

source('http://ohi-science.org/ohiprep_v2021/workflow/R/common.R')

goal     <- 'globalprep/hab_seagrass/v2021'
dir_git  <- file.path('~/github/ohiprep_v2021', goal)
dir_waycott <- file.path(dir_git, "waycott_data_extract")
dir_iucn <- file.path("/home/shares/ohi/git-annex/globalprep/_raw_data/iucn_spp/d2020-3")
ohi_rasters()

```

# Summary


We use data from Short et al. 2011, Waycott et al. 2009, and the IUCN 2020 to determine the trends of seagrass for our OHI regions. First, we determine the population trends for seagrasses (increasing, stable, decreasing), based on Short and IUCN data. Then we replace the population trends with numerical values from Waycott et al. 2009. Following we gapfill missing trends first by species family average, and then by the global average if there is no species family average. Then, we calculate the trend per OHI region using a species distribution weighted mean. The final round of gapfilling we gapfill any region that has extent data but no trend data by the georegional average. 

## Updates from previous assessment



***
## Data Sources

**Reference**: Waycott, M., Duarte, C.M., Carruthers, T.J.B., Orth, R.J., Dennison, W.C., Olyarnik, S., Calladine, A., Fourqurean, J.W., Heck, K.L., Hughes, A.R., Kendrick, G.A., Kenworthy, W.J., Short, F.T., Williams, S.L., 2009. Accelerating loss of seagrasses across the globe threatens coastal ecosystems. PNAS 106, 12377–12381. https://doi.org/10.1073/pnas.0905620106

**Description**: Contains data regarding the trends of seagrasses since 1879. 

**Time range**: 1879 - 2007



**Reference**: Short, F.T., Polidoro, B., Livingstone, S.R., Carpenter, K.E., Bandeira, S., Bujang, J.S., Calumpong, H.P., Carruthers, T.J.B., Coles, R.G., Dennison, W.C., Erftemeijer, P.L.A., Fortes, M.D., Freeman, A.S., Jagtap, T.G., Kamal, A.H.M., Kendrick, G.A., Judson Kenworthy, W., La Nafie, Y.A., Nasution, I.M., Orth, R.J., Prathep, A., Sanciangco, J.C., Tussenbroek, B. van, Vergara, S.G., Waycott, M., Zieman, J.C., 2011. Extinction risk assessment of the world’s seagrass species. Biological Conservation 144, 1961–1971. https://doi.org/10.1016/j.biocon.2011.04.010

**Description**: Contains general population trends of seagrass species.

**Time range**: 1975 - 2010



**Reference**: IUCN 2020. The IUCN Red List of Threatened Species. Version 2020-3. <http://www.iucnredlist.org>. 

Shapefiles available from: https://www.iucnredlist.org/resources/spatial-data-download

**Downloaded**: Feb 1, 2021

**Description**: Shapefiles containing the species distribution of seagrass.




# Methods

We will use data from Short et al. 2011, Waycott et al. 2009, and the IUCN 2020 to determine the trends of seagrass for our OHI regions. First, we determine the population trends for seagrasses (increasing, stable, decreasing), based on Short and IUCN data. Then we replace the population trends with numerical values from Waycott et al. 2009. Following we gapfill missing trends first by species family average, and then by the global average if there is no species family average. Then, we calculate the trend per OHI region using a species distribution weighted mean. The final round of gapfilling we gapfill any region that has extent data but no trend data by the georegional average. 


## Use IUCN species maps, Short et al. 2011, and Waycott et al. 2009

This is basically recreating the analysis done in Short et al. 2011: https://www.researchgate.net/publication/235433892_Extinction_Risk_Assessment_of_the_World's_Seagrass_Species

 1. Use the IUCN maps to map the location of all species of seagrass (similar to the Short paper). We gapfill the iucn population trend data with short data and vice versa when it is missing in the IUCN data. 

 2. Then: we replace stable species with 0, increasing with 0.0845, and decreasing with -0.0767. These numbers come from Table S2 in Waycott et al. 2009: https://www.pnas.org/content/pnas/suppl/2009/07/08/0905620106.DCSupplemental/ST2_PDF.pdf
 
 3. Then: gapfill the missing trends by species family average, and then by the global average if there is no species family average
 
 4. Then: rasterize and use zonal statistics to calculate the trend and amount of species distribution area in each region
 
 5. Then: calculate the trend using an species distribution area weighted mean. Create a gapfilling flag based on the proportion of species trends that are gapfilled within a region. 
 
 6. Then: match to our extent data, and gapfill any region that has extent data but not trend data by georegion. 
 

## Steps 1 - 3
```{r, eval = FALSE}
seagrasses <- st_read(file.path(dir_iucn, "SEAGRASSES.shp"))

# mapview(head(seagrasses$geometry)) # take a look - this takes a couple of minutes to process

seagrass_ids <- unique(seagrasses$id_no)

## read in the trend data from spp subgoal
iucn_trend <- read_csv(file.path("~/github/ohiprep_v2021/globalprep/spp/v2021/_data/iucn_trend_by_spp_2020-3.csv")) %>%
  dplyr::filter(iucn_sid %in% seagrass_ids) ## grab the population trends from our spp dataprep


## read in data from Short et al. 

short_trend <- read_csv(file.path("/home/shares/ohi/git-annex/globalprep/hab_seagrass/v2021/trends/short_pop_trends.csv")) %>% 
  dplyr::select("binomial" = `Species name`, "population_trend" = `Pop. trend`) %>%
  dplyr::mutate(population_trend = case_when(population_trend == "Unknown" ~ "unknown", 
                                             population_trend == "Stable" ~ "stable", 
                                             population_trend == "Increasing" ~ "increasing",
                                             population_trend == "Decreasing" ~ "decreasing")) ## fix the names 

seagrasses_trends <- seagrasses %>%
  left_join(iucn_trend, by = c("id_no" = "iucn_sid")) %>% ## join with iucn data
  left_join(short_trend) %>% # join with short data 
  mutate(pop_trend_final = case_when(is.na(pop_trend) & is.na(population_trend) ~ "Unknown",
                                     !is.na(pop_trend) & is.na(population_trend) ~ pop_trend,
                                     is.na(pop_trend) & !is.na(population_trend) ~ population_trend,
                                     TRUE ~ pop_trend
                                     )) %>% # gapfill the population trends that are missing from either short OR iucn, using Short as the baseline
    mutate(waycott_trend = case_when(
    pop_trend_final == "decreasing" ~ -0.0767,
    pop_trend_final == "increasing" ~ 0.0845, 
    pop_trend_final == "stable" ~ 0
  )) ## now assign the global waycott trends 

summary(seagrasses_trends)

unique(seagrasses_trends$pop_trend)

colnames(seagrasses_trends)

## calculate an overall mean to use for gapfilling if necessary
overall_mean <- mean(seagrasses_trends$waycott_trend, na.rm = TRUE)

seagrasses_trends_gf <- seagrasses_trends %>%
  dplyr::group_by(family) %>%
  dplyr::mutate(family_trend_gf = mean(waycott_trend, na.rm = TRUE)) %>% ## calculate family level mean
  dplyr::ungroup() %>%
  dplyr::mutate(waycott_trend_gf = ifelse(is.na(waycott_trend) & !is.na(family_trend_gf), family_trend_gf, waycott_trend),
                gapfill = ifelse(is.na(waycott_trend) & !is.na(family_trend_gf), "family", NA)) %>% ## gapfill trend based on family mean
  mutate(gapfill = ifelse(is.na(waycott_trend_gf), "global", gapfill), 
         waycott_trend_gf = ifelse(is.na(waycott_trend_gf), overall_mean, waycott_trend_gf)) %>% # gapfill trend based on global mean
  mutate(gapfill = ifelse(is.na(gapfill), "none", gapfill))
  
  
test999 <- st_drop_geometry(seagrasses_trends_gf) ## it worked 
 
```


## Step 4: 
Rasterize and use zonal statistics to calculate the trend and amount of species distribution area in each region.

```{r, eval = FALSE}
## use a eez and 25mile inland raster; if no 25km inland use just EEZ
ocean_25_mile <- raster(file.path("/home/shares/ohi/git-annex/globalprep/spatial/v2019/ocean_plus25mile_inland.tif"))

## fasterize per seagrass species 
sg_moll <- st_transform(seagrasses_trends_gf, crs = "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")

for(i in 1:80){ ## there are 80 observations 
 # i = 1

sg_moll_i <- sg_moll[i, ]  
  
id <- sg_moll_i$id_no

pop_trend <- sg_moll_i$pop_trend_final

gapfill <- sg_moll_i$gapfill

fasterize_file = file.path(dir_M, "git-annex/globalprep/hab_seagrass/v2021/trends/int", sprintf("sg_trend_id_%s_%s_%s_%s.tif", i, id, pop_trend, gapfill))
  
if(!file.exists(fasterize_file)) {

trend_raster <- fasterize(sg_moll_i, ocean_25_mile, field = "waycott_trend_gf")

writeRaster(trend_raster, file.path("/home/shares/ohi/git-annex/globalprep/hab_seagrass/v2021/trends/int", sprintf("sg_trend_id_%s_%s_%s_%s.tif", i, id, pop_trend, gapfill)))


}else{
  print("file exists, skipping")
}
}

## stack all values and do zonal stats summed per rgn 

sg_trend_raster_files <- list.files(file.path("/home/shares/ohi/git-annex/globalprep/hab_seagrass/v2021/trends/int"), 
                                    pattern = 'sg_trend_id_', 
                                    full.names = TRUE)

sg_trend_raster_stack <- raster::stack(sg_trend_raster_files)

zonal_means_combined_sg <- raster::zonal(sg_trend_raster_stack, 
                                     ocean_25_mile,
                                     fun = "mean",
                                     na.rm = TRUE,
                                     progress="text") #mean of all seagrass trend cells for each ohi zone
zonal_means_combined_sg_df <- data.frame(zonal_means_combined_sg)

zonal_means_df <- zonal_means_combined_sg_df %>%
  pivot_longer(cols = starts_with("sg_trend"),
               names_to = "observation",
               names_prefix = "sg_trend_id_", 
               values_to = "trend_score", 
               values_drop_na = TRUE) %>%
  separate(observation, c("row", "species_id_no", "pop_trend", "gapfill")) 

write.csv(zonal_means_df, "int/seagrass_trends_no_distribution_area.csv", row.names = FALSE)


### Now we need to extract the species distribution area per each region
for(i in 1:80){ ## there are 80 observations 
 # i = 1

sg_moll_i <- sg_moll[i, ]  
  
id <- sg_moll_i$id_no

pop_trend <- sg_moll_i$pop_trend_final

gapfill <- sg_moll_i$gapfill

fasterize_file = file.path(dir_M, "git-annex/globalprep/hab_seagrass/v2021/trends/int/species_areas", sprintf("sg_area_id_%s_%s_%s_%s.tif", i, id, pop_trend, gapfill))
  
if(!file.exists(fasterize_file)) {

area_raster <- fasterize(sg_moll_i, ocean_25_mile, field = NULL)

writeRaster(area_raster, file.path("/home/shares/ohi/git-annex/globalprep/hab_seagrass/v2021/trends/int/species_areas", sprintf("sg_area_id_%s_%s_%s_%s.tif", i, id, pop_trend, gapfill)))


}else{
  print("file exists, skipping")
}
}

######### read in just one of the files and check to see if it works correctly... ##########
# raster_1 <- raster(file.path("/home/shares/ohi/git-annex/globalprep/hab_seagrass/v2021/trends/int/species_areas/sg_area_id_1_153534_decreasing_none.tif"))

# zonal_sum_area_combined_sg <- raster::zonal(raster_1, 
#                                      ocean_25_mile,
#                                      fun = "sum",
#                                      na.rm = TRUE,
#                                      progress="text") # sum of all seagrass area cells for each ohi zone
# zonal_sum_area_combined_sg_df <- data.frame(zonal_sum_area_combined_sg)


# zonal_sums_km2 <- zonal_sum_area_combined_sg_df %>%
#   mutate(habitat = "seagrass",
#          km2 = (0.934478877011218970**2*sum)) %>% #one cell is equal to ~0.93 km
#   rename("rgn_id" = "zone") %>%
#   select(-sum)
# 
# sum(zonal_sums_km2$km2)
# [1] 246956.2
# st_area(sg_moll_i)*0.000001 # 246964.7 perfect....



## stack all area values and do zonal stats summed per rgn 

sg_area_raster_files <- list.files(file.path("/home/shares/ohi/git-annex/globalprep/hab_seagrass/v2021/trends/int/species_areas"), 
                                    pattern = 'sg_area_id_', 
                                    full.names = TRUE)

sg_area_raster_stack <- raster::stack(sg_area_raster_files)

zonal_sum_area_combined_sg <- raster::zonal(sg_area_raster_stack, 
                                     ocean_25_mile,
                                     fun = "sum",
                                     na.rm = TRUE,
                                     progress="text") # sum of all seagrass area cells for each ohi zone
zonal_sum_area_combined_sg_df <- data.frame(zonal_sum_area_combined_sg)


zonal_sum_df <- zonal_sum_area_combined_sg_df %>%
  pivot_longer(cols = starts_with("sg_area"),
               names_to = "observation",
               names_prefix = "sg_area_id_", 
               values_to = "sum", 
               values_drop_na = TRUE) %>%
  separate(observation, c("row", "species_id_no", "pop_trend", "gapfill")) %>%
  mutate(habitat = "seagrass",
          km2 = (0.934478877011218970**2*sum))

write.csv(zonal_sum_df, "int/seagrass_distributions_area.csv", row.names = FALSE)



## combine the trends and area together 

zonal_means_df <- read_csv("int/seagrass_trends_no_distribution_area.csv")

zonal_sum_df <- read_csv("int/seagrass_distributions_area.csv")

zonal_combined_area_trends <- zonal_means_df %>%
  left_join(zonal_sum_df) %>%
  dplyr::select(-row, -sum)

write.csv(zonal_combined_area_trends, "int/seagrass_trends_distributions_area_final.csv", row.names = FALSE)

## check to see if the numbers are correct.. filter for one polygon (species) and see if the sums match
# test <- zonal_combined_area_trends %>%
#   filter(row == 2)
# 
# sum(test$km2) # 2269069
# 
# test2 <- sg_moll[2, ]
# 
# sum(st_area(test2))*0.000001 # 2269069 - matches really well... i think it worked. 


```

## Step 5
Calculate the trend using an area weighted mean per each region. Create a gapfilling flag based on the proportion of species trends that are gapfilled within a region. 

```{r, eval = FALSE}
## now lets calculate the trend based on a area weighted average

trends_area_weighted <- zonal_combined_area_trends %>%
  group_by(zone, habitat) %>%
  dplyr::summarise(trend = weighted.mean(trend_score, km2)) %>%
  mutate(year = 2021) %>%
  rename("rgn_id" = "zone") %>%
  filter(rgn_id <= 250)

write.csv(trends_area_weighted, "int/habitat_trend_seagrass_updated_all_regions.csv", row.names = FALSE)


## create a proportion species of gapfilled file

trends_gf <- zonal_combined_area_trends %>%
  group_by(zone, habitat, gapfill) %>%
  summarise(n_obs = n()) %>%
  mutate(total_obs = sum(n_obs)) %>%
  ungroup() %>%
  filter(gapfill != "none") %>%
  group_by(zone, habitat, total_obs) %>%
  summarise(n_obs = sum(n_obs)) %>%
  ungroup() %>%
  mutate(prop_gapfilled = n_obs/total_obs)

trends_no_gf <- zonal_combined_area_trends %>%
  group_by(zone, habitat, gapfill) %>%
  summarise(n_obs = n()) %>%
  mutate(total_obs = sum(n_obs)) %>%
  ungroup() %>%
  filter(gapfill == "none") %>%
  mutate(prop_gapfilled = n_obs/total_obs) %>%
  filter(prop_gapfilled == 1) %>%
  mutate(prop_gapfilled = prop_gapfilled - 1) %>%
  dplyr::select(-gapfill)

trends_gf_final <- rbind(trends_gf, trends_no_gf) %>%
  dplyr::select("rgn_id" = "zone", habitat, "prop_gapfill" = "prop_gapfilled") %>%
  mutate(year = 2021) %>%
  dplyr::filter(rgn_id <= 250) %>%
  mutate(gapfill_type = ifelse(prop_gapfill != 0, "family average", "none")) %>%
  mutate(gapfill_type = ifelse(rgn_id == 16, "family and global average", gapfill_type))

write.csv(trends_gf_final, "int/trend_seagrass_gf_all_rgns.csv", row.names = FALSE)

```

## Step 6
Match to our extent data, and gapfill any region that has extent data but not trend data by georegion. Any of the regions gapfilled here will be have a prop_gapfill = 1.

```{r, eval = FALSE}

## now match to the regions which have extent data and gapfill

regions <- read.csv(file.path(here('globalprep/supplementary_information/v2018/rgn_georegions_wide_2013b.csv'))) ## read in georegions data


seagrass_extent <- read_csv("data/habitat_extent_seagrass_updated.csv") %>%
  dplyr::select(-year)

trends_area_weighted <- read_csv("int/habitat_trend_seagrass_updated_all_regions.csv")

trends_gf_final <- read_csv("int/trend_seagrass_gf_all_rgns.csv")

setdiff(trends_area_weighted$rgn_id, seagrass_extent$rgn_id) # rgns with trend data and not extent
setdiff(seagrass_extent$rgn_id, trends_area_weighted$rgn_id) # rgns with extnet data and not trend.. only 4, not so bad


trends_extent_rgns <- seagrass_extent %>%
  left_join(trends_area_weighted, by = c("rgn_id", "habitat")) %>% 
  left_join(regions) %>%
  group_by(r2) %>%
  mutate(avg_trend_r2 = mean(trend, na.rm=TRUE)) %>%
  ungroup() %>%
  group_by(r1) %>%
  mutate(avg_trend_r1 = mean(trend, na.rm=TRUE)) %>%
  ungroup() %>%
  mutate(trend2 = ifelse(is.na(trend), avg_trend_r2, trend)) %>%
  mutate(trend3 = ifelse(is.na(trend2), avg_trend_r1, trend2)) %>%
  mutate(habitat = "seagrass") %>%
  mutate(gap_fill = ifelse(is.na(trend), "r2_gap_fill", NA)) %>%
  mutate(gap_fill = ifelse(is.na(trend) & is.na(avg_trend_r2), "r1_gap_fill", gap_fill)) %>%
  mutate(gap_fill = ifelse(is.na(gap_fill), 0, gap_fill)) %>%
  dplyr::select(rgn_id, habitat, gap_fill, trend=trend3) %>%
  mutate(trend = trend*5) # multiply by 5 to get what the decrease/increase will be in 5 years 


trends_extent_rgns_gf <- trends_extent_rgns %>%
  left_join(trends_gf_final) %>%
  mutate(prop_gapfill = ifelse(is.na(prop_gapfill), 1, prop_gapfill)) %>%
  mutate(gapfill_type = ifelse(is.na(gapfill_type), gap_fill, gapfill_type)) %>%
  mutate(year = 2021) %>%
  dplyr::select(rgn_id, habitat, year, prop_gapfill, "gap_fill" = "gapfill_type")

write.csv(trends_extent_rgns_gf, "data/trend_seagrass_gf.csv", row.names = FALSE)

trends_rgns_final <- trends_extent_rgns %>%
  mutate(year = 2021) %>%
  dplyr::select(rgn_id, habitat, year, trend)

write.csv(trends_rgns_final, "data/habitat_trend_seagrass_updated.csv", row.names = FALSE)


hist(trends_rgns_final$trend)

ggplot(trends_rgns_final, aes(trend)) +
  geom_histogram()
```

```{r, eval = FALSE}

seagrass_trend <- read_csv("data/habitat_trend_seagrass_updated.csv")

old_seagrass_trend <- read_csv(file.path(here(), "globalprep/hab_seagrass/v2012/data/habitat_trend_seagrass_updated.csv"))

compare_habitat_trend <- seagrass_trend %>%
  left_join(old_seagrass_trend, by = "rgn_id") %>%
  # mutate(km2.y = ifelse(
  #   km2.x >0 & is.na(km2.y) ,0, #assign 0 values to old data km2 that have new data so that we can properly graph these differences.
  #   km2.y
  # )) %>%
  mutate(difference = trend.x - trend.y) %>%
  left_join(rgns_eez)
```



