---
title: 'OHI 2021 - Clean Water - Pathogen Pollution: Preparing Sanitation data'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---

# Summary

This document describes the steps for preparing the pathogen pollution and pathogen pollution trend data layers for the 2023 global assessment.

The percentage of the population with access to improved sanitation facilities (World Health Organization and United Nations Children's Fund, Joint Monitoring Programme, 2011) was used in combination with measurements of coastal population as a proxy for pathogens in coastal waters. Access to improved sanitation facilities is defined as the percentage of the population in a country with at least adequate access to disposal facilities that can effectively prevent human, animal, and insect contact with excreta. These data are a country-wide average (not specific to the coastal region).

# Updates from previous assessment
1. Updated data download through the SDMX REST API to programmatically access data, instead of downloading an Excel file. We are using the wash_households data flow. This resulted in small changes to the data: for example in the previous data some values for the at least basic sanitation proportion were denoted as >99, which we previously set as 99.5. The total at least basic indicator for these countries (and some others) were often not available from the API, but we were able to calculate the at least basic indicator by summing the limited, unimproved, and open defecation proportions for a given country and subtracting them from 100. The values were previously rounded to the nearest whole percentage, use of the API resulted in more precise values.The previous data set included a population column, which was not available from the API. We instead accessed the total population for each region from the WDI package (World Bank World Development Indicators). Other changes involved updating data wrangling to get the data from the API into the same format as the previous data. 

**Consider for future assessments:** make the "Safely managed" data more complete, and reconsider whether these data would be better to use.

Definition of each variable according to the data source: "At least basic": Use of improved facilities that are not shared with other households.

"Safely managed": Use of improved facilities that are not shared with other households and where excreta are safely disposed of in situ or transported and treated offsite.

------------------------------------------------------------------------

# Data Source

**Reference**: 
WHO-UNICEF. 2023. “Joint Monitoring Programme (JMP) for Water Supply and Sanitation - Household Data.” Joint Monitoring Programme (JMP) for Water Supply and Sanitation - Household Data. 2023. https://washdata.org/data/household#!/.

**Data Download** As of v2023 the data is now downloaded through the SDMX REST API which allows programmatic access to data that is available in UNICEF's Data Warehouse Explorer. The API can be found at: <https://sdmx.data.unicef.org/webservice/data.html>

For information on querying the api visit: <https://data.unicef.org/sdmx-api-documentation/#understanding>. To download data manually in Excel format use this site: <https://washdata.org/data>.

**Download Date**: Downloaded 7/18/2023

**Description**: Percentage of the national population that has access to at improved sanitation facilities that are not shared with other households.

Access to improved sanitation facilities is defined as the percentage of the population within a country with at least basic access to excreta disposal facilities that can effectively prevent human, animal, and insect contact with excreta.

Further information: <https://washdata.org/monitoring/sanitation> 

**Native data resolution**: country-wide average (not specific to the coastal region)

**Time range**: 2000 - 2022

**Format**: SDMX

------------------------------------------------------------------------



# Setup

```{r setup, results="hide", message = FALSE, warning = FALSE}
library(ohicore)  #devtools::install_github('ohi-science/ohicore@dev')
library(zoo)     # for na.locf: Last Observation Carried Forward
library(tidyverse)
library(here)
library(plotly)
library(rsdmx)
library(countrycode)
library(WDI) #reinstall to access new years of population data

source('https://raw.githubusercontent.com/OHI-Science/ohiprep_v2021/gh-pages/workflow/R/common.R')

#update these each year
version_year <- paste0("v", 2023) 
previous_version_year <- paste0("v", 2021)
data_start <- 2000
data_end <- 2022
```

# Methods

Percentage of people without sanitation was multiplied by the coastal population (within 25 miles of the coast) to estimate the number of people located along the coast without sanitation. This value was rescaled between 0 and 1 by dividing by the 99th quantile across all regions from years 2000 to 2009. 

## Download the data

```{r read data, results="hide", message = FALSE, warning = FALSE, eval=FALSE}
#data download url
url <- "https://sdmx.data.unicef.org/ws/public/sdmxapi/rest/data/UNICEF,WASH_HOUSEHOLDS,1.0/.WS_PPL_S-ALB+WS_PPL_S-L+WS_PPL_S-OD+WS_PPL_S-UI.SAN.._T?format=sdmx-compact-2.1"

#read in data from the api and save as a tibble
sanitation <- readSDMX(url) %>% as_tibble()

#save to the raw data folder for reproduceability in case the data changes
write_csv(sanitation,  here("globalprep/prs_cw_pathogen", version_year, "raw/2023_WASH.csv"))

```

## Data wrangling

Some countries are missing a value for the at least basic category indicator (ws_ppl_s\_alb). This can be calculated by adding the proportion of the population using limited sanitation services, the proportion of population using unimproved sanitation services, and the proportion of the population using open defecation, and subtractring that sum from 100 to get the value if it doesn't exist.

Scale the percentage of population with access to improved sanitation to proportion (from 0-1). Transform population and percentage into a numeric variable.

```{r data_wrangling, results="hide", message = FALSE, warning = FALSE, eval=FALSE}

#pivot so that each indicator has its own column
sanitation_wider <- sanitation %>% 
  pivot_wider(names_from = INDICATOR, values_from = OBS_VALUE) %>% 
  janitor::clean_names() %>% 
  select(country_code = ref_area, year = time_period, basic = ws_ppl_s_alb, ws_ppl_s_l, ws_ppl_s_ui, ws_ppl_s_od)

sanitation_clean <- sanitation_wider %>%
  group_by(country_code,year) %>% 
  mutate(not_basic = sum(as.numeric(ws_ppl_s_l), as.numeric(ws_ppl_s_ui), as.numeric(ws_ppl_s_od), na.rm = TRUE),
         basic = as.numeric(basic), year = as.numeric(year)) %>% 
  mutate(alb = 100 - not_basic) %>% #calculate at least basic from the other 3 indicators
    mutate(basic = ifelse(is.na(basic), alb, basic)) %>% #replace na values with the calculated value
  ungroup() %>% 
  dplyr::mutate(basic_prop = basic/100) %>% #calculate proportion
  dplyr::filter(!is.na(year)) %>% 
  tidyr::complete(country_code, year) #make implicit NA values explicit


#add in the name for each country
#receive some warnings because some of the codes included in the data are not actually country codes
sanitation_clean$country <- countrycode(sanitation_clean$country_code, "iso3c", "country.name")

#add in channel islands because country code not recognized by the function
sanitation_clean <- sanitation_clean %>% 
  mutate(country = case_when(country_code == "CHI" ~ "Channel Islands", TRUE ~ country))
#add the population data

population_weights <- WDI(
  country = "all",
  indicator = "SP.POP.TOTL", 
  start = data_start, end= data_end) %>%
  select(country_code = iso3c, population = SP.POP.TOTL, year)

#add population data
sanitation_clean <- sanitation_clean %>% 
  left_join(population_weights, by= c("country_code", "year")) %>% 
  select(country_code, country, year, basic, basic_prop, population) %>% 
  mutate(population = case_when(str_detect(country, "Martinique") ~368000,
         str_detect(country, "Guadeloupe") ~ 396000,
         is.na(population) ~ 1,
         TRUE ~ population)) #add static value for Guadeloupe and Martinique since these are not reported, found through online search
```

Change names of regions to match names in ohicore and filter out regions that are not part of the OHI assesment or do not have data.

If after running 'name_2\_rgn' (see next r chunk), there are some coastal regions that are not identified by `name_2rgn` function. They must be checked to determine how to best include them (or not include them).

```{r, results="hide", message = FALSE, warning = FALSE, eval=FALSE}

#manually clean country names 
sanitation_clean <- sanitation_clean %>%
   mutate(country = str_replace(country, "&", "and")) %>% 
  mutate(country = case_when(
                           str_detect(country, "Helena") ~"Saint Helena",
                           str_detect(country, "São Tomé and Príncipe")~ "Sao Tome and Principe",
                          TRUE ~ country
                           )) 

## v2021: Reporting Caribbean Netherlands regions (Bonaire, Sint Eustatius, and Saba) at a higher resolution:

CN <- filter(sanitation_clean, country=="Caribbean Netherlands") %>%
   rename(country_old = country)

CN_subregions <- data.frame(country_old = "Caribbean Netherlands",
                            country = c("Bonaire", "Sint Eustatius", "Saba")) %>%
  left_join(CN) %>%
  select(-country_old)

sanitation_clean <- sanitation_clean %>%
   filter(country != "Caribbean Netherlands") %>%
   rbind(CN_subregions)  

# Channel Islands correspond to OHI regions of Guernsey and Jersey. Here the data reported for Channel Islands is used for these two regions.

CI <- dplyr::filter(sanitation_clean, country=="Channel Islands") %>%
  dplyr::rename(country_old = country)
CI_subregions <- data.frame(country_old = "Channel Islands",
                            country = c("Guernsey", "Jersey")) %>%
  dplyr::left_join(CI) %>%
  dplyr::select(-country_old)

sanitation_clean <- sanitation_clean %>%
  dplyr::filter(country != "Channel Islands") %>%
  rbind(CI_subregions) 

```

Add rgn_id and merge duplicate regions using a mean weighted by population.

```{r regions, results="hide", message = FALSE, warning = FALSE, eval=FALSE}
rgn_sani <- name_2_rgn(sanitation_clean, 
                        fld_name     = 'country',
                        flds_unique  = c('year'))

#Check list in warning to make sure all countries removed are not of interest for the OHI global assesment. 
# v2023 non-matched regions: Eswatini (landlocked), Isle of Man (not reported in OHI), North Macedonia (landlocked), Palestinian Territories (disputed), Saint Barthelemy (not reported in OHI)
#Check for duplicate regions.

 rgn_sani %>%
  dplyr::group_by(rgn_id) %>%
  dplyr::summarise(n= dplyr::n()) %>% 
  filter(n > (data_end - data_start +1)) 

#Each region has 23 years of data in this case. 
#If there are regions with more than 23 values it is because in this database they are reported in a higher resolution than the OHI global assessment (eg: China, Hong Kong and Macao are just one region for the OHI global). For these regions a weighted average will be calculated as its final score.
 #v2023 regions 13, 116, 140 and 2009 had duplicates



# Calculating weighted means for the duplicated regions.

sani_final <- rgn_sani %>% 
  dplyr::group_by(rgn_id, rgn_name, year) %>% 
  dplyr::summarise(basic_sani_prop= 
                     ifelse(n() >1, weighted.mean(basic_prop, population, na.rm = TRUE), basic_prop)) %>% 
   mutate(basic_sani_prop = ifelse(is.nan(basic_sani_prop), NA, basic_sani_prop))


#Check for duplicate regions. At this point, all regions should have the same sample size.
year_check <- sani_final %>%
  dplyr::group_by(rgn_id) %>%
  dplyr::summarise(n= dplyr::n())

unique(year_check$n)
```

## Gapfilling

First step is to get an idea of what needs to be gapfilled.

```{r check, message = FALSE, warning = FALSE, eval=FALSE}

sani_gf <- sani_final %>% 
  dplyr::group_by(rgn_id, rgn_name) %>%
  dplyr::mutate(gf_count = sum(is.na(basic_sani_prop))) %>% # create column to count # of NAs (# of data points to gapfill)
  ungroup()

summary(sani_gf) 
# 161 NAs in v2019
# 212 NAs in v2021; makes sense, there are more years of data
# 191 NAS in v2023

#list of regions that need gapfilling
dplyr::filter(sani_gf, gf_count>0) %>% 
  dplyr::select(rgn_id, rgn_name, gf_count) %>% 
  unique() %>%
  data.frame()

#Some regions have no data - we will filter them out to be gapfilled later.

#Define number of years of data
years_data <- mean(table(sani_gf$rgn_id))

sani_gf <- sani_gf %>%
  dplyr::filter(gf_count != years_data)

```

### Gapfilling 1: Linear model

Use a linear model within country data to estimate missing years.

```{r gf_lm, results="hide", message = FALSE, warning = FALSE, eval=FALSE}

sani_gf_lm <- sani_gf %>%
  dplyr::group_by(rgn_id, rgn_name) %>%
  dplyr::do({
    mod <- lm(basic_sani_prop ~ year, data = .)
    gf_lm <- predict(mod, newdata = .[c('year')])
    data.frame(., gf_lm)
  }) %>%
  dplyr::ungroup()


sani_gf_lm <- sani_gf_lm %>%
  mutate(gf_lm = ifelse(gf_lm > 1, 1, gf_lm)) %>% # constrain predictions to <=1 
  mutate(method = ifelse(is.na(basic_sani_prop), "lm prediction based on year", NA)) %>%
  mutate(basic_sani_prop = ifelse(is.na(basic_sani_prop), gf_lm, basic_sani_prop))

```

### Gapfilling 2: Georegional averages

Georegional gapfilling for regions that do not have data.

```{r gf georgn, results="hide", message = FALSE, warning = FALSE, eval=FALSE}

UNgeorgn()
UNgeorgn <- UNgeorgn %>%
  dplyr::select(rgn_id, rgn_label, r1=r1_label, r2=r2_label)


year <- min(sani_gf_lm$year):max(sani_gf_lm$year) #defines the year range

sani_georgn_gf <- UNgeorgn %>%
  expand(year, UNgeorgn) %>%
  dplyr::left_join(sani_gf_lm, by = c('rgn_id', 'year'))


#Calculate two different gapfill columns using r2 and r1 UN geopolitical classification
sani_georgn_gf <- sani_georgn_gf %>%
  dplyr::group_by(year, r2) %>%
  dplyr::mutate(basic_sani_r2 = mean(basic_sani_prop, na.rm=TRUE)) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(year, r1) %>%
  dplyr::mutate(basic_sani_r1 = mean(basic_sani_prop, na.rm=TRUE)) %>%
  dplyr::ungroup()%>%
  dplyr::arrange(rgn_id, year)


#First gapfill with r2, if no value available use r1; create column indicating whether value was gapfilled and if so, by what method. Give NA to inhabited regions
sani_georgn_gf <- sani_georgn_gf %>%
  dplyr::mutate(method = ifelse(is.na(basic_sani_prop) & !is.na(basic_sani_r2), "UN georegion avg. (r2)", method)) %>%
  dplyr::mutate(method = ifelse(is.na(basic_sani_prop) & is.na(basic_sani_r2) & !is.na(basic_sani_r1), "UN georegion avg (r1)", method))%>%
  dplyr::mutate(basic_sani_prop = ifelse(is.na(basic_sani_prop) & !is.na(basic_sani_r2), basic_sani_r2, basic_sani_prop)) %>%
  dplyr::mutate(basic_sani_prop = ifelse(is.na(basic_sani_prop) & !is.na(basic_sani_r1), basic_sani_r1, basic_sani_prop)) %>%
  dplyr::select(rgn_id, rgn_label, year, basic_sani_prop, method)

#See regions that have not been gapfilled. 
dplyr::filter(sani_georgn_gf, is.na(basic_sani_prop)) %>% 
  dplyr::select(rgn_id, basic_sani_prop) %>% 
  unique() %>%
  data.frame() #NA values for uninhabitated regions. 

```

### Gapfilling 3: Replace scores for uninhabited regions

Uninhabited regions get a perfect score.

```{r, results="hide", message = FALSE, warning = FALSE, eval=FALSE}

#Identify uninhabited regions 

low_pop() # requires common.R to be loaded 
low_pop <- low_pop %>%
  dplyr::filter(est_population < 100 | is.na(est_population)) 



#Fill in all inhabited in rgn_inhab with perfect sanitation prop
sani_complete <- sani_georgn_gf %>% 
  dplyr::mutate(basic_sani_prop = ifelse(rgn_id %in% low_pop$rgn_id, 1, basic_sani_prop)) %>% 
  dplyr::mutate(gapfill = ifelse(is.na(method), 0, 1)) %>%
  dplyr::mutate(gapfill = ifelse(rgn_id %in% low_pop$rgn_id, 0, gapfill)) %>%
  dplyr::mutate(method = ifelse(rgn_id %in% low_pop$rgn_id, "No est. human population", method)) %>%
  dplyr::select(rgn_id, rgn_name = rgn_label, year, basic_sani_prop, gapfill, method)

summary(sani_complete) # should be no more NAs 
table(sani_complete$gapfill, sani_complete$method) # no pop is not considered gapfilled, should be N years * 20 regions

```

### Save gapfilled data records

```{r, results="hide", message = FALSE, warning = FALSE, eval=FALSE}

write_csv(sani_complete, here("globalprep/prs_cw_pathogen/",version_year, "/intermediate/sani_complete.csv"))


# Quick check with previous year of data
sani_complete_old <- read_csv(here("globalprep/prs_cw_pathogen",previous_version_year, "intermediate/sani_complete.csv")) %>% 
  rename(basic_sani_prop_2021 = basic_sani_prop) %>%
  rename(gapfill_2021 = gapfill) %>% 
  rename(method_2021 = method) %>% 
  left_join(sani_complete, by=c("rgn_id", "year", "rgn_name")) %>%
  filter(year == 2020)

ggplotly(ggplot(sani_complete_old, aes(y = basic_sani_prop, x = basic_sani_prop_2021, labels = rgn_id)) +
           labs(x= paste(previous_version_year, "basic_sani_prop"),
                y = paste(version_year, "basic_sani_prop"), title = "2020 Data Comparison") +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "red"))

# save gapfilling info
gf_data <- sani_complete %>%
  dplyr::select(rgn_id, year, gapfill, method)

write_csv(gf_data, here("globalprep/prs_cw_pathogen", version_year, "output/po_pathogen_popdensity25mi_gf.csv"))

gf_data_trend <- sani_complete %>%
  dplyr::arrange(rgn_id, year) %>%
  dplyr::group_by(rgn_id) %>%
  dplyr::mutate(gapfill_5year = rollsum(gapfill, 5, align="right", fill=NA)) %>%
  dplyr::mutate(method = paste(na.exclude(unique(method)), collapse = ", ")) %>%
  dplyr::mutate(gapfill = gapfill_5year/5) %>%
  dplyr::select(rgn_id, year, gapfill, method)

write_csv(gf_data, here("globalprep/prs_cw_pathogen", version_year, "output/po_pathogen_popdensity25mi_trend_gf.csv"))

```

## Standarizing sanitation data by population density

First calculate coastal population density (people/km^2^) is calculated by dividing the population within 25 miles of the coast by km^2^ within the 25 mile inland coastal area (yes! This is confusing because area is in km\^2, despite the boundary being 25 miles inland).

```{r, results="hide", message = FALSE, warning = FALSE, eval=FALSE}

# Population within 25 miles of coastline - v2021; last updated in 2021
population <- read_csv(here("globalprep/mar_prs_population/v2021/output/mar_pop_25mi.csv")) %>%
  dplyr::arrange(rgn_id, year)
#since this file has not been updated for v2023 yet we will fill 2021 and 2022 values based on the 2020 value for each region

unique_regions <- unique(population$rgn_id)

# Step 2: Create a new DataFrame with two rows for each region, one for 2021 and one for 2022
new_years <- c(2021, 2022)
new_data <- expand.grid(rgn_id = unique_regions, year = new_years) %>% 
  mutate(popsum = NA)

population <- population %>% 
  rbind(new_data) %>% 
  group_by(rgn_id) %>% 
  fill(popsum, .direction = "down")



#Read area 25mi inland to calculate population density
# (NOTE: this is confusing because it calculates the area in km2 for the 25 mile inland area)
area <- read_csv(here("globalprep/mar_prs_population/v2021/int/area_km2_25mi.csv"))


# People per km2 (for the 25 mile inland area)
pop_density <- population %>% 
  dplyr::left_join(area, by = 'rgn_id') %>% 
  dplyr::mutate(pop_per_km2 = popsum/area_km2) %>% 
  dplyr::select(rgn_id, year, pop_per_km2)

#Save population density data
write_csv(pop_density, here("globalprep/prs_cw_pathogen", version_year, "/intermediate/pathogen_pop_density_25mi.csv"))

```

These data are transformed to a pressure, with a zero score indicating no pressure and 1 indicating the highest possible pressure. Given this we want to determine the number of people *without* access.

The number of people per km\^2 without access to sanitation is calculated by:

1.  converting proportion with access to sanitation to proportion without access to sanitation (i.e., 1 - proportion_with_access).
2.  The proportion without access is multiplied by the coastal population density.
3.  Number of people without access are log transformed (ln(x+1))

```{r, message = FALSE, warning = FALSE, eval=FALSE}

unsani_pop <- sani_complete %>%  
    dplyr::select(rgn_id, rgn_name, year, basic_sani_prop) %>%
    dplyr::left_join(pop_density, 
              by=c('rgn_id', 'year')) %>%
    dplyr::mutate(propWO_x_pop = (1 - basic_sani_prop) * pop_per_km2, # this calculates the population density of people without access (WO)
           propWO_x_pop_log = log(propWO_x_pop + 1)) # log is important because the skew was high otherwise

hist(unsani_pop$propWO_x_pop, main = "people without access")

hist(unsani_pop$propWO_x_pop_log, main = "log of people without access")

```

## Pressure Score

The reference point is the 99th quantile across all countries and years 2000-2009 as a reference point.

```{r pressure, eval=FALSE}

#Calculate reference point
ref_calc <- unsani_pop %>% 
  dplyr::filter(year %in% 2000:2009) %>% #years of reference
  ##summarise(ref= max(propWO_x_pop_log, na.rm = TRUE)*1.1) %>%  # old method
  dplyr::summarise(ref= quantile(propWO_x_pop_log, probs=c(0.99), na.rm = TRUE)) %>% 
  .$ref

ref_calc
#save to the master reference point list - new folder might need to be created for assessment year if this file does not already exist.
master_refs <- read.csv(here("globalprep/supplementary_information", version_year, "/reference_points_pressures.csv"), stringsAsFactors = FALSE)

#create a row for sanitation if it doesn't exist yet, uncomment if it does not ei
# row <- tibble(pressure = "Sanitation", data_years = paste0(data_start, "-", data_end), data_description = "Estimated coastal population with inadequate sanitation", method = "99th quantile across regions/years", ref_year = "2000-2009", ref_point = ref_calc, min= NA, max = NA, units = "number of people", notes = "Reference point is log value")

#master_refs <- master_refs %>% rbind(row)


master_refs$ref_point[master_refs$pressure == "Sanitation"] <- ref_calc

write.csv(master_refs, here("globalprep/supplementary_information", version_year, "/reference_points_pressures.csv"), row.names=FALSE)

master_refs <- read.csv(here("globalprep/supplementary_information", version_year, "/reference_points_pressures.csv")) 
ref_value <- as.numeric(as.character(master_refs$ref_point[master_refs$pressure == "Sanitation"])) 
ref_value #7.10

unsani_prs <- unsani_pop %>%
  dplyr::mutate(pressure_score = propWO_x_pop_log / ref_value) %>% 
  dplyr::mutate(pressure_score = ifelse(pressure_score>1, 1, pressure_score)) %>% #limits pressure scores not to be higher than 1
  dplyr::select(rgn_id, year, pressure_score) 

summary(unsani_prs)

#Save data pressure scores 
write_csv(unsani_prs, here("globalprep/prs_cw_pathogen", version_year, "output/po_pathogen_popdensity25mi.csv"))

# Compare to v22021 data
unsani_prs_old <- read_csv(here("globalprep/prs_cw_pathogen", previous_version_year,"output/po_pathogen_popdensity25mi.csv")) %>%
  rename(pressure_score_2021 = pressure_score) %>% 
  left_join(unsani_prs, by=c("rgn_id", "year")) %>%
  filter(year == 2020)

filter(unsani_prs_old, rgn_id %in% c(185, 208))

  ggplotly(ggplot(unsani_prs_old, aes(y = pressure_score, x = pressure_score_2021, labels = rgn_id)) +
  geom_point() +
    labs(x = paste("pressure score", previous_version_year),
         y = paste("pressure score", version_year), title = "2020 Pressure Comparison") +
  geom_abline(slope = 1, intercept = 0, color = "red"))

```

## Model Trend

Using `CalculateTrend` function form the `ohicore`, trend is calculated by applying a linear regression model to the pressuere scores using a window of 5 years of data. The slope of the linear regression (annual change in pressure) is then divided by the earliest year to get proportional change and then multiplied by 5 to get estimate trend on pressure in the next five years.

```{r trend, eval=FALSE}
#Calculate trend using CalculateTrend()

#Define relevant years: Min and max year of data to calculate trend
first_assess_year <- 2012 
current_assess_year <- 2022
current_data_year <- max(unsani_prs$year) ##max year of data
first_data_year <- first_assess_year - (current_assess_year - current_data_year)

trend_data <- data.frame() #create a data.frame to save trend socores


#For loop: calculates trend for all assess years within the corresponding 5 year window.
#focal_year is the year for which the trend is being calculated.
for(focal_year in first_data_year:current_data_year){ #focal_year = 2017 

  trend_years <- (focal_year-4):focal_year #defines the 5 years window to calculate trend
  
  data_new <- unsani_prs %>% #format data to work in CalculateTrend()
    select(rgn_id, year, status=pressure_score)
  
trend_data_focal_year <- CalculateTrend(data_new, trend_years)

trend_data_focal_year <- trend_data_focal_year %>%
  mutate(year = focal_year) %>%
  select(rgn_id = region_id, year, trend=score) %>%
  data.frame()

trend_data <- rbind(trend_data, trend_data_focal_year) #add trend calculation to data-frame created outside the loop
}
summary(trend_data)

#Save trend data
write_csv(trend_data, here("globalprep/prs_cw_pathogen", version_year, "output/po_pathogen_popdensity25mi_trend.csv"))

```

## Compare to previous years

### Trend data

```{r, eval=FALSE}

new <- read_csv(here("globalprep/prs_cw_pathogen", version_year, "output/po_pathogen_popdensity25mi_trend.csv"))

compare <- read_csv(here("globalprep/prs_cw_pathogen", previous_version_year, "output/po_pathogen_popdensity25mi_trend.csv")) %>%
  select(rgn_id, year, trend_2021 = trend) %>%
  left_join(new, by=c('rgn_id', 'year')) %>%
  filter(year == 2020)


ggplotly(ggplot(compare, aes(y = trend, x = trend_2021, labels = rgn_id)) +
  geom_point() +
    labs(x = paste("Trend",previous_version_year),
         y = paste("Trend", version_year), title = "Data year 2020 Trend Comparison") +
  geom_abline(slope = 1, intercept = 0, color = "red"))

# A few regions have large variation in trend between v2021 and v2019
# positive outliers: rgn_id 161, rgn_id 154
# negative outliers: 48, 8 


# Outlier investigation 

data_old <- read_csv(here("globalprep/prs_cw_pathogen", previous_version_year, "output/po_pathogen_popdensity25mi.csv")) %>% 
  rename(old_pressure_score=pressure_score)

data_new <- read_csv(here("globalprep/prs_cw_pathogen", version_year, "output/po_pathogen_popdensity25mi.csv"))

outlier <- data_new %>% 
  left_join(data_old, by=c("rgn_id", "year")) %>% 
  filter(rgn_id == 224)

plot(outlier$pressure_score, outlier$old_pressure_score)
abline(0,1, col="red")

```

### Compare results

We checked the main discrepancies and these were due to changes in source data. Fairly small changes in access can lead to fairly large changes in pressure scores, depending on the population. There are slightly higher pressure scores this year (indicated by points tending to be above the 1-1 red line) due to modifications of reference point calculations.

### Outlier exploration

```{r outliers, eval=FALSE}

#Comparison of pressure scores

unsani_prs_old <- read_csv(here("globalprep/prs_cw_pathogen", previous_version_year, "output/po_pathogen_popdensity25mi.csv")) %>%
  rename(pressure_score_2019 = pressure_score) %>% 
  left_join(unsani_prs, by=c("rgn_id", "year")) %>% 
  filter(rgn_id %in% c(95, 16, 54, 29))

unsani_prs_old_all <- read_csv(here("globalprep/prs_cw_pathogen", previous_version_year, "output/po_pathogen_popdensity25mi.csv")) %>%
  rename(pressure_score_2019 = pressure_score) %>% 
  left_join(unsani_prs, by=c("rgn_id", "year")) %>% 
  mutate(diff = pressure_score-pressure_score_2019)
```
