---
title: 'OHI 2021: Saltmarsh trend'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

# Summary

This script generates the trend of saltmarsh. 

"The global loss rate of threatened coastal ecosystems is estimated at 4–9% year−1 for corals (Gardner et al. 2003, Bellwood et al. 2004), a minimum of 1–2% year−1 for saltmarshes (Adam 2002; Lotze et al. 2006)"
https://link.springer.com/article/10.1007%2Fs12237-008-9038-7

Apply a decreasing rate of 2% per year, or -0.015 over five years to each region.


## Updates from previous assessment

***
## Data Source 
Duarte, C.M., Dennison, W.C., Orth, R.J.W., Carruthers, T.J.B., 2008. The Charisma of Coastal Ecosystems: Addressing the Imbalance. Estuaries and Coasts: J CERF 31, 233–238. https://doi.org/10.1007/s12237-008-9038-7


**Downloaded**: October 5, 2021

**Description**:  Estimate of a global percentage of loss of saltmarsh habitat per year. 


**Time range**: NA


***
# Methods



## Setup
``` {r setup, echo = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
library(readr)      # for read_csv()
library(raster)
library(here)
library(sf)
library(fasterize)
library(tidyverse)
library(mapview)
library(sp)
library(rgeos)


source(file.path('~/github/ohiprep_v2021/workflow/R/common.R'))

goal     <- 'globalprep/hab_saltmarsh/v2021'
dir_git  <- file.path('~/github/ohiprep_v2021', goal)
region_data()

dir_git_old  <- file.path('~/github/ohiprep_v2021', "globalprep/hab_saltmarsh/v2012")

```


Apply a global trend for each region.. not satisfying. 

```{r, eval = FALSE}
## read in extent data
all <- read.csv(file.path(here("globalprep/hab_saltmarsh/v2021/data/habitat_extent_saltmarsh_updated.csv")))

# specify trend
all <- all %>%
  filter(km2 > 0) %>%
  filter(rgn_id < 255) %>%
  dplyr::select(-year) %>%
  mutate(trend = -0.015)


# Save gapfilling 

trend_gaps <- all %>%
  mutate(variable = "trend") %>%
  mutate(habitat = "saltmarsh") %>%
  mutate(gapfill = 1)
write.csv(trend_gaps, here('globalprep/hab_saltmarsh/v2021/data/trend_saltmarsh_gf.csv'), row.names=FALSE)

### save trend data:
trend <- trend_gaps %>%
  dplyr::select(rgn_id, habitat, trend) %>%
  mutate(year = 2021)
write.csv(trend, here('globalprep/hab_saltmarsh/v2021/data/habitat_trend_saltmarsh_updated.csv'), row.names=FALSE)

```

