---
title: 'OHI `r format(Sys.time(), "%Y")`: Natural Products: Ornamentals Data Prep'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: false
    toc_float: yes
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

[REFERENCE RMD FILE: <http://ohi-science.org/ohiprep_v2024/globalprep/np/v2024/STEP1a_np_ornamentals_prep.html>]

# Summary

This analysis calculates the score for ornamentals per region based on FAO commodities data. This calculation process includes creating a sustainability layer based off of risk and exposure for ornamental fishing.

# Updates from previous assessment

### v2025
- New year of FAO data(2023)

### v2024
*   Complete overhaul of methodology from previous years. Original data prep file: `STEP1a_np_ornamental_prep.Rmd`
*   New year of FAO data (2022)


------------------------------------------------------------------------

# Data Source

**Reference**:\
<https://www.fao.org/fishery/en/statistics/software/fishstatj> App release date: March 2024 FAO raw commodities quantity 1976_2023 FAO raw commodities value 1976_2023 [FAO metadata](http://www.fao.org/fishery/statistics/global-commodities-production/en)

- Global aquatic trade - All partners aggregated - Quantities and Values - 1976-2023 (Release date: July 2025)


**Downloaded**: 2025-07-14

**Description**: Quantity (tonnes) and value (USD) of raw commodities (Exports only) for each country, taxa, year. The FAO data is subset to include commodities in these categories: ornamental fish, fish oil, seaweed and plants (see: raw/commodities2products_weighting.csv for details).

**Time range**: 1976-2023

Files can be found on Mazu: `home/shares/ohi/git-annex/globalprep/_raw_data/FAO_commodities/d2025` accompanied by README.md with detailed download instructions. 

------------------------------------------------------------------------

# Methods


Although we do not know actual sustainable levels of harvest for ornamentals, $S_p$, in each region, we estimated it based on exposure and risk components:

$$S_c = 1 - average(E_c + R_c)$$

where $E_c$ is the exposure term and $R_c$ is the risk term for ornamentals.

The exposure term, $E_c$, is the intensity of harvest for ornamental fish calculated as tonnes of harvest per km2 of coral and rocky reef. This value is scaled by the global 95th quantile of the previous 10 yeas. We scale the harvest because the distribution of values is highly skewed; thus, because we do not know the true threshold of sustainable harvest, nearly all values would be considered highly sustainable otherwise. To estimate rocky reef extent area (km2) we used data from Halpern et al. (2008) ([Halpern et al. 2008](https://ohi-science.org/ohi-methods/goals/goal-models-data.html#ref-halpern2008global)), which assumes rocky reef habitat exists in all cells within 1 km of shore. Coral extent area (km2) are from UNEP-WCMC et al. (2018) ([UNEP-WCMC et al. 2018](https://ohi-science.org/ohi-methods/goals/goal-models-data.html#ref-unep-wcmc2018global)).

The risk term, $R_c$, is based on whether ornamental fishing has unsustainable harvest practices. In specific, we used the intensity of cyanide and dynamite fishing as a proxy. Risk for ornamental fish was set based on assessments of cyanide or dynamite fishing by [Reefs at Risk Revisited](www.wri.org/publication/reefs-at-risk-revisited) under the assumption that most ornamental fishes are harvested from coral reefs.

## Setup

```{r}

knitr::opts_chunk$set(eval = FALSE)

# ======= Load packages ============
if (!require(librarian)) {
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  ohicore, #devtools::install_github('ohi-science/ohicore@dev') #if relevant: restart session after reinstalling
  here,
  tidyverse,
  zoo,
  ggplot2,
  plotly,
  tictoc,
  RColorBrewer
)

# ======= Set directories ===========
# Update scenario year, set up programmatic scenario year updates
scen_year_number <- 2025
scen_year <- as.character(scen_year_number)
prev_scen_year <- as.character(scen_year_number - 1)
data_dir_year <- paste0("d", scen_year)
prev_data_dir_year <- paste0("d", prev_scen_year)
v_scen_year <- paste0("v", scen_year)
prev_v_scen_year <- paste0("v", prev_scen_year)

current_np_dir <- here::here("globalprep", "np", v_scen_year) 
prev_np_dir <- here::here("globalprep", "np", paste0(prev_v_scen_year, "_new")) # update this if name changes

# ======== Source functions ===========
source(here::here("workflow", "R", "common.R")) # directory locations
# Load FAO-specific user-defined functions
source(here::here("workflow", "R", "fao_fxn.R")) # function for cleaning FAO files
# function for handling FAO commodity data specific to NP
source(here::here(current_np_dir, "R", "np_fxn.R")) 
```

STEP 2: Ornamentals
1.    Using data A, filter the tonnes and USD for ornamentals.  
2.    Calculate exposure and risk
3.    then use these to estimate sustainability: mutate(sustainability = 1 - mean(c(exposure, risk), na.rm = TRUE))
4.    Calculate sustainable tonnes: tonnes * sustainability
5.    Rolling average of sustainable tonnes
6.    Identify max value from rolling average of sustainable tonnes (maybe limit to most recent 10 years of data to ID max). 
7.    get relative tonnes by dividing the rolling average of sustainable tonnes by max(rolling average of sustainable tonnes)


So you can then save an intermediate dataset that looks like this:

`rgn_id`\
`year`,\
`product`, (ornamentals, in this case)\
`tonnes`: from gapfilled FAO commodity data\
`usd`: from gapfilled FAO commodity data\
`sustainability`: `1 - mean(c(exposure, risk), na.rm = TRUE)`\
`sustainable_tonnes`: `tonnes` $*$ `sustainability`\
`max_sustainable_tonnes`: `max(sustainable_tonnes)`\
`rel_sust_tonnes`: `tonnes_rollmean` / `peak_sust_tonnes`

Then, in final data, save `rgn_id`, `product`, `year`, `rel_sust_tonnes` 


```{r}
# ---- Subset np harvest ornamentals ----

# read in FAO commodities usd tonnes int data
np_harvest_usd_tonnes <- read_csv(here(current_np_dir, "int", "np_harvest_usd_tonnes.csv"))

# filter FAO commodities int data to ornamentals
np_ornamentals_base <- np_harvest_usd_tonnes %>% 
  dplyr::filter(product %in% c("ornamentals"))

# check
table(np_ornamentals_base$product)
# 3938 -- v2024
# 4142 -- v2025

# ---- tonnes only ----
# select relevant columns
np_harvest_orn <- np_ornamentals_base %>%
  dplyr::select(rgn_id, rgn_name, year, product, tonnes) %>% 
  # hard-coded year cutoff point to ensure just enough years of data for rolling average calculations
  dplyr::filter(year >= 2004) %>% 
  arrange(rgn_id)

# write out tonnes subset to int
readr::write_csv(np_harvest_orn, here(current_np_dir, "int", "np_harvest_ornamentals_tonnes.csv"))

# ---- usd and tonnes ----
# select relevant columns (including usd this time)
np_harvest_orn_full <- np_ornamentals_base %>%
  dplyr::select(rgn_id, rgn_name, year, product, tonnes, usd) %>% 
  # hard-coded year cutoff point to ensure just enough years of data for rolling average calculations
  dplyr::filter(year >= 2004) %>% 
  arrange(rgn_id)

# write out tonnes subset to int
readr::write_csv(np_harvest_orn_full, here(current_np_dir, "int", "np_harvest_ornamentals_tonnes_usd.csv"))

```

Gapfill risk and exposure, everything. 

## Calculate Exposure

Calculate ornamentals exposure based on habitats. 
exposure: get tonnes / area, find global 95th quantile from past 10 years, 
use 95th quantile value to rescale exposure from 0-1

Returns the first input data frame with a new column for exposure:
[`rgn_id` `year` `product` `exposure`]

```{r}
# ====== Determine Habitat Areas for Exposure ========
# rocky reef habitat area
hab_rocky <- readr::read_csv(here::here("globalprep", "hab_rockyreef", "v2012", 
                           "data", "habitat_extent_rocky_reef_updated.csv")) %>%
  dplyr::select(rgn_id, km2) %>%
  dplyr::filter(km2 > 0)

# coral habitat area
hab_coral <- readr::read_csv(here::here("globalprep", "hab_coral", "v2021", 
                           "data", "habitat_extent_coral_updated.csv")) %>%
  dplyr::select(rgn_id, km2) %>%
  dplyr::filter(km2 > 0)

# find total area in coral and rocky reef habitats
area_dual_hab <- np_ornamentals_base %>%
  select(rgn_name, rgn_id, product, year, tonnes) %>%
  dplyr::left_join(hab_coral %>%
                     dplyr::rename(coral_km2 = km2),
                   by = c('rgn_id')) %>%
  dplyr::left_join(hab_rocky %>%
                     dplyr::rename(rocky_km2 = km2),
                   by = c('rgn_id')) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(km2 = sum(c(rocky_km2, coral_km2), na.rm = TRUE)) %>%
  dplyr::filter(km2 > 0) %>%
  dplyr::select(-coral_km2, -rocky_km2) %>% 
  dplyr::arrange(rgn_id)
  

# ======== Determine Exposure ========
### exposure: get tonnes / area, find global 95th quantile from past 10 years, 
#### use 95th quantile value to rescale exposure from 0-1

# calculate raw exposure (tonnes / area) ----
np_exp_raw <- area_dual_hab %>%
  # find regions and years with harvest in rocky reef and coral habitat
  dplyr::mutate(expos_raw = ifelse(tonnes > 0 & km2 > 0,
                                   (tonnes / km2), # raw exposure: harvest tonnes / habitat area
                                   0))  # 0 exposure if no harvest or no special habitat
  

# define most recent 10 year range ----
max_exp_year <- max(np_exp_raw$year)
exp_decade_range <- seq((max_exp_year - 9), max_exp_year, 1)


# ---- find 95th quantile of the past 10 years ----

# subset of past 10 years
np_exp_recent_decade <- np_exp_raw %>% 
  dplyr::filter(year %in% c(exp_decade_range))

global_95th_quantile <- stats::quantile(np_exp_recent_decade$expos_raw, probs = c(0.95))
global_95th_quantile
# v2024: 95th is 0.151; 99th is 45.90351 
# v2025: 95th is 0.136; 99th is 40.86775 


# ---- calculate rescaled exposure ----
# new rescale method: 
# if raw exposure >= 95th quantile, exposure = 1
# else, exposure = raw exposure / global 95th quantile

np_exposure_new <- np_exp_raw %>% 
  # bring in global 95th quantile value 
  dplyr::mutate(global_95th = global_95th_quantile) %>% 
  # rescale following new method outlined above
  dplyr::mutate(exposure = case_when(
    expos_raw >= global_95th ~ 1,
    expos_raw < global_95th ~ (expos_raw / global_95th))
  )

# There are no NAs across the data frame
colSums(is.na(np_exposure_new))

summary(np_exposure_new$exposure)

# write out exposure status
readr::write_csv(np_exposure_new, here(current_np_dir, "int", "np_exposure.csv"))

```


```{r}
# datacheck:
# plot histogram to show distribution of rescaled points

ggplot(data = np_exposure_new) +
  geom_histogram(aes(x = exposure),
                 color = "blue",
                 fill = "cornflowerblue") +
  theme_bw()

```

# Gapfilling

**UPDATE**: The 2025 data has no NAs, so this process was halted. The code will be kept for future assessments in case gapfilling is needed. 

The 2024 assessment only gapfilled with the global mean score. For the 2025 assessment, this method will be slightly improved to also include gapfilling by region first, before gapfilling by global mean. 

1. Group by region and fill with the mean exposure score
2. Group by georegion and fill with the mean exposure score
3. Fill with the global mean exposure score

```{r save-exposure, eval=FALSE}
# ======== Save intermediate data products ==========
# Create gapfilling record keeping columns to use later
np_exp_gf <- np_exposure_new %>%
  dplyr::mutate(gapfilled = ifelse(is.na(exposure), 1, 0)) 
  
# Add georegion column for gapfilling
np_exposure_rgn <- np_exp_gf %>%
  add_georegion_id() %>%
  dplyr::select(-level)

# add exposure for countries with (habitat extent == NA)
np_exp_orn <- np_exposure_rgn %>% 
  # Gapfill by region
  dplyr::group_by(rgn_id, product) %>% 
  dplyr::mutate(rgn_exp = mean(exposure, na.rm = TRUE)) %>% 
  dplyr::mutate(exposure = ifelse(is.na(exposure), rgn_exp, exposure)) %>% 
  mutate(method = ifelse(gapfilled == 1 & exposure == rgn_exp, "region_avg", NA)) %>%
  dplyr::ungroup() %>% 
  # Gapfill by georegion
  dplyr::group_by(georgn_id, product) %>% 
  dplyr::mutate(georgn_exp = mean(exposure, na.rm = TRUE)) %>% 
  dplyr::mutate(exposure = ifelse(is.na(exposure), georgn_exp, exposure)) %>% 
  mutate(method = ifelse(gapfilled == 1 & exposure == georgn_exp, "georegion_avg", NA)) %>%
  dplyr::ungroup() %>% 
  # Gapfill with global mean
  dplyr::group_by(product) %>% 
  dplyr::mutate(mean_exp = mean(exposure, na.rm = TRUE)) %>%
  dplyr::mutate(exposure = ifelse(is.na(exposure), mean_exp, exposure)) %>%
  mutate(method = ifelse(gapfilled == 1 & exposure == mean_exp, "global_avg", NA)) %>%
  dplyr::ungroup() %>%
  # coerce data type
  dplyr::mutate(product = as.character(product))

np_exp_orn_final <- np_exp_orn %>% 
  # select relevant columns
  dplyr::select(rgn_id, year, product, exposure)

# save ornamentals exposure to int 
readr::write_csv(np_exp_orn_final, here(current_np_dir,
                                  "int",
                                  "np_exposure_ornamentals.csv"))

# Gapfilling record keeping
np_exp_gf <- np_exp_orn %>%
  dplyr::select(rgn_id, product, year, gapfilled, method)

# write out to int folder
readr::write_csv(np_exp_gf, here(current_np_dir, 
                                 "int",
                                 "np_exposure_ornamentals_gf.csv"))
```

```{r}
# datacheck:
library(RColorBrewer)
np_orn_exposure_plot <- ggplot(data = np_exp_orn, aes(x = year, y = exposure)) +
  geom_line(aes(color = rgn_name)) +
  theme_bw()

plotly::ggplotly(np_orn_exposure_plot)
```


## Calculate Risk


Calculate NP risk based on blast fishing and cyanide fishing. For ornamentals:

risk = 1 if blast fishing or cyanide fishing 

Create a dataframe of risk per product per region.

```{r}

### calculates NP risk based on:
###   ornamentals:      risk = 1 if blast or cyanide fishing
### Returns a data frame of risk, by product, by region:
###
#########################################.


# ----- Determine Risk --------

r_cyanide <- read_csv(here("globalprep", "np_prs_poison_blast_fishing",
                           "v2013", "data", "gl_thr_poison_3nm_rgn2013.csv")) %>%
  #AlignDataYears(layer_nm = "np_cyanide", layers_obj = layers) %>%
  dplyr::filter(!is.na(score) & score > 0) %>%
  dplyr::select(rgn_id,
                #year = scenario_year,
                cyanide = score)

r_blast <- read_csv(here("globalprep", "np_prs_poison_blast_fishing",
                         "v2013", "data", "gl_thr_blast_3nm_rgn2013.csv")) %>%
  #AlignDataYears(layer_nm = "np_blast", layers_obj = layers)  %>%
  dplyr::filter(!is.na(score) & score > 0) %>%
  dplyr::select(rgn_id,
         #year = scenario_year,
         blast = score)


# risk for ornamentals set to 1 if blast or cyanide fishing present, based on Nature 2012 code
##  despite Nature 2012 Suppl saying Risk for ornamental fish is set to the "relative intensity of cyanide fishing"
risk_orn <- r_cyanide %>%
  full_join(r_blast, by = c("rgn_id")) %>%
  mutate(ornamentals = 1) %>%
  dplyr::select(rgn_id, ornamentals)

# define risk as binary if blast or cyanide fishing
np_risk_orn <-
  # create a dataframe from all combinations of rgn ID and year
  expand.grid(
    rgn_id  = unique(np_ornamentals_base$rgn_id),
    year = unique(np_ornamentals_base$year)
  ) %>%
  # join with ornamentals
  left_join(risk_orn, by = c("rgn_id"))  %>%
  mutate(ornamentals = ifelse(is.na(ornamentals), 0, ornamentals)) %>%
#  gather(product, risk, -rgn_id, -year) %>%
#  mutate(product = as.character(product))
  mutate(product = "ornamentals") %>% 
  rename(risk = ornamentals) %>% 
  dplyr::select(rgn_id, year, product, risk)

# track gapfilling for records
np_risk_gf <- np_risk_orn %>%
  mutate(gapfilled = 0, method = NA) %>%
  dplyr::select(-risk)


# write out risk gapfilling status
readr::write_csv(np_risk_gf, here(current_np_dir, "int", "np_risk_ornamentals_gf.csv"))

# write this out to int (no longer a risk layer for ohi-global)
readr::write_csv(np_risk_orn, here(current_np_dir, "int", "np_risk_ornamentals.csv"))
```


## Calculate Sustainability

`1 - mean(c(exposure, risk), na.rm = TRUE)`


```{r}
# read in exposure
exposure <- readr::read_csv(here(current_np_dir, "int", "np_exposure.csv"))

# read in risk
risk <- readr::read_csv(here(current_np_dir, "int", "np_risk_ornamentals.csv"))

# ---- calculate sustainability ----

# join dataframes
orn_exposure_risk <- exposure %>% 
  full_join(risk, by = join_by("rgn_id", "year", "product"))

# calculate sustainability
orn_sust <- orn_exposure_risk %>% 
  ungroup() %>% 
  group_by(rgn_id, year) %>% 
  mutate(sustainability = 1 - mean(c(exposure, risk), na.rm = TRUE)) %>% 
  ungroup() %>%
  select(rgn_id, year, exposure, risk, sustainability)

```


## Calculate Sustainable Harvest

`usd * sustainability`
This is going to be based on USD. The tonnes data seems pretty suspect in many cases and the USD data is better. And, in this case, since the USD and tonnes data are the same source they should be equivalent (vs. fishoil/fishmeal and seaweed).

```{r}
# read in ornamentals harvest (same as earlier chunk)
np_harvest_ornamentals <- readr::read_csv(here(current_np_dir, "int", "np_harvest_ornamentals_tonnes_usd.csv"))

# join sustainability and tonnes
sust_harvest_join <- np_harvest_ornamentals %>% 
  full_join(orn_sust, by = join_by("rgn_id", "year")) %>% 
  filter(year >= 2004) # ensure that proper start year is set

# check summary
summary(sust_harvest_join)
# note: 548 NAs in exposure, 0 NAs in risk 
# (and none in sustainability, since we took the mean and set na.rm = TRUE) 
#str(sust_tonnes) # checking datatypes

# calculate sustainable harvest
sust_harvest_orn <- sust_harvest_join %>% 
  mutate(sust_usd = usd * sustainability) %>%
  filter(!is.na(sust_usd))


# check
summary(sust_harvest_orn)

```



## Rolling Average of Sustainable Tonnes

```{r}
# Find max year in the summarized data table
year_max <- max(sust_harvest_orn$year)

# for each region and year, calculate rolling mean in a 5 year window (year + 4 years, right-aligned (year, look at the 4 years previous))
sust_usd_rolling <- sust_harvest_orn %>%
  arrange(rgn_id, product, year) %>%
  group_by(rgn_id, product) %>%
  mutate(
      sust_usd_rollmean = zoo::rollapply(sust_usd, width = 4,
                                            FUN = mean, align = 'right',
                                            partial = TRUE, na.rm = FALSE)) %>%
  dplyr::select(rgn_id, rgn_name, year, product, tonnes, usd, sustainability, sust_usd, sust_usd_rollmean)


```


## Identify Max Sustainable Tonnes

Find the maximum sustainable tonnes per region within the past 10 years.

```{r}
# subset to past 10 years, find max sustainable tonnes 

# define 10 year range
sust_decade <- c((year_max - 10):year_max)

# find peak sustainable tonnes 
max_sust_usd <- sust_usd_rolling %>% 
  filter(year %in% c(sust_decade)) %>% 
  ungroup() %>% 
  group_by(rgn_id) %>% 
  summarize(peak_usd = max(sust_usd_rollmean, na.rm = TRUE))
  

```


## Relative Sustainable Harvest

Will evaluate this in USD. Idea is that sustainable dollar value does not decrease relative to historical harvest value. 

```{r}
# ---- Determine relative sustainable usd value ----
orn_relative_sust_usd <- 
  # join peak sustainable tonnes with full sustainable tonnes df
  sust_usd_rolling %>% 
  ungroup() %>% 
  ## filling in peak value per region ID across full year range
  full_join(max_sust_usd, by = join_by("rgn_id")) %>% 
  # rescale to find relative sustainable tonnes
  mutate(rel_sust_usd = ifelse(sust_usd_rollmean > peak_usd, 1,
                             sust_usd_rollmean / peak_usd)) %>%
  filter(!is.na(rel_sust_usd))
 
# check
summary(orn_relative_sust_usd)


# save full df to int folder
readr::write_csv(orn_relative_sust_usd, here(current_np_dir, "int", "np_ornamentals_full.csv"))


# save selected columns to output folder
final_orn_relative_sust_usd <- orn_relative_sust_usd %>% 
  dplyr::select(rgn_id, product, year, score= rel_sust_usd)

readr::write_csv(final_orn_relative_sust_usd, here(current_np_dir, "output", "np_ornamentals.csv"))
```

