---
title: 'OHI `r format(Sys.time(), "%Y")`: Natural Products: weighting data prep'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
#  chunk_output_type: inline
  markdown: 
    wrap: 72
---

[REFERENCE RMD FILE:
<http://ohi-science.org/ohiprep_v2021/globalprep/np/v2021/STEP2_np_weighting_prep.html>]

# Summary
[note for future analyses: I think this information would be better incorporated into the individual product scripts.
I just realized that different commodities will have different rates of updates and years.]
This analysis converts FAO commodities tonnes data for ornamental fish
trade, SAUP fisheries tonnes data for FOFM, and FAO mariculture tonnes
data for seaweeds into a weighting scheme per each ohi region, based on
relative USD value of these products. This
weighting scheme will be applied to the scores to determine how much of
each natural products layer affects the overall NP score when updating
in ohi global.

Data is all generated in previous NP scripts.

# Methods

1. Read in the intermediate files with absolute tonnes of production (raw data)
2. Read in intermediate data B (USD/tonnes)
3. Merge data B and tonnes
4. Calculate total USD for each product
5. Do 5 year averages (or similar)
6. Calculate general weights for each product (products should sum to one for a region) 


## Setup

```{r setup}

knitr::opts_chunk$set(eval = FALSE)

# ======= Load packages ============
if (!require(librarian)) {
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  ohicore, #devtools::install_github('ohi-science/ohicore@dev') #if relevant: restart session after reinstalling
  here,
  purr,
  tidyverse,
  zoo,
  plotly,
  tictoc,
  RColorBrewer
)

# ======= Set directories ===========
# Update scenario year, set up programmatic scenario year updates
current_year <- 2025 # update this!!!
```

## Tonnes per Product data

Create the overall region product data frame that contains the 5 years of interest and the three products: ornamentals, seaweeds, and fish oil fish meal.


```{r}


# get USD per tonne averaged across georegion for each product and year
t_usd <- read_csv(here(sprintf("globalprep/np/v%s/int/np_usd_per_tonne.csv", current_year))) %>%
    add_georegion_id() %>%
    group_by(product, georgn_id, year) %>%
    mutate(avg_usd_per_tonne = mean(usd_per_tonne, na.rm=TRUE)) %>%
    select(rgn_name, rgn_id, product, year, georgn_id, usd, avg_usd_per_tonne) %>%
  ungroup()

```


### Tonnes harvest data
The data above only include the tonnes of product that is globally traded. However, much of the product may be consumed in the country. For that reason we use the sea around us catch data and mariculture data for fofm and seaweed.

```{r}
fofm <- read_csv(here(sprintf("globalprep/np/v%s/int/tonnes_fofm.csv", current_year))) %>%
  mutate(product = "fish_oil") %>%
  select(product, rgn_id, year, tonnes)
seaweed <- read_csv(here(sprintf("globalprep/np/v%s/int/np_relative_sust_tonnes_seaweeds_full.csv", current_year))) %>%
  mutate(tonnes = tonnes_sustainable*1/0.79) %>% #want total tonnes, not just sustainable tonnes
  select(product, rgn_id, year, tonnes)

tonnes <- rbind(fofm, seaweed) # for ornamentals the FAO trade data alone is used to estimate USD

rates <- left_join(t_usd, tonnes, by=c("rgn_id", "product", "year")) %>%
  filter(year >= 1990) %>%
  mutate(tonnes = ifelse(is.na(tonnes), 0, tonnes)) 

total_usd <- rates %>%
  mutate(est_usd = avg_usd_per_tonne*tonnes) %>%
  mutate(usd_weight = ifelse(usd>est_usd, usd, est_usd)) %>%
  select(rgn_id, year, usd, product)

fish <- total_usd %>%
  filter(product == "fish_oil") %>%
  filter(year %in% 1990:2019)

sw <- total_usd %>%
  filter(product == "seaweeds") %>%
  filter(year %in% 2010:2023)

orns <- total_usd %>%
  filter(product == "ornamentals") %>%
  filter(year %in% 2010:2023)

write_csv(fish, here("globalprep/np/v2025/output/np_fofm_weights.csv"))
write_csv(sw, here("globalprep/np/v2025/output/np_seaweed_weights.csv"))
write_csv(orns, here("globalprep/np/v2025/output/np_ornamentals_weights.csv"))

```
