---
title: 'OHI `r format(Sys.time(), "%Y")`: Natural Products, Preparing catch and B/Bmsy data for forage fish species'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

[REFERENCE RMD FILE: <http://ohi-science.org/ohiprep_v2024/globalprep/np/v2024/STEP1c_np_fishfeed_prep.html>]

# Summary

This script takes the SAUP 2023 catch data, a list of fish oil/fish meal (FOFM) species, and B/Bmsy estimates to calculate a score for FOFM species per OHI region. The FIS data prep must be completed prior to this step.

## Updates from previous assessment

- New RAM data v4.66

------------------------------------------------------------------------

# Data Source

## Sea Around us

**Reference**: Pauly D., Zeller D., Palomares M.L.D. (Editors), 2020. Sea Around Us Concepts, Design and Data (seaaroundus.org).

**Downloaded**: September 27, 2022

**Description**: Tons per year and SAUP region with information on sector type, industry type, fishing entitity, reporting status and taxonomic information.

**Time range**: 1950 - 2019

**Format**: CSV

**Additional Information**: [Methods](http://www.seaaroundus.org/sea-around-us-methods-index/)


## RAM Legacy Data

**Reference**: [RAM Legacy Stock Assessment Database](http://ramlegacy.org) v4.66

**Downloaded**: 06/26/2025

**Description**: B/Bmsy value by stock and year (other data, which we do not use, are also available in the database)

**Native data resolution**: stock (fish stock, species and region specific)

**Time range**: 1800 - 2023

**Format**: CSV format

**Additional Information**: We use the finalized b/bmsy layer from OHI-global for this data prep. We do not actually read in the raw RAM data here.

------------------------------------------------------------------------

# Methods

INCOMPLETE. See this folder's `README.md` for details. 


## Setup

```{r}
knitr::opts_chunk$set(eval = FALSE)

# ======= Load packages ============
if (!require(librarian)) {
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  ohicore, #devtools::install_github('ohi-science/ohicore@dev') #if relevant: restart session after reinstalling
  here,
  tidyverse,
  zoo,
  plotly,
  tictoc,
  RColorBrewer
)

# ======= Set directories ===========
# Update scenario year, set up programmatic scenario year updates
current_year <- 2025 # update this!!!
previous_year <- current_year - 1

# Directories
data_dir <- paste0("d", current_year)
prev_data_dir <- paste0("d", previous_year)

# Scenarios
scen_year <- paste0("v", current_year)
prev_scen_year <- paste0("v", previous_year)

data_years <- c("1950", "2023") # update to correct data years

# Natural Products directory on GitHub
current_np_dir <- here::here("globalprep", "np", scen_year) 
previous_np_dir <- here::here("globalprep", "np", paste0(prev_scen_year, "_new")) # update this if name changes

# Fisheries directory on GitHub
current_fis_dir <- here::here("globalprep", "fis", scen_year)
previous_fis_dir <- here::here("globalprep", "fis", prev_scen_year)

# ======== Source functions ===========
source(here::here("workflow", "R", "common.R")) # directory locations
# Load FAO-specific user-defined functions
source(here::here("workflow", "R", "fao_fxn.R")) # function for cleaning FAO files
# function for handling FAO commodity data specific to NP
source(here::here(current_np_dir, "R", "np_fxn.R")) 
# functions specific to mariculture dealing with compound countries
source(here("globalprep", "mar", scen_year, "mar_fxs.R"))

```

# STEP 1

Identify catch tonnes used for B/Bmsy. We use the catch average catch since 1990 to weight the yearly B/Bmsy scores for these taxa. We use the SAUP designations of whether catch is used for fishoil fishmeal.
```{r, eval=FALSE}

saup_taxonkey <- readr::read_csv(here(current_fis_dir, "int", "taxon_key_v2022.csv"))

fish <- read_csv("/home/shares/ohi/git-annex/globalprep/fis/v2022/int/ohi_rgn_eez_prod.csv") 

catch <- fish %>%
  select(rgn_id, fao_id, year, end_use_type, scientific_name, common_name, tonnes) %>%
  filter(end_use_type == "Fishmeal and fish oil") %>%
  filter(year >= 1990) %>%
  mutate(stock_id = paste(scientific_name, fao_id, sep = "-")) %>%
  select(rgn_id, year, stock_id, tonnes)


catch <- catch %>%
  group_by(rgn_id, year, stock_id) %>%
  summarize(tonnes = sum(tonnes))


## add in 2020 - 2023 (using the last value for year 2019)
copies_2020_2023 <- purrr::map_dfr(2020:2023, ~ filter(catch, year==2019) %>% mutate(year = .x))

catch <- bind_rows(catch, copies_2020_2023) %>%
  arrange(rgn_id, stock_id, year)


```

Gapfill with trade data if no evidence of fishoil fishmeal catch in SAUP data.

```{r}

catch_rgns <- catch %>%
  group_by(rgn_id, year) %>%
  summarize(stocks = sum(tonnes)) %>%
  mutate(rgn_year = paste(rgn_id, year, sep="_"))

fao <- read_csv(here(sprintf("globalprep/np/v%s/int/np_usd_per_tonne.csv", current_year))) %>%
  filter(product == "fish_oil") %>%
  mutate(rgn_year = paste(rgn_id, year, sep="_"))


missing_catch_id <- setdiff(fao$rgn_year, catch_rgns$rgn_year)

missing_catch <- fao %>%
  filter(rgn_year %in% missing_catch_id) %>%
  mutate(stock_id = "Marine fishes not identified") %>%
  select(rgn_id, year, stock_id, tonnes)


catch <- rbind(catch, missing_catch)

```


# B/Bmsy

Convert the B/Bmsy values to scores (this is done in functions.R for FIS subgoal, but we do it in the dataprep for NP), apply the underharvest penalty, and take a catch weighted average of B/Bmsy scores for each region/year.

```{r}

fis_bbmsy <- read_csv(here(current_fis_dir, "output", "fis_bbmsy.csv")) %>%
  select(stock_id, rgn_id, year, bbmsy)


b <- fis_bbmsy %>%
  dplyr::mutate(bbmsy = ifelse(bbmsy > 1 , 1, bbmsy)) %>%
  dplyr::mutate(bbmsy = as.numeric(bbmsy),
                year = as.numeric(as.character(year)),
                stock_id = as.character(stock_id)) %>%
  dplyr::mutate(stock_id = str_replace(stock_id, "_", " "))


####
# ==== Calculate scores for Bbmsy values ==== ####
#  *************NOTE *****************************
#  These values can be altered
#  ***********************************************

## The upper limit can be deleted, since we already cap scores at 1 anyways. We still want to keep the lower buffer.

alpha <- 0.5
beta <- 0.25
lowerBuffer <- 0.95
upperBuffer <- 1.05

# Rename to stock score

b$stock_score = ifelse(
  b$bbmsy < lowerBuffer,
  b$bbmsy,
  ifelse (b$bbmsy >= lowerBuffer &
            b$bbmsy <= upperBuffer, 1, NA)
)
b$stock_score = ifelse(!is.na(b$stock_score),
                 b$stock_score,
                 ifelse(
                   1 - alpha * (b$bbmsy - upperBuffer) > beta,
                   1 - alpha * (b$bbmsy - upperBuffer),
                   beta
                 ))

```

# Merge B/Bmsy and catch data

```{r}

data_fis <- catch %>%
  dplyr::left_join(b, by = c('stock_id', 'year', 'rgn_id')) 

setdiff(catch$stock_id, b$stock_id) ## need to do a lot of gapfilling here.
setdiff(b$stock_id, catch$stock_id)
intersect(b$stock_id, catch$stock_id) 

tmp <- data_fis %>%
 group_by(rgn_id, year) %>%
 summarize(total_bbmsy = sum(!is.na(stock_score)),
           total = n())

###
#  Estimate scores for taxa without b/bmsy values --------------------
# Mean score of other fish in the region is the starting point
# Then a penalty is applied based on the level the taxa are reported at
###

## this takes the mean score within each region and year
data_fis_gf <- data_fis %>%
  dplyr::group_by(rgn_id, year) %>%
  dplyr::mutate(mean_score = mean(stock_score, na.rm = TRUE)) %>%
  dplyr::ungroup()

## this takes the mean score across all regions within a year
# (when no stocks have scores within a region)
data_fis_gf <- data_fis_gf %>%
  dplyr::group_by(year) %>%
  dplyr::mutate(mean_score_global = mean(stock_score, na.rm = TRUE)) %>%
  dplyr::ungroup() 

data_fis_gf <- data_fis_gf %>%
  mutate(stock_score = ifelse(is.na(stock_score), mean_score, stock_score)) %>%
  mutate(stock_score = ifelse(is.na(stock_score), mean_score_global, stock_score))


## get tonnes per region for weighting
tonnes <- data_fis_gf %>%
  group_by(year, rgn_id) %>%
  summarize(tonnes = sum(tonnes, na.rm=TRUE)) %>%
  mutate(product = "fish_oil")
write_csv(tonnes, here(current_np_dir, "int", "tonnes_fofm.csv"))

```

## Calculate score

```{r}

# We use the mean catch for each region/stock to weight the yearly B/Bmsy values.

score_data <- data_fis_gf %>%
  group_by(rgn_id, stock_id) %>%
  mutate(mean_tonnes = mean(tonnes, na.rm=TRUE)) %>%
  filter(!is.nan(stock_score)) %>% # only have stock scores from 1990 in these data.
  ungroup()
# Calculate status for each region
###

## Take a catch weighted average of B/Bmsy scores for each region/year.


score_data <- score_data %>%
  dplyr::group_by(rgn_id, year) %>%
  dplyr::summarize(score = weighted.mean(stock_score, mean_tonnes)) %>%
  dplyr::ungroup() %>% 
  mutate(product = "fish_oil")

summary(score_data)

readr::write_csv(score_data, here(current_np_dir, "output", "np_fofm.csv"))

old <- read_csv(here(current_np_dir, "output", "np_fofm_scores.csv"))

new <- read_csv(here(current_np_dir, "output", "np_fofm.csv"))

check <- new %>%
  rename("new_score" = "score") %>%
  left_join(old, by = c("rgn_id", "year")) %>%
  select(rgn_id, year, new_score, score) %>%
  mutate(diff = new_score - score) %>%
  filter(year == 2019)

ab_scores <- ggplot() + 
  geom_abline(slope = 1, intercept = 0, col = "red") +
  geom_point(data = check, aes(x = score, y = new_score)) +
  labs(x = "Old Score", 
       y = "New Score",
       title = "FOFM New score vs old score for 2019") +
  theme_bw();ggplotly(ab_scores)

```

