---
title: 'OHI `r format(Sys.time(), "%Y")`: Natural Products: weighting data prep'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
#  chunk_output_type: inline
  markdown: 
    wrap: 72
---

[REFERENCE RMD FILE:
<http://ohi-science.org/ohiprep_v2021/globalprep/np/v2021/STEP2_np_weighting_prep.html>]

# Summary
[note for future analyses: I think this information would be better incorporated into the individual product scripts.
I just realized that different commodities will have different rates of updates and years.]
This analysis converts FAO commodities tonnes data for ornamental fish
trade, SAUP fisheries tonnes data for FOFM, and FAO mariculture tonnes
data for seaweeds into a weighting scheme per each ohi region, based on
relative USD value of these products. This
weighting scheme will be applied to the scores to determine how much of
each natural products layer affects the overall NP score when updating
in ohi global.

Data is all generated in previous NP scripts.

# Methods

1. Read in the intermediate files with absolute tonnes of production (raw data)
2. Read in intermediate data B (USD/tonnes)
3. Merge data B and tonnes
4. Calculate total USD for each product
5. Do 5 year averages (or similar)
6. Calculate general weights for each product (products should sum to one for a region) 


## Setup

```{r setup}

knitr::opts_chunk$set(eval = FALSE)

# ======= Load packages ============
if (!require(librarian)) {
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  ohicore, #devtools::install_github('ohi-science/ohicore@dev') #if relevant: restart session after reinstalling
  here,
  purr,
  tidyverse,
  zoo,
  plotly,
  tictoc,
  RColorBrewer
)

# ======= Set directories ===========
# Update scenario year, set up programmatic scenario year updates
current_year <- 2025 # update this!!!
```

## Tonnes per Product data

Create the overall region product data frame that contains the 5 years of interest and the three products: ornamentals, seaweeds, and fish oil fish meal.


```{r region-product}
# All commodities is only used here to get the max year. It gets used later on for the ornamentals tonnes
harvest_tonnes_usd <- read_csv(here(current_np_dir, "int", "np_harvest_usd_tonnes.csv"))

# Make a data frame with every region and every product ----
# Define years
max_year <- max(harvest_tonnes_usd$year)
max_year # 2023
min_year <- max_year - 4 
years <- c(min_year:max_year) 
years_df <- data.frame(year = min_year:max_year) # so that we can get 5 year average

# Load OHI region data from common.R
region_data() # gets rgns_eez

# Define subset of OHI EEZ regions
rgns_eez <- rgns_eez %>%
  dplyr::select(rgn_id) %>%
  filter(rgn_id != 213) # filter out Antarctica

# Define products dataframe
products <- data.frame(product = c("seaweeds", "ornamentals", "fish_oil"))

# Create full rgn x product x years dataframe
region_product <- rgns_eez %>% 
  # select rgn_id from rgns_eez
  dplyr::select(rgn_id) %>%  
  # cross join with products (match every row in x with every row in y)
  cross_join(products) %>% 
  # cross join with years
  cross_join(years_df) 
```

Now that the skeleton of region_product has been defined, we will fill in the individual commodity tonnes.

```{r individual-prep}
# Subset ornamentals product
orn_fill_df <- region_product %>% 
  filter(product == "ornamentals")

# Use the original FAO commodity data to get ornamentals quantities
orn_tonnes <- harvest_tonnes_usd %>%
  # filter to 5 year range defined in the beginning of the first methods chunk
  dplyr::filter(year %in% years, 
                product == "ornamentals") %>% 
  # full join to fill in the rest of the regions with 0
  full_join(orn_fill_df, by = c("rgn_id", "year", "product")) %>% 
  # gapfill the NAs to be 0
  mutate(tonnes = ifelse(is.na(tonnes), 0, tonnes),
         usd = ifelse(is.na(usd), 0, usd)) %>% 
  dplyr::select(rgn_id, year, product, tonnes)

write_csv(orn_tonnes, here(current_np_dir, "int", "orn_tonnes.csv"))

# Subset seaweed products
sw_fill_df <- region_product %>%
  filter(product == "seaweeds")

# Read in the seaweed MAR data for tonnes
sw_tonnes_raw <- read_csv(here(current_np_dir, "int", "np_seaweeds_tonnes_weighting.csv"))

sw_tonnes <- sw_tonnes_raw %>%
  mutate(product = "seaweeds") %>%
  group_by(rgn_id, year, product) %>% 
  summarise(tonnes = sum(tonnes, na.rm = TRUE)) %>% # sum across all species of seaweed
  dplyr::filter(year %in% years) %>% # filter to 5 year range 
  full_join(sw_fill_df, by = c("rgn_id", "year", "product")) %>%
    mutate(tonnes = ifelse(is.na(tonnes), 0, tonnes)) %>% # gapfill the NAs to be 0
  dplyr::select(rgn_id, year, product, tonnes)

write_csv(sw_tonnes, here(current_np_dir, "int", "sw_tonnes.csv"))

# Subset fofm products
fofm_fill_df <- region_product %>% 
  filter(product == "fish_oil")

## Read in FOFM tonnes data
fofm_tonnes <- read_csv(here(current_np_dir, "int", "raw_mean_catch_FOFM.csv")) %>%
  dplyr::filter(year %in% years) %>% 
  group_by(rgn_id, year, product) %>%
  # Note: multiply by 0.3 to account for water loss when converting fish to fish oil.. 
  ## about 30% of fish are water and 70% are fish oil
  summarise(tonnes = sum(catch_non_human) * 0.3) %>% 
  ungroup() %>%
  full_join(fofm_fill_df, by = c("rgn_id", "year", "product")) %>%
    mutate(tonnes = ifelse(is.na(tonnes), 0, tonnes)) %>% # gapfill the NAs to be 0
  dplyr::select(rgn_id, year, product, tonnes)


# EXTRA STEP FOR FOFM: gapfill for the missing years 2020-2023
# Use map dfr to create gapfill data frame
new_fofm <- map_dfr(2020:2023, function(y) {
  fofm_tonnes %>%
    filter(year == 2019) %>% 
    mutate(year = y, gapfilled = 1)
})

# Add gapfilled flag to the original
fofm_tonnes_gf <- fofm_tonnes %>%
  filter(year == 2019) %>% 
  mutate(gapfilled = 0)

# Combine original with new rows
fofm_tonnes <- bind_rows(fofm_tonnes_gf, new_fofm)

write_csv(fofm_tonnes, here(current_np_dir, "int", "fofm_tonnes.csv"))

```

## Dollars Per Tonne

```{r gapfill-value-data}
# All commodities
harvest_tonnes_usd <- read_csv(here(current_np_dir, "int", "np_harvest_usd_tonnes.csv")) %>% 
  filter(rgn_id != 213)


# Calculate $/tonne for each OHI region and product and gapfill based on UN geopolitical region ----
harvest_tonnes_usd_geo_rgn <- harvest_tonnes_usd %>%
  full_join(region_product, by = join_by("rgn_id", "product", "year")) %>%
  add_georegion_id() %>%
  dplyr::filter(year %in% years) %>%
  arrange(rgn_id)


# calculate georgn values (for each region, year, and product)
geo_rgn_values_df <- harvest_tonnes_usd_geo_rgn %>%
  group_by(georgn_id, year, product) %>%
  mutate(georgn_usd_per_tonne = usd / tonnes) %>%
  mutate(georgn_usd_per_tonne = ifelse(is.infinite(georgn_usd_per_tonne), NA, georgn_usd_per_tonne)) %>%
  summarise(georgn_usd_per_tonne = mean(georgn_usd_per_tonne, na.rm = TRUE)) %>% 
  ungroup()


# calculate global values (for each product across all regions for each year)
global_values_df <- harvest_tonnes_usd_geo_rgn %>%
  group_by(year, product) %>%
  mutate(global_usd_per_tonne = usd / tonnes) %>%
  mutate(global_usd_per_tonne = ifelse(is.infinite(global_usd_per_tonne), NA, global_usd_per_tonne)) %>%
  summarise(global_usd_per_tonne = mean(global_usd_per_tonne, na.rm = TRUE)) %>% 
  ungroup()


# Create final gapfilled data frame
harvest_tonnes_usd_values_gf <- harvest_tonnes_usd_geo_rgn %>%

  # Join the georegion and global dfs
  left_join(geo_rgn_values_df, by = join_by("georgn_id", "product", "year")) %>%
  left_join(global_values_df, by = join_by("product", "year")) %>%

  # Calculate the final usd/tonne
  mutate(usd_per_tonne = usd/tonnes) %>%

   # Assign Inf (n/0) and NaNs (0/0) as NA
  mutate(usd_per_tonne = ifelse(usd_per_tonne == "Inf", NA,
                                ifelse(usd_per_tonne == "NaN", NA, usd_per_tonne))) %>%

  # Gapfill so that when there is no regional value, take the geo regional value,
  mutate(usd_per_tonne = ifelse(is.na(usd_per_tonne) & georgn_usd_per_tonne != "NaN",
                                georgn_usd_per_tonne,
                                ifelse(is.na(usd_per_tonne) & georgn_usd_per_tonne == "NaN",
                                       global_usd_per_tonne, usd_per_tonne))) %>%
  # and if there is no geo regional value, take the global value; else, keep original values.

  # Note if gapfilling occurred in the previous step
  mutate(values_gf_description = ifelse(is.na(usd_per_tonne) & georgn_usd_per_tonne != "NaN",
                                        "geo_region",
                                        ifelse(is.na(usd_per_tonne) & georgn_usd_per_tonne == "NaN",
                                               "global", "none"))) %>%
  
  mutate(values_gf = ifelse(is.na(usd_per_tonne) & georgn_usd_per_tonne != "NaN", 
                            1,
                            ifelse(is.na(usd_per_tonne) & georgn_usd_per_tonne == "NaN", 
                                   1, 0))) 
  
nrow(harvest_tonnes_usd_values_gf) # 3300
colnames(harvest_tonnes_usd_values_gf)
colSums(is.na(harvest_tonnes_usd_values_gf))

# get USD per tonne averaged across georegion for each product and year
t_usd <- read_csv(here(sprintf("globalprep/np/v%s/int/np_usd_per_tonne.csv", current_year))) %>%
    add_georegion_id() %>%
    group_by(product, georgn_id, year) %>%
    mutate(avg_usd_per_tonne = mean(usd_per_tonne, na.rm=TRUE)) %>%
    select(rgn_name, rgn_id, product, year, georgn_id, usd, avg_usd_per_tonne) %>%
  ungroup()

```


### Tonnes harvest data
The data above only include the tonnes of product that is globally traded. However, much of the product may be consumed in the country. For that reason we use the sea around us catch data and mariculture data for fofm and seaweed.

```{r}
fofm <- read_csv(here(sprintf("globalprep/np/v%s/int/tonnes_fofm.csv", current_year))) %>%
  mutate(product = "fish_oil") %>%
  select(product, rgn_id, year, tonnes)
seaweed <- read_csv(here(sprintf("globalprep/np/v%s/int/np_relative_sust_tonnes_seaweeds_full.csv", current_year))) %>%
  mutate(tonnes = tonnes_sustainable*1/0.79) %>% #want total tonnes, not just sustainable tonnes
  select(product, rgn_id, year, tonnes)

tonnes <- rbind(fofm, seaweed) # for ornamentals the FAO trade data alone is used to estimate USD

rates <- left_join(t_usd, tonnes, by=c("rgn_id", "product", "year")) %>%
  filter(year >= 1990) %>%
  mutate(tonnes = ifelse(is.na(tonnes), 0, tonnes)) 

# Now calculate 5 year averages
weighting_usd_values_avg <- harvest_tonnes_usd_values_gf %>%
  dplyr::select(rgn_id, product, year, usd_per_tonne) %>%
  group_by(rgn_id, product) %>%
  summarise(mean_usd_per_tonne = mean(usd_per_tonne)) %>%
  ungroup()

total_usd <- rates %>%
  mutate(est_usd = avg_usd_per_tonne*tonnes) %>%
  mutate(usd_weight = ifelse(usd>est_usd, usd, est_usd)) %>%
  select(rgn_id, year, usd, product)
```

## Total USD Value

Combine individual commodities tonnes and the USD/Tonne data to find overall USD value. 

```{r combine-individual-and-value}
# Read in data from previous code chunk
orn_tonnes <- read_csv(here(current_np_dir, "int", 'orn_tonnes.csv'))
sw_tonnes <- read_csv(here(current_np_dir, "int", 'sw_tonnes.csv'))
fofm_tonnes <- read_csv(here(current_np_dir, "int", 'fofm_tonnes.csv')) %>% 
  select(!gapfilled)

# Read in final usd/tonnes
usd_tonnes_raw <- read_csv(here(current_np_dir, "int", "harvest_weighting_values.csv"))

# Filter to selected columns
usd_tonnes <- usd_tonnes_raw %>% 
  select(rgn_id, product, usd, year, usd_per_tonne, mean_usd_per_tonne, georgn_usd_per_tonne, global_usd_per_tonne)

# Bind the product dfs together
np_tonnes <- bind_rows(orn_tonnes, sw_tonnes, fofm_tonnes) 

dim(np_tonnes)
summary(np_tonnes)
colSums(is.na(np_tonnes))

# Join the usd/tonne data to the full NP dataframe
np_value <- np_tonnes %>% 
  left_join(usd_tonnes, by = join_by(year, product, rgn_id)) 

# Gapfilling
np_value_final <- np_value %>% 
  mutate(usd_product = case_when(
    product == "ornamentals" ~ georgn_usd_per_tonne * tonnes,
    product == "seaweeds" ~ mean_usd_per_tonne * tonnes,
    product == "fish_oil" ~ global_usd_per_tonne * tonnes,
    # is.na(georgn_usd_per_tonne) ~ global_usd_per_tonne * tonnes,
    .default = global_usd_per_tonne * tonnes
    )) %>% 
  # There are three regions with no georegion usd per tonne for ornamentals. Gapfill with global value
  mutate(usd_product = ifelse(is.na(usd_product), global_usd_per_tonne * tonnes, usd_product)) %>%
  # Group by rgn and year to find the overall dollar value
  group_by(rgn_id, year) %>%
  mutate(overall_usd_value = sum(usd_product)) %>%
  ungroup() %>%
  # calculate weight: proportion that each product contributes to that region's NP
  mutate(weight = usd_product/overall_usd_value) %>%
  # Some tonnes are 0 which leads to a total_usd of 0, which makes NAs
  mutate(weight = ifelse(is.na(weight), 0, weight)) %>% 
  # mutate(year = scen_year) %>%
  dplyr::select(rgn_id, year, product, tonnes, usd, mean_usd_per_tonne, georgn_usd_per_tonne, global_usd_per_tonne, usd_product, overall_usd_value, weight) %>%
  filter(rgn_id != 213) 


# Check
colSums(is.na(np_value_final)) # 0
nrow(np_value_final) # 3300

readr::write_csv(np_value_final, here(current_np_dir, "output", "np_product_weights.csv"))

```


## Datacheck

Note: we're not comparing data for a specific year, e.g., 2019, as we do in many data checks. The data processing that creates `np_product_weights.csv` fills the `year` column with the scenario year (e.g., 2024), which means that the previous year's `np_product_weights.csv` and the current year's weights cannot be compared by filtering to a specific year. Instead, we compare all weights for all regions and find the difference between this year's weights and last year's (called `diff`). We also plot the changes by having one year's weights as x-value and the other year's weights as y-values in a few scatterplots.


```{r fig.height=5, fig.width=8}

# Read in previous and current years' np_product_weights to compare
old_prod_weights <- read_csv(here(previous_np_dir, "output", "np_product_weights.csv"))
prod_weights <- read_csv(here(current_np_dir, "output", "np_product_weights.csv"))

check <- prod_weights %>%
  rename("new_weight" = "weight") %>%
  left_join(old_prod_weights, by = c("rgn_id", "product")) %>%
  mutate(diff = new_weight - weight) %>%
  left_join(rgns_eez)

# note -- these are across all years...

plot(check$new_weight, check$weight, main = "All Products")
abline(0,1, col = "red")

check_sw <- check %>%
  filter(product == "seaweeds")

plot(check_sw$new_weight, check_sw$weight, main = "Seaweeds")
abline(0,1, col = "red")

check_orn <- check %>%
  filter(product == "ornamentals") %>%
  filter(year %in% 2010:2023)

plot(check_orn$new_weight, check_orn$weight, main = "Ornamentals")
abline(0,1, col = "red")


check_fofm <- check %>%
  filter(product == "fish_oil")

plot(check_fofm$new_weight, check_fofm$weight, main = "Fish Oil / Fish Meal (FOFM)")
abline(0, 1, col = "red") ## these change because of new SAUP fisheries catch data... 

top_10_diffs <- check %>% arrange(desc(abs(diff))) %>% head(n = 10)
top_10_diffs %>% relocate(diff, .after = "rgn_id")

```


### New datacheck: interactive plot

```{r}
# ======== Setup ==========================================================
# (mostly copied from earlier setup with addition of library(plotly))
# Read in packages necessary for visualization (in case you cleared your environment)
library(ggplot2)
library(plotly)
library(dplyr)
library(tidyr)
library(here)

# Set scenario year, reproducible file paths
scen_year_number <- 2025 # update this!!
scenario_year <- as.character(scen_year_number)
v_scen_year <- paste0("v", scenario_year)

prev_year_number <- scen_year_number - 1
prev_2_year_num <- scen_year_number - 2
v_prev_year <- paste0("v", prev_year_number)
v_prev_2_year <- paste0("v", prev_2_year_num)

current_np_dir <- here::here("globalprep", "np", v_scen_year)
previous_np_dir <- here::here("globalprep", "np", paste0("v", scen_year_number - 1))

# Read in current, previous, and 2 years previous data
old_prod_weights <- read_csv(here(previous_np_dir, "output", "np_product_weights.csv")) 
# %>% 
#   select(!c(georgn_id, level))
prod_weights <- read_csv(here(current_np_dir, "output", "np_product_weights.csv"))  
# %>% 
#   select(!c(georgn_id, level))

colnames(prod_weights)
colnames(old_prod_weights)

old_prod_weights <- old_prod_weights %>% 
  rename(old_mean_usd = mean_usd_per_tonne, old_tonnes = tonnes, old_usd = usd, old_georgn = georgn_usd_per_tonne, old_global = global_usd_per_tonne, old_usd_per_tonne = usd_per_tonne, old_values_desc = values_gf_description, old_values_gf = values_gf)


check <- prod_weights %>%
  # rename("new_weight" = "weight") %>%
  left_join(old_prod_weights, by = c("rgn_id", "product", "year")) %>%
  # mutate(diff = new_weight - weight) %>%
  left_join(rgns_eez) %>% 
  select(!c(eez_iso3, territory, admin_rgn_id, admin_country_name, Notes))
  # filter(year == 2021) %>% 
  # select(rgn_id, rgn_name, year, product, new_weight, weight, diff, tonnes.x, tonnes.y, usd_product.x, usd_product.y, overall_usd_value.x, overall_usd_value.y) 


# ----- Read in NP weights data  -----
# Read in current, previous, and 2 years previous data
old_prod_weights <- read_csv(here("globalprep", "np", "v2024", "test", "np_product_weights.csv"))
prod_weights <- read_csv(here("globalprep", "np", "v2025", "output", "np_product_weights.csv"))

check <- prod_weights %>%
  rename("new_weight" = "weight") %>%
  left_join(old_prod_weights, by = c("rgn_id", "product", "year")) %>%
  mutate(diff = new_weight - weight) %>%
  left_join(rgns_eez) %>% 
  # filter(year == 2021) %>% 
  select(rgn_id, rgn_name, year, product, new_weight, weight, diff, tonnes.x, tonnes.y, usd_product.x, usd_product.y, overall_usd_value.x, overall_usd_value.y)

different <- check %>% 
  filter(diff != 0) %>% 
  mutate(diff_tonnes = tonnes.x - tonnes.y,
         diff_usd = usd_product.x - usd_product.y,
         diff_total = overall_usd_value.x - overall_usd_value.y)

# 1737 rows where there is no difference in weight
nrow(filter(check, diff == 0)) # 1737
nrow(filter(check, diff != 0)) # 903
nrow(check) # 3300
# 27% different...

## It looks like most of the difference is coming from a difference in USD. 

# ======== Plot new NP weights data vs. previous year's data ==================

# ---- Create ggplot object with `text` aes argument to prep for plotly -----
g_new <- ggplot(check, aes(x = weight, y = new_weight, color = product,
                             text = paste("<b>Region:</b>", rgn_name,
                                          "<br><b>Product:</b>", product,
                                          "<br><b>Year:</b>", year
                                          #"<br>Old weight:", older_weight,
                                          # "<br>New weight:", weight
                                          ))) +
  geom_point(alpha = 0.8) +
  geom_abline(intercept = 0, slope = 1, alpha = 0.7,
              color = "red", linetype = "solid") +
  scale_color_manual(values = c("fish_oil" = "#edae49",
                                "seaweeds" = "#386641",
                                "ornamentals" = "#83c5be")) +
  # labs(title = paste("2022 Data Year New Methodology", scenario_year, "vs", prev_year_number, "Natural Product Weights by Product"),
  #      # x = paste(prev_year_number, "Weights"),
  #      # y = paste(scenario_year, "Weights"),
  #      color = "Product") +
  theme_minimal()

# ---- Convert to plotly ------------------------------
p_new <- ggplotly(g_new, tooltip = "text") %>%
  layout(
    hovermode = "closest",
    xaxis = list(hoverformat = ".4f"), # customize number formatting 
    yaxis = list(hoverformat = ".4f") # (number of places after decimal) 
  )

# ------- Create custom hover template ----------------
p_new$x$data <- lapply(p_new$x$data, function(trace) {
  if (trace$mode == "markers") {
    trace$hovertemplate <- paste(
      "%{text}<br>",
      "<b>Old Weight</b>: %{x:.4f}<br>",
      "<b>New Weight</b>: %{y:.4f}<br>"
      )
  }
  return(trace)
})

# Display customized plot
p_new
=======
write_csv(fish, here("globalprep/np/v2025/output/np_fish_weights.csv"))
write_csv(sw, here("globalprep/np/v2025/output/np_seaweed_weights.csv"))
write_csv(orns, here("globalprep/np/v2025/output/np_ornamentals_weights.csv"))
>>>>>>> 57354bfcc8e3a05b26e2bd30c829e8bc50ab1646

```
