---
title: 'OHI `r format(Sys.time(), "%Y")`: Natural Products, Preparing catch and B/Bmsy data for forage fish species'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

[REFERENCE RMD FILE: <http://ohi-science.org/ohiprep_v2024/globalprep/np/v2024/STEP1c_np_fishfeed_prep.html>]

# Summary

This script takes the SAUP 2023 catch data, a list of fish oil/fish meal (FOFM) species, and B/Bmsy estimates to calculate a score for FOFM species per OHI region. The FIS data prep must be completed prior to this step.

## Updates from previous assessment

- New RAM data v4.66
- Used the old NP FOFM Rmd to fill out this document under the new framework. See this [google doc](https://docs.google.com/document/d/1ubCTW7ZrvvdckvY2zpBzCLHLgoIlCLW8kDMj-KWda9U/edit?tab=t.0) for reference. 
- Cut out steps 1-2 of the FOFM workflow. As the SAUP data has not been updated since 2019, we use the old data sources for the forage reference list and start directly at step 3. 

------------------------------------------------------------------------

# Data Source

## Sea Around us

**Reference**: Pauly D., Zeller D., Palomares M.L.D. (Editors), 2020. Sea Around Us Concepts, Design and Data (seaaroundus.org).

**Downloaded**: September 27, 2022

**Description**: Tons per year and SAUP region with information on sector type, industry type, fishing entitity, reporting status and taxonomic information.

**Time range**: 1950 - 2019

**Format**: CSV

**Additional Information**: [Methods](http://www.seaaroundus.org/sea-around-us-methods-index/)

## Avoiding the ecological limits of forage fish

**Reference**: Froehlich, H.E., Jacobsen, N.S., Essington, T.E., Clavelle, T., and Halpern, B.S. (2018). Avoiding the ecological limits of forage fish for fed aquaculture. Nature Sustainability 1, 298.

**Downloaded**: July 7, 2020. Obtained from Melanie Frazier (NCEAS).

**Description**: List of FOFM species from Watson v3 data.

**Native data resolution**:

**Format**: CSV format

## RAM Legacy Data

**Reference**: [RAM Legacy Stock Assessment Database](http://ramlegacy.org) v4.66

**Downloaded**: 06/26/2025

**Description**: B/Bmsy value by stock and year (other data, which we do not use, are also available in the database)

**Native data resolution**: stock (fish stock, species and region specific)

**Time range**: 1800 - 2023

**Format**: CSV format

**Additional Information**: We use the finalized b/bmsy layer from OHI-global for this data prep. We do not actually read in the raw RAM data here.

------------------------------------------------------------------------

# Methods

INCOMPLETE. See this folder's `README.md` for details. 


## Setup

```{r}
knitr::opts_chunk$set(eval = FALSE)

# ======= Load packages ============
if (!require(librarian)) {
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  ohicore, #devtools::install_github('ohi-science/ohicore@dev') #if relevant: restart session after reinstalling
  here,
  tidyverse,
  zoo,
  plotly,
  tictoc,
  RColorBrewer
)

# ======= Set directories ===========
# Update scenario year, set up programmatic scenario year updates
current_year <- 2025 # update this!!!
previous_year <- current_year - 1

# Directories
data_dir <- paste0("d", current_year)
prev_data_dir <- paste0("d", previous_year)

# Scenarios
scen_year <- paste0("v", current_year)
prev_scen_year <- paste0("v", previous_year)

data_years <- c("1950", "2023") # update to correct data years

# Natural Products directory on GitHub
current_np_dir <- here::here("globalprep", "np", scen_year) 
previous_np_dir <- here::here("globalprep", "np", paste0(prev_scen_year, "_new")) # update this if name changes

# Fisheries directory on GitHub
current_fis_dir <- here::here("globalprep", "fis", scen_year)
previous_fis_dir <- here::here("globalprep", "fis", prev_scen_year)

# ======== Source functions ===========
source(here::here("workflow", "R", "common.R")) # directory locations
# Load FAO-specific user-defined functions
source(here::here("workflow", "R", "fao_fxn.R")) # function for cleaning FAO files
# function for handling FAO commodity data specific to NP
source(here::here(current_np_dir, "R", "np_fxn.R")) 
# functions specific to mariculture dealing with compound countries
source(here("globalprep", "mar", scen_year, "mar_fxs.R"))
```

**v2025 updated steps**

This year's analysis cut steps 1 and 2 and started straight with step 3. We are using the forage fish catch stocks from the previous years analysis (as the data has not been updated since 2019).

1. Join with the final B/Bmsy layer used in the FIS model
2. Convert the B/Bmsy values to scores (this is done in functions.R)
3. Apply the underharvest penalty
4. Take a catch weighted average of B/Bmsy scores for each region/year.

*Old steps to this*

1.  Subset the forage fish catch stocks for each region/year (keep FAO and OHI rgn ids)
2.  Multiply by 0.70 to reflect the amount going to feed/oils
3.  Join with the final B/Bmsy layer used in the FIS model
4.  Convert the B/Bmsy values to scores (this is done in functions.R)
5.  Apply the underharvest penalty
6.  Take a catch weighted average of B/Bmsy scores for each region/year.


# Steps 3/4/5/6

## Join with the final B/Bmsy layer from the fis model, convert the B/Bmsy values to scores (this is done in functions.R for FIS subgoal, but we do it in the dataprep for NP), apply the underharvest penalty, and take a catch weighted average of B/Bmsy scores for each region/year.

```{r}
fis_bbmsy <- read_csv(here(current_fis_dir, "output", "fis_bbmsy.csv"))

fis_bbmsy_gf <- read_csv(here(current_fis_dir, "output", "fis_bbmsy_gf.csv"))

# This data comes from the previous year's analysis and has been copied over into the most recent scenario directory
catch_FOFM <- read_csv(here(current_np_dir, "int", "mean_catch_FOFM.csv"))

b <- fis_bbmsy %>%
  dplyr::mutate(bbmsy = ifelse(bbmsy > 1 , 1, bbmsy)) # no underharvest penalty anymore! 

c <- catch_FOFM %>%
    dplyr::mutate(stock_id_taxonkey = as.character(stock_id_taxonkey)) %>%
    dplyr::mutate(taxon_key = sub('.*_', '', stock_id_taxonkey)) %>%
    dplyr::mutate(stock_id = sub('_[^_]*$', '', stock_id_taxonkey)) %>%
    dplyr::mutate(catch = as.numeric(catch)) %>%
    dplyr::mutate(year = as.numeric(as.character(year))) %>%
    dplyr::mutate(region_id = as.numeric(as.character(rgn_id))) %>%
    dplyr::mutate(taxon_key = as.numeric(as.character(taxon_key))) %>%
    dplyr::select(region_id, year, stock_id, taxon_key, catch)


b <- b %>%
  dplyr::mutate(bbmsy = as.numeric(bbmsy),
                region_id = as.numeric(as.character(rgn_id)),
                year = as.numeric(as.character(year)),
                stock_id = as.character(stock_id))


####
# ==== Calculate scores for Bbmsy values ==== ####
#  *************NOTE *****************************
#  These values can be altered
#  ***********************************************

## The upper limit can be deleted, since we already cap pedscores at 1 anyways. We still want to keep the lower buffer.

alpha <- 0.5
beta <- 0.25
lowerBuffer <- 0.95
upperBuffer <- 1.05

b$score = ifelse(
  b$bbmsy < lowerBuffer,
  b$bbmsy,
  ifelse (b$bbmsy >= lowerBuffer &
            b$bbmsy <= upperBuffer, 1, NA)
)
b$score = ifelse(!is.na(b$score),
                 b$score,
                 ifelse(
                   1 - alpha * (b$bbmsy - upperBuffer) > beta,
                   1 - alpha * (b$bbmsy - upperBuffer),
                   beta
                 ))


####
# Merge the b/bmsy data with catch data ------------------------------
####
data_fis <- c %>%
  dplyr::left_join(b, by = c('region_id' = 'rgn_id', 'stock_id', 'year')) %>%
  dplyr::select(rgn_id = region_id, stock_id, year, taxon_key, catch, bbmsy, score)

###
#  Estimate scores for taxa without b/bmsy values --------------------
# Mean score of other fish in the region is the starting point
# Then a penalty is applied based on the level the taxa are reported at
###

## this takes the mean score within each region and year
data_fis_gf <- data_fis %>%
  dplyr::group_by(rgn_id, year) %>%
  dplyr::mutate(mean_score = mean(score, na.rm = TRUE)) %>%
  dplyr::ungroup()

## this takes the mean score across all regions within a year
# (when no stocks have scores within a region)
data_fis_gf <- data_fis_gf %>%
  dplyr::group_by(year) %>%
  dplyr::mutate(mean_score_global = mean(score, na.rm = TRUE)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(mean_score = ifelse(is.na(mean_score), mean_score_global, mean_score)) %>%
  dplyr::select(-mean_score_global)

#  *************NOTE *****************************
#  In some cases, it may make sense to alter the
#  penalty for not identifying fisheries catch data to
#  species level.
#  ***********************************************

penaltyTable <- data.frame(TaxonPenaltyCode = 1:6,
                           penalty = c(0.1, 0.25, 0.5, 0.8, 0.9, 1))

data_fis_gf <- data_fis_gf %>%
  dplyr::mutate(TaxonPenaltyCode = as.numeric(substring(taxon_key, 1, 1))) %>%
  dplyr::left_join(penaltyTable, by = 'TaxonPenaltyCode') %>%
  dplyr::mutate(score_gf = mean_score * penalty) %>%
  dplyr::mutate(method = ifelse(is.na(score), "Mean gapfilled", NA)) %>%
  dplyr::mutate(gapfilled = ifelse(is.na(score), 1, 0)) %>%
  dplyr::mutate(score = ifelse(is.na(score), score_gf, score))


gap_fill_data <- data_fis_gf %>%
  dplyr::select(rgn_id,
         stock_id,
         taxon_key,
         year,
         catch,
         score,
         gapfilled,
         method) 

readr::write_csv(gap_fill_data, here::here(current_np_dir, "output", "NP_bbmsy_summary_gf.csv"))

score_data <- data_fis_gf %>%
  dplyr::select(rgn_id, stock_id, year, catch, score)

 ###
# Calculate status for each region
###

## Take a catch weighted average of B/Bmsy scores for each region/year.

score_data <- score_data %>%
  dplyr::group_by(year, rgn_id) %>%
  dplyr::mutate(SumCatch = sum(catch)) %>%
  dplyr::ungroup()  

score_data <- score_data %>%
  dplyr::group_by(rgn_id, year) %>%
  dplyr::summarize(score = weighted.mean(score, catch)) %>%
  dplyr::ungroup() 

## v2025: add a product column to the final csv
score_data <- score_data %>% 
  mutate(product = "FOFM")

summary(score_data)

readr::write_csv(score_data, here(current_np_dir, "output", "np_fofm_scores.csv"))

```

### Testing and comparison

```{r}
# Read in the old data
old <- readr::read_csv(here::here("globalprep", "np", prev_scen_year, "output", "np_fofm_scores.csv"))
summary(old)

# Read in the new data
new <- readr::read_csv(here::here(current_np_dir, "output", "np_fofm_scores.csv"))
summary(new)

# load ohi regions
region_data()

# Investigating changes in data for same year in different data/scenario years

check <- new %>%
  rename("new_score" = "score") %>%
  left_join(old, by = c("rgn_id", "year")) %>%
  left_join(rgns_all, by = "rgn_id") %>%
  select(rgn_id, year, new_score, score, rgn_name) %>%
  mutate(diff = new_score - score) %>%
  filter(year == 2019)

ab_scores <- ggplot() + 
  geom_abline(slope = 1, intercept = 0, col = "red") +
  geom_point(data = check, aes(x = score, y = new_score, text = paste("Region:", rgn_name), alpha = 0.7)) +
  labs(x = "Old Score", 
       y = "New Score",
       title = "FOFM New score vs old score for 2019") +
  theme_bw();ggplotly(ab_scores)

```


