---
title: 'OHI `r format(Sys.time(), "%Y")`: Natural Products: weighting data prep'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
#  chunk_output_type: inline
  markdown: 
    wrap: 72
---

[REFERENCE RMD FILE:
<http://ohi-science.org/ohiprep_v2021/globalprep/np/v2021/STEP2_np_weighting_prep.html>]

# Summary

This analysis converts FAO commodities tonnes data for ornamental fish
trade, SAUP fisheries tonnes data for FOFM, and FAO mariculture tonnes
data for seaweeds into a weighting scheme per each ohi region, based off
of 5 year averages of tonnes and values (in usd) for each product. This
weighting scheme will be applied to the scores to determine how much of
each natural products layer affects the overall NP score when updating
in ohi global.

## Updates from previous year
- Complete the change in NP framework started in 2024. 

# Data Source

## FAO: Commodities

**Reference**:\
<https://www.fao.org/fishery/en/statistics/software/fishstatj> App release date: March 2024 FAO raw commodities quantity 1976_2023 FAO raw commodities value 1976_2023 [FAO metadata](http://www.fao.org/fishery/statistics/global-commodities-production/en)

- Global aquatic trade - All partners aggregated - Quantities and Values - 1976-2022 (Release date: July 2024)


**Downloaded**: 2025-07-14

**Description**: Quantity (tonnes) and value (USD) of raw commodities (Exports only) for each country, taxa, year. The FAO data is subset to include commodities in these categories: ornamental fish, fish oil, seaweed and plants (see: raw/commodities2products_weighting.csv for details).

**Time range**: 1976-2023

Files can be found on Mazu: `home/shares/ohi/git-annex/globalprep/_raw_data/FAO_commodities/d2025` accompanied by README.md with detailed download instructions. 

## Seafood Watch sustainability data

Obtained upon request of SFW.

**Reference**:\
[Seafood Watch Complete Recommendation documentation](https://www.seafoodwatch.org/globalassets/sfw/pdf/whats-new/seafood-watch-complete-recommendation-list.pdf)\
[Seafood Watch Scoring Theory](https://www.seafoodwatch.org/globalassets/sfw/pdf/standards/aquaculture/seafood-watch-aquaculture-standard-version-a4.pdf)\
[Seafood Watch Global Seaweed Assessment](https://www.seafoodwatch.org/recommendation/seaweed/seaweed-31568) (provides gapfilling constants)

**Downloaded**: June 27, 2025

**Description**: Monterrey Bay Aquarium Seafood Watch aquaculture recommendations. Sustainability scored from 0-10. Rescaled to 0-1. There is only one value for seaweeds in the data... 0.67


## Sea Around Us Concepts

**Reference**: Pauly D., Zeller D., Palomares M.L.D. (Editors), 2020.
Sea Around Us Concepts, Design and Data (seaaroundus.org).

**Downloaded**: September 27, 2022

**Description**: Tons per year and SAUP region with information on
sector type, industry type, fishing entitity, reporting status and
taxonomic information.

**Time range**: 1950 - 2019

**Format**: CSV

**Additional Information**:
[Methods](http://www.seaaroundus.org/sea-around-us-methods-index/)

## Avoiding the ecological limits of forage fish for fed aquaculture

**Reference**: Froehlich, H.E., Jacobsen, N.S., Essington, T.E.,
Clavelle, T., and Halpern, B.S. (2018). Avoiding the ecological limits
of forage fish for fed aquaculture. Nature Sustainability 1, 298.

**Downloaded**: July 7, 2020. Obtained from Melanie Frazier (NCEAS).

**Description**: List of FOFM species from Watson v3 data.

**Native data resolution**:

**Format**: CSV format

## RAM Legacy Data

**Reference**: [RAM Legacy Stock Assessment Database](http://ramlegacy.org) v4.66

**Downloaded**: 06/26/2025

**Description**: B/Bmsy value by stock and year (other data, which we do not use, are also available in the database)

**Native data resolution**: stock (fish stock, species and region specific)

**Time range**: 1800 - 2023

**Format**: CSV format

**Additional Information**: We use the finalized b/bmsy layer from OHI-global for this data prep. We do not actually read in the raw RAM data here.

## FAO: Production

**Reference**:\
<https://www.fao.org/fishery/statistics-query/en/aquaculture/aquaculture_quantity> FAO Global Aquaculture Production Quantity 1950_2022 FAO metadata found [here](http://www.fao.org/fishery/statistics/global-aquaculture-production/en)

**Downloaded**: July 26th, 2024

**Last updated**: March 29th, 2024

**Description**: Quantity (tonnes) of mariculture for each country, species, year.

**Time range**: 1950-2022


------------------------------------------------------------------------

# Methods

1. Read in the intermediate files with absolute tonnes of production (raw data)
2. Read in intermediate data B (USD/tonnes)
3. Merge data B and tonnes
4. Calculate total USD for each product
5. Do 5 year averages (or similar)
6. Calculate yearly weights for each product (products should sum to one for a region) 

The difference in years between data is a challenge. You can deal with this in a variety of ways, but I would probably repeat the FOFM 2019 data to match the years you have for seaweed and ornamentals. (and we will duplicate this in the scenario_data_years.csv). 

## Setup

```{r setup}

knitr::opts_chunk$set(eval = FALSE)

# ======= Load packages ============
if (!require(librarian)) {
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  ohicore, #devtools::install_github('ohi-science/ohicore@dev') #if relevant: restart session after reinstalling
  here,
  purr,
  tidyverse,
  zoo,
  plotly,
  tictoc,
  RColorBrewer
)

# ======= Set directories ===========
# Update scenario year, set up programmatic scenario year updates
current_year <- 2025 # update this!!!
previous_year <- current_year - 1

# Directories
data_dir <- paste0("d", current_year)
prev_data_dir <- paste0("d", previous_year)

# Scenarios
scen_year <- paste0("v", current_year)
prev_scen_year <- paste0("v", previous_year)

data_years <- c("1950", "2023") # update to correct data years

# Natural Products directory on GitHub
current_np_dir <- here::here("globalprep", "np", scen_year) 
previous_np_dir <- here::here("globalprep", "np", prev_scen_year) 

# Mariculture directory on GitHub
current_mar_dir <- here::here("globalprep", "mar", scen_year)
previous_mar_dir <- here::here("globalprep", "mar", prev_scen_year)

# ======== Source functions ===========
source(here::here("workflow", "R", "common.R")) # directory locations
# Load FAO-specific user-defined functions
source(here::here("workflow", "R", "fao_fxn.R")) # function for cleaning FAO files
# function for handling FAO commodity data specific to NP
source(here::here(current_np_dir, "R", "np_fxn.R")) 
# functions specific to mariculture dealing with compound countries
source(here("globalprep", "mar", scen_year, "mar_fxs.R"))

options(scipen = 999)
```

## Tonnes per Product

Create the overall region product data frame that contains the 5 years of interest and the three products: ornamentals, seaweeds, and fish oil fish meal.

> NOTE: future assessments will need to change the **year range for the FOFM** SAUP data because it peaks at 2019. This should still work for ornamentals and seaweeds, but a new methodology will need to be created for FOFM

```{r region-product}
# All commodities is only used here to get the max year. It gets used later on for the ornamentals tonnes
harvest_tonnes_usd <- read_csv(here(current_np_dir, "int", "np_harvest_usd_tonnes.csv"))

# Make a data frame with every region and every product ----
# Define years
max_year <- max(harvest_tonnes_usd$year)
max_year # 2023
min_year <- max_year - 4 
years <- c(min_year:max_year) 
years_df <- data.frame(year = min_year:max_year) # so that we can get 5 year average

# Load OHI region data from common.R
region_data() # gets rgns_eez

# Define subset of OHI EEZ regions
rgns_eez <- rgns_eez %>%
  dplyr::select(rgn_id) %>%
  filter(rgn_id != 213) # filter out Antarctica

# Define products dataframe
products <- data.frame(product = c("seaweeds", "ornamentals", "fish_oil"))

# Create full rgn x product x years dataframe
region_product <- rgns_eez %>% 
  # select rgn_id from rgns_eez
  dplyr::select(rgn_id) %>%  
  # cross join with products (match every row in x with every row in y)
  cross_join(products) %>% 
  # cross join with years
  cross_join(years_df) 
```

Now that the skeleton of region_product has been defined, we will fill in the individual commodity tonnes.

```{r individual-prep}
# Subset ornamentals product
orn_fill_df <- region_product %>% 
  filter(product == "ornamentals")

orn_tonnes_raw <- read_csv(here(current_np_dir, "int", "np_relative_sust_tonnes_ornamentals_full.csv"))

setdiff(orn_tonnes_raw$tonnes, harvest_tonnes_usd$tonnes) # NA (not sure what that means)
nrow(orn_tonnes_raw) # 2560
nrow(harvest_tonnes_usd) # 10860

# Use the original FAO commodity data to get ornamentals quantities
orn_tonnes <- harvest_tonnes_usd %>%
  # filter to 5 year range defined in the beginning of the first methods chunk
  dplyr::filter(year %in% years, 
                product == "ornamentals") %>% 
  # full join to fill in the rest of the regions with 0
  full_join(orn_fill_df, by = c("rgn_id", "year", "product")) %>% 
  # gapfill the NAs to be 0
  mutate(tonnes = ifelse(is.na(tonnes), 0, tonnes),
         usd = ifelse(is.na(usd), 0, usd)) %>% 
  dplyr::select(rgn_id, year, product, tonnes)

write_csv(orn_tonnes, here(current_np_dir, "int", "orn_tonnes.csv"))

# Subset seaweed products
sw_fill_df <- region_product %>%
  filter(product == "seaweeds")

# Read in the seaweed MAR data for tonnes
# Not sure if this is the right file???
sw_tonnes_raw_old <- read_csv(here(current_np_dir, "int", "np_seaweeds_tonnes_weighting.csv"))
sw_tonnes_raw <- read_csv(here(current_np_dir, "int", "np_relative_sust_tonnes_seaweeds_full.csv")) 
colnames(sw_tonnes_raw)

setdiff(sw_tonnes_raw$tonnes, sw_tonnes_raw_old$tonnes) # 439 different...
nrow(sw_tonnes_raw) # 843
nrow(sw_tonnes_raw_old) # 3202

sw_tonnes <- sw_tonnes_raw %>%
  mutate(product = "seaweeds") %>%
  group_by(rgn_id, year, product) %>% 
  summarise(tonnes = sum(tonnes, na.rm = TRUE)) %>% # sum across all species of seaweed
  dplyr::filter(year %in% years) %>% # filter to 5 year range 
  full_join(sw_fill_df, by = c("rgn_id", "year", "product")) %>%
    mutate(tonnes = ifelse(is.na(tonnes), 0, tonnes)) %>% # gapfill the NAs to be 0
  dplyr::select(rgn_id, year, product, tonnes)

write_csv(sw_tonnes, here(current_np_dir, "int", "sw_tonnes.csv"))

# Subset fofm products
fofm_fill_df <- region_product %>% 
  filter(product == "fish_oil")

## Read in FOFM tonnes data
# I think the FOFM data is correct here because I didn't change this export (only the data inside the export)
test <- read_csv(here(current_np_dir, "int", "raw_mean_catch_FOFM.csv"))
colnames(test)
view(test)
fofm_tonnes <- read_csv(here(current_np_dir, "int", "raw_mean_catch_FOFM.csv")) %>%
  dplyr::filter(year %in% years) %>% 
  group_by(rgn_id, year, product) %>%
  # Note: multiply by 0.3 to account for water loss when converting fish to fish oil.. 
  ## about 30% of fish are water and 70% are fish oil
  summarise(tonnes = sum(catch_non_human) * 0.3) %>% 
  ungroup() %>%
  full_join(fofm_fill_df, by = c("rgn_id", "year", "product")) %>%
    mutate(tonnes = ifelse(is.na(tonnes), 0, tonnes)) %>% # gapfill the NAs to be 0
  dplyr::select(rgn_id, year, product, tonnes)


# EXTRA STEP FOR FOFM: gapfill for the missing years 2020-2023
# Use map dfr to create gapfill data frame
new_fofm <- map_dfr(2020:2023, function(y) {
  fofm_tonnes %>%
    filter(year == 2019) %>% 
    mutate(year = y, gapfilled = 1)
})

# Add gapfilled flag to the original
fofm_tonnes_gf <- fofm_tonnes %>%
  filter(year == 2019) %>% 
  mutate(gapfilled = 0)

# Combine original with new rows
fofm_tonnes <- bind_rows(fofm_tonnes_gf, new_fofm) 

write_csv(fofm_tonnes, here(current_np_dir, "int", "fofm_tonnes.csv"))

```

## Dollars Per Tonne

```{r gapfill-value-data}
# All commodities
harvest_tonnes_usd <- read_csv(here(current_np_dir, "int", "np_harvest_usd_tonnes.csv")) %>% 
  filter(rgn_id != 213)


# Calculate $/tonne for each OHI region and product and gapfill based on UN geopolitical region ----
harvest_tonnes_usd_geo_rgn <- harvest_tonnes_usd %>%
  full_join(region_product, by = join_by("rgn_id", "product", "year")) %>%
  add_georegion_id() %>%
  dplyr::filter(year %in% years) %>%
  arrange(rgn_id)


# calculate georgn values (for each region, year, and product)
geo_rgn_values_df <- harvest_tonnes_usd_geo_rgn %>%
  group_by(georgn_id, year, product) %>%
  mutate(georgn_usd_per_tonne = usd / tonnes) %>%
  mutate(georgn_usd_per_tonne = ifelse(is.infinite(georgn_usd_per_tonne), NA, georgn_usd_per_tonne)) %>%
  summarise(georgn_usd_per_tonne = mean(georgn_usd_per_tonne, na.rm = TRUE)) %>% 
  ungroup()


# calculate global values (for each product across all regions for each year)
global_values_df <- harvest_tonnes_usd_geo_rgn %>%
  group_by(year, product) %>%
  mutate(global_usd_per_tonne = usd / tonnes) %>%
  mutate(global_usd_per_tonne = ifelse(is.infinite(global_usd_per_tonne), NA, global_usd_per_tonne)) %>%
  summarise(global_usd_per_tonne = mean(global_usd_per_tonne, na.rm = TRUE)) %>% 
  ungroup()


# Create final gapfilled data frame
harvest_tonnes_usd_values_gf <- harvest_tonnes_usd_geo_rgn %>%

  # Join the georegion and global dfs
  left_join(geo_rgn_values_df, by = join_by("georgn_id", "product", "year")) %>%
  left_join(global_values_df, by = join_by("product", "year")) %>%

  # Calculate the final usd/tonne
  mutate(usd_per_tonne = usd/tonnes) %>%

   # Assign Inf (n/0) and NaNs (0/0) as NA
  mutate(usd_per_tonne = ifelse(usd_per_tonne == "Inf", NA,
                                ifelse(usd_per_tonne == "NaN", NA, usd_per_tonne))) %>%

  # Gapfill so that when there is no regional value, take the geo regional value,
  mutate(usd_per_tonne = ifelse(is.na(usd_per_tonne) & georgn_usd_per_tonne != "NaN",
                                georgn_usd_per_tonne,
                                ifelse(is.na(usd_per_tonne) & georgn_usd_per_tonne == "NaN",
                                       global_usd_per_tonne, usd_per_tonne))) %>%
  # and if there is no geo regional value, take the global value; else, keep original values.

  # Note if gapfilling occurred in the previous step
  mutate(values_gf_description = ifelse(is.na(usd_per_tonne) & georgn_usd_per_tonne != "NaN",
                                        "geo_region",
                                        ifelse(is.na(usd_per_tonne) & georgn_usd_per_tonne == "NaN",
                                               "global", "none"))) %>%
  
  mutate(values_gf = ifelse(is.na(usd_per_tonne) & georgn_usd_per_tonne != "NaN", 
                            1,
                            ifelse(is.na(usd_per_tonne) & georgn_usd_per_tonne == "NaN", 
                                   1, 0))) 
  
view(harvest_tonnes_usd_values_gf)
nrow(harvest_tonnes_usd_values_gf) # 3300
colnames(harvest_tonnes_usd_values_gf)
colSums(is.na(harvest_tonnes_usd_values_gf))
```

```{r save-weighting-data}
# Save this data frame to int
readr::write_csv(harvest_tonnes_usd_values_gf,
                 here(current_np_dir, "int", "harvest_tonnes_usd_weighting_gf.csv"))

# v2025: Save a version without the 5 year averages
weighting_usd_values_raw <- harvest_tonnes_usd_values_gf %>%
  dplyr::select(rgn_id, product, year, usd_per_tonne)

nrow(weighting_usd_values_raw) # 3300

readr::write_csv(weighting_usd_values_raw,
                 here(current_np_dir, "int", "harvest_weighting_values_raw.csv"))

# OLD: Now calculate 5 year averages
# weighting_usd_values_final <- harvest_tonnes_usd_values_gf %>%
#   dplyr::select(rgn_id, product, year, usd_per_tonne) %>%
#   group_by(rgn_id, product) %>%
#   summarise(mean_usd_per_tonne = mean(usd_per_tonne)) %>%
#   ungroup()
# 
# #save this df
# readr::write_csv(weighting_usd_values_final,
#                  here(current_np_dir, "int", "harvest_weighting_values.csv"))

```

## Total USD Value

Combine individual commodities tonnes and the USD/Tonne data to find overall USD value. 

```{r combine-individual-and-value}
# Read in data from previous code chunk
orn_tonnes <- read_csv(here(current_np_dir, "int", 'orn_tonnes.csv'))
sw_tonnes <- read_csv(here(current_np_dir, "int", 'sw_tonnes.csv'))
fofm_tonnes <- read_csv(here(current_np_dir, "int", 'fofm_tonnes.csv')) %>% 
  select(!gapfilled)

view(orn_tonnes)
view(sw_tonnes)
view(fofm_tonnes)

# Read in final usd/tonnes
usd_tonnes <- read_csv(here(current_np_dir, "int", "harvest_weighting_values_raw.csv"))
# usd_tonnes_raw <- read_csv(here(current_np_dir, "int", "np_harvest_usd_tonnes.csv"))

# usd_tonnes <- usd_tonnes_raw %>% 
#   mutate(usd_per_tonne = usd / tonnes) %>% 
#   rename(tonnes_commod = tonnes)

colnames(orn_tonnes)
colnames(sw_tonnes)
colnames(fofm_tonnes)
colnames(usd_tonnes)

# Bind the product dfs together
np_tonnes <- bind_rows(orn_tonnes, sw_tonnes, fofm_tonnes) 

np_tonnes %>% 
  filter(rgn_id == 80) %>% 
  view()

dim(np_tonnes)
summary(np_tonnes)
colSums(is.na(np_tonnes))

# Join the usd/tonne data to the full NP dataframe
np_value <- np_tonnes %>% 
  left_join(usd_tonnes, by = join_by(year, product, rgn_id)) %>% 
  mutate(total_usd_value = usd_per_tonne * tonnes)

# Check
colSums(is.na(np_value)) # 0
nrow(np_value) # 3300
view(np_value)

write_csv(np_value, here(current_np_dir, "int", "all_commodities_value.csv"))

```


## 5 Year Averages

Import prepped FAO commodities data set and find 5 year average \$USD / Tonne average value per product per region. Gapfill these based on the UN geopolitical region/global values.

```{r averages}
# Read in the commodities value data
np_value <- read_csv(here(current_np_dir, "int", "all_commodities_value.csv"))

colnames(np_value)
view(np_value)

# Calculate 5 year averages
# This gives an averages 
np_averages <- np_value %>% 
  group_by(rgn_id, product) %>% 
  summarize(mean_tonnes = mean(tonnes),
            mean_usd_per_tonne = mean(usd_per_tonne))

np_comb <- np_value %>% 
  left_join(np_averages, by = join_by(rgn_id, product)) %>% 
  view()

view(np_averages)
colnames(np_comb)
```

## Weights per Product

Now we will calculate the weights per each product. To do this we need to multiply our average \$ value for each product \* tonnes of each product, and then divide by the total per each region. We will also assign year = (YYYY, e.g., 2022) so that this can be read into OHI-global (YYYY corresponds to the year that these weights were calculated).

```{r weights}
colnames(np_comb)
prod_weights <- np_comb %>% 
  mutate(usd_product = tonnes * mean_usd_per_tonne) %>% 
  group_by(rgn_id, year) %>%
  mutate(total_usd = sum(usd_product)) %>%
  ungroup() %>%
  # calculate weight: proportion that each product contributes to that region's NP
  mutate(weight = usd_product/total_usd) %>%
  # Some tonnes are 0 which leads to a total_usd of 0, which makes NAs
  mutate(weight = ifelse(is.na(weight), 0, weight)) %>% 
  # dplyr::select(-3, -4, -5, -6) %>%
  # mutate(year = scen_year) %>%
  # dplyr::select(rgn_id, year, product, weight) %>%
  filter(rgn_id != 213)

nrow(prod_weights)
view(prod_weights)
colSums(is.na(prod_weights))

prod_weights %>% 
  filter(rgn_id == 80) %>% 
  view()

colSums(is.na(np_comb))


readr::write_csv(prod_weights, here(current_np_dir, "output", "np_product_weights.csv"))

```

## Datacheck

Note: we're not comparing data for a specific year, e.g., 2019, as we do in many data checks. The data processing that creates `np_product_weights.csv` fills the `year` column with the scenario year (e.g., 2024), which means that the previous year's `np_product_weights.csv` and the current year's weights cannot be compared by filtering to a specific year. Instead, we compare all weights for all regions and find the difference between this year's weights and last year's (called `diff`). We also plot the changes by having one year's weights as x-value and the other year's weights as y-values in a few scatterplots.


```{r fig.height=5, fig.width=8}

# Read in previous and current years' np_product_weights to compare
old_prod_weights <- read_csv(here(previous_np_dir, "output", "np_product_weights.csv"))
prod_weights <- read_csv(here(current_np_dir, "output", "np_product_weights.csv"))

check <- prod_weights %>%
  rename("new_weight" = "weight") %>%
  left_join(old_prod_weights, by = c("rgn_id", "product")) %>%
  mutate(diff = new_weight - weight) %>%
  left_join(rgns_eez)

# note -- these are across all years...

plot(check$new_weight, check$weight, main = "All Products")
abline(0,1, col = "red")

check_sw <- check %>%
  filter(product == "seaweeds")

plot(check_sw$new_weight, check_sw$weight, main = "Seaweeds")
abline(0,1, col = "red")

check_orn <- check %>%
  filter(product == "ornamentals") %>%
  mutate(difference = new_weight - weight)

plot(check_orn$new_weight, check_orn$weight, main = "Ornamentals")
abline(0,1, col = "red")


check_fofm <- check %>%
  filter(product == "fish_oil")

plot(check_fofm$new_weight, check_fofm$weight, main = "Fish Oil / Fish Meal (FOFM)")
abline(0, 1, col = "red") ## these change because of new SAUP fisheries catch data... 

top_10_diffs <- check %>% arrange(desc(abs(diff))) %>% head(n = 10)
top_10_diffs %>% relocate(diff, .after = "rgn_id")

```


### New datacheck: interactive plot

```{r}
# ======== Setup ==========================================================
# (mostly copied from earlier setup with addition of library(plotly))
# Read in packages necessary for visualization (in case you cleared your environment)
library(ggplot2)
library(plotly)
library(dplyr)
library(tidyr)
library(here)

# Set scenario year, reproducible file paths
scen_year_number <- 2025 # update this!!
scenario_year <- as.character(scen_year_number)
v_scen_year <- paste0("v", scenario_year)

prev_year_number <- scen_year_number - 1
prev_2_year_num <- scen_year_number - 2
v_prev_year <- paste0("v", prev_year_number)
v_prev_2_year <- paste0("v", prev_2_year_num)

current_np_dir <- here::here("globalprep", "np", v_scen_year)
previous_np_dir <- here::here("globalprep", "np", paste0("v", scen_year_number - 1))

# ----- Read in NP weights data  -----
# Read in current, previous, and 2 years previous data
old_prod_weights <- read_csv(here(previous_np_dir, "output", "np_product_weights.csv"))
prod_weights <- read_csv(here(current_np_dir, "output", "np_product_weights.csv"))

check <- prod_weights %>%
  rename("new_weight" = "weight") %>%
  left_join(old_prod_weights, by = c("rgn_id", "product")) %>%
  mutate(diff = new_weight - weight) %>%
  left_join(rgns_eez)


# ======== Plot new NP weights data vs. previous year's data ==================


# ---- Create ggplot object with `text` aes argument to prep for plotly -----
g_new <- ggplot(check, aes(x = weight, y = new_weight, color = product,
                             text = paste("<b>Region:</b>", rgn_name,
                                          "<br><b>Product:</b>", product #,
                                          #"<br>Old weight:", older_weight,
                                          # "<br>New weight:", weight
                                          ))) +
  geom_point(alpha = 0.8) +
  geom_abline(intercept = 0, slope = 1, alpha = 0.7,
              color = "red", linetype = "solid") +
  scale_color_manual(values = c("fish_oil" = "#edae49",
                                "seaweeds" = "#386641",
                                "ornamentals" = "#83c5be")) +
  labs(title = paste(scenario_year, "vs", prev_year_number, "Natural Product Weights by Product"),
       x = paste(prev_year_number, "Weights"),
       y = paste(scenario_year, "Weights"),
       color = "Product") +
  theme_minimal()

# ---- Convert to plotly ------------------------------
p_new <- ggplotly(g_new, tooltip = "text") %>%
  layout(
    hovermode = "closest",
    xaxis = list(hoverformat = ".4f"), # customize number formatting 
    yaxis = list(hoverformat = ".4f") # (number of places after decimal) 
  )

# ------- Create custom hover template ----------------
p_new$x$data <- lapply(p_new$x$data, function(trace) {
  if (trace$mode == "markers") {
    trace$hovertemplate <- paste(
      "%{text}<br>",
      "<b>Old Weight</b>: %{x:.4f}<br>",
      "<b>New Weight</b>: %{y:.4f}<br>"
      )
  }
  return(trace)
})

# Display customized plot
p_new

```
