---
title: 'OHI `r format(Sys.time(), "%Y")`: Natural Products: Seaweeds Data Prep'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: false
    toc_float: yes
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---



[REFERENCE RMD FILE: <http://ohi-science.org/ohiprep_v2022/globalprep/np/v2021/STEP1b_np_seaweeds_prep.Rmd>]

# Summary

This analysis converts FAO mariculture data into one of the data layers used to calculate OHI `r format(Sys.time(), "%Y")` global natural products (NP) scores. We will conduct the overall NP data prep on seaweeds, fish oil/fish meal (FOFM), and ornamentals, however, our final layer from this data prep will only consist of seaweeds.

# Updates from previous assessment

-   New year of FAO mariculture data (1950-2023).
-   Explore idiosyncrasies around relative sustainable tonnes of seafood harvest and add context for the strange data check. 
-   Fix the way rolling mean is calculated (previously was only considering 4 total years of data when we want 5).


------------------------------------------------------------------------

# Data Source

## Production data

**Reference**:\
<https://www.fao.org/fishery/en/collection/aquaculture?lang=en> Release date: March 2023 FAO Global aquaculture production Quantity (1950 - 2023) Global aquaculture production (Value) (1950-2023)

**Downloaded**: June 23rd, 2025

**Instructions on downloading the FAO aquaculture data:**

**2025 download from Melanie**: 
Navigate to this link: <https://www.fao.org/fishery/statistics-query/en/global_production/global_production_quantity>

Click Download full dataset. Also downloaded the file below!

Navigate to this link: <https://www.fao.org/fishery/statistics-query/en/home>.

Click browse on Global aquaculture production. Click browse on Global aquaculture production (Quantity). Scroll to the bottom of the dimensions section, and de-select all of the pre-selected years. Under the query definition section, drag Country Name En (if not already present), ASFIS species Name En, FAO major fishing area Name En, ASFIS species Family scientific name, and Environment Name En, from the possible fields column into the selected rows column. Click on show data and confirm that data is present for 1950- two years prior to current year. Click download and select yes to include Null Values. Rename the file in the format FAO_GlobalAquacultureProduction_Quantity_1950_last_data_year. Upload the file to the appropriate year folder in the FAO_mariculture folder on mazu.

**Description**: Quantity (tonnes) of mariculture for each country, species, year.

**Time range**: 1950-2023

## Seafood Watch sustainability data

Obtained upon request of SFW.

**Reference**:\
[Seafood Watch Complete Recommendation documentation](https://www.seafoodwatch.org/globalassets/sfw/pdf/whats-new/seafood-watch-complete-recommendation-list.pdf)\
[Seafood Watch Scoring Theory](https://www.seafoodwatch.org/globalassets/sfw/pdf/standards/aquaculture/seafood-watch-aquaculture-standard-version-a4.pdf)\
[Seafood Watch Global Seaweed Assessment](https://www.seafoodwatch.org/recommendation/seaweed/seaweed-31568) (provides gapfilling constants)

**Downloaded**: June 27, 2025

**Description**: Monterrey Bay Aquarium Seafood Watch aquaculture recommendations. Sustainability scored from 0-10. Rescaled to 0-1. There is only one value for seaweeds in the data... 0.67

------------------------------------------------------------------------

# Methods



## Setup

```{r}
knitr::opts_chunk$set(eval = FALSE)

# ======= Load packages ============
if (!require(librarian)) {
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  ohicore, #devtools::install_github('ohi-science/ohicore@dev') #if relevant: restart session after reinstalling
  here,
  tidyverse,
  zoo,
  ggplot2,
  plotly,
  tictoc,
  RColorBrewer
)

# ======= Set directories ===========
# Update scenario year, set up programmatic scenario year updates
scen_year_number <- 2025 # update this!!
scen_year <- as.character(scen_year_number)
prev_scen_year <- as.character(scen_year_number - 1)

# Directory years
data_dir_year <- paste0("d", scen_year)
prev_data_dir_year <- paste0("d", prev_scen_year)

# Version years
v_scen_year <- paste0("v", scen_year)
prev_v_scen_year <- paste0("v", prev_scen_year)

data_years <- c("1950", "2023") # update to correct data years

# Natural Products directory on GitHub
current_np_dir <- here::here("globalprep", "np", v_scen_year) 
previous_np_dir <- here::here("globalprep", "np", paste0(prev_v_scen_year, "_new")) # update this if name changes


# Mariculture directory on GitHub
current_mar_dir <- here::here("globalprep", "mar", v_scen_year)
rgn_labels <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohi-global/refs/heads/draft/eez/layers/rgn_labels.csv")

# ======== Source functions ===========
source(here::here("workflow", "R", "common.R")) # directory locations
# Load FAO-specific user-defined functions
source(here::here("workflow", "R", "fao_fxn.R")) # function for cleaning FAO files
# function for handling FAO commodity data specific to NP
source(here::here(current_np_dir, "R", "np_fxn.R")) 
# functions specific to mariculture dealing with compound countries
source(here("globalprep", "mar", v_scen_year, "mar_fxs.R"))

```



### Import Raw Data: FAO Mariculture data

Mariculture production in tonnes. Ultimately used to find tonnes of non-food seaweed.

```{r}
# Read in mariculture data
mar_raw <- readr::read_csv(here(
  dir_M, "git-annex", "globalprep", "_raw_data", "FAO_mariculture", data_dir_year,
  paste0("FAO_GlobalAquacultureProduction_Quantity_1950_",
         data_years[2], ".csv")))

head(mar_raw) 
length(table(mar_raw$`Country Name En`))

```


# Data Wrangling

## Tidy mariculture data

Filter freshwater mariculture, make long format, and clean FAO codes.

```{r}
# Preliminary cleaning
mar <- mar_raw %>%
  dplyr::select(-`Unit Name`) %>% # should all be "Tonnes - live weight"
  rename(country = `Country Name En`,
         FAO_name = `ASFIS species Name En`, 
         fao = `FAO major fishing area Name En`, 
         environment = `Environment Name En`,
         family_scientific = `Family Scientific name`) %>%
  rename_at(vars(matches("\\[")), ~ str_remove(., "\\[")) %>%
  rename_at(vars(matches("\\]")), ~ str_remove(., "\\]"))

# check environments
table(mar$environment)  

# Include only marine and brackishwater environments
mar <- mar %>%
  filter(environment %in% c("Brackishwater", "Marine"))  

# Change to latest *DATA* year
latest_year <- data_years[2]


#As of 2023 we no longer need to run the fao_clean_data_new function
# NAs are no longer saved as ... and the flags are in a separate column
#however we should still replace columns where the flag is N w/.1
sub_N = 0.1

# add an index
mar <- mar %>% mutate(row_id = row_number())

# note: the following code is very similar to the function in 
# ohiprep_v2024/workflow/R/fao_online_portal_clean.R
# and can be updated to just source the function. 
## more notes on this at the end of the chunk

# pivot all of the year/value columns 
mar_values <- mar %>% 
  dplyr::select(-c(paste("1950":latest_year, "Flag"))) %>% 
  pivot_longer(cols = paste0("1950":latest_year),
               names_to = "year",
               values_to = "value")

# pivot all of the flag columns   
mar_flags <- mar %>% 
  dplyr::select(-paste0("1950":latest_year)) %>% 
  pivot_longer(cols = paste("1950":latest_year, "Flag"),
               names_to = "flag_year",
               values_to = "flag") %>% 
  mutate(year = str_remove(flag_year, " Flag")) %>% 
  dplyr::select(year, flag, row_id)
  
# combine flag and row id 
mar <- mar_values %>% left_join(mar_flags, by = c("row_id", "year"))

mar <- mar %>% 
  mutate(value = case_when(str_detect(flag, "N") ~ sub_N,
                               TRUE ~ value)) %>%
  dplyr::select(-row_id) %>% 
  mutate(FAO_name = ifelse(!is.na(FAO_name), FAO_name, family_scientific)) 

# Change NEI to nei for possible species match fix... This was a new change in the FAO data for v2025
mar <- mar %>% 
  mutate(FAO_name = str_replace(FAO_name, "NEI", "nei"))

# filter out family seaweeds:
check <- mar %>%
filter(family_scientific %in% toupper(c("Alariaceae", "Laminariaceae", "Sargassaceae", "Bangiaceae", "Bonnemaisoniaceae", "Chordariaceae", "Gelidiaceae", "Gigartinaceae", "Gracilariaceae", "Palmariaceae", "Solieriaceae", "Caulerpaceae", "Codiaceae", "Monostromataceae", "Ulotrichaceae", "Ulvaceae", "Spirulinaceae", "DUNALIELLACEAE")))

```



## Update seaweed species information

Filter out seaweed species from 'raw/species_list.csv' (from 'globalprep/mar/v2020/raw'), rename columns, and assign proportions to "include" column determined by research of non-human food vs human food seaweed species cultivated in mariculture. For NP, we are only including non-human food seaweed species. Since some species are used for both non-human food and human food purposes, a proportion is assigned based on research and best guess.

**Mariculture data prep must be completed before you complete this part**

```{r}
# Read in 'species_list.csv' (originally from 'globalprep/mar/v2025/raw'). 
#> Filter for 'Taxon_code = AL' (only seaweed species). 
#> Rename 'exclude' columns to 'include' since we're now including 
#> seaweed species that were excluded in the MAR dataprep (not primarily used as human food). 
#> Therefore, "0" means exclude completely (0%), and "1" means include completely (100%).

#> Note, this calls out to the mariculture data prep. 
#> The mariculture data prep for your year should be done before you retrieve this file. 

seaweed_sp <- readr::read_csv(here(current_mar_dir, "raw", "species_list.csv")) %>%
  filter(Taxon_code == 'AL') %>%
  rename(include = exclude)

## Save seaweed species list
readr::write_csv(seaweed_sp, here(current_np_dir, "raw", "species_list_np_seaweeds.csv"))

```




## Update species names

Update species name in the `raw/species_list_np_seaweeds_edited.csv` file with names in the `mar` dataset. Simplified the species list and cut the "species" name columns because it wasn't clear what this was trying to accomplish and created potential error.

```{r}
# Read in edited 'species_list_np_seaweeds_edited.csv'.
seaweeds <- read_csv(here(current_np_dir, "raw", "species_list_np_seaweeds.csv"))

# select specific columns
seaweeds_sp <- seaweeds %>%
  dplyr::select(FAO_name, include, Taxon_code, family)


setdiff(seaweed_sp$FAO_name, check$FAO_name) # fine to have extra things in here.
setdiff(check$FAO_name, seaweed_sp$FAO_name) #if extra things are in here...check.

# REMOVE SPECIES not relevant to natural products goal (i.e., human food species)
seaweed_np <- seaweeds_sp %>% 
  left_join(mar, by = "FAO_name") %>% 
  filter(include > 0) %>%
  filter(!is.na(value)) %>%
  rename(species = FAO_name)



```


## Convert country names to OHI regions

```{r}

# ---- Divide mariculture from countries that we report as separate regions (assume equal production in all regions) ----
# Netherlands Antilles: Conch restoration among Aruba, Bonaire, Curacao
# Channel Islands: Jersey and Guernsey
# Bonaire/S.Eustatius/Saba
# Yugoslavia SFR: no longer a country after 1992
seaweed_np <- seaweed_np %>%
  mutate(country = ifelse(country == "Réunion", "Reunion", country)) %>%  # this one is hard to get right; v2020: last year it was "R\xe9union", but this year it was "Réunion" - not present in v2021 data, not present in v2025 data
  ohicore::split_regions()  # function in mar_fxs.R

mar_rgn_all <- name_2_rgn(df_in = seaweed_np, 
                       fld_name = 'country', 
                       flds_unique = c('species', 'fao', 'environment', 'Taxon_code', 'year', 'include'))

# ---- Duplicates ----
# v2025:
# 1 United Republic of Tanzania          
# 2 United Republic of Tanzania, Zanzibar
# okay because this gets resolved in next step

# ---- Sum values of regions with multiple subregions ----
mar_rgn <- mar_rgn_all %>%
  group_by(species, year, rgn_id, include) %>% 
  summarize(value = sum(value)) %>%
  ungroup()
dim(mar_rgn)
#about 52 countries that rear algae
```


Remove species-region-environment time series with less than four years of seaweed mariculture production \> 0 tonnes (assume these are not established seaweed mariculture programs).

```{r}
mar_rgn <- mar_rgn %>% 
  group_by(rgn_id, species) %>%
  mutate(not_0 = length(value[value > 0])) %>% # length of vector of years greater than 0
  filter(not_0 > 3) %>% # filter for groups that have at least four years of seaweed mariculture production 
  ungroup() %>% 
  dplyr::select(rgn_id, species, year, include, value) 

```

# summarize tonnes of production by country/year
```{r}

mar_rgn2 <- mar_rgn %>%
  mutate(tonnes_nonfood = include * value) %>%
  group_by(rgn_id, year) %>%
  summarize(tonnes = sum(tonnes_nonfood))

```

# Add in sustainability

We use the Seafood Watch general seaweed sustainability score. Currently, this is just a placeholder because we use the same score for each country across all years (6.72 out of 10). Eventually Seafood Watch might provide country or taxon specific values, in which case this should be edited to take this into account. 

```{r}

mar_rgn2 <- mar_rgn2 %>%
   mutate(sustainability = 0.67) %>%
  mutate(tonnes_sustainable = tonnes * sustainability) %>%
  select(rgn_id, year, tonnes_sustainable)

```

# Gapfilling

## Fill in missing years after first year of harvest data with 0 values

```{r}
# Pivot wider mar_rgn to create a value for every year-species-region (if missing it will be given a NA)
mar_rgn_wide <- mar_rgn2 %>% 
  pivot_wider(names_from = year, values_from = tonnes_sustainable)

dim(mar_rgn_wide)
# 143  81

# Turn data frame back into long format
mar_rgn_gf <- mar_rgn_wide %>% 
  pivot_longer(cols = -(rgn_id), 
               names_to = "year", 
               values_to = "tonnes_sustainable") 

# NA values are converted to zero
mar_rgn_gf <- mar_rgn_gf %>%
  mutate(year = as.numeric(as.character(year))) %>%
  mutate(gf = ifelse(is.na(tonnes_sustainable), "NA_to_zero", NA)) %>%
  mutate(tonnes_sustainable = ifelse(is.na(tonnes_sustainable), 0, tonnes_sustainable)) %>%
  group_by(rgn_id) %>%
  mutate(cum_value = cumsum(tonnes_sustainable)) %>%
  ungroup() %>%
  filter(cum_value > 0) %>% # eliminates years before mariculture began
  filter(year >= 2004) %>%
  dplyr::select(-cum_value)

```



## Rolling average of sustainable tonnes


```{r}

# for each region and year, calculate rolling mean in a 5 year window (year + 4 years, right-aligned (year, look at the 4 years previous))
seaweed_st_rolling <- mar_rgn_gf %>%
  dplyr::ungroup() %>% 
  dplyr::arrange(rgn_id, year) %>%
  dplyr::group_by(rgn_id) %>%
  dplyr::mutate(
      sust_tonnes_rollmean = zoo::rollapply(tonnes_sustainable, width = 4,
                                            FUN = mean, align = 'right',
                                            partial = TRUE, na.rm = TRUE)) %>%
  dplyr::select(rgn_id, year, tonnes_sustainable, sust_tonnes_rollmean)


summary(seaweed_st_rolling)
```


## Peak sustainable tonnes

Identify max value from rolling average of sustainable tonnes (maybe limit to most recent 10 years of data to ID max). 


```{r}
# define decade range for max value 

end_data_year <- max(seaweed_st_rolling$year)
seaweed_decade <- c((end_data_year - 9):end_data_year)
seaweed_decade # 2014-2023

# find peak sustainable tonnes 
seaweed_st_max <- seaweed_st_rolling %>% 
  filter(year %in% c(seaweed_decade)) %>% 
  group_by(rgn_id) %>% 
  summarize(peak_sust_tonnes = max(sust_tonnes_rollmean, na.rm = TRUE)) %>%
  dplyr::select(rgn_id, peak_sust_tonnes)


head(seaweed_st_max)
```


## Relative sustainable tonnes

Bring max sustainable tonnes value back into sustainable tonnes full dataframe. Use this value to calculate relative sustainable tonnes: if `sust_tonnes > peak_sust_tonnes`, then set to 1. Else, `tonnes_rollmean / peak_sust_tonnes`.  

> NOTE! Because relative sustainability is calculated using a rolling mean and the maximum sustainable tonnes from the previous ten years, the data check for this value will be a little strange. The temporal reference point means that a change in the `peak_sust_tonnes` will change relative sustainability for everything. 

When `sust_tonnes` is less than or equal to `peak_sust_tonnes`,

$$RelSustTonnes = TonnesRollmean / PeakSustTonnes$$
As `peak_sust_tonnes` increases, `rel_sust_tonnes` decreases. And as `peak_sust_tonnes` decreases, `rel_sust_tonnes` increases. 

So, if `peak_sust_tonnes` for the current data year is greater than `peak_sust_tonnes` for the previous data year, when comparing scores for the previous data year, it will seem like the score has gone down. And vice verse, if `peak_sust_tonnes` for the current data year is less than `peak_sust_tonnes` for the previous data year, it will seem like the score has increased. 


```{r}
# ---- Determine relative sustainable tonnes ----
seaweed_relative_sust_tonnes <- 
  # join peak sustainable tonnes with full sustainable tonnes df
  seaweed_st_rolling %>% 
  ungroup() %>% 
  ## filling in peak value per region ID across full year range
  full_join(seaweed_st_max, by = join_by("rgn_id")) %>% 
  # rescale to find relative sustainable tonnes
  mutate(rel_sust_tonnes = ifelse(sust_tonnes_rollmean > peak_sust_tonnes, 1,
                             sust_tonnes_rollmean / peak_sust_tonnes)) %>% 
  mutate(rel_sust_tonnes = ifelse(is.na(rel_sust_tonnes), 0, rel_sust_tonnes)) %>%
  mutate(product = "seaweeds")
 
# check
summary(seaweed_relative_sust_tonnes)


# save full df to int folder
readr::write_csv(seaweed_relative_sust_tonnes, here(current_np_dir, "int", "np_relative_sust_tonnes_seaweeds_full.csv"))


# save selected columns to output folder
final_sw_relative_sust_tonnes <- seaweed_relative_sust_tonnes %>% 
  dplyr::select(rgn_id, product, year, rel_sust_tonnes)

readr::write_csv(final_sw_relative_sust_tonnes, here(current_np_dir, "output", "np_seaweeds.csv"))

```

