---
title: 'OHI `r format(Sys.time(), "%Y")`: Natural Products, Preparing catch and B/Bmsy data for forage fish species'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

[REFERENCE RMD FILE: <http://ohi-science.org/ohiprep_v2024/globalprep/np/v2024/STEP1c_np_fishfeed_prep.html>]

# Summary

This script takes the SAUP 2023 catch data, a list of fish oil/fish meal (FOFM) species, and B/Bmsy estimates to calculate a score for FOFM species per OHI region. The FIS data prep must be completed prior to this step.

## Updates from previous assessment

- New RAM data v4.66
- Used the old NP FOFM Rmd to fill out this document under the new framework. See this [google doc](https://docs.google.com/document/d/1ubCTW7ZrvvdckvY2zpBzCLHLgoIlCLW8kDMj-KWda9U/edit?tab=t.0) for reference. 
- There were a few confusing errors from the previous years' calculations that were streamlined and simplified. Primarily, exporting an intermediate data of raw catch for step 5, correctly using mean catch (instead of raw catch) for the bbmsy weighting, and correcting the weighting process to use mean catch instead of sum catch.

------------------------------------------------------------------------

# Data Source

## Sea Around us

**Reference**: Pauly D., Zeller D., Palomares M.L.D. (Editors), 2020. Sea Around Us Concepts, Design and Data (seaaroundus.org).

**Downloaded**: September 27, 2022

**Description**: Tons per year and SAUP region with information on sector type, industry type, fishing entitity, reporting status and taxonomic information.

**Time range**: 1950 - 2019

**Format**: CSV

**Additional Information**: [Methods](http://www.seaaroundus.org/sea-around-us-methods-index/)

## Avoiding the ecological limits of forage fish

**Reference**: Froehlich, H.E., Jacobsen, N.S., Essington, T.E., Clavelle, T., and Halpern, B.S. (2018). Avoiding the ecological limits of forage fish for fed aquaculture. Nature Sustainability 1, 298.

**Downloaded**: July 7, 2020. Obtained from Melanie Frazier (NCEAS).

**Description**: List of FOFM species from Watson v3 data.

**Native data resolution**:

**Format**: CSV format

## RAM Legacy Data

**Reference**: [RAM Legacy Stock Assessment Database](http://ramlegacy.org) v4.66

**Downloaded**: 06/26/2025

**Description**: B/Bmsy value by stock and year (other data, which we do not use, are also available in the database)

**Native data resolution**: stock (fish stock, species and region specific)

**Time range**: 1800 - 2023

**Format**: CSV format

**Additional Information**: We use the finalized b/bmsy layer from OHI-global for this data prep. We do not actually read in the raw RAM data here.

------------------------------------------------------------------------

# Methods

INCOMPLETE. See this folder's `README.md` for details. 


## Setup

```{r}
knitr::opts_chunk$set(eval = FALSE)

# ======= Load packages ============
if (!require(librarian)) {
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  ohicore, #devtools::install_github('ohi-science/ohicore@dev') #if relevant: restart session after reinstalling
  here,
  tidyverse,
  zoo,
  plotly,
  tictoc,
  RColorBrewer
)

# ======= Set directories ===========
# Update scenario year, set up programmatic scenario year updates
current_year <- 2025 # update this!!!
previous_year <- current_year - 1

# Directories
data_dir <- paste0("d", current_year)
prev_data_dir <- paste0("d", previous_year)

# Scenarios
scen_year <- paste0("v", current_year)
prev_scen_year <- paste0("v", previous_year)

data_years <- c("1950", "2023") # update to correct data years

# Natural Products directory on GitHub
current_np_dir <- here::here("globalprep", "np", scen_year) 
previous_np_dir <- here::here("globalprep", "np", paste0(prev_scen_year, "_new")) # update this if name changes

# Fisheries directory on GitHub
current_fis_dir <- here::here("globalprep", "fis", scen_year)
previous_fis_dir <- here::here("globalprep", "fis", prev_scen_year)

# ======== Source functions ===========
source(here::here("workflow", "R", "common.R")) # directory locations
# Load FAO-specific user-defined functions
source(here::here("workflow", "R", "fao_fxn.R")) # function for cleaning FAO files
# function for handling FAO commodity data specific to NP
source(here::here(current_np_dir, "R", "np_fxn.R")) 
# functions specific to mariculture dealing with compound countries
source(here("globalprep", "mar", scen_year, "mar_fxs.R"))
```

**v2025 updated steps**

This year's analysis cut steps 1 and 2 and started straight with step 3. We are using the forage fish catch stocks from the previous years analysis (as the data has not been updated since 2019).

1. Join with the final B/Bmsy layer used in the FIS model
2. Convert the B/Bmsy values to scores (this is done in functions.R)
3. Apply the underharvest penalty
4. Take a catch weighted average of B/Bmsy scores for each region/year.

*Old steps to this*

1.  Subset the forage fish catch stocks for each region/year (keep FAO and OHI rgn ids)
2.  Multiply by 0.70 to reflect the amount going to feed/oils
3.  Join with the final B/Bmsy layer used in the FIS model
4.  Convert the B/Bmsy values to scores (this is done in functions.R)
5.  Apply the underharvest penalty
6.  Take a catch weighted average of B/Bmsy scores for each region/year.

```{r, eval=FALSE}

## Read in Froehlich list of forage fish species
forage <- read_csv(here(current_np_dir, "raw", "msleckman.61.1-CatchMSY_Nis_FAOAreas.csv"))
sort(unique(forage$Species)) #238 forage fish groups listed

# This was not updated in v2023 so we continue to use the 2022 version.
saup_taxonkey <- readr::read_csv(here(current_fis_dir, "int", "taxon_key_v2022.csv"))
  
## Combined list:
foragefish_list <- data.frame(forage_fish = sort(unique(c(unique(forage$Species), saup_taxonkey$TaxonName[saup_taxonkey$forage_fish %in% 1]))))

foragefish_list <- data.frame(forage_fish = sort(unique(forage$Species)))

# Update column names
foragefish_list <- foragefish_list %>%
  mutate(forage_fish = case_when(
    forage_fish == "Ammodytes spp" ~ "Ammodytes",
    forage_fish == "Clupea pallasii" ~ "Clupea pallasii pallasii",
    forage_fish == "Clupeoidei" ~ "Clupeoids", 
    forage_fish == "Decapterus spp" ~ "Decapterus", 
    forage_fish == "Diplodus argenteus" ~ "Diplodus argenteus argenteus",
    forage_fish == "Etrumeus teres" ~ "Etrumeus sadina",
    forage_fish == "Gadiculus argenteus" ~ "Gadiculus argenteus thori",
    forage_fish == "Gymnocephalus cernuus" ~ "Gymnocephalus cernua",
    forage_fish == "Mullus barbatus" ~ "Mullus barbatus barbatus",
    forage_fish == "Patagonotothen brevicauda" ~ "Patagonotothen brevicauda brevicauda",
    forage_fish == "Rioraja agassizi" ~ "Rioraja agassizii",
    forage_fish == "Sardinella spp" ~ "Sardinella", 
    forage_fish == "Trachurus spp" ~ "Trachurus",
    TRUE ~ forage_fish
  )) %>% # fix some name issues in the forage fish list to match the names in SAUP 
  left_join(saup_taxonkey, by = c("forage_fish" = "TaxonName")) %>% 
   mutate(inSAUP = ifelse(is.na(TaxonKey), NA, "yes")) ## Now only 9 missing spp.. this is good enough! Note: some of the species that are listed as in the SAUP data, are actually not in the SAUP data. This is because this forage fish list was originally derived from the Watson data, and there are some Watson species which are not in the SAUP data. This is not a problem. 
# sum(is.na(foragefish_list$inSAUP))
## 10 missing spp. in v2024 

readr::write_csv(foragefish_list, here(current_np_dir, "int", "master_taxa_list_SAUP.csv"))

```

## Read in (v2022) catch data

```{r}
# This was not updated in v2023 or 2024 so we call out to the v2022 year on Mazu
file <- file.path(dir_M, 'git-annex/globalprep/fis/v2022/int/stock_catch_by_rgn_taxa.csv')

# read in Sea Around Us catch data
catch_raw <- readr::read_csv(file) 

# update column names
catch <- catch_raw %>%
  rename(common = CommonName, fao_id = fao_rgn, species = TaxonName)

# view summary
summary(catch)
colnames(catch)

## filter out non ohi eez regions if there are any
catch <- catch %>%
  filter(!is.na(rgn_id)) %>%
  filter(!is.na(fao_id)) %>%
  filter(rgn_id <= 250) %>%
  filter(rgn_id != 213) # drop Antarctica

```

## Subset the (v2022) catch data for our forage fish species

```{r}
## need to get TaxonKey's for each species to join with catch

# read in list we wrote out earlier
foragefish_list <- readr::read_csv(here::here(current_np_dir, "int", "master_taxa_list_SAUP.csv"))

# find missing species 
missing_spp <- foragefish_list %>%
  filter(is.na(TaxonKey))

missing_spp$forage_fish
# v2025 missing species:
#  [1] "Ariomma indica"          "Blicca bjoerkna"         "Centrolabrus exoletus"  
#  [4] "Citharidae"              "Hemiramphus balao"       "Liza saliens"           
#  [7] "Percarina demidoffi"     "Sardinops caeruleus"     "Sardinops melanostictus"
# [10] "Sardinops ocellatus

# test <- catch %>%
#   filter(str_detect(paste(missing_spp$forage_fish, collapse = "|"), species)) # just make sure the NA species are actually not in the SAUP data
# sort(unique(test$species)) # no full matches..
# # v2024: [1] "Hemiramphus" "Liza" 


# ---- Join forage fish species with catch data ----
forage_fish_taxa_list <- foragefish_list %>%
  # subset to only keep catch data for which we have forage species data
  left_join(catch, by = c("forage_fish" = "species", "TaxonKey")) %>%
  dplyr::select(forage_fish, inSAUP, TaxonKey) %>% # drop stock, common name, and extra catch df columns
  unique() %>% # return unique combinations of grouping variables in the data frame
  mutate(inSAUP = ifelse(
    # if TaxonKey is NA, set inSAUP to NA
   is.na(TaxonKey), NA, inSAUP)) 

# save a list of species not in SAUP data for bookkeeping
forage_fish_no_saup <- forage_fish_taxa_list %>%
  filter(is.na(TaxonKey)) 

# write out missing species list to int
readr::write_csv(forage_fish_no_saup, here(current_np_dir, "int", "forage_fish_not_in_SAUP.csv"))


# join catch data set with forage fish taxa list 
catch_fishfeed <- catch %>%
  left_join(forage_fish_taxa_list, by = c("TaxonKey")) %>%
  # subset to only include forage fish species
  dplyr::filter(!is.na(forage_fish))

nrow(catch_fishfeed) # 600864 - v2025

# write output to int folder in np/scenario year directory
readr::write_csv(catch_fishfeed, here::here(current_np_dir, "int", "saup_catch_forage_fish.csv"))

test <- catch_fishfeed %>%
  group_by(year) %>%
  summarise(sum = sum(tons)) 

summary(test)[4, 2]
## looks good.. ~30 million per year in v2023 and v2024 and v2025 
```

# Step 2

## Multiply by 0.90 to reflect the amount going to feed/oils, and not the amount going to human consumption (that is accounted for in food provision)

Justification from the [Froelich et al. 2018](https://www.nature.com/articles/s41893-018-0077-1#Sec11): "Currently, it is estimated about 10% of forage fish enter the human diet directly, but the notoriously tiny-boned fish are labour intensive (thus expensive) to process for human consumption, are the foundation of several industries and thus jobs (creating inertia to change) and are not the preferred fish type for most people."

```{r}
# read in SAUP catch data subsetted to forage fish species 
catch_fishfeed <- readr::read_csv(here::here(current_np_dir, "int", "saup_catch_forage_fish.csv"))

# Multiply by 0.90 to reflect the amount going to feed/oils, not human consumption
catch_non_human <- catch_fishfeed %>% 
  mutate(tons_non_human = tons * 0.9) 

# for each region, FAO ID, Taxon Key, stock ID, and year, 
# find sum of non-human-consumption fish harvest
catch_non_human <- catch_non_human %>%
  dplyr::select(year, rgn_id, fao_id, stock_id, TaxonKey, tons_non_human) %>%
  group_by(rgn_id, fao_id, TaxonKey, stock_id, year) %>%
  summarize(catch_non_human = sum(tons_non_human)) %>%
  ungroup()

test <- catch_non_human %>%
  group_by(year) %>%
  summarise(sum = sum(catch_non_human)) 

summary(test)[4, 2]
# ~ 26.6 mil per year -- v2024 and v2025

# --------- Zero-fill -----------------------

## these data have no zero catch values, so add years with no reported catch to data table:
catch_year_zeros <- catch_non_human %>%
  pivot_wider(names_from = year, values_from = catch_non_human,
              names_prefix = "x_",
              values_fill = 0) %>% 
  pivot_longer(cols = starts_with("x_"),
               names_to = "year",
               values_to = "catch_non_human") %>% 
  mutate(year = as.numeric(gsub("x_", "", year))) %>% 
  arrange(year)

## this part eliminates the zero catch values prior to the first reported non-zero catch  
catch_year_zeros <- catch_year_zeros %>%
  group_by(fao_id, TaxonKey, stock_id, rgn_id) %>%
  arrange(year) %>%
  mutate(cum_catch = cumsum(catch_non_human)) %>%
  filter(cum_catch > 0) %>%
  dplyr::select(-cum_catch) %>%
  ungroup()

# Calculate mean catch for ohi regions (using data from 1980 onward). These data are used to weight the RAM b/bmsy values 
# mean_catch <- catch_year_zeros %>%
#   filter(year >= 1980) %>%
#   group_by(rgn_id, fao_id, TaxonKey, stock_id) %>%
#   summarize(mean_catch_non_human = mean(catch_non_human, na.rm=TRUE)) %>% # mean catch for each stock (in a specific ohi-fao region)
#   filter(mean_catch_non_human != 0)  %>% # some stocks have no reported catch for time period
#   ungroup()

mean_catch <- catch_year_zeros %>%
  filter(year >= 1980) %>%
  group_by(rgn_id, fao_id, TaxonKey, stock_id) %>%
  mutate(mean_catch_non_human = mean(catch_non_human, na.rm=TRUE)) %>% # mean catch for each stock (in a specific ohi-fao region)
  filter(mean_catch_non_human != 0)  %>% # some stocks have no reported catch for time period
  ungroup()

mean_catch_dist <- mean_catch %>% 
  select(!catch_non_human) %>% 
  dplyr::distinct(rgn_id, fao_id, year, TaxonKey, stock_id, mean_catch_non_human)


options(scipen = 999) # to prevent taxonkey from turning into scientific notation

mean_catch_FOFM <- mean_catch_dist %>%
  mutate(stock_id_taxonkey = paste(stock_id, TaxonKey, sep="_")) %>%
  dplyr::select(rgn_id, stock_id_taxonkey, mean_catch_non_human, year) %>%
  # filter(year >= 2001) %>%  # filter to include only analysis years
  data.frame() 
  # rename(catch = mean_catch_non_human)

readr::write_csv(mean_catch_FOFM, here(current_np_dir, "int", "mean_catch_FOFM.csv"))

## Separate file to save the raw data as well as the mean catch data. Used in step 5 weighting
mean_raw_catch <- mean_catch %>% 
  mutate(stock_id_taxonkey = paste(stock_id, TaxonKey, sep="_"),
         product = "fish_oil") %>%
  dplyr::select(rgn_id, stock_id_taxonkey, mean_catch_non_human, catch_non_human, year, product) %>%
  filter(year >= 2001) %>%  # filter to include only analysis years
  data.frame() 

readr::write_csv(mean_raw_catch, here(current_np_dir, "int", "raw_mean_catch_FOFM.csv"))


```


# Steps 3/4/5/6

Join with the final B/Bmsy layer from the fis model, convert the B/Bmsy values to scores (this is done in functions.R for FIS subgoal, but we do it in the dataprep for NP), apply the underharvest penalty, and take a catch weighted average of B/Bmsy scores for each region/year.

```{r}
fis_bbmsy <- read_csv(here(current_fis_dir, "output", "fis_bbmsy.csv"))

fis_bbmsy_gf <- read_csv(here(current_fis_dir, "output", "fis_bbmsy_gf.csv"))

# This data comes from the previous year's analysis and has been copied over into the most recent scenario directory
catch_FOFM <- read_csv(here(current_np_dir, "int", "mean_catch_FOFM.csv"))

b <- fis_bbmsy %>%
  dplyr::mutate(bbmsy = ifelse(bbmsy > 1 , 1, bbmsy)) # no underharvest penalty anymore! 

c <- catch_FOFM %>%
    dplyr::mutate(stock_id_taxonkey = as.character(stock_id_taxonkey)) %>%
    dplyr::mutate(taxon_key = sub('.*_', '', stock_id_taxonkey)) %>%
    dplyr::mutate(stock_id = sub('_[^_]*$', '', stock_id_taxonkey)) %>%
    dplyr::mutate(mean_catch_non_human = as.numeric(mean_catch_non_human)) %>%
    dplyr::mutate(year = as.numeric(as.character(year))) %>%
    dplyr::mutate(region_id = as.numeric(as.character(rgn_id))) %>%
    dplyr::mutate(taxon_key = as.numeric(as.character(taxon_key))) %>%
    dplyr::select(region_id, stock_id, taxon_key, mean_catch_non_human, year)


b <- b %>%
  dplyr::mutate(bbmsy = as.numeric(bbmsy),
                region_id = as.numeric(as.character(rgn_id)),
                year = as.numeric(as.character(year)),
                stock_id = as.character(stock_id))


####
# ==== Calculate scores for Bbmsy values ==== ####
#  *************NOTE *****************************
#  These values can be altered
#  ***********************************************

## The upper limit can be deleted, since we already cap pedscores at 1 anyways. We still want to keep the lower buffer.

alpha <- 0.5
beta <- 0.25
lowerBuffer <- 0.95
upperBuffer <- 1.05

# Rename to stock score

b$stock_score = ifelse(
  b$bbmsy < lowerBuffer,
  b$bbmsy,
  ifelse (b$bbmsy >= lowerBuffer &
            b$bbmsy <= upperBuffer, 1, NA)
)
b$stock_score = ifelse(!is.na(b$stock_score),
                 b$stock_score,
                 ifelse(
                   1 - alpha * (b$bbmsy - upperBuffer) > beta,
                   1 - alpha * (b$bbmsy - upperBuffer),
                   beta
                 ))

# ISSUE: Conceptual check. The score here is representative of sustainability?
colnames(c)
####
# Merge the b/bmsy data with catch data ------------------------------
####
data_fis <- c %>%
  dplyr::left_join(b, by = c('region_id' = 'rgn_id', 'stock_id', 'year')) %>%
  dplyr::select(rgn_id = region_id, stock_id, year, taxon_key, mean_catch_non_human, bbmsy, stock_score)

###
#  Estimate scores for taxa without b/bmsy values --------------------
# Mean score of other fish in the region is the starting point
# Then a penalty is applied based on the level the taxa are reported at
###

data_fis <- data_fis %>% 
  filter(year > 2000)

## this takes the mean score within each region and year
data_fis_gf <- data_fis %>%
  dplyr::group_by(rgn_id, year) %>%
  dplyr::mutate(mean_score = mean(stock_score, na.rm = TRUE)) %>%
  dplyr::ungroup()

## this takes the mean score across all regions within a year
# (when no stocks have scores within a region)
data_fis_gf <- data_fis_gf %>%
  dplyr::group_by(year) %>%
  dplyr::mutate(mean_score_global = mean(stock_score, na.rm = TRUE)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(mean_score = ifelse(is.na(mean_score), mean_score_global, mean_score)) %>%
  dplyr::select(-mean_score_global)

#  *************NOTE *****************************
#  In some cases, it may make sense to alter the
#  penalty for not identifying fisheries catch data to
#  species level.
#  ***********************************************

penaltyTable <- data.frame(TaxonPenaltyCode = 1:6,
                           penalty = c(0.1, 0.25, 0.5, 0.8, 0.9, 1))

data_fis_gf <- data_fis_gf %>%
  dplyr::mutate(TaxonPenaltyCode = as.numeric(substring(taxon_key, 1, 1))) %>%
  dplyr::left_join(penaltyTable, by = 'TaxonPenaltyCode') %>%
  dplyr::mutate(stock_score_gf = mean_score * penalty) %>%
  dplyr::mutate(method = ifelse(is.na(stock_score), "Mean gapfilled", NA)) %>%
  dplyr::mutate(gapfilled = ifelse(is.na(stock_score), 1, 0)) %>%
  dplyr::mutate(stock_score = ifelse(is.na(stock_score), stock_score_gf, stock_score))


gap_fill_data <- data_fis_gf %>%
  dplyr::select(rgn_id,
         stock_id,
         taxon_key,
         year,
         mean_catch_non_human,
         stock_score,
         gapfilled,
         method) 

readr::write_csv(gap_fill_data, here::here(current_np_dir, "output", "NP_bbmsy_summary_gf.csv"))

score_data <- data_fis_gf %>%
  dplyr::select(rgn_id, stock_id, year, mean_catch_non_human, stock_score)

colSums(is.na(score_data)) # 2723
nrow(score_data) # 24923
# ~ 10% NA where we have SAUP catch data but no FAO BBMSY data

 ###
# Calculate status for each region
###

## Take a catch weighted average of B/Bmsy scores for each region/year.


score_data1 <- score_data %>%
  dplyr::group_by(rgn_id, year) %>%
  dplyr::summarize(score = weighted.mean(stock_score, mean_catch_non_human)) %>%
  dplyr::ungroup() %>% 
  mutate(product = "fish_oil")

summary(score_data1)

readr::write_csv(score_data1, here(current_np_dir, "output", "np_fofm_scores_new.csv"))

old <- read_csv(here(current_np_dir, "output", "np_fofm_scores.csv"))
nrow(old)

new <- read_csv(here(current_np_dir, "output", "np_fofm_scores_new.csv"))
nrow(new)
nrow(filter(new, !is.na(year))) # 3242 non-na values

check <- new %>%
  rename("new_score" = "score") %>%
  left_join(old, by = c("rgn_id", "year")) %>%
  left_join(rgns_all, by = "rgn_id") %>%
  select(rgn_id, year, new_score, score, rgn_name) %>%
  mutate(diff = new_score - score) %>%
  filter(year == 2019)

ab_scores <- ggplot() + 
  geom_abline(slope = 1, intercept = 0, col = "red") +
  geom_point(data = check, aes(x = score, y = new_score, text = paste("Region:", rgn_name), alpha = 0.7)) +
  labs(x = "Old Score", 
       y = "New Score",
       title = "FOFM New score vs old score for 2019") +
  theme_bw();ggplotly(ab_scores)
```

## Data Check

```{r}
# Read in the old data
old <- readr::read_csv(here::here("globalprep", "np", prev_scen_year, "output", "np_fofm_scores.csv"))
summary(old)

# Read in the new data
new <- readr::read_csv(here::here(current_np_dir, "output", "np_fofm_scores.csv"))
summary(new)

# load ohi regions
region_data()

# Investigating changes in data for same year in different data/scenario years

check <- new %>%
  rename("new_score" = "score") %>%
  left_join(old, by = c("rgn_id", "year")) %>%
  left_join(rgns_all, by = "rgn_id") %>%
  select(rgn_id, year, new_score, score, rgn_name) %>%
  mutate(diff = new_score - score) %>%
  filter(year == 2019)

ab_scores <- ggplot() + 
  geom_abline(slope = 1, intercept = 0, col = "red") +
  geom_point(data = check, aes(x = score, y = new_score, text = paste("Region:", rgn_name), alpha = 0.7)) +
  labs(x = "Old Score", 
       y = "New Score",
       title = "FOFM New score vs old score for 2019") +
  theme_bw();ggplotly(ab_scores)

```


