---
title: 'OHI `r format(Sys.time(), "%Y")`: Natural Products, Preparing catch and B/Bmsy data for forage fish species'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

[REFERENCE RMD FILE: <http://ohi-science.org/ohiprep_v2024/globalprep/np/v2024/STEP1c_np_fishfeed_prep.html>]

# Summary

This script takes the SAUP 2023 catch data, a list of fish oil/fish meal (FOFM) species, and B/Bmsy estimates to calculate a score for FOFM species per OHI region. The FIS data prep must be completed prior to this step.

## Updates from previous assessment

- New RAM data v4.66

------------------------------------------------------------------------

# Data Source

## Sea Around us

**Reference**: Pauly D., Zeller D., Palomares M.L.D. (Editors), 2020. Sea Around Us Concepts, Design and Data (seaaroundus.org).

**Downloaded**: September 27, 2022

**Description**: Tons per year and SAUP region with information on sector type, industry type, fishing entitity, reporting status and taxonomic information.

**Time range**: 1950 - 2019

**Format**: CSV

**Additional Information**: [Methods](http://www.seaaroundus.org/sea-around-us-methods-index/)


## RAM Legacy Data

**Reference**: [RAM Legacy Stock Assessment Database](http://ramlegacy.org) v4.66

**Downloaded**: 06/26/2025

**Description**: B/Bmsy value by stock and year (other data, which we do not use, are also available in the database)

**Native data resolution**: stock (fish stock, species and region specific)

**Time range**: 1800 - 2023

**Format**: CSV format

**Additional Information**: We use the finalized b/bmsy layer from OHI-global for this data prep. We do not actually read in the raw RAM data here.

------------------------------------------------------------------------

# Methods

INCOMPLETE. See this folder's `README.md` for details. 


## Setup

```{r}
knitr::opts_chunk$set(eval = FALSE)

# ======= Load packages ============
if (!require(librarian)) {
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  ohicore, #devtools::install_github('ohi-science/ohicore@dev') #if relevant: restart session after reinstalling
  here,
  tidyverse,
  zoo,
  plotly,
  tictoc,
  RColorBrewer
)

# ======= Set directories ===========
# Update scenario year, set up programmatic scenario year updates
current_year <- 2025 # update this!!!
previous_year <- current_year - 1

# Directories
data_dir <- paste0("d", current_year)
prev_data_dir <- paste0("d", previous_year)

# Scenarios
scen_year <- paste0("v", current_year)
prev_scen_year <- paste0("v", previous_year)

data_years <- c("1950", "2023") # update to correct data years

# Natural Products directory on GitHub
current_np_dir <- here::here("globalprep", "np", scen_year) 
previous_np_dir <- here::here("globalprep", "np", paste0(prev_scen_year, "_new")) # update this if name changes

# Fisheries directory on GitHub
current_fis_dir <- here::here("globalprep", "fis", scen_year)
previous_fis_dir <- here::here("globalprep", "fis", prev_scen_year)

# ======== Source functions ===========
source(here::here("workflow", "R", "common.R")) # directory locations
# Load FAO-specific user-defined functions
source(here::here("workflow", "R", "fao_fxn.R")) # function for cleaning FAO files
# function for handling FAO commodity data specific to NP
source(here::here(current_np_dir, "R", "np_fxn.R")) 
# functions specific to mariculture dealing with compound countries
source(here("globalprep", "mar", scen_year, "mar_fxs.R"))
```

**v2025 updated steps**

We are using the forage fish catch stocks from the previous years analysis (as the data has not been updated since 2019).

1. Join with the final B/Bmsy layer used in the FIS model
2. Convert the B/Bmsy values to scores (this is done in functions.R)
4. Take a catch weighted average of B/Bmsy scores for each region/year.



# STEP 1

Identify catch tonnes used for B/Bmsy. We use the catch average catch since 1990 to weight the yearly B/Bmsy scores for these taxa. We use the SAUP designations of whether catch is used for fishoil fishmeal.
```{r, eval=FALSE}

saup_taxonkey <- readr::read_csv(here(current_fis_dir, "int", "taxon_key_v2022.csv"))

fish <- read_csv("/home/shares/ohi/git-annex/globalprep/fis/v2022/int/ohi_rgn_eez_prod.csv") 

catch <- fish %>%
  select(rgn_id, fao_id, year, end_use_type, scientific_name, common_name, tonnes) %>%
  filter(end_use_type == "Fishmeal and fish oil") %>%
  filter(year >= 1990) %>%
  mutate(stock_id = paste(scientific_name, fao_id, sep = "-")) %>%
  select(rgn_id, year, stock_id, tonnes)


catch <- catch %>%
  filter(!is.na(rgn_id)) %>%
  filter(!is.na(fao_id)) %>%
  filter(rgn_id <= 250) %>%
  filter(rgn_id != 213) # drop Antarctica

```

## Subset the (v2022) catch data for our forage fish species

```{r}
## need to get TaxonKey's for each species to join with catch

# read in list we wrote out earlier
foragefish_list <- readr::read_csv(here::here(current_np_dir, "int", "master_taxa_list_SAUP.csv"))

# find missing species 
missing_spp <- foragefish_list %>%
  filter(is.na(TaxonKey))

missing_spp$forage_fish
# v2025 missing species:
#  [1] "Ariomma indica"          "Blicca bjoerkna"         "Centrolabrus exoletus"  
#  [4] "Citharidae"              "Hemiramphus balao"       "Liza saliens"           
#  [7] "Percarina demidoffi"     "Sardinops caeruleus"     "Sardinops melanostictus"
# [10] "Sardinops ocellatus

# test <- catch %>%
#   filter(str_detect(paste(missing_spp$forage_fish, collapse = "|"), species)) # just make sure the NA species are actually not in the SAUP data
# sort(unique(test$species)) # no full matches..
# # v2024: [1] "Hemiramphus" "Liza" 


# ---- Join forage fish species with catch data ----
forage_fish_taxa_list <- foragefish_list %>%
  # subset to only keep catch data for which we have forage species data
  left_join(catch, by = c("forage_fish" = "species", "TaxonKey")) %>%
  dplyr::select(forage_fish, inSAUP, TaxonKey) %>% # drop stock, common name, and extra catch df columns
  unique() %>% # return unique combinations of grouping variables in the data frame
  mutate(inSAUP = ifelse(
    # if TaxonKey is NA, set inSAUP to NA
   is.na(TaxonKey), NA, inSAUP)) 

# save a list of species not in SAUP data for bookkeeping
forage_fish_no_saup <- forage_fish_taxa_list %>%
  filter(is.na(TaxonKey)) 

# write out missing species list to int
readr::write_csv(forage_fish_no_saup, here(current_np_dir, "int", "forage_fish_not_in_SAUP.csv"))


# join catch data set with forage fish taxa list 
catch_fishfeed <- catch %>%
  left_join(forage_fish_taxa_list, by = c("TaxonKey")) %>%
  # subset to only include forage fish species
  dplyr::filter(!is.na(forage_fish))

nrow(catch_fishfeed) # 600864 - v2025

# write output to int folder in np/scenario year directory
readr::write_csv(catch_fishfeed, here::here(current_np_dir, "int", "saup_catch_forage_fish.csv"))

test <- catch_fishfeed %>%
  group_by(year) %>%
  summarise(sum = sum(tons)) 

summary(test)[4, 2]
## looks good.. ~30 million per year in v2023 and v2024 and v2025 
```

# Step 2

## Multiply by 0.90 to reflect the amount going to feed/oils, and not the amount going to human consumption (that is accounted for in food provision)

Justification from the [Froelich et al. 2018](https://www.nature.com/articles/s41893-018-0077-1#Sec11): "Currently, it is estimated about 10% of forage fish enter the human diet directly, but the notoriously tiny-boned fish are labour intensive (thus expensive) to process for human consumption, are the foundation of several industries and thus jobs (creating inertia to change) and are not the preferred fish type for most people."

```{r}
# read in SAUP catch data subsetted to forage fish species 
catch_fishfeed <- readr::read_csv(here::here(current_np_dir, "int", "saup_catch_forage_fish.csv"))

# Multiply by 0.90 to reflect the amount going to feed/oils, not human consumption
catch_non_human <- catch_fishfeed %>% 
  mutate(tons_non_human = tons * 0.9) 

# for each region, FAO ID, Taxon Key, stock ID, and year, 
# find sum of non-human-consumption fish harvest
catch_non_human <- catch_non_human %>%
  dplyr::select(year, rgn_id, fao_id, stock_id, TaxonKey, tons_non_human) %>%
  group_by(rgn_id, fao_id, TaxonKey, stock_id, year) %>%
  summarize(catch_non_human = sum(tons_non_human)) %>%
  ungroup()

test <- catch_non_human %>%
  group_by(year) %>%
  summarise(sum = sum(catch_non_human)) 

summary(test)[4, 2]
# ~ 26.6 mil per year -- v2024 and v2025

# --------- Zero-fill -----------------------

## these data have no zero catch values, so add years with no reported catch to data table:
catch_year_zeros <- catch_non_human %>%
  pivot_wider(names_from = year, values_from = catch_non_human,
              names_prefix = "x_",
              values_fill = 0) %>% 
  pivot_longer(cols = starts_with("x_"),
               names_to = "year",
               values_to = "catch_non_human") %>% 
  mutate(year = as.numeric(gsub("x_", "", year))) %>% 
  arrange(year)

## this part eliminates the zero catch values prior to the first reported non-zero catch  
catch_year_zeros <- catch_year_zeros %>%
  group_by(fao_id, TaxonKey, stock_id, rgn_id) %>%
  arrange(year) %>%
  mutate(cum_catch = cumsum(catch_non_human)) %>%
  filter(cum_catch > 0) %>%
  dplyr::select(-cum_catch) %>%
  ungroup()

# Calculate mean catch for ohi regions (using data from 1980 onward). These data are used to weight the RAM b/bmsy values 
# mean_catch <- catch_year_zeros %>%
#   filter(year >= 1980) %>%
#   group_by(rgn_id, fao_id, TaxonKey, stock_id) %>%
#   summarize(mean_catch_non_human = mean(catch_non_human, na.rm=TRUE)) %>% # mean catch for each stock (in a specific ohi-fao region)
#   filter(mean_catch_non_human != 0)  %>% # some stocks have no reported catch for time period
#   ungroup()

mean_catch <- catch_year_zeros %>%
  filter(year >= 1980) %>%
  group_by(rgn_id, fao_id, TaxonKey, stock_id) %>%
  mutate(mean_catch_non_human = mean(catch_non_human, na.rm=TRUE)) %>% # mean catch for each stock (in a specific ohi-fao region)
  filter(mean_catch_non_human != 0)  %>% # some stocks have no reported catch for time period
  ungroup()

mean_catch_dist <- mean_catch %>% 
  select(!catch_non_human) %>% 
  dplyr::distinct(rgn_id, fao_id, year, TaxonKey, stock_id, mean_catch_non_human)


options(scipen = 999) # to prevent taxonkey from turning into scientific notation

mean_catch_FOFM <- mean_catch_dist %>%
  mutate(stock_id_taxonkey = paste(stock_id, TaxonKey, sep="_")) %>%
  dplyr::select(rgn_id, stock_id_taxonkey, mean_catch_non_human, year) %>%
  # filter(year >= 2001) %>%  # filter to include only analysis years
  data.frame() 
  # rename(catch = mean_catch_non_human)

readr::write_csv(mean_catch_FOFM, here(current_np_dir, "int", "mean_catch_FOFM.csv"))

## Separate file to save the raw data as well as the mean catch data. Used in step 5 weighting
mean_raw_catch <- mean_catch %>% 
  mutate(stock_id_taxonkey = paste(stock_id, TaxonKey, sep="_"),
         product = "fish_oil") %>%
  dplyr::select(rgn_id, stock_id_taxonkey, mean_catch_non_human, catch_non_human, year, product) %>%
  filter(year >= 2001) %>%  # filter to include only analysis years
  data.frame() 

readr::write_csv(mean_raw_catch, here(current_np_dir, "int", "raw_mean_catch_FOFM.csv"))

```



# B/Bmsy

Convert the B/Bmsy values to scores (this is done in functions.R for FIS subgoal, but we do it in the dataprep for NP), apply the underharvest penalty, and take a catch weighted average of B/Bmsy scores for each region/year.

```{r}

fis_bbmsy <- read_csv(here(current_fis_dir, "output", "fis_bbmsy.csv")) %>%
  select(stock_id, rgn_id, year, bbmsy)


b <- fis_bbmsy %>%
  dplyr::mutate(bbmsy = ifelse(bbmsy > 1 , 1, bbmsy)) %>%
  dplyr::mutate(bbmsy = as.numeric(bbmsy),
                year = as.numeric(as.character(year)),
                stock_id = as.character(stock_id)) %>%
  dplyr::mutate(stock_id = str_replace(stock_id, "_", " "))


####
# ==== Calculate scores for Bbmsy values ==== ####
#  *************NOTE *****************************
#  These values can be altered
#  ***********************************************

## The upper limit can be deleted, since we already cap scores at 1 anyways. We still want to keep the lower buffer.

alpha <- 0.5
beta <- 0.25
lowerBuffer <- 0.95
upperBuffer <- 1.05

# Rename to stock score

b$stock_score = ifelse(
  b$bbmsy < lowerBuffer,
  b$bbmsy,
  ifelse (b$bbmsy >= lowerBuffer &
            b$bbmsy <= upperBuffer, 1, NA)
)
b$stock_score = ifelse(!is.na(b$stock_score),
                 b$stock_score,
                 ifelse(
                   1 - alpha * (b$bbmsy - upperBuffer) > beta,
                   1 - alpha * (b$bbmsy - upperBuffer),
                   beta
                 ))

```

# Merge B/Bmsy and catch data

```{r}
data_fis <- catch %>%
  dplyr::left_join(b, by = c('stock_id', 'year', 'rgn_id')) 

setdiff(catch$stock_id, b$stock_id) ## need to do a lot of gapfilling here.
setdiff(b$stock_id, catch$stock_id)
intersect(b$stock_id, catch$stock_id) 

tmp <- data_fis %>%
 group_by(rgn_id, year) %>%
 summarize(total_bbmsy = sum(!is.na(stock_score)),
           total = n())

###
#  Estimate scores for taxa without b/bmsy values --------------------
# Mean score of other fish in the region is the starting point
# Then a penalty is applied based on the level the taxa are reported at
###

## this takes the mean score within each region and year
data_fis_gf <- data_fis %>%
  dplyr::group_by(rgn_id, year) %>%
  dplyr::mutate(mean_score = mean(stock_score, na.rm = TRUE)) %>%
  dplyr::ungroup()

## this takes the mean score across all regions within a year
# (when no stocks have scores within a region)
data_fis_gf <- data_fis_gf %>%
  dplyr::group_by(year) %>%
  dplyr::mutate(mean_score_global = mean(stock_score, na.rm = TRUE)) %>%
  dplyr::ungroup() 

data_fis_gf <- data_fis_gf %>%
  mutate(stock_score = ifelse(is.na(stock_score), mean_score, stock_score)) %>%
  mutate(stock_score = ifelse(is.na(stock_score), mean_score_global, stock_score))


## get tonnes per region for weighting
tonnes <- data_fis_gf %>%
  group_by(year, rgn_id) %>%
  summarize(tonnes = sum(tonnes, na.rm=TRUE))
  mutate(product = "fish_oil")
write_csv(tonnes, here(current_np_dir, "int", "tonnes_fofm.csv"))
```

## Calculate score

```{r}

# We use the mean catch for each region/stock to weight the yearly B/Bmsy values.

score_data <- data_fis_gf %>%
  group_by(rgn_id, stock_id) %>%
  mutate(mean_tonnes = mean(tonnes, na.rm=TRUE)) %>%
  filter(!is.nan(stock_score)) %>% # only have stock scores from 1990 in these data.
  ungroup()
# Calculate status for each region
###

## Take a catch weighted average of B/Bmsy scores for each region/year.


score_data <- score_data %>%
  dplyr::group_by(rgn_id, year) %>%
  dplyr::summarize(score = weighted.mean(stock_score, mean_tonnes)) %>%
  dplyr::ungroup() %>% 
  mutate(product = "fish_oil")

summary(score_data)

readr::write_csv(score_data, here(current_np_dir, "output", "np_fofm.csv"))

old <- read_csv(here(current_np_dir, "output", "np_fofm_scores.csv"))

new <- read_csv(here(current_np_dir, "output", "np_fofm.csv"))

check <- new %>%
  rename("new_score" = "score") %>%
  left_join(old, by = c("rgn_id", "year")) %>%
  select(rgn_id, year, new_score, score) %>%
  mutate(diff = new_score - score) %>%
  filter(year == 2019)

ab_scores <- ggplot() + 
  geom_abline(slope = 1, intercept = 0, col = "red") +
  geom_point(data = check, aes(x = score, y = new_score)) +
  labs(x = "Old Score", 
       y = "New Score",
       title = "FOFM New score vs old score for 2019") +
  theme_bw();ggplotly(ab_scores)
```

