---
title: 'OHI: Clean Waters/Trash trend'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

# Summary

The coastal & marine trash trend score is calculated based on the most recently available data sources on a few different forms of marine trash: marine macroplastics/debris, marine microplastics, and coastal debris. 

Different data sources were investigated and compiled in v2025, and information on them was combined here to create an overall trash trend score analysis.

# Microplatics

Source: https://www.ncei.noaa.gov/products/microplastics

There is microplastics data from NOAA's National Center for Environmental Information 'NCEI Marine Microplastics product'. The data is available as a csv, shapefiles, GeoJSON, JSON, FGDB, or KML. The data on microplastic concentrations contains measurements in pieces/cubic meter,and were ranked on a scale of 'Very Low', 'Low', 'Medium', and 'High'. 

Observations were not specified at the country level, but merely grouped by ocean, or in some cases subregion. However, the point observations could be rasterized and grouped into the eez boundaries of each country. The concentration values could be rescaled for a 0-1 range. 



```{r setup, message = FALSE, warning = FALSE, verbose = FALSE}
#set options for all chunks in code
knitr::opts_chunk$set(message = FALSE, warning = FALSE, eval=FALSE)

# load packages
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  here,
  janitor,
  raster,
  terra,
  sf,
  stars,
  maps,
  httr,
  jsonlite,
  purrr,
  tictoc,
  sf, 
  googleVis,
  RColorBrewer,
  foreach,
  doParallel, # for using multiple cores, if needed
  tidyverse, 
  ohicore,
  ggplot2,
  ncmeta,
  ncdf4,
  plotly,
  zoo,
 measurements
)

# ---- sources! ----
source(here("workflow", "R", "common.R")) # file creates objects to process data

# ---- set year and file path info ----
current_year <- 2025 # Update this in the future!!
version_year <- paste0("v",current_year)
data_dir_version_year <- paste0("d", current_year)

# ---- data directories ----

# Raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")

# prs_oa dir
cw_dir <- here::here(dir_M, "git-annex", "globalprep", "cw_trend_trash", version_year)

# output data dir for intermediate data products
int_dir <- here::here(cw_dir, "int")
# dir.create(int_dir) # to create the path on Mazu if it has not already been done

# final output dir
output_dir <- here::here("globalprep","cw_trend_trash", version_year, "output")

# set colors
cols = rev(colorRampPalette(brewer.pal(9, 'Spectral'))(255)) # rainbow color scheme
```


# Method 1: Use vector data and intersect for each region

```{r}
# Load in the microplastics data from the Mazu server
microplastics_data <- st_read(here::here(raw_data_dir , "NOAA_NEI_Marine_Microplastics", "Microplastic_Data_MasterList_GISTeam_XYTableToPoint.shp"))
```
# Rescale the microplastics data on a 0-1 scale

```{r}
# Check the microplastics columns
colnames(microplastics_data)
```

```{r}
# Create dataframes for each sample setting
ocean_subsurface <- microplastics_data %>%
  filter(str_detect(microplastics_data$Symbology, "Ocean Subsurface"))

ocean_surface <- microplastics_data %>%
  filter(str_detect(microplastics_data$Symbology, "Ocean Surface"))

ocean_sediment <- microplastics_data %>%
  filter(str_detect(microplastics_data$Symbology, "Ocean Sediment"))

beach <- microplastics_data %>%
  filter(str_detect(microplastics_data$Symbology, "Beach"))

nurdle_patrol <- microplastics_data %>%
  filter(str_detect(microplastics_data$Symbology, "Nurdle Patrol"))
```

```{r}
# Create function to rescale each df separately based on their 95th percentile value
rescale_95 <- function(x) {
  x_min <- min(x)
  x_95 <- quantile(x, 0.95)
  x_scaled <- (x - x_min) / (x_95 - x_min)
  x_scaled[x_scaled > 1] <- 1
  return(x_scaled)
}

# Rescale all the dfs
ocean_subsurface$concentration_rescaled <- rescale_95(ocean_subsurface$Microplast)
ocean_surface$concentration_rescaled <- rescale_95(ocean_surface$Microplast)
ocean_sediment$concentration_rescaled <- rescale_95(ocean_sediment$Microplast)
beach$concentration_rescaled <- rescale_95(beach$Microplast)
nurdle_patrol$concentration_rescaled <- rescale_95(nurdle_patrol$Microplast)

# Create a column for setting
ocean_subsurface <- ocean_subsurface %>%
  mutate(Setting = "Subsurface")

ocean_surface <- ocean_surface %>%
  mutate(Setting = "Surface")

ocean_sediment <- ocean_sediment %>%
  mutate(Setting = "Sediment")

beach <- beach %>%
  mutate(Setting = "Beach")

nurdle_patrol <- nurdle_patrol %>%
  mutate(Setting = "Nurdle Patrol")

# Concatenate all the df together
microplastics_rescaled <- rbind(ocean_subsurface,
                                ocean_surface,
                                ocean_sediment,
                                beach,
                                nurdle_patrol)
```

# Rasterize the Microplastics observations for masking

```{r}
# Rasterize the microplastic observations:
# Isolate points from the data set
points <- st_as_sf(microplastics_rescaled, coords = c("Longitude_", "Latitude_"), remove = FALSE, crs = 4326) 

# Get the bounding box (extent) of the points
microplastics_bbox <- ext(points)

# Expand slightly to ensure full 1-degree cell coverage
microplastics_bbox <- extend(microplastics_bbox, 0.5)  # adds 0.5 degree padding on all sides

# Create a raster with 1-degree resolution
rast_template <- rast(extent = microplastics_bbox, 
                   resolution = 1, 
                   crs = "EPSG:4326")

# Rasterize using mean value per cell (you can also use "sum", "max", etc.)
microplastics_rast <- rasterize(points, rast_template, field = "concentration_rescaled", fun = mean)
```

```{r}
# Plot the raster
plot(microplastics_rast)
```

# Get the eez regions for each country in OHI

```{r}
# Grab regional data from OHI:
# Get the region boundaries for all countries we will calculate scores for
ohi_rasters()
```

```{r}
# Check to see if crs are matched
# Match the crs of the data
microplastics_rast <- terra::project(microplastics_rast, crs(zones))

# Match the extent of the data by resampling
zones_aligned <- resample(zones, microplastics_rast, method = "near")

# Check to see if everything matches
crs(microplastics_rast) == crs(zones_aligned) #v2025: TRUE
ext(microplastics_rast) == ext(zones_aligned) # v2025: TRUE
res(microplastics_rast) == res(zones_aligned) # v2025: TRUE
```

# Find observations that overlap with eez regions

```{r}
# Use zonal to compute mean of point values per zone
zonal_stats <- zonal(microplastics_rast, zones_aligned, fun = "mean", na.rm = TRUE)
```

```{r}
# Check for NAs
sum(is.na(zonal_stats$mean))

# Subset the rows with NAs in the mean column for zonal_stats to view which countries are missing
missing_countries <- zonal_stats %>%
  filter(is.na(mean) == TRUE) %>%
  filter(regions_eez_with_fao_ant <= 277) # Remove zones for continents and subregions
```

# Potential Gapfilling: 


```{r}
# Rename the columns for ease of use
zonal_stats_clean <- zonal_stats %>% 
  rename_at('mean', ~'avg_concentration') %>%
  rename_at('regions_eez_with_fao_ant', ~'region')

# Get the mean of avg_concentration
mean_conc <- mean(zonal_stats_clean$avg_concentration, na.rm= TRUE)

# Gapfill the 59 missing values with the global average concentration
zonal_stats_filled <- zonal_stats_clean %>%
   mutate(avg_concentration = case_when(
    is.na(avg_concentration) ~ mean_conc,
    TRUE ~ avg_concentration)) # Return values that are present; do not replace with NA

# Add in the region names
microplastics_eez <- left_join(rgns_all, zonal_stats_filled, by= c("rgn_ant_id" = "region"))
```

# Plot the results

```{r}
# Find missing values in the zonal_statistics
plot <- ggplot(microplastics_eez, aes(x= avg_concentration, y = rgn_id, colour = rgn_name)) +
  geom_point() +
  labs(title= "Microplastics Concentration by Region", 
       x= "Microplastics Concentration (pieces/m^3)",
y= "Region ID")

ggplotly(plot)
```


