---
title: 'Set up IUCN spatial data'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

library(dplyr)
library(raster)
library(here)
library(ggplot2)
library(sf)
library(fasterize)
library(stringr)
library(mapview)
library(terra)

source(here('workflow/R/common.R'))

goal     <- 'spp'
scenario <- 'v2025'
dir_goal <- here('globalprep', goal, scenario)

### goal specific folders and info
dir_anx  <- file.path(dir_M, 'git-annex/globalprep')
dir_setup   <- file.path(dir_goal, '_setup')
dir_goal_anx <- file.path(dir_anx, goal, scenario, 'spp_risk_dists')
dir_raw_data_anx <- file.path(dir_anx, '_raw_data')

source(file.path(dir_setup, 'common_fxns.R'))
```

# Summary

Examine IUCN Red List regional assessments for extinction risk categories.

To define spatial areas more accurately, create rasters of WDPA and ocean area at Gall-Peters resolution of 0.25 km^2. These can then be aggregated back to Gall Peters 10 km^2 working rasters, summing the total ocean area and total protected area within each working cell.

* Ocean area - use Natural Earth 10 m land polygons; mask over cell ID raster to exclude land.

# Data Sources

## MEOW
**Reference**: Spalding, Mark D. et al. 2007. “Marine Ecoregions of the World: A Bioregionalization of Coastal and Shelf Areas.” BioScience 57(7): 573–83. [URL](https://www.worldwildlife.org/publications/marine-ecoregions-of-the-world-a-bioregionalization-of-coastal-and-shelf-areas)

**Downloaded**: May 08, 2019

**Description**:  Shapefile of Marine Ecoregions

**Format**:  Shapefile

**File location**: `Mazu:git-annex/globalprep/_raw_data/wdpa_mpa/d2025/WDPA_Jun2025_Public_shp/`

## WDPA
**Reference**: IUCN and UNEP-WCMC (2025), The World Database on Protected Areas (WDPA) [On-line], June 2025. Cambridge, UK: UNEP-WCMC. Available at: www.protectedplanet.net.

**Downloaded**: June 22, 2025

**Description**:  Shapefile of World Database on Protected Areas

**Time range**: 1800 - 2025; some protected areas do not have an associated "status year" and are reported as year 0.

**Format**:  Shapefile

**File location**: `Mazu:git-annex/globalprep/_raw_data/wdpa_mpa/d2025/WDPA_Jun2025_Public_shp/`

# Methods

## Update geographic scope of regional assessments

The following descriptions may be useful for assigning MEOW and LME regions:

**Europe ('europe'):**

from European marine fishes: 

> The geographic scope of the European Red List assessment
encompasses the Mediterranean Sea, the Black Sea, the
Baltic Sea, the North Sea and/or the European part of
the Atlantic Ocean (i.e., the territorial waters and the
Exclusive Economic Zones (EEZs) of all European
countries in the eastern part of the Atlantic Ocean,
also including the EEZs of the Macaronesian islands
belonging to Portugal and Spain), the North Sea and
the Northeastern Atlantic Ocean, excluding the EEZs of
Greenland, Morocco and Western Sahara.

Options: 

* LME 20-26 (26 = Mediterranean) and 59 (Iceland)?
* MEOW 2 (Northern Iceland/Svalbard), [21:27, 30:31, 34:36] (Mediterranean), 44 (Black Sea)
* Alternately: Use EEZs for all countries listed as Europe by UN classification

**Northern Africa**

* LME 33 (Red Sea), 26 (Mediterranean), 27 (West Coast of North Africa)
* MEOW 87-88 (Red Sea), [32, 33, 35, 36] (Mediterranean), 28, 29 (Lusitanean upwelling)

**Southern Africa**

* LME 29 (Western Southern Africa) and 30 (Madagascar)
* MEOW 192:194 (Temperate southern Africa)

**Eastern Africa** 

* LME 30 (Madagascar) and 31 (Horn of Africa)
* MEOW 94:102 (Western Indian Ocean)

**Western Africa**   

* LME 28
* MEOW 79:86 (West African Transition, Gulf of Guinea)

**Central Africa**  

* LME 28, 29
* If this is truly "Central" then no marine anyway! ignore

**Northeastern Africa** 

* LME 32, 33
* MEOW 87:93 (Red Sea, Gulf of Aden, Somali/Arabian)

**Pan-Africa** 

* LME 26-33
* MEOW 28, 29, 32, 33, 35, 36 (Med and N), 
    79:86 (West), 87:89, 93 (NE), 94:102 (E), 192:194 (S)

**Gulf of Mexico** 

* LME 5
* MEOW 43 (N Gulf), 68:70 (S Gulf)

**Mediterranean**

* LME 26 and 62 for Mediterranean and Black Seas
* MEOW 30:36, 44

**Persian Gulf**

* LME 32 (NOTE: this encompasses the entire Arabian Sea)
* MEOW 90:91

### Spatial: `iucn_rgn_to_meow.csv`

```{r}
# Load region risk csv to view regions
rgn_risk_ts <- read_csv(file.path(dir_goal, '_data',
                                 sprintf('iucn_risk_ts_rgn_%s.csv', api_version)), show_col_types = FALSE)

# View regions
unique(rgn_risk_ts$iucn_rgn)

# Assign MEOWS to regions
rgn_to_meow <- c('europe'             = paste(c(2,21:27, 30:31, 34:36, 44), 
                                              collapse = ','), #updated 2025
                'gulf_of_mexico'      = paste0(c(43, 69:70), 
                                               collapse = ','), #updated 2025
                'pan-africa'          = paste0(c(28:29, 32:33, 35:36, 79:89, 93:102, 190:194), 
                                               collapse = ','), #updated 2025
                'mediterranean'       = paste0(c(30:36, 44), 
                                               collapse = ','), #updated 2025
                'arabian_sea'         = paste0(c(89, 91:92, 103, 105:106), 
                                               collapse = ','), #updated 2025
                'caribbean'           = paste0(c(63:68), 
                                               collapse = ","), #updated 2025
                'western_africa'      = paste0(79:83, 
                                               collapse = ','), #updated 2025
                'central_africa'      = paste0(c(81:85), 
                                               collapse = ','), #updated 2025
                'persian_gulf'        = paste0(c(90:91), 
                                               collapse = ','), #updated 2025
                'global'              = paste0(c(1:232), 
                                               collapse = ','))

# Create data frame for layer
rgn_to_meow_df <- data.frame(iucn_rgn = names(rgn_to_meow),
                             meow_id  = rgn_to_meow) %>%
  mutate(meow_id = str_split(meow_id, ',')) %>%
  unnest(meow_id) %>%
  group_by(iucn_rgn) %>%
  mutate(meow_id = as.integer(meow_id),
         priority = n()) %>%
  arrange(priority)

# Write csv
write_csv(rgn_to_meow_df, file.path(dir_goal, sprintf('_spatial/iucn_rgn_to_meow.csv')))

# Check
if(setdiff(unique(rgn_to_meow_df$iucn_rgn), unique(rgn_risk_ts$iucn_rgn)) != "global") {
  warning("Extra regions not accounted for in MEOW assignments.")
}
```

## WDPA

- Classification 
- Year of protection
- Proportion of protection

Copy over the following 10 km^2 Gall Peters projections from last year:
- ocean area raster
- cell ID raster
- ocean 500m raster

General steps:

1. Project Natural Earth 10M ocean polygons into Gall-Peters equal area projection
2. Rasterize at 500 m resolution
3. Aggregate up to 10 km resolution

### MPA Classification

Marine regions are any protected area with MARINE flag of 1 or 2, or non-zero GIS_M_AREA.  Classify by IUCN category (I-VI -> 1-6; non-classified get 8).  Include only designated, adopted, and established parks.  Omit non-MPA management plans from U.S.  Rasterize to CHI resolution, calculate protected cells vs total cells to get percent of protection for each IUCN protection category.

Because we are rasterizing the protected areas down to 500 m resolution, we risk dropping MPAs smaller than about 1 km^2. We assume these are not likely to have a significant impact on biodiversity protection, but see the sensitivity analysis.

#### Git-annex: `wdpa_mar_by_cat_Jun2025_500m.tif`

```{r}
# Copy from last year (already in _spatial folder if copied previous folder)
ocean_area_file <- file.path(dir_spatial, 'ocean_area_rast.tif')
cell_id_file <- file.path(dir_spatial, 'cell_id_rast.tif')
ocean_500m_file  <- file.path(dir_anx, 'spp', 'v2024', 'spp_risk_dists', 'spatial', 'ocean_500m.tif')

# Read in rasters
rast_cell_id <- rast(cell_id_file)
rast_ocean_500m <- rast(ocean_500m_file)

# Goal files
wdpa_marine_file <- file.path(dir_raw_data_anx, 'wdpa_mpa/d2025/wdpa_mar_by_cat_Jun2025_500m.tif')
wdpa_marine_file_int <- file.path(dir_raw_data_anx, 'wdpa_mpa/d2025/wdpa_mar_by_cat_Jun2025_500m_int.tif')

# File paths to WDPA files (change to current year)
wdpa_poly_file_0   <- file.path(dir_raw_data_anx, 'wdpa_mpa/d2025/WDPA_Jun2025_Public_shp/WDPA_Jun2025_Public_shp_0', 
                              'WDPA_Jun2025_Public_shp-polygons.shp')
wdpa_poly_file_1   <- file.path(dir_raw_data_anx, 'wdpa_mpa/d2025/WDPA_Jun2025_Public_shp/WDPA_Jun2025_Public_shp_1', 
                              'WDPA_Jun2025_Public_shp-polygons.shp')
wdpa_poly_file_2   <- file.path(dir_raw_data_anx, 'wdpa_mpa/d2025/WDPA_Jun2025_Public_shp/WDPA_Jun2025_Public_shp_2', 
                              'WDPA_Jun2025_Public_shp-polygons.shp')

# Read in WDPA files
wdpa_poly_0 <- sf::st_read(wdpa_poly_file_0)
wdpa_poly_1 <- sf::st_read(wdpa_poly_file_1)
wdpa_poly_2 <- sf::st_read(wdpa_poly_file_2)

wdpa_poly <- rbind(wdpa_poly_0, wdpa_poly_1, wdpa_poly_2)

#wdpa_poly$IUCN_CAT %>% unique() 
##only grab ones that are reported/assigned/applicable
iucn_cats <- c('Ia'  = 1,
               'Ib'  = 1,
               'II'  = 2,
               'III' = 3,
               'IV'  = 4,
               'V'   = 5,
               'VI'  = 6)
wdpa_poly$STATUS %>% table()
# Adopted   Designated  Established    Inscribed Not Reported     Proposed 
#      34       254937         2401          250           34         1219 
# wdpa_poly$NO_TAKE %>% unique()
# x <- wdpa_poly %>% filter(NO_TAKE %in% c('All', 'Part'))
# x <- wdpa_poly %>% filter(WDPAID == 309888)

wdpa_marine <- wdpa_poly %>%
  filter(MARINE > 0 | GIS_M_AREA > 0) %>%
  filter(STATUS %in% c('Designated', 'Adopted', 'Established')) %>%
  ### no paper parks!
  filter(!str_detect(tolower(MANG_PLAN), 'non-mpa')) %>%
  ### omit non-MPA fisheries or species management plans!
  mutate(NO_TAKE = ifelse(WDPAID == 309888, TRUE, NO_TAKE),
         ### patch PIPA to be no take - currently "not reported"
         no_take = (NO_TAKE == 'All') | (NO_TAKE == 'Part' & NO_TK_AREA > 0.75 * GIS_M_AREA),
         ### if NO_TK_AREA is 75% or more of GIS area, count it...
         cat = iucn_cats[IUCN_CAT],
         cat = ifelse(no_take & !cat %in% 1:2, -1, cat), ### use -1 as a "no take" flag
         cat = ifelse(is.na(cat), 8, cat)) %>%           ### use 8 as an "other protected" flag
  arrange(cat) %>%
  st_transform(crs = gp_proj4)

wdpa_marine_vect <- vect(wdpa_marine)

# Free up space
rm(wdpa_marine)
rm(wdpa_poly_0)
rm(wdpa_poly_1)
rm(wdpa_poly_2)
rm(wdpa_poly)

# Rasterize all protected areas to 500m global raster;
# Use min value: will prioritize no-take and higher levels of protection
terra::rasterize(x = wdpa_marine_vect, 
                 y = rast_ocean_500m,
                 field  = 'cat', 
                 fun    = 'min', 
                 filename = wdpa_marine_file_int,
                 overwrite = TRUE)   

wdpa_all_rast_500m <- rast(wdpa_marine_file_int)

#plot(wdpa_all_rast_500m)

# Free up space
rm(wdpa_marine_vect)

rast_ocean_500m <- terra::classify(rast_ocean_500m ,cbind(-Inf, 0.005, NA), right=FALSE)

# plot(rast_ocean_500m)

wdpa_marine_rast_500m <- terra::mask(wdpa_all_rast_500m, rast_ocean_500m)

# plot(wdpa_marine_rast_500m)
# plot(rast(file.path(dir_raw_data_anx, 'wdpa_mpa/d2021/wdpa_mar_by_cat_Feb2021_500m.tif')))

writeRaster(wdpa_marine_rast_500m, wdpa_marine_file, 
            overwrite = TRUE)
```

#### Git-annex: `wdpa_mar_by_prot_yr_Jun2025_500m.tif`

```{r}
# Read in 500m raster
rast_ocean_500m <- rast(ocean_500m_file)
#plot(rast_ocean_500m) #check to see if loaded properly 

# Goal file
wdpa_prot_yr_file <- file.path(dir_raw_data_anx, 'wdpa_mpa/d2025', 
                               'wdpa_mar_by_prot_yr_Jun2025_500m.tif')

wdpa_prot_yr_file_int <- file.path(dir_raw_data_anx, 'wdpa_mpa/d2025/wdpa_mar_by_prot_yr_Jun2025_500m_int.tif')

# Existing files
wdpa_poly_file_0   <- file.path(dir_raw_data_anx, 'wdpa_mpa/d2025/WDPA_Jun2025_Public_shp/WDPA_Jun2025_Public_shp_0', 
                              'WDPA_Jun2025_Public_shp-polygons.shp')
wdpa_poly_file_1   <- file.path(dir_raw_data_anx, 'wdpa_mpa/d2025/WDPA_Jun2025_Public_shp/WDPA_Jun2025_Public_shp_1', 
                              'WDPA_Jun2025_Public_shp-polygons.shp')
wdpa_poly_file_2   <- file.path(dir_raw_data_anx, 'wdpa_mpa/d2025/WDPA_Jun2025_Public_shp/WDPA_Jun2025_Public_shp_2',
                                'WDPA_Jun2025_Public_shp-polygons.shp')

wdpa_poly_0 <- sf::st_read(wdpa_poly_file_0)
wdpa_poly_1 <- sf::st_read(wdpa_poly_file_1)
wdpa_poly_2 <- sf::st_read(wdpa_poly_file_2)

wdpa_poly <- rbind(wdpa_poly_0, wdpa_poly_1, wdpa_poly_2)

#wdpa_poly$IUCN_CAT %>% unique()
iucn_cats <- c('Ia'  = 1,
               'Ib'  = 1,
               'II'  = 2,
               'III' = 3,
               'IV'  = 4,
               'V'   = 5,
               'VI'  = 6)
#wdpa_poly$STATUS %>% table()
# Adopted   Designated  Established    Inscribed Not Reported     Proposed 
#  34       242940         1148          249           34         1201 
# wdpa_poly$NO_TAKE %>% unique()
# x <- wdpa_poly %>% filter(NO_TAKE %in% c('All', 'Part'))

wdpa_no_take <- wdpa_poly %>%
  filter(MARINE > 0 | GIS_M_AREA > 0) %>%
  filter(STATUS %in% c('Designated', 'Adopted', 'Established')) %>%
  ### no paper parks!
  filter(!str_detect(tolower(MANG_PLAN), 'non-mpa')) %>%
  ### omit non-MPA fisheries or species management plans!
  mutate(NO_TAKE = ifelse(WDPAID == 309888, TRUE, NO_TAKE),
         ### patch PIPA to be no take - currently "not reported"
         no_take = (NO_TAKE == 'All') | (NO_TAKE == 'Part' & NO_TK_AREA > 0.75 * GIS_M_AREA),
         ### if NO_TK_AREA is 75% or more of GIS area, count it...
         cat = iucn_cats[IUCN_CAT],
         cat = ifelse(no_take & !cat %in% 1:2, -1, cat), ### use -1 as a "no take" flag
         cat = ifelse(is.na(cat), 8, cat)) %>%           ### use 8 as an "other protected" flag
  arrange(cat) %>%
  filter(cat <= 2) %>% ### this limits to categories Ia, Ib, II, and no-take
  st_transform(crs = gp_proj4)

rm(wdpa_poly_0 , wdpa_poly_1, wdpa_poly_2, wdpa_poly)

wdpa_no_take_vect <- vect(wdpa_no_take)

rm(wdpa_no_take, wdpa_marine_rast_500m)

# Rasterize all protected areas to 500 m global raster;
# use min value: will prioritize no-take and higher levels of protection
wdpa_all_rast_500m <-   terra::rasterize(x = wdpa_no_take_vect, 
                                         y = rast_ocean_500m,
                                         field  = 'STATUS_YR', 
                                         fun    = 'min',
                                         overwrite = TRUE)

writeRaster(wdpa_all_rast_500m, wdpa_prot_yr_file_int, overwrite = TRUE)

rm(wdpa_no_take_vect)

rast_ocean_500m <- terra::classify(rast_ocean_500m ,cbind(-Inf, 0.005, NA), right=FALSE)

### in case of floating point not-quite-zeros...

wdpa_all_rast_500m <- rast(wdpa_prot_yr_file_int)

terra::mask(wdpa_all_rast_500m, 
            rast_ocean_500m, 
            filename = wdpa_prot_yr_file,
            overwrite = TRUE)


# wdpa_yr_rast_500m <- rast(wdpa_prot_yr_file)
#plot(wdpa_yr_rast_500m)
#plot(raster(file.path(dir_raw_data_anx, 'wdpa_mpa/d2020/wdpa_mar_by_prot_yr_Apr2020_500m.tif')))

# writeRaster(wdpa_yr_rast_500m, wdpa_prot_yr_file, 
#             overwrite = TRUE)
# 
#}

# View
plot(wdpa_all_rast_500m)
```

### MPA Protected Area

Crosstab WDPA map to cell IDs

Be cautious when using `raster::crosstab()` as it returns factors that need to be coerced into integers... This code results in a file of cell IDs, IUCN protected area categories (1 = Ia and Ib, 2-6 = II - VI, 7 = no take, -1 = other), and proportion of cell protected by that category.

We can't simply aggregate upwards here, since we need to keep track of different protection levels (for the first) and year (for the second).

#### Spatial: `wdpa_mpa_area.csv`

Area of protection (to calculate proportion of protection).

``` {r crosstab mpa areas}

wdpa_marine_rast_500m <- rast(wdpa_marine_file)
rast_cell_id    <- rast(cell_id_file)    ### cell IDs at 10 km^2
rast_ocean_500m <- rast(ocean_500m_file)  ### ocean presence at 1 km^2
rast_ocean_area <- rast(ocean_area_file) ### ocean proportional area at 10 km^2

cell_id_500m_file <- file.path(dir_goal_anx, 'spatial/cell_id_500m.tif')

### Need a 500 m raster of cell IDs for crosstabbing...
rast_cell_id_500m <- terra::resample(x = rast_cell_id, y = rast_ocean_500m,
                                     filename = cell_id_500m_file,
                                     method = 'near',
                                     overwrite = TRUE)

#plot(rast_cell_id_500m)
#plot(raster(file.path("/home/shares/ohi/git-annex/globalprep/spp/v2020/spp_risk_dists/spatial/cell_id_500m.tif")))

# Goal file
wdpa_cell_area_file <- file.path(dir_spatial, 'wdpa_mpa_area.csv')

### now break into chunks and mclapply across them to crosstab WDPA categories to cell IDs

### Break into smaller chunks for faster processing and eventual parallelizing
n_chunks <- 60

crosstab_chunk <- function(rast1, rast2, chunk, n_chunks) {
  ## chunk <- 2
  # rast1 = wdpa_marine_rast_500m
  # rast2 = rast_cell_id_500m
  
  rast1 <- raster(rast1)
  rast2 <- raster(rast2)
  
  chunk_size <- ceiling(ncol(rast1) / n_chunks)
  left_bound <- (chunk - 1) * chunk_size + 1
  right_bound <- min(left_bound + chunk_size - 1, ncol(rast1))
  
  # chunk_ext <- terra::ext(xFromCol(rast1, left_bound), xFromCol(rast1, right_bound), yFromRow(rast1, nrow(rast1)), yFromRow(rast1, 1))
  
  chunk_ext <- extent(rast1, 1, nrow(rast1), left_bound, right_bound)
  
  message('Processing ', chunk, ': ', paste0(as.character(round(chunk_ext)), collapse = ', '))
  wdpa_chunk    <- raster::crop(rast1, chunk_ext)
  cell_id_chunk <- raster::crop(rast2, chunk_ext)
  
  wdpa_cells <- raster::crosstab(wdpa_chunk, cell_id_chunk, 
                                 long = TRUE) %>%
    setNames(c('wdpa_category', 'cell_id', 'n_prot')) %>%
    mutate(wdpa_category = as.integer(as.character(wdpa_category)),
           cell_id       = as.integer(as.character(cell_id)))
  ### wdpa_category and cell_id are crosstabbed as factors - first
  ### convert to character (to unfactorize it) then to integer.  Otherwise
  ### you end up with factor index, not actual cell ID or category.
  
  return(wdpa_cells)
}

### Use the function in an mclapply call
system.time({
  wdpa_cells_list <- parallel::mclapply(1:n_chunks, mc.cores = 12,
                                        FUN = function(x) crosstab_chunk(wdpa_marine_rast_500m, 
                                                                         rast_cell_id_500m, 
                                                                         chunk = x, n_chunks))
  
#    user   system  elapsed 
# 5041.803  971.565  553.962 
})

wdpa_cells_df <- bind_rows(wdpa_cells_list)

area_df <- data.frame(cell_id  = values(rast_cell_id),
                      area_km2 = values(rast_ocean_area)) %>%
  rename(cell_id = 1, area_km2 = 2)

wdpa_area_df <- wdpa_cells_df %>%
  full_join(area_df, by = 'cell_id') %>%
  filter(!is.na(wdpa_category)) %>%
  mutate(prot_area_km2 = n_prot / 4, ### count of cells in 500 x 500 m raster, div by 4
         mpa_pct = prot_area_km2 / area_km2) %>%
  dplyr::select(wdpa_category, cell_id, mpa_pct, prot_area_km2) %>%
  ### mpa pct is relative to *ocean* area - not *total cell* area
  group_by(wdpa_category, cell_id) %>%
  summarize(mpa_pct = sum(mpa_pct), prot_area_km2 = sum(prot_area_km2)) %>%
  ungroup()

# Last year's file
old <- read_csv(file.path(here(), "globalprep/spp/v2024/_spatial/wdpa_mpa_area.csv"), show_col_types = FALSE)

sum(old$prot_area_km2) 
# 2023: 26841457
# 2024: 24690672

sum(wdpa_area_df$prot_area_km2) 
# 2024: 24690672 
# 2025: 30000565

# Write csv
write_csv(wdpa_area_df, wdpa_cell_area_file) 
```

### MPA Year of Protection

#### Spatial: `wdpa_mpa_year.csv` 

Year of protection. 

``` {r crosstab mpa years}
wdpa_yr_rast_500m <- raster(wdpa_prot_yr_file)
rast_cell_id    <- raster(cell_id_file)    ### cell IDs at 10 km^2
rast_ocean_500m  <- raster(ocean_500m_file)  ### ocean presence at 1 km^2
rast_ocean_area <- raster(ocean_area_file) ### ocean proportional area at 10 km^2

cell_id_500m_file <- file.path(dir_goal_anx, 'spatial/cell_id_500m.tif')
rast_cell_id_500m <- raster(cell_id_500m_file)

# Goal file
wdpa_cell_year_file <- file.path(dir_spatial, 'wdpa_mpa_year.csv')

### now break into chunks and mclapply across them to crosstab WDPA categories to cell IDs
### Break into smaller chunks for faster processing and eventual parallelizing
n_chunks <- 60

crosstab_chunk2 <- function(rast1, rast2, chunk, n_chunks) {
  ### chunk <- 30
  chunk_size <- ceiling(ncol(rast1) / n_chunks)
  left_bound <- (chunk - 1) * chunk_size + 1
  right_bound <- min(left_bound + chunk_size - 1, ncol(rast1))
  
  chunk_ext <- extent(rast1, 1, nrow(rast1), left_bound, right_bound)
  message('Processing ', chunk, ': ', paste0(as.character(round(chunk_ext)), collapse = ', '))
  wdpa_chunk    <- crop(rast1, chunk_ext)
  cell_id_chunk <- crop(rast2, chunk_ext)
  
  wdpa_cells <- crosstab(wdpa_chunk, cell_id_chunk, 
                         progress = 'text',
                         long = TRUE) %>%
    setNames(c('mpa_year', 'cell_id', 'n_prot')) %>%
    mutate(mpa_year = as.integer(as.character(mpa_year)),
           cell_id  = as.integer(as.character(cell_id)))
  ### mpa_year and cell_id are crosstabbed as factors - first
  ### convert to character (to unfactorize it) then to integer.  Otherwise
  ### you end up with factor index, not actual cell ID or category.
  
  return(wdpa_cells)
}

### Use the function in an mclapply call
system.time({
  wdpa_cells_list <- parallel::mclapply(1:n_chunks, mc.cores = 12,
                                          FUN = function(x) crosstab_chunk2(wdpa_yr_rast_500m, 
                                                                            rast_cell_id_500m, 
                                                                            chunk = x, n_chunks))
#     user   system  elapsed 
# 5015.587 1036.650  571.401 
})
  
  
wdpa_cells_df <- bind_rows(wdpa_cells_list)

area_df <- data.frame(cell_id  = raster::values(rast_cell_id),
                      area_km2 = raster::values(rast_ocean_area))

wdpa_year_df <- wdpa_cells_df %>%
  full_join(area_df, by = 'cell_id') %>%
  filter(!is.na(mpa_year)) %>%
  mutate(prot_area_km2 = n_prot / 4, 
         ### count of cells in 500 x 500 m^2 raster, times .25 km^2 area per cell
         mpa_pct = prot_area_km2 / area_km2) %>%
  dplyr::select(mpa_year, cell_id, mpa_pct, prot_area_km2) %>%
  ### mpa pct is relative to *ocean* area - not *total cell* area
  group_by(mpa_year, cell_id) %>%
  summarize(mpa_pct = sum(mpa_pct), prot_area_km2 = sum(prot_area_km2)) %>%
  ungroup()

# Write csv
write_csv(wdpa_year_df, wdpa_cell_year_file)
```

### Plot WDPA protected area proportions as raster

Proportional protection can be multiplied by ocean area raster to get area of protection.  Collect by category (to avoid duplicate "by" values):

* No Take (categories 1 & 2 and specifically designated "no take")
    * classified as 1, 2, and -1 respectively
* Conservation (categories 1-4 plus no take)
    * classified as 1, 2, 3, 4, and -1 (no take)
* All protection (non-NA)
    * classified as 1-6, -1 (no take), and 8 (other protection)

**Note:** these are not saved as separate rasters at this point... they are easily created from the dataframe of cell ID to protected proportion.

``` {r}

wdpa_cell_area_file <- file.path(dir_spatial, 'wdpa_mpa_area.csv')

# Load ocean cell area
wdpa_area_df <- read_csv(wdpa_cell_area_file, show_col_types = FALSE)

notake_df <- wdpa_area_df %>%
  filter(wdpa_category <= 2) %>%
  group_by(cell_id) %>%
  summarize(mpa_pct = sum(mpa_pct)) %>%
  rename(from = cell_id, to = mpa_pct)

conservation_df <- wdpa_area_df %>%
  filter(wdpa_category <= 4) %>%
  group_by(cell_id) %>%
  summarize(mpa_pct = sum(mpa_pct)) %>%
  rename(from = cell_id, to = mpa_pct)

all_prot_df <- wdpa_area_df %>%
  group_by(cell_id) %>%
  summarize(mpa_pct = sum(mpa_pct)) %>%
  rename(from = cell_id, to = mpa_pct)

# notake_raster <- classify(x = rast(rast_cell_id), notake_df, others = NA)
# plot(notake_raster, main = 'No take protection (Ia, Ib, II + no take)')

system.time({
    notake_raster <- classify(x = rast(rast_cell_id), notake_df, others = NA)
#     user   system  elapsed 
# 5304.982 1057.941  597.739 
  })
plot(notake_raster, main = 'No take protection (Ia, Ib, II + no take)')

# conservation_raster <- classify(rast(rast_cell_id), conservation_df, others = NA)
# plot(conservation_raster, main = 'Conservation protection (Ia - IV + no take)')

system.time({
    conservation_raster <- classify(rast(rast_cell_id), conservation_df, others = NA)
#    user  system elapsed 
# 561.630   7.179 551.211 
  })
plot(conservation_raster, main = 'Conservation protection (Ia - IV + no take)')

# all_prot_raster <- classify(rast(rast_cell_id), all_prot_df, others = NA)
# plot(all_prot_raster, main = 'All protection (any category)')

system.time({
    all_prot_raster <- classify(rast(rast_cell_id), all_prot_df, others = NA)
#     user   system  elapsed 
# 1070.091   13.821 1051.193
  })
plot(all_prot_raster, main = 'All protection (any category)')
```

