---
title: 'Set up IUCN spatial data'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

library(raster)
library(sf)
library(fasterize)
library(dplyr)
library(here)
library(ggplot2)

source(here('workflow/R/common.R'))

goal     <- 'spp'
scenario <- 'v2025'
dir_goal <- here('globalprep', goal, scenario)

### goal specific folders and info
dir_data  <- file.path(dir_goal, '_data')
dir_setup <- file.path(dir_goal, '_setup')
dir_anx  <- file.path(dir_M, 'git-annex/globalprep')
dir_goal_anx <- file.path(dir_anx, goal, scenario, 'spp_risk_dists')
dir_raw_data_anx <- file.path(dir_anx, '_raw_data')

source(file.path(dir_setup, 'common_fxns.R'))

dir_bli <- file.path(dir_M, 'git-annex/globalprep/_raw_data/birdlife_intl/d2025')
dir_shp <- file.path(dir_M, 'git-annex/globalprep/_raw_data/iucn_spp/d2025-1')
  ### in this folder are shapefiles at a taxonomic level, both from
  ### IUCN spatial data download (version 2025-1)

```

# Summary

Filter BirdLife International for marine species, saved in the git-annex:

- `bli_marine_v2025.shp`

Assign MEOW codes to IUCN regions, saved in the `_spatial` folder:

- `iucn_rgn_to_meow.csv`

Set up species ranges for all populations and subpopulations, saved in the `_data` folder:

- `spp_marine_maps_2025-1.csv`

# Data sources

[MEOW Regions](https://www.worldwildlife.org/publications/marine-ecoregions-of-the-world-a-bioregionalization-of-coastal-and-shelf-areas)

[IUCN Red List: Spatial Data Download](https://www.iucnredlist.org/resources/spatial-data-download)

[BirdLife International](https://datazone.birdlife.org/contact-us/request-our-data) 

# Methods

## Update geographic scope of regional assessments

The following descriptions may be useful for assigning MEOW regions:

| Region              | MEOW                                                   |
| ------------------- | ------------------------------------------------------ |
| Northern Africa     | 87-88, 32-33, 35-36, 28-29                             |
| Southern Africa     | 192-194                                                |
| Eastern Africa      | 94-102                                                 |
| Western Africa      | 79-86                                                  |
| Central Africa      | no marine                                              |
| Northeastern Africa | 87-93                                                  |
| Pan-Africa          | 28-29, 32-33, 35-36, 79-86, 87-89, 93, 94-102, 192-194 |
| Gulf of Mexico      | 43, 68-70                                              |
| Mediterranean       | 30-36, 44                                              |
| Persian Gulf        | 90-91                                                  |

### Spatial: `iucn_rgn_to_meow.csv`

```{r}
# Load region risk csv to view regions
rgn_risk_ts <- read_csv(file.path(dir_goal, '_data',
                                 sprintf('iucn_risk_ts_rgn_%s.csv', api_version)), show_col_types = FALSE)

# View regions
unique(rgn_risk_ts$iucn_rgn)

# Assign MEOWS to regions
rgn_to_meow <- c('europe'             = paste(c(2,21:27, 30:31, 34:36, 44), 
                                              collapse = ','), #updated 2025
                'gulf_of_mexico'      = paste0(c(43, 69:70), 
                                               collapse = ','), #updated 2025
                'pan-africa'          = paste0(c(28:29, 32:33, 35:36, 79:89, 93:102, 190:194), 
                                               collapse = ','), #updated 2025
                'mediterranean'       = paste0(c(30:36, 44), 
                                               collapse = ','), #updated 2025
                'arabian_sea'         = paste0(c(89, 91:92, 103, 105:106), 
                                               collapse = ','), #updated 2025
                'caribbean'           = paste0(c(63:68), 
                                               collapse = ","), #updated 2025
                'western_africa'      = paste0(79:83, 
                                               collapse = ','), #updated 2025
                'central_africa'      = paste0(c(81:85), 
                                               collapse = ','), #updated 2025
                'persian_gulf'        = paste0(c(90:91), 
                                               collapse = ','), #updated 2025
                'global'              = paste0(c(1:232), 
                                               collapse = ','))

# Create data frame for layer
rgn_to_meow_df <- data.frame(iucn_rgn = names(rgn_to_meow),
                             meow_id  = rgn_to_meow) %>%
  mutate(meow_id = str_split(meow_id, ',')) %>%
  unnest(meow_id) %>%
  group_by(iucn_rgn) %>%
  mutate(meow_id = as.integer(meow_id),
         priority = n()) %>%
  arrange(priority)

# Write csv
write_csv(rgn_to_meow_df, file.path(dir_goal, sprintf('_spatial/iucn_rgn_to_meow.csv')))

# Check
if(setdiff(unique(rgn_to_meow_df$iucn_rgn), unique(rgn_risk_ts$iucn_rgn)) != "global") {
  warning("Extra regions not accounted for in MEOW assignments.")
}
```

## Generate species map list with subpops

Filter the BirdLife Int'l data for marine species.

**Note:** there appear to be no subpopulation polygons within the BirdLife International map set; there is no "subpop" column in that attribute table.  It appears that the individual files may not have "subpop" column either.

``` {r convert BLI to shp-dbf-etc}

# Convert BirdLife Int'l .gdb to shapefile, after filtering to species in the "marine" list.
# Here we also standardize the column names to match the IUCN column names.

# Goal file
bli_marine_shp <- file.path(dir_bli, 'bli_marine_v2025.shp')

# List bli files
bli_gpkg <- list.files(path = dir_bli, pattern = "\\.gpkg$", full.names = TRUE)

bli_sf <- st_read(bli_gpkg)

marine_spp <- read_csv(file.path(dir_data, sprintf('spp_marine_from_api_%s.csv', api_version)),
                       col_types = 'dcc')

names(bli_sf)

# Filter for marine species
bli_marine_sf <- bli_sf %>%
  filter(sisid %in% marine_spp$iucn_sid) %>%
  clean_df_names()

names(bli_marine_sf)

#if the above code ^^ does not work, use this. It functions the same way, just more tedious. 
#marine_colnames <- colnames(marine_spp[2:3])
#bli_marine_sf_check <- bli_sf %>%
#  inner_join(marine_spp, by = c("SISID" = "iucn_sid")) %>%
#  dplyr::select(-marine_colnames) %>%
#    clean_df_names() 

#st_geometry(bli_marine_sf) <- NULL #gets rid of "Shape" column. Since geometry is sticky, clean_df_names() retains the original geometry column "Shape". Now lets rename it to "shape" and retain the geometry format.

#st_geometry(bli_marine_sf) <- bli_marine_sf$shape  #have to reestablish that "shape" is the geometry column. For some reason renaming the geometry column results in r not recognizing the column as the geometry column. 

#class(bli_marine_sf)

st_geometry(bli_marine_sf) <- "geom"

# Save shapefile
st_write(bli_marine_sf, bli_marine_shp)

# Load shapefile
#bli_marine_sf <- st_read(bli_marine_shp)
```

Create subpopulation lookup to match maps with IUCN SIDs for subpops. From available shapefiles, reclassify subpopulation polygons into subpop ID numbers for matching to risk and trend. 

``` {r create_subpop_lookup}
iucn_dbfs <- list.files(dir_shp, pattern = '\\.dbf$', full.names = TRUE)
bli_dbfs  <- list.files(dir_bli, pattern = '\\.dbf$', full.names = TRUE)
shps_dbfs <- c(iucn_dbfs, bli_dbfs)

# This is all marine species with habitat info
marine_spp <- read_csv(file.path(dir_data, 
                                 sprintf('spp_marine_from_api_%s.csv', 
                                         api_version)), col_types = 'dcc')

### Define helper function to cleanly get the info from DBF files,
### including colname checks
get_dbf <- function(x) {
  
  spp_dbf_info <- foreign::read.dbf(x, as.is = TRUE) %>%
    clean_df_names()
  
  ### Individual files call the id number 'iucn_sid'; bli has been fixed to
  ### 'iucn_sid' as well; Red List Data Download files are 'id_no'... So:
  ### fix that here!
  names(spp_dbf_info)[names(spp_dbf_info) == 'id_no'] <- 'iucn_sid'

  ### if 'binomial', set to 'sciname'
  if(!'sciname' %in% names(spp_dbf_info)) {
    names(spp_dbf_info)[names(spp_dbf_info) == 'binomial'] <- 'sciname'
  }
  
  ### Some files (e.g. bli and individual files) don't have 'subpop':
  ### add it in as NAs.
  if(!'subpop' %in% names(spp_dbf_info)) spp_dbf_info$subpop <- NA
  
  ### some files don't have 'presence' field: if not present or NA,
  ### set to presence = 1.  Some have presence = 0?
  if(!'presence' %in% names(spp_dbf_info)) spp_dbf_info$presence <- 1
  
  spp_dbf_info <- spp_dbf_info %>%
    mutate(presence = ifelse(presence == 0, 1, presence))
  
  return(spp_dbf_info)
}

# IUCN data df
shps_df <- lapply(iucn_dbfs, get_dbf) %>%
  setNames(iucn_dbfs) %>%
  bind_rows(.id = 'dbf_file') %>%
  dplyr::select(dbf_file, iucn_sid, sciname = sci_name, presence, subpop) %>%
  # Filter for marine species
  filter(iucn_sid %in% marine_spp$iucn_sid)

# bli data, change column names to match
spp_dbf_info <- foreign::read.dbf(bli_dbfs, as.is = TRUE) %>%
  clean_df_names()
  
colnames(spp_dbf_info)
#  [1] "objectid"   "sisid"      "sci_name"   "presence"   "origin"     "seasonal"   "source"     "compiler"  
#  [9] "data_sens"  "sens_comm"  "dist_comm"  "tax_comm"   "generalisd" "citation"   "yrcompiled" "yrmodified"
# [17] "version"  

# Rename 'ssid' in bli to 'iucn_sid' to match
names(spp_dbf_info)[names(spp_dbf_info) == 'sisid'] <- 'iucn_sid'

# If 'sciname' isn't detected, rename 'sci_name' to match
if(!'sciname' %in% names(spp_dbf_info)) {
  names(spp_dbf_info)[names(spp_dbf_info) == 'sci_name'] <- 'sciname'
}
  
### Some files (e.g. bli and individual files) don't have 'subpop':
### add it in as NAs.
if(!'subpop' %in% names(spp_dbf_info)) spp_dbf_info$subpop <- NA

# Some files don't have 'presence' field: if not present or NA, set to presence = 1.
if(!'presence' %in% names(spp_dbf_info)) spp_dbf_info$presence <- 1

# unique(spp_dbf_info$presence)
# spp_dbf_info <- spp_dbf_info %>%
#   mutate(presence = ifelse(presence == 0, 1, presence))

# bli data df
shps_bli_df <- spp_dbf_info %>%
  cbind(bli_dbfs) %>%
  dplyr::select("dbf_file" = "bli_dbfs", iucn_sid, sciname, presence, subpop) %>%
  filter(iucn_sid %in% marine_spp$iucn_sid)  
  
# Combine IUCN and bli data dfs
shps_df <- rbind(shps_df, shps_bli_df)

# Isolate subpopulations from shapes
shp_subpops <- shps_df %>%
  dplyr::select(shp_iucn_sid = iucn_sid, sciname, shp_subpop = subpop) %>%
  # remove lines with no subpops
  filter(!is.na(shp_subpop)) %>%
  filter(shp_subpop != "0") %>%
  filter(shp_subpop != "<Null>") %>%
  distinct() %>%
  mutate(shp_subpop_clean = str_replace(tolower(shp_subpop), ' subpopulation| ocean', ''),
         shp_subpop_clean = str_replace_all(shp_subpop_clean, '[^a-z]', ' '),
         shp_subpop_clean = str_trim(shp_subpop_clean),
         # Fix typo
         shp_subpop_clean = str_replace(shp_subpop_clean, 'noth', 'north'))

# Load IUCN subpop information
sub_pops <- read_csv(file.path(dir_goal, sprintf('_data/iucn/sub_pop_%s.csv', api_version)), show_col_types = FALSE)

# Clean up IUCN subpops
api_subpops <- sub_pops %>%
  dplyr::select(api_iucn_sid = sis_id,
                sciname,
         api_subpop = subpopulation_name) %>%
  filter(!is.na(api_subpop)) %>%
  filter(sciname %in% shp_subpops$sciname) %>%
  distinct() %>%
  mutate(api_subpop_clean = str_replace_all(tolower(api_subpop), ' subpopulation| ocean', ''),
         api_subpop_clean = str_replace_all(api_subpop_clean, '[^a-z]', ' ') %>% str_trim)

# Find IUCN subpops that overlap with shapes
subpops_match_raw <- shp_subpops %>%
  full_join(api_subpops, by = 'sciname') %>%
  group_by(api_iucn_sid) %>%
  mutate(subpop_match = str_detect(shp_subpop_clean, api_subpop_clean),
         n_match = sum(subpop_match)) %>%
  filter(subpop_match | sum(subpop_match) == 0) %>%
  ungroup()

caretta_subpops <- subpops_match_raw %>%
  filter(shp_iucn_sid == 3897) %>%
  mutate(api_subpop_single = str_split(api_subpop_clean, ' ')) %>%
  unnest(api_subpop_single) %>%
  group_by(api_subpop_clean, shp_subpop_clean) %>%
  mutate(match = str_detect(shp_subpop_clean, api_subpop_single),
         n_match = sum(match),
         n_words = n()) %>%
  filter(sum(match) == n()) %>%
  ungroup() %>%
  dplyr::select(shp_iucn_sid, sciname, shp_subpop, api_iucn_sid, api_subpop) %>%
  distinct()

subpops_match <- subpops_match_raw %>%
  filter(shp_iucn_sid != 3897 & subpop_match == TRUE) %>%
  dplyr::select(shp_iucn_sid, sciname, shp_subpop, api_iucn_sid, api_subpop) %>%
  bind_rows(caretta_subpops)

### Check for missed matches:
api_subpops$api_iucn_sid[!api_subpops$api_iucn_sid %in% subpops_match$api_iucn_sid]

# 2025:
# 103290570 Acipenser transmontanus - columbia river upper  middle and lower

# Previous:
### API subpop, not matched: 16369383 Tursiops truncatus 
###                                   Mediterranean subpopulation
### * OK - no polygon for this subpopulation.  Regional assessment?

### API subpop, not matched: 44187 Orcaella brevirostris 
###                                Malampaya Sound subpopulation
### * Doesn't look like there is a shape for this subpop... Sorry dude!

### API subpop, not matched: 123095978 Orcaella brevirostris
###                                    loilo-Guimaras Subpopulation  
### * Doesn't look like there is a shape for this subpop... Sorry dude!

### API subpop, not matched: 156206333, 189865869  Delphinus delphis
###                                    Gulf of Corinth subpopulation,
###                                    Inner Mediterranean
### * It does look like there is a shape for this one... the medittarean sea.. but it seems like it is already accounted for in the matched subpops... so i think it is ok. 

### API subpop, not matched: 181208820 Tursiops truncatus
###                                    Gulf of Ambracia subpopulation
### * It doesn't look like there is a shape for this subpopulation...

### API subpop, not matched: 218582754 Dugong dugon 
###                                    New Caledonia subpopulation
### It doesn't look like there is a shape for this subpop ... 

### API subpop, not matched: 218582764 Dugong dugon 
###                                    Eastern Africa subpopulation
### It doesn't look like there is a shape for this subpop

shp_subpops[!shp_subpops$shp_subpop %in% subpops_match$shp_subpop, ]
# all matched
# 2025:
# 2476 Balaenoptera edeni - gulf of mexico
# rl_sis(2476) subpopulation = FALSE *so listed as a subpop in the shapefiles but not in the IUCN api?
# 225 Acipenser nudiventris - lake?
# 234 Acipenser transmontanus - river?
 
write_csv(subpops_match, 
          file.path(dir_setup, sprintf('int/subpops_match_api_to_shp_%s.csv', 
                                     api_version)))
```

#### Data: `spp_marine_maps_2025-1.csv`

From the species range map shapefiles, pull map info from the .dbf files.  For species listed in multiple files (e.g. sea snakes and reptiles), remove duplicates.  For species with polygons differentiated by subpopulation, adjust the iucn_sid to match the subpop info.

Add a flag for comprehensively-assessed spp - fix BONEFISH_TARPONS to all be assessed - either it's comprehensive or not - and add sea turtle species (6 of 7 assessed).

``` {r generate map list from dbfs}

iucn_dbfs  <- list.files(dir_shp, pattern = '\\.dbf$', full.names = TRUE)
bli_dbfs   <- list.files(dir_bli, pattern = '\\.dbf$', full.names = TRUE)

map_files  <- c(iucn_dbfs, bli_dbfs)

# Load comprehensive assessment species
comp_spp <- read_csv(file.path(dir_data, 
                       sprintf('iucn_comp_assessed_%s.csv', api_version)), show_col_types = FALSE)

# spp_group_names <- map_files %>%
#   basename() %>%
#   str_replace_all('\\.dbf$|_PART_.+', '') %>%
#   unique()

# IUCN files
map_info_raw_iucn <- lapply(iucn_dbfs, get_dbf) %>%
  setNames(iucn_dbfs) %>%
  bind_rows(.id = 'dbf_file') %>%
  dplyr::select(shp_iucn_sid = iucn_sid, sciname=sci_name, presence, subpop, dbf_file)

# bli files
## now run on birdlife data here, and change column names to match
spp_dbf_info <- foreign::read.dbf(bli_dbfs, as.is = TRUE) %>%
  clean_df_names()

colnames(spp_dbf_info)
# [1] "objectid"   "sisid"      "sci_name"   "presence"   "origin"     "seasonal"   "source"     "compiler"   "data_sens"  "sens_comm" 
# [11] "dist_comm"  "tax_comm"   "generalisd" "citation"   "yrcompiled" "yrmodified" "version"  
  
# Rename to match iucn data
names(spp_dbf_info)[names(spp_dbf_info) == 'sisid'] <- 'iucn_sid'

if(!'sciname' %in% names(spp_dbf_info)) {
  names(spp_dbf_info)[names(spp_dbf_info) == 'sci_name'] <- 'sciname'
}

# Some files (e.g. bli and individual files) don't have 'subpop', add it in as NAs.
if(!'subpop' %in% names(spp_dbf_info)) spp_dbf_info$subpop <- NA

# Some files don't have 'presence' field: if not present or NA, set to presence = 1. 
if(!'presence' %in% names(spp_dbf_info)) spp_dbf_info$presence <- 1

spp_dbf_info <- spp_dbf_info %>%
  mutate(presence = ifelse(presence == 0, 1, presence))

# Bind IUCN and bli dfs
map_info_raw_bli <- spp_dbf_info %>%
  cbind(bli_dbfs) %>%
  dplyr::select(shp_iucn_sid = iucn_sid, sciname, presence, subpop, "dbf_file" = "bli_dbfs")

map_info_raw <- rbind(map_info_raw_iucn, map_info_raw_bli)

### Notes on presence, origin, seasonal fields:
### * presence = 5 is extinct; 4 = probably extinct; others are extant-ish or 
###   uncertain. We will drop field and include all polygons for now.
### * origin is native, introduced, etc.  We will drop this field and not
###   worry about origin.
### * seasonal is breeding/non-breeding/passage.  We will drop this field
###   and not worry about seasonality.

# subspp <- map_info_raw %>%
#   filter(!is.na(subpop)) %>%
#   dplyr::select(shp_iucn_sid, presence) %>%
#   distinct()

# No subspecies vary by IUCN code; drop field for simplicity and include all at species level
map_info_sans_dupes <- map_info_raw %>%
  group_by(shp_iucn_sid) %>%
  filter(!(str_detect(dbf_file, 'MARINEFISH|REPTILE|REEF_FORMING_CORALS_PART2|REEF_FORMING_CORALS_PART1|MANGROVES') & 
             length(unique(dbf_file)) > 1)) %>%
  filter(!(shp_iucn_sid == 196026 & str_detect(dbf_file, 'SEASNAKE'))) %>%
    ### in REPTILES, SEASNAKES, and HAGFISH - ditch any dupes
  ungroup() %>%
  distinct()

subpops_match <- read_csv(file.path(dir_setup, 'int',
                                    sprintf('subpops_match_api_to_shp_%s.csv', 
                                            api_version)),
                          col_types = 'dccdc')

map_info_add_subpops <- map_info_sans_dupes %>%
  full_join(subpops_match, by = c('shp_iucn_sid', 'sciname', 'subpop' = 'shp_subpop')) %>%
  mutate(iucn_sid = ifelse(is.na(api_iucn_sid), shp_iucn_sid, api_iucn_sid)) %>%
  dplyr::select(shp_iucn_sid, iucn_sid, sciname, presence, subpop, dbf_file) %>%
  distinct()

# Filter to marine habitat species (to drop terrestrial reptiles e.g.)
marine_spp_ids <- read_csv(file.path(dir_data,
                                     sprintf('spp_marine_from_api_%s.csv', 
                                             api_version)),
                           col_types = 'dcc') %>%
  dplyr::select(iucn_sid, max_depth)

marine_map_info <- map_info_add_subpops %>%
  inner_join(marine_spp_ids, by = 'iucn_sid') %>%
  mutate(comp_assessed = iucn_sid %in% comp_spp$iucn_sid)
  
comp_assess_check <- marine_map_info %>%
  mutate(dbf_short = basename(dbf_file)) %>%
  group_by(dbf_short, comp_assessed) %>%
  summarize(n_in_gp = n()) %>%
 group_by(dbf_short) %>%
  mutate(both = n() == 2)

### at this point, fix a few "comprehensively assessed" issues:
### * add turtles as comprehensive
### * some species from Spatial Data Download groups do not
###   show up as comprehensively assessed despite text on the IUCN site
turtles <- c('Dermochelys', 'Caretta', 'Lepidochelys', 
             'Natator', 'Chelonia', 'Eretmochelys') %>%
  paste0(collapse = '|')
assessed_gps <- c('BONEFISH', 'CORALS', 'SEACUCUMBERS') %>%
  paste0(collapse = '|')
  ### NOTE: Damselfish removed at recommendation of Gina Ralph
  ### Others, assume that availability of maps on Spatial Data Download Page
  ### is indication of comprehensive assessment despite API call

marine_map_info <- marine_map_info %>%
  mutate(comp_assessed = ifelse(str_detect(dbf_file, assessed_gps), 
                                TRUE, comp_assessed),
         comp_assessed = ifelse(str_detect(sciname, turtles), 
                                TRUE, comp_assessed))
# Save layer
write_csv(marine_map_info, 
          file.path(dir_data,
                    sprintf('spp_marine_maps_%s.csv', api_version)))
```

#### Output: 

Not essential.

- `spp_included_summary.csv`
- `spp_included_by_family.csv`

``` {r generate table of species}

# marine_map_info <- read_csv(file.path(dir_data,
#   sprintf('spp_marine_maps_%s.csv', api_version)), show_col_types = FALSE) %>%
#   dplyr::select(-presence) %>%
#   distinct()
# 
# ## 2025: problems with 2236, 15131, 15137, and 15143 (?)
# 
# taxa_info <- read_csv(file.path(dir_goal, sprintf('_data/iucn/taxa_%s.csv', api_version)), show_col_types = FALSE) %>%
#   dplyr::select(iucn_sid, kingdom:family) %>%
#   distinct()
# 
# #sub_pops <- read_csv(file.path(dir_goal, sprintf('_data/iucn/sub_pop_%s.csv', api_version)), show_col_types = FALSE)
# 
# cat_info <- read_csv(file.path(dir_data, 
#                                sprintf('iucn_risk_current_%s.csv', api_version)),
#                      show_col_types = FALSE) %>%
#   dplyr::select(iucn_sid, cat_score) %>%
#   filter(!is.na(cat_score))
# 
# included_maps <- marine_map_info %>%
#   filter(comp_assessed) %>%
#   filter(iucn_sid %in% cat_info$iucn_sid) %>%
#   dplyr::select(iucn_sid, shp_iucn_sid, subpop, sciname, dbf_file) %>%
#   left_join(taxa_info, by = 'iucn_sid') %>%
#   mutate(spp_gp = basename(dbf_file),
#          spp_gp = spp_gp %>% tolower() %>% str_replace('.dbf|_part.+', '')) %>%
#   mutate(kingdom = tools::toTitleCase(tolower(kingdom)),
#          phylum  = tools::toTitleCase(tolower(phylum)),
#          order   = tools::toTitleCase(tolower(order)),
#          class   = tools::toTitleCase(tolower(class)),
#          family  = tools::toTitleCase(tolower(family)))
# 
# all_in_the_fam <- included_maps %>%
#   group_by(kingdom, phylum, class, order, family, spp_gp) %>%
#   summarize(species = length(unique(shp_iucn_sid)),
#             subpops = sum(!is.na(subpop)))  
# 
# basic_sum <- included_maps %>%
#   filter(!str_detect(tolower(class), 'actinopt|mammal|aves|rept')) %>%
#   group_by(kingdom, phylum, class, spp_gp) %>%
#   summarize(species = length(unique(shp_iucn_sid)),
#             subpops = sum(!is.na(subpop)))
# 
# nonfish_sum <- included_maps %>%
#   filter(str_detect(tolower(class), 'mammal|aves|rept')) %>%
#   group_by(kingdom, phylum, class, order, spp_gp) %>%
#   summarize(species = length(unique(shp_iucn_sid)),
#             subpops = sum(!is.na(subpop)))
# 
# fish_sum <- included_maps %>%
#   filter(str_detect(tolower(class), 'actinopt')) %>%
#   group_by(kingdom, phylum, class, order, family, spp_gp) %>%
#   summarize(species = length(unique(shp_iucn_sid)),
#             subpops = sum(!is.na(subpop)))
# 
# all_sum <- bind_rows(fish_sum, nonfish_sum, basic_sum) %>%
#   group_by(spp_gp, kingdom, phylum, class, order) %>%
#   summarize(family = paste(unique(family), collapse = ', '),
#             species = sum(species),
#             subpops = sum(subpops)) %>%
#   ungroup() %>%
#   mutate(family = ifelse(family == 'NA', NA, family),
#          subpops = ifelse(subpops == 0, NA, subpops),
#          spp_gp = ifelse(spp_gp == "cone_snails", "conus", spp_gp))
# 
# ## Check against last year to see if you have all of the same species. 
# 
# ## 2024: We have the same species, but the setdiff is going to return a list of those that are not matching because the all_sum df has them grouped into "marinefish" as the spp_gp. So I went through and cross-referenced the class, order, and family to make sure that they were all present - they are present 
# 
# ## 2025: Same, blennies, butterflyfish_angelfish_allies, bonefishes_tarpons_ladyfishes, and pufferfishes_and_relatives have been combined into marinefishes
# 
# old_sum <- read_csv(file.path(dir_goal, '_output', 
#                              'spp_included_summary.csv'))
# 
# old_sum <- spp_included_summary
# 
# setdiff(old_sum$spp_gp, all_sum$spp_gp)
# 
# ## end check
# 
# # Save layers
# write_csv(all_sum, file.path(dir_goal, '_output', 
#                              'spp_included_summary.csv'))
# write_csv(all_in_the_fam, file.path(dir_goal, '_output', 
#                                     'spp_included_by_family.csv'))
```

