---
title: 'Process IUCN Spatial Data'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

library(raster)
library(sf)
library(fasterize)
library(dplyr)
library(here)
library(ggplot2)
library(doParallel)
library(terra)

source(here('workflow/R/common.R'))

goal     <- 'spp'
scenario <- 'v2025'
dir_goal <- here('globalprep', goal, scenario)

### goal specific folders and info
dir_data  <- file.path(dir_goal, '_data')
dir_setup <- file.path(dir_goal, '_setup')
dir_anx  <- file.path(dir_M, 'git-annex/globalprep')
dir_goal_anx <- file.path(dir_anx, goal, scenario, 'spp_risk_dists')
dir_raw_data_anx <- file.path(dir_anx, '_raw_data')
dir_spatial <- file.path(dir_goal, '_spatial')

source(file.path(dir_setup, 'common_fxns.R'))

dir_bli <- file.path(dir_M, 'git-annex/globalprep/_raw_data/birdlife_intl/d2025')
dir_shp <- file.path(dir_M, 'git-annex/globalprep/_raw_data/iucn_spp/d2025-1')
  ### in this folder are shapefiles at a taxonomic level, both from
  ### IUCN spatial data download (version 2025-1)
```

# Summary

Using a set of IUCN species range maps, rasterize each species to 10 km x 10 km raster using `fasterize`.  Use `presence` field from shapefile.

* Subpopulation polygons must be identified and rasterized separately from the parent polygon; this must be done by sciname and subpop fields since the polygon IDs are based upon the parent ID.
* Regional assessments need not be determined at this stage - the ID numbers match the global ID numbers (including subpops).

# Data source

IUCN Red List: Spatial Data Download
IUCN: Gina Ralph direct communication

# Methods

## Read spp shapes, correct subpop IDs, `fasterize()`, depth clip, save to csv

We will loop over each species in each shapefile and rasterize separately, using `sf` and `fasterize` packages.  

* From the full map list, filter to a single shapefile
* Load shapefile using `st_read`, and correct subpop IDs from `shp_iucn_sid` to `iucn_sid`
* Loop over each `iucn_sid` in the shapefile, rasterizing (`fasterize()`) to 10 km^2 resolution, using "presence" field. 
    * clip to neritic (<=200 m) and shallow (<=60 m) depth raster if appropriate.  Otherwise mask to bathy raster.  Since bathy raster was created by masking to area raster, cells with any marine presence will be kept but any non-marine cells will be dropped.
    * Save as .tif and .csv, and compare average file sizes.  .csv easier to work with, but might be significantly larger than .tifs in the long run.
        * note: no longer saving as .tif for speed and file size - just use .csv instead!
    * use `mclapply()` to speed this up.
    
#### Annex: `spp_risk_dists/spp_rasters/iucn_sid_%s.csv`
    
``` {r rasterize iucn and clip and save to csv}

reload <- FALSE

maps_to_rasterize <- read_csv(file.path(dir_data,
                                        sprintf('spp_marine_maps_%s.csv', api_version)),
                              col_types = 'ddcicccl') %>%
  mutate(shp_file = str_replace(dbf_file, 'dbf$', 'shp'))

# Rast_base for cell IDs
rast_base <- raster(file.path(dir_spatial, 'cell_id_rast.tif'))

### If some species already processed, remove from the list to process.
if(reload == FALSE) {
  maps_already_rasterized <- list.files(file.path(dir_goal_anx, 'spp_rasters'),
                                        pattern = '.csv') %>%
    str_replace_all('iucn_sid_|.csv$', '') %>%
    as.integer()

  maps_to_rasterize <- maps_to_rasterize %>%
    filter(!iucn_sid %in% maps_already_rasterized)
}

### for selectively rerunning taxonomic groups
# maps_to_rasterize <- maps_to_rasterize %>%
#   filter(str_detect(dbf_file, 'birdlife'))

# These will be used as masks
rast_bathy <- raster(file.path(dir_spatial,
                               'bathy_rast.tif'))
rast_neritic <- raster(file.path(dir_spatial,
                                 'bathy_rast_neritic.tif'))
rast_shallow <- raster(file.path(dir_spatial,
                                 'bathy_rast_shallow.tif'))

if(nrow(maps_to_rasterize) == 0) { ### all maps accounted for as .csvs
  
  cat_msg('reload == ', reload, '... No maps to process...')
  
} else {
  
  cat_msg('reload == ', reload, '... Maps to process: ', nrow(maps_to_rasterize))
  
  ################################################################.
  ### Loop over each distinct shapefile with species range maps
  ################################################################.

  # bli_already_done <- read_csv(file.path("globalprep/spp/v2025/_setup", "int/bli_already_done.csv"))  
  # bli_done_vec <- bli_already_done$iucn_sid
  # 
  # maps_to_rasterize <- maps_to_rasterize %>%
  #   filter((iucn_sid %in% bli_done_vec))
  
  shps <- maps_to_rasterize$shp_file %>% unique()
  # shps <- shps[1]
  for(i in seq_along(shps)) {
    # i <- 1
    
    shp <- shps[i]
    
    maps_in_shp <- maps_to_rasterize %>%
      filter(shp_file == shp)
    
    id_fix <- maps_in_shp %>%
      dplyr::select(shp_iucn_sid, iucn_sid, subpop, max_depth) %>% 
      distinct()
    
    cat_msg(i, ' of ', length(shps), ': reading ', basename(shp), ' from: \n  ', shp)
    
    polys_all_geometry <- read_sf(shp, type = 6) %>%
      clean_df_names()
    
    polys_all <- polys_all_geometry %>% 
      st_set_geometry(NULL)
    
      ### we will check geoms and fix them inside the mclapply, 
      ### and then reproject, one species at a time
    
    if('sci_name' %in% names(polys_all)) {
      polys_all <- polys_all %>% 
        rename(sciname = sci_name) 
    } 

    if('binomial' %in% names(polys_all)) {
      polys_all <- polys_all %>%
        rename(sciname = binomial)
    }
    
    if(!'subpop' %in% names(polys_all)) {
      polys_all$subpop <- NA_character_
      ### if shape doesn't have subpop column, add it as NA
    }
    if('id_no' %in% names(polys_all)) {
      polys_all <- polys_all %>%
        rename(iucn_sid = id_no)
    }
    
    if('sisid' %in% names(polys_all)){
      polys_all <- polys_all %>%
        rename(iucn_sid = sisid)
    }
    
    if('presenc' %in% names(polys_all)){
      polys_all <- polys_all %>%
        rename(presence = presenc) 
    }
    
    if(!'presence' %in% names(polys_all)) {
      polys_all <- polys_all %>%
        mutate(presence = 1) 
    }
    
    polys_all <- polys_all %>%
        st_sf(geometry = st_geometry(polys_all_geometry))
    
    if(identical(st_geometry(polys_all), st_geometry(polys_all_geometry)) == FALSE) {
      cat_msg("HALT: GEOMETRIES DO NOT MATCH FOR", i, ' of ', length(shps), 'in', basename(shp), ' from: \n  ', shp)
    }
    
    polys_match <- polys_all %>%
      dplyr::select(shp_iucn_sid = iucn_sid, sciname, subpop, presence, geometry) %>%
      mutate(presence = ifelse(presence == 0, 1, presence),
             subpop   = as.character(subpop)) %>%
      inner_join(id_fix, by = c('shp_iucn_sid', 'subpop')) 
    
    spp_ids <- maps_in_shp$iucn_sid %>% 
      sort() %>% 
      unique()
    
  
    ####################################################################.
    ### In each shapefile, loop over each species ID using mclapply().
    ####################################################################.
    
    cat_msg('Processing ', basename(shp), ' with ', length(spp_ids), ' species...')
    
    # system.time({
      tmp <- parallel::mclapply(seq_along(spp_ids),
                                mc.cores = 12, 
                                FUN = function(x) {
         #x <- 81:128
        spp <- spp_ids[x]
        
        cat_msg(x, ' of ', length(spp_ids), ': Processing ', spp, ' in ', basename(shp),
                ' (group ', i, ' of ', length(shps), ')...\n')
        
        spp_shp <- polys_match %>%
          filter(iucn_sid == spp)
        
        spp_shp <- valid_check(spp_shp)
          ### if invalid geom, and bounds exceeded, buffer to 0
        spp_shp <- spp_shp %>%
          clip_to_globe() %>%
            ### indiv files should have adjusted boundaries already; only 
            ### files from Spatial Data Download should run into this
          st_transform(gp_proj4)
        
        spp_rast <- fasterize::fasterize(spp_shp, rast_base, field = 'presence', fun = 'min')
        
        ### depth clip if necessary; otherwise clip to bathy raster (which previously
        ### was clipped to area raster - so cells with any marine area will be kept,
        ### and non-marine cells will be dropped)
        if(unique(spp_shp$max_depth) == '< 20 m') {
          spp_rast <- mask(spp_rast, rast_shallow)
        } else if(unique(spp_shp$max_depth) == '< 200 m') {
          spp_rast <- mask(spp_rast, rast_neritic)
        } else {
          spp_rast <- mask(spp_rast, rast_bathy)
        }
        ### write out as a raster:
        # rast_file <- file.path(dir_o_anx, sprintf('spp_rasters/iucn_sid_%s.tif', spp))
        # raster::writeRaster(spp_rast, rast_file, overwrite = TRUE)
        ### See note below about file size; rasters are larger than csvs in general
        
        ### convert to dataframe and write out as a csv:
        spp_present <- data.frame(cell_id  = values(rast_base),
                                  presence = values(spp_rast)) %>%
          filter(!is.na(presence))
        
        if(nrow(spp_present) == 0) {
    cat_msg('Species ID ', spp, ' resulted in a zero-length dataframe.')
        }
        
        write_csv(spp_present, file.path(dir_goal_anx, 'spp_rasters',
                                         sprintf('iucn_sid_%s.csv', spp)))
        
        return(NA)
      }) ### end of mclapply FUN definition
    # }) ### end of system.time call
  } ### end of for loop over each species group
} ### end of "if" check to make sure there are any maps to rasterize

## Note: v2020: This got to the end and didn't indicate it was finished. I only knew because I periodically stopped the loop and checked after it got to the last couple of sfs and then looked at maps_to_rasterize that it was finished. 
## v2021: I had to add some extra if statements to fix column name issues with the new bli data.

```

## Generate species group maps - equal weighting per cell

From species maps data.frame, identify the IDs of species within each taxon. For each taxon:

* Collect all taxon species into a single data.frame.
* Join species extinction risk and trend data.frame to species-cell data.frame, including regional assessments.
* By cell, calculate mean extinction risk, variance of extinction risk, trend, threatened species count, n_spp, n_spp_trend
    * When aggregating further, use n_spp to weight values.
    
__NOTE:__ Because the summary files are likely to be very large for globe-spanning taxa (e.g. MARINE_MAMMALS is 275 MB), save these outputs outside of GitHub.

``` {r aggregate species maps to taxa}
# Read in lots of data
spp_maps <- read_csv(file.path(dir_data, 
                               sprintf('spp_marine_maps_%s.csv', api_version)),
                     col_types = 'ddcicccl')

spp_risk <- read_csv(file.path(dir_data, 
                               sprintf('iucn_risk_current_%s.csv', api_version)), show_col_types = FALSE)

spp_risk_rgn <- read_csv(file.path(dir_data, 
                                   sprintf('iucn_risk_rgn_current_%s.csv', api_version)),
                         col_types = 'dcccdc')

spp_trend <- read_csv(file.path(dir_data, sprintf('iucn_trend_by_spp_%s.csv', api_version)),
                      col_types = 'dccdc') %>%
  dplyr::select(iucn_sid, trend_score, iucn_rgn)

# Make a dataframe of species risk, trend, and regional risk
spp_risk_trend <- spp_risk %>%
  mutate(iucn_rgn = 'global') %>%
  bind_rows(spp_risk_rgn) %>%
  dplyr::select(iucn_sid, iucn_rgn, cat_score) %>%
  left_join(spp_trend, by = c('iucn_sid', 'iucn_rgn'))

# Raster for cell IDs
rast_cell_ids <- terra::rast(file.path(dir_spatial, 'cell_id_rast.tif'))

### Make a dataframe of cell ID to MEOW for regional assessments... also 
### make a lookup of MEOW to region
meow_rgns_rast <- terra::rast(file.path(dir_spatial, 'meow_rast.tif'))
meow_cells_all <- data.frame(cell_id = terra::values(rast_cell_ids),
                             meow_id = terra::values(meow_rgns_rast)) %>%
  dplyr::rename(cell_id = cell_id_rast, meow_id = meow_rast) %>%
  filter(!is.na(meow_id))

meow_to_rgn <- read_csv(file.path(dir_spatial, 'iucn_rgn_to_meow.csv'), show_col_types = FALSE)

```

### Loop over comprehensively assessed species

``` {r loop comprehensively assessed}
# Make a list of taxonomic groups to loop over:
taxa <- spp_maps %>%
  filter(comp_assessed) %>%
  .$dbf_file %>%
  basename() %>%
  unique() %>%
  str_replace('\\.dbf$', '')

#taxa <- taxa[str_detect(taxa, 'SEAGRASS')]
reload <- FALSE

##########################################################.
### Looping over taxonomic groups -----
##########################################################.

registerDoParallel(cores = 4) ## parallelize this so it will go quicker

foreach(taxon = taxa) %dopar% {
   #taxon <- taxa[2]
  
  taxon_sum_file <- file.path(dir_goal_anx, 'taxa_summaries',
                                   sprintf('%s_cell_sum_comp_%s.csv', tolower(taxon),
                                           api_version))
  
  if(!file.exists(taxon_sum_file) | reload == TRUE) {
    
    ### Filter to just this taxon, and only assessed within this taxon
    taxon_maps <- spp_maps %>%
      filter(str_detect(dbf_file, taxon)) %>%
      filter(comp_assessed == TRUE)
    
    taxon_risk_trend <- spp_risk_trend %>%
      filter(iucn_sid %in% taxon_maps$iucn_sid) %>%
      filter(!is.na(cat_score)) %>%
      arrange(iucn_sid)
    
    ### Using the iucn_sid field, generate a vector of all species range files for
    ### this taxon.
    taxon_ids <- taxon_risk_trend$iucn_sid %>%
      unique()
    
    cat_msg('Processing ', length(taxon_ids),
            ' comp-assessed species maps in ', taxon, '...')

    ##########################################################.
    ### Looping over species within group -----
    ##########################################################.

  taxon_cells_list <- list()
    
    #x <- 45433088
                                          
      for(x in taxon_ids){
        
        csv_file <- file.path(dir_goal_anx, 'spp_rasters', 
                              sprintf('iucn_sid_%s.csv', x))
        
        spp_risk_map <- read_csv(csv_file, col_types = 'di') %>%
          mutate(iucn_sid = x) %>%
          left_join(taxon_risk_trend, by = 'iucn_sid') 
        
        ### Identify regional assessments if any
        meow_rgns <- meow_to_rgn %>%
          filter(iucn_rgn %in% spp_risk_map$iucn_rgn)
        
        non_global_rgns <- meow_rgns %>%
          filter(iucn_rgn != 'global')
  
        if(nrow(non_global_rgns) > 0) {
          ### If any regional assessments, clip the MEOW cells down to the appropriate region...
          meow_cells <- meow_cells_all %>%
            inner_join(meow_rgns, by = 'meow_id') %>%
            rename(rgn_name = iucn_rgn)
          ### ... then filter out non-matching overlapped cells
          spp_risk_map <- spp_risk_map %>%
            left_join(meow_cells, by = 'cell_id') %>%
            mutate(rgn_name = ifelse(is.na(meow_id), 'global', rgn_name),
                   priority = ifelse(rgn_name == 'global', 100, priority)) %>%
            filter(iucn_rgn == rgn_name) %>%
            group_by(cell_id) %>%
            filter(priority == min(priority)) %>%
            ungroup()
          ### NOTE: at this point, still possible to have multiple regional
          ### assessments, if priorities are the same (e.g. Europe and Pan Africa,
          ### or overlaps around Africa).  Those values will be averaged in
          ### the group_by() below.
        }
        
        ### select down to main columns; also, if presence == 5 (extinct), 
        ### adjust category and trend scores.
        spp_risk_sum <- spp_risk_map %>%
          dplyr::select(cell_id, presence, iucn_sid, iucn_rgn, cat_score, trend_score) %>%
          distinct() %>%
          mutate(cat_score   = ifelse(presence == 5, 1, cat_score),        
                   ### fix local extinctions
                 trend_score = ifelse(presence == 5, NA, trend_score)) %>% 
                   ### fix local extinctions
          group_by(cell_id, iucn_sid) %>%
          summarize(cat_score   = mean(cat_score, na.rm = TRUE),
                    trend_score = mean(trend_score, na.rm = TRUE)) %>%
          ungroup() %>%
          mutate(trend_score = ifelse(is.nan(trend_score), NA, trend_score))
       
       taxon_cells_list[[length(taxon_cells_list)+1]] = spp_risk_sum
       
      } ### end of loop over all species in taxonomic group
    
    ##########################################################.
    ### Processing cell calculations for group -----
    ##########################################################.
  
    ### Set up for keyed data.table merging: key for iucn_sid, cell_id
    cat_msg('...binding rows to data.frame...')

    taxon_risk_map <- bind_rows(taxon_cells_list) ### this is pretty fast!
    
    cat_msg('...summarizing...')
    taxon_risk_summary <- taxon_risk_map %>%
      group_by(cell_id) %>%
      summarize(mean_risk   = mean(cat_score), 
                  ### NA categories already filtered out
                var_risk    = var(cat_score),
                n_spp_threatened = sum(cat_score >= 0.4 & cat_score < 1),  
                n_spp_risk  = n(),
                mean_trend  = mean(trend_score, na.rm = TRUE), 
                  ### not every species has a trend; na.rm = TRUE
                mean_trend  = ifelse(is.nan(mean_trend), NA, mean_trend),
                n_spp_trend = sum(!is.na(trend_score)))
    
    cat_msg('...writing file', taxon_sum_file, '...')
    write_csv(taxon_risk_summary, taxon_sum_file)

  } else { ### end of if statement checking whether file exists for this taxon
    cat_msg('Found file ', taxon_sum_file, '... skipping process...')
  }
  
} ### end of taxonomic group loop

```

### Loop over non-comprehensively-assessed species 

``` {r loop non-comprehensively assessed, eval = FALSE}
# 
# ### Make a list of taxonomic groups to loop over:
# taxa <- spp_maps %>%
#   filter(!comp_assessed) %>%
#   .$dbf_file %>%
#   basename() %>%
#   unique() %>%
#   str_replace('\\.dbf$', '')
# 
# ##########################################################.
# ### Looping over taxonomic groups -----
# ##########################################################.
# for(taxon in taxa) {
#   ### taxon <- taxa[6]
#   
#   taxon_sum_file <- file.path(dir_goal_anx, 'taxa_summaries',
#                                    sprintf('%s_cell_sum_noncomp_%s.csv', 
#                                            tolower(taxon), api_version))
#   
#   reload <- FALSE
#   if(!file.exists(taxon_sum_file) | reload == TRUE) {
#     
#     ### Filter to just this taxon, and only non-assessed within this taxon
#     taxon_maps <- spp_maps %>%
#       filter(str_detect(dbf_file, taxon)) %>%
#       filter(comp_assessed == FALSE) %>%
#       filter(max_depth != 'deep oceanic')
#     
#     taxon_risk_trend <- spp_risk_trend %>%
#       filter(iucn_sid %in% taxon_maps$iucn_sid) %>%
#       filter(!is.na(cat_score)) %>%
#       arrange(iucn_sid)
#     
#     if(nrow(taxon_risk_trend) == 0) {
#       ### if empty set, create a dummy data frame so there
#       ### will at least be a file...
#       cat_msg('Creating dummy data frame for empty spp set ', taxon)
#       taxon_risk_summary <- data.frame(
#         cell_id = -1,
#         mean_risk = NA)
#     } else {
#         
#       ### Using the iucn_sid field, generate a vector of all species range files for
#       ### this taxon.
#       taxon_ids <- taxon_risk_trend$iucn_sid %>%
#         unique()
#       
#       cat_msg('Processing ', length(taxon_ids), 
#               ' non-comp-assessed species maps in ', taxon, '...')
#   
#       ##########################################################.
#       ### Looping over species within group -----
#       ##########################################################.
#       ### Collect all species ranges for this taxon into a single data.frame.
#       ### Use mclapply since we're reading many large-ish files.  For MARINE_MAMMALS (85 assessed spp)
#       ### this takes about 30-40 seconds
#       taxon_cells_list <- parallel::mclapply(taxon_ids, mc.cores = 12,
#                                           FUN = function(x) {
#           ### x <- taxon_ids[16]
#           ### x <- 6336                            
#           csv_file <- file.path(dir_goal_anx, 'spp_rasters', 
#                                 sprintf('iucn_sid_%s.csv', x))
#           
#           spp_risk_map <- read_csv(csv_file, col_types = 'di') %>%
#             mutate(iucn_sid = x) %>%
#             left_join(taxon_risk_trend, by = 'iucn_sid') 
#           
#           ### Identify regional assessments if any
#           meow_rgns <- meow_to_rgn %>%
#             filter(iucn_rgn %in% spp_risk_map$iucn_rgn)
#           
#           non_global_rgns <- meow_rgns %>%
#             filter(iucn_rgn != 'global')
#     
#           if(nrow(non_global_rgns) > 0) {
#             ### If any regional assessments, clip the MEOW cells down to the appropriate region...
#             meow_cells <- meow_cells_all %>%
#               inner_join(meow_rgns, by = 'meow_id') %>%
#               rename(rgn_name = iucn_rgn)
#             ### ... then filter out non-matching overlapped cells
#             spp_risk_map <- spp_risk_map %>%
#               left_join(meow_cells, by = 'cell_id') %>%
#               mutate(rgn_name = ifelse(is.na(meow_id), 'global', rgn_name),
#                      priority = ifelse(rgn_name == 'global', 100, priority)) %>%
#               filter(iucn_rgn == rgn_name) %>%
#               group_by(cell_id) %>%
#               filter(priority == min(priority)) %>%
#               ungroup()
#             ### NOTE: at this point, still possible to have multiple regional
#             ### assessments, if priorities are the same (e.g. Europe and Pan Africa,
#             ### or overlaps around Africa).  Those values will be averaged in
#             ### the group_by() below.
#           }
#           
#           ### select down to main columns; also, if presence == 5 (extinct), 
#           ### adjust category and trend scores.
#           spp_risk_sum <- spp_risk_map %>%
#             dplyr::select(cell_id, presence, iucn_sid, iucn_rgn, cat_score, trend_score) %>%
#             distinct() %>%
#             mutate(cat_score   = ifelse(presence == 5, 1, cat_score),        
#                      ### fix local extinctions
#                    trend_score = ifelse(presence == 5, NA, trend_score)) %>% 
#                      ### fix local extinctions
#             group_by(cell_id, iucn_sid) %>%
#             summarize(cat_score   = mean(cat_score, na.rm = TRUE),
#                       trend_score = mean(trend_score, na.rm = TRUE)) %>%
#             ungroup() %>%
#             mutate(trend_score = ifelse(is.nan(trend_score), NA, trend_score))
#           
#           return(spp_risk_sum)
#         }) ### end of mclapply over all species in taxonomic group
#     
#       ##########################################################.
#       ### Processing cell calculations for group -----
#       ##########################################################.
#     
#       ### Set up for keyed data.table merging: key for iucn_sid, cell_id
#       cat_msg('...binding rows to data.frame...')
#   
#       taxon_risk_map <- bind_rows(taxon_cells_list) ### this is pretty fast!
#       
#       cat_msg('...summarizing...')
#       taxon_risk_summary <- taxon_risk_map %>%
#         group_by(cell_id) %>%
#         summarize(mean_risk   = mean(cat_score), 
#                     ### NA categories already filtered out
#                   var_risk    = var(cat_score),
#                   n_spp_threatened = sum(cat_score >= 0.4 & cat_score < 1),  
#                   n_spp_risk  = n(),
#                   mean_trend  = mean(trend_score, na.rm = TRUE), 
#                     ### not every species has a trend; na.rm = TRUE
#                   mean_trend  = ifelse(is.nan(mean_trend), NA, mean_trend),
#                   n_spp_trend = sum(!is.na(trend_score)))
#     } ### end of empty set check
#     cat_msg('...writing file', taxon_sum_file, '...')
#     write_csv(taxon_risk_summary, taxon_sum_file)
# 
#   } else { ### end of if statement checking whether file exists for this taxon
#     cat_msg('Found file ', taxon_sum_file, '... skipping process...')
#   }
#   
# } ### end of taxonomic group loop
#   
```

### Setup: `'int/polygon_areas_2025-1.csv`

*This is just a sanity-check for raster-calculated ranges!*

This may *over* predict range for many species whose range is limited by depth, due to polygon buffering.

``` {r calculate range from polygons, eval = FALSE}
# 
# spp_maps <- read_csv(file.path(dir_data, sprintf('spp_marine_maps_%s.csv', api_version)),
#                      col_types = 'ddcicccl') 
# # %>% dplyr::filter(!shp_iucn_sid %in% c(246014055))
# 
# spp_dbfs <- spp_maps$dbf_file %>%
#   unique()
# 
# spp_shps <- spp_dbfs %>%
#   str_replace('\\.dbf$', '.shp')
# 
# # Goal file
# polygon_areas_file <- file.path(dir_setup, sprintf('int/polygon_areas_%s.csv', api_version))
# 
# reload <- FALSE
# 
# if(!file.exists(polygon_areas_file) | reload) {
#   
#   poly_range_list <- list()
#   for(i in seq_along(spp_shps)){
#      x <- spp_shps[i]
#                             cat_msg(i, ': Processing ', x)
#                             shp <- st_read(x)
# 
#                             if('id_no' %in% names(shp)) {
#                               shp <- shp %>%
#                                 rename(iucn_sid = id_no)
#                             }
# 
#                             if('sisid' %in% names(shp)){
#                               shp <- shp %>%
#                                 rename(iucn_sid = sisid)
#                             }
# 
#                             shp <- shp %>%
#                               filter(iucn_sid %in% spp_maps$iucn_sid)
#                                 ### NOTE: this will miss subpops
#                             
#                             ## Checking and attempting to fix invalid geometries
#                             if(any(!st_is_valid(shp))) {
#                               invalid <- which(!st_is_valid(shp))
#                               
#                               shp[invalid,] <- st_make_valid(shp[invalid,])
#                             }
#                             ## If any are still invalid after using st_make_valid, don't use s2 and
#                             ## hopefully it will work
#                             if(any(!st_is_valid(shp))) {
#                               sf_use_s2(FALSE)
#                               invalid <- which(!st_is_valid(shp))
#                               
#                               shp[invalid,] <- st_make_valid(shp[invalid,])
#                             }
#                             
#                             range_df <- data.frame(iucn_sid = shp$iucn_sid,
#                                                    poly_range_km2 = st_area(shp)) %>%
#                               mutate(poly_range_km2 = as.numeric(poly_range_km2 / 1e6))
#                             cat_msg(i, ': Finished ', x)
#                             
#                       poly_range_list[[length(poly_range_list)+1]] = range_df     
#     
#   }
#   
#   
#   poly_range_df <- poly_range_list %>%
#     setNames(basename(spp_shps)) %>%
#     bind_rows(.id = 'shp_file') %>%
#     group_by(iucn_sid) %>%
#     summarize(poly_range_km2 = round(max(poly_range_km2), 3))
# 
#   write_csv(poly_range_df, polygon_areas_file)
# }
#     
```

### Check: plot raster-calculated ranges vs polygon-calculated ranges

Where dots fall on the line, species ranges by raster and by polygon agree.  Where dots fall above the line, polygon range is larger than raster range - likely due to clipping of land-based ranges, and mostly birds.  Where dots fall below the line, polygon range is smaller than the raster range - in these cases, the polygon range may be a combination of multiple polygons, e.g. subpops or seasonal areas; the code above simply takes the largest polygon per ID as a rough check.

``` {r, eval = FALSE}
# 
# polygon_areas_file <- file.path(dir_setup, sprintf('int/polygon_areas_%s.csv', api_version))
# 
# range_df <- read_csv(polygon_areas_file, col_types = 'dd') %>%
#   full_join(read_csv(spp_range_file, col_types = 'ddcc') %>%
#               mutate(shp_file = str_replace(dbf_file, 'dbf', 'shp')), 
#             by = 'iucn_sid') %>%
#   mutate(birds = str_detect(shp_file, 'bli'))
# 
# raster_zeros <- range_df %>% 
#   filter(range_km2 %in% c(0))
# 
# zeros_ids <- data.frame(iucn_sid = raster_zeros$iucn_sid, poly_range_km2 = raster_zeros$poly_range_km2, rast_range_km2 = raster_zeros$range_km2)
# 
# ######## Here I left off with adding my polygon range estimates to a dataframe for the species that had '0' ranges when calculated as rasters. I am going to use the polygon range estimates for the raster range estimates and potentially 
# write that out or see how the scores change when it is added versus not added back to the original raster dataframe. 
# 
# range_compare_scatter <- ggplot(range_df, aes(x = range_km2, y = poly_range_km2)) +
#   ggtheme_plot() +
#   geom_abline(slope = 1, intercept = 0, color = 'green4') + 
#   geom_point(aes(color = birds, label = shp_file, label2 = iucn_sid), alpha = .6)
# 
# print(range_compare_scatter)
```

