---
title: 'Process IUCN Spatial Data'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

library(raster)
library(sf)
library(fasterize)
library(dplyr)
library(here)
library(ggplot2)

source(here('workflow/R/common.R'))

goal     <- 'spp'
scenario <- 'v2025'
dir_goal <- here('globalprep', goal, scenario)

### goal specific folders and info
dir_data  <- file.path(dir_goal, '_data')
dir_setup <- file.path(dir_goal, '_setup')
dir_anx  <- file.path(dir_M, 'git-annex/globalprep')
dir_goal_anx <- file.path(dir_anx, goal, scenario, 'spp_risk_dists')
dir_raw_data_anx <- file.path(dir_anx, '_raw_data')
dir_spatial <- file.path(dir_goal, '_spatial')

source(file.path(dir_setup, 'common_fxns.R'))

dir_bli <- file.path(dir_M, 'git-annex/globalprep/_raw_data/birdlife_intl/d2025')
dir_shp <- file.path(dir_M, 'git-annex/globalprep/_raw_data/iucn_spp/d2025-1')
  ### in this folder are shapefiles at a taxonomic level, both from
  ### IUCN spatial data download (version 2025-1)
```

# Summary

Using a set of IUCN species range maps, rasterize each species to 10 km x 10 km raster using `fasterize`.  Use `presence` field from shapefile.

* Subpopulation polygons must be identified and rasterized separately from the parent polygon; this must be done by sciname and subpop fields since the polygon IDs are based upon the parent ID.
* Regional assessments need not be determined at this stage - the ID numbers match the global ID numbers (including subpops).

# Data source

IUCN Red List: Spatial Data Download
IUCN: Gina Ralph direct communication

# Methods

## Read spp shapes, correct subpop IDs, `fasterize()`, depth clip, save to csv

We will loop over each species in each shapefile and rasterize separately, using `sf` and `fasterize` packages.  

* From the full map list, filter to a single shapefile
* Load shapefile using `st_read`, and correct subpop IDs from `shp_iucn_sid` to `iucn_sid`
* Loop over each `iucn_sid` in the shapefile, rasterizing (`fasterize()`) to 10 km^2 resolution, using "presence" field. 
    * clip to neritic (<=200 m) and shallow (<=60 m) depth raster if appropriate.  Otherwise mask to bathy raster.  Since bathy raster was created by masking to area raster, cells with any marine presence will be kept but any non-marine cells will be dropped.
    * Save as .tif and .csv, and compare average file sizes.  .csv easier to work with, but might be significantly larger than .tifs in the long run.
        * note: no longer saving as .tif for speed and file size - just use .csv instead!
    * use `mclapply()` to speed this up.
    
#### Annex: `spp_risk_dists/spp_rasters/iucn_sid_%s.csv`
    
``` {r rasterize iucn and clip and save to csv}

reload <- FALSE

maps_to_rasterize <- read_csv(file.path(dir_data,
                                        sprintf('spp_marine_maps_%s.csv', api_version)),
                              col_types = 'ddcicccl') %>%
  mutate(shp_file = str_replace(dbf_file, 'dbf$', 'shp'))

# Rast_base for cell IDs
rast_base <- raster(file.path(dir_spatial, 'cell_id_rast.tif'))

### If some species already processed, remove from the list to process.
if(reload == FALSE) {
  maps_already_rasterized <- list.files(file.path(dir_goal_anx, 'spp_rasters'),
                                        pattern = '.csv') %>%
    str_replace_all('iucn_sid_|.csv$', '') %>%
    as.integer()

  maps_to_rasterize <- maps_to_rasterize %>%
    filter(!iucn_sid %in% maps_already_rasterized)
}

### for selectively rerunning taxonomic groups
# maps_to_rasterize <- maps_to_rasterize %>%
#   filter(str_detect(dbf_file, 'birdlife'))

# These will be used as masks
rast_bathy <- raster(file.path(dir_spatial,
                               'bathy_rast.tif'))
rast_neritic <- raster(file.path(dir_spatial,
                                 'bathy_rast_neritic.tif'))
rast_shallow <- raster(file.path(dir_spatial,
                                 'bathy_rast_shallow.tif'))

if(nrow(maps_to_rasterize) == 0) { ### all maps accounted for as .csvs
  
  cat_msg('reload == ', reload, '... No maps to process...')
  
} else {
  
  cat_msg('reload == ', reload, '... Maps to process: ', nrow(maps_to_rasterize))
  
  ################################################################.
  ### Loop over each distinct shapefile with species range maps
  ################################################################.

  # bli_already_done <- read_csv(file.path("globalprep/spp/v2025/_setup", "int/bli_already_done.csv"))  
  # bli_done_vec <- bli_already_done$iucn_sid
  # 
  # maps_to_rasterize <- maps_to_rasterize %>%
  #   filter((iucn_sid %in% bli_done_vec))
  
  shps <- maps_to_rasterize$shp_file %>% unique()
  # shps <- shps[1]
  for(i in seq_along(shps)) {
    # i <- 1
    
    shp <- shps[i]
    
    maps_in_shp <- maps_to_rasterize %>%
      filter(shp_file == shp)
    
    id_fix <- maps_in_shp %>%
      dplyr::select(shp_iucn_sid, iucn_sid, subpop, max_depth) %>% 
      distinct()
    
    cat_msg(i, ' of ', length(shps), ': reading ', basename(shp), ' from: \n  ', shp)
    
    polys_all_geometry <- read_sf(shp, type = 6) %>%
      clean_df_names()
    
    polys_all <- polys_all_geometry %>% 
      st_set_geometry(NULL)
    
      ### we will check geoms and fix them inside the mclapply, 
      ### and then reproject, one species at a time
    
    if('sci_name' %in% names(polys_all)) {
      polys_all <- polys_all %>% 
        rename(sciname = sci_name) 
    } 

    if('binomial' %in% names(polys_all)) {
      polys_all <- polys_all %>%
        rename(sciname = binomial)
    }
    
    if(!'subpop' %in% names(polys_all)) {
      polys_all$subpop <- NA_character_
      ### if shape doesn't have subpop column, add it as NA
    }
    if('id_no' %in% names(polys_all)) {
      polys_all <- polys_all %>%
        rename(iucn_sid = id_no)
    }
    
    if('sisid' %in% names(polys_all)){
      polys_all <- polys_all %>%
        rename(iucn_sid = sisid)
    }
    
    if('presenc' %in% names(polys_all)){
      polys_all <- polys_all %>%
        rename(presence = presenc) 
    }
    
    if(!'presence' %in% names(polys_all)) {
      polys_all <- polys_all %>%
        mutate(presence = 1) 
    }
    
    polys_all <- polys_all %>%
        st_sf(geometry = st_geometry(polys_all_geometry))
    
    if(identical(st_geometry(polys_all), st_geometry(polys_all_geometry)) == FALSE) {
      cat_msg("HALT: GEOMETRIES DO NOT MATCH FOR", i, ' of ', length(shps), 'in', basename(shp), ' from: \n  ', shp)
    }
    
    polys_match <- polys_all %>%
      dplyr::select(shp_iucn_sid = iucn_sid, sciname, subpop, presence, geometry) %>%
      mutate(presence = ifelse(presence == 0, 1, presence),
             subpop   = as.character(subpop)) %>%
      inner_join(id_fix, by = c('shp_iucn_sid', 'subpop')) 
    
    spp_ids <- maps_in_shp$iucn_sid %>% 
      sort() %>% 
      unique()
    
  
    ####################################################################.
    ### In each shapefile, loop over each species ID using mclapply().
    ####################################################################.
    
    cat_msg('Processing ', basename(shp), ' with ', length(spp_ids), ' species...')
    
    # system.time({
      tmp <- parallel::mclapply(seq_along(spp_ids),
                                mc.cores = 12, 
                                FUN = function(x) {
         #x <- 81:128
        spp <- spp_ids[x]
        
        cat_msg(x, ' of ', length(spp_ids), ': Processing ', spp, ' in ', basename(shp),
                ' (group ', i, ' of ', length(shps), ')...\n')
        
        spp_shp <- polys_match %>%
          filter(iucn_sid == spp)
        
        spp_shp <- valid_check(spp_shp)
          ### if invalid geom, and bounds exceeded, buffer to 0
        spp_shp <- spp_shp %>%
          clip_to_globe() %>%
            ### indiv files should have adjusted boundaries already; only 
            ### files from Spatial Data Download should run into this
          st_transform(gp_proj4)
        
        spp_rast <- fasterize::fasterize(spp_shp, rast_base, field = 'presence', fun = 'min')
        
        ### depth clip if necessary; otherwise clip to bathy raster (which previously
        ### was clipped to area raster - so cells with any marine area will be kept,
        ### and non-marine cells will be dropped)
        if(unique(spp_shp$max_depth) == '< 20 m') {
          spp_rast <- mask(spp_rast, rast_shallow)
        } else if(unique(spp_shp$max_depth) == '< 200 m') {
          spp_rast <- mask(spp_rast, rast_neritic)
        } else {
          spp_rast <- mask(spp_rast, rast_bathy)
        }
        ### write out as a raster:
        # rast_file <- file.path(dir_o_anx, sprintf('spp_rasters/iucn_sid_%s.tif', spp))
        # raster::writeRaster(spp_rast, rast_file, overwrite = TRUE)
        ### See note below about file size; rasters are larger than csvs in general
        
        ### convert to dataframe and write out as a csv:
        spp_present <- data.frame(cell_id  = values(rast_base),
                                  presence = values(spp_rast)) %>%
          filter(!is.na(presence))
        
        if(nrow(spp_present) == 0) {
    cat_msg('Species ID ', spp, ' resulted in a zero-length dataframe.')
        }
        
        write_csv(spp_present, file.path(dir_goal_anx, 'spp_rasters',
                                         sprintf('iucn_sid_%s.csv', spp)))
        
        return(NA)
      }) ### end of mclapply FUN definition
    # }) ### end of system.time call
  } ### end of for loop over each species group
} ### end of "if" check to make sure there are any maps to rasterize

## Note: v2020: This got to the end and didn't indicate it was finished. I only knew because I periodically stopped the loop and checked after it got to the last couple of sfs and then looked at maps_to_rasterize that it was finished. 
## v2021: I had to add some extra if statements to fix column name issues with the new bli data.

```

``` {r file size testing}
# 
# maps <- read_csv(file.path(dir_data,
#                            sprintf('spp_marine_maps_%s.csv', api_version)),
#                  col_types = 'ddciccc') %>%
#   dplyr::select(spp_group = dbf_file, iucn_sid, sciname, subpop) %>%
#   mutate(spp_group = str_replace(spp_group, '.dbf', '')) %>%
#   distinct()
# 
# csvs <- list.files(file.path(dir_goal_anx, 'spp_rasters')) %>%
#   str_replace_all('iucn_sid_|.csv', '')
# 
# x <- list.files(file.path(dir_goal_anx, 'spp_rasters'),
#                 full.names = TRUE)
# x <- x[str_detect(x, paste0(csvs, collapse = '|'))]
# 
# y <- file.info(x) %>%
#   mutate(f = basename(rownames(.)),
#          type = ifelse(str_detect(f, 'csv$'), 'csv', 'tif'),
#          iucn_sid = str_replace_all(f, 'iucn_sid_|.(csv|tif)', ''),
#          iucn_sid = as.integer(iucn_sid)) %>%
#   left_join(maps, by = 'iucn_sid')
# 
# z <- y %>%
#   group_by(type, spp_group) %>%
#   summarize(mean_size = mean(size), n_spp = n()) %>%
#   spread(type, mean_size) %>%
#   # mutate(ratio = tif / csv) %>%
#   ungroup() %>%
#   arrange(desc(csv))
# 
#  #sum(z$tif * z$n_spp) ### 0
#  #sum(z$csv * z$n_spp) ### 3449809892
# 
# 
# zz <- data.frame(spp_group = 'TOTAL: all available spp maps',
#                 n_spp = sum(z$n_spp),
#                 csv   = sum(z$csv * z$n_spp)) %>%
#   bind_rows(z)
# 
# DT::datatable(zz, caption = 'Mean file size by group and type')

```

## Generate species group maps - range-rarity weighting per cell

* Determine species ranges using ocean-area data.frame and cell location data.frame.  Create an overall data.frame and save as .csv.
    * Compare to polygon range areas
* Collect all taxon species into a single data.frame.
* Join species extinction risk and trend data.frame to species-cell data.frame, including regional assessments.
* Join species range area 
* By cell, calculate range-rarity-weighted mean extinction risk, variance of extinction risk, trend, threatened species count, rr_richness, rr_trend_richness (where rr_richness is range-rarity weighted species richness)
    * When aggregating further, use rr_richness to weight values.
    
**NOTE:** as above, because the summary files are likely to be very large for globe-spanning taxa, save these outputs outside of GitHub.

This calculated range includes ONLY ocean area. For species such as birds whose range is primarily over land but extends somewhat out over the ocean, this will underpredict range according to the polygons.  For marine species whose range is extremely tiny, the raster may miss the presence of a species entirely.  These ranges will be checked against the polygon areas as a quality check.

### Data: `iucn_spp_range_area_2025-1.csv`

``` {r calculate_ranges_from_raster_maps}
# Goal file
spp_range_file <- file.path(dir_data,
                            sprintf('iucn_spp_range_area_%s.csv', api_version))

reload <- FALSE

# Load area raster, convert to data frame
area_rast <- rast(file.path(dir_spatial, 'ocean_area_rast.tif'))
area_df <- data.frame(cell_id = 1:(nrow(area_rast)*ncol(area_rast)),
                      area_km2 = terra::values(area_rast)) %>%
  rename(area_km2 = ocean_area_rast)

# Load csvs of marine species range maps
spp_maps <- read_csv(file.path(dir_data, 
                               sprintf('spp_marine_maps_%s.csv', api_version)),
                     col_types = 'ddcicccl') 
# %>% dplyr::filter(!shp_iucn_sid %in% c(246014055))

spp_ids <- spp_maps$iucn_sid %>%
  unique() %>%
  sort()

if(reload == FALSE & file.exists(spp_range_file)) {
  cat('dont redo spp already calculated...  remove those from the running')
  spp_range_list_done <- read_csv(spp_range_file, col_types = 'ddcc')
  spp_ids <- spp_ids[!spp_ids %in% spp_range_list_done$iucn_sid]
}

message('Processing ', length(spp_ids), ' species range maps.')

## Note: if the birdlife international data is not updated (and stored in mazu under the name iucn_sid_xxxxxxxxx), this function will not work properly. 

## If the if() statement is not working properly (getting Error: file iucn_sid_x.csv does not exist), see the below chunk for troubleshooting which files are causing issues. 
if(length(spp_ids) > 0) { ### any more spp to calculate?
  system.time({
  spp_range_list <- parallel::mclapply(seq_along(spp_ids), mc.cores = 12,
                                       FUN = function(i) {
    x <- spp_ids[i]
    cat_msg(i, ': Processing species ', x)
    spp_map <- read_csv(file.path(dir_goal_anx, 'spp_rasters',
                                  sprintf('iucn_sid_%s.csv', x)),
                        col_types = 'di')
    
    spp_map_area <- spp_map %>% 
      # filter(presence != 5) %>% ### exclude extinct areas
      inner_join(area_df, by = 'cell_id')
    
    spp_range <- data.frame(iucn_sid = x,
                            range_km2 = sum(spp_map_area$area_km2, na.rm = TRUE))
    # spp_range_list[[i]] <- spp_range
    return(spp_range)
  # }
  }) ### end of mclapply
  }) ### end of system.time

  spp_range_df <- bind_rows(spp_range_list) %>%
    left_join(spp_maps %>% 
                dplyr::select(iucn_sid, sciname, dbf_file) %>%
                distinct(),
              by = 'iucn_sid')
} else {
  spp_range_df <- data.frame() ### return an empty data.frame
}
  
if(reload == FALSE & exists('spp_range_list_done')) {
  ### add new area calcs at the end of the dataframe of spp already done
  spp_range_df <- spp_range_df %>%
    bind_rows(spp_range_list_done)
}

write_csv(spp_range_df, spp_range_file)

```

### Chunk for Troubleshooting and Identifying file issues from Chunk 3 

```{r}
# This function determines which iucn_sids exist in spp_maps (from API) but not in v202X/spp_risk_dists/spp_rasters (from raw data download. It does virtually the same thing as the above function but will store the ID's that don't exist to identify which polygon download they come from (IUCN Red List)
fix_this <- function(spp_nums) { # spp_nums are the spp_ids from the above chunk
  results <- list() # Create empty list
  spp_list <- spp_nums # Define variable
  for(i in seq_along(spp_list)) { # spp_list contains species' "iucn_sid" which is what "x" becomes
    # obtain iucn_sid
    x <- spp_list[i] 
    # Set file path to spp_rasters for above iucn_sid
    spp_file <- file.path(dir_goal_anx, 'spp_rasters',
                         sprintf('iucn_sid_%s.csv', x)) # Create a file for each iucn_sid
    # Check if file exists and attempt to read it if it does
    if(file.exists(spp_file)) {
      spp_map <- read_csv(spp_file, col_types = 'di')
      
      spp_map_area <- spp_map %>% 
        # filter(presence != 5) %>% ### exclude extinct areas
        inner_join(area_df, by='cell_id')
      
      spp_range <- data.frame(iucn_sid = x,
                              range_km2 = sum(spp_map_area$area_km2, na.rm=TRUE))
    } else { # If it doesn't exist, still include it so that we can identify where it should come from
      cat_msg("Iucn_sid:", x, "file does not exist")
      spp_range <- data.frame(iucn_sid = x, 
                              range_km2 = NA)
    }
    # Store output in results list 
    results[[i]] <- spp_range
  }
  return(results)
}

fxn_test_results <- fix_this(spp_ids) # according to the math, 50 items took about 20 seconds to run, wich = 0.4s/file*12240 files = 4896s/60s/m = 81.6m / 60m/h = 1.36 hrs
## Started at 5:59pm
## Finish at 6:48

# Following above chunk and creating a dataframe of the bound list from above
results_df <- bind_rows(fxn_test_results)

# Identify which iucn_sid's files did not exist in current v202X/spp_risk_dists/spp_rasters
results_na <- results_df %>% 
  filter(range_km2 %in% c(NA))

# Define old spp_maps data to see if these iucn_sids are present in previous year 
old_map_info <- read_csv(file.path(dir_data,
  sprintf('spp_marine_maps_%s.csv', "2022-1"))) %>%
  dplyr::select(-presence) %>%
  distinct()

# Filter down old map information to determine the files from which the missing iucn_sids came from (for 2024 these missing iucn_sids came from the birdlife international 2021 which makes sense because we didn't update that data for this assessment).
# For future reference, you can use this dataframe to see which taxa names iucn_sids came from that are not present in your data. If you don't recognize them then you should go back to the IUCN Redlist website and download them, then go back to `5c_rasterize_sppp_shps.Rmd`
file_check_df <- old_map_info %>% 
  filter(iucn_sid %in% results_na$iucn_sid)



# This function was used for for rewriting birdlife international files from v2022 spp_rasters to v2024 spp_rasters. It can be used for future years by utilizing the prev_year variable


prev_year = paste0("v", 2022) # Year that you're referencing spp_rasters for 
vector = results_na$iucn_sid # Vector of iucn_sids that you've identified need to be rewritten from prev_year

rewrite_raster_function <- function(vector, prev_year) {
  files_to_rewrite <- vector
  for(i in seq_along(files_to_rewrite)) {
    x <- files_to_rewrite[i] # obtain iucn_sid
    # Identify file 
    spp_file_in <- file.path(sprintf("/home/shares/ohi/git-annex/globalprep/spp/%s/spp_risk_dists/spp_rasters/iucn_sid_%s.csv", prev_year, x))
    
    if(!file.exists(spp_file_in)) { # Send message if it doesn't exist
      
      cat("Iucn sid", x, "not found in", prev_year, "folder", "\n")
      
    } else { # otherwise read it in and rewrite it to current year directory 
      
      cat("Success! Reading and writing iucn sid:", x, "\n")
    # Read file in 
    spp_file <- read_csv(spp_file_in)
    
    spp_file_out <- file.path(dir_goal_anx, 'spp_rasters',
                         sprintf('iucn_sid_%s.csv', x)) # Create a file for each iucn_sid
    
  write_csv(spp_file, spp_file_out)
    }
  }
}

# Run! 
rewrite_raster_function(vector = vector, prev_year = prev_year)

# Once this is done, you can go back up and rerun the above function
```

