---
title: 'Set up IUCN marine species data'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE, eval = FALSE)
```

```{r}
#install.packages("rredlist")
library(rredlist)
library(dplyr)
library(tidyr)
library(tidyverse)
library(here)

#source('https://raw.githubusercontent.com/oharac/src/master/R/common.R')
source(here('workflow/R/common.R'))

goal     <- 'spp'
scenario <- 'v2025'

dir_goal <- here('globalprep', goal, scenario)

### goal specific folders and info
dir_anx  <- file.path(dir_M, 'git-annex/globalprep')
dir_setup   <- file.path(dir_goal, '_setup')
dir_goal_anx <- file.path(dir_anx, goal, scenario, 'spp_risk_dists')

source(file.path(dir_setup, 'common_fxns.R'))

```

# Summary

Get IUCN current risk assessments and historical assessments over time, for all IUCN marine species.  This method is not reliant on any information from AquaMaps.

# Data Sources

## IUCN Red List

To access the API v4, you will need to sign up for a key on the [IUCN website](https://api.iucnredlist.org/). This key can be added to the local .renviron on your machine and retrieved using the code below.

```{r set up API}
# Save API key
rl_use_iucn()
api_version <- rl_version()

# or
rl_use_api_key("YOUR_API_KEY_HERE")
```

# Methods

This file sets up the following .csv files, saved in the `_data` folder:

For *global* assessments:
- `iucn_risk_current_2025-1.csv`
- `iucn_risk_timeseries_2025-1.csv`
- `spp_marine_from_api_2025-1.csv`

For *regional* assessments:
- `iucn_risk_rgn_current_2025-1.csv`
- `iucn_risk_ts_rgn_2025-1.csv`

For trends:
- `iucn_trend_by_spp_2025-1.csv`

## Marine Species List

First, we need to establish the list of marine species IDs we are interested in. We do this by using rl_systems() to search for all species coded for the marine system ("2").

```{r marine_species_list}
# Save IUCN systems
iucn_systems <- rl_systems()$systems

# Unlist description column
iucn_systems$description <- iucn_systems$description$en

# Search for species by marine system
marine_system <- rl_systems("2")$assessments 

# Clean up data frame
marine_system <- marine_system %>%
  rename(iucn_sid = sis_taxon_id, 
         sciname = taxon_scientific_name,
         category = red_list_category_code,
         year = year_published)

# Save list of unique taxon IDs
marine_species_list <- unique(marine_system$iucn_sid)

# Save for time
write_csv(marine_system, file.path(dir_goal, sprintf('_data/iucn/marine-system-2025.csv')))

# Load for time
#marine_system <- read_csv(file.path(dir_goal, sprintf('_data/iucn/marine-system-2025.csv')), show_col_types = FALSE)

# 2025: 19,944 species
```

## *Global* Assessments

### Data: `iucn_risk_current_2025-1.csv`

Save the IUCN categories for all current *global* assessments of marine species.

```{r}
# Reclassify old codes to current codes
cat_lookup <- read_csv(file.path(dir_setup, 'raw', 'risk_code_lookup.csv'), show_col_types = FALSE)

# Create data frame for layer
marine_species_risk <- marine_system %>%
  # Get current assessments
  filter(latest == TRUE) %>%
  # Unnest the scopes variable
  unnest(scopes, names_sep = "_") %>%
  # Filter for global scopes
  filter (scopes_code == "1") %>%
  # Reclassify codes
  left_join(cat_lookup, by = c("category" = "code"), relationship = "many-to-many") %>%
  rename(old_cat = category,
         cat = code_current) %>%
  # Add version column
  mutate(iucn_version = "2025-1") %>%
  select(iucn_sid, sciname, year, old_cat, cat, cat_score, iucn_version) #criteria left out for now

# Check that no IDs are repeated (should return 0 rows)
marine_species_risk %>%
  group_by(iucn_sid) %>%
  filter(n > 1)

# Write csv
write_csv(marine_species_risk, file.path(dir_goal, '_data', 
                                         sprintf('iucn_risk_current_%s.csv', api_version)))
```

### Data: `iucn_risk_timeseries_2025-1.csv`

Save the IUCN categories for all available *global* assessments over time of marine species.

```{r}
# iucn_risk_timeseries_2025-1.csv
marine_species_risk_timeseries <- marine_system %>%
  # Unnest the scopes variable
  unnest(scopes, names_sep = "_") %>%
  # Filter for global scopes
  filter (scopes_code == "1") %>%
  # Reclassify codes
  left_join(cat_lookup, by = c("category" = "code"), relationship = "many-to-many") %>%
  rename(old_cat = category,
         cat = code_current) %>%
  # Add version column
  mutate(iucn_version = "2025-1") %>%
  select(iucn_sid, year, cat_ts = cat, cat_ts_score = cat_score, iucn_version) #criteria left out for now
# filter for more than one assessment?
  #group_by(iucn_sid) %>%
  #filter(n() >= 2) %>%
  #ungroup()

# Write csv
write_csv(marine_species_risk_timeseries, file.path(dir_goal, '_data', 
                                         sprintf('iucn_risk_timeseries_%s.csv', api_version)))
```

Check out the results for 2025, assess differences from 2024.

```{r}
# Compare to last year
nrow(marine_species_risk) # 19,357

# Import last year (change year as needed)
marine_species_risk_2024 <- read_csv(file.path(dir_goal, sprintf('_data/iucn_risk_current_%s.csv', "2024-1")), show_col_types = FALSE)

nrow(marine_species_risk_2024) # 18,599

length(setdiff(unique(marine_species_risk$iucn_sid), unique(marine_species_risk_2024$iucn_sid))) # 2,648 new species

# Look at species added in new OHI assessment
new_species <- marine_species_risk %>%
  anti_join(marine_species_risk_2024, by = "iucn_sid") %>%
  # Check to see if new species are all from same habitat? published year?
  left_join(marine_species, by = c("iucn_sid", "year", "sciname"))
```

### Data: `spp_marine_from_api_2025-1.csv`

Using the `rl_habitats()` function, get the species list for species coded with the following marine habitat codes: 9-12, plus 15.11, 15.12, 15.13.  Note category 13 is Marine Coastal/Supratidal, but includes many species whose "marine" dependence is only incidental. We include seacliffs and offshore islands (13.1) from category 13. If these species do not show up in category 12 (marine intertidal) then they are assumed to not actually depend on marine habitats.

The list can be found here (make sure to compare the webpage to the final df, to make sure it matches up): https://www.iucnredlist.org/resources/habitat-classification-scheme

Some species show up over multiple habitats (and thus depth ratings), so I chose to assign the deepest rating.

| habitat type   | habitat code   | depth classification  |
| -------------- | -------------- | --------------------- |
| deep benthic   | 10_3, 10_4     | "deep oceanic"        |
| deep oceanic   | 10_ , 11_      | "200 m +"             |
| neritic        | 9_, 15_        | "< 200 m"             |
| intertidal     | 12_, 13_       | "<20 m"               |

```{r}
# Get a list of habitats
iucn_habitats <- rl_habitats()$habitats

# Unlist the description column
iucn_habitats$description <- iucn_habitats$description$en

# Select desired marine habitats (noted above)
marine_habitats <- iucn_habitats %>%
  filter(str_detect(code, "^(9_|10_|11_|12_)") | 
           code %in% c("13_1", "15_11", "15_12", "15_13"))

# marine_habitats <- iucn_habitats %>%
#   dplyr::filter(str_detect(tolower(description), "marine|reef|lagoon") &
#                   !str_detect(tolower(description), "supratidal")) # change to include 13_1

# Initiate empty list
marine_habitats_spp <- list()

# Pull species for each marine habitat
for (i in marine_habitats$code) { 
  
  # Save habitat name
  habitat_name <- marine_habitats$description[marine_habitats$code == i]
  
  # Get list of assesments
  habitat_assessments <- rl_habitats(code = i)$assessments
  
  # Append to list
  marine_habitats_spp[[habitat_name]] <- dplyr::as_tibble(habitat_assessments)
}

# Combine individual habitat species lists into one master list
marine_species_habitats <- bind_rows(marine_habitats_spp) %>%
  rename(iucn_sid = sis_taxon_id, 
         sciname = taxon_scientific_name,
         category = red_list_category_code,
         year = year_published) 

# Save for time
write_csv(marine_species_habitats, file.path(dir_goal, sprintf('_data/iucn/marine-species-habitats-2025.csv')))

# Load for time
#marine_species_habitats <- read_csv(file.path(dir_goal, sprintf('_data/iucn/marine-species-habitats-2025.csv')), show_col_types = FALSE)

# Filter for species identified in marine system search, global code
marine_species_habitats_global <- marine_species_habitats %>%
  # Filter for latest assessments
  filter(latest == TRUE) %>%
  # Unnest scopes variable
  unnest(scopes, names_sep = "_") %>%
  # Filter for global
  filter(scopes_code == "1") %>%
  # Filter for marine species
  filter(iucn_sid %in% marine_species_list) %>%
  separate(code, c('cat', 'subcat1', 'subcat2'),
           remove = FALSE, convert = TRUE, fill = "right") %>%
  arrange(cat, subcat1, subcat2) %>%
  # Assign descriptions to each cat
  mutate(
    max_depth = factor(case_when(
      cat == 10 & subcat1 %in% 3:4         ~ "deep oceanic",
      cat %in% c(10, 11)                   ~ "200 m +",
      cat %in% c(9, 15)                    ~ "< 200 m",
      cat %in% c(12, 13)                   ~ "< 20 m",
      TRUE                                 ~ "unknown"),
    # To make it easier to pick the deepest (min) depth rating
    levels = c("deep oceanic", "200 m +", "< 200 m", "< 20 m", "unknown"))) %>%
  left_join(iucn_habitats, by = "code")

# spp_marine_from_api_2025-1.csv
habs <- marine_species_risk %>%
  # Add depth descriptions
  left_join(marine_species_habitats_global, by = c("iucn_sid", "sciname")) %>%
  select(iucn_sid, code, max_depth) %>%
  group_by(iucn_sid) %>%
  reframe(
    habs = paste(unique(code), collapse = ", "),
    # Pick the deepest rating (the lowest (min) level)
    max_depth = as.character(max_depth[which.min(as.integer(max_depth))])) %>%
  ungroup()

# Write csv
write_csv(habs, file.path(dir_goal, '_data',
                          sprintf('spp_marine_from_api_%s.csv', api_version)))
```

Check out the results for 2025, assess differences from 2024.

```{r}
nrow(habs) # 19,899
nrow(marine_species_risk) # 19,529

# Check species that are in habs but not in global
check <- anti_join(habs, marine_species_risk, by = "iucn_sid")

## Interesting: iucn_sid 155168 appears in marine_species_list and marine_species (obtained from rl_habitats(), but does not appear in global_marine and thus marine_species_risk despite having a scope of 1 (global) listed in marine_species)

habs_2024 <- read_csv(file.path(dir_goal, sprintf('_data/spp_marine_from_api_%s.csv', "2024-1")), show_col_types = FALSE)
nrow(habs_2024) # 18,590
```

## Regional Assessments

Using the `rl_scopes()` function, get the species list for species coded with the following regions: Europe (2), Mediterranean (4), Western Africa (8), S. Africa FW (9), Pan-Africa (10), Central Africa (16), Northeastern Africa (17), Eastern Africa (18), Northern Africa (21), Gulf of Mexico (45433062), Caribbean (45433063), Persian Gulf (45433064), Arabian Sea (100765562). 

Essentially all *but* Global (1).

```{r}
# Get list of scopes (regions)
iucn_scopes <- rl_scopes()$scopes

# Unlist description
iucn_scopes$description <- iucn_scopes$description$en

# Remove "Global" (see above)
scopes <- iucn_scopes %>%
  filter(!description == "Global")

# Initiate empty list
scopes_assessments <- list()

# Pull species for each region
for (i in scopes$code) { 
  
  # Save region name
  scope_name <- scopes$description[scopes$code == i]
  
  # Get list of assesments
  scope_assessments <- rl_scopes(code = i)$assessments 
  
  # Append to list
  scopes_assessments[[scope_name]] <- dplyr::as_tibble(scope_assessments)
}

# Combine assessments
rgn_assess <- bind_rows(scopes_assessments) %>%
  rename(iucn_sid = sis_taxon_id, 
                sciname = taxon_scientific_name, 
                category = red_list_category_code,
                year = year_published)

# Save for time
write_csv(rgn_assess, file.path(dir_goal, sprintf('_data/iucn/rgn-spp-2025.csv')))

# Load for time
#rgn_assess <- read_csv(file.path(dir_goal, sprintf('_data/iucn/rgn-spp-2025.csv')), show_col_types = FALSE)
```

### Data: `iucn_risk_rgn_current_2025-1.csv`

Save the IUCN categories for all available *regional* assessments for marine species.

```{r}
# iucn_risk_rgn_current_2025-1.csv
rgn_assess_marine <- rgn_assess %>%
  # Get current assesments
  filter(latest == TRUE)  %>%
  # filter for marine species
  filter(iucn_sid %in% marine_species_list) %>%
  # Add region descriptions
  left_join(scopes, by = "code") %>%
  mutate(iucn_rgn = str_replace_all(tolower(description), " ", '_')) %>%
  # Reclassify old codes
  left_join(cat_lookup, by = c("category" = "code"), relationship = "many-to-many") %>%
  select(iucn_sid, 
         sciname,
         old_cat = category,
         iucn_rgn, 
         cat_score, 
         cat_current = code_current) 

# Write csv
write_csv(rgn_assess_marine, file.path(dir_goal, sprintf('_data/iucn_risk_rgn_current_%s.csv', api_version)))
```

Check out the results for 2025, assess differences from 2024.

```{r}
# Check
nrow(rgn_assess_marine) #5032
unique(rgn_assess_marine$iucn_rgn)
length(unique(rgn_assess_marine$iucn_rgn)) #13

rgn_assess_marine_2024 <- read_csv(file.path(dir_goal, sprintf('_data/iucn_risk_rgn_current_%s.csv', "2024-1")), show_col_types = FALSE)
nrow(rgn_assess_marine_2024) #2181
unique(rgn_assess_marine_2024$iucn_rgn)
length(unique(rgn_assess_marine_2024$iucn_rgn))

print(paste0("The following regions are in v2025 and not v2024: ", setdiff(unique(rgn_assess_marine$iucn_rgn), unique(rgn_assess_marine_2024$iucn_rgn))))
```

Identify which species have different regional IUCN categories from global categories.

```{r}
# Species with different regional categories from global categories
cat_mismatch <- marine_species_risk %>%
  rename(global_cat = cat) %>%
  left_join(rgn_assess_marine, by = c("iucn_sid", "sciname")) %>%
  select(iucn_sid, sciname, year, global_cat, global_cat_score = cat_score.x, iucn_rgn, rgn_cat = cat_current, rgn_cat_score = cat_score.y) %>%
  mutate(match = ifelse(global_cat == rgn_cat, TRUE, FALSE)) %>%
  filter(match == FALSE)

print(paste0("There are ", length(unique(cat_mismatch$iucn_sid)), " species whose regional IUCN category differs from global category.")) #557 species
```

### Data: `iucn_risk_ts_rgn_2025-1.csv`

Save the IUCN categories for all available *regional* assessments over time for all marine species.

```{r}
# iucn_risk_ts_rgn_2025-1.csv
rgn_risk_ts <- rgn_assess %>%
  # Filter for species in the marine species list
  filter(iucn_sid %in% marine_species_list) %>%
  rename(old_cat = category) %>%
  # Add new updated cat codes
  left_join(cat_lookup, by = c('old_cat' = 'code')) %>%
  distinct() %>%
  filter(!is.na(cat_score) & !is.na(year)) %>%
  # Add region descriptions
  left_join(scopes, by = "code") %>%
  arrange(iucn_sid, year) %>%
  select(iucn_sid, 
         year, 
         rgn_cat_ts = code_current, 
         rgn_cat_ts_score = cat_score, 
         iucn_rgn = description) %>%
  mutate(iucn_version = api_version) %>%
  group_by(iucn_sid, iucn_rgn) %>%
  filter(n() >= 2) %>%
  ungroup() %>%
  mutate(iucn_rgn = str_replace_all(tolower(iucn_rgn), " ", "_"))

# Write csv
write_csv(rgn_risk_ts, file.path(dir_goal, '_data',
                                 sprintf('iucn_risk_ts_rgn_%s.csv', api_version)))
```

```{r}
# Check
nrow(rgn_risk_ts) #1,330
unique(rgn_risk_ts$iucn_rgn)
length(unique(rgn_risk_ts$iucn_rgn)) #10

rgn_risk_ts_2024 <- read_csv(file.path(dir_goal, sprintf('_data/iucn_risk_ts_rgn_%s.csv', "2024-1")), show_col_types = FALSE)
nrow(rgn_risk_ts_2024) #388
unique(rgn_risk_ts_2024$iucn_rgn)
length(unique(rgn_risk_ts_2024$iucn_rgn)) #2
```

## Trends assessments

### Int: `int/trend_calcs/iucn_risk_ts_and_trends_2025-1.csv`

```{r}
# Get pop trends
iucn_pop_trends <- rl_pop_trends()$population_trends

# Unlist the description column
iucn_pop_trends$description <- iucn_pop_trends$description$en

# Initiate empty list
pop_trends_spp <- list()

# Pull species for each marine habitat
for (i in iucn_pop_trends$code) { 
  
  # Save habitat name
  trend_name <- iucn_pop_trends$description[iucn_pop_trends$code == i]
  
  # Get list of assesments
  pop_assessments <- rl_pop_trends(code = i)$assessments
  
  # Append to list
  pop_trends_spp[[trend_name]] <- as_tibble(pop_assessments)
}

# Combine individual habitat species lists into one master list
pop_trends <- bind_rows(pop_trends_spp) %>%
  rename(iucn_sid = sis_taxon_id, 
         sciname = taxon_scientific_name,
         category = red_list_category_code,
         year = year_published)

# Save for time
write_csv(pop_trends, file.path(dir_goal, sprintf('_data/iucn/pop-trends-2025.csv')))

# Load for time
#pop_trends <- read_csv(file.path(dir_goal, sprintf('_data/iucn/pop-trends-2025.csv')), show_col_types = FALSE)

pop_trends_marine <- pop_trends %>% 
  # Filter for marine species
  filter(iucn_sid %in% marine_species_list) %>%
  unnest(scopes, names_sep = "_") %>%
  # Filter for latest assessment
  filter(latest == TRUE) %>%
  left_join(iucn_pop_trends, by = "code")

# Clean up region description
pop_trends_marine$iucn_rgn <- tolower(str_replace(pop_trends_marine$scopes_description$en, " ", "_"))

# iucn_risk_ts_and_trends_2025-1.csv
iucn_risk_ts_and_trends <- pop_trends_marine %>%
  mutate(iucn_version = "2025-1") %>%
  select(iucn_sid, pop_trend = description, iucn_version, iucn_rgn)

# Save csv
write_csv(iucn_risk_ts_and_trends, file.path(dir_setup, 'int/trend_calcs',
                         sprintf('iucn_risk_ts_and_trends_%s.csv', api_version)))


all_ts <- marine_species_risk_timeseries %>%
  mutate(iucn_rgn = "global") %>%
  bind_rows(rgn_risk_ts %>%
              rename(cat_ts = rgn_cat_ts,
                     cat_ts_score = rgn_cat_ts_score))
```

### Data: `iucn_trend_by_spp_2025-1.csv`

Regress species with multiple recent historic assessments against population trends

From data gathered in the execution of SPP v2022, collect species data on past assessments (category) and population trend.  Combine and save.  Note that trend has to be assumed to be related to the most recent assessment only - not included in historical assessment API call.  We will need both a text description of trend (i.e. stable, increasing, or decreasing) and multiple assessments.

For this analysis we will focus only on assessments performed since 1991, the year of v1.0 Red List.

Since 1991:

``` {r}
cat_trend_91 <- iucn_risk_ts_and_trends %>%
  left_join(all_ts, by = c('iucn_sid', 'iucn_version', 'iucn_rgn')) %>%
  distinct() %>%
  filter(year >= 1991 & !is.na(cat_ts_score) & !is.na(pop_trend)) %>%
  group_by(iucn_sid, iucn_rgn) %>%
  mutate(n_assess = n()) %>%
  ungroup() %>%
  mutate(year = as.numeric(year))

lm_91 <- cat_trend_91 %>%
  filter(n_assess >= 2) %>%
  group_by(iucn_sid, iucn_rgn) %>%
  do(calc_trend = lm(cat_ts_score ~ year, data = .)[['coefficients']][['year']]) %>%
  mutate(calc_trend = round(calc_trend, 5))

trend_91 <- cat_trend_91 %>%
  filter(n_assess > 1) %>%
  group_by(iucn_sid, iucn_rgn) %>%
  summarize(years  = paste(year, collapse = ', '),
            cat_ts = paste(cat_ts, collapse = ', '),
            scores = paste(cat_ts_score, collapse = ', '),
            pop_trend_desc = first(pop_trend)) %>%
  ungroup() %>%
  left_join(lm_91, by = c('iucn_sid', 'iucn_rgn'))

write_csv(trend_91, file.path(dir_setup, 'int/trend_calcs/trend_lm_vs_pop_91.csv'))

DT::datatable(trend_91)

trend_91 <- trend_91 %>%
  filter(!is.na(pop_trend_desc)) %>%
  mutate(pop_trend_desc = tolower(pop_trend_desc)) %>%
  mutate(pop_trend = case_when(pop_trend_desc == 'decreasing' ~ -1,
                               pop_trend_desc == 'increasing' ~  1,
                               pop_trend_desc == 'stable'     ~  0,
                               TRUE                           ~ NA_real_))
  ### NOTE: here we're coding the trend in *population* not risk.

trend_regr_91 <- lm(calc_trend ~ pop_trend, data = trend_91)

# trend_coeffs_91 <- summary(trend_regr_91)
# 
# trend_plot_91 <- ggplot(trend_91, aes(x = pop_trend, y = calc_trend, group = pop_trend)) +
#   geom_violin(color = 'grey60', fill = 'grey80') +
#   geom_jitter(color = 'blue', alpha = .4) +
#   geom_abline(intercept = trend_regr_91$coefficients[['(Intercept)']],
#               slope = trend_regr_91$coefficients[['pop_trend']],
#               color = 'darkred') +
#   scale_x_continuous(breaks = c(-1, 0, 1), labels = c('decreasing', 'stable', 'increasing')) +
#   coord_cartesian(ylim = c(-.07, .07)) +
#   labs(x = 'text population trend',
#        y = 'calculated trend from category time series',
#        title = 'Pop trend vs change in assessed risk since 1991') +
#   annotate('text', x = 0.1, y = -.06, 
#            label = paste0('R^2: ',          round(trend_coeffs_91$r.squared, 5), 
#                           '\nslope = ',     round(trend_coeffs_91$coefficients[2, 1], 5),
#                           '\np (slope) = ', signif(trend_coeffs_91$coefficients[2, 4], 5)),
#            hjust = 0)
# 
# print(trend_plot_91)

slope <- trend_coeffs_91$coefficients[2, 1] %>% round(5)
trend_score_lookup <- c('decreasing' = -slope, 'stable' = 0, 'increasing' = slope)

trend_91 <- trend_91 %>%
  dplyr::select(iucn_sid, iucn_rgn, calc_trend) %>%
  distinct()

# iucn_trend_by_spp_2025-1.csv
trend_df <- iucn_risk_ts_and_trends %>%
  left_join(all_ts, by = c('iucn_sid', 'iucn_rgn', 'iucn_version')) %>%
  distinct() %>%
  left_join(trend_91, by = c('iucn_sid', 'iucn_rgn')) %>%
  mutate(trend_score  = ifelse(is.na(calc_trend), trend_score_lookup[pop_trend], calc_trend),
         trend_source = case_when(!is.na(calc_trend)  ~ 'lm', 
                                  !is.na(trend_score) ~ 'regr',
                                  TRUE ~ NA_character_)) %>%
  select(iucn_sid, iucn_rgn, pop_trend, trend_score, trend_source) %>%
  distinct()

# Write csv
write_csv(trend_df, file.path(dir_data, sprintf('iucn_trend_by_spp_%s.csv', api_version)))
```

