---
title: 'Set up IUCN marine species data'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE, eval = FALSE)
```

```{r}
#install.packages("rredlist")
library(rredlist)
library(dplyr)
library(tidyr)
library(tidyverse)
library(here)

#source('https://raw.githubusercontent.com/oharac/src/master/R/common.R')
source(here('workflow/R/common.R'))

goal     <- 'spp'
scenario <- 'v2025'

dir_goal <- here('globalprep', goal, scenario)

### goal specific folders and info
dir_anx  <- file.path(dir_M, 'git-annex/globalprep')
dir_setup   <- file.path(dir_goal, '_setup')
dir_goal_anx <- file.path(dir_anx, goal, scenario, 'spp_risk_dists')

source(file.path(dir_setup, 'common_fxns.R'))

```

# Summary

Get IUCN current risk assessments and historical assessments over time, for all IUCN marine species.  This method is not reliant on any information from AquaMaps.

# Data Sources

## IUCN Red List

To access the API v4, you will need to sign up for a key on the [IUCN website](https://api.iucnredlist.org/). This key can be added to the local .renviron on your machine and retrieved using the code below.

```{r}
# Save API key
rl_use_iucn()
api_version <- rl_version()

# or
rl_use_api_key("YOUR_API_KEY_HERE")
```

# Methods

This file sets up the following .csv files, saved in the `_data` folder:
- `iucn_risk_current_2025-1.csv`
- `spp_marine_from_api_2025-1.csv`
- `iucn_risk_rgn_current_2025-1.csv`
- `iucn_risk_ts_rgn_2025-1.csv`
- `iucn_risk_timeseries_2025-1.csv`
- `iucn_risk_ts_rgn_2025-1.csv`
- `iucn_trend_by_spp_2025-1.csv`

## Marine Habitat Assessment

Using the `rl_habitats()` function, get the species list for species coded with the following marine habitat codes: 9-12, plus 15.11, 15.12, 15.13.  Note category 13 is Marine Coastal/Supratidal, but includes many species whose "marine" dependence is only incidental. We include seacliffs and offshore islands (13.1) from category 13. If these species do not show up in category 12 (marine intertidal) then they are assumed to not actually depend on marine habitats.

The list can be found here (make sure to compare the webpage to the final df, to make sure it matches up): https://www.iucnredlist.org/resources/habitat-classification-scheme

```{r}
# Get a list of habitats
iucn_habitats <- rl_habitats()$habitats

# Unlist the description column
iucn_habitats$description <- iucn_habitats$description$en

# Select desired marine habitats (noted above)
marine_habitats <- iucn_habitats %>%
  filter(str_detect(code, "^(9_|10_|11_|12_)") | 
           code %in% c("13_1", "15_11", "15_12", "15_13"))

# marine_habitats <- iucn_habitats %>%
#   dplyr::filter(str_detect(tolower(description), "marine|reef|lagoon") &
#                   !str_detect(tolower(description), "supratidal")) # change to include 13_1

# Initiate empty list
marine_habitats_spp <- list()

# Pull species for each marine habitat
for (i in marine_habitats$code) { 
  
  # Save habitat name
  habitat_name <- marine_habitats$description[marine_habitats$code == i]
  
  # Get list of assesments
  habitat_assessments <- rl_habitats(code = i)$assessments
  
  # Append to list
  marine_habitats_spp[[habitat_name]] <- dplyr::as_tibble(habitat_assessments)
}

# Combine individual habitat species lists into one master list
marine_species <- bind_rows(marine_habitats_spp) %>%
  #filter(latest == TRUE) %>%
  rename(iucn_sid = sis_taxon_id, 
         sciname = taxon_scientific_name,
         category = red_list_category_code,
         year = year_published) 

# List of marine species (all unique IDs)
marine_species_list <- unique(marine_species$iucn_sid)

# 2025: 20,236 species
```

### Data: `iucn_risk_current_2025-1.csv`

Using `rl_scopes()`, pull all global species assessments and filter for species from `marine_species_list`. 

*This takes approximately 3 hours.*

```{r}
# Clean up the time series: reclassify old codes to current codes
cat_lookup <- read_csv(file.path(dir_setup, 'raw', 'risk_code_lookup.csv'), show_col_types = FALSE)

# Save global assessments for all species
global <- rl_scopes(code = "1")$assessments

# Filter for marine species
global_marine <- global %>%
  rename(iucn_sid = sis_taxon_id, 
         sciname = taxon_scientific_name, 
         category = red_list_category_code,
         year = year_published) %>%
  filter(iucn_sid %in% marine_species_list)

# Prepare layer
marine_species_risk <- global_marine %>%
  # Current assessments
  filter(latest == TRUE) %>%
  # Translate to current codes
  left_join(cat_lookup, by = c("category" = "code"), relationship = "many-to-many") %>%
  rename(old_cat = category,
         # Check that rl_sis() retrieves "global" rating (does rating change with habitat and region?)
         cat = code_current) %>%
  # Add version column
  mutate(iucn_version = "2025-1") %>%
  select(iucn_sid, sciname, year, old_cat, cat, cat_score, iucn_version) #criteria left out for now

# Write csv
write_csv(marine_species_risk, file.path(dir_goal, '_data', 
                                         sprintf('iucn_risk_current_%s.csv', api_version)))
```

OLD (DELETE SOON) 
Using `rl_sis()`, query assessments for each species in the `marine_species_list`. **Warning**: this can take over 6 hours.

```{r}
#marine_species_list_test <- c(222, 224, 225)

# Initiate empty dataframe
marine_spp <- data.frame()

# Retrieve assessments for each species (this takes like ~6+ hours!)
for (i in marine_species_list) {
  
  # Get global ICUN information for each marine species
  species_assessment <- rl_sis(id = i)$assessment
  
  # Append to the dataframe
  marine_spp <- bind_rows(marine_spp, species_assessment)
  
  # Rest between queries
  Sys.sleep(1)
  
}

# Save just in case
write_csv(marine_spp, file.path(dir_goal, sprintf('_data/iucn/marine-spp-2025.csv')))


# Clean up dataframe
marine_species_risk <- marine_spp %>%
  # Filter for current rating
  filter(latest == TRUE) %>%
  rename(iucn_sid = sis_taxon_id, 
         sciname = taxon_scientific_name,
         category = red_list_category_code,
         year = year_published) %>%
  # Translate to current codes
  left_join(cat_lookup, by = c("category" = "code"), relationship = "many-to-many") %>%
  rename(old_cat = category,
         # Check that rl_sis() retrieves "global" rating (does rating change with habitat and region?)
         cat = code_current) %>%
  # Add version column
  mutate(iucn_version = "2025-1") %>%
  select(iucn_sid, sciname, year, old_cat, cat, cat_score, iucn_version) #criteria

# Check to see how many species had their categories fixed
check <- marine_species_risk %>%
  mutate(match = case_when(old_cat == cat ~ TRUE,
                           old_cat != cat ~ FALSE)) %>%
  filter(match == FALSE)

#2025: only 4 species had their categories fixed

# criteria?

# rl_categories("E")
# rl_sis(id = 222)

# Write csv
write_csv(marine_species_risk, file.path(dir_goal, '_data', 
                                         sprintf('iucn_risk_current_%s.csv', api_version)))
```

Check out the results for 2025, assess differences from 2024.

```{r}
# Compare to last year
nrow(marine_species_risk) # 19, 529

# Shouldn't have any duplicates
marine_species_risk %>%
  count(iucn_sid) %>%
  filter(n > 1)

# Import last year (change year as needed)
marine_species_risk_2024 <- read_csv(file.path(dir_goal, sprintf('_data/iucn_risk_current_%s.csv', "2024-1")), show_col_types = FALSE)

nrow(marine_species_risk_2024) # 18,599

length(setdiff(unique(marine_species_risk$iucn_sid), unique(marine_species_risk_2024$iucn_sid))) # 2,648 new species

# Look at species added in new assessment
new_species <- marine_species_risk %>%
  anti_join(marine_species_risk_2024, by = "iucn_sid") %>%
  # Check to see if new species are all from same habitat? published year?
  left_join(marine_species, by = c("iucn_sid", "year", "sciname"))
```

### Data: `spp_marine_from_api_2025-1.csv`

Some species show up over multiple habitats (and thus depth ratings), so I chose to assign the deepest rating.

| habitat type   | habitat code   | depth classification  |
| -------------- | -------------- | --------------------- |
| deep benthic   | 10_3, 10_4     | "deep oceanic"        |
| deep oceanic   | 10_ , 11_      | "200 m +"             |
| neritic        | 9_, 15_        | "< 200 m"             |
| intertidal     | 12_, 13_       | "<20 m"               |

```{r}
# Categorize by depths
marine_species_depths <- marine_species %>%
  filter(latest == TRUE) %>%
  # Break down the codes into cat, subcat1, and subcat2
  separate(code, c('cat', 'subcat1', 'subcat2'),
           remove = FALSE, convert = TRUE, fill = "right") %>%
  arrange(cat, subcat1, subcat2) %>%
  # Assign descriptions to each cat
  mutate(
    max_depth = factor(case_when(
      cat == 10 & subcat1 %in% 3:4         ~ "deep oceanic",
      cat %in% c(10, 11)                   ~ "200 m +",
      cat %in% c(9, 15)                    ~ "< 200 m",
      cat %in% c(12, 13)                   ~ "< 20 m",
      TRUE                                 ~ "unknown"),
    # To make it easier to pick the deepest (min) depth rating
    levels = c("deep oceanic", "200 m +", "< 200 m", "< 20 m", "unknown")))%>%
  left_join(iucn_habitats, by = "code")

# spp_marine_from_api_2025-1.csv
habs <- marine_species_depths %>%
  group_by(iucn_sid) %>%
  summarise(
    habs = paste(unique(code), collapse = ", "),
    # Pick the deepest rating (the lowest (min) level)
    max_depth = as.character(max_depth[which.min(as.integer(max_depth))])) %>%
  ungroup()

# Write csv
write_csv(habs, file.path(dir_goal, '_data',
                          sprintf('spp_marine_from_api_%s.csv', api_version)))
```

Check out the results for 2025, assess differences from 2024.

```{r}
nrow(habs) # 19,899
nrow(marine_species_risk) # 19,529

# Check species that are in habs but not in global
check <- anti_join(habs, marine_species_risk, by = "iucn_sid")

## Interesting: iucn_sid 155168 appears in marine_species_list and marine_species (obtained from rl_habitats(), but does not appear in global_marine and thus marine_species_risk despite having a scope of 1 (global) listed in marine_species)

habs_2024 <- read_csv(file.path(dir_goal, sprintf('_data/spp_marine_from_api_%s.csv', "2024-1")), show_col_types = FALSE)
nrow(habs_2024) # 18,590
```

## Regional Assessment

Using the `rl_scopes()` function, get the species list for species coded with the following regions: Europe (2), Mediterranean (4), Western Africa (8), S. Africa FW (9), Pan-Africa (10), Central Africa (16), Northeastern Africa (17), Eastern Africa (18), Northern Africa (21), Gulf of Mexico (45433062), Caribbean (45433063), Persian Gulf (45433064), Arabian Sea (100765562). 

Essentially all *but* Global (1).

```{r}
# Get list of scopes (regions)
scopes <- rl_scopes()$scopes

# Unlist description
scopes$description <- scopes$description$en

# Remove "Global" (see above)
scopes <- scopes %>%
  filter(!description == "Global")

# Initiate empty list
scopes_assessments <- list()

# Pull species for each region
for (i in scopes$code) { 
  
  # Save region name
  scope_name <- scopes$description[scopes$code == i]
  
  # Get list of assesments
  scope_assessments <- rl_scopes(code = i)$assessments 
  
  # Append to list
  scopes_assessments[[scope_name]] <- dplyr::as_tibble(scope_assessments)
}

# Combine assessments
rgn_assess <- bind_rows(scopes_assessments) %>%
  # Current assessments
  filter(latest == TRUE) %>%
  rename(iucn_sid = sis_taxon_id, 
                sciname = taxon_scientific_name, 
                category = red_list_category_code,
                year = year_published) %>%
  select(iucn_sid, sciname, year, category, code) %>%
  mutate(year = as.integer(year))
```

### Data: `iucn_risk_rgn_current_2025-1.csv`

```{r}
# Keep the species that overlap with regions of interest
rgn_assess_marine <- rgn_assess %>%
  # filter for marine species
  filter(iucn_sid %in% marine_species_list) %>%
  # Add region description
  left_join(scopes, by = "code") %>%
  mutate(iucn_rgn = str_replace_all(tolower(description), " ", '_')) %>%
  left_join(cat_lookup, by = c("category" = "code")) %>%
  select(iucn_sid, 
         sciname,
         old_cat = category,
         iucn_rgn, 
         cat_score, 
         cat_current = code_current) 

# Write csv
write_csv(rgn_assess_marine, file.path(dir_goal, sprintf('_data/iucn_risk_rgn_current_%s.csv', api_version)))
```

Check out the results for 2025, assess differences from 2024.

```{r}
# Check
nrow(rgn_assess_marine) #5032
unique(rgn_assess_marine$iucn_rgn)
length(unique(rgn_assess_marine$iucn_rgn)) #13

rgn_assess_marine_2024 <- read_csv(file.path(dir_goal, sprintf('_data/iucn_risk_rgn_current_%s.csv', "2024-1")), show_col_types = FALSE)
nrow(rgn_assess_marine_2024) #2181
unique(rgn_assess_marine_2024$iucn_rgn)
length(unique(rgn_assess_marine_2024$iucn_rgn))

print(paste0("The following regions are in v2025 and not v2024: ", setdiff(unique(rgn_assess_marine$iucn_rgn), unique(rgn_assess_marine_2024$iucn_rgn))))
```

Once regional assessments are gleaned from the API, check them against the global marine species list. Species are matched by ID numbers.  Species whose regional category differs from the global category are kept and scored.

* note, quick check of scientific names showed they all match
* 2025: 582 spp fall into this...
* 2024: 113 spp fall into this...

```{r}
# Species with different regional categories from global categories
cat_mismatch <- marine_species_risk %>%
  rename(global_cat = cat) %>%
  left_join(rgn_assess_marine, by = c("iucn_sid", "sciname")) %>%
  select(iucn_sid, sciname, year, global_cat, global_cat_score = cat_score.x, iucn_rgn, rgn_cat = cat_current, rgn_cat_score = cat_score.y) %>%
  mutate(match = ifelse(global_cat == rgn_cat, TRUE, FALSE)) %>%
  filter(match == FALSE)

length(unique(cat_mismatch$iucn_sid)) #582 species
```

## Get IUCN regional historical assessments

### Data: `iucn_risk_ts_rgn_2025-1.csv`

**2025 NOTE:** This .csv file has species from 10 of the regions, while v2024 only included "europe" and "mediterranean". This appears to be why v2024 has fewer rows than v2025. 

```{r}
# Combine species lists by region (this time don't filter for latest)
rgn_risk_ts <- bind_rows(scopes_assessments) %>%
  rename(iucn_sid = sis_taxon_id, 
         sciname = taxon_scientific_name, 
         old_cat = red_list_category_code,
         year = year_published,
         iucn_rgn = code) %>%
  # Filter for species in the marine species list
  filter(iucn_sid %in% marine_species_list) %>%
  # Add new updated cat codes
  left_join(cat_lookup, by = c('old_cat' = 'code')) %>%
  distinct() %>%
  filter(!is.na(cat_score) & !is.na(year)) %>%
  arrange(iucn_sid, year) %>%
  select(iucn_sid, 
         year, 
         rgn_cat_ts = code_current, 
         rgn_cat_ts_score = cat_score, 
         iucn_rgn) %>%
  mutate(iucn_version = api_version) %>%
  group_by(iucn_sid, iucn_rgn) %>%
  filter(n() >= 2) %>%
  ungroup() %>%
  # Add region description
  left_join(scopes, by = c("iucn_rgn" = "code")) %>%
  mutate(iucn_rgn = str_replace_all(tolower(description), " ", "_")) %>%
  select(-description)

# Write csv
write_csv(rgn_risk_ts, file.path(dir_goal, '_data',
                                 sprintf('iucn_risk_ts_rgn_%s.csv', api_version)))
```

```{r}
# Check
nrow(rgn_risk_ts) #1,330
unique(rgn_risk_ts$iucn_rgn)
length(unique(rgn_risk_ts$iucn_rgn)) #10

rgn_risk_ts_2024 <- read_csv(file.path(dir_goal, sprintf('_data/iucn_risk_ts_rgn_%s.csv', "2024-1")), show_col_types = FALSE)
nrow(rgn_risk_ts_2024) #388
unique(rgn_risk_ts_2024$iucn_rgn)
length(unique(rgn_risk_ts_2024$iucn_rgn)) #2
```

## Get IUCN trends assessments

### Data: 

```{r}
# Remove "Global" (see above)
global <- rl_scopes(code = "1")$assessments


global_marine <- global %>%
  rename(iucn_sid = sis_taxon_id, 
         sciname = taxon_scientific_name, 
         category = red_list_category_code,
         year = year_published) %>%
  filter(iucn_sid %in% marine_species_list)

write_csv(global_marine, file.path(dir_goal, sprintf('_data/iucn/global-2025.csv')))
```

