---
title: 'Set up IUCN marine species data'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE, eval = FALSE)
```

```{r}
#install.packages("rredlist")
library(rredlist)
library(dplyr)
library(tidyr)
library(tidyverse)
library(here)

#source('https://raw.githubusercontent.com/oharac/src/master/R/common.R')
source(here('workflow/R/common.R'))

goal     <- 'spp'
scenario <- 'v2025'

dir_goal <- here('globalprep', goal, scenario)

### goal specific folders and info
dir_anx  <- file.path(dir_M, 'git-annex/globalprep')
dir_setup   <- file.path(dir_goal, '_setup')
dir_goal_anx <- file.path(dir_anx, goal, scenario, 'spp_risk_dists')

source(file.path(dir_setup, 'common_fxns.R'))

```

# Summary

Get IUCN current risk assessments and historical assessments over time, for all IUCN marine species.  This method is not reliant on any information from AquaMaps.

# Data Sources

## IUCN Red List

To access the API v4, you will need to sign up for a key on the [IUCN website](https://api.iucnredlist.org/). This key can be added to the local .renviron on your machine and retrieved using the code below.

```{r}
# Save API key
rl_use_iucn()
api_version <- rl_version()

# or
rl_use_api_key("YOUR_API_KEY_HERE")
```

# Methods

This file sets up the following .csv files, saved in the `_data` folder:
- `iucn_risk_current_2025-1.csv`
- `spp_marine_from_api_2025-1.csv`
- `iucn_risk_rgn_current_2025-1.csv`
- `iucn_risk_ts_rgn_2025-1.csv`
- `iucn_risk_timeseries_2025-1.csv`
- `iucn_risk_ts_rgn_2025-1.csv`
- `iucn_trend_by_spp_2025-1.csv`

## Marine Species List

First, we need to establish the list of marine species IDs we are interested in. We do this by using rl_systems() to search for species coded for the marine system ("2").

```{r}
# Save IUCN systems
iucn_systems <- rl_systems()$systems

# Unlist description column
iucn_systems$description <- iucn_systems$description$en

# Search for species by marine system
marine_system <- rl_systems("2")$assessments 

# Clean up data frame
marine_system <- marine_system %>%
  rename(iucn_sid = sis_taxon_id, 
         sciname = taxon_scientific_name,
         category = red_list_category_code,
         year = year_published)

# Save list of taxon IDs
marine_species_list <- unique(marine_system$iucn_sid)

# Save for time
write_csv(marine_system, file.path(dir_goal, sprintf('_data/iucn/marine-system-2025.csv')))

# 2025: 19,944 species
```

### Data: `iucn_risk_current_2025-1.csv`

```{r}
# Clean up the time series: reclassify old codes to current codes
cat_lookup <- read_csv(file.path(dir_setup, 'raw', 'risk_code_lookup.csv'), show_col_types = FALSE)

# Create data frame for layer
marine_species_risk <- marine_system %>%
  filter(latest == TRUE) %>%
  unnest(scopes, names_sep = "_") %>%
  filter (scopes_code == "1") %>%
  left_join(cat_lookup, by = c("category" = "code"), relationship = "many-to-many") %>%
  rename(old_cat = category,
         cat = code_current) %>%
  # Add version column
  mutate(iucn_version = "2025-1") %>%
  select(iucn_sid, sciname, year, old_cat, cat, cat_score, iucn_version) #criteria left out for now

# Check that no IDs are repeated
marine_species_risk %>%
  group_by(iucn_sid) %>%
  filter(n > 1)

# Write csv
write_csv(marine_species_risk, file.path(dir_goal, '_data', 
                                         sprintf('iucn_risk_current_%s.csv', api_version)))
```

OLD DELETE SOON

Using `rl_scopes()`, pull all global species assessments and filter for species from `marine_species_list`. 

*This takes approximately 3 hours.*

```{r}
# # Clean up the time series: reclassify old codes to current codes
# cat_lookup <- read_csv(file.path(dir_setup, 'raw', 'risk_code_lookup.csv'), show_col_types = FALSE)
# 
# # Save global assessments for all species
# global <- rl_scopes(code = "1")$assessments
# 
# # Save for time
# write_csv(global, file.path(dir_goal, sprintf('_data/iucn/global-2025.csv')))
# 
# # OR once it's saved
# #global <- read_csv(file.path(dir_goal, "_data/iucn/global-2025.csv"), show_col_types = FALSE)
# 
# # Filter for marine species
# global_marine <- global %>%
#   rename(iucn_sid = sis_taxon_id,
#          sciname = taxon_scientific_name,
#          category = red_list_category_code,
#          year = year_published) %>%
#   filter(iucn_sid %in% marine_species_list)
# 
# # Prepare layer
# marine_species_risk <- global_marine %>%
#   # Current assessments
#   filter(latest == TRUE) %>%
#   # Translate to current codes
#   left_join(cat_lookup, by = c("category" = "code"), relationship = "many-to-many") %>%
#   rename(old_cat = category,
#          # Check that rl_sis() retrieves "global" rating (does rating change with habitat and region?)
#          cat = code_current) %>%
#   # Add version column
#   mutate(iucn_version = "2025-1") %>%
#   select(iucn_sid, sciname, year, old_cat, cat, cat_score, iucn_version) #criteria left out for now
# 
# # Write csv
# write_csv(marine_species_risk, file.path(dir_goal, '_data', 
#                                          sprintf('iucn_risk_current_%s.csv', api_version)))
# ```
# 
# OLD (DELETE SOON) 
# Using `rl_sis()`, query assessments for each species in the `marine_species_list`. **Warning**: this can take over 6 hours.
# 
# ```{r}
# #marine_species_list_test <- c(222, 224, 225)
# 
# # Initiate empty dataframe
# marine_spp <- data.frame()
# 
# # Retrieve assessments for each species (this takes like ~6+ hours!)
# for (i in marine_species_list) {
#   
#   # Get global ICUN information for each marine species
#   species_assessment <- rl_sis(id = i)$assessment
#   
#   # Append to the dataframe
#   marine_spp <- bind_rows(marine_spp, species_assessment)
#   
#   # Rest between queries
#   Sys.sleep(1)
#   
# }
# 
# # Save just in case
# write_csv(marine_spp, file.path(dir_goal, sprintf('_data/iucn/marine-spp-2025.csv')))
# 
# 
# # Clean up dataframe
# marine_species_risk <- marine_spp %>%
#   # Filter for current rating
#   filter(latest == TRUE) %>%
#   rename(iucn_sid = sis_taxon_id, 
#          sciname = taxon_scientific_name,
#          category = red_list_category_code,
#          year = year_published) %>%
#   # Translate to current codes
#   left_join(cat_lookup, by = c("category" = "code"), relationship = "many-to-many") %>%
#   rename(old_cat = category,
#          # Check that rl_sis() retrieves "global" rating (does rating change with habitat and region?)
#          cat = code_current) %>%
#   # Add version column
#   mutate(iucn_version = "2025-1") %>%
#   select(iucn_sid, sciname, year, old_cat, cat, cat_score, iucn_version) #criteria
# 
# # Check to see how many species had their categories fixed
# check <- marine_species_risk %>%
#   mutate(match = case_when(old_cat == cat ~ TRUE,
#                            old_cat != cat ~ FALSE)) %>%
#   filter(match == FALSE)
# 
# #2025: only 4 species had their categories fixed
# 
# # criteria?
# 
# # rl_categories("E")
# # rl_sis(id = 222)
# 
# # Write csv
# write_csv(marine_species_risk, file.path(dir_goal, '_data', 
#                                          sprintf('iucn_risk_current_%s.csv', api_version)))
```

Check out the results for 2025, assess differences from 2024.

```{r}
# Compare to last year
nrow(marine_species_risk) # 19, 529

# Shouldn't have any duplicates
marine_species_risk %>%
  count(iucn_sid) %>%
  filter(n > 1)

# Import last year (change year as needed)
marine_species_risk_2024 <- read_csv(file.path(dir_goal, sprintf('_data/iucn_risk_current_%s.csv', "2024-1")), show_col_types = FALSE)

nrow(marine_species_risk_2024) # 18,599

length(setdiff(unique(marine_species_risk$iucn_sid), unique(marine_species_risk_2024$iucn_sid))) # 2,648 new species

# Look at species added in new assessment
new_species <- marine_species_risk %>%
  anti_join(marine_species_risk_2024, by = "iucn_sid") %>%
  # Check to see if new species are all from same habitat? published year?
  left_join(marine_species, by = c("iucn_sid", "year", "sciname"))
```

### Data: `spp_marine_from_api_2025-1.csv`

Using the `rl_habitats()` function, get the species list for species coded with the following marine habitat codes: 9-12, plus 15.11, 15.12, 15.13.  Note category 13 is Marine Coastal/Supratidal, but includes many species whose "marine" dependence is only incidental. We include seacliffs and offshore islands (13.1) from category 13. If these species do not show up in category 12 (marine intertidal) then they are assumed to not actually depend on marine habitats.

The list can be found here (make sure to compare the webpage to the final df, to make sure it matches up): https://www.iucnredlist.org/resources/habitat-classification-scheme

Some species show up over multiple habitats (and thus depth ratings), so I chose to assign the deepest rating.

| habitat type   | habitat code   | depth classification  |
| -------------- | -------------- | --------------------- |
| deep benthic   | 10_3, 10_4     | "deep oceanic"        |
| deep oceanic   | 10_ , 11_      | "200 m +"             |
| neritic        | 9_, 15_        | "< 200 m"             |
| intertidal     | 12_, 13_       | "<20 m"               |

```{r}
# Get a list of habitats
iucn_habitats <- rl_habitats()$habitats

# Unlist the description column
iucn_habitats$description <- iucn_habitats$description$en

# Select desired marine habitats (noted above)
marine_habitats <- iucn_habitats %>%
  filter(str_detect(code, "^(9_|10_|11_|12_)") | 
           code %in% c("13_1", "15_11", "15_12", "15_13"))

# marine_habitats <- iucn_habitats %>%
#   dplyr::filter(str_detect(tolower(description), "marine|reef|lagoon") &
#                   !str_detect(tolower(description), "supratidal")) # change to include 13_1

# Initiate empty list
marine_habitats_spp <- list()

# Pull species for each marine habitat
for (i in marine_habitats$code) { 
  
  # Save habitat name
  habitat_name <- marine_habitats$description[marine_habitats$code == i]
  
  # Get list of assesments
  habitat_assessments <- rl_habitats(code = i)$assessments
  
  # Append to list
  marine_habitats_spp[[habitat_name]] <- dplyr::as_tibble(habitat_assessments)
}

# Combine individual habitat species lists into one master list
marine_species_habitats <- bind_rows(marine_habitats_spp) %>%
  #filter(latest == TRUE) %>%
  rename(iucn_sid = sis_taxon_id, 
         sciname = taxon_scientific_name,
         category = red_list_category_code,
         year = year_published) 

# Save for time
write_csv(marine_species_habitats, file.path(dir_goal, sprintf('_data/iucn/marine-species-habitats-2025.csv')))

# Filter for species identified in marine system search, global code
marine_species_habitats_global <- marine_species_habitats %>%
  filter(latest == TRUE) %>%
  unnest(scopes, names_sep = "_") %>%
  # Filter for global
  filter(scopes_code == "1") %>%
  # Filter for marine species
  filter(iucn_sid %in% marine_species_list) %>%
  separate(code, c('cat', 'subcat1', 'subcat2'),
           remove = FALSE, convert = TRUE, fill = "right") %>%
  arrange(cat, subcat1, subcat2) %>%
  # Assign descriptions to each cat
  mutate(
    max_depth = factor(case_when(
      cat == 10 & subcat1 %in% 3:4         ~ "deep oceanic",
      cat %in% c(10, 11)                   ~ "200 m +",
      cat %in% c(9, 15)                    ~ "< 200 m",
      cat %in% c(12, 13)                   ~ "< 20 m",
      TRUE                                 ~ "unknown"),
    # To make it easier to pick the deepest (min) depth rating
    levels = c("deep oceanic", "200 m +", "< 200 m", "< 20 m", "unknown")))%>%
  left_join(iucn_habitats, by = "code")

# spp_marine_from_api_2025-1.csv
habs <- marine_species_risk %>%
  left_join(marine_species_habitats_global, by = c("iucn_sid", "sciname")) %>%
  select(iucn_sid, code, max_depth) %>%
  group_by(iucn_sid) %>%
  reframe(
    habs = paste(unique(code), collapse = ", "),
    # Pick the deepest rating (the lowest (min) level)
    max_depth = as.character(max_depth[which.min(as.integer(max_depth))])) %>%
  ungroup()

# Write csv
write_csv(habs, file.path(dir_goal, '_data',
                          sprintf('spp_marine_from_api_%s.csv', api_version)))
```



```{r}
# # Categorize by depths
# marine_species_depths <- marine_species %>%
#   filter(latest == TRUE) %>%
#   # Break down the codes into cat, subcat1, and subcat2
#   separate(code, c('cat', 'subcat1', 'subcat2'),
#            remove = FALSE, convert = TRUE, fill = "right") %>%
#   arrange(cat, subcat1, subcat2) %>%
#   # Assign descriptions to each cat
#   mutate(
#     max_depth = factor(case_when(
#       cat == 10 & subcat1 %in% 3:4         ~ "deep oceanic",
#       cat %in% c(10, 11)                   ~ "200 m +",
#       cat %in% c(9, 15)                    ~ "< 200 m",
#       cat %in% c(12, 13)                   ~ "< 20 m",
#       TRUE                                 ~ "unknown"),
#     # To make it easier to pick the deepest (min) depth rating
#     levels = c("deep oceanic", "200 m +", "< 200 m", "< 20 m", "unknown")))%>%
#   left_join(iucn_habitats, by = "code")
# 
# # spp_marine_from_api_2025-1.csv
# habs <- marine_species_depths %>%
#   group_by(iucn_sid) %>%
#   summarise(
#     habs = paste(unique(code), collapse = ", "),
#     # Pick the deepest rating (the lowest (min) level)
#     max_depth = as.character(max_depth[which.min(as.integer(max_depth))])) %>%
#   ungroup()
# 
# # Write csv
# write_csv(habs, file.path(dir_goal, '_data',
#                           sprintf('spp_marine_from_api_%s.csv', api_version)))
```

Check out the results for 2025, assess differences from 2024.

```{r}
nrow(habs) # 19,899
nrow(marine_species_risk) # 19,529

# Check species that are in habs but not in global
check <- anti_join(habs, marine_species_risk, by = "iucn_sid")

## Interesting: iucn_sid 155168 appears in marine_species_list and marine_species (obtained from rl_habitats(), but does not appear in global_marine and thus marine_species_risk despite having a scope of 1 (global) listed in marine_species)

habs_2024 <- read_csv(file.path(dir_goal, sprintf('_data/spp_marine_from_api_%s.csv', "2024-1")), show_col_types = FALSE)
nrow(habs_2024) # 18,590
```

## Regional Assessment

Using the `rl_scopes()` function, get the species list for species coded with the following regions: Europe (2), Mediterranean (4), Western Africa (8), S. Africa FW (9), Pan-Africa (10), Central Africa (16), Northeastern Africa (17), Eastern Africa (18), Northern Africa (21), Gulf of Mexico (45433062), Caribbean (45433063), Persian Gulf (45433064), Arabian Sea (100765562). 

Essentially all *but* Global (1).

```{r}
# Get list of scopes (regions)
iucn_scopes <- rl_scopes()$scopes

# Unlist description
iucn_scopes$description <- iucn_scopes$description$en

# Remove "Global" (see above)
scopes <- iucn_scopes %>%
  filter(!description == "Global")

# Initiate empty list
scopes_assessments <- list()

# Pull species for each region
for (i in scopes$code) { 
  
  # Save region name
  scope_name <- scopes$description[scopes$code == i]
  
  # Get list of assesments
  scope_assessments <- rl_scopes(code = i)$assessments 
  
  # Append to list
  scopes_assessments[[scope_name]] <- dplyr::as_tibble(scope_assessments)
}

# Combine assessments
rgn_assess <- bind_rows(scopes_assessments) %>%
  rename(iucn_sid = sis_taxon_id, 
                sciname = taxon_scientific_name, 
                category = red_list_category_code,
                year = year_published)

# Save for time
write_csv(rgn_assess, file.path(dir_goal, sprintf('_data/iucn/rgn-spp-2025.csv')))

# Load for time
#rgn_assess <- read_csv(file.path(dir_goal, sprintf('_data/iucn/rgn-spp-2025.csv')), show_col_types = FALSE)
```

### Data: `iucn_risk_rgn_current_2025-1.csv`

```{r}
# Keep the species that overlap with regions of interest
rgn_assess_marine <- rgn_assess %>%
  # Get current assesments
  filter(latest == TRUE)  %>%
  # filter for marine species
  filter(iucn_sid %in% marine_species_list) %>%
  # Add region descriptions
  left_join(scopes, by = "code") %>%
  mutate(iucn_rgn = str_replace_all(tolower(description), " ", '_')) %>%
  # Reclassify old codes
  left_join(cat_lookup, by = c("category" = "code"), relationship = "many-to-many") %>%
  select(iucn_sid, 
         sciname,
         old_cat = category,
         iucn_rgn, 
         cat_score, 
         cat_current = code_current) 

# Write csv
write_csv(rgn_assess_marine, file.path(dir_goal, sprintf('_data/iucn_risk_rgn_current_%s.csv', api_version)))
```

Check out the results for 2025, assess differences from 2024.

```{r}
# Check
nrow(rgn_assess_marine) #5032
unique(rgn_assess_marine$iucn_rgn)
length(unique(rgn_assess_marine$iucn_rgn)) #13

rgn_assess_marine_2024 <- read_csv(file.path(dir_goal, sprintf('_data/iucn_risk_rgn_current_%s.csv', "2024-1")), show_col_types = FALSE)
nrow(rgn_assess_marine_2024) #2181
unique(rgn_assess_marine_2024$iucn_rgn)
length(unique(rgn_assess_marine_2024$iucn_rgn))

print(paste0("The following regions are in v2025 and not v2024: ", setdiff(unique(rgn_assess_marine$iucn_rgn), unique(rgn_assess_marine_2024$iucn_rgn))))
```

Once regional assessments are gleaned from the API, check them against the global marine species list. Species are matched by ID numbers.  Species whose regional category differs from the global category are kept and scored.

* note, quick check of scientific names showed they all match
* 2025: 582 spp fall into this...
* 2024: 113 spp fall into this...

```{r}
# Species with different regional categories from global categories
cat_mismatch <- marine_species_risk %>%
  rename(global_cat = cat) %>%
  left_join(rgn_assess_marine, by = c("iucn_sid", "sciname")) %>%
  select(iucn_sid, sciname, year, global_cat, global_cat_score = cat_score.x, iucn_rgn, rgn_cat = cat_current, rgn_cat_score = cat_score.y) %>%
  mutate(match = ifelse(global_cat == rgn_cat, TRUE, FALSE)) %>%
  filter(match == FALSE)

length(unique(cat_mismatch$iucn_sid)) #795 species
```

## Get IUCN regional historical assessments

### Data: `iucn_risk_ts_rgn_2025-1.csv`

**2025 NOTE:** This .csv file has species from 10 of the regions, while v2024 only included "europe" and "mediterranean". This appears to be why v2024 has fewer rows than v2025. 

```{r}
# Combine species lists by region (this time don't filter for latest)
rgn_risk_ts <- bind_rows(scopes_assessments) %>%
  rename(iucn_sid = sis_taxon_id, 
         sciname = taxon_scientific_name, 
         old_cat = red_list_category_code,
         year = year_published,
         iucn_rgn = code) %>%
  # Filter for species in the marine species list
  filter(iucn_sid %in% marine_species_list) %>%
  # Add new updated cat codes
  left_join(cat_lookup, by = c('old_cat' = 'code')) %>%
  distinct() %>%
  filter(!is.na(cat_score) & !is.na(year)) %>%
  arrange(iucn_sid, year) %>%
  select(iucn_sid, 
         year, 
         rgn_cat_ts = code_current, 
         rgn_cat_ts_score = cat_score, 
         iucn_rgn) %>%
  mutate(iucn_version = api_version) %>%
  group_by(iucn_sid, iucn_rgn) %>%
  filter(n() >= 2) %>%
  ungroup() %>%
  # Add region description
  left_join(scopes, by = c("iucn_rgn" = "code")) %>%
  mutate(iucn_rgn = str_replace_all(tolower(description), " ", "_")) %>%
  select(-description)

# Write csv
write_csv(rgn_risk_ts, file.path(dir_goal, '_data',
                                 sprintf('iucn_risk_ts_rgn_%s.csv', api_version)))
```

```{r}
# Check
nrow(rgn_risk_ts) #1,330
unique(rgn_risk_ts$iucn_rgn)
length(unique(rgn_risk_ts$iucn_rgn)) #10

rgn_risk_ts_2024 <- read_csv(file.path(dir_goal, sprintf('_data/iucn_risk_ts_rgn_%s.csv', "2024-1")), show_col_types = FALSE)
nrow(rgn_risk_ts_2024) #388
unique(rgn_risk_ts_2024$iucn_rgn)
length(unique(rgn_risk_ts_2024$iucn_rgn)) #2
```

## Get IUCN trends assessments

### Data: 

```{r}
# Get pop trends
pop_trends <- rl_pop_trends()$population_trends

# Unlist the description column
pop_trends$description <- pop_trends$description$en

# Initiate empty list
pop_trends_spp <- list()

# Pull species for each marine habitat
for (i in pop_trends$code) { 
  
  # Save habitat name
  trend_name <- pop_trends$description[pop_trends$code == i]
  
  # Get list of assesments
  pop_assessments <- rl_pop_trends(code = i)$assessments
  
  # Append to list
  pop_trends_spp[[trend_name]] <- as_tibble(pop_assessments)
}

# Combine individual habitat species lists into one master list
pop_trends_marine <- bind_rows(pop_trends_spp) %>%
  filter(latest == TRUE) %>%
  rename(iucn_sid = sis_taxon_id, 
         sciname = taxon_scientific_name,
         category = red_list_category_code,
         year = year_published) %>%
  filter(iucn_sid %in% marine_species_list) %>%
  left_join(pop_trends, by = c("code"))

# Save just in case
write_csv(pop_trends_marine, file.path(dir_goal, sprintf('_data/iucn/pop-trends-marine-2025.csv')))

# filter for global
pop_trends_marine_global <- pop_trends_marine %>% 
  unnest(scopes, names_sep = "_") %>%
  filter(scopes_code == "1")

# filter for regions
pop_trends_marine_rgn <- pop_trends_marine %>%
  unnest(scopes, names_sep = "_") %>%
  filter(scopes_code != "1")

# marine spp search by system
marine_system <- rl_systems("2")$assessments 

marine_system <- marine_system %>%
  rename(iucn_sid = sis_taxon_id, 
         sciname = taxon_scientific_name,
         category = red_list_category_code,
         year = year_published)
  
marine_system_species_list <- unique(marine_system$sis_taxon_id)


length(unique(global_marine$iucn_sid))

setdiff(marine_species_list, unique(marine_system$sis_taxon_id))

```

Comprehensive marine species list assessment

```{r}
# rl_habitats()
rl_habs <- marine_species %>%
  select(iucn_sid, sciname) %>%
  distinct() 
habitats_IDs <- unique(rl_habs$iucn_sid)
length(habitats_IDs) #20,236

# rl_scopes(code = "1")
rl_scopes <- global_marine %>%
  select(iucn_sid, sciname) %>%
  distinct() 
scopes_IDs <- unique(rl_scopes$iucn_sid)
length(scopes_IDs) #19,800

# rl_systems(code = "2")
rl_sys <- marine_system %>%
  select(iucn_sid, sciname) %>%
  distinct() 
systems_IDs <- unique(rl_sys$iucn_sid)
length(systems_IDs) #19,944

# species in habitats but not systems
length(setdiff(habitats_IDs, systems_IDs)) #1520 species
habitat_system_extra <- anti_join(rl_habs, rl_sys)
nrow(habitat_system_extra)

# species in systems but not habitats
length(setdiff(systems_IDs, habitats_IDs)) #1228 species
system_habitat_extra <- anti_join(rl_sys, rl_habs)
nrow(system_habitat_extra)

left_join(system_habitat_extra, marine_species, by = c("iucn_sid", "sciname"))

h_fulva <- rl_sis(10016)
```

