---
title: 'Set up IUCN marine species ranges by regional assessment'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)


library(dplyr)
# library(rgdal)
library(raster)
library(here)
#library(ggridges)
library(ggplot2)
library(sf)
library(raster)
library(sf)
library(fasterize)
library(stringr)
library(mapview)
library(terra)

source(here('workflow/R/common.R'))

### provenance tracking
#library(provRmd); prov_setup()

goal     <- 'spp'
scenario <- 'v2025'
dir_goal <- here('globalprep', goal, scenario)

### goal specific folders and info
dir_anx  <- file.path(dir_M, 'git-annex/globalprep')
dir_setup   <- file.path(dir_goal, '_setup')
dir_goal_anx <- file.path(dir_anx, goal, scenario, 'spp_risk_dists')
dir_raw_data_anx <- file.path(dir_anx, '_raw_data')

source(file.path(dir_setup, 'common_fxns.R'))

```

# Summary

Examine IUCN Red List regional assessments for extinction risk categories.

# Data Sources

### IUCN Red List

# Methods: Geographic scope of regional assessments

For each regional assessment, spatially classify by [Large Marine Ecosystems](https://upload.wikimedia.org/wikipedia/commons/4/4d/Global_map_of_large_marine_ecosystems.jpg) (other potential options: EEZs and/or FAO Major Fishing Areas).  Use the [https://en.wikipedia.org/wiki/United_Nations_geoscheme](UN geoscheme) to roughly identify subregional zones.  Note there may be some overlaps.

## Europe ('europe'):

from European marine fishes: 

> The geographic scope of the European Red List assessment
encompasses the Mediterranean Sea, the Black Sea, the
Baltic Sea, the North Sea and/or the European part of
the Atlantic Ocean (i.e., the territorial waters and the
Exclusive Economic Zones (EEZs) of all European
countries in the eastern part of the Atlantic Ocean,
also including the EEZs of the Macaronesian islands
belonging to Portugal and Spain), the North Sea and
the Northeastern Atlantic Ocean, excluding the EEZs of
Greenland, Morocco and Western Sahara.

Options: 

* LME 20-26 (26 = Mediterranean) and 59 (Iceland)?
* MEOW 2 (Northern Iceland/Svalbard), [21:27, 30:31, 34:36] (Mediterranean), 44 (Black Sea)
* Alternately: Use EEZs for all countries listed as Europe by UN classification

## Northern Africa 

* LME 33 (Red Sea), 26 (Mediterranean), 27 (West Coast of North Africa)
* MEOW 87-88 (Red Sea), [32, 33, 35, 36] (Mediterranean), 28, 29 (Lusitanean upwelling)

## Southern Africa 

* LME 29 (Western Southern Africa) and 30 (Madagascar)
* MEOW 192:194 (Temperate southern Africa)

## Eastern Africa 

* LME 30 (Madagascar) and 31 (Horn of Africa)
* MEOW 94:102 (Western Indian Ocean)

## Western Africa   

* LME 28
* MEOW 79:86 (West African Transition, Gulf of Guinea)

## Central Africa  

* LME 28, 29
* If this is truly "Central" then no marine anyway! ignore

## Northeastern Africa 

* LME 32, 33
* MEOW 87:93 (Red Sea, Gulf of Aden, Somali/Arabian)

## Pan-Africa 

* LME 26-33
* MEOW 28, 29, 32, 33, 35, 36 (Med and N), 
    79:86 (West), 87:89, 93 (NE), 94:102 (E), 192:194 (S)

## Gulf of Mexico 

* LME 5
* MEOW 43 (N Gulf), 68:70 (S Gulf)

## Mediterranean

* LME 26 and 62 for Mediterranean and Black Seas
* MEOW 30:36, 44

## Persian Gulf

* LME 32 (NOTE: this encompasses the entire Arabian Sea)
* MEOW 90:91


```{r}
meow <- st_read(file.path(dir_anx, "_raw_data", "meow"))

plot(meow["ECOREGION"])
```


```{r}
unique(rgn_risk_ts$iucn_rgn)

rgn_to_meow <- c('europe'             = paste(c(2,21:27, 30:31, 34:36, 44), collapse = ','), #updated
                'western_africa'      = paste0(79:83, collapse = ','), #updated
                's._africa_fw'     = paste0(192:194, collapse = ','),
                'pan-africa'          = paste0(c(28, 29, 32, 33, 35, 36, 79:89, 93:102, 192:194), #updated
                                                 collapse = ','),
                'gulf_of_mexico'      = paste0(c(43, 69:70), collapse = ','), #updated
                'mediterranean'       = paste0(c(30:36, 44), collapse = ','), #updated
                'persian_gulf'        = paste0(c(90:91), collapse = ','), #updated
                'global'              = paste0(c(1:232), collapse = ','),
                'arabian_sea'         = paste0(c(89, 91:92, 103, 105:106), collapse = ','), #updated
                'caribbean'           = paste0(c(63:68), collapse = ","), #updated
                'central_africa'      = paste0(c(84:86), collapse = ',')) #updated

rgn_to_meow_df <- data.frame(iucn_rgn = names(rgn_to_meow),
                             meow_id  = rgn_to_meow) %>%
  mutate(meow_id = str_split(meow_id, ',')) %>%
  unnest(meow_id) %>%
  group_by(iucn_rgn) %>%
  mutate(meow_id = as.integer(meow_id),
         priority = n()) %>%
  arrange(priority)

write_csv(rgn_to_meow_df, file.path(dir_goal, sprintf('_spatial/iucn_rgn_to_meow.csv')))
```


UPDATE 


``` {r code LMEs}

rgn_to_lme <- c('europe'              = '20,21,22,23,24,25,26,59',
                'northern_africa'     = '26,27,33',
                'western_africa'      = '28', 
                'central_africa'      = '28,29',
                'northeastern_africa' = '32,33',
                'southern_africa'     = '29,30',
                'eastern_africa'      = '30,31',
                'pan-africa'          = '26,27,28,29,30,31,32,33',
                'gulf_of_mexico'      = '5',
                'mediterranean'       = '26,62', 
                'persian_gulf'        = '32',
                'global'              = paste0(1:66, collapse = ','))

rgn_to_lme_df <- data.frame(iucn_rgn = names(rgn_to_lme),
                            lme_id = rgn_to_lme) %>%
  mutate(lme_id = str_split(lme_id, ',')) %>%
  unnest(lme_id) %>%
  group_by(iucn_rgn) %>%
  mutate(lme_id = as.integer(lme_id),
         priority = n()) %>%
  arrange(priority)

write_csv(rgn_to_lme_df, file.path(dir_goal, sprintf('_spatial/iucn_rgn_to_lme.csv')))

```

Where an LME is assigned to multiple regions, use a prioritization then averaging method:

* Prioritize by the region comprising the smallest number of LMEs; e.g. Mediterranean has 1 LME, North Africa has 3, Europe has 8; they all overlap in LME 26.  So for the score for LME 26 for a species assessed in all three regions, keep the score for the Mediterranean.
* If after prioritizing, multiple regions still overlap in an LME, take the average score of the overlapping regional assessments for that LME.


# Summary

To define spatial areas more accurately, create rasters of WDPA and ocean area at Gall-Peters resolution of 0.25 km^2. These can then be aggregated back to Gall Peters 10 km^2 working rasters, summing the total ocean area and total protected area within each working cell.

* Ocean area - use Natural Earth 10 m land polygons; mask over cell ID raster to exclude land.

# Data Sources

### WDPA
# Data Source
**Reference**: IUCN and UNEP-WCMC (2025), The World Database on Protected Areas (WDPA) [On-line], June 2025. Cambridge, UK: UNEP-WCMC. Available at: www.protectedplanet.net.

**Downloaded**: June 22, 2025

**Description**:  Shapefile of World Database on Protected Areas

**Time range**: 1800 - 2025; some protected areas do not have an associated "status year" and are reported as year 0.

**Format**:  Shapefile

**File location**: `Mazu:git-annex/globalprep/_raw_data/wdpa_mpa/d2025/WDPA_Jun2025_Public_shp/`

# Methods

## Ocean area raster

* Project Natural Earth 10M ocean polygons into Gall-Peters equal area projection
* rasterize at 500 m resolution
* aggregate up to 10 km resolution

### Create a 10 km^2 Gall Peters projection cell ID raster

 - Note: Just copy this over from last years, it doesn't change. 
 
```{r}
ocean_area_file <- file.path(dir_spatial, 'ocean_area_rast.tif')

cell_id_file <- file.path(dir_spatial, 'cell_id_rast.tif')

# copy from last year
ocean_500m_file  <- file.path(dir_anx, 'spp', 'v2024', 'spp_risk_dists', 'spatial', 'ocean_500m.tif')
```

## Create WDPA MPA percent protection data frame

Marine regions are any protected area with MARINE flag of 1 or 2, or non-zero GIS_M_AREA.  Classify by IUCN category (I-VI -> 1-6; non-classified get 8).  Include only designated, adopted, and established parks.  Omit non-MPA management plans from U.S.  Rasterize to CHI resolution, calculate protected cells vs total cells to get percent of protection for each IUCN protection category.

Because we are rasterizing the protected areas down to 500 m resolution, we risk dropping MPAs smaller than about 1 km^2.  We assume these are not likely to have a significant impact on biodiversity protection, but see the sensitivity analysis.

### Create WDPA map at GP 500 m (.25 km^2) resolution

Actually, create two maps: one by protection category, another by year for no-take only (categories Ia, Ib, II, and no-take).

```{r}
rast_cell_id <- rast(cell_id_file)

# Goal file
wdpa_marine_file <- file.path(dir_raw_data_anx, 'wdpa_mpa/d2025/wdpa_mar_by_cat_Jun2025_500m.tif')
wdpa_marine_file_int <- file.path(dir_raw_data_anx, 'wdpa_mpa/d2025/wdpa_mar_by_cat_Jun2025_500m_int.tif')


wdpa_poly_file_0   <- file.path(dir_raw_data_anx, 'wdpa_mpa/d2025/WDPA_Jun2025_Public_shp/WDPA_Jun2025_Public_shp_0', 
                              'WDPA_Jun2025_Public_shp-polygons.shp')
wdpa_poly_file_1   <- file.path(dir_raw_data_anx, 'wdpa_mpa/d2025/WDPA_Jun2025_Public_shp/WDPA_Jun2025_Public_shp_1', 
                              'WDPA_Jun2025_Public_shp-polygons.shp')
wdpa_poly_file_2   <- file.path(dir_raw_data_anx, 'wdpa_mpa/d2025/WDPA_Jun2025_Public_shp/WDPA_Jun2025_Public_shp_2', 
                              'WDPA_Jun2025_Public_shp-polygons.shp')

if(!file.exists(wdpa_marine_file)) {
  
  wdpa_poly_0 <- sf::st_read(wdpa_poly_file_0)
  wdpa_poly_1 <- sf::st_read(wdpa_poly_file_1)
  wdpa_poly_2 <- sf::st_read(wdpa_poly_file_2)
  
  wdpa_poly <- rbind(wdpa_poly_0, wdpa_poly_1, wdpa_poly_2)

   #wdpa_poly$IUCN_CAT %>% unique() 
  ##only grab ones that are reported/assigned/applicable
  iucn_cats <- c('Ia'  = 1,
                 'Ib'  = 1,
                 'II'  = 2,
                 'III' = 3,
                 'IV'  = 4,
                 'V'   = 5,
                 'VI'  = 6)
   wdpa_poly$STATUS %>% table()
     # Adopted   Designated  Established    Inscribed Not Reported     Proposed 
     #      34       254937         2401          250           34         1219 
  # wdpa_poly$NO_TAKE %>% unique()
  # x <- wdpa_poly %>% filter(NO_TAKE %in% c('All', 'Part'))
  # x <- wdpa_poly %>% filter(WDPAID == 309888)
  
  wdpa_marine <- wdpa_poly %>%
    filter(MARINE > 0 | GIS_M_AREA > 0) %>%
    filter(STATUS %in% c('Designated', 'Adopted', 'Established')) %>%
      ### no paper parks!
    filter(!str_detect(tolower(MANG_PLAN), 'non-mpa')) %>%
      ### omit non-MPA fisheries or species management plans!
    mutate(NO_TAKE = ifelse(WDPAID == 309888, TRUE, NO_TAKE),
             ### patch PIPA to be no take - currently "not reported"
           no_take = (NO_TAKE == 'All') | (NO_TAKE == 'Part' & NO_TK_AREA > 0.75 * GIS_M_AREA),
             ### if NO_TK_AREA is 75% or more of GIS area, count it...
           cat = iucn_cats[IUCN_CAT],
           cat = ifelse(no_take & !cat %in% 1:2, -1, cat), ### use -1 as a "no take" flag
           cat = ifelse(is.na(cat), 8, cat)) %>%           ### use 8 as an "other protected" flag
    arrange(cat) %>%
    st_transform(crs = gp_proj4)

  
  wdpa_marine_vect <- vect(wdpa_marine)

  rm(wdpa_marine)
  rm(wdpa_poly_0)
  rm(wdpa_poly_1)
  rm(wdpa_poly_2)
  rm(wdpa_poly) ## free up some space
    
  ### rasterize all protected areas to 500m global raster;
  ### use min value: will prioritize no-take and higher levels of protection
  terra::rasterize(x = wdpa_marine_vect, 
                                        y = rast_ocean_500m,
                                        field  = 'cat', 
                                        fun    = 'min', 
                                        filename = wdpa_marine_file_int)   

  #plot(wdpa_all_rast_500m)
  
  wdpa_all_rast_500m <- rast(wdpa_marine_file_int)
  
  rm(wdpa_marine_vect) ## free up some space
  
  rast_ocean_500m <- terra::classify(rast_ocean_500m ,cbind(-Inf, 0.005, NA), right=FALSE)
  # plot(rast_ocean_500m)
  
  wdpa_marine_rast_500m <- terra::mask(wdpa_all_rast_500m, rast_ocean_500m)
  
  # plot(wdpa_marine_rast_500m)
  # plot(rast(file.path(dir_raw_data_anx, 'wdpa_mpa/d2021/wdpa_mar_by_cat_Feb2021_500m.tif')))
  
  writeRaster(wdpa_marine_rast_500m, wdpa_marine_file, 
              overwrite = TRUE)
  
}


```
```{r}

rast_ocean_500m <- rast(ocean_500m_file)
#plot(rast_ocean_500m) #check to see if loaded properly 

# Goal file
wdpa_prot_yr_file <- file.path(dir_raw_data_anx, 'wdpa_mpa/d2025', 
                               'wdpa_mar_by_prot_yr_Jun2025_500m.tif')

wdpa_prot_yr_file_int <- file.path(dir_raw_data_anx, 'wdpa_mpa/d2025/wdpa_mar_by_prot_yr_Jun2025_500m_int.tif')

# Existing files
wdpa_poly_file_0   <- file.path(dir_raw_data_anx, 'wdpa_mpa/d2025/WDPA_Jun2025_Public_shp/WDPA_Jun2025_Public_shp_0', 
                              'WDPA_Jun2025_Public_shp-polygons.shp')
wdpa_poly_file_1   <- file.path(dir_raw_data_anx, 'wdpa_mpa/d2025/WDPA_Jun2025_Public_shp/WDPA_Jun2025_Public_shp_1', 
                              'WDPA_Jun2025_Public_shp-polygons.shp')
wdpa_poly_file_2   <- file.path(dir_raw_data_anx, 'wdpa_mpa/d2025/WDPA_Jun2025_Public_shp/WDPA_Jun2025_Public_shp_2',
                                'WDPA_Jun2025_Public_shp-polygons.shp')

if(!file.exists(wdpa_prot_yr_file)) {
  
  wdpa_poly_0 <- sf::st_read(wdpa_poly_file_0)
  wdpa_poly_1 <- sf::st_read(wdpa_poly_file_1)
  wdpa_poly_2 <- sf::st_read(wdpa_poly_file_2)
  
  wdpa_poly <- rbind(wdpa_poly_0, wdpa_poly_1, wdpa_poly_2)

   #wdpa_poly$IUCN_CAT %>% unique()
  iucn_cats <- c('Ia'  = 1,
                 'Ib'  = 1,
                 'II'  = 2,
                 'III' = 3,
                 'IV'  = 4,
                 'V'   = 5,
                 'VI'  = 6)
    #wdpa_poly$STATUS %>% table()
         # Adopted   Designated  Established    Inscribed Not Reported     Proposed 
         #  34       242940         1148          249           34         1201 
  # wdpa_poly$NO_TAKE %>% unique()
  # x <- wdpa_poly %>% filter(NO_TAKE %in% c('All', 'Part'))
  
  wdpa_no_take <- wdpa_poly %>%
    filter(MARINE > 0 | GIS_M_AREA > 0) %>%
    filter(STATUS %in% c('Designated', 'Adopted', 'Established')) %>%
      ### no paper parks!
    filter(!str_detect(tolower(MANG_PLAN), 'non-mpa')) %>%
      ### omit non-MPA fisheries or species management plans!
    mutate(NO_TAKE = ifelse(WDPAID == 309888, TRUE, NO_TAKE),
             ### patch PIPA to be no take - currently "not reported"
           no_take = (NO_TAKE == 'All') | (NO_TAKE == 'Part' & NO_TK_AREA > 0.75 * GIS_M_AREA),
             ### if NO_TK_AREA is 75% or more of GIS area, count it...
           cat = iucn_cats[IUCN_CAT],
           cat = ifelse(no_take & !cat %in% 1:2, -1, cat), ### use -1 as a "no take" flag
           cat = ifelse(is.na(cat), 8, cat)) %>%           ### use 8 as an "other protected" flag
    arrange(cat) %>%
    filter(cat <= 2) %>% ### this limits to categories Ia, Ib, II, and no-take
    st_transform(crs = gp_proj4)
  
  rm(wdpa_poly_0 , wdpa_poly_1, wdpa_poly_2, wdpa_poly)
  
  wdpa_no_take_vect <- vect(wdpa_no_take)
  
  rm(wdpa_no_take, wdpa_marine_rast_500m)
  
  ### rasterize all protected areas to 500 m global raster;
  ### use min value: will prioritize no-take and higher levels of protection
           wdpa_all_rast_500m <-   terra::rasterize(x = wdpa_no_take_vect, 
                                        y = rast_ocean_500m,
                                        field  = 'STATUS_YR', 
                                        fun    = 'min')
           
           writeRaster(wdpa_all_rast_500m, wdpa_prot_yr_file_int)
  
  rm(wdpa_no_take_vect)
  
  rast_ocean_500m <- terra::classify(rast_ocean_500m ,cbind(-Inf, 0.005, NA), right=FALSE)
  
    ### in case of floating point not-quite-zeros...
  
  wdpa_all_rast_500m <- rast(wdpa_prot_yr_file_int)
  
   terra::mask(wdpa_all_rast_500m, rast_ocean_500m, filename = wdpa_prot_yr_file)
   
   
   # wdpa_yr_rast_500m <- rast(wdpa_prot_yr_file)
    #plot(wdpa_yr_rast_500m)
  #plot(raster(file.path(dir_raw_data_anx, 'wdpa_mpa/d2020/wdpa_mar_by_prot_yr_Apr2020_500m.tif')))
  
  # writeRaster(wdpa_yr_rast_500m, wdpa_prot_yr_file, 
  #             overwrite = TRUE)
  # 
}

# View
plot(wdpa_all_rast_500m)
```
### Crosstab WDPA map to cell IDs

Be cautious when using `raster::crosstab()` as it returns factors that need to be coerced into integers... This code results in a file of cell IDs, IUCN protected area categories (1 = Ia and Ib, 2-6 = II - VI, 7 = no take, -1 = other), and proportion of cell protected by that category.

We can't simply aggregate upwards here, since we need to keep track of different protection levels (for the first) and year (for the second).

#### Spatial: `wdpa_mpa_area.csv`

``` {r crosstab mpa areas}

wdpa_marine_rast_500m <- rast(wdpa_marine_file)
rast_cell_id    <- rast(cell_id_file)    ### cell IDs at 10 km^2
rast_ocean_500m <- rast(ocean_500m_file)  ### ocean presence at 1 km^2
rast_ocean_area <- rast(ocean_area_file) ### ocean proportional area at 10 km^2

cell_id_500m_file <- file.path(dir_goal_anx, 'spatial/cell_id_500m.tif')

### Need a 500 m raster of cell IDs for crosstabbing...
#if(!file.exists(cell_id_500m_file)) {
rast_cell_id_500m <- terra::resample(x = rast_cell_id, y = rast_ocean_500m,
                                     filename = cell_id_500m_file,
                                     method = 'near',
                                     overwrite = TRUE)
# } else {
#   rast_cell_id_500m <- rast(cell_id_500m_file)
# }
#plot(rast_cell_id_500m)
#plot(raster(file.path("/home/shares/ohi/git-annex/globalprep/spp/v2020/spp_risk_dists/spatial/cell_id_500m.tif")))

wdpa_cell_area_file <- file.path(dir_spatial, 'wdpa_mpa_area.csv')

### now break into chunks and mclapply across them to crosstab WDPA categories to cell IDs

### Break into smaller chunks for faster processing and eventual parallelizing
n_chunks <- 60

crosstab_chunk <- function(rast1, rast2, chunk, n_chunks) {
  ## chunk <- 2
  # rast1 = wdpa_marine_rast_500m
  # rast2 = rast_cell_id_500m
  
  rast1 <- raster(rast1)
  rast2 <- raster(rast2)
  
  chunk_size <- ceiling(ncol(rast1) / n_chunks)
  left_bound <- (chunk - 1) * chunk_size + 1
  right_bound <- min(left_bound + chunk_size - 1, ncol(rast1))
  
  # chunk_ext <- terra::ext(xFromCol(rast1, left_bound), xFromCol(rast1, right_bound), yFromRow(rast1, nrow(rast1)), yFromRow(rast1, 1))
  
  chunk_ext <- extent(rast1, 1, nrow(rast1), left_bound, right_bound)
  
  message('Processing ', chunk, ': ', paste0(as.character(round(chunk_ext)), collapse = ', '))
  wdpa_chunk    <- raster::crop(rast1, chunk_ext)
  cell_id_chunk <- raster::crop(rast2, chunk_ext)
  
  wdpa_cells <- raster::crosstab(wdpa_chunk, cell_id_chunk, 
                                 long = TRUE) %>%
    setNames(c('wdpa_category', 'cell_id', 'n_prot')) %>%
    mutate(wdpa_category = as.integer(as.character(wdpa_category)),
           cell_id       = as.integer(as.character(cell_id)))
  ### wdpa_category and cell_id are crosstabbed as factors - first
  ### convert to character (to unfactorize it) then to integer.  Otherwise
  ### you end up with factor index, not actual cell ID or category.
  
  return(wdpa_cells)
}

### Use the function in an mclapply call
system.time({
  wdpa_cells_list <- parallel::mclapply(1:n_chunks, mc.cores = 12,
                                        FUN = function(x) crosstab_chunk(wdpa_marine_rast_500m, 
                                                                         rast_cell_id_500m, 
                                                                         chunk = x, n_chunks))
  #     user   system  elapsed 
  # 5335.773  588.892  550.586  - about ten minutes! 
  
})


wdpa_cells_df <- bind_rows(wdpa_cells_list)

area_df <- data.frame(cell_id  = values(rast_cell_id),
                      area_km2 = values(rast_ocean_area)) %>%
  rename(cell_id = 1, area_km2 = 2)

wdpa_area_df <- wdpa_cells_df %>%
  full_join(area_df, by = 'cell_id') %>%
  filter(!is.na(wdpa_category)) %>%
  mutate(prot_area_km2 = n_prot / 4, ### count of cells in 500 x 500 m raster, div by 4
         mpa_pct = prot_area_km2 / area_km2) %>%
  dplyr::select(wdpa_category, cell_id, mpa_pct, prot_area_km2) %>%
  ### mpa pct is relative to *ocean* area - not *total cell* area
  group_by(wdpa_category, cell_id) %>%
  summarize(mpa_pct = sum(mpa_pct), prot_area_km2 = sum(prot_area_km2)) %>%
  ungroup()

# old <- read_csv(file.path(here(), "globalprep/spp/v2021/_spatial/wdpa_mpa_area.csv"))

sum(old$prot_area_km2) # 26841457
sum(wdpa_area_df$prot_area_km2) # 24690672 

write_csv(wdpa_area_df, wdpa_cell_area_file)

```


``` {r crosstab mpa years}

wdpa_yr_rast_500m <- raster(wdpa_prot_yr_file)
rast_cell_id    <- raster(cell_id_file)    ### cell IDs at 10 km^2
rast_ocean_500m  <- raster(ocean_500m_file)  ### ocean presence at 1 km^2
rast_ocean_area <- raster(ocean_area_file) ### ocean proportional area at 10 km^2

cell_id_500m_file <- file.path(dir_goal_anx, 'spatial/cell_id_500m.tif')
rast_cell_id_500m <- raster(cell_id_500m_file)

wdpa_cell_year_file <- file.path(dir_spatial, 'wdpa_mpa_year.csv')

### now break into chunks and mclapply across them to crosstab WDPA categories to cell IDs
if(!file.exists(wdpa_cell_year_file)) {
  ### Break into smaller chunks for faster processing and eventual parallelizing
  n_chunks <- 60
  
  crosstab_chunk2 <- function(rast1, rast2, chunk, n_chunks) {
    ### chunk <- 30
    chunk_size <- ceiling(ncol(rast1) / n_chunks)
    left_bound <- (chunk - 1) * chunk_size + 1
    right_bound <- min(left_bound + chunk_size - 1, ncol(rast1))
    
    chunk_ext <- extent(rast1, 1, nrow(rast1), left_bound, right_bound)
    message('Processing ', chunk, ': ', paste0(as.character(round(chunk_ext)), collapse = ', '))
    wdpa_chunk    <- crop(rast1, chunk_ext)
    cell_id_chunk <- crop(rast2, chunk_ext)
    
    wdpa_cells <- crosstab(wdpa_chunk, cell_id_chunk, 
                           progress = 'text',
                           long = TRUE) %>%
      setNames(c('mpa_year', 'cell_id', 'n_prot')) %>%
      mutate(mpa_year = as.integer(as.character(mpa_year)),
             cell_id  = as.integer(as.character(cell_id)))
        ### mpa_year and cell_id are crosstabbed as factors - first
        ### convert to character (to unfactorize it) then to integer.  Otherwise
        ### you end up with factor index, not actual cell ID or category.
    
    return(wdpa_cells)
  }

  ### Use the function in an mclapply call
  system.time({
    wdpa_cells_list <- parallel::mclapply(1:n_chunks, mc.cores = 12,
      FUN = function(x) crosstab_chunk2(wdpa_yr_rast_500m, 
                                        rast_cell_id_500m, 
                                        chunk = x, n_chunks))
#     user   system  elapsed 
# 5401.628  619.072  562.306 - about 10 mins
  })
  
  
  wdpa_cells_df <- bind_rows(wdpa_cells_list)
  
  area_df <- data.frame(cell_id  = raster::values(rast_cell_id),
                        area_km2 = raster::values(rast_ocean_area))
  
  wdpa_year_df <- wdpa_cells_df %>%
    full_join(area_df, by = 'cell_id') %>%
    filter(!is.na(mpa_year)) %>%
    mutate(prot_area_km2 = n_prot / 4, 
              ### count of cells in 500 x 500 m^2 raster, times .25 km^2 area per cell
           mpa_pct = prot_area_km2 / area_km2) %>%
    dplyr::select(mpa_year, cell_id, mpa_pct, prot_area_km2) %>%
      ### mpa pct is relative to *ocean* area - not *total cell* area
    group_by(mpa_year, cell_id) %>%
    summarize(mpa_pct = sum(mpa_pct), prot_area_km2 = sum(prot_area_km2)) %>%
    ungroup()
  
  
  write_csv(wdpa_year_df, wdpa_cell_year_file)
  
} else {
  # git_prov(wdpa_cell_year_file, 'output')
}

```
