---
title: 'Set up IUCN marine species data'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE, eval = FALSE)
```

```{r}
#install.packages("rredlist")
library(rredlist)
library(dplyr)
library(tidyr)
library(tidyverse)
library(here)

#source('https://raw.githubusercontent.com/oharac/src/master/R/common.R')
source(here('workflow/R/common.R'))

goal     <- 'spp'
scenario <- 'v2025'

dir_goal <- here('globalprep', goal, scenario)

### goal specific folders and info
dir_anx  <- file.path(dir_M, 'git-annex/globalprep')
dir_setup   <- file.path(dir_goal, '_setup')
dir_goal_anx <- file.path(dir_anx, goal, scenario, 'spp_risk_dists')

source(file.path(dir_setup, 'common_fxns.R'))

```

### Set up API

To access the API v4, you will need to sign up for a key on the [IUCN website](https://api.iucnredlist.org/). This key can be added to the local .renviron on your machine and retrieved using the code below.

```{r}
# Save API key
rl_use_iucn()
api_version <- rl_version()

# Clean up the time series: reclassify old codes to current codes
cat_lookup <- read_csv(file.path(dir_setup, 'raw', 'risk_code_lookup.csv'), show_col_types = FALSE)
```

# Summary

Get IUCN current risk assessments and historical assessments over time, for all IUCN marine species.  This method is not reliant on any information from AquaMaps.

# Data Sources

### IUCN Red List

# Methods

## Marine Habitat Assessment

Using the `rl_habitats()` function, get the species list for species coded with the following marine habitat codes: 9-12, plus 15.11, 15.12, 15.13.  Note category 13 is Marine Coastal/Supratidal, but includes many species whose "marine" dependence is only incidental. We include seacliffs and offshore islands (13.1) from category 13. If these species do not show up in category 12 (marine intertidal) then they are assumed to not actually depend on marine habitats.

The list can be found here (Make sure to compare the webpage to the final df, to make sure it matches up.): https://www.iucnredlist.org/resources/habitat-classification-scheme

```{r}
# Get a list of habitats
iucn_habitats <- rl_habitats()$habitats

# Unlist the description column
iucn_habitats$description <- iucn_habitats$description$en

# Select desired marine habitats (noted above)
# marine_habitats <- iucn_habitats %>%
#   dplyr::filter(str_detect(tolower(description), "marine|reef|lagoon") &
#                   !str_detect(tolower(description), "supratidal"))
marine_habitats <- iucn_habitats %>%
  filter(str_detect(code, "^(9_|10_|11_|12_)") | 
           code %in% c("13_1", "15_11", "15_12", "15_13"))

# Initiate empty list
marine_habitats_spp <- list()

# Pull species for each habitat
for (i in marine_habitats$code) { 
  
  # Save habitat name
  habitat_name <- marine_habitats$description[marine_habitats$code == i]
  
  # Get list of assesments
  habitat_assessments <- rl_habitats(code = i)$assessments
  
  # Append to list
  marine_habitats_spp[[habitat_name]] <- dplyr::as_tibble(habitat_assessments)
}

# Combine individual habitat species lists into one
marine_species <- bind_rows(marine_habitats_spp) %>%
  #filter(latest == TRUE) %>%
  rename(iucn_sid = sis_taxon_id, 
         sciname = taxon_scientific_name,
         category = red_list_category_code,
         year = year_published) 

# List of marine species
marine_species_list <- unique(marine_species$iucn_sid)

# 2025: 20,236 species
```

### Data: `iucn_risk_current_2025-1.csv`

Since our initial download has multiple entries (multiple years) for some species that show up in different habitat codes, I chose to keep only the most recent entry between the two habitat codes.

**2025 NOTE:** About ~907 rows of species have been assessed in 2025. This could account for some of the increase of about 1,500 species from last year.

```{r}

#marine_species_list_test <- c(222, 224, 225)
start <- proc.time()

# Global IUCN species ratings
marine_spp <- data.frame()

for (i in marine_species_list) {
  
  # Get global ICUN information for each marine species
  species_assessment <- rl_sis(id = i)$assessment
  
  # Append to the dataframe
  marine_spp <- bind_rows(marine_spp, species_assessment)
  
  # Rest between queries
  Sys.sleep(1)
  
}

end <- proc.time()

print("Species retrieval time: ", end-start)

# Clean up dataframe
marine_species_risk <- marine_spp %>%
  # Filter for current rating
  filter(latest == TRUE) %>%
  rename(iucn_sid = sis_taxon_id, 
         sciname = taxon_scientific_name,
         category = red_list_category_code,
         year = year_published) %>%
  # Translate to current codes
  left_join(cat_lookup, by = c("category" = "code")) %>%
  rename(old_cat = category,
         cat = code_current) %>%
  # Add version column
  mutate(iucn_version = "2025-1") %>%
  select(iucn_sid, sciname, year, old_cat, cat, cat_score, iucn_version)

## add another column? criteria?




marine_species_coded <- marine_species %>%
  # We just want species for this csv
  select(-code) %>%
  # Keep category only from most recent year DECISION POINT
  group_by(iucn_sid, sciname) %>%
  slice_max(order_by = year, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  left_join(cat_lookup, by = c("category" = "code")) %>%
  # Add version column
  mutate(iucn_version = "2025-1")

# Create df for iucn_risk_current_2025-1.csv
marine_species_risk <- marine_species_coded %>%
  rename(old_cat = category,
         cat = code_current) %>%
  select(iucn_sid, sciname, year, old_cat, cat, cat_score, iucn_version)

# Write csv
write_csv(marine_species_risk, file.path(dir_goal, '_data', 
                                         sprintf('iucn_risk_current_%s.csv', api_version)))
```

### Data: `spp_marine_from_api_2025-1.csv`

Since some species show up over multiple habitats (and thus depth ratings), I chose to assign the deepest rating.

| habitat type   | habitat code   | depth classification  |
| -------------- | -------------- | --------------------- |
| deep benthic   | 10_3, 10_4     | "deep oceanic"        |
| deep oceanic   | 10_ , 11_      | "200 m +"             |
| neritic        | 9_, 15_        | "< 200 m"             |
| intertidal     | 12_            | "<20 m"               |

```{r}
# Categorize by depths
marine_species_depths <- marine_species %>%
  filter(latest == TRUE) %>%
  # Break down the codes into cat, subcat1, and subcat2
  separate(code, c('cat', 'subcat1', 'subcat2'),
           remove = FALSE, convert = TRUE) %>%
  arrange(cat, subcat1, subcat2) %>%
  # Assign descriptions to each cat
  mutate(
    max_depth = factor(case_when(
      cat == 10 & subcat1 %in% 3:4         ~ "deep oceanic",
      cat %in% c(10, 11)                   ~ "200 m +",
      cat %in% c(9, 15)                    ~ "< 200 m",
      cat == 12                            ~ "< 20 m",
      TRUE                                ~ "unknown"),
    # To make it easier to pick the deepest (min) depth rating
    levels = c("deep oceanic", "200 m +", "< 200 m", "< 20 m", "unknown")))%>%
  left_join(iucn_habitats, by = "code")

# spp_marine_from_api_2025-1.csv
habs <- marine_species_depths %>%
  group_by(iucn_sid) %>%
  summarise(
    habs = paste(unique(code), collapse = ", "),
    # Pick the deepest rating (the lowest (min) level)
    max_depth = as.character(max_depth[which.min(as.integer(max_depth))])) %>%
  ungroup()

# Write csv
write_csv(habs, file.path(dir_goal, '_data',
                          sprintf('spp_marine_from_api_%s.csv', api_version)))
```

## Regional Assessment

Using the `rl_scopes()` function, get the species list for species coded with the following regions: Europe (2), Mediterranean (4), Western Africa (8), S. Africa FW (9), Pan-Africa (10), Central Africa (16), Northeastern Africa (17), Eastern Africa (18), Northern Africa (21), Gulf of Mexico (45433062), Caribbean (45433063), Persian Gulf (45433064), Arabian Sea (100765562). 

Essentially all but Global (1).

```{r}
# Get list of scopes (regions)
scopes <- rl_scopes()$scopes

# Unlist description
scopes$description <- scopes$description$en

# Remove "Global" (see above)
scopes <- scopes %>%
  filter(!description == "Global")

# Initiate empty list
scopes_assessments <- list()

# Pull species for each region
for (i in scopes$code) { 
  
  # Save region name
  scope_name <- scopes$description[scopes$code == i]
  
  # Get list of assesments
  scope_assessments <- rl_scopes(code = i)$assessments 
  
  # Append to list
  scopes_assessments[[scope_name]] <- dplyr::as_tibble(scope_assessments)
}

# Combine assessments
rgn_assess <- dplyr::bind_rows(scopes_assessments) %>%
  # filters for most recent assessment?
  filter(latest == TRUE) %>%
  dplyr::rename(iucn_sid = sis_taxon_id, 
                sciname = taxon_scientific_name, 
                category = red_list_category_code,
                year = year_published) %>%
  dplyr::select(iucn_sid, sciname, year, category, code) %>%
  mutate(year = as.integer(year))

```

### Data: `iucn_risk_rgn_current_2025-1.csv`

```{r}
# Keep the species that overlap with regions of interest
rgn_assess_marine <- rgn_assess %>%
  # filter for marine species
  filter(iucn_sid %in% marine_species_list) %>%
  # Add region description
  left_join(scopes, by = "code") %>%
  mutate(iucn_rgn = str_replace_all(tolower(description), " ", '-')) %>%
  left_join(cat_lookup, by = c("category" = "code")) %>%
  select(iucn_sid, 
         sciname,
         old_cat = category,
         iucn_rgn, 
         cat_score, 
         cat_current = code_current) 

# Write csv
write_csv(rgn_assess_marine, file.path(dir_goal, sprintf('_data/iucn_risk_rgn_current_%s.csv', api_version)))

# Check
nrow(rgn_assess_full) #5032
nrow(read_csv(file.path(dir_goal, sprintf('_data/iucn_risk_rgn_current_%s.csv', "2024-1")), show_col_types = FALSE)) #2181
```

Once regional assessments are gleaned from the API, check them against the global marine species list.  Species are matched by ID numbers.  Species whose regional category differs from the global category are kept and scored.

* note, quick check of scientific names showed they all match
* 113 spp fall into this...

```{r}
marine_spp %>%
  rename(global_cat = )


rgn_assess_match <- rgn_assess %>%
  # Add region description
  left_join(scopes, by = "code") %>%
  mutate(iucn_rgn = str_replace_all(tolower(description), " ", '-')) %>%
  select(-c(code, description)) %>%
  # load marine species risk again? keep only species in common
  inner_join(marine_species, by = c("iucn_sid", "sciname", "year")) %>%
  select(iucn_sid, 
         sciname, 
         rgn_cat = category, 
         iucn_rgn, cat_score, 
         global_cat = cat) 

rgn_spp_list <- rgn_assess_match %>%
  mutate(match = ifelse(global_cat == rgn_cat, TRUE, FALSE)) %>%
  filter(match == FALSE)

#265 differences
# this may not reflect "global" cat rather most rescent assessment from one habitat
```

## Get IUCN regional historical assessments

### Data: `iucn_risk_ts_rgn_2025-1.csv`

**2025 NOTE:** This .csv file has species from all the regions, while v2024 only included "europe" and "mediterranean". This appears to be why v2024 has only 388 rows while v2025 has 1019 rows. 

Upon filtering v2025 for med and europe there are ~800 values

```{r}
# Combine species lists by region (this time don't filter for latest)
rgn_assess_hist <- bind_rows(scopes_assessments) %>%
  rename(iucn_sid = sis_taxon_id, 
         sciname = taxon_scientific_name, 
         category = red_list_category_code,
         year = year_published) %>%
  # Filter for species in the marine species list
  filter(iucn_sid %in% marine_species_list) %>%
  select(iucn_sid, 
         year,
         sciname,
         old_cat = category, 
         iucn_rgn = code) %>%
  # Add new updated cat codes
  left_join(cat_lookup, by = c('old_cat' = 'code')) %>%
  distinct() %>%
  filter(!is.na(cat_score) & !is.na(year)) %>%
  arrange(iucn_sid, year) %>%
  select(iucn_sid, 
         year, 
         rgn_cat_ts = code_current, 
         rgn_cat_ts_score = cat_score, 
         iucn_rgn) %>%
  mutate(iucn_version = api_version) %>%
  group_by(iucn_sid, iucn_rgn) %>%
  filter(n() >= 2) %>%
  ungroup() %>%
  # Add region description
  left_join(scopes, by = c("iucn_rgn" = "code")) %>%
  mutate(iucn_rgn = str_replace_all(tolower(description), " ", "-")) %>%
  select(-description)

# Write csv
write_csv(rgn_assess_hist, file.path(dir_goal, '_data',
                                 sprintf('iucn_risk_ts_rgn_%s.csv', api_version)))

```


```{r}
# Remove "Global" (see above)
global <- rl_scopes(code = "1")$assessments

global_marine <- global %>%
  rename(iucn_sid = sis_taxon_id, 
         sciname = taxon_scientific_name, 
         category = red_list_category_code,
         year = year_published) %>%
  filter(iucn_sid %in% marine_species_list)
```

