---
title: "OHI 2025 - Species subgoal (Biodiversity goal and resilience)"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE, eval = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

library(dplyr)
library(raster)
library(here)
library(ggridges)
library(ggplot2)
library(tidyverse)
library(terra)

goal     <- 'spp'
scenario <- 'v2025'
dir_goal <- here('globalprep', goal, scenario)

source(here('workflow/R/common.R'))

region_data()
rgns_global <- rgns_eez  %>%
  dplyr::select(rgn_id) %>%
  dplyr::filter(rgn_id != 213)
```

# Source

We use the methods and code from Casey O'Hara: https://github.com/oharac/spp_risk_dists
to prepare the data for the species subgoal.

# Summary

Spatial species range data and extinction risk data from IUCN is used to generate regional scores for the Species subgoal (part of the Biodiversity goal) and resilience layers.

**Mean risk status per cell:**
Species ranges are converted to a global spatial raster of 10 km resolution.  

The mean extinction risk for each cell, $\bar{R}_{cell}$,  is calculated by averaging the IUCN extinction risk of the species with ranges overlapping the cell.  

*Risk* is a scaled value representing the species extinction risk category: 
    * 'LC' = 0.0, 'NT' = 0.2, 'VU' = 0.4, 'EN' = 0.6, 'CR' = 0.8, 'EX' = 1.0


$$\bar{R}_{cell} = \frac{\displaystyle\sum_{species}(Risk)}{n_{spp}}$$

**Mean risk status per region:**
The mean extinction risk for a region, $\bar{R}_{SPP}$, is estimated by averaging the risk values of the raster cells falling within each OHI region, with each cell's contribution weighted by the number of species in the cell.

**Species goal model**

The regional risk values are converted to species status scores by subtracting the risk values from 1 and rescaling so a risk value of $\ge 0.75 $ receives a score of zero. 

From Halpern et al (2012):

> The target for the Species sub-goal is to have all species at a risk status of Least Concern. We scaled the lower end of the biodiversity goal to be 0 when 75% species are extinct, a level comparable to the five documented mass extinctions and would constitute a catastrophic loss of biodiversity. 


$$X_{SPP} = \frac{((1 - \bar{R}_{SPP}) - 0.25)}{(1 - 0.25)} * 100%$$

where:

* $X_{SPP}$ is Species goal status
* $\bar{R}_{SPP}$ is mean extinction risk for a region

Trend is calculated using time series of risk categories based on current and past assessments.

# Data Sources

* __Reference__: 
    * IUCN 2025. The IUCN Red List of Threatened Species. Version 2025-1. <http://www.iucnredlist.org>.
        * Shapefiles available from: https://www.iucnredlist.org/resources/spatial-data-download
        * __Downloaded__: July, 2025
    * BirdLife International and Handbook of the Birds of the World (2024) Bird species distribution maps of the world. Version 2024.2. Available at http://datazone.birdlife.org/species/requestdis.
        * Zipped shapefile available from BirdLife International.  
        * __Downloaded__: July 2, 2025
* __Description__:  Shapefiles containing polygons of assessed species ranges; each shapefile represents all assessed species within a comprehensively-assessed (i.e. >90% assessed) taxonomic group.
* __Native data resolution__: NA
* __Time range__: NA
* __Format__:  Shapefile

***

# Methods

There are several steps that need to be taken to get to this point in the data prep.  

Here is an overview of the organization of files and data that are run prior to this:

### Code

#### Run this first! Setup directory: `spp/v20XX/_setup`

1. `1_query_iucn_species_data.Rmd`

2. `2_set_up_spatial_data.Rmd`

3. `3_process_iucn_spatial_data.Rmd`


#### Then run this!  Root directory: `v2025`

1. `1a_biodiversity_maps_comp_assessed.Rmd`
2. `1c_biodiversity_maps_all_spp.Rmd`

***
## Compare all vs. comprehensively assessed species

- Comprehensively-assessed species analyzed in `1a_biodiversity_maps_comp_assessed.Rmd`
- All species analyzed in `1c_biodiversity_maps_all_spp.Rmd`

In the past, we have used all species with IUCN risk assessments to calculate the species subgoal.  However, some of Casey's work suggests it is better to use the taxa groups that have been comprehensively assessed by IUCN (> 90% of species assessed). The general concern is that IUCN tends to oversample species in the Atlantic portion of the ocean, relative to other regions. This is indicated by the larger number of species with IUCN status in this region.  However, the Atlantic falls in line with the other regions when looking at the comprehensively assessed species.  

```{r, eval=FALSE}
#devtools::install_github("dill/beyonce")
library(beyonce)
cols <- beyonce_palette(129, 100, type = "continuous")

# COMP ONLY
#2025
n_comp <- terra::rast(here("globalprep/spp/v2025/_output/n_spp_risk_raster_comp.tif"))
log_n_comp <- log(n_comp)
plot(log_n_comp, col=cols)

#2024
n_comp <- terra::rast(here("globalprep/spp/v2024/_output/n_spp_risk_raster_comp.tif"))
log_n_comp <- log(n_comp)
plot(log_n_comp, col=cols)

#2022
n_comp <- terra::rast(here("globalprep/spp/v2022/_output/n_spp_risk_raster_comp.tif"))
log_n_comp <- log(n_comp)
plot(log_n_comp, col=cols)

#2021
n_comp <- terra::rast(here("globalprep/spp/v2021/_output/n_spp_risk_raster_comp.tif"))
log_n_comp <- log(n_comp)
plot(log_n_comp, col=cols)


# ALL
#2025
n_all <- terra::rast(here("globalprep/spp/v2025/_output/n_spp_risk_raster_all.tif"))
log_n_all <- log(n_all)
plot(log_n_all, col=cols)

#2024
n_all <- terra::rast(here("globalprep/spp/v2024/_output/n_spp_risk_raster_all.tif"))
log_n_all <- log(n_all)
plot(log_n_all, col=cols)

#2022
n_all <- terra::rast(here("globalprep/spp/v2022/_output/n_spp_risk_raster_all.tif"))
log_n_all <- log(n_all)
plot(log_n_all, col=cols)

#2021
n_all <- terra::rast(here("globalprep/spp/v2021/_output/n_spp_risk_raster_all.tif"))
log_n_all <- log(n_all)
plot(log_n_all, col=cols)


prop_comp <- n_comp/n_all
plot(prop_comp, col=rev(cols))
#this makes sense. For some reason Damselfish and a few reptiles were classified as "non comprehensive", but they are actually comprehensive. 
```

## SPP: Status

### Status: Extract average species risk for each region

For each cell, we multiply the average species risk by the number of species in order to weight each cells contribution by the number of species.  We sum these values for each region and calculate:

(average species risk * number species)/number of species

```{r, eval = FALSE}

mean_risk_comp <- terra::rast(here("globalprep/spp/", scenario, "/_output/mean_risk_raster_comp.tif"))

n_comp <- terra::rast(here("globalprep/spp/", scenario, "/_output/n_spp_risk_raster_comp.tif"))

regions_ohi <- terra::rast(here("globalprep/spp/", scenario, "/_spatial/eez_rast.tif"))

risk_stack_comp <- c(regions_ohi, mean_risk_comp, n_comp)
plot(risk_stack_comp)

risk_vals_comp <- terra::values(risk_stack_comp) %>%
  data.frame()
risk_vals_comp <- filter(risk_vals_comp, !is.na(eez_rast))
risk_vals_comp <- filter(risk_vals_comp, !is.na(mean_risk_raster_comp))

rgn_risk_comp <- risk_vals_comp %>%
  rowwise() %>%
  dplyr::mutate(risk_weight = mean_risk_raster_comp * n_spp_risk_raster_comp) %>%
  group_by(eez_rast) %>%
  summarize(rgn_risk_weight = sum(risk_weight),
            rgn_n_species = sum(n_spp_risk_raster_comp))


## This is the first time this object comes up in this script and its calling on itself ?? (found the above code from the spp_data_prep.Rmd in the v2021 root directory)
rgn_risk_comp <- rgn_risk_comp %>%
  dplyr::rowwise() %>%
  dplyr::mutate(mean_risk = rgn_risk_weight/rgn_n_species) %>%
  dplyr::select(rgn_id = eez_rast, mean_risk)


```

### Status: estimate for previous years

We use the trend data to estimate risk values for previous years (vs. using the same values for all assessment years).  The change in species status across years is based on a linear model.

Trend is calculated using the same method as the risk calculation. For each cell, we multiply the average species trend by the number of species in order to weight each cell's contribution by the number of species in the cell.  We sum these values for each OHI region and calculate for each region:

(average species trend * number species)/number of species

```{r, eval = FALSE}

trend_comp <- terra::rast(here("globalprep/spp/", scenario, "/_output/trend_raster_comp.tif"))

n_trend_comp <- terra::rast(here("globalprep/spp/", scenario, "/_output/n_trend_raster_comp.tif"))

regions_ohi <- terra::rast(here("globalprep/spp/", scenario, "/_spatial/eez_rast.tif"))

trend_stack_comp <- c(regions_ohi, trend_comp, n_trend_comp)
plot(trend_stack_comp)

trend_vals_comp <- terra::values(trend_stack_comp) %>%
  data.frame()

trend_vals_comp <- filter(trend_vals_comp, !is.na(eez_rast))
trend_vals_comp <- filter(trend_vals_comp, !is.na(trend_raster_comp))

rgn_trend_comp <- trend_vals_comp %>%
  rowwise() %>%
  dplyr::mutate(trend_weight = trend_raster_comp * n_trend_raster_comp) %>%
  group_by(eez_rast) %>%
  summarize(rgn_trend_weight = sum(trend_weight),
            rgn_n_species = sum(n_trend_raster_comp)) %>%
  rename(rgn_id = eez_rast) %>%
  ungroup()

rgn_trend_comp <- rgn_trend_comp %>%
  dplyr::rowwise() %>%
  dplyr::mutate(mean_trend = rgn_trend_weight/rgn_n_species) %>%
  dplyr::select(rgn_id, mean_trend)

```

### Status: Get yearly risk scores based on trend

We estimate previous risk for each region, using the trend data.  We assume change in risk is linear.  

```{r, eval = FALSE}
assess_years <- 2012:2024 # should be current year or one year behind?
years <- expand.grid(rgn_id = unique(rgn_risk_comp$rgn_id), year=assess_years)

# this is what the trend will be multiplied by to get a risk estimate for each year:
year_multiplier <- data.frame(year=assess_years, multiplier = rev(0:(length(assess_years)-1))) 

rgn_risk_comp_yrs <- rgn_risk_comp %>%
  dplyr::left_join(rgn_trend_comp, by = "rgn_id") %>%
  dplyr::left_join(years, by = "rgn_id") %>%
  dplyr::left_join(year_multiplier, by="year") %>%
  dplyr::rowwise() %>%
  dplyr::mutate(mean_risk_per_year = mean_risk - mean_trend*multiplier) %>%
  dplyr::select(rgn_id, year, mean_risk = mean_risk_per_year)

```

### Status: Converting regional mean risk to status

We rescale the data so a risk factor of 0.75 is equal to zero.

```{r, eval = FALSE}
rgn_status <- rgn_risk_comp_yrs %>%
  mutate(spp_status = (0.75 - mean_risk)/0.75)
```

### Status: Gapfill missing regions

Region 232 (Bosnia) does not have a value, which is not surprising because their coast is super small and results are erratic for this region. We gapfill with surrounding regions.

```{r, eval = FALSE}

status_gf <- rgns_global %>%
  left_join(rgn_status) %>%
  dplyr::select(-mean_risk)

summary(status_gf)
filter(status_gf, is.na(spp_status))

croatia <- filter(status_gf, rgn_id == 187)
mont <- filter(status_gf, rgn_id == 186) 

bosnia <- bind_rows(croatia, mont) %>%
  group_by(year) %>%
  summarize(spp_status = mean(spp_status)) %>%
  mutate(rgn_id = 232)

status_gf <- status_gf %>%
  filter(rgn_id !=232) %>%
  bind_rows(bosnia)

```

### Status: Final formatting for ohi-global

`sp_status_global.csv`

```{r, eval = FALSE}
status <- status_gf %>%
  dplyr::select(rgn_id, year, score = spp_status)

dim(status)   #220*length(assess_years)
summary(status) # should be no NA values

sum(is.na(status))

write.csv(status, here("globalprep/spp/", scenario, "/output/sp_status_global.csv"), row.names=FALSE)  

```

### Status: Compare to last year

```{r, eval = FALSE}
status <- read_csv(here("globalprep/spp/", scenario, "/output/sp_status_global.csv"))

old_spp <- read.csv(here("globalprep/spp/v2024/output/sp_status_global.csv")) %>%
  filter(year == max(year)) %>%
  rename(old_score = score) %>%
  mutate(year = 2024) %>%
  left_join(status) %>%
  rename(new_score = score)

plot(old_spp$old_score, old_spp$new_score, xlab = "2024 Spp Status", ylab = "2025 Spp Status")
abline(0,1, col="red")

old_spp_gather <- old_spp %>%
  dplyr::select(rgn_id, old_score, new_score) %>%
  tidyr::gather("assessment", "score", -1) %>%
  dplyr::filter(rgn_id <= 250)

ggplot(old_spp_gather, aes(y=assessment, x=score)) + 
  geom_density_ridges()

ggplot(old_spp, aes(x = old_score, y = new_score)) + 
  geom_point() + 
  geom_abline(intercept = 0)

```

## SPP: Trend

### Trend: calculating

Getting proportional trend requires the status data (trend/status). Proportional trend is multiplied by 5 to get estimated change in five years.

```{r, eval = FALSE}
# proportional trend requires status data
status <- read.csv(here("globalprep/spp/", scenario, "/output/sp_status_global.csv")) %>%
  dplyr::filter(year==max(year)) %>%
  dplyr::select(rgn_id, score)

# Calculated in above section: Trend data
rgn_trend_score <- rgn_trend_comp %>%
  mutate(spp_trend_adj = -mean_trend/0.75) %>%  # puts in comparable units to status
  left_join(status, by="rgn_id") %>%
  dplyr::mutate(trend_score = spp_trend_adj/score * 5)

```

### Trend: Gapfilling missing data

Check there are data for every region.  Region 232 (Bosnia) does not have a value which is not surprising because their coast is super small and results are erratic for this region.  We estimate this using the mean of the 2 surrounding regions.  

```{r, eval = FALSE}

trend <- rgns_global %>%
  left_join(rgn_trend_score) 

summary(trend)

filter(trend, is.na(trend_score))

croatia <- filter(trend, rgn_id == 187)
mont <- filter(trend, rgn_id == 186)
bosnia <- mean(c(croatia$trend_score, mont$trend_score))

trend$trend_score[trend$rgn_id == 232] <- bosnia 
```

## Trend: Final formatting for ohi-global

`sp_trend_global.csv`

```{r, eval = FALSE}
trend <- trend %>%
  dplyr::select(rgn_id, score = trend_score)
dim(trend) # should be 220
summary(trend) # should be no NAs

write.csv(trend, here("globalprep/spp/", scenario, "/output/sp_trend_global.csv"), row.names=FALSE)  
```

### Trend: Compare to last year

Previously we used proxy data rather than actual change in IUCN status over time.  Our previous method overestimated the magnitude of the trend.  It is not surprising there is poor correlation with trend estimates in previous years (before 2019), but it is reassuring the the values mainly fall in the same quadrant, and trends are largely the same now (2019 - present)

```{r, eval = FALSE}

trend <- read.csv(here("globalprep/spp/", scenario, "/output/sp_trend_global.csv"))  

trend <- read.csv(here("globalprep/spp/v2024/output/sp_trend_global.csv"))

old_spp <- read.csv(here("globalprep/spp/v2024/output/sp_trend_global.csv")) %>%
  rename(old_score = score) %>%
  left_join(trend) 

plot(old_spp$old_score, old_spp$score)# xlim=c(-0.35, 0.05))
abline(h=0, col="red")
abline(v=0, col="red")
abline(0,1, col="red")

ggplot(old_spp, aes(x = old_score, y = score)) +
  geom_point() + 
  geom_abline(intercept = 0) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  labs(x = "2024 Trend Assessment",
       y = "2025 Trend Assessment")
```

### Status/trend: save a record of gapfilling

```{r, eval = FALSE}
status <- read.csv(here("globalprep/spp/", scenario, "/output/sp_status_global.csv")) %>%
  mutate(gapfilled = ifelse(rgn_id == 232, 1, 0)) %>%
  mutate(method = ifelse(rgn_id == 232, "mean of neighbors", NA)) %>%
  dplyr::select(rgn_id, year, gapfilled, method)

write.csv(status, here("globalprep/spp/", scenario, "/output/sp_status_global_gf.csv"), row.names=FALSE)          

trend <- read.csv(here("globalprep/spp/", scenario, "/output/sp_trend_global.csv")) %>%
  mutate(gapfilled = ifelse(rgn_id == 232, 1, 0)) %>%
  mutate(method = ifelse(rgn_id == 232, "mean of neighbors", NA)) %>%
  dplyr::select(rgn_id, gapfilled, method)

write.csv(trend, here("globalprep/spp/", scenario, "/output/sp_trend_global_gf.csv"), row.names=FALSE)          
```

## Resilience data

We use species condition data as a resilience measure as well.  We also calculate species condition at 3nm of shoreline, because for some goals, nearshore species condition is the relevant metric.   

### Resilience: Prepare rasters for 3nm extraction
We reproject the data to have higher resolution in order to more easily extract the data at the 3nm scale.  

We modify the method a bit from above due to size of the rasters.  

```{r, eval=FALSE}
#3nm raster file
rgns <- raster(file.path(dir_M, "git-annex/globalprep/spatial/v2018/rgns_3nm_offshore_mol.tif"))
plot(rgns)

# relevant species files
mean_risk_comp <- raster(here("globalprep/spp/", scenario, "/_output/mean_risk_raster_comp.tif"))
plot(mean_risk_comp)

n_comp <- raster(here("globalprep/spp/", scenario, "/_output/n_spp_risk_raster_comp.tif"))
plot(n_comp)


risk_x_n <- mean_risk_comp*n_comp
plot(risk_x_n)


# project rasters to moll
# saved in Mazu:spp/v2022
raster::projectRaster(risk_x_n, rgns, method="ngb", overwrite=TRUE, 
              filename=file.path(dir_M, "git-annex/globalprep/spp", scenario, "int/risk_x_n_comp_mol.tif"), progress = "text")

raster::projectRaster(n_comp, rgns, method="ngb", overwrite=TRUE, 
              filename=file.path(dir_M, "git-annex/globalprep/spp", scenario, "int/n_comp_mol.tif"), progress = "text")

```

Compare

```{r}
risk_x_n_new <- rast(file.path(dir_M, "git-annex/globalprep/spp", scenario, "int/risk_x_n_comp_mol.tif"))
plot(risk_x_n_new)

# Load old raster
risk_x_n_old <- rast(file.path(dir_M, "git-annex/globalprep/spp", "v2024", "int/risk_x_n_comp_mol.tif"))
# Check
plot(risk_x_n_old)

n_comp_new <- rast(file.path(dir_M, "git-annex/globalprep/spp", scenario, "int/n_comp_mol.tif"))
plot(n_comp_new)

n_comp_old <- rast(file.path(dir_M, "git-annex/globalprep/spp", "v2024", "int/n_comp_mol.tif"))
plot(n_comp_old)
```

### Resilience: Extract data

Extract species risk data that corresponds to 3nm regions.

```{r, eval = FALSE}

#3nm raster file
rgns <- rast(file.path(dir_M, "git-annex/globalprep/spatial/v2018/rgns_3nm_offshore_mol.tif"))
# plot(rgns)

risk_x_n_mol <- rast(file.path(dir_M, "git-annex/globalprep/spp", scenario, "int/risk_x_n_comp_mol.tif"))
# plot(risk_x_n_mol)
n_mol <- rast(file.path(dir_M, "git-annex/globalprep/spp", scenario, "int/n_comp_mol.tif"))
# plot(n_mol)

risk_stack <- c(rgns, risk_x_n_mol, n_mol)

risk_df <- terra::values(risk_stack)
 
risk_df_2 <- as.data.frame(risk_df) %>%
  filter(!is.na(rgns_3nm_offshore_mol)) %>%
  filter(!is.na(risk_x_n_comp_mol))


rgn_3nm_risk <- risk_df_2 %>%
  rowwise() %>%
  group_by(rgns_3nm_offshore_mol) %>%
  summarize(risk_x_n_comp_mol = sum(risk_x_n_comp_mol),
            n_comp_mol = sum(n_comp_mol)) %>%
  ungroup() %>%
  rename(zone = 1) %>%
  dplyr::mutate(rgn_wt_risk = risk_x_n_comp_mol/n_comp_mol) %>%
  dplyr::select(rgn_id = zone, rgn_wt_risk)
  

```

### Resilience: estimate for previous years

We use the trend data to estimate risk values for previous years (vs. using the same values for all assessment years).  The change in species status across years is based on a linear model.

Trend is calculated using the same method as the risk calculation. For each cell, we multiply the average species trend by the number of species in order to weight each cell's contribution by the number of species in the cell.  We sum these values for each OHI region and calculate for each region:

(average species trend * number species)/number of species

#### Resilience: Prepare rasters for 3nm extraction (trend to estimate previous years' data)

We reproject the data to have higher resolution in order to more easily extract the data at the 3nm scale.  

```{r, eval=FALSE}
rgns <- raster(file.path(dir_M, "git-annex/globalprep/spatial/v2018/rgns_3nm_offshore_mol.tif"))

trend_comp <- raster::raster(here("globalprep/spp/", scenario, "/_output/trend_raster_comp.tif"))

n_trend_comp <- raster::raster(here("globalprep/spp/", scenario, "/_output/n_trend_raster_comp.tif"))

trend_x_n <- trend_comp*n_trend_comp

# project rasters to moll
# saved in Mazu:spp/v2022
projectRaster(trend_x_n, rgns, method="ngb", overwrite=TRUE, 
              filename=file.path(dir_M, "git-annex/globalprep/spp", scenario, "int/trend_x_n_comp_mol.tif"),
              progress="text")

projectRaster(n_trend_comp, rgns, method="ngb", over=TRUE, 
              filename=file.path(dir_M, "git-annex/globalprep/spp", scenario, "int/n_trend_comp_mol.tif"),
              progress="text")

```

```{r}
trend_x_n_new <- rast(file.path(dir_M, "git-annex/globalprep/spp", scenario, "int/trend_x_n_comp_mol.tif"))
plot(trend_x_n_new)

# Load old raster
trend_x_n_old <- rast(file.path(dir_M, "git-annex/globalprep/spp", "v2024", "int/trend_x_n_comp_mol.tif"))
plot(trend_x_n_old)

n_trend_comp_new <- rast(file.path(dir_M, "git-annex/globalprep/spp", scenario, "int/n_trend_comp_mol.tif"))
plot(n_trend_comp_new)

n_trend_comp_old <- rast(file.path(dir_M, "git-annex/globalprep/spp", "v2024", "int/n_trend_comp_mol.tif"))
plot(n_trend_comp_old)
```

### Resilience: Extract trend data (used to estimate previous years' data)

Extract species risk data that corresponds to 3nm regions.

```{r, eval = FALSE}
#3nm raster file
rgns_3nm <- rast(file.path(dir_M, "git-annex/globalprep/spatial/v2018/rgns_3nm_offshore_mol.tif"))
plot(rgns_3nm)

trend_3nm_mol <- rast(file.path(dir_M, "git-annex/globalprep/spp", scenario, "int/trend_x_n_comp_mol.tif"))

n_trend_3nm_mol <- rast(file.path(dir_M, "git-annex/globalprep/spp", scenario, "int/n_trend_comp_mol.tif"))


trend_stack <- c(rgns_3nm, trend_3nm_mol, n_trend_3nm_mol)

# trend_df <- terra::zonal(trend_stack, rgns_3nm, fun='sum')

trend_df <- terra::values(trend_stack)
 
trend_df_2 <- as.data.frame(trend_df) %>%
  filter(!is.na(rgns_3nm_offshore_mol)) %>%
  filter(!is.na(trend_x_n_comp_mol))


rgn_3nm_trend <- trend_df_2 %>%
  data.frame() %>%
  rowwise() %>%
  group_by(rgns_3nm_offshore_mol) %>%
  summarize(trend_x_n_comp_mol = sum(trend_x_n_comp_mol),
            n_trend_comp_mol = sum(n_trend_comp_mol)) %>%
  ungroup() %>%
  rename(zone = 1) %>%
  dplyr::mutate(rgn_wt_trend = trend_x_n_comp_mol/n_trend_comp_mol) %>%
  dplyr::select(rgn_id = zone, rgn_wt_trend)

```
  
### Resilience: Calculate yearly risk scores based on trend

We estimate previous risk for each region, using the trend data.  We assume change in risk is linear.  

```{r, eval = FALSE}
assess_years <- 2012:2024
years <- expand.grid(rgn_id = unique(rgn_3nm_trend$rgn_id), year=assess_years)

# this is what the trend will be multiplied by to get a risk estimate for each year:
year_multiplier <- data.frame(year=assess_years, multiplier = rev(0:(length(assess_years)-1))) 

rgn_risk_3nm <- rgn_3nm_risk %>%
  left_join(rgn_3nm_trend, by = "rgn_id") %>%
  left_join(years, by = "rgn_id") %>%
  left_join(year_multiplier, by="year") %>%
  rowwise() %>%
  mutate(mean_risk_per_year = rgn_wt_risk - rgn_wt_trend*multiplier) %>%
  dplyr::select(rgn_id, year, mean_risk = mean_risk_per_year)

```

### Resilience: Converting regional mean risk to status

We rescale the data so a risk factor of 0.75 is equal to zero.

```{r, eval = FALSE}

rgn_3nm_res <- rgn_risk_3nm %>%
  mutate(spp_status = (0.75 - mean_risk)/0.75)

# quick check
hist(rgn_3nm_res$spp_status)

```

### Resilience: Gapfill missing regions

Region 19 (Tuvalu) does not have a value.  This is an island.  We gapfill with the value from the entire eez.

```{r, eval = FALSE}

res_gf <- rgns_global %>%
  left_join(rgn_3nm_res) %>%
    dplyr::select(-mean_risk)

summary(res_gf)
filter(res_gf, is.na(spp_status))

# get eez value:
eez_status <- read.csv(here("globalprep/spp/v2024/output/sp_status_global.csv")) %>%
  filter(rgn_id == 19) %>%
  rename(spp_status = score)

res_gf <- res_gf %>%
  filter(!is.na(spp_status)) %>%
  bind_rows(eez_status)


summary(res_gf)
```

### Resilience: Final formatting for ohi-global

`sp_status_3nm.csv`

```{r, eval = FALSE}
resilience <- res_gf %>%
  dplyr::select(rgn_id, year, score = spp_status)

summary(resilience) # should be no NA values

write.csv(resilience, here("globalprep/spp", scenario, "output/sp_status_3nm.csv"), row.names=FALSE)  

```

### Resilience: save a record of gapfilling

```{r, eval = FALSE}
res <- read.csv(here("globalprep/spp", scenario, "output/sp_status_3nm.csv")) %>%
  mutate(gapfilled = ifelse(rgn_id == 19, 1, 0)) %>%
  mutate(method = ifelse(rgn_id == 19, "eez scale data used", NA)) %>%
  dplyr::select(rgn_id, year, gapfilled, method)

write.csv(res, here("globalprep/spp", scenario, "output/sp_status_3nm_gf.csv"), row.names=FALSE)          
```

### Resilience: Compare

Compared to the entire EEZ, most (but not all) coastal areas have higher species condition scores. The correlation with last year is better than it was last year. 

```{r, eval = FALSE}

# compare to eez values
eez_status <- read.csv(here("globalprep/spp", scenario, "output/sp_status_global.csv")) %>%
  filter(year == max(year)) %>%
  dplyr::select(rgn_id, eez_score = score)

status <- read.csv(here("globalprep/spp", scenario, "output/sp_status_3nm.csv")) %>%
  dplyr::select(rgn_id, year, nm3_score = score) %>%
  filter(year == max(year)) %>%
  left_join(eez_status, by = "rgn_id")

plot(status$nm3_score, status$eez_score)
abline(0,1)

ggplot(status, aes(x = nm3_score, y = eez_score)) + 
  geom_point() + 
  geom_abline(intercept = 0) +
  labs(x = "3nm score", y = "eez score")

# compare to last year's values
status <- read.csv(here("globalprep/spp", scenario, "output/sp_status_3nm.csv")) %>%
  filter(year == max(year)) %>%
  dplyr::select(-year)

old_spp <- read.csv(here("globalprep/spp/v2024/output/sp_status_3nm.csv")) %>%
  filter(year == max(year)) %>%
  rename(old_score = score) %>%
  # mutate(year = 2020) %>%
  dplyr::select(-year) %>%
  left_join(status) %>%
  rename(new_score = score) %>%
  mutate(difference = new_score - old_score) %>%
  left_join(rgns_eez)

plot(old_spp$old_score, old_spp$new_score)
abline(0,1, col="red")

ggplot(old_spp, aes(x = old_score, y = new_score)) + 
  geom_point() + 
  geom_abline(intercept = 0) +
  labs(x = "v2024 resilience", y = "v2025 resilience")

old_spp_gather <- old_spp %>%
  dplyr::select(rgn_id, old_score, new_score) %>%
  tidyr::gather("assessment", "score", -1) %>%
  filter(rgn_id <= 250)

ggplot(old_spp_gather, aes(y=assessment, x=score)) + 
  geom_density_ridges()

check_3nm_old <- old_spp %>%
  filter(new_score > 0.85, old_score > 0.79, old_score <0.87) %>%
  mutate(diff = old_score - new_score)

```

