---
title: "OHI 2022: Exclude non-surface water areas"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---

## Summary

Multiply our leaching and volatized rasters by the appropriate surface water rasters to get an estimate of the amount of N that will reach the coast. See below for more instructions. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here);library(janitor);library("future");library(furrr);library(vroom);library(raster); library(tidyverse);library(readr); library(fasterize); library(sf); library(terra)


git_manure <- file.path(here(), "globalprep/prs_land-based_nutrient/v2022/STEP2_manure")
location_tifs_path <- file.path("/home/shares/ohi/git-annex/globalprep/_raw_data/halpern_food_systems/d2021/animal_farm/location_tifs") # doesn't change
prep <- file.path("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2022")
git_crop <- file.path(here(), "globalprep/prs_land-based_nutrient/v2022/STEP1_crop")
git_compile <- file.path(here(), "globalprep/prs_land-based_nutrient/v2022/STEP3_compile_prep_layers")

source(here(git_crop, "R/common_spatial.R")) # read in necessary spatial files and functions. 

```

## Methods 

Loop that will multiply the yearly leaching raster by the surface water raster and save. Multiply the volatilization rasters by a raster proportion of cell covered by surface water + 1km from coast. Additionally save the total cellStats sum for each of the multiplied rasters so that we can compare to the total nutrient layers. 

Do this for: 

 - Crop Leaching N * 1km coast + 1km surface water
 - Manure Leaching N * 1km coast + 1km surface water
 - Manure volatilization N * 1km coast + prop rivers 
 - Crop volatilitzation N * 1km coast + prop rivers 

We could also do this for P, but we are excluding P for this analysis. 

```{r}

## load in the 1km surface and coast raster
sw_reproj <- rast(file.path("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2021/surface_water_reproj.tif"))

## load in 1km coast + prop river 
prop_river <- rast(file.path("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2021/", "river_coastline.tif"))

years <- as.character(c(2005:2020))
nutrients <- c("N")
types <- c("leached", "volt")
types <- "volt"

sum_df <- data.frame(nutrient = NA, year = NA, process_type = NA, category_type = NA, before_exclusion_sum = NA, after_exclusion_sum = NA)

registerDoParallel(cores = 6)

foreach(yr = years) %:% 
  foreach(nut = nutrients) %:% 
    foreach(type = types) %dopar% {

# yr <- 2005
# nut <- "N"
# type <- "leached"

      if(type == "leached"){
crop_raster <- rast(file.path(paste0(prep, "/crop_stressors_yearly/", "crop_", type, "_nutrient_", nut, "_", yr, ".tif")))
manure_raster <- rast(file.path(paste0(prep, "/manure_stressors_yearly/", "manure_", type, "_nutrient_", nut, "_", yr, ".tif")))

combined_rast <- crop_raster + manure_raster

before_surface_combined <- global(combined_rast, "sum", na.rm = TRUE)

combined_surface <- combined_rast * sw_reproj

after_surface_combined <- global(combined_surface, "sum", na.rm = TRUE)

for_df_combined <- data.frame(nutrient = nut, year = yr, process_type = type, category_type = "crop_manure", before_exclusion_sum = before_surface_combined$sum, after_exclusion_sum = after_surface_combined$sum)


writeRaster(combined_surface, file.path(paste0(prep, "/exclude_surface_water/", "surface_water_1km_crop_manure_", type, "_nutrient_", nut, "_", yr, ".tif" )), overwrite=TRUE)



before_surface_crop <- global(crop_raster, "sum", na.rm = TRUE)

crop_surface <- crop_raster * sw_reproj

after_surface_crop <- global(crop_surface, "sum", na.rm = TRUE)

for_df_crop <- data.frame(nutrient = nut, year = yr, process_type = type, category_type = "crop", before_exclusion_sum = before_surface_crop$sum, after_exclusion_sum = after_surface_crop$sum)


writeRaster(crop_surface, file.path(paste0(prep, "/exclude_surface_water/", "surface_water_1km_crop_", type, "_nutrient_", nut, "_", yr, ".tif" )), overwrite=TRUE)



before_surface_manure <- global(manure_raster, "sum", na.rm = TRUE)

manure_surface <- manure_raster * sw_reproj

after_surface_manure <- global(manure_surface, "sum", na.rm = TRUE)

for_df_manure <- data.frame(nutrient = nut, year = yr, process_type = type, category_type = "manure", before_exclusion_sum = before_surface_manure$sum, after_exclusion_sum = after_surface_manure$sum)


writeRaster(manure_surface, file.path(paste0(prep, "/exclude_surface_water/", "surface_water_1km_manure_", type, "_nutrient_", nut, "_", yr, ".tif" )), overwrite=TRUE)



sum_df <- rbind(sum_df, for_df_combined, for_df_crop, for_df_manure) %>%
  dplyr::filter(!is.na(nutrient))

write.csv(sum_df, file.path(paste0(git_compile, "/tmp/", "sum_df_", type, "_", nut, "_", yr, ".csv")), row.names = FALSE)

      }else{
        
crop_raster <- rast(file.path(paste0(prep, "/crop_stressors_yearly/", "crop_", type, "_nutrient_", nut, "_", yr, ".tif")))
manure_raster <- rast(file.path(paste0(prep, "/manure_stressors_yearly/", "manure_", type, "_nutrient_", nut, "_", yr, ".tif")))

combined_rast <- crop_raster + manure_raster

before_surface_combined <- global(combined_rast, "sum", na.rm = TRUE)

combined_surface <- combined_rast * prop_river

after_surface_combined <- global(combined_surface, "sum", na.rm = TRUE)

for_df_combined <- data.frame(nutrient = nut, year = yr, process_type = type, category_type = "crop_manure", before_exclusion_sum = before_surface_combined$sum, after_exclusion_sum = after_surface_combined$sum)


writeRaster(combined_surface, file.path(paste0(prep, "/exclude_surface_water/", "surface_water_1km_crop_manure_", type, "_nutrient_", nut, "_", yr, ".tif" )), overwrite=TRUE)



before_surface_crop <- global(crop_raster, "sum", na.rm = TRUE)

crop_surface <- crop_raster * prop_river

after_surface_crop <- global(crop_surface, "sum", na.rm = TRUE)

for_df_crop <- data.frame(nutrient = nut, year = yr, process_type = type, category_type = "crop", before_exclusion_sum = before_surface_crop$sum, after_exclusion_sum = after_surface_crop$sum)


writeRaster(crop_surface, file.path(paste0(prep, "/exclude_surface_water/", "surface_water_1km_crop_", type, "_nutrient_", nut, "_", yr, ".tif" )), overwrite=TRUE)



before_surface_manure <- global(manure_raster, "sum", na.rm = TRUE)

manure_surface <- manure_raster * prop_river

after_surface_manure <- global(manure_surface, "sum", na.rm = TRUE)

for_df_manure <- data.frame(nutrient = nut, year = yr, process_type = type, category_type = "manure", before_exclusion_sum = before_surface_manure$sum, after_exclusion_sum = after_surface_manure$sum)


writeRaster(manure_surface, file.path(paste0(prep, "/exclude_surface_water/", "surface_water_1km_manure_", type, "_nutrient_", nut, "_", yr, ".tif" )), overwrite=TRUE)



sum_df <- rbind(sum_df, for_df_combined, for_df_crop, for_df_manure) %>%
  dplyr::filter(!is.na(nutrient))

write.csv(sum_df, file.path(paste0(git_compile, "/tmp/", "sum_df_", type, "_", nut, "_", yr, ".csv")), row.names = FALSE)
        
}

}

files <- list.files(path = file.path(git_compile, "tmp"), full = TRUE)
sum_df_all <- 
  do.call(rbind,
          lapply(files, read.csv))

write.csv(sum_df_all, file.path(git_compile, "int/all_nutrient_layers_sw_sums.csv"), row.names = FALSE)

```

3. Join the surface water nutrient layer sums with the total nutrient sums, and compare the percentages of each. Just for data checking funsies. 

```{r}
surface_water_sums <- read.csv(file.path(git_compile, "int/all_nutrient_layers_sw_sums.csv"))

total_nutrient_sums_crop <- read.csv(file.path(git_compile, "int/crop_nutrient_by_year.csv")) %>%
  mutate(type = "crop")

total_nutrient_sums_manure <- read.csv(file.path(git_compile, "int/manure_nutrient_by_year.csv"))

total_nutrient_sums <- rbind(total_nutrient_sums_crop, total_nutrient_sums_manure) %>% 
  group_by(year, nutrient) %>%
  summarise(nutrient_sum = sum(nutrient_sum)) %>%
  mutate(type = "crop_manure") %>%
  ungroup() %>%
  rbind(total_nutrient_sums_crop, total_nutrient_sums_manure) 


all_dfs <- surface_water_sums %>%
  left_join(total_nutrient_sums, by = c("category_type" = "type", "year", "nutrient")) %>%
  rename("total_nutrient_applied" = "nutrient_sum") %>%
  mutate(percent_excluded_before = before_exclusion_sum/total_nutrient_applied) %>%
  mutate(percent_excluded_after = after_exclusion_sum/total_nutrient_applied)

write.csv(all_dfs, file.path(git_compile, "int/all_layers_sw_percents.csv"), row.names = FALSE)


test <- read.csv(file.path(git_compile, "int/all_layers_sw_percents.csv")) #### NOTE: show Mel this ####

test2 <- test %>%
  group_by(nutrient, process_type, category_type) %>%
  summarise(mean_pre_sw = mean(percent_excluded_before)*100,
            mean_post_sw = mean(percent_excluded_after)*100)

```




