---
title: 'OHI 2022: Natural Products: weighting data prep'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

[REFERENCE RMD FILE: http://ohi-science.org/ohiprep_v2021/globalprep/np/v2021/STEP2_np_weighting_prep.html]

# Summary
This analysis converts FAO commodities tonnes data for ornamental fish trade, SAUP fisheries tonnes data for FOFM, and FAO mariculture tonnes data for seaweeds into a weighting scheme per each ohi region, based off of 5 year averages of tonnes and values (in usd) for each product. This weighting scheme will be applied to the scores to determine how much of each natural products layer affects the overall NP score when updating in ohi global. 


***

# Data Source

**Reference**:      
    http://www.fao.org/fishery/statistics/software/fishstatj/en#downlApp
    App release date: July 2021 
FAO raw commodities quantity 1976_2019
FAO raw commodities value 1976_2019
[FAO metadata](http://www.fao.org/fishery/statistics/global-commodities-production/en)

**Downloaded**: June, 23, 2022

**Description**:  Quantity (tonnes) and value (USD) of raw commodities (Exports only) for each country, taxa, year.  The FAO data is subset to include commodities in these categories: ornamental fish, fish oil, seaweed and plants (see: raw/commodities2products_weighting.csv for details).

**Time range**: 1976-2019


**Reference**: Pauly D., Zeller D., Palomares M.L.D. (Editors), 2020. Sea Around Us Concepts, Design and Data (seaaroundus.org).

**Downloaded**: September 27, 2022

**Description**: Tons per year and SAUP region with information on sector type, industry type, fishing entitity, reporting status and taxonomic information. 

**Time range**: 1950 - 2019

**Format**: CSV

**Additional Information**: [Methods](http://www.seaaroundus.org/sea-around-us-methods-index/)


**Reference**: Froehlich, H.E., Jacobsen, N.S., Essington, T.E., Clavelle, T., and Halpern, B.S. (2018). Avoiding the ecological limits of forage fish for fed aquaculture. Nature Sustainability 1, 298.

**Downloaded**: July 7, 2020. Obtained from Melanie Frazier (NCEAS). 

**Description**:  List of FOFM species from Watson v3 data. 

**Native data resolution**:   

**Format**:  CSV format


**Reference**: [RAM Legacy Stock Assessment Database](http://ramlegacy.org) v4.491

**Downloaded**: 06/03/2021

**Description**: B/Bmsy value by stock and year (other data, which we do not use, are also available in the database)

**Native data resolution**: stock (fish stock, species and region specific)

**Time range**: 1950 - 2024

**Format**: CSV format 

**Additional Information**: We use the finalized b/bmsy layer from OHI-global for this data prep. We do not actually read in the raw RAM data here. We only use up to year == 2018.

* DOI: 10.5281/zenodo.2542919


**Reference**:      
http://www.fao.org/fishery/statistics/software/fishstatj/en#downlApp
FAO Global Aquaculture Production Quantity 1950_2020
FAO metadata found [here](http://www.fao.org/fishery/statistics/global-aquaculture-production/en)

**Downloaded**:  April 25, 2022

**Description**:  Quantity (tonnes) of mariculture for each country, species, year.

**Time range**: 1950-2020

***

## Setup

```{r setup, warning=FALSE, message=FALSE}

knitr::opts_chunk$set(eval=FALSE)

library(here)
library(tidyverse)
source(here('workflow/R/common.R'))
source(here('globalprep/np/v2022/R/np_fxn.R'))

prep <- file.path("globalprep/np/v2022/")

prep_year <- 2022

```
  
## Methods

Import prepped FAO commodities data set and find 5 year average $USD / Tonne average value per product per region. Gapfill these based on the UN geopolitical region/global values. 
```{r}
## read in the prepped FAO commodities data set 
harvest_tonnes_usd <- read_csv(file.path(prep, "int/np_harvest_tonnes_usd.csv")) %>%
  filter(rgn_id != 213)

max_year = max(harvest_tonnes_usd$year)
min_year = max_year - 4 
years = c(min_year:max_year)
years_df = data.frame(year = min_year:max_year) ## so that we can get 5 year average

region_data()

rgns_eez <- rgns_eez %>%
  dplyr::select(rgn_id) %>%
  filter(rgn_id != 213) # filter out antarctica

## make a data frame with every region and every product
products <- data.frame(product = c("seaweeds", "ornamentals", "fish_oil"))

region_product <- full_join(rgns_eez, products, by = character()) %>%
  full_join(years_df, by = character())

## calculate $/tonne for each ohi region and product and gapfill based on UN geopolitical region 

harvest_tonnes_usd_geo_rgn <- harvest_tonnes_usd %>%
    full_join(region_product, by = c("rgn_id", "product", "year")) %>%
  add_georegion_id() %>%
  dplyr::filter(year %in% years) %>%
  arrange(rgn_id)
  
  
geo_rgn_values_df <- harvest_tonnes_usd_geo_rgn %>%
  dplyr::group_by(georgn_id, year, product) %>%
  dplyr::mutate(values = usd/tonnes) %>%
    mutate(values = ifelse(values == "Inf", NA, values)) %>% ## assign Inf as NA
  summarise(georgn_values = mean(values, na.rm = TRUE)) %>% 
  ungroup() ## now we have a 5 year average value for geo region

global_values_df <- harvest_tonnes_usd_geo_rgn %>%
  dplyr::group_by(product, year) %>%
  mutate(values = usd/tonnes) %>%
  mutate(values = ifelse(values == "Inf", NA, values)) %>% ## assign Inf as NA
  summarise(global_values = mean(values, na.rm = TRUE)) ## now we have a 5 year average global value

harvest_tonnes_usd_values_gf <- harvest_tonnes_usd_geo_rgn %>%
  left_join(geo_rgn_values_df, by = c("georgn_id", "product", "year")) %>%
  left_join(global_values_df, by = c("product", "year")) %>%
  mutate(values = usd/tonnes) %>%
  mutate(values = ifelse(values == "Inf", NA, 
                         ifelse(values == "NaN", NA, values))) %>% ## assign Inf and NaNs as NA
  mutate(values_final = ifelse(is.na(values) & georgn_values != "NaN", georgn_values,
                               ifelse(is.na(values) & georgn_values == "NaN", global_values, values))) %>%## gapfill so that when there is no regional value, take the geo regional value, and if there is no geo regional value, take the global value. 
    mutate(values_gf_description = ifelse(is.na(values) & georgn_values != "NaN", "geo_region",
                               ifelse(is.na(values) & georgn_values == "NaN", "global", "none"))) %>%
  mutate(values_gf = ifelse(is.na(values) & georgn_values != "NaN", 1,
                               ifelse(is.na(values) & georgn_values == "NaN", 1, 0))) ## now we have a dataset with gapfilled average values and calculated yearly values... now we will save this, and then calculate the 5 year averages

## save this data frame to int
write.csv(harvest_tonnes_usd_values_gf, file.path(prep, "int/harvest_tonnes_usd_weighting_gf.csv"), row.names = FALSE)

## now calculate 5 year averages and assign year = 2020
weighting_usd_values_final <- harvest_tonnes_usd_values_gf %>%
  dplyr::select(rgn_id, product, year, values_final) %>%
  group_by(rgn_id, product) %>%
  summarise(value_per_tonne = mean(values_final)) %>%
  ungroup() 

#save this df
write.csv(weighting_usd_values_final, file.path(prep, "int/harvest_weighting_values.csv"), row.names = FALSE)

```

Now prep the tonnes data from each product to be per each region for the 6 year average (2013-2018). 
```{r}
## Read in FOFM tonnes data
fofm_tonnes <- read_csv(file.path(prep, "int/mean_catch_FOFM.csv")) %>%
  dplyr::filter(year %in% years) %>%
  mutate(product = "fish_oil") %>%
  group_by(rgn_id, year, product) %>%
  summarise(tonnes = sum(catch)*0.3) %>% ## Note: multiply by 0.3 to account for water loss when converting fish to fish oil.. about 30% of fish are water and 70% are fish oil
  ungroup() %>%
  group_by(rgn_id, product) %>%
  summarise(tonnes = mean(tonnes)) %>% ##calculate 5 year mean of catch
  ungroup() %>%
  full_join(rgns_eez) %>%
  dplyr::select(1:3) %>%
  mutate(product = "fish_oil") %>%
  mutate(tonnes = ifelse(is.na(tonnes), 0, tonnes)) 


## Read in ornamentals tonnes data
np_tonnes <- read_csv(file.path(prep, "int/np_harvest_tonnes_usd.csv")) %>% 
  filter(rgn_id != 213)

orn_fill_df <- region_product %>%
  filter(product == "ornamentals")
  
orn_tonnes <- np_tonnes %>%
  dplyr::filter(year %in% years, product == "ornamentals") %>%
  full_join(orn_fill_df, by = c("rgn_id", "year", "product")) %>% ## full join to fill in the rest of the regions with 0
  mutate(tonnes = ifelse(is.na(tonnes), 0, tonnes),
         usd = ifelse(is.na(usd), 0, usd)) %>% ## gapfill the NAs to be 0
  select(rgn_id, year, product, tonnes) %>%
  group_by(rgn_id, product) %>%
  summarise(tonnes = mean(tonnes)) %>% ## calculate 5 year mean of catch
  ungroup()
  
## Read in seaweed tonnes data
sw_fill_df <- region_product %>%
  filter(product == "seaweeds")

sw_tonnes <- read_csv(file.path(prep, "int/np_seaweeds_tonnes_weighting.csv")) %>%
  mutate(product = "seaweeds") %>%
  group_by(rgn_id, year, product) %>%
  summarise(tonnes = sum(tonnes, na.rm = TRUE)) %>%
  dplyr::filter(year %in% years) %>%
  full_join(sw_fill_df, by = c("rgn_id", "year", "product")) %>%
    mutate(tonnes = ifelse(is.na(tonnes), 0, tonnes)) %>% ## gapfill the NAs to be 0
  dplyr::select(rgn_id, year, product, tonnes) %>%
  ungroup() %>%
  group_by(rgn_id, product) %>%
  summarise(tonnes = mean(tonnes)) %>% ## calculate 5 year average
  ungroup()


```

Now we will calculate the weights per each product. To do this we need to multiply our average $ value for each product * tonnes of each product, and then divide by the total per each region. We will also assign year = 2022 so that this can be read into OHI-global (2022 corresponds to the year that these weights were calculated).

```{r}
harvest_weighting_values <- read_csv(file.path(prep, "int/harvest_weighting_values.csv")) %>%
  dplyr::select(rgn_id, product, value_per_tonne)

prod_weights <- orn_tonnes %>%
  bind_rows(sw_tonnes) %>%
  bind_rows(fofm_tonnes) %>% 
  left_join(harvest_weighting_values, by = c("rgn_id", "product")) %>%
  arrange(rgn_id) %>%
  mutate(usd_product = tonnes*value_per_tonne) %>%
  group_by(rgn_id) %>%
  mutate(total_usd = sum(usd_product)) %>%
  ungroup() %>%
  mutate(weight = usd_product/total_usd) %>%
  dplyr::select(-3, -4, -5, -6) %>%
  mutate(year = prep_year) %>%
  dplyr::select(rgn_id, year, product, weight) %>%
  filter(rgn_id != 213)

write.csv(prod_weights, file.path(prep, "output/np_product_weights.csv"), row.names = FALSE)
```

```{r}
##datacheck
old_prod_weights <- read_csv(file.path(prep, "../v2021/output/np_product_weights.csv"))

prod_weights <- read_csv(file.path(prep, "output/np_product_weights.csv"))

check <- prod_weights %>%
  rename("new_weight" = "weight") %>%
  left_join(old_prod_weights, by = c("rgn_id", "product")) %>%
  mutate(diff = new_weight - weight) %>%
  left_join(rgns_eez)

plot(check$new_weight, check$weight)
abline(0,1, col="red")

check_sw <- check %>%
  filter(product == "seaweeds")

plot(check_sw$new_weight, check_sw$weight)
abline(0,1, col="red")

check_orn <- check %>%
  filter(product == "ornamentals") %>%
  mutate(difference = new_weight - weight)

plot(check_orn$new_weight, check_orn$weight)
abline(0,1, col="red")

## check saint martin, Niue, Bonaire, Sint Eustasius - should decrease in production
## check djibouti - should increase in production

test <- harvest_tonnes_usd %>%
  filter(rgn_id == 46)

harvest_tonnes_usd_old <- read_csv(file.path(prep, "../v2021/int/np_harvest_tonnes_usd.csv")) %>%
  filter(rgn_id != 213)

test2 <- harvest_tonnes_usd_old %>%
  filter(rgn_id == 46)


gf_orn <- read_csv(file.path( prep, "output/np_ornamentals_harvest_tonnes_gf.csv"))

check_fofm <- check %>%
  filter(product == "fish_oil")

plot(check_fofm$new_weight, check_fofm$weight)
abline(0,1, col="red") ## these change because of new SAUP fisheries catch data... 

```

