---
title: "Learning about OHI scores 2022 key"
output: html_document
---

```{r setup, include=FALSE}
## loading packages
library(dplyr)
library(tidyr)
library(ggplot2)
```


## OH NO!
While you are working at your computer at NCEAS, you receive a frantic call from someone working on an OHI project:

THEY NEED HELP!

"No problem" you say, "I'll get Julie or Melanie or Gage"...but they are no where to be found!

"Hold on, I'll get Jamie or Casey or Courtney"...but they are also gone.

It is clearly going to be up to you to save the day!!

## You've got this!

Please help them answer the following questions.  You will work from this document, using the code chunks as a work space to work explore the data. Do NOT bother keeping the code chunks neat and organized, we want to see your work process in all its messy glory.  However, I DO recommend including plenty of comments, mostly because I have found this helps me organize my thoughts.   

You can ask us any questions along the way (slack or issues)!

Good luck!

## Questions

### Getting the data

*Goal* Read the latest OHI global data into the R working space (call the data object "scores").

*Hint* Here is a link to the data: 
https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/eez/scores.csv

```{r, include=FALSE}
## Working space

scores <- read.csv("https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/eez/scores.csv")

```

### Region explore

*Question* Why are there N=222 regions (see code below)...aren't there supposed to be 220? Can you make this data more human readable by including the full goal name (vs. the 2 letter abbreviation) and the country name.

*Hint 1* Here is a link to the official region list (the rgn_id column in this file matches the region_id column in the scores data): https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/eez/spatial/regions_list.csv

*Hint 2* Here is a table that includes goal names: https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/metadata_documentation/ohi_model/tables/ten_goals.csv

*Hint 3* Which region has all NA values for scores?

*Answer* This includes region 0 (the overall ohi global scores), and region 213 (antarctica, not included in global assessment)

```{r, include=FALSE}
## Working space

length(unique(scores$region_id))

regions <- scores %>%
  group_by(region_id) %>%
  summarise(n = length(region_id))

na_df1 <- scores %>% 
  group_by(region_id) %>%
  summarize(n = length(score[is.na(score)])) %>%
  arrange(-n)
#region 213

```

*Question* Which region id corresponds to the United States?
163

### Goal explore

*Question * When I look at the number of goals, N=19 show up....I thought there were supposed to be 10 goals!!! Can you explain what all these are? Also, why do some of the goal abbreviations have 2 letters and others 3 letters?  

*Hint 1* Review the methods doc here: https://raw.githack.com/OHI-Science/ohi-global/published/global2018/Results/Supplement_Results.html

*Answer* The 10 main goals have 2 letter abbreviations, while the 8 sub goals have 3 letter abbreviations. Each region also have an Index category, which is the overall score for that region based off of all of the goals.

```{r, include=FALSE}
## Working space

unique(scores$goal)

```




### Dimension explore

*Question* There are 6 dimensions reported in this dataset (future, pressures, resilience, score, status, and trend).  When I look at the number of records in each of these categories, I notice that there are more "score" values reported than "status" values (see working space below). Given that scores are calculated using status, it seems like these should have the same number of records. Can you figure out what is going on here. Is something going wrong?  

*Answer* There are more score values than status values, because of the “Index” scores. The Index score for each region does not have a corresponding status value, only a score. This makes there appear to be more score values.

```{r, include=FALSE}
## Working space

unique(scores$dimension)
table(scores$dimension)

table(scores$goal)
table(scores$goal, scores$dimension)

# ok...but why are the different values for pressures/resilience/trend and future?
# addition of region 0!
ao <- filter(scores, goal=="AO")
table(scores$region_id, scores$dimension)
```


### Missing data

*Question* Figure out which goals/subgoals have the most NA values for scores in 2019.  Which ones have the least?  Can you discern any reason why some goals have lots of missing values and others do not?

*Hint* Include only dimension = score, and year = 2019, and cut region_id = 0.

*Answer* Natural Products (NP) has the most NA values for 2019 with 96. Carbon Storage (CS) has the second most NA values, with 73. The reason for NP having so many NA values could be that some regions do not harvest any of the 6 natural product categories included in the score calculation. The goals with the most NA values could be harder to measure than the goals with the least NA values. 

```{r, include=FALSE}
## Working space

na_data <- scores %>% 
  filter(dimension == "score") %>%
  filter(region_id != 0) %>%
  filter(year == 2019) %>%
  group_by(goal) %>%
  summarise(n = length(score[is.na(score)]))

```


*Question* If we have a goal with a future status of 80 and status of 90...what is the score?  

*Hint* Isolate the future, status, and score values for one region and one goal and see if you can identify a pattern.

*Answer* They are averaged. The score would be 85. 

```{r, include=FALSE}
## Working space
df_future_status <- scores %>%
  filter(dimension %in% c("future", "status", "score"), region_id == 131) #filter for dimension in future, status, score and region_id = mexico. Looks like the future and status scores are just averaged.  


```


### Metadata

*Project* Based on your data exploration and other resources, provide some metadata that describes each variable.  Write it so it would be useful to you in the future as you are looking through these data.

Write it in the following table.  NOTE: Knit the document to see if the table is formatting correctly (but don't bother if you do not know how to knit a document or if you are running into errors!).

Variable   | Description                  | Values
---------- | ---------------------------- | ------------------------------
goal       | The category of ocean-derived benefit. Abbreviations with 2 letters are goals, while abbreviations with 3 letters are subgoals, which are used to calculate the goals.                              | AO, BD, CP, CS, CW, ECO, FIS, FP, HAB, ICO, Index, LE, LIV, LSP, MAR, NP, SP, SPP, TR
dimension  | Variables influencing goal calculation.                             | future, pressures, resilience, score, status, trend
region_id  | Numerical value which identifies the region or country in the assessment. | 0 to 250 
score      | The OHI score associated with an observation.                             |    0 to 100      
year       |  Scenario year               | 2012 to 2019 


Abbreviation   | Name                  |    Type 
---------- | ----------------------------| ---------------------------- 
 AO        |  Artisanal Fishing Opportunity    | Goal
 BD        |          Biodiversity              | Goal
  CP       |           Coastal Protection                       |  Goal
  CS       |    Carbon Storage                          |     Goal      
   CW      |       Clean Waters                      |   Goal 
 ECO       |         Economies                       | Subgoal
 FIS       |           Fisheries                       | Subgoal
  FP       |          Food Provision                        |  Goal
  HAB      |       Habitat                       |  Subgoal      
   ICO     |     Iconic Species                        |   Goal 
 Index     |      Index                   | Global Average
 LE        |          Coastal Livelihoods and Economies                        | Goal
    LIV    |       Livelihoods                           |  Subgoal
  LSP      |      Lasting Special Places                        |  Subgoal      
   MAR     |     Mariculture                        |   Subgoal
    NP     |        Natural Products                          | Goal
    SP     |        Sense of Place                          |  Goal
  SPP      |        Species Condition                      |  Subgoal      
   TR      |       Tourism and Recreation                      |   Goal


```{r, include=FALSE}
## Working space

summary(scores)
unique(scores$goal) # get all goal values
unique(scores$dimension)

```


### Plot

*Project* Create a scatterplot that compares 2012 and 2019 *scores* for each region for the artisanal opportunities goal. Based on this, do scores for the artisanal opportunities goal appear to have increased or decreased over time? 

*Answer* Write your answer here!


```{r, include=FALSE}
## Working space

scores_12_18 <- scores %>%
  filter(year %in% c(2012, 2019), goal == "AO", dimension == "score")

ggplot(scores_12_18, aes(x = region_id, y = score)) +
  geom_point(aes(color = factor(year))) #<< this one seems to be the most inciteful graph among the three. 

ggplot(scores_12_18, aes(x = year, y = score)) +
  geom_point(aes(color = factor(region_id))) + 
  theme(legend.position = "none") #doesnt tell us anything at all 

scores_2012 <- scores %>%
  filter(year == 2012, goal == "AO", dimension == "score")

scores_2018 <- scores %>%
  filter(year == 2019, goal == "AO", dimension == "score")

plot(scores_2012$score, scores_2018$score) #doesnt tell us much 

# make two new columns for the scores corresponding to 2012 and 2018
new_scores_df <- scores_12_18 %>%
  spread(year, score) %>%
  mutate(difference = `2019` - `2012`)  

ggplot(new_scores_df, aes(x = `2012`, y = `2019`, size = difference)) +
  geom_point() +
  theme()
```


```{r, echo = FALSE, warning = FALSE}
ggplot(new_scores_df, aes(x = region_id, y = difference)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red")+
  theme_bw() +
  coord_flip() + 
  labs(y = "Score Difference between 2018 and 2012", x = "Region")

scatter <- ggplot(new_scores_df, aes(x = difference, y = region_id, color = case_when(
  difference > 0 ~ "red",
  difference < 0 ~ "blue",
  difference == 0 ~ "black"
))) +
  geom_point() +
  geom_vline(xintercept = 0) +
  theme_bw() +
  labs(x = "Score Difference between 2018 and 2012", y = "Region") +
  theme(legend.position = "none") 

library(plotly)

ggplotly(scatter)
```


Create a plot that shows the distribution (e.g., histogram or barplot or ??) of *Index* scores of all countries.

```{r across year plot}

## Code to create a plot showing distribution of country Index scores



```

