---
title: 'OHI 2021 - Fix pourpoints which get masked from the plume model '
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../workflow/templates/ohi_hdr.html' 
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---


[REFERENCE RMD FILE](http://ohi-science.org/ohiprep_v2021/globalprep/prs_land-based_nutrient/v2021/STEP1_pourpoints_prep.Rmd)


# Summary
This prep assigns the pourpoints which are excluded from the plume model due to the ocean mask to their nearest coastal point, and rewrites the appropriate shapefiles to run through the plume model. The resulting files will be the files which should be run through the plume model.

## The following data are used:

* 



## Initial set-up code

```{r setup, eval = FALSE}
library(tidyverse)
library(raster)
library(sf)
library(mapview)
library(janitor)
library(here)
library(parallel)
library(doParallel)
# install.packages("nngeo")
library(nngeo)


source(here('workflow/R/common.R'))

ww_raw_dir <- "/home/shares/ohi/git-annex/land-based/wastewater/data/raw"
ww_intermediate_dir <- "/home/shares/ohi/git-annex/land-based/wastewater/data/interim"
prs_int_dir <- "/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2021/int"
prep <- file.path("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2021")


```

## 1. Correct the global pourpoints file so that it will not be masked out of the plume model

NOTE: for v2022, you can likely skip step 1 of this script and move directly to step 2. 

Create list of missing pourpoints 

```{r}
global_plume <- st_read("/home/shares/ohi/git-annex/land-based/wastewater/data/raw/pour_points/global_plume_2007_2010.shp")
# plot(global_plume$geometry)
mapview(tail(global_plume, 10000))

global_plume <- st_transform(global_plume, 4326) 

## exclude the coastal pourpoints from the global pourpoint file
coastal_pourpoints <- st_read(file.path(ww_intermediate_dir, "watersheds_coastal.shp"))

coastal_pp_ids <- unique(coastal_pourpoints$basin_id)

global_plume_coastal <- global_plume %>%
  filter(basin_id %in% coastal_pp_ids)

## now grab a list of the pourpoints which were included in the first plume model run of v2021
pp_2005_all <- st_read(file.path(prs_int_dir, "pourpoints/pourpoints_crop_manure_leached_volt_N_2005.shp"))

fils_try2 <- data.frame(x = list.files("/home/sgclawson/plumes/archive/output1_try2/", pattern="tif$", full.names = FALSE, recursive = TRUE))

fils_try1 <- data.frame(x = list.files("/home/sgclawson/plumes/archive/output1_try1/", pattern = "tif$", full.names = FALSE, recursive = TRUE))

all_files_done <- rbind(fils_try2, fils_try1)

pourpoint_names_files <- data.frame(fils = sub('.*plume_effluent_', '', all_files_done$x))

pourpoint_names <- data.frame(basin_id = sub('.tif', '', pourpoint_names_files$fils)) %>%
  mutate(basin_id = as.character(basin_id)) %>%
  dplyr::filter(str_detect(basin_id, "subsets", negate = TRUE))


## now filter out these pourpoints from our global plume to get the ones which were excluded
global_plume_exclude <- global_plume_coastal %>%
  filter(!(basin_id %in% pourpoint_names$basin_id)) # many of these values 


setdiff(pp_2005_remaining$basin_id, global_plume_exclude$basin_id)
zero_vals <- c(setdiff(global_plume_exclude$basin_id, pp_2005_remaining$basin_id))

global_plume_exclude_zero <- global_plume_exclude %>%
  filter(!(basin_id %in% zero_vals)) # now we have a list of the pourpoints that were excluded from the first v2021 plume model run

```

Now take the pourpoints which are excluded from the plume model and find the nearest shore line point. Assign the masked out pourpoints to the nearest shoreline and save that file. 

```{r}
shoreline_points <- st_read(file.path(prs_int_dir, "global_pourpoints_fix/coastline_1km_points.shp"))


## try this on a subset of points
shoreline_points_subset <- head(shoreline_points, 1000) %>%
  mutate(row = row_number())
mapview(shoreline_points_subset)

global_plume_exclude_zero_test <- global_plume_exclude_zero %>%
  mutate(row = row_number()) %>%
  filter(row %in% c(2418:2559)) %>%
  mutate(row = row_number())

mapview(global_plume_exclude_zero_test)

nearest_neighbor <- st_nn(global_plume_exclude_zero_test, shoreline_points_subset, sparse = TRUE)


test <- data.frame(row_id_pours = 1:142, row_id_shoreline = unlist(nearest_neighbor))

test_global_plume_fix <- global_plume_exclude_zero_test %>%
  st_drop_geometry() %>%
  left_join(test, by = c("row" = "row_id_pours")) %>%
  left_join(shoreline_points_subset, by = c("row_id_shoreline" = "row")) %>%
  st_as_sf() # %>%
 # st_jitter(0.05)

mapview(test_global_plume_fix) # it worked! 

## now try on all points...


shoreline_points_nn <- shoreline_points %>%
  mutate(row_id_shoreline = row_number())
mapview(head(shoreline_points_nn, 100))

global_plume_exclude_zero_nn <- global_plume_exclude_zero %>%
  mutate(row_id_pours = row_number())

mapview(global_plume_exclude_zero_nn)

nearest_neighbor <- st_nn(global_plume_exclude_zero_nn, shoreline_points_nn, sparse = TRUE, parallel = 12)


matched <- data.frame(row_id_pours = 1:nrow(global_plume_exclude_zero_nn), row_id_shoreline = unlist(nearest_neighbor))

global_plume_fix <- global_plume_exclude_zero_nn %>%
  st_drop_geometry() %>%
  left_join(matched, by = c("row_id_pours")) %>%
  left_join(shoreline_points_nn, by = c("row_id_shoreline")) %>%
  st_as_sf() # %>%
 # st_jitter(0.05)

st_write(global_plume_fix, file.path(prs_int_dir, "global_pourpoints_fix/excluded_fixed.shp"))

mapview(global_plume_fix)
mapview(global_plume_exclude_zero_nn)

## it worked! 

global_plume_included <- global_plume_coastal %>% 
  filter((basin_id %in% pourpoint_names$basin_id)) %>%
  dplyr::select(basin_id, geometry)

global_plume_fix_final <- global_plume_fix %>%
  dplyr::select(basin_id, geometry)  %>%
  rbind(global_plume_included)

st_write(global_plume_fix_final, file.path(prs_int_dir, "global_pourpoints_fix/global_pourpoints_coastal_fix.shp"))

```

2. Correct each yearly pourpoint file to match the geometries in the fixed one we created above.

```{r}
files_list <- str_subset(string = c(list.files(file.path(prs_int_dir, "pourpoints")), full = TRUE), ".shp")

excluded_shp <- st_read(file.path(prs_int_dir, "global_pourpoints_fix/excluded_fixed.shp")) %>%
  dplyr::select("basin_id" = "basin_d", geometry)

excluded_pp_list <- c(excluded_shp$basin_id)

registerDoParallel(cores = 5)
foreach(file = files_list) %dopar% {
  
  # file = files_list[1]
  
  pp <- st_read(file.path(prs_int_dir, "pourpoints", file)) 
  
  
  pp_included <- pp %>%
    filter(!(basin_id %in% excluded_pp_list))
  
  pp_excluded_values <- pp %>% 
    filter((basin_id %in% excluded_pp_list)) %>%
    st_drop_geometry()
  
  ## now join excluded values with the fixed pp dataset
  pp_excluded_fix <- pp_excluded_values %>%
    left_join(excluded_shp) %>%
    st_as_sf() %>%
    dplyr::select(basin_id, geometry, effluent)
  
  # mapview(pp_excluded_fix)
  
  ## now rbind back with the included 
  
  pp_all_fix <- rbind(pp_included, pp_excluded_fix)
  

 # final_name <- paste0("FIXED_", file)
  
  st_write(pp_all_fix, file.path(prs_int_dir, "pourpoints_FINAL", file), delete_dsn = TRUE) # this is what you will run through the plume! 
}

```

