---
title: 'OHI 2021 - Land based nutrient ocean mask prep '
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../workflow/templates/ohi_hdr.html' 
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---


[REFERENCE RMD FILE](http://ohi-science.org/ohiprep_v2021/globalprep/prs_land-based_nutrient/v2021/STEP1_pourpoints_prep.Rmd)


# Summary
This prep does the following:
 - creates an ocean mask buffered 1km inland to use with the plume model (this way we will grab many pourpoints that would otherwise be masked out in the plume model, but are still coastal).
 - Create a shapefile of points 1km inland along the coast. We use this in the next prep step to assign any pourpoints that were masked out of the original plume run to their nearest coastal point. This way we won't lose any pourpoints in the diffusion model. 
 
## The following data are used:

* An ocean raster  


# Updates from previous assessment
This is a new layer for the 2021 assessment year.

## Initial set-up code

```{r setup, eval = FALSE}
library(tidyverse)
library(raster)
library(sf)
library(mapview)
library(janitor)
library(here)
library(parallel)
library(doParallel)


source(here('workflow/R/common.R'))

ww_raw_dir <- "/home/shares/ohi/git-annex/land-based/wastewater/data/raw"
ww_intermediate_dir <- "/home/shares/ohi/git-annex/land-based/wastewater/data/interim"
prs_int_dir <- "/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2021/int"
prep <- file.path("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2021")


```

```{r}
ocean_mask <- raster(file.path("/home/shares/ohi/git-annex/globalprep/cw_nutrient_o_chem/raw", "ocean_mask_landnull.tif"))


# ocean_mask <- raster(file.path(prs_int_dir, "ocean_mask_50km.tif"))

plot(ocean_mask)

## Gapfill using mean of surrounding cells that are NA
gf_raster <- function(x){raster::focal(x, w = matrix(1,3,3), fun = mean, na.rm=TRUE, pad = TRUE, NAonly=TRUE)}

## Repeat 1 time... we want to create a 1km inland buffer
i <- 0
while (i < 1){
r <- gf_raster(ocean_mask)
i <- i + 1
print(i)
}

## write the 1km ocean mask buffered
writeRaster(r, file.path(prs_int_dir, "ocean_masks/ocean_mask_1km.tif"))


## now find the difference between the original ocean mask and the 1km buffer so that we can get a line of 1km inland cells

ocean_mask[ocean_mask ==0] <- 1

r[r == 0] <- 1 

line_1km <- r - ocean_mask

line_1km <- overlay(r,
                      ocean_mask,
                      fun=function(r1, r2, na.rm = TRUE){return(r1-r2)})

# 
# beginCluster(n=8)
# z <- clusterR(stack(r,ocean_mask), overlay, arg = list(fun = function(r1, r2, na.rm = TRUE){return(r1-r2)}))
# endCluster()

beginCluster(n=12)
z2 <- clusterR(stack(r,ocean_mask), overlay, arg = list(fun = sum, na.rm = TRUE))
endCluster()

z2
plot(z2)

stacker <- stack(r,ocean_mask)

calc(stacker, fun="sum", na.rm=TRUE, filename = "something")

beginCluster(n=12)
z <- clusterR(stacker, calc, args=list(sum, na.rm=T))
endCluster()

cellStats(r, "sum", na.rm = TRUE) # 531466807
cellStats(ocean_mask, "sum", na.rm = TRUE) # 529550427
plot(z)
unique(z) # [1] 0 1 2 ; good

z2 <- z

z2[z2 != 1] <- NA 

z2
plot(z2)
test <- raster::select(z2)
plot(test)

writeRaster(z2, file.path(prs_int_dir, "global_pourpoints_fix/ocean_1km_line_rast.tif"))


## now convert to spatial points
points <- rasterToPoints(z2)

points_sf <- st_as_sf(data.frame(points), coords = c("x", "y"), crs = 4326)

mapview(tail(points_sf, 10000))


st_write(points_sf, file.path(prs_int_dir, "global_pourpoints_fix/coastline_1km_points.shp"))

```

