---
title: 'OHI 2021: Saltmarsh trend'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

# Summary

This script generates the trend of saltmarsh.

"We show a net salt marsh loss globally, equivalent to an area double the size of Singapore (719 km2), with a loss rate of 0.28% year from 2000 to 2019" ([Campbell et al. 2022](https://doi.org/10.1038/s41586-022-05355-z)).

Apply a decreasing rate of .28% per year.

## Updates from previous assessment

In 2023 we updated the rate of salt marsh decrease per year from the previous assessment. Previously we applied a decreasing rate of 1.5% per year, or -0.015 over each region. This was based on the reference: "The global loss rate of threatened coastal ecosystems is estimated at 4--9% year for corals (Gardner et al. 2003, Bellwood et al. 2004), a minimum of 1--2% year for saltmarshes (Adam 2002; Lotze et al. 2006)" <https://link.springer.com/article/10.1007%2Fs12237-008-9038-7>. We are now using a value of -0.28% per year of lost salt marsh habitat, based on the findings of Campbell et al. 2022.

------------------------------------------------------------------------

## Data Source

Campbell, A.D., Fatoyinbo, L., Goldberg, L. et al. Global hotspots of salt marsh change and carbon emissions. Nature 612, 701--706 (2022). <https://doi.org/10.1038/s41586-022-05355-z>

**Date Acessed**: July 31 2023

**Description**: Estimate of a global percentage of loss of saltmarsh habitat per year.

**Time range**: NA, based on a study from 2000 to 2019

------------------------------------------------------------------------

# Methods

## Setup

```{r setup, echo = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
library(readr)      # for read_csv()
library(raster)
library(here)
library(sf)
library(fasterize)
library(tidyverse)
library(mapview)
library(sp)
library(rgeos)


source(here('workflow/R/common.R'))

current_year <- 2023
version_year <- paste0("v",current_year)
previous_version_year <- paste0("v", 2021)

goal     <- 'globalprep/hab_saltmarsh/v2023'
dir_git  <- file.path('~/github/ohiprep_v2023', goal)
region_data()

dir_git_old  <- file.path(dir_git, "globalprep/hab_saltmarsh/v2021")

```

Apply a global trend for each region.. not satisfying.

```{r, eval = FALSE}
## read in extent data
all <- read.csv(file.path(here("globalprep/hab_saltmarsh/", version_year, "data/habitat_extent_saltmarsh_updated.csv")))

# specify trend
all <- all %>%
  filter(km2 > 0) %>%
  filter(rgn_id < 255) %>%
  dplyr::select(-year) %>%
  mutate(trend = -0.0028)


# Save gapfilling 

trend_gaps <- all %>%
  mutate(variable = "trend") %>%
  mutate(habitat = "saltmarsh") %>%
  mutate(gapfill = 1)
write.csv(trend_gaps, here("globalprep/hab_saltmarsh/", version_year, "data/trend_saltmarsh_gf.csv"), row.names=FALSE)

### save trend data:
trend <- trend_gaps %>%
  dplyr::select(rgn_id, habitat, trend) %>%
  mutate(year = current_year)
write.csv(trend, here("globalprep/hab_saltmarsh/", version_year, "/data/habitat_trend_saltmarsh_updated.csv"), row.names=FALSE)

```
