---
title: 'OHI: Clean Waters/Trash trend'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

# Summary

The coastal & marine trash trend score is calculated based on the most recently available data sources on a few different forms of marine trash: marine macroplastics/debris, marine microplastics, and coastal debris. 

Different data sources were investigated and compiled in v2025, and information on them was combined here to create an overall trash trend score analysis.

# Microplatics

Source: https://www.ncei.noaa.gov/products/microplastics

Our microplastics data is from NOAA's National Center for Environmental Information 'NCEI Marine Microplastics product'. The data is available as a csv, shapefiles, GeoJSON, JSON, FGDB, or KML. The data on microplastic concentrations contains measurements in pieces/cubic meter,and were ranked on a scale of 'Very Low', 'Low', 'Medium', and 'High'. 

Observations were not specified at the country level, but merely grouped by ocean, or in some cases subregion. However, the point observations could be rasterized and grouped into the eez boundaries of each country. The concentration values could be rescaled for a 0-1 range. 


```{r setup, message = FALSE, warning = FALSE, verbose = FALSE, eval= FALSE}
#set options for all chunks in code
knitr::opts_chunk$set(message = FALSE, warning = FALSE, eval=FALSE)

# load packages
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  here,
  janitor,
  raster,
  terra,
  sf,
  stars,
  maps,
  httr,
  jsonlite,
  purrr,
  tictoc,
  sf, 
  googleVis,
  RColorBrewer,
  foreach,
  doParallel, # for using multiple cores, if needed
  tidyverse, 
  ohicore,
  ggplot2,
  ncmeta,
  ncdf4,
  plotly,
  zoo,
gstat
)

# ---- sources! ----
source(here("workflow", "R", "common.R")) # file creates objects to process data

# ---- set year and file path info ----
current_year <- 2025 # Update this in the future!!
version_year <- paste0("v",current_year)
data_dir_version_year <- paste0("d", current_year)

# ---- data directories ----

# Raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")

# prs_oa dir
cw_dir <- here::here(dir_M, "git-annex", "globalprep", "cw_pressure_trash", version_year)

# final output dir
output_dir <- here::here("globalprep","cw_pressure_trash", version_year, "output")

# set colors
cols = rev(colorRampPalette(brewer.pal(9, 'Spectral'))(255)) # rainbow color scheme
```


# The data was manually downloaded and uploaded to the mazu server. In year's after v2025, the NOAA_NEI_Marine_Microplastics can be deleted and replaced to maintain the below code.

# Method 1: Use vector data and intersect for each region

```{r, eval= FALSE}
# Load in the microplastics data from the Mazu server
microplastics_data <- st_read(here::here(raw_data_dir , "NOAA_NEI_Marine_Microplastics", "Microplastic_Data_MasterList_GISTeam_XYTableToPoint.shp"))
```
# Rescale the microplastics data on a 0-1 scale

```{r, eval= FALSE}
# Check the microplastics columns
colnames(microplastics_data)

# Check the unique methods of sampling
unique(microplastics_data$Medium)

# Check to see the different ways of scaling for each medium
unique(microplastics_data$Symbology)

# We are gonna rescale using the Symbology column, grouping by general category so that the different concentration units can be plotted together.
```


```{r, eval= FALSE}
# Again, we will be rescaling by sampling method, each of which had different units 
# Create data frames for each sample setting
ocean_subsurface <- microplastics_data %>%
  filter(str_detect(microplastics_data$Symbology, "Ocean Subsurface"))

ocean_surface <- microplastics_data %>%
  filter(str_detect(microplastics_data$Symbology, "Ocean Surface"))

ocean_sediment <- microplastics_data %>%
  filter(str_detect(microplastics_data$Symbology, "Ocean Sediment"))

beach <- microplastics_data %>%
  filter(str_detect(microplastics_data$Symbology, "Beach"))

nurdle_patrol <- microplastics_data %>%
  filter(str_detect(microplastics_data$Symbology, "Nurdle Patrol"))
```

```{r, eval= FALSE}
# Create function to rescale each df separately based on their 95th percentile value
rescale_95 <- function(x) {
  x_min <- min(x)
  x_95 <- quantile(x, 0.95)
  x_scaled <- (x - x_min) / (x_95 - x_min)
  x_scaled[x_scaled > 1] <- 1
  return(x_scaled)
}

# Rescale all the dfs
ocean_subsurface$concentration_rescaled <- rescale_95(ocean_subsurface$Microplast)
ocean_surface$concentration_rescaled <- rescale_95(ocean_surface$Microplast)
ocean_sediment$concentration_rescaled <- rescale_95(ocean_sediment$Microplast)
beach$concentration_rescaled <- rescale_95(beach$Microplast)
nurdle_patrol$concentration_rescaled <- rescale_95(nurdle_patrol$Microplast)

# Create a column for setting so that we could view observations by setting if so desired
ocean_subsurface <- ocean_subsurface %>%
  mutate(Setting = "Subsurface")

ocean_surface <- ocean_surface %>%
  mutate(Setting = "Surface")

ocean_sediment <- ocean_sediment %>%
  mutate(Setting = "Sediment")

beach <- beach %>%
  mutate(Setting = "Beach")

nurdle_patrol <- nurdle_patrol %>%
  mutate(Setting = "Nurdle Patrol")

# Concatenate all the df together for our total rescaled data
microplastics_rescaled <- rbind(ocean_subsurface,
                                ocean_surface,
                                ocean_sediment,
                                beach,
                                nurdle_patrol)
```


# Rasterize the Microplastics observations for masking

```{r, eval= FALSE}
# Rasterize the microplastic observations:
# Isolate points from the data set
points <- st_as_sf(microplastics_rescaled, coords = c("Longitude_", "Latitude_"), remove = FALSE, crs = 4326) 

# Get the bounding box (extent) of the points
microplastics_bbox <- ext(points)

# Expand slightly to ensure full 1-degree cell coverage
microplastics_bbox <- extend(microplastics_bbox, 0.5)  # adds 0.5 degree padding on all sides

# Create a raster with 1-degree resolution
rast_template <- rast(extent = microplastics_bbox, 
                   resolution = 1, # use a 1-degree resolution
                   crs = "EPSG:4326")

# Rasterize using mean value per cell (you can also use "sum", "max", etc.)
microplastics_rast <- rasterize(points, rast_template, field = "concentration_rescaled", fun = mean)
```

```{r, eval= FALSE}
# Plot the raster
plot(microplastics_rast)
```

# Run an Inverse-Distance Weighted Correlation to gapfill our data

```{r, eval= FALSE}
# Get cell coordinates
coords <- as.data.frame(xyFromCell(microplastics_rast, 1:ncell(microplastics_rast)))
coords$value <- values(microplastics_rast)

# Separate known and NA cells
known <- coords[!is.na(coords$value), ]
unknown <- coords[is.na(coords$value), ]

# Convert to sf objects for gstat
known_sf <- st_as_sf(known, coords = c("x", "y"), crs = crs(microplastics_rast))
unknown_sf <- st_as_sf(unknown, coords = c("x", "y"), crs = crs(microplastics_rast))

# Create a gstat IDW model
idw_model <- gstat(formula = value ~ 1, data = known_sf, nmax = 15, set = list(idp = 2))

# Predict missing values
predicted <- predict(idw_model, newdata = unknown_sf)

# Add predicted values back to raster
coords[is.na(coords$value), "value"] <- predicted$var1.pred

# Fill in the values for our raster
filled_r <- microplastics_rast
values(filled_r) <- coords$value

# Plot the results
plot(filled_r)
```

# Load in the land mass rasters

```{r, eval= FALSE}
# Load in the land mass data from the Mazu server
landmasses <- terra::rast("/home/shares/ohi/git-annex/globalprep/_raw_data/landmass_raster/landmasses.tif")
```

```{r, eval= FALSE}
plot(landmasses)
```
```{r, eval= FALSE}
# Check the total number of layers in the raster and their names
nlyr(landmasses) 
names(landmasses)

# Grab only a single one of the landmass rasters
landmasses_single <- landmasses[[3]]

# Plot the third layer to see how we need to assign values 
plot(landmasses_single)
```

# Mask out the oceans and seas

```{r, eval= FALSE}
# Set values higher than 250 to NA
landmasses_single[landmasses_single <= 250] <- NA # This radiance level was selected based on the third raster layer values. This may have to be updated if the rasters change

# Resample the landmass raster to match the filler raster
microplastics_re <- resample(filled_r, landmasses_single, method = "near")

# Mask out raster1 where raster2 has non-NA values
masked_microplastics <- mask(microplastics_re, landmasses_single)
```

```{r, eval= FALSE}
# Plot the newly masked microplastics data to check if it worked correctly
plot(masked_microplastics)
```

# Get the eez regions for each country in OHI

```{r, eval= FALSE}
# Grab regional data from OHI:
# Get the region boundaries for all countries we will calculate scores for
ohi_rasters()
```

```{r, eval= FALSE}
# Check to see if crs are matched
# Match the crs of the data
microplastics_rast <- terra::project(masked_microplastics, crs(zones))

# Match the extent of the data by resampling
zones_aligned <- resample(zones, microplastics_rast, method = "near")

# Check to see if everything matches
crs(microplastics_rast) == crs(zones_aligned) #v2025: TRUE
ext(microplastics_rast) == ext(zones_aligned) # v2025: TRUE
res(microplastics_rast) == res(zones_aligned) # v2025: TRUE
```

# Find observations that overlap with eez regions

```{r, eval= FALSE}
# Use zonal to compute mean of point values per zone
zonal_stats <- zonal(microplastics_rast, zones_aligned, fun = "mean", na.rm = TRUE)
```

```{r, eval= FALSE}
region_data()
```

```{r, eval= FALSE}
# Check for NAs
sum(is.na(zonal_stats$mean))

# Subset the rows with NAs in the mean column for zonal_stats to view which countries are missing
missing_countries <- zonal_stats %>%
  filter(is.na(mean) == TRUE) %>%
  filter(regions_eez_with_fao_ant <= 277) # Remove zones for continents and subregions
```

# Potential Gapfilling: 

```{r, eval= FALSE}
# Rename the columns for ease of use
zonal_stats_clean <- zonal_stats %>% 
  rename_at('mean', ~'avg_concentration') %>%
  rename_at('regions_eez_with_fao_ant', ~'region')

# Get the mean of avg_concentration
mean_conc <- mean(zonal_stats_clean$avg_concentration, na.rm= TRUE)

# Load the regions all data for data names
region_data()

# Gapfill the 59 missing values with the global average concentration
zonal_stats_filled <- zonal_stats_clean %>%
   mutate(avg_concentration = case_when(
    is.na(avg_concentration) ~ mean_conc,
    TRUE ~ avg_concentration)) # Return values that are present; do not replace with NA

# Add in the region names
microplastics_eez <- left_join(rgns_all, zonal_stats_filled, by= c("rgn_ant_id" = "region"))
```

# Plot the results

```{r, eval= FALSE}
# Find missing values in the zonal_statistics
plot <- ggplot(microplastics_eez, aes(x= avg_concentration, y = rgn_id, colour = rgn_name)) +
  geom_point() +
  labs(title= "Microplastics Concentration by Region", 
       x= "Microplastics Concentration (pieces/m^3)",
y= "Region ID")

ggplotly(plot)
```

# Save the data as a csv

```{r, eval= FALSE}
# Filter and rename the microplastics_eez data
microplastics_save <- microplastics_eez %>%
  mutate(year = 2025) %>%
  rename(pressure_score = avg_concentration) %>%
  select(rgn_id, year, pressure_score)
```

```{r, eval= FALSE}
# Create file name for csv
file_name <- paste0("trash_pressure.csv")

# Create the full path
output_path <- file.path("output", file_name)

# Save the data frame
write.csv(microplastics_save, file = output_path, row.names = FALSE)
```


# Plotting our results

```{r, eval= FALSE}
# Rename the 2025 scores
trash_pressure <- trash_pressure %>%
  rename(pressure_score_2025 = pressure_score)

# Rename the 2015 scores
trash_eez_2015_updated <- trash_eez_2015_updated %>%
  rename(pressure_score_2015 = pressure_score)

# Retreive region names from region_data()
region_data()
region_names <- rgns_eez %>%
  select(rgn_id, rgn_name)
  
# Join the datasets to one another, then add region names
trash_compare <- full_join(trash_pressure_filtered, trash_eez_2015_updated, by= "rgn_id")
trash_compare <- full_join(trash_compare, region_names, by= "rgn_id")
```


# Plot the scores between 2025 and 2015 (different metrics of recording trash pressure)

```{r, eval= FALSE}
plot <- ggplot(trash_compare, aes(x= pressure_score_2015, y= pressure_score_2025, color= as.factor(rgn_name))) + 
geom_point() +
  geom_smooth() +
  labs(color= "Region Name",
       title= "Comparison of Trash Pressures from 2015 to 2025")

ggplotly(plot)
```


# Find the difference between scores and plot

```{r, eval= FALSE}
trash_compare <- trash_compare %>%
  mutate(diff = pressure_score_2025 - pressure_score_2015)
```

```{r, eval= FALSE}
plot <- ggplot(trash_compare, aes(x= diff, y= rgn_name, color= as.factor(rgn_name))) + 
geom_col() +
  labs(color= "Region Name",
       title= "Comparison of Trash Pressures from 2015 to 2025")+
  theme(axis.text.y= element_blank(),
        legend.position= "none")

ggplotly(plot)
```

