---
title: "Learning about OHI Scores"
output: html_document
date: "2024-06-07"
---

```{r setup, include=FALSE}
## loading packages.. you might need to install them first
library(dplyr)
library(tidyr)
library(ggplot2)
```


## OH NO!
While you are working at your computer at NCEAS, you receive a frantic call from someone working on an OHI project:

THEY NEED HELP!

It is clearly going to be up to you to save the day!!

## You've got this!

Please help them answer the following questions.  You will work from this document, using the code chunks as a work space to work explore the data. Do NOT bother keeping the code chunks neat and organized.   

You can ask us any questions along the way (slack or issues)!

Good luck!

## Questions

### Getting the data

*Goal* Read the latest OHI global data into the R working space (call the data object "scores"). These are the scores that were calculated for the 2023 assessment. This is the final output after running the scripts in ohi-global. Poke around at the data to get a feel for what is there.

*Hint* Here is a link to the data: 
https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/eez/scores.csv


```{r, include=FALSE}
## Working space
scores = read_csv('https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/eez/scores.csv')
scores = as.data.frame(scores) %>% 
  rename(rgn_id = region_id)
str(scores) #'score','region_id',
rows_na <- which(is.na(scores[, c('score','region_id', 'goal')]))
test = scores[rows_na, ]

goals = unique(test$goal)

test2 = test %>% 
  filter(dimension == "score") %>% 
  group_by(region_id, goal) %>% 
  summarise(score = is.na(score))

scores[rows_na, ]
```

### Region explore

*Question* Why are there N=222 regions (see code below)...aren't there supposed to be 220? 

Can you make this data more human readable by including the full goal name (vs. the 2 letter abbreviation) and the country name.

*Hint 1* Here is a link to the official region list (the rgn_id column in this file matches the region_id column in the scores data): https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/eez/spatial/regions_list.csv

*Hint 2* Here is a table that includes goal names: https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/metadata_documentation/ohi_model/tables/ten_goals.csv

*Hint 3* Which region has all NA values for scores?

*Answer* Write your answer here!

```{r, include=FALSE}
## Working space
regions = read_csv('https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/eez/spatial/regions_list.csv')
goals = read_csv('https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/metadata_documentation/ohi_model/tables/ten_goals.csv')

regions2 = regions %>% 
  select(rgn_id, rgn_name)

## Join the dataframes by keys to obtain long versions of each region id and goal abbreviation
scores_long = scores %>% 
  left_join(goals, join_by(goal == Abbreviation)) %>% 
  # mutate(rgn_id = as.character(rgn_id)) %>% 
  left_join(regions2, join_by(rgn_id == rgn_id))

```

*Question* Which region id corresponds to the United States?

```{r}
united_states = regions %>% 
  filter(rgn_name == "United States")
united_states$rgn_id
```


### Goal explore

*Question * When I look at the number of goals, N=19 show up....I thought there were supposed to be 10 goals!!! Can you explain what all these are? Also, why do some of the goal abbreviations have 2 letters and others 3 letters?  

*Hint 1* Review the methods doc here: https://raw.githack.com/OHI-Science/ohi-global/published/documents/methods/Supplement.html

*Answer* Write your answer here!

```{r, include=FALSE}
## Working space
unique(scores_long$Goal)
## There are 10 goals with 8 subgoals, then we calculate the 10 goals with equal weighting. The 19th is the "index" goal which is the average for a given region
```


### Dimension explore

*Question* There are 6 dimensions reported in this dataset (future, pressures, resilience, score, status, and trend).  When I look at the number of records in each of these categories, I notice that there are more "score" values reported than "status" values (see working space below). Given that scores are calculated using status, it seems like these should have the same number of records. Can you figure out what is going on here. Is something going wrong?

*Answer* Write your answer here!

```{r, include=FALSE}
## Working space

unique(scores$dimension)
table(scores$dimension)

table(scores$goal)
table(scores$goal, scores$dimension)

## The 'index' goal which is the region average of all scores doesn't have pressures, resilience, status or trend calculated for it. Scores calculated using present and future status. 
```



### Missing data

*Question* Figure out which goals/subgoals have the most NA values for scores in 2021.  Which ones have the least? Can you discern any reason why some goals have lots of missing values and others do not?

*Hint* Include only dimension = score, and year = 2023, and cut region_id = 0.

*Answer* Write your answer here!

```{r, include=FALSE}
## Working space
na_check = test %>% 
  filter(dimension == "score" & year == 2021)

na_goals = na_check %>% 
  group_by(goal) %>% 
  count(nrow(score)) %>% 
  ungroup() %>% 
  filter(n %in% max(n) | n %in% min(n))

## Tourism and recreation has a lot of NA values because it includes uninhabited places, and there is no way to really gapfill it in a meaningful way.
## The ones that have the least tend to be calculated with raster data 

# na_goals[which.max(na_goals$n), ]
# na_goals[which.min(na_goals$n), ]

```


### Scores

*Question* If we have a goal with a future status of 80 and status of 90...what is the score?  

*Hint* Isolate the future, status, and score values for one region and one goal and see if you can identify a pattern.

*Answer* Write your answer here!

```{r, include=FALSE}
## Working space
scores3 = scores %>% 
  filter(dimension == "future" | dimension == "status" | dimension == "score") %>% 
  filter(rgn_id == 163 & goal == "BD" & year == 2021)
scores3
## Calculate the score by taking the average between the status and future status 
(70.22+68.58)/2
```


### Metadata

*Project* Based on your data exploration and other resources, provide some metadata that describes each variable.  Write it so it would be useful to you in the future as you are looking through these data.

Write it in the following table.  NOTE: Knit the document to see if the table is formatting correctly (but don't bother if you do not know how to knit a document or if you are running into errors!).

Variable   | Description                  | Values
---------- | ---------------------------- | ------------------------------
goal       | 2 or 3 digit code for goal.  | 
dimension  | parameters used to calculate scores as well as score itself | "future": likely future status; "status": current status; "pressures": potential limitors weighting down future status; "resilience": potential benefits limiting effect of pressure on future status; "score": the score calculated for each region and goal/subgoal; "trend": recent changes in current status 
region_id  | numeric region identifier                             | 
score      | metric defining how well a country is managing an individual goal (0-100)                             |
year       |  Scenario year               | 2012 to 2022


```{r, include=FALSE}
## Working space
unique(scores$dimension)
```


### Plot

*Project* Create a scatterplot that compares 2012 and 2023 *scores* for each region for the artisanal opportunities goal. Based on this, do scores for the artisanal opportunities goal appear to have increased or decreased over time? 

Why might this be? How could you begin to figure this out?

*Answer* Write your answer here!

```{r, include=FALSE, warning=FALSE}
## Working space
plot = scores %>% 
  filter(goal == "AO" & year == 2012 | year == 2023) %>% 
  filter(dimension == "score")

plot_wide = plot %>% 
  pivot_wider( names_from = year, values_from = score) %>% 
  filter(goal == "AO") %>% 
  drop_na()

max(plot_wide$rgn_id)

ggplot(plot_wide, aes(y = rgn_id, x = (`2023` - `2012`), color = (`2023` - `2012`))) + 
  geom_point() +
  theme_bw() + 
  labs(color = "Difference") +
  geom_segment(aes(x = 0, xend = 0, y = 0, yend = 250), color = "red", linetype = "dashed")

 
```


Create a plot that shows the distribution (e.g., histogram or barplot or ??) of *Index* scores of all countries.

```{r across year plot}

## Code to create a plot showing distribution of country Index scores

tmp = plot %>% 
  filter(goal == "Index") %>% 
  drop_na() %>%
  group_by(rgn_id)

ggplot(tmp, aes(x = score)) + 
  geom_histogram(color = "black", binwidth = 2) + 
  theme_bw() + 
  labs(x = "Score", y = "Frequency")

```
