---
title: 'OHI 2021 - Tourism and Recreation '
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html' 
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---


[REFERENCE RMD FILE](http://ohi-science.org/ohiprep_v2021/globalprep/tr/v2021/tr_data_prep.html)


# Summary
This document describes the steps for obtaining the data used to calculate the tourism and recreation goal for the 2021 global assessment.

The general calculation is:
tr = Ep * Sr 
and
Xtr = tr/90th quantile across regions

* Ep = Proportion of workforce directly employed in tourism
* Sr = (S-1)/5; Sustainability of tourism


## The following data are used:

* Tourism sustainability: World Economic Forum. The Travel & Tourism Competitiveness Index 2019 dataset (version 03 September 2019). 2019.
* Proportion of workforce directly employed in tourism: World Travel & Tourism Council ([WTTC](https://www.wttc.org/datagateway))
* Per capita GDP: (World Bank with gaps filled using CIA data), used to gapfill missing values in Tourism sustainability (in previous years)


# Updates from previous assessment

## Tourism sustainability
Data was updated to include the 2019 WEF, TTCI report 

## Tourism employment
WTTC data includes projections 10 years in the future, and in 2018 it was unclear when these projections began, so they used 2017 as their maximum data year. When downloading data for the 2020 assessment it was clear from the WTTC data gateway where the real data ended and projections began, so 2020 was used as the maximum data year.


**We were able to update the following data:**

* Tourism sustainability data from the WEC Travel and Tourism Competitiveness Report were updated to include the September 2019 report update. (downloaded on 03/11/2020)
* Proportion of jobs in tourism - WTTC data reported until 2030, but 2020 is most recent year of real data (year_max) (downloaded from WTTC on 03/12/2020)

## Initial set-up code

```{r setup, message=FALSE, warning=FALSE, results="hide"}

#library(devtools)
#devtools::install_github("ohi-science/ohicore@dev") # dont worry about devtools
library(ohicore)
library(tidyverse)
library(stringr)
library(WDI)
library(here)
library(janitor)
library(plotly)

source('http://ohi-science.org/ohiprep_v2021/workflow/R/common.R')

## maximum year of wttc data:
year_max <- 2020

source(here("globalprep/tr/v2021/R/tr_fxns.R"))

setwd(here("globalprep/tr/v2021"))

```


# Ep: Proportion of workforce directly employed in tourism

These data are from the [World Travel & Tourism Council](http://www.wttc.org/datagateway).  We use "direct" employment data (see mazu: git-annex/globalprep/_raw_data/WTTC/d2019/README.md for instructions on obtaining data). The data extend to 2030, which includes 10 years of projections. The actual data goes to 2020 (projected/real data are differentiated on the data gateway chart).

These data are cleaned and formatted using the R/process_WTTC.R script. Missing values are gapfilled using the UN georegion information.

```{r wttc prop tourism, message=FALSE, warning=FALSE, echo=FALSE, results="hide"}

## describe where the raw data are located:
scenario_yr <- "v2020"
dir_wttc <- file.path(dir_M, 'git-annex/globalprep/_raw_data/WTTC/d2020/raw')
dir_github <- here("globalprep/tr", scenario_yr)

## processing script that formats the WTTC for OHI, saves the following: intermediate/wttc_empd_rgn.csv
source(here("globalprep/tr/v2020/R/process_WTTC.R"))
##### v2020: Duplicates = Guadeloupe/Martinique, PR/VI, China (accounted for in the script)

## read in the dataset created by above function:
tr_jobs_pct_tour <- read_csv(here('globalprep/tr/v2020/intermediate/wttc_empd_rgn.csv')) %>% 
 dplyr::select(rgn_id, year, jobs_pct)

## format data to have complete years/regions and convert percentage of jobs to proportion of jobs
rgn_names <- read_csv('https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/eez/spatial/regions_list.csv') %>%
    dplyr::select(rgn_id)

# create data frame with a row for each combination of region id and year
rgn_names <- expand.grid(rgn_id = rgn_names$rgn_id, 
                             year= min(tr_jobs_pct_tour$year):max(tr_jobs_pct_tour$year)) 
      
tr_data_raw <- rgn_names %>%
  full_join(tr_jobs_pct_tour %>%
                rename(Ep = jobs_pct) %>%
                mutate(Ep = Ep/100) %>%
                mutate(Ep = ifelse(Ep > 1, NA, Ep)),
              by = c('rgn_id', 'year')) %>%
  filter(!rgn_id == 213) %>% 
  filter(!rgn_id == 255) # ditch disputed regions and Antarctica

## v2019: >2500 NAs in Ep column
## v2020: 2738 NAs in Ep column
# summary(tr_data_raw)


## gapfill missing data using UN georegion data:
georegions       <- georegions
georegion_labels <- georegion_labels

tr_data_raw <- tr_data_raw %>%
  left_join(georegions, by = 'rgn_id') %>%
  left_join(georegion_labels, by = 'rgn_id') %>%
  select(-r0)

# Calculate two different gapfill columns using r2 and r1
tr_data_raw_gf <- tr_data_raw %>%
  group_by(year, r2) %>%
  mutate(Ep_pred_r2 = mean(Ep, na.rm=TRUE)) %>%
  ungroup() %>%
  group_by(year, r1) %>%
  mutate(Ep_pred_r1 = mean(Ep, na.rm=TRUE)) %>%
  ungroup()

# first gapfill with r2, if no value available use r1; create column indicating whether value was gapfilled and if so, by what method.
tr_data_raw_gf <- tr_data_raw_gf %>%
  mutate(Ep_all = ifelse(is.na(Ep), Ep_pred_r2, Ep)) %>%
  mutate(Ep_all = ifelse(is.na(Ep_all), Ep_pred_r1, Ep_all)) %>% 
  mutate(gapfilled = ifelse(is.na(Ep) & !is.na(Ep_all), "gapfilled", NA)) %>%
  mutate(method = ifelse(is.na(Ep) & !is.na(Ep_pred_r2), "UN georegion (r2)", NA)) %>%
  mutate(method = ifelse(is.na(Ep) & is.na(Ep_pred_r2) & !is.na(Ep_pred_r1), "UN georegion (r1)", method)) 

######################################

### After gap-filling, make sure low/uninhabited regions are NA
# Create df for unpopulated/low populated regions
low_pop()
low_pop <- low_pop %>%
  filter(est_population < 3000 | is.na(est_population)) %>%  #filter out regions that have populations > 3000 and keep NA values 
  rename(rgn_label = rgn_nam)

#summary(tr_data_raw_gf)
# v2020 371 NAs

# make sure all the NAs are uninhabited regions
tr_data_nas <- tr_data_raw_gf %>% 
  filter(is.na(Ep_all)) %>% 
  select(rgn_id, year, r1_label, r2_label, rgn_label) %>% 
  left_join(low_pop, by = c("rgn_id", "rgn_label"))
#  filter(!duplicated(rgn_id))

max(tr_data_nas$est_population, na.rm=TRUE)<3000 # should be true

# make sure all the uninhabited regions are NA (along with gapfill and method if they were gapfilled above)
tr_data_raw_gf <- tr_data_raw_gf %>% 
  mutate(Ep_all = ifelse(rgn_id %in% low_pop$rgn_id, NA, Ep_all)) %>% 
  mutate(gapfilled = ifelse(is.na(Ep_all), NA, gapfilled)) %>% 
  mutate(method = ifelse(is.na(Ep_all), NA, method))


# check NAs once more 
 #summary(tr_data_raw_gf)
# v2019: Adding the low pop df identifies 13 additional regions that should be NA instead of gapfilled, taking the total number of NAs in the data set from 245 to 700
# v2020: Adding the low pop df takes the total number of NAs in the data set from 371 to 832


# save the gapfill report data
tr_data_gf <- tr_data_raw_gf %>%
  select(rgn_id, year, gapfilled, method) 

write_csv(tr_data_gf, here("globalprep/tr/v2020/output/tr_jobs_pct_tourism_gf.csv"))

tr_data <- tr_data_raw_gf %>%
  select(rgn_id, year, Ep=Ep_all) 



# Save gap-filled data
write_csv(tr_data, here("globalprep/tr/v2020/output/tr_jobs_pct_tourism.csv"))

```

## Data check and outlier investigation
```{r data check, message=FALSE, echo=FALSE}


## A quick check to make sure last year's values aren't too crazy different
## (NOTE: the source data has been updated, so there are some changes, but they shouldn't be super different)

old <- read_csv(here('globalprep/tr/v2019/output/tr_jobs_pct_tourism.csv')) %>%
  select(rgn_id, year, ep_old=Ep)

new <- read_csv(here('globalprep/tr/v2020/output/tr_jobs_pct_tourism.csv')) %>%
  left_join(old) %>%
  filter(year==2019) %>%
  arrange(ep_old)

# Visualize data comparison 
ggplotly(ggplot(new, aes(x = Ep, y = ep_old, labels = rgn_id)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 20, color = "red"))


######## investigate outliers (if applicable)

# new_diff <- new %>%
#   mutate(difference = ep_old-Ep) %>% 
#   filter(!is.na(ep_old)) %>% 
#   filter(!is.na(Ep))
# 
# outliers <- tr_data_raw %>% 
#   filter(rgn_id %in% c(15,116,117,118,119,122,125,127,173,250)) %>% 
#   filter(year == 2018) %>% 
#   select(rgn_id, rgn_label) %>% 
#   left_join(new, by = "rgn_id", "year") %>% 
#   mutate(difference = Ep-ep_old) %>% 
#   mutate(avg_diff = mean(new_diff$difference))

```

# Ts: Tourism sustainability

These data are from the World Economic Forum's "Travel and Tourism Competitiveness Report" (http://reports.weforum.org/travel-and-tourism-competitiveness-report-2019/downloads/) See mazu: _raw_data/WEF-Economics/ for more details and the raw data.

These data are not compatible across years, so only the most recent year of data is across scenarios.

These data are gapfilled using gdppcppp and UN georegion information (see next section for obtaining and preparing these data).

```{r WEF processing, eval=FALSE}

# read in files
ttci_raw <- read.csv(file.path(dir_M, "git-annex/globalprep/_raw_data/WEF-Economics/d2020/WEF_TTCR19_data_for_download.csv")) %>%
  dplyr::filter(row_number() %in% c(1,592)) %>% #Select the rows of relevance to us... TTCI values and the countries
  dplyr::select(11:158) %>% ## select only columns with countries in them  # check colnames(ttci_raw) to see what column number zimbabwe is... that is where we want to stop
  t() %>% # transpose the data to tidy
  data.frame() %>% 
  remove_rownames() %>%
  rename(country = "X1", score= "X2") %>% 
  mutate(score = as.numeric(as.character(score))) %>% 
  mutate(country = as.character(country))
  
#Checking the imported data vs 2017! 
ttci_v2017 <-read.csv(file.path(dir_M, "git-annex/globalprep/_raw_data/WEF-Economics/d2017/wef_ttci.csv")) %>% 
  rename(score_old= score) %>%
  arrange(country) %>% 
  left_join(ttci_raw)

plot(ttci_v2017$score, ttci_v2017$score_old)
abline(0,1, col="red") #Were the data actually updated? YES! 


#Changing names that are not recognized by ohicore
ttci <- ttci_raw %>%
    mutate(country = ifelse(country == "Congo, Democratic Rep.", "Democratic Republic of the Congo", country)) %>%
    mutate(country = ifelse(str_detect(country, "Ivoire"), "Ivory Coast", country))
  
  
ttci_rgn <- name_2_rgn(df_in = ttci, 
                       fld_name='country')

##Duplicated regions weighted mean
weight_data <- data.frame(country = c("China", "Hong Kong SAR"),
                          population = c(1439323774, 7496988))

ttci_rgn <- ttci_rgn %>%
  arrange(country) %>%
  left_join(weight_data, by = "country") %>%
  mutate(population = ifelse(is.na(population), 1, population)) %>%
  group_by(rgn_id, rgn_name) %>%
  summarize(score = weighted.mean(score, population)) %>%
  select(rgn_id, rgn_name, score)

head(ttci_rgn, 10)

### Save TTCI data file
write.csv(ttci_rgn, file.path(here(), 'globalprep/tr/v2020/intermediate/wef_ttci.csv'), row.names = FALSE)

```

## Preparing the gdppcppp data:
These data are used to gapfill missing values in tourism sustainability.  Most of the data are from the World Bank, but CIA data fill some gaps (CIA data is available for only the most recent year).

The Artisanal Opportunities goal uses gdppcppp data, so we will get the data that was processed for that goal.


```{r worldbank, eval=FALSE}
wb <- read.csv("../../ao/v2020/intermediate/gdppcppp_ohi.csv") %>%
  dplyr::select(rgn_id, year, value)

```

CIA data are used to fill in missing gaps in the gdppcppp data (https://www.cia.gov/library/publications/the-world-factbook/rankorder/2004rank.html)

Downloaded: 04/01/2020

See README on the raw folder for instructions on how to download this data. 

The following code is used to prepare these data for OHI:

```{r cia gdp, eval=FALSE}

cia_gdp <- read.delim('raw/cia_gdp_pc_ppp.txt', stringsAsFactors = FALSE, header = FALSE) %>% 
  separate(V1, c("number", "other"), sep = " ", extra = "merge") %>% 
  separate(other, c("country", "gdppcppp"), sep = "\\$") %>%
  mutate(country = str_trim(country, side = "both")) %>% 
  mutate(country = noquote(country)) %>% 
  mutate(country = as.character(country)) %>% 
  mutate(gdppcppp = as.numeric(gsub(",", "", gdppcppp))) %>% 
  select(-number) %>% 
  data.frame()

##Data reported in a lower resolution than OHI regions
splits <- data.frame(country = "Saint Helena, Ascension, and Tristan da Cunha", country2 = c("Saint Helena",                                                                "Ascension","Tristan da Cunha")) %>%
  mutate(country = as.character(country),
         country2 = as.character(country2))

cia_gdp <- cia_gdp %>%
  left_join(splits, by='country') %>%
  mutate(country2 = ifelse(is.na(country2), country, country2)) %>%
  select(country=country2, pcgdp_cia = gdppcppp)


cia_gdp_rgn <- name_2_rgn(df_in = cia_gdp, 
                       fld_name='country')

### Duplicated regions: Collapse regions after weighting by population (regions we include as a single region) - 

population_weights <- data.frame(country = c("Virgin Islands", "Puerto Rico",
                                             "China", "Hong Kong", "Macau",
                                             "Guam", "Northern Mariana Islands"),
                                 population = c(104423, 2860840,
                                         1439323774, 7496988, 649342,
                                         168775, 55144))

cia_gdp_rgn <- cia_gdp_rgn %>%
  left_join(population_weights, by="country") %>%
  mutate(population = ifelse(is.na(population), 1, population)) %>%
  group_by(rgn_id) %>%
  summarize(pcgdp_cia = weighted.mean(pcgdp_cia, population)) %>%
  ungroup() %>%
  filter(rgn_id <= 250) %>%
  select(rgn_id, pcgdp_cia)

write.csv(cia_gdp_rgn, "intermediate/wb_rgn_cia_GDPPCPPP.csv", row.names=FALSE)

```

The following code combines the two gdp datasets and gapfills missing regions using UN georegions.

If there is no World Bank gdppcppp data (pcgdp), the CIA data is used (pcgdp_cia).  The pcgdp2 variable includes both the World Bank and CIA data (with CIA data only used if there is not World Bank data).  The remaining data are estimated using UN geopolitical regions.  Ideally, the mean gdppcppp value is calculated at the r2 scale (gdp_pred_r2) using regions within each class with gdppcppp data.  If there were not enough regions with data at the r2 scale, the average at the r1 scale was used (gdp_pred_r1). The gdp_all variable combines all estimates using the following heirarchy:  World Bank -> CIA -> estimated using mean from r2 UN geopolitical regions -> estimated using mean from r1 UN geopolitical regions.    

```{r gapfill gdp, eval=FALSE}

### world bank gdp data
gdppcppp <- wb %>%
  select(rgn_id, year, pcgdp = value)

### cia gdp data
gdppcppp2 <- read.csv('intermediate/wb_rgn_cia_GDPPCPPP.csv')


### Use WB data, but if missing, use pcgdp_cia.
### combine with UN georegion data
years <- data.frame(year = min(gdppcppp$year):max(gdppcppp$year))

georegions <- ohicore::georegions

regions <- georegions %>%
  left_join(georegion_labels, by = 'rgn_id')

gdp_raw <- merge(years, regions, by=NULL) %>%
   left_join(gdppcppp, by = c('rgn_id', 'year')) %>%
  left_join(gdppcppp2, by = c("rgn_id")) 

## quick compare to make sure the CIA and World Bank data are compatible
plot(gdp_raw$pcgdp[gdp_raw$year==2018], gdp_raw$pcgdp_cia[gdp_raw$year==2018])
abline(0,1, col="red")
# a bit of scatter for a few regions, but overall looks good

gdp_raw <- gdp_raw %>%
  mutate(pcgdp2 = ifelse(is.na(pcgdp), pcgdp_cia, pcgdp))

##Calculating the means across different geopolitical levels (e.g. r2, r1)
gdp_raw <- gdp_raw %>%
  group_by(r2, year) %>%
  mutate(gdp_pred_r2 = mean(pcgdp2, na.rm=TRUE)) %>%
  ungroup() %>%
  group_by(r1, year) %>%
  mutate(gdp_pred_r1 = mean(pcgdp2, na.rm=TRUE)) %>%
  ungroup() 

gdp_raw_gf <- gdp_raw %>%
  mutate(gdp_all = ifelse(is.na(pcgdp2), gdp_pred_r2, pcgdp2)) %>%
  mutate(gdp_all = ifelse(is.na(gdp_all), gdp_pred_r1, gdp_all)) %>%
  mutate(gapfilled = ifelse(is.na(pcgdp2) & !is.na(gdp_all), "gapfilled", NA)) %>%
  mutate(method = ifelse(is.na(pcgdp2) & !is.na(gdp_pred_r2), "UN georegion (r2)", NA)) %>%
  mutate(method = ifelse(is.na(pcgdp2) & is.na(gdp_pred_r2) & !is.na(gdp_pred_r1), "UN georegion (r1)", method)) 

write_csv(gdp_raw_gf, "intermediate/gdp_raw_gf.csv")

gdp_data_gf <- gdp_raw_gf %>%
  select(rgn_id, year, gapfilled, method) 

write_csv(gdp_data_gf, "intermediate/gdp_gf.csv")

gdp_data <- gdp_raw_gf %>%
  select(rgn_id, year, pcgdp = gdp_all)

write_csv(gdp_data, "intermediate/gdp.csv")

```


The final step is gapfilling the Sustainability data using a linear model with gdppcppp and UN geopolitical regions as predictor variables.  

```{r, eval=FALSE}

sust  <- read.csv('intermediate/wef_ttci.csv', stringsAsFactors = FALSE)

### don't need to gapfill data without tourism data:
## Most recent tourism data is 2019.  

ep_gf <- read.csv("output/tr_jobs_pct_tourism.csv") %>%
  filter(year == 2019) %>%
  select(rgn_id, Ep) %>%
  filter(!is.na(Ep))

# gdp dataframe prepared above (World Bank, CIA, and gapfilled gdp data)
gdp_raw_gf <- read.csv("intermediate/gdp_raw_gf.csv", stringsAsFactors = FALSE) %>% 
  filter(year == 2018) %>%
  select(rgn_id, r0_label, r1_label, r2_label, rgn_label, pcgdp, pcgdp_cia, pcgdp2, gdp_all) 

tr_sust <- gdp_raw_gf %>%
           left_join(sust, by = c("rgn_id")) %>%
          left_join(ep_gf, by = c("rgn_id")) %>%  
          rename(S_score = score) %>%
          filter(rgn_id != 213)

### Add gapfill flag variable 
##Reminder:
## pcgdp2: includes both the World Bank and CIA data (with CIA data only used if there is not World Bank data)
## Ep: Proportion of workforce directly employed in tourism
##S_score: tourism sustainability score

tr_sust_gf <- tr_sust %>%
  mutate(gapfilled = ifelse(is.na(S_score) & !is.na(Ep), "gapfilled", NA)) %>%
  mutate(method = ifelse(is.na(S_score) & !is.na(Ep) & is.na(pcgdp2), "lm georegion + gdppcppp, with est. gdppcppp", NA)) %>%
  mutate(method = ifelse(is.na(S_score) & !is.na(Ep) & !is.na(pcgdp2), "lm georegion + gdppcppp", method)) %>%
  select(rgn_id, gapfilled, method)

write.csv(tr_sust_gf, "output/tr_sustainability_gf.csv", row.names=FALSE)

```
  

### Gapfilling
Linear models using gdppcppp and UN geopolitical regions as predictor variables. However if there is no gdppc data we estimate the gdppc using the UN georegions and then used in the linear model to gapfill the sustainability score.

```{r, eval=FALSE}

### Gapfill S using r1 and/or r2 regional data and PPP-adjusted per-capita GDP
### Looked at models with a year variable, but wasn't significant and decided to exclude

mod3 <- lm(S_score ~ as.factor(r2_label) + gdp_all, data=tr_sust, na.action = na.exclude)
summary(mod3)
anova(mod3)

mod4 <- lm(S_score ~ as.factor(r1_label) + gdp_all, data=tr_sust, na.action = na.exclude)
summary(mod4)
anova(mod4)

plot(predict(mod3), tr_sust$S_score)
abline(0,1)
plot(predict(mod4), tr_sust$S_score)
abline(0,1)


## Estimate missing data and gapfill
# Some of the r1 levels do not have data and consequently causes a fail. This chink of code  drop these levels so an NA is returned

# Select only r2 column
new_data <- tr_sust %>% 
  dplyr::select(r2_label, gdp_all)

unique(tr_sust$r2_label)

r2_w_data <- unique(tr_sust$r2_label[!is.na(tr_sust$S_score)])
  
new_data_r2 <- new_data %>%
  mutate(r2_label = ifelse(r2_label %in% r2_w_data, r2_label, NA))

# Predict sustainability scores using linear model 3 (using r2 data)
tr_sust <- tr_sust %>% 
  dplyr::mutate(S_score_pred_r2 = predict(mod3, newdata = new_data_r2))


# Select only r1 column
new_data <- tr_sust %>% 
  dplyr::select(r1_label, gdp_all)

unique(tr_sust$r1_label)

r1_w_data <- unique(tr_sust$r1_label[!is.na(tr_sust$S_score)])

new_data_r1 <- new_data %>%
  mutate(r1_label = ifelse(r1_label %in% r1_w_data, r1_label, NA))

# Predict sustainability scores using linear model 4 (using r1 data)
tr_sust <- tr_sust %>% 
  dplyr::mutate(S_score_pred_r1 = predict(mod4, newdata = new_data_r1))



## some are missing the r1 predictions, but none of these have Ep scores, so not relevant
filter(tr_sust, is.na(S_score_pred_r1))

tr_sust <- tr_sust %>%
  mutate(S_score_2 = ifelse(is.na(S_score), S_score_pred_r2, S_score)) %>%
  mutate(S_score_2 = ifelse(is.na(S_score_2), S_score_pred_r1, S_score_2)) %>%
  mutate(year = '2019') %>%
#  filter(!is.na(Ep)) %>%
  select(rgn_id, year, S_score=S_score_2)

summary(tr_sust)

write_csv(tr_sust, "output/tr_sustainability.csv")
```


## Compare with previous year of data 

```{r, eval=FALSE}
tr_sust <- read_csv("output/tr_sustainability.csv")

compare <- read.csv("../v2018/output/tr_sustainability.csv") %>% # change year
  select(rgn_id, old_S_score = S_score) %>%
  left_join(tr_sust, by = "rgn_id") %>%
  mutate(dif = old_S_score - S_score)

plot(compare$S_score, compare$old_S_score)
abline(0, 1, col="red")
#looks reasonable 

```


# Tw: Travel warnings

 - Travel warnings were deleted from the v2020 assessment. 

