---
title: "OHI 2022: Crop nutrient fertilizer raters part 2: fubc 2017 wrangling"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---

## Description
FUBC rates for 2014/15 were extracted from Heffer et al. (2017, FUBC rates provided by request) and harmonised with our study’s crop categories. 

## Data Source

**Reference**: [ifa 2017 fubc](https://www.fertilizer.org/images/Library_Downloads/2017_ifa_AgCom_17_134%20rev_fubc%20assessment%202014.pdf); IFA FUBC 2017 data for 2014/15. Placeholder dataset for 2021 report dataset.

**Description**: Fertilizer use by crop (FUBC) rates from the International Fertilizer Association’s (IFA) FUBC series. The FUBC series provides national application rates, measured in kilograms per hectare, of N, P2O5, and K2O for all crop categories.

**Downloaded**: 4/14/2021

## Setup
```{r setup, include = FALSE}
# getting packages we want
library(here)
library(janitor)
library(vroom)



source(here('workflow/R/common.R'))

crop_nutrient <- file.path(here(), "globalprep/prs_land-based_nutrient/v2021/STEP1_crop")

region_data()

food_rgns <- read_csv(here("globalprep/spatial/v2021/food_rgns.csv"), col_types = "cdc")
``` 

## Methods


### Wrangling

```{r}
# Read-in ifa 2014/15 data
fubc_2017 <- 
  vroom(here(crop_nutrient, "raw/fubc_2017_rates.csv"),
        skip = 2,
        .name_repair = make_clean_names) %>% 
  remove_empty(., which = c("rows", "cols"), quiet = TRUE)

colnames(fubc_2017)[1:2] <- c("fubc_2017_country", "nutrient")

fubc_2017 <- 
  fubc_2017 %>%
  mutate(fubc_2017_country = case_when(
    !is.na(fubc_2017_country) ~ fubc_2017_country,
    (is.na(fubc_2017_country) & !is.na(lag(fubc_2017_country, n = 1L))) ~ lag(fubc_2017_country, n = 1L),
    (is.na(fubc_2017_country) & !is.na(lag(fubc_2017_country, n = 2L))) ~ lag(fubc_2017_country, n = 2L),
    (is.na(fubc_2017_country) & !is.na(lag(fubc_2017_country, n = 3L))) ~ lag(fubc_2017_country, n = 3L))) %>% 
  filter(!nutrient %in% "N+P+K") %>% 
  mutate_all(funs(str_replace_all(., "#DIV/0!", NA_character_))) %>% 
  mutate_at(3:17, funs(as.numeric)) %>% 
  filter(!fubc_2017_country == "World")

# Gather crops.
fubc_2017 <- 
  fubc_2017 %>% 
  gather(fubc_2017_crop, 
         rate, 
         -c(fubc_2017_country, nutrient)) 

fubc_2017 <- 
  fubc_2017 %>% 
  group_by(fubc_2017_crop, nutrient) %>% 
  group_modify(~ mutate(.x, rate = if_else(is.na(rate),
                                       .x$rate[.x$fubc_2017_country == "ROW"],
                                       rate)))
```

-------------------------------------------------------------------------------

### CROPS

#### import crop name datasets
```{r}
crop_categories <- 
  vroom(here(crop_nutrient, "raw/crop_codes_updated.csv")) %>% 
  dplyr::select(split_rest_names) %>% 
  unique()

crop_codes <- 
  vroom(here(crop_nutrient, "raw/crop_codes_updated.csv")) %>% 
  dplyr::select(split_rest_names, item_code, SPAM_full_name) %>% 
  unique()

crop_defs <- 
  vroom(here(crop_nutrient, "raw/FAO_crop_definitions.csv"), 
        .name_repair = make_clean_names)
```

#### Match IFA crop categories with item_codes
```{r}
crop_match <- 
  tibble(fubc_2017_crop = c(rep("oth_cereals", 5),
                            rep("fiber_crops", 2),
                            rep("roots_tubers", 5),
                            rep("fruits", 5),
                            "vegetables",
                            rep("sugar_crops", 2),
                            rep("oth_oilseeds", 5),
                            "soybeans",
                            "oil_palm",
                            "rice",
                            "maize",
                            "wheat"),
       split_rest_names = c("barl", "ocer", "pmil", "smil", "sorg", #oth_ce
                            "ofib", "cott",# fibre
                            "pota", "cass","swpo", "yams", "orts", #roots_tubers
                            "temf", "trof", "bana", "plnt", "cnut", # fruits
                            "vege", # veg
                            "sugb", "sugc",# sugar
                            "sesa", "rape", "grou", "ooil", "sunf", # oth_os
                            "soyb",  "oilp", "rice", "maiz", "whea")) %>% #grassland
  left_join(., crop_categories)
crop_match

write_csv(crop_match,
          here(crop_nutrient, "int/fubc_2017_crop_list.csv"))
```

#### Match crop names
```{r}
# Match names
fubc_2017 <- 
  left_join(fubc_2017, crop_match) %>%
  na.omit() %>% 
  dplyr::select(-fubc_2017_crop)
```

-------------------------------------------------------------------------------

### COUNTRIES

#### Match iso3c codes with fubc_2017_country names

```{r}
other_fubc_countries <- 
  unique(fubc_2017$fubc_2017_country)[unique(fubc_2017$fubc_2017_country) %in% c("EU-27", "USA", "New-Zealand", "ROW", "World", "China") == FALSE]

EU_countries <- 
  c("Austria", "Belgium", "Bulgaria", "Croatia", "Cyprus", "Czech Republic",
    "Denmark","Estonia","Finland","France","Germany","Greece","Hungary",
    "Ireland","Italy","Latvia","Lithuania","Luxembourg","Malta","Netherlands",
    "Poland","Portugal","Slovakia","Slovenia","Spain","Sweden","Romania",
    "United Kingdom","Guernsey","Faroe Islands","Gibraltar","Åland",
    "Svalbard and Jan Mayen")

china_countries <- c("China", "Macao", "Hong Kong")

rest_of_the_world <- 
  food_rgns$Country[!food_rgns$Country %in% c(other_fubc_countries, 
                                              china_countries, 
                                              EU_countries,
                                              "United States",
                                              "New Zealand")]

# Renaming 
country_match <- 
  tibble(fubc_2017_country = c(rep("EU-27", 33),
                               rep("China", 3),
                               "USA",
                               "New-Zealand",
                               other_fubc_countries,
                               rep("ROW", length(rest_of_the_world))),
         Country = c(EU_countries,
                     china_countries,
                     "United States",
                     "New Zealand",
                     other_fubc_countries,
                     rest_of_the_world))

# Merge with country names list and populate iso3c for ROW
setdiff(fubc_2017$fubc_2017_country, country_match$fubc_2017_country)
setdiff(country_match$fubc_2017_country, fubc_2017$fubc_2017_country)
```

#### Match country names
```{r}
fubc_2017 <- 
  left_join(fubc_2017, country_match) %>% 
  left_join(., food_rgns) %>% 
  dplyr::select(-Country, -fubc_2017_country, -ID_0)
unique(fubc_2017$iso3c)
```

### Check

```{r}
244 * 3 * 30 == 
  length(unique(fubc_2017$split_rest_names)) *
  length(unique(fubc_2017$iso3c)) * 
  length(unique(fubc_2017$nutrient)) 

244 * 3 * 30 == nrow(fubc_2017)
```

## Save

```{r}
write_csv(fubc_2017, here(crop_nutrient, "int/fubc_2017_gf.csv"))
```

