---
title: "OHI 2022: STEP7: Combine all leaching sources and all volatization sources"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---

## Summary 
Combines all the sources of manure volatization and leaching for each livestock system 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(doParallel)
library(tidyverse)
library(readr)
library(janitor)

nutrient_d2021_anx <- file.path("/home/shares/ohi/git-annex/globalprep/_raw_data/FAOSTAT/manure_nutrient/d2021/")
git_manure <- file.path(here(), "globalprep/prs_land-based_nutrient/v2021/STEP2_manure")
location_tifs_path <- file.path("/home/shares/ohi/git-annex/globalprep/_raw_data/halpern_food_systems/d2021/animal_farm/location_tifs")
prep <- file.path("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2021")
git_crop <- file.path(here(), "globalprep/prs_land-based_nutrient/v2021/STEP1_crop")

source(here(git_crop, "R/common_spatial.R")) # read in necessary spatial files 

un <- read_csv(here("globalprep/spatial/v2021/UNSD_Methodology.csv")) %>% 
  dplyr::select(iso3c, Global_Name, Region_Name, Sub_region_Name, Intermediate_Region_Name)
```

## All Leaching
```{r}

category_list <- c("chickens_industrial_meat", "chickens_industrial_eggs", "chickens_backyard_eggs&meat", "goats_grassland_milk", "goats_grassland_meat", "goats_mixed_milk", "goats_mixed_meat", "sheep_grassland_milk", "sheep_grassland_meat", "sheep_mixed_milk", "sheep_mixed_meat", "cows_grassland_milk", "cows_grassland_meat", "cows_mixed_milk", "cows_mixed_meat", "cows_feedlot_meat", "pigs_backyard_meat", "pigs_intermediate_meat", "pigs_industrial_meat", "buffaloes_grassland_milk", "buffaloes_mixed_milk")

year_list <- c(2005:2019)


registerDoParallel(4)   

foreach(category = category_list) %:% 
  foreach(yr = year_list) %dopar% {
  
  # category = category_list[13]
  # yr = 2017
  
  ## Grab the leaching N from manure first
 # N_excretion <- raster(file.path(paste(prep, "/animal_farm_int/N_from_excretion_manure/", category, "_N_excretion_", yr, ".tif", sep = "")))
  # plot(log(N_excretion + 1))
  
  N_excretion <- raster(file.path(paste(prep, "/animal_farm_int/N_from_excretion_manure/", category, "_N_excretion_", yr, ".tif", sep = ""))) %>% 
    raster_df() %>% 
    dplyr::select(-cellindex) 
 colnames(N_excretion) <- c("x", "y", "exc") 
 N_excretion <- N_excretion %>% 
   mutate(exc = ifelse(is.na(exc), 0, exc)) %>%
     arrange(x,y) %>%
  mutate(cell_id = 1:9331200) # create unique cell_id to join by 
  
  mms_prop <- read_csv(list.files(file.path(prep, "animal_farm_int/STEP3_N_leaching_prop_mms"), pattern = category, full = TRUE), col_types = c("ddcn")) %>% 
    mutate(frac_leach_mms = ifelse(is.na(frac_leach_mms), 0, frac_leach_mms)) %>%
      arrange(x,y) %>%
  mutate(cell_id = 1:9331200) # create unique cell_id to join by 


  applied_crop_prop <- read_csv(list.files(file.path(prep, "animal_farm_int/STEP5_N_leached_prop_applied_crops"), pattern = category, full = TRUE), col_types = c("ddcn")) %>% 
    mutate(N_prop_leach_applied = ifelse(is.na(N_prop_leach_applied), 0, N_prop_leach_applied)) %>%
      arrange(x,y) %>%
  mutate(cell_id = 1:9331200) # create unique cell_id to join by 

  
    applied_pasture_prop <- read_csv(list.files(file.path(prep, "animal_farm_int/STEP5.5_N_leached_prop_applied_pastures"), pattern = category, full = TRUE), col_types = c("ddcn")) %>% 
    mutate(N_prop_leach_applied_pastures = ifelse(is.na(N_prop_leach_applied_pastures), 0, N_prop_leach_applied_pastures)) %>%
        arrange(x,y) %>%
  mutate(cell_id = 1:9331200) # create unique cell_id to join by 
    
  
  pasture_prop <- read_csv(list.files(file.path(prep, "animal_farm_int/STEP6.5_N_leaching_prop_pasture"), pattern = category, full = TRUE), col_types = c("ddcn")) %>%  
    mutate(prop_leaching = ifelse(is.na(prop_leaching), 0, prop_leaching)) %>%
    arrange(x,y) %>%
  mutate(cell_id = 1:9331200) # create unique cell_id to join by 
  
  tonnes_N_leached_manure <- left_join(mms_prop, pasture_prop, by = "cell_id") %>% 
    left_join(applied_crop_prop, by = "cell_id") %>% 
    left_join(applied_pasture_prop, by = "cell_id") %>% 
    left_join(N_excretion, by = "cell_id") %>% 
    mutate(prop_leach_all = (frac_leach_mms + N_prop_leach_applied + N_prop_leach_applied_pastures + prop_leaching),
           leached_amount = exc*prop_leach_all) %>% 
    dplyr::select(x,y, leached_amount) # check selection; good
  
 tonnes_N_leached_manure_raster <- rasterFromXYZ(tonnes_N_leached_manure, crs = food_crs)
 
 # plot(log(tonnes_N_leached_manure_raster+1))

 #  tonnes_N_leached <- N_excretion *(mms_prop +pasture_prop + applied_prop)
  
   names(tonnes_N_leached_manure_raster) <- "tonnes_N"
    tonnes_N_leached_manure_raster[is.na(tonnes_N_leached_manure_raster)] <- 0
  
  ## Now lets grab the grassland fertilizer leached if relevant
    if(str_detect(category, "grassland")){
      
    fertilizer_leached <- raster(file.path(prep, paste0("animal_farm_int/grassland_nutrient_pollution/", category, "_leached_N_", yr, ".tif", sep = "")))
   #  plot(log(fertilizer_leached+1))
    tonnes_N_leached_raster <- sum(stack(tonnes_N_leached_manure_raster,fertilizer_leached), na.rm = TRUE)
    
    }else{
      tonnes_N_leached_raster <- tonnes_N_leached_manure_raster 
      }
    
    png(filename=file.path(paste0(git_manure, "/plot_check/", category, "_leached_", yr, ".png")))
        plot(log(1+ tonnes_N_leached_raster), main = paste0(category, yr, "_leached"))
    dev.off()

  
  raster::writeRaster(tonnes_N_leached_raster, file.path(prep, paste0("animal_farm_int/STEP7_total_N_leaching/", category, "_N_leaching_", yr, ".tif", sep = "")), format = "GTiff", overwrite = TRUE)
  
}


```

Numbers check

```{r}
## check 2017
yrs = c(2005:2019)
for(yr in yrs){
list_files <- list.files(file.path(prep, "animal_farm_int/STEP7_total_N_leaching"), full = TRUE, pattern = paste0(yr))

global_leaching_total <- sum(stack(lapply(list_files, raster)))
plot(log(1+global_leaching_total))
total_leach <- cellStats(global_leaching_total, stat = 'sum', na.rm = TRUE)
total_leach # 36328420

df_check <- data.frame(animal=NA,
                       global_n = NA)

animals_list <- c("cows", "chickens", "goats", "pigs", "buffaloes", "sheep")
for(animal_oi in animals_list) {

  animal_files <- str_subset(list_files, pattern = animal_oi)
  leaching <- sum(stack(lapply(animal_files, raster)), na.rm = TRUE)
  
  total <- as_tibble(cellStats(leaching, stat = 'sum', na.rm = TRUE)) %>% 
    mutate(animal = animal_oi) %>% 
    rename(global_n = value)
  
  df_check <- rbind(df_check, total) %>%  filter(!is.na(animal))
  
}

## check to see if the total leaching is 20-30%ish of total N excretion

list_files <- list.files(file.path(prep, "animal_farm_int/N_from_excretion_manure"), full = TRUE, pattern = paste0(yr))

global_N_total <- sum(stack(lapply(list_files, raster)), na.rm = TRUE)
total_exc <- cellStats(global_N_total, stat = 'sum', na.rm = TRUE)

print(total_leach/total_exc)
## 0.1897774 # woohooooo 

}
```

Look at all the USA cows 
```{r}

applied_crop <- read_csv(here::here(git_manure, "data_check/usa_cows_leached_applied_crops.csv")) %>% 
  rename(prop_leach_applied_crop = prop_applied_leaching)
applied_past <- read_csv(here::here(git_manure, "data_check/usa_cows_leached_applied_pastures.csv")) %>% 
    rename(prop_leach_applied_pasture = prop_applied_leaching)
mms <- read_csv(here::here(git_manure, "data_check/usa_cows_leached_mms.csv"))
 pasture <- read_csv(here::here(git_manure, "data_check/usa_cows_leached_pasture.csv"))

usa_cows <- left_join(applied_crop, mms) %>%
  left_join(pasture) %>%
  left_join(applied_past) %>% 
  select(iso3c, animal_system_product,
         prop_leach_mms,
         prop_leach_pasture = prop_pasture_leaching,
         prop_leach_applied_crop,
         prop_leach_applied_pasture)
```
 

## All Volatization
```{r}

category_list <- c("chickens_industrial_meat", "chickens_industrial_eggs", "chickens_backyard_eggs&meat", "goats_grassland_milk", "goats_grassland_meat", "goats_mixed_milk", "goats_mixed_meat", "sheep_grassland_milk", "sheep_grassland_meat", "sheep_mixed_milk", "sheep_mixed_meat", "cows_grassland_milk", "cows_grassland_meat", "cows_mixed_milk", "cows_mixed_meat", "cows_feedlot_meat", "pigs_backyard_meat", "pigs_intermediate_meat", "pigs_industrial_meat", "buffaloes_grassland_milk", "buffaloes_mixed_milk")

year_list <- c(2005:2019)

registerDoParallel(5)   

foreach(category = category_list) %:% 
  foreach(yr = year_list) %dopar% {
  
  # category = category_list[13]
  # yr = 2017
    

  ## lets deal with the manure nutrient volatization
  N_excretion <-  raster(file.path(paste(prep, "/animal_farm_int/N_from_excretion_manure/", category, "_N_excretion_", yr, ".tif", sep = ""))) %>%
    raster_df() %>% 
    dplyr::select(-cellindex) 
  
 colnames(N_excretion) <- c("x", "y", "exc") 
 
 N_excretion <- N_excretion %>% 
   mutate(exc = ifelse(is.na(exc), 0, exc)) %>%
          arrange(x,y) %>%
  mutate(cell_id = 1:9331200) # create unique cell_id to join by 
 
  mms_volt_prop <- read_csv(list.files(file.path(prep, "animal_farm_int/STEP3.5_N_volatizing_prop_mms"), pattern = category, full = TRUE), col_types = c("ddcn")) %>% 
    mutate(prop_volt = ifelse(is.na(prop_volt), 0, prop_volt)) %>%
              arrange(x,y) %>%
  mutate(cell_id = 1:9331200) # create unique cell_id to join by 
  
  applied_volt_crop_prop <- read_csv(list.files(file.path(prep, "animal_farm_int/STEP5_N_volatizing_prop_applied_crops"), pattern = category, full = TRUE), col_types = c("ddcn")) %>% 
    mutate(N_prop_volt_applied = ifelse(is.na(N_prop_volt_applied), 0, N_prop_volt_applied)) %>%
              arrange(x,y) %>%
  mutate(cell_id = 1:9331200) # create unique cell_id to join by 
  
    applied_volt_pasture_prop <- read_csv(list.files(file.path(prep, "animal_farm_int/STEP5.5_N_volatizing_prop_applied_pasture"), pattern = category, full = TRUE), col_types = c("ddcn")) %>% 
    mutate(N_prop_volt_applied_pastures = ifelse(is.na(N_prop_volt_applied_pastures), 0, N_prop_volt_applied_pastures)) %>%
                arrange(x,y) %>%
  mutate(cell_id = 1:9331200) # create unique cell_id to join by 
  
  pasture_volt_prop <- read_csv(list.files(file.path(prep, "animal_farm_int/STEP6.5_N_volatizing_prop_pasture"), pattern = category, full = TRUE), col_types = c("ddcn")) %>% 
    rename(prop_volt_pasture = prop_volt) %>% 
    mutate(prop_volt_pasture = ifelse(is.na(prop_volt_pasture), 0, prop_volt_pasture)) %>%
              arrange(x,y) %>%
  mutate(cell_id = 1:9331200) # create unique cell_id to join by 
  
  
  tonnes_N_volatized <- left_join(mms_volt_prop, pasture_volt_prop, by = "cell_id") %>% 
    left_join(applied_volt_crop_prop, by = "cell_id") %>% 
    left_join(applied_volt_pasture_prop, by = "cell_id") %>% 
    left_join(N_excretion, by = "cell_id") %>% 
    mutate(prop_volt_all = (prop_volt + N_prop_volt_applied + N_prop_volt_applied_pastures + prop_volt_pasture),
           volt_amount = exc*prop_volt_all) %>% 
    select(x,y, volt_amount)
  
 tonnes_N_manure_volatized_raster <- rasterFromXYZ(tonnes_N_volatized, crs = food_crs)
 
 # plot(log(tonnes_N_manure_volatized_raster+1))
 
    names(tonnes_N_manure_volatized_raster) <- "tonnes_N"
    tonnes_N_manure_volatized_raster[is.na(tonnes_N_manure_volatized_raster)] <- 0
    
    ## now lets deal with the volatization from grassland fertlizer
      ## Now lets grab the grassland fertilizer leached if relevant
    if(str_detect(category, "grassland")){
      
    fertilizer_volt <- raster(file.path(prep, paste0("animal_farm_int/grassland_nutrient_pollution/", category, "_volatilized_N_", yr, ".tif", sep = "")))
    
    
    tonnes_N_volatized_raster <- sum(stack(tonnes_N_manure_volatized_raster,fertilizer_volt), na.rm = TRUE)
    
    }else{
      tonnes_N_volatized_raster <- tonnes_N_manure_volatized_raster}
    
    
        png(filename=file.path(paste0(git_manure, "/plot_check/volt/", category, "_volt_", yr, ".png")))
        plot(log(1+ tonnes_N_volatized_raster), main = paste0(category, yr, "_volt"))
    dev.off()
  
  raster::writeRaster(tonnes_N_volatized_raster, file.path(prep, paste0("animal_farm_int/STEP7_total_N_volatizing/", category, "_N_volatizing_", yr, ".tif", sep = "")), format = "GTiff", overwrite = TRUE)
  
  }


## check 2017 totals; amount of volatization should be anywhere from 30-40%

list_files_1 <- list.files(file.path(prep, "animal_farm_int/N_from_excretion_manure"), full = TRUE, pattern = "2017")

global_N_total <- sum(stack(lapply(list_files_1, raster)), na.rm = TRUE)
total_exc <- cellStats(global_N_total, stat = 'sum', na.rm = TRUE)

list_files <- list.files(file.path(prep, "animal_farm_int/STEP7_total_N_volatizing/"), full = TRUE, pattern = "2017")

global_total_volt <-  sum(stack(lapply(list_files, raster)))
total_volt <- cellStats(global_total_volt, stat = 'sum', na.rm = TRUE)
total_volt/total_exc
## 0.3460552 ; from global food project
## 0.3620408 ; OHI2021 with newer FAO data

```


Calculate the totals for each system for leaching 
```{r}
raster_list <- list.files(file.path(prep,"animal_farm_int/STEP7_total_N_leaching/"), full.names = TRUE)

df_check <- data.frame(iso3c = NA, animal = NA, system = NA, product = NA, total_N_leached = NA, year = NA)
  
#registerDoParallel(4)
for(rast in raster_list) {
  
#  rast <- raster_list[21]
    ## grab the names of all the parts from the file
    animal_name <- str_split(str_split(rast, pattern = "//")[[1]][2], pattern = "_")[[1]][1]
    system_name <- str_split(str_split(rast, pattern = "//")[[1]][2], pattern = "_")[[1]][2]
    product_name <- str_split(str_split(rast, pattern = "//")[[1]][2], pattern = "_")[[1]][3]
    yr = str_split(str_split(rast, pattern = "//")[[1]][2], pattern = "_")[[1]][6]
    yr = gsub(".tif", "", yr)
     
  ## values are the same throughout the country so a median or mean function won't make a difference
    country_sum <- zonal(raster(rast), food_rgns_tif, fun="sum", progress="text", na.rm=TRUE) %>% 
    data.frame() %>%
    rename(ID_0 = zone, total_N_leached = sum) %>%
    left_join(food_rgns, by="ID_0") %>% 
      mutate(animal = animal_name,
             system = system_name,
             product = product_name, 
             year = yr) %>% 
       select(iso3c,total_N_leached, animal, system, product, year)
  
    df_check <- rbind(df_check, country_sum) %>% 
      filter(!is.na(iso3c)) 
    
    cat_msg("finished",   match(rast, raster_list), "out of", length(raster_list))
}


global_check <- df_check %>% 
  group_by(animal, system, product, year) %>% 
  dplyr::summarise(total_N_leached = sum(total_N_leached)) %>% 
  ungroup()

write_csv(global_check, here::here(git_manure, "data_check/leaching_global_check.csv"))
## look at it

ggplot(df_check, aes(x = animal, y = total_N_leached, color = system) ) +
  geom_point(position = "jitter") +
  labs(title = "Total N Leached")
```


Calculate the totals for each system for volatized
```{r}
raster_list <- list.files(file.path(prep,"animal_farm_int/STEP7_total_N_volatizing/"), full.names = TRUE)

df_check <- data.frame(iso3c = NA, animal = NA, system = NA, product = NA, total_N_volt = NA, year = NA)
  
for(rast in raster_list){
  
#  rast <- raster_list[20]
    ## grab the names of all the parts from the file
    animal_name <- str_split(str_split(rast, pattern = "//")[[1]][2], pattern = "_")[[1]][1]
    system_name <- str_split(str_split(rast, pattern = "//")[[1]][2], pattern = "_")[[1]][2]
    product_name <- str_split(str_split(rast, pattern = "//")[[1]][2], pattern = "_")[[1]][3]
    yr = str_split(str_split(rast, pattern = "//")[[1]][2], pattern = "_")[[1]][6]
    yr = gsub(".tif", "", yr)
     
  ## values are the same throughout the country so a median or mean function won't make a difference
    country_sum <- zonal(raster(rast), food_rgns_tif, fun="sum", progress="text", na.rm=TRUE) %>% 
    data.frame() %>%
    rename(ID_0 = zone, total_N_volt = sum) %>%
    left_join(food_rgns, by="ID_0") %>% 
      mutate(animal = animal_name,
             system = system_name,
             product = product_name,
             year = yr) %>% 
       select(iso3c,total_N_volt, animal, system, product, year)
  
    df_check <- rbind(df_check, country_sum) %>% 
      filter(!is.na(iso3c)) 
    
    cat_msg("finished",   match(rast, raster_list), "out of", length(raster_list))
    
}


global_check <- df_check %>% 
  group_by(animal, system, product, year) %>% 
  dplyr::summarise(total_N_volt = sum(total_N_volt)) %>% 
  ungroup()

write_csv(global_check, here::here(git_manure, "data_check/volatized_global_check.csv"))
## look at it

ggplot(df_check, aes(x = animal, y = total_N_volt, color = system) ) +
  geom_point(position = "jitter") +
  labs(title = "Total N volatized")
```


Calculate the proportion of total N excreted that is leached and also volatized by animal-system-product 

## proportion of n excreted per animal system that volatized
```{r}

category_list <- sort(c("chickens_industrial_meat", "chickens_industrial_eggs", "chickens_backyard_eggs&meat", "goats_grassland_milk", "goats_grassland_meat", "goats_mixed_milk", "goats_mixed_meat", "sheep_grassland_milk", "sheep_grassland_meat", "sheep_mixed_milk", "sheep_mixed_meat", "cows_grassland_milk", "cows_grassland_meat", "cows_mixed_milk", "cows_mixed_meat", "cows_feedlot_meat", "pigs_backyard_meat", "pigs_intermediate_meat", "pigs_industrial_meat", "buffaloes_grassland_milk", "buffaloes_mixed_milk"))

years <- sort(c(2005:2019))

volt_n_check <- data.frame(food_system = NA, N_excreted = NA, N_volatized = NA, prop_volatized=NA, year = NA)
  

for(category in category_list){
  for(yr in years)  {
  
  # category = category_list[13] 
  # yr = 2005  
  
  ## lets deal with the manure nutrient volatization
  N_excretion <- raster(file.path(paste0(prep, "/animal_farm_int/N_from_excretion_manure/", category, "_N_excretion_", yr, ".tif"))) %>% 
    raster_df() %>% 
    select(-cellindex) 
 colnames(N_excretion) <- c("x", "y", "exc") 
 
 
 N_excretion_global <- N_excretion %>% 
   dplyr::summarise(N_excreted = sum(exc, na.rm = TRUE))
   # buffaloes_grassland_milk_N_volatizing_2005.tif
  N_volat <- raster(file.path(paste0(prep, "/animal_farm_int/STEP7_total_N_volatizing/", category, "_N_volatizing_", yr, ".tif"))) %>% 
    raster_df() %>% 
    select(-cellindex)  
  colnames(N_volat) <- c("x", "y", "N_volt") 
  
  N_volat_global <- N_volat %>% 
   dplyr::summarise(N_volatized = sum(N_volt, na.rm = TRUE))
  
  df_total <- cbind(N_excretion_global, N_volat_global) %>% 
    mutate(food_system = category,
           year = yr, 
           prop_volatized = N_volatized/N_excreted) 
  
 volt_n_check  <- rbind(volt_n_check, df_total)
 
 
  N_excretion_list <- list.files(file.path(prep, "animal_farm_int/N_from_excretion_manure/"),  full = TRUE)
 
 matched <- match(file.path(paste0(prep, "/animal_farm_int/N_from_excretion_manure//", category, "_N_excretion_", yr, ".tif")), N_excretion_list)
 
     cat_msg("finished",   matched, "out of", length(N_excretion_list))

 
  }
}

 write_csv(volt_n_check, here::here(git_manure, "data_check/volatized_prop_global_check.csv"))

```

## proportion of n excreted per animal system that leached
```{r}

category_list <- sort(c("chickens_industrial_meat", "chickens_industrial_eggs", "chickens_backyard_eggs&meat", "goats_grassland_milk", "goats_grassland_meat", "goats_mixed_milk", "goats_mixed_meat", "sheep_grassland_milk", "sheep_grassland_meat", "sheep_mixed_milk", "sheep_mixed_meat", "cows_grassland_milk", "cows_grassland_meat", "cows_mixed_milk", "cows_mixed_meat", "cows_feedlot_meat", "pigs_backyard_meat", "pigs_intermediate_meat", "pigs_industrial_meat", "buffaloes_grassland_milk", "buffaloes_mixed_milk"))

years <- c(2005:2019)

n_leach_check <- data.frame(food_system = NA, N_excreted = NA, N_leached = NA, prop_leached=NA, year = NA)
  

for(category in category_list) {
  for(yr in years) {
  
  # category = category_list[12]
  
  ## lets deal with the manure nutrient volatization
  N_excretion <- raster(file.path(paste0(prep, "/animal_farm_int/N_from_excretion_manure/", category, "_N_excretion_", yr, ".tif"))) %>% 
    raster_df() %>% 
    select(-cellindex) 
 colnames(N_excretion) <- c("x", "y", "exc") 
 
 N_excretion_global <- N_excretion %>% 
   dplyr::summarise(N_excreted = sum(exc, na.rm = TRUE))
   
   N_leach <- raster(file.path(paste0(prep, "/animal_farm_int/STEP7_total_N_leaching/", category, "_N_leaching_", yr, ".tif")))  %>% 
    raster_df() %>% 
    select(-cellindex)  
  colnames(N_leach) <- c("x", "y", "N_leached") 
  
  N_leach_global <- N_leach %>% 
   dplyr::summarise(N_leached = sum(N_leached, na.rm = TRUE))
  
  df_total <- cbind(N_excretion_global, N_leach_global) %>% 
    mutate(food_system = category,
           year = yr,
           prop_leached = N_leached/N_excreted) 
  
 n_leach_check  <- rbind(n_leach_check, df_total)
 
   N_excretion_list <- list.files(file.path(prep, "animal_farm_int/N_from_excretion_manure/"),  full = TRUE)
 
 matched <- match(file.path(paste0(prep, "/animal_farm_int/N_from_excretion_manure//", category, "_N_excretion_", yr, ".tif")), N_excretion_list)
 
     cat_msg("finished",   matched, "out of", length(N_excretion_list))
 

 
  }
}


   write_csv(n_leach_check, here::here(git_manure, "data_check/leached_n_prop_global_check.csv"))

```


## proportion of np excreted per animal system that leached
```{r}

category_list <- sort(c("chickens_industrial_meat", "chickens_industrial_eggs", "chickens_backyard_eggs&meat", "goats_grassland_milk", "goats_grassland_meat", "goats_mixed_milk", "goats_mixed_meat", "sheep_grassland_milk", "sheep_grassland_meat", "sheep_mixed_milk", "sheep_mixed_meat", "cows_grassland_milk", "cows_grassland_meat", "cows_mixed_milk", "cows_mixed_meat", "cows_feedlot_meat", "pigs_backyard_meat", "pigs_intermediate_meat", "pigs_industrial_meat", "buffaloes_grassland_milk", "buffaloes_mixed_milk"))

years <- c(2005:2019)

p_leach_check <- data.frame(food_system = NA, P_excreted = NA, P_leached = NA, prop_leached=NA, year = NA)
  

for(category in category_list) {
  
  for(yr in years){
  
  # category = category_list[12]
  #  yr = 2005
  
  ## lets deal with the manure nutrient volatization
  P_excretion <- raster(file.path(paste0(prep, "/animal_farm_int/P_from_excretion_manure/", category, "_P_excretion_", yr, ".tif"))) %>% 
    raster_df() %>% 
    select(-cellindex) 
 colnames(P_excretion) <- c("x", "y", "exc") 
 
 P_excretion_global <- P_excretion %>% 
   dplyr::summarise(P_excreted = sum(exc, na.rm = TRUE))
 
  p_leach_manure <- raster(file.path(paste0(prep,"/animal_farm_int/STEP2_P_leach/", category, "_", yr, ".tif")))

  if(str_detect(category, "grassland")){
      p_leach_fertilizer <- raster(file.path(paste0(prep,"/animal_farm_int/grassland_nutrient_pollution/", category, "_leached_P_", yr, ".tif")))
 
  }else{
    p_leach_fertilizer <- calc(food_rgns_tif, fun = function(x){ifelse(!is.na(x), 0, x)})
  }
  
 all_P_leach <- sum(stack(p_leach_manure, p_leach_fertilizer), na.rm = TRUE) %>% 
   raster_df() %>% 
    select(-cellindex)  
  colnames(all_P_leach) <- c("x", "y", "P_leached") 
  
  P_leach_global <- all_P_leach %>% 
   dplyr::summarise(P_leached = sum(P_leached, na.rm = TRUE))
  
  df_total <- cbind(P_excretion_global, P_leach_global) %>% 
    mutate(food_system = category,
           year = yr,
           prop_leached = P_leached/P_excreted) 
  
 p_leach_check  <- rbind(p_leach_check, df_total)
 
   P_excretion_list <- list.files(file.path(prep, "animal_farm_int/P_from_excretion_manure/"),  full = TRUE)
 
 matched <- match(file.path(paste0(prep, "/animal_farm_int/P_from_excretion_manure//", category, "_P_excretion_", yr, ".tif")), P_excretion_list)
 
     cat_msg("finished",   matched, "out of", length(P_excretion_list))
 
 
  }
}


   write_csv(p_leach_check, here::here(git_manure, "data_check/leached_p_prop_global_check.csv"))

```


Combine to look at them all together

```{r}
n_leach <- read_csv(here::here(git_manure, "data_check/leached_n_prop_global_check.csv")) %>% rename(prop_n_leached = prop_leached)
p_leach <- read_csv(here::here(git_manure, "data_check/leached_p_prop_global_check.csv"))%>% rename(prop_p_leached = prop_leached)
n_volt <- read_csv(here::here(git_manure, "data_check/volatized_prop_global_check.csv"))

all <- left_join(n_leach, n_volt, by = c("food_system", "N_excreted")) %>% left_join(p_leach) %>% 
  filter(!is.na(food_system)) %>% 
  select(food_system, year, prop_n_leached, prop_volatized, prop_p_leached)
```

