---
title: 'OHI 2021: Kelp extent'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---

# Summary

This script generates the extent of kelp for each OHI region for the latest year of data. We do this using polygon data. We extract the area of each polygon per each ohi region using our combined land and eez raster. 


## Updates from previous assessment
This is an entirely new layer for the 2021 assessment! 

***
## Data Source 

**Reference**: Jayathilake, D.R.M., Costello, M.J., 2020. A modelled global distribution of the kelp biome. Biological Conservation 252, 108815. https://doi.org/10.1016/j.biocon.2020.108815


**Downloaded**: 04/28/2021 

**Description**:  
This is a MaxEnt model map of the global distribution of the
Laminarian kelp biome. This was modelled using 44,265 cleaned
primary occurrence records from the Global Biodiversity
Information Facility (GBIF), and the Ocean Biodiversity
Information System (OBIS) and 13 environmental var

https://data.unep-wcmc.org/pdfs/49/A_Modelled_Global_Distribution_of_the_Kelp_Biome_metadata.pdf?1609770367


**Time range**: 1900 - 2020 


***
# Methods


## Setup
``` {r setup, echo = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
library(readr)      # for read_csv()
library(raster)
library(here)
library(sf)
library(fasterize)
library(tidyverse)
library(mapview)
library(sp)
library(rgeos)

source(file.path('~/github/ohiprep_v2021/workflow/R/common.R'))

goal     <- 'globalprep/hab_kelp/v2021'
dir_git  <- file.path('~/github/ohiprep_v2021', goal)
dir_wcmc <- file.path(file.path(dir_M, 'git-annex/globalprep/_raw_data/wcmc_kelp'))
dir_kelp <- file.path(dir_wcmc, "d2020/UniAuk-004-ModelledDistributionKelpBiome")
ohi_rasters() # call the region zones raster
regions_shape()
region_data()

land_ocean <- raster("/home/shares/ohi/git-annex/globalprep/spatial/v2017/regions_land_ocean.tif")

land_ocean
```

Read in kelp data and explore
```{r, eval = FALSE}

kelp_shp <- st_read(file.path(dir_kelp, "01_Data/Kelp UNEP.shp"))

mapview(head(kelp_shp))

```


Match the data to OHI regions 

```{r, eval = FALSE}
  
kelp_moll <- st_transform(kelp_shp, crs = "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")

fasterize_py <- fasterize::fasterize(kelp_moll, raster = land_ocean, 
                                     field = NULL) # all polygons given value of 1

fasterize_py
plot(fasterize_py)

print(cellStats(fasterize_py, "sum", na.rm=TRUE)*0.934478877011218970**2) # 1,403,760 km2 - paper says 1,469,900 km2 
sum(kelp_moll$Shape_Area) # 1.469908e+12 - we are losing about 60000 km2... this seems reasonable though. 

writeRaster(fasterize_py, filename = file.path(dir_M, "/git-annex/globalprep/hab_kelp/v2021/int/py_kelp_rast.tif"), overwrite=TRUE) 


# Calculate zonal stats with zones raster and new combined kelp. Convert to km^2 and save int/output files*

zonal_sums_combined <- raster::zonal(fasterize_py, 
                                     land_ocean,
                                     fun = "sum",
                                     na.rm = TRUE,
                                     progress="text") #sum all kelp cells for each ohi zone
zonal_sums_combined_df <- data.frame(zonal_sums_combined)

zonal_sums_km2 <- zonal_sums_combined_df %>%
  mutate(year = 2020, habitat = "kelp",
         km2 = (0.934478877011218970**2*sum)) %>% #one cell is equal to ~0.93 km
  rename("rgn_id" = "zone") %>%
  dplyr::select(-sum) %>% 
  filter(rgn_id <= 250)

print(sum(zonal_sums_km2$km2)) # 1394773


write.csv(zonal_sums_km2, file.path(here("globalprep/hab_kelp/v2021/data/habitat_extent_kelp.csv")), row.names = FALSE)

## make gapfilling file

kelp_extent_gf <- read_csv(file.path(dir_git, "data/habitat_extent_kelp.csv")) %>%
  mutate(variable = "extent") %>%
  dplyr::select(-km2, -year) %>%
  mutate(gap_fill = 0)

write.csv(kelp_extent_gf, file.path(dir_git, "data/extent_kelp_gf.csv"), row.names = FALSE)
```


