---
title: 'OHI 2021: Kelp condition'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---

# Summary

This script generates the health condition of kelp for each OHI region for the latest year of data. We do this by applying a 2% decrease to kelp globally for 50 years, and calculate the condition based on our extent data. 


## Updates from previous assessment
This is an entirely new layer for the 2021 assessment! 

***
## Data Source 

**Reference**: Wernberg, T., Krumhansl, K., Filbee-Dexter, K., Pedersen, M.F., 2019. Status and Trends for the World’s Kelp Forests, in: World Seas: An Environmental Evaluation. Elsevier, pp. 57–78. https://doi.org/10.1016/B978-0-12-805052-1.00003-6

**Description**: "In the past half century, threats to kelp forests have increased in number and severity, leading to a global decline of kelp abundances of ~2% per year."

**Time range**: 2019


***
# Methods


## Setup
``` {r setup, echo = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
library(readr)      # for read_csv()
library(raster)
library(here)
library(sf)
library(fasterize)
library(tidyverse)
library(mapview)
library(sp)
library(rgeos)

source(file.path('~/github/ohiprep_v2021/workflow/R/common.R'))

goal     <- 'globalprep/hab_kelp/v2021'
dir_git  <- file.path('~/github/ohiprep_v2021', goal)
dir_wcmc <- file.path(file.path(dir_M, 'git-annex/globalprep/_raw_data/wcmc_kelp'))
dir_kelp <- file.path(dir_wcmc, "d2020/UniAuk-004-ModelledDistributionKelpBiome")
ohi_rasters() # call the region zones raster
regions_shape()
region_data()

land_ocean <- raster("/home/shares/ohi/git-annex/globalprep/spatial/v2017/regions_land_ocean.tif")

land_ocean
```


## Methods 

Calculate a global average condition based on a 2% loss per year. 

Paper: https://www.researchgate.net/publication/327606143_Status_and_Trends_for_the_World's_Kelp_Forests#:~:text=Kelps%20exhibit%20a%20great%20diversity,of%20~2%25%20per%20year.

Excerpt: "In the past half century, threats to kelp forests have increased in number and severity, leading to a global decline of kelp abundances of ~2% per year."


To do this, we will calculate the estimated loss based on a 2% loss per year using a compound interest formula and the total global extent we have extracted from [this paper](https://www.sciencedirect.com/science/article/pii/S0006320720308739).


Any regions that have a negative trend will receive the condition calculated here. Any those that have stable or increasing trends will receive a condition of 1. 

```{r, eval = FALSE}

## read in extent data

kelp_extent <- read_csv(file.path(dir_git, "data/habitat_extent_kelp.csv"))

sum(kelp_extent$km2) # 1394773 km2 total

## interest rate condition calculation

p = 1394773 # current balance of kelp extent km2 (2020)

r = 0.02 #interest rate (gaining 2% per year, since we are trying to figure out how much extent there would be 50 years ago)

n = 1 #yearly 

t = 50 # 50 years 

p*(1 + (r/n))**(n*t) # 3754154 # This is how much you should gain over 50 years 

## so 50 years ago there would be
3754154 + 1394773 # 5148927 km2 of global extent

## Now lets calculate the decline from 50 years ago extent 

p = 5148927 #initial balance of kelp extent km2 (from 50 years ago, calculated above)

r = -0.02 #interest rate (losing 2% per year)

n = 1 #yearly 

t = 50 # 50 years 

p*(1 + (r/n))**(n*t) # 1875083 # This is how much you should lose over 50 years 


5148927 - 1875083 # 3273844 # this is what would be remaining
 
(3273844)/(5148927) # 0.6358303 condition


## Based on this, we will assign a condition of 0.64 to all countries with a negative trend. 


## read in trend data 

kelp_trend <- read_csv(file.path(dir_git, "data/habitat_trend_kelp.csv"))

## assign conditions
kelp_condition <- kelp_trend %>%
  mutate(health = ifelse(trend < 0, 0.64, 1)) %>%
  dplyr::select(-trend)

kelp_condition_gf <- kelp_condition %>%
  mutate(gap_fill = ifelse(health == 0.64, "global value", "none")) %>%
  dplyr::select(-health)

write.csv(kelp_condition, file.path(dir_git, "data/habitat_health_kelp.csv"), row.names = FALSE)
write.csv(kelp_condition_gf, file.path(dir_git, "data/health_kelp_gf.csv"), row.names = FALSE)

```


