---
title: 'OHI `r format(Sys.Date(), "%Y")`: Sea Surface Temperature Pressure Layer'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

# Summary

This script creates the annual sst anomaly rasters for the global Ocean Health Index assessment. For more guidance please refer to the readme.md.

------------------------------------------------------------------------

# Data Source

Data come from NOAA's NCEI: [CoRTAD version 6](https://www.ncei.noaa.gov/products/coral-reef-temperature-anomaly-database)

See `prs_sst/v2015/dataprep.R` for preparation of the "annual_pos_anomalies" data.

**Native Data Resolution**: \~4km\
**Description**: Cortadv6_SSTA.nc = SST anomalies (weekly SST minus weekly climatological SST), weekly data for all years, degrees Kelvin Cortadv6_weeklySST.nc = SST, weekly data for all years, degrees Kelvin\
**Time Range**: 1982 - 2021 (weekly averages across all years)\
**Format**: NetCDF **Downloaded**: June 22, 2022

------------------------------------------------------------------------

## Setup

```{r setup, message=F,warning=F, eval = FALSE}

# Set options for code chunks
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'figs/', message = FALSE, warning = FALSE)

# Load packages
library(raster)
library(RColorBrewer)
library(tidyverse)
library(rgdal)
library(doParallel)
library(foreach)
library(sf)
library(ncdf4)
library(httr)
library(lubridate)
library(animation)
library(plotly)
library(here)
library(snow)
library(terra)

# suppress progress bars for terra
terraOptions(progress=0)

# OHI spatial files, directories, etc
source(here("workflow/R/common.R"))

# Update scenario year, set up programmatic scenario year updating
scen_year_number <- 2023
scen_year <- as.character(scen_year_number)
prev_scen_year <- as.character(scen_year_number - 1)

# suppress progress bars for terra
terraOptions(progress=0)

# Standard OHI file paths
dir_data <- paste0(dir_M, "/git-annex/globalprep/_raw_data/CoRTAD_sst/d", scen_year)
dir_int  <- paste0(dir_M, "/git-annex/globalprep/prs_sst/v", scen_year, "/int")
dir_output  <- paste0(dir_M, "/git-annex/globalprep/prs_sst/v", scen_year, "/output")

# Change this file path for full update to correct update scenario year
dir_rasters <- paste0(dir_M, 
                      "/git-annex/globalprep/prs_sst/prs_sst_calculated_rasters/v2023_update")

# Load in OHI spatial data
ohi_rasters()
regions_shape()

# Specify years for calculations
yrs <- 1982:2022

# Rainbow color scheme and load in OHI regions
cols <- rev(colorRampPalette(brewer.pal(11, 'Spectral'))(255)) 
land <- regions %>% subset(rgn_type %in% c("land", "land-disputed", "land-noeez"))

```

------------------------------------------------------------------------

## Get new data if available

------------------------------------------------------------------------

## Generate annual positive anomalies

We consider anomalies the mean plus one standard deviation; these are the thresholds used to identify 'extreme events'. Since the sea surface temperature anomaly data downloaded from CoRTAD is just the mean, we calculate standard deviation and count cases where the anomaly data exceeds the standard deviation (?)

```{r read in and fomat data, eval = FALSE}

# Load netcdf for SSTA radiation data
ssta <- rast(list.files(dir_data, 
                              pattern = "SSTA.nc", 
                              full.names = TRUE), 
                   subds = "SSTA")


# Rename each layer, terra autonames so we take the names from layer dates
names(ssta) <- time(ssta)

# Assign names to an object
names_weekly <- names(ssta)

# Convert the names into a data frame
ssta_df <- data.frame(name = names_weekly) %>%
    # Convert the name to a date
    mutate(date = as.Date(name)) %>%
    # Extract the year, month, day, and week
    mutate(
        year = year(date),
        month = month(date),
        day = day(date),
        week = week(date))

```

```{r create anomaly rasters}

# Plan for a multicore future
cores <- 4
registerDoParallel(cores)

# Loop runs each week in a year past the weekly SD rasters to identify and sum anomalies
foreach(j = yrs, .packages = c("terra", "dplyr")) %dopar% {
  
  # Start timer
  start_time <- Sys.time()
  
  # Print a time stamp
  print(paste("calculating anomaly for", j, "-- started at", start_time))
  
  # Filter 'ssta_df' for the current year 'j' and select the 'week' column. 
  # The result is stored in 'wks'.
  wks <- ssta_df %>% 
    filter(year == j) %>% 
    select(week)
  
  # initialize s
  s <- list()
  
  for(i in wks$week) {
    
    # Load a raster layer 'sd_sst' from a TIFF file specific to the current week 'i'.
    sd_sst <- terra::rast(file.path(dir_rasters, 
                                    "weekly_sd_30yr_rasters", 
                                    sprintf("sst_sd_week_%s.tif", i)))
    
    # Find the index 'w' of the current week 'i' in the 'names_ssta' vector for the current year 'j'.
    w <- which(substr(names_weekly, 1, 4) == j)[i]
    
    # Get the 'w'-th raster layer from the 'ssta' stack.
    w_ssta <- ssta[[w]]
    
    # Find anomalies using custom function
    anomaly_raster <- terra::lapp(c(w_ssta, sd_sst),
                                  fun = function(x, y){ifelse(is.na(x) | is.na(y), 0,
                                                              ifelse(x > y, 1, 0))})
    
    # Fill s with anomaly rasters
    s[[i]] <- anomaly_raster
    
    # Print a message when it is at the 28th i loop
    if (i == 30) {
      print(paste("Halfway done with year", j))
    }
    
  } # End of inner loop

  # Stack the raster layers in 's' to create a multi-layered raster
  s_stack <- rast(s)

  # Use the 'sum_anom' function defined earlier to calculate the sum of the raster stack 's'.
  yr <- terra::app(s_stack, fun = sum)
  
  # Write the raster layer 'yr' to a TIFF file named "annual_pos_anomalies_sd_<year>.tif".
  writeRaster(yr, filename = file.path(dir_rasters,
                                       "annual_positive_anomalies",
                                       sprintf("annual_pos_anomalies_sd_%s.tif", j)),
              overwrite=TRUE)
  
  # End timer and calculate elapsed time
  end_time <- Sys.time()
  elapsed_time <- end_time - start_time
  
  # Print time message
  print(paste("Anom for year", j, "started at", start_time, "and took", elapsed_time, "minutes to complete"))
  
} # End of outer loop

```
