---
title: 'OHI `r format(Sys.Date(), "%Y")`: Sea Surface Temperature Pressure Layer'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

# Summary

This takes the positive anomaly rasters created in earlier steps and groups them into 5 year segments. Then uses the reference 5 year period of 1985 - 1989 to find the surplus positive anomalies in each 5 year period. These values are then rescaled to be between 0 and 1 and then saved to disk. Once complete, you can return to the main prs_sst folder to finish the data prep.

------------------------------------------------------------------------

## Setup

```{r setup, message=F,warning=F, eval = FALSE}

# Set options for code chunks
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'figs/', message = FALSE, warning = FALSE)

# Load packages
library(raster)
library(RColorBrewer)
library(tidyverse)
library(rgdal)
library(doParallel)
library(foreach)
library(sf)
library(ncdf4)
library(httr)
library(lubridate)
library(animation)
library(plotly)
library(here)
library(snow)
library(terra)

# suppress progress bars for terra
terraOptions(progress=0)

# OHI spatial files, directories, etc
source(here("workflow/R/common.R"))

# Update scenario year, set up programmatic scenario year updating
scen_year_number <- 2023
scen_year <- as.character(scen_year_number)
prev_scen_year <- as.character(scen_year_number - 1)

# Standard OHI file paths
dir_data <- paste0(dir_M, "/git-annex/globalprep/_raw_data/CoRTAD_sst/d", scen_year)
dir_int  <- paste0(dir_M, "/git-annex/globalprep/prs_sst/v", scen_year, "/int")
dir_output  <- paste0(dir_M, "/git-annex/globalprep/prs_sst/v", scen_year, "/output")

# Change this file path for full update to correct update scenario year
dir_rasters <- paste0(dir_M, 
                      "/git-annex/globalprep/prs_sst/prs_sst_calculated_rasters/v2023_update")

# Specify years for calculations, the data set's years, often 1 year behind scenario year
yrs <- 1982:2022

# Load in OHI spatial data
ohi_rasters()
regions_shape()

# Rainbow color scheme and load in OHI regions
cols <- rev(colorRampPalette(brewer.pal(11, 'Spectral'))(255)) 
land <- regions %>% subset(rgn_type %in% c("land", "land-disputed", "land-noeez"))

```

------------------------------------------------------------------------

## Five Year Reference Raster

This chunk creates and saves the 5 year reference period anomaly raster that sums all the positive anomalies from 1985 - 1989. This file is then subtracted from the 5 year anomaly periods starting from 1986 - 1990 and going to the most recent 5 year period.

```{r create reference 5 year anomalies}

# List all pos anom files 
anom_files <- list.files(file.path(dir_rasters, "annual_positive_anomalies"), pattern = "annual_pos_anomalies", full.names = TRUE)

# Initialize an empty vector 'ref_years'
ref_years <- c()

# This loop puts all the anom files between 1985 - 1989 together
for(i in 1985:1989){ref_years = c(ref_years, grep(i, anom_files))}

# Take the file names in ref_years and sum them into a 1 layer raster
ref <- rast(anom_files[ref_years]) %>% app(., fun=sum)

# Write the raster to a TIFF file named "85_89_ref_anom.tif".
writeRaster(ref, filename = file.path(file.path(dir_rasters, "referenc_period_5yr_positive_anomalies"), "85_89_ref_anom.tif"),
            overwrite=TRUE)
```

------------------------------------------------------------------------

## 5 Year Surplus Anomalies Rasters

This code chunk sums positive anomalies in every 5 year period starting from 1986 - 1990 and going to the most recent 5 year period and subtracts the reference period anomalies to create a surplus anomaly file representing how many more positive anomalies happens in each cell during that 5 year period than the 5 year reference period.

```{r 5 yr Surplus Anom}

# Specify the projection
anom_proj = "+proj=longlat +ellps=WGS84 +no_defs"

# Load OHI land sf object then transform to Mollweide
land_mask <- land %>% st_geometry %>% st_transform(anom_proj)

# Convert the 'land_mask' to a terra vector object to use later
land_mask_terra <- terra::vect(land_mask)

# This loop groups pos anom rasters into 5 year chunks and subtracts the reference year
foreach(i = seq(1986, yrs[length(yrs)-4])) %do% {
  
  # Start timer
  start_time <- Sys.time()

  # Generate a sequence of 5 consecutive years starting from 'i'
  years <- i:(i + 4)
  
  # Load reference anomaly period raster
  ref <- rast(file.path(dir_rasters, "referenc_period_5yr_positive_anomalies", "85_89_ref_anom.tif"))
  
  # Load OHI five year anomaly period raster stack
  five_year_rast <- rast(c(anom_files[substr(anom_files, 137, 140) %in% years]))

  # Sum pos anoms in 5 year period then subtract reference period 
  r <- app(five_year_rast, fun=sum) - ref
  
  # Then apply the land mask to the resulting raster
  r <- mask(r, land_mask_terra, inverse = TRUE) 
  
  # Save raster file
  terra::writeRaster(r, file=sprintf("%s/sst_diff_ocean_%s-%s.tif", 
                                      file.path(dir_rasters, "surplus_5yr_positive_anomalies"), years[1], years[5]), 
                      overwrite=TRUE)
  
  # End timer and calculate elapsed time
  end_time <- Sys.time()
  elapsed_time <- end_time - start_time
  
  # Print time message
  print(paste("Anom for years", years[1], "through", years[5], "started at", 
              start_time, "and took", elapsed_time, "to complete"))
}

```

------------------------------------------------------------------------

## Calculate Rescaling Value

This code chunk extracts all the values from all the 5 year surplus anomaly files and calculates the 99.99th percentile value. This percentile value is then used to rescale the data between 0 and 1.

```{r}
# List all surplus pos anom files
diffs <- list.files(file.path(dir_rasters, "surplus_5yr_positive_anomalies"), pattern = "diff", full.names = TRUE)

# Initialize an empty vector to store raster values
vals <- c()

# Loop over the list of surplus anom files
for(i in 1:length(diffs)){
  
  # Extract values from each raster file in list
  m = rast(diffs[i]) %>% values()
  
  # Concatenate the values into a single vector 
  vals = c(vals, m)
}

# Calculate minimum (-142 v2018; -159 v2021; -157 v2022; -166 v2023)
min_val   <- min(vals, na.rm = TRUE) 

# Calculate maximum (182 v2018; 228 v2021; 247 v2023)
max_val   <- max(vals, na.rm = TRUE) 

# Calculate 99.99th percentile (128 v2018; 148 v2021; 148 v2022, 159 v2023)
resc_num  <- quantile(vals, prob = 0.9999, na.rm = TRUE)

# Specify path to previous 99.99th percentile
sup_info  <- here(paste0("globalprep/supplementary_information/v", scen_year))

# Read the reference points pressures data
rescale <- read.csv(file.path(sup_info, "reference_points_pressures.csv"))

# Update the stored reference point with the calculated 99.99th percentile
rescale$ref_point[rescale$pressure == "Sea Surface Temperature"] <- resc_num 

# Write the updated data back to the file
write.csv(rescale, file.path(sup_info, "reference_points_pressures.csv"), row.names = FALSE)

```

------------------------------------------------------------------------

## Rescaling

This code chunk takes the 99.99th percentile value and uses it to rescale all the surplus anomaly files to values between 0 and 1. It then resamples to the OHI standard resolution and saves to disk

```{r rescale surplus pos anoms}

# List surplus pos anom files
diffs <- list.files(file.path(dir_rasters, "surplus_5yr_positive_anomalies"), pattern = "diff.*tif", full.names = TRUE)

# Read in 99.99th percentile rescaling value
resc_num <- read.csv(file.path(sup_info, "reference_points_pressures.csv")) %>%
  filter(pressure == "Sea Surface Temperature") %>%
  .$ref_point

# Transition to numeric
resc_num <- as.numeric(as.character(resc_num))

# Create output directory if it does not exist
if(!file.exists(dir_output)){dir.create(path = dir_output)}

# Loop over each file in the diffs list, 2.5 minutes per file
for(i in 1:length(diffs)) {
  
  # Start timer with tic()
  start_time <- Sys.time()

  # Load the ith raster file from diffs
  r = terra::rast(diffs[i])

  # Extract the year from the filename
  years = substr(diffs[i], 133, 141)
  
  # Check if output file for this year already exists
  if(file.exists(sprintf("%s/sst_%s_1985-1989.tif", dir_output, years))){
    print(paste("skipping", years))
    next
  } else {
    
    # Set the Coordinate Reference System (CRS) for the raster
    crs(r) <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
    
    # Reproject the raster to Mollweide projection (mollCRS)
    out = terra::project(x = r, y = mollCRS, threads = TRUE)
    
    # Apply a function to rescale positive values between 0 and 1
    out <- terra::app(out,
                      fun = function(x){ifelse(x > 0,
                                               ifelse(x > resc_num, 1, x/resc_num), 0)})
    
    # Resameple to resolution of ocean raster and save to disk
    terra::resample(out, ocean, method = "near",
                    filename = sprintf("%s/sst_%s_1985-1989.tif", dir_output, years),
                    overwrite = TRUE,
                    threads = TRUE)
    
  } # End if statement
  
  # End timer and calculate elapsed time
  end_time <- Sys.time()
  elapsed_time <- end_time - start_time
  
  # Print time message
  print(paste("Nomalized anoms for years", years, 
              "started at", start_time, "and took", 
              elapsed_time, "minutes to complete"))
  
  # Free up memory
  gc()
  
} # End loop

```

## Next Steps

Now that you have successfully recalculated all the SST rasters, you can return to the main prs_sst folder and work through the sst_layer_prep.rmd notebook to complete this layer.
