---
title: "OHI `r format(Sys.Date(), '%Y')` - Livelilihoods and Economies Data Cleaning"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This script was created in 2023 to progress towards updating the
livelihoods and economies goal.

It is used to read in and wrangle all available livelihoods and
economies data as of September 2023. When newer updated data is not
available, the original data used in the livelihoods and economies goal in 2013 is
included.

# Setup

```{r, eval = FALSE}
# load libraries, set directories
library(ohicore)  #devtools::install_github('ohi-science/ohicore@dev')
library(tidyverse)
library(plotly)
library(here)
library(janitor)
library(pdftools)
library(modelr)
library(WDI)
source(here('workflow/R/common.R')) # directory locations

source(here('workflow/R/fao_fxn.R')) # function for cleaning old version of FAO data 

source(here('workflow/R/fao_online_portal_clean.R'))

current_year <- 2023
version_year <- paste0("v",current_year)

data_path <- paste0("globalprep/le/",version_year)
```

# Download Data

**Dynamic Data Sources:**

These data sources are updated regularly. If new years of data have been
released, an updated version of the data should be downloaded and used.

-   FAO Capture Production Quantity

    -   The [FAO capture
        production](https://www.fao.org/fishery/en/collection/capture?lang=en)
        data was downloaded via the Statistical Query Panel. This data
        is updated annually and can be downloaded using the following
        instructions:

        -   Navigate to the [online query
            portal](https://www.fao.org/fishery/statistics-query/en/capture/capture_quantity)
            for FAO Global Capture Production Quantity. Deselect all
            pre-selected years. Drag these fields into selected rows:
            ASFIS species name, FAO major fishing area name, ASFIS
            species ISSCAP group name En. ASFIS species Family
            scientific name, FAO major fishing areas, Inland/Marine
            areas Name en. Click on show data and confirm that data is
            present for 1950- two years prior to current year. Click
            download and select yes to include Null Values.

-   FAO Commodities Data

    -   Data from [FAO Global Aquatic Trade
        Statistics](https://www.fao.org/fishery/en/collection/global_commodity_prod):
        Aquatic trade - All partners aggregated (Value) is used to
        estimate revenue from aquarium fishing. This is the original
        data source used, but it has been updated since the livelihoods
        and economies goal was originally calculated.

    -   To download this data, navigate to the Aquatic trade - All
        partners aggregated (Value) [online query
        portal](https://www.fao.org/fishery/statistics-query/en/trade/trade_value).
        De-select all pre-selected years. Under query definition move
        "Reporting Name Country (En) ,"Commodity Name (En), and "Trade
        Flow Name (En)" to selected rows. Click on show data and confirm
        that data is present for 1976- two years prior to current year.
        Click download and select yes to include Null Values.

-   FAO Aquaculture Production

    -   Data from the [Global Aquaculture
        Production](https://www.fao.org/fishery/en/collection/aquaculture?lang=en)
        dataset: Global Aquaculture production Value, is used for
        mariculture revenue.

    -   To download this data, navigate to the Global aquaculture
        production Value [online query
        portal](https://www.fao.org/fishery/statistics-query/en/aquaculture/aquaculture_value).
        De-select all pre-selected years. Under query definition move
        "Country (En) ,"ASFIS species Name", "ASFIS species Family
        Scientific Name", "FAO Major Fishing area Name en" and
        "Environment name en" to selected rows. Click on show data and
        confirm that data is present for the entire time period
        available in the portal, the end date is typically two years
        prior to current year. Click download and select yes to include
        Null Values.

-   [OECD Sustainable Ocean
    Economy](https://stats.oecd.org/Index.aspx?QueryId=95228#) data

    -   This platform is being retired, this data set will likely be
        located in the [OECD Data
        Explorer](https://data-explorer.oecd.org/) in future years.

    -   metadata is available for download through the portal.

**Static Data Sources:**

-   ex-vessel price database.

    -   This database contains ex-vessel price per tonne. It was
        downloaded from the environmental markets lab on request. It is
        possible that this database may be updated in future years.

    -   Recommended citation (note that this is for the original
        database, we use a version of this that emlab updated):
        Melnychuk, M. C., Clavelle, T., Owashi, B., and
        Strauss, K. 2016. Reconstruction of global ex-vessel prices of
        fished species. - ICES Journal of Marine Science.
        <doi:10.1093/icesjms/fsw169>.

-   Fao 2019 Yearbook

    -   This data was published in pdf format
        [here](https://www.fao.org/fishery/static/Yearbook/YB2019_USBcard/navigation/index_intro_e.htm)

    -   Although this report is published annually, newer versions of
        the report do not separate fish, aquaculture and fishery
        processing employees in a way that is usable for our analysis.

-   O'Connor et al 2009

    -   This data was published in a report from 2009. It is available
        in pdf format.
        [here](https://www.cms.int/sites/default/files/document/BackgroundPaper_Aus_WhaleWatchingWorldwide_0.pdf)

-   OWW

    -   The last year of data in this dataset is 2008

    -   Data can be downloaded
        [here](https://www.nber.org/research/data/occupational-wages-around-world-oww-database)

# Read in Data

```{r, eval = FALSE}
# OECD Data 
oecd_raw <- read_csv(file.path(dir_M, "/git-annex/globalprep/_raw_data/OECD/d2023/ocean_economy.csv")) %>% 
  select(Country, Year, VARIABLE, Variable, Unit, Value, PowerCode) 

#FAO commodities data 
#used for aquarium fish
commodities_value <- read_csv(file.path(dir_M, "git-annex/globalprep/_raw_data/FAO_commodities/d2023/online_query_portal/FAO_raw_commodities_value_1976_2021.csv"))

#FAO capture production data, used for fishing revenue
fao_capture <- read_csv(file.path(dir_M, "git-annex/globalprep/_raw_data/FAO_capture/d2023/online_query_portal/Global_capture_production_Quantity_1950-2021.csv"))

#ex-vessel prices, used for fishing revenue 
exvessel_prices <- read_csv(file.path(dir_M, "git-annex/globalprep/_raw_data/ex-vessel-price-database-updated/price-db-results/exvessel_price_database_1976_2019.csv"))

#FAO aquaculture value
aquaculture_value <- read_csv(file.path(dir_M, "git-annex/globalprep/_raw_data/FAO_mariculture/d2023/online_query_portal/FAO_GlobalAquacultureProduction_Value_1984_2021.csv"))

#FAO yearbook pdf tables
fao_pdf <- pdf_text(file.path(dir_M, "/git-annex/globalprep/_raw_data/FAO/d2023/fao_fishers_table.pdf"))

#whale watching pdf
whale_pdf <- pdf_text(file.path(dir_M, "/git-annex/globalprep/_raw_data/IFAW_MarineMammalWatching/whale_watching_worldwide.pdf"))

#oww
oww <- read_csv(file.path(dir_M,"/git-annex/globalprep/_raw_data/OWW/d2023/oww3.csv"))
#unwto -come back to tourism later when we know which data want to use
```

# Data Cleaning and prep

## Revenue

All values for revenue are formatted to US dollars.

### Ocean Energy

```{r, eval = FALSE}
#get the ocean energy data
offshore <- oecd_raw %>% 
  filter(VARIABLE == "RD_TOTAL_OCEAN_OFFSHORE") %>% 
  select(-VARIABLE) %>% 
    clean_names() %>% 
  mutate(sector = "wte",
         unit = "USD (1)",
         data_source = "OECD sustainable ocean economy",
         value = (value * 1000000)
         ) %>% 
  select(-c(power_code)) %>% 
  filter(!country %in% c("OECD - Total",
         "European Union – 27 countries (from 01/02/2020)")) %>%
  select(country, year, value, unit, sector, data_source) 

write_csv(offshore, here(data_path, "int/wte_revenue.csv"))
```

### Aquarium Fishing

From the methods:

-   approximate revenue from aquarium fishing we used **export** data
    from the FAO Global Commodities database for 'Ornamental fish' for
    all available years.

-   used data from two of the four subcategories listed, excluding the
    subcategory 'Fish for culture including ova, fingerlings, etc.'
    because it is not specific to ornamental fish, and the subcategory
    'Ornamental freshwater fish' because it is not from marine systems.

```{r, eval = FALSE}
fao_latest_data_year <- 2021 #update to the last year available in FAO data, likely the same for all FAO data included. 
#clean up data for ornamental fishing revenue
ornamental_value <- commodities_value %>%
  filter(str_detect(`Commodity Name En`, "Ornamental")) %>% 
   rename(country = 'Reporting country Name En',
          commodity = `Commodity Name En`,
           trade = `Trade flow Name En`) %>%
    filter(trade == "Exports" & commodity != "Ornamental freshwater fish") %>% #remove imports and all other fish besides ornamentals 
  fao_online_portal_clean (initial_data_year = 1976, last_data_year = 2021) %>% #function to clean and pivot the data 
  mutate(value = (value * 1000), #convert from thousands to single dollar 
         unit = "USD (1)")

#sum for each country 
ornamental_revenue <-
  ornamental_value %>% 
  group_by(country, year) %>% 
  summarize(value = sum(value, na.rm = TRUE), unit = first(unit)) %>% 
  mutate(sector = "aqf",#code for aquarium fishing sector
         data_source = "FAO global trade") %>% 
  select(country, year, value, unit, sector, data_source)

#write as intermediate file
write_csv(ornamental_revenue, here(data_path, "int/aqf_revenue.csv"))
```

### Fishing

FAO capture data is combined with ex-vessel price data to estimate total
revenue from fishing for each country.

Here we do some basic cleanup of the global capture production database.

```{r, eval = FALSE}

fao_capture_clean <- fao_capture %>% 
   dplyr::rename(country = "Country Name En",
                asfis_species = "ASFIS species Name En",
                area = "FAO major fishing area Name En",
                area_type = "Inland/Marine areas Name En",
                isscaap_group = "ISSCAAP group Name En",
                family_scientific = "Family Scientific name")

fao_capture_clean <- fao_capture_clean %>% 
  fao_online_portal_clean(initial_data_year = 1950, last_data_year = fao_latest_data_year) %>%  #function to clean fao data
  mutate(FAO_name = ifelse(!is.na(asfis_species), asfis_species, family_scientific)) %>% 
  filter(area_type == "Marine areas") %>% #we only want marine capture
  mutate(year = as.numeric(year)) %>% 
  filter(year >= 1976 & year <= 2019) %>% #filter to years in the price database
  filter(!is.na(value) & value > 0) %>% # we do not need the nas or 0, only want the ones with value for tonnes
  filter(`Unit Name` == "Tonnes - live weight") # we don't have price data for species measured in whole animals
  
```

Next we'll clean up and gapfill the exvessel price data. This has
exvessel price in usd/metric tonne.

```{r, eval = FALSE}
exvessel_prices_clean <- exvessel_prices %>% 
  clean_names() %>% 
  mutate(year = as.numeric(year)) %>% 
  group_by(asfis_species, scientific_name, pooled_commodity, isscaap_group, isscaap_division, year) %>% 
  summarize(exvessel = mean(exvessel, na.rm = TRUE)) %>%   #there are some duplicate prices for a species-average when this is the case
   filter(!(is.na(year) & is.na(exvessel))) %>% #we can't use rows that are na for price and year
 
  dplyr::group_by(asfis_species) %>%
  #counts the numbers of non-missing values for each country (logical TRUEs regarded as one)
  dplyr::mutate(value_num = sum(!is.na(exvessel))) %>% 
  filter(value_num > 0) #the minimum we have is 8 years

#look at the data to see how price and year are related overall
model <- lm(exvessel~year, data = exvessel_prices_clean)
summary(model)
#very significant though does not explain a high % of the variation 

#look at linear model by individual species

#define all of the unique species 
species <- unique(exvessel_prices_clean$asfis_species)

#take a look at it at the individual species level
model_list <- list()

#loop through all of the unique species and run the model, store the results in a list
for (i in seq_along(species)) {
  
  species_current <- species[i]
   
  new <- exvessel_prices_clean %>% 
    filter(asfis_species == species_current)
  
  model <- lm(exvessel~year, data = new)
  summary <-summary(model)
  coefficients <- summary$coefficients %>% 
    as.data.frame()
  model_list[[i]] <-coefficients
}

model_test <- bind_rows(model_list)

# Filter rows based on row names containing "year", only want significance of year
filtered_data <- model_test[grep("year", rownames(model_test)), ] %>% 
  clean_names() 
not_sig <- filtered_data %>% filter(pr_t > 0.05) 

(nrow(filtered_data) - nrow(not_sig)) / nrow(filtered_data)
#roughly 85% of these have a significant relationship, appears to be a reasonable gapfilling method

#use lm to predict the value for years which are missing data 
#based on code from ao_need data prep
price_model <- function(df) {lm(exvessel~year, data = df)}


exvessel_prices_gapfilled <- exvessel_prices_clean %>% 
  dplyr::group_by(asfis_species) %>% 
  tidyr::nest() %>% 
  dplyr::mutate(
    ### Apply the model to all country groupings
    model = purrr::map(data, price_model),
    #Use the trained model to get predicted values
    predictions = purrr::map2(data, model, add_predictions)) %>% 
  tidyr::unnest(cols = c(predictions)) %>% 
  dplyr::select(-data, -model, prediction = pred) %>%
  dplyr::mutate(
    gapfilled = dplyr::case_when(is.na(exvessel) | value_num == 1 ~ 1, T ~ 0),
    exvessel = dplyr::case_when(is.na(exvessel) ~ prediction, T ~ exvessel),
    method = dplyr::case_when(
      value_num == 1 ~ "gapfilled using one year of data",
      gapfilled == 1 & value_num > 1 ~ paste0("lm based on N years data: ", value_num),
      T ~ as.character(NA))) %>% 
  dplyr::ungroup() %>% 
  filter(exvessel > 0) #remove price less than 0. This doesn't make sense and never occurs after 1998 anyway so would not be used to calculate any scores. 
```

```{r, eval = FALSE}

#combine the price data with the capture data 
fao_capture_price <- fao_capture_clean %>% 
  left_join(exvessel_prices_gapfilled, by = c("asfis_species", "year", "isscaap_group")) %>% 
  select(-prediction)

#7% of observations still have NAS
summary(fao_capture_price)

#if still NA fill with average price for isscaap species group for that year

#find the average price for each group/year in the exvessel data
group_average_price_year <- exvessel_prices_gapfilled %>% 
  group_by(isscaap_group,year) %>% 
  summarize(mean_group_price = mean(exvessel))


fao_capture_price_final <- 
  fao_capture_price %>% 
  left_join(group_average_price_year, by = c("isscaap_group", "year")) %>% 
  mutate(final_price = ifelse(!is.na(exvessel), exvessel, mean_group_price)) %>%   
  mutate(gapfilled = ifelse((is.na(exvessel) & !is.na(final_price)), 1, gapfilled)) %>% 
  mutate(method = ifelse((is.na(exvessel) & !is.na(final_price)), "filled based on iscaap group average", method)) 

summary(fao_capture_price_final)
#there are still some NAS here, 2% 
nas <- fao_capture_price_final %>% 
  filter(is.na(final_price))
unique(nas$isscaap_group)
#after group discussion we decided to leave these as NA, due to lack of available data
#some of these species including turtles and corals are also considered unsustainable


fao_capture_price_final <- fao_capture_price_final %>% 
  select(country, year, value, asfis_species, final_price, gapfilled, method) %>% 
  mutate(revenue = final_price * value) # multiply price per tonne by number of tonnes

summary(fao_capture_price_final)

#sum prices by country for each year
fishing_revenue <- fao_capture_price_final %>% 
  group_by(country, year) %>% 
  summarize(value = sum(revenue, na.rm = TRUE)) %>% 
  mutate(unit = "USD (1)",
         sector = "cf",
         data_source = "FAO capture production and ex-vessel prices")

write_csv(fishing_revenue, here(data_path, "int/cf_revenue.csv"))
```

### Mariculture

```{r, eval = FALSE}
#clean the aquaculture value data
#this will also be used in the mariculture jobs data section, so need to keep in all environments in the first step
aquaculture_value_clean <- aquaculture_value %>% 
   dplyr::rename(country = "Country Name En",
                species =  "ASFIS species Name En",
                area = "FAO major fishing area Name En",
                environment = "Environment Name En") %>%
  dplyr::select(-c("Unit Name")) %>%
  fao_online_portal_clean(initial_data_year = 1984, last_data_year = fao_latest_data_year) #clean data, note that we are still using a sub_n value of .1 (100 dollars)

#filter to only marine habitats
 mariculture_revenue <- aquaculture_value_clean %>% 
   filter(environment %in% c("Brackishwater", "Marine"))

#sum by region and year (we need total revenue)
total_mar_rev <- mariculture_revenue %>%
  group_by(country, year) %>% 
  summarize(value = sum(value, na.rm = T)) %>% 
  mutate(sector = "mar", data_source = "FAO aquaculture value", unit = "USD (1)",
          value = value *1000) %>% #multiply value times 1000 to get dollars
  select(country, year, value, unit, sector, data_source)

write_csv(total_mar_rev, here(data_path, "int/mar_revenue.csv"))
```

### Marine Mammal Watching

The whale watching revenue is stored in a separate table for each
country throughout the paper. We will create function to extract each of
the tables and then combine them.

Key takeaways from the OHI methods (7.19.0.4 Marine mammal watching and
7.61.1.3 Marine mammal watching) to guide cleaning

-   total expenditures are used as a close proxy for total revenue.

-   We used total expenditure data (direct and indirect expenditures) to
    avoid using a literature derived multiplier effect.

-   When IFAW reported "minimal" revenue from whale watching, we
    converted this description to a 0 for lack of additional
    information.

-   For countries with both marine and freshwater cetacean viewing, we
    adjusted by the proportion of marine revenue as described for the
    jobs dataset.

    -   Jobs Dataset description: "For countries with both marine and
        freshwater cetacean viewing, we adjusted by the proportion of
        marine revenue accordingly. Because some of the whale watching
        in O'Connor et al.focused on freshwater cetacean viewing, we
        categorized the target species listed for each country as
        freshwater or marine. For countries with both marine and
        freshwater species, we categorized the whale watching in those
        countries as either 50% or 90% marine, based on the number of
        marine versus freshwater target species and information provided
        in the report narrative. For Colombia and Indonesia, more
        detailed information in the report narrative allowed for a more
        precise determination of the percentage of marine-based whale
        watching. We applied these marine proportions to data on the
        number of whale watchers before converting these estimates into
        employment estimates."

First we will extract all the tables with summary of country results.
This will mainly be used for the total number of whale watchers which is
part of the jobs calculation, but we will extract it now to get the list
of countries included in the paper.

```{r, eval = FALSE}
#number of whale watchers per country

# there are 7 different tables (one per larger region) in the paper with this information, need to extract all of them and combine

page <- 199
slice_set <-  7:10
region <- "North America"
new_name <- "na_data"
#make a function to extract each one 
whale_pdf_convert_1 <- function(page, slice_set, region, new_name){
  
  columns <- c("country", "1998", "2008", "growth") #create columns
  
  final_data <- whale_pdf[page] %>%  #call the entire page
  as.data.frame() %>% 
  separate_rows('.', sep = "\n") %>% #split into rows based on page breaks
  rename(main_column = ".") %>% 
  slice(slice_set) %>% #input rows to select
  mutate(main_column = str_trim(main_column)) %>% 
  separate(main_column, into = columns, sep = " {2,}", extra = "merge") %>% #split into columns when there are two or more spaces
    select(-c("growth")) %>% 
     pivot_longer(cols = c("1998", "2008"), names_to = "year") %>% 
    mutate(value = case_when(value == "None" ~ "0",
                           value == "Minimal"~ "0",
                           T ~ value)) %>% 
  mutate(value = str_remove_all(value, ",")) %>% #remove commas
  mutate(value = str_remove(value, " ")) %>% #remove spaces between numbers
  mutate(value = ifelse(value == "N/A", NA, value)) %>% 
  mutate(value = as.numeric(value)) %>% 
  mutate(region = region)
  
  assign(new_name, final_data, envir = .GlobalEnv)
}

#africa and middle east
whale_pdf_convert_1(42, 5:26, "Africa and Middle East", "africa_data")

#europe
whale_pdf_convert_1(82, 5:26, "Europe", "europe_data")

#asia
whale_pdf_convert_1(121, 7:25, "Asia", "asia_data")

#oceania
whale_pdf_convert_1(157, 5:22, "Oceania, Pacific Islands and Antarctica", "oceania_data")
  
oceania_data <- oceania_data %>%
  filter(country != "Micronesia") %>% 
  mutate(country = str_replace(country, "Federated States of", "Federated States of Micronesia"))

#north america
whale_pdf_convert_1(199, 7:10, "North America", "na_data")

#mexico had a subscript that got combined with value, replace manually
na_data <- na_data %>% 
  mutate(value = ifelse(country == "México" & year == "2008", 190184, value))

#north america
whale_pdf_convert_1(238, 5:27, "Central America and Caribbean", "ca_data")

#north america
whale_pdf_convert_1(269, 17:27, "South America", "sa_data")

#now that we have all of the available countries, combine into one data frame
whale_watching_numbers <- rbind(africa_data, europe_data, asia_data, oceania_data, na_data, ca_data, sa_data)
```

For whale watching revenue each of the countries included in the table
have their own report. We'll use the list of countries from the previous
step to locate and convert all of these tables.

```{r, eval = FALSE}
#country <- country_list[15] # for testing
#region <- region_list[15]

#test <- whale_extract(country,region)

#create a function to find all of the country tables with revenue and extract
whale_extract <- function(country, region){

#the summary table has some of the countries written a bit differently 
#this replaces them with the way they are written in the separate tables when needed
replacements <- c("The Gambia" = "Gambia, The",
                  "Portugal – Madeira Archipegalo" = "Portugal ‐ Madeira Archipelago",
                  "China ‐ Mainland" = "China (Mainland)",
                  "Myanmar" = "Myanmar (Burma)",
                  "Guadeloupe and islands" = "Guadeloupe and islands (including St. Martin and St. Barthélemy)",
                  "Falkland Islands" = "Falkland Islands (Las Malvinas)", 
                  "Georgia, Ukraine – Black Sea" = "Georgia, Ukraine and Russia – Black Sea",
                  "St. Vincent and the Grenadines" = "St Vincent and the Grenadines",
                  "Netherlands Antilles" = "Netherlands Antilles – Aruba, Bonaire, Curaçao and St. Maarten")

country_new <- str_replace_all(country, replacements)

# Escape special characters in the country name
country_new <- str_replace_all(country_new, "[\\(\\)]", "\\\\\\0") #remove special characters


match_character <- paste0(country_new, "\n") # add /n to help find the locations 

matches <- gregexpr(match_character, whale_pdf) #search the pdf for the country

#set each item to true or false depending on if it was a match
matches_good <- sapply(matches, function(match) any(match != -1))

#get the page number with the table (where match is true)
page_number_original <- which(matches_good)

#if there is more than one match use another method to narrow it down
if(length(page_number_original) >1) {
 updated_match <- paste(country_new, "Year")
  #search again
 page_search <- whale_pdf[page_number_original]
 
 squish <- str_squish(page_search)
 
  matches <- gregexpr(updated_match, squish)
  matches_good <- sapply(matches, function(match) any(match != -1))

#get the page number with the table
page_number_new <- which(matches_good)

#update the page number to just the correct one
page_number_original <- page_number_original[page_number_new]

#if it is still giving us more than one possible page, select the one we want by page (based on looking at the pdf)
if (length(page_number_original) > 1){
  
page_number_original <- subset(page_number_original, page_number_original %in% c("118", "114", "160", "162", "194"))
  
}

}

#subset to the page with the table 
page <- whale_pdf[page_number_original]

#set columns 
column_jobs <- c("Year", "Number of whale watchers", "AAGR", "Number of operators",   "Direct expenditure", "Indirect expenditure", "Total Expenditure")

#turn the page into a table 
data <- page %>% 
  as.data.frame() %>% 
  separate_rows('.', sep = "\n") %>% 
  rename(main_column = ".") %>% 
  mutate(main_column = str_trim(main_column)) %>% 
  separate(main_column, into = column_jobs, sep = " {2,}", extra = "merge")

# Find the index of the row with value of country
index <- which(data$Year == country_new)
if(length(index) == 0){
  index <- 1
}

#filter the page to just the table 
data_clean <- data %>%
  slice((index + 1):n()) %>% #remove any other tables above 
  filter(Year %in% c("1991", "1994", "1998", "2003", "2004", "2005", "2006", "2007", "2008")) %>% 
  mutate(first_2008_index = min(which(Year == 2008, arr.ind = TRUE))) %>%
  filter(ifelse(is.na(first_2008_index), TRUE, row_number() <= first_2008_index)) %>%
  select(-first_2008_index) %>% 
  clean_names() %>% 
  mutate_all(~ str_replace_all(., ",", "")) %>% 
  mutate_all(~ str_replace_all(., "\\$", "")) %>% 
  mutate_all(~ str_replace_all(., "None", "0")) %>% #set none to zero 
  mutate_all(~ str_replace_all(., "Minimal", "0")) %>% #set minimal to zero based on methods 
  mutate(country = country, 
         region = region,
         page = page_number_original) 
#add a column with the country and one with page number (used to get other tables later on)
return(data_clean)
}

#identify the countries based on data we created earlier
#118 countries
country_regions <- whale_watching_numbers %>% 
  group_by(country, region) %>% 
  summarize(n=n())

country_list <- country_regions$country
region_list <- country_regions$region

#create an empty list
whale_list <- list()

#run the function and extract data for each country 
for (i in seq_along(country_list)) {
  
  country <- country_list[i]
  region <- region_list[i]
  data <- whale_extract(country,region)
  
  whale_list[[i]] <- data 
  
}

#combine all of the individual tables into one and clean 
whale_revenue_data <- bind_rows(whale_list) %>% 
  mutate_all(~ str_replace_all(., "N/A", NA_character_)) %>% 
   mutate(number_of_whale_watchers = as.numeric(str_extract(number_of_whale_watchers, "^[0-9]+")),
          aagr = str_extract(aagr, "[-‐]?[0-9]+\\.*[0-9]*"),
          total_expenditure = as.numeric(str_extract(total_expenditure, "^[0-9]+"))) %>% 
  mutate(direct_expenditure = as.numeric(direct_expenditure),
         total_expenditure = as.numeric(total_expenditure),
         indirect_expenditure = as.numeric(indirect_expenditure),
         number_of_whale_watchers = as.numeric(number_of_whale_watchers)) %>% 
  mutate(page = as.numeric(page))
```

We now need to subset the revenue to only revenue related to freshwater
species. The first step is identifying all of the main species used for
each country. These are also stored in various tables scattered
throughout the pdf. We will primarily follow the methods outlined in the
OHI original methods, however we will programmatically query for the
habitat of each species, and assign an exact percentage of marine whale
watching.

```{r, eval = FALSE}

#testing 
#page <- 135
#country <- 	"Japan"

key_species_country <- function(page, country){
  #the first few countries were split into multiple regions and had to be processed separately
  if (country %in% c("New Zealand")){
    filtered_data <- data.frame(species = c("Bryde’s whale",
                                            "bottlenose dolphin",
                                            "short‐beaked common dolphin",
                                            "orca",
                                            "sperm whale",
                                            "bottlenose dolphin",
                                            "dusky dolphin",
                                            "Hector’s dolphin")) %>% 
      mutate(country = country)
  } else if (country %in% c("Japan", "Canada", "Australia", "USA")){
   
    start_page <- page
    list <- sort(species_country$page)
    
    end_page <- which(list == page)
    
    end_page <- list[end_page+1]
    
    next_country <- species_country %>% filter(page == end_page)
    next_country <- next_country$country
    
   pages_search <- whale_pdf[start_page:end_page]
    
  whale_data <- pages_search %>% 
   as.data.frame() %>% #turn into data frame
   separate_rows('.', sep = "\n") %>% 
  rename(main_column = ".") 
  
  end_row <- which(whale_data$main_column == next_country) -1
  
  whale_data <- whale_data %>% 
    slice(1:end_row) %>% mutate(country = country)
  
  #filter to just rows that have main speicies
 index <- which(str_detect(whale_data$main_column, "Main species"))

  temp_list <- list()
 for (i in seq_along(index)){
   start_row <- index[i] + 1
   data_subset <- whale_data[(start_row):nrow(whale_data), ]
   end_row <- which(str_detect(data_subset$main_column,"Tourists:"))[1] -1
   # Filtering the data frame to include rows between 'start_row + 1' and 'end_row'
filtered_data <- data_subset %>% 
  slice(1:end_row)

filtered_data <- filtered_data %>% 
  filter(!str_detect(main_column, "Small cetaceans:")) %>% 
  mutate(main_column = str_squish(main_column)) %>% 
  rename(species = main_column) %>% 
  mutate(country = country) %>% 
  separate_rows(species, sep = ",") %>% 
  mutate(species = str_squish(species)) %>% 
  filter(!(species == ""))
   temp_list[[i]] <-filtered_data   
 }
  filtered_data <- bind_rows(temp_list) %>% 
     mutate(
    species = ifelse(species == "false", paste("false", lead(species)), species)
  ) %>%
  filter(species != "false" | lead(species) != "false") %>% 
    filter(species!= "Small Cetaceans:")
  } else if (country %in% c("Kenya")){
    filtered_data <- data.frame(species = "Indo‐Pacific bottlenose dolphin",
                                   country = country)
    #these countries are included but don't have formal whale watching
  } else if (country %in% c("Morocco", "Senegal", "Seychelles", "Turkey")){
    filtered_data <- data.frame(species = NA,
                                   country = country)
  } else{ # the rest of the countries can be processed in the same manner
  #subset the whale pdf to the page and the two below
 pages_search <-whale_pdf[page:(page+2)]
 
 whale_data <- pages_search %>% 
   as.data.frame() %>% #turn into data frame
   separate_rows('.', sep = "\n") %>% 
  rename(main_column = ".") 
 
  #filter to just rows below where the country name is
 index <- which(whale_data$main_column == country)
 if (length(index) <1) {
   index <-1
 }

 filtered_data <- whale_data[(index):nrow(whale_data), ]
 
 #Find the row index where the condition is met
start_row <- which(str_detect(filtered_data$main_column,"Main species:"))[1] +1

# Calculate the ending row index
end_row <- which(str_detect(filtered_data$main_column,"Tourists:"))[1] -1

# Filtering the data frame to include rows between 'start_row + 1' and 'end_row'
filtered_data <- filtered_data[(start_row):end_row, ]

filtered_data <- filtered_data %>% 
  filter(!str_detect(main_column, "Small cetaceans:")) %>% 
  mutate(main_column = str_squish(main_column)) %>% 
  rename(species = main_column) %>% 
  mutate(country = country) %>% 
  separate_rows(species, sep = ",") %>% 
  mutate(species = str_squish(species)) %>% 
  filter(!(species == "")) 

}

return(filtered_data)

  
}

#make a data frame of unique country and page number combos
species_country <- whale_revenue_data %>% 
  group_by(country,page) %>% 
  summarize(n =n())

#loop through all of the countries and the page numbers we located earlier
species_list <- list()
for (i in seq_along(species_country$country)) {
  
  country <- species_country$country[i]
  page <- species_country$page[i]
  data <- key_species_country(page, country)
  
  species_list[[i]] <- data 
  
}

#in the description instead of table, need to add separataly
pakistan <- data.frame(country = "Pakistan", species = "Indus River dolphin")

species_final <- bind_rows(species_list) %>%
  mutate(
    species = case_when(
      lead(species) == "dolphin" ~ paste(species, "dolphin"),
      lead(species) == "headed" ~ paste(species, "headed"),
      lead(species) == "whale" ~ paste(species, "whale"),
      lead(species) == "whale)" ~ paste(species, "whale)"),
      lag(species) == "southern" ~ paste("southern", species),
      lag(species) == "short‐" ~ paste0("short‐", species),
      lag(species) == "beaked" ~ paste0("beaked", species),
      lag(species) == "false" ~ paste0("false", species),
      lag(species) == "gray" ~ paste("gray", species),
      lag(species) == "Atlantic" ~ paste("Atlantic", species),
      lag(species) == "melon-" ~ paste("melon-", species),
      lag(species) == "long‐finned" ~ paste("long‐finned", species),
      lag(species) == "Indo‐Pacific" ~ paste("Indo‐Pacific", species),
      lag(species) == "short‐beaked" ~ paste("short‐beaked", species),
      lag(species) == "short‐finned" ~ paste("short‐finned", species),
      lag(species) == "harbour" ~ paste("harbour", species),
      lag(species) == "various" ~ paste("various", species),
      lag(species) == "Pacific" ~ paste("pacific", species),
      lag(species) == "Dall’s" ~ paste("dall’s", species),
      TRUE ~ species
    )) %>% 
  filter(!species %in% c("dolphin", "southern", "short‐", "beaked", "short‐beaked", "false", "gray", "whale", "Atlantic", "melon‐", "species", "others occasionally sighted", "as the Chinese white dolphin)",
"None", "101", "long‐finned", "headed whale", "Indo‐Pacific", "Small cetaceans","harbour",
"short‐finned", "various", "Pacific", "Dall’s",
"whale)")) %>% 
  filter(country != "Pakistan") %>% 
  bind_rows(pakistan) %>% 
  mutate(species = str_remove(species, "known locally")) %>% 
  mutate(species = str_to_lower(species)) %>% 
  mutate(species = case_when(species == "risso’s dolphin dolphin" ~ "risso's dolphin",
                             T~species))
```

Now that we have a list of all the main species for each country, we'll
match as many of these species as possible to the IUCN Red List and
query the IUCN api for habitat information.

```{r, eval = FALSE}
#check the iucn red list for species matches 

#convert some species to match the iucn red list 
species_final <- species_final %>% 
  mutate(species = str_remove(species, "‐")) %>% 
   mutate(species = str_remove(species, "[-’']")) %>% 
         mutate(species = str_remove(species, "’")) %>% 
  mutate(species = str_remove(species, "'")) %>% 
   mutate(species = str_remove_all(species, "[()]"))%>% 
  mutate(species = str_remove(species, "–")) %>% 
  mutate(species = str_squish(species)) %>% 
  mutate(species = str_replace(species, "dolphins", "dolphin")) %>% 
  mutate(species = case_when(species == "orca"~ "killer whale",
                             species == "bottlenose dolphin" ~ "common bottlenose dolphin",
                             species == "pantropical spotted dolphins" ~ "pantropical spotted dolphins",
                             species == "sperm whale (north west of crete)"  ~ "sperm whale",
                             species == "falsekiller whale" ~ "false killer whale",
                             species %in%  c("rissos dophin", "rissos dolphin dolphin") ~ "rissos dolphin",
                             species %in% c("longbeaked common dolphin",
                                           "shortbeaked common dolphin") ~ "common dolphin",
                             species == "dwarf or pygmy sperm whale"  ~ "pygmy sperm whale",
                             species == "beakedwhales especially blainvilles." ~ "blainvilles beaked whale",
                             species %in% c("minke whale", "minke whales",
                                            "dwarf minke whale subspecies of minke whale" )  ~ "common minke whale",
                             species == "sperm whale north west of crete"~ "sperm whale",
                             species == "amazon bolivian river dolphin" ~ "amazon river dolphin",                    species == "beluga" ~ "beluga whale",
                             species == "humpback" ~ "humpback whale",
                             (species == "finless porpoise" & country == "Bangladesh")~ "indopacific finless porpoise",
                             (species == "finless porpoise" & country == "Japan")~ "narrowridged finless porpoise",
                             species == "sperm whale occasional" ~ "sperm whale",
                             
                             TRUE ~species))

#read in the list of all the species IUCN has, created in iconic species data prep
iucn_available <- read_csv(here("globalprep/ico/v2023/raw/spp_list_from_api.csv")) %>% 
  filter(class == "MAMMALIA") 

#create a new column join_name, where we modify the list to match our species
iucn_available <- iucn_available %>% 
  filter(!is.na(main_common)) %>% 
  distinct(sciname, main_common) %>% 
  mutate(join_name = str_remove(main_common, "-")) %>% 
  mutate(join_name = str_to_lower(join_name)) %>% 
  mutate(join_name = str_remove(join_name, "'")) %>% 
  mutate(join_name = str_remove(join_name, "’")) %>% 
  mutate(join_name = str_remove_all(join_name, "[()]")) %>% 
  distinct(sciname, join_name, .keep_all = T)


#check which species we can't get from iucn
setdiff(species_final$species, iucn_available$join_name)

#according to the paper, all the species with no match are associated with marine watching
#various dolphin- marine
#tucuxi marine & marine tucuxi -marine
#various tropical dolphin -marine
#dolphin various species - marine
#NA - doesn't have any significant whale watching


#add the iucn data to the final species list 
species_iucn <- species_final %>% 
  left_join(iucn_available, by = c("species" = "join_name")) %>% 
  filter(!is.na(species) & !is.na(sciname))

query_list <- unique(species_iucn$sciname)

#query the iucn red list for habitat information

# Loop through each species and query the API for the habitats for these species

results_list <- list()
api_file <- "/home/shares/ohi/git-annex/globalprep/ico/api_key_gc.csv"
api_key <- scan(api_file, what = 'character') 
for (species_name in query_list) {
  # Construct the API endpoint URL for the current species
  spp_page_url <- paste0('https://apiv3.iucnredlist.org/api/v3/habitats/species/name/', 
                         URLencode(species_name), '?token=', api_key)
  
  # Make the API request and parse the JSON response
  species_data <-jsonlite::fromJSON(spp_page_url) %>% as_tibble()
  
  #add back to the list
  results_list[[species_name]] <- species_data
   }


#turn into a single dataframe
species_results <- bind_rows(results_list) %>%
  unnest(col = result)

#check if any of them are found in freshwater habitats
unique(species_results$habitat)

potentially_freshwater <- species_results %>% 
  filter(str_detect(habitat,"Wetlands")| str_detect(habitat, "Artificial/Aquatic"))

#join back to get species iucn to get common name
freshwater_species <- species_iucn %>% 
  left_join(potentially_freshwater, by = c("sciname" = "name")) %>% 
  filter(!is.na(habitat)) %>% 
  filter(!suitability == "Marginal") #assume they're not common, so unlikely to be tourists watching for them in marginal habitat

#make a list of all the freshwater species 
freshwater_list <- unique(freshwater_species$species)

#here we will calculate percent marine
#note that when a species is listed twice (because the country is divided into smaller regions, it is counted twice)

species_final_hab <- species_final %>% 
  filter(!is.na(species)) %>% #remove countries with no species
  mutate(habitat = case_when(species %in% freshwater_list ~"Freshwater",
         TRUE ~ "Marine")) %>% #classify as marine or freshwater
  group_by(country, habitat) %>% 
  summarize(count = n()) %>% #count the total number of species in each habitat
  ungroup() %>% 
  pivot_wider(names_from = habitat, values_from = count) %>% 
  replace_na(list(Freshwater = 0, Marine = 0)) %>% 
  mutate(percent_marine = Marine/(Marine + Freshwater)) #find percent of total species that are marine

species_final_hab <- species_final_hab %>% 
  select(country, percent_marine)
```

Now that we have percent marine mammal watching we can calculate the
final marine mammal watching revenue.

```{r, eval = FALSE}
whale_revenue_final <- whale_revenue_data %>% 
  left_join(species_final_hab, by = "country") %>% 
  mutate(value = percent_marine * total_expenditure) %>%
  mutate(value = ifelse(is.na(value), 0, value)) %>% #nas are 0, countries with 0 documented revenue
  mutate(unit ="USD (1)",
         sector = "mmw",
         data_source = "O’Connor et al 2009") %>% 
  select(country, year, value, unit, sector, data_source)

write_csv(whale_revenue_final,  here(data_path, "int/mmw_revenue.csv"))
```

### Tourism

The final data source for tourism revenue was not finalized in 2023.
This component will need to be added before proceeding with the
livelihoods and economies update. See the README for discussion of
available data sources.

## Jobs

### Fishing, Mariculture, Fish Processing

#### FAO Yearbook Number of Fishers

Turn the number of fishers table from the FAO yearbook into a data frame
and clean.

```{r, eval = FALSE}
#turn the fao pdf into a format we can use
fishers_table <- fao_pdf[[3]] %>% #page number of the pdf
  as.data.frame()

#specify columns 
columns <- c("country", "1995", "2000", "2005", "2010", "2014", "2015", "2016", "2017", "2018", "2019")

#turn the pdf page into a clean dataframe
fao_fisher_jobs <- fishers_table %>%
  separate_rows('.', sep = "\n") %>% #separate rows based on line breaks
  rename(main_column = ".") %>%
  mutate(main_column = str_remove_all(main_column, "\\*")) %>% #remove the * 
  mutate(main_column = str_replace_all(main_column, "F(?!\\p{L})", " ")) %>% #replace all the Fs not followed by letters (used to indicate estimate)
  slice(7:69) %>% #remove all rows not in the table
  separate(main_column, into = columns, sep = " {2,}", extra = "merge") %>% #separate into columns based on double spaces
  pivot_longer(cols = -c("country"), names_to = "year") %>% 
  mutate(value = str_squish(value)) %>% #remove extra spaces
  mutate(value = str_remove_all(value, " ")) %>% 
  mutate(value = if_else(value == "…" , NA, value)) %>% 
  mutate(value = as.numeric(value)) %>% 
  mutate(Unit = "persons (1)",
         sector = "cf",
         data_source = "FAO number of fishers") %>% 
  clean_names() %>% 
  select(country, year, value, unit, sector, data_source)
```

#### Fao Yearbook Fish Farmers

```{r, eval = FALSE}
fish_farm_table <- fao_pdf[[5]] %>% #page number of the pdf
  as.data.frame()

#specify columns
columns <- c("country", "1995", "2000", "2005", "2010", "2014", "2015", "2016", "2017", "2018", "2019")

#turn the pdf page into a clean dataframe
fao_farm_jobs <- fish_farm_table %>%
  separate_rows('.', sep = "\n") %>% #separate into rows based on line breaks
  rename(main_column = ".") %>%
  mutate(main_column = str_remove_all(main_column, "\\*")) %>% #remove the * 
  mutate(main_column = str_replace_all(main_column, "F(?!\\p{L})", " "))  %>% 
  slice(7:36) %>% #select the part of the page with the table
  separate(main_column, into = columns, sep = " {2,}", extra = "merge") %>%  #split into columns if there is more than two spaces
  pivot_longer(cols = -c("country"), names_to = "year") %>% 
  mutate(value = str_squish(value)) %>% 
  mutate(value = str_remove_all(value, " ")) %>% 
  mutate(value = if_else(value == "…" , NA, value)) %>% 
  mutate(value = as.numeric(value)) %>% 
  mutate(unit = "persons (1)",
         sector = "mar",
         data_source = "FAO number of fish farmers") %>% 
  clean_names() %>% 
   select(country, year, value, unit, sector, data_source)
```

#### OECD Jobs Data

```{r, eval = FALSE}
#number of people employed in th fishing sector 

#note that this is data for the same category as the FAO fishers data, FAO data will be used to gapfill
oecd_fisher_jobs <- oecd_raw %>% 
  filter(VARIABLE == "FISH_EMP_MARINE_FISH_TOT")  %>% 
  select(- c(PowerCode, VARIABLE)) %>% 
  filter(!Country == "OECD - Total") %>% 
  clean_names() %>% 
  mutate(unit = ("persons (1)"),
         value = (value * 1000),
         sector = "cf",
         data_source = "oecd sustainable economies") %>% 
  select(country, year, value, unit, sector, data_source)

#number of people employed in aquaculture, both marine and inland

#note that this is data for the same category as the FAO fish farmers data, FAO data will be used to gapfill
oecd_aqua_jobs <- oecd_raw %>% 
  filter(VARIABLE == "FISH_EMP_AQUACULTURE_TOT")  %>% 
  select(- c(PowerCode, VARIABLE)) %>% 
  filter(!Country == "OECD - Total")  %>% 
  clean_names() %>% 
  mutate(Unit = ("persons (1)"),
         value = (value * 1000),
         sector = "mar",
         data_source = "oecd sustainable economies") %>% 
    filter(! country %in% c("OECD - Total")) %>% 
  select(country, year, value, unit, sector, data_source)

```

Clean data for the Fish Processing Sector.

Due to timing constraints we did not determine a method to subset this
data to only marine related fishery processing in 2023. **This will
needed to be added once a method is determined.**

```{r, eval = FALSE}
#fishery processing jobs
oecd_fishproc_jobs <- oecd_raw %>% 
  filter(VARIABLE == "FISH_EMP_PROCESSING_TOT") %>% 
    select(- c(PowerCode, VARIABLE)) %>% 
    clean_names() %>% 
  mutate(unit = ("persons (1)"),
         value = (value * 1000), 
         sector = "fp",
         data_source = "oecd sustainable economies") %>% 
  select(country, year, value, unit, sector, data_source)

write_csv(oecd_fishproc_jobs, here(data_path, "int/fp_jobs.csv"))
```

Gapfill and save files.

For the fishers jobs data we will use the OECD data set as the primary
data source and gapfill using data from FAO. To update this data in
future years, we will need to consider gapfilling OECD data using
another method, as FAO no longer publishes disaggregated data on the
number of people employed in fishing. Another possibility is contacting
FAO or ILOSTAT to see if they are able to provide disaggregated data for
this sector.

```{r, eval = FALSE}
oecd_fisher_jobs <-oecd_fisher_jobs %>% 
  mutate(country = ifelse(country == "Korea", "South Korea", country)) #south korea is an oecd member

#rename countries in both datasets to match
fao_fisher_jobs <- fao_fisher_jobs %>% 
  mutate(country = case_when(country == "USA" ~ "United States",
                   country == "Turkey" ~ "Türkiye",
                   country == "China" ~ "China (People's Republic of)",
                   country == "China,Taiwan" ~  "Chinese Taipei",
                    country == "UK" ~ "United Kingdom",
                   country == "Korea Rep" ~ "South Korea",
                   country == "Korea D P Rp" ~ "North Korea",
                   T~ country)) %>% 
  mutate(year = as.numeric(year))

#look at the difference to see if we need to rename any other countries
setdiff(oecd_fisher_jobs$country, fao_fisher_jobs$country)

missing_oecd <- setdiff(fao_fisher_jobs$country, oecd_fisher_jobs$country)
missing_oecd
#make a list of the countries in the fao data but not oecd, these are the ones we want to add
fao_missing_oecd <- fao_fisher_jobs %>% 
  filter(country %in% c(missing_oecd))

fishing_jobs_final <- oecd_fisher_jobs %>% 
  rbind(fao_missing_oecd)

write_csv(fishing_jobs_final, here(data_path, "int/cf_jobs.csv"))
```

For the mariculture jobs data we will use the OECD data set as the
primary data source and gapfill using data from FAO. To update this data
in future years, we will need to consider gapfilling OECD data using
another method, as FAO no longer publishes disaggregated data on the
number of people employed in aquacutlure. Another possibility is
contacting FAO or ILOSTAT to see if they are able to provide
disaggregated data for this sector.

```{r, eval = FALSE}
oecd_aqua_jobs <- oecd_aqua_jobs %>% 
  mutate(country = ifelse(country == "Korea", "South Korea", country)) #south korea is an oecd member

fao_farm_jobs <- fao_farm_jobs %>% 
  mutate(country = case_when(country == "USA" ~ "United States",
                   country == "Turkey" ~ "Türkiye",
                   country == "China" ~ "China (People's Republic of)",
                   country == "China, Taiwan" ~  "Chinese Taipei",
                    country == "UK" ~ "United Kingdom",
                   country == "Korea Rep" ~ "South Korea",
                   T~ country)) %>% 
  mutate(year = as.numeric(year))

setdiff(oecd_aqua_jobs$country, fao_farm_jobs$country)

#countries in fao data not in oecd
oecd_missing <- setdiff(fao_farm_jobs$country, oecd_aqua_jobs$country)
oecd_missing

fao_farm_oecd_missing <- fao_farm_jobs %>% 
  filter(country %in% c(oecd_missing))

#combine the two datasets
aqua_jobs_combined <- oecd_aqua_jobs %>% 
  rbind(fao_farm_oecd_missing) %>% 
  mutate(year = as.numeric(year))
```

Both data-sets combine mariculture and other aquaculture. Because of
this we need to subset the aquaculture jobs to just those related to
marine and brackish mariculture.

In the methods document it states: "In order to estimate the proportion
of total aquaculture jobs that can be attributed to marine and brackish
aquaculture, we used country-specific proportions of marine and brackish
aquaculture revenues (compared to total revenues) calculated from FAO
aquaculture production data, assuming that numbers of jobs approximately
scale with production in terms of revenue. For country-years with no
data for the proportion of marine/brackish production because of gaps in
the FAO production data, we used the proportion from the most recent
year for which data were available. For countries without proportion
estimates from any years, we used the average proportion from the
country's geographic region (e.g., Caribbean, Polynesia, Eastern Asia),
with the exception of American Samoa, for which we used the proportion
value from Guam."

To replicate these methods we will use the fao global aquaculture
production value database. We will find the proportion of revenue for
each country that is associated with marine/brackish species.

```{r, eval = FALSE}
#classify brackish as marine for summing 
 aquaculture_value_proportions <- aquaculture_value_clean %>% 
   filter(!is.na(environment)) %>% 
   mutate(environment = ifelse(environment == "Brackishwater", "Marine", environment)) %>% #assign environment to marine or freshwater
  group_by(country, year, environment) %>% 
  summarize(value = sum(value, na.rm = T)) %>%  # find the total for each environment by year
   pivot_wider(names_from = environment, values_from = value) %>% 
   mutate(Marine = ifelse((is.na(Marine) & !is.na(Freshwater)), 0, Marine)) %>% #fill in 0 for marine if it's NA and there is freshwater data
   mutate(Freshwater = ifelse((is.na(Freshwater) & !is.na(Marine)), 0, Freshwater)) %>% #fill in 0 for freshwater if it's NA and there is marine data 
   mutate(prop_marine = (Marine/sum(Freshwater,Marine))) %>% 
   mutate(prop_marine = ifelse(Marine == 0 & Freshwater == 0, NA, prop_marine)) #if both are 0, we don't know the proportions
 
 aquaculture_value_proportions <- aquaculture_value_proportions %>% 
   select(country, year, prop_marine) %>% 
   mutate(year = as.numeric(year))
 

#change names in the jobs data to match names in the prop data
aqua_jobs_clean <- aqua_jobs_combined %>% 
  mutate(country = case_when(
    country == "Netherlands" ~ "Netherlands (Kingdom of the)",
    country == "Russian Fed"  ~ "Russian Federation",
    country ==  "United States" ~ "United States of America",
    country == "China (People's Republic of)" ~ "China",
    country == "Czech Republic" ~ "Czechia",
    country == "Chinese Taipei" ~ "Taiwan Province of China",
    country == "South Korea"  ~ "Korea, Republic of",
    country == "Slovak Republic" ~"Slovakia",
    country == "Iran"  ~ "Iran (Islamic Rep. of)",
    TRUE ~ country
  ))

#make sure countries are named the same
 setdiff(aqua_jobs_clean$country, aquaculture_value_proportions$country)
 
setdiff(aquaculture_value_proportions$country, aqua_jobs_combined$country)
 
#combine prop_marine with the jobs data set
mariculture_jobs_final <- aqua_jobs_clean %>% 
   left_join(aquaculture_value_proportions, by = c("year", "country"))
#check if we need to gapfilll
summary(mariculture_jobs_final) # no nas in prop_marine for the countries we have jobs data for, don't have to gapfill as described above
  
mariculture_jobs_final <- mariculture_jobs_final %>% 
  mutate(jobs = round(prop_marine * value)) %>% 
  select(country, year, value= jobs, unit, sector, data_source)
 

write_csv(mariculture_jobs_final, here(data_path, "int/mar_jobs.csv"))
```

### Tourism

The final data source for tourism jobs was not finalized in 2023. This
component will need to be added before proceeding with the livelihoods
and economies update. See the README for discussion of available data
sources.

### Marine Mammal Watching Jobs

Key takeaways from methods:

-   Jobs based on number of whale watchers in a country and a regional
    average number of whale watchers per employee. Includes all marine
    mammal watching.

-   When IFAW reported "minimal" numbers of whale watchers, we converted
    this description to a 0 for lack of additional information.

-   For countries with both marine and freshwater species, we
    categorized the whale watching in those countries as either 50% or
    90% marine, based on the number of marine versus freshwater target
    species and information provided in the report narrative. For
    Colombia and Indonesia, more detailed information in the report
    narrative allowed for a more precise determination of the percentage
    of marine-based whale watching. We applied these marine proportions
    to data on the number of whale watchers before converting these
    estimates into employment estimates.

We roughly followed these methods, however percent marine was calculated
programmatically, and an exact percentage was used. See revenue section
for methods.

```{r, eval = FALSE}
#get table with regional average number of whale watchers per employee out of pdf
columns <- c("region", "job_number", "watch_per_employee")

regional_whale_employee <- whale_pdf[26] %>% #page number
  as.data.frame() %>% 
  separate_rows('.', sep = "\n") %>% #separate into rows based on page break
  rename(main_column = ".") %>% 
  slice(23:31) %>% #select rows with table
  slice(1,2,3,4, 6,7,9) %>% #remove rows that get split into two
  mutate(main_column = str_trim(main_column)) %>% 
  separate(main_column, into = columns, sep = " {2,}", extra = "merge") %>% #separate into columns if there is a double or more space
  mutate(region = case_when(str_detect(region, "Oceania") ~ "Oceania and the Pacific
Islands",
str_detect(region, "Central America") ~  "Central America and Caribbean",
T ~ region)) %>% #fix a few regions that were split up 
  mutate(watch_per_employee = as.numeric(str_remove(watch_per_employee, ","))) %>% #remove commas from numbers
  select(-job_number)
```

Apply proportion marine to the number of whale watchers (table extracted
in marine mammal watching revenue section)

```{r, eval = FALSE}
whale_watching_marine <- whale_watching_numbers %>% 
  left_join(species_final_hab, by = "country") %>% 
  mutate(watcher = value * percent_marine) %>% 
  select(watcher,country, region, year)
```

```{r, eval = FALSE}
whale_jobs <- whale_watching_marine %>% 
  left_join(regional_whale_employee, by = c("region")) %>% 
  mutate(jobs = round(watcher/watch_per_employee,0)) %>% 
  mutate(value = ifelse(is.na(jobs), 0, jobs)) %>%  #nas are 0s in this case
  mutate(unit ="persons (1)",
         sector = "mmw",
         data_source = "O’Connor et al 2009") %>% 
  select(country, year,value, unit, sector, data_source)
write_csv(whale_jobs, here(data_path, "int/mmw_jobs.csv"))

```

## Wages Data

### Tourism, Fishing, Ports and Harbors, Ship and Boat-building

**From the OHI methods:** 

-   We used the [Occupational Wages around the
World](http://www.nber.org/oww/) (OWW) database produced by Remco H. Oostendorp and Richard B. Freeman in 2005.

-   "The database provides several different calibrations, and we use
    the"x3wl calibration", described as a "country-specific and uniform
    calibration with lexicographic weighting," and recommended as being
    the preferred calibration in most cases."

-   Although significant gaps exist in this database, it contains
    country-specific information on average wages in many industries for
    more than 150 countries from 1983-2003.

-   Data represent average monthly wages of a male worker. Wage data
    were divided by the inflation conversion factor for 2010 so that
    wage data across years would be comparable
    (<http://oregonstate.edu/cla/polisci/sahr/sahr>), and then
    multiplied by the purchasing power parity-adjusted per capita gdp
    (ppppcgdp, WorldBank).

-   The adjusted wage data were then multiplied by 12 to get annual
    wages.

-   We used the industry and occupation classifications reported in the
    OWW to estimate wages for marine-related sectors.

**Update Notes:**

-   The available data from OWW goes through 2008, the database has been
    discontinued and will not be updated in the future.

-   Since we have not decided what year we will be using the inflation
    conversion factor for, data are cleaned only to the point of
    selecting the appropriate sectors and occupations.

-   Although the methods say we used "xw3l", it appears this is a typo
    and the correct calibraiton is w3wl. OWW documentation says: "The
    standardized wage with country-specific calibration and imputation
    and lexicographic weighting (w3wl) is recommended for use." The
    variable **mw3wl_us** is selected: **monthly: country-specific
    calibration with imputation (lex), US\$.** Occupation codes are
    defined in Technical_document_OWW_1953-2008_release1.0.docx under
    the documentation section.

-   All data are in monthly wages (USD)

From the methods we know the occupations we need:

-   Fishing

    -   Industry: deep sea & coastal fishing; (y3 Industry code AE)

    -   Occupations: Deep-sea fisherman; inshore (coastal) maritime
        fisherman (y4 Occupation codes 9 & 10)

-   Ports & harbors

    -   Industry: supporting services to maritime transport; (y3: NE)

    -   Occupation: dock worker (y4: 117)

-   Transportation and shipping

    -   Industry: maritime transport (y3 : ND)

    -   occupations: ship's chief engineer; ship's passenger stewards;
        able seaman (y4: 114, 115, 116)

-   Ship & boat building

    -   Industry: shipbuilding and repairing; (y3: JD)

    -   Occupation: ship plater (y4: 75)

-   Tourism

    -   Industry: restaurants and hotels; (y3: MC)

    -   Occupations: hotel receptionist; cook; waiter; room attendant or
        chambermaid. (y4: 97, 98, 99, 100)

    -   These data are not specific to coastal/marine tourism jobs, and
        thus we assumed that wages in these jobs are equal in coastal
        and non-coastal areas

Process and save the oww data

```{r, eval = FALSE}
oww_clean <- oww %>% 
  filter(y4 %in% c(9, 10, 75, 97, 98, 99, 100, 114,115, 116, 117)) %>% #select the relevant occupation codes
  select(y0, y1, y3, y4, country_name, mw3wl_us) %>% 
  rename(country = country_name,
         year = y0, 
         industry_code = y3,
         occupation_code = y4,
         value = mw3wl_us) %>% 
  mutate(unit = "monthly wage USD (1)",
         sector = case_when(occupation_code %in% c(9, 10) ~ "cf",
                            occupation_code %in% c(117) ~"ph",
                            occupation_code %in% c(114, 115, 116) ~ "tran",
                            occupation_code == 75 ~ "sb", 
                            occupation_code %in% c(97, 98, 99, 100) ~ "tour"),
         data_source = "oww database") %>% 
  select(country, year, value, unit, occupation_code, sector, data_source)


write_csv(oww_clean, here(data_path, "int/all_sectors_wages.csv"))
```
