---
title: "OHI `r format(Sys.Date(), '%Y')` - Economies Tourism Dataprep"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

------------------------------------------------------------------------


# Economies Dataprep 
        
### v2025 Updates

- New WTTC data to use for tourism revenue
- Added data from the IMF to gapfill a few missing countries in World Bank GDP data
- Overhauled framework such that scoring and coastal population proportion happens in `ohi_global`, not here. 

## Overview


### Tourism revenue from World Trade and Tourism Council (WTTC) 

This data does not have the usual download steps. It was acquired via a request directly to the WTTC via the enquiry form below. More detailed instructions around the acquisition of this data can be found in [issue 358](https://github.com/OHI-Science/globalfellows-issues/issues/358). 
  
- Data enquiry form  <https://wttc.org/research/data-enquiries>
- Make sure it's clear that this is a request from an academic institution, as the WTTC seems to make allowances for providing free data to researchers. Also make sure you communicate that the raw data will not be publicly available.



### World Bank: GDP (constant 2015 US$) 

**Reference**: World Bank. (n.d.). *GDP (constant 2015 US$)*. World Bank Open Data. <https://data.worldbank.org/indicator/NY.GDP.MKTP.KD> .

License: CC BY-4.0

**Downloaded**: 2025-08-20

**Last updated**: No precise date listed on website. Found in the first few pages of the .csv when opened locally: 2024-06-28. 

**Description**: “GDP at purchaser's prices is the sum of gross value added by all resident producers in the economy plus any product taxes and minus any subsidies not included in the value of the products. It is calculated without making deductions for depreciation of fabricated assets or for depletion and degradation of natural resources. Data are in constant 2015 prices, expressed in U.S. dollars. Dollar figures for GDP are converted from domestic currencies using 2015 official exchange rates. For a few countries where the official exchange rate does not reflect the rate effectively applied to actual foreign exchange transactions, an alternative conversion factor is used.” Aggregation method: Gap-filled total. 

**Download Instructions**: Navigate to the World Bank GDP (constant 2015 US$) data page <https://data.worldbank.org/indicator/NY.GDP.MKTP.KD>. The full time range should be selected on the slider bar by default. Select “CSV” under the “Download” options on the lower right-hand side of the page. Save in the appropriate folder on Mazu: 
`home/shares/ohi/git-annex/globalprep/_raw_data/WorldBank/dYYYY/WorldBank_global_annual_GDP_2015_constant_USD/WorldBank_global_annual_GDP_2015_constant_USD.csv`

**Time range**: 1960-2023

**Native data resolution**: Country-level (not spatial data). 

**Format**: `.csv.` Note: Data is also available in XML and EXCEL formats. 

**Metadata**: Click on “Details” (in the upper right hand corner of the plot on the main page), scroll to the bottom of the popup page, and select “All Metadata”.
<http://databank.worldbank.org/data/reports.aspx?source=2&type=metadata&series=NY.GDP.MKTP.KD&_gl=1*10r25gl*_gcl_au*MTI5NjExNDc0NS4xNzE2NTY4ODAx> 

## IMF Total GDP Data

**Download Instructions**: Navigate to this link <https://data.imf.org/en/Data-Explorer?datasetUrn=IMF.RES:WEO(6.0.0)>. 

Choose these options for the selectors on the left-hand side:
  - Indicator: Gross Domestic Product (GDP), current prices, US Dollar
  - Time Period: Custom --> select years of interest (for v2025, this was 2008-2023)
- Hit apply
- Click the blue download button dropdown on the right side of the page 
  - Select "Data on this page"
  
This should give you a .csv file of total annual GDP by country. 

### Setup

```{r}
knitr::opts_chunk$set(eval = FALSE)

# ---- Load packages ----

if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  ohicore, # name_2_rgn
  tidyverse,
  here,
  countrycode,
  janitor,
  priceR, # for inflation adjustment
  RColorBrewer,
  foreach,
  doParallel, # for using multiple cores
  httr,
  scales,
  plotly,
  readxl,
  zoo # for gapfilling
)


# ---- Source common.R ----
source(here::here("workflow", "R", "common.R"))

# ---- Set scenario year, reproducible file paths ----
current_year <- 2025 # Update this in the future!!
previous_year <- 2024

data_year <- 2023

version_year <- paste0("v",current_year)
previous_vyear <- paste0("v", previous_year)

current_data_dir <- paste0("d", current_year)
previous_data_dir <- paste0("d", previous_year)

# ---- data directories
# Raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")

# Livelihoods and Economies directory
current_le_dir <- here("globalprep", "le", version_year)
previous_le_dir <- here("globalprep", "le", previous_vyear)
```



# Tourism Revenue

## Visitor Exports

```{r}
# International visitor vis_spending or Visitor exports, US$, real 2024 prices and exchange rates, billions
visitor_exports <- read_xlsx(here(raw_data_dir, "WTTC", current_data_dir, 
                              "WTTC-Data-Request-OHI-13.08.2025.xlsx"),
                              sheet = 3, # make sure sheet num matches
                              skip = 2)

# Pivot longer and rename
visitor_long <- visitor_exports %>% 
  pivot_longer(cols = -1, # everything except the first column
               values_to = "vis_spending",
               names_to = "year") %>% 
  rename(country = Economy) %>% 
  mutate(year = as.numeric(year))

# Fix name_2_rgn error with Netherland Antillies
neth_ant <- visitor_long %>%
    filter(country == 'Former Netherland Antillies') %>% 
    mutate(vis_spending = (vis_spending / 5),  
           'Curacao' = vis_spending,
           'Sint Maarten' = vis_spending,
           'Bonaire' = vis_spending,
           'Saint Eustatius' = vis_spending,
           'Saba' = vis_spending) %>%
    select(-c(vis_spending, country)) %>%
    pivot_longer(cols = c("Curacao", "Sint Maarten", "Bonaire",
                          "Saint Eustatius", "Saba"),
                 names_to = "country",
                 values_to = "vis_spending") %>%
    mutate(country = as.character(country),
           vis_spending = as.numeric(vis_spending)) %>% 
    select(c(country, vis_spending, year))
  
visitor_long <- visitor_long %>%
    filter(country != 'Former Netherland Antillies') %>%
    bind_rows(neth_ant)

# run ohicore name_2_rgn 
visitor_rgn <- name_2_rgn(df_in = visitor_long, 
                       fld_name = 'country',
                       flds_unique = c("year"))

# Check duplicates
tour_duplicates <- visitor_rgn[duplicated(visitor_rgn[, c("rgn_name", "year")]),]
unique(tour_duplicates$rgn_name)

# aggregate: sum by rgn_name etc.
visitor_clean <- visitor_rgn %>%
  group_by(rgn_id, rgn_name, year) %>%
  dplyr::summarize(vis_spending = sum(vis_spending, na.rm = TRUE) )%>%
  ungroup() %>% 
  # turn 0s into NAs (appropriate flag for lack of data)
  mutate(vis_spending = case_when(vis_spending == 0 ~ NA,
                          .default = vis_spending),
         unit = "US$, real 2024 prices and exchange rates, billions",
         source = "WTTC")

# Confirm duplicates have been removed
tour_duplicates <- visitor_clean[duplicated(visitor_clean[, c("rgn_name", "year")]),]
unique(tour_duplicates$rgn_name)

# Check years for possible gapfilling
visitor_clean <- visitor_clean %>% 
  group_by(rgn_id, rgn_name) %>% 
  complete(year = 2008:data_year) %>% 
  ungroup()

visitor_clean %>% 
  filter(is.na(vis_spending))
# Missing two observations for 2021

# ---- Visitor vis_spending gapfilling ----
visitor_gf <- visitor_clean %>% 
  # First, create gapfilled boolean and method columns
  mutate(vis_gapfilled = ifelse(is.na(vis_spending), "1", NA_character_),
         vis_gf_method = ifelse(is.na(vis_spending), "1", NA_character_)) %>%
  # Regroup for gapfilling
  group_by(rgn_id, rgn_name) %>% 
  arrange(rgn_id, year) %>%   
  # Linear interpolation
  mutate(vis_spending = na.approx(vis_spending, x = year, na.rm = FALSE)) %>% 
  mutate(vis_gapfilled = ifelse(vis_gapfilled == "1" & !is.na(vis_spending), 
                               "gapfilled", vis_gapfilled),
         vis_gf_method = ifelse(vis_gf_method == "1" & !is.na(vis_spending), 
                               "Interpolation", vis_gf_method)) %>% 
  # Upfilling and Downfiling
  tidyr::fill(vis_spending, .direction = "updown") %>% 
  mutate(vis_gapfilled = ifelse(vis_gapfilled == "1" & !is.na(vis_spending), 
                               "gapfilled", vis_gapfilled),
         vis_gf_method = ifelse(vis_gf_method == "1" & !is.na(vis_spending), 
                               "Up/Downfilling", vis_gf_method))

table(visitor_gf$vis_gf_method)
colSums(is.na(visitor_gf)) # 0 NAs for vis_spending now
length(unique(visitor_gf$rgn_id)) # 147 regions

```

## Domestic Spending

```{r}
# Domestic T&T expenditure, US$, real 2024 prices and exchange rates, billions
domesting_spending <- read_xlsx(here(raw_data_dir, "WTTC", current_data_dir, 
                              "WTTC-Data-Request-OHI-13.08.2025.xlsx"),
                              sheet = 4, # make sure sheet num matches
                              skip = 2)

# Pivot longer and rename
domestic_long <- domesting_spending %>% 
  pivot_longer(cols = -1, # everything except the first column
               values_to = "dom_spending",
               names_to = "year") %>% 
  rename(country = Economy) %>% 
  mutate(year = as.numeric(year))

# Fix name_2_rgn error with Netherland Antillies
neth_ant <- domestic_long %>%
    filter(country == 'Former Netherland Antillies') %>% 
    mutate(dom_spending = (dom_spending / 5),  
           'Curacao' = dom_spending,
           'Sint Maarten' = dom_spending,
           'Bonaire' = dom_spending,
           'Saint Eustatius' = dom_spending,
           'Saba' = dom_spending) %>%
    select(-c(dom_spending, country)) %>%
    pivot_longer(cols = c("Curacao", "Sint Maarten", "Bonaire",
                          "Saint Eustatius", "Saba"),
                 names_to = "country",
                 values_to = "dom_spending") %>%
    mutate(country = as.character(country),
           dom_spending = as.numeric(dom_spending)) %>% 
    select(c(country, dom_spending, year))
  
domestic_long <- domestic_long %>%
    filter(country != 'Former Netherland Antillies') %>%
    bind_rows(neth_ant)

# run ohicore name_2_rgn 
domestic_rgn <- name_2_rgn(df_in = domestic_long, 
                       fld_name = 'country',
                       flds_unique = c("year"))

# Check duplicates
tour_duplicates <- domestic_rgn[duplicated(domestic_rgn[, c("rgn_name", "year")]),]
unique(tour_duplicates$rgn_name)

# aggregate: sum by rgn_name etc.
domestic_clean <- domestic_rgn %>%
  group_by(rgn_id, rgn_name, year) %>%
  dplyr::summarize(dom_spending = sum(dom_spending, na.rm = TRUE) )%>%
  ungroup() %>% 
  # turn 0s into NAs (appropriate flag for lack of data)
  mutate(dom_spending = case_when(dom_spending == 0 ~ NA,
                          .default = dom_spending),
         unit = "US$, real 2024 prices and exchange rates, billions",
         source = "WTTC")

# Confirm duplicates have been removed
tour_duplicates <- domestic_clean[duplicated(domestic_clean[, c("rgn_name", "year")]),]
unique(tour_duplicates$rgn_name)

# Check years for possible gapfilling
domestic_clean <- domestic_clean %>% 
  group_by(rgn_id, rgn_name) %>% 
  complete(year = 2008:data_year) %>% 
  ungroup()

colSums(is.na(domestic_clean))
# No missing data, so no need for gapfilling
length(unique(domestic_clean$rgn_id)) # 147 regions

```

## Combine Visitor and Domestic Spending

```{r}
# Combine spending data
tourism_spending <- full_join(visitor_gf, domestic_clean, 
                              by = join_by(rgn_id, rgn_name, year, unit, source)) %>% 
  mutate(total_spending = vis_spending + dom_spending)

colSums(is.na(tourism_spending)) # No NAs

# Save tourism spending as an int data file
tour_spending_export <- tourism_spending %>% 
  select(rgn_id, rgn_name, year, total_spending, unit)

write_csv(tour_spending_export, here(current_le_dir, "int", "eco_tour_spending.csv"))

# Save gapfilling info
write_csv(tourism_spending, here(current_le_dir, "int", "eco_tour_spending_gf.csv"))
```


# Total GDP 

Current unit of "total" in dataframe:

- "Tourism direct GDP as a proportion of total GDP (%)" 

## World Bank Data

This serves as the primary data source for total GDP


```{r}
# ---- World Bank GDP in 2015 Constant USD data ----
wb_gdp_raw <- read_csv(here(raw_data_dir, "WorldBank", current_data_dir,
                            "WorldBank_global_annual_GDP_2015_constant_USD",
                            "API_NY.GDP.MKTP.KD_DS2_en_csv_v2_366142.csv"),
                       skip = 4) # skip first 4 rows

# Preliminary cleaning 
wb_gdp_long <- wb_gdp_raw %>% 
  pivot_longer(cols = `1960`:`2024`,
               names_to = "year",
               values_to = "gdp") %>% 
  clean_names() %>% 
  # Need to use fixed bc of parentheses
  mutate(country_name = str_replace_all(country_name, fixed("Puerto Rico (US)"), "Puerto Rico")) %>% 
  select(country = country_name, country_code, 
         year, gdp, unit = indicator_name) %>% 
  # Filter to years of interest
  filter(year >= 2008 & year < 2024) 


# Add OHI regions
wb_gdp_rgn <- name_2_rgn(wb_gdp_long,
                        fld_name = 'country',
                        flds_unique = c("year")) 

# Check duplicates
duplicates <- wb_gdp_rgn[duplicated(wb_gdp_rgn[, c("rgn_name", "year")]),]
unique(duplicates$rgn_name)

# aggregate: sum by rgn_name etc.
wb_gdp_agg <- wb_gdp_rgn %>%
  group_by(rgn_id, rgn_name, year, unit) %>%
  dplyr::summarize(gdp = sum(gdp, na.rm = TRUE) )%>%
  ungroup() %>% 
  # turn 0s into NAs (appropriate flag for lack of data)
  mutate(gdp = case_when(gdp == 0 ~ NA,
                          .default = gdp),
         source = "WorldBank",
         year = as.integer(year))

# Confirm duplicates have been removed
duplicates <- wb_gdp_agg[duplicated(wb_gdp_agg[, c("rgn_name", "year")]),]
unique(duplicates$rgn_name)

# Check years for possible gapfilling
wb_gdp_comp <- wb_gdp_agg %>% 
  group_by(rgn_id, rgn_name) %>% 
  complete(year = 2008:data_year) %>% 
  ungroup()


# ---- Total GDP gapfilling ----
wb_gdp_gf <- wb_gdp_comp %>% 
  # First, create gapfilled boolean and method columns
  mutate(wb_gapfilled = ifelse(is.na(gdp), "1", NA_character_),
         wb_gf_method = ifelse(is.na(gdp), "1", NA_character_)) %>%
  # Regroup for gapfilling
  group_by(rgn_id, rgn_name) %>% 
  arrange(rgn_id, year) %>%   
  # Linear interpolation
  mutate(gdp = na.approx(gdp, x = year, na.rm = FALSE)) %>% 
  mutate(wb_gapfilled = ifelse(wb_gapfilled == "1" & !is.na(gdp), 
                               "gapfilled", wb_gapfilled),
         wb_gf_method = ifelse(wb_gf_method == "1" & !is.na(gdp), 
                               "Interpolation", wb_gf_method)) %>% 
  # Upfilling and Downfiling
  tidyr::fill(gdp, .direction = "updown") %>% 
  mutate(wb_gapfilled = ifelse(wb_gapfilled == "1" & !is.na(gdp), 
                               "gapfilled", wb_gapfilled),
         wb_gf_method = ifelse(wb_gf_method == "1" & !is.na(gdp), 
                               "Up/Downfilling", wb_gf_method))

table(wb_gdp_gf$wb_gf_method)

wb_gdp_gf %>% 
  filter(is.na(gdp)) %>% 
  distinct(rgn_id, rgn_name)
# 5 regions with no data. We'll drop these

wb_gdp_gf <- wb_gdp_gf %>% 
  drop_na(gdp)

colSums(is.na(wb_gdp_gf)) # No NAs
nrow(wb_gdp_gf) # 2576
summary(wb_gdp_gf)
length(unique(wb_gdp_gf$rgn_id)) # 161 regions

```

## IMF Data

Secondary gapfilling dataset for total GDP

```{r}
# Read in the IMF data
imf_gdp_raw <- read_csv(here(raw_data_dir, "IMF", current_data_dir,
                      "dataset_2025-08-27T15_17_19.726948394Z_DEFAULT_INTEGRATION_IMF.RES_WEO_6.0.0.csv"))


# Preliminary cleaning 
imf_gdp_long <- imf_gdp_raw %>% 
  pivot_longer(cols = `2008`:`2023`,
               names_to = "year",
               values_to = "gdp") %>% 
  clean_names() %>% 
  # Remove all text after commas in country name. This is for name 2 rgn later
  mutate(country = case_when(country == "Congo, Democratic Republic of the" ~ "Democratic Republic of the Congo",
                             country == "Congo, Republic of" ~ "Republique du Congo",
                             country == "São Tomé and Príncipe, Democratic Republic of" ~ "Sao Tome and Principe",
                             .default = country),
         country = str_remove_all(country, ",.*"),
         year = as.integer(year)) %>% 
  select(country, year, gdp, indicator, unit = scale)

# Add OHI regions
imf_gdp_rgn <- name_2_rgn(imf_gdp_long,
                        fld_name = 'country',
                        flds_unique = c("year"))

# Check duplicates
duplicates <- imf_gdp_rgn[duplicated(imf_gdp_rgn[, c("rgn_name", "year")]),]
unique(duplicates$rgn_name)
# No duplicates here

# Check years for possible gapfilling
imf_gdp_comp <- imf_gdp_rgn %>% 
  group_by(rgn_id, rgn_name) %>% 
  complete(year = 2008:data_year) %>% 
  ungroup() %>% 
  select(!country)

# ---- Total GDP gapfilling ----
imf_gdp_gf <- imf_gdp_comp %>% 
  # First, create gapfilled boolean and method columns
  mutate(imf_gapfilled = ifelse(is.na(gdp), "1", NA_character_),
         imf_gf_method = ifelse(is.na(gdp), "1", NA_character_)) %>%
  # Regroup for gapfilling
  group_by(rgn_id, rgn_name) %>% 
  arrange(rgn_id, year) %>%   
  # Linear interpolation
  mutate(gdp = na.approx(gdp, x = year, na.rm = FALSE)) %>% 
  mutate(imf_gapfilled = ifelse(imf_gapfilled == "1" & !is.na(gdp), 
                               "gapfilled", imf_gapfilled),
         imf_gf_method = ifelse(imf_gf_method == "1" & !is.na(gdp), 
                               "Interpolation", imf_gf_method)) %>% 
  # Upfilling and Downfiling
  tidyr::fill(gdp, .direction = "updown") %>% 
  mutate(imf_gapfilled = ifelse(imf_gapfilled == "1" & !is.na(gdp), 
                               "gapfilled", imf_gapfilled),
         imf_gf_method = ifelse(imf_gf_method == "1" & !is.na(gdp), 
                               "Up/Downfilling", imf_gf_method))

table(imf_gdp_gf$imf_gf_method)
colSums(is.na(imf_gdp_gf)) # No NAs
nrow(imf_gdp_gf) # 2533
length(unique(imf_gdp_gf$rgn_id)) # 149 regions

```

## Join WB and IMF data

The conversion from IMF to WB dollar units is trickier because the IMF isn't using a constant value, but rather the real price for each year. This means we need a separate deflation factor for each year of our data

```{r}
# Make df of year and gdp deflators
deflators <- data.frame(
  year = 2008:2023,
  gdp_deflator = c(
    88.646, 87.729, 89.676, 91.652, 94.099, 95.776, 97.042,
    97.316, 99.275, 101.711, 104.134, 106.748, 109.606,
    112.378, 116.057, 119.908, 125.230
  )
)

deflator_wb <- 2015

# Convert from real prices to constant 2015
imf_gdp_conv <- imf_gdp_gf %>%
  left_join(deflators, by = "year") %>%
  mutate(deflator_wb = deflators$gdp_deflator[deflators$year == deflator_wb],
         gdp_const_wb = gdp * (deflator_wb / gdp_deflator)) %>% 
  mutate(source = "IMF") %>% 
  select(rgn_id, rgn_name, year, gdp = gdp_const_wb, source, 
         imf_gapfilled, imf_gf_method)

# Filter out only the new regions
new_rgns <- anti_join(imf_gdp_conv, wb_gdp_gf, 
                      by = join_by(rgn_id, rgn_name, year)) %>% 
  select(-c(imf_gapfilled, imf_gf_method))

# Bind new regions to WB data
total_gdp <- bind_rows(wb_gdp_gf, new_rgns) %>% 
  mutate(unit = "constant 2015 US$")

colSums(is.na(total_gdp))
summary(total_gdp)
length(unique(total_gdp$rgn_id)) # 163 regions

# Export intermediate file
gdp_export <- total_gdp %>% 
  select(rgn_id, rgn_name, year, gdp, unit)

write_csv(gdp_export, here(current_le_dir, "int", "eco_total_gdp.csv"))

# Save gapfilling information
write_csv(total_gdp, here(current_le_dir, "int", "eco_total_gdp_gf.csv"))
```


# Combine GDP with Tourism Revenue

This is the conversion we're using to go between 2024 real prices to 2015 constant prices.

$$Value_{2015} = Value_{2024} \times \frac {GDP Deflator_{2015}} {GDP Deflator_{2024}}$$
Price deflator data comes from [FRED](https://fred.stlouisfed.org/data/A191RD3A086NBEA). Where GDP Deflator 2015: 97.316 and GDP Deflator 2024: 125.230

Update these values as needed for different prices and years. 

```{r}
# Read in clean data
total_gdp <- read_csv(here(current_le_dir, "int", "eco_total_gdp.csv"))
tour_spending <- read_csv(here(current_le_dir, "int", "eco_tour_spending.csv"))

# Need to fix units between total gdp and tourism spending
# UPDATE AS NEEDED
deflator_wb <- 97.316 # 2015 value
deflator_wttc <- 125.230 # 2024 value

deflation_factor <- deflator_wb / deflator_wttc

# Convert WTTC units to match WB units
tour_convert <- tour_spending %>% 
  # First, change from billions to regular $
  mutate(total_spending = total_spending * 1e9) %>% 
  # Then, apply the conversion factor
  mutate(spending_2015 = total_spending * deflation_factor) %>% 
  select(rgn_id, rgn_name, year, tourism_spending = spending_2015) 

# Join WB and WTTC data
eco_tourism <- left_join(tour_convert, total_gdp, 
                         by = join_by(rgn_id, rgn_name, year)) %>% 
  mutate(tour_prop = tourism_spending / gdp,
         unit = "constant 2015 US$")

colSums(is.na(eco_tourism)) # 96 missing data points
nrow(eco_tourism) # 2352
length(unique(eco_tourism$rgn_id)) # 147 rgns

eco_tourism %>% 
  filter(is.na(tour_prop)) %>% 
  distinct(rgn_id, rgn_name)
# 6 Regions with WTTC data and no WB data

# Drop NAs
eco_tourism <- eco_tourism %>% 
  drop_na(tour_prop) 

# Save final tourism revenue proportion
eco_tour_export <- eco_tourism %>% 
  select(rgn_id, year, tour_rev_prop = tour_prop)

write_csv(eco_tour_export, here(current_le_dir, "output", "eco_tourism_proportion.csv"))

# Save info with units
write_csv(eco_tourism, here(current_le_dir, "int", "eco_tourism_prop_units.csv"))

```

# Data check

```{r}
eco_tour_plot <- eco_tourism %>%
  plot_ly(x = ~year, y = ~tour_prop, color = ~rgn_name,
          type = "scatter", mode = "lines") %>%
  layout(title = "Economies: Tourism Proportion",
         xaxis = list(title = "Year"),
         yaxis = list(title = "Tourism revenue proportion"))
eco_tour_plot


# hist_max <- ggplot() +
#   geom_histogram(data = liv_rescaled_final, aes(x = liv_norm),
#                  color = "grey10", fill = "indianred")+
#   labs(title = "Distribution of rescaled livelihood: Maximum",
#        x = "Livelihood proportion: Max rescale",
#        y = "Count") +
#   theme_bw();ggplotly(hist_max)

# htmlwidgets::saveWidget(liv_final_plot, here::here(current_tr_dir, "figs", "ap_rescaled.html")) # save plot

```

```{r}
# Get max data years for plotting
# current_data_year <- max(new_tr$year)
# current_data_year
# previous_data_year <- max(old_tr$year)
# previous_data_year

# Combine data frames
# comb_tr <- new_tr %>% 
#   rename(new_Ap = Ap) %>% 
#   left_join(old_tr, by = join_by(rgn_id, year))

# Plot
# comp_plot1 <- ggplot() + 
#   geom_jitter(data = filter(comb_tr, year == previous_data_year),
#               aes(x = Ap, y = new_Ap, 
#                   text = paste("Year:", year, "<br>Region:", rgn_name)),
#               alpha = 0.7) +
#   labs(x = paste("Previous Value", previous_data_year),
#        y = paste("Current Value", previous_data_year),
#        title = paste("Sustainable tourism and recreation", previous_data_year, "Data", previous_year, "Assessment vs.", previous_data_year, "Data", current_year, "assessment")) +
#   geom_abline(slope = 1, intercept = 0, col = "red") +
#   theme_bw();ggplotly(comp_plot1)
```


