---
title: 'OHI 2025: Beaches Condition'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---

# Beach Health: Summary

This document serves to calculate the condition of sandy beaches around the globe, by applying predicted rates of degredation from this [article](https://www.nature.com/articles/s41598-018-24630-6). This 2018 article review geospatial data from 1984 to 2016, and determined that 24% of all sandy coastline was eroding at a rate of 0.5m/yr. This is the second document that should be calculated in the hab_beaches folder.

This document applies a simple equation to the beaches_extent data, that calculates the predicted rate of loss, and applies it in relation to how many years have passed since the time of assessment (from 2016 to now) assuming a linear relationship and constant rate of change. If/when further information or updated information is compiled on coastal erosion, it should be applied here and the methods altered to reflect that. Condition is then scaled from 0-1.

## Data Source

**Downloaded**: 09/01/2025

**Description**:\
Global Distribution of Beaches <https://www.nature.com/articles/s41598-018-24630-6>.

**Time range**: 2010-2020

```{r}
# Load in packages
# Load relevant libraries
library(readr) 
library(raster)
library(here)
library(sf)
library(fasterize)
library(tidyverse)
library(mapview)
library(sp)
library(terra)

source(here::here('workflow', 'R', 'common.R'))
goal <- 'globalprep/hab_beaches/v2025'
dir_git  <- file.path('~/github/ohiprep_v2025', goal)
```

```{r}
# Load in the extent data 
beaches_extent <- read.csv(here(goal, "data", "hab_beaches_extent.csv"))

# Apply linear equation to the values of extent
beaches_health <- beaches_extent %>%
  mutate(loss = 0.24 * 0.005 * km2)

# Multipy this loss by number of years since approximation of rate of change
beaches_health <- beaches_health %>%
  mutate(total_loss = loss*9) # From 2016-2025 the difference is 9 years

# Scale the total loss from 0-1
beaches_health <- beaches_health %>%
  mutate(health = (total_loss - min(total_loss)) / (max(total_loss) - min(total_loss)))
```

```{r}
# Save only relevent columns for the final csv
final_beaches_health <- beaches_health %>%
  select(rgn_id, habitat, year, health)

# Save the health data as a csv file
write.csv(final_beaches_health, here(goal, "data", "hab_beaches_health.csv"), row.names = FALSE)
```
