---
title: "beaches_extent"
author: "Maddy Enda"
date: "2025-08-25"
output: html_document
echo: False
---

```{r}
# Load relevant libraries
library(readr) 
library(raster)
library(here)
library(sf)
library(fasterize)
library(tidyverse)
library(mapview)
library(sp)
library(terra)

source(here::here('workflow', 'R', 'common.R'))

goal <- 'globalprep/hab_beaches/v2025'

dir_git  <- file.path('~/github/ohiprep_v2025', goal)

dir_beaches<- file.path(file.path(dir_M, 'git-annex/globalprep/_raw_data/Global_Coastline_Data/v2025')) #note: folder should be made for each year of new data, and changed

ohi_rasters() # call the region zones raster
```

# Load in the data from the server

```{r}
# Get all reefextent_.gpkg files
beaches_2010 <- st_read(file.path(dir_beaches, "GCL2010.shp"))

beaches_2015 <- st_read(file.path(dir_beaches, "GCL2015.shp"))

beaches_2020 <- st_read(file.path(dir_beaches, "GCL2020.shp"))
```

```{r}
# Combine all of the beaches data into one object
beaches_all <- bind_rows(beaches_2010, beaches_2015, beaches_2020)
```

```{r}
# Find the length of all coastline
# vector of lengths for each feature
lengths <- st_length(beaches_all)

# Total length of coastal data
total_length <- sum(lengths)

# Total length in km
total_length_km <- units::set_units(total_length, "km")
```

```{r}
# Confirm reprojection
beaches_transform <- st_transform(beaches_all, crs= "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")
```

## Rasterize the lines

```{r}
# Take the land rasters from OHI
land_ocean <- raster("/home/shares/ohi/git-annex/globalprep/spatial/v2017/regions_land_ocean.tif")

# Convert raster to a Terra() package type raster (SpatRaster). This is because lines can only be rasterized through the terra package. In order to stick with OHI convention, we will be converting this back to a RasterLayer. 
eez_raster <- rast(land_ocean)

# Make valid, Vectorize the lines
beaches_transform <- st_make_valid(beaches_transform)
beaches_transform <- beaches_transform[!st_is_empty(beaches_transform), ]# remove empty geometries

# Vectorize the lines (a terra specific package)
lines_vect <- vect(beaches_transform)

# Rasterize lines to the same grid as zones
beaches_raster <- rasterize(lines_vect, eez_raster, field = 1, background = 0)

writeRaster(beaches_raster, filename = file.path(dir_M, "git-annex/globalprep/hab_beaches/beaches_raster.tif"), overwrite=TRUE)
```


# Plot the raster of beaches data
```{r}
plot(beaches_raster)
```

# Run Zonal Statistics to Get Average Length Per EEZ

I loaded in the raster from the server in order to save time, and decrease the GiB used by the environment. By loading it in using raster(), we also convert it back into a RasterLayer, which will allow us to continue in the raster package from here.

```{r}
# Load in the beaches raster
beaches_raster <- raster(file.path(dir_M, "git-annex/globalprep/hab_beaches/beaches_raster.tif"))
```

```{r}
# Calculate zonal stats with zones raster and new coral. Convert to km^2 and save int/output files*
zonal_sums <- terra::zonal(beaches_raster,
                                     land_ocean,
                                     fun = "sum",
                                     na.rm = TRUE,
                                     progress="text") #sum all coral cells for each ohi zone
# Create a zonal_sums dataframe
zonal_sums_df <- data.frame(zonal_sums)

zonal_sums_km2 <- zonal_sums_df %>%
  mutate(year = 2025, habitat = "beaches",
         km2 = (0.934478877011218970**2*sum)) %>% #one cell is equal to ~0.93 km
  rename("rgn_id" = "zone") %>%
  dplyr::select(-sum) %>%
  filter(rgn_id <= 250)

print(sum(zonal_sums_km2$km2)) # v2025: 1,056,968 km2

# Write the data frame to CSV locally (on Github). Be sure to check file path using here() 
write.csv(zonal_sums_km2, here("int", "beaches_zonal.csv"), row.names = FALSE)
```

