---
title: 'OHI 2022 - Land based nutrient pourpoints prep '
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../workflow/templates/ohi_hdr.html' 
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: inline
---


[REFERENCE RMD FILE](http://ohi-science.org/ohiprep_v2021/globalprep/prs_land-based_nutrient/v2021/STEP4_pourpoints/STEP1_pourpoints_prep.Rmd)


# Summary
This document describes the steps for obtaining and prepping data used for the land based nutrient data layer for clean waters. In specific, this data prep focuses on prepping the N pourpoint shapefiles from crop fertilizer and manure excretion, which will be fed into the plume model. Here we aggregate each N raster to the appropriate watershed, and eventually the appropriate pourpoint.

## The following data are used:

* N leaching and volatilization rasters created in previous steps of this prep.
* Global watersheds dataset 
* Global pourpoints dataset (associated with each watershed)


# Updates from previous assessment
Added a new year of data for the fertilizers by nutrient (2020) for v2022

## Initial set-up code

```{r setup, eval = FALSE}
library(tidyverse)
library(raster)
library(sf)
library(mapview)
library(janitor)
library(here)
library(parallel)
library(doParallel)
library(terra)


source(here('workflow/R/common.R'))

ww_raw_dir <- "/home/shares/ohi/git-annex/land-based/wastewater/data/raw"
ww_intermediate_dir <- "/home/shares/ohi/git-annex/land-based/wastewater/data/interim"
prs_int_dir <- "/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2022/int"
prep <- file.path("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2022")


```

# Methods 

**Read in watersheds files, transform the CRS to 4326, and bind together to make one file.**
These watershed files are taken from the wastewater project at NCEAS (Tulhoske & Halpern 2021). These will be combined with the pourpoints data and the nutrient data, and evenutally be plugged into the plume model.
```{r, eval = FALSE}
## Watersheds; read in, reproject to EPSG 4326, and combine into one file. 

watersheds_af <- st_read(file.path(ww_raw_dir, "basins_laea/af_bas.shp")) %>%
  clean_names() %>%
  dplyr::select(id, gridcode, area, basin_id) ## select columns of interest
crs(watersheds_af) ## check CRS
# plot(watersheds_af$geometry)
watersheds_af <- st_transform(watersheds_af, 4326) ## change CRS to 4326
# plot(watersheds_af$geometry)
crs(watersheds_af)
st_write(watersheds_af, file.path(prs_int_dir, "watersheds/watersheds_af_4326.shp")) ## write transformed file to mazu. 

watersheds_au <- st_read(file.path(ww_raw_dir, "basins_laea/au_bas.shp")) %>%
    clean_names() %>%
  dplyr::select(id, gridcode, area, basin_id)
plot(watersheds_au$geometry)
watersheds_au <- st_transform(watersheds_au, 4326)
plot(watersheds_au$geometry)
crs(watersheds_au)
st_write(watersheds_au, file.path(prs_int_dir, "watersheds/watersheds_au_4326.shp"))

watersheds_eu <- st_read(file.path(ww_raw_dir, "basins_laea/eu_bas.shp")) %>%
    clean_names() %>%
  dplyr::select(id, gridcode, area, basin_id)
plot(watersheds_eu$geometry)
watersheds_eu <- st_transform(watersheds_eu, 4326) 
crs(watersheds_eu)
st_write(watersheds_eu, file.path(prs_int_dir, "watersheds/watersheds_eu_4326.shp"))

watersheds_na <- st_read(file.path(ww_raw_dir, "basins_laea/na_bas.shp")) %>%
  #filter(ID == 79890) %>% 
    clean_names() %>%
  dplyr::select(id, gridcode, area, basin_id)
plot(watersheds_na$geometry)
watersheds_na <- st_transform(watersheds_na, 4326) 
plot(watersheds_na$geometry)
crs(watersheds_na)
st_write(watersheds_na, file.path(prs_int_dir, "watersheds/watersheds_na_4326.shp"))

watersheds_pa <- st_read(file.path(ww_raw_dir, "basins_laea/pa_bas.shp")) %>%
    clean_names() %>%
  dplyr::select(id, gridcode, area, basin_id)
plot(watersheds_pa$geometry)
watersheds_pa <- st_transform(watersheds_pa, 4326) 
plot(watersheds_pa$geometry)
crs(watersheds_pa)
st_write(watersheds_pa, file.path(prs_int_dir, "watersheds/watersheds_pa_4326.shp"))

watersheds_sa <- st_read(file.path(ww_raw_dir, "basins_laea/sa_bas.shp")) %>%
    clean_names() %>%
  dplyr::select(id, gridcode, area, basin_id)
plot(watersheds_sa$geometry)
watersheds_sa <- st_transform(watersheds_sa, 4326) 
crs(watersheds_sa)
st_write(watersheds_sa, file.path(prs_int_dir, "watersheds/watersheds_sa_4326.shp"))

watersheds_as <- st_read(file.path(ww_raw_dir, "basins_laea/as_bas.shp")) %>%
    clean_names() %>%
  dplyr::select(id, gridcode, area, basin_id)
plot(watersheds_as$geometry)
watersheds_as <- st_transform(watersheds_as, 4326) 
crs(watersheds_as)
st_write(watersheds_as, file.path(prs_int_dir, "watersheds/watersheds_as_4326.shp"))

watersheds_all <- rbind(watersheds_eu, watersheds_au, watersheds_na, watersheds_pa, watersheds_af, watersheds_sa, watersheds_as) ## bind all of these files together to make the global watershed files with CRS 4326. 
#plot(watersheds_all$geometry)

st_write(watersheds_all, file.path(prs_int_dir, "watersheds/watersheds_all_4326.shp"))

# mapview(tail(watersheds_all, 10000))
```

**Read in pourpoints file and transform crs to be 4326**
```{r, eval = FALSE}
## Pourpoints

global_plume <- st_read("/home/shares/ohi/git-annex/land-based/wastewater/data/raw/pour_points/global_plume_2007_2010.shp")
# plot(global_plume$geometry)
mapview(tail(global_plume, 10000))

global_plume <- st_transform(global_plume, 4326) 

```

**Write a function or for loop to read in nutrient data per year and disaggregate by 15 to make higher resolution.**
```{r, eval = FALSE}

dis_rast_area <- rast(file.path(prs_int_dir, "disagg_raster_area.tif")) # made inside the loop, only need to do once

years <- as.character(c(2005:2020))

# takes 12 mins per file doing 4 at a time.. so (12*15)/4 = 45 mins to run

registerDoParallel(cores=2)
foreach(yr = years) %dopar% {
## N/P fertilizer databanks 
  
 # yr = 2005
  
  volt_file <- rast(list.files(file.path(prep, "exclude_surface_water"), pattern = paste0("crop_manure_volt_nutrient_N_", yr), full = TRUE))
  
  leached_file <- rast(list.files(file.path(prep, "exclude_surface_water"), pattern = paste0("crop_manure_leached_nutrient_N_", yr), full = TRUE))
  
 # file = files_list[1]
 rast <- volt_file + leached_file

# cellStats(rast, "sum") # 18172110

# divide nutrients by area of cell.  to get nutrients/km2 (or, whatever); 2, Disagregate the data; 3. multiply the disagregated raster by the new cell areas; 4. sum across all raster cells and compare to the original raster…the total should be about the same

 
# rast_div <- rast/raster::area(rast)
# test2 <- raster::area(rast)
  
rast_div <- rast/terra::cellSize(rast, unit = "km")

# test <- terra::cellSize(rast_fix, unit = "km")

dis_rast <- disaggregate(rast_div, fact = 15, method = 'near', progress = TRUE) # this about 2 mins for 1 file... so if we do 4 at a time should be fairly quick


# only needs to be done once... you can ignore this.. or if you do want to run it, don't run it in the for loop. We just need an area raster so that we can multiply by the cell area, without having to calculate the cell area every time again
# dis_rast_area <- terra::cellSize(dis_rast, unit = "km")
# writeRaster(dis_rast_area, file.path(prs_int_dir, "disagg_raster_area.tif"))



dis_rast_fix <- dis_rast*dis_rast_area # takes ~4 mins per raster

# global(dis_rast_fix, "sum", na.rm = TRUE) # 18150910 it worked! 
# global(rast, "sum", na.rm = TRUE) # 18150911
 
file_final <- paste0(prs_int_dir, "/disagg_rasts/", "disagg_crop_manure_leached_volt_N_", yr, ".tif") 

writeRaster(x = dis_rast_fix, filename = file.path(file_final), overwrite = TRUE) # save the file takes 6 mins


}


```



**Calculate zonal stats per watershed and save.**
Take the watersheds shapefile and rasterize it so that we can run zonal stats on it. We will calculate the zonal stats per each watershed to obtain an estimate of the N concentration per each watershed. 
```{r, eval = FALSE}
## Zonal stats per watershed 
rast_files <- list.files(file.path(prs_int_dir, "disagg_rasts"), full = TRUE) ## read in disaggregated raster

## rasterize the watershed file (only need to do this once.. doesn't need to be in a loop)
rast <- raster(rast_files[1])

watersheds_all <- st_read(file.path(prs_int_dir, "watersheds/watersheds_all_4326.shp")) ## read in watersheds file

wat_length <- length(watersheds_all$basin_id)

watersheds_all <- watersheds_all %>%
  arrange(id) %>%
  mutate(id_new = 1:wat_length) # produce a new id for watersheds so that we can fasterize correctly... I don't believe fasterize would work with basin_id since it is a character variable. 


raster_watersheds <- fasterize(watersheds_all, rast, field = "id_new") ## make watershed raster

plot(raster_watersheds)
# length(unique(raster_watersheds)) # 
raster_watersheds

writeRaster(x = raster_watersheds, filename = file.path(prs_int_dir, "raster_watersheds.tif"), overwrite = TRUE)

raster_watersheds <- raster(file.path(prs_int_dir, "raster_watersheds.tif"))
plot(raster_watersheds)

## Start for loop here for(file in files){}

rast_files <- list.files(file.path(prs_int_dir, "disagg_rasts"), full = TRUE) ## read in disaggregated raster

registerDoParallel(cores = 6)
foreach(file = rast_files) %dopar% {

 # file <- rast_files[15]
  
  rast <- raster(file)
  
 # cellStats(rast, "sum") 

zs <- raster::zonal(rast, raster_watersheds, fun = 'sum', na.rm = TRUE) ## calculate sum of N or P per watershed
zs_df <- data.frame(zs)

file_base <- gsub(".tif", "", basename(file))

write_csv(zs_df, file.path(paste0(prs_int_dir, "/zonal_df", "/zonal_", file_base, ".csv")))


watersheds_all_zs <- watersheds_all %>%
  left_join(zs_df, by = c("id_new" = "zone"))

# test <- watersheds_all_zs %>%
#   filter(is.na(sum))

st_write(watersheds_all_zs, file.path(paste0(prs_int_dir, "/zonal_shp", "/zonal_watersheds_", file_base, ".shp")))

}

## 43 NAs of sum after joining... This means that 43 watersheds are being left out for some reason... likely because they are too small for fasterizing/zonal statistics. 
```

**Fix the missing 43 watersheds that zonal stats missed and add back into the data**
```{r, eval = FALSE}
watersheds_all <- st_read(file.path(prs_int_dir, "watersheds/watersheds_all_4326.shp")) %>%
  arrange(id)  ## read in watersheds file


files_list <- str_subset(string = c(list.files(file.path(prs_int_dir, "zonal_shp")), full = TRUE), ".shp")

registerDoParallel(cores = 5)
foreach(file = files_list) %dopar% {
  
#   file <- files_list[1]
watersheds_zonal <- st_read(file.path(prs_int_dir, "zonal_shp/", file))

watersheds_zonal_NA <- watersheds_zonal %>%
  dplyr::filter(is.na(sum)) ## filter for NAs

watersheds_zonal_NA$centroid <- st_centroid(watersheds_zonal_NA) %>%
  st_geometry() ## check the centroids

## take a look to compare and see if it worked correctly
# plot(st_geometry(watersheds_zonal_NA))
# plot(st_set_geometry(watersheds_zonal_NA, 'centroid')[, 0], add = T, col = 'red', pch = 19) ## it worked..

## now lets extract from our N/P raster disaggregated
watersheds_zonal_NA_centroid <- watersheds_zonal_NA %>%
  dplyr::select(-centroid) %>%
  st_centroid() %>%
  dplyr::select(-sum)


dis_rast_file <- gsub("^.*?_", "", file)
dis_rast_file <- gsub("^.*?_", "", dis_rast_file)
dis_rast_file <- gsub(".shp", ".tif", dis_rast_file)

dis_rast <- rast(file.path(prs_int_dir, "disagg_rasts", dis_rast_file))

nutrient_type <- str_extract(dis_rast_file, "N")


 ## extract from N
zonal_NA_sum <- extract(dis_rast, vect(watersheds_zonal_NA_centroid), method = 'simple') %>% ## extract zonal stats based on centroid point data 
  dplyr::select(tonnes_N)

# join extracted values with NA df
watersheds_zonal_NA_fixed <- data.frame(watersheds_zonal_NA, sum_new = zonal_NA_sum$tonnes_N) %>%
  mutate(sum = ifelse(is.na(sum), sum_new, sum)) %>%
  dplyr::select(-centroid, -sum_new) %>%
  st_as_sf() %>%
  mutate(sum = sum*area)


##  bind back together with overall zonal sf
watersheds_zonal_all_fixed <- watersheds_zonal %>%
  dplyr::filter(!is.na(sum)) %>%
  rbind(watersheds_zonal_NA_fixed)  %>%
  mutate(nutrient = nutrient_type) %>%
  dplyr::select(basin_id, nutrient, total_nutrient = sum) ## now we have a dataset with the columns we want

# sum(watersheds_zonal_all_fixed$total_nutrient, na.rm = TRUE) # 18014415

st_write(watersheds_zonal_all_fixed, file.path(paste0(prs_int_dir, "/zonal_shp_fix/", file)), delete_dsn  = TRUE)


}
```

**Combine watersheds zonal stats with pourpoints data**
```{r, eval = FALSE}
files_list <- str_subset(string = c(list.files(file.path(prs_int_dir, "zonal_shp_fix")), full = TRUE), ".shp")

coastal_pourpoints <- st_read(file.path(ww_intermediate_dir, "watersheds_coastal.shp"))

coastal_pp_ids <- unique(coastal_pourpoints$basin_id)


registerDoParallel(cores = 5)
foreach(file = files_list) %dopar% {
  # file <- files_list[15]
  
watersheds_zonal_all_fixed <- as.data.frame(st_read(file.path(prs_int_dir, "zonal_shp_fix", file))) %>%
  dplyr::select(-geometry, basin_id = 1, nutrient = 2, total_nutrient = 3) ## fix the column names, for some reason they were mispelled in the writing process

# length(unique(global_plume$basin_id)) # 142642
# length(unique(watersheds_zonal_all_fixed$basin_id)) # 142641

## there is one more watershed present in the pourpoint data than in the watershed data

## join these datasets together 
ws_pp_zonal_all <- left_join(watersheds_zonal_all_fixed, global_plume, by = "basin_id") %>%
  dplyr::select(-4, -5, -6)

# unique(ws_pp_zonal_all$basin_id)

n_occur <- data.frame(table(ws_pp_zonal_all$basin_id)) %>% ## check the duplicates (or triplicates)
      filter(Freq >1) ## filter for the duplicates in the data

dups <- ws_pp_zonal_all %>%
  filter(basin_id %in% n_occur$Var1) %>% ## duplicates... there appear to be multiple geometries for some basin_ids
st_as_sf()
  
# mapview(dups)

# dups_orig <- watersheds_zonal_all_fixed %>%
#   filter(basin_id %in% n_occur$Var1) ## compare to original data
# 
# dups_pourpoints <- global_plume %>%
#   filter(basin_id %in% n_occur$Var1) ## compare to pourpoints data 

## I've convinced myself this is nothing to worry about. Just select one of the multiple geometries and use that. They are all very close together. 

dups_fix <- dups %>%
  group_by(basin_id) %>%
  filter(row_number() == 1) %>% ## filter for the first row in each basin id group.
  ungroup() %>%
  as.data.frame()

`%notin%` <- Negate(`%in%`)

## remove the dups from joined data 
ws_pp_zonal_all_no_dups <- ws_pp_zonal_all %>%
  filter(basin_id %notin% dups$basin_id)

## add the dups back into the no dups file
ws_pp_zonal_all_fix <- ws_pp_zonal_all_no_dups %>%
  rbind(dups_fix) %>%
  st_as_sf() %>%
  rename("effluent" = "total_nutrient") %>%
  dplyr::select(-nutrient) %>%
  filter(basin_id %in% c(coastal_pp_ids)) %>% ## filter for only coastal pourpoints
  filter(!is.na(effluent)) %>%
  filter(effluent > 0) # filter for effluent greater than 0... no reason to run the plume model on no nutrients


  
# str(ws_pp_zonal_all_fix)
# 
# summary(ws_pp_zonal_all_fix)
# sum(ws_pp_zonal_all_fix$effluent, na.rm = TRUE) # 18035161


final_name <- gsub("^.*?disagg_", "pourpoints_", file)

st_write(ws_pp_zonal_all_fix, file.path(prs_int_dir, "pourpoints", final_name), delete_dsn = TRUE) 

# mapview(ws_pp_zonal_all_fix)
}


```

Now we need to copy over the appropriate ocean mask file from the wastewater project.. this won't change. So for next year, you can just copy over from the v2022 folder. 

``` {bash }

mkdir /home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2022/int/ocean_masks

cp /home/sgclawson/grassdata/ocean_mask.tif /home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2022/int/ocean_masks

```


