---
title: "OHI `r format(Sys.Date(), '%Y')` - Livelilihoods (liv): Labor Force Data Preparation"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

# Livelihoods: Proportion of Tourism Jobs per Country/Region per Year

# v2025 Updates:
- Added ILOSTAT data for gapfilling missing regions in World Bank data
- Added general gapfilling methods for World Bank and ILOSTAT data to this processing step. In previous assessments, there either was no gapfilling, or gapfilling was done in individual sector preps. Now, it get's completed here. 

## Overview

* **Data:** Labor Force & Employment Data

  * Labor Force data from World Bank, last updated 7/1/25 (downloaded August 7, 2025). This serves as the primary dataset for total labor force population data

      - https://data.worldbank.org/indicator/SL.TLF.TOTL.IN
      - Click the csv button on the bottom/middle right under the download heading.
      
  * ILOSTAT Labor Force data, (downloaded August 20, 2025). Secondary dataset to gapfill missing countries in world bank data

      - https://rshiny.ilo.org/dataexplorer38/?lang=en&segment=indicator&id=SDG_0831_SEX_ECO_RT_A 
      - Select indicator "Labour force by sex and age (thousands) - annual"
      - Left hand selection menus:
        - Sex: Total
        - Age: Aggregate bands: Total
        - Adjust time period slider to assessment years of interest. i.e. in v2025, we used years 2008-2023
      - Click the export button. Default selections should be fine, but you want "filtered", "label", and ".csv". Hit download. 


### Setup

```{r}
knitr::opts_chunk$set(eval = FALSE, warning = FALSE, message = FALSE)

# load packages
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  ohicore,
  here,
  janitor,
  terra,
  countrycode,
  readxl,
  RColorBrewer,
  foreach,
  tidyverse, 
  httr,
  plotly,
  zoo
)


# source 
source(here("workflow", "R", "common.R"))

# ---- set year and file path info ----
current_year <- 2025 # Update this in the future!!
previous_year <- 2024
data_year <-  2023

version_year <- paste0("v",current_year)
previous_vyear <- paste0("v", previous_year)

current_data_dir <- paste0("d", current_year)
previous_data_dir <- paste0("d", previous_year)

data_year <-  2023

# ---- data directories ----

# Raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")

# WorldBank raw data directory
current_wb_dir <- here(raw_data_dir, "WorldBank", current_data_dir)

# Livelihoods and Economies directory
current_le_dir <- here("globalprep", "le", version_year)
previous_le_dir <- here("globalprep", "le", previous_vyear)
```

Create country code --> iso3 function

```{r}
country_regex_to_iso3c <- function(country_string) {
  country_string %>%
    countrycode::countrycode(origin = "country.name", destination = "iso3c", origin_regex = TRUE)
}
```

# Labor Force Data

## World Bank 

```{r}
# ===================== Read in Data ==============================
# World Bank labor force data ----
labor_raw <- read_csv(here(current_wb_dir, "WorldBank_total_labor_force_1990-2024",
                                   "API_SL.TLF.TOTL.IN_DS2_en_csv_v2_38674.csv"),
                               skip = 4, # How many rows to skip before data starts,
                              # Manually open the file to confirm. changes year to year
                               na = "")

# OHI regions data ----
# read in OHI regions for joining
region_names <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/eez/spatial/regions_list.csv") 



# ==================== Tidy Data ===================================

# Labor force data 
labor_clean <- labor_raw %>% 
  # conver to lower_snake_case
  janitor::clean_names() %>% 
  # tidy data from wide to long (years are currently columns)
  pivot_longer(cols = 5:length(names(labor_raw)),
               names_to = "year", 
               values_to = "labor_force") %>% 
  # remove unnecessary columns
  select(-c(indicator_name, indicator_code)) %>% 
  # clean up year column (currently in the form of xYYYY)
  mutate(year = str_remove_all(year, pattern = "x")) %>% 
  # fix data types
  mutate(year = as.numeric(year),
         country_name = as.factor(country_name),
         labor_force = as.numeric(labor_force)) %>% 
  filter(year >= 2008) # starting 5 years back from first year of evaluation


# clean regions data
region_clean <- region_names %>% 
  janitor::clean_names() %>% 
  select(-c("notes"))


# Run ohicore name_2_rgn 
# check names
labor_rgn <- name_2_rgn(df_in = labor_clean, 
                       fld_name='country_name',
                      flds_unique = c("year"))

duplicates <- labor_rgn[duplicated(labor_rgn[, c("rgn_name", "year")]),]
unique(duplicates$country_name)
unique(duplicates$rgn_name)

# aggregating to address duplicates ----
labor_rgn_agg <- labor_rgn %>%
  group_by(rgn_id, rgn_name, year) %>%
  dplyr::summarize(labor_force = sum(labor_force, na.rm=TRUE)
            )%>%
  ungroup() %>% 
  dplyr::mutate(year = as.numeric(year)) %>% 
  mutate(labor_force = case_when(
    labor_force == 0 ~ NA, # replace 0s with NAs, because there are no true 0s in our dataset -- NAs became 0s in the summarize step earlier
    .default = labor_force
  ))

```

## Gapfill World Bank 

```{r}
# Labor force data
# created in liv_labor_force_dataprep.Rmd
# labor_force <- read_csv(here(current_le_dir, "int", "liv_labor_force.csv"))

# Gapfill total labor force values for missing years
world_bank_gf <- labor_rgn_agg %>% 
  # this data often goes one year past the current data year for other sources
  filter(year <= data_year) %>% 
  group_by(rgn_id, rgn_name) %>% 
  # Ensure we're hitting all years for joining later on
  complete(year = 2008:data_year) %>% 
  ungroup() %>%
  # First, create gapfilled boolean and method columns
  mutate(lf_gapfilled = ifelse(is.na(labor_force), "1", NA_character_),
         lf_gf_method = ifelse(is.na(labor_force), "1", NA_character_)) %>%
  # need to group again before gapfilling 
  group_by(rgn_id, rgn_name) %>% 
  arrange(rgn_id, year) %>% 
  # Linear interpolation
  mutate(labor_force = na.approx(labor_force, x = year, na.rm = FALSE)) %>% 
  mutate(lf_gapfilled = ifelse(lf_gapfilled == "1" & !is.na(labor_force), 
                               "gapfilled", lf_gapfilled),
         lf_gf_method = ifelse(lf_gf_method == "1" & !is.na(labor_force), 
                               "Interpolation", lf_gf_method)) %>% 
  # Upfilling and downfilling
  tidyr::fill(labor_force, .direction = "updown") %>% 
  mutate(lf_gapfilled = ifelse(lf_gapfilled == "1" & !is.na(labor_force), 
                               "gapfilled", lf_gapfilled),
         lf_gf_method = ifelse(lf_gf_method == "1" & !is.na(labor_force), 
                               "Up/Downfilling", lf_gf_method)) %>% 
  # Add data source
  mutate(source = "World Bank")

table(world_bank_gf$lf_gf_method) 
# 3 downfilling, 142 upfilling, (1 is a flag for missing here), no interpolation
colSums(is.na(world_bank_gf)) # 384 still missing

world_bank_gf %>% 
  filter(is.na(labor_force)) %>% 
  distinct(rgn_id, rgn_name) %>% 
  nrow() 
# 24 regions with NO labor force data 

world_bank_gf %>% 
  filter(!is.na(labor_force)) %>% 
  distinct(rgn_id, rgn_name) %>% 
  nrow()
# 142 regions WITH labor force data

# Drop NAs before combining with ILOSTAT data
world_bank_gf <- world_bank_gf %>% 
  drop_na(labor_force)

```


## ILOSTAT

```{r}
# Read in raw ILOSTAT
ilostat_labor <- read_csv(here(raw_data_dir, "ILOSTAT", current_data_dir,
                               "EAP_TEAP_SEX_AGE_NB_A-filtered-2025-08-21.csv"))

# Select columns and adjust units
ilostat_labor_clean <- ilostat_labor %>% 
  select(country = ref_area.label, year = time, labor_force = obs_value) %>% 
  mutate(labor_force = labor_force * 1000)
         # unit = "Persons (1)",
         # source = "ILOSTAT")

# Fix name_2_rgn error with Netherland Antillies
neth_ant <- ilostat_labor_clean %>%
    filter(country == 'Netherlands Antilles') %>% 
    mutate(labor_force = (labor_force / 5),  
           'Curacao' = labor_force,
           'Sint Maarten' = labor_force,
           'Bonaire' = labor_force,
           'Saint Eustatius' = labor_force,
           'Saba' = labor_force) %>%
    select(-c(labor_force, country)) %>%
    pivot_longer(cols = c("Curacao", "Sint Maarten", "Bonaire",
                          "Saint Eustatius", "Saba"),
                 names_to = "country",
                 values_to = "labor_force") %>%
    mutate(country = as.character(country),
           labor_force = as.numeric(labor_force)) %>% 
    select(c(country, labor_force, year))
  
ilostat_labor_clean <- ilostat_labor_clean %>%
    filter(country != 'Netherlands Antilles') %>%
    bind_rows(neth_ant)

# Add rgn names/ids
ilostat_rgn <- name_2_rgn(df_in = ilostat_labor_clean, 
                       fld_name='country',
                      flds_unique = c("year"))

# Check duplicates
duplicates <- ilostat_rgn[duplicated(ilostat_rgn[, c("rgn_name", "year")]),]
unique(duplicates$rgn_name)

# Aggregate duplicates
ilostat_rgn_agg <- ilostat_rgn %>%
  group_by(rgn_id, rgn_name, year) %>%
  dplyr::summarize(labor_force = sum(labor_force, na.rm=TRUE)
            )%>%
  ungroup() %>% 
  dplyr::mutate(year = as.numeric(year)) %>% 
  mutate(labor_force = case_when(
    labor_force == 0 ~ NA, # replace 0s with NAs, because there are no true 0s in our dataset -- NAs became 0s in the summarize step earlier
    .default = labor_force
  ))

# Confirm duplicates have been aggregated
duplicates <- ilostat_rgn_agg[duplicated(ilostat_rgn_agg[, c("rgn_name", "year")]),]
unique(duplicates$rgn_name)
```

## Gapfill ILOSTAT

```{r}
# Gapfill total labor force values for missing years
ilostat_gf <- ilostat_rgn_agg %>% 
  # this data often goes one year past the current data year for other sources
  filter(year <= data_year) %>% 
  group_by(rgn_id, rgn_name) %>% 
  # Ensure we're hitting all years for joining later on
  complete(year = 2008:data_year) %>% 
  ungroup() %>%
  # First, create gapfilled boolean and method columns
  mutate(lf_gapfilled = ifelse(is.na(labor_force), "1", NA_character_),
         lf_gf_method = ifelse(is.na(labor_force), "1", NA_character_)) %>%
  # need to group again before gapfilling 
  group_by(rgn_id, rgn_name) %>% 
  arrange(rgn_id, year) %>% 
  # Linear interpolation
  mutate(labor_force = na.approx(labor_force, x = year, na.rm = FALSE)) %>% 
  mutate(lf_gapfilled = ifelse(lf_gapfilled == "1" & !is.na(labor_force), 
                               "gapfilled", lf_gapfilled),
         lf_gf_method = ifelse(lf_gf_method == "1" & !is.na(labor_force), 
                               "Interpolation", lf_gf_method)) %>% 
  # Upfilling and downfilling
  tidyr::fill(labor_force, .direction = "updown") %>% 
  mutate(lf_gapfilled = ifelse(lf_gapfilled == "1" & !is.na(labor_force), 
                               "gapfilled", lf_gapfilled),
         lf_gf_method = ifelse(lf_gf_method == "1" & !is.na(labor_force), 
                               "Up/Downfilling", lf_gf_method)) %>% 
  # Add data source column 
  mutate(source = "ILOSTAT")

table(ilostat_gf$lf_gf_method) 
 # Interpolation Up/Downfilling 
 #           379            745 
colSums(is.na(ilostat_gf)) # 0 missing

ilostat_gf %>% 
  filter(!is.na(labor_force)) %>% 
  distinct(rgn_id, rgn_name) %>% 
  nrow()
# 164 regions WITH labor force data
```


# Join datasets

```{r}
# Find only the observations in ilostat NOT in wb
new_rgns <- anti_join(ilostat_gf, world_bank_gf, 
                      by = join_by(rgn_id, rgn_name, year))

# Bind new region data to world bank
total_labor_force <- bind_rows(world_bank_gf, new_rgns)

length(unique(total_labor_force$rgn_id)) # 170 regions total!
colSums(is.na(total_labor_force)) # no missing data

# Add final columns for export
labor_export <- total_labor_force %>%
  mutate(rgn_id = as.character(rgn_id),
         unit = "Persons (1)",
         sector = "liv") %>% 
  select(rgn_id, rgn_name, year, labor_force, unit, sector, source)

write_csv(labor_export, here(current_le_dir, "int", "liv_labor_force.csv"))

# Save gapfilling information
labor_gf <- total_labor_force %>% 
  select(rgn_id, rgn_name, year, lf_gapfilled, lf_gf_method, source)

write_csv(labor_gf, here(current_le_dir, "int", "liv_labor_force_gf.csv"))

```

# Data check

```{r}
# Read in data
new_le <- read_csv(here(current_le_dir, "int", "liv_labor_force.csv"))
old_le <- read_csv(here(previous_le_dir, "int", "liv_labor_force.csv"))

# Get max data years for plotting
current_data_year <- max(new_le$year)
current_data_year
previous_data_year <- max(old_le$year)
previous_data_year

colSums(is.na(new_le)) # 390
nrow(new_le) # 2656

colSums(is.na(old_le)) # 362
nrow(old_le) # 2490


# Combine data frames
comb_le <- new_le %>% 
  rename(new_labor_force = labor_force) %>% 
  left_join(old_le, by = join_by(rgn_id, year, rgn_name))

# Plot
comp_plot1 <- ggplot() + 
  geom_jitter(data = filter(comb_le, year == previous_data_year), aes(x = labor_force_total, y = new_labor_force, text = paste("Year:", year, "<br>Region:", rgn_name)), alpha = 0.7) +
  labs(x = paste("Previous Value", previous_data_year),
       y = paste("Current Value", previous_data_year),
       title = paste("Total Labor Population", previous_data_year, "Data", previous_year, "Assessment vs.", previous_data_year, "Data", current_year, "assessment")) +
  geom_abline(slope = 1, intercept = 0, col = "red", alpha = 0.7) +
  theme_bw();ggplotly(comp_plot1)
```


