---
title: "OHI `r format(Sys.Date(), '%Y')` - Livelilihoods Data Preparation"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

# Livelihoods: Proportion of Tourism Jobs per Country/Region per Year

Livelihoods includes two equally important sub-components, the **number of jobs**, which is a proxy for livelihood quantity, and the **per capita average annual wages**, which is a proxy for job quality. 

## Overview

* **Data:** Labor Force & Employment Data

  * Labor Force data from World Bank, last updated 7/1/25 (downloaded August 7, 2025).

      - <https://data.worldbank.org/indicator/SL.TLF.TOTL.IN>
      - Located in `globalprep/_raw_data/

  * Jobs data from UN Tourism / UNWTO (downloaded August 7, 2025)
  
      - Key Tourism Statistics <https://www.unwto.org/tourism-statistics/key-tourism-statistics>
      - According to the website, the latest update of the dataset took place in 31 January 2024. 
      
  * Quality/Wage data from ILOSTAT (downloaded August 7, 2025) -- for tour/cf wage data
  
      - <https://rshiny.ilo.org/dataexplorer46/?lang=en&id=EAR_4MTH_SEX_ECO_CUR_NB_A> 
      
  [OECD data](https://ec.europa.eu/eurostat/data/database)
  - This is the file path on the OECD website: `Data navigation tree/Detailed datasets/Industry, trade, and services/Tourism (tour)/Tourism industries - annual data (tour_inda)/Annual and quarterly data on employment for selected tourism industries (source: Labor force survey)(tour_lfsa)/Employed persons by full-time/part-time and NACE Rev. 2 activity (tour_lfs1r2).csv`

## Methods

### Setup

```{r}
knitr::opts_chunk$set(eval = FALSE, warning = FALSE, message = FALSE)

# ---- Load packages ----
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}

librarian::shelf(
  ohicore,
  here,
  janitor,
  terra,
  countrycode,
  readxl,
  RColorBrewer,
  foreach,
  tidyverse, 
  httr,
  plotly,
  zoo # for gapfilling
)

# ---- Source common.R ----
source(here("workflow", "R", "common.R"))

region_data() # for rgns_all and rgns_eez

# OHI regions data ----
region_names <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/eez/spatial/regions_list.csv") 

# ---- set year and file path info ----
current_year <- 2025 # Update this in the future!!
previous_year <- 2024

data_year <- 2023

version_year <- paste0("v",current_year)
previous_vyear <- paste0("v", previous_year)

current_data_dir <- paste0("d", current_year)
previous_data_dir <- paste0("d", previous_year)

# ---- data directories ----

# Raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")

# Livelihoods and Economies directory
current_le_dir <- here("globalprep", "le", version_year)
previous_le_dir <- here("globalprep", "le", previous_vyear)

# Create country code --> iso3 function
country_regex_to_iso3c <- function(country_string) {
  country_string %>%
    countrycode::countrycode(origin = "country.name", destination = "iso3c", origin_regex = TRUE)
}
```

labor_force: 2009 - 2024
un_jobs_data_raw: 2008 - 2023
ilo_wage_data: 2015 - 2025

v2025: Let's aim for 2008-2023


## Labor force gapfilling

```{r}
# Labor force data
# created in liv_labor_force_dataprep.Rmd
labor_force <- read_csv(here(current_le_dir, "int", "liv_labor_force.csv"))

# Gapfill total labor force values for missing years
labor_force_gf <- labor_force %>% 
  # this data often goes one year past the current data year for other sources
  filter(year <= data_year) %>% 
  group_by(rgn_id, rgn_name) %>% 
  # Ensure we're hitting all years for joining later on
  complete(year = 2008:data_year) %>% 
  ungroup() %>%
  # First, create gapfilled boolean and method columns
  mutate(lf_gapfilled = ifelse(is.na(labor_force_total), "1", NA_character_),
         lf_gf_method = ifelse(is.na(labor_force_total), "1", NA_character_),
         # Add some metadata back after expanding
         unit = "Persons (1)",
         sector = "liv",
         data_source = "World Bank") %>%
  # need to group again before gapfilling 
  group_by(rgn_id, rgn_name) %>% 
  arrange(rgn_id, year) %>% 
  # Linear interpolation
  mutate(labor_force_total = na.approx(labor_force_total, x = year, na.rm = FALSE)) %>% 
  mutate(lf_gapfilled = ifelse(lf_gapfilled == "1" & !is.na(labor_force_total), 
                               "gapfilled", lf_gapfilled),
         lf_gf_method = ifelse(lf_gf_method == "1" & !is.na(labor_force_total), 
                               "Interpolation", lf_gf_method)) %>% 
  # Upfilling
  tidyr::fill(labor_force_total, .direction = "up") %>% 
  mutate(lf_gapfilled = ifelse(lf_gapfilled == "1" & !is.na(labor_force_total), 
                               "gapfilled", lf_gapfilled),
         lf_gf_method = ifelse(lf_gf_method == "1" & !is.na(labor_force_total), 
                               "Upfilling", lf_gf_method)) %>% 
  # Downfilling
  tidyr::fill(labor_force_total, .direction = "down") %>% 
  mutate(lf_gapfilled = ifelse(lf_gapfilled == "1" & !is.na(labor_force_total), 
                               "gapfilled", lf_gapfilled),
         lf_gf_method = ifelse(lf_gf_method == "1" & !is.na(labor_force_total), 
                               "Downfilling", lf_gf_method))

# view(labor_force_gf)
table(labor_force_gf$lf_gf_method) 
# 3 downfilling, 142 upfilling, (1 is a flag for missing here), no interpolation
colSums(is.na(labor_force_gf)) # 384 still missing 

labor_force_gf %>% 
  filter(is.na(labor_force_total)) %>% 
  distinct(rgn_id, rgn_name) %>% 
  nrow() # view()
# 24 regions with NO labor force data 

labor_force_gf %>% 
  filter(!is.na(labor_force_total)) %>% 
  distinct(rgn_id, rgn_name) %>% 
  nrow()
# 142 regions WITH labor force data
```


# Employment (number of jobs)

(put this within previous chunk, read in data etc. for number of tourism jobs, join with labor_regions, make new proportion of tourism jobs column, group by year, do prelim plotting etc. and evaluate data quality etc.)

### UNWTO Data

```{r}
# Jobs data
# total number of people employed by the tourism sector, in thousands
un_jobs_data_raw <- read_xlsx(here(raw_data_dir, "UNWTO", current_data_dir,
                                   "UN_Tourism_8_9_2_employed_persons_04_2025",
                                   "UN_Tourism_8_9_2_employed_persons_04_2025.xlsx"),
                              # skip = 2, # skip first 2 rows
                              sheet = 2, # the sheet name has changed over time, confirm this
                              col_names = TRUE, # use first row as column names
                              na = c("","..")) # define NA markers

# Select columns of interest and add a unit column
un_jobs_data <- un_jobs_data_raw %>% 
  select(country = GeoAreaName, year = TimePeriod, jobs = Value) %>% 
  mutate(jobs = as.numeric(jobs)) 

# run ohicore name_2_rgn 
tour_jobs_rgn <- name_2_rgn(df_in = un_jobs_data, 
                       fld_name = 'country',
                       flds_unique = c("year"))

# Duplicates
# [1] "China, Hong Kong Special Administrative Region"
# [2] "China, Macao Special Administrative Region"
# [1] "China"

# aggregate: sum by rgn_name etc.
unwto_jobs <- tour_jobs_rgn %>%
  group_by(rgn_id, rgn_name, year) %>%
  dplyr::summarize(jobs = sum(jobs, na.rm = TRUE) 
            )%>%
  ungroup() %>% 
  filter(year >= 2008) %>% 
  # turn 0s into NAs (appropriate flag for lack of data)
  mutate(jobs = case_when(jobs == 0 ~ NA,
                          .default = jobs),
         unit = "Persons (1)",
         source = "UNWTO")

# Confirm duplicates are gone
tour_duplicates <- unwto_jobs[duplicated(unwto_jobs[, c("rgn_name", "year")]),]
unique(tour_duplicates$rgn_name)

length(unique(unwto_jobs$rgn_id)) # 73
```

### OECD Data

```{r}
# OECD data
oecd_jobs_data_raw <- read_csv(here(raw_data_dir, "OECD", 
                                   current_data_dir,
                                   "OECD_2009_2022_tourism_employment.csv"))

# Clean raw data
oecd_clean <- oecd_jobs_data_raw %>% 
  janitor::clean_names() %>% 
  # Filter to persons employed, not enterprises
  # AND total tourism activity, not sub categories
  filter(unit_measure == "PS" & activity == "_T") %>% 
  select(country_code = ref_area, country = reference_area, 
         unit = unit_of_measure, year = time_period, jobs = obs_value)

# run ohicore name_2_rgn
oecd_job_rgn <- name_2_rgn(df_in = oecd_clean, 
                       fld_name = 'country',
                       flds_unique = c("year"))

oecd_jobs <- oecd_job_rgn %>% 
  select(rgn_id, rgn_name, year, jobs, unit) %>% 
  mutate(unit = "Persons (1)",
         source = "OECD")

# view(oecd_jobs_clean)
unique(oecd_jobs$rgn_name)

# Adds Netherlands, United States, and Morocco!
# setdiff(oecd_jobs_clean$rgn_name, tourism_jobs_gf$rgn_name)
# 
# # Let's see how the magnitude changes between countries with data in both
# compare_oecd_unwto <- tourism_jobs_gf %>% 
#   full_join(oecd_jobs_clean, by = join_by(rgn_id, rgn_name, year))
# 
# shared_countries <- compare_oecd_unwto %>% 
#   filter(!is.na(tourism_jobs) & !is.na(value))
# view(shared_countries)
# 
# unique(shared_countries$rgn_name) # 10 countries shared
# 
# comp_plot <- ggplot(shared_countries) +
#   geom_point(aes(x = year, y = tourism_jobs), color = "red", alpha = 0.7) +
#   geom_point(aes(x = year, y = jobs), color = "blue", alpha = 0.7) +
#   facet_wrap(~rgn_name, #) +
#              scales = "free_y") +
#   labs(title = "UNWTO vs. OECD Data Shared Countries: Free y-axis",
#        y = "Jobs",
#        x = "Year") +
#   theme(panel.spacing.x = unit(0.01, "cm")) +
#   theme_bw();ggplotly(comp_plot)
```

### Eurostat Data

```{r}
# Eurostat data uses a weird mix of tabs and commas as delimiters, 
# so we do a bit of pre-processing here
txt <- readLines(here(raw_data_dir, "Eurostat", 
                      current_data_dir, "estat_tour_lfs1r2.tsv"))

# Replace commas or tabs with a single tab
txt_clean <- gsub("[\t,]", "\t", txt)

# Read in with consistent delimiter
eurostat_raw <- read.delim(textConnection(txt_clean), sep = "\t", header = TRUE)

# Preliminary cleaning
eurostat_clean <- eurostat_raw %>% 
  janitor::clean_names() %>% 
  mutate(country = countrycode(geo_time_period, 
                               origin = "eurostat", "country.name")) %>% 
  # Filtering down to specific employment codes and times that consider total tourism employment. This metadata was not easy to find. 
  filter(worktime == "TOTAL", wstatus == "EMP", nace_r2 == "TOTAL") %>% 
  # Remove the aggregated EU countries
  filter(!(geo_time_period %in% c("EA20", "EU27_2020"))) %>% 
  relocate(country, .after = unit) %>% 
  select(-c(freq, geo_time_period)) %>% 
  # tidy up year column names
  rename_with(~ gsub("x", "", .)) %>% 
  pivot_longer(cols = -c(nace_r2, wstatus, worktime, unit, country),
              names_to = "year", values_to = "value") %>% 
  # Fix weirdness in values
  mutate(value = str_remove_all(value, "[bd]"),
         value = str_replace_all(value, ":", NA_character_),
         country = str_replace_all(country, "&", "and"),
         value = as.numeric(value),
         year = as.numeric(year))
# Don't worry about the warning in countrycode, we drop those aggregated regions with no matches

# run ohicore name_2_rgn 
eurostat_rgn <- name_2_rgn(df_in = eurostat_clean, 
                       fld_name = 'country',
                       flds_unique = c("year"))

# Fix units and add some metadata
eurostat_jobs <- eurostat_rgn %>%
  mutate(jobs = value * 1000,
         unit = "Persons (1)",
         source = "Eurostat") %>% 
  select(rgn_id, rgn_name, year, jobs, unit, source)

# check what countries we've added:
# setdiff(eurostat_jobs$rgn_name, tourism_jobs_gf$rgn_name)
# # [1] "Germany"     "Estonia"     "Ireland"    
# # [4] "Iceland"     "Montenegro"  "Netherlands"
# # [7] "Norway" 
# # That's pretty good!
# 
# # Let's see how the magnitude changes between countries with data in both
# compare_euro_unwto <- tourism_jobs_gf %>% 
#   full_join(eurostat_jobs, by = join_by(rgn_id, rgn_name, year))
# 
# shared_countries <- compare_euro_unwto %>% 
#   filter(!is.na(tourism_jobs) & !is.na(jobs))
# 
# unique(shared_countries$rgn_name) # 21 countries shared
# 
# comp_plot <- ggplot(shared_countries) +
#   geom_point(aes(x = year, y = tourism_jobs), color = "red", alpha = 0.7) +
#   geom_point(aes(x = year, y = jobs), color = "blue", alpha = 0.7) +
#   facet_wrap(~rgn_name) +
#              # scales = "free_y") +
#   labs(title = "UNWTO vs. Eurostat Data Shared Countries: Free y-axis",
#        y = "Jobs",
#        x = "Year") +
#   theme(panel.spacing.x = unit(0.01, "cm")) +
#   theme_bw();ggplotly(comp_plot)
```

## Join all tourism jobs together

```{r}
# view(unwto_jobs)
# view(eurostat_jobs)
# view(oecd_jobs)
# 
# colSums(is.na(unwto_jobs)) # 10
# colSums(is.na(eurostat_jobs)) # 25 
# colSums(is.na(oecd_jobs)) # 0
# 
# colnames(unwto_jobs)
# colnames(eurostat_jobs)
# colnames(oecd_jobs)

# Find all observations in oecd data NOT in unwto data
new_oecd_jobs <- anti_join(oecd_jobs, unwto_jobs, by = join_by(rgn_id, rgn_name, year)) 
view(new_oecd_jobs)

# Bind unwto and new oecd data
unwto_oecd <- bind_rows(unwto_jobs, new_oecd_jobs) 

# Find all observations in eurostat data NOT in unwto+oeced data
new_eurostat_jobs <- anti_join(eurostat_jobs, unwto_oecd, by = join_by(rgn_id, rgn_name, year))

# Bind eurostat with previous two
tourism_jobs <- bind_rows(unwto_oecd, new_eurostat_jobs) 

# Check NAs
colSums(is.na(tourism_jobs)) # 10
nrow(tourism_jobs) # 1007
length(unique(tourism_jobs$rgn_name)) # 82
# Increased regions from 73 to 82, not bad
```

### Gapfill tourism jobs

```{r}
# Gapfill total labor force values for missing years
tourism_jobs_gf <- tourism_jobs %>% 
  group_by(rgn_id, rgn_name) %>% 
  # Ensure we're hitting all years for joining later on
  complete(year = 2008:data_year) %>% 
  ungroup() %>%
  # First, create gapfilled boolean and method columns
  mutate(tr_gapfilled = ifelse(is.na(jobs), "1", NA_character_),
         tr_gf_method = ifelse(is.na(jobs), "1", NA_character_),
         unit = "Persons (1)") %>%
  # Regroup for gapfilling
  group_by(rgn_id, rgn_name) %>% 
  arrange(rgn_id, year) %>%   
  # Linear interpolation
  mutate(jobs = na.approx(jobs, x = year, na.rm = FALSE)) %>% 
  mutate(tr_gapfilled = ifelse(tr_gapfilled == "1" & !is.na(jobs), 
                               "gapfilled", tr_gapfilled),
         tr_gf_method = ifelse(tr_gf_method == "1" & !is.na(jobs), 
                               "Interpolation", tr_gf_method)) %>% 
  # Upfilling
  tidyr::fill(jobs, .direction = "up") %>% 
  mutate(tr_gapfilled = ifelse(tr_gapfilled == "1" & !is.na(jobs), 
                               "gapfilled", tr_gapfilled),
         tr_gf_method = ifelse(tr_gf_method == "1" & !is.na(jobs), 
                               "Upfilling", tr_gf_method)) %>% 
  # Downfilling
  tidyr::fill(jobs, .direction = "down") %>% 
  mutate(tr_gapfilled = ifelse(tr_gapfilled == "1" & !is.na(jobs), 
                               "gapfilled", tr_gapfilled),
         tr_gf_method = ifelse(tr_gf_method == "1" & !is.na(jobs), 
                               "Downfilling", tr_gf_method))

view(tourism_jobs_gf)
table(tourism_jobs_gf$tr_gf_method)
# Downfilling Interpolation     Upfilling 
#          94            58           191 
colSums(is.na(tourism_jobs_gf)) # 0 NAs
# Source is gonna have NAs due to the gapfilling
nrow(tourism_jobs_gf) # 1340 observations
```

## Labor force join

```{r}
# join with labor force data 
# COME BACK -- rejoin once labor force has been updated
tour_jobs_labor <- full_join(tourism_jobs_gf, labor_force_gf , by = c("rgn_id", "rgn_name", "year", "unit"))  %>% 
  select(-c(sector, data_source)) # these currently only show Labor Force data-related values. Will add back on before exporting.

view(tour_jobs_labor)

# calculate tourism jobs proportion of total labor force 
tour_jobs_prop <- tour_jobs_labor %>% 
  mutate(tour_jobs_prop = (jobs / labor_force_total)) %>% 
  mutate(rgn_id = as.numeric(rgn_id)) 

summary(tour_jobs_prop$tour_jobs_prop)

colSums(is.na(tour_jobs_prop)) 
 #         jobs      labor_force_total      tour_jobs_prop 
 #         1344                   412                 1420

nrow(tour_jobs_prop) # 2684
nrow(tourism_jobs_gf) # 1340
nrow(labor_force_gf) # 2656
# ~ 71% NA
# After first round of gapfilling, dropped to 56% 
# After adding new data sources, down to 52%

tour_jobs_prop %>% 
  filter(is.na(jobs) & is.na(labor_force_total)) %>% 
  distinct(rgn_id, rgn_name) %>% 
  nrow()

# 21 countries with no data for jobs or labor force, we'll end up dropping these for now
tour_jobs_final <- tour_jobs_prop %>% 
  filter(!is.na(jobs),
         !is.na(labor_force_total)) 

length(unique(tour_jobs_final$rgn_id)) # 79 final regions

colSums(is.na(tour_jobs_final))
nrow(tour_jobs_final) # 1264
view(tour_jobs_prop)

# write_csv(tour_jobs_final, here(current_le_dir, "int", "tourism_jobs_unwto_gf.csv"))
```

```{r}
df_with_missing <- tour_jobs_prop %>%
  group_by(rgn_id, rgn_name) %>%
  filter(any(is.na(jobs))) %>%
  arrange(rgn_name, year)

View(df_with_missing)
df_with_missing %>% 
  filter(!is.na(jobs)) %>% view()

test <- tour_jobs_prop %>% 
  # group_by(rgn_id, rgn_name) %>% 
  # Ensure we're hitting all years for joining later on
  # complete(year = 2008:data_year) %>% 
  # ungroup() %>%
  # First, create gapfilled boolean and method columns
  # mutate(tr_gapfilled = ifelse(is.na(jobs), "1", NA_character_),
  #        tr_gf_method = ifelse(is.na(jobs), "1", NA_character_),
  #        unit = "Persons (1)") %>%
  # Regroup for gapfilling
  group_by(rgn_id, rgn_name) %>% 
  arrange(rgn_id, year) %>%   
  # Linear interpolation
  mutate(jobs = na.approx(jobs, x = year, na.rm = FALSE)) %>% 
  mutate(tr_gapfilled = ifelse(tr_gapfilled == "1" & !is.na(jobs), 
                               "gapfilled", tr_gapfilled),
         tr_gf_method = ifelse(tr_gf_method == "1" & !is.na(jobs), 
                               "Interpolation", tr_gf_method)) %>% 
  # Upfilling
  tidyr::fill(jobs, .direction = "up") %>% 
  mutate(tr_gapfilled = ifelse(tr_gapfilled == "1" & !is.na(jobs), 
                               "gapfilled", tr_gapfilled),
         tr_gf_method = ifelse(tr_gf_method == "1" & !is.na(jobs), 
                               "Upfilling", tr_gf_method)) %>% 
  # Downfilling
  tidyr::fill(jobs, .direction = "down") %>% 
  mutate(tr_gapfilled = ifelse(tr_gapfilled == "1" & !is.na(jobs), 
                               "gapfilled", tr_gapfilled),
         tr_gf_method = ifelse(tr_gf_method == "1" & !is.na(jobs), 
                               "Downfilling", tr_gf_method))
```


Create the proportion between tourism jobs and the total labor force

```{r}
# tourism_job_proportion <- tourism_labor_join %>%
#   dplyr::relocate(tourism_jobs, .before = labor_force_total) %>%
#   mutate(tourism_proportion = tourism_jobs/labor_force_total) %>%
#   rename(labor_force_total = labor_force_total) %>%
#   select(rgn_id, rgn_name.x, country, year, labor_force_total, tourism_jobs, tourism_proportion)
colnames(tour_jobs_prop_df)
 max(tour_jobs_prop_df$year)

# plot it, interactively!
line_plot <- plotly::plot_ly(tour_jobs_prop, x = ~year, y = ~tour_jobs_prop, color = ~rgn_name, type = "scatter", mode = "lines") %>% 
  layout(title = "All Regions: Proportional Tourism Employment Within Total Labor Force", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Percent of people by region employed by the tourism sector"))

line_plot

# htmlwidgets::saveWidget(line_plot, file = "prop_tourism_laborforce.html")
```

### Saving intermediate data

```{r}
old_tour_jobs <- read_csv(here(previous_le_dir, "int", "tour_jobs.csv"))
colSums(is.na(old_tour_jobs)) # 1856
nrow(old_tour_jobs) # 2671
# ~ 70% NA

write_csv(tourism_job_proportion, here(int_dir, "tour_jobs.csv"))
```


# Quality (ILO wage data)
  - note: this data came from many sources and was compiled by ILO

```{r}
# ILO wage data ----
ilo_wage_data <- read_csv(here(raw_data_dir, "ILOSTAT", current_data_dir, "EAR_4MTH_SEX_ECO_CUR_NB_A-filtered-2025-08-07.csv"))


# preliminary cleanings
wage_data_clean <- ilo_wage_data %>% 
  # lower_snake_case
  janitor::clean_names() %>% 
  # more intuitive value name
  rename(monthly_wage = obs_value) %>% 
  # filter to PPP adjusted data (this accounts for inflation, diff in currencies globally, cost of living etc.) NEED TO CHECK WHAT THE YEAR IS ON YOUR DATASET
  # Why are we using 2017 here?
  filter(classif2_label == "Currency: 2021 PPP $") %>%
  # group by country/region
  group_by(ref_area_label) %>%
  # set column name to year
  rename(year = time) %>%
  # filter to the cap date range from the jobs data --
  filter(year %in% c(2008:max(tour_jobs_prop_df$year))) # %>%
  # add iso3 column, tourism sector label
  # mutate(iso3 = country_regex_to_iso3c(ref_area_label),
         # sector = "tour") %>%
  # select(c(ref_area_label, iso3, year, monthly_wage, classif2_label))

#   CHECK KOSOVO FOR COUNTRY REGEX TO ISO3C
view(wage_data_clean)
table(ilo_wage_data$classif2.label)
colnames(ilo_wage_data)

# make sequence of years for gapfilling ----
# note: data does not have wage data for some years in some countries -- we want to account for that by leaving them as NAs, then gapfilling with the average between the pre and post years

years_df <- tibble(iso3 = wage_data_clean$iso3) %>% 
  group_by(iso3) %>% 
  summarize(year = seq(2008, 2023))


wage_data_years <- left_join(years_df, wage_data_clean, by = c("iso3", "year"))

# filling in country names when possible --- 
wage_years_filled <- wage_data_years %>% 
  group_by(iso3) %>% 
  fill(ref_area_label, .direction = "downup") %>% 
  ungroup()

# =================

# test gapfilling
# gap_fill_test <- wage_years_filled %>% 
#   mutate(ref_area_label = as.factor(ref_area_label)) %>% 
#   mutate(lm_est = list(lm(monthly_wage ~ year + ref_area_label)))
# 
# 
# lm_test <- lm(monthly_wage ~ year + ref_area_label, data = wage_years_filled)
# summary(lm_test)
# 
# lm_test$coefficients 

# wage_years_filled$lm_values <- lm_test$fitted.values

# gap filling  ----
wage_gf <- wage_years_filled %>% 
  group_by(ref_area_label) %>% 
  # interpolate (fill missing values between 2 values)
  mutate(appx_wage = zoo::na.approx(monthly_wage, # using values in this column
                                    na.rm = FALSE, # don't replace (internal) NAs in new column that can't be approximated
                                    #  extrapolate using rule = 2 from approx(),
                                    # uses closest data extreme to extrapolate 
                                    # for leading and trailing NAs
                                    rule = 2)) %>% 
  
  # fill NA values with the only other observed value (making sure to note the list of countries/regions we did this for)
  mutate(appx_wage_fill = zoo::na.aggregate(appx_wage, by = 1, FUN = mean, na.rm = FALSE))
# note: made this a new column so that we can still identify each country that we are using this fill (na.aggregate function) method for.
  
# other notes: 
# need more than 1 data point to do any approximation. 
# also, extremes (leading and trailing) are just copied data points from nearest extreme

```



```{r}
# find regions with only 1 data point (still have NAs in appx_wage column)

na_regions <- wage_gf %>% 
  filter(is.na(appx_wage))


unique(na_regions$ref_area_label)
num_na <- length(unique(na_regions$ref_area_label))
num_tot <- length(unique(wage_gf$ref_area_label))

paste0("proportion of countries/regions with only 1 data point: ", round(((num_na / num_tot) * 100), 3), "%")
```




```{r}
# preliminary plot 
line_plot <- plotly::plot_ly(wage_gf, x = ~year, y = ~appx_wage_fill, color = ~ref_area_label, type = "scatter", mode = "lines") %>% 
  layout(title = "All Regions: monthly Service Wages (USD (2017) PPP-Adjusted)", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "monthly Wages (USD)"))

line_plot

# htmlwidgets::saveWidget(line_plot, file = "prop_tourism_laborforce.html")

```




In this plot we can see that Sao Tome and Principe is a clear high outlier with no variation -- it is one of the countries for which we gapfilled with copied values from a single observation. We will continue to join the data with the OHI regions, then plot again and reinvestigate.


Clean up ILO data, join with OHI regions

```{r}
# join with OHI regions
wage_region_join <- left_join(region_clean, wage_gf, by = c("eez_iso3" = "iso3"))

# come back ---- stopping point for Mel to take a look!

# Currency: 2017 PPP $

wage_regions <- wage_region_join %>% 
  mutate(unit = "Currency: 2017 PPP $") %>% 
  select(-classif2_label) %>% 
  mutate(sector = "tour",
         data_source = "ILO")

```


Plot

```{r}
# interactive plot after joining with OHI Regions
plotly::plot_ly(wage_regions, x = ~year, y = ~appx_wage_fill, color = ~admin_country_name, type = "scatter", mode = "lines") %>% 
  layout(title = "All Regions: Monthly Service Wages (USD)", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Monthly Wages (USD)"))

```

We can see that Sao Tome and Principe is still included and is still this high, unvarying outlier. 
This lack of data combined with how high of an outlier is led us to reconsider our methods for this country and consider dropping it from this intermediate data product. 

Let's drop & plot what that looks like. 

```{r}
wage_regions_no_sp <- wage_regions %>% 
  filter(!ref_area_label %in% c("Sao Tome and Principe"))

# plot to see what it looks like after dropping Sao Tome and Principe
plotly::plot_ly(wage_regions_no_sp, x = ~year, y = ~monthly_wage, color = ~admin_country_name, type = "scatter", mode = "lines") %>% 
  layout(title = "All Regions: Monthly Service Wages (USD)", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Monthly Wages (USD)"))

```

Dropping Sao Tome and Principe (STP) adjusted the y-axis scale significantly to better visualize the monthly service wage data (USD) for other geographic areas.  Investigating further, it looks as though the Sao Tome and Principe observation from 2017 (133814.9 USD) may have been inputted incorrectly, measured as annual income instead of monthly, or there was an error when it was adjusted by PPP.  Dustin noticed that the data source from this observation was "HIES - Household Budget Survey".  This was the only observation from this data source. Upon further investigation, we discovered the average annual income in STP is ~2,400 USD [WorldData]("https://www.worlddata.info/africa/sao-tome-and-principe/index.php#:~:text=With%20an%20average%20annual%20income,the%20lower%20middle%2Dincome%20countries") and a GDP per capita of ~2,817 USD [World Bank]("https://www.worldbank.org/en/country/saotome/overview"). Therefore, we have decided to drop it for the intermediate data set.


## Saving intermediate data products

```{r}
# write_csv(wage_regions_no_sp, here(int_dir, "tour_quality.csv"))

```

