---
title: "OHI `r format(Sys.Date(), '%Y')` - Livelilihoods Data Preparation"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

# Livelihoods: Proportion of Tourism Jobs per Country/Region per Year

Livelihoods includes two equally important sub-components, the **number of jobs**, which is a proxy for livelihood quantity, and the **per capita average annual wages**, which is a proxy for job quality. 

## Overview

* **Data:** Labor Force & Employment Data

  * Labor Force data from World Bank, last updated 7/1/25 (downloaded August 7, 2025).

      - <https://data.worldbank.org/indicator/SL.TLF.TOTL.IN>
      - Located in `globalprep/_raw_data/

  * Jobs data from UN Tourism / UNWTO (downloaded August 7, 2025)
  
      - Key Tourism Statistics <https://www.unwto.org/tourism-statistics/key-tourism-statistics>
      - According to the website, the latest update of the dataset took place in 31 January 2024. 
      
  * Quality/Wage data from ILOSTAT (downloaded August 7, 2025) -- for tour/cf wage data
  
      - <https://rshiny.ilo.org/dataexplorer46/?lang=en&id=EAR_4MTH_SEX_ECO_CUR_NB_A> 
      
  - navigate to ILOSTAT explorer, select:

  - “Sex”: select “none” (to un-select all options), then select “Total”

  - “Economic Activity”: select “none” and then select “ISIC-Rev.4: I. Accommodation and food service activities” AND "ISIC-Rev.4: I. Arts, entertainment and recreation"

  - “Currency”: keep all selected (we end up filtering this in R)

  - set Time Frame (sliding range bar) to 2008-present year

  - Then click “Export”, and for the file type option (appears as a drop-down menu), select “csv”.
      
  [OECD data](https://ec.europa.eu/eurostat/data/database)
  - This is the file path on the OECD website: `Data navigation tree/Detailed datasets/Industry, trade, and services/Tourism (tour)/Tourism industries - annual data (tour_inda)/Annual and quarterly data on employment for selected tourism industries (source: Labor force survey)(tour_lfsa)/Employed persons by full-time/part-time and NACE Rev. 2 activity (tour_lfs1r2).csv`

## Methods

### Setup

```{r}
knitr::opts_chunk$set(eval = FALSE, warning = FALSE, message = FALSE)

# ---- Load packages ----
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}

librarian::shelf(
  ohicore,
  here,
  janitor,
  terra,
  countrycode,
  readxl,
  RColorBrewer,
  foreach,
  tidyverse, 
  httr,
  plotly,
  zoo # for gapfilling
)

# ---- Source common.R ----
source(here("workflow", "R", "common.R"))

region_data() # for rgns_all and rgns_eez

# OHI regions data ----
region_names <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/eez/spatial/regions_list.csv") 

# ---- set year and file path info ----
current_year <- 2025 # Update this in the future!!
previous_year <- 2024

data_year <- 2023

version_year <- paste0("v",current_year)
previous_vyear <- paste0("v", previous_year)

current_data_dir <- paste0("d", current_year)
previous_data_dir <- paste0("d", previous_year)

# ---- data directories ----

# Raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")

# Livelihoods and Economies directory
current_le_dir <- here("globalprep", "le", version_year)
previous_le_dir <- here("globalprep", "le", previous_vyear)

# Create country code --> iso3 function
country_regex_to_iso3c <- function(country_string) {
  country_string %>%
    countrycode::countrycode(origin = "country.name", destination = "iso3c", origin_regex = TRUE)
}
```

labor_force: 2009 - 2024
un_jobs_data_raw: 2008 - 2023
ilo_wage_data: 2015 - 2025

v2025: Let's aim for 2008-2023


## Labor force gapfilling

```{r}
# Labor force data
# created in liv_labor_force_dataprep.Rmd
labor_force <- read_csv(here(current_le_dir, "int", "liv_labor_force.csv"))

# Gapfill total labor force values for missing years
labor_force_gf <- labor_force %>% 
  # this data often goes one year past the current data year for other sources
  filter(year <= data_year) %>% 
  group_by(rgn_id, rgn_name) %>% 
  # Ensure we're hitting all years for joining later on
  complete(year = 2008:data_year) %>% 
  ungroup() %>%
  # First, create gapfilled boolean and method columns
  mutate(lf_gapfilled = ifelse(is.na(labor_force_total), "1", NA_character_),
         lf_gf_method = ifelse(is.na(labor_force_total), "1", NA_character_),
         # Add some metadata back after expanding
         unit = "Persons (1)",
         sector = "liv",
         data_source = "World Bank") %>%
  # need to group again before gapfilling 
  group_by(rgn_id, rgn_name) %>% 
  arrange(rgn_id, year) %>% 
  # Linear interpolation
  mutate(labor_force_total = na.approx(labor_force_total, x = year, na.rm = FALSE)) %>% 
  mutate(lf_gapfilled = ifelse(lf_gapfilled == "1" & !is.na(labor_force_total), 
                               "gapfilled", lf_gapfilled),
         lf_gf_method = ifelse(lf_gf_method == "1" & !is.na(labor_force_total), 
                               "Interpolation", lf_gf_method)) %>% 
  # Upfilling and downfilling
  tidyr::fill(labor_force_total, .direction = "updown") %>% 
  mutate(lf_gapfilled = ifelse(lf_gapfilled == "1" & !is.na(labor_force_total), 
                               "gapfilled", lf_gapfilled),
         lf_gf_method = ifelse(lf_gf_method == "1" & !is.na(labor_force_total), 
                               "Up/Downfilling", lf_gf_method)) 

# view(labor_force_gf)
table(labor_force_gf$lf_gf_method) 
# 3 downfilling, 142 upfilling, (1 is a flag for missing here), no interpolation
colSums(is.na(labor_force_gf)) # 384 still missing 

labor_force_gf %>% 
  filter(is.na(labor_force_total)) %>% 
  distinct(rgn_id, rgn_name) %>% 
  nrow() 
# 24 regions with NO labor force data 

labor_force_gf %>% 
  filter(!is.na(labor_force_total)) %>% 
  distinct(rgn_id, rgn_name) %>% 
  nrow()
# 142 regions WITH labor force data
```


# Quantity (number of jobs)

(put this within previous chunk, read in data etc. for number of tourism jobs, join with labor_regions, make new proportion of tourism jobs column, group by year, do prelim plotting etc. and evaluate data quality etc.)

### UNWTO Data

```{r}
# Jobs data
# total number of people employed by the tourism sector, in thousands
un_jobs_raw <- read_xlsx(here(raw_data_dir, "UNWTO", current_data_dir,
                              "UN_Tourism_8_9_2_employed_persons_04_2025",
                              "UN_Tourism_8_9_2_employed_persons_04_2025.xlsx"),
                          # skip = 2, # skip first 2 rows
                          sheet = 2, # sheet name has changed over time, check this
                          col_names = TRUE, # use first row as column names
                          na = c("","..")) # define NA markers

# Select columns of interest and add a unit column
un_jobs_data <- un_jobs_raw %>% 
  select(country = GeoAreaName, year = TimePeriod, jobs = Value) %>% 
  mutate(jobs = as.numeric(jobs)) 

# run ohicore name_2_rgn 
tour_jobs_rgn <- name_2_rgn(df_in = un_jobs_data, 
                       fld_name = 'country',
                       flds_unique = c("year"))

# Duplicates
# [1] "China, Hong Kong Special Administrative Region"
# [2] "China, Macao Special Administrative Region"
# [3] "China"

# aggregate: sum by rgn_name etc.
unwto_jobs <- tour_jobs_rgn %>%
  group_by(rgn_id, rgn_name, year) %>%
  dplyr::summarize(jobs = sum(jobs, na.rm = TRUE) 
            )%>%
  ungroup() %>% 
  filter(year >= 2008) %>% 
  # turn 0s into NAs (appropriate flag for lack of data)
  mutate(jobs = case_when(jobs == 0 ~ NA,
                          .default = jobs),
         unit = "Persons (1)",
         source = "UNWTO")

# Confirm duplicates are gone
tour_duplicates <- unwto_jobs[duplicated(unwto_jobs[, c("rgn_name", "year")]),]
unique(tour_duplicates$rgn_name)

length(unique(unwto_jobs$rgn_id)) # 73
```

### OECD Data

```{r}
# OECD data
oecd_jobs_data_raw <- read_csv(here(raw_data_dir, "OECD", 
                                   current_data_dir,
                                   "OECD_2009_2022_tourism_employment.csv"))

# Clean raw data
oecd_clean <- oecd_jobs_data_raw %>% 
  janitor::clean_names() %>% 
  # Filter to persons employed, not enterprises
  # AND total tourism activity, not sub categories
  filter(unit_measure == "PS" & activity == "_T") %>% 
  select(country_code = ref_area, country = reference_area, 
         unit = unit_of_measure, year = time_period, jobs = obs_value)

# run ohicore name_2_rgn
oecd_job_rgn <- name_2_rgn(df_in = oecd_clean, 
                       fld_name = 'country',
                       flds_unique = c("year"))

# Select columns of interest
oecd_jobs <- oecd_job_rgn %>% 
  select(rgn_id, rgn_name, year, jobs, unit) %>% 
  mutate(unit = "Persons (1)",
         source = "OECD")

# Countries added
setdiff(oecd_jobs$rgn_name, unwto_jobs$rgn_name)
# [1] "Netherlands"   "United States"
# [3] "Morocco"  
```

### Eurostat Data

```{r}
# Eurostat data uses a weird mix of tabs and commas as delimiters, 
# so we do a bit of pre-processing here
txt <- readLines(here(raw_data_dir, "Eurostat", 
                      current_data_dir, "estat_tour_lfs1r2.tsv"))

# Replace commas or tabs with a single tab
txt_clean <- gsub("[\t,]", "\t", txt)

# Read in with consistent delimiter
eurostat_raw <- read.delim(textConnection(txt_clean), sep = "\t", header = TRUE)

# Preliminary cleaning
eurostat_clean <- eurostat_raw %>% 
  janitor::clean_names() %>% 
  mutate(country = countrycode(geo_time_period, 
                               origin = "eurostat", "country.name")) %>% 
  # Filtering down to specific employment codes and times that consider total tourism employment. This metadata was not easy to find. 
  filter(worktime == "TOTAL", wstatus == "EMP", nace_r2 == "TOTAL") %>% 
  # Remove the aggregated EU countries
  filter(!(geo_time_period %in% c("EA20", "EU27_2020"))) %>% 
  relocate(country, .after = unit) %>% 
  select(-c(freq, geo_time_period)) %>% 
  # tidy up year column names
  rename_with(~ gsub("x", "", .)) %>% 
  pivot_longer(cols = -c(nace_r2, wstatus, worktime, unit, country),
              names_to = "year", values_to = "value") %>% 
  # Fix weirdness in values
  mutate(value = str_remove_all(value, "[bd]"),
         value = str_replace_all(value, ":", NA_character_),
         country = str_replace_all(country, "&", "and"),
         value = as.numeric(value),
         year = as.numeric(year))
# Don't worry about the warning in countrycode, we drop those aggregated regions with no matches

# run ohicore name_2_rgn 
eurostat_rgn <- name_2_rgn(df_in = eurostat_clean, 
                       fld_name = 'country',
                       flds_unique = c("year"))

# Fix units and add some metadata
eurostat_jobs <- eurostat_rgn %>%
  mutate(jobs = value * 1000,
         unit = "Persons (1)",
         source = "Eurostat") %>% 
  select(rgn_id, rgn_name, year, jobs, unit, source)

# check what countries we've added:
setdiff(eurostat_jobs$rgn_name, unwto_jobs$rgn_name)
# [1] "Germany"     "Estonia"     "Ireland"
# [4] "Iceland"     "Montenegro"  "Netherlands"
# [7] "Norway"

```

## Join all tourism jobs together

```{r}
# Find all observations in oecd data NOT in unwto data
new_oecd_jobs <- anti_join(oecd_jobs, unwto_jobs, by = join_by(rgn_id, rgn_name, year)) 

# Bind unwto and new oecd data
unwto_oecd <- bind_rows(unwto_jobs, new_oecd_jobs) 

# Find all observations in eurostat data NOT in unwto+oeced data
new_eurostat_jobs <- anti_join(eurostat_jobs, unwto_oecd, by = join_by(rgn_id, rgn_name, year))

# Bind eurostat with previous two
tourism_jobs <- bind_rows(unwto_oecd, new_eurostat_jobs) 

# Check NAs
colSums(is.na(tourism_jobs)) # 10
nrow(tourism_jobs) # 1007
# Only ~10% NA
length(unique(tourism_jobs$rgn_name)) # 82
# Increased regions from 73 to 82, not bad
```

### Gapfill tourism jobs

```{r}
# Gapfill total labor force values for missing years
tourism_jobs_gf <- tourism_jobs %>% 
  group_by(rgn_id, rgn_name) %>% 
  # Ensure we're hitting all years for joining later on
  complete(year = 2008:data_year) %>% 
  ungroup() %>%
  # First, create gapfilled boolean and method columns
  mutate(tr_gapfilled = ifelse(is.na(jobs), "1", NA_character_),
         tr_gf_method = ifelse(is.na(jobs), "1", NA_character_),
         unit = "Persons (1)") %>%
  # Regroup for gapfilling
  group_by(rgn_id, rgn_name) %>% 
  arrange(rgn_id, year) %>%   
  # Linear interpolation
  mutate(jobs = na.approx(jobs, x = year, na.rm = FALSE)) %>% 
  mutate(tr_gapfilled = ifelse(tr_gapfilled == "1" & !is.na(jobs), 
                               "gapfilled", tr_gapfilled),
         tr_gf_method = ifelse(tr_gf_method == "1" & !is.na(jobs), 
                               "Interpolation", tr_gf_method)) %>% 
  # Upfilling
  tidyr::fill(jobs, .direction = "updown") %>% 
  mutate(tr_gapfilled = ifelse(tr_gapfilled == "1" & !is.na(jobs), 
                               "gapfilled", tr_gapfilled),
         tr_gf_method = ifelse(tr_gf_method == "1" & !is.na(jobs), 
                               "Up/Downfilling", tr_gf_method))


table(tourism_jobs_gf$tr_gf_method)
# Interpolation Up/Downfilling 
#            58            285 

colSums(is.na(tourism_jobs_gf)) # 0 NAs
# Source is gonna have NAs due to the gapfilling

```

## Labor force join

```{r}
# join with labor force data 
tour_jobs_labor <- full_join(tourism_jobs_gf, labor_force_gf , by = c("rgn_id", "rgn_name", "year", "unit"))  %>% 
  select(-c(sector, data_source)) # these currently only show Labor Force data-related values. Will add back on before exporting.


# calculate tourism jobs proportion of total labor force 
tour_jobs_prop <- tour_jobs_labor %>% 
  mutate(tour_jobs_prop = (jobs / labor_force_total)) %>% 
  mutate(rgn_id = as.numeric(rgn_id)) 

summary(tour_jobs_prop$tour_jobs_prop)

colSums(is.na(tour_jobs_prop)) 
 #         jobs      labor_force_total      tour_jobs_prop 
 #         1344                   412                 1420

nrow(tour_jobs_prop) # 2684
nrow(tourism_jobs_gf) # 1340
nrow(labor_force_gf) # 2656
# ~ 71% NA
# After first round of gapfilling, dropped to 56% 
# After adding new data sources, down to 52%

tour_jobs_prop %>% 
  filter(is.na(jobs) & is.na(labor_force_total)) %>% 
  distinct(rgn_id, rgn_name) %>% 
  nrow()

# 21 countries with no data for jobs or labor force, we'll end up dropping these for now
tour_jobs_final <- tour_jobs_prop %>% 
  filter(!is.na(jobs),
         !is.na(labor_force_total)) 

length(unique(tour_jobs_final$rgn_id)) # 79 final regions

colSums(is.na(tour_jobs_final))
nrow(tour_jobs_final) # 1264

write_csv(tour_jobs_final, here(current_le_dir, "int", "liv_tour_quantity.csv"))
```


Create the proportion between tourism jobs and the total labor force

```{r}
# plot it, interactively!
# line_plot <- plotly::plot_ly(tour_jobs_prop, x = ~year, y = ~tour_jobs_prop, color = ~rgn_name, type = "scatter", mode = "lines") %>% 
#   layout(title = "All Regions: Proportional Tourism Employment Within Total Labor Force", 
#          xaxis = list(title = "Year"),
#          yaxis = list(title = "Percent of people by region employed by the tourism sector"))
# 
# line_plot

# htmlwidgets::saveWidget(line_plot, file = "prop_tourism_laborforce.html")
```



# Quality (wages)

  - note: this data came from many sources and was compiled by ILO

```{r}
# ILO wage data
ilo_wage_data <- read_csv(here(raw_data_dir, "ILOSTAT", current_data_dir, "EAR_4MTH_SEX_ECO_CUR_NB_A-filtered-2025-08-14.csv"))


# preliminary cleanings
wage_data_clean <- ilo_wage_data %>% 
  janitor::clean_names() %>% 
  rename(monthly_wage = obs_value) %>% 
  # filter to PPP adjusted data (this accounts for inflation, diff in currencies globally, cost of living, etc.) 
  # NEED TO CHECK WHAT THE YEAR IS ON YOUR DATASET
  filter(classif2_label == "Currency: 2021 PPP $") %>%
  # group by country/region
  group_by(ref_area_label) %>%
  # set column name to year
  rename(year = time) %>%
  # filter to the cap date range from the jobs data
  filter(year %in% c(2008:data_year))  %>%
  # add iso3 column, tourism sector label
  mutate(iso3 = country_regex_to_iso3c(ref_area_label), 
          sector = "tour") %>%
  select(ref_area_label, iso3, year, monthly_wage, 
         classif2_label, classif1_label)

# v2025: we updated the categories to include arts, entertainment, and recreation
wage_pivot <- wage_data_clean %>% 
  pivot_wider(names_from = classif1_label,
              values_from = monthly_wage) %>%
  rename(accomodation_food = `Economic activity (ISIC-Rev.4): I. Accommodation and food service activities`,
         arts_ent_rec = `Economic activity (ISIC-Rev.4): R. Arts, entertainment and recreation`) %>% 
          # need to change NAs to 0 for adding
  mutate(arts_ent_rec = ifelse(is.na(arts_ent_rec), 0, arts_ent_rec),
         accomodation_food = ifelse(is.na(accomodation_food), 0, accomodation_food),
         # Sum the two sectors 
         total_monthly_wage = accomodation_food + arts_ent_rec,
         # turn back to NAs for ease of gapfilling
         total_monthly_wage = ifelse(total_monthly_wage == 0, NA_real_,
                                     total_monthly_wage))


colSums(is.na(wage_pivot)) 
# No NAs in the total, which is good!

# make sequence of years for gapfilling
# note: data does not have wage data for some years in some countries -- we want to account for that by leaving them as NAs, then gapfilling with the average between the pre and post years

years_df <- tibble(iso3 = wage_pivot$iso3) %>% 
  group_by(iso3) %>% 
  reframe(year = seq(2008, data_year))

wage_data_years <- left_join(years_df, wage_pivot, by = c("iso3", "year"))

# filling in country names when possible --- 
wage_years_filled <- wage_data_years %>% 
  group_by(iso3) %>% 
  fill(ref_area_label, .direction = "downup") %>% 
  ungroup()

# gap filling
wage_gf <- wage_years_filled %>% 
  group_by(ref_area_label) %>% 
  # interpolate (fill missing values between 2 values)
  mutate(appx_wage = zoo::na.approx(total_monthly_wage, # using values in this column
                                    na.rm = FALSE, # don't replace (internal) NAs in new column that can't be approximated
                                    #  extrapolate using rule = 2 from approx(),
                                    # uses closest data extreme to extrapolate 
                                    # for leading and trailing NAs
                                    rule = 2)) %>% 
  
  # fill NA values with the only other observed value (making sure to note the list of countries/regions we did this for)
  mutate(appx_wage_fill = zoo::na.aggregate(appx_wage, by = 1, FUN = mean, na.rm = FALSE))

colSums(is.na(wage_gf))

```

notes: 
- made this a new column so that we can still identify each country that we are using this fill (na.aggregate function) method for.
- need more than 1 data point to do any approximation. 
- also, extremes (leading and trailing) are just copied data points from nearest extreme

```{r}
# find regions with only 1 data point (still have NAs in appx_wage column)
na_regions <- wage_gf %>% 
  filter(is.na(appx_wage))

unique(na_regions$ref_area_label)
num_na <- length(unique(na_regions$ref_area_label))
num_tot <- length(unique(wage_gf$ref_area_label))

paste0("proportion of countries/regions with only 1 data point: ", round(((num_na / num_tot) * 100), 3), "%")
```

```{r}
# preliminary plot 
# line_plot <- plotly::plot_ly(wage_gf, x = ~year, y = ~appx_wage_fill, color = ~ref_area_label, type = "scatter", mode = "lines") %>% 
#   layout(title = "All Regions: monthly Service Wages (USD (2017) PPP-Adjusted)", 
#          xaxis = list(title = "Year"),
#          yaxis = list(title = "monthly Wages (USD)"))
# 
# line_plot

# htmlwidgets::saveWidget(line_plot, file = "prop_tourism_laborforce.html")

```


Clean up ILO data, join with OHI regions

```{r}
# join with OHI regions
wage_region_join <- left_join(rgns_eez, wage_gf, by = c("eez_iso3" = "iso3"))

# Currency: 2021 PPP $
wage_regions <- wage_region_join %>% 
  mutate(unit = "Currency: 2021 PPP $") %>% 
  select(-classif2_label) %>% 
  mutate(sector = "tour",
         data_source = "ILO")

```


Plot

```{r}
# interactive plot after joining with OHI Regions
# plotly::plot_ly(wage_regions, x = ~year, y = ~appx_wage_fill, color = ~admin_country_name, type = "scatter", mode = "lines") %>% 
#   layout(title = "All Regions: Monthly Service Wages (USD)", 
#          xaxis = list(title = "Year"),
#          yaxis = list(title = "Monthly Wages (USD)"))

```

# Join quantity and quality data

```{r}
# Narrow wage columns before joining
wage_small <- wage_regions %>% 
  select(rgn_id, rgn_name, year, total_monthly_wage, appx_wage, 
         appx_wage_fill, wage_unit = unit, sector)

# Join quantity and quality data
tour_job_wage <- tour_jobs_final %>% 
  left_join(wage_small, by = join_by(rgn_id, rgn_name, year))
         
summary(tour_job_wage)

nrow(tour_job_wage) # 1264
colSums(is.na(tour_job_wage))
```


```{r}
# Gapfill with regional means for wage data
georegions <- ohicore::georegions %>% 
  left_join(rgns_eez, by = "rgn_id") %>% 
   # not needed
  select(-c(Notes, territory, admin_rgn_id, admin_country_name, r0, eez_iso3))

# Find avg wage by georegion
wage_georgn <- tour_job_wage %>% 
  group_by(rgn_id, rgn_name) %>% 
  complete(year = 2008:data_year) %>% # ensures that there is a row for every year for each region, to ensure there are no gaps
  left_join(georegions, by = c("rgn_id", "rgn_name")) %>%
  ungroup() %>%
  group_by(r2) %>%
  mutate(mean_wage_r2 = mean(appx_wage_fill, na.rm = TRUE),
         # to make NaN = NAs
         mean_wage_r2 = ifelse(is.nan(mean_wage_r2), NA, mean_wage_r2)) %>%
  ungroup() %>%
  group_by(r1) %>%
  mutate(mean_wage_r1 = mean(appx_wage_fill, na.rm = TRUE),
         mean_wage_r1 = ifelse(is.nan(mean_wage_r1), NA, mean_wage_r1)) %>%
  ungroup()

# Gapfill with R2 then R1 regions
wage_georgn_gf <- wage_georgn %>% 
  mutate(
    # Gapfilling record
    georgn_gapfill = case_when(
    is.na(appx_wage_fill) & !is.na(mean_wage_r2) ~ "r2",
    is.na(appx_wage_fill) & is.na(mean_wage_r2) & !is.na(mean_wage_r1) ~ "r1",
    .default = NA
  ),
  # Actual gapfilling
    appx_wage_fill = case_when(
    is.na(appx_wage_fill) & !is.na(mean_wage_r2) ~ mean_wage_r2,
    is.na(appx_wage_fill) & is.na(mean_wage_r2) & !is.na(mean_wage_r1) ~ mean_wage_r1,
    TRUE ~ appx_wage_fill)
  )

colSums(is.na(wage_georgn_gf))
# appx wage fill now has no NAs!

```


# Livelihoods Calculation

```{r}
# Quantity times quality
liv_tourism <- wage_georgn_gf %>% 
  mutate(livelihood = tour_jobs_prop * appx_wage_fill) 

liv_tourism_max <- liv_tourism %>%
  # look at previous ten years
  filter(year >= (max(year) - 10)) %>% 
  group_by(rgn_id, rgn_name) %>% 
  mutate(ref_point = max(livelihood)) %>% 
  ungroup() %>% 
  mutate(liv_norm = livelihood / ref_point,
         liv_norm = ifelse(liv_norm > 1, 1, liv_norm)) %>% 
  select(rgn_id, rgn_name, year, liv_norm, livelihood, ref_point,
         tour_jobs_prop, appx_wage_fill)

```

## Coastal Proportion

Now, lets bring in our tourism data and calculate the coastal proportion! The coastal proportion will then be multiplied by the arrivals value per region and year. Finally, the coastal area will be divided, giving us a density of tourism per region (individuals/km\^2).

```{r}
# ------- coastal population data -----------
coastal_pop_data <- read_csv(here("globalprep","mar_prs_population","v2025","output","mar_pop_25mi.csv")) ## read in coastal population data from other data layer

# Checking out the coastal population data, the only year with missing population is 2023. We'll gapfill 2023 with the 2022 value

coastal_pop_data_fill <- coastal_pop_data %>%
  filter(year == 2022) %>%
  mutate(year = 2023) %>%
  rbind(coastal_pop_data) # add 2023 data year, just using year 2022

length(unique(coastal_pop_data_fill$rgn_id)) # v2024 and v2025: 220 regions available
```

```{r}
# bring in World Bank Global population data 
total_population_wb <- read_csv(here(raw_data_dir, "WorldBank", current_data_dir,
                                     # UPDATE FILE NAME TO MATCH
                                     "WorldBank_global_country_population_1960-2024",
                                     "API_SP.POP.TOTL_DS2_en_csv_v2_38144.csv"), 
                                skip = 4) %>% # check how many rows to skip in excel
  pivot_longer(cols = 5:(ncol(.) - 1), names_to = "year",
               values_to = "total_pop") %>% 
  janitor::clean_names()

# give ohi rgns to world bank data
total_population_rgns <- name_2_rgn(df_in = total_population_wb, 
                                        fld_name = 'country_name',
                                        keep_fld_name = TRUE)

# this shows rows that have at least one duplicate, including both the first and subsequent occurrences of each duplicated row. using the tidyverse allows us to see all instances of duplicated data, not just the second and subsequent occurrences (which is what just a duplicated() call would give you).
duplicates_wb <- total_population_rgns %>%
  group_by(rgn_id, rgn_name, year) %>% 
  filter(n() > 1) %>%
  ungroup()
unique(duplicates_wb$rgn_name)
# [1] "China"                                              
# [2] "Northern Mariana Islands and Guam"                  

# drop Macao and Hong Kong, since we only need mainland China
# Northern Mariana Islands and Guam, Puerto Rico and Virgin Islands of the United States: aggregate!
total_population_final <- total_population_rgns %>%
  filter(country_name != "Hong Kong SAR, China" & country_name != "Macao SAR, China") %>% # only keep mainland, not city specific data
  group_by(rgn_id, rgn_name, year) %>% 
  summarise(
    total_pop = sum(total_pop, na.rm = TRUE), # aggregate by rgn and year
    .groups = "drop" # ungroup
  ) %>% 
  mutate(year = as.numeric(year)) 

length(unique(total_population_final$rgn_name)) # v2024: 166 regions available of total population data
```


## Coastal Area

Now, lets bring in the $$km^2$$ of coastal area for all OHI regions by aggregating the area protected 1 kilometer inland, and 3 nautical miles out into the eez.

```{r}
# read in area of coastline data
inland_data <- read_csv(here("globalprep", "lsp", version_year, "output", "rgn_area_inland1km.csv"))
offshore_data <- read_csv(here("globalprep", "lsp", version_year, "output", "rgn_area_offshore3nm.csv"))

# get combined value of inland and offshore for each ohi region
inland_offshore <- inland_data %>%
  left_join(offshore_data, by = join_by(rgn_id)) %>%
  select(rgn_id, area.x, area.y) %>%
  group_by(rgn_id) %>%
  mutate(total_inland_offshore_area = area.x + area.y) %>%
  select(rgn_id, total_inland_offshore_area)
```

# Tourism Livelihoods Proportion

Now we can join all of these data sources and conduct $$\frac{TourismLivelihoods * (CoastalPop/TotalPop)}{CoastalArea}$$.

```{r}
# join the tourism data, coastal population, and total population data 

liv_prop_df <- left_join(liv_tourism_max, total_population_final, by = c("rgn_id","rgn_name","year")) %>% 
  mutate(year = as.numeric(year)) %>% 
  left_join(., coastal_pop_data_fill, by = c("rgn_id","year")) %>% 
  left_join(., inland_offshore, by = c("rgn_id")) %>% 
  filter(year %in% 2008:data_year) %>% # to keep consistent with the coastal pop data
  mutate(coastal_prop = (popsum/total_pop)) %>% # values that are greater than 1 are because of inconsistencies between World Bank and the coastal population data.  However, those with a value greater than 1 are mostly islands or small regions with a large coastal area.  In that case, we could say 100% of the tourism could be coastally related.
  mutate(coastal_prop = ifelse(coastal_prop > 1, 1, coastal_prop)) # for rgns and years in which the total_population is greater than 1, we cap it at 1 

colnames(liv_prop_df)

liv_final <- liv_prop_df %>% 
  mutate(liv_prop = (livelihood * coastal_prop)/(total_inland_offshore_area))  %>% 
  drop_na(liv_prop) %>% # regions that did not have tourism arrival values but were present in the coastal population data
  select(rgn_id, rgn_name, year, liv_prop) # this final `Ap` (arrivals proportion) is in the units of (tourists/km2).  Aka, a higher proportion indicates that there is a larger number of tourists per km2 of coastal area, and there is therefore a higher tourism density.  This number is now ready to be divided by `Sr`, or the sustainability of tourism.

options(scipen = 999)
view(liv_final)
hist(liv_final$liv_prop)

# save the Ap proportions dataframe
# write_csv(ap_final, here::here(current_le_dir, "int", "ap.csv"))

# # save which observations were gapfilled and how into output -- from v2023 version
tr_arrivals_props_tourism_gf <- arrivals_prop_df %>%
  select(rgn_id, rgn_name, year, data_source, arrivals_gf)

# write_csv(tr_arrivals_props_tourism_gf, here::here(current_tr_dir, "output", "tr_arrivals_props_tourism_gf.csv"))
```

```{r}
# interactive plot of the arrival proportions before rescaling
ap_plot <- ap_final %>%
  plot_ly(x = ~year, y = ~Ap, color = ~rgn_name, 
          type = "scatter", mode = "lines") %>%
  layout(title = "Global Arrival proportions (Ap) by Region",
         xaxis = list(title = "Year"),
         yaxis = list(title = "Arrival proportion (Ap)"))
ap_plot

htmlwidgets::saveWidget(ap_plot, here::here(current_tr_dir, "figs", "ap_interactive_plot.html")) # save plot
```

## Rescale Ap proportion using the 90th quantile

As according to the [OHI methods](https://ohi-science.org/ohi-methods/goals/goal-models-data.html#tourism-and-recreation), the Ap values for each year and region is rescaled using the 90th quartile for each region.

$$\frac{T_r}{T_{90th}}$$

```{r}
# to make the Ap scaled between 0 and 1, we can divide the current Ap by the 90th quartile values by region.
liv_rescaled <- liv_final %>% 
  group_by(rgn_id, rgn_name) %>% 
  summarize(value_90th_percentile = quantile(liv_prop, 0.9, na.rm = TRUE))

liv_rescaled_final <- left_join(liv_final, liv_rescaled, by = c("rgn_id", "rgn_name")) %>% 
  mutate(
    liv_rescaled = liv_prop / value_90th_percentile,
    liv_rescaled = if_else(liv_rescaled < 0, 0, liv_rescaled), # if less than 0, that is due to gapfilling where linear models drew the tourism arrival values less than 0.  
    liv_rescaled = if_else(liv_rescaled > 1, 1, liv_rescaled) # if more than 1, then the Ap value for that year was above the 90th percentile.  So the value is adjusted to 1
         ) %>% 
  select(rgn_id, rgn_name, year, liv_rescaled) 

# save the final tourism arrivals csv
# write_csv(tr_arrivals_props_tourism, here::here(current_tr_dir, "output", "tr_arrivals_props_tourism.csv"))
view(liv_rescaled_final)

colSums(is.na(liv_rescaled_final))

hist(liv_rescaled_final$liv_rescaled)
colSums(is.na(liv_rescaled_final))
length(unique(liv_rescaled_final$rgn_id)) # 79 regions
```

```{r}
# interactive plot of the arrival proportions after rescaling
ap_final_plot <- liv_rescaled_final %>%
  plot_ly(x = ~year, y = ~liv_rescaled, color = ~rgn_name, 
          type = "scatter", mode = "lines") %>%
  layout(title = "Global Arrival proportions (Ap) by Region, Rescaled",
         xaxis = list(title = "Year"),
         yaxis = list(title = "Arrival proportion (Ap)"))
ap_final_plot

htmlwidgets::saveWidget(ap_final_plot, here::here(current_tr_dir, "figs", "ap_rescaled.html")) # save plot
```


# Data check

```{r}
year_check <- liv_rescaled_final %>% 
  filter(year == 2023)
view(year_check)

ggplot(year_check, aes(x = liv_rescaled)) +
  geom_boxplot()
```

