---
title: "OHI `r format(Sys.Date(), '%Y')` - Livelilihoods Data Preparation"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

# Livelihoods: Proportion of Tourism Jobs per Country/Region per Year

Livelihoods includes two equally important sub-components, the **number of jobs**, which is a proxy for livelihood quantity, and the **per capita average annual wages**, which is a proxy for job quality. 

## Overview

* **Data:** Labor Force & Employment Data

  * Labor Force data from World Bank, last updated 7/1/25 (downloaded August 7, 2025).

      - <https://data.worldbank.org/indicator/SL.TLF.TOTL.IN>
      - Located in `globalprep/_raw_data/

  * Jobs data from UN Tourism / UNWTO (downloaded August 7, 2025)
  
      - Key Tourism Statistics <https://www.unwto.org/tourism-statistics/key-tourism-statistics>
      - According to the website, the latest update of the dataset took place in 31 January 2024. 
      
  * Quality/Wage data from ILOSTAT (downloaded August 7, 2025) -- for tour/cf wage data
  
      - <https://rshiny.ilo.org/dataexplorer46/?lang=en&id=EAR_4MTH_SEX_ECO_CUR_NB_A> 

## Methods

### Setup

```{r}
knitr::opts_chunk$set(eval = FALSE, warning = FALSE, message = FALSE)

# ---- Load packages ----
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}

librarian::shelf(
  ohicore,
  here,
  janitor,
  terra,
  countrycode,
  readxl,
  RColorBrewer,
  foreach,
  tidyverse, 
  httr,
  plotly,
  zoo # for gapfilling
)

# ---- Source common.R ----
source(here("workflow", "R", "common.R"))

region_data() # for rgns_all and rgns_eez

# ---- set year and file path info ----
current_year <- 2025 # Update this in the future!!
previous_year <- 2024

version_year <- paste0("v",current_year)
previous_vyear <- paste0("v", previous_year)

current_data_dir <- paste0("d", current_year)
previous_data_dir <- paste0("d", previous_year)

# ---- data directories ----

# Raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")

# WorldBank raw data directory
current_wb_dir <- here(raw_data_dir, "WorldBank", current_data_dir)

# ILOSTAT raw data directory
current_ilostat_dir <- here(raw_data_dir, "ILOSTAT", current_data_dir)

# UNWTO raw data directory
current_unwto_dir <- here(raw_data_dir, "UNWTO", current_data_dir)

# Livelihoods and Economies directory
current_le_dir <- here("globalprep", "le", version_year)
previous_le_dir <- here("globalprep", "le", previous_vyear)

# Create country code --> iso3 function
country_regex_to_iso3c <- function(country_string) {
  country_string %>%
    countrycode::countrycode(origin = "country.name", destination = "iso3c", origin_regex = TRUE)
}
```

### Read in Data

```{r}
# ===================== Read in Data ==============================


# Labor force data ----
# created in liv_labor_force_dataprep.Rmd
labor_force <- read_csv(here(current_le_dir, "int", "liv_labor_force.csv"))

# Jobs data ----
# total number of people employed by the tourism sector, in thousands
un_jobs_data_raw <- read_xlsx(here(current_unwto_dir,
                                   "UN_Tourism_8_9_2_employed_persons_04_2025",
                                   "UN_Tourism_8_9_2_employed_persons_04_2025.xlsx"),
                              # skip = 2, # skip first 2 rows
                              sheet = 2, # the sheet name has changed over time, confirm this
                              col_names = TRUE, # use first row as column names
                              na = c("","..")) # define NA markers



# ILO wage data ----
ilo_wage_data <- read_csv(here(current_ilostat_dir, "EAR_4MTH_SEX_ECO_CUR_NB_A-filtered-2025-08-07.csv"))

# OHI regions data ----
region_names <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/eez/spatial/regions_list.csv") 
```


## Labor force gapfilling

```{r}
summary(labor_force)
# 2009 - 2024
summary(un_jobs_data_raw)
# 2008 - 2023
summary(ilo_wage_data)
# 2015 - 2025

colSums(is.na(labor_force))

filter(labor_force, is.na(labor_force_total)) %>% view()

# Gapfill total labor force values for missing years
labor_force <- labor_force %>% 
  group_by(rgn_id, rgn_name) %>% 
  # Ensure we're hitting all years for joining later on
  complete(year = 2008:max(un_jobs_data_raw$TimePeriod)) %>% 
  filter(year < 2024) %>% 
  ungroup() %>% 
  # First, create gapfilled boolean and method columns
  mutate(lf_gapfilled = ifelse(is.na(labor_force_total), 1, NA),
         lf_gf_method = ifelse(is.na(labor_force_total), 1, NA)) %>%
   # Upfilling: fill in any values that are empty from early years with values from the nearest year. 
  tidyr::fill(labor_force_total, .direction = "up") %>% 
  mutate(lf_gapfilled = ifelse(lf_gapfilled == 1 & !is.na(labor_force_total), 
                            "gapfilled", lf_gapfilled),
         lf_gf_method = ifelse(lf_gf_method == 1 & !is.na(labor_force_total), 
                            "Upfilling", lf_gf_method)) %>% 
  # Downfilling
  tidyr::fill(labor_force_total, .direction = "down") %>% 
  mutate(lf_gapfilled = ifelse(lf_gapfilled == 1 & !is.na(labor_force_total), 
                            "gapfilled", lf_gapfilled),
         lf_gf_method = ifelse(lf_gf_method == 1 & !is.na(labor_force_total), 
                            "Downfilling", lf_gf_method)) %>% 
  # Fix missing static metadata from the expand 
  mutate(unit = "Persons (1)",
         sector = "liv",
         data_source = "World Bank")

view(labor_force)
table(labor_force$lf_gf_method) # 16 downfilling, 513 upfilling
colSums(is.na(labor_force))
```

v2025: Let's aim for 2008-2023

# Employment (number of jobs)

(put this within previous chunk, read in data etc. for number of tourism jobs, join with labor_regions, make new proportion of tourism jobs column, group by year, do prelim plotting etc. and evaluate data quality etc.)

### UNWTO Data

```{r}
# ==================== Tidy Data ===================================

# Select columns of interest and add a unit column
un_jobs_data <- un_jobs_data_raw %>% 
  select(country = GeoAreaName, year = TimePeriod, tourism_jobs = Value) %>% 
  mutate(unit = "Persons (1)",
         tourism_jobs = as.numeric(tourism_jobs)) 

# run ohicore name_2_rgn --------
tour_jobs_rgn <- name_2_rgn(df_in = un_jobs_data, 
                       fld_name = 'country',
                       flds_unique = c("year"))

# identify duplicates
tour_duplicates <- tour_jobs_rgn[duplicated(tour_jobs_rgn[, c("rgn_name", "year")]),]
unique(tour_duplicates$country)
# [1] "China, Hong Kong Special Administrative Region"
# [2] "China, Macao Special Administrative Region"
unique(tour_duplicates$rgn_name)
# [1] "China"

# aggregate: sum by rgn_name etc. ----
tour_jobs_rgn_agg <- tour_jobs_rgn %>%
  group_by(rgn_id, rgn_name, year) %>%
  dplyr::summarize(tourism_jobs = sum(tourism_jobs, na.rm = TRUE) 
            )%>%
  ungroup() %>% 
  filter(year >= 2008) %>% 
  # turn 0s into NAs (appropriate flag for lack of data)
  mutate(tourism_jobs = case_when(
    tourism_jobs == 0 ~ NA,
    .default = tourism_jobs
  ))

# Confirm duplicates are gone
tour_duplicates2 <- tour_jobs_rgn_agg[duplicated(tour_jobs_rgn_agg[, c("rgn_name", "year")]),]
unique(tour_duplicates2$rgn_name)

# Gapfill total labor force values for missing years
tourism_jobs_gf <- tour_jobs_rgn_agg %>% 
  group_by(rgn_id, rgn_name) %>% 
  # Ensure we're hitting all years for joining later on
  complete(year = 2008:max(un_jobs_data_raw$TimePeriod)) %>% 
  ungroup() %>%
  group_by(rgn_id, rgn_name) %>% 
  arrange(rgn_id, year) %>% 
  # First, create gapfilled boolean and method columns
  mutate(tr_gapfilled = ifelse(is.na(tourism_jobs), "1", NA_character_),
         tr_gf_method = ifelse(is.na(tourism_jobs), "1", NA_character_)) %>%
  arrange(rgn_id, year) %>% 
  # Linear interpolation
  mutate(tourism_jobs = na.approx(tourism_jobs, x = year, na.rm = FALSE)) %>% 
  mutate(tr_gapfilled = ifelse(tr_gapfilled == "1" & !is.na(tourism_jobs), 
                               "gapfilled", tr_gapfilled),
         tr_gf_method = ifelse(tr_gf_method == "1" & !is.na(tourism_jobs), 
                               "Interpolation", tr_gf_method)) %>% 
  # Upfilling
  tidyr::fill(tourism_jobs, .direction = "up") %>% 
  mutate(tr_gapfilled = ifelse(tr_gapfilled == "1" & !is.na(tourism_jobs), 
                               "gapfilled", tr_gapfilled),
         tr_gf_method = ifelse(tr_gf_method == "1" & !is.na(tourism_jobs), 
                               "Upfilling", tr_gf_method)) %>% 
  # Downfilling
  tidyr::fill(tourism_jobs, .direction = "down") %>% 
  mutate(tr_gapfilled = ifelse(tr_gapfilled == "1" & !is.na(tourism_jobs), 
                               "gapfilled", tr_gapfilled),
         tr_gf_method = ifelse(tr_gf_method == "1" & !is.na(tourism_jobs), 
                               "Downfilling", tr_gf_method))

view(tourism_jobs_gf)
table(tourism_jobs_gf$tr_gf_method)
# Downfilling Interpolation     Upfilling 
#          94            58           204 

colSums(is.na(tourism_jobs_gf))
nrow(tourism_jobs_gf)

# No U.S.
colnames(rgns_eez)
unique(tourism_jobs_gf$rgn_name) %>% view()
```

### OECD Data

```{r}
# OECD data
oecd_jobs_data_raw <- read_csv(here(raw_data_dir, "OECD", 
                                   current_data_dir,
                                   "OECD_2009_2022_tourism_employment.csv"))

# Clean raw data
oecd_clean <- oecd_jobs_data_raw %>% 
  janitor::clean_names() %>% 
  # Filter to persons employed, not enterprises
  # AND total tourism activity, not sub categories
  filter(unit_measure == "PS" & activity == "_T") %>% 
  select(country_code = ref_area, country = reference_area, 
         unit = unit_of_measure, year = time_period, value = obs_value)

# run ohicore name_2_rgn --------
oecd_job_rgn <- name_2_rgn(df_in = oecd_clean, 
                       fld_name = 'country',
                       flds_unique = c("year"))

oecd_jobs_clean <- oecd_job_rgn %>% 
  select(rgn_id, rgn_name, year, value, unit) %>% 
  mutate(unit = "Persons (1)",
         source = "OECD")

view(oecd_jobs_clean)
unique(oecd_jobs_clean$rgn_name)

# Adds Netherlands, United States, and Morocco!
setdiff(oecd_jobs_clean$rgn_name, tourism_jobs_gf$rgn_name)

# Let's see how the magnitude changes between countries with data in both
compare_oecd_unwto <- tourism_jobs_gf %>% 
  full_join(oecd_jobs_clean, by = join_by(rgn_id, rgn_name, year))

shared_countries <- compare_oecd_unwto %>% 
  filter(!is.na(tourism_jobs) & !is.na(value))
view(shared_countries)

unique(shared_countries$rgn_name) # 10 countries shared

comp_plot <- ggplot(shared_countries) +
  geom_point(aes(x = year, y = tourism_jobs), color = "red", alpha = 0.7) +
  geom_point(aes(x = year, y = value), color = "blue", alpha = 0.7) +
  facet_wrap(~rgn_name, #) +
             scales = "free_y") +
  labs(title = "UNWTO vs. OECD Data Shared Countries: Free y-axis",
       y = "Jobs",
       x = "Year") +
  theme(panel.spacing.x = unit(0.01, "cm")) +
  theme_bw();ggplotly(comp_plot)
```

### Eurostat Data

```{r}
# Eurostat data uses a weird mix of tabs and commas as delimiters, 
# so we do a bit of pre-processing here
txt <- readLines(here(raw_data_dir, "Eurostat", 
                      current_data_dir, "estat_tour_lfs1r2.tsv"))

# Replace commas or tabs with a single tab
txt_clean <- gsub("[\t,]", "\t", txt)

# Read in with consistent delimiter
eurostat_raw <- read.delim(textConnection(txt_clean), sep = "\t", header = TRUE)

# Preliminary cleaning
eurostat_clean <- eurostat_raw %>% 
  janitor::clean_names() %>% 
  mutate(country = countrycode(geo_time_period, 
                               origin = "eurostat", "country.name")) %>% 
  # Filtering down to specific employment codes and times that consider total tourism employment. This metadata was not easy to find. 
  filter(worktime == "TOTAL", wstatus == "EMP", nace_r2 == "TOTAL") %>% 
  # Remove the aggregated EU countries
  filter(!(geo_time_period %in% c("EA20", "EU27_2020"))) %>% 
  relocate(country, .after = unit) %>% 
  select(-c(freq, geo_time_period)) %>% 
  # tidy up year column names
  rename_with(~ gsub("x", "", .)) %>% 
  pivot_longer(cols = -c(nace_r2, wstatus, worktime, unit, country),
              names_to = "year", values_to = "value") %>% 
  # Fix weirdness in values
  mutate(value = str_remove_all(value, "[bd]"),
         value = str_replace_all(value, ":", NA_character_),
         country = str_replace_all(country, "&", "and"),
         value = as.numeric(value),
         year = as.numeric(year))

# run ohicore name_2_rgn --------
eurostat_rgn <- name_2_rgn(df_in = eurostat_clean, 
                       fld_name = 'country',
                       flds_unique = c("year"))

# Fix units and add some metadata
eurostat_jobs <- eurostat_rgn %>%
  mutate(jobs = value * 1000,
         unit = "Persons (1)",
         source = "Eurostat") %>% 
  select(rgn_id, rgn_name, year, jobs, unit, source)

# check what countries we've added:
setdiff(eurostat_jobs$rgn_name, tourism_jobs_gf$rgn_name)
# [1] "Germany"     "Estonia"     "Ireland"    
# [4] "Iceland"     "Montenegro"  "Netherlands"
# [7] "Norway" 
# That's pretty good!

# Let's see how the magnitude changes between countries with data in both
compare_euro_unwto <- tourism_jobs_gf %>% 
  full_join(eurostat_jobs, by = join_by(rgn_id, rgn_name, year))

shared_countries <- compare_euro_unwto %>% 
  filter(!is.na(tourism_jobs) & !is.na(jobs))

unique(shared_countries$rgn_name) # 21 countries shared

comp_plot <- ggplot(shared_countries) +
  geom_point(aes(x = year, y = tourism_jobs), color = "red", alpha = 0.7) +
  geom_point(aes(x = year, y = jobs), color = "blue", alpha = 0.7) +
  facet_wrap(~rgn_name) +
             # scales = "free_y") +
  labs(title = "UNWTO vs. Eurostat Data Shared Countries: Free y-axis",
       y = "Jobs",
       x = "Year") +
  theme(panel.spacing.x = unit(0.01, "cm")) +
  theme_bw();ggplotly(comp_plot)
```


```{r}
# Join UNWTO and OECD
test <- left_join(oecd_jobs_clean, tourism_jobs_gf,
                                 by = join_by(rgn_id, rgn_name, year))
view(test)

# join with labor force data ------------
# COME BACK -- rejoin once labor force has been updated
tour_jobs_labor <- full_join(tourism_jobs_gf, labor_force , by = c("rgn_id", "rgn_name", "year"))  %>% 
  select(-c(sector, data_source)) # these currently only show Labor Force data-related values. Will add back on before exporting.

summary(tourism_jobs_gf)
summary(labor_force)

# calculate tourism jobs proportion of total labor force -----
tour_jobs_prop_df <- tour_jobs_labor %>% 
  mutate(tour_jobs_prop = (tourism_jobs / labor_force_total)) %>% 
  mutate(rgn_id = as.numeric(rgn_id)) 

colSums(is.na(tour_jobs_prop_df)) 
 # tourism_jobs      labor_force_total      tour_jobs_prop 
 #         1488                     0                 1488

nrow(tour_jobs_prop_df) # 2656
nrow(tour_jobs_rgn_agg) # 773
nrow(labor_force) # 2656
# ~ 71% NA
# After first round of gapfilling, dropped to 56% 

tour_jobs_prop_df %>% 
  group_by(rgn_id, rgn_name) %>% 
  filter(is.na(tourism_jobs)) %>% 
  view()

tour_jobs_na <- tour_jobs_prop_df %>%
  group_by(rgn_id, rgn_name) %>%
  filter(any(is.na(tourism_jobs))) %>%
  arrange(rgn_name, year)

length(unique(tour_jobs_na$rgn_name)) # 93

tour_jobs_non_na <- tour_jobs_prop_df %>% 
filter(any(!is.na(tourism_jobs)))

length(unique(tour_jobs_non_na$rgn_name)) # 73

length(unique(tour_jobs_prop_df$rgn_name)) # 73


# 73 countries with 

View(df_with_missing) 


rm(tour_duplicates, tour_duplicates2)
```

```{r}
georegions <- ohicore::georegions %>% 
  left_join(rgns_eez, by = "rgn_id") %>% 
  select(-c(Notes)) # not needed

georegions_label <- ohicore::georegion_labels %>% 
  left_join(rgns_eez, by = "rgn_id") %>% 
  select(-c(Notes)) # not needed
table(georegions_label$r1_label)

colnames(georegions)

test <- tour_jobs_prop_df %>% 
  left_join(georegions_label, by = join_by(rgn_id, rgn_name))
view(test)

test %>% 
  distinct(rgn_id, rgn_name) %>% view()

table(test$r1_label)

```


I think this is old code...
```{r}
# join w OHI regions
# tourism_un_jobs_ohi <- left_join(region_clean, un_jobs_cc, by = c("eez_iso3" = "iso3c"))
# 
# 
# # ----------Join with Labor Force Data--------------
# tourism_labor_join <- left_join(labor_regions, tourism_un_jobs_ohi, by = c("rgn_id", "year")) 
```

Create the proportion between tourism jobs and the total labor force
```{r}
# tourism_job_proportion <- tourism_labor_join %>%
#   dplyr::relocate(tourism_jobs, .before = labor_force_total) %>%
#   mutate(tourism_proportion = tourism_jobs/labor_force_total) %>%
#   rename(labor_force_total = labor_force_total) %>%
#   select(rgn_id, rgn_name.x, country, year, labor_force_total, tourism_jobs, tourism_proportion)
colnames(tour_jobs_prop_df)
 max(tour_jobs_prop_df$year)

# plot it, interactively!
line_plot <- plotly::plot_ly(tour_jobs_prop_df, x = ~year, y = ~labor_force_total, color = ~rgn_name, type = "scatter", mode = "lines") %>% 
  layout(title = "All Regions: Proportional Tourism Employment Within Total Labor Force", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Percent of people by region employed by the tourism sector"))

line_plot

# htmlwidgets::saveWidget(line_plot, file = "prop_tourism_laborforce.html")
```

### Saving intermediate data

```{r}
old_tour_jobs <- read_csv(here(previous_le_dir, "int", "tour_jobs.csv"))
colSums(is.na(old_tour_jobs)) # 1856
nrow(old_tour_jobs) # 2671
# ~ 70% NA

write_csv(tourism_job_proportion, here(int_dir, "tour_jobs.csv"))
```


# Quality (ILO wage data)
  - note: this data came from many sources and was compiled by ILO

```{r}
# preliminary cleanings
wage_data_clean <- ilo_wage_data %>% 
  # lower_snake_case
  janitor::clean_names() %>% 
  # more intuitive value name
  rename(monthly_wage = obs_value) %>% 
  # filter to PPP adjusted data (this accounts for inflation, diff in currencies globally, cost of living etc.) NEED TO CHECK WHAT THE YEAR IS ON YOUR DATASET
  # Why are we using 2017 here?
  filter(classif2_label == "Currency: 2021 PPP $") %>%
  # group by country/region
  group_by(ref_area_label) %>%
  # set column name to year
  rename(year = time) %>%
  # filter to the cap date range from the jobs data --
  filter(year %in% c(2008:max(tour_jobs_prop_df$year))) # %>%
  # add iso3 column, tourism sector label
  # mutate(iso3 = country_regex_to_iso3c(ref_area_label),
         # sector = "tour") %>%
  # select(c(ref_area_label, iso3, year, monthly_wage, classif2_label))

#   CHECK KOSOVO FOR COUNTRY REGEX TO ISO3C
view(wage_data_clean)
table(ilo_wage_data$classif2.label)
colnames(ilo_wage_data)

# make sequence of years for gapfilling ----
# note: data does not have wage data for some years in some countries -- we want to account for that by leaving them as NAs, then gapfilling with the average between the pre and post years

years_df <- tibble(iso3 = wage_data_clean$iso3) %>% 
  group_by(iso3) %>% 
  summarize(year = seq(2008, 2023))


wage_data_years <- left_join(years_df, wage_data_clean, by = c("iso3", "year"))

# filling in country names when possible --- 
wage_years_filled <- wage_data_years %>% 
  group_by(iso3) %>% 
  fill(ref_area_label, .direction = "downup") %>% 
  ungroup()

# =================

# test gapfilling
# gap_fill_test <- wage_years_filled %>% 
#   mutate(ref_area_label = as.factor(ref_area_label)) %>% 
#   mutate(lm_est = list(lm(monthly_wage ~ year + ref_area_label)))
# 
# 
# lm_test <- lm(monthly_wage ~ year + ref_area_label, data = wage_years_filled)
# summary(lm_test)
# 
# lm_test$coefficients 

# wage_years_filled$lm_values <- lm_test$fitted.values

# gap filling  ----
wage_gf <- wage_years_filled %>% 
  group_by(ref_area_label) %>% 
  # interpolate (fill missing values between 2 values)
  mutate(appx_wage = zoo::na.approx(monthly_wage, # using values in this column
                                    na.rm = FALSE, # don't replace (internal) NAs in new column that can't be approximated
                                    #  extrapolate using rule = 2 from approx(),
                                    # uses closest data extreme to extrapolate 
                                    # for leading and trailing NAs
                                    rule = 2)) %>% 
  
  # fill NA values with the only other observed value (making sure to note the list of countries/regions we did this for)
  mutate(appx_wage_fill = zoo::na.aggregate(appx_wage, by = 1, FUN = mean, na.rm = FALSE))
# note: made this a new column so that we can still identify each country that we are using this fill (na.aggregate function) method for.
  
# other notes: 
# need more than 1 data point to do any approximation. 
# also, extremes (leading and trailing) are just copied data points from nearest extreme

```



```{r}
# find regions with only 1 data point (still have NAs in appx_wage column)

na_regions <- wage_gf %>% 
  filter(is.na(appx_wage))


unique(na_regions$ref_area_label)
num_na <- length(unique(na_regions$ref_area_label))
num_tot <- length(unique(wage_gf$ref_area_label))

paste0("proportion of countries/regions with only 1 data point: ", round(((num_na / num_tot) * 100), 3), "%")
```




```{r}
# preliminary plot 
line_plot <- plotly::plot_ly(wage_gf, x = ~year, y = ~appx_wage_fill, color = ~ref_area_label, type = "scatter", mode = "lines") %>% 
  layout(title = "All Regions: monthly Service Wages (USD (2017) PPP-Adjusted)", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "monthly Wages (USD)"))

line_plot

# htmlwidgets::saveWidget(line_plot, file = "prop_tourism_laborforce.html")

```




In this plot we can see that Sao Tome and Principe is a clear high outlier with no variation -- it is one of the countries for which we gapfilled with copied values from a single observation. We will continue to join the data with the OHI regions, then plot again and reinvestigate.


Clean up ILO data, join with OHI regions

```{r}
# join with OHI regions
wage_region_join <- left_join(region_clean, wage_gf, by = c("eez_iso3" = "iso3"))

# come back ---- stopping point for Mel to take a look!

# Currency: 2017 PPP $

wage_regions <- wage_region_join %>% 
  mutate(unit = "Currency: 2017 PPP $") %>% 
  select(-classif2_label) %>% 
  mutate(sector = "tour",
         data_source = "ILO")

```


Plot

```{r}
# interactive plot after joining with OHI Regions
plotly::plot_ly(wage_regions, x = ~year, y = ~appx_wage_fill, color = ~admin_country_name, type = "scatter", mode = "lines") %>% 
  layout(title = "All Regions: Monthly Service Wages (USD)", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Monthly Wages (USD)"))

```

We can see that Sao Tome and Principe is still included and is still this high, unvarying outlier. 
This lack of data combined with how high of an outlier is led us to reconsider our methods for this country and consider dropping it from this intermediate data product. 

Let's drop & plot what that looks like. 

```{r}
wage_regions_no_sp <- wage_regions %>% 
  filter(!ref_area_label %in% c("Sao Tome and Principe"))

# plot to see what it looks like after dropping Sao Tome and Principe
plotly::plot_ly(wage_regions_no_sp, x = ~year, y = ~monthly_wage, color = ~admin_country_name, type = "scatter", mode = "lines") %>% 
  layout(title = "All Regions: Monthly Service Wages (USD)", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Monthly Wages (USD)"))

```

Dropping Sao Tome and Principe (STP) adjusted the y-axis scale significantly to better visualize the monthly service wage data (USD) for other geographic areas.  Investigating further, it looks as though the Sao Tome and Principe observation from 2017 (133814.9 USD) may have been inputted incorrectly, measured as annual income instead of monthly, or there was an error when it was adjusted by PPP.  Dustin noticed that the data source from this observation was "HIES - Household Budget Survey".  This was the only observation from this data source. Upon further investigation, we discovered the average annual income in STP is ~2,400 USD [WorldData]("https://www.worlddata.info/africa/sao-tome-and-principe/index.php#:~:text=With%20an%20average%20annual%20income,the%20lower%20middle%2Dincome%20countries") and a GDP per capita of ~2,817 USD [World Bank]("https://www.worldbank.org/en/country/saotome/overview"). Therefore, we have decided to drop it for the intermediate data set.


## Saving intermediate data products

```{r}
# write_csv(wage_regions_no_sp, here(int_dir, "tour_quality.csv"))

```

