---
title: 'OHI `r format(Sys.Date(), "%Y")` - World Governance Indicator'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

[REFERENCE RMD FILE](http://ohi-science.org/ohiprep_v2025/globalprep/prs_res_wgi/v2025/WGI_dataprep.html)

# Summary
This script downloads WGI data and prepares it for a pressures (1 - WGI) and resilience data layer.  

## Updates from previous assessment
- New data download
- Added `librarian` to handle package imports
- Added directory, scenario, and year variables to make the script more reproducible
- Updated the exploratory code and data import at the top to be less redundant


## Data Source 
**Reference**:  http://info.worldbank.org/governance/wgi/index.aspx#home    

**Downloaded**: 

July 29, 2025 (source data updated Sep 2023)

**Description**:  
The Worldwide Governance Indicators (WGI) project reports aggregate and individual governance indicators for 215 economies over the period 1996–2022, for six dimensions of governance:

* Voice and Accountability
* Political Stability and Absence of Violence
* Government Effectiveness
* Regulatory Quality
* Rule of Law
* Control of Corruption

**Time range**: 1996-2023


```{r}
knitr::opts_chunk$set(eval = FALSE)

# ======= Load packages ============
if (!require(librarian)) {
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  ohicore, #devtools::install_github('ohi-science/ohicore@dev') #if relevant: restart session after reinstalling
  WDI, # install.packages('WDI')  # used to extract World Development Indicator (World Bank) data 
  here,
  janitor,
  tools,
  tidyverse,
  plotly,
  strex
)

# check website to see what years are available: http://info.worldbank.org/governance/wgi/index.aspx#home


# ======= Set directories ===========
# Update scenario year, set up programmatic scenario year updates
current_year <- 2025 # update this!!!
previous_year <- current_year - 1

# Directories
data_dir <- paste0("d", current_year)
prev_data_dir <- paste0("d", previous_year)

# Scenarios
scen_year <- paste0("v", current_year)
prev_scen_year <- paste0("v", previous_year)

start_year <- 1996
end_year <- 2023 # Update depending on available data from Worldbank website

# WGI directory on GitHub
current_wgi_dir <- here::here("globalprep", "prs_res_wgi", scen_year) 
previous_wgi_dir <- here::here("globalprep", "prs_res_wgi", prev_scen_year) 

# ======== Source functions ===========
source(here::here("workflow", "R", "common.R")) # directory locations
# functions specific to mariculture dealing with compound countries, updated in 2024
source(here::here("globalprep", "mar", scen_year, "mar_fxs.R"))

# Avoid scientific notation for numbers with many sig figs
options(scipen = 999)

```

# Obtain the WGI data
Download each of the 6 WGI indicators. Note: This didn't work in 2023 for some reason. For the 2024 assessment, check to see if it will work again. Instead. I manually downlaoded the data from here: https://databank.worldbank.org/source/worldwide-governance-indicators#. Use the uncommented chunk below for this. 

```{r indicator-decsription}
# Import WDI data via WDI package
indicators <-  data.frame(WDI_data[[1]])

ind_names <- list("VA.EST", "PV.EST", "GE.EST", "RQ.EST", "RL.EST", "CC.EST")

# View indicator names and descriptions
for (ind_name in ind_names){
  
  name <- indicators[grep(ind_name, indicators$indicator), 2] 
  description <- indicators[grep(ind_name, indicators$indicator), 3] 
  cat("\nIndicator name:", name)
  cat("\nDescription:", description)
  cat("\n") # space between each indicator

}
```


```{r extract-data}
# Create list of indicator names and short names
ind_dict <- list(
  "voice_acct" = "Voice and Accountability: Estimate",
  "pol_stb" = "Political Stability and Absence of Violence/Terrorism: Estimate",
  "govt_eff" = "Government Effectiveness: Estimate",
  "reg_qual" = "Regulatory Quality: Estimate",
  "rule_law" = "Rule of Law: Estimate",
  "cont_corr" = "Control of Corruption: Estimate"
)

# Query the catalog for each of the 6 indicators
indicator_data <- list()

for (key in names(ind_dict)) {
  value <- ind_dict[[key]]
  
  indicator_code <- WDIsearch(value, field = 'name')[1, 'indicator']
  
  df <- WDI(indicator = indicator_code,country = 'all',
            start = start_year, end = end_year)
  
  # Rename the indicator column to the key
  colnames(df)[which(names(df) == indicator_code)] <- key
  
  # Keep relevant columns only
  df <- df %>%
    select(country, iso3c, year, all_of(key))
  
  indicator_data[[key]] <- df
  
}

# Join the individual dfs together
final_data <- Reduce(function(x, y) {
  left_join(x, y, by = c("country", "iso3c", "year"))
}, indicator_data)

# Check
# view(final_data)
colSums(is.na(final_data))
nrow(final_data)

```

Combine the indicators into a single table, with a column for each indicator, and rows for each country-year pair.

```{r access-data-without-api}
wgi_raw <- read_csv(here::here(current_wgi_dir, "raw", "P_Data_Extract_From_Worldwide_Governance_Indicators_2025.csv"))

d_web <- wgi_raw %>%
  clean_names() %>%
  rename(iso3c = country_code, country = country_name) %>%
  filter(str_detect(series_code, "EST")) %>%
  dplyr::select(-series_name) %>%
  pivot_longer(cols = starts_with("x"), values_to = "est", names_to = "year") %>%
  mutate(year = as.numeric(str_after_last(year, "yr"))) %>%
  mutate(est = ifelse(est == "..", NA, est)) %>%
  pivot_wider(values_from = "est", names_from = "series_code")

view(d_web)

write_csv(d_web, here(current_wgi_dir, "raw", paste0("worldbank_wgi_from_web_", current_year, ".csv")))
```


Combine the indicators into a single table, with a column for each indicator, and rows for each country-year pair.

## Save a record of any new raw data for archival purposes

Uncomment the code chunk lines when updating WGI data, this will most likely occur when calculating for new assessment year:

```{r}
# Update file name and pathing
write_csv(final_data, here(current_wgi_dir, "raw", paste0("worldbank_wgi_from_api_", current_year, ".csv")))
# This dataset currently has non-OHI regions included
```


# Gapfill, part 1: filling missing years of data for indicators, within countries
The first gapfilling occurs when we use the average of previous years data within each region/indicator.  This occurs when a region has data for an indicator, but not for all years.


**Read in WGI data - change appended date in file name to reflect the most recent version of the saved WGI data:**
```{r}
d_api <- read_csv(here(current_wgi_dir, "raw", paste0("worldbank_wgi_from_api_", current_year, ".csv")))
colnames(d_api)

d_web <- write_csv(d_web, here(current_wgi_dir, "raw", paste0("worldbank_wgi_from_web_", current_year, ".csv")))
colnames(d_web)



```

```{r}
# Comparing the API data and the web download data
colSums(is.na(d_web)) # 117-231 NAs across all 6 inds
nrow(d_web) # 530
# ~ 38% NA
view(d_web)

colSums(is.na(d_api)) # 2370 - 2482 NAs across all 6 inds
nrow(d_api) # 7448
# ~ 32% NA
view(d_api)

diff_countries <- setdiff(d_api$country, d_web$country)

diff_df <- filter(d_api, country %in% diff_countries) %>% 
  view()

length(unique(diff_df$country))
# Puerto Rico (US) is the only country different between the API and website data that actually has data for all 6 indicators. 

d_web <- d_web %>% 
  rename(voice_acct = VA.EST, pol_stb = PV.EST, govt_eff = GE.EST, 
         reg_qual = RQ.EST, rule_law = RL.EST, cont_corr = CC.EST)

list(
  "voice_acct" = "Voice and Accountability: Estimate",
  "pol_stb" = "Political Stability and Absence of Violence/Terrorism: Estimate",
  "govt_eff" = "Government Effectiveness: Estimate",
  "reg_qual" = "Regulatory Quality: Estimate",
  "rule_law" = "Rule of Law: Estimate",
  "cont_corr" = "Control of Corruption: Estimate"
)
ind_names <- list("VA.EST", "PV.EST", "GE.EST", "RQ.EST", "RL.EST", "CC.EST")

setdiff(d_api$voice_acct, d_web$voice_acct) # 0
setdiff(d_api$pol_stb, d_web$pol_stb) # 0 
setdiff(d_api$govt_eff, d_web$govt_eff) # 0 
setdiff(d_api$reg_qual, d_web$reg_qual) # 0
setdiff(d_api$rule_law, d_web$rule_law) # 0
setdiff(d_api$cont_corr, d_web$cont_corr) # 0

columns <- d_api[,4:9]

for (col in columns){
  # print(paste("Indicator:", col))
  print(setdiff(d_api$col, d_web$col))
}
diff_scores <- setdiff(d_api$country, d_web$country)

diff_df <- filter(d_api, country %in% diff_countries) %>% 
  view()

length(unique(diff_df$country))
```


```{r}
# CC.EST:VA.EST
# voice_acct:cont_corr
d_long <- d_web %>% 
  pivot_longer(cols = CC.EST:VA.EST, names_to = "indicator", values_to = "value") %>% 
  mutate(value = as.numeric(value))


## each country has 20 years of data
d_gap_fill  <- d_long %>%
  group_by(country, iso3c, year) %>%
  mutate(NA_count_c_y = sum(is.na(value))) %>% # gf record: NA values within a region/year prior to gapfilling, max value is 6 (meaning that a country has no data)
  ungroup() %>%
  group_by(country, iso3c, indicator) %>% # gapfill missing data with mean of values across years within the same region/indicator
  mutate(ind_mean_c_i = mean(value, na.rm=TRUE)) %>%
  ungroup() %>%
  mutate(value = ifelse(is.na(value), ind_mean_c_i, value)) %>%
  group_by(country, iso3c, year) %>% 
  mutate(NA_count_post_gf1 = sum(is.na(value))) # gf record: NA values within a region/year after within region/indicator gapfilling (i.e. indicator is gapfilled by other years of data), used to cut regions <4 indicators (below)    

```


## Safeguard: cut regions with < 4 indicators (if any) to calculate score.
Once gapfilling is complete, the WGI scores are calculated as an average of the 6 indicators.  However, if a country is missing 4 or more of the indicators within a year the average would be very biased.  In these cases, a different method should be used to gapfill these data 

(NOTE: for the 2024 assessment all regions had at least 4 of the 6 indicators).

```{r purge data poor region}

countries_no_data <- d_gap_fill %>%
  filter(NA_count_post_gf1 > 3)

countries_no_data <- unique(countries_no_data$country)
countries_no_data # API data: 61 countries... a lot compared to 2024
# Hmm, when I pull the data from the website and not the API, there are no countries with more than 3 NA columns

## In this case, the countries with minimal data (< 4 indicators ever calculated) are deleted.  
## These will be gap-filled later on if they are deleted now.
d_gap_fill <- d_gap_fill %>%
  filter(!(country %in% countries_no_data))

```

# Calculate overall WGI score for each country

This involves:

* taking the average of the 6 indicators (assuming there are at least 4 of the 6 indicators)
* rescaling the data from 0 to 1 

```{r}

d_calcs <- d_gap_fill %>%
  group_by(country, iso3c, year) %>%
  summarize(score_wgi_scale = mean(value, na.rm=T),
            NA_start = mean(NA_count_c_y), # initial mean number of NA across indicators, pre-gapfill 
            NA_post_gf_1 = mean(NA_count_post_gf1)) %>% # number of NA across indicators, post-gapfill across year gapfill within region/indicator
  ungroup() 

```

## Check that the values in `scores_wgi_scale` fall within the wgi range specified below:
```{r}

summary(d_calcs) # checking to make sure NA values make sense
  ## v2025: NA_start max = 6, NA_post_gf_1 max = 0

## v2025: score_wgi_scale min = -2.409, max = 1.947... perfect
wgi_range = c(-2.5, 2.5) # historic values have been between -2.5 and 2.5

```

```{r}

d_calcs <- d_calcs %>%
  mutate(score =  (score_wgi_scale - wgi_range[1]) / (wgi_range[2] - wgi_range[1])) %>%
  ungroup(); head(d_calcs); summary(d_calcs)
# converts the scores between 0 and 1, using min/max to scale the data

# document gapfilling
d_calcs <- d_calcs %>%
  mutate(gap_fill = NA_start - NA_post_gf_1,   # if there are values in NA_post_gf_1, it means these weren't gapfilled
         gap_fill = ifelse(is.na(score), 0, gap_fill)) %>% # number of values that were gapfilled
  select(-NA_start, -NA_post_gf_1)

```

## Explore & check intermediate d_calcs data table

```{r}

d_calcs[d_calcs$gap_fill>0, ] %>% View()
# No "Niue" in this data...
d_calcs[d_calcs$country == "Niue", ] %>% View() # should have gap-fill values between 0-6
d_calcs[d_calcs$country == "American Samoa",] %>% View() # should have gap-fill values between 0-6

```

## Save intermediate file 

```{r}

## save intermediate file of wgi scores pre-gapfilling (for OHI+ use)
# write.csv(d_calcs %>%
#             select(country, iso3c, year, score_wgi_scale, score_ohi_scale = score, gap_fill), 
#           here(paste0('globalprep/prs_res_wgi/v', scen_year, '/intermediate/wgi_combined_scores_by_country.csv')),
#           row.names = FALSE)
# Write out gapfill data
d_gf_final <- d_calcs %>% 
  select(country, iso3c, year, score_wgi_scale, score_ohi_scale = score, gap_fill)

write_csv(d_gf_final, here(current_wgi_dir, "int", "wgi_combined_scores_by_country.csv"))

```


# Convert country names to ohi regions
v2025: we alreayd have an `ohicore` function that splits the names of countries that get reported as a group. I've removed the manual version of this and instead added the `split_regions()` function

```{r}

# Special case: user-defined function deals with 
# Breaking up Antilles into separate reported rgns
d_test <- d_calcs %>% 
  rename(value = score)

d_split <- ohicore::split_regions(d_test, 
                                  country_column = 'country', 
                                  value_column = 'score')

d_rgn <- ohicore::name_2_rgn(df_in = d_split,
                    fld_name = 'country', 
                    flds_unique = c('year'))

## We report these regions at a greater spatial resolution:

## Aruba is part of the Netherlands Antilles, but it is reported separately
country_split_1 <- data.frame(country = "Netherlands Antilles", region = c('Bonaire', 'Curacao', 'Saba', 'Sint Maarten', 'Sint Eustatius'))
country_split_2 <- data.frame(country = "Jersey, Channel Islands", region = c('Jersey', 'Guernsey'))
country_split <- rbind(country_split_1, country_split_2)

country_split_data <- country_split %>%
  left_join(d_calcs) %>%
  select(-country) %>%
  rename(country = region)

d_calcs <- d_calcs %>%
  filter(!(country %in% c("Netherlands Antilles", "Jersey, Channel Islands"))) %>%
  rbind(country_split_data)  %>%
  mutate(country = as.character(country)) %>%
  dplyr::select(-iso3c)

d_calcs$country[grep("Korea, Dem.", d_calcs$country)] <- "North Korea"

d_calcs$country[grep("Turkiye", d_calcs$country)] <- "Turkey"
# Maybe in future update package with country synonym list


## Function to convert to OHI region ID
d_calcs_rgn <- name_2_rgn(df_in = d_calcs, 
                       fld_name='country', 
                       flds_unique=c('year'))


## Combine the duplicate regions (we report these at lower resolution)
## In this case, we take the weighted average
population_weights <- data.frame(country = c("Virgin Islands (U.S.)", "Puerto Rico",
                                             "China", "Hong Kong SAR, China", "Macao SAR, China"),
                                 population = c(104337, 3203295, 
                                                1408975000, 7524100, 687000))
# updated population values on 30 July 2025 (source: World Bank website, 2024 values); https://databank.worldbank.org/source/population-estimates-and-projections


d_calcs_rgn <- d_calcs_rgn %>%
  left_join(population_weights, by="country") %>%
  mutate(population = ifelse(is.na(population), 1, population)) %>% 
  group_by(rgn_id, year) %>%
  summarize(score = weighted.mean(score, population),
            gapfill_within_rgn = weighted.mean(gap_fill, population)) %>%
  ungroup() %>%
  filter(rgn_id <= 250)

summary(d_calcs_rgn)

```


# Gapfill, part 2: Filling in missing territorial region value

Assigning territorial region value to be the mean of parent country value and territorial regions with data (using same `sov_id`).

```{r}

## data that describes territories of countries
territory = rgn_master %>% 
  select(rgn_id = rgn_id_2013,
         sov_id) %>%               
  group_by(rgn_id) %>% # remove duplicated countries from this rgn_id list
  summarize(sov_id = mean(sov_id, na.rm=T)) %>% # duplicates always have the same sov_id (r2 value)
  filter(rgn_id <= 250, rgn_id != 213)

	
## expand to include all years of data
territory <- data.frame(year = start_year:end_year) %>% 
  merge(territory, by=NULL) 


## assign territories the values of their sovereign country
d_sovs = d_calcs_rgn %>% 
  full_join(territory, by = c('rgn_id', 'year')) %>%
  group_by(sov_id, year) %>%
  mutate(score_gf_territory = mean(score, na.rm=TRUE),
         gapfill_within_rgn = mean(gapfill_within_rgn, na.rm=TRUE))%>%
   filter(!is.na(gapfill_within_rgn)) %>%
  ungroup()

# filter(d_sovs, rgn_id %in% c(1,2,3,16) & year == 2018) %>% View() # check in console to make sure Australia and territories have the same score and sov_id
```


Define new data object from d_sovs which includes gapfill method and gapfilled scores:

```{r}

d_gf2 <- d_sovs %>%
  mutate(gapfill_territory = ifelse(is.na(score) & !is.na(score_gf_territory), "territory", "NA")) %>%
  mutate(score = ifelse(is.na(score), score_gf_territory, score)) %>%
  select(rgn_id, year, score, gapfill_within_rgn, gapfill_territory)

summary(d_gf2)
```

Add region names and clean the region data, and make sure we have all the regions:

```{r}

# get region names
regions <- rgn_master %>%
  filter(rgn_typ == "eez") %>%
  filter(rgn_id_2013 <= 250) %>% # > 250 are either FAO or a disputed region
  filter(rgn_id_2013 != 213) %>% # 213 is Antarctica
  select(rgn_id = rgn_id_2013, rgn_name=rgn_nam_2013) %>%
  unique() %>%
  arrange(rgn_id) 

d_gf2 <- regions %>%
  left_join(d_gf2)
  
```


## Look at data table for the territories (gapfilled)

```{r}

head(d_sovs)
summary(d_sovs) 
# v2021: score has 1029 NAs
# v2022: 1127 NAs
# v2024: 1176 NAs
# v2025: 1736 NAs Hmm, more here than expected

```

## Look at table with scores and gapfilling methods

```{r}

## check for NA values within "score" variable
## if so, need to gapfill using UN geopolitical regions
summary(d_gf2) 
d_gf2[is.na(d_gf2$score), ]
# v2021: score has 0 NAs
# v2022: 0 NAs
# v2023: 0 NAs
# v2024: 0 NAs! 
# v2025: Taiwan oof

```

# Check data 

Comparing this year's values against last year's.  These should be the same unless there have been updates to WGI source data or a change to methods.  

Plot most recent shared year between last year and this year's data, and look for a relationship close to a 1:1 relationship. If data are significantly off the line, look at the original (raw) data to investigate.

```{r}

# Most recent shared year should be three years before prior to the assessment year (scen_year)
recent_shared_year <- as.numeric(scen_year) - 2 
prev_assessment_year <- as.numeric(scen_year) - 1

new_2022_scores <- d_gf2 %>% # rename variable with respective year for clarity
  filter(year == 2023) %>% 
  select(rgn_id, score)


old_wgi <- read_csv(here(previous_wgi_dir, "output", "wgi_res.csv")) %>% # rename variable
  filter(year == 2022) %>%
  select(rgn_id, old_score=resilience_score)

colnames(d_gf2)
colnames(old_wgi)

score_compare_2022 <- d_gf2 %>% 
  filter(year == 2022) %>% 
  left_join(old_wgi, by = join_by('rgn_id'))

ab_scores <- ggplot() + 
  geom_abline(slope = 1, intercept = 0, col = "red") +
  geom_point(data = score_compare_2022, aes(x = old_score, y = score, text = paste("Region:", rgn_name))) +
  labs(x = "Old Score", 
       y = "New Score",
       title = "WGI 2025 score vs. 2024 score data year 2022") +
  theme_bw();ggplotly(ab_scores)

score_compare_plot <- ggplot(score_compare_2022, aes(x = old_score, y = score, text = rgn_id)) +
  geom_point() +
  geom_abline(slope=1, intercept = 0, col = "red")
   
score_compare_plot
# v2024: looks good

```

```{r}
# Read in the old data
old <- readr::read_csv(here::here("globalprep", "np", prev_scen_year, "output", "np_fofm_scores.csv"))
summary(old)

# Read in the new data
new <- readr::read_csv(here::here(current_np_dir, "output", "np_fofm_scores.csv"))
summary(new)

# load ohi regions
region_data()

# Investigating changes in data for same year in different data/scenario years
check <- new %>%
  rename("new_score" = "score") %>%
  left_join(old, by = c("rgn_id", "year")) %>%
  left_join(rgns_all, by = "rgn_id") %>%
  select(rgn_id, year, new_score, score, rgn_name) %>%
  mutate(diff = new_score - score) %>%
  filter(year == 2019)

ab_scores <- ggplot() + 
  geom_abline(slope = 1, intercept = 0, col = "red") +
  geom_point(data = check, aes(x = score, y = new_score, text = paste("Region:", rgn_name), alpha = 0.7)) +
  labs(x = "Old Score", 
       y = "New Score",
       title = "FOFM New score vs old score for 2019") +
  theme_bw();ggplotly(ab_scores)
```


Look at top/bottom 10 regions to make sure these seem reasonable:

```{r}

## Top/Bottom 10 scorers:

tmp <- d_gf2 %>%
  filter(year==2022) %>%
  arrange(score) %>%
  select(rgn_id, score) %>%
  left_join(regions)

tmp[1:10, ] # mostly countires with politcal instability 
tmp[211:220, ] # makes sense... mostly countries with democratic socialist policies or small islands

hist(tmp$score)

```

Look at a summary to confirm scores are between 0 and 1, there are 220 regions, and there are no NAs (for this particular dataset):

```{r}

summary(d_gf2) # scores are between 0-1, gapfilled info present, no NAs
length(unique(d_gf2$rgn_id)) # 220 regions
c(min(d_gf2$score), max(d_gf2$score)) # checking for score min and max; making sure they're between 0-1

```

# Save the data
Save gapfilling and data for this assessment year.


```{r}

tmp_data_res <- d_gf2 %>%
  select(rgn_id, year, resilience_score = score)
write.csv(tmp_data_res, here(paste0("globalprep/prs_res_wgi/", scen_year, "/output/wgi_res.csv")), row.names=FALSE)

tmp_data_prs <- d_gf2 %>%
  mutate(score = 1 - score) %>%
  select(rgn_id, year, pressure_score = score)
write.csv(tmp_data_prs, here(paste0("globalprep/prs_res_wgi/", scen_year, "/output/wgi_prs.csv")), row.names=FALSE)

# gapfilled data
tmp_gf <- d_gf2 %>%
  select(rgn_id, year, gapfill_within_rgn, gapfill_territory) %>%
  mutate(gapfill_within_rgn = ifelse(gapfill_within_rgn == 0, NA, gapfill_within_rgn)) %>%
  mutate(gapfill_within_rgn = ifelse(!is.na(gapfill_within_rgn), 
                                            paste0("gapfill_within_rgn: ", gapfill_within_rgn), NA)) %>%
  mutate(gapfill_territory = ifelse(gapfill_territory == "territory", "territory, mean of admin countries", NA)) %>%
  mutate(method = paste(gapfill_within_rgn, gapfill_territory, sep="; ")) %>%
  mutate(method = gsub("NA; ", "", method)) %>%
  mutate(method = gsub("; NA", "", method)) %>%
  mutate(method = gsub("NA", NA, method)) %>%
  mutate(gapfilled = ifelse(is.na(method), 0, 1)) %>%
  select(rgn_id, year, gapfilled, method)


write.csv(tmp_gf, here(paste0("globalprep/prs_res_wgi/", scen_year, "/output/wgi_gf.csv")), row.names=FALSE)


```



