---
title: 'OHI `r format(Sys.Date(), "%Y")` - World Governance Indicator'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

[REFERENCE RMD FILE](http://ohi-science.org/ohiprep_v2025/globalprep/prs_res_wgi/v2025/WGI_dataprep.html)

# Summary
This script downloads WGI data and prepares it for a pressures (1 - WGI) and resilience data layer.  

## Updates from previous assessment
One thing to note is that there is a large discrepancy in the available countires depending on whether you get the data from the online repository versus the API. The API provides ~60 extra regions that in the v2025 assessment all had no data. So, in this case, it didn't make a huge difference. However, to keep things in the v2025 assessment inline with the v2024 assessment, the dataset downloaded from the online repository was used instead of the API dataset. The code to acccess the API will be kept, but commented out in case future fellows would like to use it. 

- New data download
- Added `librarian` to handle package imports
- Added directory, scenario, and year variables to make the script more reproducible
- Updated the exploratory code and data import at the top to be less redundant


## Data Source 
**Reference**:  http://info.worldbank.org/governance/wgi/index.aspx#home    

**Downloaded**: 

July 29, 2025 (source data updated Sep 2023)

**Description**:  
The Worldwide Governance Indicators (WGI) project reports aggregate and individual governance indicators for 215 economies over the period 1996–2023, for six dimensions of governance:

* Voice and Accountability
* Political Stability and Absence of Violence
* Government Effectiveness
* Regulatory Quality
* Rule of Law
* Control of Corruption

**Time range**: 1996-2023


```{r setup}
knitr::opts_chunk$set(eval = FALSE)

# ======= Load packages ============
if (!require(librarian)) {
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  ohicore, #devtools::install_github('ohi-science/ohicore@dev') #if relevant: restart session after reinstalling
  WDI, # install.packages('WDI')  # used to extract World Development Indicator (World Bank) data 
  here,
  janitor,
  tools,
  tidyverse,
  plotly,
  strex
)

# check website to see what years are available: http://info.worldbank.org/governance/wgi/index.aspx#home


# ======= Set directories ===========
# Update scenario year, set up programmatic scenario year updates
current_year <- 2025 # update this!!!
previous_year <- current_year - 1

# Directories
data_dir <- paste0("d", current_year)
prev_data_dir <- paste0("d", previous_year)

# Scenarios
scen_year <- paste0("v", current_year)
prev_scen_year <- paste0("v", previous_year)

start_year <- 1996
end_year <- 2023 # Update depending on available data from Worldbank website

# WGI directory on GitHub
current_wgi_dir <- here::here("globalprep", "prs_res_wgi", scen_year) 
previous_wgi_dir <- here::here("globalprep", "prs_res_wgi", prev_scen_year) 

# ======== Source functions ===========
source(here::here("workflow", "R", "common.R")) # directory locations
# functions specific to mariculture dealing with compound countries, updated in 2024
source(here::here("globalprep", "mar", scen_year, "mar_fxs.R"))

# Avoid scientific notation for numbers with many sig figs
options(scipen = 999)

```

# Obtain the WGI data
Download each of the 6 WGI indicators. Data was manually downlaoded the data from here: https://databank.worldbank.org/source/worldwide-governance-indicators#. The API should also work, but v2025 assessment found the web download data was slightly cleaner and easier to work with. It could be worth checking both for future assessment years. 

Combine the indicators into a single table, with a column for each indicator, and rows for each country-year pair.

```{r web-access-data}
# Read in manually downloaded data
wgi_raw <- read_csv(here::here(current_wgi_dir, "raw", "P_Data_Extract_From_Worldwide_Governance_Indicators_2025.csv"))

# Clean column names and pivot
d_web <- wgi_raw %>%
  clean_names() %>%
  rename(iso3c = country_code, country = country_name) %>%
  filter(str_detect(series_code, "EST")) %>%
  dplyr::select(-series_name) %>%
  pivot_longer(cols = starts_with("x"), values_to = "est", names_to = "year") %>%
  mutate(year = as.numeric(str_after_last(year, "yr"))) %>%
  mutate(est = ifelse(est == "..", NA, est)) %>%
  pivot_wider(values_from = "est", names_from = "series_code")

# Quick check
colSums(is.na(d_web))
summary(d_web)
# view(d_web)

# Write out data for archival purposes
write_csv(d_web, here(current_wgi_dir, "raw", paste0("worldbank_wgi_from_web_", current_year, ".csv")))
```

These two chunks are the old code to access the API. They do not need to be run if you are working with the data download.

```{r api-indicator-decsription, eval=FALSE}
# Import WDI data via WDI package
# indicators <-  data.frame(WDI_data[[1]])
# 
# ind_names <- list("VA.EST", "PV.EST", "GE.EST", "RQ.EST", "RL.EST", "CC.EST")
# 
# # View indicator names and descriptions
# for (ind_name in ind_names){
#   
#   name <- indicators[grep(ind_name, indicators$indicator), 2] 
#   description <- indicators[grep(ind_name, indicators$indicator), 3] 
#   cat("\nIndicator name:", name)
#   cat("\nDescription:", description)
#   cat("\n") # space between each indicator
# 
# }
```


```{r api-extract-data, eval=FALSE}
# Create list of indicator names and short names
# ind_dict <- list(
#   "VA.EST" = "Voice and Accountability: Estimate",
#   "PV.EST" = "Political Stability and Absence of Violence/Terrorism: Estimate",
#   "GE.EST" = "Government Effectiveness: Estimate",
#   "RQ.EST" = "Regulatory Quality: Estimate",
#   "RL.EST" = "Rule of Law: Estimate",
#   "CC.EST" = "Control of Corruption: Estimate"
# )
# 
# # Query the catalog for each of the 6 indicators
# indicator_data <- list()
# 
# for (key in names(ind_dict)) {
#   value <- ind_dict[[key]]
#   
#   indicator_code <- WDIsearch(value, field = 'name')[1, 'indicator']
#   
#   df <- WDI(indicator = indicator_code,country = 'all',
#             start = start_year, end = end_year)
#   
#   # Rename the indicator column to the key
#   colnames(df)[which(names(df) == indicator_code)] <- key
#   
#   # Keep relevant columns only
#   df <- df %>%
#     select(country, iso3c, year, all_of(key))
#   
#   indicator_data[[key]] <- df
#   
# }
# 
# # Join the individual dfs together
# final_data <- Reduce(function(x, y) {
#   left_join(x, y, by = c("country", "iso3c", "year"))
# }, indicator_data)
# 
# # Check
# # view(final_data)
# colSums(is.na(final_data))
# nrow(final_data)

# Update file name and pathing
# write_csv(final_data, here(current_wgi_dir, "raw", paste0("worldbank_wgi_from_api_", current_year, ".csv")))
# This dataset currently has non-OHI regions included

```


# Gapfill, part 1: filling missing years of data for indicators, within countries
The first gapfilling occurs when we use the average of previous years data within each region/indicator.  This occurs when a region has data for an indicator, but not for all years.

```{r gapfill-missing-years}
# Read in WGI data
d_web <- read_csv(here(current_wgi_dir, "raw", paste0("worldbank_wgi_from_web_", current_year, ".csv")))

# Pivot for gapfilling
d_long <- d_web %>% 
  pivot_longer(cols = CC.EST:VA.EST, names_to = "indicator", values_to = "value") %>% 
  mutate(value = as.numeric(value))


# Each country has 20 years of data
d_gap_fill  <- d_long %>%
  group_by(country, iso3c, year) %>%
  mutate(NA_count_c_y = sum(is.na(value))) %>% # gf record: NA values within a region/year prior to gapfilling, max value is 6 (meaning that a country has no data)
  ungroup() %>%
  group_by(country, iso3c, indicator) %>% # gapfill missing data with mean of values across years within the same region/indicator
  mutate(ind_mean_c_i = mean(value, na.rm=TRUE)) %>%
  ungroup() %>%
  mutate(value = ifelse(is.na(value), ind_mean_c_i, value)) %>%
  group_by(country, iso3c, year) %>% 
  mutate(NA_count_post_gf1 = sum(is.na(value))) # gf record: NA values within a region/year after within region/indicator gapfilling (i.e. indicator is gapfilled by other years of data), used to cut regions <4 indicators (below)    

```


## Safeguard: cut regions with < 4 indicators (if any) to calculate score.
Once gapfilling is complete, the WGI scores are calculated as an average of the 6 indicators.  However, if a country is missing 4 or more of the indicators within a year the average would be very biased.  In these cases, a different method should be used to gapfill these data 

(NOTE: for the 2024 assessment all regions had at least 4 of the 6 indicators).

```{r drop-data-poor-rgn}

countries_no_data <- d_gap_fill %>%
  filter(NA_count_post_gf1 > 3)

countries_no_data <- unique(countries_no_data$country)
countries_no_data # v2025 0 countries

## In this case, the countries with minimal data (< 4 indicators ever calculated) are deleted.  
## These will be gap-filled later on if they are deleted now.
d_gap_fill <- d_gap_fill %>%
  filter(!(country %in% countries_no_data))

```

# Calculate overall WGI score for each country

This involves:

* taking the average of the 6 indicators (assuming there are at least 4 of the 6 indicators)
* rescaling the data from 0 to 1 

```{r calculate-score}
d_calcs <- d_gap_fill %>%
  group_by(country, iso3c, year) %>%
  summarize(score_wgi_scale = mean(value, na.rm=T),
            NA_start = mean(NA_count_c_y), # initial mean number of NA across indicators, pre-gapfill 
            NA_post_gf_1 = mean(NA_count_post_gf1)) %>% # number of NA across indicators, post-gapfill across year gapfill within region/indicator
  ungroup() 

# scaled scores should be between -2.5 to 2.5
summary(d_calcs) # checking to make sure NA values make sense
# v2025: NA_start max = 6, NA_post_gf_1 max = 0
# v2025: score_wgi_scale min = -2.409, max = 1.947... perfect

```


```{r convert-score}
# Define the wgi score range
wgi_range = c(-2.5, 2.5) # historic values have been between -2.5 and 2.5

# Convert scores between 0 and 1, using min/max to scale the data
d_calcs <- d_calcs %>%
  mutate(score =  (score_wgi_scale - wgi_range[1]) / (wgi_range[2] - wgi_range[1])) %>%
  ungroup()
summary(d_calcs)

# Document gapfilling
d_calcs <- d_calcs %>%
  # if there are values in NA_post_gf_1, it means these weren't gapfilled
  mutate(gap_fill = NA_start - NA_post_gf_1,   
         gap_fill = ifelse(is.na(score), 0, gap_fill)) %>% # number of values that were gapfilled
  select(-NA_start, -NA_post_gf_1)

```

## Save intermediate file 

```{r save-int-file}
# Explore & check intermediate d_calcs data table
d_calcs[d_calcs$gap_fill>0, ] %>% View()
d_calcs[d_calcs$country == "Niue", ] %>% View() # should have gap-fill values between 0-6
d_calcs[d_calcs$country == "American Samoa",] %>% View() # should have gap-fill values between 0-6

# Save intermediate file of wgi scores pre-gapfilling (for OHI+ use)
d_gf_final <- d_calcs %>% 
  select(country, iso3c, year, score_wgi_scale, score_ohi_scale = score, gap_fill)

write_csv(d_gf_final, here(current_wgi_dir, "int", "wgi_combined_scores_by_country.csv"))

```

# Convert country names to ohi regions

v2025 note: this part is very messy and future assessment could look into applying the `ohicore` function `split_regions()`, [here](https://github.com/OHI-Science/ohicore/blob/master/R/split_regions.R). Did not have a chance to get to it this year, but it's worth doing in the future. 

```{r country-to-region}
## We report these regions at a greater spatial resolution:

## Aruba is part of the Netherlands Antilles, but it is reported separately
country_split_1 <- data.frame(country = "Netherlands Antilles", region = c('Bonaire', 'Curacao', 'Saba', 'Sint Maarten', 'Sint Eustatius'))
country_split_2 <- data.frame(country = "Jersey, Channel Islands", region = c('Jersey', 'Guernsey'))
country_split <- rbind(country_split_1, country_split_2)

country_split_data <- country_split %>%
  left_join(d_calcs) %>%
  select(-country) %>%
  rename(country = region)

d_calcs <- d_calcs %>%
  filter(!(country %in% c("Netherlands Antilles", "Jersey, Channel Islands"))) %>%
  rbind(country_split_data)  %>%
  mutate(country = as.character(country)) %>%
  dplyr::select(-iso3c)

d_calcs$country[grep("Korea, Dem.", d_calcs$country)] <- "North Korea"

d_calcs$country[grep("Turkiye", d_calcs$country)] <- "Turkey"
# Maybe in future update package with country synonym list


## Function to convert to OHI region ID
d_calcs_rgn <- name_2_rgn(df_in = d_calcs, 
                       fld_name='country', 
                       flds_unique=c('year'))


## Combine the duplicate regions (we report these at lower resolution)
## In this case, we take the weighted average
population_weights <- data.frame(country = c("Virgin Islands (U.S.)", "Puerto Rico",
                                             "China", "Hong Kong SAR, China", "Macao SAR, China"),
                                 population = c(104337, 3203295, 
                                                1408975000, 7524100, 687000))
# updated population values on 30 July 2025 (source: World Bank website, 2024 values); https://databank.worldbank.org/source/population-estimates-and-projections


d_calcs_rgn <- d_calcs_rgn %>%
  left_join(population_weights, by="country") %>%
  mutate(population = ifelse(is.na(population), 1, population)) %>% 
  group_by(rgn_id, year) %>%
  summarize(score = weighted.mean(score, population),
            gapfill_within_rgn = weighted.mean(gap_fill, population)) %>%
  ungroup() %>%
  filter(rgn_id <= 250)

summary(d_calcs_rgn)

```

# Gapfill, part 2: Filling in missing territorial region value

Assigning territorial region value to be the mean of parent country value and territorial regions with data (using same `sov_id`).

```{r gapfill-territories}
# data that describes territories of countries
territory = rgn_master %>% 
  select(rgn_id = rgn_id_2013,
         sov_id) %>%               
  group_by(rgn_id) %>% # remove duplicated countries from this rgn_id list
  summarize(sov_id = mean(sov_id, na.rm=T)) %>% # duplicates always have the same sov_id (r2 value)
  filter(rgn_id <= 250, rgn_id != 213)

# expand to include all years of data
territory <- data.frame(year = start_year:end_year) %>% 
  merge(territory, by=NULL) 


# assign territories the values of their sovereign country
d_sovs = d_calcs_rgn %>% 
  full_join(territory, by = c('rgn_id', 'year')) %>%
  group_by(sov_id, year) %>%
  mutate(score_gf_territory = mean(score, na.rm=TRUE),
         gapfill_within_rgn = mean(gapfill_within_rgn, na.rm=TRUE))%>%
   filter(!is.na(gapfill_within_rgn)) %>%
  ungroup()

# filter(d_sovs, rgn_id %in% c(1,2,3,16) & year == 2018) %>% view() # check in console to make sure Australia and territories have the same score and sov_id
```

Define new data object from d_sovs which includes gapfill method and gapfilled scores, and add region names and clean the region data, and make sure we have all the regions.

```{r add-regions}
d_gf2 <- d_sovs %>%
  mutate(gapfill_territory = ifelse(is.na(score) & !is.na(score_gf_territory), "territory", "NA")) %>%
  mutate(score = ifelse(is.na(score), score_gf_territory, score)) %>%
  select(rgn_id, year, score, gapfill_within_rgn, gapfill_territory)

# get region names
regions <- rgn_master %>%
  filter(rgn_typ == "eez") %>%
  filter(rgn_id_2013 <= 250) %>% # > 250 are either FAO or a disputed region
  filter(rgn_id_2013 != 213) %>% # 213 is Antarctica
  select(rgn_id = rgn_id_2013, rgn_name=rgn_nam_2013) %>%
  unique() %>%
  arrange(rgn_id) 

d_gf2 <- regions %>%
  left_join(d_gf2)

summary(d_gf2)
```


## Gapfill check

1. Look at table with scores and gapfilling methods
2. Look at data table for the territories (gapfilled)

```{r na-check}
colSums(is.na(d_sovs))['score'] 
summary(d_sovs) 
# v2021: score has 1029 NAs
# v2022: 1127 NAs
# v2024: 1176 NAs
# v2025: 1225 NAs


# check for NA values within "score" variable
# if so, need to gapfill using UN geopolitical regions
summary(d_gf2) 
colSums(is.na(d_sovs))['score'] 
# v2021: score has 0 NAs
# v2022: 0 NAs
# v2023: 0 NAs
# v2024: 0 NAs! 
# v2025: 0 NAs
```

# Check data 

Comparing this year's values against last year's.  These should be the same unless there have been updates to WGI source data or a change to methods.  

Plot most recent shared year between last year and this year's data, and look for a relationship close to a 1:1 relationship. If data are significantly off the line, look at the original (raw) data to investigate.

```{r plot-data-check}
# Read in the old data
old_wgi <- read_csv(here(previous_wgi_dir, "output", "wgi_res.csv")) %>% 
  # Find the max year to compare with the new data
  filter(year == max(year)) %>%
  select(rgn_id, old_score = resilience_score, year)

summary(old_wgi)

# Read in the new data
new_wgi <- d_gf2 %>% 
  # Filter to old data's max year
  filter(year == max(old_wgi$year)) %>%
  select(rgn_id, new_score = score, year)

summary(new_wgi)

# load ohi regions
region_data()

# Investigating changes in data for same year in different data/scenario years
check <- new_wgi %>%
  left_join(old_wgi, by = c("rgn_id", "year")) %>%
  left_join(rgns_all, by = "rgn_id") %>%
  select(rgn_id, year, old_score, new_score, rgn_name) %>%
  mutate(diff = new_score - old_score)

wgi_plot <- ggplot() + 
  geom_abline(slope = 1, intercept = 0, col = "red") +
  geom_point(data = check, aes(x = old_score, y = new_score, text = paste("Region:", rgn_name)),  alpha = 0.7) +
  labs(x = "Old Score", 
       y = "New Score",
       title = paste0("WGI New score vs old score for data year ", max(old_wgi$year))) +
  theme_bw();ggplotly(wgi_plot)
```

Look at top/bottom 10 regions to make sure these seem reasonable and look at a summary to confirm scores are between 0 and 1, there are 220 regions, and there are no NAs (for this particular dataset):

```{r data-check}
# Top/Bottom 10 scorers:

tmp <- d_gf2 %>%
  filter(year==2022) %>%
  arrange(score) %>%
  select(rgn_id, score) %>%
  left_join(regions)

tmp[1:10, ] # mostly countires with politcal instability 
tmp[211:220, ] # makes sense... mostly countries with democratic socialist policies or small islands

hist(tmp$score)

summary(d_gf2) # scores are between 0-1, gapfilled info present, no NAs
length(unique(d_gf2$rgn_id)) # 220 regions
c(min(d_gf2$score), max(d_gf2$score)) # checking for score min and max; making sure they're between 0-1

```


# Save the data
Save gapfilling and data for this assessment year.

```{r export-data}
# ---- Resilience ----
tmp_data_res <- d_gf2 %>%
  select(rgn_id, year, resilience_score = score)

write_csv(tmp_data_res, here(current_wgi_dir, "output", "wgi_res.csv"))

# ---- Pressure ----
tmp_data_prs <- d_gf2 %>%
  mutate(score = 1 - score) %>%
  select(rgn_id, year, pressure_score = score)

write_csv(tmp_data_prs, here(current_wgi_dir, "output", "wgi_prs.csv"))

# ---- Gapfill ----
tmp_gf <- d_gf2 %>%
  select(rgn_id, year, gapfill_within_rgn, gapfill_territory) %>%
  mutate(gapfill_within_rgn = ifelse(gapfill_within_rgn == 0, NA, gapfill_within_rgn)) %>%
  mutate(gapfill_within_rgn = ifelse(!is.na(gapfill_within_rgn), 
                                            paste0("gapfill_within_rgn: ", gapfill_within_rgn), NA)) %>%
  mutate(gapfill_territory = ifelse(gapfill_territory == "territory", "territory, mean of admin countries", NA)) %>%
  mutate(method = paste(gapfill_within_rgn, gapfill_territory, sep="; ")) %>%
  mutate(method = gsub("NA; ", "", method)) %>%
  mutate(method = gsub("; NA", "", method)) %>%
  mutate(method = gsub("NA", NA, method)) %>%
  mutate(gapfilled = ifelse(is.na(method), 0, 1)) %>%
  select(rgn_id, year, gapfilled, method)

write_csv(tmp_gf, here(current_wgi_dir, "output", "wgi_gf.csv"))
```

## Comparing Web vs. API

```{r web-vs-api}
# Comparing the API data and the web download data
d_api <- read_csv(here(current_wgi_dir, "raw", paste0("worldbank_wgi_from_api_", current_year, ".csv")))
colnames(d_api)

d_web <- write_csv(d_web, here(current_wgi_dir, "raw", paste0("worldbank_wgi_from_web_", current_year, ".csv")))
colnames(d_web)

colSums(is.na(d_web)) # 117-231 NAs across all 6 inds
nrow(d_web) # 530
# ~ 38% NA

colSums(is.na(d_api)) # 2370 - 2482 NAs across all 6 inds
nrow(d_api) # 7448
# ~ 32% NA

diff_countries <- setdiff(d_api$country, d_web$country)

diff_df <- filter(d_api, country %in% diff_countries) %>% 
  view()

length(unique(diff_df$country))
# Puerto Rico (US) is the only country different between the API and website data that actually has data for all 6 indicators. 
```


