---
title: 'OHI 2021: Mangrove extent'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---

# Summary

This script generates the extent of mangrove for each OHI region for the latest year of data (year == 2016). We do this using polygon data. We extract the area of each polygon per each ohi region using our combined land and eez raster.


## Updates from previous assessment
Using a new dataset from global mangrove watch. 

***
## Data Source 

**Reference**: Bunting, P., Rosenqvist, A., Lucas, R., Rebelo, L.-M., Hilarides, L., Thomas, N., Hardy, A., Itoh, T., Shimada, M., Finlayson, C., 2018. The Global Mangrove Watch—A New 2010 Global Baseline of Mangrove Extent. Remote Sensing 10, 1669. https://doi.org/10.3390/rs10101669

**Downloaded**: 03/09/2021

**Description**:  
Global Mangrove Watch (1996 - 2016)	
https://data.unep-wcmc.org/datasets/45
Reported at spatial cell scale. 

The GMW aims to provide geospatial information about mangrove extent and
changes to the Ramsar Convention, national wetland practitioners, decision makers
and NGOs. It is part of the Ramsar Science and Technical Review Panel (STRP) work
plan for 2016-2018 and a Pilot Project to the Ramsar Global Wetlands Observation
System (GWOS), which is implemented under the GEO-Wetlands Initiative. The
primary objective of the GMW has been to provide countries lacking a national
mangrove monitoring system with first cut mangrove extent and change maps, to
help safeguard against further mangrove forest loss and degradation.
The GMW has generated a global baseline map of mangroves for 2010 using ALOS
PALSAR and Landsat (optical) data, and changes from this baseline for seven epochs
between 1996 and 2017 derived from JERS-1, ALOS and ALOS-2. Annual maps are
planned from 2018 and onwards

**Time range**: 1996-2016


***
# Methods


## Setup
``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(readr)      # for read_csv()
library(raster)
library(here)
library(sf)
library(fasterize)
library(tidyverse)
library(mapview)
library(sp)
library(rgeos)

source(file.path('~/github/ohiprep_v2021/workflow/R/common.R'))

goal     <- 'globalprep/hab_mangrove/v2021'
dir_git  <- file.path('~/github/ohiprep_v2021', goal)
dir_wcmc <- file.path(file.path(dir_M, 'git-annex/globalprep/_raw_data/wcmc_mangrove'))
ohi_rasters() # call the region zones raster
regions_shape()
region_data()

land_ocean <- raster("/home/shares/ohi/git-annex/globalprep/spatial/v2017/regions_land_ocean.tif")
plot(log(land_ocean+1)) # i think this is the one i want

land_ocean
```

Read in GMW data and explore
```{r, eval = FALSE}

files <- list.files("/home/shares/ohi/git-annex/globalprep/_raw_data/wcmc_mangrove/01_Data/")

files_2 <- files[grepl(".shp", files) & !grepl(".xml", files)]
```


Loop over all years of data and match their regions to OHI regions 

```{r, eval = FALSE}
  
for(i in 1:7){ ## there are 7 files 
 # i = 1
  
gmw_file = files_2[i]  

gmw_year = gsub("_v2.shp", "", gmw_file)

d_year <- gsub("GMW_", "", gmw_year)

fasterize_file = file.path(dir_M, "/git-annex/globalprep/Mangrove/v2021/int", sprintf("py_mangrove_rast_%s.tif", gmw_year))
  
if(!file.exists(fasterize_file)) {

gmw_shp <- st_read(file.path(dir_wcmc, "01_Data", gmw_file))
  
## transform 2016 data to our projections
gmw_moll <- st_transform(gmw_shp, crs = "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")


#fasterize our polygon dataset into a raster
fasterize_py <- fasterize::fasterize(gmw_moll, raster = land_ocean, 
                                     field = NULL) # all polygons given value of 1

fasterize_py

print(cellStats(fasterize_py, "sum", na.rm=TRUE)*0.934478877011218970**2) 

writeRaster(fasterize_py, filename = file.path(dir_M, "/git-annex/globalprep/Mangrove/v2021/int", sprintf("py_mangrove_rast_%s.tif", gmw_year)), overwrite=TRUE) 




# Calculate zonal stats with zones raster and new combined seagrass. Convert to km^2 and save int/output files*

zonal_sums_combined <- raster::zonal(fasterize_py, 
                                     land_ocean,
                                     fun = "sum",
                                     na.rm = TRUE,
                                     progress="text") #sum all mangrove cells for each ohi zone
zonal_sums_combined_df <- data.frame(zonal_sums_combined)

zonal_sums_km2 <- zonal_sums_combined_df %>%
  mutate(year = d_year, habitat = "mangrove",
         km2 = (0.934478877011218970**2*sum)) %>% #one cell is equal to ~0.93 km
  rename("rgn_id" = "zone") %>%
  select(-sum)

print(sum(zonal_sums_km2$km2))


## Testing year 2016:
#136804.3 km2.. worked perfectly!

# From GMW website: 
# "The area of mangrove habitat in Indonesia was 26,508.12  km² in 2016"

# Our estimates say indonesia has 26684.8 km2... nearly perfect... i believe it worked. Now I will do this with every year... 

write.csv(zonal_sums_km2, file.path(here("globalprep/hab_mangrove/v2021/int", sprintf("habitat_extent_mangrove_%s.csv", d_year))), row.names = FALSE)


}else{
  print("file exists, skipping")
}
}
```

Save 2016 as most current year to use in global

```{r, eval = FALSE}
gmw_ohi_2016 <- read.csv(file.path(here("globalprep/hab_mangrove/v2021/int/habitat_extent_mangrove_2016.csv"))) 

gmw_ohi_2016 <- gmw_ohi_2016 %>%
  filter(rgn_id <= 250)

write.csv(gmw_ohi_2016, file.path(here("globalprep/hab_mangrove/v2021/data/habitat_extent_mangrove_updated.csv")), row.names = FALSE)
```



Datacheck: 

```{r, eval = FALSE}
gmw_2016 <- st_read(file.path(dir_wcmc, "01_Data/GMW_2016_v2.shp"))

sum(st_area(gmw_2016))*0.000001 # 135927.3 km2

colnames(gmw_2016) # [1] "ogc_fid"  "pxlval"   "geometry"

head(gmw_2016)

# Simple feature collection with 6 features and 2 fields
# geometry type:  POLYGON
# dimension:      XY
# bbox:           xmin: -62.86778 ymin: 9.999333 xmax: -62.45511 ymax: 10
# geographic CRS: WGS 84
#   ogc_fid pxlval                       geometry
# 1       1      1 POLYGON ((-62.86778 10, -62...
# 2       2      1 POLYGON ((-62.866 10, -62.8...
# 3       3      1 POLYGON ((-62.846 10, -62.8...
# 4       4      1 POLYGON ((-62.84311 10, -62...
# 5       5      1 POLYGON ((-62.45556 10, -62...
# 6       6      1 POLYGON ((-62.74844 10, -62...


gmw_1996 <- st_read(file.path(dir_wcmc, "01_Data/GMW_1996_v2.shp"))

sum(st_area(gmw_1996))*0.000001 # 142004.9 km2


gmw_2007 <- st_read(file.path(dir_wcmc, "01_Data/GMW_2007_v2.shp"))

sum(st_area(gmw_2007))*0.000001 # 138110.5 km2


gmw_2008 <- st_read(file.path(dir_wcmc, "01_Data/GMW_2008_v2.shp"))

sum(st_area(gmw_2008))*0.000001 # 138324.3  km2


gmw_2009 <- st_read(file.path(dir_wcmc, "01_Data/GMW_2009_v2.shp"))

sum(st_area(gmw_2009))*0.000001 # 138098.9 km2


gmw_2010 <- st_read(file.path(dir_wcmc, "01_Data/GMW_2010_v2.shp"))

sum(st_area(gmw_2010))*0.000001 # 136839.3 km2


gmw_2015 <- st_read(file.path(dir_wcmc, "01_Data/GMW_2015_v2.shp"))

sum(st_area(gmw_2015))*0.000001 # 135968.2 km2


## Now compare to what we calculated above 

gmw_ohi_2016 <- read.csv(file.path(here("globalprep/hab_mangrove/v2021/int/habitat_extent_mangrove_2016.csv")))
sum(gmw_ohi_2016$km2) # 136804.3

gmw_ohi_1996 <- read.csv(file.path(here("globalprep/hab_mangrove/v2021/int/habitat_extent_mangrove_1996.csv")))
sum(gmw_ohi_1996$km2) # 142919.7

gmw_ohi_2007 <- read.csv(file.path(here("globalprep/hab_mangrove/v2021/int/habitat_extent_mangrove_2007.csv")))
sum(gmw_ohi_2007$km2) # 139059.1

gmw_ohi_2008 <- read.csv(file.path(here("globalprep/hab_mangrove/v2021/int/habitat_extent_mangrove_2008.csv")))
sum(gmw_ohi_2008$km2) # 139264.3

gmw_ohi_2009 <- read.csv(file.path(here("globalprep/hab_mangrove/v2021/int/habitat_extent_mangrove_2009.csv")))
sum(gmw_ohi_2009$km2) # 139033.7

gmw_ohi_2010 <- read.csv(file.path(here("globalprep/hab_mangrove/v2021/int/habitat_extent_mangrove_2010.csv")))
sum(gmw_ohi_2010$km2) # 137775.4

gmw_ohi_2015 <- read.csv(file.path(here("globalprep/hab_mangrove/v2021/int/habitat_extent_mangrove_2015.csv")))
sum(gmw_ohi_2015$km2) # 136855.9

## All of our estimates are a little bit bigger than the raw data estimates, probably due to reprojections. I am ok with this amount of error. 

old_extent <- read.csv(file.path(here("globalprep/hab_mangrove/v2015/data/habitat_extent_mangrove_updated.csv"))) %>%
  filter(habitat == "mangrove")
sum(old_extent$km2) # 82292.24 km2

## Our new extent is much larger than our old extent 


compare_habitat_extent <- gmw_ohi_2016 %>%
  filter(rgn_id <= 250) %>%
  left_join(rgns_eez) %>%
  left_join(old_extent, by = "rgn_id") %>%
  mutate(km2.y = ifelse(
    km2.x >0 & is.na(km2.y) ,0, #assign 0 values to old data km2 that have new data so that we can properly graph these differences.
    km2.y
  )) %>%
  mutate(difference = km2.x - km2.y)
sum(compare_habitat_extent$km2.x) - sum(compare_habitat_extent$km2.y) # 54397.7 km2 increase 

test <- compare_habitat_extent %>%
  filter(km2.x == 0, km2.y > 0) ## we lost 11 regions but gained 4... net loss of 7

ggplot(compare_habitat_extent, aes(x = km2.y, y = km2.x)) +
  geom_point() +
  geom_abline() +
  labs(title = "mangrove Habitat version old vs new", x = "old extent", y=  "new extent") +
  theme_bw()

filter(compare_habitat_extent, km2.y == 0 & km2.x > 0)


compare_habitat_extent <- gmw_ohi_2016 %>%
  left_join(gmw_ohi_2015, by = "rgn_id") %>%
  mutate(km2.y = ifelse(
    km2.x >0 & is.na(km2.y) ,0, #assign 0 values to old data km2 that have new data so that we can properly graph these differences.
    km2.y
  )) %>%
  mutate(difference = km2.x - km2.y)
sum(compare_habitat_extent$km2.x) - sum(compare_habitat_extent$km2.y) # -51.5218 km2 increase 
sum(compare_habitat_extent$difference)

ggplot(compare_habitat_extent, aes(x = km2.y, y = km2.x)) +
  geom_point() +
  geom_abline() +
  labs(title = "Global mangrove watch Habitat 2015 vs 2016", x = "old extent", y=  "new extent") +
  theme_bw()

filter(compare_habitat_extent, km2.y == 0 & km2.x > 0)


compare_habitat_extent <- gmw_ohi_2016 %>%
  left_join(gmw_ohi_1996, by = "rgn_id") %>%
  mutate(km2.y = ifelse(
    km2.x >0 & is.na(km2.y) ,0, #assign 0 values to old data km2 that have new data so that we can properly graph these differences.
    km2.y
  )) %>%
  mutate(difference = km2.x - km2.y)
sum(compare_habitat_extent$km2.x) - sum(compare_habitat_extent$km2.y) # -51.5218 km2 increase 
sum(compare_habitat_extent$difference)

ggplot(compare_habitat_extent, aes(x = km2.y, y = km2.x)) +
  geom_point() +
  geom_abline() +
  labs(title = "Global mangrove watch Habitat 1996 vs 2016", x = "old extent", y=  "new extent") +
  theme_bw()

filter(compare_habitat_extent, km2.y == 0 & km2.x > 0)
```


