---
title: "OHI 2021: Combine nutrient rasters"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---

## Summary
Grab all the relevant nutrient files and compile into a couple of summarized per year files, for manure and crops 

 - Total P applied per year for manure 
    - only leaching files
 - Total P applied per year for crops 
    - only leaching files
 - Total P for crops and manure 
   - only leaching files
   
 - Total N for manure 
    - volatized files together
    - Leaching files together
 - Total N for crops 
    - volatized files together
    - Leaching files together

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(raster)
library(readr)
library(tidyverse)
library(doParallel)

nutrient_d2021_anx <- file.path("/home/shares/ohi/git-annex/globalprep/_raw_data/FAOSTAT/manure_nutrient/d2021/")
git_manure <- file.path(here(), "globalprep/prs_land-based_nutrient/v2021/STEP2_manure")
location_tifs_path <- file.path("/home/shares/ohi/git-annex/globalprep/_raw_data/halpern_food_systems/d2021/animal_farm/location_tifs")
prep <- file.path("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2021")
git_crop <- file.path(here(), "globalprep/prs_land-based_nutrient/v2021/STEP1_crop")
git_compile <- file.path(here(), "globalprep/prs_land-based_nutrient/v2021/STEP3_compile_prep_layers")


source(here(git_crop, "R/common_spatial.R")) # read in necessary spatial files 

```


Combine and save the appropriate files so that we have yearly volatilized rasters and yearly leaching files per each nutrient.
 - make sure to run through this this with one nutrient and one year first... i.e. specify `nutrient <- "N"` and `yr <- 2005` and run line by line so you make sure to understand what is happening.
```{r}

years <- as.character(c(2005:2019))

# nutrients <- c("N", "P")
nutrients <- c("N")


for(yr in years){
  
  for(nutrient in nutrients){
    
 
  # yr <- "2005"
  # nutrient <- "N"

    if(nutrient == "N"){
      
    
  n_leach_manure <- list.files(file.path(prep, "animal_farm_int/STEP7_total_N_leaching"), pattern = yr, full = TRUE) # list leached manure files
  
  n_volt_manure <- list.files(file.path(prep, "animal_farm_int/STEP7_total_N_volatizing"), pattern = yr, full = TRUE) # list volatilized manure files
  
  crop_pat <- paste0(".+leached_N_", yr)
 
  leach_crop <- list.files(file.path(prep, "crop_predatalayers"), pattern = crop_pat, full = TRUE) # list leach crop files

  crop_pat_volt <- paste0(".+volatilized_N_", yr)
  volt_crop <- list.files(file.path(prep, "crop_predatalayers"), pattern = crop_pat_volt, full = TRUE) # list leach crop volatilized files

  
  ## Now save a file for just crops leached N
    all_nutrients_stack <- stack(leach_crop) 

    rasterOptions(todisk = FALSE)

 ## try with parallel processing 8 cores.. this will speed it up significantly. 10 cores is a lot though, so if you NOT are running it overnight, maybe decrease to 6-8 cores (don't wanna hog cores during the workday).

 raster::beginCluster(n = 8)

 parallel_sum <- raster::clusterR(all_nutrients_stack, fun = calc,
                            args = list(fun = sum, na.rm = TRUE))

endCluster()


dis_rast_area <- raster(file.path(prs_int_dir, "disagg_raster_area.tif"))


    names(parallel_sum) <- "tonnes_N"
    parallel_sum[is.na(parallel_sum)] <- 0


       png(filename=file.path(paste0(git_compile, "/plot_check/crops_leached/", "tonnes_N_", yr, ".png")))
        plot(log(1+ parallel_sum), main = paste0(yr))
         dev.off()

      print(cellStats(parallel_sum, "sum")) #


  raster::writeRaster(parallel_sum, file.path(prep,  "crop_stressors_yearly", paste0("crop_leached_nutrient_N_", yr, ".tif", sep = "")), format = "GTiff", overwrite = TRUE)



  ## Now save a file for just manure leached N
    all_nutrients_stack <- stack(n_leach_manure)

    rasterOptions(todisk = FALSE)

 ## try with parallel processing 8 cores.. this will speed it up significantly. 10 cores is a lot though, so if you NOT are running it overnight, maybe decrease to 6-8 cores (don't wanna hog cores during the workday).

 raster::beginCluster(n = 8)

 parallel_sum <- raster::clusterR(all_nutrients_stack, fun = calc,
                            args = list(fun = sum, na.rm = TRUE))

endCluster()


    names(parallel_sum) <- "tonnes_N"
    parallel_sum[is.na(parallel_sum)] <- 0


       png(filename=file.path(paste0(git_compile, "/plot_check/manure_leached/", "tonnes_N_", yr, ".png")))
        plot(log(1+ parallel_sum), main = paste0(yr))
         dev.off()

      print(cellStats(parallel_sum, "sum")) #


  raster::writeRaster(parallel_sum, file.path(prep,  "manure_stressors_yearly", paste0("manure_leached_nutrient_N_", yr, ".tif", sep = "")), format = "GTiff", overwrite = TRUE)
  
  
    ## Now save a file for manure volatized N
    all_nutrients_stack <- stack(c(n_volt_manure))
  
    rasterOptions(todisk = FALSE) 
 
 ## try with parallel processing 8 cores.. this will speed it up significantly. 10 cores is a lot though, so if you NOT are running it overnight, maybe decrease to 6-8 cores (don't wanna hog cores during the workday). 
  
 raster::beginCluster(n = 8)

 parallel_sum <- raster::clusterR(all_nutrients_stack, fun = calc,
                            args = list(fun = sum, na.rm = TRUE))

endCluster()
  

    names(parallel_sum) <- "tonnes_N"
    parallel_sum[is.na(parallel_sum)] <- 0
    
    
       png(filename=file.path(paste0(git_compile, "/plot_check/manure_volt/", "tonnes_N_", yr, ".png")))
        plot(log(1+ parallel_sum), main = paste0(yr))
         dev.off()
         
      print(cellStats(parallel_sum, "sum")) # 
    
  
  raster::writeRaster(parallel_sum, file.path(prep,  "manure_stressors_yearly", paste0("manure_volt_nutrient_N_", yr, ".tif", sep = "")), format = "GTiff", overwrite = TRUE)
  
  
      ## Now save a file for crops volatized N
    all_nutrients_stack <- stack(c(volt_crop))
  
    rasterOptions(todisk = FALSE) 
 
 ## try with parallel processing 8 cores.. this will speed it up significantly. 10 cores is a lot though, so if you NOT are running it overnight, maybe decrease to 6-8 cores (don't wanna hog cores during the workday). 
  
 raster::beginCluster(n = 8)

 parallel_sum <- raster::clusterR(all_nutrients_stack, fun = calc,
                            args = list(fun = sum, na.rm = TRUE))

endCluster()
  
  
 # stack_sum <- calc(all_nutrients_stack, sum, na.rm = TRUE)

    names(parallel_sum) <- "tonnes_N"
    parallel_sum[is.na(parallel_sum)] <- 0
    
    
       png(filename=file.path(paste0(git_compile, "/plot_check/crops_volt/", "tonnes_N_", yr, ".png")))
        plot(log(1+ parallel_sum), main = paste0(yr))
         dev.off()
         
      print(cellStats(parallel_sum, "sum")) # 
    
  
  raster::writeRaster(parallel_sum, file.path(prep,  "crop_stressors_yearly", paste0("crop_volt_nutrient_N_", yr, ".tif", sep = "")), format = "GTiff", overwrite = TRUE)
  
  
    }else{
      
   #   nutrient = "P"
  p_leach_manure <- list.files(file.path(prep, "animal_farm_int/STEP2_P_leach/"), pattern = yr, full = TRUE)

  fert_pat <- paste0(".+leached_P_", yr)
  
  p_leach_fertilizer_grassland <-  list.files(file.path(prep, "animal_farm_int/grassland_nutrient_pollution"), pattern = fert_pat, full = TRUE)
  
  crop_pat <- paste0(".+leached_P_", yr)
 
  leach_crop <- list.files(file.path(prep, "crop_predatalayers"), pattern = crop_pat, full = TRUE)
  
    ## Do only crop leaching P  
  all_nutrients_stack <- stack(leach_crop)
  

    rasterOptions(todisk = FALSE) 
 
 ## try with parallel processing 10 cores.. this will speed it up significantly. 10 cores is a lot though, so if you NOT are running it overnight, maybe decrease to 6-8 cores (don't wanna hog cores during the workday). 
  
 raster::beginCluster(n = 8)

 parallel_sum <- raster::clusterR(all_nutrients_stack, fun = calc,
                            args = list(fun = sum, na.rm = TRUE))

endCluster()
  
  
    names(parallel_sum) <- "tonnes_P"
    parallel_sum[is.na(parallel_sum)] <- 0
    
    
       png(filename=file.path(paste0(git_compile, "/plot_check/crops_leached/", "tonnes_P_", yr, ".png")))
        plot(log(1+ parallel_sum), main = paste0(yr))
         dev.off()
         
      print(cellStats(parallel_sum, "sum")) # 
    
  
  raster::writeRaster(parallel_sum, file.path(prep,  "crop_stressors_yearly", paste0("crop_leached_nutrient_P_", yr, ".tif", sep = "")), format = "GTiff", overwrite = TRUE)
  
  
  
      ## Do only manure leaching P  
  all_nutrients_stack <- stack(c(p_leach_manure, p_leach_fertilizer_grassland))
  

    rasterOptions(todisk = FALSE) 
 
 ## try with parallel processing 10 cores.. this will speed it up significantly. 10 cores is a lot though, so if you NOT are running it overnight, maybe decrease to 6-8 cores (don't wanna hog cores during the workday). 
  
 raster::beginCluster(n = 8)

 parallel_sum <- raster::clusterR(all_nutrients_stack, fun = calc,
                            args = list(fun = sum, na.rm = TRUE))

endCluster()
  
  
    names(parallel_sum) <- "tonnes_P"
    parallel_sum[is.na(parallel_sum)] <- 0
    
    
       png(filename=file.path(paste0(git_compile, "/plot_check/manure_leached/", "tonnes_P_", yr, ".png")))
        plot(log(1+ parallel_sum), main = paste0(yr))
         dev.off()
         
      print(cellStats(parallel_sum, "sum")) # 
    
  
  raster::writeRaster(parallel_sum, file.path(prep,  "manure_stressors_yearly", paste0("manure_leached_nutrient_P_", yr, ".tif", sep = "")), format = "GTiff", overwrite = TRUE)

    }
  
  }
}


```


Calculate total nutrient application per year for manure and crops and save to dataframes (for checking).

```{r}

## Read manure leaching and volt N nutrient layers and manure leaching P nutrient layers and create a dataframe columns year, nutrient, total nutrient sum


years <- as.character(c(2005:2019))
manure_total_nutrients_df <- data.frame(year = NA, nutrient = NA, nutrient_sum = NA)



for(yr in years){

  # yr <- 2005

  list_files <- list.files(file.path(prep, "animal_farm_int/N_from_excretion_manure"), full = TRUE, pattern = paste0(yr))

global_N_total <- sum(stack(lapply(list_files, raster)), na.rm = TRUE)
total_exc_N <- cellStats(global_N_total, stat = 'sum', na.rm = TRUE) # 190397192

N_df <- data.frame(year = yr, nutrient = "N", nutrient_sum = total_exc_N)

  list_files_P <- list.files(file.path(prep, "animal_farm_int/P_from_excretion_manure"), full = TRUE, pattern = paste0(yr))
global_P_total <- sum(stack(lapply(list_files_P, raster)), na.rm = TRUE)
total_exc_P <- cellStats(global_P_total, stat = 'sum', na.rm = TRUE) # 40600526


P_df <- data.frame(year = yr, nutrient = "P", nutrient_sum = total_exc_P)


manure_total_nutrients_df <- rbind(manure_total_nutrients_df, N_df, P_df)

}

manure_total_nutrients_df <- manure_total_nutrients_df %>%
  filter(!is.na(nutrient_sum)) %>%
  mutate(type = "manure")
write.csv(manure_total_nutrients_df, file.path(git_compile, "int/manure_nutrient_by_year.csv"), row.names = FALSE)



## read all crop application layers to estimate total N and total P applied

sum_df <- data.frame(year = NA, nutrient = NA, nutrient_sum = NA)

years <- as.character(c(2011:2019))

registerDoParallel(cores = 5)

foreach(yr = years) %dopar% {

# yr = 2011

N_pat <- paste0(".+application_N_", yr)
    
total_N <- 
  list.files(file.path(prep, "/crop_int", fsep = ""), pattern = N_pat, full.names = TRUE) %>% 
  .[!grepl(c("toba|teas|cof|othr"), .)] %>% 
  stack() %>% 
  stackApply(., 1, fun = "sum") %>% 
  cellStats(., "sum")

P_pat <- paste0(".+application_P2O5_", yr)

total_P <- 
  list.files(file.path(prep, "/crop_int", fsep = ""), pattern = P_pat, full.names = TRUE) %>% 
  .[!grepl(c("toba|teas|cof|othr"), .)] %>% 
  stack() %>% 
  stackApply(., 1, fun = "sum") %>% 
  cellStats(., "sum")

total_P <- total_P/2.29 

sum_df_P <- data.frame(year = yr, nutrient = "P", nutrient_sum = total_P)

sum_df_N <- data.frame(year = yr, nutrient = "N", nutrient_sum = total_N)

sum_df <- rbind(sum_df_P, sum_df_N, sum_df)

}

sum_df <- sum_df %>%
  dplyr::filter(!is.na(nutrient_sum))

write.csv(sum_df, file.path(git_compile, "int/crop_nutrient_by_year.csv"), row.names = FALSE)

```


