---
title: "OHI `r format(Sys.Date(), '%Y')` - FIS Revenue Preparation"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
eval: FALSE
---

# FIS Economies: USD Value of Marine (Commercial) Fishing per Country/Region per Year

## Overview

**FAO Capture Data (downloaded August 8, 2025)**

-   Data Source FAO Global Capture Production (in metric tonnes)

-   This version of the value database was downloaded from the
    Statistical Query Panel. Data from [FAO Global Capture
    Production](https://www.fao.org/fishery/en/collection/capture?lang=en)

-   Citation: FAO 2025. Global Capture Production. Fisheries and
    Aquaculture Division
    <https://www.fao.org/fishery/en/collection/capture?lang=en>

-   Source information: Navigate to the [online query
    portal](https://www.fao.org/fishery/statistics-query/en/capture/capture_quantity)
    for FAO Global Capture Production Quantity. Deselect all
    pre-selected years. Drag these fields into selected rows: ASFIS
    species name, FAO major fishing area name, ASFIS species ISSCAP
    group name En. ASFIS species Family scientific name, FAO major
    fishing areas, Inland/Marine areas Name en. Click on show data and
    confirm that data is present for 1950- two years prior to current
    year. Click download and select yes to include Null Values.

-   Date: August 11th, 2025

-   Time range: 1950-2023

-   Native data resolution: Country level

-   Format: csv

-   Description: Global Capture Production Quantity

**Ex-Vessel Price Data (downloaded August 24, 2023)**

-   Ex-vessel price data is in USD/metric tonne.

    -   ex-vessel-prices: ex-vessel prices for fishery caught species
        from 1976-2019

        -   exvessel_price_database_1976_2019.csv: ex-vessel price data
            gathered from [Melnychuk et al.
            2016](https://doi.org/10.1093/icesjms/fsw169) and updated to
            2019 using methods described in the public-facing [github
            repo](https://github.com/SFG-UCSB/price-db-sfg) associated
            with the Melnychuk et al. 2016 paper

    -   \*\*Citation for paper\*\*

    -   Melnychuk, M. C., Clavelle, T., Owashi, B., and
        Strauss, K. 2016. Reconstruction of global ex-vessel prices of
        fished species. - ICES Journal of Marine Science.
        <doi:10.1093/icesjms/fsw169>.

v2025: Migrated from "cf" to "fis" to better encompass all fishing rather than just commercial fishing.

### Setup

```{r setup}

knitr::opts_chunk$set(message = FALSE, warning = FALSE, eval=FALSE)

# load packages
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  here,
  janitor,
  foreach,
  doParallel, # for using multiple cores, if needed
  tidyverse, 
  plotly,
  zoo, # for gapfilling
  pdftools, # for FAO yearbook
  ohicore,
  modelr, # needed for gapfilling exvessel price data
  ggplot2,
  readxl
)

# ---- sources! ----
source(here("workflow/R/common.R")) # file creates objects to process data

source(here('workflow/R/fao_fxn.R')) # fxn for cleaning old version of FAO data 

source(here('workflow/R/fao_online_portal_clean.R')) # fxn for obtaining FAO data

# ---- set year and file path info ----
current_year <- 2025 # Update this in the future!!
version_year <- paste0("v",current_year)
data_dir_version_year <- paste0("d", current_year)
data_path <- here("globalprep", "eco", version_year)

# ---- data directories ----

# Raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")

# World Bank raw data directory
wb_dir <- here(raw_data_dir, "WorldBank", data_dir_version_year)

# ILO raw data directory
ilo_dir <- here(raw_data_dir, "ILOSTAT", data_dir_version_year)

# FAO raw data directory
fao_dir <- here(raw_data_dir, "FAO", data_dir_version_year)

# OECD raw data directory
oecd_dir <- here(raw_data_dir, "OECD", data_dir_version_year)

# FAO capture raw data directory
fao_capture_dir <- here(raw_data_dir, "FAO_capture", data_dir_version_year)

# Ex-vessel raw data directory
ex_vess_dir <- here(raw_data_dir, "ex-vessel-price-database-updated")

# output data dir for intermediate data products
int_dir <- here(data_path, "int")

# output data dir for final data products
output_dir <- here(data_path, "output")
```

# Methods

## Calculate estimated exvessel prices (based off Melnycuk's methods and emLab's code)

### Read in data

```{r}
# Linkage tables
link1<-read.csv(here("globalprep/eco/v2025/int/link1_Nov202025.csv"), stringsAsFactors = F)
link2<-read.csv(file.path(ex_vess_dir, 'price-db-linkage-tables/exvessel_tableS2.csv'), stringsAsFactors = F)
link3<-read.csv(file.path(ex_vess_dir, 'price-db-linkage-tables/exvessel_tableS3.csv'), stringsAsFactors = F)

# Updated commodity value data
cvalue <- read.csv(file.path(ex_vess_dir, 'price-db-data/Commodity_value_76to23.csv'), stringsAsFactors = F, na.strings = c('...','0 0','-')) %>% # value in 1000s
  clean_names() %>%
  filter(trade_flow_name == "Exports") %>%
  mutate(row_id = row_number())

# Updated commodity quantity data (v2025)
cquantity <- read.csv(file.path(ex_vess_dir, 'price-db-data/Commodity_quantity_76to23.csv'), stringsAsFactors = F,na.strings = c('...','0 0','-')) %>% # quantity in tons
  clean_names() %>%
  filter(trade_flow_name == "Exports") %>%
  mutate(row_id = row_number())

# FAO estimated ex-vessel data by ISSCAAP group
faoprices <- read_excel(file.path(ex_vess_dir, 'price-db-data/FAO exvessel estimates from archives.xlsx'), sheet = 4)[1:6]

## Set options for database
lowprocess <- FALSE # for using all commodities or just low processed commodities
commod_min <- 100 # minimum quantity required for inclusion in export price calculation
```

### Process FAO data

```{r}
process_fao <- function(data) {
  
  # Deal with year columns 
  yr_col_names <- names(data)[startsWith(names(data), "x_")]
  year_cols <- grep("x_", names(data), value = TRUE)
  years <- as.numeric(str_remove(year_cols, "x_"))
  initial_data_year <- min(years)
  last_data_year <- max(years)
  new_year_names <- paste0(initial_data_year:last_data_year)
  names(data)[names(data) %in% yr_col_names] <- new_year_names
  
  # Deal with flag columns
  s_cols <- names(data)[startsWith(names(data), "s")]
  new_names <- paste0(initial_data_year:last_data_year, " Flag")
  names(data)[names(data) %in% s_cols] <- new_names
  
  # Pivot all of the year/value columns 
  fao_values <- data %>% 
    # Remove flag columns
    dplyr::select(-c(contains("Flag"))) %>% 
    pivot_longer(cols = paste0(initial_data_year:last_data_year),
                 names_to = "year",
                 values_to = "value")
  
  # Pivot all of the flag columns   
  fao_flags <- data %>%
    select(-matches("^\\d{4}$")) %>% 
    pivot_longer(cols = paste(initial_data_year:last_data_year, "Flag"),
                 names_to = "flag_year",
                 values_to = "flag") %>% 
    mutate(year = str_remove(flag_year, " Flag")) %>% 
    select(year, flag, row_id) 
  
  # Combine flag and row id 
  fao_new <- fao_values %>%
    left_join(fao_flags, by = c("row_id", "year"))
  
  # Replace 
  fao_processed <- fao_new %>% 
    mutate(value = case_when((str_detect(flag, "N") & value == 0) ~ 0.1, #sub_N = 0.1
                             TRUE ~ value)) %>% #replace values that are 0 and have the flag N with sub_N
    select(-c(row_id, flag))
  
  
  return(fao_processed)
  
}

cvalue_processed <- process_fao(cvalue) %>%
  select(reporting_country_name, commodity_name, year, value_usd = value)

cquantity_processed <- process_fao(cquantity) %>%
  select(reporting_country_name, commodity_name, year, quantity_tonnes = value) 

```

### Process commodities

```{r}

# join value and quantity tables so they only include matched pairs
cboth <- inner_join(cvalue_processed, cquantity_processed) %>%
  # remove contested commodity "Miscellaneous crustaceans, not frozen, nei" due to apparent errors in data
  filter(commodity_name != 'Miscellaneous crustaceans, not frozen, nei') %>%
  filter(year >= 2005)

# standardize missing values to NAs
fix<-cboth$quantity_tonnes==0 | cboth$value_usd==0 | is.na(cboth$quantity_tonnes) | is.na(cboth$value_usd)

cboth$quantity_tonnes[fix]<-NA
cboth$value_usd[fix]<-NA

## Make table to fill in extra FAO commodity categories by hand
#commod_cats <- data.frame(FAO_commodity = sort(unique(cboth$commodity_name)))
#link_cats <- link1$FAO_commodity
#setdiff(commod_cats$FAO_commodity, link_cats$FAO_commodity)

#tmp <- full_join(commod_cats, link1, by = "FAO_commodity") 
#write_csv(tmp, here("globalprep/eco/v2025/int/link1_new.csv"))

# Calculate country export prices
cboth <- cboth %>%
  left_join(link1, by=c("commodity_name" = "FAO_commodity")) %>%
  mutate(tonnes = quantity_tonnes * exvessel_estimated) %>% ## this cuts out categories we aren't including, such as algae, an multiplies dried fish by 3.3
  mutate(export_price = 1000 * (value_usd) / tonnes) %>%
  filter(!is.na(export_price)) %>%
  filter(export_price != "Inf")

tmp <- cboth %>%
  group_by(reporting_country_name, year) %>%
  summarize(total_tonnes = sum(tonnes),
            total_usd = sum(value_usd))
filter(tmp, reporting_country_name == "United States of America")

# aggregate by ISSCAAP_group to link with fisheries data
isscaap_country_cost <- cboth %>%
  group_by(reporting_country_name, ISSCAAP_group, year) %>%
  summarize(tonnes = sum(tonnes, na.rm=TRUE),
            value_usd=sum(value_usd, na.rm=TRUE)) %>%
  ungroup() %>%
    mutate(export_price = 1000 * (value_usd) / tonnes) 

# aggregate across pooled commodities
isscaap_global_cost <- isscaap_country_cost %>%
  group_by(ISSCAAP_group, year) %>%
  summarize(pooled_quantity = sum(tonnes, na.rm=T),
            pooled_value    = sum(value_usd, na.rm=T)) %>%
  ungroup() %>%
  mutate(pooled_price       = pooled_value/pooled_quantity,
         year = as.numeric(year)) 

```


## Join with catch data

### Read in data

```{r}
# FAO Capture raw data, downloaded August 8, 2025.  Data goes from 1950 - 2023. 
fao_capture_raw <- read_csv(here(fao_capture_dir, "global_capture_production_quantity_1950-2023.csv"))

key <- read_csv(here("globalprep/eco/v2025/int/ASFIS_sp_2025_new.csv")) %>%
  select(ISSCAAP_Group, Taxonomic_Code, Scientific_Name, English_name, Family, `Order or higher taxa`) %>%
  filter(!is.na(English_name)) %>%
  unique()

dups <- key[duplicated(key$English_name), ]
setdiff(fao_capture_clean$asfis_species, key$English_name)


## add missing to the list to add by hand
missing <- data.frame(English_name = setdiff(fao_capture_clean$asfis_species, key$English_name))
find_missing <- full_join(missing, key) 
# saved as ASFIS_sp_2025_new.csv

iscaap_key <- read_csv(file.path(data_path, "int", "species-key.csv"), show_col_types = FALSE) %>%
  select(isscaap, isscaap_group) %>% unique()

```

### Process FAO capture data

```{r}
# v2025 updated function to clean FAO data
fao_capture_clean <- fao_online_portal_clean(fao_capture_raw) %>%
  # Filter to years in the price database
  filter(year >= 2005 & year <= 2023) %>% 
  rename(quantity_tonnes = value) %>%
  # Remove 0 tonnes
  filter(quantity_tonnes > 0) %>%
  # Remove "inland waters"
  filter(!str_detect(area, "Inland waters"))

fao_capture_isscaap <- fao_capture_clean %>%
  # Add ISSCAAP group to species for join with exvessel
  left_join(key, by = c('asfis_species' = 'English_name')) %>%
  left_join(iscaap_key, by=c("ISSCAAP_Group" = "isscaap")) %>%
  rename(ISSCAAP_group=isscaap_group, reporting_country_name=country)

setdiff(fao_capture_isscaap$reporting_country_name, isscaap_country_cost$reporting_country_name)
setdiff(fao_capture_isscaap$ISSCAAP_group, isscaap_country_cost$ISSCAAP_group)

#combine the price data with the capture data 
fao_capture_price <- fao_capture_isscaap %>% 
  left_join(isscaap_country_cost, by=c("reporting_country_name", "year", "ISSCAAP_group")) %>%
  select(-tonnes, -value_usd) %>%
  mutate(total_usd = quantity_tonnes*export_price)
filter(fao_capture_price, year==2023 & reporting_country_name == "United States of America")
  
tmp <- fao_capture_price %>%
  group_by(reporting_country_name, year) %>%
  summarize(total_usd = sum(total_usd, na.rm=TRUE))
tmp_us <- filter(fao_capture_price, year==2023 & reporting_country_name == "United States of America")
filter(tmp, reporting_country_name == "United States of America")
# v2025: 9.8% of observations still have NAS
summary(fao_capture_price)
sum(is.na(fao_capture_price$exvessel))/nrow(fao_capture_price)

#if still NA, fill with average price for isscaap species group for that year

#find the average price for each group/year in the exvessel data
group_average_price_year <- exvessel_prices_gapfilled %>%
  group_by(isscaap_group, year) %>% 
  summarise(mean_group_price = mean(exvessel))
 
# finish the join
fao_capture_price_final <- fao_capture_price %>% 
  left_join(group_average_price_year, by = c("isscaap_group", "year")) %>%
  mutate(final_price = ifelse(!is.na(exvessel), exvessel, mean_group_price)) %>%   
  mutate(gapfilled = ifelse((is.na(exvessel) & !is.na(final_price)), 1, gapfilled)) %>% 
  mutate(method = ifelse((is.na(exvessel) & !is.na(final_price)), "filled based on iscaap group average", method)) 

summary(fao_capture_price_final)

#there are still some NAS here, 2% 
nas <- fao_capture_price_final %>% 
  filter(is.na(final_price))

unique(nas$isscaap_group)
#after group discussion we decided to leave these as NA, due to lack of available data
#some of these species including turtles and corals are also considered unsustainable
# v2025:
# [1] "Corals"                            NA                                  "Miscellaneous aquatic plants"     
#  [4] "Blue-whales, fin-whales"           "Sperm-whales, pilot-whales"        "Brown seaweeds"                   
#  [7] "Red seaweeds"                      "Green seaweeds"                    "Sponges"                          
# [10] "Pearls, mother-of-pearl, shells"   "Turtles"                           "Eared seals, hair seals, walruses"

nrow(nas)/nrow(fao_capture_price_final)
# v2025: now 3.6%

fao_capture_price_final <- fao_capture_price_final %>% 
  select(country, year, asfis_species, quantity_tonnes, final_price, gapfilled, method) %>% 
  mutate(revenue = final_price * quantity_tonnes) # multiply price per tonne by number of tonnes

summary(fao_capture_price_final)

#sum prices by country for each year
fishing_revenue <- fao_capture_price_final %>% 
  group_by(country, year) %>% 
  summarize(value = sum(revenue, na.rm = TRUE)) %>% 
  mutate(unit = "USD (1)",
         sector = "fis",
         data_source = "FAO capture production and ex-vessel prices")

# save intermediate data!
write_csv(fishing_revenue, here(int_dir, "fis_revenue.csv"))
```

### Sanity check the revenue data

```{r}
# Compare to NOAA numbers
usa <- fishing_revenue %>%
  filter(country == "United States of America") %>%
  mutate(revenue_mil = value/1000000)

# Compare to state of the world fisheries numbers
global_revenue <- fishing_revenue %>%
  group_by(year) %>%
  summarise(revenue = sum(value),
            revenue_mil = revenue/1000000,
            revenue_bil = revenue/1000000000)
```

### Run `ohicore::name_2_rgn`

```{r name_2_rgn}
# check how many countries are available from the v2023 .csv before running names_2_rgn
length(unique(fishing_revenue$country)) 
# v2025: 203
# v2024: 203

# run names_2_rgn, which identifies country names from the cf_revenue and determines which OHI region it belongs to
fis_revenue_rgn <- ohicore::name_2_rgn(df_in = fishing_revenue, 
                             fld_name='country',
                             flds_unique = c("year"))

# to determine the correct OHI regions and compare the function's actions:
region_names <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/eez/spatial/regions_list.csv", show_col_types = FALSE) %>% select(-Notes)
```

When names_2_rgn has been run, there are some countries from fis_revenue
that are removed due to not having any match in the lookup tables. This
includes: bonaire/s.eustatius/saba, british indian ocean ter, channel
islands, french southern terr, isle of man, netherlands antilles, other
nei, saint barthélemy, saint helena/asc./trist., saint-martin (french),
yugoslavia sfr.

Some could be considered OHI regions, but they are aggregated within
another name. This includes: Bonaire/S.Eustatius/Saba, which can be
split into Bonaire, Sint Eustatius, and Saba; and the Channel Islands,
which should be split into Jersey and Guernsey.

Others can be dropped because they do not belong to an OHI region. This
includes yugoslavia sfr, saint barthélemy, other nei, isle of man
(self-governed), and french southern terr.

```{r fix_rgns}
# fix the country names so they can be recognized by names_2_rgn
fis_revenue_fix <- fishing_revenue %>%
  mutate(country = case_when(
    country=="British Indian Ocean Ter" ~ "British Indian Ocean Territory",
    country=="Saint Helena/Asc./Trist." ~ "Saint Helena", 
    country=="Saint-Martin (French)" ~ "Northern Saint-Martin",
    country=="Netherlands Antilles" ~ "Curacao", #Sint Maarten has already been covered in a separate country name, so this must be referring to Curacao
    country=="Puerto Rico" ~ "Puerto Rico and Virgin Islands of the United States", #to ensure the fxn aggregated correctly
    country=="United States Virgin Islands"~"Puerto Rico and Virgin Islands of the United States", #to ensure the fxn aggregated correctly
    country=="China, Hong Kong SAR" ~ "China", #to ensure the fxn aggregated correctly
    country=="China, Macao SAR" ~ "China", #to ensure the fxn aggregated correctly
    country=="United Republic of Tanzania, Zanzibar" ~ "Tanzania", #to ensure the fxn aggregated correctly
    country=="United Republic of Tanzania" ~ "Tanzania", #to ensure the fxn aggregated correctly
    TRUE ~ country # Everything else, leave it be
  ))
```

Re-run names_2_rgn after fixing names.

```{r fix_names}
# re-run names_2_rgn, check which countries are still dropped
fis_revenue_rgn_fix <- ohicore::name_2_rgn(df_in = fis_revenue_fix, 
                             fld_name='country',
                             flds_unique = c("year"))

# we are left with regions that are not associated with OHI, and also some that need to be disaggregated so that name_2_rgn can recognize them. Let's do that:
```

## Disaggregation

**Regions to disaggregate:**

Channel Islands ➞ Guernsey (228) and Jersey (227)

Bonaire/S.Eustatius/Saba ➞ Bonaire (245), Sint Eustatius (249), and Saba
(248)

To do this, look at the rasterized fishing effort calculated in 2017
within the FIS subgoal, to determine the relative proportion of
mean_catch within each region's eez. Afterwards, distribute revenue
relative to that proportion.

```{r proportions}
# load in data on the mean catch for each region, used as a proxy to determine which region has a greater commercial fishing economy
mean_catch_2017 <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohiprep_v2024/gh-pages/globalprep/fis/v2020/int/mean_catch.csv", show_col_types = FALSE)

# function to calculate average catch for certain regions and certain years
calculate_avg_catch <- function(data, region_ids, years) {
  data %>% 
    select(-stock_id_taxonkey) %>% 
    filter(rgn_id %in% region_ids, year %in% years) %>% # for the specified region id and year:
    group_by(rgn_id, year) %>%
    summarize(catch = sum(mean_catch, na.rm = TRUE)) %>% # summarize the mean catch for all species per rgn_id and year within the sequence of years
    ungroup() %>%
    group_by(rgn_id) %>% #for all of the years per region, find the average catch using the mean catch of each year
    summarize(avg_catch = mean(catch)) %>%
    select(rgn_id, avg_catch) %>% 
    dplyr::distinct() %>%  # ensures that there's only one row per region, removing any potential duplicates
    pivot_wider(names_from = "rgn_id", values_from = "avg_catch") # to call on specific observations more easily
}

# define parameters to input into the functions
years_to_analyze <- 2014:2017
channel_islands_ids <- c("228", "227")
bes_islands_ids <- c("245", "249", "248")

# calculate proportion of average catch for Channel Island regions for the 3 most recent years
mean_catch_ci_agg <- calculate_avg_catch(mean_catch_2017, channel_islands_ids, years_to_analyze) %>%
  rename(jersey = `227`, guernsey = `228`) %>% 
  pivot_longer(cols = c(jersey, guernsey),
             names_to = "rgn_id",
             values_to = "avg_catch") %>% 
  mutate(prop = avg_catch/sum(avg_catch, na.rm = TRUE))

# ccalculate proportion of average catch for Bonaire/S.Eustatius/Saba
mean_catch_bes_agg <- calculate_avg_catch(mean_catch_2017, bes_islands_ids, years_to_analyze) %>%
  rename(bonaire = `245`, sint_eustatius = `249`, saba = `248`) %>% 
  pivot_longer(cols = c(bonaire, sint_eustatius, saba),
             names_to = "rgn_id",
             values_to = "avg_catch") %>% 
  mutate(prop = avg_catch/sum(avg_catch, na.rm = TRUE))

# calculate proportions for each disaggregated region
## Channel Islands
prop_j <- mean_catch_ci_agg$prop[1]
prop_g <- mean_catch_ci_agg$prop[2]

## Bonaire/S. Eustatius/Saba
prop_b <- mean_catch_bes_agg$prop[1]
prop_e <- mean_catch_bes_agg$prop[2]
prop_s <- mean_catch_bes_agg$prop[3]
```

Now, use the proportions to disaggregate.

```{r disaggregation}
# use proportions to disaggregate Channel Islands 
split_ci <- fishing_revenue %>%
  filter(country == "Channel Islands") %>%
  mutate(split_col = "Guernsey;Jersey") %>%
  separate_rows(split_col, sep = ";") %>%
  group_by(country, year, value, unit) %>%
  mutate(new_value = case_when(
    split_col == "Guernsey" ~ value * prop_g,
    split_col == "Jersey" ~ value * prop_j,
    TRUE ~ NA_real_
  )) %>% # multiply each value by its proportion of avg catch in the area to split the value appropriately
  ungroup() %>%
  mutate(country = split_col) %>%
  select(-split_col)

# split for Bonaire/S.Eustatius/Saba
split_bes <- fishing_revenue %>%
  filter(country == "Bonaire/S.Eustatius/Saba") %>%
  mutate(split_col = "Bonaire;Sint Eustatius;Saba") %>%
  separate_rows(split_col, sep = ";") %>%
  group_by(country, year, value, unit) %>%
  mutate(new_value = case_when(
    split_col == "Bonaire" ~ value * prop_b,
    split_col == "Sint Eustatius" ~ value * prop_e,
    split_col == "Saba" ~ value * prop_s,
    TRUE ~ NA_real_
  )) %>% # multiply each value by its proportion of avg catch in the area to split the value appropriately
  ungroup() %>%
  mutate(country = split_col) %>%
  select(-split_col)

# merge with cf_revenue
fis_revenue_disaggregated <- fis_revenue_fix %>%
  bind_rows(split_ci, split_bes) %>% # bind with the disaggregated dataframes
  mutate(value = case_when(
    country %in% c("Guernsey", "Jersey") ~ new_value,
    country %in% c("Bonaire", "Sint Eustatius", "Saba") ~ new_value,
    TRUE ~ value
  )) %>% # if disaggregated, use the new value
  select(-new_value) # not needed anymore
```

Use a final names_2_rgn after disaggregation.

```{r final_names}
# using disaggregated revenue data within the name_2_rgn fxn: looks good!!
# seeing the aggregated versions of regions that we already disaggregated drop out. Great!
fis_revenue_rgn_final <- name_2_rgn(df_in = fis_revenue_disaggregated, 
                             fld_name='country',
                             flds_unique = c("year"))

#========== Indexing to check fxn ===============

# to see how many duplicates there are after names_2_rgn
rev_duplicates <- fis_revenue_rgn_final[duplicated(fis_revenue_rgn_final[, c("country", "year")]),]

# look at the unique duplicates for country and rgn_name to see if the function worked well
# setdiff to determine which observations are not the same in the country and rgn_id duplicates
rev_diff <- setdiff(unique(rev_duplicates$rgn_name), unique(rev_duplicates$country))
rev_diff # none! Move on

#=========== Aggregate by rgn_id and year ==============
fis_revenue_rgn_agg <- fis_revenue_rgn_final %>% 
  select(rgn_id, rgn_name, year, value, unit, sector, data_source) %>% 
  group_by(rgn_id, rgn_name, year) %>% # because china needs to be aggregated and has duplicate years
  dplyr::summarize(value = sum(value, na.rm=TRUE)) %>% # sum the value for each region and year
  ungroup() %>% 
  dplyr::mutate(year = as.numeric(year)) # to ensure it is numeric
```

## CF Revenue Plot

```{r rev_plot}
fis_rev_plot <- plotly::plot_ly(fis_revenue_rgn_agg, x = ~year, y = ~value, color = ~rgn_name, type = "scatter", mode = "lines") %>% 
  layout(title = "All Regions: Yearly Revenue for the Marine Fishing Sector", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Revenue of all catch (USD)"))
fis_rev_plot

# looks great! China has a very large revenue, which makes sense.  Indonesia and Peru are the have the next highest revenue in 2019.  Interesting to see a decrease in the early 2000s for most regions.
```

## Join with a "year" tibble df --in the future, replace this with `reframe()`!!

```{r year_join}
# make all geographic areas have the same date range (and fill missing values with NAs) to avoid problems downstream
years_df <- tibble(rgn_name = fis_revenue_rgn_agg$rgn_name) %>% 
  mutate(rgn_id = fis_revenue_rgn_agg$rgn_id) %>% 
  group_by(rgn_id, rgn_name) %>% 
  summarize(year = seq(min(fis_revenue_rgn_agg$year),
                       max(fis_revenue_rgn_agg$year)))

# ======== join with fis revenue aggregated data ========
fis_revenue_yrs <- left_join(years_df, fis_revenue_rgn_agg,
                          by = c("rgn_id", "rgn_name", "year"))

```

## Heatmap before Gapfilling

Typically, this would be used to visualize the data before and after
gapfilling, to get a better picture of what areas needed to be filled
and whether it was done correctly. However, here we are taking from
v2023's .csv that had already been gapfilled, so we are only going to
visualize how well the gapfilling worked in this case.

```{r interactive_heatmap}
# make a df where the "status" of the data (present or not) is defined.  Here, it would be 
heatmap_df <- fis_revenue_yrs %>% 
  mutate(status = !is.na(value))  #so that if there is data, the heatmap will be filled in

# make an interactive heatmap using ggplotly
heatmap_plot <- ggplot(data = heatmap_df) +
  geom_tile(aes(x = year,
                y = rgn_name,
                fill = status, 
                color = "black")) +
  scale_fill_manual(values = c("cornflowerblue", "gainsboro")) +
  labs(x = "Year",
       y = "Region Names",
       fill = "Data Status",
       color = " ") +
  theme_bw()
interactive_heatmap <- plotly::ggplotly(heatmap_plot)

interactive_heatmap
```

It looks as though Norfolk Island may need to be cut, and a couple of
other regions need to be gap filled now that we joined with a data frame
of all possible years. Let's use a rule that we defined as a team, so
that cutting regions is consistent across sectors.

## Cleaning for GF

Rule in order to be preserved:

-   must have at least one data point in the last five years &

-   must have more than three data points over time

Otherwise, the region needs to be removed.

All regions preserved can be gapfilled if needed.

```{r cut_rgns}
# ======= Determining regions that need to be cut based on the above rule ========
start_year <- 2008
stop_year <- 2023
date_range <- start_year:stop_year
last_five_years <- (stop_year - 4):stop_year

# ---- first sub-rule: must have at least one data pt in the last five years ----
one_data_pt <- heatmap_df %>%
  group_by(rgn_name) %>%
  summarize(
    has_recent_data = any(year %in% last_five_years & status == "TRUE"),
    total_data_points = sum(status == "TRUE"),
  ) %>%
  ungroup() %>% 
  filter(has_recent_data == "FALSE") %>%
  pull(rgn_name)

# ---- second sub-rule: must have more than three data points over time ----
three_pts <- heatmap_df %>%
  group_by(rgn_name) %>%
  summarize(
    has_recent_data = any(year %in% last_five_years & status == "TRUE"),
    total_data_points = sum(status == "TRUE"),
  ) %>%
  ungroup() %>% 
  filter(total_data_points < 3) %>%
  pull(rgn_name)

# see which and how many regions must be filtered out
one_data_pt 
# v2025: none
# v2024: Norfolk Island
length(one_data_pt) 
# v2024: 1
# v2025: 0

three_pts 
# v2025: 0
# v2024: NA
length(three_pts) 
# v2024: 0
# v2025: 0

# determine regions to keep
all_regions <- unique(heatmap_df$rgn_name) # a vector of all regions
regions_to_keep <- setdiff(all_regions, one_data_pt) # remove the regions that do not follow the rule

# filter the original dataframe to keep only the valid regions
filtered_heatmap_df <- heatmap_df %>%
  filter(rgn_name %in% regions_to_keep)
```

## Gapfilling

Here we'll use `na.approx()` from {zoo} to interpolate missing values
and extrapolate missing extremes (NA values that do not fall between
non-NA values) by copying the nearest extreme value.

Since we already filtered out any region that does not adhere to the
rule above, any regions left that have NAs can be gapfilled.

```{r}
# estimate value for regions with missing data
value_filled <- filtered_heatmap_df %>% 
  group_by(rgn_name) %>% 
  # interpolate (fill missing values between 2 values)
  mutate(appx_value = zoo::na.approx(value, # using values in this column
                                     na.rm = FALSE, # don't replace (internal) NAs in new column that can't be approximated
                                     #  extrapolate using rule = 2 from approx(),
                                     # which uses closest data extreme to
                                     #  extrapolate for leading and trailing NAs
                                     rule = 2
  )) %>% 
  ungroup()

# for the British Indian Ocean Territory in 2019, it copied the value from 2018.
# for Saba, anything before 2011 was filled with the 2011 value.  The exact same thing happened for Bonaire and Sint Eustatius, which makes sense because those three were disaggregated and therefore it makes sense that they would have the same data availability, as they came from the same source.
```

### Heatmap to check gapfilling

```{r heatmap_check}
heatmap_gf_df <- value_filled %>% 
  select(-c(value, status)) %>% 
  mutate(status = !is.na(appx_value))

heatmap_gf <- ggplot(data = heatmap_gf_df) +
  geom_tile(aes(x = year,
                y = rgn_name,
                fill = status, 
                color = "black")) +
  scale_fill_manual(values = c("cornflowerblue", "gainsboro")) +
  labs(x = "Year",
       y = "Region Names",
       fill = "Data Status",
       color = " ") +
  theme_bw()
interactive_heatmap_gf <- plotly::ggplotly(heatmap_gf)
#interactive_heatmap_gf

# YAY! it worked the way we wanted it to.  The heatmap is a solid color; there are only TRUEs, which indicates that there is a value present for every observation.
```

## Implementing Sector Multipliers

Refer to the OHI Methods
<https://ohi-science.org/ohi-methods/goals/goal-models-data.html#livelihoods-and-economies>
or [Ben Halpern's Supplementary
Information](https://static-content.springer.com/esm/art%3A10.1038%2Fnature11397/MediaObjects/41586_2012_BFnature11397_MOESM79_ESM.pdf)
**(page 29)** for the multipliers to apply towards revenue values by
sector. For commercial fishing revenue, the multiplier is 1.568 for
developed and developing countries.

```{r}
# As described in table 6.10: "Sector-specific multipliers are used to calculate total jobs and total revenue created by sector-based employment in developing and developed nations."

# -------- implement multiplier ------------
fis_multiplier <- heatmap_gf_df %>% 
  # Multiply all values with the multiplier to account for ripple effects commercial fishing may have on revenue
  mutate(mult_value = appx_value * 1.568) %>% 
  select(-appx_value)
```

## Saving final output

```{r}
# columns should be: rgn_id, rgn_name, year, usd (previously value), unit, sector, usd_yr (in this case = year)

fis_revenue <- fis_multiplier %>% 
  mutate(usd = mult_value,
         unit = "USD (1)",
         sector = "fis") %>%
  select(rgn_id, year, revenue = usd)

write_csv(fis_revenue, here(output_dir, "fis_revenue.csv"))
```

## Calculating proportion
