---
title: 'OHI 2022 - Land based nutrient plume exploration '
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../workflow/templates/ohi_hdr.html' 
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

[REFERENCE RMD FILE](http://ohi-science.org/ohiprep_v2020/globalprep/cw_nutrient_o_chem/v2020/plume_output_exploration.html)


# Summary
This document is used to explore the output of the plume modeling done for this data layer.  

## The following data are used:

* 



## Initial set-up code

```{r setup, eval = FALSE}
library(tidyverse)
library(raster)
library(sf)
library(mapview)
library(janitor)
library(here)


source(here('workflow/R/common.R'))

ww_raw_dir <- "/home/shares/ohi/git-annex/land-based/wastewater/data/raw"
ww_intermediate_dir <- "/home/shares/ohi/git-annex/land-based/wastewater/data/interim"
prs_int_dir <- "/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2022/int"
prep <- file.path("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2022")

```

**Take a look at the output from the full plume model**
 - We need to check this to see if the model ran on all of the basin_id
 - It looks like the model did not write rasters for everything. I think this is due to very small effluent values. 
 - Check the plumes that were run, and see if the ones with really small values are included. 
 
```{r, eval = FALSE}
test_plumes <- list.files("/home/sgclawson/grassdata/location/PERMANENT/cats", pattern = "plume_effluent")
test_plumes


## now read in total shp file

files_shp <- list.files(file.path("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2022/int/pourpoints/"), pattern = "shp", full = TRUE)

ws_pp_zonal_all_fix <- st_read(files_shp[1])   %>%
  mutate(row_num = row_number()) 

test_ws <- ws_pp_zonal_all_fix %>%
  filter(effluent !=0)
sum(test_ws$effluent) # 16198472


head(test_ws)

plumes_names <- unique(test_ws$basin_id)

plumes_names

test_plumes <- gsub("plume_effluent_", "", test_plumes)

missing_plumes <- setdiff(plumes_names, test_plumes) # missing 217 plumes 

missing_plumes


check_2 <- ws_pp_zonal_all_fix %>%
  filter(basin_id %in% missing_plumes)

sum(check_2$effluent) # 2382.453 the missing plumes only represent this many tonnes


test <- raster("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2022/output/N_plume/pourpoints_crop_manure_leached_volt_N_2005_joined.tif")
cellStats(test, "sum") # 16083871

16198472 - 16083871 # 114601 so we lose that much in tonnes... but much of that is probably excluded from the plume model anyways. 

plot(log(test+1))

N_effluent_log <- raster::calc(test, fun = function(x){log(x + 1)})
plot(N_effluent_log)

test <- raster::select(N_effluent_log)
plot(test)




```






