---
title: "OHI `r format(Sys.Date(), '%Y')` - Mariculture Economies Data Preparation"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

# Overview

## Mariculture Revenue Data

-   Data came from FishStatJ, FAO's application to obtain different
    fishery-related metrics by country as well as by sector
-   Citation: © FAO 2024. Global Aquaculture Production. In:
    Fisheries and Aquaculture. Rome. [Cited Tuesday, July 9th 2024].
    <https://www.fao.org/fishery/en/collection/aquaculture>

-   **Instructions for download from FishStatJ**
    -   Go to FAO website for download
        [FAO](https://www.fao.org/fishery/en/statistics/software/fishstatj)
    -   Also open the user manual found on that page, linked
        [here](https://www.fao.org/fishery/static/FishStatJ/FishStatJ_4.03.05-Manual.pdf)
    -   Once downloaded, open FishStatJ on your computer and follow the
        instructions to set it up.
    -   Then click file -\> manage workspaces -\> click 'FAO Global
        Fishery and Aquaculture Production Statistics' -\> click
        'Import' -\> Next -\> Next; until it opens
    -   This should import the workspace and allow you to access the
        'FAO Global Fishery and Aquaculture Production Value' data
    -   Once it opens, click 'File' -\> 'Export Selection (CSV File)'
    -   Store it somewhere you can find on your local drive and from
        there move it into the 'FAO_mariculture' folder under
        /home/shares/ohi/git-annex/globalprep/\_raw_data/FAO_mariculture/*your
        data year*

## Mariculture Jobs Data

-   Partially obtained from [FAO Fisheries and Aquaculture Statistical Yearbook](https://openknowledge.fao.org/server/api/core/bitstreams/2be6c2fa-07b1-429d-91c5-80d3d1af46a6/content)
-   Also brought in OECD data from their online [OECD Data Explorer](https://data-explorer.oecd.org/vis?df%5Bds%5D=DisseminateFinalDMZ&df%5Bid%5D=DSD_SOE%40DF_SOE&df%5Bag%5D=OECD.ENV.EPI&dq=.A....&pd=1995%2C2024&to%5BTIME_PERIOD%5D=false&vw=tb)

# Methods

## Setup

```{r render-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)

# options(scipen=99) # for number of digits printed
```

```{r library-setup}
# load packages ---- 
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  ohicore,
  here,
  janitor,
  terra,
  readxl,
  RColorBrewer,
  foreach,
  doParallel, # for using multiple cores
  priceR,
  pdftools,
  threadr,
  tidyverse, 
  httr,
  plotly,
  zoo # for gapfilling
  
)

# source common.R ----
source(here::here("workflow/R/common.R"))

# set year and file path info ----
current_year <- 2024 # Update this!!

version_year <- paste0("v",current_year) # Assessment year
data_dir_version_year <- paste0("d", current_year) # Data folder for assessment year
data_path <- here::here("globalprep", "le", version_year) # Path to assessment folder (v202X) 

# Raw data directory (on Mazu) ---- 
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")

# world bank raw data directory
wb_dir <- here::here(raw_data_dir, "WorldBank", data_dir_version_year)

# FAO Mariculture raw data directory 
fao_dir <- here::here(raw_data_dir, "FAO_mariculture", data_dir_version_year)

# output data directory for intermediate data products
int_dir <- here::here(data_path, "int")
```

## Read in Data

```{r read data}
# OHI regions data ----
# read in OHI regions for joining
region_names <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/eez/spatial/regions_list.csv") %>% 
  janitor::clean_names() %>% 
  select(-notes)

# FAO Aquaculture data ----
aquaculture_value <- read_csv(file.path(fao_dir, "/FAO_GlobalAquacultureProduction_Value_1984_2022.csv")) %>% 
  janitor::clean_names()

## str to view file structure (it can be very messy)
# str(aquaculture_value)
```

## Clean Aquaculture value data

**Note:** This process was done manually because instead of reading in the data with "flag" as the header for each observation's symbol, 2024's year was read in with "symbol" as the header for each observation's symbol

-   As a result, the data cleaning function fao_online_portal_clean.R
    was not applicable. We instead used components of the
    cleaning function and manually cleaned the data. See v2023
    livelihoods_economies_dataprep.Rmd to see what they did.
    -   With the data in the correct format (drawn from online FAO
        portal), this function should apply

```{r clean fao data}
#clean the aquaculture value data ----
#this will also be used in the mariculture jobs data section, so need to keep in all environments in the first step

# Cleaning aquaculture value data
aquaculture_value_clean <- aquaculture_value %>% 
   dplyr::rename(country = country_name, # Assigning relevant column names 
                species =  asfis_species_name, 
                area = fao_major_fishing_area_name,
                environment = environment_name) %>%
  dplyr::select(-c(unit_name)) %>% #clean data, note that we are still using a sub_n value of .1 (100 dollars) 
  mutate(row_id = 1:nrow(.))

# This section adjusts columns simultaneously so that you dont have to ---- 

# Check these! These are for the "year" columns
initial_data_year = 1984
last_data_year = 2022

# Testing to make sure the "year" specs follow the same naming format as the df
test = paste0("x", initial_data_year:last_data_year)

# Check these! These are for the "symbol" or "flag" columns (2024 they were in increments of 2)
flag_start = 8
flag_end = 84

# Testing to make sure the symbol or flag specs are the same naming format as the df 
test = paste0("symbol_", seq(flag_start, flag_end, by = 2))

# Pivoting columns to tidy the data ---- 

# This creates a dataframe of just the values in the FAO data 
fao_values <- aquaculture_value_clean %>% 
  dplyr::select(-c(paste0("symbol_", seq(flag_start, flag_end, by = 2)))) %>% # remove "flag" columns
  pivot_longer(cols = paste0("x", initial_data_year:last_data_year), 
               names_to = "year",
               values_to = "value") %>% # Pivot value colunns
  mutate(year = str_remove(year, "x")) # clean_names() put an 'x' in front of years. Removing the x

# This is a little bit more messy than the values because my column headers contained symbol instead of flag 
fao_flags1 <- aquaculture_value_clean %>% 
  dplyr::select(-c(paste0("x", initial_data_year:last_data_year))) # select out year

### MAYBE MOVE THIS TO APPLY TO THE AQUACULTURE_VALUE_CLEAN DF RATHER THAN FAO_FLAGS1

# Rename the columns in the Flag dataframe to actually have "flag" rather than symbol AND** getting the year for each flag so that we can rejoin the flags df with the values df 
colnames(fao_flags1) = c("country", "species", "area", "environment", "unit", paste0("flag", initial_data_year:last_data_year), "row_id")

# Because values were already selected out, this line of code is just pivoting the flags df and cleaning
fao_flags <- fao_flags1 %>% 
  pivot_longer(cols = paste0("flag", initial_data_year:last_data_year),
               names_to = "flag_year",
               values_to = "flag") %>% # Pivot longer 
  mutate(year = str_remove(flag_year, "flag")) %>% # Remove the "flag" from flag year so that we can have year to join by in the final flag df
  dplyr::select(year, row_id, flag) # selecting relevant columns to join by 

# Creating final df ---- 

# Joining fao values dataframe with the fao flags dataframe so that we can account for "N" flag and assign a value to it 
fao_new <- fao_values %>%
  left_join(fao_flags, by = c("row_id", "year")) # This is why we need year from each flag. If we just pivoted the values and then the flags in the aquaculture_values_clean df then it would double the values for each observation. 

# Assign value to substitute "N" flags for 
sub_N = 0.1

# Replace "N" flags 
fao_new <- fao_new %>% 
  mutate(value = case_when((str_detect(flag, "N") & value == 0) ~ sub_N,
                           TRUE ~value)) %>% #replace values that are 0 and have the flag N with sub_N
  select(-c(row_id, flag)) # Getting rid of temporary indexing row and flag row 


# Filter to only marine habitats
 mariculture_revenue <- fao_new %>% 
   filter(environment %in% c("Brackishwater", "Marine")) %>% 
   mutate(species = case_when(
     species == "Mexican spiny loster" ~ "Mexican spiny lobster",
     TRUE ~ species
   ))
 
# 
# mar_sp <- read.csv(here::here('globalprep/mar/v2023/raw/species_list.csv'), stringsAsFactors=FALSE) %>%
#   select(FAO_name, exclude,Taxon_code, family, notes, notes_2)
# 
# mar_sp_fix <- read.csv(here::here('globalprep/mar/v2022/raw/species_list.csv'), stringsAsFactors=FALSE)
# 
# new.spp <- setdiff(mariculture_revenue$species, mar_sp$FAO_name)
# 
# new.spp # I suppose there are 17 new species in the 2024 data
# 
# 
# 
# mariculture_revenue <- mariculture_revenue %>%
#   dplyr::rename(FAO_name = species)
# 
# mar_test <- mariculture_revenue %>% left_join(mar_sp, by="FAO_name")
# 
# mar_test <- mariculture_test %>% dplyr::filter(exclude < 1)
# 
# mar_test <- mar_test %>%
#   dplyr::mutate(include = 1 - exclude)
# 
# mar_test <- mar_test %>%
#   mutate(value_include = value*include)

# Sum by region and year (we need total revenue)
total_mar_rev <- mariculture_revenue %>%
  group_by(country, year) %>% 
  dplyr::summarize(value = sum(value, na.rm = TRUE)) %>% 
  mutate(sector = "mar", data_source = "FAO aquaculture value", unit = "USD (1)",
          value = value *1000) %>% # Multiply value times 1000 to get dollars
  select(country, year, value, unit, sector, data_source) # Select relevant columns

# Plotting to gut check ---- 
total_mar_rev %>% 
  # filter(!country == "China") %>% # If needed, filter out China because it tends to squish the lower values 
  ggplot(aes(x = as.numeric(year), y = value, color = country)) + 
  geom_line() + 
  theme(legend.position = "none") + 
  scale_y_continuous(labels = scales::label_dollar(scale = 0.000000001, suffix = "B"))
```

## Assigning country names to their respective regions in OHI

-   **NOTE**: This year used the names_2_rgn function to find out which
    country names needed to be hard-coded to align with the reference
    regions.
    - Run the names_2_rgn function on your data, and the output should tell you any countries that    were removed for not having a match in the look-up table 
    - Depending on which regions show up as not having a match, you may be required to hard-code the names to those of synonyms using case_when() (see chunk below for an example)
    - call 'rgn_synonyms <- ohicore::rgn_synonyms' in a code chunk to load the OHI Region synonyms for review 

```{r fix country names}
# Need to join with regions and assign each country its respective OHI Region ----

# Checking the unique country names for anything that needs to be hard-coded
unique(total_mar_rev$country)

# This is hard coding regions with to their synonyms so that the names_2_rgn function can assign them region IDs
mar_rev_rgn <- total_mar_rev %>% 
  mutate(Country = case_when(
    country=="Antigua and Barbuda" ~ "Antigua & Barbuda",
    country=="Bonaire, Sint Eustatius and Saba" ~ "Sint Eustasius",
    country=="Bosnia and Herzegovina" ~ "Bosnia Herzegovina",
    country=="China, Hong Kong SAR" ~ "China",
    country=="Democratic People's Republic of Korea" ~ "North Korea",
    country=="Republic of Korea" ~ "South Korea",
    country=="R<e9>union" ~ "Reunion",
    country=="Saint Kitts and Nevis" ~ "Saint Kitts &amp Nevis",
    country=="Saint Pierre and Miquelon" ~ "Saint Pierre and Miqelon",
    country=="Saint Vincent and the Grenadines" ~ "Saint Vincent/Grenadines",
    country=="Türkiye" ~ "Turkey",
    country=="Yugoslavia SFR" ~ "The Former Yugoslav Republic of Macedonia",
    country=="Netherlands Antilles" ~ "Curacao",
    TRUE ~ country
  ))

## Need to split the Channel Islands into Jersey and Guernsey 
m_CI <- mar_rev_rgn %>%
  filter(country == 'Channel Islands') %>%
  mutate(
    value            = value/2,
    'Guernsey'        = value,
    'Jersey'           = value) %>%
  ungroup() %>% 
  dplyr::select(-c(value, Country)) %>% 
  pivot_longer(cols = c(Guernsey, Jersey), values_to = "value", names_to = "Country") %>% 
  mutate(Country = as.character(Country),
         country = Country) %>% 
    dplyr::relocate(value, .before = unit)

mar_rev_rgn <- mar_rev_rgn %>%
  filter(country != 'Channel Islands') %>%
      bind_rows(m_CI)
  arrange(country, fao, environment, species, year, value) 

# Using name_2_rgn function to populate values for rgn_id and rgn_name for each of the countries listed in the dataframe
mar_rev_rgn_temp <- ohicore::name_2_rgn(df_in = mar_rev_rgn, 
                       fld_name='Country', # Check to make sure that the case matches
                      flds_unique = c("year"))


## Received a warning for a duplicated country. Using the following code to display the portion of the dataframe that's duplicated 
    ## Can see that in this case the duplicated country is China, with both a "China" and "China, Hong Kong SAR" both being considered china 

########## Left off here, I will need to do some aggregation on the duplicated rgn_id's, add the values together and then bring them back into the actual dataframe ###################
duplicates <- mar_rev_rgn_temp[duplicated(mar_rev_rgn_temp[, c('rgn_id','year')]),  ]

# 209, 140, 13, 186, 73, 202 --> Duplicate rgn_id's
# Reducing redundancy by creating a new df with the "China" regions so I can combine them (see line 254)
mar_rev_rgn_dupes <- mar_rev_rgn_temp %>% 
  filter(rgn_id %in% c(209, 140, 13, 186, 73, 202)) %>% 
  ungroup()

# Because I combined hong kong and china, I need to remove them from the original df so that I can rbind() my combined values to the original df 
mar_rev_rgn_no_dupes <- mar_rev_rgn_temp %>% 
  filter(!rgn_id %in% c(209, 140, 13, 186, 73, 202))

# The source code was used to aggregate duplicates by population for the "Social Progress Index", This would likely be useful for another metric of progress, but since we're dealing with hard values for revenue from aquaculture, my thought process is that we can just add them together, and it will represent the combined value from both Hong Kong as well as China

mar_rev_dupes_temp <- mar_rev_rgn_dupes %>% 
  mutate(year = as.character(year),
         rgn_id = as.character(rgn_id)) %>% 
  # group_by(rgn_id) %>%
  dplyr::summarize(value = sum(value), .by = c(year, rgn_id))

rgn_209 <- mar_rev_rgn_temp %>% 
  filter(rgn_id %in% c(209)) %>% 
  filter(country %in% "China")

joined_209 <- mar_rev_dupes_temp %>% 
  filter(rgn_id %in% 209) %>% 
  mutate(country = rgn_209$country,
         unit = rgn_209$unit,
         sector = rgn_209$sector,
         data_source = rgn_209$data_source,
         Country = rgn_209$Country,
         rgn_name = rgn_209$rgn_name,
         rgn_id = as.numeric(rgn_id)
         ) %>% 
  relocate(country, .before = year) %>% 
  relocate(rgn_id, .before = rgn_name)

rgn_140 <- mar_rev_rgn_temp %>% 
  filter(rgn_id %in% c(140)) %>% 
  filter(country %in% "Martinique")

joined_140 <- mar_rev_dupes_temp %>% 
  filter(rgn_id %in% 140) %>% 
  mutate(country = rgn_140$country,
         unit = rgn_140$unit,
         sector = rgn_140$sector,
         data_source = rgn_140$data_source,
         Country = rgn_140$Country,
         rgn_name = rgn_140$rgn_name,
         rgn_id = as.numeric(rgn_id)
         ) %>% 
  relocate(country, .before = year) %>% 
  relocate(rgn_id, .before = rgn_name)

rgn_13 <- mar_rev_rgn_temp %>% 
  filter(rgn_id %in% c(13)) %>% 
  filter(country %in% "Northern Mariana Islands")

joined_13 <- mar_rev_dupes_temp %>% 
  filter(rgn_id %in% 13) %>% 
  mutate(country = rgn_13$country,
         unit = rgn_13$unit,
         sector = rgn_13$sector,
         data_source = rgn_13$data_source,
         Country = rgn_13$Country,
         rgn_name = rgn_13$rgn_name,
         rgn_id = as.numeric(rgn_id)
         ) %>% 
  relocate(country, .before = year) %>% 
  relocate(rgn_id, .before = rgn_name)

rgn_186 <- mar_rev_rgn_temp %>% 
  filter(rgn_id %in% c(186)) %>% 
  filter(country %in% "Serbia and Montenegro")

joined_186 <- mar_rev_dupes_temp %>% 
  filter(rgn_id %in% 186) %>% 
  mutate(country = rgn_186$country,
         unit = rgn_186$unit,
         sector = rgn_186$sector,
         data_source = rgn_186$data_source,
         Country = rgn_186$Country,
         rgn_name = rgn_186$rgn_name,
         rgn_id = as.numeric(rgn_id)
         ) %>% 
  relocate(country, .before = year) %>% 
  relocate(rgn_id, .before = rgn_name)

rgn_73 <- mar_rev_rgn_temp %>% 
  filter(rgn_id %in% c(73)) %>% 
  filter(country %in% "Un. Sov. Soc. Rep.")

joined_73 <- mar_rev_dupes_temp %>% 
  filter(rgn_id %in% 73) %>% 
  mutate(country = rgn_73$country,
         unit = rgn_73$unit,
         sector = rgn_73$sector,
         data_source = rgn_73$data_source,
         Country = rgn_73$Country,
         rgn_name = rgn_73$rgn_name,
         rgn_id = as.numeric(rgn_id)
         ) %>% 
  relocate(country, .before = year) %>% 
  relocate(rgn_id, .before = rgn_name)

rgn_202 <- mar_rev_rgn_temp %>% 
  filter(rgn_id %in% c(202)) %>% 
  filter(country %in% "United Republic of Tanzania, Zanzibar")

joined_202 <- mar_rev_dupes_temp %>% 
  filter(rgn_id %in% 202) %>% 
  mutate(country = rgn_202$country,
         unit = rgn_202$unit,
         sector = rgn_202$sector,
         data_source = rgn_202$data_source,
         Country = rgn_202$Country,
         rgn_name = rgn_202$rgn_name,
         rgn_id = as.numeric(rgn_id)
         ) %>% 
  relocate(country, .before = year) %>% 
  relocate(rgn_id, .before = rgn_name)

# Now rejoining all of my un-duplicated data back into the df without duplicates 
mar_rev_fixed <- mar_rev_rgn_no_dupes %>% 
  rbind(., joined_13) %>% 
  rbind(., joined_202) %>% 
  rbind(., joined_209) %>% 
  rbind(., joined_140) %>% 
  rbind(., joined_186) %>% 
  rbind(., joined_73)

# Checking for duplicates one more times --> No observations = no duplicates 
duplicates <- mar_rev_fixed[duplicated(mar_rev_fixed[, c('rgn_id','year')]),  ]

# Binding duplicate data and ensuring dataframe is in the CORRECT format to write out to /int folder ----
eco_mar_usd_raw <- mar_rev_fixed %>% 
  relocate(c(rgn_id, rgn_name), .before = country) %>% # Relocating columns in the order specified by README
  ungroup() %>% # Getting rid of any leftover groups 
  select(-c(country, sector, data_source, Country)) %>% # Getting rid of extra columns
  dplyr::rename(usd = value) %>% # Renaming value to specified format name 
  mutate(sector = "mar", # Adding sector
         usd_yr = year, # UPDATE THIS BASED ON THE USD METRIC CONVERSION (should specify in metadata, but for the data for v2024 the data was just converted to that year's USD value, so the usd_yr is just year)
         year = as.numeric(year)) # Making sure its numeric per file structure. Could check this with class()

mar_heatmap <- eco_mar_usd_raw %>% 
  mutate(status = case_when(
    usd == 0 ~ FALSE,
    usd > 0 ~ TRUE
  )) %>% 
  filter(year %in% 2009:2019)

heatmap <- ggplot(mar_heatmap, aes(x = year, y = rgn_name, fill = as.factor(status))) + 
  geom_tile(color = "black") + 
  scale_fill_manual(values = c("darkgrey", "lightgreen")) + 
  theme_bw()
heatmap

# plotly::ggplotly(heatmap) # rgn_id 249, 244, 98, and 116 have missing data for almost all years, aside from that what's in there is pretty good and those countries likely are present from freshwater aquaculture data im assuming 

total_na <- mar_heatmap %>% 
  group_by(rgn_name) %>% 
  dplyr::summarize(sum = sum(status))

# Check here if you need to update the years in your dataframe. If there are  

## Checking how many unique countries are within the revenue data - 142 (2024 Assessment)
duplicates <- eco_mar_usd_raw[duplicated(eco_mar_usd_raw[, c('rgn_id','year')]),  ]


```

## Find out which countries can be gapfilled versus the countries that cant be

```{r gapfill identification}
# For 2024: The countries that can be gapfilled are: Yemen, Uruguay, Togo, Sint Eustatius, Seychelles (needs > three data points over time), Qatar, Puerto Rico, Poland, Libya, Jordan, Falkland Islands, Estonia, Dominica, Curacao, Cape Verde

eco_mar_usd_raw_cut <- eco_mar_usd_raw %>% 
  filter(!rgn_name %in% c("Yemen", "Uruguay", "Togo", "Sint Eustatius", "Seychelles", "Qatar", "Puerto Rico and Virgin Islands of the United States", "Poland", "Libya", "Jordan", "Falkland Islands", "Estonia", "Dominica", "Curacao", "Cape Verde"))

mar_heatmap <- eco_mar_usd_raw_cut %>% 
  mutate(status = case_when(
    usd == 0 ~ FALSE,
    usd > 0 ~ TRUE
  )) %>% 
  filter(year %in% 2009:2019)
heatmap <- ggplot(mar_heatmap, aes(x = year, y = rgn_name, fill = as.factor(status))) + 
  geom_tile(color = "black") + 
  scale_fill_manual(values = c("darkgrey", "lightgreen")) + 
  theme_bw()

heatmap
```


### Gapfilling

Here we'll use `na.approx()` from {zoo} to interpolate missing values and extrapolate missing extremes (NA values that do not fall between non-NA values) by copying the nearest extreme value. After speaking with Melanie Frazier about our methodology, she recommended to not fill in any values for countries with only 1 data point. If you choose to do this in future years, we've provided commented-out code following the chunk below for assessing how much of your data this impacts.

```{r gapfilling}
# gapfilling ----
# estimate tourism GDP proportion for geo areas with missing data
eco_mar_usd_gf <- eco_mar_usd_raw_cut %>% 
  mutate(usd = case_when(
    usd == 0 ~ NA,
    TRUE ~ usd
  )) %>% 
  group_by(rgn_name) %>% 
  # interpolate (fill missing values between 2 values)
  mutate(appx_usd = zoo::na.approx(usd, # using values in this column
                                       na.rm = FALSE, # don't replace (internal) NAs in new column that can't be approximated
                                       #  extrapolate using rule = 2 from approx(),
                                       # which uses closest data extreme to
                                       #  extrapolate for leading and trailing NAs
                                       rule = 2
  ))

mar_heatmap <- eco_mar_usd_gf %>% 
  mutate(status = case_when(
    appx_usd == 0 ~ FALSE,
    appx_usd > 0 ~ TRUE
  )) %>% 
  filter(year %in% 2009:2019)

heatmap <- ggplot(mar_heatmap, aes(x = year, y = rgn_name, fill = as.factor(status))) + 
  geom_tile(color = "black") + 
  scale_fill_manual(values = c("lightgreen")) + 
  theme_bw()

heatmap

# Doing some data prep so that we have both the raw gapfilled data as well as the one that we're writing out ----
eco_mar_usd_write <- eco_mar_usd_gf %>% 
  select(-usd) %>% 
  rename(usd = appx_usd) %>% 
  relocate(usd, .before = unit)

# Gut check ---- 
check <- is.na(eco_mar_usd_write$usd)
summary(check) # Cool no NAs
```

## Weighting by Developed vs Developing Country 

Refer to the OHI Methods or [Ben Halpern's Supplementary Information](https://static-content.springer.com/esm/art%3A10.1038%2Fnature11397/MediaObjects/41586_2012_BFnature11397_MOESM79_ESM.pdf) **(page 29)** for the multipliers to apply towards revenue values by sector for developing versus developed countries. For mariculture revenue, the multipliers are 1.59 for developing countries and 2.377 for developed countries.

```{r apply multiplier and finalize}
# Finally: Applying the multiplier to mariculture for developed versus developing countries ----
# Setting filepath to unsd data on developing versus developed countries 
fp_unsd <- file.path(dir_M,"git-annex/globalprep/_raw_data/UNSD/d2024/historical-classification-of-developed-and-developing-regions.xlsx")

# Reading in unsd data 
unsd_class <- read_excel(fp_unsd,col_names=TRUE, sheet = "Distinction as of May 2022") %>% 
  janitor::clean_names()

# Reading in OHI Regions data for gut checks/individual country checks 
ohi_regions <- readr::read_csv("https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/eez/spatial/regions_list.csv") %>% 
  janitor::clean_names() %>% 
  dplyr::select(-notes)

# Fixing a couple of the countries that are present under different synonyms in the unsd dataset OR Duplicated. Do this AFTER running name_2_rgn!!! 
unsd_class_fix <- unsd_class %>% 
  mutate(country_or_area = case_when(
  country_or_area == "Bonaire, Sint Eustatius and Saba" ~ "Sint Eustasius",
  country_or_area == "China, Hong Kong Special Administrative Region" ~ "China",
  country_or_area == "China, Macao Special Administrative Region" ~ "China",
  country_or_area == "Åland Islands" ~ "Finland",
  TRUE ~ country_or_area
  ))
  # dplyr::filter(!m49_code %in% c(630, 316, 474))

# Using name to rgn function to assign rgn_id to country names 
class_temp <- ohicore::name_2_rgn(df_in = unsd_class_fix, 
                       fld_name='country_or_area')

# Identifying duplicates  
duplicates <- class_temp[duplicated(class_temp$rgn_id), ]

# Removing countries with duplicated rgn_id by using their iso2 code 
class_temp <- class_temp %>% 
  dplyr::filter(!iso_alpha2_code %in% c("HK", "MO", "AX", "MP", "VI", "MQ")) %>% 
  dplyr::select(rgn_id, rgn_name, developed_developing_regions)

# Left joining the new developed/developing country classes to their respective rgn_id
eco_mar_usd_class <- eco_mar_usd_write %>% 
  left_join(y = class_temp, by = c("rgn_id"))

# 140, 116, 13 Were duplicated so hard removing those above with their m49_code (Puerto Rico and the US virgin Islands, Guam and Nothern Mariana Islands, Guadaloupe and Martinique)

# Removing unecessary columns from the joined df
eco_mar_usd_class <- eco_mar_usd_class %>% 
  dplyr::select(-c(rgn_name.y)) %>% 
  dplyr::rename(rgn_name = rgn_name.x)

# Applying weights to mariculture revenue data -----
eco_mar_usd_final <- eco_mar_usd_class %>% 
  mutate(usd_weighted = ifelse(developed_developing_regions == "Developing", usd*1.59, usd*2.377))

# Tidying up data for export to the 'int' folder 
eco_mar_usd_final <- eco_mar_usd_final %>% 
  mutate(usd = usd_weighted) %>% 
  select(-c(usd_weighted, developed_developing_regions))

#One last ggplot to gut check that the weighting did what we wanted it to do 

# ggplot(eco_mar_usd_final, aes(x = as.numeric(year), y = usd, color = rgn_name)) +
#   geom_line() +
#   theme(legend.position = "none") +
#   scale_y_continuous(labels = scales::label_dollar(scale = 0.000000001, suffix = "B"))


# Writing out csv FINALLY -----
# write_csv(eco_mar_usd_final, here::here(data_path, "int", "eco_mar_usd_pre.csv"))
```

