---
title: "OHI `r format(Sys.Date(), '%Y')` - CF Livelihoods Job Preparation"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

# CF Livelihoods: Proportion of Tourism Jobs per Country/Region per Year

## Overview

* **Data:** Labor Force & Employment Data

  * Labor Force data from World Bank (downloaded June 28. 2024)

      - https://data.worldbank.org/indicator/SL.TLF.TOTL.IN
      
  * OECD (Employment in fisheries, aquaculture and processing, 2009 - 2021) (downloaded July 2, 2024) -- for cf job data
  
    - <https://data-explorer.oecd.org/vis?fs[0]=Topic%2C1%7CAgriculture%20and%20fisheries%23AGR%23%7CFisheries%20and%20aquaculture%23AGR_FSA%23&pg=0&fc=Topic&bp=true&snb=6&vw=ov&df[ds]=dsDisseminateFinalDMZ&df[id]=DSD_FISH_EMP%40DF_FISH_EMPL&df[ag]=OECD.TAD.ARP&df[vs]=1.0&dq=.A...._T._T&pd=2009%2C&to[TIME_PERIOD]=false&ly[cl]=TIME_PERIOD&ly[rs]=REF_AREA&ly[rw]=DOMAIN>
    
  * FAO Yearbook (downloaded July 2, 2024) -- for cf job gapfilling if needed
  
    - <https://openknowledge.fao.org/server/api/core/bitstreams/2be6c2fa-07b1-429d-91c5-80d3d1af46a6/content>
      
  * ILOSTAT (downloaded July 2, 2024) -- for cf wage data
  
      - https://rshiny.ilo.org/dataexplorer46/?lang=en&id=EAR_4MTH_SEX_ECO_CUR_NB_A 
      - select Rev 3.1.B: Fishing
      
  * Ex-Vessel Price Data (downloaded August 24, 2023) -- for cf wage data
      - from emlab, see README.md


### Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, eval=FALSE)

# load packages
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  here,
  janitor,
  terra,
  readxl,
  RColorBrewer,
  foreach,
  doParallel, # for using multiple cores
  tidyverse, 
  httr,
  plotly,
  zoo, # for gapfilling
  pdftools
)
#remotes::install_github("skgrange/threadr") # for na_extrapolate
library(threadr)


# source 
source(here("workflow/R/common.R"))

source(here('workflow/R/fao_fxn.R')) # function for cleaning old version of FAO data 

source(here('workflow/R/fao_online_portal_clean.R'))

# set year and file path info
current_year <- 2024 # Update this!!

version_year <- paste0("v",current_year)
data_dir_version_year <- paste0("d", current_year)
data_path <- here("globalprep", "le", version_year)

# Raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")

# world bank raw data directory
wb_dir <- here(raw_data_dir, "WorldBank", data_dir_version_year)

# ILO raw data directory
ilo_dir <- here(raw_data_dir, "ILOSTAT", data_dir_version_year)

# FAO raw data directory
fao_dir <- here(raw_data_dir, "FAO", data_dir_version_year)

# OECD raw data directory
oecd_dir <- here(raw_data_dir, "OECD", data_dir_version_year)

# FAO capture raw data directory
fao_capture_dir <- here(raw_data_dir, "FAO_capture", data_dir_version_year)

# Ex-vessel raw data directory
ex_vess_dir <- here(raw_data_dir, "ex-vessel-price-database-updated")

# output data dir for intermediate data products
int_dir <- here(data_path, "int")
```


Create country code --> iso3 function
```{r}
library(countrycode)

country_regex_to_iso3c <- function(country_string) {
  country_string %>%
    countrycode::countrycode(origin = "country.name", destination = "iso3c", origin_regex = TRUE)
}
```

## Read in Data

```{r}
#----------OHI regions data--------------------
# read in OHI regions for joining
region_names <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/eez/spatial/regions_list.csv") %>% 
  janitor::clean_names() %>% 
  select(-notes)

#----------Labor Force Data (from earlier)-----
labor_force <- read_csv(here(data_path, "int", "labor_force.csv"))

############################################
# Livelihoods: Number of fisher jobs
############################################

#---------- cf OECD data---------------------------
oecd_cf_job_raw <- read_csv(here(oecd_dir, "OECD_2009-2021_employment_fish_aqua_processing_raw.csv"))

#------cf FAO Yearbook Number of Fishers-------
# turn the number of fishers table from the FAO yearbook into a data frame and clean.
# note: world total includes countries and regions not within the table.  Unsure of where that data comes from; look for methodology.
fao_pdf <- pdf_text(here(fao_dir, "fao_statistics_yearbook_07_03_24.pdf"))

fishers_table <- fao_pdf[[169]] %>% #page number of the pdf
  as.data.frame()

#specify columns 
columns <- c("country", "1995", "2000", "2005", "2010", "2015", "2016", "2017", "2018", "2019", "2020","2021")

#turn the pdf page into a clean dataframe
fao_fisher_jobs <- fishers_table %>%
  separate_rows('.', sep = "\n") %>% #separate rows based on line breaks
  rename(main_column = ".") %>%
  mutate(main_column = str_remove_all(main_column, "\\*")) %>% #remove the * 
  mutate(main_column = str_replace_all(main_column, "E(?!\\p{L})", " ")) %>% #replace all the Es not followed by letters (used to indicate estimate)
  slice(7:69) %>% #remove all rows not in the table
  separate(main_column, into = columns, sep = " {2,}", extra = "merge") %>% #separate into columns based on double spaces
  pivot_longer(cols = -c("country"), names_to = "year") %>% 
  mutate(value = str_squish(value)) %>% #remove extra spaces
  mutate(value = str_remove_all(value, " ")) %>% 
  mutate(value = if_else(value == "…" , NA, value)) %>% 
  mutate(value = as.numeric(value)) %>% 
  mutate(unit = "persons (1)",
         sector = "cf",
         data_source = "FAO number of fishers") %>% 
  slice(-c(1:44))

length(unique(fao_fisher_jobs$country)) # v2024: 59

# --------- run names_2_rgn to see how many OHI regions we have from this data -----
fao_fisher_jobs_rgn <- name_2_rgn(df_in = fao_fisher_jobs, 
                             fld_name='country',
                             flds_unique = c("year"))

## nothing was dropped out.  let's see how many OHI regions we have:
length(unique(fao_fisher_jobs_rgn$rgn_id)) # v2024: 50

```

# Livelihood: Number of Jobs -- cf

(put this within previous chunk, read in data etc. for number of tourism jobs, join with labor_regions, make new proportion of tourism jobs column, group by year, do prelim plotting etc. and evaluate data quality etc.)

### Cleaning and wrangling OECD data
```{r}
# total number of people employed by the fishing (previously commercial fishing, cf -- see README.md) sector

# no NAs, which means it may have been gapfilled already
unique(is.na(oecd_cf_job_raw$OBS_VALUE))

# ============== cf OECD number of fishers data ============== 
#note that this is data for the same category as the FAO fishers data, FAO data will be used to gapfill
oecd_fisher_jobs <- oecd_cf_job_raw %>% 
  janitor::clean_names() %>% 
  select(ref_area, reference_area, working_domain, time_period, obs_value, unit_of_measure) %>% 
  filter(working_domain %in% c("Marine fishing, nei","Marine deep-sea fishing", "Marine coastal fishing","Subsistence"))  %>% 
  mutate(year = as.numeric(time_period),
         value = obs_value,
         unit = ("persons (1)"),
         sector = "cf",
         data_source = "oecd sustainable economies") %>% 
  select(-c(unit_of_measure, obs_value, time_period, working_domain))

unique(is.na(oecd_fisher_jobs$value))
# FALSE -- no NAs

# ============== Run ohicore name_2_rgn ==================================

# check names
#china, taiwan province of or china (provice of china) should pick it up as taiwan
# "Chinese Taipei", "China (People’s Republic of)", "Korea"

oecd_fisher_jobs_fix <- oecd_fisher_jobs %>%
  mutate(reference_area = case_when(
    reference_area=="Chinese Taipei" ~ "Taiwan",
    reference_area=="China (People’s Republic of)" ~ "China",
    reference_area=="Korea" ~ "South Korea",
    TRUE ~ reference_area # Everything else, leave it be
  ))

oecd_fisher_jobs_rgn <- name_2_rgn(df_in = oecd_fisher_jobs_fix, 
                       fld_name='reference_area',
                      flds_unique = c("year"))

oecd_duplicates <- oecd_fisher_jobs_rgn[duplicated(oecd_fisher_jobs_rgn[, c("reference_area", "year")]),]

oecd_ref_duplicates <- unique(oecd_duplicates$reference_area)
oecd_rgn_duplicates <- unique(oecd_duplicates$rgn_name)

oecd_diff <- setdiff(oecd_rgn_duplicates, oecd_ref_duplicates)
oecd_diff # Vietnam is the only difference, and name_2_rgn works, so we are good to move forward! The raw data had duplicates in observation by year and region, so now we have to aggregate.
```

It should be noted that for the OECD number of fishers data, it has no NAs. Additionally, for the reference area of the Philippines, for 2009-2020 the observed value is consistently 120,000.  Therefore, the values have either been rounded or there has already been some sort of gapfilling. Within `oecd_cf_job_raw`, there is a column named "DECIMAL", which could allude to some sort of rounding.  Investigation was done to find OECD's metadata, however the link failed (below).  **WILL RETURN!**

Page with metadata: <https://www.oecd-ilibrary.org/economics/data/oecd-stat/data-warehouse_data-00900-en?parentId=http%3A%2F%2Finstance.metastore.ingenta.com%2Fcontent%2Fcollection%2Fdata-00285-en>
Failed link: <https://stats.oecd.org/wbos/default.aspx?checksum=40e0edc46aec7aee1cce210b898d5bd6&timestamp=1720045544&cid=id5845&oecdstat=data-00285-en&doi=data-00900-en&return_url=https%3A%2F%2Fwww.oecd-ilibrary.org%3A443%2Fcontent%2Fdata%2Fdata-00900-en&institution_name=UC+at+Santa+Barbara&lang=en>

```{r}
# preliminary plot
prelim_plot <- plotly::plot_ly(oecd_fisher_jobs_rgn, x = ~year, y = ~value, color = ~rgn_name, type = "scatter", mode = "lines") %>% 
  layout(title = "All Regions: Marine Fishing Employment", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Number of people by region employed by the fishing sector"))
prelim_plot
```

### Aggregating data
```{r}
# ============== Aggregating data ==================================
length(unique(oecd_fisher_jobs_rgn$reference_area)) #44
length(unique(oecd_fisher_jobs_rgn$rgn_name)) #44

duplicate_exists <- oecd_fisher_jobs_rgn %>% 
  group_by(rgn_name, year) %>% 
  filter(n() > 1)

bang_filt <- oecd_fisher_jobs_rgn %>% 
  filter(rgn_name %in% "Bangladesh")
head(bang_filt) # still have multiple entries per region and year from raw oecd data, need to aggregate

cf_jobs_rgn_agg <- oecd_fisher_jobs_rgn %>%
  group_by(rgn_id, rgn_name, year) %>%
  dplyr::summarize(value = sum(value, na.rm=TRUE)) %>%
  ungroup() %>% 
  dplyr::mutate(year = as.numeric(year)) 
```

```{r}
# plot of aggregated data
prelim_plot_agg <- plotly::plot_ly(cf_jobs_rgn_agg, x = ~year, y = ~value, color = ~rgn_name, type = "scatter", mode = "lines") %>% 
  layout(title = "All Regions: Marine Fishing Employment", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Number of people by region employed by the fishing sector"))
prelim_plot_agg

#plot without the large outliers Bangladesh and China
remove_values <- c("Bangladesh", "China")
cf_job_subset <- cf_jobs_rgn_agg %>% 
  filter(!(rgn_name %in% remove_values))

prelim_plot_lim <- plotly::plot_ly(cf_job_subset, x = ~year, y = ~value, color = ~rgn_name, type = "scatter", mode = "lines") %>%
  layout(title = "All Regions: Marine Fishing Employment", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Number of people by region employed by the fishing sector"))
prelim_plot_lim
```

### Create the proportion between tourism jobs and the total labor force
```{r}
# join labor force data with cf employment
cf_job_labor_join <- full_join(labor_force, cf_jobs_rgn_agg)

# make proportion of people employed in the commercial fishing sector
cf_job_prop <- cf_job_labor_join %>% 
  dplyr::relocate(value, .before = labor_force_total) %>%
  mutate(cf_prop_percent = (value/labor_force_total)*100) %>%
  select(-c(data_source, sector))
```

Preliminary plot of the percent of labor force working in the fishing sector (cf)
```{r}
cf_prop_plot <- plotly::plot_ly(cf_job_prop, x = ~year, y = ~cf_prop_percent, color = ~rgn_name, type = "scatter", mode = "lines") %>% 
  layout(title = "All Regions: Percent of Labor Force Employed by Marine Fishing Sector", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Percent by region employed by the fishing sector"))
cf_prop_plot
```

Saving intermediate data
```{r}
# %>% 
#   mutate(sector = "cf",
#          data_source = "ILOSTAT")

# write_csv(cf_job_prop_agg, here(int_dir, "cf_jobs.csv"))
```

### Gapfilling using FAO Yearbook
```{r}
# ============== Run ohicore name_2_rgn ==================================

# This has since been fixed by updating the synonyms within the ohicore repo! The code here will be left to provide an example of geo area names that may need to be fixed so that they can be associated with the appropriate OHI region.  However, in the future, the actual function should be updated with any new synonym found so that this case_when code never has to be run.

fao_fisher_jobs_fix <- fao_fisher_jobs %>%
  filter(country != "World total") %>% 
  mutate(country = case_when(
    country=="China,Taiwan" ~ "Taiwan",
    country=="Dem R Congo" ~ "Democratic Republic of the Congo",
    country=="DPR Korea" ~ "North Korea",
    country=="Faroe Is" ~ "Faeroe Islands",
    country=="Russian Fed" ~ "Russia",
    country=="USA" ~ "United States",
    country=="Rep.of Korea" ~ "South Korea",
    TRUE ~ country # Everything else, leave it be
  ))

# use names_2_rgn again after fixing the countries that were not recognized by the function
fao_fisher_jobs_rgn <- name_2_rgn(df_in = fao_fisher_jobs_fix, 
                       fld_name='country',
                      flds_unique = c("year"))

#determine which of the regions are duplicated by country and year
fao_duplicates <- fao_fisher_jobs_rgn[duplicated(fao_fisher_jobs_rgn[, c("country", "year")]),]
# none!

length(unique(fao_fisher_jobs_rgn$rgn_name)) # v2024: 57 regions available in this data

# see where there is overlapping between FAO and OECD data
fao_rgns <- unique(fao_fisher_jobs_rgn$rgn_name)
oecd_rgns <- unique(cf_jobs_rgn_agg$rgn_name) # use the data from before labor_force join

fao_oecd_diff <- setdiff(fao_rgns, oecd_rgns)
fao_oecd_diff # 27 regions could be filled! If the methodology makes sense.
```

Only years 2015-2021 could be gapfilled by the FAO Yearbook. This is still a good amount.
```{r}
# ============= Evaluating FAO Yearbook Methodology =========

unique(fao_fisher_jobs_rgn$year)
# years are weird: has 2010, then jumps to 2015 and increases sequentially to 2020 from there.
 # [1] "1995" "2000" "2005" "2010" "2015" "2016" "2017" "2018" "2019" "2020" "2021"

#make a list of the countries in the fao data but not oecd, these are the ones we want to add
missing_cf_oecd <- setdiff(fao_fisher_jobs_rgn$rgn_name, cf_jobs_rgn_agg$rgn_name)
missing_cf_oecd

fao_missing_oecd <- fao_fisher_jobs_rgn %>% 
  filter(rgn_name %in% c(missing_cf_oecd)) %>% 
  select(rgn_id, rgn_name, year, value)

fishing_jobs_final <- cf_jobs_rgn_agg %>% 
  rbind(fao_missing_oecd)

# evluate binded dataset containing oecd and fao data
length(unique(fishing_jobs_final$rgn_name)) #72!

# create proportion using the binded data
cf_job_final_join <- full_join(labor_force, fishing_jobs_final)

# make proportion of people employed in the commercial fishing sector
cf_job_prop <- cf_job_labor_join %>% 
  dplyr::relocate(value, .before = labor_force_total) %>%
  mutate(cf_prop_percent = (value/labor_force_total)*100) %>%
  select(-c(data_source, sector))
```



### Update on OECD and FAO email:

Mikecz, Orsolya (NFISS)
[Orsolya.Mikecz\@fao.org](mailto:Orsolya.Mikecz@fao.org){.email} Thu,
Jul 18, 2024 at 2:45 AM To: Sophia Lecuona
[sophiamireille\@ucsb.edu](mailto:sophiamireille@ucsb.edu){.email},
"[Fabiana.Cerasa\@oecd.org](mailto:Fabiana.Cerasa@oecd.org){.email}"
[Fabiana.Cerasa\@oecd.org](mailto:Fabiana.Cerasa@oecd.org){.email} Cc:
Melanie Frazier
[frazier\@nceas.ucsb.edu](mailto:frazier@nceas.ucsb.edu){.email}

Hello Sophia,

We thank you for your message and for your interest in our data. I am
attaching the data we have for 53 countries that are part of the joint
collection carried out by FAO and the OECD. These data are publicly
available through the OECD.

Regarding additional countries only collected by FAO, you can find some
information in the FAO Fisheries and Aquaculture Yearbooks
<https://www.fao.org/cwp-on-fishery-statistics/handbook/tools-and-resources/list-of-fao-yearbooks-of-fishery-statistics/en/>
and the recently published State of the World Fisheries and Aquaculture
2024
<https://www.fao.org/publications/home/fao-flagship-publications/the-state-of-world-fisheries-and-aquaculture/en>.
The dataset is not public and cannot be shared, but if you have a more
specific request, we may be able to share a subset of aggregated data.

We hope this helps and please do not hesitate to contact us if you have
any questions regarding the data shared in the attachment.

Best regards,

Orsolya and Fabiana

#### Read in FAO OECD joint collection data

Though this script is in regards to LIV::quality of jobs using wages, the data provided by the OECD and FAO joint collection is only employment data, aka number of jobs.  It will be evaluated in this script, since this script mentions the initial email to Orsolya and Fabiana.

```{r}
# read in joint data and see how many usable countries are available for marine fishing
jobs_joint_data_raw <- read_csv(here(oecd_fao_dir, "Fisheries_employment_jointFAOOECD_all_data.csv"))

# clean and subset for marine fishing only, will need to aggregate by year and region for all sexes.
# U = unspecified, by assumption
cf_jobs_oecd_fao <- joint_data_raw %>% 
  janitor::clean_names() %>% 
  select(-c(oc3, working_time, flag_est_country, flag_est_faooecd, flag, comment)) %>% 
  rename(type = oc2) %>% 
  # only have marine fishing for commercial fishing, since mariculture will cover aquaculture and processing is considered a different sector
  filter(type %in% "Marine fishing")

# aggregate by year and geographic_area
cf_jobs_oecd_fao_agg <- cf_jobs_oecd_fao %>% 
  dplyr::group_by(geographic_area, year) %>% 
  dplyr::summarize(value = sum(value, na.rm = TRUE))

length(unique(cf_jobs_oecd_fao_agg$geographic_area)) # v2024: 47

# --------- run names_2_rgn to see how many OHI regions we have from this data -----
cf_rgn_jobs_oecd_fao <- name_2_rgn(df_in = cf_jobs_oecd_fao_agg, 
                             fld_name='geographic_area',
                             flds_unique = c("year"))

## nothing was dropped out.  let's see how many OHI regions we have:
length(unique(cf_rgn_jobs_oecd_fao$rgn_id)) # v2024: 47

## it makes sense that there should be the same number of regions, since nothing was aggregated or disaggregated when name_2_rgn was run. Let's setdiff with the FAO yearbook data and see how many regions we can have overall.

# --------------- set diff with FAO yearbook job numbers ---------
length(setdiff(cf_rgn_jobs_oecd_fao$rgn_id, fao_fisher_jobs_rgn$rgn_id)) # v2024: 21 regions differ
```

The difference between the FAO yearbook data and the joint collection data is too small; the maximum number of regions available would be in the 70s, which is still too low to be usable (as determined earlier, after rbinding the FAO yearbook and OECD data).  Perhaps future years can find more comprehensive publicly available data.

