---
title: "OHI `r format(Sys.Date(), '%Y')` - CF LIV Quality Preparation"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

# CF Livelihoods: Quality of Tourism Jobs per Country/Region per Year

## Overview

**Data:**

-   ILOSTAT (downloaded July 2, 2024) -- for cf wage data

    ```         
    - <https://rshiny.ilo.org/dataexplorer46/?lang=en&id=EAR_4MTH_SEX_ECO_CUR_NB_A>
    - select Rev 3.1.B: Fishing, Sex: Total, all regions, all years
    - `EAR_4MTH_SEX_ECO_CUR_NB_A-filtered-2024-07-03-cf-wage.csv`
    ```

-   OECD (downloaded July 8, 2024) -- for ppp adj

    ```         
    - <https://data-explorer.oecd.org/vis?lc=en&tm=ppp&pg=0&snb=111&vw=ov&df[ds]=dsDisseminateFinalDMZ&df[id]=DSD_NAMAIN10%40DF_TABLE4&df[ag]=OECD.SDD.NAD&df[vs]=1.0&dq=A.AUS%2BAUT%2BBEL%2BCAN%2BCHL%2BCOL%2BCRI%2BCZE%2BDNK%2BEST%2BFIN%2BFRA%2BDEU%2BGRC%2BHUN%2BISL%2BIRL%2BISR%2BITA%2BJPN%2BKOR%2BLVA%2BLTU%2BLUX%2BMEX%2BNLD%2BNZL%2BNOR%2BPOL%2BPRT%2BSVK%2BSVN%2BESP%2BSWE%2BCHE%2BTUR%2BGBR%2BUSA...PPP_B1GQ.......&pd=2009%2C2020&to[TIME_PERIOD]=false>
    - `OECD_2009-2020_wage_quality_annual_ppp_raw.csv`
    ```

### Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, eval=FALSE)

# load packages
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  here,
  janitor,
  terra,
  readxl,
  RColorBrewer,
  foreach,
  doParallel, # for using multiple cores
  tidyverse, 
  httr,
  plotly,
  zoo, # for gapfilling
  pdftools
)
#remotes::install_github("skgrange/threadr") # for na_extrapolate
library(threadr)


# source 
source(here("workflow/R/common.R"))

source(here('workflow/R/fao_fxn.R')) # function for cleaning old version of FAO data 

source(here('workflow/R/fao_online_portal_clean.R'))

# set year and file path info
current_year <- 2024 # Update this!!

version_year <- paste0("v",current_year)
data_dir_version_year <- paste0("d", current_year)
data_path <- here("globalprep", "le", version_year)

# Raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")

# world bank raw data directory
wb_dir <- here(raw_data_dir, "WorldBank", data_dir_version_year)

# ILO raw data directory
ilo_dir <- here(raw_data_dir, "ILOSTAT", data_dir_version_year)

# FAO raw data directory
fao_dir <- here(raw_data_dir, "FAO", data_dir_version_year)

# OECD raw data directory
oecd_dir <- here(raw_data_dir, "OECD", data_dir_version_year)

# FAO capture raw data directory
fao_capture_dir <- here(raw_data_dir, "FAO_capture", data_dir_version_year)

# Ex-vessel raw data directory
ex_vess_dir <- here(raw_data_dir, "ex-vessel-price-database-updated")

# OECD and FAO joint collection raw data directory
oecd_fao_dir <- here(raw_data_dir, "OECD_FAO_joint_collection", data_dir_version_year)

# output data dir for intermediate data products
int_dir <- here(data_path, "int")
```

### Read in Data

```{r}
#-------cf ILO wage data (2009 - 2021) -------
# this data has limited entries, so local currency will need to be ppp adjusted to increase the number of observations per geo area
cf_ilo_wage_raw <- read_csv(here(ilo_dir, "EAR_4MTH_SEX_ECO_CUR_NB_A-filtered-2024-07-03-cf-wage.csv"))

cf_wage_ppp_raw <- read_csv(here(oecd_dir, "OECD_2009-2020_wage_quality_annual_ppp_raw.csv"))
```

# Livelihood: Quality of Jobs -- cf

-   note: this data came from many sources and was compiled by ILO. PPP
    GDP adjustments come from OECD

Cleaning and wrangling data

```{r}
# separating into the different currencies
cf_ilo_wage <- cf_ilo_wage_raw %>% 
  janitor::clean_names() %>% 
  rename(currency = classif2_label, 
         year = time) %>% 
  select(ref_area_label, year, currency,obs_value, note_indicator_label, source_label)

length(unique(cf_ilo_wage$ref_area_label)) #32

unique(cf_ilo_wage$currency)
# "Currency: Local currency" "Currency: 2017 PPP $"     "Currency: U.S. dollars" 

# Local currency -- needs to be PPP adjusted
cf_wage_local <- cf_ilo_wage %>% 
  filter(currency %in% "Currency: Local currency")

# Currency: U.S. dollars -- needs to be PPP adjusted
cf_wage_usd <- cf_ilo_wage %>% 
  filter(currency %in% "Currency: U.S. dollars")

# 2017 PPP $
cf_wage_2017ppp <- cf_ilo_wage %>% 
  filter(currency %in% "Currency: 2017 PPP $")

############# Bring in OECD data to adjust monthly earnings to 2017 PPP
```

## Next Steps

-   ILOSTAT data does not have enough regions/countries. There are
    currently only 32 reported geo areas. To combat this, Fabiana Cerasa
    (OECD), and Orsolya Mikecz (FAO) were contacted via email to obtain
    more comprehensive data.
    -   We came across their June 2023 [powerpoint
        presentation](%22https://www.fao.org/fishery/services/storage/fs/fishery/documents/CWP/presentations/2023/CWP_IS_2023_Pr.6.2.pdf%22)
        on [Fisheries and aquaculture employment
        statistics](%22https://www.fao.org/fishery/en/collection/global_fishers%22),
        a discussion on different topics related to the current joint
        collection of employment statistics by OECD and FAO.
-   OECD data was brought in to account for the fact that the ILOSTAT
    data has three different currency types that need to be
    standardized. This would not be needed if Fabiana Cerasa and Orsolya
    Mikecz fail to respond.

### Update on OECD and FAO email:

Mikecz, Orsolya (NFISS)
[Orsolya.Mikecz\@fao.org](mailto:Orsolya.Mikecz@fao.org){.email} Thu,
Jul 18, 2024 at 2:45 AM To: Sophia Lecuona
[sophiamireille\@ucsb.edu](mailto:sophiamireille@ucsb.edu){.email},
"[Fabiana.Cerasa\@oecd.org](mailto:Fabiana.Cerasa@oecd.org){.email}"
[Fabiana.Cerasa\@oecd.org](mailto:Fabiana.Cerasa@oecd.org){.email} Cc:
Melanie Frazier
[frazier\@nceas.ucsb.edu](mailto:frazier@nceas.ucsb.edu){.email}

Hello Sophia,

We thank you for your message and for your interest in our data. I am
attaching the data we have for 53 countries that are part of the joint
collection carried out by FAO and the OECD. These data are publicly
available through the OECD.

Regarding additional countries only collected by FAO, you can find some
information in the FAO Fisheries and Aquaculture Yearbooks
<https://www.fao.org/cwp-on-fishery-statistics/handbook/tools-and-resources/list-of-fao-yearbooks-of-fishery-statistics/en/>
and the recently published State of the World Fisheries and Aquaculture
2024
<https://www.fao.org/publications/home/fao-flagship-publications/the-state-of-world-fisheries-and-aquaculture/en>.
The dataset is not public and cannot be shared, but if you have a more
specific request, we may be able to share a subset of aggregated data.

We hope this helps and please do not hesitate to contact us if you have
any questions regarding the data shared in the attachment.

Best regards,

Orsolya and Fabiana

#### Read in FAO OECD joint collection data

Though this script is in regards to LIV::quality of jobs using wages, the data provided by the OECD and FAO joint collection is only employment data, aka number of jobs.  It will be evaluated in this script, since this script mentions the initial email to Orsolya and Fabiana.

```{r}
# read in joint data and see how many usable countries are available for marine fishing
jobs_joint_data_raw <- read_csv(here(oecd_fao_dir, "Fisheries_employment_jointFAOOECD_all_data.csv"))

# clean and subset for marine fishing only, will need to aggregate by year and region for all sexes.
# U = unspecified, by assumption
cf_jobs_oecd_fao <- joint_data_raw %>% 
  janitor::clean_names() %>% 
  select(-c(oc3, working_time, flag_est_country, flag_est_faooecd, flag, comment)) %>% 
  rename(type = oc2) %>% 
  # only have marine fishing for commercial fishing, since mariculture will cover aquaculture and processing is considered a different sector
  filter(type %in% "Marine fishing")

# aggregate by year and geographic_area
cf_jobs_oecd_fao_agg <- cf_jobs_oecd_fao %>% 
  dplyr::group_by(geographic_area, year) %>% 
  dplyr::summarize(value = sum(value, na.rm = TRUE))

length(unique(cf_jobs_oecd_fao_agg$geographic_area)) # v2024: 47
```

It appears that we only have 47 geo areas. This will be taken into `liv_cf_jobs_prep.Rmd` so that it can be worked with.
