---
title: "OHI `r format(Sys.Date(), '%Y')` - Mariculture Livelihoods & Economies Data Preparation"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

### As of July 2024, we are currently shelving Livelihoods portion of Livelihoods and Economies because it has proven difficult to find data on jobs for certain sectors (cough cough tourism) as well as find their wages to infer some reference point of quality compared to quantity. So This is as far as we got in the current dataprep for Mariculture Jobs before pivoting to focus only on updating economies.

# Livelihoods: Proportion of Tourism Jobs per Country/Region per Year

### Mariculture Jobs Data

-   Partially obtained from [FAO Fisheries and Aquaculture Statistical Yearbook](https://openknowledge.fao.org/server/api/core/bitstreams/2be6c2fa-07b1-429d-91c5-80d3d1af46a6/content)
-   Also brought in OECD data from their online [OECD Data Explorer](https://data-explorer.oecd.org/vis?df%5Bds%5D=DisseminateFinalDMZ&df%5Bid%5D=DSD_SOE%40DF_SOE&df%5Bag%5D=OECD.ENV.EPI&dq=.A....&pd=1995%2C2024&to%5BTIME_PERIOD%5D=false&vw=tb)


### Setup

```{r}
# load packages ---- 
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  ohicore,
  here,
  janitor,
  terra,
  readxl,
  RColorBrewer,
  foreach,
  doParallel, # for using multiple cores
  priceR,
  pdftools,
  threadr,
  tidyverse, 
  httr,
  plotly,
  zoo # for gapfilling
  
)

# source common.R ----
source(here::here("workflow/R/common.R"))

# set year and file path info ----
current_year <- 2024 # Update this!!

version_year <- paste0("v",current_year) # Assessment year
data_dir_version_year <- paste0("d", current_year) # Data folder for assessment year
data_path <- here::here("globalprep", "le", version_year) # Path to assessment folder (v202X) 

# Raw data directory (on Mazu) ---- 
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")

# world bank raw data directory
wb_dir <- here::here(raw_data_dir, "WorldBank", data_dir_version_year)

# FAO Mariculture raw data directory 
fao_dir <- here::here(raw_data_dir, "FAO_mariculture", data_dir_version_year)

# output data directory for intermediate data products
int_dir <- here::here(data_path, "int")
```

## This was copied from Sophia's pdf processing she did for the fao yearbook

```{r}
#------cf FAO Yearbook Number of Fishers-------
# turn the number of fishers table from the FAO yearbook into a data frame and clean.
# note: world total includes countries and regions not within the table.  Unsure of where that data comes from; look for methodology.

# Need To update the fao_dir from above with the pdf of the FAO Fisheries and Aquaculture Statistics Yearbook
fao_dir <- here::here(raw_data_dir, "FAO", data_dir_version_year)

## Loading the pdf in as an object 
fao_pdf <- pdf_text(here::here(fao_dir, "fao_statistics_yearbook_07_03_24.pdf"))

## Selecting pdf table 
fishers_table <- fao_pdf[[170]] %>% # page number of the pdf that the desired table is on 
  as.data.frame()

## Specify columns 
columns <- c("country", "1995", "2000", "2005", "2010", "2015", "2017", "2018", "2019", "2020","2021")

## Setting the columns name because mutate wasn't picking up on the "." as a column name.
colnames(fishers_table) = c("main_column")
## Checking 
# colnames(fishers_table)


## Turn the pdf page into a clean dataframe
fao_mar_jobs <- fishers_table %>%
  separate_rows(main_column, sep = "\n") %>%  #separate rows based on line breaks
  # rename(main_column = '.') %>%  ## Removed because of above mention of column name 
  mutate(main_column = str_remove_all(main_column, "\\*")) %>% # Remove the * 
  mutate(main_column = str_replace_all(main_column, "E(?!\\p{L})", " ")) %>% # Replace all the Es not followed by letters (used to indicate estimate)
  slice(7:69) %>% #remove all rows not in the table
  separate(main_column, into = columns, sep = " {2,}", extra = "merge") %>% #separate into columns based on double spaces
  pivot_longer(cols = -c("country"), names_to = "year") %>% 
  mutate(value = str_squish(value)) %>% #remove extra spaces
  mutate(value = str_remove_all(value, " ")) %>% 
  mutate(value = if_else(value == "…" , NA, value)) %>% 
  mutate(value = as.numeric(value)) %>% 
  mutate(unit = "persons (1)",
         sector = "mar/cf",
         data_source = "FAO number of aquaculture jobs") %>% 
  slice(-c(1:44))

## Final cleaning. Removing extraneous rows not necessary 
fao_mar_jobs_clean <- fao_mar_jobs %>% 
  dplyr::slice(-c(1:6, 327:516)) %>% # %>% view() # Add view() to check df to make sure its tidy after 
  mutate(year = as.numeric(year))

########### OECD data on aquaculture employment ###########

oecd.fp <- file.path(dir_M, "git-annex/globalprep/_raw_data/OECD", data_dir_version_year, "OECD.ENV.EPI,DSD_SOE@DF_SOE,+.A.FISH_EMP_AQUACULTURE_TOT....csv")

#### BRING THIS JAWN IN #####

oecd_mar_jobs <- read.csv(oecd.fp) %>% janitor::clean_names()

length(unique(oecd_mar_jobs$reference_area))

oecd_mar_jobs_clean <- oecd_mar_jobs %>% 
  ## Getting rid of all unnecessary columns 
  select(-c(obs_status_2, observation_status_2, price_base, price_base_2, ref_year_price, price_reference_year, observation_value, time_period_2, iucn_management_categories, ret_period, freq, measure, structure, structure_id, action, measure_2, unit_measure, designation, return_periods, decimals, decimals_2, obs_status, observation_status, unit_mult, unit_multiplier)) %>% 
  ## Moving columns to more intuitive order 
  relocate(c(ref_area, reference_area, time_period, obs_value), .before = structure_name) %>% 
  ## Multiplying obs_value by 1000 to obtain raw value of jobs per country 
  mutate(value = obs_value*1000) %>% 
  ## One more relocate
  relocate(value, .after = obs_value) %>% 
  ## Renaming structure_name to source
  dplyr::rename(data_source = structure_name) %>% 
  ## Removing redundant/unneeded columns 
  select(-c(obs_value, frequency_of_observation)) %>% 
  ## Renaming columns to join with other jobs data from FAO
  dplyr::rename(year = time_period,
                country = reference_area) %>% 
  ## Relocating one last time 
  relocate(unit_of_measure, .before=data_source)

oecd_mar_jobs_join <- oecd_mar_jobs_clean %>% 
  select(-c(unit_of_measure, ref_area)) %>% 
  mutate(unit = "persons (1)",
         sector = "mar") %>% 
  relocate(c(unit, sector), .before=data_source)

## With both datasets our total unique regions comes out to 82
length(unique(fao_mar_jobs_clean$country)) + length(unique(oecd_mar_jobs_join$country))

mar_jobs_tot <- rbind(fao_mar_jobs_final, oecd_mar_jobs_join)

## Joined the unique country names comes out to 65
length(unique(mar_jobs_tot$country))

duplicates <- as.data.frame(mar_jobs_tot[duplicated(mar_jobs_tot[, c('country', 'year')]),  ])

length(unique(duplicates$country))

```

## This piece of code is for assigning countries to region id when there are dis-aggregated country names (i.e China, Hong Kong, Macao) that are considered one country for the OHI Regions


```{r}
## Hard coding a few of the country names that are either synonyms of countries or difficult to identify with name_2_rgn() function
mar_jobs_rgn <- fao_mar_jobs_clean %>%
  filter(!country %in% "World total") %>% 
  dplyr::rename(Country = country) %>% 
  mutate(Country = case_when(
   Country=="China,Taiwan" ~ "China",
    Country=="Russian Fed" ~ "Russia",
    Country=="Rep.of Korea" ~ "South Korea",
    Country=="DPR Korea" ~ "North Korea",
    Country=="Türkiye" ~ "Turkey",
   Country=="USA" ~ "United States of America",
    TRUE ~ Country # Everything else, leave it be
    ))

# Channel Islands are Jersey and Guernsey, but these are already in the data

mar_jobs_rgn_temp <- name_2_rgn(df_in = mar_jobs_rgn, 
                       fld_name='Country',
                      flds_unique = c("year"))

## Received a warning for a duplicated country. Using the following code to display the portion of the dataframe that's duplicated 
    ## Can see that in this case it was China, even though they actually didn't have any duplicated values. Interesting 
mar_jobs_rgn_temp[duplicated(mar_jobs_rgn_temp[, c('Country','year')]),  ] 

# Weight the following duplicates by population
# China, Guadeloupe, Guam, Hong Kong, Macao, Martinique, Northern Mariana Islands, Puerto Rico, Virgin Islands (U.S.)... Populations taken from google

# pop_weights <- data.frame(Country = c("China", "Hong Kong", "Macao",
#                  "Guadeloupe", "Martinique",
#                  "Guam", "Northern Mariana Islands",
#                  "Puerto Rico", "Virgin Islands (U.S.)"),
#                  pop = c(1452883897, 7632334, 669037,
#                          400276, 367507,
#                          172069, 49551,
#                          3252407, 103910))

# mar_jobs_rgn_final <- mar_jobs_rgn_temp %>%
#   # left_join(pop_weights, by = "Country") %>%
#   mutate(pop = ifelse(is.na(pop), 1, pop)) %>%
#   group_by(rgn_id, rgn_name, year) %>%
#   dplyr::summarize(score = weighted.mean(score, pop, na.rm=TRUE),
#             method = paste(unique(method), collapse=" "),
#             gapfill = weighted.mean(gapfill, pop, na.rm=TRUE))%>%
#   ungroup() %>% 
#   dplyr::mutate(year = as.numeric(year))

## Only twenty nine countries with data 
# length(unique(mar_jobs_rgn_temp$Country))
```

