---
title: "Softbottom Habitat Destruction"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---

# Data Sources

## Global Fishing Effort Model Data and Shiny App

**Reference**: [Global Fishing Effort Model Data and Shiny App](https://zenodo.org/records/15758813)

Clawson, G., Blanchard, J., & Novaglio, C. (2025). Global Fishing Effort Model Data and Shiny App. Zenodo. https://doi.org/10.5281/zenodo.15758813

**Downloaded**: 4 August 2025

**Description**:
We employ a statistical remapping approach similar to that of McDonald et al. 2024 that spatially allocates industrial fishing effort based on multiple data sources, including country-level effort data from Rousseau et al. (2019), AIS data from Global Fishing Watch, and environmental and geographical factors (e.g., sea surface temperature, chlorophyll-a concentration, bathymetry, distance to productive fishing grounds) as well as governance variables (e.g., EEZ boundaries, FAO Major Fishing Areas, etc.) from different sources. Our approach ensures that the spatial allocation of fishing effort reflects real-world constraints and ecological drivers and enables us to allocate country-level fishing effort data across the global ocean in a way that aligns with observed vessel behavior, known ecological preferences, and regulatory constraints. The resulting dataset offers a 1 degree resolution. 

**Native data resolution**: 1 degree

**Time range**: 2010-2017

**Format**: .csv

```{r}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE, eval = FALSE)
```

```{r}
library(dplyr)
library(tidyr)
library(tidyverse)
library(here)
library(sf)
library(countrycode)

#source('https://raw.githubusercontent.com/oharac/src/master/R/common.R')
source(here('workflow/R/common.R'))

goal     <- 'hab_prs_hd_subtidal_soft_bottom'
scenario <- 'v2025'

dir_goal <- here('globalprep', goal, scenario)

### goal specific folders and info
dir_anx  <- file.path(dir_M, 'git-annex/globalprep')
dir_goal_anx <- file.path(dir_anx, goal, scenario, 'spp_risk_dists')

```

Download data for all years here: https://data.mapping-global-fishing.cloud.edu.au/shiny/mapping_fishing_effort_app/

Find the reference point:

```{r}
#2010
# Read csv
fishing_effort_2010 <- read_csv(file.path(dir_anx, "_raw_data", "global_fishing_effort", "fishing_effort_trawl_bottom_2010.csv"), show_col_types = FALSE)

# Sum fishing hours by lon, lat, eez_sovereign_iso3c, flag_country_name pixel_area_km2
fishing_effort_2010_summarized <- fishing_effort_2010 %>%
  group_by(lon, lat, eez_sovereign_iso3c, pixel_area_km2) %>%
  summarise(sum_eff_active_fishing_hours = sum(eff_active_fishing_hours, na.rm = TRUE))

check_2010 <- fishing_effort_2010_summarized %>%
  group_by(lon, lat) %>%
  summarise(country_count = n())

unique(check_2010$country_count) # only 1

# Log pressure
fishing_effort_2010_logged <- fishing_effort_2010_summarized %>%
  mutate(pressure = log(sum_eff_active_fishing_hours/pixel_area_km2 +1)) %>%
  filter(!eez_sovereign_iso3c == "High seas")

#hist(fishing_effort_logged$pressure)
quant_99_2010 <- quantile(fishing_effort_2010_logged$pressure, probs = 0.99)
quant_90_2010 <- quantile(fishing_effort_2010_logged$pressure, probs = 0.9)
quant_86_2010 <- quantile(fishing_effort_2010_logged$pressure, probs = 0.86)

rast_2010 <- fishing_effort_2010_logged %>%
  group_by(lat, lon) %>%
  summarise(count = n())

sum(rast_2010$count) #15006

fishing_effort_2010_sf <- fishing_effort_2010_logged %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) 

fishing_effort_2010_grid <- fishing_effort_2010_sf %>%
  st_make_grid(cellsize = c(1,1),
               what = "polygons") %>%
  st_sf() %>%
  # Fix invalid geometries
  st_make_valid()

# Spatially join emissions data to the grid
fishing_effort_2010_grid_sf <- st_join(fishing_effort_2010_grid, fishing_effort_2010_sf, left = FALSE) 

plot(fishing_effort_2010_grid_sf["pressure"])

#2017
# Read csv
fishing_effort_2017 <- read_csv(file.path(dir_anx, "_raw_data", "global_fishing_effort", "fishing_effort_trawl_bottom_2017.csv"), show_col_types = FALSE)

# Sum fishing hours by lon, lat, eez_sovereign_iso3c, flag_country_name pixel_area_km2
fishing_effort_2017_summarized <- fishing_effort_2017 %>%
  group_by(lon, lat, eez_sovereign_iso3c, pixel_area_km2) %>%
  summarise(sum_eff_active_fishing_hours = sum(eff_active_fishing_hours, na.rm = TRUE))

check_2017 <- fishing_effort_2017_summarized %>%
  group_by(lon, lat) %>%
  summarise(country_count = n())

unique(check_2017$country_count) # only 1

# Log pressure
fishing_effort_2017_logged <- fishing_effort_2017_summarized %>%
  mutate(pressure = log(sum_eff_active_fishing_hours/pixel_area_km2 +1)) %>%
  filter(!eez_sovereign_iso3c == "High seas")

#hist(fishing_effort_logged$pressure)
quant_99_2017 <- quantile(fishing_effort_2017_logged$pressure, probs = 0.99)
quant_90_2017 <- quantile(fishing_effort_2017_logged$pressure, probs = 0.9)
quant_86_2017 <- quantile(fishing_effort_2017_logged$pressure, probs = 0.86)

rast_2017 <- fishing_effort_2017_logged %>%
  group_by(lat, lon) %>%
  summarise(count = n())

sum(rast_2017$count) #9235

fishing_effort_2017_sf <- fishing_effort_2017_logged %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) 

fishing_effort_2017_grid <- fishing_effort_2017_sf %>%
  st_make_grid(cellsize = c(1,1),
               what = "polygons") %>%
  st_sf() %>%
  # Fix invalid geometries
  st_make_valid()

# Spatially join emissions data to the grid
fishing_effort_2017_grid_sf <- st_join(fishing_effort_2017_grid, fishing_effort_2017_sf, left = FALSE) 

plot(fishing_effort_2017_grid_sf["pressure"])

```


```{r}
trawl_bottom_files <- list.files(file.path(dir_anx, "_raw_data", "global_fishing_effort"), pattern = ".csv", full.names = TRUE)

region_data()

fishing_effort_trawl_bottom <- list()

for (i in seq_along(trawl_bottom_files)) {
  
  #i <- 1
  
  # Read csv
  fishing_effort <- read_csv(trawl_bottom_files[i], show_col_types = FALSE)
  
  # Extract year
  year <- str_extract(trawl_bottom_files[i], "\\d{4}")
  
  message("Processing data for ", year, ".")
  
  # Sum fishing hours by lon, lat, eez_sovereign_iso3c, flag_country_name pixel_area_km2
  fishing_effort_summarized <- fishing_effort %>%
    filter(!eez_sovereign_iso3c == "High seas") %>%
    group_by(lon, lat, pixel_area_km2) %>%
    summarise(sum_eff_active_fishing_hours = sum(eff_active_fishing_hours, na.rm = TRUE))
  
  # Log pressure
  fishing_effort_logged <- fishing_effort_summarized %>%
    mutate(pressure = log(sum_eff_active_fishing_hours/pixel_area_km2 +1)) 

  fishing_effort_sf <- fishing_effort_logged %>%
    mutate(pressure = pressure/quant_86_2010) %>%
    st_as_sf(coords = c("lon", "lat"), crs = 4326) 
  
  fishing_effort_grid <- fishing_effort_sf %>%
    st_make_grid(cellsize = c(1,1),
                 what = "polygons") %>%
    st_sf() %>%
    # Fix invalid geometries
    st_make_valid()
  
  # Spatially join emissions data to the grid
  fishing_effort_grid_sf <- st_join(fishing_effort_grid, fishing_effort_sf, left = FALSE) 
  
  fishing_effort_trawl_bottom[[year]] <- fishing_effort_grid_sf
}

# Plot
for (i in seq_along(c(2010:2017))) {
  
  #i <- 1
  
  year <- names(fishing_effort_trawl_bottom)[i]
  
  png(file.path(dir_goal, "int/plots", paste0("fishing_effort_", year, "_q86_2010.png")), width = 800, height = 600)
  plot(fishing_effort_trawl_bottom[[i]]['pressure'], main = year)
  
  dev.off()
  
}


plot(fishing_effort_trawl_bottom[["2014"]]['pressure'], main = "2014")
```

1. prepare for reference point

- 2010 vs. 2017
- 99 vs. 90
- raster cells: 15,006

save intermediate maps to github in "int" for each year



zonal mean for each year is final final output (same naming conventions from before)


hist plots

how much of coastal areas highly degraded due to distructive fishing practices to help set referenc point (10%? 1%?)

https://www.pnas.org/doi/10.1073/pnas.1802379115

Overall, 14% of the 7.8 million-km2 study area was trawled, and 86% was not trawled.


at 3, one value for every raster
majority? centroid? how did they assign eez? are pixel repeated? more than one country eez?

```{r}
check <- fishing_effort_2017 %>%
  group_by(lat, lon) %>%
  summarise(count_countries = n_distinct(eez_sovereign_iso3c))

unique(check$count_countries)
```


