---
title: "Softbottom Habitat Destruction"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---
# Summary

Find mean bottom trawling pressure (in terms of fishing effort) for each EEZ.

# Data Sources

## Global Fishing Effort Model Data and Shiny App

**Reference**: [Global Fishing Effort Model Data and Shiny App](https://zenodo.org/records/15758813)

Clawson, G., Blanchard, J., & Novaglio, C. (2025). Global Fishing Effort Model Data and Shiny App. Zenodo. https://doi.org/10.5281/zenodo.15758813

**Downloaded**: 4 August 2025

**Description**:
We employ a statistical remapping approach similar to that of McDonald et al. 2024 that spatially allocates industrial fishing effort based on multiple data sources, including country-level effort data from Rousseau et al. (2019), AIS data from Global Fishing Watch, and environmental and geographical factors (e.g., sea surface temperature, chlorophyll-a concentration, bathymetry, distance to productive fishing grounds) as well as governance variables (e.g., EEZ boundaries, FAO Major Fishing Areas, etc.) from different sources. Our approach ensures that the spatial allocation of fishing effort reflects real-world constraints and ecological drivers and enables us to allocate country-level fishing effort data across the global ocean in a way that aligns with observed vessel behavior, known ecological preferences, and regulatory constraints. The resulting dataset offers a 1 degree resolution. 

**Native data resolution**: 1 degree

**Time range**: 2010-2017 (downloaded individually)

**Format**: .csv

# Methods

```{r}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE, eval = FALSE)
```

```{r}
library(dplyr)
library(tidyr)
library(tidyverse)
library(here)
library(sf)
library(terra)
library(plotly)

#source('https://raw.githubusercontent.com/oharac/src/master/R/common.R')
source(here('workflow/R/common.R'))

goal     <- 'hab_prs_hd_subtidal_soft_bottom'
scenario <- 'v2025'

dir_goal <- here('globalprep', goal, scenario)

### goal specific folders and info
dir_anx  <- file.path(dir_M, 'git-annex/globalprep')
dir_goal_anx <- file.path(dir_anx, goal, scenario, 'spp_risk_dists')
```

### Find the reference point:

Download data for all years here: https://data.mapping-global-fishing.cloud.edu.au/shiny/mapping_fishing_effort_app/

For this analysis, we chose to use the 99.9 percentile (0.1% of area) from 2010 to represent total seafloor destruction: `quant_99_9_2010`. 

```{r}
#2010
# Read csv
fishing_effort_2010 <- read_csv(file.path(dir_anx, "_raw_data", "global_fishing_effort", "fishing_effort_trawl_bottom_2010.csv"), show_col_types = FALSE)

# Sum fishing hours by lon, lat, eez_sovereign_iso3c, flag_country_name pixel_area_km2
fishing_effort_2010_summarized <- fishing_effort_2010 %>%
  group_by(lon, lat, eez_sovereign_iso3c, pixel_area_km2) %>%
  summarise(sum_eff_active_fishing_hours = sum(eff_active_fishing_hours, na.rm = TRUE))

check_2010 <- fishing_effort_2010_summarized %>%
  group_by(lon, lat) %>%
  summarise(country_count = n())

unique(check_2010$country_count) # only 1

# Log pressure
fishing_effort_2010_logged <- fishing_effort_2010_summarized %>%
  mutate(pressure = log(sum_eff_active_fishing_hours/pixel_area_km2 +1)) %>%
  filter(!eez_sovereign_iso3c == "High seas")

#hist(fishing_effort_logged$pressure)
#quant_99_99_2010 <- quantile(fishing_effort_2010_logged$pressure, probs = 0.9999)
quant_99_9_2010 <- quantile(fishing_effort_2010_logged$pressure, probs = 0.999)
#quant_99_2010 <- quantile(fishing_effort_2010_logged$pressure, probs = 0.99)
#quant_90_2010 <- quantile(fishing_effort_2010_logged$pressure, probs = 0.9)
#quant_86_2010 <- quantile(fishing_effort_2010_logged$pressure, probs = 0.86)

#plot(fishing_effort_2010_grid_sf["pressure"])
```

### Rasterize trawling effort data

Calculate fishing effort in terms of log(`sum_eff_active_fishing_hours`/`pixel_area_km2` +1). Cap values at 1. 

```{r}
# Load bottom trawling filenames
trawl_bottom_files <- list.files(file.path(dir_anx, "_raw_data", "global_fishing_effort"), pattern = ".csv", full.names = TRUE)

# Initiate an empty list
fishing_effort_trawl_bottom <- list()

# Rasterize each year's data
for (i in seq_along(trawl_bottom_files)) {
  
  #i <- 1
  
  # Read csv
  fishing_effort <- read_csv(trawl_bottom_files[i], show_col_types = FALSE)
  
  # Extract year
  year <- str_extract(trawl_bottom_files[i], "\\d{4}")
  
  message("Processing data for ", year, ".")

  # Sum fishing hours by lon, lat, eez_sovereign_iso3c, flag_country_name pixel_area_km2
  fishing_effort_summarized <- fishing_effort %>%
    filter(!eez_sovereign_iso3c == "High seas") %>%
    group_by(lon, lat, pixel_area_km2) %>%
    summarise(sum_eff_active_fishing_hours = sum(eff_active_fishing_hours, na.rm = TRUE))
  
  # Log pressure
  fishing_effort_logged <- fishing_effort_summarized %>%
    mutate(pressure = log(sum_eff_active_fishing_hours/pixel_area_km2 +1)) 
  
  # Convert to spatial points
  pts <- vect(fishing_effort_logged, geom = c("lon", "lat"), crs = "EPSG:4326")
  
  #plot(pts)
  
  # Calculate extent and set 1-degree resolution
  ext <- ext(min(fishing_effort_logged$lon), max(fishing_effort_logged$lon), min(fishing_effort_logged$lat), max(fishing_effort_logged$lat))
  
  # Create raster template
  r_template <- rast(ext, resolution = 1, crs = "EPSG:4326")
  
  # Rasterize fishing effort
  fishing_effort_raster <- rasterize(pts, r_template, field = "pressure")
  
  #plot(fishing_effort_raster)
  
  # Divide by reference point
  fishing_effort_raster <- fishing_effort_raster/quant_99_9_2010
  
  #plot(fishing_effort_raster)
  
  # Cap pressure value at 1
  fishing_effort_raster <- ifel(fishing_effort_raster > 1, 1, fishing_effort_raster)
  
  # Append to list
  fishing_effort_trawl_bottom[[year]] <- fishing_effort_raster
}

data_years <- as.numeric(names(fishing_effort_trawl_bottom))

# # Plots
# for (i in seq_along(c(min(data_years):max(data_years)))) {
#   
#   #i <- 1
#   
#   year <- names(fishing_effort_trawl_bottom)[i]
#   
#   png(file.path(dir_goal, "int/plots", paste0("fishing_effort_", year, "_q99_99_2010.png")), width = 800, height = 600)
#   plot(fishing_effort_trawl_bottom[[i]]['pressure'], main = year)
#   
#   dev.off()
#   
# }
```

### Calculate mean pressure

Use `zonal()` to calculate mean pressure scores for each EEZ. Make sure to fill all NA values with 0. 

```{r}
# Turn off scientific notation
options(scipen = 999)

# Load regions data
region_data()

# Load the EEZ raster
eez_ocean_raster <- rast(file.path('globalprep/spp/v2024', '_spatial', 'eez_rast.tif')) 

#plot(eez_ocean_raster)

# Initiate empty list
fishing_pressure <- list()

# Calculate zonal stats for each year
for (i in seq_along(c(min(data_years):max(data_years)))) {
  
  # Pull year
  year <- names(fishing_effort_trawl_bottom)[i]
  
  # Pull raster for that year
  fishing_effort_raster <- project(fishing_effort_trawl_bottom[[year]], eez_ocean_raster)
  
  # Fill NA with 0 (for zonal stats)
  fishing_effort_raster <- ifel(is.na(fishing_effort_raster), 0, fishing_effort_raster)

  # Calculate zonal stats
  zonal_stats <- zonal(fishing_effort_raster, eez_ocean_raster, fun = "mean") %>% # na.rm = TRUE
    rename(rgn_id = eez_rast,
           pressure = last) %>%
    mutate(year = year)
  
  # Append to list
  fishing_pressure[[year]] <- zonal_stats
  
}

# Combine all years for output
total_fishing_pressure <- bind_rows(fishing_pressure) %>%
  select(rgn_id, year, pressure_score = pressure)

# Save output layer
write_csv(total_fishing_pressure, file.path(dir_goal, 'output', 'hd_sb_subtidal.csv'))

# Histogram
# ggplot(total_fishing_pressure, aes(x = pressure_score, fill = factor(year))) +
#   geom_histogram(bins = 30, color = "black", alpha = 0.8) + #position = "dodge",
#   labs(fill = "Year", x = "Pressure", y = "Count") +
#   scale_fill_brewer(palette = "Blues") +
#   theme_minimal()
# 
# ggplot(total_fishing_pressure, aes(x = pressure_score, fill = factor(year))) +
#   geom_histogram(bins = 30, color = "black", alpha = 0.8) +
#   labs(x = "Pressure", y = "Count", fill = "Year") +
#   scale_fill_brewer(palette = "Blues") +
#   facet_wrap(~year, ncol = 1) +
#   theme_minimal() +
#   theme(legend.position = 'none')
```

### Check with v2017

```{r}
# Import old data
hd_sb_subtidal_2017 <- read_csv(here('globalprep', goal, "v2017", "output", "hd_sb_subtidal.csv"), show_col_types = FALSE) 

# Combine data
hd_sb_subtidal_compare <- total_fishing_pressure %>%
  mutate(year = as.numeric(year)) %>%
  rename(pressure_v2025 = pressure_score) %>%
  left_join(hd_sb_subtidal_2017, by = c("rgn_id", "year")) %>%
  rename(pressure_v2017 = pressure_score) %>%
  left_join(rgns_eez, by = "rgn_id")

# Static plot
ggplot(hd_sb_subtidal_compare, aes(pressure_v2017, pressure_v2025)) +
  geom_point() +
  geom_abline(intercept = 0, color = "red") +
  theme_minimal()

# Interactive plot
p <- ggplot(hd_sb_subtidal_compare, aes(x = pressure_v2017, y = pressure_v2025, group = rgn_name)) + #color = factor(rgn_name)
  geom_abline(intercept = 0, slope = 1, color = "red") +
  geom_point(size = 2, alpha = 0.8) +
  labs(color = "Region ID", x = "Pressure v2017", y = "Pressure v2025", title = "Years: 2010-2014") +
  theme_minimal() +
  #scale_color_viridis_d() +
  theme(legend.position = 'none')

ggplotly(p)

# Compare 2010 and 2017 from v2025
total_fishing_pressure %>%
  filter(year %in% c(2010,2017)) %>%
  pivot_wider(names_from = "year", values_from = "pressure_score") %>%
  ggplot(aes(`2010`, `2017`)) +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  geom_point() +
  labs(x = "Year: 2010", y = "Year: 2017") +
  theme_minimal()

```

