---
title: 'OHI 2022 - Targeted harvest pressure'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html' 
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

[REFERENCE RMD FILE](http://ohi-science.org/ohiprep_v2021/globalprep/prs_targetedharvest/v2021/targetharvest_dataprep.html)

# Summary

This analysis converts FAO capture production data into the OHI 2021 targeted harvest pressure data.  

# Updates from previous assessment

2022 - One more year of data

***

# Data Source 

    http://www.fao.org/fishery/statistics/software/fishstatj/en#downlApp
     Release date: March 2021  
FAO Global Capture Production Quantity 1950_2019  
Information: http://www.fao.org/fishery/statistics/global-capture-production/en  

**Reference**: United Nations, 2021. FAO Fisheries & Aquaculture - Fishery Statistical Collections - Global Capture Production [WWW Document]. URL http://www.fao.org/fishery/statistics/global-capture-production/en (accessed 4.29.21).


**Downloaded**: April 25, 2022

**Description**:  Quantity (tonnes) of fisheries capture for each county, species, year.

**Time range**: 1950-2020

***

# Methods

## Setup

```{r, eval=FALSE}

# load libraries, set directories
library(ohicore)  #devtools::install_github('ohi-science/ohicore@dev')
library(tidyverse)
library(plotly)
library(here)
library(janitor)

### Load FAO-specific user-defined functions
source(here('workflow/R/fao_fxn.R')) # function for cleaning FAO files (not combined into common.R like most other functions have been at this point)
source(here('workflow/R/common.R')) # directory locations

version_year <- 2022
latest_data_yr <- version_year - 2
```

## Read in the raw data

This includes the FAO capture production data and a list of the "target" species.

```{r, eval=FALSE}

## FAO capture production data - all columns being parsed as characters and producing error in one column, but not sure which? (read.csv might help avoid this error?)
fis_fao_raw <- read_csv(
  file.path(dir_M, 'git-annex/globalprep/_raw_data/FAO_capture',
            paste0('d', version_year),
            paste0('Global_capture_production_Quantity_1950-', latest_data_yr,'.csv'))
  # , na = "..."
)

# List of species included as cetaceans or marine turtles (this probably won't change at all)
sp2grp <- read_csv(here('globalprep', 'prs_targetedharvest', paste0('v', version_year), 'raw', 'species2group.csv')) %>%
  dplyr::filter(incl_excl == 'include') %>%
  dplyr::select(target, species); head(sp2grp)

```

## Clean the FAO data

```{r, eval=FALSE}

# Rename columns and remove unit column
fao_clean <- fis_fao_raw %>% 
  dplyr::rename(country = "Country (Name)",
                species = "ASFIS species (Name)",
                area = "FAO major fishing area (Name)") %>%
  dplyr::select(-"Unit (Name)") %>%
  dplyr::rename_with(~ base::gsub("\\[", "", .)) %>% 
  dplyr::rename_with(~ base::gsub("\\]", "", .))
  # rename_at(vars(matches("\\[")), ~ str_remove(., "\\[")) %>%
  # rename_at(vars(matches("\\]")), ~ str_remove(., "\\]"))

# Gather by year and value to expand and make each line a single observation for country, species and year (tidy data!) 

# Pivot_longer by year and value to expand and make each line a single observation for country, species and year (tidy data!)
fao_clean <- fao_clean %>%
  tidyr::pivot_longer(cols = -(1:3), names_to = 'year', values_to = 'value', values_drop_na = T) %>%
  # tidyr::gather("year", "value", -(1:3)) %>%
  dplyr::mutate(year = gsub("X", "", year)) %>%
    fao_clean_data_new() 

fao_clean <- fao_clean %>%
  dplyr::mutate(
    species = as.character(species),
    species = stringr::str_replace_all(
      string = species, 
      pattern = "Henslow.*s swimming crab", 
      replacement = "Henslow's swimming crab")
    )
  # dplyr::mutate(species = ifelse(stringr::str_detect(species, "Henslow.*s swimming crab"), "Henslow's swimming crab", species))


```

## Identify the target species

This analysis only includes target species.  The warning messages need to be checked and, if necessary, changes should be made to the raw/species2group.csv  

```{r, eval=FALSE}

# check for discrepancies in species list
spgroups <-  sort(as.character(unique(fao_clean$species))) # species groups in FAO data 
groups <-  c('turtle', 'whale', 'dolphin', 'porpoise') # seals and sea lions removed from vector (pinnipeds no longer included) 

# Going through FAO data species and seeing if they're in our master list of species
## Looking to see if we need to add species that have changed name
for (group in groups) {# group='dolphin'
  possibles <- spgroups[grep(group, spgroups)]
  d_missing_l <-  setdiff(possibles, sp2grp$species)
  if (length(d_missing_l) > 0){
    cat(sprintf("\nMISSING in the lookup the following species in target='%s'.\n    %s\n", 
                group, paste(d_missing_l, collapse='\n    ')))
  }
}

# check for species in lookup not found in data
l_missing_d <-  setdiff(sp2grp$species, spgroups)
if (length(l_missing_d) > 0){
  cat(sprintf('\nMISSING: These species in the lookup are not found in the FAO data \n'))
  print(l_missing_d)
}

#### v2022
# MISSING in the lookup the following species in target='turtle'.
#     Chinese softshell turtle - not a marine turtle
#     River and lake turtles nei - not a marine turtle
# 
# MISSING in the lookup the following species in target='whale'.
#     Creek whaler - shark, not a whale
#     Velvet whalefish - fish, not a whale
# 
# MISSING in the lookup the following species in target='dolphin'.
#     Common dolphinfish - fish not a cetacean
#     Pompano dolphinfish - fish not a cetacean

## filter data to include only target species ----
target_spp <- fao_clean %>%
  dplyr::filter(species %in% sp2grp$species) # this goes from 2384 spp in FAO list to just 72

unique(target_spp$area) # confirm these are all marine regions
unique(fao_clean$species) # 2384 species

```

## Summarize data

```{r, eval=FALSE}

# pivot wider to expand years
wide <- target_spp %>%
  tidyr::pivot_wider(names_from = year, values_from = value) %>%
  dplyr::select(-`NA`) %>%
  # tidyr::spread(year, value) %>%
  dplyr::left_join(sp2grp, by='species'); head(wide) 

# pivot longer long by target
long <- wide %>%
  dplyr::select(-area) %>%
  tidyr::pivot_longer(cols = c(-country, -species, -target),
                      names_to = 'year',
                      values_to = 'value',
                      values_drop_na = T) %>%
  # tidyr::gather(year, value, -country, -species, -target, na.rm=T) %>%
  dplyr::mutate(year = as.integer(as.character(year))) %>%
  dplyr::arrange(country, target, year); head(long)


# explore Japan[210] as an example
japan <- long %>% 
  dplyr::group_by(country, target, year) %>%
  dplyr::summarize(value = sum(value)) %>% 
  dplyr::filter(country == 'Japan', target == 'cetacean', year >= 2000) 

# summarize totals per region per year - number of individual animals from each spp group? 
sum <- long %>%
  dplyr::group_by(country, year) %>%
  dplyr::summarize(value = sum(value, na.rm=TRUE)) %>%
  dplyr::filter(value != 0) %>%
  dplyr::ungroup(); head(sum) 

```

## Assign country names to OHI regions

```{r, eval=FALSE}

sum <- sum %>%
  dplyr::mutate(country = as.character(country)) %>%
  dplyr::mutate(country = ifelse(stringr::str_detect(country, "C.*te d'Ivoire"), "Ivory Coast", country))

### Function to convert to OHI region ID
m_sum_rgn <- name_2_rgn(df_in = sum, 
                       fld_name='country', 
                       flds_unique=c('year'))

# Check out duplicates based on error message from previous step
dplyr::filter(m_sum_rgn, country %in% c("Guadeloupe", "Martinique")) 
# this is ok, we report these two together, so this will be fixed with the summarize in the next step 

# They will be summed:
m_sum_rgn <- m_sum_rgn %>%
  dplyr::group_by(rgn_id, rgn_name, year) %>%
  dplyr::summarize(value = sum(value)) %>%
  dplyr::ungroup()

```

## Scale the data and save files

Data is rescaled by dividing by the 95th quantile of values across all regions from 2011 to 2020 (most recent year of FAO data).

```{r, eval=FALSE}

target_harvest <- m_sum_rgn %>%
  dplyr::mutate(
    quant_95 = quantile(value[year %in% 2011:latest_data_yr], 0.95, na.rm = TRUE),
    score = value / quant_95,
    score = ifelse(score > 1, 1, score)) %>%
  dplyr::select(rgn_id, year, pressure_score = score) %>%
  dplyr::arrange(rgn_id, year); head(target_harvest); summary(target_harvest)
  
# v2021 quant_95 = 3409.4
# v2022 quant_95 = 3450.05

# any regions that did not have a catch should have score = 0 
rgns <-  rgn_master %>%
  dplyr::filter(rgn_typ == "eez") %>%
  dplyr::select(rgn_id = rgn_id_2013) %>%
  dplyr::filter(rgn_id < 255) %>%
  base::unique() %>%
  dplyr::arrange(rgn_id)

# This is just a list of rgn IDS - do we want to update it to a rgn list more recent than 2013?

# Add year; for v2022, min year is 1950, and max year is 2020 
rgns <- expand.grid(rgn_id = rgns$rgn_id, year = min(target_harvest$year):max(target_harvest$year))

# Change NAs in pressure_score column to 0s
target_harvest <-  rgns %>%
  dplyr::left_join(target_harvest) %>%
  dplyr::mutate(pressure_score = ifelse(is.na(pressure_score), 0, pressure_score)) %>%
  dplyr::arrange(rgn_id); head(target_harvest); summary(target_harvest)

# Write target_harvest to "fao_targeted.csv" in output folder
write_csv(
  target_harvest, 
  here('globalprep', 'prs_targetedharvest', paste0('v', version_year), 'output', 'fao_targeted.csv')
)

# Create gapfill dataframe
target_harvest_gf <- target_harvest %>%
  dplyr::mutate(gapfill = 0) %>%
  dplyr::select(rgn_id, year, gapfill)
# all zeroes for gapfill column; nothing being gapfilled but need to have a record 

# Write target_harvest_gf to "fao_targeted_gf.csv" in output folder
write_csv(
  target_harvest_gf,
  here('globalprep', 'prs_targetedharvest', paste0('v', version_year), 'output', 'fao_targeted_gf.csv')
)
```

## Data check

The data from last year and this year should be the same unless there were changes to underlying FAO data or the master species list.

In this case, all of the regions looked very similar.

```{r, eval=FALSE}

# pull just 2019 data from target_harvest df, since its the most recent year in common
common_year <- 2019

new <- readr::read_csv(here("globalprep", "prs_targetedharvest", paste0('v', version_year), "output", "fao_targeted.csv")) %>% 
  dplyr::filter(year == common_year)

old <- readr::read_csv(here("globalprep", "prs_targetedharvest", paste0('v', version_year - 1), "output", "fao_targeted.csv")) %>%
  dplyr::filter(year == common_year) %>%
  dplyr::select(rgn_id, year, pressure_score_old = pressure_score) %>%
  dplyr::left_join(new, by = c("rgn_id", "year"))

# Compare pressure_score between last year and this year's assessments
compare_plot <- ggplot(data = old, aes(x = pressure_score_old, y = pressure_score, label = rgn_id)) +
  geom_point() +
  geom_abline(color = "red")

plot(compare_plot)
ggplotly(compare_plot)

### v2022: Explore outliers

### explore United States [163]
# out_country <- "United States of America"
### explore Indonesia [216]
out_country <- "Indonesia"

outlier_country <- long %>% 
  dplyr::group_by(country, target, year) %>%
  dplyr::summarize(value = sum(value)) %>% 
  dplyr::filter(country == out_country, year >= 2000) 


region_data() ## read in regions 


new <- readr::read_csv(here("globalprep", "prs_targetedharvest", paste0('v', version_year), "output", "fao_targeted.csv")) %>% 
  dplyr::filter(year == 2020) %>%
  dplyr::select(-year)

old <- readr::read_csv(here("globalprep", "prs_targetedharvest", paste0('v', version_year - 1), "output", "fao_targeted.csv")) %>%
  dplyr::filter(year == common_year) %>%
  dplyr::select(-year) %>%
  dplyr::select(rgn_id, pressure_score_2019 = pressure_score) %>%
  dplyr::left_join(new, by = c("rgn_id")) %>%
  dplyr::rename(pressure_score_2020 = pressure_score) %>%
  left_join(rgns_eez)
  

# Compare pressure_score between last year and this year's assessments
compare_plot <- ggplot(data = old, aes(x = pressure_score_2019, y = pressure_score_2020, label = rgn_name)) +
  geom_point() +
  geom_abline(color = "red")

plot(compare_plot)
ggplotly(compare_plot)


```










