---
title: 'OHI `r format(Sys.time(), "%Y")`: Natural Products: Seaweeds Data Prep'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: false
    toc_float: yes
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---



[REFERENCE RMD FILE: <http://ohi-science.org/ohiprep_v2022/globalprep/np/v2021/STEP1b_np_seaweeds_prep.Rmd>]

# Summary

This analysis converts FAO mariculture data into one of the data layers used to calculate OHI `r format(Sys.time(), "%Y")` global natural products (NP) scores. We will conduct the overall NP data prep on seaweeds, fish oil/fish meal (FOFM), and ornamentals, however, our final layer from this data prep will only consist of seaweeds.

# Updates from previous assessment

-   New year of FAO mariculture data (1950-2023).


------------------------------------------------------------------------

# Data Source

## Production data

**Reference**:\
<https://www.fao.org/fishery/en/collection/aquaculture?lang=en> Release date: March 2023 FAO Global aquaculture production Quantity (1950 - 2023) Global aquaculture production (Value) (1950-2023)

**Downloaded**: June 23rd, 2025

**Instructions on downloading the FAO aquaculture data:**

**2025 download from Melanie**: 
Navigate to this link: <https://www.fao.org/fishery/statistics-query/en/global_production/global_production_quantity>

Click Download full dataset. Also downloaded the file below!

Navigate to this link: <https://www.fao.org/fishery/statistics-query/en/home>.

Click browse on Global aquaculture production. Click browse on Global aquaculture production (Quantity). Scroll to the bottom of the dimensions section, and de-select all of the pre-selected years. Under the query definition section, drag Country Name En (if not already present), ASFIS species Name En, FAO major fishing area Name En, ASFIS species Family scientific name, and Environment Name En, from the possible fields column into the selected rows column. Click on show data and confirm that data is present for 1950- two years prior to current year. Click download and select yes to include Null Values. Rename the file in the format FAO_GlobalAquacultureProduction_Quantity_1950_last_data_year. Upload the file to the appropriate year folder in the FAO_mariculture folder on mazu.

**Description**: Quantity (tonnes) of mariculture for each country, species, year.

**Time range**: 1950-2023

## Seafood Watch sustainability data

Obtained upon request of SFW.

**Reference**:\
[Seafood Watch Complete Recommendation documentation](https://www.seafoodwatch.org/globalassets/sfw/pdf/whats-new/seafood-watch-complete-recommendation-list.pdf)\
[Seafood Watch Scoring Theory](https://www.seafoodwatch.org/globalassets/sfw/pdf/standards/aquaculture/seafood-watch-aquaculture-standard-version-a4.pdf)\
[Seafood Watch Global Seaweed Assessment](https://www.seafoodwatch.org/recommendation/seaweed/seaweed-31568) (provides gapfilling constants)

**Downloaded**: June 27, 2025

**Description**: Monterrey Bay Aquarium Seafood Watch aquaculture recommendations. Sustainability scored from 0-10. Rescaled to 0-1. There is only one value for seaweeds in the data... 0.67

------------------------------------------------------------------------

# Methods



## Setup

```{r}

knitr::opts_chunk$set(eval = FALSE)

# ======= Load packages ============
if (!require(librarian)) {
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  ohicore, #devtools::install_github('ohi-science/ohicore@dev') #if relevant: restart session after reinstalling
  here,
  tidyverse,
  zoo,
  ggplot2,
  plotly,
  tictoc,
  RColorBrewer
)

# ======= Set directories ===========
# Update scenario year, set up programmatic scenario year updates
scen_year_number <- 2025 # update this!!
scen_year <- as.character(scen_year_number)
prev_scen_year <- as.character(scen_year_number - 1)
data_dir_year <- paste0("d", scen_year)
prev_data_dir_year <- paste0("d", prev_scen_year)
v_scen_year <- paste0("v", scen_year)
prev_v_scen_year <- paste0("v", prev_scen_year)

data_years <- c("1950", "2023") # update to correct data years

# Natural Products directory on GitHub
current_np_dir <- here::here("globalprep", "np", paste0(v_scen_year, "_new")) # update this if name changes
# Mariculture directory on GitHub
current_mar_dir <- here::here("globalprep", "mar", v_scen_year)
rgn_labels <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohi-global/refs/heads/draft/eez/layers/rgn_labels.csv")

# ======== Source functions ===========
source(here::here("workflow", "R", "common.R")) # directory locations
# Load FAO-specific user-defined functions
source(here::here("workflow", "R", "fao_fxn.R")) # function for cleaning FAO files
# function for handling FAO commodity data specific to NP
source(here::here(current_np_dir, "R", "np_fxn.R")) 
# functions specific to mariculture dealing with compound countries
source(here("globalprep", "mar", v_scen_year, "mar_fxs.R"))
```



### Import Raw Data: FAO Mariculture data

Mariculture production in tonnes. Ultimately used to find tonnes of non-food seaweed.

```{r}
# Read in mariculture data
mar_raw <- readr::read_csv(here(
  dir_M, "git-annex", "globalprep", "_raw_data", "FAO_mariculture", data_dir_year,
  paste0("FAO_GlobalAquacultureProduction_Quantity_1950_",
         data_years[2], ".csv")))

head(mar_raw) 
```


# Data Wrangling

## Tidy mariculture data

Filter freshwater mariculture, make long format, and clean FAO codes.

```{r}
# Preliminary cleaning
mar <- mar_raw %>%
  dplyr::select(-`Unit Name`) %>% # should all be "Tonnes - live weight"
  rename(country = `Country Name En`,
         FAO_name = `ASFIS species Name En`, 
         fao = `FAO major fishing area Name En`, 
         environment = `Environment Name En`,
         family_scientific = `Family Scientific name`) %>%
  rename_at(vars(matches("\\[")), ~ str_remove(., "\\[")) %>%
  rename_at(vars(matches("\\]")), ~ str_remove(., "\\]"))

# check environments
table(mar$environment)  

# Include only marine and brackishwater environments
mar <- mar %>%
  filter(environment %in% c("Brackishwater", "Marine"))  

# Change to latest *DATA* year
latest_year <- data_years[2]

#As of 2023 we no longer need to run the fao_clean_data_new function
# NAs are no longer saved as ... and the flags are in a separate column
#however we should still replace columns where the flag is N w/.1
sub_N = 0.1

# add an index
mar <- mar %>% mutate(row_id = row_number())

# note: the following code is very similar to the function in 
# ohiprep_v2024/workflow/R/fao_online_portal_clean.R
# and can be updated to just source the function. 
## more notes on this at the end of the chunk

# pivot all of the year/value columns 
mar_values <- mar %>% 
  dplyr::select(-c(paste("1950":latest_year, "Flag"))) %>% 
  pivot_longer(cols = paste0("1950":latest_year),
               names_to = "year",
               values_to = "value")

# pivot all of the flag columns   
mar_flags <- mar %>% 
  dplyr::select(-paste0("1950":latest_year)) %>% 
  pivot_longer(cols = paste("1950":latest_year, "Flag"),
               names_to = "flag_year",
               values_to = "flag") %>% 
  mutate(year = str_remove(flag_year, " Flag")) %>% 
  dplyr::select(year, flag, row_id)
  
# combine flag and row id 
mar <- mar_values %>% left_join(mar_flags, by = c("row_id", "year"))

mar <- mar %>% 
  mutate(value = case_when(str_detect(flag, "N") ~ sub_N,
                               TRUE ~ value)) %>%
  dplyr::select(-row_id) %>% 
  mutate(FAO_name = ifelse(!is.na(FAO_name), FAO_name, family_scientific)) 


# # ---- test: using fao_online_portal_clean.R function ----
# # note: this function drops the "flag" column!
# source(here("workflow", "R", "fao_online_portal_clean.R"))
# mar_portal_clean <- fao_online_portal_clean(fao = mar, initial_data_year = data_years[1],
#                         last_data_year = data_years[2], sub_N = sub_N) %>% 
#   mutate(FAO_name = ifelse(!is.na(FAO_name), FAO_name, family_scientific)) 
```



## Update seaweed species information

Filter out seaweed species from 'raw/species_list.csv' (from 'globalprep/mar/v2020/raw'), rename columns, and assign proportions to "include" column determined by research of non-human food vs human food seaweed species cultivated in mariculture. For NP, we are only including non-human food seaweed species. Since some species are used for both non-human food and human food purposes, a proportion is assigned based on research and best guess.

**Mariculture data prep must be completed before you complete this part**

```{r}
# Read in 'species_list.csv' (originally from 'globalprep/mar/v2025/raw'). 
#> Filter for 'Taxon_code = AL' (only seaweed species). 
#> Rename 'exclude' columns to 'include' since we're now including 
#> seaweed species that were excluded in the MAR dataprep (not primarily used as human food). 
#> Therefore, "0" means exclude completely (0%), and "1" means include completely (100%).

#> Note, this calls out to the mariculture data prep. 
#> The mariculture data prep for your year should be done before you retrieve this file. 


seaweed_sp <- readr::read_csv(here(current_mar_dir, "raw", "species_list.csv")) %>%
  filter(Taxon_code == 'AL') %>%
  rename(include = exclude)

## Save seaweed species list
readr::write_csv(seaweed_sp, here(current_np_dir, "raw", "species_list_np_seaweeds.csv"))

```




## Update species names

Update species name in the `raw/species_list_np_seaweeds_edited.csv` file with names in the `mar` dataset. Simplified the species list and cut the "species" name columns because it wasn't clear what this was trying to accomplish and created potential error.

```{r}
# Read in edited 'species_list_np_seaweeds_edited.csv'.
seaweeds <- read_csv(here(current_np_dir, "raw", "species_list_np_seaweeds.csv"))

# select specific columns
seaweeds_sp <- seaweeds %>%
  dplyr::select(FAO_name, include, Taxon_code, family)

# REMOVE SPECIES not relevant to natural products goal (i.e., human food species)
seaweed_np <- seaweeds_sp %>% 
  left_join(mar, by = "FAO_name") %>% 
  filter(include > 0)

spp_1 <- unique(sort(seaweed_sp$FAO_name))
length(spp_1) # 62
spp_2 <- unique(sort(seaweed_np$FAO_name))
length(spp_2) # 53
# v2025 went from 62 species to 53 species (9 completely human food species removed)
rgn_1 <- unique(seaweed_np$country)
length(rgn_1) # 51

# Sum production values for each group to account for duplicate rows after name change (remove NA values)
seaweed_np <- seaweed_np %>%
  filter(!is.na(value)) %>%
  group_by(country, fao, environment, FAO_name, year, Taxon_code, family, include) %>% 
    summarize(value = sum(value)) %>% 
  ungroup()

spp_3 <- unique(sort(seaweed_np$FAO_name))
length(spp_3) # 32
rgn_2 <- unique(seaweed_np$country)
length(rgn_2) # 50

setdiff(spp_2, spp_3) # 25 species lost 
# v2024:
#  [1] "BANGIACEAE"        "Bright green nori"
#  [3] "CHORDARIACEAE"     "DUNALIELLACEAE"   
#  [5] "Giant kelps nei"   "GIGARTINACEAE"    
#  [7] "Kelp nei"          "LESSONIACEAE"     
#  [9] "Mozuku"            "OSCILLATORIACEAE" 
# [11] "SARGASSACEAE"      "SOLIERIACEAE"     
# [13] "ULOTRICHACEAE"     "Wakame nei" 

# v2025
#  [1] "[Eucheuma isiforme]"        "[Macrocystis integrifolia]"
#  [3] "[Porphyra columbina]"       "[Spirulina maxima]"        
#  [5] "Aquatic plants nei"         "BANGIACEAE"                
#  [7] "Bright green nori"          "CHORDARIACEAE"             
#  [9] "DUNALIELLACEAE"             "Eucheuma seaweeds nei"     
# [11] "Giant kelps nei"            "GIGARTINACEAE"             
# [13] "Green sea feather"          "Kelp nei"                  
# [15] "LESSONIACEAE"               "Mozuku"                    
# [17] "Nori nei"                   "OSCILLATORIACEAE"          
# [19] "SARGASSACEAE"               "Sea lettuces nei"          
# [21] "Seaweeds nei"               "SOLIERIACEAE"              
# [23] "Spirulina nei"              "ULOTRICHACEAE"             
# [25] "Wakame nei"     

setdiff(rgn_1, rgn_2)
# v2025: NA/50 unique regions?

# Eliminate country-species data with zero production throughout the time-series (1950-recent)
seaweed_np <- seaweed_np %>%
  group_by(country, FAO_name) %>%
  mutate(total_value = sum(value)) %>%
  filter(total_value > 0) %>%
  dplyr::select(-total_value) %>%
  ungroup()

```

## Convert country names to OHI regions

```{r}
# ---- rename FAO_name to species ----
# this is done because these functions require 'species' because they were made for an older data format
seaweed_np <- seaweed_np %>% rename(species = FAO_name)

# ---- Divide mariculture from countries that we report as separate regions (assume equal production in all regions) ----
# Netherlands Antilles: Conch restoration among Aruba, Bonaire, Curacao
# Channel Islands: Jersey and Guernsey
# Bonaire/S.Eustatius/Saba
# Yugoslavia SFR: no longer a country after 1992
seaweed_np <- seaweed_np %>%
  mutate(country = ifelse(country == "Réunion", "Reunion", country)) %>%  # this one is hard to get right; v2020: last year it was "R\xe9union", but this year it was "Réunion" - not present in v2021 data, not present in v2025 data
  ohicore::split_regions()  # function in mar_fxs.R

mar_rgn_all <- name_2_rgn(df_in = seaweed_np, 
                       fld_name = 'country', 
                       flds_unique = c('species', 'fao', 'environment', 'Taxon_code', 'year', 'include'))

# ---- Duplicates ----
# v2025:
# 1 United Republic of Tanzania          
# 2 United Republic of Tanzania, Zanzibar
# okay because this gets resolved in next step

# ---- Sum values of regions with multiple subregions ----
mar_rgn <- mar_rgn_all %>%
  group_by(fao, environment, species, year, Taxon_code, family, rgn_id, include) %>% 
  summarize(value = sum(value)) %>%
  ungroup()
# went from 3150 to 3121 obs - v2021 - This is correct; Un. Sov. Soc. Rep. and Russian Federation were combined.
# went from 3233 to 3173 obs - v2022 - Same as 2021 and Zanzibar / Tanzania were combined
# went from 3162 to 3130 obs - v2023
# went from 3483 to 3450 obs - v2024 - decrease of 33 indicates expected trend from when Zanzibar / Tanzania were combined
# v2025 - went from 2922 to 2888 decrease of 34 indicates expected trend from when Zanzibar / Tanzania were combined
nrow(mar_rgn_all)
nrow(mar_rgn)

# There was no diff in 2024 or 2025
setdiff(mar_rgn$rgn_id, mar_rgn_all$rgn_id)
```

**Take a look at the tidied data for a single year and region**

```{r}
data.frame(
  filter(mar_rgn, rgn_id == 182) %>%
  filter(year == 2016) %>%
  arrange(species)
  )
```


# Gapfilling

## Fill in missing years after first year of harvest data with 0 values

Checked to make sure that there weren't instances in which it made more sense to carry the previous year's data forward as a method of gapfilling. This didn't seem to be the case.

```{r}
# Pivot wider mar_rgn to create a value for every year-species-region (if missing it will be given a NA)
mar_rgn_wide <- mar_rgn %>% 
  pivot_wider(names_from = year, values_from = value)

dim(mar_rgn_wide)

# Turn data frame back into long format
mar_rgn_gf <- mar_rgn_wide %>% 
  pivot_longer(cols = -(fao:include), 
               names_to = "year", 
               values_to = "value") %>% 
  arrange(rgn_id, species, year, Taxon_code, fao, environment) 

# NA values are converted to zero
mar_rgn_gf <- mar_rgn_gf %>%
  mutate(year = as.numeric(as.character(year))) %>%
  mutate(value_w_0 = ifelse(is.na(value), 0, value)) %>%
  group_by(fao, environment, species, Taxon_code, rgn_id) %>% 
  mutate(cum_value = cumsum(value_w_0)) %>%
  ungroup() %>%
  filter(cum_value > 0) %>% # eliminates years before mariculture began
  mutate(gap_0_fill = ifelse(is.na(value), "NA_to_zero", "0")) %>% # record gapfill
  mutate(value = ifelse(is.na(value), 0, value)) %>% # finally, convert all NAs in original column to 0
  dplyr::select(-cum_value, -value_w_0)

```

See how may NA values were converted to 0

```{r}
table(mar_rgn_gf$gap_0_fill)
## 422 of these out of 2448+442 cases had NA converted to 0 - v2021 - seems reasonable given the added year of production
## 460 out of 2500+460 had NA converted to 0 - v2022
## 419 out of 2543+419 had NA converted to 0 - v2023
## 524 out of 2781+514 had NA converted to 0 - v2024 ≈ 15.9%
## 463 out of 2740 had NA converted to 0 - v2025 ~ 16.9%
```

Remove species-region-environment time series with less than four years of seaweed mariculture production \> 0 tonnes (assume these are not established seaweed mariculture programs).

```{r}

mar_rgn_gf <- mar_rgn_gf %>% 
  group_by(rgn_id, species, fao, environment) %>%
  mutate(not_0 = length(value[value > 0])) %>% # length of vector of years greater than 0
  filter(not_0 > 3) %>% # filter for groups that have at least four years of seaweed mariculture production 
  ungroup() %>% 
  dplyr::select(rgn_id, species, fao, environment, year, include, value, Taxon_code, gap_0_fill) 

```

Add a unique identifier per cultivated stock that describes each species, fao region, and environment grouping.

```{r}
# Add a unique identifier per cultivated stock
identifier <- mar_rgn_gf %>% 
  dplyr::select(rgn_id, species, fao, environment) %>% 
  unique() %>% 
  mutate(species_code = 1:n())
max(identifier$species_code)
# 86 unique identifiers - v2025

mar_rgn_gf <- left_join(mar_rgn_gf, identifier)
maric <- mar_rgn_gf

# Sanity check
colSums(is.na(maric))
```

# Calculate and save tonnes of seaweed

Find the tonnes per each region/year per each seaweed type (multiplied by "include" proportions).

Used to estimate total seaweed mariculture yield per country.

```{r}
# Multiply "include" column by "value" column to find tonnes per region/year for each seaweed species
maric <- maric %>% 
  mutate(tonnes = include * value)

# Save for weighting purposes later on
write_csv(maric, here(current_np_dir, "int", "np_seaweeds_tonnes_weighting.csv"))
```

----

# Seafood Watch sustainability

Below is the old code for reading in and wrangling Seafood Watch data. Unfortunately, they only have a single value (0.67) for seaweed sustainability, so we'll just be mass applying that. 

----

## Import data: Seafood Watch sustainability scores

These data describe the sustainability country/species combinations. In cases where these data were not available for a specific county/species, we just used the seafood watch seaweed sustainability score (6.72) (this was all of the seaweed species listed).

```{r sw-scores}
# ---- Load in Seafood Watch sustainability scores data from Mazu ----
# Go look at the file path, the file name changes from year to year
sw_sus <- read.csv(file.path(dir_M, paste0('git-annex/globalprep/_raw_data/seafood_watch_mar_sustainability/d',
                                           #scen_year, 
                                           "2025",
                                           
                                           '/SFW_Aquaculture_ratings_062725.csv')), # update this file name
                   check.names = FALSE, stringsAsFactors = FALSE, na.strings = c("NA", ""))
head(sw_sus)
```

## Wrangle

### Tidy Seafood Watch sustainability data

Rename columns to match with MAR data and fill in species column

```{r tidy_sw-sus}
# ---- Rename columns ----
sw_sus <- sw_sus %>%
  rename(report_title = 'ReportTitle',
         published_date = 'PublishedDate',
         sw_species = 'CommonNames',
         fao_species = 'FAOCommonName',
         fda_species = 'FDACommonName',
         water_body = 'BOWs',
         country = 'Countries',
          scientific_name = 'ScientificName',
         method = 'Methods',
         score = 'AssessmentScore', 
         escapes_score = 'C6Score',
         rec = 'AssessmentColor' 
         ) %>% 
  dplyr::select(report_title, published_date, sw_species, scientific_name, fao_species, fda_species, country, water_body, method, escapes_score, score, rec)

# ---- Change species names ----
# using FAO species name (fao_species); if NA, use common name (sw_species)
sw_sus$species <- ifelse(!is.na(sw_sus$fao_species), sw_sus$fao_species, sw_sus$sw_species)

```

### Keep NA countries

```{r keep-na-countries_sw-sus}
# These need to be re-added later (get cut when associated with region ids)
sw_sus_no_rgn <- filter(sw_sus, is.na(country)|country == "Worldwide")
nrow(sw_sus_no_rgn)
# 115 entries in v2024
# 114 entries in v2025
```

### Convert country names to OHI region IDs.

```{r sw-sus_convert-region}
# ---- Update country names ----
# Change country names to match OHI region names
sw_sus_multiple <- sw_sus %>% 
  filter(str_detect(country, "\\|")) %>%
  separate_rows(country, sep = " \\| ")

sw_sus_df <- sw_sus %>%
  filter(!str_detect(country, "\\|")) %>%
  rbind(sw_sus_multiple) %>%
  filter(!is.na(country), country!= "Worldwide")
  

# Convert country names to OHI region IDs. 
sw_sus_rgn <- name_2_rgn(df_in = sw_sus_df, 
                       fld_name = 'country', 
                       flds_unique = c('fao_species', 'fda_species', 'sw_species', 'score'),
                       keep_fld_name = TRUE) # 'country' now shows the original Seafood Watch data name; 'rgn_name' is what we want to use from now on

# duplicates found, checked out 
# v2025 duplicates found:
# 1 United States 
# 2 Chile         
# 3 Indonesia     
# 4 Norway        
# 5 United Kingdom
# 6 Canada 
# 7 Chine
# after investigating the data further, these duplicates don't seem to be an issue -- each entry is an unique species and/or based on an unique report

# Re-add NA countries
sw_sus_rgn <- bind_rows(sw_sus_rgn, sw_sus_no_rgn) %>%
  unique()
nrow(sw_sus_rgn)
summary(sw_sus_rgn)
  # 254 obs - v2021
  # 273 obs - v2022 
  # 269 obs - v2023
  # 269 obs - v2024
  # 277 obs - v2025
```

**Join the seaweed sustainability data with the mariculture data**

```{r sw-fao-joining}
maric <- readr::read_csv(here(current_np_dir, "int", "np_seaweeds_tonnes_weighting.csv"))

maric <- maric %>%
  group_by(environment, species, year, Taxon_code, rgn_id) %>% 
  summarize(value = sum(value), gap_0_fill = first(gap_0_fill)) %>%
  ungroup()

# run this code to check that there are no duplicates for any species, year, rgn_id combinations. If there are they will be unintentionally deleted later, so you may need to change family or taxon codes to match and rerun the above group by
duplicate_check <- maric %>% group_by(environment, species, year, rgn_id) %>% 
  summarize(n = n())

unique(duplicate_check$n) # should be one

# ---- Add a unique identifier per cultivated stock ----
identifier <- maric %>% 
  dplyr::select(rgn_id, species, environment) %>% 
  unique() %>% 
  mutate(species_code = 1:n())
max(identifier$species_code)
# 93 unique identifiers - v2021
# 96 unique identifiers - v2023
# 105 unique identifiers - v2024
# 86 unique identifiers - v2025


# add the unique identifier back to the dataset
mar_rgn_gf <- left_join(maric, identifier) 

maric <- mar_rgn_gf
setdiff(mar_rgn_gf$species, sw_sus_rgn$species)


# don't be alarmed by the score column being populated by all NAs after this step!
# we'll fill with a set value in the next step
mar_sw_sus_join <- maric %>%
  left_join(sw_sus_rgn, # has score column
            by = c("species", "rgn_id")) %>%
  dplyr::select(rgn_id, rgn_name, year, species, Taxon_code, score, value, gap_0_fill, species_code) %>% ## none of the specific species match
  rename(tonnes = value)
summary(mar_sw_sus_join)
colSums(is.na(mar_sw_sus_join))
```

**Since there are no sustainability scores for any of the species listed, we will gapfill with the seafood watch "Seaweed (Global)" score, which is 6.72.**


```{r}
# "calculate" (define) sustainability score for all seaweed 
mar_sw_sus_calc <- mar_sw_sus_join %>%
  mutate(sust = round(6.72 / 10,2)) %>% ## since none of the species match, we will give the general worldwide seaweed score from seafood watch (6.72)
  dplyr::select(-score)

# check: sust (sustainability score) should be 0.67 for all
unique(mar_sw_sus_calc$sust)
```

Since some regions have multiple sustainability scores for the same species due to multiple aquaculture methods, but we don't know what proportions of which methods are used, we take the average of the sustainability scores in these instances.

Average sustainability scores within regions with more than score (due to more than one aquaculture method):

```{r sw-sus-avg}
# aggregation: average sustainability per species per region
mar_sw_sus_avg <- mar_sw_sus_calc %>% 
  dplyr::group_by(rgn_id, species) %>% 
  dplyr::mutate(sust_avg = mean(sust, na.rm = TRUE)) %>% 
  dplyr::ungroup()
```

Get rid of duplicates for region/species/year:

```{r sw-sus-dup}

mar_sw_sus <- mar_sw_sus_avg %>% 
  # keep only unique rows from the data frame
  dplyr::distinct(rgn_id, species, year,
                  .keep_all = TRUE) %>% # keep all variables in .data. If a combination of the variables (rgn_id, species, year) is not distinct, this keeps the first row of values.
  dplyr::select(-sust, sust_coeff = sust_avg, taxon_group = Taxon_code) %>%
  dplyr::mutate(taxa_code = paste(species, species_code, sep = "_"))
```

**Now look at a summary after appending all the Seafood Watch data**

```{r sw-sus-summary}
summary(mar_sw_sus)
colSums(is.na(mar_sw_sus))
# No NAs in sust! (sust_coeff) 
```


## Save Data

Save intermediates and document gapfilling

```{r}
# Save seaweed mariculture sustainability dataset ----
seaweed_sust <- mar_sw_sus %>%
  dplyr::select(rgn_id, taxa_code, year, sust_coeff)

readr::write_csv(seaweed_sust, here(current_np_dir, "int", "np_seaweed_sust.csv"))

# Save seaweed mariculture harvest tonnes data ----
## ("tonnes" column already incorporated include proportions)
seaweed_harvest_tonnes <- mar_sw_sus %>%
  dplyr::select(rgn_id, taxa_code, year, tonnes)

anyDuplicated(seaweed_harvest_tonnes) # check for duplication
#> [1] 0

readr::write_csv(seaweed_harvest_tonnes, here(current_np_dir, "int", "np_seaweed_harvest_tonnes.csv"))

# Save a gapfilled dataset for FAO tonnes data ----
mar_FAO_gf <- mar_sw_sus %>% 
  rename("gapfill_fao" = "gap_0_fill") %>%
  mutate(method = ifelse(gapfill_fao == 0, "none", gapfill_fao), 
         gapfilled = ifelse(gapfill_fao == 0, 0, 1)) %>%
  dplyr::select(rgn_id, taxa_code, year, gapfilled, method)
 
readr::write_csv(mar_FAO_gf, here(current_np_dir, "output", "np_seaweed_harvest_tonnes_gf.csv"))


# Save a gapfilled dataset for mar sustainability dataset ----
mar_sust_gf <- mar_sw_sus %>%
  mutate(method = "sfw_seaweed_score",
         gapfilled = 1) %>%
  dplyr::select(rgn_id, year, taxa_code, gapfilled, method)
 
readr::write_csv(mar_sust_gf, here(current_np_dir, "output", "np_seaweed_sust_gf.csv"))
```


# Calculate Sustainable Tonnes

For each region and year, calculate total sustainable tonnes (of non-food seaweed harvested):

`tonnes * sustainability`


```{r}
# ==== Calculate sustainability ====

# filter data to 2004 onwards ----
mar_sw_sus_filter <- mar_sw_sus %>% 
  dplyr::filter(year >= 2004)

# Quick check
begin_data_year <- min(mar_sw_sus_filter$year)
begin_data_year
end_data_year <- max(mar_sw_sus_filter$year)
end_data_year


# Group by region and year to get summary (total tonnes) ----
mar_seaweed_total_tonnes <- mar_sw_sus_filter %>% 
  dplyr::group_by(rgn_id, year) %>% 
  summarise(tonnes = sum(tonnes, na.rm = TRUE)) %>% 
  ungroup() %>% 
  dplyr::arrange(rgn_id, year)

summary(mar_seaweed_total_tonnes)
#head(mar_seaweed_total_tonnes)


# Multiply total tonnes by 0.67 (sustainability coefficient) ----
mar_sw_sust_tonnes <- mar_seaweed_total_tonnes %>% 
  dplyr::mutate(sustainability = 0.67) %>% 
  dplyr::mutate(sust_tonnes = tonnes * sustainability)

summary(mar_sw_sust_tonnes)
#str(mar_sw_sust_tonnes)


# write out raw sust_tonnes to int
readr::write_csv(mar_sw_sust_tonnes, here(current_np_dir, "int", "seaweed_sust_tonnes_raw.csv"))

```




## Rolling average of sustainable tonnes


```{r}

# for each region and year, calculate rolling mean in a 5 year window (year + 4 years, right-aligned (year, look at the 4 years previous))
seaweed_st_rolling <- mar_sw_sust_tonnes %>%
  dplyr::ungroup() %>% 
  dplyr::arrange(rgn_id, year) %>%
  dplyr::group_by(rgn_id) %>%
  dplyr::mutate(
      sust_tonnes_rollmean = zoo::rollapply(sust_tonnes, width = 4,
                                            FUN = mean, align = 'right',
                                            partial = TRUE, na.rm = FALSE)) %>%
  dplyr::rename(sust_tonnes_orig = sust_tonnes) %>% 
  dplyr::mutate(sust_tonnes = ifelse(!is.na(sust_tonnes_rollmean), sust_tonnes_rollmean, sust_tonnes_orig)) %>%
  dplyr::select(rgn_id, year, tonnes, sustainability, sust_tonnes, sust_tonnes_orig)

summary(seaweed_st_rolling)
```



## Peak sustainable tonnes

Identify max value from rolling average of sustainable tonnes (maybe limit to most recent 10 years of data to ID max). 


```{r}
# define decade range for max value 

end_data_year <- max(mar_sw_sus_filter$year)
seaweed_decade <- c((end_data_year - 10):end_data_year)
seaweed_decade

# find peak sustainable tonnes 
seaweed_st_max <- seaweed_st_rolling %>% 
  filter(year %in% c(seaweed_decade)) %>% 
  ungroup() %>% 
  group_by(rgn_id) %>% 
  mutate(peak_sust_tonnes = max(sust_tonnes, na.rm = TRUE)) %>% 
  ungroup() %>% 
  # keep first value for each 
  # (it's copied a few times otherwise which causes issues down the line)
  dplyr::distinct(rgn_id, peak_sust_tonnes) %>% 
  dplyr::select(rgn_id, peak_sust_tonnes)


head(seaweed_st_max)
```


## Relative sustainable tonnes

Bring max sustainable tonnes value back into sustainable tonnes full dataframe. Use this value to calculate relative sustainable tonnes: if `sust_tonnes > peak_sust_tonnes`, then set to 1. Else, `tonnes_rollmean / peak_sust_tonnes`.  


```{r}
# ---- Determine relative sustainable tonnes ----
seaweed_relative_sust_tonnes <- 
  # join peak sustainable tonnes with full sustainable tonnes df
  seaweed_st_rolling %>% 
  ungroup() %>% 
  ## filling in peak value per region ID across full year range
  full_join(seaweed_st_max, by = join_by("rgn_id")) %>% 
  # rescale to find relative sustainable tonnes
  mutate(rel_sust_tonnes = ifelse(sust_tonnes > peak_sust_tonnes, 1,
                             sust_tonnes / peak_sust_tonnes)) %>% 
  mutate(product = "seaweeds")
 
# check
summary(seaweed_relative_sust_tonnes)


# save full df to int folder
readr::write_csv(seaweed_relative_sust_tonnes, here(current_np_dir, "int", "np_relative_sust_tonnes_seaweeds_full.csv"))


# save selected columns to output folder
final_sw_relative_sust_tonnes <- seaweed_relative_sust_tonnes %>% 
  dplyr::select(rgn_id, product, year, rel_sust_tonnes)

readr::write_csv(final_sw_relative_sust_tonnes, here(current_np_dir, "output", "np_relative_sust_tonnes_seaweeds.csv"))

```

# Data check
Compare seaweed sustainability from previous years

```{r}
# Read in previous assessment data
old_seaweed <- read_csv(here("globalprep", "np", paste0(prev_v_scen_year, "_new"), "int", "np_relative_sust_tonnes_seaweeds_full.csv"))

# Read in current assessment data
new_seaweed <- read_csv(here("globalprep", "np", paste0(v_scen_year, "_new"), "int", "np_relative_sust_tonnes_seaweeds_full.csv"))
max(new_seaweed$year)

previous_data_year <- 2022
current_data_year <- 2023

# ---- Compare ornamental sust data for all countries: ----
# previous data year (2022) 
old_sw1 <- old_seaweed %>% 
  filter(year == previous_data_year) %>% 
  dplyr::select(rgn_id, product, old_rel_sust_tonnes = rel_sust_tonnes, old_sust = sustainability, old_tonnes = tonnes, old_sust_tonnes = sust_tonnes, old_peak = peak_sust_tonnes)

new_sw1 <- new_seaweed %>% 
  filter(year == previous_data_year) %>% 
  dplyr::select(rgn_id, product, rel_sust_tonnes, sustainability, tonnes, sust_tonnes, peak_sust_tonnes)

sw_total1 <- old_sw1 %>% 
  full_join(new_sw1, by = c("rgn_id", "product"))

# Join region labels for easier viewing
sw_total1 <- sw_total1 %>% 
  left_join(rgn_labels, by = "rgn_id") %>% 
  select(-type, rgn_name = label)

###### Sustainable Tonnes ######
comp_plot1 <- ggplot() + 
  geom_jitter(data = sw_total1, aes(x = old_sust_tonnes, y = sust_tonnes, text = paste("Product:", product, "<br>Region:", rgn_name)), alpha = 0.7) +
  labs(x = paste("Previous Value", previous_data_year), 
       y = paste("Current Value", previous_data_year),
       title = paste("Sustainable tonnes of seaweed", previous_data_year, "Data", prev_scen_year, "Assessment vs.", previous_data_year, "Data", scen_year_number, "assessment")) +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  theme_bw();ggplotly(comp_plot1)

###### Relative Sustainable tonnes ######
comp_plot2 <- ggplot() + 
  geom_jitter(data = sw_total1, aes(x = old_rel_sust_tonnes, y = rel_sust_tonnes, text = paste("Product:", product, "<br>Region:", rgn_name)), alpha = 0.7) +
  labs(x = paste("Previous Value", previous_data_year), 
       y = paste("Current Value", previous_data_year),
       title = paste("Relative sustainable tonnes of seaweed", previous_data_year, "Data", prev_scen_year, "Assessment vs.", previous_data_year, "Data", scen_year_number, "assessment")) +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  theme_bw();ggplotly(comp_plot2)

###### Harvest Tonnes ######
comp_plot3 <- ggplot() + 
  geom_jitter(data = sw_total1, aes(x = old_tonnes, y = tonnes, text = paste("Product:", product, "<br>Region:", rgn_name)), alpha = 0.7) +
  labs(x = paste("Previous Value", previous_data_year), 
       y = paste("Current Value", previous_data_year),
       title = paste("Seaweed harvest tonnes", previous_data_year, "Data", prev_scen_year, "Assessment vs.", previous_data_year, "Data", scen_year_number, "assessment")) +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  theme_bw();ggplotly(comp_plot3)




# ---- Compare relative seaweed sust data for all countries: ----
# previous data year (2022) vs. current data year (2023) 
old_sw2 <- old_seaweed %>% 
  filter(year == previous_data_year) %>% 
  dplyr::select(rgn_id, product, old_rel_sust_tonnes = rel_sust_tonnes, old_sust = sustainability, old_tonnes = tonnes, old_sust_tonnes = sust_tonnes, old_peak = peak_sust_tonnes)

new_sw2 <- new_seaweed %>% 
  filter(year == current_data_year) %>% 
  dplyr::select(rgn_id, product, rel_sust_tonnes, sustainability, tonnes, sust_tonnes, peak_sust_tonnes)

sw_total2 <- old_sw2 %>% 
  full_join(new_sw2, by = c("rgn_id", "product"))

# Join region labels for easier viewing
sw_total2 <- sw_total2 %>% 
  left_join(rgn_labels, by = "rgn_id") %>% 
  select(-type, rgn_name = label)

###### Sustainable Tonnes ######
comp_plot5 <- ggplot() + 
  geom_jitter(data = sw_total2, aes(x = old_sust_tonnes, y = sust_tonnes, text = paste("Product:", product, "<br>Region:", rgn_name)), alpha = 0.7) +
  labs(x = paste("Previous Value", previous_data_year), 
       y = paste("Current Value", current_data_year),
       title = paste("Sustainable tonnes of seaweed", previous_data_year, "Data", prev_scen_year, "Assessment vs.", current_data_year, "Data", scen_year_number, "assessment")) +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  theme_bw();ggplotly(comp_plot5)

###### Relative Sustainable tonnes ######
comp_plot6 <- ggplot() + 
  geom_jitter(data = sw_total2, aes(x = old_rel_sust_tonnes, y = rel_sust_tonnes, text = paste("Product:", product, "<br>Region:", rgn_name)), alpha = 0.7) +
  labs(x = paste("Previous Value", previous_data_year), 
       y = paste("Current Value", current_data_year),
       title = paste("Relative sustainable tonnes of seaweed", previous_data_year, "Data", prev_scen_year, "Assessment vs.", current_data_year, "Data", scen_year_number, "assessment")) +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  theme_bw();ggplotly(comp_plot6)

colnames(sw_total2)

###### Harvest Tonnes ######
comp_plot7 <- ggplot() + 
  geom_jitter(data = sw_total2, aes(x = old_tonnes, y = tonnes, text = paste("Product:", product, "<br>Region:", rgn_name)), alpha = 0.7) +
  labs(x = paste("Previous Value", previous_data_year), 
       y = paste("Current Value", current_data_year),
       title = paste("Seaweed harvest tonnes", previous_data_year, "Data", prev_scen_year, "Assessment vs.", current_data_year, "Data", scen_year_number, "assessment")) +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  theme_bw();ggplotly(comp_plot7)


```

