---
title: "OHI `r format(Sys.Date(), '%Y')`: Mariculture"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---
# Summary

This analysis converts FAO mariculture and seafood watch sustainability data into data used to calculate the OHI global mariculture status score. This also calculates the genetic escapee from mariculture pressure data.

# 2025 updates from 2024 assessment
- We got updated Seafood Watch data from the Monterey Bay Aquarium in 2025. Because of this, we wanted to combine the SFW data from 2024 and 2025 together as a way to gapfill and as a way to change the data year in the `scenario_data_years.csv` for the OHI-global score checking. Gapfilling by a different data year occurs BEFORE all of the other gapfilling methods and has not been documented in the gapfill csv yet. This is a change that future assessments could add.


## Setup 

```{r setup, message=FALSE, warning=FALSE}

knitr::opts_chunk$set(eval = FALSE)
  
# ---- Load packages ----
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
 librarian::shelf(
  ohicore,  #devtools::install_github('ohi-science/ohicore@dev')
  dplyr,
  tidyverse, # loads readr, stringr, tidyr, etc.
  tictoc,
  knitr,
  kableExtra,
  here,
  plotly
)

#library(taxize) had issues installing taxize because its dependency phanghorn is not compatible with this version of R


# ---- Set directories & source functions ----
current_year <- 2025 # update these!
previous_year <-  2024 # update these!
version_year <- paste0("v", current_year)
old_version_year <- paste0("v", previous_year)
current_mar_dir <- here::here("globalprep", "mar", version_year)
previous_mar_dir <- here::here("globalprep", "mar", old_version_year)


# ---- Load FAO-specific user-defined functions ----
# functions specific to mariculture dealing with compound countries, updated in 2024
source(here::here("globalprep", "mar", version_year, "mar_fxs.R"))
# functions for cleaning fao data
source(here::here("workflow", "R", "fao_fxn.R")) 
#source(here::here("workflow", "R", "fao_online_portal_clean.R")) # fao cleaning function # switched to new version in 2024
source(here::here("workflow", "R", "fao_query_data_tidy.R")) # 2024 version of FAO data tidying function 
source(here::here("workflow", "R", "common.R")) # directory locations, OHI essentials

# general raw data directory in Mazu
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")
# FAO data directory in Mazu
FAO_mariculture_dir <- here::here(raw_data_dir, "FAO_mariculture")
# Seafood Watch data directory in Mazu
seafood_watch_dir <- here::here(raw_data_dir, "seafood_watch_mar_sustainability")

```

# Import Raw Data: FAO Mariculture data

Mariculture production in tonnes.

```{r import-data}
## UPDATE THESE!
current_year <- current_year # update in previous chunk
last_year <- 2024 #last year of analysis

#last year of data available
#update if not two years behind the current assessment year
latest_data_year <- current_year - 2 # most up to date year in the data aquaculture data
previous_data_year <- last_year - 2 #last year of data in the last assessment, for comparison plots

current_data_folder <- paste0("d", current_year) # assessment year for file paths
previous_data_folder <- paste0("d", last_year)  # previous assessment year
```

## Data check mariculture and sustainability scores

Comparing this year's data to previous year's data. Expect small variation from year to year. Plot to view differences.

```{r data-check, eval = FALSE}
options(scipen = 999)

# ============== Mariculture Yield =============================================

# ---- Read in old and new data for plots ----
# mar_old_data <- readr::read_csv(here::here("globalprep", "mar", paste0( "v", last_year), "output", "MAR_FP_data.csv")) 
# mar_new_data <- readr::read_csv(here::here(current_mar_dir, "output", "MAR_FP_data.csv"))

# Comparing 2023 data to the combined data
mar_old_data <- readr::read_csv(here::here("globalprep", "mar", "v2023", "output", "MAR_FP_data.csv"))
mar_new_data <- readr::read_csv(here::here(current_mar_dir, "output", "MAR_FP_data_comb.csv"))

previous_data_year <- 2021
latest_data_year

# ---- Compare mariculture yield data for all countries: ----
# previous data year (2022) vs. current data year (2023) 
mar_old_1 <- mar_old_data %>% 
  filter(year == previous_data_year) %>% 
  dplyr::select(rgn_id, species, old_value = value, environment)

mar_new_1 <- mar_new_data %>% 
  filter(year == latest_data_year) %>% 
  dplyr::select(rgn_id, species, value, environment)


yield_1 <- mar_old_1 %>% 
  full_join(mar_new_1, by = c("rgn_id", "species", "environment")); View(yield_1)

# Plot
comp_plot_1 <- ggplot() + 
  geom_point(data = yield_1, aes(x = old_value, y = value, text = paste("species:", species, "<br>region ID:", rgn_id))) +
  labs(x = paste("Previous Value", previous_data_year),
       y = paste("Current Value", latest_data_year),
       title = paste0(current_year, " (",latest_data_year, ") vs. ", last_year, " (",previous_data_year, ")", " All Countries")) +
  geom_abline(slope = 1, intercept = 0, col = "red") 

ggplotly(comp_plot_1)


# ---- Compare mariculture yield data for all countries: ----
# 2022 last year to 2022 this year (previous_data_year)
mar_old_2 <- mar_old_data %>% 
  filter(year == previous_data_year) %>% 
  dplyr::select(rgn_id, species, old_value = value, environment, fao)

mar_new_2 <- mar_new_data %>% 
  filter(year == previous_data_year) %>% 
  dplyr::select(rgn_id, species, value, environment, fao)

yield_2 <- mar_old_2 %>% 
  full_join(mar_new_2, by = c("rgn_id","species", "environment", "fao")); View(yield_2)


comp_plot_2 <- ggplot() + 
  geom_point(data = yield_2, aes(x = old_value, y = value, text = paste("Species:", species, "<br>Region ID:", rgn_id))) +
  labs(x = paste("Previous Value", previous_data_year), 
       y = paste("Current Value", previous_data_year),
       title = paste(previous_data_year, "data", last_year, "assessment vs.", previous_data_year, "data", current_year, "assessment")) +
  geom_abline(slope = 1, intercept = 0, col = "red")


ggplotly(comp_plot_2)


# ---- Compare mariculture yield data for Russia: ----
# 2022 data from 2024 assessment to 2023 data from 2025 assessment
mar_old_3 <- mar_old_data %>% 
  filter(rgn_id == 73, year == previous_data_year) %>% 
  dplyr::select(rgn_id, species, fao, old_value = value, environment)

mar_new_3 <- mar_new_data %>% 
  filter(rgn_id == 73, year == latest_data_year) %>% 
  dplyr::select(rgn_id, species, fao, value, environment)

yield_3 <- mar_old_3 %>% 
  full_join(mar_new_3, by = c("rgn_id","species","fao"))
View(yield_3)

# new plot code
comp_plot_3 <- ggplot() + 
  geom_point(data = yield_3,
             aes(x = old_value, y = value,
                 text = paste("species:", species,
                              "<br>region ID:", rgn_id))) +
  labs(x = paste("Previous Value", previous_data_year),
       y = paste("Current Value", latest_data_year),
       title = "Russia Mariculture Yield Data Comparison") +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  theme_bw()

ggplotly(comp_plot_3)


# ---- Compare mariculture yield data for Iceland: ----
# 2022 data from 2024 assessment to 2023 data from 2025 assessment
# one Atlantic Salmon observation was doubled (copied) in 2022 so I added "environment" which resolved the issue (Brackish vs Marine issue)
mar_old_4 <- mar_old_data %>% 
  filter(rgn_id == 143, year == previous_data_year) %>% 
  dplyr::select(rgn_id, species, fao, old_value = value, environment) %>% 
  ungroup()

mar_new_4 <- mar_new_data %>% 
  filter(rgn_id == 143, year == latest_data_year) %>% 
  dplyr::select(rgn_id, species, fao, value, environment) %>% 
  ungroup()

yield_4 <- mar_old_4 %>% 
  full_join(mar_new_4, by = c("rgn_id","species","fao", "environment"))
View(yield_4)

# new plot code
comp_plot_4 <- ggplot() + 
  geom_point(data = yield_4,
             aes(x = old_value, y = value,
                 text = paste("species:", species,
                              "<br>region ID:", rgn_id))) +
  labs(x = paste("Previous Value", previous_data_year),
       y = paste("Current Value", latest_data_year),
       title = "Iceland Mariculture Yield Data Comparison") +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  theme_bw()

ggplotly(comp_plot_4)


# ---- Compare mariculture yield data for Vietnam ----
mar_old_5 <- mar_old_data %>% 
  filter(rgn_id == 207, year == previous_data_year) %>% 
  dplyr::select(rgn_id, species, fao, environment, old_value = value)
mar_new_5 <- mar_new_data %>% 
  filter(rgn_id == 207, year == latest_data_year) %>% 
  dplyr::select(rgn_id, species, fao, environment, value)

yield_5 <- mar_old_5 %>% 
  full_join(mar_new_5, by = c("rgn_id","species","fao", "environment"))
View(yield_5)

comp_plot_5 <- ggplot() + 
  geom_point(data = yield_5,
             aes(x = old_value, y = value,
                 text = paste("species:", species, "<br>region ID:", rgn_id))) +
  labs(x = paste("Previous Value", previous_data_year),
       y = paste("Current Value", latest_data_year),
       title = "Vietnam Mariculture Yield Data Comparison") +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  theme_bw()

ggplotly(comp_plot_5)


# ---- Compare yield data for Belize; ----
# yield of Whiteleg shrimp was halved between 2019 and 2020 assessment years.
# Whiteleg shrimp doubled between 2021 and 2022
mar_old_6 <- mar_old_data %>% 
  filter(rgn_id == 164, year == previous_data_year) %>% 
  dplyr::select(rgn_id, species, fao, old_value = value)
mar_new_6 <- mar_new_data %>% 
  filter(rgn_id == 164, year == latest_data_year) %>% 
  dplyr::select(rgn_id, species, fao, value)

yield_6 <- mar_old_6 %>% 
  full_join(mar_new_6, by = c("rgn_id","species","fao"))
View(yield_6)

comp_plot_6 <- ggplot() + 
  geom_point(data = yield_6,
             aes(x = old_value, y = value,
                 text = paste("species:", species, "<br>region ID:", rgn_id))) +
  labs(x = paste("Previous Value", previous_data_year),
       y = paste("Current Value", latest_data_year),
       title = "Belize Mariculture Yield Data Comparison") +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  theme_bw()

ggplotly(comp_plot_6)


# ---- Compare yield data for Tanzania ----
mar_old_7 <- mar_old_data %>% 
  filter(rgn_id == 202, year == previous_data_year) %>% 
  dplyr::select(rgn_id, species, fao, old_value = value, environment)
mar_new_7 <- mar_new_data %>% 
  filter(rgn_id == 202, year == previous_data_year) %>% 
  dplyr::select(rgn_id, species, fao, value, environment)

yield_7 <- mar_old_7 %>% 
  full_join(mar_new_7, by = c("rgn_id","species","fao", "environment"))

comp_plot_7 <- ggplot() + 
  geom_point(data = yield_7,
             aes(x = old_value, y = value,
                 text = paste("species:", species, "<br>region ID:", rgn_id))) +
  labs(x = paste("Previous Value", previous_data_year),
       y = paste("Current Value", latest_data_year),
       title = "Tanzania Mariculture Yield Data Comparison") +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  theme_bw()

ggplotly(comp_plot_7)


# ---- Compare yield data for Denmark ----
# pressure score decreased (not sure which version this comment was about)
mar_old_8 <- mar_old_data %>% 
  filter(rgn_id == 175, year == previous_data_year) %>% 
  dplyr::select(rgn_id, species, fao, old_value = value, environment)
mar_new_8 <- mar_new_data %>% 
  filter(rgn_id == 175, year == latest_data_year) %>% 
  dplyr::select(rgn_id, species, fao, value, environment)

yield_8 <- mar_old_8 %>% 
  full_join(mar_new_8, by = c("rgn_id","species","fao", "environment"))

comp_plot_8 <- ggplot() + 
  geom_point(data = yield_8,
             aes(x = old_value, y = value,
                 text = paste("species:", species, "<br>region ID:", rgn_id))) +
  labs(x = paste("Previous Value", previous_data_year),
       y = paste("Current Value", latest_data_year),
       title = "Denmark Mariculture Yield Data Comparison") +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  theme_bw()

ggplotly(comp_plot_8)


# ---- Compare yield data for Ecuador ----
# score increased by a lot (again, not sure which version this comment is from)
# v2024: (2022 data) whiteleg shrimp yield has notably increased 
mar_old_9 <- mar_old_data %>% 
  filter(rgn_id == 137, year == previous_data_year) %>% 
  dplyr::select(rgn_id, species, fao, old_value = value, environment)

mar_new_9 <- mar_new_data %>% 
  filter(rgn_id == 137, year == latest_data_year) %>% 
  dplyr::select(rgn_id, species, fao, value, environment)

yield_9 <- mar_old_9 %>% 
  full_join(mar_new_9, by = c("rgn_id","species","fao"))
View(yield_9)

comp_plot_9 <- ggplot() + 
  geom_point(data = yield_9,
             aes(x = old_value, y = value,
                 text = paste("species:", species, "<br>region ID:", rgn_id))) +
  labs(x = paste("Previous Value", previous_data_year),
       y = paste("Current Value", latest_data_year),
       title = "Ecuador Mariculture Yield Data Comparison") +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  theme_bw()

ggplotly(comp_plot_9)

```

## Data checking for score check error 2025 assessment

```{r}
previous_data_year <- 2024

# Read in the final outputted csv
sust_old <- readr::read_csv(here::here("globalprep", "mar", paste0("v", previous_data_year),
                                       "output", "mar_sustainability.csv")) %>% 
  dplyr::select(rgn_id, taxa_code, sust_old = sust_coeff, year)

sust_new <- readr::read_csv(here::here(current_mar_dir, "output", "mar_sustainability.csv")) %>% 
  dplyr::select(rgn_id, taxa_code, sust_new = sust_coeff, year)

# tidy up species names (such as "Abalones_1"), rename as species column
sust_old$species <- str_remove(sust_old$taxa_code, "\\_[^.]*$") 
sust_new$species <- str_remove(sust_new$taxa_code, "\\_[^.]*$") 

view(sust_old) 
view(sust_new) 

sust_new %>% 
  filter(rgn_id == 182 & year == 2024) %>% 
  view()

# Filter down to just taiwan (country with largest decrease in score) for checking
old_taiwan <- sust_old %>% 
  filter(rgn_id == 182)

new_taiwan <- sust_new %>% 
  filter(rgn_id == 182)

view(old_taiwan)
view(new_taiwan)

nrow(old_taiwan) # 59
nrow(new_taiwan) # 62

setdiff(new_taiwan$sust_new, old_taiwan$sust_old) # rows in the new data that aren't in the old data
# [1] 0.90909091 0.88311688 0.70129870 0.81818182
# [5] 0.87012987 0.76623377 0.09090909 0.61038961

setdiff(new_taiwan$species, old_taiwan$species) # rows in the new data that aren't in the old data
# [1] "Largemouth black bass" "Crimson seabream"     
setdiff(old_taiwan$species, new_taiwan$species) # rows in the old data that aren't in the new data
# character(0)

sust_new24 <- sust_new %>% 
  filter(year == 2024)

# Compare across the entire dataset
sust_total <- full_join(sust_new24, sust_old, by = c("rgn_id", "species", "year"))

sust_2024 <- sust_total %>% 
  filter(year == 2024)

view(sust_total)

# Compare sust_coeff between 2025 and 2024
diff_plot <- ggplot() +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  geom_jitter(data = sust_total, aes(x = sust_new, y = sust_old, text = paste("Region ID:", rgn_id, "<br>", "Species:", species)), alpha = 0.6) +
  labs(title = "Sustainability coefficient 2025 vs. 2024",
       x = "Sust Coeff 2025",
       y = "Sust Coeff 2024") +
  theme_bw();plotly::ggplotly(diff_plot)

# Compare for just taiwan
total_spain <- sust_total %>% 
  filter(rgn_id == 182)

nrow(total_spain) # 100

spain_plot <- ggplot() +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  # Added jitter and alpha to help with the overlap
  geom_jitter(data = total_spain, aes(y = sust_new, x = sust_old, text = paste("Region ID:", rgn_id, "<br>", "Species:", species)), alpha = 0.6) +
  labs(title = "Sustainability Spain 2025 vs. 2024",
       x = "Sust Coeff 2025",
       y = "Sust Coeff 2024") +
  theme_bw();plotly::ggplotly(spain_plot)

```

```{r}
# Changing the data years to account for the new SFW data
latest_data_year <- 2025
previous_data_year <- 2024

# Compare new/old Mariculture sustainability scores
sust_old <- read_csv(here("globalprep", "mar", paste0("v", previous_data_year), "output", "mar_sustainability.csv")) %>%
  dplyr::select(rgn_id, taxa_code_old = taxa_code, sust_old = sust_coeff)

# tidy up species names (such as "Abalones_1"), rename as species column
sust_old$species <- str_remove(sust_old$taxa_code_old, "\\_[^.]*$")

# Drop the unneeded taxa code
sust_old <- sust_old %>%
  mutate(year_old = 2024)

mar_sw_sus_final <- mar_sw_sus_final %>%
  mutate(year = 2025)

# Join previous and current year data
sust_comb <- full_join(mar_sw_sus_final, sust_old, by = c("rgn_id", "species"))

# Pivot longer so we only have one year column
# Check NAs across entire df
colSums(is.na(sust_comb))

# Fill NAs in sust_old (previous year sustainability) with Sust from this year

# Don't fill year, just score.

sust_comb_clean <- sust_comb %>%
  mutate(sust_old = ifelse(is.na(sust_old), Sust, sust_old),
         Sust = ifelse(is.na(Sust), sust_old, Sust),
         year_old = ifelse(is.na(year_old), year, year_old),
         year = ifelse(is.na(year), year_old, year),
         taxa_code = ifelse(is.na(taxa_code), taxa_code_old, taxa_code),
         taxa_code_old = ifelse(is.na(taxa_code_old), taxa_code, taxa_code_old))

colSums(is.na(sust_comb_clean))

# Now that we've filled the NAs, reseparate the data frames so we can do a simple bindrows on them
sust_new_clean <- sust_comb_clean %>%
  select(rgn_id, species, taxa_code, year, sust = Sust)

sust_old_clean <- sust_comb_clean %>%
  select(rgn_id, species, taxa_code = taxa_code_old, year = year_old, sust = sust_old)

sust_clean <- bind_rows(sust_new_clean, sust_old_clean)

# # define the maximum sustainability score, check with:
summary(mar_sw_sus_final)

# Issue: the 2024 data was already scaled. Should I pull from the unscaled version of the 2024 data and recalculate score? Or, I can scale the 2025 data before I combine it with the 2024 data?
max_sus <- 0.77

# Save sustainability scores data for 2023 (SFW data year)
mar_sustainability_score <- sust_clean %>%
  mutate(year = current_year) %>% #change year if data not updated to current year
  dplyr::select(rgn_id, year, taxa_code, sust_coeff = sust) %>%
  unique() %>%
  mutate(sust_coeff = ifelse(sust_coeff > max_sus, 1, sust_coeff/max_sus)) # rescale to the highest sust score

# Check for any duplicated rows
sum(duplicated(mar_sustainability_score))

# Check for NAs in the data frame
colSums(is.na(mar_sustainability_score))

# Confirm that the right balance of years is present in the data frame
table(mar_sustainability_score$year)

view(mar_sustainability_score)

# Saves the appropriate Sustainability Score file
 write.csv(mar_sustainability_score, here(current_mar_dir, "output", "mar_sustainability.csv"), row.names = F)
readr::write_csv(mar_sustainability_score, here(current_mar_dir, "output", "mar_sustainability_new.csv"))
```


```{r sust-scores}
# ============= Sustainability scores ==========================================

# Changing the data years to account for the new SFW data
latest_data_year <- 2025
previous_data_year <- 2023

# Compare new/old Mariculture sustainability scores

sust_old <- readr::read_csv(here::here("globalprep", "mar", paste0("v", previous_data_year),
                                       "output", "mar_sustainability.csv")) %>% 
  dplyr::select(rgn_id, taxa_code, sust_old = sust_coeff)

# tidy up species names (such as "Abalones_1"), rename as species column
sust_old$species <- str_remove(sust_old$taxa_code, "\\_[^.]*$") 

# sust_new <- readr::read_csv(here::here(current_mar_dir, "int", "mar_sustainability_check_comb.csv")) %>% 
#   dplyr::select(rgn_id, species, sust_new = sust_coeff)
sust_new <- readr::read_csv(here::here(current_mar_dir, "output", "mar_sustainability.csv")) %>% 
  dplyr::select(rgn_id, taxa_code, sust_new = sust_coeff)

sust_compare <- sust_new %>% 
  full_join(sust_old, by = c("rgn_id", "species")) %>%
  mutate(difference = sust_new - sust_old) %>%
  group_by(rgn_id) %>%
  summarize(mean_rgn_sus_diff = mean(difference, na.rm = TRUE))

mean_diff_sus <- ggplot(sust_compare, aes(x = rgn_id, y = mean_rgn_sus_diff)) + 
  geom_point() +
  theme_bw() +
  labs(title = paste("Mariculture Sustainability: Mean Difference", latest_data_year, "Assessment - ", previous_data_year, "Assessment"),
       x = "Region ID",
       y = "Average region sustainability difference")

ggplotly(mean_diff_sus)

# Another way to compare sust scores, this is optional

sust_total <- full_join(sust_new, sust_old, by = c("rgn_id", "species"))

# Compare sust_coeff between 2025 and 2024
diff_plot <- ggplot() +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  geom_point(data = sust_total, aes(x = sust_new, y = sust_old, text = paste("Region ID:", rgn_id, "<br>", "Taxa Code:", taxa_code))) +
  labs(title = "Sustainability coefficient 2025 vs. 2024",
       x = "Sust Coeff 2025",
       y = "Sust Coeff 2024") +
  theme_bw();plotly::ggplotly(diff_plot)
```

## Data check genetic escapes

Comparing combined genetic escape data to uncombined data

```{r}
# Read in uncombined data
gen_esc <- readr::read_csv(here::here(current_mar_dir, "output", "GenEsc.csv"))

# Read in combined data
gen_esc_comb <- readr::read_csv(here::here(current_mar_dir, "output", "GenEsc_comb.csv"))

gen_esc24 <- gen_esc_comb %>% 
  filter(vyear == 2025) %>% 
  rename(prs_score_old = pressure_score) %>% 
  select(-vyear)

gen_compare <- full_join(gen_esc, gen_esc24, by = c("rgn_id", "year"))

# plot current data of previous and current year
comp_gen_years_plot <- ggplot() + 
  geom_point(data = gen_compare, aes(x = prs_score_old, y = pressure_score, text = paste("Region ID:", rgn_id))) +
  labs(x = paste("Previous Pressure", previous_data_year), y = paste("Current Pressure", latest_data_year)) +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  theme_bw();ggplotly(comp_gen_years_plot)

```

Comparing this year's data to previous year's data. Expect small variation from year to year. Plot to view differences.

```{r data-check-gen-esc}

# ==== Compare genetic escapes pressure score; ====
# saw large changes between 2018 and 2017 assessment years. No big changes between 2019 and 2018 assessment years. Big changes from 2019 to 2020 due to a data update. Small changes between 2020 and 2021 due to updating the FAO data, gapfilling caused this. Also fixed some name errors. Small changes between 2021 and 2022, added a few name conversions. Some changes between 2022 and 2023, fixed a handful of errors. 

# Changing the years back because the genetic escapes data only goes up until 2023 
latest_data_year <- 2023
previous_data_year <- 2022

current_gen_esc_data <- readr::read_csv(here::here(current_mar_dir, "output", "GenEsc.csv"))

gen_old <- current_gen_esc_data %>%
  filter(year == previous_data_year) %>% 
  dplyr::select(rgn_id, year, prs_score_old = pressure_score)

gen_new <- current_gen_esc_data %>%
  filter(year == latest_data_year) %>%
  dplyr::select(rgn_id, year, pressure_score)

gen_compare <- full_join(gen_old, gen_new, by = c("rgn_id"))

# plot current data of previous and current year
comp_gen_years_plot <- ggplot() + 
  geom_point(data = gen_compare, aes(x = prs_score_old, y = pressure_score, text = paste("Region ID:", rgn_id))) +
  labs(x = paste("Previous Pressure", previous_data_year), y = paste("Current Pressure", latest_data_year)) +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  theme_bw()

ggplotly(comp_gen_years_plot)


# ---- Compare old genetic escape pressure scores for the same year ----
old_gen_esc_data <- readr::read_csv(here::here("globalprep", "mar", "v2022",
                                  "output", "GenEsc.csv")) 

previous_data_year <- 2020

old_last_year <- old_gen_esc_data %>%
  filter(year == previous_data_year) %>% 
  select(rgn_id, year, prs_score_old = pressure_score)

new_last_year <- current_gen_esc_data %>% 
  filter(year == previous_data_year) %>%
  select(rgn_id, year, pressure_score)

compare_gen_data_years <- full_join(old_last_year, new_last_year, by = c("rgn_id"))

View(compare_gen_data_years)

comp_gen_data_years_plot <- ggplot() + 
  geom_point(data = compare_gen_data_years, aes(x = prs_score_old, y = pressure_score, text = paste("Region ID:", rgn_id))) +
  labs(x = paste("Previous Pressure", previous_data_year), y = paste("Current Pressure", previous_data_year)) +
  geom_abline(slope = 1, intercept = 0, col = "red") + theme_bw()


ggplotly(comp_gen_data_years_plot)



# Compare genetic escapes pressure scores for Vietnam, data was updated by FAO
old_gen_vietnam <- old_gen_esc_data %>%
  filter(rgn_id == 207) %>% 
  select(rgn_id, year, prs_score_old = pressure_score)

new_gen_vietnam <- current_gen_esc_data %>%
  filter(rgn_id == 207) %>% 
  select(rgn_id, year, pressure_score)

compare_gen_vietnam <- full_join(old_gen_vietnam, new_gen_vietnam, 
                                 by = c("rgn_id", "year"))

View(compare_gen_vietnam)

compare_gen_vietnam_plot <- ggplot() + 
  geom_point(data = compare_gen_vietnam, aes(x = prs_score_old, y = pressure_score, 
                                             text = paste("Region ID:", rgn_id))) +
  labs(x = paste("Previous Pressure", previous_data_year), 
       y = paste("Current Pressure", previous_data_year),
       title = "Vietnam MAR current and previous pressure") +
  geom_abline(slope = 1, intercept = 0, col = "red") + theme_bw()

ggplotly(compare_gen_vietnam_plot)

# Compare genetic escapes pressure scores for Denmark
old_gen_denmark <- old_gen_esc_data %>%
  filter(rgn_id == 175) %>% 
  select(rgn_id, year, prs_score_old = pressure_score)

new_gen_denmark <- current_gen_esc_data %>%
  filter(rgn_id == 175) %>% 
  select(rgn_id, year, pressure_score)

compare_gen_denmark <- full_join(old_gen_denmark, new_gen_denmark, 
                                 by = c("rgn_id","year"))

View(compare_gen_denmark)

compare_gen_denmark_plot <- ggplot() + 
  geom_point(data = compare_gen_denmark, aes(x = prs_score_old, y = pressure_score, 
                                             text = paste("Region ID:", rgn_id))) +
  labs(x = paste("Previous Pressure", previous_data_year), 
       y = paste("Current Pressure", previous_data_year),
       title = "Denmark MAR current and previous pressure") +
  geom_abline(slope = 1, intercept = 0, col = "red") + theme_bw()

ggplotly(compare_gen_denmark_plot)

```
