---
title: "OHI `r format(Sys.Date(), '%Y')`: Mariculture"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

# Summary

This analysis converts FAO mariculture and seafood watch sustainability data into data used to calculate the OHI global mariculture status score. This also calculates the genetic escapee from mariculture pressure data.

Part 1 works **only** with the FAO mariculture data and prepares it for later analysis. This script will generate *2 versions* of the MAR data: clean version from the current year's assessment, and a version that combines the current year's assessment with the previous year's assessment. This is to accomodate a change in the SFW dataset that occurred in 2025 where we combined the 2024 and 2025 SFW data together. The clean data get's used for  

# 2025 updates from 2024 assessment
- We got updated Seafood Watch data from the Monterey Bay Aquarium in 2025. Because of this, we wanted to combine the SFW data from 2024 and 2025 together as a way to gapfill and as a way to change the data year in the `scenario_data_years.csv` for the OHI-global score checking. Gapfilling by a different data year occurs BEFORE all of the other gapfilling methods and has not been documented in the gapfill csv yet. This is a change that future assessments could add.


------------------------------------------------------------------------

# Data Source

## Production data

**Reference**:\
<https://www.fao.org/fishery/en/collection/aquaculture?lang=en> Release date: March 2023 FAO Global aquaculture production Quantity (1950 - 2023) Global aquaculture production (Value) (1950-2023)

**Downloaded**: June 23rd, 2025

**Instructions on downloading the FAO aquaculture data:**

**2025 download from Melanie**: 
Navigate to this link: <https://www.fao.org/fishery/statistics-query/en/global_production/global_production_quantity>

Click Download full dataset. Also downloaded the file below!

Navigate to this link: <https://www.fao.org/fishery/statistics-query/en/home>.

Click browse on Global aquaculture production. Click browse on Global aquaculture production (Quantity). Scroll to the bottom of the dimensions section, and de-select all of the pre-selected years. Under the query definition section, drag Country Name En (if not already present), ASFIS species Name En, FAO major fishing area Name En, ASFIS species Family scientific name, and Environment Name En, from the possible fields column into the selected rows column. Click on show data and confirm that data is present for 1950- two years prior to current year. Click download and select yes to include Null Values. Rename the file in the format FAO_GlobalAquacultureProduction_Quantity_1950_last_data_year. Upload the file to the appropriate year folder in the FAO_mariculture folder on mazu.

**Description**: Quantity (tonnes) of mariculture for each country, species, year.

**Time range**: 1950-2023

------------------------------------------------------------------------

## Setup 

```{r setup, message=FALSE, warning=FALSE}

knitr::opts_chunk$set(eval = FALSE)
  
# ---- Load packages ----
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
 librarian::shelf(
  ohicore,  #devtools::install_github('ohi-science/ohicore@dev')
  dplyr,
  tidyverse, # loads readr, stringr, tidyr, etc.
  tictoc,
  knitr,
  kableExtra,
  here,
  plotly
)

#library(taxize) had issues installing taxize because its dependency phanghorn is not compatible with this version of R


# ---- Set directories & source functions ----
current_year <- 2025 # update these!
previous_year <-  2024 # update these!
version_year <- paste0("v", current_year)
old_version_year <- paste0("v", previous_year)
current_mar_dir <- here::here("globalprep", "mar", version_year)
previous_mar_dir <- here::here("globalprep", "mar", old_version_year)


# ---- Load FAO-specific user-defined functions ----
# functions specific to mariculture dealing with compound countries, updated in 2024
source(here::here("globalprep", "mar", version_year, "mar_fxs.R"))
# functions for cleaning fao data
source(here::here("workflow", "R", "fao_fxn.R")) 
#source(here::here("workflow", "R", "fao_online_portal_clean.R")) # fao cleaning function # switched to new version in 2024
source(here::here("workflow", "R", "fao_query_data_tidy.R")) # 2024 version of FAO data tidying function 
source(here::here("workflow", "R", "common.R")) # directory locations, OHI essentials

# general raw data directory in Mazu
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")
# FAO data directory in Mazu
FAO_mariculture_dir <- here::here(raw_data_dir, "FAO_mariculture")
# Seafood Watch data directory in Mazu
seafood_watch_dir <- here::here(raw_data_dir, "seafood_watch_mar_sustainability")

```

# Import Raw Data: FAO Mariculture data

Mariculture production in tonnes.

```{r import-data}
## UPDATE THESE!
current_year <- current_year # update in previous chunk
last_year <- 2024 #last year of analysis

#last year of data available
#update if not two years behind the current assessment year
latest_data_year <- current_year - 2 # most up to date year in the data aquaculture data
previous_data_year <- last_year - 2 #last year of data in the last assessment, for comparison plots

current_data_folder <- paste0("d", current_year) # assessment year for file paths
previous_data_folder <- paste0("d", last_year)  # previous assessment year

# update the data file name to the current last year of data 
mar_raw <- readr::read_csv(here::here(FAO_mariculture_dir, current_data_folder, paste0("FAO_GlobalAquacultureProduction_Quantity_1950_", latest_data_year, ".csv")))

```

# Wrangle:

## Tidy mariculture data

Filter freshwater mariculture, make long format, and clean FAO codes.

```{r fao-data-prep}
# source 2025 FAO query data tidying function here if you skipped it in the setup chunk
# source(here::here("workflow", "R", "fao_query_data_tidy.R"))

# ---- preliminary tidying of FAO data ----
mar_tidy <- fao_query_data_tidy(fao = mar_raw, # FAO data frame
                                initial_data_year = 1950, # first year of data in mar df
                                last_data_year = latest_data_year, # most recent year of data in df
                                sub_N = 0.1 # what flagged values are substituted with (same as default value in function)
                                )

#names(mar_tidy) # run to easily copy+paste column names

#table(mar_tidy$unit_name)

# ---- clean tidied FAO data ----
mar_clean <- mar_tidy %>% 
  dplyr::select(-c(unit_name)) %>% # all are "Tonnes - live weight"
  dplyr::rename(
    country = country_name_en,
    FAO_name = asfis_species_name_en,
    fao = fao_major_fishing_area_name_en,
    environment = environment_name_en,
    family_scientific = family_scientific_name
  ) %>% 
  dplyr::mutate(FAO_name = ifelse(!is.na(FAO_name), FAO_name, # if species name is not NA, keep species name
                                  family_scientific)) #if species name is NA, use the family name


# investigate environment types
table(mar_clean$environment)

# filter out Freshwater, only keep Brackishwater and Marine
mar_filter <- mar_clean %>% 
  dplyr::filter(environment %in% c("Brackishwater", "Marine"))

# check
table(mar_filter$environment)

# rename cleaned, tidied, and filtered data to "mar"
mar <- mar_filter
```


## Update species names

Update species name in the `raw/species_list.csv` file with names in the `mar` dataset (exclude non-food species).

**Instructions for updating the species list:**

More information can be found in issues #8, #81, #241, #246 for reference.

The species_list.csv is a master list of all the mariculture species in the FAO data. Every year there are potential changes to the FAO data, additional species may be added or names can change, etc. When a species is in the FAO data, but not the species_list.csv it will need to be added to the species_list.csv. To do this, do a quick search within the species_list.csv to make sure it wasn't some small change to the name (e.g. "long-back eel" vs. "longback eel"). Otherwise, do some research to see if any of the synonyms for the FAO species match the names in species_list.csv. If it is one of these scenarios, the information can be copied to a new row with the name modification. If it seems to be a new taxa, You need to add the species output from this (new_spp) to species_list.csv and fill out the corresponding information. Add the new taxa name, taxon_code (categories defined in the README in the /raw folder), and family (general family from Google) to the species_list.csv. Do some research to make sure the species is being used for food and is marine/brackish. If it is not used for food or is freshwater then it is excluded from the analysis (i.e., a 1 in the exclude column). I.e. add FAO_name, exclude (whether to include in assessment or not. If they are only harvested for medicinal purposes, do not include, or not eaten by humans do not include). Because seaweeds are generally used for both human and non-human consumption, we quantify whether to exclude or not for seaweed species as proportions. For example, if a seaweed species is 80% used for non-human consumption, we would exclude 80% of the harvest (exclude = 0.8). However, for fish species, we only classify as 0 or 1.

After checking the species list, it is recommended to download the list as a csv, open up in excel, and manually add the new species and associated data that way.

Some helpful resources for researching new species:

FAO SPECIES IDENTIFICATION GUIDE FOR FISHERY PURPOSES -- THE LIVING MARINE RESOURCES OF THE
WESTERN CENTRAL PACIFIC; 

*   [Volume 2. Cephalopods, crustaceans, holothurians and sharks](https://decapoda.nhm.org/pdfs/27678/27678.pdf)

*   [Volume 3. Batoid fishes, chimaeras and Bony fishes part 1 (Elopidae to Linophrynidae)](https://openknowledge.fao.org/server/api/core/bitstreams/83754bd7-e58a-4fbe-914a-5c12f112e957/content)

*   [Volume 4. Bony fishes part 2 (Mugilidae to Carangidae) ](https://www.fao.org/4/x2400e/x2400e00.htm)

*   [Volume 5. Bony fishes part 3 (Menidae to Pomacentridae)](https://www.fao.org/4/y0770e/y0770e00.htm)

[SeaLifeBase search tool](https://www.sealifebase.se/search.php)

```{r update-species}

## Commented out 'read.csv' lines are different versions of the dataset, representing the three different seaweed exclusion methods (exclude some seaweed, all seaweed, all seaweed nei (nei = not elsewhere identified)). Use at the end for data checking. The first one is the one which is saved to the "output" folder, and the main analysis to run. 
# Read in species list
mar_sp <- readr::read_csv(here::here(current_mar_dir, "raw", "species_list.csv")) %>%
  dplyr::select(FAO_name, exclude, taxon_code = Taxon_code, family, notes, notes_2)

# to compare to previous year's csv (in v2024)
# old_mar_sp <- readr::read_csv(here::here("globalprep/mar/v2023/raw/species_list.csv")) %>%
#   dplyr::select(FAO_name, exclude, taxon_code = Taxon_code, family, notes, notes_2)
# old_test_spp <- setdiff(mar$FAO_name, old_mar_sp$FAO_name)
# old_test_spp

# v2025 Change to lowercase on the FAO data frame 
mar <- mar %>% 
  mutate(FAO_name = str_replace_all(mar$FAO_name, "NEI", "nei"))

# find differences between species names in new FAO data vs. species list
new_spp <- setdiff(mar$FAO_name, mar_sp$FAO_name)

new_spp # Check: if dim has 0 rows it means all match

## If there is a list of species, hand check species_list.csv to see whether to keep (exclude seaweeds and species harvested for ornamental/medicinal). See instructions above.

### NOTE: if the list of species to add is huge, and it is unmanageable to do by hand, see the mar_dataprep.Rmd in the v2020 folder, and it will have code to use Taxize. Otherwise just do it by hand, because taxize is a pain, and most of them won't be found in taxize anyways... 


# Remove species not relevant to mariculture goal (i.e., non-food species)
mar_sp_join <- mar %>% dplyr::left_join(mar_sp, by = "FAO_name")

# Create intermediate table, perform data wrangling and prep for analysis
mar_join_int <- mar_sp_join %>%
  dplyr::filter(exclude < 1) %>% # filter out species that should be excluded from the MAR subgoal
  dplyr::mutate(include = 1 - exclude) %>% # create an "include" column (1-exclude) 
  dplyr::rename(species = FAO_name) %>%  # rename FAO_name to species
  dplyr::mutate(value_include = value * include) # multiply value by proportion of harvest to include

## Check to see if 85% of seaweed goes to human consumption... filter for seaweeds, sum the tonnes included, divide by total tonnes seaweed
# seaweed_check2 <- mar_join_int %>%
#   filter(!is.na(value)) %>%
#   filter(taxon_code == "AL") %>%
#   group_by(year) %>%
#   summarise(include_total = sum(value_include),
#             overall_total = sum(value)) %>%
#   mutate(percent = include_total/overall_total)
# 
#  mean(seaweed_check2$percent)
 ## v2023: 81% .... matches pretty darn well.
 # v2024: 80.1%
 # v2025: 79.8%

# ---- Eliminate country-species data with zero production across the entire time-series (1950-recent) ----
mar_value_filter <- mar_join_int %>%
  dplyr::filter(!is.na(value_include)) %>% # remove NA values 
  dplyr::select(country, fao, environment, species, year, # select relevant columns
                taxon_code, family, value = value_include) %>% 
  dplyr::group_by(country, species) %>% # for each country-species combination
  dplyr::mutate(total_value = sum(value)) %>% # create total_value column
  dplyr::filter(total_value > 0) %>% # omit data with zero production
  dplyr::select(-total_value) %>% # drop temporary column
  dplyr::ungroup() # remove grouping

```

## Convert country names to OHI regions

```{r convert-regions}
# ---- update names before using name_to_rgn ----
mar_name_split <- mar_value_filter %>%
  mar_split()  # function in mar_fxs.R 

# Figure out if any of the [former] countries are actually being split
split_countries <- setdiff(mar_value_filter$country, mar_name_split$country)
split_countries

# Divides mariculture from countries that we report as separate regions (assume equal production in all regions)
  # Netherlands Antilles: Conch restoration among Aruba, Bonaire, Curacao
  # Channel Islands: Jersey and Guernsey
  # Bonaire/S.Eustatius/Saba
  # Yugoslavia SFR: no longer a country after 1992, no values 
  # v2024: added new formatting of "Bonaire, Sint Eustatius and Saba"
  # v2025: Added formatting for USSR, serbia and montenegro, and added [former] tag to netherlands and antilles


# ---- Run name_2_rgn ----
mar_rgn <- name_2_rgn(df_in = mar_name_split, 
                       fld_name ='country', 
                       flds_unique = c('species', 'fao', 'environment', 'taxon_code', 'year')) 

# v2024 notes:
## deletes Yugoslavia SFR -- this is okay because after 1991 (starting in 1992), the dataset differentiates between the countries that used to make up Yugoslavia (like Croatia, Serbia and Montenegro (Serbia is landlocked and we'll need to deal with the duplicates from this group), Bosnia and Herzegovina, Slovenia) starting in 1992. 

## Serbia and Montenegro: separate data for Montenegro beginning in 2006; Serbia is landlocked, so these values only apply to Montenegro (data is from FAO major fishing area classifications: Mediterranean and Black Sea) -> just call montenegro. 

# ---- Addressing duplicates, warnings ----

## China & Hong Kong: sum Hong Kong values within China OHI region classification. Call China, Hong Kong SAR. Then group with China and sum. 
# mar_sum <- mar_name_split %>% 
#   group_by(country, year, species) %>% 
#   mutate(country == China, country == China + China, Hong Kong SAR)

## Guadeloupe and Martinique: have data for each separately -- need to aggregate (sum) values for OHI grouped region

## Northern Mariana Islands and Guam: have data for Guam for full range, started having data for Northern Mariana Islands in 2007. Can sum both for OHI grouped region. 

## United Republic of Tanzania + United Republic of Tanzania, Zanzibar: According to Wikipedia, "Zanzibar is an insular semi-autonomous region which united with Tanganyika in 1964 to form the United Republic of Tanzania." 

### in past years, these values have been summed under the OHI region "Tanzania" 

mar_sum <- mar_name_split %>%
  mutate(country = case_when(
    country %in% c("China", "China, Hong Kong SAR") ~ "China",
    country %in% c("Guadeloupe", "Martinique") ~ "Martinique",
    country %in% c("Guam", "Northern Mariana Islands") ~ "Guam",
    country %in% c("United Republic of Tanzania", "United Republic of Tanzania, Zanzibar") ~ "United Republic of Tanzania",
    TRUE ~ country
  )) %>%
  dplyr::group_by(country, year, species, fao, environment, taxon_code, family) %>%
  dplyr::summarise(value = sum(value, na.rm = TRUE)) %>% 
  dplyr::ungroup()

# ---- Run name_2_rgn Again ----
mar_rgn <- name_2_rgn(df_in = mar_sum, 
                       fld_name ='country', 
                       flds_unique = c('species', 'fao', 'environment', 'taxon_code', 'year')) 

# Confirmed that duplicate regions have been handled

# ---- Aggregation: sum values of regions with multiple subregions ----
mar_rgn_agg <- mar_rgn %>%
  dplyr::group_by(fao, environment, species, year, taxon_code, rgn_id, rgn_name, family) %>% 
  dplyr::summarize(value = sum(value)) %>%
  dplyr::ungroup()
```

Take a look at the tidied data for a single year and region

```{r see-tidy-data eval=FALSE}
test <- data.frame(
  dplyr::filter(mar_rgn_agg, rgn_id == 207) %>% # Vietnam
  dplyr::filter(year == 2020) %>%
  arrange(species))
# View(test)
rm(test)
```

# Intermediate Mariculture Data Joining 

This step is ONLY used for the SFW `mar_sustainability_score.csv` later on. Mariculture tonnes does not get exported in that file, but mariculture and sustainability data do get combined before that final csv is written out, hence the reason to also combine the mariculture data.  

```{r}
# Join 2024 data before gapfilling
int_mar_old <- read_csv(here(previous_mar_dir, "int", "int_mar.csv")) %>% 
  janitor::clean_names()

# Trying to determine what I need to join on to avoid duplicates (i.e. what is the primary key)
int_mar_old %>%
  count(rgn_id, species, year, environment, family, fao) %>%
  filter(n > 1)

mar_rgn_agg %>%
  count(rgn_id, species, year, environment, family, fao) %>%
  filter(n > 1)

# Add a new year column that is the version year
mar_rgn_agg_vyear <- mar_rgn_agg %>% 
  mutate(vyear = 2025) 

int_mar_old <- int_mar_old %>% 
  mutate(vyear = 2024,
         year = as.character(year)) 

# Join previous and current year data
mar_rgn_agg_comb <- full_join(mar_rgn_agg_vyear, int_mar_old, by = join_by(rgn_id, species, year, environment, family, fao))

# Clean names
mar_rgn_agg_comb <- mar_rgn_agg_comb %>% 
  select(rgn_id, species, year, environment, family, fao, value_new = value.x,
         taxon_code_new = taxon_code.x, taxon_code_old = taxon_code.y, 
         vyear_new = vyear.x, vyear_old = vyear.y, value_old = value.y)

# Check NAs for gap filling
colSums(is.na(mar_rgn_agg_comb))

# Gap fill with other year data
mar_rgn_agg_clean <- mar_rgn_agg_comb %>%
  mutate(
    # Tonnes of production
    value_old = ifelse(is.na(value_old), value_new, value_old),
    value_new = ifelse(is.na(value_new), value_old, value_new),
    # Version year
    vyear_new = ifelse(is.na(vyear_new), current_year, vyear_new),
    vyear_old = ifelse(is.na(vyear_old), previous_year, vyear_old),
    # Taxon code
    taxon_code_new = ifelse(is.na(taxon_code_new), taxon_code_old, taxon_code_new),
    taxon_code_old = ifelse(is.na(taxon_code_old), taxon_code_new, taxon_code_old),
    )

# Confirm NAs are gone for columns of interest
colSums(is.na(mar_rgn_agg_clean))

# Filter down to desired columns
mar_rgn_agg_filt <- mar_rgn_agg_clean %>% 
  select(rgn_id, species, year, fao, environment, value_old, value_new, taxon_code = taxon_code_new, family, vyear_old, vyear_new)

# Pivot longer so data is in tidy format
mar_rgn_agg_long <- mar_rgn_agg_filt %>%
  pivot_longer(
    cols = c(value_old, value_new, vyear_old, vyear_new),
    names_to = c(".value", "year_old"),
    names_sep = "_"
  ) %>% 
  select(-year_old)

table(mar_rgn_agg_long$vyear)

# Reassign the combined mariculture data to maric for further processing
mar_rgn_agg_comb <- mar_rgn_agg_long

# Remove unneeded variables
rm(mar_rgn_agg_vyear, mar_rgn_agg_clean, mar_rgn_agg_filt, mar_rgn_agg_long)
```

# Gapfilling

## Fill in missing years after first year of harvest data with 0 values

For example: Production of blue shrimp in Maine starts in 1983 -- don't include years before that.

Checked to make sure that there weren't instances in which it made more sense to carry the previous year's data forward as a method of gapfilling. This didn't seem to be the case.


```{r gapfill-missing-0}
# Uncombined data
# ---- fill "years" column for gapfilling ----
# Confirm there are no duplicates in the combined data frame
mar_rgn_agg %>%
  dplyr::count(rgn_id, species, fao, environment, taxon_code, family, year) %>%
  dplyr::filter(n > 1)

# Spread mar_rgn to create a value for every year-species-region (if missing it will be given a NA)
mar_rgn_spread <- tidyr::pivot_wider(mar_rgn_agg,
                                     names_from = c(year),
                                     values_from = value)
dim(mar_rgn_spread)

# Turn data frame back into long format
mar_rgn_yrs_filled <- tidyr::pivot_longer(mar_rgn_spread,
                           cols = num_range("",paste0(1950:latest_data_year)),
                           names_to = "year",
                           values_to = "value") %>% 
  arrange(rgn_id, species, year, taxon_code, family, fao, environment)


## NA values are converted to zero.
mar_rgn_gf <- mar_rgn_yrs_filled %>%
  mutate(year = as.numeric(year)) %>%
  mutate(value_w_0 = ifelse(is.na(value), 0, value)) %>%
  group_by(fao, environment, species, taxon_code, family, rgn_id) %>% 
  mutate(cum_value = cumsum(value_w_0)) %>%
  ungroup() %>%
  filter(cum_value > 0) %>% # Eliminates years before mariculture began
  mutate(gap_0_fill = ifelse(is.na(value), "NA_to_zero", "0")) %>% # Record gapfill
  mutate(value = ifelse(is.na(value), 0, value)) %>% # Finally, convert all NAs in original column to 0
  dplyr::select(-cum_value, -value_w_0)


## Combined data
# ---- fill "years" column for gapfilling ----
# Confirm there are no duplicates in the combined data frame
mar_rgn_agg_comb %>%
  dplyr::count(rgn_id, species, fao, environment, taxon_code, family, year, vyear) %>%
  dplyr::filter(n > 1)

# Spread mar_rgn to create a value for every year-species-region (if missing it will be given a NA)
mar_rgn_spread <- tidyr::pivot_wider(mar_rgn_agg_comb,
                                     names_from = c(year, vyear),
                                     values_from = value)
dim(mar_rgn_spread)

# Turn data frame back into long format
mar_rgn_yrs_filled <- mar_rgn_spread %>%
  pivot_longer(
    cols = matches("^[0-9]{4}_[0-9]{4}$"), # 2025: regex to identify the new column names
    names_to = c("year", "vyear"),
    names_sep = "_",
    values_to = "value"
  ) %>%
  arrange(rgn_id, species, year, taxon_code, family, fao, environment)


## NA values are converted to zero.
mar_rgn_comb_gf <- mar_rgn_yrs_filled %>%
  mutate(year = as.numeric(year)) %>%
  mutate(vyear = as.numeric(vyear)) %>%
  mutate(value_w_0 = ifelse(is.na(value), 0, value)) %>%
  group_by(fao, environment, species, taxon_code, family, rgn_id) %>% 
  mutate(cum_value = cumsum(value_w_0)) %>%
  ungroup() %>%
  filter(cum_value > 0) %>% # Eliminates years before mariculture began
  mutate(gap_0_fill = ifelse(is.na(value), "NA_to_zero", "0")) %>% # Record gapfill
  mutate(value = ifelse(is.na(value), 0, value)) %>% # Finally, convert all NAs in original column to 0
  dplyr::select(-cum_value, -value_w_0)
```

Check how may NA values were converted to 0

```{r table-NA-0, eval=FALSE}
# Sanity check for combing 24 and 25 data
old <- mar_rgn_comb_gf %>% 
  filter(vyear == 2024)

new <- mar_rgn_comb_gf %>% 
  filter(vyear == 2025)

# See how many MAR values differ between them
length(setdiff(old$value, new$value)) # 261
length(setdiff(new$value, old$value)) # 302

# Check NAs gapfilled
table(mar_rgn_gf$gap_0_fill)
## 3790 of these out of 27810+3790 cases had NA values converted to 0 - 2018 assessment
## 3763 of these out of 28694+4291 cases had NA values converted to 0 - 2019 assessment
## 5932 of these out of 29305 +5932 cases had NA values converted to 0 - 2020 assessment
## 6441 of these out of 30170 + 6441 cases had NA values converted to 0 - 2021 assessment
## 7037 of these out of 30975 + 7037 cases had NA values converted to 0 - 2022 assessment
##7367 of these out of 34187 + 7367 cases had NA values converted to 0 -2023 assessment
## 8201 of these out of 35029 + 8201 (43230) cases had NA values converted to 0 = 0.1897 ≈ 19% of values - 2024 assessment

# 2025 Assessment
# After changing mar_fxs region splitting: 11101 out of 47619 (~23%) cases had NA values converted to 0

# After combining with 2024 data: 22046 out of 96271 (~23%) cases had NA values converted to 0
```

Remove family-region-environment time series with less than four years of mariculture production \> 0 tonnes (assume these are not established mariculture programs).

```{r remove-less-four-years}
# Uncombined data
mar_rgn_gf_filter <- mar_rgn_gf %>% 
  group_by(rgn_id, family, fao, environment) %>%
  mutate (not_0 = length(value[value > 0])) %>% # Length of vector of years greater than 0
  filter(not_0 > 3) %>% # Filter for groups that have at least four years of mariculture production 
  ungroup() %>% 
  dplyr::select(rgn_id, species, fao, environment, year, value, taxon_code, family, gap_0_fill) 

maric <- mar_rgn_gf_filter

# Combined data
mar_rgn_gf_filter <- mar_rgn_comb_gf %>% 
  group_by(rgn_id, family, fao, environment) %>%
  mutate (not_0 = length(value[value > 0])) %>% # Length of vector of years greater than 0
  filter(not_0 > 3) %>% # Filter for groups that have at least four years of mariculture production 
  ungroup() %>% 
  dplyr::select(rgn_id, species, fao, environment, year, vyear, value, taxon_code, family, gap_0_fill) 

maric_comb <- mar_rgn_gf_filter
```

# Save mariculture file

Used to estimate total mariculture yield per country.

Saves the Mariculture-FP file

```{r save-mariculture-yield, eval=FALSE}
# Write out only 2025 mariculture data
readr::write_csv(maric, file = here::here(current_mar_dir, "output", "MAR_FP_data.csv"))

# Change name here for the combined data
readr::write_csv(maric_comb, file = here::here(current_mar_dir, "output", "MAR_FP_data_comb.csv"))
```