---
title: "OHI `r format(Sys.Date(), '%Y')`: Mariculture"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

# Summary

This analysis converts FAO mariculture and seafood watch sustainability data into data used to calculate the OHI global mariculture status score. This also calculates the genetic escapee from mariculture pressure data.

# 2025 updates from 2024 assessment
- We got updated Seafood Watch data from the Monterey Bay Aquarium in 2025. Because of this, we wanted to combine the SFW data from 2024 and 2025 together as a way to gapfill and as a way to change the data year in the `scenario_data_years.csv` for the OHI-global score checking. Gapfilling by a different data year occurs BEFORE all of the other gapfilling methods and has not been documented in the gapfill csv yet. This is a change that future assessments could add.


## Setup 

```{r setup, message=FALSE, warning=FALSE}

knitr::opts_chunk$set(eval = FALSE)
  
# ---- Load packages ----
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
 librarian::shelf(
  ohicore,  #devtools::install_github('ohi-science/ohicore@dev')
  dplyr,
  tidyverse, # loads readr, stringr, tidyr, etc.
  tictoc,
  knitr,
  kableExtra,
  here,
  plotly
)

#library(taxize) had issues installing taxize because its dependency phanghorn is not compatible with this version of R


# ---- Set directories & source functions ----
current_year <- 2025 # update these!
previous_year <-  2024 # update these!
version_year <- paste0("v", current_year)
old_version_year <- paste0("v", previous_year)
current_mar_dir <- here::here("globalprep", "mar", version_year)
previous_mar_dir <- here::here("globalprep", "mar", old_version_year)


# ---- Load FAO-specific user-defined functions ----
# functions specific to mariculture dealing with compound countries, updated in 2024
source(here::here("globalprep", "mar", version_year, "mar_fxs.R"))
# functions for cleaning fao data
source(here::here("workflow", "R", "fao_fxn.R")) 
#source(here::here("workflow", "R", "fao_online_portal_clean.R")) # fao cleaning function # switched to new version in 2024
source(here::here("workflow", "R", "fao_query_data_tidy.R")) # 2024 version of FAO data tidying function 
source(here::here("workflow", "R", "common.R")) # directory locations, OHI essentials

# general raw data directory in Mazu
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")
# FAO data directory in Mazu
FAO_mariculture_dir <- here::here(raw_data_dir, "FAO_mariculture")
# Seafood Watch data directory in Mazu
seafood_watch_dir <- here::here(raw_data_dir, "seafood_watch_mar_sustainability")

```

# Import Raw Data: FAO Mariculture data

Mariculture production in tonnes.

```{r import-data}
## UPDATE THESE!
current_year <- current_year # update in previous chunk
last_year <- 2024 #last year of analysis

#last year of data available
#update if not two years behind the current assessment year
latest_data_year <- current_year - 2 # most up to date year in the data aquaculture data
previous_data_year <- last_year - 2 #last year of data in the last assessment, for comparison plots

current_data_folder <- paste0("d", current_year) # assessment year for file paths
previous_data_folder <- paste0("d", last_year)  # previous assessment year



# FAO mariculture and Seafood Watch sustainability scores

```{r match-type_sw-sus-data}
# 2025: this code doesn't seem necessary for the 2025 data. Running it produces no sw_sp_w

colSums(is.na(sw_sus_rgn))

## Add column "match_type" to categorize obs. that 1) have a country associated, 2) are "global" Seafood Watch data, or 3) only distinguished by water body
sw_sus_rgn <- sw_sus_rgn %>%
  dplyr::mutate(match_type =
                  dplyr::case_when(!is.na(rgn_name) ~ "sw_sp_c", # Add match type specific to species/country
                                  country == "Worldwide" ~ "sw_sp_g", # Add match type specific to species/global Seafood Watch data
                                  TRUE ~ "sw_sp_w" # Add match type specific to species/water body
  ))
 # if 0, this is the issue


table(sw_sus_rgn$match_type)

# 2025 combined: 
# sw_sp_c sw_sp_g sw_sp_w 
#     316     176       2 

```

Assign a Mediterranean bordering rgn id to each waterbody row

```{r gf_sw_sp_w}
## Add a line for each country that borders each water body
  # The only water body without a rgn is the Mediterranean; add all countries that border it (source: https://www.medqsr.org/mediterranean-marine-and-coastal-environment#:~:text=Today%2021%20countries%2C%20with%20surface,Syria%2C%20Tunisia%2C%20and%20Turkey.)

# filter for waterbody specific rows 
sw_sus_rgn_water <- sw_sus_rgn %>%
  filter(match_type == 'sw_sp_w')

# define Mediterranean rgns
med_rgns <- data.frame(rgn_id = c(82, 84, 232, 187, 81, 214, 179, 80, 79, 184, 78, 67, 68, 185, 186, 62, 188, 182, 77, 61, 76))

# create new function called expand.grid.df()
expand.grid.df <- function(...) Reduce(function(...) merge(..., by=NULL), list(...))
# coming from MEDS, this ^^ was wild to read
# Reduce: uses a binary function to successively combine the elements of a given vector and a possibly given initial value.
# merge: Merge two data frames by common columns or row names, or do other versions of database join operations.

gf_sw_sp_w <- expand.grid.df(sw_sus_rgn_water, med_rgns) %>%
  dplyr::select(-rgn_id.x, rgn_id = rgn_id.y)

```

# Join mariculture and SFW data

```{r sw-fao-joining}
# Combined
## Join data specific to species/country
sw_sp_c <- sw_sus_rgn %>%
  filter(match_type == 'sw_sp_c') %>%
  select(rgn_id, species, Sust_sw_sp_c = score, esc_sw_sp_c = escapes_score, vyear)

taiwan_before <- sw_sp_c %>% 
  filter(rgn_id == 14)
colSums(is.na(taiwan_before)) # sust and escapes both have 0 NAs
nrow(taiwan_before) # 4

#there will be many to many relationship b/c maric has more than one of the same species for a rgn_id due to year and the sustainability may have more than one method for a species/rgn combination
mar_sw_sus <- maric %>%
  left_join(sw_sp_c, by = c("species", "rgn_id", "vyear"))

taiwan <- mar_sw_sus_test %>% 
  filter(rgn_id == 14) 

colSums(is.na(taiwan)) # sust and escapes both have 4680 NAs
nrow(taiwan) # 4812

# 2025: commented this out because we have no water body regions
# Join data specific to species/water body
sw_sp_w <- gf_sw_sp_w %>%
  select(rgn_id, species, Sust_sw_sp_w = score, esc_sw_sp_w = escapes_score, vyear)

mar_sw_sus <- mar_sw_sus %>%
  left_join(sw_sp_w, by= c("species", "rgn_id", "vyear"))

## Join data specific to species/global Seafood Watch data
sw_sp_g <- sw_sus_rgn %>% 
  filter(match_type == 'sw_sp_g') %>%
  select(species, Sust_sw_sp_g = score, esc_sw_sp_g = escapes_score, vyear)

mar_sw_sus <- mar_sw_sus %>%
  left_join(sw_sp_g, by= c("species", "vyear"))

test3 <- mar_sw_sus %>%
  filter(taxon_code == "GAST")

nrow(test3) # 3918
sum(is.na(test3$Sust_sw_sp_c)) # 3918
# All sust_sw_sp_c and esc_sw_sp_c scores are NA for this test...

# Checking for the entire data frame now
nrow(mar_sw_sus) # 127150
sum(is.na(mar_sw_sus$Sust_sw_sp_c)) # 104190
sum(is.na(mar_sw_sus$esc_sw_sp_c)) # 104190

# Uncombined
## Join data specific to species/country
sw_sp_c <- sw_sus_rgn %>%
  filter(match_type == 'sw_sp_c') %>%
  select(rgn_id, species, Sust_sw_sp_c = score, esc_sw_sp_c = escapes_score)

#there will be many to many relationship b/c maric has more than one of the same species for a rgn_id due to year and the sustainability may have more than one method for a species/rgn combination
mar_sw_sus_uncomb <- maric_uncomb %>%
  left_join(sw_sp_c, by = c("species", "rgn_id"))

# Join data specific to species/water body
sw_sp_w <- gf_sw_sp_w %>%
  select(rgn_id, species, Sust_sw_sp_w = score, esc_sw_sp_w = escapes_score)

mar_sw_sus_uncomb <- mar_sw_sus_uncomb %>%
  left_join(sw_sp_w, by= c("species", "rgn_id"))

## Join data specific to species/global Seafood Watch data
sw_sp_g <- sw_sus_rgn %>% 
  filter(match_type == 'sw_sp_g') %>%
  select(species, Sust_sw_sp_g = score, esc_sw_sp_g = escapes_score)

mar_sw_sus_uncomb <- mar_sw_sus_uncomb %>%
  left_join(sw_sp_g, by= c("species"))

# Checking for the entire data frame now
nrow(mar_sw_sus_uncomb) # 125468
sum(is.na(mar_sw_sus_uncomb$Sust_sw_sp_c)) # 79950
sum(is.na(mar_sw_sus_uncomb$esc_sw_sp_c)) # 79950

```

Merge the three Seafood Watch type categories into a single sustainability score and genetic escapes score column in the following order:

1.  Sust_sw_sp_c: specific to species/country
2.  Sust_sw_sp_w: specific to species/water body
3.  Sust_sw_sp_g: specific to species/global Seafood Watch data

For example, if Sust_sw_sp_c is missing, use Sust_sw_sp_w and so on.

```{r sw-fao-sust-scores}
# Combined
mar_sw_sus <- mar_sw_sus %>%
  dplyr::mutate(Sust = ifelse(!is.na(Sust_sw_sp_c), Sust_sw_sp_c, Sust_sw_sp_w)) %>%
  dplyr::mutate(Sust = ifelse(is.na(Sust), Sust_sw_sp_g, Sust)) %>%
  dplyr::mutate(gapfill_sus = case_when(Sust_sw_sp_c ==
                                          Sust ~ "none",
                                        is.na(Sust_sw_sp_c) & Sust_sw_sp_w == Sust ~ "water_body",
                                        is.na(Sust_sw_sp_c) & is.na(Sust_sw_sp_w) & Sust_sw_sp_g == Sust ~ "global_species")) %>%
  dplyr::mutate(esc_score = ifelse(!is.na(esc_sw_sp_c), esc_sw_sp_c, esc_sw_sp_w)) %>%
  dplyr::mutate(esc_score = ifelse(is.na(esc_score), esc_sw_sp_g, esc_score)) %>%
  dplyr::mutate(gapfill_esc = case_when(esc_sw_sp_c == esc_score ~ "none",
                                        is.na(esc_sw_sp_c) & esc_sw_sp_w == esc_score ~ "water_body",
                                        is.na(esc_sw_sp_c) & is.na(esc_sw_sp_w) & esc_sw_sp_g == esc_score ~ "global_species")) %>%
  dplyr::select(-c(Sust_sw_sp_c, Sust_sw_sp_w, Sust_sw_sp_g, esc_sw_sp_c, esc_sw_sp_w, esc_sw_sp_g))

summary(mar_sw_sus)

# Uncombined
mar_sw_sus_uncomb <- mar_sw_sus_uncomb %>%
  dplyr::mutate(Sust = ifelse(!is.na(Sust_sw_sp_c), Sust_sw_sp_c, Sust_sw_sp_w)) %>%
  dplyr::mutate(Sust = ifelse(is.na(Sust), Sust_sw_sp_g, Sust)) %>%
  dplyr::mutate(gapfill_sus = case_when(Sust_sw_sp_c ==
                                          Sust ~ "none",
                                        is.na(Sust_sw_sp_c) & Sust_sw_sp_w == Sust ~ "water_body",
                                        is.na(Sust_sw_sp_c) & is.na(Sust_sw_sp_w) & Sust_sw_sp_g == Sust ~ "global_species")) %>%
  dplyr::mutate(esc_score = ifelse(!is.na(esc_sw_sp_c), esc_sw_sp_c, esc_sw_sp_w)) %>%
  dplyr::mutate(esc_score = ifelse(is.na(esc_score), esc_sw_sp_g, esc_score)) %>%
  dplyr::mutate(gapfill_esc = case_when(esc_sw_sp_c == esc_score ~ "none",
                                        is.na(esc_sw_sp_c) & esc_sw_sp_w == esc_score ~ "water_body",
                                        is.na(esc_sw_sp_c) & is.na(esc_sw_sp_w) & esc_sw_sp_g == esc_score ~ "global_species")) %>%
  dplyr::select(-c(Sust_sw_sp_c, Sust_sw_sp_w, Sust_sw_sp_g, esc_sw_sp_c, esc_sw_sp_w, esc_sw_sp_g))

summary(mar_sw_sus_uncomb)

# 2025 version without water body
# mar_sw_sus <- mar_sw_sus %>%
#   dplyr::mutate(Sust = Sust_sw_sp_c,
#                 Sust = ifelse(is.na(Sust), Sust_sw_sp_g, Sust),
#                 gapfill_sus = case_when(Sust_sw_sp_c == Sust ~ "none",
#                                         is.na(Sust_sw_sp_c) & Sust_sw_sp_g == Sust ~ "global_species")) %>%
#   dplyr::mutate(esc_score = esc_sw_sp_c,
#                 esc_score = ifelse(is.na(esc_score), esc_sw_sp_g, esc_score),
#                 gapfill_esc = case_when(esc_sw_sp_c == esc_score ~ "none",
#                                         is.na(esc_sw_sp_c) & esc_sw_sp_g == esc_score ~ "global_species")) %>%
#   dplyr::select(-c(Sust_sw_sp_c, Sust_sw_sp_g, esc_sw_sp_c, esc_sw_sp_g))
# 
# summary(mar_sw_sus)
```

Since some regions have multiple sustainability scores for the same species due to multiple aquaculture methods, but we don't know what proportions of which methods are used, we take the average of the sustainability scores in these instances.

Average sustainability scores within regions with more than one score (due to more than one aquaculture method):

```{r sw-sus-avg}

# check which countries have duplicates
anyDuplicated(mar_sw_sus)
# 2022 scenario year: 532
#2023 280
# 2024 scenario year: 316
# 2025: 329
# 2025 combined: 664

mar_sw_sus <- mar_sw_sus %>% 
  dplyr::group_by(rgn_id, species, vyear) %>% 
  dplyr::mutate(Sust_avg = mean(Sust, na.rm = TRUE),
                esc_avg = mean(esc_score, na.rm = TRUE)) %>% 
  dplyr::ungroup()

# code NAN values in Sust_avg and esc_avg columns as NA 
mar_sw_sus$Sust_avg[is.nan(mar_sw_sus$Sust_avg)] <- NA
mar_sw_sus$esc_avg[is.nan(mar_sw_sus$esc_avg)] <- NA
# 2025 nrows = 58086
taiwan3 <- mar_sw_sus %>% 
  filter(rgn_id == 14)

colSums(is.na(taiwan3)) # 3422
nrow(taiwan3) # 5978

# Uncombined
mar_sw_sus_uncomb <- mar_sw_sus_uncomb %>% 
  dplyr::group_by(rgn_id, species) %>% 
  dplyr::mutate(Sust_avg = mean(Sust, na.rm = TRUE),
                esc_avg = mean(esc_score, na.rm = TRUE)) %>% 
  dplyr::ungroup()

# code NAN values in Sust_avg and esc_avg columns as NA 
mar_sw_sus_uncomb$Sust_avg[is.nan(mar_sw_sus_uncomb$Sust_avg)] <- NA
mar_sw_sus_uncomb$esc_avg[is.nan(mar_sw_sus_uncomb$esc_avg)] <- NA
```

Get rid of duplicates for region/species/year:

```{r sw-sus-dup}
# Combined
mar_sw_sus <- mar_sw_sus %>% 
  dplyr::distinct(rgn_id, species, environment, year, vyear, .keep_all = TRUE) %>%
  # remove old Sust column (which was used to produce Sust_avg column) &
  # rename Sust_avg to Sust &
  # remove old esc_score column (which was used to produce esc_avg column) &
  # rename esc_avg to esc_score :
  dplyr::select(-Sust, Sust = Sust_avg, -esc_score, esc_score = esc_avg)

taiwan4 <- mar_sw_sus %>% 
  filter(rgn_id == 14)

colSums(is.na(taiwan4)) # 3422
nrow(taiwan4) # 4812

# Uncombined
mar_sw_sus_uncomb <- mar_sw_sus_uncomb %>% 
  dplyr::distinct(rgn_id, species, environment, year, .keep_all = TRUE) %>%
  # remove old Sust column (which was used to produce Sust_avg column) &
  # rename Sust_avg to Sust &
  # remove old esc_score column (which was used to produce esc_avg column) &
  # rename esc_avg to esc_score :
  dplyr::select(-Sust, Sust = Sust_avg, -esc_score, esc_score = esc_avg)

```

Now look at a summary after appending all the Seafood Watch data

```{r sw-sus-summary, eval = FALSE}
summary(mar_sw_sus) 
# 21215 NA values out of 32051 total obs; ~68% without scores (2022)
# 16378 NA values out of 33732 total obs ~ 49% without scores (2023)
# 17126 NA values out of 34920 total obs; ~ 49% without scores (2024)
# 18875 NA values out of 38000 total obs:~ 49.7% without scores (2025)
# This makes sense, since the data is twice as long now
# 40046 NA values out of 80223 total obs: ~ 49.9% without scores (2025 + 2024 combined)
```

Gapfilling based on families and taxon_code

Steps:\
- Group by country and family and summarise to get mean sustainability/escapes values (gapfilled by country/family)\
- Group by georegion and family and summarise to get mean sustainability/escapes values (gapfilled by georegion/family)\
- Group by family (global family gapfilled)\
- Gapfill by average of the taxon_code\
- Assign any Algae with the overall SFW seaweeds score from SFW data (7.92) and the C6 Escapes score (4, a subcomponent of the overall seaweed score)\
- link to the SFW Global Assessment for Seaweed available for download [here](https://www.seafoodwatch.org/recommendation/seaweed/seaweed-31568)\
- Apply the global average to any remaining sustainability/escapes score NAs\
- Finally, we rescale the scores to the highest sustainability score, under the assumption that the highest score is the best sustainability currently out there, and should be given a score of 100.

```{r sw-gapfill-categories}
## First we need to add a georegion column that we can group by later on. 
# Combined
mar_sw_sus <- mar_sw_sus %>%
  add_georegion_id() %>%
  dplyr::select(-level)

mar_sw_sus <- mar_sw_sus %>%
  group_by(rgn_id, family, vyear) %>%
  mutate(avg_rgn_fam_sust = mean(Sust, na.rm = TRUE), # produced NaN values (2022 scenario year)
         avg_rgn_fam_esc = mean(esc_score, na.rm = TRUE)) %>% # produced NaN values (2022 secario year)
  ungroup() %>%
  group_by(georgn_id, family, vyear) %>%
  mutate(avg_gr_fam_sust = mean(Sust, na.rm = TRUE), # produced some NaN values (2023 scenario year)
         avg_gr_fam_esc = mean(esc_score, na.rm = TRUE)) %>% # produced NaN values (2023 scenario year)
  ungroup() %>%
  group_by(family, vyear) %>%
  mutate(avg_fam_sust = mean(Sust, na.rm = TRUE), # produced some NaN values (2022 scenario year)
         avg_fam_esc = mean(esc_score, na.rm = TRUE)) %>% # produced NaN values (2022 scen ario year)
  ungroup() %>%
  group_by(taxon_code, vyear) %>%
  mutate(avg_taxon_code_sust = mean(Sust, na.rm = TRUE), # produced some NaN values (2022 scecario year)
         avg_taxon_code_esc = mean(esc_score, na.rm = TRUE)) %>% # produced some NaN values (2022 scecario year)
  ungroup() %>%
  mutate(avg_global = mean(Sust, na.rm = TRUE),
         avg_global_esc = mean(esc_score, na.rm = TRUE)) %>%
  ungroup()

# code NaN as NA for the columns that produced NaN, likely due to taking the mean of a set of all NA values 
mar_sw_sus$avg_rgn_fam_sust[is.nan(mar_sw_sus$avg_rgn_fam_sust)] <- NA
mar_sw_sus$avg_rgn_fam_esc[is.nan(mar_sw_sus$avg_rgn_fam_esc)] <- NA
mar_sw_sus$avg_gr_fam_sust[is.nan(mar_sw_sus$avg_gr_fam_sust)] <- NA
mar_sw_sus$avg_gr_fam_esc[is.nan(mar_sw_sus$avg_gr_fam_esc)] <- NA
mar_sw_sus$avg_fam_sust[is.nan(mar_sw_sus$avg_fam_sust)] <- NA
mar_sw_sus$avg_fam_esc[is.nan(mar_sw_sus$avg_fam_esc)] <- NA
mar_sw_sus$avg_taxon_code_sust[is.nan(mar_sw_sus$avg_taxon_code_sust)] <- NA
mar_sw_sus$avg_taxon_code_esc[is.nan(mar_sw_sus$avg_taxon_code_esc)] <- NA

taiwan5 <- mar_sw_sus %>% 
  filter(rgn_id == 14)

colSums(is.na(taiwan5)) # 3422
nrow(taiwan5) # 4812

# Uncombined
mar_sw_sus_uncomb <- mar_sw_sus_uncomb %>%
  add_georegion_id() %>%
  dplyr::select(-level)

mar_sw_sus_uncomb <- mar_sw_sus_uncomb %>%
  group_by(rgn_id, family) %>%
  mutate(avg_rgn_fam_sust = mean(Sust, na.rm = TRUE), # produced NaN values (2022 scenario year)
         avg_rgn_fam_esc = mean(esc_score, na.rm = TRUE)) %>% # produced NaN values (2022 secario year)
  ungroup() %>%
  group_by(georgn_id, family) %>%
  mutate(avg_gr_fam_sust = mean(Sust, na.rm = TRUE), # produced some NaN values (2023 scenario year)
         avg_gr_fam_esc = mean(esc_score, na.rm = TRUE)) %>% # produced NaN values (2023 scenario year)
  ungroup() %>%
  group_by(family) %>%
  mutate(avg_fam_sust = mean(Sust, na.rm = TRUE), # produced some NaN values (2022 scenario year)
         avg_fam_esc = mean(esc_score, na.rm = TRUE)) %>% # produced NaN values (2022 scen ario year)
  ungroup() %>%
  group_by(taxon_code) %>%
  mutate(avg_taxon_code_sust = mean(Sust, na.rm = TRUE), # produced some NaN values (2022 scecario year)
         avg_taxon_code_esc = mean(esc_score, na.rm = TRUE)) %>% # produced some NaN values (2022 scecario year)
  ungroup() %>%
  mutate(avg_global = mean(Sust, na.rm = TRUE),
         avg_global_esc = mean(esc_score, na.rm = TRUE)) %>%
  ungroup()

# code NaN as NA for the columns that produced NaN, likely due to taking the mean of a set of all NA values 
mar_sw_sus_uncomb$avg_rgn_fam_sust[is.nan(mar_sw_sus_uncomb$avg_rgn_fam_sust)] <- NA
mar_sw_sus_uncomb$avg_rgn_fam_esc[is.nan(mar_sw_sus_uncomb$avg_rgn_fam_esc)] <- NA
mar_sw_sus_uncomb$avg_gr_fam_sust[is.nan(mar_sw_sus_uncomb$avg_gr_fam_sust)] <- NA
mar_sw_sus_uncomb$avg_gr_fam_esc[is.nan(mar_sw_sus_uncomb$avg_gr_fam_esc)] <- NA
mar_sw_sus_uncomb$avg_fam_sust[is.nan(mar_sw_sus_uncomb$avg_fam_sust)] <- NA
mar_sw_sus_uncomb$avg_fam_esc[is.nan(mar_sw_sus_uncomb$avg_fam_esc)] <- NA
mar_sw_sus_uncomb$avg_taxon_code_sust[is.nan(mar_sw_sus_uncomb$avg_taxon_code_sust)] <- NA
mar_sw_sus_uncomb$avg_taxon_code_esc[is.nan(mar_sw_sus_uncomb$avg_taxon_code_esc)] <- NA
```

Obtain a sustainability and escapes score for each record, and a book-keeping column of whether it's actual or gapfilled

For missing sustainability and escapes scores:

Steps:

1.  Group by country and family and summarise to get mean sustainability and escapes values (gapfilled by country/family)
2.  Group by georegion and family and summarise to get mean sustainability and escapes values (gapfilled by georegion/family)
3.  Group by family to get mean sustainability and escapes values (global family gapfilled)
4.  Gapfill by average of the taxon_code to get mean sustainability and escapes values (taxon_group gapfilled)
5.  Assign any Algae with the seaweeds score from SFW data (6.72 for sustainability, 4 for escapes)
6.  Apply the global average to any remaining sustainability and escapes score NAs

```{r sw-sus-gapfill-missing}
# define gapfilling scores for global seaweed and escapes (a subcomponent of the global seaweed score) found in pdf here (updated occasionally): https://www.seafoodwatch.org/recommendation/seaweed/seaweed-31568
# seaweed_gf_score <- 6.72
# escapes_gf_score <- 4 
#in 2023 this wasn't necessary because these scores were already included in the seafood watch data, and I already applied them to all seaweed species when cleaning seafood watch data. Left the code above in case it is needed in future years. 

# Combined
mar_sw_sus_final <- mar_sw_sus %>%
  
  
  #### sustainability scores ####
  mutate(Sust = ifelse(is.na(Sust), avg_rgn_fam_sust, Sust)) %>% # Gapfill with region/family level
  mutate(gapfill_sus = ifelse(is.na(gapfill_sus) & Sust == avg_rgn_fam_sust, "region_family_avg", gapfill_sus)) %>%
  ## add in rgn family gapfill record
  mutate(Sust = ifelse(is.na(Sust), avg_gr_fam_sust, Sust)) %>% # Gapfill with georegion/family level
  mutate(gapfill_sus = ifelse(is.na(gapfill_sus) & Sust == avg_gr_fam_sust, "georegion_family_avg", gapfill_sus)) %>% # Add in georegion/family gapfill record
  mutate(Sust = ifelse(is.na(Sust),  avg_fam_sust, Sust)) %>% # Gapfill with family level
  mutate(gapfill_sus = ifelse(is.na(gapfill_sus) & Sust == avg_fam_sust, "family_avg", gapfill_sus)) %>% ## Add in family gapfill record
  mutate(Sust = ifelse(is.na(Sust), avg_taxon_code_sust, Sust)) %>%
  mutate(gapfill_sus = ifelse(is.na(gapfill_sus) & Sust == avg_taxon_code_sust, "taxon_group_avg", gapfill_sus)) %>%
  #mutate(Sust = ifelse(taxon_code == "AL", seaweed_gf_score, Sust)) %>% ## assign any algae with the overall SFW seaweed score, see Steps above for reference to document
  #mutate(gapfill_sus = ifelse(is.na(gapfill_sus) & Sust == seaweed_gf_score, "seaweed_score", gapfill_sus)) %>%
  mutate(Sust = ifelse(is.na(Sust), avg_global, Sust)) %>% 
  mutate(gapfill_sus = ifelse(is.na(gapfill_sus) & Sust == avg_global, "global_avg", gapfill_sus)) %>%
  
  #### genetic escapes scores ####
  mutate(esc_score = ifelse(is.na(esc_score), avg_rgn_fam_esc, esc_score)) %>% # Gapfill with region/family level
  mutate(gapfill_esc = ifelse(is.na(gapfill_esc) & esc_score == avg_rgn_fam_esc, "region_family_avg", gapfill_esc)) %>% ## add in rgn family gapfill record
  mutate(esc_score = ifelse(is.na(esc_score), avg_gr_fam_esc, esc_score)) %>% # Gapfill with georegion/family level
  mutate(gapfill_esc = ifelse(is.na(gapfill_esc) & esc_score == avg_gr_fam_esc, "georegion_family_avg", gapfill_esc)) %>% # Add in georegion/family gapfill record
  mutate(esc_score = ifelse(is.na(esc_score),  avg_fam_esc, esc_score)) %>% # Gapfill with family level
  mutate(gapfill_esc = ifelse(is.na(gapfill_esc) & esc_score == avg_fam_esc, "family_avg", gapfill_esc)) %>% ## Add in family gapfill record
  mutate(esc_score = ifelse(is.na(esc_score), avg_taxon_code_esc, esc_score)) %>%
  mutate(gapfill_esc = ifelse(is.na(gapfill_esc) & esc_score == avg_taxon_code_esc, "taxon_group_avg", gapfill_esc)) %>%
  #mutate(esc_score = ifelse(taxon_code == "AL", escapes_gf_score, esc_score)) %>% ## assign any algae with the SFW seaweed score for C6 Escapes, see Steps above for reference to document
  #mutate(gapfill_esc = ifelse(is.na(gapfill_esc) & esc_score == escapes_gf_score, "seaweed_score", gapfill_esc)) %>%
  mutate(esc_score = ifelse(is.na(esc_score), avg_global_esc, esc_score)) %>% 
  mutate(gapfill_esc = ifelse(is.na(gapfill_esc) & esc_score == avg_global_esc, "global_avg", gapfill_esc)) %>%
  
  #### final tidying ####
mutate(taxa_code = paste(species, species_code, sep="_")) %>%
  select(rgn_id, species, species_code, taxa_code, taxa_group = taxon_code,
         family, year, vyear, gapfill_sus, gapfill_esc, gapfill_fao = gap_0_fill,
         tonnes = value, Sust, esc_score) %>%
  mutate(Sust = round(Sust/10, 2),
         esc_score = 1 - round(esc_score/10, 2))

summary(mar_sw_sus_final)
filter(mar_sw_sus_final, gapfill_sus == "georegion_family_avg")
filter(mar_sw_sus_final, gapfill_sus == "family_avg")
filter(mar_sw_sus_final, gapfill_sus == "taxon_group_avg")

taiwan6 <- mar_sw_sus_final %>% 
  filter(rgn_id == 14)

colSums(is.na(taiwan6)) # 0
nrow(taiwan6) # 4812

# Uncombined
mar_sw_sus_uncomb_final <- mar_sw_sus_uncomb %>%
  
  
  #### sustainability scores ####
  mutate(Sust = ifelse(is.na(Sust), avg_rgn_fam_sust, Sust)) %>% # Gapfill with region/family level
  mutate(gapfill_sus = ifelse(is.na(gapfill_sus) & Sust == avg_rgn_fam_sust, "region_family_avg", gapfill_sus)) %>%
  ## add in rgn family gapfill record
  mutate(Sust = ifelse(is.na(Sust), avg_gr_fam_sust, Sust)) %>% # Gapfill with georegion/family level
  mutate(gapfill_sus = ifelse(is.na(gapfill_sus) & Sust == avg_gr_fam_sust, "georegion_family_avg", gapfill_sus)) %>% # Add in georegion/family gapfill record
  mutate(Sust = ifelse(is.na(Sust),  avg_fam_sust, Sust)) %>% # Gapfill with family level
  mutate(gapfill_sus = ifelse(is.na(gapfill_sus) & Sust == avg_fam_sust, "family_avg", gapfill_sus)) %>% ## Add in family gapfill record
  mutate(Sust = ifelse(is.na(Sust), avg_taxon_code_sust, Sust)) %>%
  mutate(gapfill_sus = ifelse(is.na(gapfill_sus) & Sust == avg_taxon_code_sust, "taxon_group_avg", gapfill_sus)) %>%
  #mutate(Sust = ifelse(taxon_code == "AL", seaweed_gf_score, Sust)) %>% ## assign any algae with the overall SFW seaweed score, see Steps above for reference to document
  #mutate(gapfill_sus = ifelse(is.na(gapfill_sus) & Sust == seaweed_gf_score, "seaweed_score", gapfill_sus)) %>%
  mutate(Sust = ifelse(is.na(Sust), avg_global, Sust)) %>% 
  mutate(gapfill_sus = ifelse(is.na(gapfill_sus) & Sust == avg_global, "global_avg", gapfill_sus)) %>%
  
  #### genetic escapes scores ####
  mutate(esc_score = ifelse(is.na(esc_score), avg_rgn_fam_esc, esc_score)) %>% # Gapfill with region/family level
  mutate(gapfill_esc = ifelse(is.na(gapfill_esc) & esc_score == avg_rgn_fam_esc, "region_family_avg", gapfill_esc)) %>% ## add in rgn family gapfill record
  mutate(esc_score = ifelse(is.na(esc_score), avg_gr_fam_esc, esc_score)) %>% # Gapfill with georegion/family level
  mutate(gapfill_esc = ifelse(is.na(gapfill_esc) & esc_score == avg_gr_fam_esc, "georegion_family_avg", gapfill_esc)) %>% # Add in georegion/family gapfill record
  mutate(esc_score = ifelse(is.na(esc_score),  avg_fam_esc, esc_score)) %>% # Gapfill with family level
  mutate(gapfill_esc = ifelse(is.na(gapfill_esc) & esc_score == avg_fam_esc, "family_avg", gapfill_esc)) %>% ## Add in family gapfill record
  mutate(esc_score = ifelse(is.na(esc_score), avg_taxon_code_esc, esc_score)) %>%
  mutate(gapfill_esc = ifelse(is.na(gapfill_esc) & esc_score == avg_taxon_code_esc, "taxon_group_avg", gapfill_esc)) %>%
  #mutate(esc_score = ifelse(taxon_code == "AL", escapes_gf_score, esc_score)) %>% ## assign any algae with the SFW seaweed score for C6 Escapes, see Steps above for reference to document
  #mutate(gapfill_esc = ifelse(is.na(gapfill_esc) & esc_score == escapes_gf_score, "seaweed_score", gapfill_esc)) %>%
  mutate(esc_score = ifelse(is.na(esc_score), avg_global_esc, esc_score)) %>% 
  mutate(gapfill_esc = ifelse(is.na(gapfill_esc) & esc_score == avg_global_esc, "global_avg", gapfill_esc)) %>%
  
  #### final tidying ####
mutate(taxa_code = paste(species, species_code, sep="_")) %>%
  select(rgn_id, species, species_code, taxa_code, taxa_group = taxon_code,
         family, year, gapfill_sus, gapfill_esc, gapfill_fao = gap_0_fill,
         tonnes = value, Sust, esc_score) %>%
  mutate(Sust = round(Sust/10, 2),
         esc_score = 1 - round(esc_score/10, 2))

summary(mar_sw_sus_uncomb_final)
summary(mar_sw_sus_final)

filter(mar_sw_sus_uncomb_final, gapfill_sus == "georegion_family_avg")
filter(mar_sw_sus_uncomb_final, gapfill_sus == "family_avg")
filter(mar_sw_sus_uncomb_final, gapfill_sus == "taxon_group_avg")
```

```{r old-data-comb}
# Changing the data years to account for the new SFW data
# latest_data_year <- 2025
# previous_data_year <- 2024
# 
# # Compare new/old Mariculture sustainability scores
# sust_old <- read_csv(here("globalprep", "mar", paste0("v", previous_data_year), "output", "mar_sustainability.csv")) %>% 
#   dplyr::select(rgn_id, taxa_code_old = taxa_code, sust_old = sust_coeff)
# 
# # tidy up species names (such as "Abalones_1"), rename as species column
# sust_old$species <- str_remove(sust_old$taxa_code_old, "\\_[^.]*$")
# 
# # Drop the unneeded taxa code
# sust_old <- sust_old %>% 
#   mutate(year_old = 2024)
# 
# mar_sw_sus_final <- mar_sw_sus_final %>% 
#   mutate(year = 2025)
# 
# # Join previous and current year data
# sust_comb <- full_join(mar_sw_sus_final, sust_old, by = c("rgn_id", "species"))
# 
# # Pivot longer so we only have one year column
# # Check NAs across entire df
# colSums(is.na(sust_comb))
# 
# # Fill NAs in sust_old (previous year sustainability) with Sust from this year
# 
# # Don't fill year, just score. 
# 
# sust_comb_clean <- sust_comb %>%
#   mutate(sust_old = ifelse(is.na(sust_old), Sust, sust_old),
#          Sust = ifelse(is.na(Sust), sust_old, Sust),
#          year_old = ifelse(is.na(year_old), year, year_old),
#          year = ifelse(is.na(year), year_old, year),
#          taxa_code = ifelse(is.na(taxa_code), taxa_code_old, taxa_code),
#          taxa_code_old = ifelse(is.na(taxa_code_old), taxa_code, taxa_code_old))
# 
# colSums(is.na(sust_comb_clean))
# 
# # Now that we've filled the NAs, reseparate the data frames so we can do a simple bindrows on them
# sust_new_clean <- sust_comb_clean %>% 
#   select(rgn_id, species, taxa_code, year, sust = Sust)
# 
# sust_old_clean <- sust_comb_clean %>% 
#   select(rgn_id, species, taxa_code = taxa_code_old, year = year_old, sust = sust_old)
# 
# sust_clean <- bind_rows(sust_new_clean, sust_old_clean)
# 
# # define the maximum sustainability score, check with:
# summary(mar_sw_sus_final)
# 
# # Issue: the 2024 data was already scaled. Should I pull from the unscaled version of the 2024 data and recalculate score? Or, I can scale the 2025 data before I combine it with the 2024 data? 
# max_sus <- 0.77
# 
# ## Save sustainability scores data for 2023 (SFW data year)
# mar_sustainability_score <- sust_clean %>%
#   # mutate(year = current_year) %>% #change year if data not updated to current year
#   dplyr::select(rgn_id, year, taxa_code, sust_coeff = sust) %>%
#   unique() %>%
#   mutate(sust_coeff = ifelse(sust_coeff > max_sus, 1, sust_coeff/max_sus)) # rescale to the highest sust score
# 
# # Check for any duplicated rows
# sum(duplicated(mar_sustainability_score))
# 
# # Check for NAs in the data frame
# colSums(is.na(mar_sustainability_score))
# 
# # Confirm that the right balance of years is present in the data frame
# table(mar_sustainability_score$year)
# 
# view(mar_sustainability_score)

## Saves the appropriate Sustainability Score file 
#write.csv(mar_sustainability_score, here(current_mar_dir, "output", "mar_sustainability.csv"), row.names = F)
# readr::write_csv(mar_sustainability_score, here(current_mar_dir, "output", "mar_sustainability_new.csv"))
```

Now we need to prep the escapes pressure layer a bit more. Final formatting of the escapee data. This is used as a pressure layer.

For each region/year: take a weighted average of the genetic escape probability for each taxa based on tonnes of mariculture.

```{r gen-esc-formatting}
# subset the cleaned data to only include escapes data
# Genetic escapes will use the combined data, but we don't need to inlcude version year here
tonnes_esc <- mar_sw_sus_final %>%
  dplyr::select(rgn_id, year, species, gapfill_esc, esc_score, tonnes)

# use "tonnes" values to weight the regional average of escapes
genEscapes <- tonnes_esc %>%
  group_by(rgn_id, year) %>%
  summarize(genEscapes = weighted.mean(esc_score, tonnes, na.rm=TRUE)) %>% 
  ungroup()
```

Obtain gapfill information from `genEscapes`

Gapfill values in output file are the proportion of data that is gapfilled.

```{r obtain-gapfill-info-genEscapes}

# Obtain corresponding gapfilling information for each region (average of gapfilled data, weighted by tonnes of mariculture).
genEscapes_gf <- tonnes_esc %>%
  mutate(gapfill = ifelse(gapfill_esc == "none", 0, 1)) %>%
  group_by(rgn_id, year) %>%
  summarize(genEscapes = weighted.mean(gapfill, tonnes, na.rm = TRUE)) %>%
  ungroup() %>% 
  filter(year == latest_data_year) %>%
  select(rgn_id, pressures.score = genEscapes) %>%
  mutate(pressures.score = ifelse(pressures.score == "NaN", NA, pressures.score)) %>%
  data.frame()

```

# Save data: Genetic Escapes

## Save gapfill data

```{r save-gen-esc-gapfilled, eval = FALSE}
# Save the Genetic Escapes Gapfill file
readr::write_csv(genEscapes_gf, here(current_mar_dir, "output", "GenEsc_gf.csv"))
```

Obtain escapee data layers from `genEscapes`

```{r create-gen-esc-layers}
# Create the escapee data layers:
data <- genEscapes %>%
    dplyr::select(rgn_id, year, pressure_score = genEscapes)
```

## Save escapee pressure layer

```{r save-gen-esc-pressure-layer, eval = FALSE}
## Saves the Genetic Escapes data file 
readr::write_csv(data, here(current_mar_dir, "output", "GenEsc.csv"))  
```

# Save Data: Mariculture tonnes and sustainability scores

```{r save-mar-sw-harvest-data, eval = FALSE}

## Save mariculture harvest data
mar_harvest_tonnes <- mar_sw_sus_uncomb_final %>%
  # This data frame ONLY has MAR data, therefore this one will not use the combined 2024 and 2025 data
  dplyr::select(rgn_id, taxa_code, taxa_group, family, year, tonnes)

anyDuplicated(mar_harvest_tonnes) # Check for duplication

## Saves the appropriate Mariculture Harvest Tonnes file
# write.csv(mar_harvest_tonnes, here(current_mar_dir, "output", "mar_harvest_tonnes.csv"), row.names = F)
readr::write_csv(mar_harvest_tonnes, here::here(current_mar_dir, "output", "mar_harvest_tonnes.csv"))


## Save gapfill data for mariculture harvest
mar_harvest_tonnes_gf <- mar_sw_sus_final %>%
  dplyr::select(rgn_id, taxa_code, taxa_group, family, year, tonnes = gapfill_fao)


## Saves the appropriate Mariculture Harvest Gapfill file 
#write.csv(mar_harvest_tonnes_gf, here(current_mar_dir, "output", "mar_harvest_tonnes_gf.csv"), row.names = F)
readr::write_csv(mar_harvest_tonnes_gf, here::here(current_mar_dir, "output", "mar_harvest_tonnes_gf.csv"))

# define the maximum sustainability score, check with: summary(mar_sw_sus_final)
max_sus <- 0.77

## Save sustainability scores data for 2023 (SFW data year)
mar_sustainability_score <- mar_sw_sus_final %>%
  mutate(year = current_year) %>% # change year if data not updated to current year
  dplyr::select(rgn_id, year, taxa_code, sust_coeff = Sust) %>%
  unique() %>%
  mutate(sust_coeff = ifelse(sust_coeff > max_sus, 1, sust_coeff/max_sus)) # rescale to the highest sust score

anyDuplicated(mar_sustainability_score)

taiwan7 <- mar_sustainability_score %>% 
  filter(rgn_id == 14)

colSums(is.na(taiwan7)) # 0
nrow(taiwan7) # 62

taiwan_old24 <- mar_sustainability %>% 
  filter(rgn_id == 14) %>% 
  view()

setdiff(taiwan7$sust_coeff, taiwan_old24$sust_coeff)

## Saves the appropriate Sustainability Score file 
#write.csv(mar_sustainability_score, here(current_mar_dir, "output", "mar_sustainability.csv"), row.names = F)
readr::write_csv(mar_sustainability_score, here::here(current_mar_dir, "output", "mar_sustainability.csv"))


## Save gapfill data for sustainability scores
mar_sustainability_score_gf = mar_sw_sus_final %>%
  select(rgn_id, taxa_code, sust_coeff = gapfill_sus) %>%
  unique()


## Saves the appropriate Sustainability Score Gapfill file 
#write.csv(mar_sustainability_score_gf, here(current_mar_dir, "output", "mar_sustainability_gf.csv"), row.names = F)
readr::write_csv(mar_sustainability_score_gf, here::here(current_mar_dir, "output", "mar_sustainability_gf.csv"))


## save mar sustainability for data checking; rgn_id, species, score
mar_sustainability_score <- mar_sw_sus_final %>%
  mutate(year = current_year) %>% # Only 2023 sustainability scores exist (Seafood Watch)
  select(rgn_id, year, species, sust_coeff = Sust) %>%
  unique() %>% 
    mutate(sust_coeff = ifelse(sust_coeff > max_sus, 1, sust_coeff/max_sus)) # rescale to the highest sust score

## Saves the appropriate Sustainability Score file
#write.csv(mar_sustainability_score, here(current_mar_dir, "int", "mar_sustainability_check.csv"), row.names = F)
readr::write_csv(mar_sustainability_score, here::here(current_mar_dir, "int", "mar_sustainability_check.csv"))

```

