---
title: "OHI `r format(Sys.Date(), '%Y')`: Mariculture"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

# Summary

This analysis converts FAO mariculture and seafood watch sustainability data into data used to calculate the OHI global mariculture status score. This also calculates the genetic escapee from mariculture pressure data.

# 2025 updates from 2024 assessment
- We got updated Seafood Watch data from the Monterey Bay Aquarium in 2025. Because of this, we wanted to combine the SFW data from 2024 and 2025 together as a way to gapfill and as a way to change the data year in the `scenario_data_years.csv` for the OHI-global score checking. Gapfilling by a different data year occurs BEFORE all of the other gapfilling methods and has not been documented in the gapfill csv yet. This is a change that future assessments could add.


------------------------------------------------------------------------

# Data Source

## Seafood Watch sustainability data

Obtained upon request of SFW.

**Reference**:\
[Seafood Watch Complete Recommendation documentation](https://www.seafoodwatch.org/globalassets/sfw/pdf/whats-new/seafood-watch-complete-recommendation-list.pdf)\
[Seafood Watch Scoring Theory](https://www.seafoodwatch.org/globalassets/sfw/pdf/standards/aquaculture/seafood-watch-aquaculture-standard-version-a4.pdf)\
[Seafood Watch Global Seaweed Assessment](https://www.seafoodwatch.org/recommendation/seaweed/seaweed-31568) (provides gapfilling constants)

**Downloaded**: June 27, 2025

**Description**: Monterey Bay Aquarium Seafood Watch aquaculture recommendations. Sustainability scored from 0-10. Rescaled to 0-1 using the max score in the data.

------------------------------------------------------------------------

## Setup 

```{r setup, message=FALSE, warning=FALSE}

knitr::opts_chunk$set(eval = FALSE)
  
# ---- Load packages ----
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
 librarian::shelf(
  ohicore,  #devtools::install_github('ohi-science/ohicore@dev')
  dplyr,
  tidyverse, # loads readr, stringr, tidyr, etc.
  tictoc,
  knitr,
  kableExtra,
  here,
  plotly
)

#library(taxize) had issues installing taxize because its dependency phanghorn is not compatible with this version of R


# ---- Set directories & source functions ----
current_year <- 2025 # update these!
previous_year <-  2024 # update these!
version_year <- paste0("v", current_year)
old_version_year <- paste0("v", previous_year)
current_mar_dir <- here::here("globalprep", "mar", version_year)
previous_mar_dir <- here::here("globalprep", "mar", old_version_year)


# ---- Load FAO-specific user-defined functions ----
# functions specific to mariculture dealing with compound countries, updated in 2024
source(here::here("globalprep", "mar", version_year, "mar_fxs.R"))
# functions for cleaning fao data
source(here::here("workflow", "R", "fao_fxn.R")) 
#source(here::here("workflow", "R", "fao_online_portal_clean.R")) # fao cleaning function # switched to new version in 2024
source(here::here("workflow", "R", "fao_query_data_tidy.R")) # 2024 version of FAO data tidying function 
source(here::here("workflow", "R", "common.R")) # directory locations, OHI essentials

# general raw data directory in Mazu
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")
# FAO data directory in Mazu
FAO_mariculture_dir <- here::here(raw_data_dir, "FAO_mariculture")
# Seafood Watch data directory in Mazu
seafood_watch_dir <- here::here(raw_data_dir, "seafood_watch_mar_sustainability")

```

# Import Raw Data: FAO Mariculture data

Mariculture production in tonnes.

```{r import-data}
## UPDATE THESE!
current_year <- current_year # update in previous chunk
last_year <- 2024 #last year of analysis

#last year of data available
#update if not two years behind the current assessment year
latest_data_year <- current_year - 2 # most up to date year in the data aquaculture data
previous_data_year <- last_year - 2 #last year of data in the last assessment, for comparison plots

current_data_folder <- paste0("d", current_year) # assessment year for file paths
previous_data_folder <- paste0("d", last_year)  # previous assessment year

# update the data file name to the current last year of data 
mar_raw <- readr::read_csv(here::here(FAO_mariculture_dir, current_data_folder, paste0("FAO_GlobalAquacultureProduction_Quantity_1950_", latest_data_year, ".csv")))

```

# Sustainability Scores from Seafood Watch Data
These data will be complete combined with the 2024 scenario data. There is no need to keep them separate like there is with the mariculture data. 

## Import data: Seafood Watch sustainability scores

These data describe the sustainability country/species combinations. In cases where these data were not available for a specific county/species, we averaged the data across taxonomic groups to gapfill the missing data. We also calculate the genetic escapes pressure layer from this, based on the column `AqCriteria6`.

```{r sw-scores}

# ---- Load in Seafood Watch sustainability scores data from Mazu: ----
# Read in new Seafood Watch data 
sw_sus_raw <- read_csv(file.path(seafood_watch_dir,
                             current_data_folder,
                             "SFW_Aquaculture_ratings_062725.csv"), # update with name of new data file
                   na = c("NA", ""))


# v2025: 2024 assessment couldn't get new data, so here we're comparing to 2023.
old_sw_sus <- readr::read_csv(file.path(seafood_watch_dir,
                                 paste0("d", current_year - 2),
                                 "SFW_Aquaculture_ratings_053123.csv" #v2023 data
                               ),
                       na = c("NA", ""))


# Check what columns names might have changed between 2023 and 2025
setdiff(colnames(sw_sus_raw), colnames(old_sw_sus)) 
# [1] "StandardVersion"    "RecognizedBycatch" 
# [3] "FAOBOWs"            "ISSCFGCode"        
# [5] "ISSCFGParentCode"   "ISSCFGAbbreviation"

setdiff(colnames(old_sw_sus), colnames(sw_sus_raw))
# [1] "RecType"       "Version"       "Streamline"   
# [4] "ReportStatus"  "UpdateType"    "FAOMajorCodes"

```

# Wrangle

## Seafood Watch sustainability data background

Rename columns to match with MAR data and fill in species column

Notes on AssessmentScore column:

This column represents the Overall Score, which is a combination of 7 primary criteria and 3 "exceptional criterion that may not be relevant to all aquaculture production, yet can be a significant concern for those production practices where it is relevant" (pg. 57). These criteria are named as follows:

-   Criterion 1 (C1): Data
-   C2: Effluent
-   C3: Habitat
-   C4: Chemical Use
-   C5: Feed
-   C6: Escapes
-   C7: Disease, pathogen and parasite interaction

-   C8X: Source of Stock – Independence from wild fish stocks
-   C9X: Wildlife Mortalities
-   C10X: Introduction of Secondary Species 

The Final numerical score = 

$$\mathrm{final \:score} = \frac{\sum{(C1–C7 \mathrm{\:scores})} + (C8X + C9X + C10X)}{7}$$
 with the final score having a range of 0-10. 


Note on C6Score column -- C6: Escapes

From pages 43-52 of the [Aquaculture Sustainability Standards document](https://www.seafoodwatch.org/globalassets/sfw/pdf/standards/aquaculture/seafood-watch-aquaculture-standard-version-a4.pdf)

Criterion 6 (C6) final scores are a combination of Factor 6.1 and 6.2.

Factor 6.1: 
> "Factor 6.1 assigns a level of risk to each type of production system based on the ability of
farmed species to escape the system and enter the surrounding ecosystem. Production
system escape risks are categorized as Low to High based on openness, management
practices, escape trends, and vulnerability to environmental factors (e.g., tsunami, flood,
predator damage, etc.)." (pg. 44)

-   basically assigns a risk level based on the ability of farmed species to escape -- "Risk of escape"
-   adjustments made based on factors such as recapture of "escapes" before they have an impact

Factor 6.2:
>   "Invasiveness, referred to as the risk of competitive and genetic interactions (CGI), is defined
as “…the degree to which an organism is able to spread from site of primary introduction, to
establish a viable population in the ecosystem, to negatively affect biodiversity on the
individual, community, or ecosystem level and cause adverse socioeconomic consequence”
(Panov et al. 2008). According to this definition, Factor 6.2 considers both the short-term and
long-term ecological impacts of escape. This factor has been adapted (and greatly simplified)
from the Marine Fish Invasiveness Screening Kit (MFISK) (and other similar tools developed by
Copp et al. [2007, 2009]), and from the Global Aquaculture Performance Index (GAPI)’s similar
use and adaptation of the same tools (Volpe et al. 2013)." (pg. 45)

-   basically another quantification of risk based on the threat escapees pose on the environment -- "Competitiveness and genetic interactions"

> "The risk of impacts resulting from repeated escapes of farmed stock (regardless of their ability
to establish), or the risks resulting in the establishment of escapees differs according to
species-specific characteristics, and particularly between native and non-native species..." (pg. 45)

A C6 score of 10 indicates a very low risk of escape with low estimated competitiveness and genetic interactions. An Escape Criterion (C6) score ≤ 1 is considered "Critical" (pg. 52), and may indicate a "High Concern" risk of escape and "High Concern" potential for competitiveness and genetic impact. 

## Tidy Seafood Watch sustainability data
```{r tidy_sw-sus}
# ---- Tidy Seafood Watch (SFW) data ----
sw_sus <- sw_sus_raw %>% 
  # rename columns ----
  janitor::clean_names() %>% 
  dplyr::rename(sw_species = "common_names",
                fao_species = "fao_common_name", 
                fda_species = "fda_common_name",
                water_body = "bo_ws", # so curious as to why this is what janitor::clean_names does to BOWs...
                country = "countries",
                method = "methods",
                score = "assessment_score",
                escapes_score = "c6score", # 
                rec = "assessment_color"
                ) %>% 
  # select relevant columns
  dplyr::select(report_title, published_date, sw_species, scientific_name, fao_species, fda_species, country, water_body, method, escapes_score, score, rec) %>%
  # filter out freshwater aquaculture
  filter(!method %in% c("Freshwater net pen")) %>% 
  # set scores less than 0 equal to 0
  mutate(score = ifelse(score < 0, 0, score))

## Change species names using FAO species name (fao_species); if NA, use common name (sw_species)
sw_sus$species <- ifelse(!is.na(sw_sus$fao_species), sw_sus$fao_species, sw_sus$sw_species)
```

### Keep NA countries

```{r keep-na-countries_sw-sus}

## These need to be re-added later (get cut when associated with region ids)
sw_sus_no_rgn <- filter(sw_sus, is.na(country) | country == "Worldwide")
dim(sw_sus_no_rgn)

# in 2023 assessment there were 114 "Worldwide" rows and only 1 NA row
# 115 entries in 2024
# 114 entries in 2025
```

### Convert country names to OHI region IDs.

```{r sw-sus_convert-region}

# ---- Change/update country names to match OHI region names ----
# Separate rows that are associated with multiple countries
sw_sus_multiple <- sw_sus %>% 
  filter(str_detect(country, "\\|")) %>% # Detects strings with "|" in them
  separate_rows(country, sep = " \\| ")

# Rejoin the rows we just removed, but in clean format (1 row per country)
sw_sus_df <- sw_sus %>%
  filter(!str_detect(country, "\\|")) %>%
  rbind(sw_sus_multiple) %>%
  # remove all NA and 'Worldwide' rows in order to correctly rejoin those in the next steps, these rows are stored in the variable sw_sus_no_rgn
  filter(!is.na(country), country!="Worldwide")
  
# Convert country names to OHI region IDs. (ohicore/R/name_2_rgn.R)
sw_sus_rgn <- ohicore::name_2_rgn(df_in = sw_sus_df, 
                       fld_name='country', 
                       flds_unique=c('fao_species', 'fda_species', 'sw_species', 'score'),
                       keep_fld_name = TRUE) # 'country' now shows the original Seafood Watch data name; 'rgn_name' is what we want to use from now on
# v2023: 151 entries
# v2024: (used same data, 151 entries)

# v2024 duplicates found: 
# 1 Chile         
# 2 Norway        
# 3 United Kingdom
# 4 United States 
# 5 Canada  
# after investigating the data further, these duplicates don't seem to be an issue -- each entry is an unique species and/or based on an unique report

# v2025 duplicates found:
# 1 United States 
# 2 Chile         
# 3 Indonesia     
# 4 Norway        
# 5 United Kingdom
# 6 Canada 
# Everything is the same as above except we've added Indonesia... will look into this more. 

# Add NA countries back in 
sw_sus_rgn <- bind_rows(sw_sus_rgn, sw_sus_no_rgn) %>%
  unique()
dim(sw_sus_rgn)
# 259 entries
# v2024: same (because same exact data)
# v2025: 267 entries

#2023 switched to comparing common names because the scientific name column was two columns in the previous years data. Could compare scientific names next year. No differences in species detected. 

# 2025: 2 species in each setdiff detected

setdiff(sw_sus_rgn$sw_species, old_sw_sus$CommonNames) # "Yellowtail" "Buri"
setdiff(old_sw_sus$CommonNames, sw_sus_rgn$sw_species) # "Buri amberjack"       "Yellowtail amberjack"

# 2025 data has changed the names from Yellowtail amberjack --> Yellowtail and Buri amberjack --> Buri. The scientific names are the same for these two. Going off the 2023 note, I'll run these setdiff's again with scientific name.

setdiff(sw_sus_rgn$scientific_name, old_sw_sus$ScientificName)
setdiff(old_sw_sus$ScientificName, sw_sus_rgn$scientific_name)
# The only difference here was the overarching "seaweed" species. the scientific names are listed in different order/have slightly different composition between 2023 and 2025. 
```

Check that the non-matches between Seafood Watch sustainability data species (sw_sus_rgn) and the FAO mariculture species in the wrangled FAO Aquaculture Production data table (maric) are not due to spelling errors or slightly different names. We want to include as many species that have sustainability scores as possible

# 2025 update

Since we have new SFW data, we've decided to combine data from v2024 and v2025 to have two years of data and to assist with gapfilling. In order for the final output csv `mar_sustainability_score.csv` to have the correct final values, we also need to combine the mariculture data up here at the top. 

```{r read-mar-data}
# Read in the combined MAR data we just wrote. Most of the final ouputs use the combined data, so it will be the main file here
maric <- readr::read_csv(here::here(current_mar_dir, "output", "MAR_FP_data_comb.csv"))

# We will also use the uncombined data, but only for looking at mar_harvest_tonnes
maric_uncomb <- readr::read_csv(here::here(current_mar_dir, "output", "MAR_FP_data.csv"))

```

## 2024 assessnent SFW data

I've gone back to v2024 to export this intermediate file for ease of combining here. It's called `int_sfw.csv`

```{r combine-new-old-sfw-data}
# Save intermediate file for 2026 analysis
# readr::write_csv(sw_sus_rgn, here(current_mar_dir, "int", "int_sfw.csv"))

# Read in previous analysis data
sfw_int_old <- read_csv(here(previous_mar_dir, "int", "int_sfw.csv"))

# There are 3 duplicates present in the old data
sfw_int_old %>%
  dplyr::count(rgn_id, species, rgn_name, method, published_date, report_title, water_body, scientific_name, sw_species, fao_species, fda_species, country, rec) %>%
  dplyr::filter(n > 1) 

nrow(sfw_int_old) # 259

# Drop duplicate rows
sfw_int_old_clean <- sfw_int_old %>%
  distinct(rgn_id, species, rgn_name, method, published_date, report_title, 
           water_body, scientific_name, sw_species, fao_species, fda_species, 
           country, rec, .keep_all = TRUE)

# Check that the 4 duplicate rows got dropped
nrow(sfw_int_old_clean) # 255


# Filter down before combining
sfw_int_old <- sfw_int_old %>% 
  select(escapes_old = escapes_score, score_old = score, species, rgn_id, rgn_name, scientific_name, sw_species, fao_species, water_body, method, rec, country)


# Add a new year column that is the version year
sfw_int_old <- sfw_int_old %>% 
  mutate(vyear = 2024) 

sw_sus_rgn <- sw_sus_rgn %>% 
  mutate(vyear = 2025) %>% 
  # Rename for pivoting later on
  rename(score_new = score, escapes_new = escapes_score) %>% 
  select(-c(report_title, fda_species, published_date))

# Join previous and current year data
sw_sus_rgn_comb <- full_join(sw_sus_rgn, sfw_int_old, by = join_by(rgn_id, species, method, water_body, rec))

# Clean names
sw_sus_rgn_comb <- sw_sus_rgn_comb %>% 
  select(rgn_id, species, score_new, method, rec, water_body, score_old,
         escapes_new, escapes_old, scientific_name_new = scientific_name.x,
         scientific_name_old = scientific_name.y, rgn_name_new = rgn_name.x,
         rgn_name_old = rgn_name.y, vyear_new = vyear.x, vyear_old = vyear.y,
         country_new = country.x, country_old = country.y)

# Check NAs for gap filling
colSums(is.na(sw_sus_rgn_comb))

# Gap fill with other year data
sw_sus_rgn_clean <- sw_sus_rgn_comb %>%
  mutate(
    # Sustainability score
    score_new = ifelse(is.na(score_new), score_old, score_new),
    score_old = ifelse(is.na(score_old), score_new, score_old),
    # Escapes score
    escapes_new = ifelse(is.na(escapes_new), escapes_old, escapes_new),
    escapes_old = ifelse(is.na(escapes_old), escapes_new, escapes_old),
    # Version year
    vyear_new = ifelse(is.na(vyear_new), current_year, vyear_new),
    vyear_old = ifelse(is.na(vyear_old), previous_year, vyear_old),
    # Scientific name
    scientific_name_new = ifelse(is.na(scientific_name_new), scientific_name_old, scientific_name_new),
    scientific_name_old = ifelse(is.na(scientific_name_old), scientific_name_new, scientific_name_old),
    # Region name
    rgn_name_new = ifelse(is.na(rgn_name_new), rgn_name_old, rgn_name_new),
    rgn_name_old = ifelse(is.na(rgn_name_old), rgn_name_new, rgn_name_old),
    # Country
    country_new = ifelse(is.na(country_new), country_old, country_new),
    country_old = ifelse(is.na(country_old), country_new, country_old),
    )

# Confirm NAs are gone for columns of interest
colSums(is.na(sw_sus_rgn_clean))

# Filter down to desired columns
sw_sus_rgn_filt <- sw_sus_rgn_clean %>%
  select(rgn_id, species, method, water_body, score_new, score_old, escapes_new, escapes_old, vyear_old, vyear_new, rgn_name = rgn_name_new, country = country_new, scientific_name = scientific_name_new)

# Pivot longer so data is in tidy format
sw_sus_rgn_long <- sw_sus_rgn_filt %>%
  pivot_longer(
    cols = c(score_new, score_old, escapes_new, escapes_old, vyear_old, vyear_new),
    names_to = c(".value", "year_old"),
    names_sep = "_"
  ) %>% 
  select(-year_old) %>% 
  rename(escapes_score = escapes)

table(sw_sus_rgn_long$vyear)

# Reassign the combined mariculture data to maric for further processing
sw_sus_rgn <- sw_sus_rgn_long
```

```{r baseline-species-check}
## Make sure same species are spelled the same in the two data tables (e.g. check that there are no extra spaces)
sort(setdiff(sw_sus_rgn$species, maric$species)) # Seafood Watch species with no FAO data
sort(setdiff(maric$species, sw_sus_rgn$species)) # FAO species with no Seafood Watch data - there will probably be a long list
#this is a baseline check, run the fixes in the next chunk from previous years before hand fixing any species
```

**Instructions for updating species:**

Hand check each of the species output here, check to make sure there aren't obvious cases when the species is the same, but the names are presented slightly differently in the lists. For example, some species might have an extra space in the name somewhere. If any of these obvious differences exist, we need to go in and change the names to match. 

The FAO list has far more species than the sustainability list. But, go over them to make sure that we are getting sustainability scores for as many species as we can.

note -- Seafood Watch makes a distinction between Common sole and Dover sole

**2025 note**: A lot of the non-matches are because FAO classifies a species as freshwater and thus it gets filtered out early on. The freshwater SFW species will be dropped here because we do not want to include them in our final analysis. See [issue #338](https://github.com/OHI-Science/globalfellows-issues/issues/338) for more information about the species list. 

Fix the non-matches

```{r sw-sus_fix-non-matches}

# Rename species in Seafood Watch data to match FAO species data 
sw_sus_rgn$species <- gsub("Atlantic Bluefin tuna", "Atlantic bluefin tuna", sw_sus_rgn$species)

# ---- Drop freshwater SFW species ----
sw_sus_rgn <- sw_sus_rgn %>% 
  filter(!species %in% c("American eel", "American yellow perch", "Atlantic sturgeon", "Blue tilapia", "Channel catfish", "Danube sturgeon(=Osetr)", "Giant river prawn", "Hybrid red tilapia", "Hybrid tilapia", "Mozambique tilapia", "Nile tilapia", "NorthAfrican catfish", "Red swamp crawfish", "Sabaki tilapia", "Siberian sturgeon", "Sterlet sturgeon", "Striped catfish", "Wami tilapia", "Wels(=Som) catfish", "White sturgeon"))

# ---- Fix FAO species names ----
# rename any "mistakes" in the FAO data, such as species names whose first word is uncapitalized  
maric$species <- gsub("blood cockle", "Blood cockle", maric$species)
maric$species <- gsub("\\[Solea spp\\]", "Common sole", maric$species)

# rename FAO species to general categories when used in Seafood Watch data 
maric <- maric %>% 
  # Oysters
  mutate(species = case_when(
    family == "Cardiidae" ~ "Cockles",
    str_detect(species,"oyster") ~ "Oysters",
    # Clams
    species == "CORBICULIDAE" ~ "Clams",
    str_detect(species, "clam") ~ "Clams",
    str_detect(species, "Clam") ~ "Clams",
    str_detect(species, "mussel") ~ "Mussels",
    taxon_code == "AL" ~ "Seaweed",
    family == "Haliotidae" ~ "Abalones", 
    family == "Pectinidae" ~ "Scallops",
    TRUE ~ species)
    ) %>% 
  mutate(family = case_when(species == "Clams" ~ NA,
                            species == "Oysters"~ NA,
                            TRUE ~ family
  )) #setting family as NA when the group contains more than one family


#updating the names combines species and creates creates duplicate species which we will need to sum
#also sums duplicates due to multiple fao codes for the same region/species
maric <- maric %>%
  group_by(environment, species, year, vyear, taxon_code, rgn_id, family) %>% 
  summarize(value = sum(value), gap_0_fill = first(gap_0_fill)) %>%
  ungroup()

#run this code to check that there are no duplicates for any species, year, rgn_id combinations. If there are they will be unintentionally deleted later, so you may need to change family or taxon codes to match and rerun the above group by
duplicate_check <- maric %>% 
  group_by(rgn_id, species, year, vyear, environment) %>% 
  summarize(n = n()) 
unique(duplicate_check$n) #should be 1

# Incorrect seaweed families were causing duplicate issues for the combined data, so drop those here
maric <- maric %>% 
  filter(!(rgn_id == 164 & family == "Solieriaceae")) %>% 
  filter(!(rgn_id == 216 & family == "Caulerpaceae"))
 
# Run the duplicate check again
duplicate_check <- maric %>% 
  group_by(rgn_id, species, year, vyear, environment) %>% 
  summarize(n = n()) 
unique(duplicate_check$n) #should be 1


#Add a unique identifier per cultivated stock that describes each species, region, and environment grouping.
identifier <- maric %>% 
  select(rgn_id, species, environment) %>% 
  unique() %>% 
  mutate(species_code = 1:n())

# Not 100% sure if this is joining correctly... need to check out later
maric <- left_join(maric, identifier, by = join_by(rgn_id, species, environment))

# It looks like all the identifiers applied correctly 
length(unique(maric$species_code)) # 1501
nrow(identifier) # 1501

# run again after cleaning to check there are less mismatched species. The remaining ones will be gapfilled. 
sort(setdiff(sw_sus_rgn$species, maric$species))
sort(setdiff(maric$species, sw_sus_rgn$species))

nrow(sw_sus_rgn) # 494


## Uncombined data
# ---- Fix FAO species names ----
# rename any "mistakes" in the FAO data, such as species names whose first word is uncapitalized  
maric_uncomb$species <- gsub("blood cockle", "Blood cockle", maric_uncomb$species)
maric_uncomb$species <- gsub("\\[Solea spp\\]", "Common sole", maric_uncomb$species)

# rename FAO species to general categories when used in Seafood Watch data 
maric_uncomb <- maric_uncomb %>% 
  # Oysters
  mutate(species = case_when(
    family == "Cardiidae" ~ "Cockles",
    str_detect(species,"oyster") ~ "Oysters",
    # Clams
    species == "CORBICULIDAE" ~ "Clams",
    str_detect(species, "clam") ~ "Clams",
    str_detect(species, "Clam") ~ "Clams",
    str_detect(species, "mussel") ~ "Mussels",
    taxon_code == "AL" ~ "Seaweed",
    family == "Haliotidae" ~ "Abalones", 
    family == "Pectinidae" ~ "Scallops",
    TRUE ~ species)
    ) %>% 
  mutate(family = case_when(species == "Clams" ~ NA,
                            species == "Oysters"~ NA,
                            TRUE ~ family
  )) #setting family as NA when the group contains more than one family


#updating the names combines species and creates creates duplicate species which we will need to sum
#also sums duplicates due to multiple fao codes for the same region/species
maric_uncomb <- maric_uncomb %>%
  group_by(environment, species, year, taxon_code, rgn_id, family) %>% 
  summarize(value = sum(value), gap_0_fill = first(gap_0_fill)) %>%
  ungroup()

#run this code to check that there are no duplicates for any species, year, rgn_id combinations. If there are they will be unintentionally deleted later, so you may need to change family or taxon codes to match and rerun the above group by
duplicate_check <- maric_uncomb %>% 
  group_by(rgn_id, species, year, environment) %>% 
  summarize(n = n()) 
unique(duplicate_check$n) #should be 1


#Add a unique identifier per cultivated stock that describes each species, region, and environment grouping.
identifier <- maric_uncomb %>% 
  select(rgn_id, species, environment) %>% 
  unique() %>% 
  mutate(species_code = 1:n())

# Not 100% sure if this is joining correctly... need to check out later
maric_uncomb <- left_join(maric_uncomb, identifier, by = join_by(rgn_id, species, environment))

# It looks like all the identifiers applied correctly 
length(unique(maric_uncomb$species_code)) # 1350
nrow(identifier) # 1350

# run again after cleaning to check there are less mismatched species. The remaining ones will be gapfilled. 
sort(setdiff(sw_sus_rgn$species, maric_uncomb$species))
sort(setdiff(maric_uncomb$species, sw_sus_rgn$species))
```

# FAO mariculture and Seafood Watch sustainability scores

