---
title: 'OHI 2021: Mariculture resilience'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

**NOTE: This layer was not added to the OHI yet. We will hopefully revise and add for v2022.**

# Summary

# Updates from previous assessment
* Updated sustainability scores from seafood watch.

***

# Data Source 

## Seafood Watch sustainability data

**Reference**: 
https://www.seafoodwatch.org/-/m/sfw/pdf/whats%20new/complete%20recommendation%20list.pdf
Release date: August 3, 2020

**Downloaded**: Auguest 3, 2020

**Description**: Monterey Bay Aquarium Seafood Watch aquaculture recommendations. Sustainability scored from 0-10. Rescaled to 0-1.

***
  
```{r setup, message=FALSE, warning=FALSE}

knitr::opts_chunk$set(eval=FALSE)
  
## Load libraries, set directories
library(ohicore)  #devtools::install_github('ohi-science/ohicore@dev')
library(dplyr)
library(stringr)
library(tidyr)
library(tidyverse)
library(taxize)
library(knitr)
library(kableExtra)

## Load user-defined functions
source('../../../workflow/R/common.R') # directory locations

```



# Sustainability Scores from Seafood Watch Data

## Import data: Seafood Watch sustainability scores

These data describe the sustainability country/species combinations.

```{r sw-scores, eval = FALSE}

## Load in Seafood Watch sustainability scores data from mazu:
sw_sus_raw <- read.csv(file.path(dir_M, 'git-annex/globalprep/_raw_data/seafood_watch_mar_sustainability/d2020/Seafood-Watch_aquaculture-recs_July-2020.csv'), check.names = FALSE, stringsAsFactors = FALSE, na.strings = c("NA", ""))

head(sw_sus)

```

## Wrangle

### Tidy Seafood Watch sustainability data

Rename columns and fill in species column

```{r tidy_sw-sus, eval = FALSE}

## Rename columns
sw_sus <- sw_sus_raw %>%
  rename(report_title = 'Report Title',
         start_year = 'Start year',
         sw_species = 'Common name',
         genus = 'Genus',
         spp = 'Species',
         fao_species = 'FAO Common name',
         region = 'Region',
         country = 'Country',
         state_territory = 'State/Territory',
         sub_region = 'Sub-Region',
         water_body = 'Body of Water',
         parent_method = 'Parent Method',
         method = 'Method',
         score = 'Overall Score',
         data_score = 'AqCriteria1',
         effluent_score = 'AqCriteria2',
         habitat_score = 'AqCriteria3', 
         chemical_score = 'AqCriteria4',
         feed_score = 'AqCriteria5',
         escapes_score = 'AqCriteria6',
         pathogen_score = 'AqCriteria7',
         stock_score = 'AqCriteria8',
         mortality_score = 'AqCriteria9',
         sec_species_score = 'AqCriteria10',
         rec = 'Overall Recommendation'
         ) 

## Change species names using FAO species name (fao_species); if NA, use common name (sw_species)
sw_sus$species <- ifelse(!is.na(sw_sus$fao_species), sw_sus$fao_species, sw_sus$sw_species)

```

### Keep NA countries

```{r keep-na-countries_sw-sus, eval = FALSE}

## These need to be re-added later (get cut when associated with region ids)
sw_sus_no_rgn <- filter(sw_sus, is.na(country))
  # 203 entries with no country

```

### Convert country names to OHI region IDs.

```{r sw-sus_convert-region, eval = FALSE}

## Change country names to match OHI region names
sw_sus <- sw_sus %>% 
  mutate(country = ifelse(country=="Korea, the Republic of", "South Korea", country)) %>% # Data removed for not having a match; change name to match
  mutate(country = ifelse(country=="United Kingdom of Great Britain and Northern Ireland (the)", "United Kingdom", country))  # Data removed for not having a match; change name to match

## Convert country names to OHI region IDs. (ohicore/R/name_2_rgn.R)
sw_sus_rgn <- name_2_rgn(df_in = sw_sus, 
                       fld_name='country', 
                       flds_unique=c('fao_species', 'sw_species', 'region', 'score'),
                       keep_fld_name = TRUE) # 'country' now shows the original Seafood Watch data name; 'rgn_name' is what we want to use from now on
  # Goes from 330 obs. to 127 obs. (because 203 obs. have no country associated)


## Re-add NA countries
sw_sus_rgn <- bind_rows(sw_sus_rgn, sw_sus_no_rgn) %>%
  unique()
  # Back to 330 obs.

```

```{r match-type_sw-sus-data, eval = FALSE}

## Add column "match_type" to categorize obs. that 1) have a country associated, 2) are "global" Seafood Watch data, or 3) only distinguished by water body
sw_sus_rgn <- sw_sus_rgn %>% 
  dplyr::mutate(match_type = dplyr::case_when(!is.na(country) ~ "sw_sp_c", # Add match type specific to species/country
                                str_detect(report_title, regex("Global")) ~ "sw_sp_g", # Add match type specific to species/global Seafood Watch data
                                TRUE ~ "sw_sp_w" # Add match type specific to species/water body
  ))

table(sw_sus_rgn$match_type)

```

**Assign a mediterraen bordering rgn id to each waterbody row**
```{r gf_sw_sp_w, eval = FALSE}
## Add a line for each country that borders each water body
  # The only water body is the Mediterranean; add all countries that border it (source: https://www.medqsr.org/mediterranean-marine-and-coastal-environment#:~:text=Today%2021%20countries%2C%20with%20surface,Syria%2C%20Tunisia%2C%20and%20Turkey.)

## filter for waterbody specific rows 
sw_sus_rgn_water <- sw_sus_rgn %>%
  filter(match_type == 'sw_sp_w')

## define meditteraen rgns
med_rgns <- data.frame(rgn_id = c(82, 84, 232, 187, 81, 214, 179, 80, 79, 184, 78, 67, 68, 185, 186, 62, 188, 182, 77, 61, 76))

expand.grid.df <- function(...) Reduce(function(...) merge(..., by=NULL), list(...))

gf_sw_sp_w <- expand.grid.df(sw_sus_rgn_water, med_rgns) %>%
  dplyr::select(-rgn_id.x, rgn_id = rgn_id.y)


```


```{r sw-joining, eval = FALSE}

## add back the water region scores 

sw_sus_rgn_all <- sw_sus_rgn %>%
  filter(match_type != "sw_sp_w") %>%
  rbind(gf_sw_sp_w)

test <- sw_sus_rgn_all %>%
  filter(!is.na(rgn_id)) %>%
  group_by(rgn_id) %>%
  summarise(sust = mean(score))

final_data <- sw_sus_rgn_all %>%
  filter(!is.na(rgn_id)) %>%
  group_by(rgn_id) %>%
  summarise(resilience.score = mean(score)*0.1) %>%
  ungroup()

write.csv(final_data, "output/res_sfw.csv", row.names = FALSE)
```


