---
title: 'OHI 2025 - Mariculture Population Pressure Layers'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---

[REFERENCE RMD FILE](https://ohi-science.org/ohiprep_v2021/globalprep/mar_prs_population/v2021/mar_prs_pop_dataprep.html)


# Summary

This document describes the steps for obtaining and wrangling the data used to calculate the mariculture population pressure sub-dimension for the 2025 global assessment. The general data preparation calculations are summarized [here](http://ohi-science.org/ohi-global/layers.html#inland_coastal_population). For context and explanation see the mariculture (subgoal of food provision) [model summary](http://ohi-science.org/ohi-global/goals.html#food_provision:_mariculture).

# Updates from previous assessments

* Since SEDAC was terminated (see below), we opted to use [Oak Ridge National Library LandScan Global](https://landscan.ornl.gov/) data for a couple of reasons:

  1. Data are still being updated
  2. There are data from year 2000 onward (can recalculate old scores)

* [Posted June 17, 2025](https://www.earthdata.nasa.gov/news/data-from-sedac-available-again-earthdata-search): On April 30, 2025, the contract for NASA's Socioeconomic Data and Applications Center (SEDAC), managed by Columbia University's Center for Integrated Earth System Information (CIESIN), was terminated and the datasets SEDAC curated were no longer available. Since then, the Earth Science Data and Information System (ESDIS) project has been working to get the datasets rehoused. Starting today, all SEDAC datasets have been relocated to Earthdata Cloud and can be found in Earthdata Search. All of the datasets are complete and will no longer be updated with new versions or collections. The Dataset Lifecycle Data Basics page explains collection progress states for more information.

* We are no longer reprojecting and resampling population data to World Mollweide projection, as that affected the population data for "border" regions (i.e. those on the border of where the projection wraps around) far too much. 
  
  * We have instead modified the eez_25mi_inland raster to WGS 1984 projection using "nearest neighbor" method for resampling and not modified the population rasters. 

* In addition, because the previous script used the raster package, this new script has been changed to utilize the terra package instead. 
  
  * Some processes are less iterative in the terra package so must be done in sequence, however the processes that can function iteratively have been made so. 

***

# Data Sources

## Oak Ridge National Library LandScan

As recommended by [Yin X, Li P, Feng Z, Yang Y, You Z, Xiao C., 2021](https://www.mdpi.com/2220-9964/10/10/681).

**Reference**: [LandScan Global](https://landscan.ornl.gov/)

Lebakula, V., Epting, J., Moehl, J., Stipek, C., Adams, D., Reith, A., Kaufman, J., Gonzales, J., Reynolds, B., Basford, S., Martin, A., Buck, W., Faxon, A., Cunningham, A., Roy, A., Barbose, Z., Massaro, J., Walters, S., Woody, C., … Urban, M. (2024). LandScan Silver Edition [Data set]. Oak Ridge National Laboratory. https://doi.org/10.48690/1531770

**Downloaded**: 30 June 2025

**Description**:
The LandScan Program was initiated at Oak Ridge National Laboratory (ORNL) in 1997 to address the need for improved estimates of population for consequence assessment. For example, natural and manmade disasters across the globe place vast populations at risk, often with little or no advance warning. This motivated the need for a more realistic population distribution that could capture where people are likely to be throughout the course of a day, not just residential locations as are reported in official censuses. Further, it was critical to develop highly resolved estimates so that they were useful to evaluate to events across multiple geographic scales. This initial effort is what is known as LandScan Global and has been updated annually since the first version in 1998. [See more.](https://landscan.ornl.gov/about)

**Native data resolution**: 30 arc-seconds

**Time range**: 2000-2023

**Format**: GeoTiff

**2025 NOTE:**
The population trends for some of the smaller (potentially more data-deficient) countries are a little odd. We explored this a little in issue [#336](https://github.com/OHI-Science/globalfellows-issues/issues/336). We concluded that it was likely a modeling issue on ORNL's end.

## UN Population

UN population data is used in this data prep routine to check population counts; spatially distributed counts derived from the gridded world population are aggregated by region and checked against the UN population estimates.

**Reference**: https://population.un.org/wpp/

- Data were downloaded for the following options:

  - File type: Standard Projections
  - Major Topic/Special Groupings: Most Used
  - Sub Group: Demographic Indicators
  - Files: Compact

**Downloaded**: 23 June 2025

**Description**: Population (in thousands) for countries.

**Native data resolution**: Country scores

**Time range**: 1950-2023

**Format**: Excel file

**Data cleaning process**: 

Values from the "ESTIMATES" tab of `WPP2024_GEN_F01_DEMOGRAPHIC_INDICATORS_COMPACT.xlsx` were copied for all countries and years (1950-2023) and pasted into new `UN_pop_clean_v2025.csv`. 

**2025 NOTE:**
The format of the UN population data changed, providing population estimates for January and July. So, I decided to take the total population from the "Total Population January" column. These numbers are similar to past data (i.e. `v2024/output/UN_population.csv`) but not exactly the same. 

***

# Setup

Load all relevant libraries including parallel processing packages
```{r setup, message=FALSE, warning=FALSE, verbose=FALSE, eval=FALSE}

# Set options for all chunks in code
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.width = 6, fig.height = 4, fig.path = "figs/")

# Load packages, installing them first where necessary
pkg <- c("raster", "sp", "sf", "fasterize", "tidyverse", "foreach", "parallel","doParallel")
new_pkg <- pkg[!(pkg %in% installed.packages())]
if (length(new_pkg)){install.packages(new_pkg)}
if (!("ohicore" %in% installed.packages())){devtools::install_github("ohi-science/ohicore")}

library(ohicore)

# Spatial libraries
library(raster)
library(sp)
library(fasterize)
library(sf) # Vector spatial data
library(terra) # Raster spatial data

# Data wrangling libraries
library(tidyverse)
library(here)
library(plotly)
library(dplyr)

# Parallel processing libraries
library(parallel)
library(foreach)
library(doParallel)

```

Define frequently used pathnames. Change scenario and data years in file pathnames code chunk to reflect the most recent data (d) and current assessment year (v).
```{r file paths, eval=FALSE}

# Source common and spatial common files
source('http://ohi-science.org/ohiprep_v2020/workflow/R/common.R')

## Update these!
scenario_yr <- "v2025" # change to reflect assessment year!
data_yr_gpw <- "d2025" # change to reflect year of most recently downloaded Gridded Population of the World (GPW) data! (2025 for 2025 assessment)
data_yr_un_pop <- "d2025" # change to reflect year of most recently downloaded UN population data! (changed to 2025 for 2025 assessment)

# Define commonly used file paths (matched to what we did in ico_data_prep.Rmd)
dir_server <- file.path(dir_M, "git-annex/globalprep/_raw_data")
dir_github <- here("/globalprep/mar_prs_population", scenario_yr)

## DOUBLE CHECK
# Checking to see if there is a README in the mar_prs_pop and mar_prs_pop/v20?? folders - all goal prep files will need a README!
if(!file.exists(file.path(dir_github, 'README.md'))) {
  warning(sprintf('No README detected in %s', dir_github))
}
if(!file.exists(file.path(dirname(dir_github), 'README.md'))) {
  warning(sprintf('No README detected in %s', file.path(dirname(dir_github))))
}

```

## Import Raw Data

```{r import raw data, eval=FALSE}

# Read in the raw population count data
raw <- list.files(file.path(dir_server, "LandScan_global_population"),
                  full.names = TRUE, pattern = "\\.tif$",
                  recursive = TRUE)

raw <- raw[!grepl("-colorized", raw)] # filter out the "colorized" files

raw # check that this looks correct; can double check with the folder on Mazu server folder

# Check if the directory exists, and create it if it doesn't
if (!dir.exists(file.path(dir_M, "git-annex/globalprep/mar_prs_population/", scenario_yr, "int/"))) {
  dir.create(file.path(dir_M, "git-annex/globalprep/mar_prs_population/", scenario_yr, "int/"))
} else {
  cat("Directory already exists.")
}

# Rename files for simplicity
for (i in seq_along(raw)) {
  
  # Identify variable name
  name <- paste0("raw", i)
  
  # Identify year for file name
  year <- 2000 + (i-1)
  
  # Load data to variable name and save to Mazu with year included
  assign(name, terra::rast(raw[i]) %>% 
  terra::writeRaster(., filename = file.path(dir_M, sprintf("git-annex/globalprep/mar_prs_population/%s/int/human_count_%s_mol.tif", scenario_yr, year)), overwrite = TRUE))
  
}

ohi_rasters() # loads ohi rasters

# Import raw, cleaned UN population data to be wrangled then used to confirm gpw-derived spatial population
pop <- read.csv(file.path(dir_server, 
                          sprintf("UnitedNations_population/%s/UN_pop_clean_v2025.csv",
                                  data_yr_un_pop)), # change year if updated
                strip.white = TRUE, stringsAsFactors = FALSE, 
                # Add NA to blank cells
                na.strings = c(""))
```
## Crop rasters to common extent

The rasters from LandScan vary *slightly* in extent between years. So, extents were compiled into a data frame and then the largest common extent was determined. Rasters were cropped to this extent rather than filled with NA values, but it would be worth keeping an eye on the extents for future years to make sure not significant areas are being cut out.

```{r Crop Rasters, eval=FALSE}

# Initiate an empty list
rasters <- list()

# Add each raw raster to the list to iterate over
for (i in 1:24) {
  rasters[[i]] <- get(paste0("raw", i))
}

# Build data frame of extents from the rasters list
extents_df <- do.call(rbind, lapply(seq_along(rasters), function(i) {
  ex <- ext(rasters[[i]])
  data.frame(
    raster = paste0("raster_", i),
    xmin = ex[1],
    xmax = ex[2],
    ymin = ex[3],
    ymax = ex[4]
  )
}))

# Compute common overlapping extent
common_extent <- ext(
  max(extents_df$xmin),  # max of left edges
  min(extents_df$xmax),  # min of right edges
  max(extents_df$ymin),  # max of bottom edges
  min(extents_df$ymax)   # min of top edges
)

# Initiate empty list for cropped rasters
cropped_list <- list()

# Crop each raster and save to the cropped list
for (i in seq_along(rasters)) {
  cropped_list[[i]] <- crop(rasters[[i]], common_extent)
}
```

# Methods and Calculations

## Append or Extract by OHI Regions

Stack rasters of population counts for each year, sum within regions (using `eez_25mi_inland`) for each year, and save this as `coastal_pop_zonal_sums.csv` in the intermediate folder. Gather this into long format dataframe, and save as `output/mar_pop_25mi.csv`.

Note: The resampled and reprojected `eez_25mi_inland` raster can be sourced from `git-annex/globalprep/spatial/v2024/eez_25mi_inland_wgs1984.tif`. The steps taken to resample and reproject this raster can be found below in the Appendix.

```{r extract population data from all population layers, eval=FALSE}

# Load files and create a SpatRaster of the human_count files (each layer is going to be a year) ---- (these should be each of the "raw#" rasters)

pop_all_years <- terra::rast(cropped_list) # Use terra::rast() on list files to make a spatraster with as many layers as there are files 

# Check it out 
(pop_all_years) # Has 24 layers, one for each of our raster files 

# Read in EEZ plus 25mi inland WGS 1984*** raster and tidy for zonal statistics ---- 
# This was reprojected in 2024, so we can load from the 2024 assessment. See Appendix for more methods.
zone_raster <- terra::rast(file.path(dir_M, "git-annex/globalprep/spatial/v2024/eez_25mi_inland_wgs1984.tif"))

# Need to make it a categorical raster for zonal statistics. See what it loads in as?
terra::datatype(zone_raster) # [1] "INT4U"

zone_raster <- terra::as.factor(zone_raster) # Ensuring R knows its a categorical raster. Won't have a datatype anymore though

# Need to check names to make sure they are 'zone' for zonal statistics
names(zone_raster) # [1] "ocean_plus25mile_inland"
names(zone_raster) = c("zone") # Ensuring R knows that the layer we want defines ZONES!

# Check it out --> resolution, layer name
(zone_raster)

# Conduct zonal statistics, tidy up and write out ----
# Crop to same common extent as population count rasters
zone_raster <- crop(zone_raster, common_extent)

# Conduct zonal statistics
coastal_pop <- terra::zonal(pop_all_years, zone_raster, "sum", na.rm=TRUE) 

# Check it to see the column names --> fix these in following step
View(coastal_pop)

# Rename just in case
coastal_pop2 <- coastal_pop

# Fix column names to "pop_" and year
colnames(coastal_pop2)[2:ncol(coastal_pop2)] <- paste0("pop_", str_extract(colnames(coastal_pop2)[2:ncol(coastal_pop2)], "\\d+"))

# Writing to 'int' folder in project to have zonal sums 
write_csv(coastal_pop2, file.path(here("globalprep/mar_prs_population/", scenario_yr, "/int/coastal_pop_zonal_sums.csv")))

# Pivoting longer to have one "rgn_id" column, one "year" column, and one "popsum" column
coastal_pop3 <- coastal_pop2 %>% 
  pivot_longer(cols = -c(zone), names_to = "year", values_to = "popsum") %>% 
  mutate(year = str_remove_all(year, "pop_"),
         year = as.numeric(year)) %>% 
  dplyr::rename(rgn_id = zone) %>%
  dplyr::mutate(rgn_id = as.numeric(rgn_id)) %>% 
  dplyr::filter(rgn_id <= 250) 

# Converting NaN's to 0's (mostly occur with uninhabited islands)
coastal_pop_nona <- coastal_pop3 %>% 
  dplyr::mutate(popsum = case_when(
    popsum == "NaN" ~ 0,
    .default=popsum 
  ))

# Writing it out to 'ouput' folder to have the coastal population of each country for all years 
write_csv(coastal_pop_nona, file.path(here("globalprep/mar_prs_population/", scenario_yr, "/output/mar_pop_25mi.csv")))
```

## Calculate Area of 25mi Inland Buffers

```{r calculate area and mar. pressure, eval=FALSE}

# Load in data calculated in spatial folder 
area <- read_csv(here("globalprep/spatial/v2019/output/area_km2_25mi_inland.csv")) # last updated v2019

```

## Calculate Mariculture Population Pressure

```{r calculate mar. population pressure, eval=FALSE}

# Rescale organized coastal population data
pop_rescaled <- coastal_pop_nona %>%
  left_join(area, by="rgn_id") %>%
  mutate(density = popsum/area_km2) %>%
  mutate(ln_density = log(density + 1)) %>%
  mutate(scalar = max(ln_density, na.rm = TRUE)) %>%
  mutate(dens_rescaled = ln_density/scalar) %>%
  mutate(dens_rescaled = ifelse(dens_rescaled > 1, 1, dens_rescaled))

filter(pop_rescaled, is.na(area_km2)) # no NA values for the area column (Antarctica not included)

pressure_data <- pop_rescaled %>%
  dplyr::select(rgn_id, year, pressure_score = dens_rescaled)

write_csv(pressure_data, file.path(here("globalprep/mar_prs_population/", scenario_yr, "/output/prs_pop_density.csv")))

```

## Gapfilling

No gapfilling was completed as part of this data prep methodology. Datasets are saved with "_gf" appended just to indicate they are finalized versions.

```{r save prs_pop_density and mar_pop_25mi layers, eval=FALSE}

prs <- read.csv(file.path("globalprep/mar_prs_population/", scenario_yr, "/output/prs_pop_density.csv")) %>%
  dplyr::mutate(pressure_score = 0)

write.csv(prs, file.path(here("globalprep/mar_prs_population/", scenario_yr, "/output/prs_pop_density_gf.csv")), row.names = FALSE)

mar <- read.csv(file.path("globalprep/mar_prs_population/", scenario_yr, "/output/mar_pop_25mi.csv")) %>%
  dplyr::mutate(popsum = 0)

write.csv(mar, file.path(here("globalprep/mar_prs_population/", scenario_yr, "/output/mar_pop_25mi_gf.csv")), row.names = FALSE)

```

## Data Checks and/or Meta-Analysis

### Create `UN_population.csv` for Data-checking

```{r create un_population.csv, eval=FALSE}

# Filter for countries (added for 2025 analysis)
pop_clean <- pop %>%
  filter(Type == "Country/Area")

# Select population estimates for January
pop_gather <- dplyr::select(pop_clean, c(matches("Region"), "Year", matches("Total.*January"))) %>%
  setNames(c("country", "year", "population")) %>%
  dplyr::mutate(population = gsub(" ", "", population)) %>%
  dplyr::mutate(population = as.numeric(as.character(population)) * 1000)

# Ignore Jersey and Guernsey (Channel Islands) for now
pop_gather_rename <- pop_gather %>%
  dplyr::mutate(country = ifelse(str_detect(country,"C\\Ste d'Ivoire"), "Cote d'Ivoire", country)) %>%
  dplyr::mutate(country = ifelse(str_detect(country,"R\\Sunion"), "Reunion", country)) %>%
  dplyr::mutate(country = ifelse(str_detect(country,"Cura\\Sao"), "Curacao", country)) %>%
  dplyr::mutate(country = ifelse(country=="China, Taiwan Province of China", "Taiwan", country)) %>%
  dplyr::mutate(country = ifelse(country=="Dem. People's Republic of Korea", "North Korea", country))
  
# Organize the data into regions used in OHI, and save
pop_rgn <- name_2_rgn(df_in = pop_gather_rename, 
                      fld_name='country', 
                      flds_unique=c('year'))

# Sum population by region and year
pop_rgn <- pop_rgn %>%
  dplyr::group_by(rgn_id, year) %>%
  dplyr::summarize(population = sum(population)) %>%
  data.frame()

write.csv(pop_rgn, file.path("globalprep/mar_prs_population/", scenario_yr, "/output/UN_population.csv"), row.names = FALSE)

# OHI regions
ohi_rgns <- ohicore::rgn_master
```

### Compare `UN_population` with Calculated Count

```{r check full calculated population against latest UN population data, eval=FALSE}

compare_yr <- 2023 # comparison between raster and UN population data for a specific year (most recent common year)

# Sum counts regionally for the scenario year
pop_rast <- terra::rast(file.path(dir_M, sprintf("git-annex/globalprep/mar_prs_population/%s/int/human_count_%s_mol.tif", scenario_yr, compare_yr)))
names(pop_rast) = c(paste0("pop_", compare_yr))

zones <- terra::rast(file.path(dir_M, "git-annex/globalprep/spatial/v2024/eez_25mi_inland_wgs1984.tif")) # pulling from 2024 server data
names(zones) = c("zone")
zones <- terra::as.factor(zones)

pop_counts <- terra::zonal(pop_rast, zones, fun = "sum", na.rm = TRUE)

pop_UN <- pop_rgn %>%
  dplyr::filter(year == compare_yr) %>%
  dplyr::select(rgn_id, pop_UN = population)

# Join filtered UN population and summed calculated counts
compare_pop <- data.frame(pop_counts) %>%
  dplyr::rename(sum = paste0("pop_", compare_yr)) %>%
  dplyr::select(rgn_id = zone, pop_rast = sum) %>%
  dplyr::mutate(rgn_id = as.numeric(rgn_id)) %>%
  dplyr::left_join(pop_UN, by="rgn_id") %>% 
  filter(!is.na(pop_UN))

options(scipen = 999)

compare_pop$diff <- compare_pop$pop_rast - compare_pop$pop_UN

# Check out the dataframe
View(compare_pop)

# Check plot - investigate outliers (post to issue first)
pop_compare_plot <- ggplot(compare_pop) +
  geom_point(aes(pop_rast, pop_UN, label = rgn_id), alpha = .6) +
  geom_abline(slope = 1, intercept = 0, color = 'red') +
  labs(x = sprintf('Calculated Population %s', compare_yr),
       y = sprintf('UN Population %s', compare_yr),
       title = 'Population Comparison')

# View
ggplotly(pop_compare_plot)

##Plot log values of same comparison
pop_compare_log <- ggplot(compare_pop) +
  geom_point(aes(log(pop_rast), log(pop_UN), label = rgn_id), alpha = .6) +
  geom_abline(slope = 1, intercept = 0, color = 'red') +
  labs(x = sprintf('Log of Calculated Population %s', compare_yr),
       y = sprintf('Log of UN Population %s', compare_yr),
       title = 'Population Comparison (log values)')

# View
ggplotly(pop_compare_log)

```

### Compare to Previous Year for Coastal Population

```{r comparison coastal, eval=FALSE}

# Assessment year to pull data from
prev_scen_yr <- paste0("v", as.numeric(substr(scenario_yr, 2, 5)) -1) 

# Year to compare
comp_yr <- 2020

old <- read_csv(here('globalprep/mar_prs_population/', prev_scen_yr, '/output/mar_pop_25mi.csv')) %>% 
  dplyr::select(rgn_id, year, popsum_old=popsum)

tmp <- coastal_pop_nona %>%
  dplyr::left_join(old, by=c("rgn_id", "year")) %>%
  # Look at 2020 data
  filter(year == comp_yr)

# Compare data for a specific year between assessments
years_compare_plot <- ggplot(tmp) +
  geom_point(aes(log(popsum), log(popsum_old), label = rgn_id), alpha = .6) +
  geom_abline(slope = 1, intercept = 0, color = 'red') +
  labs(x = sprintf('Log of Calculated Population %s', scenario_yr),
       y = sprintf('Log of Calculated Population %s', prev_scen_yr),
       title = paste0('Population comparison (25 mi) between assessments for year ', comp_yr))

# View
ggplotly(years_compare_plot)

```

### Compare to Previous Year for Population Density

```{r comparison density, eval=FALSE}

# Assessment year to pull data from
prev_scen_yr <- paste0("v", as.numeric(substr(scenario_yr, 2, 5)) -1) 

# Year to compare
comp_yr <- 2020

old <- read_csv(here('globalprep/mar_prs_population/', prev_scen_yr, '/output/prs_pop_density.csv')) %>% 
  dplyr::select(rgn_id, year, pressure_old=pressure_score)

tmp <- pressure_data %>%
  dplyr::left_join(old, by=c("rgn_id", "year")) %>%
  # Look at 2020 data
  filter(year == comp_yr)

# Compare data for a specific year between assessments
years_compare_plot_density <- ggplot(tmp) +
  geom_point(aes(log(pressure_score), log(pressure_old), label = rgn_id), alpha = .6) +
  geom_abline(slope = 1, intercept = 0, color = 'red') +
  labs(x = sprintf('Log of Calculated Population Density %s', scenario_yr),
       y = sprintf('Log of Calculated Population Density %s', prev_scen_yr),
       title = paste0('Population comparison (density) between assessments for year ', comp_yr))

# View
ggplotly(years_compare_plot_density)

```

# Appendix

## Reproject EEZ + 25-mile Inland Raster

Gridded world population counts from CIESE and CIAT are in WGS 1984 Projection, and the ocean_plus25mile_inland.tif is projected in the World Mollweide coordinate reference system. We will resample the ocean raster using nearest neighbor method to match the population count raster 0.008333333 x 0.008333333 degree resolution. 

Why reproject the ocean raster? It is a categorical raster defining zones, whereas the population raster is a continuous raster with highly variable cell values. As a result, changing the projection of the categorical raster would minimize errors propagating through calculations because it both 1. minimizes direct modifications to the data that we're interested in, and 2. is a categorical raster, indicating that the faster "nearest neighbor" method of resampling won't realistically change the underlying structure of the regions - only borders, which are typically less developed and less populated [And There Was Light: Trade and the Development of Border Regions](https://people.unil.ch/mariusbrulhart/files/2023/06/borderlights.pdf). 

Therefore, when choosing which raster to manipulate, we saw 1. that modifying the population raster resulted in some pretty large errors propagating, 2. manipulating the ocean raster is methodologically more sound, and 3. because we're not manipulating the population raster scale, it doesn't need to be population density, so we can use an unequal areas projection to run zonal statistics. 

```{r create eez + 25-mile inland raster, eval=FALSE}

# No longer using shape files, raster only
count_rast <- raster::raster(raw[1]) # load one raw raster to use as template for reprojection using raster package
count_rast <- terra::rast(count_rast) # Convert it to SpatRaster using terra::rast()

# Load EEZ + 25mile inland raster using raster package
eez_25mi_inland <- raster::raster(file.path(dir_M, "git-annex/globalprep/spatial/v2019/ocean_plus25mile_inland.tif"))
eez_25mi_inland <- terra::rast(eez_25mi_inland) # Convert it to SpatRaster using terra::rast()
(eez_25mi_inland) # Check out the object 

# View the datatype of the raster (should be categorical raster to conduct zonal statistics properly)
terra::datatype(eez_25mi_inland) # 'FLT4S'

# Converting it to 'int' before factorizing so that any numbers are rounded to whole 
ocean_rast_int <- terra::as.int(eez_25mi_inland) # Changing from float raster to integer raster 
plot(ocean_rast_int)  # Checking to make sure that it makes sense 

# Reproject integer raster to WGS 1984 (same as population count raster)
ocean_int_reproj <- terra::project(x = ocean_rast_int, count_rast, method="near")
plot(ocean_int_reproj) # Checking to make sure that it makes sense

# Factorizing the regions raster to ensure that it is categorical 
ocean_fact_reproj <- terra::as.factor(ocean_int_reproj) 
plot(ocean_fact_reproj) # Checking to make sure it makes sense 

# intersecting the two rasters to obtain TRUE where they overlap and FALSE where they do not 
pop_intersect <- terra::intersect(count_rast, ocean_fact_reproj)
plot(pop_intersect) # Making sure that I know where R thinks the two overlap. This makes sense!! 

# Checking names of the layers --> The layer name for ocean raster should be 'zone' for zonal statistics
names(ocean_fact_reproj) # [1] "ocean_plus25mile_inland"
names(ocean_fact_reproj) = c("zone") # [1] "zone"

# Another command can be used to set the layer name of ocean raster 
# terra::set.names(ocean_fact_reproj, "zone")

## Writing the ocean raster out to Mazu --> for future reference, you only need to load this raster and then rename the layer to 'zone'

# zonal_rast <- terra::writeRaster(ocean_fact_reproj, filename = file.path(dir_M, "git-annex/globalprep/spatial/v2024/eez_25mi_inland_wgs1984.tif"), overwrite = TRUE, datatype='INT4U')
```
