---
title: "OHI `r format(Sys.Date(), '%Y')`: Food Provision/Fisheries, Identify FAO and OHI regions for RAM stocks"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

# Summary

Generates a dataset describing the OHI and FAO major fishing regions that correspond to each stock's range.  This combines last year's dataset of stock-spatial information from RAM v4.61 database and Christopher M. Free's spatial work at Rutgers as well as newly added stocks from RAM v4.65 with manually assigned ohi and fao region ids. Dataset only contains stocks with B/Bmsy data.

### Updates from previous assessment

- The RAM data changed for v2025 so we re-ran this

## Setup

```{r}
## Libraries
library(tidyverse)
library(sf)
library(here) 

source(here("workflow", "R", "common.R"))

# ---- set year and file path info ----
current_year <- 2025 # Update this in the future!!
version_year <- paste0("v",current_year)
data_dir_version_year <- paste0("d", current_year)
```

## Identify Newly Added Stocks

Compare stocks found in previous version of RAM data to new data to see what additional stocks have been added.

Old Stocks: Used `timeseries` table, because in RAM v3.80, this table only contained single assessment for each unique stock. Now we use `timeseries_values_view, since v4.44 (assessment year 2019). 

```{r}
## old stocks
load(here::here(dir_M, "git-annex", "globalprep","_raw_data","RAM","d2024","RAMLDB v4.65","R Data","DBdata[asmt][v4.65].RData"))

ram_bmsy_old <- timeseries_values_views %>%
dplyr::select(stockid, stocklong, year, TBdivTBmsy, SSBdivSSBmsy, TBdivTBmgt, SSBdivSSBmgt) %>%
  mutate(ram_bmsy = 
           ifelse(!is.na(TBdivTBmsy), TBdivTBmsy, SSBdivSSBmsy)) %>%
  mutate(ram_bmsy =
           ifelse(is.na(TBdivTBmsy) & is.na(SSBdivSSBmsy), TBdivTBmgt, ram_bmsy)) %>%
  mutate(ram_bmsy = 
           ifelse(is.na(TBdivTBmsy) & is.na(SSBdivSSBmsy) & is.na(TBdivTBmgt), SSBdivSSBmgt, ram_bmsy)) %>%
  dplyr::filter(year > 1979) %>%
  filter(!is.na(ram_bmsy)) %>%
  dplyr::select(stockid, stocklong, year, ram_bmsy)

## check number of unique entries
old_stockid <- unique(ram_bmsy_old$stockid) 
old_stockid
length(unique(ram_bmsy_old$stockid)) # v2025: 474 unique entries

```

New Stocks: Used `timeseries_values_views` table, because in RAM v4.44, this is the table that contains the most recent assessment for each unique stock. The `timeseries` table has all assessments conducted so there are multiple `assessid` per unique `stockid`.

```{r}
## new stocks
load(here::here(dir_M, "git-annex", "globalprep","_raw_data","RAM","d2025","RAMLDB v4.66","R Data","DBdata[asmt][v4.66].RData"))

ram_bmsy <- timeseries_values_views %>%
dplyr::select(stockid, stocklong, year, TBdivTBmsy, SSBdivSSBmsy, TBdivTBmgt, SSBdivSSBmgt) %>%
  mutate(ram_bmsy = 
           ifelse(!is.na(TBdivTBmsy), TBdivTBmsy, SSBdivSSBmsy)) %>%
  mutate(ram_bmsy =
           ifelse(is.na(TBdivTBmsy) & is.na(SSBdivSSBmsy), TBdivTBmgt, ram_bmsy)) %>%
  mutate(ram_bmsy = 
           ifelse(is.na(TBdivTBmsy) & is.na(SSBdivSSBmsy) & is.na(TBdivTBmgt), SSBdivSSBmgt, ram_bmsy)) %>% 
  filter(!is.na(ram_bmsy)) %>% 
  dplyr::filter(year > 1979) %>%
  dplyr::select(stockid, stocklong, year, ram_bmsy)

## check number of unique entries
new_stockid <- unique(ram_bmsy$stockid)
new_stockid
length(unique(ram_bmsy$stockid)) # v2025: 476 unique entries

```


### Investigate Differences


```{r}
## find newly added stocks, when using timeseries_values_views
new_stocks <- setdiff(new_stockid, old_stockid) # v2024: there are 4 new stocks
new_stocks
unused_stocks <- setdiff(old_stockid, new_stockid) # these are no longer included in the RAM data, will be deleted
unused_stocks

# Fix the stock name error in the old stock
# ram_bmsy_old <- ram_bmsy_old %>% 
#   mutate(stockid = str_replace_all(stockid, "TARAKNZ", "TARAKENZ"))

```

There were 4 stocks included in the new data but not the old data.
These include: 

[1] "PRGURNGUR3" "PRGURNGUR7"
[3] "TARAKENZ"   "TARAKSA"  


There were 2 stocks present in the old data but not the new.
These include:

[1] "COD2J3KL" "TARAKNZ" 

v2025:
- "TARAKNZ" is sub-antarctic and "TARAKENZ" is eastern new zealand.
- "COD2J3KL" is not anywhere in the new data
 

Subset for just the additional stocks in RAM v4.65 data

```{r}
ram_bmsy_new <- ram_bmsy %>% 
  filter(stockid %in% new_stocks) 

# Find just the unique new stocks
id_new <- ram_bmsy_new %>% 
  select(stockid, stocklong) %>% 
  distinct()

# view(id_new)
```


## Add Stock Info to Old RAM_fao_ohi_rgns

Each stock needs the corresponding ohi and fao regions associated with its range. Make sure there are no duplicates here

```{r}
## Grab last year's fao-ohi-assessid data table
## (we need to get the stock id from the old RAM data)
RAM_fao_ohi_rgns <- read.csv(here("globalprep","fis", "v2024","int","RAM_fao_ohi_rgns_final.csv"))


RAM_fao_ohi_rgns_fix <- RAM_fao_ohi_rgns %>%
  distinct() %>% ## fixes any duplicates. Now lets check for mispellings or other errors 
  group_by(rgn_id, fao_id, stockid) %>%
  dplyr::mutate(dupe = n()>1) %>% # find the rest of the duplicates 
  ungroup()
  
#filter for duplicates
dupes <- RAM_fao_ohi_rgns_fix %>%
  filter(dupe == "TRUE") # v2025: 0 good

dupes$stockid # no duplicates

## fix the dupes 
dupes_fix <- dupes %>%
  filter(!is.na(RAM_area_m2))

##filter out dupes from RAM_fao_ohi_rgns_fix and add fixed dupes back in
#RAM_fao_ohi_rgns_fix_final 

RAM_rgns_old <- RAM_fao_ohi_rgns_fix %>%
  filter(dupe == "FALSE") %>%
  rbind(dupes_fix) %>%
  select(-dupe) ## now we have fixed the duplicate problem (if there was one...). 
```


## Remove stocks that are no longer in the RAM data
(but were in the previous version)

```{r}
filter(RAM_rgns_old, stockid %in% unused_stocks) # quick check to see which stocks are not being used

# filter out any stocks that are unused
RAM_rgns_old_filt <- RAM_rgns_old %>%
  filter(!(stockid %in% unused_stocks)) 
```

## Combine New and Old Stock Data

Essentially adds new stocks to the bottom of the old stocks data table.  The regions of the new stocks are blank at this point. 
```{r}
## Make sure these new species aren't in the old data
setdiff(id_new$stockid, RAM_rgns_old_filt$stockid)
# 2025 
# Pretty sure these first three are just all new zealand??
# "PRGURNGUR3" Pacific red gurnard New Zealand Area GUR 3
# "PRGURNGUR7" Pacific red gurnard New Zealand Area GUR 7
# "TARAKENZ" Tarakihi Eastern New Zealand

RAM_rgns_new <- RAM_rgns_old_filt %>% 
  full_join(id_new, by = c("stockid","stocklong")) ## added the 4 new stocks, still don't have rgn_ids etc
nrow(RAM_rgns_new)
```

**2025 stocks**

[PRGURNGUR3](https://fs.fish.govt.nz/Page.aspx?pk=8&stock=GUR3)
[PRGURNGUR7](https://fs.fish.govt.nz/Page.aspx?pk=8&stock=GUR7)
[TARAKENZ (NZ)](https://fs.fish.govt.nz/Page.aspx?pk=8&stock=TAR1)
[TARAKENZ (E AUS)](https://www.fishsource.org/stock_page/1911)
[TARAKSA](https://www.fishsource.org/stock_page/1912)
[gurnard nei](https://www.fishsource.org/stock_page/2662)

## Fill Region IDs for New Stocks

Here we need to figure out what ohi region id and FAO region id each of the new stocks fall in (there are 4 new stocks for v2024). 

1. Filter for all of the new stocks (all those with NA rgn and fao ids)
2. Join with the "stock" and "area" dataset, which contain more information on the distribution of each stock
3. For those stocks that are not labelled as multinational, we will assume that they only fall in the FAO and ISO3Cs that are attached to those stocks. Then we will join with our OHI regions dataset by ISO3C to get the OHI region ids. 
4. For those that are labelled as "multinational" we will have to figure the regions out through google and other means (see below and the code for a complete description)


Investigate what FAO/OHI region each stock belongs to: 
* Primarily used www.fishsource.org to find stock distribution information (just search the stocklong name and usually something will come up). Can just put these regions/countries down if the fish shows up. 
* Another good source to use is https://ices-library.figshare.com/ which will often have the specific name of the fish and then list the regions (FAO/other areas) in the title.
* If the fish or a close enough match is on fishsource take the fishing countries listed at the top for the OHI regions. Then, use the FAO areas or other regional division (and find the FAO equivalent) to use for the associated FAO codes. If not, use the FAO areas or other regional divisions for both OHI regions and FAO codes.
* For cross referencing with ICES sub-region codes, use these maps: https://www.researchgate.net/figure/Stock-units-defined-by-ICES-for-white-anglerfish-southern-stock-in-Divisions-VIIIc-and_fig1_31198841 and http://gis.ices.dk/sf/
* Referenced F_CODE/FAO_AREAS (filtered for MAJOR in F_LEVEL) in ohi_fao_rgns shapefile (fis/v2017/int) from mazu (can use other map visuals to identify stock extent). 
* FAO major fishing areas (fao_id): https://upload.wikimedia.org/wikipedia/commons/3/3a/FAO_Major_Fishing_Areas.svg

Some regions have more than 1 `fao_id` or `rgn_id` (e.g. Alaska skate Bering Sea and Aleutian Islands,	Kamchatka flounder Bering Sea and Aleutian Islands have `fao_id` 61 and 67).

```{r}
## v2025: All stocks are from new zealand, so our job is pretty simple 

## join with stock and area
# first we join our stocks that have NAs in FAO id and rgn id (there should be 4 in v2025) to the "stock" and "area" dataframes that were loaded from the RAM database

RAM_nas <- RAM_rgns_new %>%
  filter(is.na(fao_id)) %>% 
  left_join(as.data.frame(stock)) %>%
  left_join(area)

# Checking out RAM_nas, we can see that all 4 new stocks fall inside new zealand with the same primary FAO area of 81. New Zealand's OHI region ID is 162. 

## filter for not multinational - we will use what is notated
# Now we will filter for those stocks that have country not labeled as "multinational". We will assume that the FAO id and country names are the true ranges for these stocks.

region_data()

RAM_na_non_multi <- RAM_nas %>%
  dplyr::select(-rgn_id) %>%
  filter(country != "multinational") %>%
  mutate(fao_id = primary_FAOarea) %>%
  left_join(rgns_eez, by = c("ISO3_code" = "eez_iso3"), relationship =
  "many-to-many") %>%
  dplyr::select(rgn_id, fao_id, RAM_area_m2, stockid, stocklong, region) %>%
  mutate(fao_id = as.double(fao_id)) 

unique(RAM_na_non_multi$region) # v2025: 4 are from new zealand

## filter for multinational
RAM_na_multi <- RAM_nas %>%
  filter(country == "multinational")
 
unique(RAM_na_multi$region) # v2025: none are multinational

# Because we have no multinational stocks, everything already has an FAO and OHI ID. We'll use RAM_rgns_new later on to complete the cleaning


```

For reference, here is what v2024 did: 

```{r v2024-regions}
# v2024: we have two larger areas that require manual data entry
# [1] "South America"          
# [2] "Mediterranean-Black Sea"

#### We know from the RAM excel sheet "Stock Changes History" that these stocks used to be named something else. So we will look at last years "RAM_fao_ohi_rgns" to figure out the ohi region ids for these stocks. Not relevant for v2024.

####  now lets google each stocklong individually to figure out what OHI regions they are in

# ---- South America ----
# [1] "CHTRACCH"; Chilean jack mackerel Chilean EEZ and offshore
# https://www.fishsource.org/stock_page/756 
# OHI regions: Chile (224), Peru (138), Ecuador (137)
# fao_id: 87
# See: `https://upload.wikimedia.org/wikipedia/commons/3/3a/FAO_Major_Fishing_Areas.svg` 

# ram_fao_areaname_multinational_south_america <- RAM_na_multi %>%
#   dplyr::filter(region %in% c("South America")) %>%
#   distinct(country, region, areaname, stocklong, stockid, RAM_area_m2, primary_country) %>% # these are all just general rules, taken from an emLAB project: https://github.com/emlab-ucsb/sustain-sci-subsidy-reform-paper/blob/master/data/lookup-tables/assign_fao_areas_to_ram_v4.491.Rmd
#   mutate(areaname_fao_area = case_when(
#     areaname == "Chilean EEZ and offshore" ~ "87",
#     TRUE ~ "87" # Otherwise we assume it's 87, because we are making a new column named areaname_fao_area
#   )) %>% 
#   mutate(ohi_rgn = case_when(
#     stockid == "CHTRACCH" ~ "224;138;137",
#     TRUE ~ "NA"
#   )) %>%
#   separate_rows(c(ohi_rgn, areaname_fao_area), sep = ";", convert = TRUE) %>%
#   dplyr::select("rgn_id" = "ohi_rgn", "fao_id" = "areaname_fao_area", RAM_area_m2, stockid, stocklong)

# ----- Mediterranean-Black Sea -----
# [2] "RMULLMEDGSA7"; Red mullet Gulf of Lions
# https://firms.fao.org/firms/resource/13618/en
# OHI regions: France (179), Spain (182)
# fao_id: 37 
# See: `https://upload.wikimedia.org/wikipedia/commons/3/3a/FAO_Major_Fishing_Areas.svg` 

# [3] "SARDMEDGSA17-18";  Sardine Adriatic Sea (GSA 17,18)
# stocklong: Sardine Adriatic Sea (GSA 17,18)
# https://www.fishsource.org/stock_page/779
# OHI regions: Albania (82), Montenegro (186),  Croatia (187), Italy (184), Slovenia (188)
# fao_id: 37 
# See: `https://upload.wikimedia.org/wikipedia/commons/3/3a/FAO_Major_Fishing_Areas.svg` 

# [4] "SPRBLKGSA29"; Sprat Black Sea
# stocklong: Sprat Black Sea
# https://www.fishsource.org/stock_page/2323 
# OHI regions: Turkey (76), Bulgaria (71), Romania (72)
# fao_id: 37
# See: `https://upload.wikimedia.org/wikipedia/commons/3/3a/FAO_Major_Fishing_Areas.svg` 

# ram_fao_areaname_multinational_mbs <- RAM_na_multi %>%
#   dplyr::filter(region %in% c("Mediterranean-Black Sea")) %>%
#   distinct(country, region, areaname, stocklong, stockid, RAM_area_m2, primary_country) %>% # these are all just general rules, taken from an emLAB project: https://github.com/emlab-ucsb/sustain-sci-subsidy-reform-paper/blob/master/data/lookup-tables/assign_fao_areas_to_ram_v4.491.Rmd
#   mutate(areaname_fao_area = case_when(
#     areaname == "Gulf of Lions" ~ "37",
#     areaname == "Adriatic Sea (GSA 17,18)" ~ "37",
#     areaname == "Black Sea" ~ "37",
#     TRUE ~ "37" # Otherwise we assume it's 37, because we are making a new column named areaname_fao_area
#   )) %>% 
#   mutate(ohi_rgn = case_when(
#     stockid == "RMULLMEDGSA7" ~ "179;182",
#     stockid == "SARDMEDGSA17-18" ~ "82;186;187;184;188",
#     stockid == "SPRBLKGSA29" ~ "76;71;72",
#     TRUE ~ "NA"
#   )) %>%
#   separate_rows(c(ohi_rgn, areaname_fao_area), sep = ";", convert = TRUE) %>%
#   dplyr::select("rgn_id" = "ohi_rgn", "fao_id" = "areaname_fao_area", RAM_area_m2, stockid, stocklong)
# 
# # ====== Join them all together =======
# 
# RAM_na_all <- rbind(RAM_na_non_multi, ram_fao_areaname_multinational_mbs, ram_fao_areaname_multinational_south_america) ## now all of the NA ones have rgn_id and fao_id! 
# 
# summary(RAM_na_all)

# v2024: 
 #     rgn_id          fao_id     
 # Min.   : 71.0   Min.   :37.00  
 # 1st Qu.: 82.0   1st Qu.:37.00  
 # Median :179.0   Median :37.00  
 # Mean   :146.6   Mean   :48.54  
 # 3rd Qu.:186.0   3rd Qu.:37.00  
 # Max.   :224.0   Max.   :87.00  
 #                                
 #  RAM_area_m2 
 # Min.   : NA  
 # 1st Qu.: NA  
 # Median : NA  
 # Mean   :NaN  
 # 3rd Qu.: NA  
 # Max.   : NA  
 # NA's   :13   
 #   stockid         
 # Length:13         
 # Class :character  
 # Mode  :character  
                
 #  stocklong        
 # Length:13         
 # Class :character  
 # Mode  :character 
```

## Tidy Final Datatable

### Fix Incorrect FAO-OHI Matches

Each row must have both an FAO id and a rgn id. Fix duplicate/incorrect fao-ohi matches by using ohi_fao_rgns spatial file

```{r}
## Spatial file with fao and ohi regions, F_CODE is the FAO id
fao_ohi <- st_read(dsn = here::here(dir_M, "git-annex","globalprep","fis","v2017","int"),
                   layer = "ohi_fao_rgns") # Geodetic CRS:  WGS 84
st_geometry(fao_ohi) <- NULL # removes geometry

fao_ohi_id <- fao_ohi %>%
  select(rgn_id, fao_id = F_CODE) %>% 
  arrange(rgn_id) %>% 
  mutate(fao_id = as.numeric(as.character(fao_id))) %>% 
  mutate(rgn_id = as.numeric(as.character(rgn_id))) %>% 
  distinct() 

## Filter for correct fao-ohi pairs in the RAM regions table 
RAM_rgns_new_final <- fao_ohi_id %>% 
  left_join(RAM_na_non_multi, by = c("rgn_id", "fao_id")) %>% 
  # 2024 version of this line
  # left_join(RAM_na_all, by = c("rgn_id", "fao_id")) %>% 
  filter(!is.na(stockid)) %>% # keep matches only
 arrange(stocklong, fao_id)

# Confirm everything has an ID
colSums(is.na(RAM_rgns_new_final))
```

Compare with unfiltered RAM regions table to check on fao-ohi pairs that were dropped

```{r}
## Number of unique stocks after joining
nrow(RAM_rgns_new_final) #v2025: 4

## Number of unique stocks before joining
# v2025 Use RAM_na_non_multi instead of RAM_na_all because we had no multinational stocks
RAM_temp <- RAM_na_non_multi %>% select(rgn_id, fao_id, stockid, stocklong) %>% distinct() %>% arrange(stocklong, fao_id)
nrow(RAM_temp) #v2025: 4

## add identifier for the two RAM table versions
RAM_temp$idtemp <- 1:nrow(RAM_temp)
RAM_rgns_new_final$idtemp2 <- 1:nrow(RAM_rgns_new_final)

## view the ohi-fao pairs that would be removed
RAM_temp %>% 
  full_join(RAM_rgns_new_final, by = c("rgn_id", "fao_id", "stockid", "stocklong")) %>% 
  filter(is.na(idtemp2)) ## v2025: we did not lose any!
```

Tidy up RAM data table.  
```{r}
# v2024: none hand-added
## directory for intermediate outputs
int_dir <- here::here("globalprep","fis", version_year, "int")

write_csv(RAM_rgns_new_final, here(int_dir, "RAM_new_stocks.csv"))

## check that there are still 4 unique stock ids
length(unique(RAM_rgns_new_final$stockid)) # v2025: 4, great!
```

### RETURN: Combine with full dataset

Combine newly added stocks with ohi and fao region information to the old stock data table. Make sure there are no NAs!

```{r, eval=FALSE}
RAM_final <- bind_rows(read_csv(here::here(int_dir,"RAM_new_stocks.csv"))) %>% 
  select(-idtemp2)

RAM_fao_ohi_rgn_final <- RAM_rgns_old %>% 
  full_join(RAM_final, by = c("rgn_id", "fao_id", "stockid", "stocklong")) %>%
  select(-RAM_area_m2.y, RAM_area_m2 = RAM_area_m2.x) 

# no NAs in IDs
colSums(is.na(RAM_fao_ohi_rgn_final))

RAM_fao_ohi_rgn_final_v2 <- RAM_rgns_old %>% 
  full_join(RAM_final, by = c("rgn_id", "fao_id", "stockid", "stocklong", "RAM_area_m2"))
 
# save the final output!
write_csv(RAM_fao_ohi_rgn_final, here(int_dir, "RAM_fao_ohi_rgns_final.csv"))
```

