---
title: "OHI `r format(Sys.Date(), '%Y')` - Mariculture Livelihoods Quantity Data Preparation"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

# MAR Livelihoods: Number of Mariculture Jobs per Country/Region per Year

### Mariculture Jobs Data

**OECD Sustainable Ocean Economy (downloaded August 11, 2025)** 

- Time period: 2009 - 2022
- Measure: Socio-economic context>Fisheries and aquaculture>People employed in aquaculture sector
- Sex: Total

[OECD Data Explorer](https://data-explorer.oecd.org/vis?df%5Bds%5D=DisseminateFinalDMZ&df%5Bid%5D=DSD_SOE%40DF_SOE&df%5Bag%5D=OECD.ENV.EPI&dq=.A....&pd=1995%2C2024&to%5BTIME_PERIOD%5D=false&vw=tb)

**FAO Fishery and Aquaculture Statistics Yearbook 2021 (downloaded August 11, 2025)**

- Table on page 170

[FAO Fisheries and Aquaculture Statistical Yearbook](https://openknowledge.fao.org/server/api/core/bitstreams/2be6c2fa-07b1-429d-91c5-80d3d1af46a6/content)

Table on page 28 [Scaling factors](https://static-content.springer.com/esm/art%3A10.1038%2Fnature11397/MediaObjects/41586_2012_BFnature11397_MOESM79_ESM.pdf)

- scaling factors were averaged for developed and developing countries

### Setup

```{r, setup}
# load packages ---- 
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  ohicore,
  here,
  janitor,
  terra,
  readxl,
  RColorBrewer,
  foreach,
  doParallel, # for using multiple cores
  priceR,
  pdftools,
  tidyverse, 
  httr,
  plotly,
  zoo # for gapfilling
)

# source common.R ----
source(here::here("workflow/R/common.R"))

# set year and file path info ----
current_year <- 2025 # Update this!!

version_year <- paste0("v",current_year) # Assessment year
data_dir_version_year <- paste0("d", current_year) # Data folder for assessment year
data_path <- here::here("globalprep", "liv", version_year) # Path to assessment folder (v202X) 

# Raw data directory (on Mazu) ---- 
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")

# world bank raw data directory
wb_dir <- here::here(raw_data_dir, "WorldBank", data_dir_version_year)

# OECD raw data directory
oecd_dir <- here(raw_data_dir, "OECD", data_dir_version_year)

# FAO Mariculture raw data directory 
fao_dir <- here::here(raw_data_dir, "FAO", data_dir_version_year)

# output data directory for intermediate data products
int_dir <- here::here(data_path, "int")

# output data directory for final data products
output_dir <- here::here(data_path, "output")
```

## OECD Data

```{r,  eval = FALSE}
#---------- fis OECD data---------------------------
oecd_mar_job_raw <- read_csv(here(oecd_dir, "OECD_1995-2022_mar_employment.csv"), show_col_types = FALSE)

# total number of people employed by aquaculture (1000s)

# Check for NA values
sum(is.na(oecd_mar_job_raw$OBS_VALUE))
# v2025: none!

# ============== mar OECD number of fish farmers data ============== 
oecd_mar_jobs <- oecd_mar_job_raw %>% 
  janitor::clean_names() %>% 
  dplyr::select(ref_area, reference_area, time_period, obs_value, unit_multiplier) %>% 
  mutate(year = as.numeric(time_period),
         value = obs_value*1000,
         unit = ("persons (1)"),
         sector = "mar",
         data_source = "oecd sustainable economies") %>% 
  dplyr::select(-c(unit_multiplier, obs_value, time_period))

unique(is.na(oecd_mar_jobs$value))
# FALSE -- no NAs
# v2025: NAs

# ============== Run ohicore name_2_rgn ==================================

# check names
#china, taiwan province of or china (provice of china) should pick it up as taiwan
# "Chinese Taipei", "China (People’s Republic of)", "Korea"

oecd_mar_jobs_fix <- oecd_mar_jobs %>%
  mutate(reference_area = case_when(
    reference_area=="Chinese Taipei" ~ "Taiwan",
    reference_area=="China (People’s Republic of)" ~ "China",
    reference_area=="Korea" ~ "South Korea",
    TRUE ~ reference_area # Everything else, leave it be
  ))

oecd_mar_jobs_rgn <- name_2_rgn(df_in = oecd_mar_jobs_fix, 
                       fld_name='reference_area',
                      flds_unique = c("year"))

oecd_duplicates <- oecd_mar_jobs_rgn[duplicated(oecd_mar_jobs_rgn[, c("reference_area", "year")]),]
# none

length(unique(oecd_mar_jobs_rgn$reference_area)) 
# v2025: 47
length(unique(oecd_mar_jobs_rgn$rgn_name)) 
# v2025: 47
```

Gapfill OECD Data

```{r,  eval = FALSE}
region_data()

# Join with OHI regions
jobs_region_join <- left_join(rgns_eez, oecd_mar_jobs_rgn, by = c("rgn_id", "rgn_name")) %>%
  dplyr::select(-c(territory, admin_rgn_id, admin_country_name, 
            Notes, eez_iso3))

data_year <- max(jobs_region_join$year, na.rm = TRUE)

# ---- Within country gapfilling ----
mar_jobs_gf <- jobs_region_join %>% 
   group_by(rgn_id, rgn_name) %>% 
  complete(year = 2008:data_year) %>% # ensures that there is a row for every year for each region, to ensure there are no gaps
  ungroup() %>%
  # First, create gapfilled boolean and method columns
  mutate(jb_gapfilled = ifelse(is.na(value), "1", NA_character_),
         jb_gf_method = ifelse(is.na(value), "1", NA_character_)) %>%
  # Regroup for gapfilling
  group_by(rgn_id, rgn_name) %>% 
  arrange(rgn_id, year) %>%   
  # Linear interpolation
  mutate(value = na.approx(value, x = year, na.rm = FALSE)) %>% 
  mutate(jb_gapfilled = ifelse(jb_gapfilled == "1" & !is.na(value), 
                               "gapfilled", jb_gapfilled),
         jb_gf_method = ifelse(jb_gf_method == "1" & !is.na(value), 
                               "Interpolation", jb_gf_method)) %>% 
  # Upfilling and Downfiling
  tidyr::fill(value, .direction = "updown") %>% 
  mutate(jb_gapfilled = ifelse(jb_gapfilled == "1" & !is.na(value), 
                               "gapfilled", jb_gapfilled),
         jb_gf_method = ifelse(jb_gf_method == "1" & !is.na(value), 
                               "Up/Downfilling", jb_gf_method))

table(mar_jobs_gf$jb_gf_method)

# Regions with no job data after all gapfilling
mar_jobs_gf %>% 
  filter(is.na(value)) %>% 
  distinct(rgn_id, rgn_name) 

# Narrow wage columns before joining
jobs_final_oecd <- mar_jobs_gf %>% 
  drop_na(value) %>% 
  mutate(unit = "Persons (1)",
         sector = "mar",
         source = "OECD") %>% 
  dplyr::select(rgn_id, rgn_name, year, number_jobs = value, unit, sector, source)

colSums(is.na(jobs_final_oecd))
length(unique(jobs_final_oecd$rgn_id)) # 47 regions
```

## FAO Yearbook Data

```{r,  eval = FALSE}
#------fis FAO Yearbook Number of Fish Farmers-------
# turn the number of fish farmers table from the FAO yearbook into a data frame and clean.
# note: world total includes countries and regions not within the table.  Unsure of where that data comes from; look for methodology.

fao_pdf <- pdf_text(here(fao_dir, "fao_statistics_yearbook_07_03_24.pdf"))

cat(fao_pdf[[170]])
table <- strsplit(fao_pdf[[170]], "\n")[[1]]
# Remove header and footer (may need to adjust with new tables)
table_lines_keep <- table[11:(length(table) - 14)]
# Convert to df
mar_table <- table_lines_keep %>%
  str_trim() %>%
  # number of columns is n
  str_split_fixed("\\s{2,}", n = 11) %>%
  as.data.frame(stringsAsFactors = FALSE)

#specify columns 
columns <- c("country", "1995", "2000", "2005", "2010", "2015", "2017", "2018", "2019", "2020", "2021")

colnames(mar_table) <- columns

fao_mar_jobs <- mar_table %>%
  mutate(country = str_remove_all(country, "\\*")) %>%  # Remove asterisks
  mutate(across(2:11, ~ str_remove_all(as.character(.), "E"))) %>%  # Remove "E", coerce to character first
  pivot_longer(cols = 2:11, names_to = "year", values_to = "value") %>%
  mutate(value = str_squish(value)) %>% #remove extra spaces
  mutate(value = str_remove_all(value, " ")) %>% 
  mutate(value = if_else(value == "…" , NA, value)) %>% 
  mutate(value = as.numeric(value)) %>% 
  mutate(unit = "persons (1)",
         sector = "cf",
         data_source = "FAO number of mariculture jobs")

length(unique(fao_mar_jobs$country)) 
# v2025: 59

# --------- run names_2_rgn to see how many OHI regions we have from this data -----
fao_mar_jobs_rgn <- name_2_rgn(df_in = fao_mar_jobs, 
                             fld_name='country',
                             flds_unique = c("year"))

#let's see how many OHI regions we have:
length(unique(fao_mar_jobs_rgn$rgn_id)) 
# v2025: 50 (Uganda dropped for being landlocked)
```

Gapfill FAO data

```{r,  eval = FALSE}
# Join with OHI regions
jobs_region_join <- left_join(rgns_eez, fao_mar_jobs_rgn, by = c("rgn_id", "rgn_name")) %>%
  dplyr::select(-c(territory, admin_rgn_id, admin_country_name, 
            Notes, eez_iso3)) %>%
  mutate(year = as.numeric(year))

data_year <- max(jobs_region_join$year, na.rm = TRUE)

# ---- Within country gapfilling ----
mar_jobs_gf <- jobs_region_join %>% 
  group_by(rgn_id, rgn_name) %>% 
  complete(year = 2008:data_year) %>% # ensures that there is a row for every year for each region, to ensure there are no gaps
  ungroup() %>%
  # First, create gapfilled boolean and method columns
  mutate(jb_gapfilled = ifelse(is.na(value), "1", NA_character_),
         jb_gf_method = ifelse(is.na(value), "1", NA_character_)) %>%
  # Regroup for gapfilling
  group_by(rgn_id, rgn_name) %>% 
  arrange(rgn_id, year) %>%   
  # Linear interpolation
  mutate(value = na.approx(value, x = year, na.rm = FALSE)) %>% 
  mutate(jb_gapfilled = ifelse(jb_gapfilled == "1" & !is.na(value), 
                               "gapfilled", jb_gapfilled),
         jb_gf_method = ifelse(jb_gf_method == "1" & !is.na(value), 
                               "Interpolation", jb_gf_method)) %>% 
  # Upfilling and Downfiling
  tidyr::fill(value, .direction = "updown") %>% 
  mutate(jb_gapfilled = ifelse(jb_gapfilled == "1" & !is.na(value), 
                               "gapfilled", jb_gapfilled),
         jb_gf_method = ifelse(jb_gf_method == "1" & !is.na(value), 
                               "Up/Downfilling", jb_gf_method))

table(mar_jobs_gf$jb_gf_method)

# Regions with no job data after all gapfilling
mar_jobs_gf %>% 
  filter(is.na(value)) %>% 
  distinct(rgn_id, rgn_name) 

# Narrow wage columns before joining
jobs_final_fao <- mar_jobs_gf %>% 
  drop_na(value) %>% 
  mutate(unit = "Persons (1)",
         sector = "fis",
         source = "FAO Yrbk") %>% 
  dplyr::select(rgn_id, rgn_name, year, number_jobs = value, unit, sector, source)

colSums(is.na(jobs_final_fao))
length(unique(jobs_final_fao$rgn_id)) # 50 regions
```

Supplement the OECD data with FAO data

```{r,  eval = FALSE}
# Find which countries OECD is missing
fao_supplemental_countries <- setdiff(unique(jobs_final_fao$rgn_name), unique(jobs_final_oecd$rgn_name))

# Filter FAO to those countries
fao_supplemental_jobs <- jobs_final_fao %>%
  filter(rgn_name %in% fao_supplemental_countries)
```

Prep to use scaling factor:

```{r}
# Finally: Applying the multiplier to mariculture for developed versus developing countries ----
# Setting filepath to unsd data on developing versus developed countries 
fp_unsd <- file.path(dir_M,"git-annex/globalprep/_raw_data/UNSD/d2024/historical-classification-of-developed-and-developing-regions.xlsx")

# Reading in unsd data 
unsd_class <- read_excel(fp_unsd,col_names=TRUE, sheet = "Distinction as of May 2022") %>% 
  janitor::clean_names()

# Reading in OHI Regions data for gut checks/individual country checks 
ohi_regions <- readr::read_csv("https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/eez/spatial/regions_list.csv") %>% 
  janitor::clean_names() %>% 
  dplyr::select(-notes)

# Fixing a couple of the countries that are present under different synonyms in the unsd dataset OR Duplicated. Do this AFTER running name_2_rgn!!! 
unsd_class_fix <- unsd_class %>% 
  mutate(country_or_area = case_when(
  country_or_area == "Bonaire, Sint Eustatius and Saba" ~ "Sint Eustasius",
  country_or_area == "China, Hong Kong Special Administrative Region" ~ "China",
  country_or_area == "China, Macao Special Administrative Region" ~ "China",
  country_or_area == "Åland Islands" ~ "Finland",
  TRUE ~ country_or_area
  ))

# Using name to rgn function to assign rgn_id to country names 
class_temp <- ohicore::name_2_rgn(df_in = unsd_class_fix, 
                       fld_name='country_or_area')

# Identifying duplicates  
duplicates <- class_temp[duplicated(class_temp$rgn_id), ]

# Removing countries with duplicated rgn_id by using their iso2 code 
class_temp <- class_temp %>% 
  dplyr::filter(!iso_alpha2_code %in% c("HK", "MO", "FI", "MP", "VI", "MQ")) %>% 
  dplyr::select(rgn_id, rgn_name, developed_developing_regions)
```

Set up final df:

```{r}
# Bind dataframes
jobs_final <- jobs_final_oecd %>%
  rbind(fao_supplemental_jobs) %>%
  filter(year >= 2008 & year <= 2022)%>% 
  left_join(y = class_temp, by = c("rgn_id")) %>%
  dplyr::select(-c(rgn_name.y)) %>% 
  dplyr::rename(rgn_name = rgn_name.x) %>%
  # Multiply by scaling factor
  mutate(number_jobs = ifelse(developed_developing_regions == "Developing", number_jobs*1.973, number_jobs*2.7))

length(unique(jobs_final$rgn_name)) #54 countries!

write_csv(jobs_final, file.path(output_dir, "mar_quantity.csv"))
```

**NOTE:** It's important to note that these jobs are for all of *aquaculture* (not just mariculture)