---
title: "OHI `r format(Sys.Date(), '%Y')` - Livelihoods Data Preparation"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

# Livelihoods: Proportion of Tourism Jobs per Country/Region per Year

Livelihoods includes two equally important sub-components, the **number of jobs**, which is a proxy for livelihood quantity, and the **per capita average annual wages**, which is a proxy for job quality. 

### v2025 Updates

- New data for tourism jobs!! A data inquiry was filed with the WTTC and they very kindly agreed to provide the data necessary for tourism jobs as well as tourism revenue (to be used in economies, tourism). This data covers the time range 2008-2023. For future assessments, I would recommend filling out the data enquiry form again. The point of contact for 2025 was Chok Tsering: `Chok.Tsering@wttc.org`. 
- Increased total regions included from ~80 to ~145.

## Overview

* **Data:** Labor Force & Employment Data

  * Jobs data from World Trade and Tourism Council (WTTC) 
  
      - Data enquiry form  <https://wttc.org/research/data-enquiries>
      - Make sure it's clear that this is a request from an academic institution, as the WTTC seems to make allowances for providing free data to researchers. Also make sure you communicate that the raw data will not be publicly available. 

      
  * Quality/Wage data from ILOSTAT (downloaded August 7, 2025) -- for tour/cf wage data
  
      - <https://rshiny.ilo.org/dataexplorer46/?lang=en&id=EAR_4MTH_SEX_ECO_CUR_NB_A> 
      
  - navigate to ILOSTAT explorer, select:

  - “Sex”: select “none” (to un-select all options), then select “Total”

  - “Economic Activity”: select “none” and then select “ISIC-Rev.4: I. Accommodation and food service activities” AND "ISIC-Rev.4: I. Arts, entertainment and recreation"

  - “Currency”: keep all selected (we end up filtering this in R)

  - set Time Frame (sliding range bar) to 2008-present year

  - Then click “Export”, and for the file type option (appears as a drop-down menu), select “csv”.
      

## Methods

### Setup

```{r}
knitr::opts_chunk$set(eval = FALSE, warning = FALSE, message = FALSE)

# ---- Load packages 
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}

librarian::shelf(
  ohicore,
  here,
  janitor,
  terra,
  countrycode,
  readxl,
  RColorBrewer,
  foreach,
  tidyverse, 
  httr,
  plotly,
  zoo # for gapfilling
)

# ---- Source common.R 
source(here("workflow", "R", "common.R"))

region_data() # for rgns_all and rgns_eez

# OHI regions data 
region_names <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/eez/spatial/regions_list.csv") 

# ---- set year and file path info
current_year <- 2025 # Update this in the future!!
previous_year <- 2024

data_year <- 2023

version_year <- paste0("v",current_year)
previous_vyear <- paste0("v", previous_year)

current_data_dir <- paste0("d", current_year)
previous_data_dir <- paste0("d", previous_year)

# ---- data directories
# Raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")

# Livelihoods and Economies directory
current_le_dir <- here("globalprep", "liv", version_year)
previous_le_dir <- here("globalprep", "liv", previous_vyear)

# Create country code --> iso3 function
country_regex_to_iso3c <- function(country_string) {
  country_string %>%
    countrycode::countrycode(origin = "country.name", destination = "iso3c", origin_regex = TRUE)
}
```


# Quantity (number of jobs)

(put this within previous chunk, read in data etc. for number of tourism jobs, join with labor_regions, make new proportion of tourism jobs column, group by year, do prelim plotting etc. and evaluate data quality etc.)

## WTTC Data

```{r}
# total number of people employed by the tourism sector, in thousands of persons
wttc_jobs_raw <- read_xlsx(here(raw_data_dir, "WTTC", current_data_dir,
                              "WTTC-Data-Request-OHI-13.08.2025.xlsx"),
                          skip = 2, # skip first 2 rows
                          sheet = 2, # sheet name has changed over time, check this
                          col_names = TRUE) # use first row as column names

# Pivot longer and rename
wttc_long <- wttc_jobs_raw %>% 
  pivot_longer(cols = -1, # everything except the first column
               values_to = "jobs",
               names_to = "year") %>% 
  rename(country = Economy) %>% 
  mutate(year = as.numeric(year))

# Fix name_2_rgn error with Netherland Antillies
neth_ant <- wttc_long %>%
    filter(country == 'Former Netherland Antillies') %>% 
    mutate(jobs = (jobs / 5),  
           'Curacao' = jobs,
           'Sint Maarten' = jobs,
           'Bonaire' = jobs,
           'Saint Eustatius' = jobs,
           'Saba' = jobs) %>%
    select(-c(jobs, country)) %>%
    pivot_longer(cols = c("Curacao", "Sint Maarten", "Bonaire",
                          "Saint Eustatius", "Saba"),
                 names_to = "country",
                 values_to = "jobs") %>%
    mutate(country = as.character(country),
           jobs = as.numeric(jobs)) %>% 
    select(c(country, jobs, year))
  
wttc_long <- wttc_long %>%
    filter(country != 'Former Netherland Antillies') %>%
    bind_rows(neth_ant)

# run ohicore name_2_rgn 
wttc_rgn <- name_2_rgn(df_in = wttc_long, 
                       fld_name = 'country',
                       flds_unique = c("year"))

# Check duplicates
tour_duplicates <- wttc_rgn[duplicated(wttc_rgn[, c("rgn_name", "year")]),]
unique(tour_duplicates$rgn_name)

# aggregate: sum by rgn_name etc.
wttc_jobs <- wttc_rgn %>%
  group_by(rgn_id, rgn_name, year) %>%
  dplyr::summarize(jobs = sum(jobs, na.rm = TRUE) )%>%
  ungroup() %>% 
  filter(year >= 2008) %>% 
  # turn 0s into NAs (appropriate flag for lack of data)
  mutate(jobs = case_when(jobs == 0 ~ NA,
                          .default = jobs),
         unit = "Persons (1)",
         source = "WTTC")

# Confirm duplicates have been removed
tour_duplicates <- wttc_jobs[duplicated(wttc_jobs[, c("rgn_name", "year")]),]
unique(tour_duplicates$rgn_name)

# Check years for possible gapfilling
wttc_jobs <- wttc_jobs %>% 
  group_by(rgn_id, rgn_name) %>% 
  complete(year = 2008:data_year) %>% 
  ungroup()

# This data is so clean that no gapfilling is necessary! The above step was just a check to ensure we didn't need to add any years
colSums(is.na(wttc_jobs))
summary(wttc_jobs)
length(unique(wttc_jobs$rgn_id)) # 147 regions!!! Much more than the ~80 we were getting between unwto, oecd, and eurostat

```

## Coastal Proportion

Now, lets bring in our tourism data and calculate the coastal proportion! The coastal proportion will then be multiplied by the arrivals value per region and year. Finally, the coastal area will be divided, giving us a density of tourism per region (individuals/km\^2).

```{r}
# -------  population data -----------
# Coastal population
coastal_pop_data <- read_csv(here("globalprep","mar_prs_population","v2025","output","mar_pop_25mi.csv")) ## read in coastal population data from other data layer

# UN total population
un_pop <- read_csv(here("globalprep", "mar_prs_population", version_year,
                        "output", "UN_population.csv"))

rgns_eez_small <- rgns_eez %>%
  select(rgn_id, rgn_name)

# Add region names
coast_pop_rgn <- left_join(coastal_pop_data, rgns_eez_small, 
                           by = join_by(rgn_id))
un_pop_rgn <- left_join(un_pop, rgns_eez_small, by = join_by(rgn_id))


# Calculate coastal proportion
coastal_prop_df <- left_join(un_pop_rgn, coast_pop_rgn, 
                             by = join_by(rgn_id, rgn_name, year)) %>% 
  # mutate(year = as.numeric(year)) %>% 
  filter(year %in% 2008:data_year) %>% 
  mutate(coastal_prop = (popsum / population),
         coastal_prop = ifelse(coastal_prop > 1, 1, coastal_prop))

colSums(is.na(coastal_prop_df))
length(unique(coastal_prop_df$rgn_id)) #183

write_csv(coastal_prop_df, here(current_le_dir, "int", "coastal_pop_proportion.csv"))

```

## Coastal Jobs 

Multiply the total number of tourism jobs by the coastal population proportion as a proxy for marine-related tourism jobs.

```{r}
colnames(wttc_jobs)
# Calculate livelihood then multiply by coastal population proportion
wttc_coastal_jobs <- wttc_jobs %>% 
  left_join(coastal_prop_df, by = c("rgn_id","rgn_name","year")) %>% 
  # Coastal livelihood
  mutate(jobs_coast = jobs * coastal_prop)

colSums(is.na(wttc_coastal_jobs))

# Regions without labor force data
wttc_coastal_jobs %>% 
  filter(is.na(coastal_prop)) %>% 
  distinct(rgn_name)

```


## Labor force join

```{r}
# Read in total labor force data. Prepped in `liv_labor_force_dataprep.Rmd`
labor_force <- read_csv(here(current_le_dir, "int", "labor_force.csv"))

# Select columns of interest
labor_force <- labor_force %>% 
  select(rgn_id, rgn_name, year, labor_force)


# join with labor force data 
tour_jobs_prop <- full_join(wttc_coastal_jobs, labor_force, 
                             by = c("rgn_id", "rgn_name", "year")) %>% 
  # calculate tourism jobs proportion of total labor force 
  mutate(jobs_coast_prop = jobs_coast / labor_force)

summary(tour_jobs_prop$jobs_coast_prop)

colSums(is.na(tour_jobs_prop)) 
 #         jobs      labor_force         tour_jobs_prop 
 #          432               64                    496  

nrow(tour_jobs_prop) # 2784
# Using v2024 method ~ 71% NA
# After first round of gapfilling, dropped to 56% 
# After adding new data sources (OECD, Eurostat), down to 52%
# With WTTC, only 25% NAs!!
# After adding ILOSTAT data to total labor force, down to 17% NA!!

# There's still room to gapfill here potentially using WTTC as base data and UNWTO, OECD, and Eurostat to gapfill. Future assessments could look into this

length(unique(tour_jobs_prop$rgn_id)) # 174 regions

# Regions without jobs data
tour_jobs_prop %>% 
  filter(is.na(jobs)) %>% 
  distinct(rgn_name)

# Regions without labor force data
tour_jobs_prop %>% 
  filter(is.na(labor_force)) %>% 
  distinct(rgn_name)

# Number of countries without propotion data
tour_jobs_prop %>% 
  filter(is.na(jobs_coast_prop)) %>% 
  distinct(rgn_name) %>% 
  nrow()

# 34 countries with no data for jobs or labor force, we'll end up dropping these for now
tour_jobs_final <- tour_jobs_prop %>% 
  #drop_na(jobs_coast_prop) %>% 
  # Save number of jobs (instead of prop, calculate prop in ohi-global)
  select(rgn_id, year, quantity = jobs_coast)

length(unique(tour_jobs_final$rgn_id)) # 140 final regions
colSums(is.na(tour_jobs_final))

# Export quantity data
write_csv(tour_jobs_final, here(current_le_dir, "output", "tour_quantity.csv"))

```


```{r, eval=FALSE}
# plot it, interactively!
line_plot <- plotly::plot_ly(tour_jobs_prop, x = ~year, y = ~jobs_coast_prop, color = ~rgn_name, type = "scatter", mode = "lines") %>%
  layout(title = "All Regions: Coastal Proportional Tourism Employment Within Total Labor Force",
         xaxis = list(title = "Year"),
         yaxis = list(title = "Percent of people by region employed by the tourism sector"))

 # line_plot

# htmlwidgets::saveWidget(line_plot, file = "prop_tourism_laborforce.html")
```



# Quality (wages)

  - note: this data came from many sources and was compiled by ILO

```{r}
# ILO wage data
ilo_wage_data <- read_csv(here(raw_data_dir, "ILOSTAT", current_data_dir,
                               "EAR_4MTH_SEX_ECO_CUR_NB_A-filtered-2025-08-14.csv"))

# preliminary cleanings
wage_data_clean <- ilo_wage_data %>% 
  janitor::clean_names() %>% 
  rename(monthly_wage = obs_value) %>% 
  # filter to PPP adjusted data (this accounts for inflation, diff in currencies globally, cost of living, etc.) 
  # NEED TO CHECK WHAT THE YEAR IS ON YOUR DATASET
  filter(classif2_label == "Currency: 2021 PPP $") %>%
  # group by country/region
  group_by(ref_area_label) %>%
  # set column name to year
  rename(year = time) %>%
  # filter to the cap date range from the jobs data
  filter(year %in% c(2008:data_year))  %>%
  # Kosovo is not included in OHI regions, so we'll drop it here
  filter(ref_area_label != "Kosovo") %>% 
  # add iso3 column
  mutate(iso3 = country_regex_to_iso3c(ref_area_label)) %>%
  # Change name to country for ease
  select(country = ref_area_label, iso3, year, monthly_wage, 
         classif2_label, classif1_label) 


# v2025: we updated the categories to include arts, entertainment, and recreation
wage_pivot <- wage_data_clean %>% 
  pivot_wider(names_from = classif1_label,
              values_from = monthly_wage) %>%
  rename(accomodation_food = `Economic activity (ISIC-Rev.4): I. Accommodation and food service activities`,
         arts_ent_rec = `Economic activity (ISIC-Rev.4): R. Arts, entertainment and recreation`) %>% 
          # need to change NAs to 0 for adding
  mutate(arts_ent_rec = ifelse(is.na(arts_ent_rec), 0, arts_ent_rec),
         accomodation_food = ifelse(is.na(accomodation_food), 0, accomodation_food),
         # Find average wage of both sectors
         total_monthly_wage = (accomodation_food + arts_ent_rec) / 2,
         # turn back to NAs for ease of gapfilling
         total_monthly_wage = ifelse(total_monthly_wage == 0, NA_real_,
                                     total_monthly_wage))

# join with OHI regions
wage_region_join <- left_join(rgns_eez, wage_pivot, by = c("eez_iso3" = "iso3")) %>%
  select(-c(territory, admin_rgn_id, admin_country_name, 
            Notes, country, eez_iso3, classif2_label ))

```

## Wages Gapfill

Start by gapfilling with existing data, then we'll gapfill using the r1 and r2 georegion averages

```{r}
# ---- Within country gapfilling ----
wage_gf <- wage_region_join %>% 
   group_by(rgn_id, rgn_name) %>% 
  complete(year = 2008:data_year) %>% # ensures that there is a row for every year for each region, to ensure there are no gaps
  ungroup() %>%
  # First, create gapfilled boolean and method columns
  mutate(wg_gapfilled = ifelse(is.na(total_monthly_wage), "1", NA_character_),
         wg_gf_method = ifelse(is.na(total_monthly_wage), "1", NA_character_)) %>%
  # Regroup for gapfilling
  group_by(rgn_id, rgn_name) %>% 
  arrange(rgn_id, year) %>%   
  # Linear interpolation
  mutate(total_monthly_wage = na.approx(total_monthly_wage, 
                                        x = year, na.rm = FALSE)) %>% 
  mutate(wg_gapfilled = ifelse(wg_gapfilled == "1" & !is.na(total_monthly_wage), 
                               "gapfilled", wg_gapfilled),
         wg_gf_method = ifelse(wg_gf_method == "1" & !is.na(total_monthly_wage), 
                               "Interpolation", wg_gf_method)) %>% 
  # Upfilling and Downfiling
  tidyr::fill(total_monthly_wage, .direction = "updown") %>% 
  mutate(wg_gapfilled = ifelse(wg_gapfilled == "1" & !is.na(total_monthly_wage), 
                               "gapfilled", wg_gapfilled),
         wg_gf_method = ifelse(wg_gf_method == "1" & !is.na(total_monthly_wage), 
                               "Up/Downfilling", wg_gf_method))

table(wage_gf$wg_gf_method)
# Interpolation Up/Downfilling 
#           159            730 

length(unique(wage_gf$rgn_id)) # 221
colSums(is.na(wage_gf)) # 1785
nrow(wage_gf) # 3641
# ~ 49% NA currently. Will gapfill with georegion next
# Keep in mind this is including ALL OHI regions, that's why NA count is so high

```


```{r}
# ---- Georegion gapfilling ----

# Gapfill with regional means for wage data

georegions <- ohicore::georegions %>% 
  left_join(rgns_eez_small, by = "rgn_id")

# Find avg wage by georegion
wage_georgn <- wage_gf %>% 
  left_join(georegions, by = c("rgn_id", "rgn_name")) %>%
  # Get average for georgn and year
  group_by(r2, year) %>%
  mutate(mean_wage_r2 = mean(total_monthly_wage, na.rm = TRUE),
         # to make NaN = NAs
         mean_wage_r2 = ifelse(is.nan(mean_wage_r2), NA, mean_wage_r2)) %>%
  ungroup() %>%
  group_by(r1, year) %>%
  mutate(mean_wage_r1 = mean(total_monthly_wage, na.rm = TRUE),
         mean_wage_r1 = ifelse(is.nan(mean_wage_r1), NA, mean_wage_r1)) %>%
  ungroup()

# ---- Georegion gapfilling ----
wage_georgn_gf <- wage_georgn %>% 
  mutate(
    # Gapfilling record
    georgn_gapfill = case_when(
    is.na(total_monthly_wage) & !is.na(mean_wage_r2) ~ "r2",
    is.na(total_monthly_wage) & is.na(mean_wage_r2) & !is.na(mean_wage_r1) ~ "r1",
    .default = NA
  ),
  # Actual gapfilling
    total_monthly_wage = case_when(
    is.na(total_monthly_wage) & !is.na(mean_wage_r2) ~ mean_wage_r2,
    is.na(total_monthly_wage) & is.na(mean_wage_r2) & !is.na(mean_wage_r1) ~ mean_wage_r1,
    TRUE ~ total_monthly_wage)
  ) %>% 
  drop_na(year)

table(wage_georgn_gf$georgn_gapfill)
#   r2 
# 1552 

# Regions with no wage data after all gapfilling
wage_georgn_gf %>% 
  filter(is.na(total_monthly_wage)) %>% 
  distinct(rgn_id, rgn_name) 

# Narrow wage columns before joining
wage_final <- wage_georgn_gf %>% 
  drop_na(total_monthly_wage) %>% 
  mutate(unit = "Currency: 2021 PPP $",
         sector = "tour",
         source = "ILOSTAT",
         annual_income = total_monthly_wage * 12) %>% 
  select(rgn_id, rgn_name, year, annual_income, unit, sector, source)

colSums(is.na(wage_final))
length(unique(wage_final$rgn_id)) # 213 final regions
# view(wage_final)

wage_export <- wage_final %>% 
  select(rgn_id, year, quality = annual_income)

write_csv(wage_export, here(current_le_dir, "output", "tour_quality.csv"))

```

# Join Quantity and Quality

```{r}
# Read in cleaned data
jobs <- read_csv(here(current_le_dir, "output", "tour_quantity.csv")) 
wages <- read_csv(here(current_le_dir, "output", "tour_quality.csv")) 

# Join quantity and quality data
# Jobs is the limiting factor here so we'll left join wages into jobs
tour_livelihood <- left_join(jobs, wages, by = join_by(rgn_id, year)) %>% 
  mutate(livelihood = jobs_coast_prop * annual_income)

tour_liv_rescaled <- tour_livelihood %>%
  # look at previous ten years
  filter(year >= (max(year) - 10)) %>% 
  group_by(rgn_id) %>% 
  mutate(ref_point = max(livelihood)) %>% 
  ungroup() %>% 
  mutate(liv_rescaled = livelihood / ref_point,
         liv_rescaled = ifelse(liv_rescaled > 1, 1, liv_rescaled))

view(tour_liv_rescaled)

# Select final livelihoods proportion data
liv_tour_final <- liv_rescaled %>% 
  select(rgn_id, rgn_name, year, livelihood = liv_rescaled)
```



```{r}
liv_norm_plot <- tour_liv_rescaled %>%
  plot_ly(x = ~year, y = ~livelihood, color = ~rgn_name,
          type = "scatter", mode = "lines") %>%
  layout(title = "Livelihood Coastal Proportion: Max Rescale",
         xaxis = list(title = "Year"),
         yaxis = list(title = "Livelihood (max rescale)"))
liv_norm_plot

# hist_max <- ggplot() +
#   geom_histogram(data = liv_rescaled_final, aes(x = liv_norm),
#                  color = "grey10", fill = "indianred")+
#   labs(title = "Distribution of rescaled livelihood: Maximum",
#        x = "Livelihood proportion: Max rescale",
#        y = "Count") +
#   theme_bw();ggplotly(hist_max)

# htmlwidgets::saveWidget(liv_final_plot, here::here(current_tr_dir, "figs", "ap_rescaled.html")) # save plot

```


# Data check

```{r}
# Get max data years for plotting
# current_data_year <- max(new_tr$year)
# current_data_year
# previous_data_year <- max(old_tr$year)
# previous_data_year

# Combine data frames
# comb_tr <- new_tr %>% 
#   rename(new_Ap = Ap) %>% 
#   left_join(old_tr, by = join_by(rgn_id, year))

# Plot
# comp_plot1 <- ggplot() + 
#   geom_jitter(data = filter(comb_tr, year == previous_data_year),
#               aes(x = Ap, y = new_Ap, 
#                   text = paste("Year:", year, "<br>Region:", rgn_name)),
#               alpha = 0.7) +
#   labs(x = paste("Previous Value", previous_data_year),
#        y = paste("Current Value", previous_data_year),
#        title = paste("Sustainable tourism and recreation", previous_data_year, "Data", previous_year, "Assessment vs.", previous_data_year, "Data", current_year, "assessment")) +
#   geom_abline(slope = 1, intercept = 0, col = "red") +
#   theme_bw();ggplotly(comp_plot1)
```


# OLD - legacy tourism data

* Jobs data from UN Tourism / UNWTO (downloaded August 7, 2025)
  
      - Key Tourism Statistics <https://www.unwto.org/tourism-statistics/key-tourism-statistics>
      - According to the website, the latest update of the dataset took place in 31 January 2024. 

  * Labor Force data from World Bank, last updated 7/1/25 (downloaded August 7, 2025).

      - <https://data.worldbank.org/indicator/SL.TLF.TOTL.IN>
      - Located in `globalprep/_raw_data/

  * Jobs data from [OECD](https://ec.europa.eu/eurostat/data/database)
  
      - Visit this site <https://ec.europa.eu/eurostat/data/database>
      - This is the file path on the OECD website: `Data navigation tree/Detailed datasets/Industry, trade, and services/Tourism (tour)/Tourism industries - annual data (tour_inda)/Annual and quarterly data on employment for selected tourism industries (source: Labor force survey)(tour_lfsa)/Employed persons by full-time/part-time and NACE Rev. 2 activity (tour_lfs1r2).csv`

### UNWTO Data

```{r, eval=FALSE}
# Jobs data
# total number of people employed by the tourism sector, in thousands
un_jobs_raw <- read_xlsx(here(raw_data_dir, "UNWTO", current_data_dir,
                              "UN_Tourism_8_9_2_employed_persons_04_2025",
                              "UN_Tourism_8_9_2_employed_persons_04_2025.xlsx"),
                          # skip = 2, # skip first 2 rows
                          sheet = 2, # sheet name has changed over time, check this
                          col_names = TRUE, # use first row as column names
                          na = c("","..")) # define NA markers

# Select columns of interest and add a unit column
un_jobs_data <- un_jobs_raw %>% 
  select(country = GeoAreaName, year = TimePeriod, jobs = Value) %>% 
  mutate(jobs = as.numeric(jobs)) 

# run ohicore name_2_rgn 
tour_jobs_rgn <- name_2_rgn(df_in = un_jobs_data, 
                       fld_name = 'country',
                       flds_unique = c("year"))

# Duplicates
# [1] "China, Hong Kong Special Administrative Region"
# [2] "China, Macao Special Administrative Region"
# [3] "China"

# aggregate: sum by rgn_name etc.
unwto_jobs <- tour_jobs_rgn %>%
  group_by(rgn_id, rgn_name, year) %>%
  dplyr::summarize(jobs = sum(jobs, na.rm = TRUE) )%>%
  ungroup() %>% 
  filter(year >= 2008) %>% 
  # turn 0s into NAs (appropriate flag for lack of data)
  mutate(jobs = case_when(jobs == 0 ~ NA,
                          .default = jobs),
         unit = "Persons (1)",
         source = "UNWTO")

# Confirm duplicates are gone
tour_duplicates <- unwto_jobs[duplicated(unwto_jobs[, c("rgn_name", "year")]),]
unique(tour_duplicates$rgn_name)

colnames(unwto_jobs)

unwto_jobs_gf <- unwto_jobs %>% 
  group_by(rgn_id, rgn_name) %>% 
  # Ensure we're hitting all years for joining later on
  complete(year = 2008:data_year) %>% 
  ungroup() %>%
  # First, create gapfilled boolean and method columns
  mutate(tr_gapfilled = ifelse(is.na(jobs), "1", NA_character_),
         tr_gf_method = ifelse(is.na(jobs), "1", NA_character_),
         unit = "Persons (1)") %>%
  # Regroup for gapfilling
  group_by(rgn_id, rgn_name) %>% 
  arrange(rgn_id, year) %>%   
  # Linear interpolation
  mutate(jobs = na.approx(jobs, x = year, na.rm = FALSE)) %>% 
  mutate(tr_gapfilled = ifelse(tr_gapfilled == "1" & !is.na(jobs), 
                               "gapfilled", tr_gapfilled),
         tr_gf_method = ifelse(tr_gf_method == "1" & !is.na(jobs), 
                               "Interpolation", tr_gf_method)) %>% 
  # Upfilling and Downfiling
  tidyr::fill(jobs, .direction = "updown") %>% 
  mutate(tr_gapfilled = ifelse(tr_gapfilled == "1" & !is.na(jobs), 
                               "gapfilled", tr_gapfilled),
         tr_gf_method = ifelse(tr_gf_method == "1" & !is.na(jobs), 
                               "Up/Downfilling", tr_gf_method))

table(unwto_jobs_gf$tr_gf_method)
# Interpolation Up/Downfilling 
#            58            298 

length(unique(unwto_jobs_gf$rgn_id)) # 73
```

### OECD Data

```{r, eval=FALSE}
# OECD data
oecd_jobs_data_raw <- read_csv(here(raw_data_dir, "OECD", 
                                   current_data_dir,
                                   "OECD_2009_2022_tourism_employment.csv"))

# Clean raw data
oecd_clean <- oecd_jobs_data_raw %>% 
  janitor::clean_names() %>% 
  # Filter to persons employed, not enterprises
  # AND total tourism activity, not sub categories
  filter(unit_measure == "PS" & activity == "_T") %>% 
  select(country_code = ref_area, country = reference_area, 
         unit = unit_of_measure, year = time_period, jobs = obs_value)

# run ohicore name_2_rgn
oecd_job_rgn <- name_2_rgn(df_in = oecd_clean, 
                       fld_name = 'country',
                       flds_unique = c("year"))

# Select columns of interest
oecd_jobs <- oecd_job_rgn %>% 
  select(rgn_id, rgn_name, year, jobs, unit) %>% 
  mutate(unit = "Persons (1)",
         source = "OECD")

oecd_jobs_gf <- oecd_jobs %>% 
  group_by(rgn_id, rgn_name) %>% 
  # Ensure we're hitting all years for joining later on
  complete(year = 2008:data_year) %>% 
  ungroup() %>%
  # First, create gapfilled boolean and method columns
  mutate(tr_gapfilled = ifelse(is.na(jobs), "1", NA_character_),
         tr_gf_method = ifelse(is.na(jobs), "1", NA_character_),
         unit = "Persons (1)") %>%
  # Regroup for gapfilling
  group_by(rgn_id, rgn_name) %>% 
  arrange(rgn_id, year) %>%   
  # Linear interpolation
  mutate(jobs = na.approx(jobs, x = year, na.rm = FALSE)) %>% 
  mutate(tr_gapfilled = ifelse(tr_gapfilled == "1" & !is.na(jobs), 
                               "gapfilled", tr_gapfilled),
         tr_gf_method = ifelse(tr_gf_method == "1" & !is.na(jobs), 
                               "Interpolation", tr_gf_method)) %>% 
  # Upfilling and Downfiling
  tidyr::fill(jobs, .direction = "updown") %>% 
  mutate(tr_gapfilled = ifelse(tr_gapfilled == "1" & !is.na(jobs), 
                               "gapfilled", tr_gapfilled),
         tr_gf_method = ifelse(tr_gf_method == "1" & !is.na(jobs), 
                               "Up/Downfilling", tr_gf_method))

# Countries added
setdiff(oecd_jobs$rgn_name, unwto_jobs$rgn_name)
# [1] "Netherlands"   "United States"
# [3] "Morocco" 

table(oecd_jobs_gf$tr_gf_method)
# Up/Downfilling 
#            39 
```

### Eurostat Data

```{r, eval=FALSE}
# Eurostat data uses a weird mix of tabs and commas as delimiters, 
# so we do a bit of pre-processing here
txt <- readLines(here(raw_data_dir, "Eurostat", 
                      current_data_dir, "estat_tour_lfs1r2.tsv"))

# Replace commas or tabs with a single tab
txt_clean <- gsub("[\t,]", "\t", txt)

# Read in with consistent delimiter
eurostat_raw <- read.delim(textConnection(txt_clean), sep = "\t", header = TRUE)

# Preliminary cleaning
eurostat_clean <- eurostat_raw %>% 
  janitor::clean_names() %>% 
  mutate(country = countrycode(geo_time_period, 
                               origin = "eurostat", "country.name")) %>% 
  # Filtering down to specific employment codes and times that consider total tourism employment. This metadata was not easy to find. 
  filter(worktime == "TOTAL", wstatus == "EMP", nace_r2 == "TOTAL") %>% 
  # Remove the aggregated EU countries
  filter(!(geo_time_period %in% c("EA20", "EU27_2020"))) %>% 
  relocate(country, .after = unit) %>% 
  select(-c(freq, geo_time_period)) %>% 
  # tidy up year column names
  rename_with(~ gsub("x", "", .)) %>% 
  pivot_longer(cols = -c(nace_r2, wstatus, worktime, unit, country),
              names_to = "year", values_to = "value") %>% 
  # Fix weirdness in values
  mutate(value = str_remove_all(value, "[bd]"),
         value = str_replace_all(value, ":", NA_character_),
         country = str_replace_all(country, "&", "and"),
         value = as.numeric(value),
         year = as.numeric(year))
# Don't worry about the warning in countrycode, we drop those aggregated regions with no matches

# run ohicore name_2_rgn 
eurostat_rgn <- name_2_rgn(df_in = eurostat_clean, 
                       fld_name = 'country',
                       flds_unique = c("year"))

# Fix units and add some metadata
eurostat_jobs <- eurostat_rgn %>%
  mutate(jobs = value * 1000,
         unit = "Persons (1)",
         source = "Eurostat") %>% 
  select(rgn_id, rgn_name, year, jobs, unit, source)

eurostat_jobs_gf <- eurostat_jobs %>% 
  group_by(rgn_id, rgn_name) %>% 
  # Ensure we're hitting all years for joining later on
  complete(year = 2008:data_year) %>% 
  ungroup() %>%
  # First, create gapfilled boolean and method columns
  mutate(tr_gapfilled = ifelse(is.na(jobs), "1", NA_character_),
         tr_gf_method = ifelse(is.na(jobs), "1", NA_character_),
         unit = "Persons (1)") %>%
  # Regroup for gapfilling
  group_by(rgn_id, rgn_name) %>% 
  arrange(rgn_id, year) %>%   
  # Linear interpolation
  mutate(jobs = na.approx(jobs, x = year, na.rm = FALSE)) %>% 
  mutate(tr_gapfilled = ifelse(tr_gapfilled == "1" & !is.na(jobs), 
                               "gapfilled", tr_gapfilled),
         tr_gf_method = ifelse(tr_gf_method == "1" & !is.na(jobs), 
                               "Interpolation", tr_gf_method)) %>% 
  # Upfilling and Downfiling
  tidyr::fill(jobs, .direction = "updown") %>% 
  mutate(tr_gapfilled = ifelse(tr_gapfilled == "1" & !is.na(jobs), 
                               "gapfilled", tr_gapfilled),
         tr_gf_method = ifelse(tr_gf_method == "1" & !is.na(jobs), 
                               "Up/Downfilling", tr_gf_method))

# check what countries we've added:
setdiff(eurostat_jobs$rgn_name, unwto_jobs$rgn_name)
# [1] "Germany"     "Estonia"     "Ireland"
# [4] "Iceland"     "Montenegro"  "Netherlands"
# [7] "Norway"

table(eurostat_jobs_gf$tr_gf_method)
# Up/Downfilling 
#             25 
```

## Join all tourism jobs together

```{r eval=FALSE}
# # Find all observations in oecd data NOT in unwto data
# new_oecd_jobs <- anti_join(oecd_jobs, unwto_jobs, by = join_by(rgn_id, rgn_name, year)) 
# 
# # Bind unwto and new oecd data
# unwto_oecd <- bind_rows(unwto_jobs, new_oecd_jobs) 
# 
# # Find all observations in eurostat data NOT in unwto+oeced data
# new_eurostat_jobs <- anti_join(eurostat_jobs, unwto_oecd, by = join_by(rgn_id, rgn_name, year))
# 
# # Bind eurostat with previous two
# tourism_jobs <- bind_rows(unwto_oecd, new_eurostat_jobs) 

# Find all observations in oecd data NOT in unwto data
new_oecd_jobs <- anti_join(oecd_jobs_gf, unwto_jobs_gf, by = join_by(rgn_id, rgn_name, year)) 

# Bind unwto and new oecd data
unwto_oecd <- bind_rows(unwto_jobs_gf, new_oecd_jobs) 

# Find all observations in eurostat data NOT in unwto+oeced data
new_eurostat_jobs <- anti_join(eurostat_jobs_gf, unwto_oecd, by = join_by(rgn_id, rgn_name, year))

# Bind eurostat with previous two
tourism_jobs <- bind_rows(unwto_oecd, new_eurostat_jobs) 

# Check NAs
colSums(is.na(tourism_jobs)) # 10
nrow(tourism_jobs) # 1007
# Only ~10% NA

table(tourism_jobs$tr_gf_method)
# Interpolation Up/Downfilling 
#            58            315 

length(unique(tourism_jobs$rgn_name)) # 82
# Increased regions from 73 to 82, not bad
```

### Gapfill tourism jobs

```{r}
# Gapfill total labor force values for missing years
# tourism_jobs_gf <- tourism_jobs %>% 
#   group_by(rgn_id, rgn_name) %>% 
#   # Ensure we're hitting all years for joining later on
#   complete(year = 2008:data_year) %>% 
#   ungroup() %>%
#   # First, create gapfilled boolean and method columns
#   mutate(tr_gapfilled = ifelse(is.na(jobs), "1", NA_character_),
#          tr_gf_method = ifelse(is.na(jobs), "1", NA_character_),
#          unit = "Persons (1)") %>%
#   # Regroup for gapfilling
#   group_by(rgn_id, rgn_name) %>% 
#   arrange(rgn_id, year) %>%   
#   # Linear interpolation
#   mutate(jobs = na.approx(jobs, x = year, na.rm = FALSE)) %>% 
#   mutate(tr_gapfilled = ifelse(tr_gapfilled == "1" & !is.na(jobs), 
#                                "gapfilled", tr_gapfilled),
#          tr_gf_method = ifelse(tr_gf_method == "1" & !is.na(jobs), 
#                                "Interpolation", tr_gf_method)) %>% 
#   # Upfilling
#   tidyr::fill(jobs, .direction = "updown") %>% 
#   mutate(tr_gapfilled = ifelse(tr_gapfilled == "1" & !is.na(jobs), 
#                                "gapfilled", tr_gapfilled),
#          tr_gf_method = ifelse(tr_gf_method == "1" & !is.na(jobs), 
#                                "Up/Downfilling", tr_gf_method))
# 
# 
# table(tourism_jobs$tr_gf_method)
# # Interpolation Up/Downfilling 
# #            58            285 
# 
# colSums(is.na(tourism_jobs_gf)) # 0 NAs
# Source is gonna have NAs due to the gapfilling

```
