---
title: "OHI `r format(Sys.Date(), '%Y')` - Aquarium Fishing Livelihoods & Economies Data Preparation"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

# Economies: Aquarium Fishing Revenue per Country per Year (2019-2021)

## Overview

* **Data:** Aquarium fishing -- export value (USD) data

-   The original data source for aquarium fishing revenue had been updated since this goal was originally calculated: [FAO global trade value data.](https://www.fao.org/fishery/en/collection/global_commodity_prod)
-   Revenue data was prepared as is described in the methods: export data from the FAO Global Commodities database for 'Ornamental fish' for all available years, ornamental freshwater fish were excluded. The global commodities database is a component of the Global Aquatic Trade Statistic Collection published by FAO.
-  **v2024:** downloaded using  on July 3rd, 2024 using the FAO status query interface/dashboard (data exploration & download portal, seems to be relatively new):
  - [Global aquatic trade - By partner country Value (2019 - 2021)](https://www.fao.org/fishery/statistics-query/en/trade_partners/trade_partners_value)
  -  under "Trade Flow" in the Dimensions filtering section, select "Exports" (alternatively, you could skip this and filter to "Export" in R)
  -  scroll to the bottom of the page, click the "download" button/icon, then select "csv", "Flag enabled" (we clean this later), then "Yes" for "Include null values"
  -  select "Preferences", then:
      - for "Show unavailable values" select "NA"
      - for "Thousands separator" select "No space"
    - (all years -- 2019, 2020, and 2021 are selected by default, no countries or commodities etc. are selected for any filtering)
    - © FAO 2024. Global Aquatic Trade Statistics. In: Fisheries and Aquaculture. Rome. [Cited Wednesday, July 3rd 2024]. 
https://www.fao.org/fishery/en/collection/global_commodity_prod
-   [Metadata](https://www.fao.org/fishery/en/collection/global_commodity_prod)

* **Data:** Labor Force & Employment Data

  * Labor Force data from World Bank (downloaded June 28. 2024)

      - https://data.worldbank.org/indicator/SL.TLF.TOTL.IN
      
  * ILOSTAT (downloaded July 2, 2024) -- for tour/cf wage data
  
      - https://rshiny.ilo.org/dataexplorer46/?lang=en&id=EAR_4MTH_SEX_ECO_CUR_NB_A 


## Setup

```{r}
# load packages
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  here, # reproducible file paths
  janitor, # data cleaning: lower_snake_case 
  terra,
  countrycode,
  readxl,
  paletteer, # custom color palettes for plotting
  RColorBrewer,
  foreach,
  doParallel, # for using multiple cores
  tidyverse, 
  httr,
  plotly, # interactive plots
  kableExtra, # for more aesthetically pleasing tables
  zoo # for gapfilling
  
)
#remotes::install_github("skgrange/threadr") # for na_extrapolate
library(threadr)


# source 
source(here("workflow", "R", "common.R"))
# used to clean FAO online portal data
# source(here("workflow", "R", "fao_online_portal_clean.R")) # old function
source(here("workflow", "R", "fao_query_data_tidy.R")) # new function

# set year and file path info
current_year <- 2024 # Update this!!

version_year <- paste0("v",current_year)
data_dir_version_year <- paste0("d", current_year)
data_path <- here("globalprep", "le", version_year)

# Raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")

# world bank raw data directory
wb_dir <- here(raw_data_dir, "WorldBank", data_dir_version_year)
fao_comm_dir <- here(raw_data_dir, "FAO_commodities", data_dir_version_year)


# output data dir for intermediate data products
int_dir <- here(data_path, "int")
```


Create country code --> iso3 function

```{r}
library(countrycode)

country_regex_to_iso3c <- function(country_string) {
  country_string %>%
    countrycode::countrycode(origin = "country.name", destination = "iso3c", origin_regex = TRUE)
}
```

### Read in Data

```{r}
# OHI regions data ----
# read in OHI regions for joining
region_names <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/eez/spatial/regions_list.csv") %>% 
  janitor::clean_names() %>% 
  select(-notes)

# labor force from World Bank (cleaned, joined with OHI regions)
labor_force <- read_csv(here(int_dir, "labor_force.csv"))

# Aquarium Fishing FAO data 
commodities_value_raw <- read_csv(here(fao_comm_dir, "query_portal", "trade_partners_value.csv"))

```
```{r}
# check and see if any values are considered negligible
commodities_value_raw %>% filter(`2021 Flag` == "N")
```


## Data cleaning & Wrangling

```{r}
# adapted from the v2023 livelihood_economies_dataprep.Rmd
fao_latest_data_year <- 2021 # update to the last year available in FAO data, likely the same for all FAO data included.

# clean up commodities value data for ornamental fishing revenue
ornamental_value <- commodities_value_raw %>%
  janitor::clean_names() %>% 
  # filter to ornamental commodities 
  filter(str_detect(commodity_name_en, "Ornamental")) %>% 
  rename(country = "reporting_country_name_en", 
         partner_country = "partner_country_name_en",
         commodity = "commodity_name_en",
         trade = "trade_flow_name_en") %>%
  filter(trade == "Exports" & commodity != "Ornamental freshwater fish") %>% # remove all other fish (commodities) that are not ornamentals (bc aquarium = ornamental), don't include the freshwater ornamental fish (interested in saltwater/ocean only)
  # fao cleaning function found in workflow/R
  fao_query_data_tidy(initial_data_year = 2019,
                          last_data_year = 2021) %>% # used function to clean and pivot data, remove Flag columns (definitely check out this script!)
  mutate(value = (value * 1000), # convert from thousands to single dollar 
         unit = "USD (1)") %>% 
  # add iso3 column
  mutate(iso3 = country_regex_to_iso3c(country)) # get iso3 code of reporting country
```

## Revenue

Sum annual revenue for each country

```{r}
# sum for each country 
ornamental_revenue <- ornamental_value %>% 
  group_by(country, year) %>% 
  summarize(value = sum(value, na.rm = TRUE),
            unit = first(unit)) %>% 
  mutate(sector = "aqf", # code for aquarium fishing sector
         data_source = "FAO global trade") %>% 
  select(country, year, value, unit, sector, data_source) %>% 
  # add iso3 column
  mutate(iso3 = country_regex_to_iso3c(country)) %>%  # get iso3 code of reporting country
  relocate(iso3, .after = country)

# tidy up a couple of country names
ornamental_revenue_renamed <- ornamental_revenue %>% 
  # rename to "United Kingdom" (mostly for plotting)
  mutate(country = str_remove_all(country,
                                  pattern = " of Great Britain and Northern Ireland")) %>% 
  # renamed to "Netherlands"
  mutate(country = str_replace_all(country,
                                   # using fixed() here is important
                                   pattern = fixed("Netherlands (Kingdom of the)"),
                                   replacement = "Netherlands"))


# why do we need fixed()? Not entirely sure, but the following 3 methods didn't work as intended. I think it may have to do with the punctuation ()
# mutate(country = str_remove_all(country, pattern = "(Kingdom of the)")) 
# yields "Netherlands ()"
# str_replace_all(country, pattern = " (Kingdom of the)", replacement = "")
# does nothing
# str_replace_all(country, pattern = "Netherlands (Kingdom of the)", replacement = "Netherlands")
# does nothing


```

## Join & Save

Join with OHI Regions

```{r}
aqf_ohi_join <- left_join(region_names, ornamental_revenue_renamed, by = c("eez_iso3" = "iso3")) %>% 
  # fill in aqf for countries without data
  mutate(sector = "aqf")

# leaving in "country" column for easy processing later (if "country" is NA, we know that the FAO didn't have any data on that country/region or didn't include that country in its dataset at all)

```
Save intermediate data (revenue table joined with OHI regions) as csv in `int` folder

```{r}
# write as intermediate file
#write_csv(aqf_ohi_join, here(int_dir, "aqf_revenue.csv"))


```


## Preliminary plotting

### Interactive line plot

```{r}
m <- list(
  l = 60,
 # r = 5,
  b = 80,
  t = 80,
  pad = 4
)

aqf_plot <- plot_ly(data = ornamental_revenue_renamed,
                    x = ~year,
                    y = ~value,
                    color = ~country,
                    type = "scatter",
                    mode = "lines") %>% 
  layout(title = "Aquarium Fishing Annual Revenue (USD)", 
         margin = m,
         xaxis = list(title = "Year"),
         yaxis = list(title = "Revenue (USD)"))

aqf_plot

```


Aquarium fishing (AQF) revenue in Spain appears to far exceed any other country. From what little exposure to the aquarium hobbyist world I've had, I expected to see Thailand, Indonesia, Philippines, and the US & UK in the top 10. I'm interested in learning more about how and why the AQF revenue is so high in Spain and the Netherlands. I know that Spain is one of the world's biggest players in the fish and seafood export market, but I have no knowledge about why their aquarium fish exports are showing up as such high values in this dataset.





### Highest AQF Revenue 
Let's look at a comparison of the countries that have the highest revenue from aquarium fishing: 

```{r fig.height=4, fig.width=8}
# ============ Data exploration, wrangling, subsetting =======================
# first: view top 10 revenue over all 3 years
ornamental_revenue_renamed %>% 
  ungroup() %>% 
  slice_max(value, n = 10) %>% # select top 10 rows based on "value" column
  select(country, year, value) %>% # simplify for readability
  kableExtra::kbl() %>% # using {kableExtra} and styling for customized table output
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"),
                full_width = F, position = "left",
                html_font = "Cambria")

# top revenue per year -------------------------------------------
# top 10
top_10_revenue_yearly <- ornamental_revenue_renamed %>% 
  ungroup() %>% 
  group_by(year) %>% 
  slice_max(value, n = 10) %>% 
  select(country, year, value) # %>% 
#  mutate(year = as.factor(year)) %>% 
#  mutate(year = fct_relevel(year, "2019", "2020", "2021"))

# top 5 
top_5_revenue_yearly <- ornamental_revenue_renamed %>% 
  ungroup() %>% 
  group_by(year) %>% 
  slice_max(value, n = 5) %>% 
  select(country, year, value)
  

# check:
unique(top_5_revenue_yearly$country)

# pull out names of top 5 countries
top_5_countries <- c(unique(top_5_revenue_yearly$country))



# ================= Plotting ==================================
# ---------- plot of top 10 countries over time ---------------
top_10_plot <- ggplot(data = top_10_revenue_yearly,
       aes(x = as.numeric(year), # needs to be numeric, otherwise geom_line() won't work
           y = value,
           color = country,
           fill = country)) +
  geom_point() +
  geom_line() + 
#  geom_col(position = "dodge") +
  # custom colorblind-friendly palette
  paletteer::scale_color_paletteer_d(palette = "colorBlindness::paletteMartin") +
  scale_fill_paletteer_d(palette = "colorBlindness::paletteMartin") +
#  paletteer::scale_color_paletteer_d("rcartocolor::Safe") +
#  scale_fill_paletteer_d("rcartocolor::Safe") +
#  geom_line() +
  theme_minimal() +
  labs(x = "",
       y = "",
       color = "",
       fill = "",
       title = "Top 10 Countries: Total Annual Value of Aquarium Fishing Exports (USD) (2019-2021)",
       subtitle = "Data Source: FAO (2024)") +
  
  # axis breaks & lables
  scale_y_continuous(labels = scales::label_dollar(scale = 0.000001, suffix = "M")
                   #  , breaks = c(25000, 50000, 75000, 100000, 125000)
                     ) +
  scale_x_continuous(breaks = c(2019, 2020, 2021))


# --------- plot of top 5 countries over time -----------------
top_5_plot <- ggplot(data = top_5_revenue_yearly,
       aes(x = as.numeric(year), # needs to be numeric, otherwise geom_line() won't work
           y = value,
           color = country,
           fill = country)) +
  geom_point() +
  geom_line() + 
#  geom_col(position = "dodge") + # couldn't get this to play nicely with fct_reorder() for the 3 years
  # custom colorblind-friendly palette
  paletteer::scale_color_paletteer_d("rcartocolor::Safe") +
  scale_fill_paletteer_d("rcartocolor::Safe") +
#  geom_line() +
  theme_minimal() +
  labs(x = "",
       y = "",
       color = "",
       fill = "",
       title = "Top 5 Countries: Total Annual Value of Aquarium Fishing Exports (USD) (2019-2021)",
       subtitle = "Data Source: FAO (2024)") +
  
  # axis breaks & lables
  scale_y_continuous(labels = scales::label_dollar(scale = 0.000001, suffix = "M")
                   #  , breaks = c(25000, 50000, 75000, 100000, 125000)
                     ) +
  scale_x_continuous(breaks = c(2019, 2020, 2021))




# ------- show plots --------
top_10_plot
top_5_plot
```




```{r fig.height=4, fig.width=8}
ggplot(data = top_5_revenue_yearly,
       aes(x = fct_reorder(country, value), 
           y = value,
           color = country,
           fill = country)) +
  geom_col(position = "dodge") +
  # custom colorblind-friendly palette
  paletteer::scale_color_paletteer_d("rcartocolor::Safe") +
  scale_fill_paletteer_d("rcartocolor::Safe") +
  facet_wrap(~year) +
   scale_y_continuous(labels = scales::label_dollar(scale = 0.000001, suffix = "M")
                   #  , breaks = c(25000, 50000, 75000, 100000, 125000)
                     ) +
  coord_flip() +
  theme_minimal() +
  labs(x = "", 
       y = "", 
       color = "",
       fill = "",
       title = "Top 5 Aquarium Fishing Countries by Exports (USD) (2019-2021)") +
  theme(
    legend.position = "none"
  )
```







