---
title: "OHI `r format(Sys.Date(), '%Y')` - CF Data Preparation"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

# CF Livelihoods: Proportion of Tourism Jobs per Country/Region per Year

## Overview

* **Data:** Labor Force & Employment Data

  * Labor Force data from World Bank (downloaded June 28. 2024)

      - https://data.worldbank.org/indicator/SL.TLF.TOTL.IN
      
  * OECD (Employment in fisheries, aquaculture and processing, 2009 - 2021) (downloaded July 2, 2024) -- for cf job data
  
    - <https://data-explorer.oecd.org/vis?fs[0]=Topic%2C1%7CAgriculture%20and%20fisheries%23AGR%23%7CFisheries%20and%20aquaculture%23AGR_FSA%23&pg=0&fc=Topic&bp=true&snb=6&vw=ov&df[ds]=dsDisseminateFinalDMZ&df[id]=DSD_FISH_EMP%40DF_FISH_EMPL&df[ag]=OECD.TAD.ARP&df[vs]=1.0&dq=.A...._T._T&pd=2009%2C&to[TIME_PERIOD]=false&ly[cl]=TIME_PERIOD&ly[rs]=REF_AREA&ly[rw]=DOMAIN>
    
  * FAO Yearbook (downloaded July 2, 2024) -- for cf job gapfilling if needed
  
    - <https://openknowledge.fao.org/server/api/core/bitstreams/2be6c2fa-07b1-429d-91c5-80d3d1af46a6/content>
      
  * ILOSTAT (downloaded July 2, 2024) -- for cf wage data
  
      - https://rshiny.ilo.org/dataexplorer46/?lang=en&id=EAR_4MTH_SEX_ECO_CUR_NB_A 
      - select Rev 3.1.B: Fishing
      
  * Ex-Vessel Price Data (downloaded August 24, 2023) -- for cf wage data
      - from emlab, see README.md


### Setup

```{r}
# load packages
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  here,
  janitor,
  terra,
  readxl,
  RColorBrewer,
  foreach,
  doParallel, # for using multiple cores
  tidyverse, 
  httr,
  plotly,
  zoo, # for gapfilling
  pdftools
)
#remotes::install_github("skgrange/threadr") # for na_extrapolate
library(threadr)


# source 
source(here("workflow/R/common.R"))

source(here('workflow/R/fao_fxn.R')) # function for cleaning old version of FAO data 

source(here('workflow/R/fao_online_portal_clean.R'))

# set year and file path info
current_year <- 2024 # Update this!!

version_year <- paste0("v",current_year)
data_dir_version_year <- paste0("d", current_year)
data_path <- here("globalprep", "le", version_year)

# Raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")

# world bank raw data directory
wb_dir <- here(raw_data_dir, "WorldBank", data_dir_version_year)

# ILO raw data directory
ilo_dir <- here(raw_data_dir, "ILOSTAT", data_dir_version_year)

# FAO raw data directory
fao_dir <- here(raw_data_dir, "FAO", data_dir_version_year)

# OECD raw data directory
oecd_dir <- here(raw_data_dir, "OECD", data_dir_version_year)

# FAO capture raw data directory
fao_capture_dir <- here(raw_data_dir, "FAO_capture", data_dir_version_year)

# Ex-vessel raw data directory
ex_vess_dir <- here(raw_data_dir, "ex-vessel-price-database-updated")

# output data dir for intermediate data products
int_dir <- here(data_path, "int")
```

Create country code --> iso3 function
```{r}
library(countrycode)

country_regex_to_iso3c <- function(country_string) {
  country_string %>%
    countrycode::countrycode(origin = "country.name", destination = "iso3c", origin_regex = TRUE)
}
```

## Read in Data

```{r}
#----------OHI regions data--------------------
# read in OHI regions for joining
region_names <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/eez/spatial/regions_list.csv") %>% 
  janitor::clean_names() %>% 
  select(-notes)

#----------Labor Force Data (from earlier)-----
labor_force <- read_csv(here(data_path, "int", "liv_labor_force.csv"))

############################################
# Livelihoods: Number of fisher jobs
############################################

#---------- cf OECD data---------------------------
oecd_cf_job_raw <- read_csv(here(oecd_dir, "OECD_2009-2021_employment_fish_aqua_processing_raw.csv"))

#------cf FAO Yearbook Number of Fishers-------
# turn the number of fishers table from the FAO yearbook into a data frame and clean.
# note: world total includes countries and regions not within the table.  Unsure of where that data comes from; look for methodology.
fao_pdf <- pdf_text(here(fao_dir, "fao_statistics_yearbook_07_03_24.pdf"))

fishers_table <- fao_pdf[[169]] %>% #page number of the pdf
  as.data.frame()

#specify columns 
columns <- c("country", "1995", "2000", "2005", "2010", "2015", "2016", "2017", "2018", "2019", "2020","2021")

#turn the pdf page into a clean dataframe
fao_fisher_jobs <- fishers_table %>%
  separate_rows('.', sep = "\n") %>% #separate rows based on line breaks
  rename(main_column = ".") %>%
  mutate(main_column = str_remove_all(main_column, "\\*")) %>% #remove the * 
  mutate(main_column = str_replace_all(main_column, "E(?!\\p{L})", " ")) %>% #replace all the Es not followed by letters (used to indicate estimate)
  slice(7:69) %>% #remove all rows not in the table
  separate(main_column, into = columns, sep = " {2,}", extra = "merge") %>% #separate into columns based on double spaces
  pivot_longer(cols = -c("country"), names_to = "year") %>% 
  mutate(value = str_squish(value)) %>% #remove extra spaces
  mutate(value = str_remove_all(value, " ")) %>% 
  mutate(value = if_else(value == "…" , NA, value)) %>% 
  mutate(value = as.numeric(value)) %>% 
  mutate(unit = "persons (1)",
         sector = "cf",
         data_source = "FAO number of fishers") %>% 
  slice(-c(1:44))

############################################
# Livelihoods: Quality of fisher jobs
############################################

#-------cf ILO wage data (2009 - 2021) -------
# this data has limited entries, so local currency will need to be ppp adjusted to increase the number of observations per geo area
cf_ilo_wage_raw <- read_csv(here(ilo_dir, "EAR_4MTH_SEX_ECO_CUR_NB_A-filtered-2024-07-03-cf-wage.csv"))

cf_wage_ppp_raw <- read_csv(here(oecd_dir, "OECD_2009-2020_wage_quality_annual_ppp_raw.csv"))

############################################
# Economies: Revenue of commercial fishing
############################################

fao_capture_raw <- read_csv(here(fao_capture_dir, "Global_capture_production_Quantity_1950-2021.csv"))

exvessel_prices_raw <- read_csv(here(ex_vess_dir, "price-db-results", "exvessel_price_database_1976_2019.csv"))
```

# Livelihood: Number of Jobs -- cf

(put this within previous chunk, read in data etc. for number of tourism jobs, join with labor_regions, make new proportion of tourism jobs column, group by year, do prelim plotting etc. and evaluate data quality etc.)

### Cleaning and wrangling data
```{r}
# total number of people employed by the fishing (previously commercial fishing, cf -- see README.md) sector

# no NAs, which means it may have been gapfilled already
unique(is.na(oecd_cf_job_raw$OBS_VALUE))

# ============== cf OECD number of fishers data ============== 
#note that this is data for the same category as the FAO fishers data, FAO data will be used to gapfill
oecd_fisher_jobs <- oecd_cf_job_raw %>% 
  janitor::clean_names() %>% 
  select(ref_area, reference_area, working_domain, time_period, obs_value, unit_of_measure) %>% 
  filter(working_domain %in% c("Marine fishing, nei","Marine deep-sea fishing", "Marine coastal fishing","Subsistence"))  %>% 
  mutate(year = as.numeric(time_period),
         value = obs_value,
         unit = ("persons (1)"),
         sector = "cf",
         data_source = "oecd sustainable economies") %>% 
  select(-c(unit_of_measure, obs_value, time_period, working_domain))

unique(is.na(oecd_fisher_jobs$value))
# FALSE -- no NAs

# ============== Run ohicore name_2_rgn ==================================

# check names
#china, taiwan province of or china (provice of china) should pick it up as taiwan
# "Chinese Taipei", "China (People’s Republic of)", "Korea"

oecd_fisher_jobs_fix <- oecd_fisher_jobs %>%
  mutate(reference_area = case_when(
    reference_area=="Chinese Taipei" ~ "Taiwan",
    reference_area=="China (People’s Republic of)" ~ "China",
    reference_area=="Korea" ~ "South Korea",
    TRUE ~ reference_area # Everything else, leave it be
  ))

oecd_fisher_jobs_rgn <- name_2_rgn(df_in = oecd_fisher_jobs_fix, 
                       fld_name='reference_area',
                      flds_unique = c("year"))

oecd_duplicates <- oecd_fisher_jobs_rgn[duplicated(oecd_fisher_jobs_rgn[, c("reference_area", "year")]),]

oecd_ref_duplicates <- unique(oecd_duplicates$reference_area)
oecd_rgn_duplicates <- unique(oecd_duplicates$rgn_name)

oecd_diff <- setdiff(oecd_rgn_duplicates, oecd_ref_duplicates)
oecd_diff # Vietnam is the only difference, and name_2_rgn works, so we are good to move forward! The raw data had duplicates in observation by year and region, so now we have to aggregate.
```

It should be noted that for the OECD number of fishers data, it has no NAs. Additionally, for the reference area of the Philippines, for 2009-2020 the observed value is consistently 120,000.  Therefore, the values have either been rounded or there has already been some sort of gapfilling. Within `oecd_cf_job_raw`, there is a column named "DECIMAL", which could allude to some sort of rounding.  Investigation was done to find OECD's metadata, however the link failed (below).  **WILL RETURN!**

Page with metadata: <https://www.oecd-ilibrary.org/economics/data/oecd-stat/data-warehouse_data-00900-en?parentId=http%3A%2F%2Finstance.metastore.ingenta.com%2Fcontent%2Fcollection%2Fdata-00285-en>
Failed link: <https://stats.oecd.org/wbos/default.aspx?checksum=40e0edc46aec7aee1cce210b898d5bd6&timestamp=1720045544&cid=id5845&oecdstat=data-00285-en&doi=data-00900-en&return_url=https%3A%2F%2Fwww.oecd-ilibrary.org%3A443%2Fcontent%2Fdata%2Fdata-00900-en&institution_name=UC+at+Santa+Barbara&lang=en>

```{r}
# preliminary plot
prelim_plot <- plotly::plot_ly(oecd_fisher_jobs_rgn, x = ~year, y = ~value, color = ~rgn_name, type = "scatter", mode = "lines") %>% 
  layout(title = "All Regions: Marine Fishing Employment", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Number of people by region employed by the fishing sector"))
prelim_plot
```

### Aggregating data
```{r}
# ============== Aggregating data ==================================
length(unique(oecd_fisher_jobs_rgn$reference_area)) #44
length(unique(oecd_fisher_jobs_rgn$rgn_name)) #44

duplicate_exists <- oecd_fisher_jobs_rgn %>% 
  group_by(rgn_name, year) %>% 
  filter(n() > 1)

bang_filt <- oecd_fisher_jobs_rgn %>% 
  filter(rgn_name %in% "Bangladesh")
head(bang_filt) # still have multiple entries per region and year from raw oecd data, need to aggregate

cf_jobs_rgn_agg <- oecd_fisher_jobs_rgn %>%
  group_by(rgn_id, rgn_name, year) %>%
  dplyr::summarize(value = sum(value, na.rm=TRUE)) %>%
  ungroup() %>% 
  dplyr::mutate(year = as.numeric(year)) 
```

```{r}
# plot of aggregated data
prelim_plot_agg <- plotly::plot_ly(cf_jobs_rgn_agg, x = ~year, y = ~value, color = ~rgn_name, type = "scatter", mode = "lines") %>% 
  layout(title = "All Regions: Marine Fishing Employment", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Number of people by region employed by the fishing sector"))
prelim_plot_agg

#plot without the large outliers Bangladesh and China
remove_values <- c("Bangladesh", "China")
cf_job_subset <- cf_jobs_rgn_agg %>% 
  filter(!(rgn_name %in% remove_values))

prelim_plot_lim <- plotly::plot_ly(cf_job_subset, x = ~year, y = ~value, color = ~rgn_name, type = "scatter", mode = "lines") %>%
  layout(title = "All Regions: Marine Fishing Employment", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Number of people by region employed by the fishing sector"))
prelim_plot_lim
```

### Create the proportion between tourism jobs and the total labor force
```{r}
# join labor force data with cf employment
cf_job_labor_join <- full_join(labor_force, cf_jobs_rgn_agg)

# make proportion of people employed in the commercial fishing sector
cf_job_prop <- cf_job_labor_join %>% 
  dplyr::relocate(value, .before = labor_force_total) %>%
  mutate(cf_prop_percent = (value/labor_force_total)*100) %>%
  select(-c(data_source, sector))
```

Preliminary plot of the percent of labor force working in the fishing sector (cf)
```{r}
cf_prop_plot <- plotly::plot_ly(cf_job_prop, x = ~year, y = ~cf_prop_percent, color = ~rgn_name, type = "scatter", mode = "lines") %>% 
  layout(title = "All Regions: Percent of Labor Force Employed by Marine Fishing Sector", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Percent by region employed by the fishing sector"))
cf_prop_plot
```

Saving intermediate data
```{r}
# %>% 
#   mutate(sector = "cf",
#          data_source = "ILOSTAT")

# write_csv(cf_job_prop_agg, here(int_dir, "cf_jobs.csv"))
```

### Gapfilling using FAO Yearbook
```{r}
# ============== Run ohicore name_2_rgn ==================================

# These data were removed for not being of the proper rgn_type (eez,ohi_region) or mismatching region names in the lookup tables:
#         tmp_type
# tmp_name landlocked
#   uganda         11

fao_fisher_jobs_fix <- fao_fisher_jobs %>%
  filter(country != "World total") %>% 
  mutate(country = case_when(
    country=="China,Taiwan" ~ "Taiwan",
    country=="Dem R Congo" ~ "Democratic Republic of the Congo",
    country=="DPR Korea" ~ "North Korea",
    country=="Rep.of Korea" ~ "South Korea",
    country=="Faroe Is" ~ "Faeroe Islands",
    country=="Russian Fed" ~ "Russia",
    country=="USA" ~ "United States",
    country=="Rep.of Korea" ~ "South Korea",
    TRUE ~ country # Everything else, leave it be
  ))

fao_fisher_jobs_rgn <- name_2_rgn(df_in = fao_fisher_jobs_fix, 
                       fld_name='country',
                      flds_unique = c("year"))


fao_duplicates <- fao_fisher_jobs_rgn[duplicated(fao_fisher_jobs_rgn[, c("country", "year")]),]

fao_ref_duplicates <- unique(fao_duplicates$country)
fao_rgn_duplicates <- unique(fao_duplicates$rgn_name)

fao_diff <- setdiff(fao_rgn_duplicates, fao_ref_duplicates)
fao_diff # none!

length(unique(fao_fisher_jobs_rgn$rgn_name)) # 57

# see where there is overlapping
fao_rgns <- unique(fao_fisher_jobs_rgn$rgn_name)
oecd_rgns <- unique(cf_jobs_rgn_agg$rgn_name) # use the data from before labor_force join

fao_oecd_diff <- setdiff(fao_rgns, oecd_rgns)
fao_oecd_diff # 27 regions could be filled! If the methodology makes sense.
```

Only years 2015-2021 could be gapfilled by the FAO Yearbook. This is still a good amount.
```{r}
# ============= Evaluating FAO Yearbook Methodology =========

unique(fao_fisher_jobs_rgn$year)
# years are weird: has 2010, then jumps to 2015 and increases sequentially to 2020 from there.
 # [1] "1995" "2000" "2005" "2010" "2015" "2016" "2017" "2018" "2019" "2020" "2021"

#make a list of the countries in the fao data but not oecd, these are the ones we want to add
missing_cf_oecd <- setdiff(fao_fisher_jobs_rgn$rgn_name, cf_jobs_rgn_agg$rgn_name)
missing_cf_oecd

fao_missing_oecd <- fao_fisher_jobs_rgn %>% 
  filter(rgn_name %in% c(missing_cf_oecd)) %>% 
  select(rgn_id, rgn_name, year, value)

fishing_jobs_final <- cf_jobs_rgn_agg %>% 
  rbind(fao_missing_oecd)

# evluate binded dataset containing oecd and fao data
length(unique(fishing_jobs_final$rgn_name)) #72!

# create proportion using the binded data
cf_job_final_join <- full_join(labor_force, fishing_jobs_final)

# make proportion of people employed in the commercial fishing sector
cf_job_prop <- cf_job_labor_join %>% 
  dplyr::relocate(value, .before = labor_force_total) %>%
  mutate(cf_prop_percent = (value/labor_force_total)*100) %>%
  select(-c(data_source, sector))
```


# Livelihood: Quality of Jobs -- cf

  - note: this data came from many sources and was compiled by ILO.  PPP GDP adjustments come from OECD

Cleaning and wrangling data
```{r}
# separating into the different currencies
cf_ilo_wage <- cf_ilo_wage_raw %>% 
  janitor::clean_names() %>% 
  rename(currency = classif2_label, 
         year = time) %>% 
  select(ref_area_label, year, currency,obs_value, note_indicator_label, source_label)

length(unique(cf_ilo_wage$ref_area_label))

unique(cf_ilo_wage$currency)

################ "Currency: Local currency" "Currency: 2017 PPP $"     "Currency: U.S. dollars" 

# Local currency -- needs to be PPP adjusted
cf_wage_local <- cf_ilo_wage %>% 
  filter(currency %in% "Currency: Local currency")

# Currency: U.S. dollars -- needs to be PPP adjusted
cf_wage_usd <- cf_ilo_wage %>% 
  filter(currency %in% "Currency: U.S. dollars")

# 2017 PPP $
cf_wage_2017ppp <- cf_ilo_wage %>% 
  filter(currency %in% "Currency: 2017 PPP $")

############# Bring in OECD data to adjust monthly earnings to 2017 PPP
cf_wage_ppp <- cf_wage_ppp_raw %>% 
  

```


Gapfilling if necessary
```{r}

```


```{r}
## preliminary plot 

```


Clean up ILO data, join with OHI regions
```{r}
# join with OHI regions

```


Plot
```{r}


```

# Economies: Revenue -- cf

```{r}
# read in cf revenue data from v2023, using FAO Capture data, Ex-vessel data, 
cf_revenue <- read_csv(here("globalprep", "le", "v2023", "int", "cf_revenue.csv"))

# ============== Run ohicore name_2_rgn ==================================

# check names
# These data were removed for not having any match in the lookup tables:
# 
# bonaire/s.eustatius/saba (need to be disaggregated) british indian ocean ter (DNE)    channel islands (split by eez area: Jersey, larger island)    french southern terr (DNE)
#                        1                        1                        1                        1 
#.   isle of man (cannot be lumped w Britain, maybe Ireland)  netherlands antilles  (curacao)      other nei (DNE)        saint barthélemy (DNE)
#                        1                        1                        1                        1 
# saint helena/asc./trist. (Saint Helena)   saint-martin (french)  (DNE)         yugoslavia sfr (DNE)
#                        1                        1                        1 
# 
# These data were removed for not being of the proper rgn_type (eez,ohi_region) or mismatching region names in the lookup tables:
#            tmp_type
# tmp_name    disputed landlocked
#   ethiopia         0         16
#   palestine       25          0

length(unique(cf_revenue$country)) #203

cf_revenue_fix <- cf_revenue %>%
  mutate(country = case_when(
    country=="Bonaire/S.Eustatius/Saba" ~ "Bonaire", # how to disaggregate instead of aggregate?
    country=="British Indian Ocean Ter" ~ "British Indian Ocean Territory",
    country=="Channel Islands" ~ "Jersey", # would be disaggregated into Guernsey and Jersey
    country=="Saint Helena/Asc./Trist." ~ "Saint Helena", 
    country=="Saint-Martin (French)" ~ "Northern Saint-Martin",
    country=="Netherlands Antilles" ~ "Curacao",
    country=="Puerto Rico" ~ "Puerto Rico and Virgin Islands of the United States",
    country=="China, Hong Kong SAR" ~ "China",
    country=="China, Macao SAR" ~ "China",
    country=="United Republic of Tanzania, Zanzibar" ~ "Tanzania",
    country=="Tanzania, United Rep. of" ~ "Tanzania",
    TRUE ~ country # Everything else, leave it be
  ))

# COME BACK!!!! look at rasterized fishing effort, to determine the relative proportion of it within each eez, and distribute catch relative to that proportion

cf_revenue_rgn <- name_2_rgn(df_in = cf_revenue_fix, 
                             fld_name='country',
                             flds_unique = c("year"))

cf_revenue_rgn_agg <- cf_revenue_rgn %>% 
  select(rgn_id, rgn_name, year, value, unit, sector, data_source) %>% 
  group_by(rgn_id, rgn_name, year) %>% # because china needs to be aggregated and has duplicate years
  dplyr::summarize(value = sum(value, na.rm=TRUE)) %>%
  ungroup() %>% 
  dplyr::mutate(year = as.numeric(year)) 

rev_duplicates <- cf_revenue_rgn[duplicated(cf_revenue_rgn[, c("country", "year")]),]
rev_ref_duplicates <- unique(rev_duplicates$country)
rev_rgn_duplicates <- unique(rev_duplicates$rgn_name)
rev_diff <- setdiff(rev_rgn_duplicates, rev_ref_duplicates)
rev_diff # none! Move on
```

CF Revenue Plot
```{r}
cf_rev_plot <- plotly::plot_ly(cf_revenue_rgn_agg, x = ~year, y = ~value, color = ~rgn_name, type = "scatter", mode = "lines") %>% 
  layout(title = "All Regions: Yearly Revenue for the Marine Fishing Sector", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Revenue of all catch (USD)"))
cf_rev_plot

cf_revenue_china <- cf_revenue_rgn_agg %>% 
  filter(rgn_name %in% "China") #before aggregation, 

plotly::plot_ly(cf_revenue_china, x = ~year, y = ~value, color = ~rgn_name, type = "scatter", mode = "lines") %>% 
  layout(title = "All Regions: Yearly Revenue for the Marine Fishing Sector", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Revenue of all catch (USD)"))
```

Saving intermediate data
```{r}
# columns should be: rgn_id, rgn_name, year, usd (previously value), unit, sector, usd_yr (in this case = year)

cf_revenue_int <- cf_revenue_rgn_agg %>% 
  mutate(usd = value,
         unit = "USD (1)",
         sector = "cf",
         usd_yr = year) %>% 
  select(-value)

# write_csv(cf_revenue_int, here(int_dir, "eco_cf_usd_pre.csv"))
```


```{r}
# ###################################################
# 
# # Bring in imports and exports from FAO yearbook
# imp_table <- fao_pdf[[190]] %>% #page number of the pdf
#   as.data.frame()
# 
# #specify columns 
# columns <- c("blank", "aggregate_geo_area", "2000", "2005", "2010", "2015", "2016", "2017", "2018", "2019", "2020","2021")
# 
# #turn the pdf page into a clean dataframe
# cf_fao_imports <- imp_table %>%
#   separate_rows('.', sep = "\n") %>% #separate rows based on line breaks
#   rename(main_column = ".") %>%
#   mutate(main_column = str_remove_all(main_column, "\\*")) %>% #remove the *
#   slice(13:49) %>% #remove all rows not in the table
#   separate(main_column, into = columns, sep = " {3,}", extra = "merge") %>% #separate into columns based on double spaces
#   pivot_longer(cols = -c("country"), names_to = "year") %>% 
#   mutate(value = str_squish(value)) %>% #remove extra spaces
#   mutate(value = str_remove_all(value, " ")) %>% 
#   mutate(value = if_else(value == "…" , NA, value)) %>% 
#   mutate(value = as.numeric(value)) %>% 
#   mutate(unit = "persons (1)",
#          sector = "cf",
#          data_source = "FAO number of fishers") %>% 
#   slice(-c(1:44))
```


