---
title: "OHI `r format(Sys.Date(), '%Y')` - Fish Processing Livelihoods & Economies Data Preparation"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

# Economies: Fish Processing: Proportion of Jobs per Country per Year (2019-2021)

## Overview

* **Data:** [OECD Employment in Fisheries, Aquaculture, and Processing Dataset](https://data-explorer.oecd.org/vis?fs[0]=Topic%2C1%7CAgriculture%20and%20fisheries%23AGR%23%7CFisheries%20and%20aquaculture%23AGR_FSA%23&pg=0&fc=Topic&bp=true&snb=6&df[ds]=dsDisseminateFinalDMZ&df[id]=DSD_FISH_EMP%40DF_FISH_EMPL&df[ag]=OECD.TAD.ARP&df[vs]=1.0&dq=.A...PROC._T._T&pd=2009%2C2021&to[TIME_PERIOD]=false&ly[cl]=TIME_PERIOD&ly[rs]=REF_AREA&vw=tb)

Filtered to select:
 -  Time Period 2009-2021 (2009-most recent year of data as of July 5th, 2024).
 -  "Working domain" --> "Processing" 
 -  "Sex" --> "Total"
 -  "Working status" --> "Total"

* **Data:** Labor Force Data

  * Labor Force data from World Bank (downloaded June 28. 2024)

      - https://data.worldbank.org/indicator/SL.TLF.TOTL.IN
      


## Setup

```{r}
# load packages
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  here, # reproducible file paths
  janitor, # data cleaning: lower_snake_case 
  terra,
  countrycode,
  readxl,
  paletteer, # custom color palettes for plotting
  RColorBrewer,
  foreach,
  doParallel, # for using multiple cores
  tidyverse, 
  httr,
  plotly, # interactive plots
  kableExtra, # for more aesthetically pleasing tables
  zoo # for gapfilling
  
)
#remotes::install_github("skgrange/threadr") # for na_extrapolate
library(threadr)


# source 
source(here("workflow", "R", "common.R"))


# set year and file path info
current_year <- 2024 # Update this!!

version_year <- paste0("v",current_year)
data_dir_version_year <- paste0("d", current_year)
data_path <- here("globalprep", "le", version_year)

# Raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")

# world bank raw data directory
oecd_data_dir <- here(raw_data_dir, "OECD", data_dir_version_year)


# output data dir for intermediate data products
int_dir <- here(data_path, "int")
```


Create country code --> iso3 function

```{r}
library(countrycode)

country_regex_to_iso3c <- function(country_string) {
  country_string %>%
    countrycode::countrycode(origin = "country.name", destination = "iso3c", origin_regex = TRUE)
}
```

### Read in Data

```{r}
# OHI regions data ----
# read in OHI regions for joining
region_names <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohi-global/draft/eez/spatial/regions_list.csv") %>% 
  janitor::clean_names() %>% 
  select(-notes)

# labor force from World Bank (cleaned, joined with OHI regions)
labor_force <- read_csv(here(int_dir, "labor_force.csv"))

# Aquarium Fishing FAO data 
processing_jobs_raw <- read_csv(here(oecd_data_dir,
                                     "OECD_2009-2021_employment_fish_aqua_processing_raw.csv"))

```


## Data cleaning & preliminary exploration

```{r}
processing_clean <- processing_jobs_raw %>% 
  # lower_snake_case
  janitor::clean_names()
# preliminary exploration 
#str(processing_jobs_raw)
# > unique(processing_clean$working_domain)
# [1] "Fishing sector"          "Marine deep-sea fishing" "Inland waters fishing"  
# [4] "Aquaculture"             "Marine fishing, nei"     "Processing"             
# [7] "Marine coastal fishing"  "Unspeficied"             "Subsistence" 
# eventually, we want to just have it filtered to processing:
# [1] "Processing" # <-- this is what we want to be extra sure of!
# > unique(processing_clean$frequency_of_observation)
# [1] "Annual"
# > unique(processing_clean$measure)
# [1] "EMP"
# > unique(processing_clean$measure_2)
# [1] "Employment"
# > unique(processing_clean$unit_of_measure)
# [1] "Persons"
# etc...

# Number of unique countries:
length(unique(processing_clean$reference_area)) 
# 50 ...without having filtered to only OHI regions& countries

# create cleaned dataframe with iso3 column -------------------
fp_clean <- processing_clean %>% 
  # select relevant columns
  select(c(reference_area, time_period, obs_value, unit_of_measure, working_domain)) %>% 
  rename(year = time_period,
         fp_jobs = obs_value) %>% 
  # double check to make sure we only have "Processing" data (don't include data entries for working domains like Aquaculture, Fishing sector: Inland waters fishing, for example, which are part of the full dataset on the OECD data portal.)
  filter(working_domain %in% c("Processing")) %>% 
  # add iso3 column
  mutate(eez_iso3 = country_regex_to_iso3c(reference_area))

length(unique(fp_clean$reference_area)) 
# only 33 now...
```


```{r}
# check names
fp_rgn <- name_2_rgn(df_in = fp_clean, 
                       fld_name='reference_area',
                      flds_unique = c("year"))

fp_clean <- fp_clean %>%
  mutate(reference_area = case_when(
    reference_area == "China (People’s Republic of)" ~ "China",
    reference_area == "Korea" ~ "South Korea",
    TRUE ~ reference_area # Everything else, leave it be
    ))

fp_rgn <- name_2_rgn(df_in = fp_clean, 
                       fld_name='reference_area',
                      flds_unique = c("year"))

length(unique(fp_rgn$rgn_name))
# 30 in 2024 (data up to 2021)


fp_rgn_2019_filter <- fp_rgn %>% 
  filter(year <= 2019)

length(unique(fp_rgn_2019_filter$rgn_name))
# 29 when filtered up to & including 2019
```


```{r}
# join with labor force data ------
# fish processing labor force dataframe
fp_labor <- left_join(labor_force, fp_rgn, by = c("rgn_name", "year")) %>% 
  filter(year %in% c(2009:2021))

na_count <- sum(is.na(fp_labor$working_domain))
not_na_count <- sum(!is.na(fp_labor$working_domain))

paste0("Percentage of NA data points in Fish Processing & Labor Force join: ", round(na_count / nrow(fp_labor) * 100, 2),  "%")
```

Given that we only have data for 18% of the countries/regions for 2009-2021, we may want to consider how we can account for this in our processing and analysis later.


This data does not distinguish between coastal and inland fish processing jobs. However, a key purpose of this index is to track how a country is doing over time -- assuming the proportion of coastal to inland fish processing jobs stays relatively stable within a 5-year window within a country over time, we can continue to use this data (spoke to Melanie Frazier about this issue and that was her reasoning).


Also, it seems like the territories without unique eez iso3 codes were treated as copies of admin country (reference area) fish processing jobs data -- for example, Portugal:

```{r}
fp_labor %>% filter(admin_country_name == "Portugal")

```
We can see that for each of these regions, because they don't have unique EEZ ISO3 codes, the values repeat/match exactly between each territory in Portugal for each year from 2009-2021. This occurred when creating the cleaned labor force dataset as well.

This is another issue or concern that we are navigating. My concern is specifically around island territories who are more immediately and severely sensitive to changes in ocean health -- and those sensitivities and fluctuations being obscured by using the same values for different metrics as their admin country. 


For now, let's create a tidier dataframe and make a proportional jobs column to track the number of fish processing jobs in a way that accounts for population growth within a country 

```{r}
# creating proportional jobs column ------
# new data frame
fp_jobs_prop <- fp_labor %>% 
  select(-c(reference_area)) %>% 
  ungroup() %>% 
  group_by() %>% 
  mutate(fp_proportion = fp_jobs / labor_force_total)
  



# prelim plot to visualize data
m <- list(
  l = 80,
  r = 60,
  b = 80,
  t = 80,
  pad = 4
)

fp_line_plot <- plotly::plot_ly(fp_jobs_prop, x = ~year, y = ~fp_proportion, color = ~admin_country_name, type = "scatter", mode = "lines") %>% 
  layout(title = "All Regions: Proportional Fish Processing Employment Within Total Labor Force", 
         margin = m,
         xaxis = list(title = "Year"),
         yaxis = list(title = "Percent of people by region employed in fish processing"))
fp_line_plot
```

Considering that both the total labor force and number of fish processing jobs values were copied over for both the mainland and territories of Portugal, I find the split line (which only shows Portugal when I hover over the upper and lower parts of the yellow line) confusing.


## Save intermediate data

Save data to `int` folder (in this project/on GitHub)

```{r}
fish_process_int <- fp_jobs_prop %>% select(-c(data_status, working_domain)) %>% 
  mutate(sector = "fp",
         data_source = "OECD")


#write_csv(fish_process_int, here(int_dir, "fp_jobs.csv"))
```

