---
title: 'OHI `r format(Sys.Date(), "%Y")`: Sea Surface Temperature Pressure Layer'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

# Summary

This script creates the Sea Surface Temperature (SST) layer for the 2022 global Ocean Health Index assessment.

------------------------------------------------------------------------

# Updates from previous assessment

Used updated 2021 data from NOAA

------------------------------------------------------------------------

# Data Source

Data come from NOAA's NCEI: [CoRTAD version 6](https://www.ncei.noaa.gov/products/coral-reef-temperature-anomaly-database)

See `prs_sst/v2015/dataprep.R` for preparation of the "annual_pos_anomalies" data.

**Native Data Resolution**: \~4km\
**Description**: Cortadv6_SSTA.nc = SST anomalies (weekly SST minus weekly climatological SST), weekly data for all years, degrees Kelvin Cortadv6_weeklySST.nc = SST, weekly data for all years, degrees Kelvin\
**Time Range**: 1982 - 2021 (weekly averages across all years)\
**Format**: NetCDF **Downloaded**: June 22, 2022

------------------------------------------------------------------------

## Setup

```{r setup, message=F,warning=F, eval = FALSE}

# Set options for code chunks
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'figs/', message = FALSE, warning = FALSE)

# Load packages
library(raster)
library(RColorBrewer)
library(tidyverse)
library(rgdal)
library(doParallel)
library(foreach)
library(sf)
library(ncdf4)
library(httr)
library(lubridate)
library(animation)
library(plotly)
library(here)
library(snow)
library(terra)
library(tictoc)
library(furrr)
library(purrr)
library(future)
library(beepr)

# suppress progress bars for terra
terraOptions(progress=0)

# OHI spatial files, directories, etc
source(here("workflow/R/common.R"))

# Update scenario year, set up programmatic scenario year updating
scen_year_number <- 2023
scen_year <- as.character(scen_year_number)
prev_scen_year <- as.character(scen_year_number - 1)

# Standard OHI file paths
dir_data <- paste0(dir_M, "/git-annex/globalprep/_raw_data/CoRTAD_sst/d", scen_year)
dir_int  <- paste0(dir_M, "/git-annex/globalprep/prs_sst/v", scen_year, "/int")
dir_output  <- paste0(dir_M, "/git-annex/globalprep/prs_sst/v", scen_year, "/output")
dir_anom_threshold  <- paste0(dir_M, "/git-annex/globalprep/prs_sst/sd_mean_30yr_rasters")

# Load in OHI spatial data
ohi_rasters()
regions_shape()

# Specify years for calculations
yrs <- 1982:2022

# Rainbow color scheme and load in OHI regions
cols <- rev(colorRampPalette(brewer.pal(11, 'Spectral'))(255)) 
land <- regions %>% subset(rgn_type %in% c("land", "land-disputed", "land-noeez"))

```

------------------------------------------------------------------------

## Standard Deviation Calculation

```{r}
# read in this new and old SD
sd_2023 <- rast("/home/shares/ohi/git-annex/globalprep/prs_sst/sd_mean_30yr_rasters/sst_sd_week_1.tif")
sd_2021 <- rast("/home/shares/ohi/git-annex/globalprep/prs_sst/v2021/int/sd_sst_week_1.tif")
            
# extract values
sd_2023_v <- terra::values(sd_2023, na.rm = TRUE)
sd_2021_v <- terra::values(sd_2021, na.rm = TRUE)

# sample down to 50000
sd_2023_sampled <- sample(sd_2023_v, size = 50000)
sd_2021_sampled <- sample(sd_2021_v, size = 50000)

# Create data frames
df_2023 <- data.frame(value = sd_2023_sampled, year = "2023")
df_2021 <- data.frame(value = sd_2021_sampled, year = "2021")

# ggplot 2023/21 SD
sd_23 <- ggplot(df_2023, aes(x=value)) +
  geom_histogram(bins=50, fill="skyblue", color="black", alpha=0.7) +
  labs(x="Week 1 SD", y="Count", title="SD 2023") + #Corrected title year from 2021 to 2023
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face="bold")) +
  xlim(0,4) +
  ylim(0,6000)

sd_21 <- ggplot(df_2021, aes(x=value)) +
  geom_histogram(bins=50, fill="skyblue", color="black", alpha=0.7) +
  labs(x="Week 1 SD", y="Count", title="SD 2021") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face="bold")) +
  xlim(0,4) +
  ylim(0,6000)


sd_21 + sd_23
```

------------------------------------------------------------------------

## Raw Anoms

```{r}
## compare raw pos anom values 
sst_v2023_test_raw <- rast(file.path(dir_int, "annual_pos_anomalies_sd_2016.tif"))
sst_v2021_test_raw <- rast("/home/shares/ohi/git-annex/globalprep/prs_sst/v2021/int/annual_pos_anomalies_sd_2016.tif")

df_tmp_raw <- data.frame(v2023 = values(sst_v2023_test_raw), 
                     v2021 = values(sst_v2021_test_raw))

colnames(df_tmp_raw) <- c("v2023", "v2021")

df_samp_raw_panom <- df_tmp_raw %>%
  mutate(row_number = row_number()) %>% 
  filter(row_number %in% sample(1:length(row_number), 7000))

ggplot(data = df_samp_raw_panom, aes(x=v2023, y = v2021)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0) +
  labs(x="Positive Anomalies 2023", y="Positive Anomalies 2021", title="Positive Annual Anomalies Count 23/21")
```

------------------------------------------------------------------------

## Raw Excess Anoms

```{r}
## compare raw diff values 
sst_v2023_test_raw <- rast(file.path(dir_int, "sst_diff_ocean_2008-2012.tif"))
sst_v2021_test_raw <- rast("/home/shares/ohi/git-annex/globalprep/prs_sst/v2021/int/sst_diff_ocean_2008-2012.tif")

df_tmp_raw <- data.frame(v2023 = values(sst_v2023_test_raw), 
                     v2021 = values(sst_v2021_test_raw))

colnames(df_tmp_raw) <- c("v2023", "v2021")

df_samp_raw <- df_tmp_raw %>%
  mutate(row_number = row_number()) %>% 
  filter(row_number %in% sample(1:length(row_number), 7000))

ggplot(data = df_samp_raw, aes(x=v2023, y = v2021)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0) +
  geom_abline(slope = 0, intercept = 0) +
  labs(x="Dif Anomalies 2023", y="Dif Anomalies 2021", title="2012 5 yr Excess Anomalies 23/21")
```

------------------------------------------------------------------------

```{r eval = FALSE}
## compare 2008-2012 extremes between v2023 and v2021 assessement 
sst_v2023_test <- rast(file.path(dir_output, "sst_2008-2012_1985-1989.tif"))
sst_v2021_test <- rast("/home/shares/ohi/git-annex/globalprep/prs_sst/v2021/output/sst_2008-2012_1985-1989.tif")

# extract values and put them into a data frame
df_tmp <- data.frame(v2023 = values(sst_v2023_test), 
                     v2021 = values(sst_v2021_test))

# rename columns, terra fails to keep good column names
colnames(df_tmp) <- c("v2023", "v2021")

# sample down due to size constraints
df_samp <- df_tmp %>%
  mutate(row_number = row_number()) %>% 
  filter(row_number %in% sample(1:length(row_number), 70000))

# create the plot
ggplot(data = df_samp, aes(x=v2023, y = v2021)) +
  geom_point()+
  geom_abline(slope = 1, intercept = 0) + 
  labs(x="Scaled Excess Anoms 2023", y="Scaled Excess Anoms 2021", title="Scaled Excess Anom Raster Values 23/21")

```

#### Violin

```{r}
# Load necessary packages
library(ggplot2)
library(reshape2)

# Convert data to long format
df_long <- melt(df_samp, id.vars = NULL, measure.vars = c("v2023", "v2021"), variable.name = "Year", value.name = "Value")

# Create violin plot
ggplot(df_long, aes(x = Year, y = Value)) +
  geom_violin(trim = FALSE, fill = 'blue', color = 'black') +
  theme_minimal() +
  labs(title = "Violin plot of v2023 and v2021",
       x = "Year", 
       y = "Value")


```

#### Statistics

```{r}
# Compute statistics
summary_v2023 <- summary(df_samp$v2023, na.rm = TRUE)
summary_v2021 <- summary(df_samp$v2021, na.rm = TRUE)

# Standard Deviation
sd_v2023 <- sd(df_samp$v2023, na.rm = TRUE)
sd_v2021 <- sd(df_samp$v2021, na.rm = TRUE)

# Create data frame for kable
stat_df <- data.frame(
  Statistic = c("Min", "1st Qu.", "Median", "Mean", "3rd Qu.", "Max", "SD"),
  v2023 = c(summary_v2023[1], summary_v2023[2], summary_v2023[3], summary_v2023[4], summary_v2023[5], summary_v2023[6], sd_v2023),
  v2021 = c(summary_v2021[1], summary_v2021[2], summary_v2021[3], summary_v2021[4], summary_v2021[5], summary_v2021[6], sd_v2021)
)

# Display with kable
knitr::kable(stat_df, row.names = FALSE, align = 'c', caption = "Summary Statistics for v2023 and v2021")

```

------------------------------------------------------------------------

## PRS Score Dot Plot

```{r}
# load in prs data from 23 and 21
prs_data_23 <- read.csv("output/sst_updated.csv")
prs_data_22 <- read.csv("../v2022/output/sst_updated.csv")

# join into one df
compare <- left_join(prs_data_23, prs_data_22, by = c("rgn_id", "year")) %>%
  rename(prs2023 = pressure_score.x, prs2022 = pressure_score.y)

# 2022 vs 2023 pressure scores dot plot
p <- ggplot(data = compare) +
  geom_point(aes(x = prs2023, y = prs2022)) +
  geom_abline(intercept = 0, slope = 1) +
  facet_wrap(~year) +
  labs(x="PRS Score 2023", y="PRS Score 2022", title="SST PRS Score 22/23")

print(p)


```

------------------------------------------------------------------------

## PRS Score Histogram

```{r}
# histogram comparing 2021 values to 2023 values
hist2022 <- ggplot(compare, aes(x = prs2022)) +
  geom_histogram(fill = "gray", color = "black") + 
  theme_bw() + 
  labs(x="PRS Score", y="Count", title="SST PRS Score 2022") +
  xlim(0, 1) +
  ylim(0, 300)

hist2023 <- ggplot(compare, aes(prs2023)) +
  geom_histogram(fill = "gray", color = "black") + 
  theme_bw() + 
  labs(x="PRS Score", y="Count", title="SST PRS Score 2023") +
  xlim(0, 1) +
  ylim(0, 300)

hist2022 + hist2023
```

```{r}

# Get list of rescaled and masked sst data files
sst_res_mask <- list.files(dir_output, "sst.*_rescaled_icemask.tif", full.names = TRUE)

# Create gif visualizing the rescaled and masked sst
# This loop outputs warning and error text but successfully creates .gif
saveGIF({
  for(i in 1:length(sst_res_mask)){
    
    # Get name for image from file path date
    n = sprintf("SST Pressure %s", 
                substr(sst_res_mask[i], 64, 72))
    
    # Plot and save raster raster
    plot(raster(sst_res_mask[i]), 
         zlim = c(0, 1), # fix zlimits
         axes = FALSE, box = FALSE, 
         main = n)}}, 
  ani.width = 750,
  ani.height = 400,
  movie.name = sprintf("%s/sst.gif", dir_output)) 

```
