---
title: "Prep gapfilling data"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(ggplot2)

set.seed(22774)

tmp <- data.frame(country=c(1:100), 
                  ecoregion=rep(c("A", "B"), each=50), 
                  indicatorA = rnorm(100, 8, sd=1), 
                  indicatorB = rnorm(100, 9, sd=0.5),
                  indicatorC = c(rnorm(50, 2, sd=1), rnorm(50, 8, sd=0.5)))

tmp <- tmp %>%
  mutate(indicatorA = ifelse(indicatorA>10, 10, indicatorA),
         indicatorB = ifelse(indicatorB>10, 10, indicatorB),
         indicatorC = ifelse(indicatorC > 10, 10, indicatorC),
         indicatorC = ifelse(indicatorC < 0, 0, indicatorC))


NAs <- sample(1:100, 40, replace=FALSE)

tmp$indicatorC[NAs] <- NA

write_csv(tmp, here("Reference/gapfilling/gf_explore.csv"))

```

The following is a dataset used to calculate an index. It is comprised of 3 indicators.  Notice that "indicatorC" has quite a bit of missing data. We are going to explore whether we should gapfill these data.

```{r}

gf <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohiprep_v2022/gh-pages/Reference/gapfilling/gf_explore.csv")

```

The overall index is calculated as the average of indicatorA, indicatorB, indicatorC.
Try calculating the index without gapfilling the missing data. Explore the results by plotting. Do you see any evidence of bias?

```{r}

gf2 <- gf %>%
  rowwise() %>%
  mutate(index_no_gf = mean(c(indicatorA, indicatorB, indicatorC), na.rm=TRUE),
         missing_data = ifelse(is.na(indicatorC), 1, 0))

ggplot(gf2, aes(y=index_no_gf, x = ecoregion, color=missing_data)) + 
  geom_jitter(width = 0.25)

```

Explore whether indicatorC appears corrlated with ecoregion.

```{r}

ggplot(gf2, aes(y=indicatorC, x = ecoregion)) + 
  geom_jitter(width = 0.25)

```

Gapfill missing data using the ecoregion mean of indicatorC for the country's respective ecoregion. Explore the results. Does this seem like a reasonable approach?

```{r}

gf2 <- gf %>%
  rowwise() %>%
  mutate(index_no_gf = mean(c(indicatorA, indicatorB, indicatorC), na.rm=TRUE),
         missing_data = ifelse(is.na(indicatorC), 1, 0),
         indicatorC_gf = ifelse(is.na(indicatorC) & ecoregion == "A", 
                                mean(gf$indicatorC[gf$ecoregion=="A"], na.rm=TRUE),
                                indicatorC)) %>%
  mutate(indicatorC_gf = ifelse(is.na(indicatorC_gf),
                                mean(gf$indicatorC[gf$ecoregion=="B"], na.rm=TRUE),
                                indicatorC_gf)) %>%
  mutate(index = mean(c(indicatorA, indicatorB, indicatorC_gf), na.rm=TRUE))
  

ggplot(gf2, aes(y=index, x = ecoregion, color=missing_data)) + 
  geom_jitter(width = 0.25)

```


Does the uncertainty associated with gapfilling (using ecoregion mean) seem like it will be higher for region A or B? What are the potential concerns of gapfilling? Can you think of any ways to improve this approach to gapfilling?