---
title: 'OHI 2021: Social Progress Index'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---


# Summary
This script gapfills the Social Progress Index (SPI) data and formats it for the OHI global assessment.

# Updates
Now 11 years of data included in SPI. Created updated spi_categories.csv in Mazu. See the 2021 Methodology Report cited below for a detailed description of the SPI changes from 2020 to 2021. This is saved in Mazu and can also be downloaded [here](https://www.socialprogress.org/index/global/methodology).

***

**Citation**: http://www.socialprogress.org/

Stern, S., Harmacek, J., Htitich, M., & Krylova, P.: 2021 Social Progress Index
Methodology Summary. Social Progress Imperative. Washington, DC.

**Source information**: 
https://www.socialprogress.org/download
https://www.socialprogress.org/index/global/methodology

**Date Downloaded**: 11/02/2021

**Time range**: 2011-2021

**Native data resolution**: country scores

**Format**:  Excel file

**Description**: Social Progress Index scores and components for countries.

```{r setup, warning=FALSE, message = FALSE, eval = FALSE}
# load libraries, set directories
library(ohicore)  #devtools::install_github('ohi-science/ohicore@dev')
library(dplyr)
library(stringr)
library(tidyr)
library(Hmisc)
library(here)
library(validate)
library(tidyverse)
library(readxl)

## comment out when knitting
setwd(here::here("globalprep","prs_res_spi","v2021"))


### Load FAO-specific user-defined functions
source('../../../workflow/R/common.R') # directory locations

set.seed(227)

```

# Social Progress Index data

Organize data and gapfill missing countries that have incomplete data.  This index is comprised of 3 indicators, which are each comprised of 4 subindicators, which are comprised of several datasets.  If one of the subindicators are missing, the SPI is not calculated.  The first round of gapfilling involves using relationships between the the subindicators to gapfill missing data. If a region is missing all subindicator data, then a second round of gapfilling is performed using relationships between UN geopolitical regions and the World Governance Indicator to gapfill the SPI score.

The following gets all years of data.


```{r, eval = FALSE}
cats <- read.csv(file.path(dir_M, 'git-annex/globalprep/_raw_data/SocialProgressIndex/d2020/spi_categories.csv')) %>%
  mutate(subcategory = gsub(" ", "", subcategory))

spi_raw <- read_excel(file.path(dir_M, "git-annex/globalprep/_raw_data/SocialProgressIndex/d2021/SPI2011-2021-dataset.xlsx"), sheet = 2)

names(spi_raw) <- gsub(" ", "", names(spi_raw))

all_spi <- spi_raw %>%
  dplyr::select(-`SPIRank`, -`SPIcountrycode`, -Status) %>%
  dplyr::filter(Country != "World") %>%
  gather('subcategory', 'score', -Country, -`SPI\r\nyear`) %>%
  dplyr::filter(subcategory %in% cats$subcategory) %>%
  left_join(cats, by = "subcategory") %>%
  rename("year" = "SPI\r\nyear")
```


## Gapfilling: Step 1
In this case, we use relationships between the subindicators to estimate missing data.


### Gapfill Basic Human Need (bhn) indicator 

```{r, eval = FALSE}

set.seed(227)
bhn_subs <- all_spi %>%
  dplyr::filter(category %in% c("bhn")) %>%
  spread(subcategory, score)

bhn_tmp <- all_spi %>%
  dplyr::filter(category %in% c("bhn_score")) %>%
  select(Country, year, bhn_score=score) %>%
  left_join(bhn_subs, by=c("Country", "year"))


bhn_tmp <- bhn_tmp %>%
  rowwise() %>%
  mutate(NA_tot = sum(is.na(NutritionandBasicMedicalCare), is.na(PersonalSafety), is.na(Shelter),
             is.na(WaterandSanitation)))

## Ideally most NA_tot values are 0
hist(bhn_tmp$NA_tot)
table(bhn_tmp$NA_tot)


imputes <- 50
bhn_gf <- aregImpute(~ WaterandSanitation + Shelter + NutritionandBasicMedicalCare + PersonalSafety, 
                         data = bhn_tmp, type = "regression", n.impute = imputes)


bhn_gf

# to get mean and sd of all imputations
impute_scores_all <- data.frame()

for (imp in 1:imputes){ 
   #imp = 1
  
  imputed <- impute.transcan(bhn_gf, imputation=imp, data=bhn_tmp, list.out=TRUE,
                           pr=FALSE, check=FALSE)
  
      subcat_data <- data.frame(imputed)
  
      impute_scores <- data.frame(Country = bhn_tmp$Country, 
                                  year = bhn_tmp$year,
                                  imputation = imp)
      
      impute_scores <- cbind(impute_scores, subcat_data)
    
      impute_scores_all <- rbind(impute_scores_all, impute_scores)
    }

## Convert class of value columns from `impute` to `numeric` to avoid "Warning message: attributes are not identical across measure variables; they will be dropped" when gathering columns
impute_scores_all$WaterandSanitation <- as.numeric(impute_scores_all$WaterandSanitation)
impute_scores_all$Shelter <- as.numeric(impute_scores_all$Shelter)
impute_scores_all$NutritionandBasicMedicalCare <- as.numeric(impute_scores_all$NutritionandBasicMedicalCare)
impute_scores_all$PersonalSafety <- as.numeric(impute_scores_all$PersonalSafety)

impute_scores_summary <- impute_scores_all %>%
  gather("subcategory", "score", -(1:3)) %>%
  group_by(Country, year, subcategory) %>%
  dplyr::summarize(score_predict = mean(score),
            sd_score_predict = sd(score)) %>% 
  ungroup()

bhn_tmp_long <- bhn_tmp %>%
  select(Country, year, bhn_score, NA_tot, NutritionandBasicMedicalCare, 
         PersonalSafety, Shelter, WaterandSanitation) %>%
  gather("subcategory", "score", -(1:4)) %>%
  left_join(impute_scores_summary, by=c("Country", "year", "subcategory"))

## Should be no NAs in column sd_score_predict or score_predict
summary(bhn_tmp_long)

bhn_data <- bhn_tmp_long %>%
  mutate(score = ifelse(is.na(score) & NA_tot < 4, score_predict, score)) %>%
  group_by(Country, year) %>%
  dplyr::summarize( 
            score = mean(score),
            NA_tot = mean(NA_tot)) %>%
  ungroup() %>% 
  mutate(category = "bhn")
summary(bhn_data)

```

### Gapfill Opportunity (op) indicator 

```{r, eval = FALSE}

set.seed(227)
op_subs <- all_spi %>%
  dplyr::filter(category %in% c("op")) %>%
  spread(subcategory, score)

op_tmp <- all_spi %>%
  dplyr::filter(category %in% c("op_score")) %>%
  select(Country, year, op_score=score) %>%
  left_join(op_subs, by=c("Country", "year"))


op_tmp <- op_tmp %>%
  rowwise() %>%
  mutate(NA_tot = sum(is.na(AccesstoAdvancedEducation), is.na(PersonalFreedomandChoice), is.na(PersonalRights),
             is.na(Inclusiveness)))

## Ideally most NA_tot values are 0
hist(op_tmp$NA_tot)
table(op_tmp$NA_tot)

#md.pattern(select(op_tmp, -(1:4))) # mice package

imputes <- 50
op_gf <- aregImpute(~ AccesstoAdvancedEducation + PersonalFreedomandChoice + PersonalRights + Inclusiveness, 
                         data = op_tmp, type = "regression", n.impute = imputes)


op_gf

# to get mean and sd of all imputations
impute_scores_all <- data.frame()

for (imp in 1:imputes){ 
  #imp = 1
  
  ## The imputed score values from 0 to 100
  imputed <- impute.transcan(op_gf, imputation=imp, data=op_tmp, list.out=TRUE,
                           pr=FALSE, check=FALSE)
  
      subcat_data <- data.frame(imputed)
  
      impute_scores <- data.frame(Country = op_tmp$Country, 
                                  year = op_tmp$year,
                                  imputation = imp)
      
      impute_scores <- cbind(impute_scores, subcat_data)
    
      impute_scores_all <- rbind(impute_scores_all, impute_scores)
    }

## Convert columns from impute to numeric just to be safe and to prevent warning message from gathering
impute_scores_all$AccesstoAdvancedEducation <- as.numeric(impute_scores_all$AccesstoAdvancedEducation)
impute_scores_all$PersonalFreedomandChoice <- as.numeric(impute_scores_all$PersonalFreedomandChoice)
impute_scores_all$PersonalRights <- as.numeric(impute_scores_all$PersonalRights)
impute_scores_all$Inclusiveness <- as.numeric(impute_scores_all$Inclusiveness)

impute_scores_summary <- impute_scores_all %>%
  gather("subcategory", "score", -(1:3)) %>%
  group_by(Country, year, subcategory) %>%
  dplyr::summarize(score_predict = mean(score),
            sd_score_predict = sd(score)) %>% 
  ungroup()


op_tmp_long <- op_tmp %>%
  select(Country, year, op_score, NA_tot, AccesstoAdvancedEducation, 
         PersonalFreedomandChoice, PersonalRights, Inclusiveness) %>%
  gather("subcategory", "score", -(1:4)) %>%
  left_join(impute_scores_summary, by=c("Country", "year", "subcategory"))


## should have no NAs in score_predict and sd_score_predict
summary(op_tmp_long)

op_data <- op_tmp_long %>%
  mutate(score = ifelse(is.na(score) & NA_tot < 4, score_predict, score)) %>%
  group_by(Country, year) %>%
  dplyr::summarize( #op_score_old = mean(op_score), # used this to test to make sure all is well, try `check_validate( )` below
            score = mean(score),
            NA_tot = mean(NA_tot)) %>%
  ungroup() %>% 
  mutate(category = "op") 

summary(op_data)
```


### Gapfill Foundations of Wellbeing (fw) indicator 

```{r, eval = FALSE}

set.seed(227)
fw_subs <- all_spi %>%
  dplyr::filter(category %in% c("fw")) %>%
  spread(subcategory, score)

fw_tmp <- all_spi %>%
  dplyr::filter(category %in% c("fw_score")) %>%
  select(Country, year, fw_score=score) %>%
  left_join(fw_subs, by=c("Country", "year"))


fw_tmp <- fw_tmp %>%
  rowwise() %>%
  mutate(NA_tot = sum(is.na(AccesstoBasicKnowledge), is.na(AccesstoInformationandCommunications), is.na(EnvironmentalQuality),is.na(HealthandWellness)))

## most values should be 0
hist(fw_tmp$NA_tot)
table(fw_tmp$NA_tot)

#md.pattern(select(fw_tmp, -(1:4))) # mice package

imputes <- 50
fw_gf <- aregImpute(~ AccesstoBasicKnowledge + AccesstoInformationandCommunications + EnvironmentalQuality + HealthandWellness, 
                         data = fw_tmp, type = "regression", n.impute = imputes)


fw_gf

# to get mean and sd of all imputations
impute_scores_all <- data.frame()

for (imp in 1:imputes){ 
  #imp = 1
  
  imputed <- impute.transcan(fw_gf, imputation=imp, data=fw_tmp, list.out=TRUE,
                           pr=FALSE, check=FALSE)
  
      subcat_data <- data.frame(imputed)
  
      impute_scores <- data.frame(Country = fw_tmp$Country, 
                                  year = fw_tmp$year,
                                  imputation = imp)
      
      impute_scores <- cbind(impute_scores, subcat_data)
    
      impute_scores_all <- rbind(impute_scores_all, impute_scores)
    }

## Convert columns from impute to numeric just to be safe and to prevent warning message from gathering. Due to some impute values having *
impute_scores_all$AccesstoBasicKnowledge <- as.numeric(impute_scores_all$AccesstoBasicKnowledge)
impute_scores_all$AccesstoInformationandCommunications <- as.numeric(impute_scores_all$AccesstoInformationandCommunications)
impute_scores_all$EnvironmentalQuality <- as.numeric(impute_scores_all$EnvironmentalQuality)
impute_scores_all$HealthandWellness <- as.numeric(impute_scores_all$HealthandWellness)

impute_scores_summary <- impute_scores_all %>%
  gather("subcategory", "score", -(1:3)) %>%
  group_by(Country, year, subcategory) %>%
  dplyr::summarize(score_predict = mean(score),
            sd_score_predict = sd(score)) %>% 
  ungroup()


fw_tmp_long <- fw_tmp %>%
  select(Country, year, fw_score, NA_tot, AccesstoBasicKnowledge, 
         AccesstoInformationandCommunications, EnvironmentalQuality, HealthandWellness) %>%
  gather("subcategory", "score", -(1:4)) %>%
  left_join(impute_scores_summary, by=c("Country", "year", "subcategory"))


summary(fw_tmp_long)

fw_data <- fw_tmp_long %>%
  mutate(score = ifelse(is.na(score) & NA_tot < 4, score_predict, score)) %>%
  group_by(Country, year) %>%
  dplyr::summarize( #fw_score_old = mean(fw_score), # used this to test to make sure all is well, try `check_that( )` below
            score = mean(score),
            NA_tot = mean(NA_tot)) %>%
  ungroup() %>% 
  mutate(category = "fw") 

```



## Combine data to get Social Progress Index

The next step averages the 3 indicators and identifies gapfilling. All the rows where all four indicators are missing `NA_tot = 4` will have an NA in the `score` column.

```{r, eval = FALSE}

spi_calc <- rbind(bhn_data, op_data, fw_data) %>%
  mutate(gapfill = paste(category, NA_tot, sep = "_")) %>%
  mutate(score = ifelse(score > 100, 100, score),
         score = ifelse(score < 0, 0, score)) %>%
  group_by(Country, year) %>%
  dplyr::summarize(
            score = mean(score),
            method = paste(unique(gapfill), collapse=", "),
            gapfill = sum(NA_tot)) %>%
  ungroup() %>% 
  mutate(method = ifelse(gapfill == 0, NA, method),
         gapfill = ifelse(gapfill >= 1, 1, 0)) %>% # change values to just yes or no
  data.frame()

summary(spi_calc)
## Some scores might still be NA because no values exist for that country

# check:
check <- all_spi %>%
  filter(subcategory=="SocialProgressIndex") %>%
  dplyr::select(Country, year, true_score=score) %>%
  left_join(spi_calc, by = c("Country", "year"))

## Perfect
plot(check$true_score, check$score)
abline(0,1, col="red")


write.csv(spi_calc, "int/Country_spi.csv", row.names=FALSE)


other_vars <-  rbind(bhn_data, op_data, fw_data) %>%
  mutate(gapfill = paste(category, NA_tot, sep = "_")) %>%
  mutate(score = ifelse(score > 100, 100, score),
         score = ifelse(score < 0, 0, score)) %>%
  mutate(method = gapfill) %>%
  dplyr::select(Country, year, score, method, category)


all_together <- spi_calc %>%
  mutate(category = "spi") %>%
  dplyr::select(Country, year, score, method, category) %>%
  rbind(other_vars)

write.csv()
```

## Assign countries to OHI regions

```{r, eval = FALSE}

spi_calc <- read.csv("int/Country_spi.csv", stringsAsFactors = FALSE)

spi_calc <- spi_calc %>%
  mutate(Country = case_when(
   Country=="Côte d'Ivoire" ~ "Ivory Coast",
    Country=="R<e9>union" ~ "Reunion",
    Country=="Cura<e7>ao" ~ "Curacao",
    Country=="Korea, Democratic Republic of" ~ "North Korea",
    Country=="St. Helena" ~ "Saint Helena",
    Country=="St. Martin" ~ "Northern Saint-Martin",
    TRUE ~ Country # Everything else, leave it be
    ))

# Channel Islands are Jersey and Guernsey, but these are already in the data

spi_rgn <- name_2_rgn(df_in = spi_calc, 
                       fld_name='Country',
                      flds_unique = c("year"))

# Weight the following duplicates by population
# China, Guadeloupe, Guam, Hong Kong, Macao, Martinique, Northern Mariana Islands, Puerto Rico, Virgin Islands (U.S.)... Populations taken from google

pop_weights <- data.frame(Country = c("China", "Hong Kong", "Macao",
                 "Guadeloupe", "Martinique",
                 "Guam", "Northern Mariana Islands",
                 "Puerto Rico", "Virgin Islands (U.S.)"),
                 pop = c(1446751367, 7690000, 661103,
                         400215, 374878,
                         170744, 58016,
                         2761319, 104219))
spi_rgn <- spi_rgn %>%
  left_join(pop_weights, by = "Country") %>%
  mutate(pop = ifelse(is.na(pop), 1, pop)) %>%
  group_by(rgn_id, rgn_name, year) %>%
  dplyr::summarize(score = weighted.mean(score, pop, na.rm=TRUE),
            method = paste(unique(method), collapse=" "),
            gapfill = weighted.mean(gapfill, pop, na.rm=TRUE))%>%
  ungroup() %>% 
  dplyr::mutate(year = as.numeric(year))


```

# Compare to WGI

WGI is a couple years behind the SPI, so we will use 2018 WGI for the 2019, 2020, and 2021 WGI There is a strong correlation betwen the WGI and SPI indicators.
```{r, eval = FALSE}

wgi <- read.csv('../../prs_res_wgi/v2020/output/wgi_res.csv') %>%
  select(rgn_id, year, wgi_score = resilience_score)

length(unique(wgi$rgn_id))

wgi_2019 <- wgi %>%
  filter(year == 2018) %>%
  mutate(year = 2019)

wgi_2020 <- wgi %>%
  filter(year == 2018) %>%
  mutate(year = 2020)

wgi_2021 <- wgi %>%
  filter(year == 2018) %>%
  mutate(year = 2021)

wgi <- rbind(wgi, wgi_2019, wgi_2020, wgi_2021)

wgi_spi <- wgi %>%
  left_join(spi_rgn, by=c("rgn_id", "year"))

plot(wgi_spi$wgi_score*100, wgi_spi$score, ylab="Social Progress Index, score", xlab="Worldwide Governance Score")
abline(0,1, col="red")

mod <- lm(score ~ wgi_score, data=wgi_spi)
summary(mod)

```

# Second round of gapfilling 
In this case, UN geopolitical regions and WGI scores are used to estimate regions with no data.  Based on this analysis, a model that includes WGI data and r2 UN geopolitical regions is the best model to predict missing SPI values. 

```{r, eval = FALSE}

years <- data.frame(year = min(spi_rgn$year):max(spi_rgn$year))

rgns_gf_un <- georegions %>%
  merge(years) %>%
  left_join(spi_rgn, by = c("rgn_id","year")) %>%
  left_join(wgi, by = c("rgn_id", "year")) %>%
  mutate(r2 = as.factor(r2)) %>%
  mutate(r1 = as.factor(r1))


## Compare models to select a gapfilling method
mod1 <- lm(score ~ r2, data = rgns_gf_un, na.action="na.exclude")
mod2 <- lm(score ~ r2 + wgi_score, data = rgns_gf_un, na.action="na.exclude")
mod3 <- lm(score ~ r1, data = rgns_gf_un, na.action="na.exclude")
mod4 <- lm(score ~ r1 + wgi_score, data = rgns_gf_un, na.action="na.exclude")
mod5 <- lm(score ~ wgi_score, data = rgns_gf_un, na.action="na.exclude")


## the lowest AIC score is likely the best model
AIC(mod1, mod2, mod3, mod4, mod5)

## plot the models with the two lowest AIC scores
plot(predict(mod2), rgns_gf_un$score)
abline(0,1)


## Estimate missing data and gapfill
## need to add this because some R2 regions have no data
r2_regions <- unique(rgns_gf_un$r2[!is.na(rgns_gf_un$score)])
rgns_gf_un$r2 <- ifelse(rgns_gf_un$r2 %in% r2_regions, rgns_gf_un$r2, NA)

r1_regions <- unique(rgns_gf_un$r1[!is.na(rgns_gf_un$score)])
rgns_gf_un$r1 <- ifelse(rgns_gf_un$r1 %in% r1_regions, rgns_gf_un$r1, NA)


## Predict scores using r2 and wgi scores
mod_gf_r2 <- lm(score ~ r2 + wgi_score, data = rgns_gf_un, na.action = "na.exclude")
rgns_gf_un$score_pred_r2 <- predict(mod_gf_r2, newdata = rgns_gf_un[, c("r2", "wgi_score")])
## Predict scores using r1 and wgi scores
mod_gf_r1 <- lm(score ~ r1 + wgi_score, data = rgns_gf_un, na.action = na.exclude)
rgns_gf_un$score_pred_r1 <- predict(mod_gf_r1, newdata = rgns_gf_un[, c("r1", "wgi_score")])
## Predict scores just using wgi scores
mod_gf_wgi <- lm(score ~ wgi_score, data = rgns_gf_un, na.action = na.exclude)
rgns_gf_un$score_pred_wgi <- predict(mod_gf_wgi, newdata = data.frame(wgi_score = rgns_gf_un$wgi_score))

## Record gapfill methods
## Combine scores with predicted model scores
rgns_gf <- rgns_gf_un %>%
  mutate(method = ifelse(is.na(score), "UN georgn & WGI", method)) %>%
  mutate(gapfill = ifelse(is.na(score), "1", gapfill)) %>%
  mutate(score = ifelse(is.na(score), score_pred_r2, score)) %>%
  mutate(score = ifelse(is.na(score), score_pred_r1, score)) %>%
  mutate(score = ifelse(is.na(score), score_pred_wgi, score)) %>%
  select(rgn_id, year, score, method, gapfill) %>%
  mutate(score = ifelse(score > 100, 100, score),
         score = ifelse(score < 0, 0, score))

summary(rgns_gf)  # should be no NA values
length(unique(rgns_gf$rgn_id)) # should be 220 regions

```

# Uninhabited regions

These regions will receive an NA for their score (when established population is < 3000 people). 

```{r, eval = FALSE}
low_pop()
low_pop <- low_pop %>%
  filter(est_population < 3000 | is.na(est_population)) 

rgns_gf_uninhab <- rgns_gf %>%
  mutate(score = ifelse(rgn_id %in% low_pop$rgn_id, NA, score)) %>%
  mutate(gapfill = ifelse(rgn_id %in% low_pop$rgn_id, NA, gapfill))

```

# Save final data

```{r, eval = FALSE}
#gf_2019 <- read_csv(file.path("../v2019/output/spi_gf.csv"))


gf_data <- rgns_gf_uninhab %>%
  select(rgn_id, year, gapfill, method) %>%
  mutate(gapfill = ifelse(is.na(method), 0, 1))
write.csv(gf_data, "output/spi_gf.csv", row.names=FALSE)


res_data <- rgns_gf_uninhab %>%
  select(rgn_id, year, resilience_score=score) %>%
  mutate(resilience_score = resilience_score/100)
write.csv(res_data, "output/spi_res.csv", row.names=FALSE)

prs_data <- rgns_gf_uninhab %>%
  select(rgn_id, year, pressure_score=score) %>%
  mutate(pressure_score = 1 - (pressure_score/100))
write.csv(prs_data, "output/spi_prs.csv", row.names=FALSE)

```

# Data check 

```{r, eval = FALSE}
## Resilience comparison, filtered for 2020
new_spi_res <- read_csv(file.path("output/spi_res.csv")) %>%
  filter(year == 2020) %>%
  select(rgn_id, new_resilience_score = resilience_score)

old_new <- read_csv("../v2020/output/spi_res.csv") %>%
  filter(year == 2020) %>% 
  left_join(new_spi_res, by = 'rgn_id')

plot(old_new$new_resilience_score, old_new$resilience_score)
abline(0, 1, col="red")

## Pressure score comparison, filtered for 2020
new_spi_prs <- read_csv(file.path("output/spi_prs.csv")) %>%
  filter(year == 2020) %>%
  select(rgn_id, new_pressure_score = pressure_score)

old_new <- read_csv("../v2020/output/spi_prs.csv") %>%
  filter(year == 2020) %>% 
  left_join(new_spi_prs, by = 'rgn_id')

plot(old_new$new_pressure_score, old_new$pressure_score)
abline(0, 1, col="red")


## Pressure score comparison, All years 
new_spi_prs <- read_csv(file.path("output/spi_prs.csv")) %>%
  #filter(year == 2018) %>%
  select(rgn_id, new_pressure_score = pressure_score)

old_new <- read_csv("../v2019/output/spi_prs.csv") %>%
  #filter(year == 2018) %>% 
  left_join(new_spi_prs, by = 'rgn_id')

plot(old_new$new_pressure_score, old_new$pressure_score)
abline(0, 1, col="red")
```


***

