---
title: 'OHI 2021: LSP, Rasterize WDPA polygons'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
---

# Summary

* Setup WDPA shapefile for lasting special places by filtering out non-"Designated" protected areas, and filtering out "non-MPA programmatic management plans" (US protected areas that distort scores).
* Rasterize the result using `fasterize::fasterize()` (not `gdal_rast2()` as in previous assessments) and save to disk using `writeRasterBlocks()` defined in this script.

View lsp_data_prep.Rmd for full description of Lasting Special Places data preparation.

# Updates from previous assessment

2021
Changed data source to 2021
The source data has now been split into 3 different files, so we will need to merge all three shapefiles together, before we can work with the data. 

We updated the script to save 3 separate files for the reordering and transforming, for ease of use and reproducibility. 

2020
Changed data source to 2020

2019:
Changing all of the readOGR to st_read so as to treat the shapefiles as simple features.
Changed spTranform to st_transform. 
Changed writeOGR to st_write. 
All of the above changes significantly reduce run time. 

Deleted the following code from LSP data prep:
wdpa_diff <- st_difference(wdpa_sf[1], wdpa_sf_18[1])

The code was included as a data check. It subtracts the old WDPA shapefile from the new WDPA shapefile to visualize the change in polygons between last year and this year. Overall, the resulting shapefile would be relatively sparse because there shouldn't be too many changes.
However, this was taking forever to run (>24 hours). It is a nice check, but not critical.

***

# Setup

```{r, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

```{r setup, echo = FALSE, message = TRUE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = FALSE, message = FALSE, warning = FALSE, eval=FALSE)
library(rgdal)
library(gdalUtils)
library(tidyverse)
library(here)
library(sf)
library(mapview)
library(fasterize)

source('http://ohi-science.org/ohiprep_v2020/workflow/R/common.R')

goal     <- 'globalprep/lsp'
scenario <- 'v2021'
dir_goal_anx   <- file.path(dir_M, 'git-annex', goal, scenario) 
dir_goal  <- file.path(here('globalprep', goal, scenario))
dir_data <- file.path(dir_M, 'git-annex/globalprep', '_raw_data',
                         'wdpa_mpa/d2021/WDPA_WDOECM_wdpa_shp')


```


# Methods

## Filter WDPA Shapefile 

Read in the polygons from the WDPA dataset; filter as needed.

``` {r filter_and_reorder_poly}

shp_raw_0     <- file.path(dir_data, 'WDPA_WDOECM_wdpa_shp0', 'WDPA_WDOECM_wdpa_shp-polygons')
shp_raw_1 <- file.path(dir_data, 'WDPA_WDOECM_wdpa_shp1', 'WDPA_WDOECM_wdpa_shp-polygons')
shp_raw_2 <- file.path(dir_data, 'WDPA_WDOECM_wdpa_shp2', 'WDPA_WDOECM_wdpa_shp-polygons')
shp_reorder_0 <- file.path(dir_data, 'shps', 'WDPA_Feb2021_shp_ordered_0') ## replace month and year with the appropriate month and year
shp_reorder_1 <- file.path(dir_data, 'shps', 'WDPA_Feb2021_shp_ordered_1') ## replace month and year with the appropriate month and year
shp_reorder_2 <- file.path(dir_data, 'shps', 'WDPA_Feb2021_shp_ordered_2') ## replace month and year with the appropriate month and year

### 2021 Note: We couldn't get the shapefile to write to mazu, so we did not write the shapefile. In future iterations, try to write it, just for reproducibility and data checking reasons. 


  ## Read in the raw shapefiles

  wdpa_poly_0 <- st_read(dsn = dirname(shp_raw_0), 
                       layer = basename(shp_raw_0),
                       stringsAsFactors = FALSE)

  
  wdpa_poly_1 <- st_read(dsn = dirname(shp_raw_1),
                       layer = basename(shp_raw_1),
                       stringsAsFactors = FALSE)

  
  wdpa_poly_2 <- st_read(dsn = dirname(shp_raw_2), 
                       layer = basename(shp_raw_2),
                       stringsAsFactors = FALSE)
 
  
  wdpa_list <- list(wdpa_poly_0, wdpa_poly_1, wdpa_poly_2)
  
  ## filter polygons
  # glimpse(wdpa_poly_0$STATUS_YR) ## check class of status_yr: checked, integer
  
## check some things:
    wdpa_poly_0$STATUS %>% unique()
  ## [1] "Designated"   "Inscribed"    "Proposed"     "Not Reported" "Established"  "Adopted"   


  x <- wdpa_poly_0 %>% filter(str_detect(tolower(MANG_PLAN), 'non-mpa program')) # 0 - good - same thing for the other two 
  
myFunc <- function(DF) {

  DF <- DF %>%
    setNames(tolower(names(.))) %>% #improve?
    dplyr::select(wdpaid, name, orig_name, 
           desig, desig_eng, desig_type,
           iucn_cat, 
           marine, no_take, no_tk_area, 
           status, status_yr, 
           mang_auth, mang_plan, verif,
           sub_loc, parent_iso, iso3) %>%
    dplyr::mutate(status_yr = as.integer(status_yr))
  
  DF <- DF[DF$status == 'Designated', ]
  DF <- DF[!str_detect(tolower(DF$mang_plan), 'non-mpa program'), ]
  
return(DF)
}  
  
wdpa_poly_list <- lapply(wdpa_list, myFunc)

## check to see if it worked
test <- wdpa_poly_list[[1]] ## it did.. only 19 columns and little less observations 
  

## now we need to unlist them, and write them to the appropriate folder
wdpa_poly_fix_0 <- wdpa_poly_list[[1]]  

wdpa_poly_fix_1 <- wdpa_poly_list[[2]]  

wdpa_poly_fix_2 <- wdpa_poly_list[[3]]  
  
  ### write the shapefile out now... didn't work for 2021, but try for 2022.
  suppressWarnings(st_write(wdpa_poly_fix_0,
           dsn = dirname(shp_reorder_0), layer = basename(shp_reorder_0),
           driver = 'ESRI Shapefile'))
  ## lots of warnings similar to (but i suppressed them to save time):
  ##   "Warning 1: Value 555593627 of field WDPAID of feature 507 not successfully written."
  ## warning ignored: WDPAID field not used in analysis
  
  suppressWarnings(st_write(wdpa_poly_fix_1,
           dsn = dirname(shp_reorder_1), layer = basename(shp_reorder_1),
           driver = 'ESRI Shapefile'))
  
    
  suppressWarnings(st_write(wdpa_poly_fix_2,
           dsn = dirname(shp_reorder_2), layer = basename(shp_reorder_2),
           driver = 'ESRI Shapefile'))
  
  rm('wdpa_poly_list') ### clean up memory
  

```

## Transform to Mollweide Projection

Transform ordered polygons to Mollweide and save as new polygons.

``` {r transform_poly}

## 2021: This also wouldnt save to mazu for some reason... try again next year. 
shp_xformed_0 <- file.path(dir_data, 'shps', 'WDPA_Feb2021_shp_xformed_0') 
shp_xformed_1 <- file.path(dir_data, 'shps', 'WDPA_Feb2021_shp_xformed_1') 
shp_xformed_2 <- file.path(dir_data, 'shps', 'WDPA_Feb2021_shp_xformed_2') 
  

    
    message('loading wdpa polygons (filtered and ordered)')
    shp_reorder_0 <- file.path(dir_data, 'shps', 'WDPA_Feb2021_shp_ordered_0')
    shp_reorder_1 <- file.path(dir_data, 'shps', 'WDPA_Feb2021_shp_ordered_1')
    shp_reorder_2 <- file.path(dir_data, 'shps', 'WDPA_Feb2021_shp_ordered_2')
    
    wdpa_poly_0 <- st_read(dsn = dirname(shp_reorder_0), 
                          layer = basename(shp_reorder_0),
                          stringsAsFactors = FALSE)
    wdpa_poly_1 <- st_read(dsn = dirname(shp_reorder_1), 
                          layer = basename(shp_reorder_1),
                          stringsAsFactors = FALSE)
    wdpa_poly_2 <- st_read(dsn = dirname(shp_reorder_2), 
                          layer = basename(shp_reorder_2),
                          stringsAsFactors = FALSE)

wdpa_list <- list(wdpa_poly_0, wdpa_poly_1, wdpa_poly_2)    
    
myFunc2 <- function(DF) {

  message('Spatial transforming WDPA polygons to Mollweide')
  #crs_mol <- CRS('+proj=moll +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs')

  DF <- st_transform(DF, crs = '+proj=moll +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs')

return(DF)
}  

wdpa_poly_list <- lapply(wdpa_list, myFunc2)
    
## lets see if it worked
wdpa_poly_fix_0 <- wdpa_poly_list[[1]]  
st_crs(wdpa_poly_fix_0) ## it worked

wdpa_poly_fix_1 <- wdpa_poly_list[[2]]  
wdpa_poly_fix_2 <- wdpa_poly_list[[3]]  
  

## Lets write the transformed files to Mazu

  st_write(wdpa_poly_fix_0, dsn = dirname(shp_xformed_0), layer = basename(shp_xformed_0),
           driver = 'ESRI Shapefile', update = TRUE)
  
  st_write(wdpa_poly_fix_1, dsn = dirname(shp_xformed_1), layer = basename(shp_xformed_1),
           driver = 'ESRI Shapefile', update = TRUE)
    
  st_write(wdpa_poly_fix_2, dsn = dirname(shp_xformed_2), layer = basename(shp_xformed_2),
           driver = 'ESRI Shapefile', update = TRUE)
  
  ## This will likely take ~7 mins per file
  ## will receive A LOT of warnings, disregard these as you did above. You can look at the file size on mazu to check if it is done writing. Warnings will continue for awhile after it has completed writing to mazu. 
  

```

### Define some raster functions.

Here we define `gdal_rast2()` function as a wrapper for `gdalUtils::gdal_rasterize()`, which has some clunky arguments.  This was originally written for OHIBC.

Here we also define `writeRasterBlocks()` to write out very large rasters chunk-by-chunk to avoid an issue in which `writeRaster()` fails due to inability to work with long vectors (over 2 billion elements?).

``` {r define_gdal_rast2}
## from OHIBC...


writeRasterBlocks <- function(rast, filename) {
  bs <- blockSize(rast)
  rast <- writeStart(rast, filename, overwrite = TRUE)
  ptm <- proc.time()
  for (i in 1:bs$n) { # i <- 1
    message('... writing block ', i, ' of ', bs$n, ' to ', basename(filename), 
        ' (', (proc.time() - ptm)[3], ' sec)')
    v <- getValues(rast, row = bs$row[i], nrows = bs$nrows[i] )
    rast <- writeValues(rast, v, bs$row[i])
  }
  rast <- writeStop(rast)
   #return(rast)
}



```


## Shapefile to Raster: `fasterize()`

`fasterize()` from the `fasterize` package takes advantage of Simple Features objects from the `sf` package, rather than objects from the `sp` package.  It is considerably faster; it returned a completed raster in ten minutes.  However, saving the very large (18GB) resulting raster proved problematic.  The `writeRasterBlocks()` function defined above helped get around that problem though still took over an hour to write the raster to disk.

``` {r fasterize_wdpa, eval = FALSE}

library(sf)
library(fasterize) ### devtools::install_github('ecohealthalliance/fasterize')
rast_wdpa_file   <- file.path(dir_goal_anx, 'rast', 'wdpa_2021_moll_500m.tif') 

if(!file.exists(rast_wdpa_file)) {
shp_xformed_file_0 <- file.path(dir_data, 'shps', 'WDPA_Feb2021_shp_xformed_0') 
shp_xformed_file_1 <- file.path(dir_data, 'shps', 'WDPA_Feb2021_shp_xformed_1') 
shp_xformed_file_2 <- file.path(dir_data, 'shps', 'WDPA_Feb2021_shp_xformed_2') 
  
  ptm <- proc.time()
  wdpa_poly_0 <- sf::st_read(dsn = dirname(shp_xformed_file_0),
                           layer = basename(shp_xformed_file_0))
  
  wdpa_poly_1 <- sf::st_read(dsn = dirname(shp_xformed_file_1),
                           layer = basename(shp_xformed_file_1))
  
  wdpa_poly_2 <- sf::st_read(dsn = dirname(shp_xformed_file_2),
                           layer = basename(shp_xformed_file_2))
  
  wdpa_poly_all <- rbind(wdpa_poly_0, wdpa_poly_1, wdpa_poly_2)
  
  cat('elapsed: ', (proc.time() - ptm)[3])
  ### ~1 min to read in each file

  rast_base <- raster::raster(file.path(dir_M, 'git-annex/globalprep/spatial/d2014',
                                'data/rgn_mol_raster_500m',
                                'rgn_inland1km_mol_500mcell.tif'))
  
  # test_poly <- sf::st_read(path.expand('~/github/ohibc/prep/spatial'),
  #                             'ohibc_offshore_3nm',
  #                             stringsAsFactors = FALSE) %>%
  #   st_transform(rast_base@crs@projargs)
  # poly_ext <- raster::extent(as.vector(st_bbox(test_poly)))
  # 
  # rast_base <- raster::crop(rast_base, poly_ext)
  # test_poly <- wdpa_poly[1:100, ]
  
  ## For any overlapping polygons, assign the oldest (minimum) status year to that area
  ptm <- proc.time()
  rast_wdpa <- fasterize(wdpa_poly_all, rast_base, field = 'status_yr', fun = 'min')
  cat('fasterize elapsed: ', (proc.time() - ptm)[3]) 
  ### 358 - 403 seconds?! holy crap, if it rasterized properly...
  #2019 fasterize elapsed 26 seconds... is this right? 
  # 2021 45.006 seconds! 
  
  # ptm <- proc.time()
  # raster::writeRaster(rast_wdpa, rast_wdpa_file)
  # cat('writeRaster elapsed: ', (proc.time() - ptm)[3])
  ### writeRaster failed on an issue with large vectors...
  ### see below about writing block by block approach.
  
  library(raster)
  # rast <- rast_wdpa
  # filename <- rast_wdpa_file
  ptm <- proc.time()
  x <- writeRasterBlocks(rast_wdpa, rast_wdpa_file)
  message('writeRaster elapsed: ', (proc.time() - ptm)[3])
  #~10 minutes
}

check <- raster::raster(rast_wdpa_file)
check2 <- raster::raster(file.path("/home/shares/ohi/git-annex/globalprep/lsp/v2020/rast/wdpa_2020_moll_500m.tif"))
## they look almost exactly the same... good.

```

## Data Checking

Compare shapfile 2019 v 2021.

```{r}

library(sf)
library(raster)

## Read in as sf
wdpa_sf_19 <- st_read(dsn = file.path("/home/shares/ohi/git-annex/globalprep/_raw_data/wdpa_mpa/d2019/shps"), layer = "WDPA_Apr2019_shp_ordered") 

shp_reorder_0 <- file.path(dir_data, 'shps', 'WDPA_Feb2021_shp_ordered_0')
shp_reorder_1 <- file.path(dir_data, 'shps', 'WDPA_Feb2021_shp_ordered_1')
shp_reorder_2 <- file.path(dir_data, 'shps', 'WDPA_Feb2021_shp_ordered_2')
    
wdpa_poly_0 <- st_read(dsn = dirname(shp_reorder_0), 
                          layer = basename(shp_reorder_0),
                          stringsAsFactors = FALSE)
wdpa_poly_1 <- st_read(dsn = dirname(shp_reorder_1), 
                          layer = basename(shp_reorder_1),
                          stringsAsFactors = FALSE)
wdpa_poly_2 <- st_read(dsn = dirname(shp_reorder_2), 
                          layer = basename(shp_reorder_2),
                          stringsAsFactors = FALSE)

wdpa_poly_all_21 <- rbind(wdpa_poly_0, wdpa_poly_1, wdpa_poly_2)

plot(wdpa_poly_all_21[1])
plot(wdpa_sf_19[1])


colnames(wdpa_poly_all_21)

unique(wdpa_poly_all_21$orig_name)


test <- wdpa_poly_all_21 %>%
  filter(parent_iso == "GBR") %>%
  filter(iso3 == "PCN")

mapview(test)

st_area(test) # 836075862002 m2 


  rast_base <- raster::raster(file.path(dir_M, 'git-annex/globalprep/spatial/d2014',
                                'data/rgn_mol_raster_500m',
                                'rgn_inland1km_mol_500mcell.tif'))

  rast_wdpa <- fasterize(test, rast_base, field = 'status_yr', fun = 'min')
  
  plot(rast_wdpa)
  
  rast_wdpa

```



