---
title: 'OHI 2021 - Land based nutrient plume exploration '
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../workflow/templates/ohi_hdr.html' 
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

[REFERENCE RMD FILE](http://ohi-science.org/ohiprep_v2020/globalprep/cw_nutrient_o_chem/v2020/plume_output_exploration.html)


# Summary
This document is used to explore the output of the plume modeling done for this data layer.  

## The following data are used:

* 



## Initial set-up code

```{r setup, eval = FALSE}
library(tidyverse)
library(raster)
library(sf)
library(mapview)
library(janitor)
library(here)


source(here('workflow/R/common.R'))

ww_raw_dir <- "/home/shares/ohi/git-annex/land-based/wastewater/data/raw"
ww_intermediate_dir <- "/home/shares/ohi/git-annex/land-based/wastewater/data/interim"
prs_int_dir <- "/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2021/int"
prep <- file.path("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2021")

```

**Take a look at the output from the full plume model**
 - We need to check this to see if the model ran on all of the basin_id
 - It looks like the model did not write rasters for everything. I think this is due to very small effluent values. 
 - Check the plumes that were run, and see if the ones with really small values are included. 
 
```{r, eval = FALSE}
## Count how many rasters the plume model created and compare to the number of pourpoints with >0 N effluent

fils <- list.files("/home/sgclawson/plumes/output1/", pattern="tif$", full.names = FALSE, recursive = TRUE)

files <- data_frame(fils)
 

pourpoint_names_files <- data.frame(fils = sub('.*plume_effluent_', '', files$fils))

pourpoint_names <- data.frame(basin_id = sub('.tif', '', pourpoint_names_files$fils)) %>%
  mutate(basin_id = as.character(basin_id)) %>%
  dplyr::filter(str_detect(basin_id, "subsets", negate = TRUE))


## now read in total shp file

files_shp <- list.files(file.path("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2021/int/pourpoints/"), pattern = "shp", full = TRUE)

ws_pp_zonal_all_fix <- st_read(files_shp[1])   %>%
  mutate(row_num = row_number()) 

test_ws <- ws_pp_zonal_all_fix %>%
  filter(effluent !=0)
sum(test_ws$effluent) # 16217349

excluded_2005 <- c(setdiff(ws_pp_zonal_all_fix$basin_id, pourpoint_names$basin_id))

included_2005 <- c(pourpoint_names$basin_id)

excluded <- paste0("plume_effluent_", excluded_2005)
write.csv(excluded, file.path(here("globalprep/prs_land-based_nutrient/v2021/STEP6_check_plumes/int/excluded_pp.csv")), row.names = FALSE)


check <- ws_pp_zonal_all_fix %>%
  dplyr::filter(basin_id %in% excluded_2005)
sum(check$effluent) # missing 3212062 tonnes N 
mapview(check, cex = "effluent")

check_0 <- check %>%
  dplyr::filter(effluent >0) ## yes, the really small values were excluded... 

# Now lets check where those small pourpoints were located in the shapefile... I bet that one of them is in row number 36522?
check_2 <- ws_pp_zonal_all_fix %>%
  arrange(basin_id) %>%
  mutate(row_num = row_number()) %>%
  filter(effluent >0)

check_1 <- ws_pp_zonal_all_fix %>%
  filter(effluent == 0)


test <- raster("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2021/output/N_plume/pourpoints_crop_manure_leached_volt_N_2005_joined.tif")
cellStats(test, "sum") # 4296690

plot(test)

test2 <- raster::select(test)
plot(test2)

N_effluent_log <- raster::calc(test, fun = function(x){log(x + 1)})
plot(N_effluent_log)


files_shp <- list.files(file.path("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2021/int/pourpoints/"), pattern = "shp", full = TRUE)

tester <- st_read(files_shp[1])
sum(tester$effluent) # 16217349

min(tester$effluent)


testing <- tester %>%
  filter(effluent != 0)


testing <- tester %>% 
  filter(str_detect(basin_id, "ai"))

mapview(testing)


test_0_val <- raster(file.path("/home/sgclawson/plumes/output2/subsets/subset4/plume_effluent_pa_00001.tif"))
plot(test_0_val)
cellStats(test_0_val, "sum")
```

See if we can combine try1 and try2 rasters
```{r}
try1 <- raster(file.path("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2021/output/try1/pourpoints_crop_manure_leached_volt_N_2005_joined.tif"))
try2 <- raster(file.path("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2021/output/N_plume/pourpoints_crop_manure_leached_volt_N_2005_joined.tif"))

plot(log(try2+1))

cellStats(try2, "sum", na.rm = TRUE) # 8576483
cellStats(try1, "sum", na.rm = TRUE) # 4296690

8576483 + 4296690 # 12873173

# 16217349... I think im missing ~ 3 million tonnes ? 

# try3 <- try1 + try2 # i think i will have to mosaic these together... shouldn't be a problem 

test <- st_read("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2021/int/pourpoints/pourpoints_crop_manure_leached_volt_N_2005.shp")

sum(test$effluent) 

## We should check the pourpoints against each other to see what are missing 


plot(log(try2 + 1))

mosaic(try1, try2, fun = "sum", filename = "/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2021/int/test_mosaic/mosaic_2005.tif", progress = "text")

test_mosaic <- raster(file.path("/home/shares/ohi/git-annex/globalprep/prs_land-based_nutrient/v2021/int/test_mosaic/mosaic_2005.tif"))

cellStats(test_mosaic, "sum", na.rm = TRUE) # 12873176 it worked i think? 

test_mosaic[test_mosaic < 0] <- 0


m <- c(0, 0.000000001, NA)
rclmat <- matrix(m)
test_mosaic_reclass <- reclassify(test_mosaic, rclmat, progress = "text")
cellStats(test_mosaic_reclass, "sum", na.rm = TRUE, progress = "text") # 12873176 - it worked

plot(test_mosaic_reclass)

mosaic_calc <- raster::calc(test_mosaic_reclass, fun=function(x){log(x+1)}, progress = "text")
plot(mosaic_calc)

par(mar=c(1,1,1,1))

library(colorspace)
library(RColorBrewer)
breakpoints <- c(0, 0.01, 1, 2, 3,4,5,6,7,8,9)
colors <- c("white", diverge_hcl(9, "Blue-Red"))
# colors <- c("white", "#11C638", "#48CA58", "#66CE70", "#7FD286", "#95D69A", "#A9D9AD", "#BDDDBF", "#D0E0D1", "#E2E2E2", "#E7D9D0", "#EBCFBD", "#EEC6A9", "#F0BC95", "#F1B37F", "#F1AA66", "#F1A047", "#8B000")
#8b0000

# diverge_hcl(9, h=c(130, 43), c =100, l = c(70,90))

testing <- raster::select(mosaic_calc)
plot(testing, breaks = breakpoints, col = colors)

```


```{r}
test_plumes <- list.files("/home/sgclawson/grassdata/location/PERMANENT/cats", pattern = "plume_effluent")
test_plumes
included_2005


test <- st_read(file.path("/home/shares/ohi/git-annex/land-based/wastewater/data/processed/N_effluent_output/effluent_N_pourpoints_plumes_tot_N.shp"))
st_crs(test)
summary(test)

st_crs(ws_pp_zonal_all_fix)

summary(ws_pp_zonal_all_fix)

```



