---
title: "Data Download"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(xlsx)
```

# New Global Fisheries Catch Data

* Industrial Catch (1950 - 2017) - reported, iuu, and discard catch data for each cell location and unique identifier
* Non-Industrial Catch (1950 - 2017) - reported, iuu, and discard catch data for each cell location and unique identifier
* DATA CODE DEFINITIONS (gear/taxa/country codes and cell lat/lon references)

```{r}

source("https://raw.githubusercontent.com/OHI-Science/ohiprep_v2019/gh-pages/workflow/R/common.R")
  
## Save download url suffixes 

web_years <- c("Catch1950_1954", "Catch1955_1959", "Catch1960_1964", "Catch1965_1969", 
               "Catch1970_1974", "Catch1975_1979", "Catch1980_1984", "Catch1985_1989",
               "Catch1990_1994", "Catch1995_1999", "Catch2000_2004", "Catch2005_2009", 
               "Catch2010_2014", "Catch2015_2019")

# reference <- c("IndexInd", "IndexNInd")

## Download reference data from web and save into mazu
# for(ref in reference){  #ref <- "IndexInd"
#   
#   data <- read.csv(sprintf("", ref))
#   
#   write.csv(data, file.path(dir_M, sprintf("git-annex/globalprep/_raw_data/IMAS_GlobalFisheriesLandings/d2019/%s.csv", ref)), row.names=F)
# }

## Download catch data from web and save into mazu
for(web_year in web_years){  
#web_year <- "Catch2015_2019"

data <- read.csv(sprintf("%s.csv", web_year)) #see readme on mazu for link...

saveRDS(data, file.path(dir_M, sprintf("git-annex/globalprep/_raw_data/IMAS_GlobalFisheriesLandings/d2020/%s.rds", web_year)))

}
```

# How to download the data code definitions. 

For this, it is saved as an .xlsx file (Codes.xlsx). This file has 4 sheets in it which contain meta data that explain columns in the Index files, and one sheet with is "Spatial Cells Reference - contains geospatial information associated wtih the Industrial Catch data". 
To download this data, go to the IMAS website: http://data.imas.utas.edu.au/portal/search?uuid=ff1274e1-c0ab-411b-a8a2-5a12eb27f2c0 and dowload it manually.

This .xlsx file contains these sheets:

* Spatial Cells Reference (Cells) - contains geospatial information associated wtih the Industrial Catch data
* Gear Reference (Gear) - contains information regarding how different fishing gear is classified in the index datasets.
* Taxa Reference (Taxa) - contains information regarding how different taxa is classified in the index datasets.
* Country Reference (Country) - contains informations regarding how different countries are labeled in the index datasets. 



