---
title: 'OHI `r format(Sys.Date(), "%Y")` - Tourism and Recreation '
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html' 
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

[REFERENCE RMD FILE](http://ohi-science.org/ohiprep_v2024/globalprep/tr/v2024/tr_data_prep.html)

# Summary

This document describes the steps for obtaining the data used to calculate the tourism and recreation goal for the 2024 global assessment.

The general calculation is: tr = Ap \* Sr and Xtr = tr/90th quantile across regions

-   Ap = Proportion of overnight tourist arrivals to total arrivals
-   Sr = (S-1)/5; Sustainability of tourism

`part1_arrivals_prep.Rmd` is the script for preparing the UNWTO data on international and domestic arrival flights as well as domestic accommodations. 

`part2_sustainability_prep.Rmd` is the script for calculating tourism sustainability using WEF data. 

## Updates

v2025: 
- Split script into two separate pieces
- Cleaned up the data sources and introduction because it was very unclear which files were needed previously
- Changed gapfilling method #4 under raw gapfilling from a poorly-fit linear model to a linear interpolation using the `zoo` package

> This goal is still fairly messy and could use a lot of tightening up in future assessments. v2025 did an in-depth check of the gapfilling methods, the results of which can be found in [issue #355](https://github.com/OHI-Science/globalfellows-issues/issues/355) for more details. 


# Data Sources

**New data sources**

- [United Nations World Tourism Organization](https://www.unwto.org/tourism-statistics/tourism-statistics-database):
  - [International Inbound arrivals](https://www.unwto.org/tourism-statistics/tourism-data-inbound-tourism): Scroll down to the bottom of this page until you see a card near the bottom titled "International Arrivals" and a "Download data" button. click download data. File name should look something like `UN_Tourism_inbound_arrivals_DATE.xlsx`
  - [Domestic Inbound arrivals](https://www.unwto.org/tourism-statistics/tourism-data-domestic-tourism): Scroll down to the bottom of this page until you see a card near the bottom titled "Total Trips" and a "Download data" button. click download data. File name should look something like `UN_Tourism_domestic_trips_DATE.xlsx`
  - [Domestic Accommodations](https://www.unwto.org/tourism-statistics/tourism-data-domestic-tourism): Scroll down to the bottom of this page until you see a card near the bottom titled "Accommodation" and a "Download data" button. click download data. File name should look something like `UN_Tourism_domestic_accommodation_DATE.xlsx`

You can also find the same files if you do the bulk data download from the homepage, although you will only need these three. 

- [World bank total population data](https://data.worldbank.org/indicator/SP.POP.TOTL): Open link. Find download button on the bottom/middle right side of the page. Click the CSV option. Upload to mazu under
 `globalprep/_raw_data/WorldBank/dYEAR/WorldBank_global_country_population_1960-20XX/API_SP.POP.TOTL_DS2_en_csv_v2_######.csv`).

**Data sources from other goals**

- `mar_prs_population` Coastal population: `mar_prs_population/vYEAR/output/mar_pop_25mi.csv`

- `lsp` Coastal area:
`lsp/vYEAR/output/rgn_area_inland1km.csv` and `lsp/vYEAR/output/rgn_area_offshore3nm.csv`


**Methodology:**

1.  Download and clean international UNWTO inbound tourism arrivals data

-   Filter to 2005:2021 and remove regions without enough data (less than or equal to three observations)

2.  Download and clean domestic UNWTO inbound tourism Accommodations and Trips data

-   Filter to 2005:2021 and remove regions without enough data (less than or equal to three observations)
-   Gapfill total guests with:
    -   "Accommodations" data: hotel guests (linear model prediction)
    -   "Trips" data: overnight guests (simple replacement)
    -   "Accommodations" data: total overnights (this is guests \* days; so determine average number of nights stayed and divide by this)
    -   "Accommodations" data: hotel overnights (linear interpolation)

3.  Gapfilling missing years in preparation for International and Domestic Aggregation

-   International: for years that do not have values (upfilling for previous or gap years, then lm for future years)
-   Domestic: for years that do not have values (upfilling for previous or gap years, then lm for future years)
-   Domestic: for regions present in international data that is not in domestic data. This is done by taking the mean of the domestic guest values by georegion (r2 and then r1 if there is not enough data for the r2). We decided that there would be less error after aggregating if we at least put a mean (by georegion) domestic guest value for regions that have no domestic data, but they have international data. That way, all international regions will be adding to a domestic value, rather than some regions only having international data.

4.  Calculating Ap (arrivals proportion)

-   Aggregating International and Domestic
-   Finding the Coastal Proportion
    -   This is done by dividing the coastal population by the total population of a region.
-   Determining regional Coastal Area
    -   The $km^2$ of coastal area for all OHI regions by aggregating the area protected 1 kilometer inland, and 3 nautical miles out into the eez.
-   Ap
    -   Join all of these data sources and conduct $$Ap = \frac{TourismArrivals * (CoastalPop/TotalPop)}{CoastalArea}$$

5.  Rescale Ap proportion using the 90th quantile

-   As according to the [OHI methods](https://ohi-science.org/ohi-methods/goals/goal-models-data.html#tourism-and-recreation), the Ap values for each year and region is rescaled using the 90th quartile for each region.

$$\frac{T_r}{T_{90th}}$$

# Setup

```{r setup, message=FALSE, warning=FALSE, results="hide"}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, eval=FALSE)

if (!require(librarian)) {
  install.packages("librarian")
  library(librarian)
}

# Load libraries
librarian::shelf(
  ohicore, #devtools::install_github('ohi-science/ohicore@dev')
  here,
  tidyverse,
  WDI,
  readxl,
  naniar,
  countrycode,
  pdftools,
  plotly,
  zoo
)

# ---- sources! ----
source(here("workflow", "R", "common.R")) # file creates objects to process data

region_data() # for rgns_all and rgns_eez

regions_shape() # returns spatial shape object named 'regions' which includes land, eez, highseas, and antarctica regions

#source(here(paste0("globalprep/tr/v", version_year, "/R/tr_fxns.R"))) # not used presently

# ---- set year and file path info ----
current_year <- 2025 # Update this in the future!!
previous_year <- 2024

version_year <- paste0("v",current_year)
previous_vyear <- paste0("v", current_year - 1)

current_data_dir <- paste0("d", current_year)
previous_data_dir <- paste0("d", (current_year - 1))

data_years <- 2005:2022 # update with latest data year (probably just add 1)

# ---- data directories ----

# Raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")

# UNWTO (UN World Tourism) raw data directory
current_unwto_dir <- here(raw_data_dir, "UNWTO", current_data_dir)
previous_unwto_dir <- here(raw_data_dir, "UNWTO", previous_data_dir)

# Tourism and recreation directory
current_tr_dir <- here("globalprep", "tr", version_year)
previous_tr_dir <- here("globalprep", "tr", previous_vyear)
```

# UNWTO Tourism Arrivals

*International data:* Click Inbound Arrivals, scroll down to "Total arrivals", click download data. File name should look something like `UN_Tourism_inbound_arrivals_DATE.xlsx`

## International Inbound Tourism Arrivals - UNWTO

```{r}
# Read in arrivals data
unwto_arrivals_int <- read_xlsx(here(current_unwto_dir, "UN_Tourism_inbound_arrivals_11_2023.xlsx"), 
                                        sheet = 2, # sheet number with actual data in it
                                        skip = 2) # number of rows to skip before reading in data. check how many by opening the sheet in excel

# Clean arrivals data
int_arrivals_clean <- unwto_arrivals_int %>% 
  select(country = `Basic data and indicators`, total_arrivals = `...6`, subdivision_1 = `...7`, subdivision_2 = `...8`, `1995`:as.character(max(data_years))) %>% 
  
  # add country name to all data associated with that country
  fill(country, .direction = "down") %>% 
  pivot_longer(cols = c("total_arrivals", "subdivision_1", "subdivision_2"),
               values_to = "metric",
               values_drop_na = TRUE) %>% # make the metrics into one column
  select(-name) %>% # get rid of the name column since it's redundant
  select(country, metric, everything()) %>% # reorder things
  
  # swap .. with NAs
  replace_with_na_all(condition = ~.x == "..") %>% 
  pivot_longer(cols = 3:ncol(.), names_to = "year",
               values_to = "tourism_arrivals_ct") %>% # make the years not columns anymore
  pivot_wider(names_from = metric, values_from = tourism_arrivals_ct) %>%
  
  # rename metrics so easier to work with and fix types and add a new column to fill with the new calculated values later
  mutate(overnights = as.numeric(`Overnights visitors (tourists)`), 
         same_day = as.numeric(`Same-day visitors (excursionists)`), 
         total_arrivals = as.numeric(`Total arrivals`),
         tourism_arrivals_ct = as.numeric(NA)) %>% 
  select(country, year, overnights, same_day, total_arrivals, tourism_arrivals_ct) %>% # select columns needed for analysis (cruise passengers seem to be included in same-day)
  group_by(country, year) # group by county and year
  

int_arrivals_gf <- int_arrivals_clean %>% 
  # Gapfill
  mutate(
    tourism_arrivals_ct = case_when(
      # if there is a value, dont gapfill
      !is.na(overnights) ~ overnights, 
      # gapfill when there is no data on overnights, fill with total_arrivals - same day
      is.na(overnights) & !is.na(same_day) & !is.na(total_arrivals) ~ total_arrivals - same_day,
      # otherwise, NA
      TRUE ~ tourism_arrivals_ct 
    ), 
    # there were 0 situations like this in v2024 or v2025
    total_arrivals = case_when(
      !is.na(total_arrivals) ~ total_arrivals,
      is.na(total_arrivals) & !is.na(same_day) & !is.na(overnights) ~ overnights + same_day,
      TRUE ~ total_arrivals
    )
  ) %>% 
  mutate(arrivals_method = ifelse(is.na(overnights) & !is.na(same_day) & !is.na(total_arrivals), "Arrivals minus same day", NA),
         arrivals_method = ifelse(is.na(total_arrivals) & !is.na(same_day) & !is.na(overnights), "Overnights plus same day", arrivals_method)) %>%
  mutate(arrivals_gapfilled = ifelse(arrivals_method == "Arrivals minus same day", "gapfilled", NA),
         arrivals_gapfilled = ifelse(arrivals_method == "Overnights plus same day", "gapfilled", arrivals_gapfilled)) %>% # prepare a "gapfilled" column to indicate "gapfilled" or NA
  ungroup()

# ----- Quick data check -----
# v2024: overnights has 1036 NAs out of 6021
# v2024: same_day has 3131 NAs out of 6021
# v2024: total_arrivals has 2363 NAs

colSums(is.na(int_arrivals_gf))
# v2025: 
#          overnights            same_day 
#                1131                3276 
#      total_arrivals tourism_arrivals_ct 
#                2612                1099 
# Pretty similar to the v2024 numbers


# Select desired columns for further processing
int_arrivals_gf <- int_arrivals_gf %>% 
  # select only needed columns
  select(country, year, tourism_arrivals_ct, total_arrivals,
         arrivals_method, arrivals_gapfilled) %>%
  mutate(country = str_to_title(country),
         tourism_arrivals_ct = round(as.numeric(tourism_arrivals_ct) * 1000),
         total_arrivals = round(as.numeric(total_arrivals) * 1000)) # fix units
```

Now, run `ohicore::name_2_rgn()` so that we can move forward with OHI regions! Ensure duplicates are dealt with.

```{r}
# Run name_2_rgn() to ensure the UNWTO data has OHI region names
int_arrivals_rgn_names <- name_2_rgn(df_in = int_arrivals_gf, # do this just for Bonaire since it is the only region not matching above
                                        fld_name = 'country',
                                        # flds_unique = c('year'),
                                        keep_fld_name = TRUE) %>%
  dplyr::select(rgn_id, rgn_name, year, tourism_arrivals_ct, arrivals_method, arrivals_gapfilled) # v2024: not losing the USA!! Yay (compare to v2023).  Also, total_arrivals was dropped at this point because tourism_arrivals_ct has already been gapfilled using totals if same-day was present (Overnights = total - same-day), and it feels like if total was used to gapfill places that don't have any same-day data, it may be overinflating the arrivals for that region.  This can be revisited.

length(unique(int_arrivals_gf$country)) # 223 

# ---- determine if there were any duplicates due to name_2_rgn ----

colSums(is.na(int_arrivals_rgn_names)) # v2025: 861 NAs in tourism_arrivals_ct
length(int_arrivals_rgn_names$tourism_arrivals_ct) # within 5012 observations for `tourism_arrivals_ct`

# this shows rows that have at least one duplicate, including both the first and subsequent occurrences of each duplicated row. using the tidyverse allows us to see all instances of duplicated data, not just the second and subsequent occurrences (which is what just a duplicated() call would give you).
duplicates_int <- int_arrivals_rgn_names %>% 
  group_by(rgn_id, rgn_name, year) %>% 
  filter(n() > 1) %>%
  ungroup()

unique(duplicates_int$rgn_name)
# v2025:
# [1] "China"                                              
# [2] "Guadeloupe and Martinique"                          
# [3] "Northern Mariana Islands and Guam"                  
# [4] "Montenegro"                                         
# [5] "Puerto Rico and Virgin Islands of the United States"

# Make a function to avoid losing NAs (they turn to 0s) when there are values to aggregate with
sum_with_na <- function(x) {
  if (all(is.na(x))) { # for x, if all of the values are NA, leave as NA.  
    return(NA)
  } else { 
    return(sum(x, na.rm = TRUE)) # if there are values for that grouped dataframe, then sum them according to the grouping (in this case, by region and year)
  }
}

# use the function within a summarize to aggregate for all regions that have duplicate years
int_arrivals_agg <- int_arrivals_rgn_names %>% 
  group_by(rgn_id, rgn_name, year) %>% 
  summarize(
    tourism_arrivals_ct = sum_with_na(tourism_arrivals_ct), # put each column in the function, so no NAs are lost
    arrivals_method = first(arrivals_method, na_rm = TRUE),
    arrivals_gapfilled = first(arrivals_gapfilled, na_rm = TRUE),
    .groups = "drop" # to ungroup, fixes warning from summarise to use reframe (when it is not applicable here because we want to reduce each group down to a single row)
  ) %>% 
  mutate(
    arrivals_method = ifelse(is.na(arrivals_method) & !is.na(tourism_arrivals_ct), "UNWTO", arrivals_method), # so that if there is data from UNWTO that was not gapfilled, the arrivals_method is accurate
    year = as.numeric(year)
  )

# double check to test
duplicates_int_2 <- int_arrivals_agg %>%
  group_by(rgn_id, year) %>% 
  filter(n() > 1) %>%
  ungroup()
unique(duplicates_int_2$rgn_name) # v2025: 0! yay -- and China/other duplicates are preserved here!!

# check out things so far
summary(int_arrivals_agg) 
# v2023: 828 NAs in arrivals (before filtering the years down and gapfilling), 1708 in `total_arrivals`
# v2024: 757 NAs in `tourism_arrivals_ct` (before filtering the years down and gapfilling)
# v2025: 825 NAs in `tourism_arrivals_ct` (before filtering the years down and gapfilling)

# see how many regions have values
check_international <- int_arrivals_agg %>% 
  filter(!is.na(tourism_arrivals_ct))
length(unique(check_international$rgn_id)) # v2025: 161 regions present!

rm(duplicates_int, duplicates_int_2, check_international)
```


```{r}
# filter international data for usable years to be used in the future
int_arrivals_counts <- int_arrivals_agg %>% 
  filter(year %in% data_years) %>% # This should be 2005 to whatever the max data year is for your assessment 
  group_by(rgn_id) %>% 
  summarize(count_values = sum(!is.na(tourism_arrivals_ct))) %>% # for each region, how many values in tourism_arrivals_ct did it have
  filter(count_values >= 3) %>% # keep only the regions that have greater than or equal to three values
  ungroup()

length(unique(int_arrivals_counts$rgn_id)) # v2025: 156 regions

# Filter down final data frame to desired regions and years
international_arrivals <- int_arrivals_agg %>% 
  filter(rgn_id %in% int_arrivals_counts$rgn_id) %>% 
  filter(year %in% data_years) # filter for regions we want in the right timeframe

length(unique(international_arrivals$rgn_id)) # 156, matches above

colSums(is.na(international_arrivals)) # 263 NAs in tourism_arrivals_ct
nrow(international_arrivals) # 2808
# Only ~9% of the data to gapfill 

# Export intermediate df for future use
write_csv(international_arrivals, here(current_tr_dir, "int", "international_arrivals_int.csv"))
```

```{r}
int_arrivals <- read_csv(here(current_tr_dir, "int", "international_arrivals_int.csv"))

missing

colnames(int_arrivals)

int_arrivals %>% 
  filter(rgn_id %in% missing$rgn_id & year %in% missing$year) %>% 
  view()

```


## Domestic Inbound Tourism Accomodation and Trips - UNWTO

*Domestic data:* Click Domestic Tourism, scroll down to "Total trips", click download. Then download from "Accommodations".

### Clean Accommodations data

```{r}
# bring in domestic accommodation data from UNWTO
unwto_arrivals_dom_acc <- read_xlsx(here(current_unwto_dir, "UN_Tourism_inbound_accommodation_11_2023.xlsx"), 
                                        sheet = 2, # sheet number with actual data in it
                                        skip = 2) # number of rows to skip before reading in data. check how many by opening the sheet in excel

# clean up the raw xlsx! We want to have the country, the source (Total:Guests, Total:Overnights, Hotel:Guests, Hotel:Overnights), then the values
unwto_clean_dom <- unwto_arrivals_dom_acc %>% 
  select(country = `Basic data and indicators`, source_arrivals = `...6`, subdivision_1 = `...7`, subdivision_2 = `...8`, `1995`:as.character(max(data_years))) %>% 
  fill(country, .direction = "down") %>% # add country name to all data associated with that country
  fill(source_arrivals, .direction = "down") %>% 
  filter(subdivision_1 %in% "Overnights" | subdivision_1 %in% "Guests") %>% 
  select(-subdivision_2) %>% # get rid of the NA column
  replace_with_na_all(condition = ~.x == "..") %>% # swap ".." with NAs
  mutate(source_arrivals_all = paste(source_arrivals, subdivision_1, sep = ":")) %>% # unite the columns that define the source for easier cleaning and pivoting
  select(country, source_arrivals_all, 4:ncol(.)) # use the . to specify the whole dataframe

# tidy up the data to be usable
dom_acc_guests <- unwto_clean_dom %>% 
  pivot_longer(cols = 3:ncol(.), names_to = "year",
               values_to = "tourism_arrivals_ct") %>% # make the years not columns anymore
  pivot_wider(names_from = source_arrivals_all, values_from = tourism_arrivals_ct) %>%
  mutate(
    total_guests = as.numeric(`Total:Guests`),
    total_overnights = as.numeric(`Total:Overnights`),
    hotel_guests = as.numeric(`Hotels and similar establishments:Guests`), 
    hotel_overnights = as.numeric(`Hotels and similar establishments:Overnights`),
    tourism_arrivals_ct = as.numeric(NA)
    ) %>% # rename metrics so easier to work with, make numeric, and add a new column to fill with the new calculated values later
  select(country, year, total_guests, total_overnights, hotel_guests, hotel_overnights, tourism_arrivals_ct) %>% # select columns needed for future raw gapfilling 
  mutate(country = str_to_title(country)) # make countries look nice
```

Now let's take a look at the raw data for Accomodations!

```{r}
# see how many NAs are within the raw data for Total:Guests
colSums(is.na(dom_acc_guests))
# v2024                                
#  total_guests      total_overnights
# NA's   :4947       NA's   :4614    
#  hotel_guests      hotel_overnights  
# NA's   :4129       NA's   :3835 

# v2025
#        total_guests    total_overnights 
#                4539                4108 
#        hotel_guests    hotel_overnights 
#                3281                3087 
 

length(unique(dom_acc_guests$country)) # 223, which includes regions that have all NA values (UNWTO always has datasets with those 223 countries)

# remove NAs to see how many regions have total_guests values
dom_acc_guests_noNA <- dom_acc_guests %>% 
  filter(!is.na(total_guests)) 

length(unique(dom_acc_guests_noNA$country)) # 83 unique countries
```

So, it looks like there are about 4947 NAs within the Total:Guests column. However, we want to preserve guests/individuals as the unit (rather than overnights), so in the Raw Gapfilling section, let's see if we can gapfill with either other columns (hotel_guests) or the "Trips" data.

Let's run ohicore::name_2_rgn() so we don't have to worry about it in the future.

```{r}
# run name_2_rgn on the domestic data so it has OHI region ids
rgns_dom_acc_guests <- name_2_rgn(df_in = dom_acc_guests, 
                                        fld_name = 'country',
                                        keep_fld_name = TRUE) # lots of landlocked countries were removed, and there are duplicates.  Let's take a look.

# we only need mainland China data, not necessarily Hong Kong or Macao.
duplicates <- rgns_dom_acc_guests %>%
  group_by(rgn_id, year) %>% 
  filter(n() > 1) %>%
  ungroup()
unique(duplicates$rgn_name)
# [1] "China"                                              
# [2] "Guadeloupe and Martinique"                          
# [3] "Northern Mariana Islands and Guam"                  
# [4] "Montenegro"                                         
# [5] "Puerto Rico and Virgin Islands of the United States"

# see which ones even have data -- if none of the columns have a value, then we should drop these regions anyways.
duplicates_no_all_NAs <- duplicates %>%
  filter(if_any(c(total_guests, total_overnights, hotel_guests, hotel_overnights), ~!is.na(.))) # only keeps the rows where at least one of the columns has a value

unique(duplicates_no_all_NAs$rgn_name) 
# All of the regions above have data

# therefore, move forward by aggregating by region id and year for puerto rico and virgin islands.
## make a function to avoid losing NAs (they turn to 0s) when there are values to aggregate with
sum_with_na <- function(x) {
  if (all(is.na(x))) {
    return(NA)
  } else {
    return(sum(x, na.rm = TRUE))
  }
}

# now use the function to aggregate by region and year!
dom_acc_guests_agg <- rgns_dom_acc_guests %>% 
  group_by(rgn_id, rgn_name, year) %>% 
  summarize(
    tourism_arrivals_ct = sum_with_na(tourism_arrivals_ct), # put each column in the function, so no NAs are lost
    total_guests = sum_with_na(total_guests),
    total_overnights = sum_with_na(total_overnights),
    hotel_guests = sum_with_na(hotel_guests),
    hotel_overnights = sum_with_na(hotel_overnights),
    .groups = "drop" # to ungroup, fixes warning from summarise to use reframe (when it is not applicable here)
  ) #### looks good!!!


# double check to test
duplicates_2 <- dom_acc_guests_agg %>%
  group_by(rgn_id, year) %>% 
  filter(n() > 1) %>%
  ungroup()
unique(duplicates_2$rgn_name) # v2025: 0! yay
```

```{r}
colSums(is.na(dom_acc_guests_agg)) # 3454 `total_guests` 
nrow(dom_acc_guests_agg) # 4844
# ~ 71% NA data for total guests

# Write out intermediate domestic accommodations data
write_csv(dom_acc_guests_agg, here(current_tr_dir, "int", "domestic_accommodations_int.csv"))
```

### Clean Trips data

Domestic trips data:

```{r}
# Read in domestic trips data
unwto_arrivals_dom_trips <- read_xlsx(here(current_unwto_dir, "UN_Tourism_domestic_trips_11_2023.xlsx"), 
                                        sheet = 2, # sheet number with actual data in it
                                        skip = 2) # number of rows to skip before reading in data. check how many by opening the sheet in excel

# clean up the Trips data!
unwto_dom_trips_clean <- unwto_arrivals_dom_trips %>% 
  select(country = `Basic data and indicators`, total_trips = `...6`, subdivision_1 = `...7`, subdivision_2 = `...8`, `1995`:as.character(max(data_years))) %>% 
  fill(country, .direction = "down") %>% # add country name to all data associated with that country
  pivot_longer(cols = c("total_trips", "subdivision_1", "subdivision_2"),
               values_to = "metric",
               values_drop_na = TRUE) %>% # make the metrics into one column
  select(-name) %>% # get rid of the name column since it's just the titles of the metrics which are already there
  select(country, metric, everything()) %>% # reorder things
  replace_with_na_all(condition = ~.x == "..") %>% # swap .. with NAs
  pivot_longer(cols = 3:ncol(.), names_to = "year",
               values_to = "tourism_arrivals_trips") %>% # make the years not columns anymore
  mutate(country = str_to_title(country)) %>% # make the country names look nicer
  pivot_wider(names_from = metric, values_from = tourism_arrivals_trips) %>%
  mutate(trips_overnights = as.numeric(`Overnights visitors (tourists)`), 
         trips_same_day = as.numeric(`Same-day visitors (excursionists)`), 
         trips_total = as.numeric(`Total trips`)) %>% # rename metrics so easier to work with, make numeric, and add a new column to fill with the new calculated values later
  select(country, year, trips_overnights, trips_same_day, trips_total) # select only the columns we will be working with
```


```{r}
# see how many NAs are within the raw data for Trips:Overnights
colSums(is.na(unwto_dom_trips_clean))
# v2025
# trips_overnights   trips_same_day 
#             5316             5694 
#      trips_total 
#             5486 

length(unique(unwto_dom_trips_clean$country)) # 223, which includes regions that have all NA values (UNWTO always has datasets with those 223 countries)

unwto_dom_trips_clean_noNA <- unwto_dom_trips_clean %>% 
  filter(!is.na(trips_overnights)) # remove NAs to see how many regions have total_guests values

length(unique(unwto_dom_trips_clean_noNA$country)) # 82 unique countries, not bad! Let's see if there is an association between trips_overnights and total_guests, to see if we can use the trips_overnights values for total_guests etc!
```

Let's run ohicore::name_2_rgn() so we don't have to worry about it in the future.

```{r}
# run name_2_rgn on the domestic data so it has OHI region ids
rgns_unwto_dom_trips_clean <- name_2_rgn(df_in = unwto_dom_trips_clean, 
                                        fld_name = 'country',
                                        keep_fld_name = TRUE) # lots of landlocked countries were removed, and there are duplicates.  Let's take a look.

# we only need mainland China data, not necessarily Hong Kong or Macao.
duplicates_trips <- rgns_unwto_dom_trips_clean %>%
  group_by(rgn_id, year) %>% 
  filter(n() > 1) %>%
  ungroup()
unique(duplicates_trips$rgn_name)
# [1] "China"                                              
# [2] "Guadeloupe and Martinique"                          
# [3] "Northern Mariana Islands and Guam"                  
# [4] "Montenegro"                                         
# [5] "Puerto Rico and Virgin Islands of the United States"

# see which ones even have data -- if none of the columns have a value, then we should drop these regions anyways.
duplicates_trips_no_all_NAs <- duplicates_trips %>%
  filter(if_any(c(trips_overnights, trips_same_day, trips_total), ~!is.na(.))) # only keeps the rows where at least one of the columns has a value

unique(duplicates_trips_no_all_NAs$rgn_name) # [1] "China" needs to be aggregated and fixed... the rest did not have data anyway.

# therefore, move forward by aggregating by region id and year for china.
## make a function to avoid losing NAs (they turn to 0s) when there are values to aggregate with
sum_with_na <- function(x) {
  if (all(is.na(x))) {
    return(NA)
  } else {
    return(sum(x, na.rm = TRUE))
  }
}

# now use the function to aggregate by region and year!
dom_trips_agg <- rgns_unwto_dom_trips_clean %>% 
  filter(!(rgn_id %in% c(140, 13, 186, 116))) %>% # removing duplicates that are all NAs. 140 is Guadeloupe and Martinique, 13 is Northern Mariana Islands and Guam, 186 is Montenegro, 116 is puerto rico and virgin islands
  group_by(rgn_id, rgn_name, year) %>% 
  summarize(
    trips_overnights = sum_with_na(trips_overnights), # put each column in the function, so no NAs are lost
    trips_same_day = sum_with_na(trips_same_day),
    trips_total = sum_with_na(trips_total),
    .groups = "drop" # to ungroup, fixes warning from summarise to use reframe (when it is not applicable here)
  ) #### looks good!!!

# double check to test
duplicates_trips_2 <- dom_trips_agg %>%
  group_by(rgn_id, year) %>% 
  filter(n() > 1) %>%
  ungroup()
unique(duplicates_trips_2$rgn_name) # v2025: 0! yay
```

```{r}
colSums(is.na(dom_trips_agg)) # 4034 NAs for trips_overnights
nrow(dom_trips_agg) # 4732 
# ~ 85% NAs

# Write out intermediate domestic trips data
write_csv(dom_trips_agg, here(current_tr_dir, "int", "domestic_trips_int.csv"))
```

# Raw Gapfilling

## Domest Accommodations

**Hotel Guests used gapfill Total Guests**

Hotel guests are registered individuals that stayed overnight at a hotel or similar establishment. Total guests includes dwellings or places one might stay in that region that is not considered a hotel, but would still be associated with tourism. Let's use hotel_guests to estimate total_guests.

```{r}
# Read in the data we cleaned earlier
dom_acc_guests_agg <- read_csv(here(current_tr_dir, "int", "domestic_accommodations_int.csv"))

colSums(is.na(dom_acc_guests_agg))

# ------- total_guests ~ hotel_guests ---------
mod1 <- lm(total_guests ~ hotel_guests, data = dom_acc_guests_agg)
summary(mod1)
# v2024
# significant model! multiple r squared = 0.9503
# total_guests = 4014.424577 + 1.136106*hotel_guests

# v2025
# multiple r squared = 0.9557
# total_guests = 426.79148 + 1.22071*hotel_guests

# plot the data to compare
mod_plot1 <- ggplot(dom_acc_guests_agg, aes(x = hotel_guests, y = total_guests)) +
  geom_point(alpha = 0.5) + 
  geom_abline(intercept = mod1$coefficients[1], 
              slope = mod1$coefficients[2], 
              color = "red", linetype = "dashed") +
  theme_minimal() + 
  labs(title = "Hotel Guests vs Total Guests, Accomodations data",
       x = "Hotel Guests",
       y = "Total Guests");ggplotly(mod_plot1) 

# log scale
ggplot(dom_acc_guests_agg, aes(x = log(hotel_guests + 1), y = log(total_guests + 1))) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  theme_minimal() +
  labs(title = "Hotel:Guests vs Total:Guests (Log Scale) within Accommodation",
       x = "Log(Hotel Guests + 1)",
       y = "Log(Total Guests + 1)") # looks a lot better!

# let's look at their correlation. see ?`cor` if you are interested in learning more about how this is calculated (`cor` computes the variance of x and the covariance or correlation of x and y).  Kendall correlation is preferred when there are small samples or some outliers (https://datascience.stackexchange.com/questions/64260/pearson-vs-spearman-vs-kendall)

# significant amt of NAs, so we need to remove them before finding the correlation:
clean_dom_acc_guests <- dom_acc_guests_agg[complete.cases(dom_acc_guests_agg[c("total_guests", "hotel_guests")]), ] # filtering the dom_acc_guests data frame to keep only the rows where both "total_guests" and "hotel_guests" have non-missing values

# now lets try to find the correlation:
cor(clean_dom_acc_guests$total_guests, clean_dom_acc_guests$hotel_guests, method = "kendall") # 0.9034738, great!! This means that there is a pretty high correlation between the two variables (cor is close to 1).  Correlation measures the strength of association between two variables and the direction of the relationship: so in this case, it is positive and they have high association.

sd(clean_dom_acc_guests$total_guests) # 12044.65 guests
sd(clean_dom_acc_guests$hotel_guests) # 9645.782 guests
# SDs are very different from each other...

# therefore, go ahead and use the linear model to conduct gapfilling in the next chunk using this data.
```

Let's move on to gapfilling. Since it is slightly more than 1, let's use the slope and coefficient from the linear model to calculate total_guests from hotel_guests. This is just to improve accuracy.

```{r hotel_guests_gf}
# ------- find coefficients for gapfilling using hotel_guests when total_guests = NA 

# total_guests = 426.79148 + 1.22071*hotel_guests

# will need to gapfill these slopes and intercepts for regions that have hotel_guests data but no total_guests data. join with georegions, group by r2, and later r1 and take the mean for those regions. Otherwise, use regional level.
georegions <- ohicore::georegions %>% 
  left_join(rgns_eez, by = "rgn_id") %>% 
  select(-c(Notes)) # not needed

# first, remove any regions that only have NAs, then calculate coefficients
dom_acc_guests_coef <- dom_acc_guests_agg %>% 
  left_join(georegions, by = c("rgn_id")) %>% 
  group_by(rgn_id) %>% # group by region
  # remove rows where there are NAs for both columns
  filter(if_all(c(total_guests, hotel_guests), ~!is.na(.))) %>% 
  mutate(
    # obtain the slope for each region that has data
    total_guests_slope = lm(total_guests ~ hotel_guests)$coefficient[2]
  ) %>% 
  ungroup() %>% 
  # get values for r1 and r2, so that if regional is available, use that... then if r2 is available, use that... finally, use r1 if you can't otherwise.
  # r2 is the more specific georegion (carribean, central africa, etc.)
  group_by(r2) %>% 
  mutate(total_guests_slope_r2 = lm(total_guests ~ hotel_guests)$coefficient[2]) %>% 
  ungroup() %>% 
  # r1 is the more general georegion (continent)
  group_by(r1) %>% 
  mutate(total_guests_slope_r1 = lm(total_guests ~ hotel_guests)$coefficient[2]) %>% 
  ungroup() %>% 
  # prep for left join, otherwise many-to-many
  distinct(rgn_id, r2, r1, total_guests_slope, total_guests_slope_r2, total_guests_slope_r1) %>% 
  # select the columns that will be needed
  select(rgn_id, r2, r1, total_guests_slope, total_guests_slope_r2, total_guests_slope_r1) 
```

```{r}
# ------- gapfill using hotel_guests when total_guests = NA ---------
dom_acc_guests_gf1 <- dom_acc_guests_agg %>%
  left_join(georegions, by = c("rgn_id","rgn_name")) %>% 
  left_join(dom_acc_guests_coef, by = c("rgn_id","r2","r1")) %>%
  group_by(r2) %>% # get values for r1 and r2 for regions that do not have data, so that if regional is available, use that... then if r2 is available, use that... finally, use r1 if you can't otherwise.
  mutate(
    total_guests_slope_r2mean = mean(total_guests_slope_r2, na.rm = TRUE),
    total_guests_slope_r2mean = ifelse(is.nan(total_guests_slope_r2mean), NA, total_guests_slope_r2mean) # any NaN values to NAs
  ) %>%
  ungroup() %>%
  group_by(r1) %>%
  mutate(
    total_guests_slope_r1mean = mean(total_guests_slope_r1, na.rm = TRUE),
    total_guests_slope_r1mean = ifelse(is.nan(total_guests_slope_r1mean), NA, total_guests_slope_r1mean) # any NaN values to NAs
  ) %>%
  ungroup() %>%  # now we have intercept and slope values for all regions (at different levels), lets use that to gapfill total_guests with hotel_overnights!
  mutate(
    arrivals_method = ifelse(is.na(total_guests) & !is.na(hotel_guests), "Hotel:Guests", NA), # do this first so that we can use when tourism_arrivals_ct is NA as a qualifier
    tourism_arrivals_ct = case_when(
      !is.na(total_guests) ~ total_guests, # if there is a value, dont gapfill
      is.na(total_guests) & !is.na(hotel_guests) & !is.na(total_guests_slope) ~ (hotel_guests*total_guests_slope),
      is.na(total_guests) & !is.na(hotel_guests) & is.na(total_guests_slope) & !is.na(total_guests_slope_r2mean) ~ (hotel_guests*total_guests_slope_r2mean), # if regional lm coefficients weren't possible, go to r2 which is second least specific, then r1
      is.na(total_guests) & !is.na(hotel_guests) & is.na(total_guests_slope) & is.na(total_guests_slope_r2mean) & !is.na(total_guests_slope_r1mean) ~ (hotel_guests*total_guests_slope_r1mean),
      TRUE ~ tourism_arrivals_ct
    ),
    arrivals_gapfilled = ifelse(arrivals_method == "Hotel:Guests", "gapfilled", NA)
  ) %>% # example: 	New Caledonia and Monaco.  The numbers make sense because the estimated total_guests were larger than hotel_guests (makes sense because total includes other establishments than just hotels) and we can compare to hotel_overnights, of which tourism_arrivals_ct is about half ish, which based on our knowledge of mean_stays (found later on in raw gf for dom) the average is between 2-3 nights.  So this makes sense!
  select(rgn_id, rgn_name, year, tourism_arrivals_ct, total_guests, total_overnights, hotel_guests, hotel_overnights, arrivals_method, arrivals_gapfilled) # select all that we need moving forward
```

```{r}
# take a look at how that changed the number of NAs after our first dom gf step:
colSums(is.na(dom_acc_guests_gf1))# now, `tourism_arrivals_ct` has 2419 NAs, versus 3454 NAs in `total_guests` from the raw clean accommodation data


dom_acc_guests_gf1_noNA <- dom_acc_guests_gf1 %>% 
  filter(!is.na(tourism_arrivals_ct)) # remove NAs to see how many regions have total_guests values

length(unique(dom_acc_guests_gf1_noNA$rgn_name)) # we now have 111 unique countries, thats okay. Let's move on and see if we can gapfill further.

```

## Trips:Overnights used to fill in data for Accommodations:Total:Guests

**Join Accommodations and Trips data frames first.**

```{r}
# Read in the domestic trips data
dom_trips_agg <- read_csv(here(current_tr_dir, "int", "domestic_trips_int.csv"))

dom_acc_trips_join <- left_join(dom_acc_guests_gf1, dom_trips_agg, by = c("rgn_id","rgn_name", "year")) # now it has all variables within one data frame
```

Now, evaluate if gapfilling is appropriate:

```{r}
# ------- total_guests ~ trips_overnights ---------
mod2 <- lm(total_guests ~ trips_overnights, data = dom_acc_trips_join)
summary(mod2)
# significant model and slope! almost sig intercept, but that is okay. multiple r-squared was 0.7657 (okay enough to move forward, not the best).
# total_guests = -2799.2542 + 0.7111*trips_overnights

# plot the data to compare
ggplot(dom_acc_trips_join, aes(x = total_guests, y = trips_overnights)) +
  geom_point(alpha = 0.5) +  
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  theme_minimal() + 
  labs(title = "Accomodation:Total:Guests vs Trips:Overnight",
       x = "Trips Overnight Visitors (from Trips)",
       y = "Total Guests (from Accomodation)") # does not look that bad, the slope is slightly less than 1

# log scale
ggplot(dom_acc_trips_join, aes(x = log(total_guests + 1), y = log(trips_overnights + 1))) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  theme_minimal() +
  labs(title = "Accomodation:Total:Guests vs Trips:Overnight (Log Scale)",
       x = "Log(Trips Overnight Visitors (from Trips) + 1)",
       y = "Log(Total Guests (from Accomodation) + 1)") # mediocre, but not awful.

# significant amt of NAs, so we need to remove them before finding the correlation:
clean_dom_acc_trips_join <- dom_acc_trips_join[complete.cases(dom_acc_trips_join[c("total_guests", "trips_overnights")]), ] #filtering the dom_acc_guests data frame to keep only the rows where both "total_guests" and "hotel_guests" have non-missing values

# now lets try to find the correlation:
cor(clean_dom_acc_trips_join$total_guests, clean_dom_acc_trips_join$trips_overnights, method = "kendall") # 0.8060229, not bad at all! This means that there is a positive correlation between the two variables (cor is relatively close to 1).  Correlation measures the strength of association between two variables and the direction of the relationship: so in this case, it is positive and they have an association.
# v2025 corr is only 0.5025098. still decent, but worse thann v2024.
sd(clean_dom_acc_trips_join$total_guests) # 59897.74 guests
sd(clean_dom_acc_trips_join$trips_overnights) # 73703.34 guests, larger standard deviation than total_guests

# go ahead and do raw gapfilling in the next chunk using this data.
```

The above analysis has shown that there's not a strong enough relationship to use a linear model to gapfill here. We'll instead gapfill with the raw values. 
Let's gf!

```{r}
# -------- gf using trips_overnights for NA values of tourism_arrivals_ct -----
dom_acc_guests_gf2 <- dom_acc_trips_join %>% 
  group_by(rgn_id, year) %>% # group by county and year
  mutate(
    tourism_arrivals_ct = case_when(
      is.na(tourism_arrivals_ct) & !is.na(trips_overnights) ~ trips_overnights, # gapfill, when there is no data on total guests or hotel guests, fill with trips_overnights visitors
      TRUE ~ tourism_arrivals_ct # otherwise, NA
    ),
    arrivals_method = ifelse(is.na(total_guests) & is.na(hotel_guests) & !is.na(trips_overnights), "Trips:Overnights", arrivals_method),
    arrivals_gapfilled = ifelse(arrivals_method == "Trips:Overnights", "gapfilled", arrivals_gapfilled) # fill the "gapfilled" column to indicate "gapfilled" or NA
    ) %>% 
  ungroup() # ungroup since not needed anymore

# take a look at how that changed the number of NAs:
summary(dom_acc_guests_gf2) # now, `tourism_arrivals_ct` has 2277 NAs
colSums(is.na(dom_acc_guests_gf2)) # now, `tourism_arrivals_ct` has 2277 NAs

dom_acc_guests_gf2_noNA <- dom_acc_guests_gf2 %>% 
  filter(!is.na(tourism_arrivals_ct)) # remove NAs to see how many regions have total_guests values

length(unique(dom_acc_guests_gf2_noNA$rgn_id)) # we now have 122 unique regions,that's better! Let's move on and see if we can gapfill even further!
```

## Mean Stay as a multiplier to convert Total:Overnights into Total:Guests

So, there are some instances in which there is data for Total:Overnights, but not Total:Guests. Based on looking at the methodology (<https://www.e-unwto.org/doi/epdf/10.18111/9789284424160?role=tab>), it seems that guests are the individuals coming to a specific place, versus overnights includes the same people over again if they stay for multiple nights.

Essentially, Guests = Overnights / Mean Stay (in days).

So, we need to find mean stays! As a reminder, this would need to be calculated using the raw `total_guests` and `total_overnights` variables.

We can do this by **calculating `mean_stay` for each region**, if possible. Using an lm, see if `mean_stay` could be correlated to the georegion (r2 ideally because it is more specific, but r1 works as well). If so, take the mean for that georegion, and use it as a way to convert Accommodation:Total:Overnights into Accommodation:Total:Guests.

```{r lm_mean_stay}
dom_mean_stay <- dom_acc_guests_gf2 %>% 
  group_by(rgn_id, rgn_name, year) %>% 
  mutate(mean_stay = total_overnights/total_guests) # by referring to UNWTO documentation

range(dom_mean_stay$mean_stay, na.rm = TRUE) # 0.9666667 8.4101144
# v2025:  0.3581081 65.0432432
sd(dom_mean_stay$mean_stay, na.rm = TRUE) # 1.234238
# v2025: 4.466338

# re-pasting the code as a reminder of what the object `georegions` is from
georegions <- ohicore::georegions %>% 
  left_join(rgns_eez, by = "rgn_id") %>% 
  select(-c(Notes)) # not needed

# join mean_stay df with georegions! so we can see if there can be an association by georgn...
dom_mean_stay_georgns <- dom_mean_stay %>% 
  left_join(georegions, by = c("rgn_id","rgn_name"))

mod3 <- lm(mean_stay ~ r2, data = dom_mean_stay_georgns)
summary(mod3)
# not significant model :(, so we have to move forward with the more general and less specific r1 groupings

mod4 <- lm(mean_stay ~ r1, data = dom_mean_stay_georgns)
summary(mod4)
# not significant model!! Multiple R-squared is 0.01365..... NOT GOOD.
```

```{r}
# lets do a quick plot to compare mean stays and r1, just to be sure that we cant use georegions to estimate mean_stay:
# plot the data to compare
ggplot(dom_mean_stay_georgns, aes(x = as.numeric(r1), y = mean_stay)) +
  geom_point(alpha = 0.5) +  # Add some transparency to points
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  theme_minimal() +  # Use a minimal theme
  labs(title = "Georegions vs Mean Stay",
       x = "Georegion (r1)",
       y = "Mean Stay (Days)") # yikesss!!

# log scale
ggplot(dom_mean_stay_georgns, aes(x = log(r1 + 1), y = log(mean_stay + 1))) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  theme_minimal() +
  labs(title = "Georegions vs Mean Stay",
       x = "Log(Georegion (r1) + 1)",
       y = "Log(Mean Stay (Days) + 1)") # yup, definitely no correlation! lets check numerically though

# let's look at their correlation. see ?`cor` if you are interested in learning more about how this is calculated (`cor` computes the variance of x and the covariance or correlation of x and y).  Kendall correlation is preferred when there are small samples or some outliers (https://datascience.stackexchange.com/questions/64260/pearson-vs-spearman-vs-kendall)
range(dom_mean_stay_georgns$r1, na.rm = TRUE) #   2 419, interesting to know
sum(is.na(dom_mean_stay_georgns$mean_stay)) # 3829 NAs
sum(is.na(dom_mean_stay_georgns$r1)) # 27 NAs: high seas regions
mean(!is.na(dom_mean_stay_georgns$mean_stay)) # 0.1706736, aka 17% NA/total obs
mean(!is.na(dom_mean_stay_georgns$r1)) # 0.994152, aka 99% NA/total obs, great!

# significant amt of NAs, so we need to remove them before finding the correlation:
clean_dom_mean_stay_georgns <- dom_mean_stay_georgns %>% 
  filter(!is.na(mean_stay)) #filtering the dom_acc_guests data frame to keep only the rows where mean_stay has non-missing values

# now lets try to find the correlation:
cor(clean_dom_mean_stay_georgns$mean_stay, clean_dom_mean_stay_georgns$r1, method = "kendall") # 0.08158826, VERY LITTLE correlation... to be expected.
```

We should not move forward using georegions to group by and take the mean. Instead, we should either use the global mean to gapfill for mean_stays, or consider using a moving average.

What I will do for now is take the mean by region, assuming that there are some instances in which there may be some years that have overnights and guest values, but in other years it could have only overnights values. In that case, it would make the most sense to use the value found within the region if possible.

For regions that only have overnights data, we will use the global mean of mean_stays as our value to divide Overnights by.

**As a reminder:**

Mean Stay = Overnights/Guests (to see how many nights they stayed on average)

Therefore, if we have Overnights, we can find Guests by dividing by the Mean Stay. (Guests = Overnights / Mean Stay)

```{r mean_stay_multiplier}
dom_mean_stay_georgns_gf <- dom_mean_stay_georgns %>%
  group_by(rgn_id, rgn_name) %>% 
  mutate(
    mean_stay_filled = mean(mean_stay, na.rm = TRUE),
    mean_stay_filled = if_else(is.nan(mean_stay_filled), NA, mean_stay_filled)
    )

# global mean multiplier (technically, divider lol)
global_mean_stay <- mean(dom_mean_stay_georgns$mean_stay, na.rm = TRUE)
global_mean_stay
# v2024: 2.893876 days
# v2025: 5.492028 days (is this too high? need to check)

dom_mean_stay_georgns_filled <- dom_mean_stay_georgns_gf %>%
  mutate(
    mean_stay_filled = ifelse(is.na(mean_stay_filled), global_mean_stay, mean_stay_filled)
    ) # if the mean_stay was already found regionally, leave it, otherwise replace with global mean_stay value
```

```{r}
# ------ use mean_stay to gapfill values --------
dom_acc_guests_gf3 <- dom_mean_stay_georgns_filled %>% 
  mutate(
    tourism_arrivals_ct = case_when(
      is.na(tourism_arrivals_ct) & !is.na(total_overnights) ~ total_overnights/mean_stay_filled,
      TRUE ~ tourism_arrivals_ct
    ),
    arrivals_method = ifelse(is.na(total_guests) & !is.na(total_overnights), "Total:Overnights / Mean Stay", arrivals_method),
    arrivals_gapfilled = ifelse(arrivals_method == "Total:Overnights / Mean Stay", "gapfilled", arrivals_gapfilled) # fill the "gapfilled" column to indicate "gapfilled" or NA
  ) %>%  # it worked!! Angola 1998 is a good example of where it may have worked.  There are others.
  select(rgn_id, rgn_name, year, total_guests, hotel_overnights, tourism_arrivals_ct, arrivals_method, arrivals_gapfilled) # only select columns needed for future analysis

# take a look at how that changed the number of NAs:
summary(dom_acc_guests_gf3) # now, `tourism_arrivals_ct` has 1993 NAs
colSums(is.na(dom_acc_guests_gf3))

dom_acc_guests_gf3_noNA <- dom_acc_guests_gf3 %>% 
  filter(!is.na(tourism_arrivals_ct)) # remove NAs to see how many regions have total_guests values

length(unique(dom_acc_guests_gf3_noNA$rgn_name)) # we now have 136
```

## Convert Accommodations:Hotel:Overnights into Acc:Total:Guests

As a reminder, within the "Clean Accommodations Data" section we did a linear model between total_guests and hotel_overnights, and found that it was a significant model with a good multiple r-squared. Therefore, we are going to move forward and use the significant intercept and slope to estimate total_guests. Let's first do some exploring:

```{r}
# --------- lm between total_guests ~ hotel_overnights --------------
mod5 <- lm(total_guests ~ hotel_overnights, data = dom_acc_guests_agg)
summary(mod5)
 # p-value: < 0.00000000000000022, multiple R squared is 0.8971, not bad!! intercept and slope is significant, at -1477.876236 and 0.694437, respectively.

# plot the variables against each other to see their relationship, see if the model runs through the data appropriately
mod_plot2 <- ggplot(dom_acc_guests_agg, aes(y = total_guests, x = hotel_overnights)) +
  geom_point(alpha = 0.5) + 
  geom_abline(intercept = mod5$coefficients[1], 
              slope = mod5$coefficients[2], 
              color = "red", linetype = "dashed") +
  theme_minimal() + 
  labs(title = "Hotel Overnights vs Total Guests, Accomodations data",
       x = "Hotel Overnights",
       y = "Total Guests");ggplotly(mod_plot2) 

# it makes sense that the hotel overnights have a larger amount of individuals, since it includes mean stay (it is hotel_guests*mean_stay_hotel).  In this case, because hotel_overnights is like two/three steps removed from the total_guests unit, let's see if we can use the lm to find values for total_guests from hotel_overnights, instead of recalculating a mean_stay for hotel_guests specifically and hoping that the resulting hotel_guest values could pass for total_guest values.
```

Let's move forward, using the intercept and slope to determine total_guests values from hotel_overnights!

```{r}
# first, remove any regions that only have NAs. Then use region and georegions to get lm coefficients
dom_gf4_coef <- dom_acc_guests_gf3 %>% 
  left_join(georegions, by = c("rgn_id")) %>% 
  group_by(rgn_id) %>% # group by region
  filter(if_all(c(total_guests, hotel_overnights), ~!is.na(.))) %>% # remove rows where there are NAs for both columns
  mutate(
    total_guests_slope = lm(total_guests ~ 0 + hotel_overnights)$coefficient[1]
  ) %>% # obtain the intercept and slope for each region that has data
  ungroup() %>% 
  group_by(r2) %>% # get values for r1 and r2, so that if regional is available, use that... then if r2 is available, use that... finally, use r1 if you can't otherwise.
  mutate(
    total_guests_slope_r2 = lm(total_guests ~ 0 + hotel_overnights)$coefficient[1]
  ) %>% 
  ungroup() %>% 
  group_by(r1) %>% 
  mutate(
    total_guests_slope_r1 = lm(total_guests ~ 0 + hotel_overnights)$coefficient[1]
  ) %>% 
  ungroup() %>% 
  distinct(rgn_id, r2, r1, total_guests_slope, total_guests_slope_r2, total_guests_slope_r1) %>% # prep for left join, otherwise many-to-many
  select(rgn_id, r2, r1, total_guests_slope, total_guests_slope_r2, total_guests_slope_r1) # select the columns that will be needed
```

```{r}
# ------- gapfill using hotel_overnights when total_guests = NA ---------
# will need to gapfill these slopes and intercepts for regions that have hotel_overnights data but no total_guests data. join with georegions, group by r2, and take the mean for those regions.  Otherwise, use regional level.
dom_acc_guests_gf4 <- dom_acc_guests_gf3 %>%
  left_join(georegions, by = c("rgn_id","rgn_name")) %>% 
  left_join(dom_gf4_coef, by = c("rgn_id","r2","r1")) %>%
  group_by(r2) %>% # get values for r1 and r2 for regions that do not have data, so that if regional is available, use that... then if r2 is available, use that... finally, use r1 if you can't otherwise.
  mutate(
    total_guests_slope_r2mean = mean(total_guests_slope_r2, na.rm = TRUE),
    total_guests_slope_r2mean = ifelse(is.nan(total_guests_slope_r2mean), NA, total_guests_slope_r2mean) # any NaN values to NAs
  ) %>%
  ungroup() %>%
  group_by(r1) %>%
  mutate(
    total_guests_slope_r1mean = mean(total_guests_slope_r1, na.rm = TRUE),
    total_guests_slope_r1mean = ifelse(is.nan(total_guests_slope_r1mean), NA, total_guests_slope_r1mean) # any NaN values to NAs
  ) %>%
  ungroup() %>%  # now we have intercept and slope values for all regions (at different levels), lets use that to gapfill total_guests with hotel_overnights!
  mutate(
    arrivals_method = ifelse(is.na(total_guests) & !is.na(hotel_overnights), "Hotel:Overnights", arrivals_method), # do this first so that we can use when tourism_arrivals_ct is NA as a qualifier
    tourism_arrivals_ct = case_when(
      is.na(tourism_arrivals_ct) & !is.na(hotel_overnights) & !is.na(total_guests_slope) ~ (hotel_overnights*total_guests_slope),
      is.na(tourism_arrivals_ct) & !is.na(hotel_overnights) & is.na(total_guests_slope) & !is.na(total_guests_slope_r2mean) ~ (hotel_overnights*total_guests_slope_r2mean), # if regional lm coefficients weren't possible, go to r2 which is second least specific, then r1
      is.na(tourism_arrivals_ct) & !is.na(hotel_overnights) & is.na(total_guests_slope) & is.na(total_guests_slope_r2mean) & !is.na(total_guests_slope_r1mean) ~ (hotel_overnights*total_guests_slope_r1mean),
      TRUE ~ tourism_arrivals_ct
    ),
    arrivals_gapfilled = ifelse(arrivals_method == "Hotel:Overnights", "gapfilled", arrivals_gapfilled)
  ) %>% # example: 	New Caledonia and Monaco.  The numbers make sense because the estimated total_guests were larger than hotel_guests (makes sense because total includes other establishments than just hotels) and we can compare to hotel_overnights, of which tourism_arrivals_ct is about half ish, which based on our knowledge of mean_stays (found later on in raw gf for dom) the average is between 2-3 nights.  So this makes sense!
  select(rgn_id, rgn_name, year, tourism_arrivals_ct, arrivals_method, arrivals_gapfilled) # select all that we need moving forward
```

Now that we can gapfilled using this method, lets see how it changed our data:

```{r}
summary(dom_acc_guests_gf4) # we now have 2688 out of 4617 observations.  This is now 58% filled!
 # v2024                         
 # tourism_arrivals_ct
 #  NA's   :2688     
colSums(is.na(dom_acc_guests_gf4))
# v2025: 1619


dom_acc_guests_gf4_noNA <- dom_acc_guests_gf4 %>% 
    filter(!is.na(tourism_arrivals_ct)) # remove NAs to see how many regions have total_guests values

length(unique(dom_acc_guests_gf4_noNA$rgn_name)) # 150 unique regions, that's great!!! 

# Let's see how many are left if we filter to 2005-2021, which is our years of interest.
dom_acc_guests_filt <- dom_acc_guests_gf4_noNA %>% 
  filter(year %in% data_years)

length(unique(dom_acc_guests_filt$rgn_name)) # 139 unique regions, NOT BAD!!! Let's move forward :) 
```

### Filter years and regions without enough data

```{r}
# filter domestic data for usable years to be used in the future. Ensure that it is between 2005 - 2021, and that there are at least three raw data values!
domestic_guests_rgns <- dom_acc_guests_gf4 %>% 
  filter(year %in% data_years) %>% 
  group_by(rgn_id) %>% 
  summarize(count_values = sum(!is.na(tourism_arrivals_ct))) %>% # for each region, how many values in tourism_arrivals_ct did it have
  filter(count_values >= 3) # keep only the regions that have greater than or equal to three values

length(unique(domestic_guests_rgns$rgn_id)) # v2024: 87 rgns
# v2025: 134 regions

domestic_guests <- dom_acc_guests_gf4 %>% 
  filter(rgn_id %in% domestic_guests_rgns$rgn_id) %>%  # filter only for regions that are in that time frame and have at least three values
  filter(year %in% data_years)
  
length(unique(domestic_guests$rgn_id)) # 134

colSums(is.na(domestic_guests)) # 339 tourism_arrivals_ct
nrow(domestic_guests) # 2412
# ~ 14% NAs

write_csv(domestic_guests, here(current_tr_dir, "int", "domestic_guests_first_gf.csv"))
```

# Gapfilling for International and Domestic Aggregation

Here is the second big gapfilling step in the script, which is the gapfilling of prepped international and domestic tourism arrivals data, for regions that are missing values for certain years.

## International - for years that do not have values

```{r}
# gapfill for international arrivals!
# downfill then upfill missing values using a linear model of the average increase per years across all years of data for 1995-2019
# for 2020 and 2021, use the global average proportion increase or decrease and add to the previous years value

# reminder: 
# `tourism_arrivals_ct` is the tourism arrivals count from the INTERNATIONAL data, which consists of overnights, and was gapfilled with (total_arrivals - same_day).

international_arrivals <- read_csv(here(current_tr_dir, "int", "international_arrivals_int.csv"))

# ---- Gapfill 1995-2019 ----
pre_pandemic_int <-  international_arrivals %>%
  # Adjust to match you want to consider 'pandemic years'
  filter(!(year %in% c(2020, 2021, 2022))) %>%
  group_by(rgn_id, rgn_name) %>%
  arrange(year) %>%
  # Linear interpolation
  mutate(tourism_arrivals_ct = na.approx(tourism_arrivals_ct, x = year, na.rm = FALSE)) %>% 
  mutate(
    arrivals_gapfilled = ifelse(is.na(arrivals_method) & !is.na(tourism_arrivals_ct), "gapfilled", arrivals_gapfilled),
    arrivals_method = ifelse(is.na(arrivals_method) & !is.na(tourism_arrivals_ct), "Interpolation", arrivals_method)
  ) %>% 
  # Upfilling # fill in any values that are empty from early years with values from the nearest year. Doing this because doesn't make sense to add earlier years based on a trend 
  tidyr::fill(tourism_arrivals_ct, .direction = "up") %>% 
  mutate(
    arrivals_gapfilled = ifelse(is.na(arrivals_method) & !is.na(tourism_arrivals_ct), "gapfilled", arrivals_gapfilled),
    arrivals_method = ifelse(is.na(arrivals_method) & !is.na(tourism_arrivals_ct), "Upfilling", arrivals_method)
         ) 

# ==== gapfill for 2020 and 2021 now ==== 
int21_22 <- international_arrivals %>%
  # Pandemic years
  filter(year %in% c(2020, 2021, 2022)) %>%
  pivot_wider(names_from = year, 
              values_from = tourism_arrivals_ct,
              names_prefix = "int_tourism_") %>%
  # 2022 compared to 2021
  mutate(tourism_ct_diff_2022 = (int_tourism_2022 - int_tourism_2021)/int_tourism_2021,
  # 2021 compared to 2020
         tourism_ct_diff_2021 = (int_tourism_2021 - int_tourism_2020)/int_tourism_2020)

# global average change for 2021 tourism_arrivals_ct
gf_2021_tourism_int <- mean(int21_22$tourism_ct_diff_2021, na.rm = TRUE) 
gf_2021_tourism_int
# v2024 and v2025: 0.1290837

# global average change for 2022 tourism_arrivals_ct
gf_2022_tourism_int <- mean(int21_22$tourism_ct_diff_2022, na.rm = TRUE) 
gf_2022_tourism_int
# v2025: 4.407749 

# ok, lets use these values to gapfill 2021 if it is NA and 2020 exists. So increase by X proportion (0.129)
int2020 <- international_arrivals %>%
  filter(year %in% c(2019, 2020)) %>%
  filter(!is.na(tourism_arrivals_ct)) %>%
  pivot_wider(names_from = year, 
              values_from = tourism_arrivals_ct,
              names_prefix = "int_tourism_") %>%
  mutate(tourism_ct_diff = (int_tourism_2020 - int_tourism_2019)/int_tourism_2019)

# global average change for 2020 toursism
gf_2020_tourism_int <- mean(int2020$tourism_ct_diff, na.rm = TRUE)
gf_2020_tourism_int
# v2024 and v2025: -0.7118338

# ok, lets use these values to gapfill 2020 if it is NA and 2019 exists. So decrease by X proportion ~70%

pandemic_int <- international_arrivals %>%
  filter(year > 2019) 

int_arrivals_gf <- pre_pandemic_int %>%
  complete(year = 2005:max(data_years)) %>% 
  rbind(pandemic_int) %>%
  group_by(rgn_id, rgn_name) %>%
  arrange(rgn_id, year) %>%
  # apply global average proportional increase or decrease for 2020 and 2021, because of covid pandemic messing up trends...
  mutate(tourism_arrivals_ct = ifelse(year == 2020 & is.na(tourism_arrivals_ct), lag(tourism_arrivals_ct, n = 1) + lag(tourism_arrivals_ct, n = 1)*gf_2020_tourism_int, tourism_arrivals_ct)) %>%
  mutate(tourism_arrivals_ct = ifelse(year == 2021 & is.na(tourism_arrivals_ct), lag(tourism_arrivals_ct, n = 1) + lag(tourism_arrivals_ct, n = 1)*gf_2021_tourism_int, tourism_arrivals_ct)) %>%
  mutate(tourism_arrivals_ct = ifelse(year == 2022 & is.na(tourism_arrivals_ct), lag(tourism_arrivals_ct, n = 1) + lag(tourism_arrivals_ct, n = 1)*gf_2022_tourism_int, tourism_arrivals_ct)) %>%
  mutate(arrivals_method = ifelse(is.na(arrivals_method) & !is.na(tourism_arrivals_ct), "2020:2022 method", arrivals_method)) %>%
  mutate(arrivals_gapfilled = ifelse(arrivals_method == "2020:2022 method", "gapfilled", arrivals_gapfilled)) %>%
  filter(year >= 2008) %>% # get only the year we need and beyond
  drop_na(tourism_arrivals_ct) %>%  # remove any remaining NAs (any remaining have all NAs for that region)
   mutate(year = as.numeric(year)) # ensure it is numeric

length(unique(int_arrivals_gf$rgn_name)) # 157 regions after gapfilling! good, same as `international_arrivals`

summary(int_arrivals_gf) 
table(int_arrivals_gf$arrivals_method) 

colSums(is.na(int_arrivals_gf)) # no more missing values
```

## Domestic - for years that do not have values

```{r}
# gapfill for domestic arrivals!
# downfill then upfill missing values using a linear model of the average increase per years across all years of data for 1995-2019
# for 2020 and 2021, use the global average proportion increase or decrease and add to the previous years value

# reminder: 
# `tourism_arrivals_ct` is the tourism arrivals count from the DOMESTIC data, which is primarily total guests (accommodation) data, and was gapfilled using domestic trip data also from UNWTO.

domestic_guests <- read_csv(here(current_tr_dir, "int", "domestic_guests_first_gf.csv"))
colnames(domestic_guests)

# ---- Gapfill 1995-2019 ----
pre_pandemic_dom <-  domestic_guests %>%
  # Adjust to match you want to consider 'pandemic years'
  filter(!(year %in% c(2020, 2021, 2022))) %>%
  group_by(rgn_id, rgn_name) %>%
  arrange(year) %>%
  # Linear interpolation
  mutate(tourism_arrivals_ct = na.approx(tourism_arrivals_ct, x = year, na.rm = FALSE)) %>% 
  mutate(
    arrivals_gapfilled = ifelse(is.na(arrivals_method) & !is.na(tourism_arrivals_ct), "gapfilled", arrivals_gapfilled),
    arrivals_method = ifelse(is.na(arrivals_method) & !is.na(tourism_arrivals_ct), "Interpolation", arrivals_method)
  ) %>% 
  # Upfilling # fill in any values that are empty from early years with values from the nearest year. Doing this because doesn't make sense to add earlier years based on a trend 
  tidyr::fill(tourism_arrivals_ct, .direction = "up") %>% 
  mutate(
    arrivals_gapfilled = ifelse(is.na(arrivals_method) & !is.na(tourism_arrivals_ct), "gapfilled", arrivals_gapfilled),
    arrivals_method = ifelse(is.na(arrivals_method) & !is.na(tourism_arrivals_ct), "Upfilling", arrivals_method)
         ) 


# ==== gapfill for 2020 and 2021 now ==== 

dom21_22 <- domestic_guests %>%
  filter(year %in% c(2020, 2021, 2022)) %>%
  filter(!is.na(tourism_arrivals_ct)) %>%
  pivot_wider(names_from = year, 
              values_from = tourism_arrivals_ct,
              names_prefix = "int_tourism_") %>%
  mutate(tourism_arrivals_ct_diff_2022 = (int_tourism_2022 - int_tourism_2021)/int_tourism_2021,
         tourism_arrivals_ct_diff_2021 = (int_tourism_2021 - int_tourism_2020)/int_tourism_2020)

# global average increase for 2021 tourism_arrivals_ct
gf_2021_tourism_dom <- mean(dom21_22$tourism_arrivals_ct_diff_2021, na.rm = TRUE)
gf_2021_tourism_dom
# v2024: 0.2361595
# v2025: 0.3409155

gf_2022_tourism_dom <- mean(dom21_22$tourism_arrivals_ct_diff_2022, na.rm = TRUE)
gf_2022_tourism_dom
# v2025: 2.24985

# ok, lets use these values to gapfill 2021 if it is NA and 2020 exists. So increase by X proportion (0.294)

dom_2020 <- domestic_guests %>%
  filter(year %in% c(2019, 2020)) %>%
  filter(!is.na(tourism_arrivals_ct)) %>%
  pivot_wider(names_from = year, 
              values_from = tourism_arrivals_ct,
              names_prefix = "int_tourism_") %>%
  mutate(tourism_arrivals_ct_diff = (int_tourism_2020 - int_tourism_2019)/int_tourism_2019)
 
# global average decrease for 2020 toursism
gf_2020_tourism_dom <- mean(dom_2020$tourism_arrivals_ct_diff, na.rm = TRUE) 
gf_2020_tourism_dom
# v2024: -0.2984019
# v2025: -0.6833755

# ok, lets use these values to gapfill 2020 if it is NA and 2019 exists. So decrease by X proportion ~35%

pandemic_dom <- domestic_guests %>%
  filter(year > 2019) 

dom_arrivals_gf <- pre_pandemic_dom %>%
  rbind(pandemic_dom) %>%
  group_by(rgn_id, rgn_name) %>%
  arrange(rgn_id, year) %>%
  # apply global average proportional increase or decrease for 2020 and 2021, because of covid pandemic messing up trends...
  mutate(tourism_arrivals_ct = ifelse(year == 2020 & is.na(tourism_arrivals_ct), lag(tourism_arrivals_ct, n = 1) + lag(tourism_arrivals_ct, n = 1)*gf_2020_tourism_dom, tourism_arrivals_ct)) %>%
  mutate(tourism_arrivals_ct = ifelse(year == 2021 & is.na(tourism_arrivals_ct), lag(tourism_arrivals_ct, n = 1) + lag(tourism_arrivals_ct, n = 1)*gf_2021_tourism_dom, tourism_arrivals_ct)) %>%
  mutate(tourism_arrivals_ct = ifelse(year == 2022 & is.na(tourism_arrivals_ct), lag(tourism_arrivals_ct, n = 1) + lag(tourism_arrivals_ct, n = 1)*gf_2022_tourism_dom, tourism_arrivals_ct)) %>%
  # record gapfilling method
  mutate(arrivals_method = ifelse(is.na(arrivals_method) & !is.na(tourism_arrivals_ct), "2020:2022 method", arrivals_method)) %>%
  mutate(arrivals_gapfilled = ifelse(arrivals_method == "2020:2022 method", "gapfilled", arrivals_gapfilled)) %>%
  filter(year >= 2008) %>% # get only the year we need and beyond
  drop_na(tourism_arrivals_ct) %>%  # remove any remaining NAs (any remaining have all NAs for that region)
   mutate(year = as.numeric(year)) # ensure it is numeric

length(unique(dom_arrivals_gf$rgn_name)) # v2025: 134

summary(dom_arrivals_gf) 
colSums(is.na(dom_arrivals_gf)) # 0 NAs
table(dom_arrivals_gf$arrivals_method)
```


```{r}
dom_missing <- dom_arrivals_gf %>% 
  filter(rgn_id %in% missing$rgn_id & year %in% missing$year)

int_missing <- int_arrivals_gf %>% 
  filter(rgn_id %in% missing$rgn_id & year %in% missing$year)

view(int_missing)
view(dom_missing)

colSums(is.na(int_missing))
colSums(is.na(dom_missing))
```

## Gapfilling -- for regions that do not have domestic data: prep for aggregation.

Let's move on, as we decided that it makes sense to aggregate the international and domestic data! This is because we are creating a score for each region, and only comparing across regions once that score has been calculated (versus ranking regions and comparing them to each other based on tourism density, not the rescaled score).

We decided that there would be less error after aggregating if we at least put a mean (by georegion) domestic guest value for regions that have no domestic data, but they have international data. That way, all international regions will be adding to a domestic value, rather than some regions only having international data.

```{r}
# first, let's join international and domestic data to see where the gaps are!
## aggregating international and domestic data by region and year
total_arrivals <- left_join(int_arrivals_gf, dom_arrivals_gf, by = c("rgn_id","rgn_name","year")) %>% 
  rename(
    tourism_arrivals_ct_int = tourism_arrivals_ct.x,
    tourism_arrivals_ct_dom = tourism_arrivals_ct.y,
    data_source = arrivals_method.x,
    arrivals_gf = arrivals_gapfilled.x,
    arrivals_method = arrivals_method.y,
    arrivals_gapfilled = arrivals_gapfilled.y
  )

length(unique(total_arrivals$rgn_name)) # v2025: 157 regions
colSums(is.na(total_arrivals))
# > colSums(is.na(total_arrivals))
#                  rgn_id                rgn_name 
#                       0                       0 
#                    year tourism_arrivals_ct_int 
#                       0                       0 
#             data_source             arrivals_gf 
#                       0                    2073 
# tourism_arrivals_ct_dom         arrivals_method 
#                     600                     600 
#      arrivals_gapfilled 
#                     600 
```

Now that we have a joined dataframe, let's adjust it to gapfill by years.

```{r}
# join with georegions again, and take the mean of the domestic guests data for r2 and r1 groups
view(mean_domestic_guests)
mean_domestic_guests <- total_arrivals %>% 
  group_by(rgn_id, rgn_name) %>% 
  complete(year = 2008:max(data_years)) %>% # ensures that there is a row for every year for each region, to ensure there are no gaps
  left_join(georegions, by = c("rgn_id")) %>%
  ungroup() %>%
  group_by(r2) %>%
  mutate(
    # take the mean of domestic arrivals by georegion
    tourism_arrivals_ct_dom_r2 = mean(tourism_arrivals_ct_dom, na.rm = TRUE),
    tourism_arrivals_ct_dom_r2 = ifelse(is.nan(tourism_arrivals_ct_dom_r2), NA, tourism_arrivals_ct_dom_r2), # to make NaN = NAs
    # take the mean of international arrivals by georegion
    tourism_arrivals_ct_int_r2 = mean(tourism_arrivals_ct_int, na.rm = TRUE),
    tourism_arrivals_ct_int_r2 = ifelse(is.nan(tourism_arrivals_ct_int_r2), NA, tourism_arrivals_ct_dom_r2) # to make NaN = NAs
  ) %>%
  ungroup() %>%
  group_by(r1) %>%
  mutate(
    tourism_arrivals_ct_dom_r1 = mean(tourism_arrivals_ct_dom, na.rm = TRUE),
    tourism_arrivals_ct_dom_r1 = ifelse(is.nan(tourism_arrivals_ct_dom_r1), NA, tourism_arrivals_ct_dom_r1),
    tourism_arrivals_ct_int_r1 = mean(tourism_arrivals_ct_int, na.rm = TRUE),
    tourism_arrivals_ct_int_r1 = ifelse(is.nan(tourism_arrivals_ct_int_r1), NA, tourism_arrivals_ct_dom_r1)
  ) %>%
  ungroup()

length(unique(mean_domestic_guests$rgn_name)) # 157 yay!
colSums(is.na(mean_domestic_guests))

length(unique(georegions$r2)) # 22 r2 regions
length(unique(mean_domestic_guests$tourism_arrivals_ct_dom_r2)) # 22, got all r2 regions

length(unique(georegions$r1)) # 7 regions
length(unique(mean_domestic_guests$tourism_arrivals_ct_dom_r1)) # 7, that means it got all the r1 regions

write_csv(mean_domestic_guests, here(current_tr_dir, "int", "mean_domestic_guests.csv"))
```

```{r}
mean_domestic_guests <- read_csv(here(current_tr_dir, "int", "mean_domestic_guests.csv"))

mean_domestic_guests %>% 
  filter(rgn_id %in% missing$rgn_id & year %in% missing$year) %>% 
  view()
```

# Calculating Arrival Proportion

## Aggregating International and Domestic

```{r}
# ----- use the mean domestic data to fill for any domestic rows that need guest values, then aggregate! -----

# Read in the data
mean_domestic_guests <- read_csv(here(current_tr_dir, "int", "mean_domestic_guests.csv"))

arrivals_aggregated <- mean_domestic_guests %>% 
  mutate(
    tourism_arrivals_ct_dom = case_when(
      is.na(tourism_arrivals_ct_dom) & !is.na(tourism_arrivals_ct_dom_r2) ~ tourism_arrivals_ct_dom_r2,
      is.na(tourism_arrivals_ct_dom) & is.na(tourism_arrivals_ct_dom_r2) & !is.na(tourism_arrivals_ct_dom_r1) ~ tourism_arrivals_ct_dom_r1,
      TRUE ~ tourism_arrivals_ct_dom
    ),
    tourism_arrivals_ct_int = case_when(
      is.na(tourism_arrivals_ct_int) & !is.na(tourism_arrivals_ct_int_r2) ~ tourism_arrivals_ct_int_r2,
      is.na(tourism_arrivals_ct_int) & is.na(tourism_arrivals_ct_int_r2) & !is.na(tourism_arrivals_ct_int_r1) ~ tourism_arrivals_ct_int_r1,
      TRUE ~ tourism_arrivals_ct_int
    ),
    tourism_arrivals_ag = case_when(
      !is.na(tourism_arrivals_ct_int) & !is.na(tourism_arrivals_ct_dom) ~ tourism_arrivals_ct_int + tourism_arrivals_ct_dom,
      is.na(tourism_arrivals_ct_int) ~ tourism_arrivals_ct_dom,
      is.na(tourism_arrivals_ct_dom) ~ tourism_arrivals_ct_int,
      TRUE ~ tourism_arrivals_ct_int
    ),
    ag_method = case_when(
      !is.na(tourism_arrivals_ct_int) & !is.na(tourism_arrivals_ct_dom) ~ "Aggregated international and domestic",
      is.na(tourism_arrivals_ct_int) ~ "International",
      is.na(tourism_arrivals_ct_dom) ~ "Domestic",
      TRUE ~ data_source
    ),
    ag_gapfilled = case_when(
      arrivals_gf == "gapfilled" & arrivals_gapfilled == "gapfilled" ~ "gapfilled - both",
      arrivals_gf == "gapfilled" & is.na(arrivals_gapfilled) ~ "gapfilled - international",
      is.na(arrivals_gf) & arrivals_gapfilled == "gapfilled" ~ "gapfilled - domestic",
      TRUE ~ arrivals_gf
    )
  )

length(!is.na(arrivals_aggregated$tourism_arrivals_ct_int)) # 2184 v2025: 2705
length(!is.na(arrivals_aggregated$tourism_arrivals_ct_dom)) # 2184 v2025: 2705
# yay! now we can aggregate international and domestic with the least amount of error with the data we have
colSums(is.na(arrivals_aggregated))
nrow(arrivals_aggregated)

# save gf information
tourism_aggregated_gf <- arrivals_aggregated %>% 
  rename(
    int_method = data_source,
    int_gapfilled = arrivals_gf,
    dom_method = arrivals_method,
    dom_gapfilled = arrivals_gapfilled
  ) %>% 
  select(rgn_id, rgn_name, year, int_method, int_gapfilled, dom_method, dom_gapfilled, ag_method, ag_gapfilled)

write_csv(tourism_aggregated_gf, here::here(current_tr_dir, "int", "tourism_aggregated_gf.csv"))
```

## Coastal Proportion

Now, lets bring in our tourism data and calculate the coastal proportion! The coastal proportion will then be multiplied by the arrivals value per region and year. Finally, the coastal area will be divided, giving us a density of tourism per region (individuals/km\^2).

```{r, eval=FALSE}
# ------- coastal population data -----------
# In 2024, the coastal population data was fixed! This is great.  We will now use `globalprep/mar_prs_population/v2024/output/mar_pop_25mi.csv`.
coastal_pop_data <- read_csv(here("globalprep","mar_prs_population","v2025","output","mar_pop_25mi.csv")) ## read in coastal population data from other data layer

# Checking out the coastal population data, the only year with missing population is 2023. We'll gapfill 2023 with the 2022 value

coastal_pop_data_fill <- coastal_pop_data %>%
  filter(year == 2022) %>%
  mutate(year = 2023) %>%
  rbind(coastal_pop_data) # add 2023 data year, just using year 2022


length(unique(coastal_pop_data_fill$rgn_id)) # v2024 and v2025: 220 regions available! Any region without data is uninhabited.
```

```{r, eval=FALSE}
# ------- total global population data -----------
# bring in World Bank Global population data 
total_population_wb <- read_csv(here(raw_data_dir, "WorldBank", current_data_dir,
                                     # UPDATE FILE NAME TO MATCH
                                     "WorldBank_global_country_population_1960-2024",
                                     "API_SP.POP.TOTL_DS2_en_csv_v2_38144.csv"), 
                                skip = 4) %>% # check how many rows to skip in excel
  pivot_longer(cols = 5:(ncol(.) - 1), names_to = "year",
               values_to = "total_pop") %>% 
  janitor::clean_names()



# give ohi rgns to world bank data
total_population_rgns <- name_2_rgn(df_in = total_population_wb, 
                                        fld_name = 'country_name',
                                        keep_fld_name = TRUE)

# this shows rows that have at least one duplicate, including both the first and subsequent occurrences of each duplicated row. using the tidyverse allows us to see all instances of duplicated data, not just the second and subsequent occurrences (which is what just a duplicated() call would give you).
duplicates_wb <- total_population_rgns %>%
  group_by(rgn_id, rgn_name, year) %>% 
  filter(n() > 1) %>%
  ungroup()
unique(duplicates_wb$rgn_name)
# [1] "China"                                              
# [2] "Northern Mariana Islands and Guam"                  

# drop Macao and Hong Kong, since we only need mainland China
# Northern Mariana Islands and Guam, Puerto Rico and Virgin Islands of the United States: aggregate!
total_population_final <- total_population_rgns %>%
  filter(country_name != "Hong Kong SAR, China" & country_name != "Macao SAR, China") %>% # only keep mainland, not city specific data
  group_by(rgn_id, rgn_name, year) %>% 
  summarise(
    total_pop = sum(total_pop, na.rm = TRUE), # aggregate by rgn and year
    .groups = "drop" # ungroup
  ) %>% 
  mutate(year = as.numeric(year)) 


length(unique(total_population_final$rgn_name)) # v2024: 166 regions available of total population data!
length(unique(arrivals_aggregated$rgn_name)) # 157
length(union(arrivals_aggregated$rgn_name, total_population_final$rgn_name)) # v2025: 177, so 10 regions do not have total population data
length(intersect(arrivals_aggregated$rgn_name, total_population_final$rgn_name)) # v2025: 146, only 146 regions have the same name
```


## Coastal Area

Now, lets bring in the $$km^2$$ of coastal area for all OHI regions by aggregating the area protected 1 kilometer inland, and 3 nautical miles out into the eez.

```{r, eval=FALSE}
# ------- coastal area data -----------
# read in area of coastline data
inland_filepath <- here::here("globalprep", "lsp", version_year, "output", "rgn_area_inland1km.csv")
inland_data <- read_csv(inland_filepath)
offshore_filepath <- here::here("globalprep", "lsp", version_year, "output", "rgn_area_offshore3nm.csv")
offshore_data <- read_csv(offshore_filepath)


# get combined value of inland and offshore for each ohi region
inland_offshore <- inland_data %>%
  left_join(offshore_data, by = join_by(rgn_id)) %>%
  select(rgn_id, area.x, area.y) %>%
  group_by(rgn_id) %>%
  mutate(total_inland_offshore_area = area.x + area.y) %>%
  select(rgn_id, total_inland_offshore_area)
```

# Arrival Proportion (AP)

Now we can join all of these data sources and conduct $$\frac{TourismArrivals * (CoastalPop/TotalPop)}{CoastalArea}$$.

```{r, eval=FALSE}
# ============= join the tourism arrival, coastal population, and total population data =================== #

arrivals_prop_df <- left_join(arrivals_aggregated, total_population_final, by = c("rgn_id","rgn_name","year")) %>% 
  mutate(year = as.numeric(year)) %>% 
  left_join(., coastal_pop_data_fill, by = c("rgn_id","year")) %>% 
  left_join(., inland_offshore, by = c("rgn_id")) %>% 
  filter(year %in% 2000:2022) %>% # to keep consistent with the coastal pop data
  mutate(coastal_prop = (popsum/total_pop)) %>% # values that are greater than 1 are because of inconsistencies between World Bank and the coastal population data.  However, those with a value greater than 1 are mostly islands or small regions with a large coastal area.  In that case, we could say 100% of the tourism could be coastally related.
  mutate(coastal_prop = ifelse(coastal_prop > 1, 1, coastal_prop)) # for rgns and years in which the total_population is greater than 1, we cap it at 1 

ap_final <- arrivals_prop_df %>% 
  mutate(Ap = (tourism_arrivals_ag * coastal_prop)/(total_inland_offshore_area))  %>% 
  drop_na(Ap) %>% # regions that did not have tourism arrival values but were present in the coastal population data
  select(rgn_id, rgn_name, year, Ap) # this final `Ap` (arrivals proportion) is in the units of (tourists/km2).  Aka, a higher proportion indicates that there is a larger number of tourists per km2 of coastal area, and there is therefore a higher tourism density.  This number is now ready to be divided by `Sr`, or the sustainability of tourism.

# save the Ap proportions dataframe
write_csv(ap_final, here::here(current_tr_dir, "int", "ap.csv"))

# # save which observations were gapfilled and how into output -- from v2023 version
tr_arrivals_props_tourism_gf <- arrivals_prop_df %>%
  select(rgn_id, rgn_name, year, data_source, arrivals_gf)

write_csv(tr_arrivals_props_tourism_gf, here::here(current_tr_dir, "output", "tr_arrivals_props_tourism_gf.csv"))
```

```{r}
# interactive plot of the arrival proportions before rescaling
ap_plot <- ap_final %>%
  plot_ly(x = ~year, y = ~Ap, color = ~rgn_name, 
          type = "scatter", mode = "lines") %>%
  layout(title = "Global Arrival proportions (Ap) by Region",
         xaxis = list(title = "Year"),
         yaxis = list(title = "Arrival proportion (Ap)"))
ap_plot

htmlwidgets::saveWidget(ap_plot, here::here(current_tr_dir, "figs", "ap_interactive_plot.html")) # save plot
```

## Rescale Ap proportion using the 90th quantile

As according to the [OHI methods](https://ohi-science.org/ohi-methods/goals/goal-models-data.html#tourism-and-recreation), the Ap values for each year and region is rescaled using the 90th quartile for each region.

$$\frac{T_r}{T_{90th}}$$

```{r}
# to make the Ap scaled between 0 and 1, we can divide the current Ap by the 90th quartile values by region.
ap_rescaled <- ap_final %>% 
  group_by(rgn_id, rgn_name) %>% 
  summarize(value_90th_percentile = quantile(Ap, 0.9, na.rm = TRUE))

tr_arrivals_props_tourism <- left_join(ap_final, ap_rescaled, by = c("rgn_id", "rgn_name")) %>% 
  mutate(
    Ap = Ap/value_90th_percentile,
    Ap = if_else(Ap < 0, 0, Ap), # if less than 1, that is due to gapfilling where linear models drew the tourism arrival values less than 0.  
    Ap = if_else(Ap > 1, 1, Ap) # if more than 1, then the Ap value for that year was above the 90th percentile.  So the value is adjusted to 1
         ) %>% 
  select(rgn_id, rgn_name, year, Ap) # keep in rgn_name when you want to plot!!!

# interactive plot of the arrival proportions after rescaling
ap_final_plot <- tr_arrivals_props_tourism %>%
  plot_ly(x = ~year, y = ~Ap, color = ~rgn_name, 
          type = "scatter", mode = "lines") %>%
  layout(title = "Global Arrival proportions (Ap) by Region, Rescaled",
         xaxis = list(title = "Year"),
         yaxis = list(title = "Arrival proportion (Ap)"))
ap_final_plot

htmlwidgets::saveWidget(ap_final_plot, here::here(current_tr_dir, "figs", "ap_rescaled.html")) # save plot

# save the final tourism arrivals csv
write_csv(tr_arrivals_props_tourism, here::here(current_tr_dir, "output", "tr_arrivals_props_tourism.csv"))

colSums(is.na(tr_arrivals_props_tourism))

test <- tr_arrivals_props_tourism %>% 
  filter(rgn_id %in% missing$rgn_id & year %in% missing$year)
view(test)

int_missing <- int_arrivals_gf %>% 
  filter(rgn_id %in% missing$rgn_id & year %in% missing$year)

```

# Data check

```{r}
# Read in data
new_tr <- read_csv(here(current_tr_dir, "output", "tr_arrivals_props_tourism.csv"))
old_tr <- read_csv(here(previous_tr_dir, "output", "tr_arrivals_props_tourism.csv"))

na_countries <- new_tr %>% 
  filter(rgn_id %in% c(67, 190, 77, 53, 197, 47, 73, 207, 43, 66, 139, 168, 
                       194, 49, 100, 9, 173, 181, 193, 6, 8, 98, 99, 151))
colnames(old_na_countries) 


old_na_countries <- old_tr %>% 
  filter(rgn_id %in% c(67, 190, 77, 53, 197, 47, 73, 207, 43, 66, 139, 168, 
                       194, 49, 100, 9, 173, 181, 193, 6, 8, 98, 99, 151))

na_diff <- old_na_countries %>% 
  rename(old_Ap = Ap) %>% 
  left_join(na_countries, by = join_by(rgn_id, year))
view(na_diff)

missing <- na_diff %>% 
  filter(is.na(Ap))

# Get max data years for plotting
current_data_year <- max(new_tr$year)
current_data_year
previous_data_year <- max(old_tr$year)
previous_data_year

# Combine data frames
comb_tr <- new_tr %>% 
  rename(new_Ap = Ap) %>% 
  left_join(old_tr, by = join_by(rgn_id, year))

# Plot
comp_plot1 <- ggplot() + 
  geom_jitter(data = filter(comb_tr, year == previous_data_year),
              aes(x = Ap, y = new_Ap, 
                  text = paste("Year:", year, "<br>Region:", rgn_name)),
              alpha = 0.7) +
  labs(x = paste("Previous Value", previous_data_year),
       y = paste("Current Value", previous_data_year),
       title = paste("Sustainable tourism and recreation", previous_data_year, "Data", previous_year, "Assessment vs.", previous_data_year, "Data", current_year, "assessment")) +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  theme_bw();ggplotly(comp_plot1)
```


