---
title: 'OHI `r format(Sys.Date(), "%Y")` - Tourism and Recreation '
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html' 
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

[REFERENCE RMD FILE](http://ohi-science.org/ohiprep_v2024/globalprep/tr/v2024/tr_data_prep.html)

# Summary

This document describes the steps for calculating tourism and recreation sustainability.

-   Sr = (S-1)/5; Sustainability of tourism

## Updates

v2025: 
- Split script into two separate pieces
- Cleaned up the data sources and introduction because it was very unclear which files were needed previously
- Changed gapfilling method #4 under raw gapfilling from a poorly-fit linear model to a linear interpolation using the `zoo` package

> This goal is still fairly messy and could use a lot of tightening up in future assessments. v2025 did an in-depth check of the gapfilling methods, the results of which can be found in [issue #355](https://github.com/OHI-Science/globalfellows-issues/issues/355) for more details. 


# Data Sources

**New data sources**

-   Tourism sustainability: World Economic Forum. The Travel & Tourism Development Index 2024 (version May 2024). 2024. [TTDI](https://www3.weforum.org/docs/WEF_Travel_and_Tourism_Development_Index_2024.pdf)
- GDP per capita, World bank: This is the main dataset used to measure GDP. It is used to gapfill missing values in Tourism sustainability
- GDP per capita, CIA: This is a secondary dataset used to fill gaps in the World Bank data. (<https://www.cia.gov/the-world-factbook/field/real-gdp-per-capita/country-comparison/>) Click the download button on the website, top right hand corner.

**Data sources from other goals**

- `ao` World Bank GDP per capita purchasing power parity: `globalprep/ao/vYEAR/intermediate/gdppcppp_ohi.csv`

**Citation**: World Bank. (2024). *Population, total*. https://data.worldbank.org/indicator/SP.POP.TOTL © 2024 The World Bank Group, All Rights Reserved.

**Methodology:**

Ts: Sustainability Index -- TTDI

-   Data is from the World Economic Forum's "Travel and Tourism Development Index" (<https://www.weforum.org/publications/travel-tourism-development-index-2024/>)
-   2023 scores are from the index. World Bank GDP per capita ppp adj from `/ohiprep_v2024/globalprep/ao/v2024/intermediate/gdppcppp_ohi.csv` and substituted with [CIA GDP per capita ppp adj](https://www.cia.gov/the-world-factbook/field/real-gdp-per-capita/country-comparison) is used to gapfill scores into previous years (2005 - 2022).


# Setup

```{r setup, message=FALSE, warning=FALSE, results="hide"}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, eval=FALSE)

if (!require(librarian)) {
  install.packages("librarian")
  library(librarian)
}

# Load libraries
librarian::shelf(
  ohicore, #devtools::install_github('ohi-science/ohicore@dev')
  here,
  tidyverse,
  WDI,
  readxl,
  naniar,
  countrycode,
  pdftools,
  plotly,
)

# ---- sources! ----
source(here("workflow", "R", "common.R")) # file creates objects to process data

region_data() # for rgns_all and rgns_eez

regions_shape() # returns spatial shape object named 'regions' which includes land, eez, highseas, and antarctica regions

#source(here(paste0("globalprep/tr/v", version_year, "/R/tr_fxns.R"))) # not used presently

# ---- set year and file path info ----
current_year <- 2025 # Update this in the future!!
previous_year <- 2024

version_year <- paste0("v",current_year)
previous_vyear <- paste0("v", previous_year)

current_data_dir <- paste0("d", current_year)
previous_data_dir <- paste0("d", previous_year)

data_years <- 2005:2022 # update with latest data year (probably just add 1)

# ---- data directories ----

# Raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "git-annex", "globalprep", "_raw_data")

# UNWTO (UN World Tourism) raw data directory
current_unwto_dir <- here(raw_data_dir, "UNWTO", current_data_dir)
previous_unwto_dir <- here(raw_data_dir, "UNWTO", previous_data_dir)

# Tourism and recreation directory
current_tr_dir <- here("globalprep", "tr", version_year)
previous_tr_dir <- here("globalprep", "tr", previous_vyear)
```


# Ts: Sustainability Index -- TTDI

These data are from the World Economic Forum's "Travel and Tourism Development Index" (<https://www.weforum.org/publications/travel-tourism-development-index-2024/>) Downloaded 08/20/2024.

See mazu: \_raw_data/WEF-Economics/ for more details and the raw data.

The TTDI was formerly the TTCI which was a similar index, but unfortunately not comparable. v2024's updated TTDI uses data up through 2023 to calculate their sustainability scores.

These data are gapfilled using gdppcppp and UN georegion information (see next section for obtaining and preparing these data).

```{r WEF processing, eval=FALSE}
# bring in TTDI pdf! Download it to your computer, place it in Mazu
ttdi_file <- "WEF_Travel_and_Tourism_Development_Index_2024.pdf" # the file name
ttdi_raw <- pdf_text(here::here(raw_data_dir, "WEF-Economics", previous_data_dir, ttdi_file)) # using the package `pdftools`, read in using pdf_text()

sr_scores_table <- ttdi_raw[[11]] %>% # page number of the pdf that has the data we want (aka the sustainability scores)
  as.data.frame() # make the page as a data frame.  Be warned, the data is all there! You just can't see it yet.

columns = c("rank","economy","score","rank_change","score_change","diff_avg","rank_2","economy_2","score_2","rank_change_2","score_change_2","diff_avg_2", "rank_3","economy_3","score_3","rank_change_3","score_change_3","diff_avg_3") # mirror the pdf and create clean column names

sr_scores <- sr_scores_table %>% 
  separate_rows('.', sep = "\n") %>%  # separate the rows: \n separates by moving down to the next line without returning to the beginning of the line
  rename(main_column = ".") %>% # rename the only column as main_column instead of just the period
  slice(16:55) %>% # choose the rows that have the table in them (disregard the page title etc)
  separate_wider_position(
    main_column, # the issue here is that there is not a common delimiter, and the three columns blur together. So, we can use `separate_wider_position()`, which allows you to instead define column breaks by the number of characters, which includes spaces
    widths = c(rank = 2, economy = 24, score = 4, rank_change = 3, score_change = 5, diff_avg = 19,
               rank_2 = 7, economy_2 = 24, score_2 = 5, rank_change_2 = 3, score_change_2 = 7, diff_avg_2 = 19,
               rank_3 = 9, economy_3 = 26, score_3 = 8, rank_change_3 = 1, score_change_3 = 1, diff_avg_3 = 1), # these numbers were found through trial and error, but are always at least the length of the longest character string
    names_sep = NULL, # default, to avoid renaming the column names
    names_repair = "check_unique", #the default), no name repair, but checks they are unique
    too_many = "debug", # will add additional columns to the output to help you locate and resolve the underlying problem
    too_few = "debug", # adds additional columns to the output to help you locate and resolve the underlying problem... these will not remain in final code
    cols_remove = FALSE # always FALSE if too_few or too_many are set to "debug"
  ) %>% 
  select(rank, economy, score,  rank_2, economy_2, score_2, rank_3, economy_3, score_3) %>% # to get rid of debugging columns and rankings
  mutate(
    economy = str_trim(economy, side = "both"), # remove the whitespace before and after the string
    rank_2 = str_trim(rank_2, side = "both"),
    economy_3 = str_trim(economy_3, side = "both"),
    rank_3 = str_trim(rank_3, side = "both"),
    score_3 = str_trim(score_3, side = "both"),
    economy_3 = str_replace(economy_3, pattern = "Trinidad and Tobago 3.5", replacement = "Trinidad and Tobago"), # this line specifically had issues with spacing, and no matter what it always had bleeding from the score.  Therefore, the economy name and score were fixed using str_replace.
    score_3 = str_replace(score_3, pattern = "2 -5", replacement = "3.52") 
    )

# now, we have a dataframe that contains three columns we would like to make into one. let's subset and create three dfs that each have the same name (they couldnt have the same name beforehand because they were in the same df), then rbind them together to have one df of all regions and their scores.
sr_scores_sub1 <- sr_scores %>% 
  select(rank, economy, score)

sr_scores_sub2 <- sr_scores %>% 
  select(rank_2, economy_2, score_2) %>% 
  rename(
    rank = rank_2,
    economy = economy_2,
    score = score_2
  )

sr_scores_sub3 <- sr_scores %>% 
  select(rank_3, economy_3, score_3) %>% 
    rename(
    rank = rank_3,
    economy = economy_3,
    score = score_3
  )

# rbind them together and clean up -- we do not need the column "rank", it was just preserved until now as an easy way to ensure a clean bind
sr_scores_clean <- rbind(sr_scores_sub1, sr_scores_sub2, sr_scores_sub3) %>% 
  select(-rank) %>% # remove rank
  filter(!is.na(score)) %>%  # remove NA, leaves 119 regions, good
  mutate(
    S_score = as.numeric(score)) # ensure it is numeric 


# run name_2_gn so each economy receives an OHI rgn name and id  
ttdi_rgn <- name_2_rgn(df_in = sr_scores_clean, 
                       fld_name = 'economy',
                       keep_fld_name = TRUE) %>% # v2024: no duplicates! looks good
  select(rgn_id, rgn_name, S_score)

### Save TTDI data file
write_csv(ttdi_rgn, here(current_tr_dir, "int","wef_ttdi.csv")) # save to int folder
```

## Preparing the gdppcppp data:

These data are used to gapfill missing values in tourism sustainability. Most of the data are from the World Bank, but CIA data fill some gaps (CIA data is available for only the most recent year).

The Artisanal Opportunities goal uses gdppcppp data, so we will get the data that was processed for that goal.

```{r worldbank, eval=FALSE}
wb <- read_csv(here::here("globalprep","ao", version_year, "intermediate","gdppcppp_ohi.csv")) %>%
  dplyr::select(rgn_id, year, value)
```

CIA data are used to fill in missing gaps in the gdppcppp data (<https://www.cia.gov/the-world-factbook/field/real-gdp-per-capita/country-comparison>)

Downloaded: 09/22/2024

See README on the raw folder for instructions on how to download this data. Place in Mazu.

The following code is used to prepare these data for OHI:

```{r cia gdp, eval=FALSE}

cia_gdp <- read_csv(here::here(raw_data_dir, "CIA", current_data_dir, "Real_GDP_per_capita.csv")) %>% 
  # remove dollar signs and commas and convert to numeric
  mutate(value = as.numeric(gsub("[$,]", "", value))) %>% 
  select(name, value) %>% 
  rename(country = name, pcgdp_cia = value)

 ## Data reported in a lower resolution than OHI regions
splits <- data.frame(country = "Saint Helena, Ascension, and Tristan da Cunha", 
                     country2 = c("Saint Helena", "Ascension","Tristan da Cunha"))

cia_gdp_splits <- cia_gdp %>%
  left_join(splits, by='country') %>%
  mutate(country2 = ifelse(is.na(country2), country, country2)) %>%
  select(country = country2, pcgdp_cia)

cia_gdp_rgn <- name_2_rgn(df_in = cia_gdp_splits, 
                       fld_name='country')

### Duplicated regions: Collapse regions after weighting by population (regions we include as a single region) - 

population_weights <- data.frame(country = c("Virgin Islands", "Puerto Rico",
                                             "China", "Hong Kong", "Macau",
                                             "Guam", "Northern Mariana Islands"),
                                 # from world bank - updated v2025 using 2024 data from `total_population_wb` read in earlier during aggregation of tourism data
                                 population = c(104377, 3203295, 1408975000,
                                                7524100, 687000, 167777, 44278))

cia_gdp_rgn_clean <- cia_gdp_rgn %>%
  left_join(population_weights, by="country") %>%
  mutate(population = ifelse(is.na(population), 1, population)) %>%
  group_by(rgn_id) %>%
  summarize(pcgdp_cia = weighted.mean(pcgdp_cia, population)) %>%
  ungroup() %>%
  filter(rgn_id <= 250) %>%
  select(rgn_id, pcgdp_cia)

# save the cia data in intermediate folder 
write_csv(cia_gdp_rgn_clean, here::here(current_tr_dir, "int", "wb_rgn_cia_GDPPCPPP.csv"))
```

The following code combines the two gdp datasets and gapfills missing regions using UN georegions.

If there is no World Bank gdppcppp data (pcgdp), the CIA data is used (pcgdp_cia). The pcgdp2 variable includes both the World Bank and CIA data (with CIA data only used if there is not World Bank data). The remaining data are estimated using UN geopolitical regions. Ideally, the mean gdppcppp value is calculated at the r2 scale (gdp_pred_r2) using regions within each class with gdppcppp data. If there were not enough regions with data at the r2 scale, the average at the r1 scale was used (gdp_pred_r1). The gdp_all variable combines all estimates using the following hierarchy: World Bank -\> CIA -\> estimated using mean from r2 UN geopolitical regions -\> estimated using mean from r1 UN geopolitical regions.

```{r gapfill gdp, eval=FALSE}
### world bank gdp data
gdppcppp <- wb %>%
  select(rgn_id, year, pcgdp = value)

### cia gdp data
gdppcppp2 <- read_csv(here(current_tr_dir, "int", "wb_rgn_cia_GDPPCPPP.csv"))

### Use WB data, but if missing, use pcgdp_cia.
### combine with UN georegion data
years <- data.frame(year = min(gdppcppp$year):max(gdppcppp$year))

# bring in georegions and join with their labels
georegions <- ohicore::georegions

regions <- georegions %>%
  left_join(georegion_labels, by = 'rgn_id')

# merge the df "regions" with years, so there is one year per region from 2005 to 2023. Then join the wb and cia data
gdp_raw <- merge(years, regions, by=NULL) %>%
  left_join(gdppcppp, by = c("rgn_id", "year")) %>%
  left_join(gdppcppp2, by = c("rgn_id")) 

## quick compare to make sure the CIA and World Bank data are compatible
plot(gdp_raw$pcgdp[gdp_raw$year==2023], gdp_raw$pcgdp_cia[gdp_raw$year==2023])
abline(0,1, col="red")
# a few minor outliers but overall looks good!

gdp_replace_raw <- gdp_raw %>%
  mutate(pcgdp2 = ifelse(is.na(pcgdp), pcgdp_cia, pcgdp)) %>% # create an object to replace WB data with CIA data if necessary
  select(-c(pcgdp, pcgdp_cia))

## Calculating the means across different geopolitical levels (e.g. r2, r1)
gdp_means_georegions <- gdp_replace_raw %>%
  group_by(r2, year) %>%
  mutate(gdp_pred_r2 = mean(pcgdp2, na.rm=TRUE)) %>%
  ungroup() %>%
  group_by(r1, year) %>%
  mutate(gdp_pred_r1 = mean(pcgdp2, na.rm=TRUE)) %>%
  ungroup() 

#if we dont have gdp data from the world bank or cia, then replace with the r2 mean.  If it still does not have data, replace with the r1 mean.  
gdp_georegions_gf <- gdp_means_georegions %>%
  mutate(gdp_all = ifelse(is.na(pcgdp2), gdp_pred_r2, pcgdp2)) %>%
  mutate(gdp_all = ifelse(is.na(gdp_all), gdp_pred_r1, gdp_all)) %>%
  mutate(gapfilled = ifelse(is.na(pcgdp2) & !is.na(gdp_all), "gapfilled", NA)) %>%
  mutate(method = ifelse(is.na(pcgdp2) & !is.na(gdp_pred_r2), "UN georegion (r2)", NA)) %>%
  mutate(method = ifelse(is.na(pcgdp2) & is.na(gdp_pred_r2) & !is.na(gdp_pred_r1), "UN georegion (r1)", method)) 

write_csv(gdp_georegions_gf, here(current_tr_dir, "int", "gdp_raw_gf.csv"))

gdp_data_gf <- gdp_georegions_gf %>%
  select(rgn_id, year, gapfilled, method) 

write_csv(gdp_data_gf, here(current_tr_dir, "int", "gdp_gf.csv"))

gdp_data <- gdp_georegions_gf %>%
  select(rgn_id, year, pcgdp = gdp_all)

write_csv(gdp_data, here(current_tr_dir, "int", "gdp.csv"))
```

The final step is gapfilling the Sustainability data using an ANCOVA model with gdppcppp and UN geopolitical regions as predictor variables.

```{r, eval=FALSE}
# tr sustainability raw TTDI scores
sust <- read_csv(here(current_tr_dir, "int", "wef_ttdi.csv"))

# bring in tourism arrivals data to see which regions need gapfilled S_scores
ap_gf <- read_csv(here(current_tr_dir, "output", "tr_arrivals_props_tourism.csv")) %>%
  # filter(year == 2021) %>%
  select(rgn_id, Ap, year) %>%
  filter(!is.na(Ap)) %>% 
  mutate(year = as.numeric(year))

# gdp dataframe prepared above (World Bank, CIA, and gapfilled gdp data)
gdp_raw_gf <- read_csv(here(current_tr_dir, "int", "gdp_raw_gf.csv")) %>% 
  # filter(year == 2021) %>%
  select(rgn_id, r0_label, r1_label, r2_label, rgn_label, pcgdp2, gdp_all, year) %>% 
  mutate(year = as.numeric(year))

tr_sust <- gdp_raw_gf %>%
  left_join(sust, by = c("rgn_id")) %>%
  left_join(ap_gf, by = c("rgn_id", "year"))
  # filter(rgn_id != 213)

### Add gapfill flag variable 
## Reminder:
## pcgdp2: includes both the World Bank and CIA data (with CIA data only used if there is not World Bank data)
## Ap: Proportion of tourism arrivals (international and domestic when available)
## S_score: tourism sustainability score

# specify whether each S_score will be gapfilled and with what method
tr_sust_gf <- tr_sust %>%
  mutate(gapfilled = ifelse(is.na(S_score) & !is.na(Ap), "gapfilled", NA)) %>%
  mutate(method = ifelse(is.na(S_score) & !is.na(Ap) & is.na(pcgdp2), "lm georegion + gdppcppp, with est. gdppcppp", NA)) %>%
  mutate(method = ifelse(is.na(S_score) & !is.na(Ap) & !is.na(pcgdp2), "lm georegion + gdppcppp", method)) %>%
  select(rgn_id, gapfilled, method, year)

write_csv(tr_sust_gf, here(current_tr_dir, "output", "tr_sustainability_gf.csv"))
```

## Gapfilling

ANOVA model using gdppcppp (GDP per capita, PPP) and UN geopolitical regions (as factors) as predictor variables. However if there is no gdppc data we estimate the gdppc using the UN georegions and then used in the linear model to gapfill the sustainability score.

```{r, eval=FALSE}

### Gapfill S using r1 and/or r2 regional data and PPP-adjusted per-capita GDP
### Looked at models with a year variable, but wasn't significant and decided to exclude

mod3 <- lm(S_score ~ as.factor(r2_label) + gdp_all, data=tr_sust, na.action = na.exclude)
summary(mod3)
anova(mod3)

mod4 <- lm(S_score ~ as.factor(r1_label) + gdp_all, data=tr_sust, na.action = na.exclude)
summary(mod4)
anova(mod4)

plot(predict(mod3), tr_sust$S_score)
abline(0,1)
plot(predict(mod4), tr_sust$S_score)
abline(0,1)


## Estimate missing data and gapfill
# Some of the r1 levels do not have data and consequently causes a fail. This chunk of code drops these levels so an NA is returned

# Select only r2 column
new_data <- tr_sust %>% 
  dplyr::select(r2_label, gdp_all)

# see what regions are 
unique(tr_sust$r2_label)

r2_w_data <- unique(tr_sust$r2_label[!is.na(tr_sust$S_score)]) # only the r2s that dont have NAs for the sustainability score
  
new_data_r2 <- new_data %>%
  mutate(r2_label = ifelse(r2_label %in% r2_w_data, r2_label, NA))

# Predict sustainability scores using linear model 3 (using r2 data)
tr_sust_mod3 <- tr_sust %>% 
  dplyr::mutate(s_score_pred_r2 = predict(mod3, newdata = new_data_r2))


# Select only r1 column
new_data <- tr_sust %>% 
  dplyr::select(r1_label, gdp_all)

unique(tr_sust$r1_label)

r1_w_data <- unique(tr_sust$r1_label[!is.na(tr_sust$S_score)])

new_data_r1 <- new_data %>%
  mutate(r1_label = ifelse(r1_label %in% r1_w_data, r1_label, NA))

# Predict sustainability scores using linear model 4 (using r1 data)
tr_sust_mod4 <- tr_sust_mod3 %>% 
  dplyr::mutate(s_score_pred_r1 = predict(mod4, newdata = new_data_r1))


## some are missing the r1 predictions, but none of these have Ap scores, so not relevant
View(filter(tr_sust_mod4, is.na(s_score_pred_r1)))

tr_sust_final <- tr_sust_mod4 %>%
  mutate(s_score_2 = ifelse(is.na(S_score), s_score_pred_r2, S_score)) %>%
  mutate(s_score_2 = ifelse(is.na(s_score_2), s_score_pred_r1, s_score_2)) %>%
  # filter(year %in% c(2019, 2021)) %>%
  select(rgn_id, year, S_score=s_score_2)

summary(tr_sust_final)

write_csv(tr_sust_final, here(current_tr_dir, "output", "tr_sustainability.csv"))
```


# Data check

```{r}
# Read in data
new_tr <- read_csv(here(current_tr_dir, "output", "tr_sustainability.csv"))
old_tr <- read_csv(here(previous_tr_dir, "output", "tr_sustainability.csv"))

# Get max data years for plotting
current_data_year <- max(new_tr$year)
current_data_year
previous_data_year <- max(old_tr$year)
previous_data_year

# Get rgn names for ease of viewing
rgn_name <- rgns_eez %>% 
  select(rgn_name, rgn_id)

# Combine data frames
comb_tr <- new_tr %>% 
  rename(new_S_score = S_score) %>% 
  left_join(old_tr, by = join_by(rgn_id, year)) %>% 
  left_join(rgn_name, by = join_by(rgn_id))

# Plot
comp_plot1 <- ggplot() + 
  geom_jitter(data = filter(comb_tr, year == previous_data_year), aes(x = S_score, y = new_S_score, text = paste("Year:", year, "<br>Region:", rgn_name)), alpha = 0.7) +
  labs(x = paste("Previous Value", previous_data_year),
       y = paste("Current Value", previous_data_year),
       title = paste("Sustainable tourism and recreation", previous_data_year, "Data", previous_year, "Assessment vs.", previous_data_year, "Data", current_year, "assessment")) +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  theme_bw();ggplotly(comp_plot1)
```

