---
title: 'OHI 2023: Food Provision'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

# Summary

This script combines the fisheries catch data with the mariculture production data to create the weights for how much of each score will affect the entire food provision score. 

# Updates from previous assessment

- Mariculture data was updated since last scenario year but fisheries wasn't
- Updated scenario year input/some file paths to be more programmatic
- Updated labeling of the graphs

# Methods

```{r, eval = FALSE}
## load libraries
library(dplyr)
library(tidyr)
library(here)

scen_year <- 2023 # change to latest year
prev_scen_year <- scen_year - 1

setwd(here::here("globalprep","fp", paste0("v", scen_year)))

## Load FAO-specific user-defined functions
source('http://ohi-science.org/ohiprep_v2021/workflow/R/common.R')
```


# Import Output Data: Mariculture data

Mariculture production in tonnes.
```{r, eval = FALSE}
mar <- read.csv(file.path("..", "..", "mar", paste0("v", scen_year), "output", "MAR_FP_data.csv")) # see metadata in its prep
```

Fisheries data.
```{r, eval = FALSE}
fis <- read.csv(file.path("..", "..", "fis", paste0("v", scen_year), "output", "FP_fis_catch.csv")) %>%
  dplyr::select(rgn_id, year, fis_t = fis_catch) # see metadata in its prep
```

# Wrangle 

## Tidy MAR data
```{r, eval = FALSE}
mar_tidy <- mar %>%
  group_by(rgn_id, year) %>%
  summarize(mar_t = sum(value, na.rm=TRUE)) %>%
  dplyr::select(rgn_id, year, mar_t) %>%
  ungroup()

# this one is turning to NA in FP
filter(mar_tidy, rgn_id == 95) # ok, this makes sense
```

## Tidy FIS data

```{r, eval = FALSE}
fis_tidy <- fis %>%
  mutate(fis_t = ifelse(fis_t == 0, NA, fis_t)) %>%  # 11 NA values is correct (there were 0 in v2023)
  group_by(rgn_id) %>%
  arrange(year) %>%
  fill(fis_t) %>% 
  ungroup()
```

## Combine MAR and FIS

Adjust years so they are equivalent.
```{r, eval = FALSE}
adjust <- max(mar_tidy$year) - max(fis_tidy$year)

mar_adjust <- mar_tidy %>%
  mutate(year = year - adjust)

tmp <- full_join(fis_tidy, mar_adjust, by = c('rgn_id', 'year')) # v2023: removed all = TRUE because was not a valid argument, did not see an equivalent

## If NA, turn it into a 0 before weighting
tmp_weights <- tmp %>%
  mutate(fis_t = ifelse(is.na(fis_t), 0, fis_t)) %>%
  mutate(mar_t = ifelse(is.na(mar_t), 0, mar_t)) %>%
  mutate(w_fis = fis_t/(fis_t + mar_t)) %>%
  mutate(w_fis = ifelse(mar_t == 0 & fis_t == 0, NA, w_fis)) %>%
  filter(year >= 2005) %>%
  dplyr::select(rgn_id, year, w_fis) 

hist(tmp_weights$w_fis,
     main = "Weights",
     xlab = "Value")
```

# Data check

Compare to previous year data
```{r, eval = FALSE}
compare <- read.csv(paste0("../../fp/v", prev_scen_year, "/output/wildcaught_weight.csv")) %>%
  rename(w_fis_old = w_fis) %>%
  left_join(tmp_weights, by = c('rgn_id', 'year'))
plot(compare$w_fis_old, compare$w_fis,
     xlab = paste0("Old data (v", prev_scen_year, ")"),
     ylab = paste0("New data (v", scen_year, ")"))
abline(0, 1, col="red")
```

# Save data

```{r, eval = FALSE}
write.csv(tmp_weights, 'output/wildcaught_weight.csv', row.names=FALSE)

## add gf file (no gapfilling)
tmp_gf <- tmp_weights %>%
  mutate(w_fis = 0) %>%
  dplyr::select(rgn_id, year, gapfilled = w_fis)

write.csv(tmp_gf, 'output/wildcaught_weight_gf.csv', row.names=FALSE)
```

