---
title: 'OHI 2019: Coral extent'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

# Summary

This script generates the extent of coral for each OHI region. 


## Updates from previous assessment
Creating an actual script to calculate this. This has not been updated since 2012. Updating the data with newest version (version 4). 

***
## Data Source 

**Downloaded**: 07/25/2019

**Description**:  
Global Distribution of Coral Reefs
https://data.unep-wcmc.org/datasets/1

This dataset shows the global distribution of coral reefs in tropical and subtropical regions. It is the most comprehensive global dataset of warm-water coral reefs to date, acting as a foundation baseline map for future, more detailed, work. This dataset was compiled from a number of sources by UNEP World Conservation Monitoring Centre (UNEP-WCMC) and the WorldFish Centre, in collaboration with WRI (World Resources Institute) and TNC (The Nature Conservancy). Data sources include the Millennium Coral Reef Mapping Project (IMaRS-USF and IRD 2005, IMaRS-USF 2005) and the World Atlas of Coral Reefs (Spalding et al. 2001).

**Time range**: 1954-2018


***
# Methods
Reclassify the coral extent data into a mask of 1 or NA.


## Setup
``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(readr)      # for read_csv()
library(raster)
library(here)
library(sf)
library(fasterize)
library(tidyverse)

source(here('workflow/R/common.R'))

dir_wcmc <- file.path(file.path(dir_M, 'git-annex/globalprep/_raw_data/wcmc_coral'))
ohi_rasters() # call the region zones raster

zones_high_res <- disaggregate(zones, fact=c(2,2),
                               filename= file.path(dir_M, "git-annex/globalprep/hab_coral/v2019/int/high_res_zones.tif"),
                               progress="text")

zones_high_res <-raster(file.path(dir_M, "git-annex/globalprep/hab_coral/v2019/int/high_res_zones.tif"))
```

```{r, echo = FALSE, eval = FALSE}

habitat_extent_coral_old <- read_csv(here("globalprep/hab_coral/v2012/data/habitat_extent_coral_updated.csv")) %>% 
  dplyr::select(rgn_id, old_km2=km2)
  

v4_coral_pts <- sf::st_read(dsn = file.path(dir_wcmc, "14_001_WCMC008_CoralReefs2018_v4/01_Data"), layer = "WCMC008_CoralReef2018_Pt_v4")
dim(v4_coral_pts)

v4_coral_py <- sf::st_read(dsn = file.path(dir_wcmc, "14_001_WCMC008_CoralReefs2018_v4/01_Data"), layer = "WCMC008_CoralReef2018_Py_v4")
dim(v4_coral_py)
```

**Convert coral shapefiles into same CRS as our region zones raster**
```{r, echo = FALSE, eval = FALSE}

moll_crs <- crs(zones, asText = TRUE)

v4_coral_pts_moll <- st_transform(v4_coral_pts, crs = moll_crs) #project points shapefile to match zones crs


crs(v4_coral_pts_moll) #check to see it worked

v4_coral_py_moll <- st_transform(v4_coral_py, crs = moll_crs)

crs(v4_coral_py_moll)
plot(v4_coral_py_moll[1])
plot(zones)
plot(v4_coral_pts_moll[1])
```

**Fasterize/rasterize: Where there is coral assign a value of 1 and NA otherwise**
```{r, echo = FALSE, eval = FALSE}

#create a column full of "1's"
v4_coral_pts_moll$raster <- 1
#rasterize our points data to raster 

raster::rasterize(v4_coral_pts_moll, 
                                   zones_high_res,
                                   "raster",
                                   fun = min,
 filename = file.path(dir_M, "git-annex/globalprep/hab_coral/v2019/int/pt_coral_rast_high_res.tif"), overwrite=TRUE,
 progress="text") 


rasterize_pts <- raster(file.path(dir_M, "/git-annex/globalprep/hab_coral/v2019/int/pt_coral_rast_high_res.tif")) 
plot(rasterize_pts)
zoom()
#cell stats sum - should be similar to dim
cellStats(rasterize_pts, stat = "sum") #887 vs. N=925 points (good). 

#fasterize our polygon dataset into a raster
fasterize_py <- fasterize::fasterize(v4_coral_py_moll, raster = zones_high_res, 
                                     field = NULL) 
                     
                     
writeRaster(fasterize_py, file.path(dir_M, "/git-annex/globalprep/hab_coral/v2019/int/py_coral_rast_high_res.tif"), overwrite=TRUE)    

fasterize_py <- raster(file.path(dir_M, "/git-annex/globalprep/hab_coral/v2019/int/py_coral_rast_high_res.tif"))

plot(fasterize_py)

#check to see if all points are 0 and 1 for polygon raster
fasterize_py
cellStats(fasterize_py, "sum", na.rm=TRUE)
# 687246
```

**Stack rasters and adjust**
```{r, echo = FALSE, eval = FALSE}
stacked_coral <- raster::stack(rasterize_pts, fasterize_py)
plot(stacked_coral)

raster::calc(stacked_coral, 
                             fun = sum,
                             na.rm = TRUE,
                             filename = file.path(dir_M, "/git-annex/globalprep/hab_coral/v2019/int/combined_coral_pt_py_high_res.tif"), overwrite=TRUE, progress=TRUE)

sum_coral <- raster(file.path(dir_M, "/git-annex/globalprep/hab_coral/v2019/int/combined_coral_pt_py_high_res.tif"))
plot(sum_coral)
sum_coral

## Need to assign all "2's" as "1's" and all "0's" as "NA". There are 2s because 1+1=2. 
combined_coral <- sum_coral

m <- c(1, 1, # if value 1, assign it 1
       2, 1, # if value 2, assign it 1
       0, NA) #if value 0, assign it NA

rclmat <- matrix(m, ncol = 2, byrow = TRUE) # make m a matrix

raster::reclassify(combined_coral, 
                                        rclmat, 
                                        filename = file.path(dir_M, "/git-annex/globalprep/hab_coral/v2019/int/combined_coral_pt_py_adjusted_high_res.tif"), overwrite = TRUE,
                                     progress=TRUE) #reclassify and write raster according the our reclassify matrix

combined_coral <- raster(file.path(dir_M, "/git-annex/globalprep/hab_coral/v2019/int/combined_coral_pt_py_adjusted_high_res.tif"))


```

### aggregate cells to have same resolution as original raster
```{r}
cellStats(combined_coral, "sum", na.rm=TRUE) #687964

combined_coral_aggregate <- aggregate(combined_coral, fact=2, fun="sum", na.rm=TRUE, progress="text")

combined_coral_aggregate_adj <- combined_coral_aggregate
combined_coral_aggregate_adj[combined_coral_aggregate_adj>1] <- 1 

writeRaster(combined_coral_aggregate_adj, file.path(dir_M, "/git-annex/globalprep/hab_coral/v2019/int/coral_high_res.tif"), overwrite=TRUE)

# for wastewater project
writeRaster(combined_coral_aggregate_adj, "/home/shares/ohi/git-annex/land-based/wastewater/habitat/habitat_rasters/coral.tif", overwrite=TRUE)

```


**Calculate zonal stats with zones raster and new combined coral. Convert to km^2 and save int/output files**
```{r, echo = FALSE, eval = FALSE}

combined_coral_aggregate_adj <- raster("/home/shares/ohi/git-annex/land-based/wastewater/habitat/habitat_rasters/coral.tif")

zonal_sums_combined <- raster::zonal(combined_coral_aggregate_adj, 
                                     zones,
                                     fun = "sum",
                                     na.rm = TRUE) #sum all seagrass cells for each ohi zone
zonal_sums_combined_df <- data.frame(zonal_sums_combined)

summary(zonal_sums_combined)

zonal_sums_km2 <- zonal_sums_combined_df %>%
  mutate(year = 2019, habitat = "coral",
         km2 = (0.934478877011218970**2*sum)) %>% #one cell is equal to ~0.93 km
  dplyr::rename("rgn_id" = "zone") %>%
  dplyr::select(-sum)

#compare new and old data
compare_habitat_extent <- zonal_sums_km2 %>%
  left_join(habitat_extent_coral_old, by = "rgn_id") %>%
  mutate(old_km2 = ifelse(
    km2 >0 & is.na(old_km2) ,0, #assign 0 values to old data km2 that have new data so that we can properly graph these differences.
    old_km2
  ))


## Compare old and new habitat extent data. We aren't using the "final" dataset here, so we can see the regions that had extent data added. 48 of them. 
ggplot(compare_habitat_extent, aes(x = old_km2, y = km2)) +
  geom_point() +
  geom_abline() +
  labs(title = "Coral Habitat version ? vs version 4", x = "version ? extent", y=  "version 4 extent") +
  theme_bw()

## Compare habitat extents
sum(compare_habitat_extent$km2, na.rm=TRUE)
check_coral_py_moll <- v4_coral_py_moll
check_coral_py_moll$area <- st_area(v4_coral_py_moll)/1000000
median(check_coral_py_moll$area) #0.1440342 this is in km2
sum(check_coral_py_moll$area) # 152,275.7 km2
sum(compare_habitat_extent$km2, na.rm=TRUE) 

## quite a bit more than the actual amount...but we would never really assume that each square has full coral coverage.

```









