---
title: 'OHI 2025: Allen Coral Atlas API'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---

# How to load in the data
This file shows how to download the reef extent data from the Allen Coral Atlas using their api.
More information on the reef extent criteria can be found at this link: This script uses Google Earth Engine backend to access the data, which will require you to make an account at this [link](https://console.cloud.google.com/earth-engine/welcome?pli=1).

```{r, eval= FALSE}
# Install packages
install.packages("remotes")
remotes::install_github("r-spatial/rgee")

# Load libraries
library(rgee)
library(tidyverse)
```

After opening the r interface with GEE with ee_Initialize(), a browser window will open asking you to sign in to your Google account and grant Earth Engine permissions. After successful authentication, you can start querying GEE.
```{r, eval= FALSE}
# Authenticate to Google Earth Engine
ee_Initialize()
```

Next, we will access the data. You will have to make a project in the Earth Engine Code Editor to do so, and this is how you R will get permissions to access this data.

```{r, eval= FALSE}
# Load the reef habitat image
reef_extent <- ee$Image("ACA/reef_habitat/v2_0")

# Select the 'reef_mask' band
reef_masked <- reef_extent$select("reef_mask")$selfMask()

# Optionally, check the image information (metadata) to confirm the result
print(reef_masked$getInfo())

# Export the masked reef extent to Google Drive
export_task <- ee$batch$Export$image$toDrive(
  image = reef_masked,
  description = 'reef_extent_masked_global',
  fileFormat = 'GeoJSON',
  region = reef_masked$geometry(),
  scale = 30,  # You can adjust resolution
  maxPixels = 1e13  # To handle large images
)

# Start the export task
export_task$start()

# Check the status of the export task
export_task$status()
```

Once the data has notified you if it is loaded on your google drive, upload it to the server via interface (like Cyberduck).
