---
title: 'OHI 2025: Coral trend'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---

# Summary

Coral trend was calculated using condition data from 1975-2006 (Bruno and Selig 2007, Schutte et al. 2010).


## Updates from previous assessment
Using a new dataset for the global distribution of coral reefs for the extent, so we are going to gapfill the trend of the new regions that have been added to the coral extent data. 


## Setup
``` {r setup, echo = FALSE, message = FALSE, warning = FALSE, eval= FALSE}
library(readr)      # for read_csv()
library(raster)
library(here)
library(sf)
library(fasterize)
library(tidyverse)
library(mapview)
library(sp)
#library(rgeos)

source(file.path(here('workflow/R/common.R'))
region_data()
goal  <- 'globalprep/hab_coral/v2025'
dir_git  <- file.path('~/github/ohiprep_v2025', goal)
dir_wcmc <- file.path(file.path(dir_M, 'git-annex/globalprep/_raw_data/wcmc_coral'))
dir_git_old  <- file.path("globalprep/hab_coral/v2012")
```

Now match the old trend data to the extent data, and gapfill any missing trend data with georegion averages. This will also exclude any regions that have trend data but no extent data. 


```{r, eval = FALSE}
trend <- read_csv(file.path(dir_git_old, "data/habitat_trend_coral_updated.csv"))

## read in georegions data
regions <- read.csv(file.path(here('globalprep/supplementary_information/v2018/rgn_georegions_wide_2013b.csv')))

## read in extent data
all <- read.csv(file.path(here("globalprep/hab_coral/v2025/data/habitat_extent_coral_updated.csv")))

all_old <- read.csv(file.path(here("globalprep/hab_coral/v2012/data/habitat_extent_coral_updated.csv")))


all <- all %>%
  filter(km2 > 0) %>%
  filter(rgn_id < 255) %>%
  dplyr::select(-year)

setdiff(all$rgn_id, trend$rgn_id) # coral extent data but no trend score # 6  10  11  12  28  38  44  45  46  48  49  50  51  53  54  56  57  58  66  79  97  98  99 106 107 114 119 125 136 137 146 148 149 150 155 157 159 190 191 195 204 212 231 237

setdiff(trend$rgn_id, all$rgn_id) # have trend but no extent... two are small islands. One is new zealand. New Zealand does not have a coral reef.. so probably not a big deal. 
# [1]   3  90 162


tmp <- all %>%      #eliminates the regions with health scores but no corals
  left_join(trend)


tmp  <- tmp %>%
  left_join(regions) %>%
  group_by(r2) %>%
  mutate(avg_trend_r2 = mean(trend, na.rm=TRUE)) %>%
  ungroup() %>%
  group_by(r1) %>%
  mutate(avg_trend_r1 = mean(trend, na.rm=TRUE)) %>%
  ungroup() %>%
  mutate(trend2 = ifelse(is.na(trend), avg_trend_r2, trend)) %>%
  mutate(trend3 = ifelse(is.na(trend2), avg_trend_r1, trend2)) %>%
  mutate(gap_fill = ifelse(is.na(trend), "georegion", NA)) %>%
  mutate(gap_fill = ifelse(is.na(trend) & is.na(avg_trend_r2), "georegion", gap_fill)) %>%
  mutate(gap_fill = ifelse(is.na(gap_fill), 0, gap_fill)) %>%
  dplyr::select(rgn_id, habitat, gap_fill, trend=trend3)
summary(tmp)

# Since several new regions had scores added this year, and a couple do not fall into georegion categories, I will manually add values for them: Madeira, and the Canary Islands. We will fill these with the mean trend of 0.012191
tmp <- tmp %>%
  mutate( trend = case_when(
    rgn_id == 57 ~ 0.012191,
    rgn_id == 58 ~ 0.012191,
    TRUE ~ trend
  ))

# Add gapfill to these areas
tmp <- tmp %>%
  mutate( gap_fill = case_when(
    rgn_id == 57 ~ "georegion",
    rgn_id == 58 ~ "georegion",
    TRUE ~ gap_fill
  ))

### save summary of gap-filling:
old_gaps <- read.csv(file.path(here("globalprep/hab_coral/v2012/data/trend_coral_gf.csv")))


trend_gaps <- tmp %>%
  left_join(old_gaps, by = c("rgn_id", "habitat")) %>%
  mutate(gap_fill = ifelse(gap_fill.x != 0, gap_fill.x, gap_fill.y)) %>%
  mutate(gap_fill = ifelse(is.na(gap_fill), 0, gap_fill)) %>%
  full_join(rgns_eez) %>% 
  dplyr::select(rgn_id, habitat, variable, gap_fill) %>%
  mutate(variable = "trend") %>%
  mutate(habitat = "coral")
write.csv(trend_gaps, 'globalprep/hab_coral/v2025/data/trend_coral_gf.csv', row.names=FALSE)

### save health data:
trend <- tmp %>%
  dplyr::select(rgn_id, habitat, trend) %>%
  mutate(year = 2025)
write.csv(trend, 'globalprep/hab_coral/v2025/data/habitat_trend_coral_updated.csv', row.names=FALSE)

```

Datacheck: compare to old health data

```{r, eval = FALSE}
# Read in the data
old_trend <- read_csv(file.path(here("globalprep/hab_coral/v2021/data/habitat_trend_coral_updated.csv")))
trend <- read_csv(file.path(here("globalprep/hab_coral/v2025/data/habitat_trend_coral_updated.csv")))

# Rename the dfs
trend_2021 <- old_trend %>%
  rename(trend_2021 = trend)

trend_2025 <- trend %>%
  rename(trend_2025 = trend)


compare_habitat_trend <- trend_2025 %>%
  filter(rgn_id <= 250) %>%
  left_join(trend_2021, by = "rgn_id") %>%
  mutate(difference = trend_2021 - trend_2025) %>%
  left_join(rgns_eez)

compare <- ggplot(compare_habitat_trend, 
                  aes(x = trend_2021, 
                      y = trend_2025,
                      text = paste("Region ID:", rgn_id))) + geom_point() + geom_abline() + 
  labs(
    title = "Compare Coral Trend v2021 to v2025",
    x = "Trend 2021",
    y = "Trend 2025"
  )

# Convert ggplot to interactive plotly
ggplotly(compare, tooltip = "text")
```
