---
title: 'OHI 2025: Coral Extent'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---

# Allen Coral Atlas

The [Allen Coral Atlas project](https://allencoralatlas.org/atlas/#4.84/13.5381/96.3750) contains much data on coral reef extent, stressors, and geoforms. This layer contains data from 2018-2020, making it the most updated extent, and therefore we decided to use it for the coral_extent section starting in v2025. We found a few errors, as there are around 10 countries (all smaller countries), that had Multicollection geometries fall within their eez borders, that upon further inspection were empty. It is currently our belief that some error is occurring in the downloading format that removes data from these countries, and we hope that this error may be rectified in the future. For more information on the criteria behind categorizing and labeling coral extent, click on this link. 

There is a separate .Rmd file in this folder that contains the code for using a google earth engine api to download the data. There is also code included to manually download the data, but it is a very time consuming process. 

Here we will download all of the new coral reef extent data from the Allen Coral Atlas project, using Google Earth Engine. The package that we will use in R to do so is called "rgee" from the r-spatial package.

```{r, eval= FALSE}
# Read in packages
library(readr)      # for read_csv()
library(raster)
library(here)
library(sf)
library(fasterize)
library(tidyverse)
library(mapview)
library(sp)
library(dplyr)


source(here::here('workflow', 'R', 'common.R'))
goal     <- 'globalprep/hab_coral/v2025'
dir_git  <- file.path('~/github/ohiprep_v2025', goal)
ohi_rasters() # call the region zones raster
dir_aca_manual <- file.path(file.path(dir_M, 'git-annex/globalprep/_raw_data/Allen_Coral_Atlas_Manual/Allen_Coral_Atlas/'))
dir_aca <- file.path(file.path(dir_M, 'git-annex/globalprep/_raw_data/allen_coral_atlas/'))
```


# Load the data in from the server

Depending on which option you are using, select code from below. 
```{r, eval= FALSE}
# Manually Downloaded----------------------------------------------------------------
# Get all reefextent_.gpkg files
aca_files <- list.files(
  dir_aca_manual, 
  pattern = "^reefextent_.*\\.gpkg$", 
  full.names = TRUE
)

# read all layers from all gpkg files
reef_list <- lapply(aca_files, function(f) {
  # list layers inside each gpkg
  layers <- st_layers(f)$name
  
  # read each layer
  lapply(layers, function(l) {
    st_read(f, layer = l, quiet = TRUE)
  })
})

# flatten if you want a single list of sf objects
reef_list <- unlist(reef_list, recursive = FALSE)

# (optional) combine into one big sf object if they have same schema
reef_all <- do.call(rbind, reef_list)

plot(reef_all)

# Save as Shapefile (.shp)
st_write(reef_all, file.path(dir_aca_manual, "global_reefextent.gpkg"))
```


```{r, eval= FALSE}
# Downloaded Via API ----------------------------------------------------------------
## read in polygon data
aca_py <- st_read(file.path(dir_aca, "d2025/global_reefextent.geojson"))
```

## Prep Polygon Data for ACA

```{r, eval= FALSE}
## read in polygon data
aca_py <- st_read(file.path(dir_aca_manual, "global_reefextent.gpkg"))

## take a look
head(aca_py)

# Make valid the raster
aca_py <- aca_py %>%
  st_make_valid()

# Reproject the coral polygons into Mollewide
aca_moll <- st_transform(aca_py, crs = "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")

# Keep only polygons
aca_polygons <- aca_moll[st_geometry_type(aca_moll) %in% c("POLYGON", "MULTIPOLYGON"), ]
```


## Rasterize the polygon extent
```{r, eval= FALSE}
# Take the land rasters from OHI
land_ocean <- raster("/home/shares/ohi/git-annex/globalprep/spatial/v2017/regions_land_ocean.tif")

# Rasterize the geometries
fasterize_aca <- fasterize::fasterize(aca_polygons, raster = land_ocean,
                                     field = NULL) # all polygons given value of 1

# Find the total sum of the area
sum(st_area(aca_polygons))*0.000001 # v2021: 152275.7 km2 using R function...
# v2025: 392,094 km2, so much more area!

# Check: Get area in km2 of all coverage
print(cellStats(fasterize_aca, "sum", na.rm=TRUE)*0.934478877011218970**2) # 149899.6 km2 - metadata says there is around 150000 km2 of cover.. this seems good to me.
# v2025: This says 391,765.6 km2, so a little has been lost, but overall very close to vector area.

plot(fasterize_aca, col = "blue")

writeRaster(fasterize_aca, filename = file.path(dir_M, "/git-annex/globalprep/hab_coral/v2025/int/aca_coral_rast.tif"), overwrite=TRUE)
```


```{r, eval= FALSE}
# Calculate zonal stats with zones raster and new coral. Convert to km^2 and save int/output files*
zonal_sums_combined <- raster::zonal(fasterize_aca,
                                     land_ocean,
                                     fun = "sum",
                                     na.rm = TRUE,
                                     progress="text") #sum all coral cells for each ohi zone
zonal_sums_combined_df <- data.frame(zonal_sums_combined)

zonal_sums_km2 <- zonal_sums_combined_df %>%
  mutate(year = 2025, habitat = "coral",
         km2 = (0.934478877011218970**2*sum)) %>% #one cell is equal to ~0.93 km
  rename("rgn_id" = "zone") %>%
  dplyr::select(-sum) %>%
  filter(rgn_id <= 250)

print(sum(zonal_sums_km2$km2)) # 146152.5

## save this
write.csv(zonal_sums_km2, file.path("globalprep/hab_coral/v2025/int/coral_aca_area_zonal.csv"), row.names = FALSE)
```

## WCMC Data Comparison

Since the point data, converted to polygons, only adds around 132 km2, this would bring the total extent from the WCMC data at 150,132. 

With the ACA data containing roughly 4 times the polygons, and the area is around 391,765 km2, we will be going forward with only ACA data. However, in case there is not data for a certain country using our new data that we know has coral, we can gapfill with previous year's data.


```{r, eval= FALSE}
# Comparing with the final 2021 data to see gaps. I loaded in the habitat_extent_coral_updated.csv from v2021 to compare to the aca results thus far
zonal_sums_km2 <- as.data.frame(zonal_sums_km2)
habitat_extent_coral_updated <- as.data.frame(habitat_extent_coral_updated)

# Join the data sets and filter
df_compare <- coral_aca_area_zonal %>%
  dplyr::select(rgn_id, area_2025= km2) %>%
  left_join( habitat_extent_coral_updated %>% dplyr::select(rgn_id, area_2021 = km2), by = "rgn_id") %>% filter(area_2025 == 0, area_2021 > 0)

# Join with the region id information
region_data()
df_compare <- df_compare %>%
  left_join(rgns_all, by= "rgn_id")
```

## Missing Regions

We are missing data on 11 regions that were recorded in 2021 but not in 2025. Upon examining the data available on the [Allen Coral Atlas Website](https://allencoralatlas.org/atlas/#7.42/7.7841/81.6158), there is data for most of these locations. Likely, because all of the reefs associated with these small islands (as well as the tiny reef tracts of Guatemala) are very fine scale in resolution, they may not be incorporated once converted to rasters. We lose around 300 km2 converting from vector to raster, and I wonder if this is why we have no data for those areas. I can try and run the zones with the vector data, but due to the large number of polygons, it may take too long. I may just gapfill these regions with data from previous years.

NOTE: I ran the check all the way at the bottom of the script to ascertain where the missing values are from. They are likely due to errors in the download process, and thus we must gapfill with previous years

## Gapfilling

```{r, eval= FALSE}
# Join 2025 area and 2021 area by region_id and update km2 column conditionally
filled_zonal_sums <- zonal_sums_km2 %>%
  mutate(km2 = case_when(
    zonal_sums_km2$rgn_id %in% df_compare$rgn_id ~ 
      df_compare$area_2021[match(zonal_sums_km2$rgn_id, df_compare$rgn_id)], 
    TRUE ~ zonal_sums_km2$km2  # Keep original km2 value if region_id is not in df_compare
    ))

# Check to see if this worked
df_compare_check <- filled_zonal_sums %>%
  dplyr::select(rgn_id, area_2025= km2) %>%
  left_join( habitat_extent_coral_updated %>% dplyr::select(rgn_id, area_2021 = km2), by = "rgn_id") %>% filter(area_2025 == 0, area_2021 > 0)

# YES! This is zero now which means we have values for all countries that have reef extent
```


## Write the final product

```{r, eval= FALSE}
# Rename coral column so that coral will not be read in as a color in perpetuity ('coral' -> 'corals'!)
coral_habitat_extent <- df_compare_check %>%
  mutate(habitat = 'corals')

# Write the zonal sums to the "habitat_extent_coral_updated.csv"
write.csv(coral_habitat_extent, here("globalprep/hab_coral/v2025/data/habitat_extent_coral_updated.csv"), row.names = FALSE)

# lets make a gapfilling file
region_data()

coral_area_final_gf <- zonal_sums_km2 %>%
  dplyr::select(rgn_id, habitat) %>%
  mutate(variable = "extent", gap_fill = 0) %>% 
  full_join(rgns_eez, by = "rgn_id") %>%
  dplyr::select(rgn_id, habitat, variable, gap_fill) %>%
  mutate(habitat = "coral", variable = "extent")

write.csv(coral_area_final_gf, here"globalprep/hab_coral/v2025/data/extent_coral_gf.csv", row.names = FALSE)
```


# Datacheck

```{r, eval = FALSE}
# Load in this year's and last year's data
coral_area_2025 <- read.csv(here("globalprep/hab_coral/v2025/data/habitat_extent_coral_updated.csv"))
coral_area_2021 <- read.csv(here("globalprep/hab_coral/v2021/data/habitat_extent_coral_updated.csv"))

# Create an area column for each 
coral_area_2025 <- coral_area_2025 %>%
  mutate(area_2025 = km2)

coral_area_2021 <- coral_area_2021 %>%
  mutate(area_2021 = km2)

# Create comparison graph
compare_habitat_extent <- coral_area_2025 %>%
  left_join(coral_area_2021, by = "rgn_id") 

compare_coral_extent <- ggplot(compare_habitat_extent, aes(x = area_2021, y = area_2025, color = as.factor(rgn_id))) +
  geom_point() +
  geom_abline() +
  labs(title = "Coral Habitat v2021 vs v2025", x = "old extent", y=  "new extent") + theme(legend.position = "none")

ggplotly(compare_coral_extent)


sum(coral_area_final$km2)
# 146279.4 total area after eez intersection - missing some, but the v4_coral_py includes regions that OHI doesnt include

sum(habitat_extent_coral_updated$km2)
#132924.5 total area for the old extent data
```


# EXTRA: Check to see if unmatched countries have any data upon download (empty multicollection check)

## Find overlap with country EEZ boundaries

```{r, eval= FALSE}
# Run the ohi function to get eez data
region_data()
regions_shape()

regions_eez <- regions %>% 
  filter(rgn_type == "eez")

eez <- regions_eez %>%
  filter(rgn_id %in% rgns_eez$rgn_id)

eez <- eez %>%
  filter(rgn_name != "Antarctica")

eez <- eez %>%
  st_make_valid()
```

```{r, eval= FALSE}
# make sure CRS matches
eez_transform <- st_transform(eez, st_crs(aca_polygons))

for (i in seq_len(nrow(eez))) {
  eez_zone <- eez_transform[i, ] # one EEZ polygon
  
  # clip: keeps parts of big_sf inside or overlapping eez_zone
  clipped <- st_intersection(aca_polygons, eez_zone)
  
 
  # create filename (use ID if available, otherwise index)
  zone_id <- if ("rgn_name" %in% names(eez_transform)) eez_zone$rgn_name else i
  zone_id <- gsub(" ", "_", zone_id)   # replace spaces
  out_file <- paste0("/home/shares/ohi/git-annex/globalprep/hab_coral/v2025/eez_", zone_id, ".gpkg")

   # check if file already exists
  if (file.exists(out_file)) {
    cat("Skipping (already exists):", out_file, "\n")
    next
  }
  
  # Make the geometries valid
  clipped <- st_make_valid(clipped)
  clipped <- st_buffer(clipped, 0)
  clipped <- st_collection_extract(clipped, "POLYGON")
  
  # save to disk
  st_write(clipped, out_file, delete_dsn = TRUE)
  
  cat("Saved:", out_file, "\n")
}
```

# Check a specific file
```{r, eval= FALSE}
eez_Guatemala <- st_read(file.path("/home/shares/ohi/git-annex/globalprep/hab_coral/v2025//eez_Guatemala.gpkg"))
```

