---
title: 'OHI 2021: Coral condition'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
---

# Summary

Coral condition was calculated using current condition data divided by reference condition. We used condition data from percent live coral cover from 12,634 surveys from 1975-2006 (Bruno and Selig 2007, Schutte et al. 2010). When multiple data points were available for the same site and year, we averaged these data, and also averaged the site data to calculate a per country per year average. However, data were missing for several countries and some countries did not have data for the reference or current year time periods or had only 1-2 surveys. Because coral cover can be highly temporally and spatially dynamic, having only a few surveys that may have been motivated by different reasons (i.e., documenting a pristine or an impacted habitat) can bias results. To calculate condition we used fitted values from a linear trend of all data per country, which was more robust to data poor situations and allowed us to take advantage of periods of intense sampling that did not always include both current and reference years. Then, we created a fitted linear model of all these data points from 1975-2010, provided that 2 or more points are in 1980-1995 and 2 or more points are in 2000-2010. We defined the ‘current’ condition (health) as the mean of the predicted values for 2008-2010, and the reference condition as the mean of the predicted values for 1985-1987. Where country data were not available, we used an average from adjacent EEZs weighted by habitat area, or a georegional average weighted by habitat area, based on countries within the same ocean basin (Figure 7.1).

# v2025 Update:
The Global Coral Reef Monitoring Network released an updated report in 2020 that showed percent of decrease in live coral coverage in comparison to the 1980s-2010s data.The percent of live coral loss for the GCRMN zones was calculated up to 2020. In this .Rmd file, we will be intersecting OHI regions with the GCRMN regions, and applying the relevant percent decreases to countries that fall within each zone. This process uses the results from each individual GCRMN Regional analysis, which can be found [here](https://gcrmn.net/2020-report/). Each pdf contains various tables, one of which has percent total decrease in coral cover from 1980, which was the value used in this assessment. 

Since we are using a new dataset for the global distribution of coral reefs (Allen Coral Atlas), we are going to gap fill the condition of the new regions that have been added to the coral extent data. 


## Setup
``` {r setup, echo = FALSE, message = FALSE, warning = FALSE, eval= FALSE}
library(readr)      # for read_csv()
library(raster)
library(here)
library(sf)
library(fasterize)
library(tidyverse)
library(mapview)
library(sp)
#library(rgeos)

source(here('workflow/R/common.R'))

goal     <- 'globalprep/hab_coral/v2025'
dir_git  <- file.path('~/github/ohiprep_v2025', goal)
dir_wcmc <- file.path(file.path(dir_M, 'git-annex/globalprep/_raw_data/wcmc_coral'))
dir_gcrmn <- file.path(file.path(dir_M, 'git-annex/globalprep/_raw_data/gcrmn-regions'))
region_data()
dir_git_old  <- file.path("globalprep/hab_coral/v2012")
```

Now match the old health data to the extent data, and gapfill any missing health data with georegion averages. This will also exclude any regions that have health data but no extent data. 


```{r, eval = FALSE}
health <- read_csv(file.path(dir_git_old, "data/habitat_health_coral_updated.csv"))

health_old_gf <- read_csv(file.path(dir_git_old, "data/health_coral_gf.csv")) %>%
  left_join(rgns_eez)

## read in georegions data
regions <- read.csv(file.path(here('globalprep/supplementary_information/v2018/rgn_georegions_wide_2013b.csv')))

## read in extent data
all <- read.csv(file.path(here("globalprep/hab_coral/v2025/data/habitat_extent_coral_updated.csv")))

all_old <- read.csv(file.path(here("globalprep/hab_coral/v2012/data/habitat_extent_coral_updated.csv")))


all <- all %>%
  filter(km2 > 0) %>%
  filter(rgn_id < 255) %>%
  dplyr::select(-year)

setdiff(all$rgn_id, health$rgn_id) # coral extent data but no health score
# v2025: These are new regions that have data this year but did not last year: 53  56  57  58  66  79  97  98  99 106 136 195

setdiff(health$rgn_id, all$rgn_id) # have health but no extent... two are small islands. One is new zealand. New Zealand does not have a coral reef.. so probably not a big deal. 
# v2025: The same regions have health data but no extent from previous years: 3  90 162


# Countries that have corals that were added this year (besides Guatemala and Israel): Pakistan, Cape Verde, Madeira, Canary Islands, Senegal, Liberia, Togo, Benin, Ghana, Ivory Coast.

# We will join the health data with the region names, to see which countries are near to these locations, and what their scores are.

health_regions <- health %>%
  left_join(rgns_all, by= "rgn_id")

# Relevant Scores:

# African Continent -
# Djibouti: 0.98453834
# Egypt: 0.98453834
# Eritrea: 0.98453834
# Kenya: 0.58067335
# Madagascar: 0.75530778
# Mozambique: 0.75530778
# Somalia: 0.58067335
# South Africa: 0.75530778
# Sudan: 0.98453834
# Tanzania: 0.58067335

# Middle East:
# Oman:0.98453834
# Kuwait:0.98453834
# Qatar:0.98453834

# Notes: Senegal, Liberia, Togo, Benin, Ghana, and the Ivory Coast have no scores near them that can be used. However, we can find the average of other countries that may be similar based on continent and climate. Do to the lush greenery of these countries, I will choose Mozambique, Tanzania, South Africa, and Kenya, which are the most lush coastal countries we have data for. For islands in the Atlantic off the coast of the continent of Africa, we will use the same scores as the African countries added this year. Pakistan will be the mean of Oman, Kuwait, and Qatar. We will give rgn 79 (israel), the same score as Egypt (214), since it is adjacent. (health = 0.9845383). We will give rgn 136 (Guatemala), the same score as Belize (164), since it is adjacent (health = 0.01995085)

african_countries_new <- sum(0.75530778, 0.58067335,0.75530778, 0.58067335)/4
# The value is 0.667990565
```


# Gapfill missing data

```{r, eval= FALSE}
tmp <- all %>%      #eliminates the regions with health scores but no coral
  left_join(health) %>%
  mutate(health = case_when(
    rgn_id == 79 ~ 0.9845383,
    rgn_id == 136 ~ 0.01995085,
    rgn_id == 53 ~ 0.9845383,
    rgn_id == 56 ~ 0.667990565,
    rgn_id == 57 ~ 0.667990565,
    rgn_id == 58 ~ 0.667990565,
    rgn_id == 66 ~ 0.667990565,
    rgn_id == 97 ~ 0.667990565,
    rgn_id == 98 ~ 0.667990565,
    rgn_id == 99 ~ 0.667990565,
    rgn_id == 106 ~ 0.667990565,
    rgn_id == 195 ~ 0.667990565,
    TRUE ~ health
  ))

### save summary of gap-filling:
old_gaps <- read.csv(file.path(here("globalprep/hab_coral/v2012/data/health_coral_gf.csv")))


health_gaps <- tmp %>%
  left_join(old_gaps, by = c("rgn_id", "habitat")) %>%
  mutate(gap_fill = case_when(rgn_id %in% c(79, 136, 53, 56, 57, 58, 66, 97, 98, 99, 106, 195 ) ~ "adjacent", 
                              rgn_id %in% c(107, 121, 149) ~ "0",
                              TRUE ~ gap_fill)) %>%
  full_join(rgns_eez) %>% 
    dplyr::select(rgn_id, habitat, variable, gap_fill) %>%
    mutate(variable = "health") %>%
  mutate(habitat = "coral", 
         year = 2012) 
write.csv(health_gaps, 'globalprep/hab_coral/v2025/data/health_coral_gf.csv', row.names=FALSE)
```


# Use GCRMN Changes

```{r, eval= FALSE}
# Load in the shapefiles for each country's eez
regions_shape()

# Load in the shapefiles for the GCRMN Regions
gcrmn_regions <- st_read(file.path(dir_gcrmn, 'gcrmn_regions.shp'))

# Project gcrmn regions into Mollewide
gcrmn_regions <- st_transform(gcrmn_regions, crs = "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")

# Run an st_intersection to determine which countries fall into which GCRMN regions:
# Compute intersections with gcrmn_regions
intersections <- st_intersection(
  regions %>% select(rgn_id, rgn_name, geometry), 
  gcrmn_regions %>% select(region, geometry)
)

# Compute area of overlap
intersections <- intersections %>%
  mutate(area_overlap = st_area(geometry))

# For each country, keep the region with largest overlap
countries_with_regions <- intersections %>%
  group_by(rgn_id) %>%
  slice_max(order_by = area_overlap, n = 1) %>%
  ungroup() %>%
  select(rgn_id, rgn_name, region)

# Join GCRMN data to health (by rgn_id)
regions_health <- left_join(countries_with_regions, health, by= "rgn_id")

# Apply modifiers to health scores (Multiply health by GCRMN modifier for each region)
regions_health <- regions_health %>%
  mutate(region_number = case_when(
    region == "Australia" ~ 0.25,
    region == "Pacific" ~ 0.17,
    region == "EAS" ~ 0.11,
    region == "WIO" ~ 0.02,
    region == "South Asia" ~ 0.34,
    region == "PERSGA" ~ 0.06,
    region == "ROPME" ~ 0.14,
    region == "ETP"  ~ 0.13,
    region == "Caribbean"  ~ 0.02,
    region == "Brazil"  ~ 0.20,
    TRUE ~ NA_real_  # default if none of the conditions match
  ))

# Use average of health for any NA values left
avg_health <- mean(regions_health$health, na.rm = TRUE)

regions_health <- regions_health %>%
   mutate(health = case_when(
    is.na(health) ~ avg_health,  # replace NA with average
    TRUE ~ health           # keep existing values
  ))
```

```{r, eval= FALSE}
# Subtract percent change (except from Brazil that increased in coral cover)
regions_health <- regions_health %>%
  mutate(adjusted_health = if_else(
    region == "Brazil",        # condition
    health + (health*region_number),            # if TRUE → add
    health - (health*region_number)             # if FALSE → subtract
  ))

# Remove geometry column
regions_health <- regions_health %>% st_drop_geometry()

# Rename columns
regions_health <- regions_health %>%
  rename(
    old_health = health,
    health = adjusted_health
  )
```


# Write final file

```{r, eval= FALSE}
### save health data:
health <- regions_health %>%
  dplyr::select(rgn_id, habitat, health) %>%
  mutate(year = 2025)

write.csv(health, 'globalprep/hab_coral/v2025/data/habitat_health_coral_updated.csv', row.names=FALSE)
```


Datacheck: compare to old health data

```{r, eval = FALSE}
# Read in previous assessment 
old_health <- read.csv(file.path(here("globalprep/hab_coral/v2021/data/habitat_health_coral_updated.csv")))

# Rename columns for ease of comparison
old_health <- old_health %>%
  rename(health_2021 = health)

health <- health %>%
  rename(health_2025 = health)

# Join and compare
compare_habitat_health <- health %>%
  filter(rgn_id <= 250) %>%
  left_join(old_health, by = "rgn_id") %>%
  mutate(difference = health_2021 - health_2025) %>%
  left_join(rgns_eez)

test <- compare_habitat_health %>%
  filter(is.na(health_2021)) ## we gained 10 regions with health data

# Graaph the differences
compare <- ggplot(compare_habitat_health, 
                  aes(x = health_2021, 
                      y = health_2025,
                      text = paste("Region ID:", rgn_id))) +
  geom_point() +
  geom_abline() +
  labs(
    title = "Compare Coral Health v2021 to v2025",
    x = "Health 2021",
    y = "Health 2025"
  )

# Convert ggplot to interactive plotly
ggplotly(compare, tooltip = "text")
```

