---
title: 'OHI 2019: Seagrass extent'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../workflow/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

# Summary

This script generates the extent of seagrass for each OHI region. 


## Updates from previous assessment
Creating an actual script to calculate this. This has not been updated since 2012. Updating the data with newest version (version 6). 

***
## Data Source 

**Downloaded**: 07/15/2019

**Description**:  
Global Distribution of Seagrasses
https://data.unep-wcmc.org/datasets/7

This dataset shows the global distribution of seagrasses, and is composed of two subsets of point and polygon occurence data. The data were compiled by UNEP World Conservation Monitoring Centre in collaboration with many collaborators (e.g. Frederick Short of the University of New Hampshire), organisations (e.g. the OSPAR Convention for the Northeast Atlantic sea), and projects (e.g. the European project Mediterranean Sensitive Habitats "Mediseh"), across the globe (full list available in "Metadata_Seagrass.dbf").

**Time range**: 1934-2015


***
# Methods
Reclassify the seagrass extent data into a mask of 1 or NA, and then compute zonal statistics for the count of cells within an OHI region that have seagrass and then convert into km2.


## Setup
``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(readr)      # for read_csv()
library(raster)
library(here)
library(sf)
library(fasterize)
library(tidyverse)

source(here('workflow/R/common.R'))

dir_wcmc <- file.path(file.path(dir_M, 'git-annex/globalprep/_raw_data/wcmc_seagrass'))
ohi_rasters() # call the region zones raster
regions_shape()

# zones_high_res <- disaggregate(zones, fact=c(2,2),
#                                filename= file.path(dir_M, "git-annex/globalprep/hab_coral/v2019/int/high_res_zones.tif"),
#                                progress="text")

zones_high_res <-raster(file.path(dir_M, "git-annex/globalprep/hab_coral/v2019/int/high_res_zones.tif"))

habitat_extent_seagrass_old <- read_csv(here("globalprep/hab_seagrass/v2012/data/habitat_extent_seagrass_updated.csv")) %>% 
  dplyr::select(rgn_id, old_km2=km2)

```

```{r, echo = FALSE, eval = FALSE}
habitat_extent_seagrass_updated <- read_csv(here("globalprep/hab_seagrass/v2012/data/habitat_extent_seagrass_updated.csv"))

v6_seagrass_pts <- sf::st_read(dsn = file.path(dir_wcmc, "014_001_WCMC013-014_SeagrassPtPy2018_v6/01_Data"), layer = "WCMC_013_014_SeagrassesPt_v6")

v6_seagrass_py <- sf::st_read(dsn = file.path(dir_wcmc, "014_001_WCMC013-014_SeagrassPtPy2018_v6/01_Data"), layer = "WCMC_013_014_SeagrassesPy_v6")

```

**Convert seagrass shapefiles into same CRS as our region zones raster**
```{r, echo = FALSE, eval = FALSE}

moll_crs <- crs(zones_high_res, asText = TRUE)
v6_seagrass_pts_moll <- st_transform(v6_seagrass_pts, crs = moll_crs) #project points shapefile to match zones crs
crs(v6_seagrass_pts_moll) #check to see it worked

v6_seagrass_py_moll <- st_transform(v6_seagrass_py, crs = moll_crs)
crs(v6_seagrass_py_moll)
plot(v6_seagrass_py_moll[1])

```

**Fasterize/rasterize: Where there is seagrass assign a value of 1 and NA otherwise**
```{r, echo = FALSE, eval = FALSE}

## convert from multipoint to point
v6_seagrass_pts_moll_cast <- st_cast(v6_seagrass_pts_moll, "POINT")
plot(v6_seagrass_pts_moll[1])
dim(v6_seagrass_pts_moll[1])
plot(v6_seagrass_pts_moll_cast[1])
dim(v6_seagrass_pts_moll_cast[1])

#convert to a spatial type 
v6_seagrass_pts_moll_spat <- as(v6_seagrass_pts_moll_cast, "Spatial")
#create a column full of "1's"
v6_seagrass_pts_moll_spat$raster <- 1
#rasterize our points data to raster 
rasterize_pts <- raster::rasterize(v6_seagrass_pts_moll_spat, 
                                   zones_high_res,
                                   "raster",
                                   fun = min,
 filename = file.path(dir_M, "/git-annex/globalprep/hab_seagrass/v2019/int/pt_seagrass_rast_high_res.tif"), overwrite=TRUE) #write raster

plot(rasterize_pts)
#cell stats sum - should be similar to dim
cellStats(rasterize_pts, stat = "sum") #10889... reassuring, it is similar to the dimensions (N=17668). 

#fasterize our polygon dataset into a raster
fasterize_py <- fasterize::fasterize(v6_seagrass_py_moll, raster = zones_high_res, 
                                     field = NULL)# all polygons given value of 1


writeRaster(fasterize_py, filename = file.path(dir_M, "/git-annex/globalprep/hab_seagrass/v2019/int/py_seagrass_rast_high_res.tif"), overwrite=TRUE) 

plot(fasterize_py)
fasterize_py
rasterize_pts
```

**Stack rasters and adjust**
```{r, echo = FALSE, eval = FALSE}

stacked_seagrass <- raster::stack(rasterize_pts, fasterize_py)
plot(stacked_seagrass)

raster::calc(stacked_seagrass, 
                             fun = sum,
                             na.rm = TRUE,
                             filename = file.path(dir_M, "/git-annex/globalprep/hab_seagrass/v2019/int/combined_pt_py.tif"), overwrite=TRUE, progress="text")

sum_seagrass <- raster(file.path(dir_M, "/git-annex/globalprep/hab_seagrass/v2019/int/combined_pt_py.tif"))

plot(sum_seagrass)
sum_seagrass


## Need to assign all "2's" as "1's" and all "0's" as "NA". There are 2s because 1+1=2. 
combined_seagrass <- sum_seagrass

m <- c(1, 1, # if value 1, assign it 1
       2, 1, # if value 2, assign it 1
       0, NA) #if value 0, assign it NA

rclmat <- matrix(m, ncol = 2, byrow = TRUE) # make m a matrix

raster::reclassify(combined_seagrass, 
                                        rclmat, 
                                        filename = file.path(dir_M, "/git-annex/globalprep/hab_seagrass/v2019/int/combined_pt_py_adjusted.tif"), overwrite = TRUE, progress="text") #reclassify and write raster according the our reclassify matrix

combined_seagrass <- raster(file.path(dir_M, "/git-annex/globalprep/hab_seagrass/v2019/int/combined_pt_py_adjusted.tif"))
combined_seagrass

cellStats(combined_seagrass, "sum", na.rm=TRUE)
1494365 * 0.4672394 * 0.4672394  # 326239
seagrass_polygon_area<- st_area(v6_seagrass_py_moll)/1000000
sum(seagrass_polygon_area)
```


### aggregate cells to have same resolution as original raster
```{r}

combined_seagrass_aggregate <- aggregate(combined_seagrass, fact=2, fun="sum", na.rm=TRUE, progress="text")

combined_seagrass_aggregate_adj <- combined_seagrass_aggregate
combined_seagrass_aggregate_adj[combined_seagrass_aggregate_adj>1] <- 1 

writeRaster(combined_seagrass_aggregate_adj, file.path(dir_M, "/git-annex/globalprep/hab_seagrass/v2019/int/seagrass_high_res.tif"), overwrite=TRUE)

# for wastewater project
writeRaster(combined_seagrass_aggregate_adj, "/home/shares/ohi/git-annex/land-based/wastewater/habitat/habitat_rasters/seagrass.tif", overwrite=TRUE)

```

**Calculate zonal stats with zones raster and new combined seagrass. Convert to km^2 and save int/output files**
```{r, echo = FALSE, eval = FALSE}

zonal_sums_combined <- raster::zonal(combined_seagrass_aggregate_adj, 
                                     zones,
                                     fun = "sum",
                                     na.rm = TRUE,
                                     progress="text") #sum all seagrass cells for each ohi zone
zonal_sums_combined_df <- data.frame(zonal_sums_combined)

summary(zonal_sums_combined_df)

zonal_sums_km2 <- zonal_sums_combined_df %>%
  mutate(year = 2019, habitat = "seagrass",
         km2 = (0.934478877011218970**2*sum)) %>% #one cell is equal to ~0.93 km
  rename("rgn_id" = "zone") %>%
  select(-sum)

#compare new and old data
compare_habitat_extent <- zonal_sums_km2 %>%
  left_join(habitat_extent_seagrass_old, by = "rgn_id") %>%
  mutate(old_km2 = ifelse(
    km2 >0 & is.na(old_km2) ,0, #assign 0 values to old data km2 that have new data so that we can properly graph these differences.
    old_km2
  ))


## Compare old and new habitat extent data. We aren't using the "final" dataset here, so we can see the regions that had extent data added. 48 of them. 
ggplot(compare_habitat_extent, aes(x = old_km2, y = km2)) +
  geom_point() +
  geom_abline() +
  labs(title = "Coral Habitat version ? vs version 4", x = "version ? extent", y=  "version 4 extent") +
  theme_bw()

sum(seagrass_polygon_area)  # 560,802
sum(compare_habitat_extent$km2, na.rm=TRUE)  # 1,218,752
sum(compare_habitat_extent$old_km2, na.rm=TRUE)  # 294,052
```



