---
title: "CRS Prep & Review"
date: "2024-08-01"
#date: "`r format(Sys.Date(), '%Y')`"
#author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
author: "Anna Ramji"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../workflow/templates/ohi_hdr.html'
pdf_document:
  toc: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# ---- Load packages ----
library(terra)
library(here)
library(tidyverse)
```


## Data Exploration

Data describing commercial landings (tonnes) in 2017.

```{r}
# ---- Read in commercial landings data ----
fish <- terra::rast(here("Reference", "CRS", "commercial_landings_2017.tif"))

# ---- preliminary data exploration -----
# print out geospatial data info (class, dimensions, resolution, extent, CRS, etc.)
fish # equivalent to (fish)

# print out summary of value layer
summary(fish)

# ---- raw plot ----
## take a look!
terra::plot(fish)
# adjust margins: par(mar = c(bottom, left, top, right))

terra::plot(fish, mar = c(0, 1.5, 0, 5))
par(oma = c(1,1,1,1))
terra::plot(fish, mar = c(0, 1.5, 0, 5) )#,
            #oma = c(0,0,0,0)) # outer margin area

# ---- log-transformed plot ----
# lets log so we can visualize this a little better
terra::plot(log(fish + 1) , mar = c(0,1.5,0,3.5))

#par(mar = c(5, 4, 4, 0))
#par(mar = c(5, 4, 4, 0))
#par(oma = c(0,1,0,1))
#par(omi = c(0, 0, 0, 0))
terra::plot(log(fish + 1), main = "Log-transformed Commercial Landings (EPSG:4326)", mar = c(0, 1.5, 0, 3.5), box = TRUE, cex.main = 1)

# library(tidyterra)
# ggplot() +
#   geom_spatraster(data = fish, aes(fill = commercial_landings_2017)) +
#   coord_sf(crs = 4326) +
#   scale_fill_grass_c(palette = "plasma")

```



### How many fish were captured?

```{r}
terra::global(fish, "sum", na.rm = TRUE)
# 86.3 million tonnes
```
Check against Sea Around Us to make sure this seems generally reasonable.

-   seems a little low but generally makes sense (close enough) -- [Sea Around Us](https://www.seaaroundus.org/data/#/global?chart=catch-chart&dimension=sector&measure=tonnage&limit=10) says around 100 million <https://www.seaaroundus.org/data/#/global?chart=catch-chart&dimension=sector&measure=tonnage&limit=10>. 87 million seems close enough to that (within reasonable expected range of values)


### What is the area of each cell?

```{r}
cell_area <- terra::cellSize(fish)
# terra::plot(cell_area, box = TRUE,
#             mar = c(2, 1.5, 1, 8)) # adjusted margins for exporting plot

terra::plot(cell_area, main = "Cell Size: Original CRS (EPSG:4326)", mar = c(0, 1.5, 0, 6))


print(cell_area)
```

### What is the plotted image saying?

- cell size is much smaller at the poles, increases as you head in towards the equator

Given this, what is kind of deceptive about these plots?

-   if density of fish landings in tonnes, pole values would appear higher than they are -- apparent tonnes per cell would be higher 
-   if count, same thing (seems higher than in reality)

```{r}
terra::plot(fish)

# lets log so we can visualize this a little better
terra::plot(log(fish + 1))

```

-   the cells in the northern and southern extremes have a much smaller area than those concentrated in the center
-   this may hide higher commercial fishing values in the center band, as those cells (pixels?) may appear smaller when plotted 
-   value associated w the cell is changing when you convert -- resampling by diff method (nearest neighbor, linear interpolation, bilinear)
-   poles may appear to have higher values than they actually do -- same value over smaller area = higher density value
-   more coarse resolution at the poles can lead to higher catch shown over a greater area than in reality
-   when we sum values across cells, the area of that cell is relevant when we do a global summation:
    -   From the documentation on `cellSize()`
    
> Compute the area covered by individual raster cells.
> Computing the surface area of raster cells is especially relevant for longitude/latitude rasters.

> But note that for both angular (longitude/latitude) and for planar (projected) coordinate reference systems raster cells sizes are generally not constant, unless you are using an equal-area coordinate reference system.

> For planar CRSs, the area is therefore not computed based on the linear units of the coordinate reference system, but on the *actual* area by transforming cells to longitude/latitude. If you do not want that correction, you can use transform=FALSE or init(x, prod(res(x)))


## Reprojection #1: to Mollweide

Let's convert these data to another coordinate reference system and take a look at how this impacts the data:
```{r}
# define Mollweide CRS specifications
moll <- "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"
fish_moll <- terra::project(fish, terra::crs(moll))

# take a look
par(mar = c(5, 4, 4, 0)) # set margins: bottom, left, top, right
terra::plot(fish_moll)
terra::plot(log(fish_moll + 1), main = "Projected CRS: Mollweide")
terra::plot(log(fish + 1), main = "Original CRS: WGS 84 EPSG 4326")

## Count the total tonnes
terra::global(fish_moll, "sum", na.rm = TRUE)
# 389.3 million
```

-   389 million tonnes is a much higher value than the raw data's ~87 million value. Reprojecting the data impacted the values to a significant degree. Cell sizes have been stretched and increased to be uniform across the globe. When we performed our global summation after transforming, the commercial fish landings values were counts per cell, not density (counts per km^2, for example). This means that when we projected the data from EPSG:4326 to Mollweide, the count associated with one cell in the original raster was copied over to potentially multiple cells assigned to the same area in the Mollweide template. 


```{r}
# calculate cell area
cell_area_moll <- terra::cellSize(fish_moll)

#par(mai = c(1, 1, 3, 5)) # set margins: bottom, left, top, right
#par(oma = c(1, 1, 3, 4))
#par(mar = c(1,1,1,1)) 
terra::plot(cell_area_moll, main = "Cell Size: Mollweide Projection", mar = c(0, 2, 0, 6))

terra::plot(cell_area, main = "Cell Size: Original CRS (EPSG:4326)", mar = c(0, 1.5, 0, 6))
```



```{r}
cell_area_moll$area
# min 0
# max is 547 million
cell_area$area
# min: 122442386 (122 million)
# max: 3077249667 (3 billion)
```


Compare to the original tonnes? What is happening? How can you improve this?
-   original tons: 86331119
-   Mollweide tons: 389271065
-   the cell size area is now relatively consistent across the globe -- old cell sizes have stretched etc., to fit a standardized cell size
-     perform global calculations before reprojecting



-   could take raster of cell size, multiply by original (if density) raster, then sum values 
-   ^ divide if count

-   tonnes / area = density, then project
-   take new cell size, multiply by density raster
-   sum at end to check again -- might need to reduce resolution, summary before projecting data


```{r}
# get density (tonnes/area)
fish_density <- fish / cell_area

par(mar = c(4, 4, 4, 0)) # set margins: bottom, left, top, right

terra::plot(log(fish_density), main = "Commercial Landings Density (original CRS: EPSG 4326)")

# reproject to Mollweide
fish_density_moll <- terra::project(fish_density, crs(moll))
fish_density_moll

terra::plot(log(fish_density_moll), main = "Commercial Landings Density (projected CRS: Mollweide)")

# get area of cells
new_cell_size <- cellSize(fish_density_moll)
terra::plot(new_cell_size)

# multiply density by area to get count (tonnes)
new_tonnes <- fish_density_moll * new_cell_size
terra::plot(log(new_tonnes + 1), main = "Density to Counts: Mollweide")

# sum to check
terra::global(new_tonnes, "sum", na.rm = TRUE)
# 85 million!
```

-   close enough :)
-   hopefully you don't care about Fiji data too much (gets distorted with Mollweide projection)


```{r}
#View(fish)

summary(fish)

fish_polygons <- terra::as.polygons(fish)

fish_df <- terra::values(fish_polygons)
```

```{r}
# Get the cell numbers
cells <- 1:terra::ncell(fish)

# Get the x and y coordinates for each cell
xy <- terra::xyFromCell(fish, cells)

# Extract the values
values <- terra::values(fish)

# Combine everything into a data frame
fish_df <- data.frame(
  x = xy[,1],
  y = xy[,2],
  value = values
)

fish_no_na <- fish_df %>% 
  na.omit()
```

-   

See what happens when we make multiple CRS conversions:

```{r}
# Robinson projection
rob <- "+proj=robin +datum=WGS84 +units=m +no_defs"

fish_moll_rob <- terra::project(fish_moll, crs(rob))
plot(log(fish_moll_rob + 1))


terra::global(fish_moll_rob, "sum", na.rm = TRUE)

# now back to lat long:
fish_moll_rob_latlon <- project(fish_moll_rob, fish)

plot(log(fish_moll_rob_latlon + 1))
plot(log(fish + 1))

# check to see if it exactly matches the start
check <- fish_moll_rob_latlon - fish
plot(log(check + 1))
check


```

What is this telling us?


-   the values have changed notably between projections

-   some are highly inflated (large positive numbers), other deflated (large negative in new projection)

-   adding error when we flip between resolutions

-   they're all just approximations (by nature)


```{r}
fish_moll_rob

fish_moll_rob_latlon <- project(fish_moll_rob, fish) # match template (convert crs and resolution)
fish_moll_rob_latlon

fish_moll_rob_latlon_v2 <- project(fish_moll_rob, crs(fish)) # match crs (just convert crs)
fish_moll_rob_latlon_v2
```

