---
title: "Untitled"
output: html_document
date: "2024-08-01"
---

```{r setup, include=FALSE}

library(terra)
library(here)

```

Data describing commercial landings (tonnes) in 2017.
```{r}
fish <- rast(here("Reference/CRS/commercial_landings_2017.tif"))

## Take a look
plot(fish)
# lets log so we can visualize this a little better
plot(log(fish+1))
```

How many fish were captured?
```{r}
global(fish, "sum", na.rm=TRUE)

```
Check against Sea Around Us to make sure this seems generally reasonable.

What is the area of each cell?
```{r}
cell_area <- cellSize(fish)
plot(cell_area)
```
Describe what the plotted image is saying?

Given this, what is kind of deceptive about these plots?
```{r}
plot(fish)
# lets log so we can visualize this a little better
plot(log(fish+1))

```

Let's convert these data to another coordinate reference system:
```{r}
moll <- "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"
fish_moll <- terra::project(fish, crs(moll))

# take a look
plot(fish_moll)
plot(fish_moll + 1)

## Count the total tonnes
global(fish_moll, "sum", na.rm=TRUE)


```

Compare to the original tonnes? What is happening? How can you improve this?

See what happens when we make multiple CRS conversions:

```{r}
#Robinson projection
rob <- "+proj=robin +datum=WGS84 +units=m +no_defs"

fish_moll_rob <-terra::project(fish_moll, crs(rob))
plot(log(fish_moll_rob + 1))

# now back to lat long:
fish_moll_rob_latlon <- project(fish_moll_rob, fish)

plot(log(fish_moll_rob_latlon + 1))

# check to see if it exactly matches the start

check <- fish_moll_rob_latlon - fish
plot(log(check+1))
check



```

What is this telling us?